### 2016-09-20 oldmort.Rnw
first commit

\documentclass[a4paper,11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\usepackage{epstopdf}
\usepackage{url}

\usepackage{natbib}

\bibliographystyle{apalike}
%\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\newcommand{\R}{{\bf R}}

\usepackage[tablesonly]{endfloat}
\usepackage{endfloat}

\renewcommand{\efloatseparator}{\vfill}

% patch \efloat@iwrite to use \protected@write
\makeatletter
\renewcommand\efloat@iwrite[1]{%
   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother

\author{Sören Edvinsson and Göran Broström \\
Centre for Demographic and Ageing Research ({\sc Cedar})\\
Umeå University, Umeå, Sweden}

\title{Social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
    Chicago, Illinois
    November 17--20, 2016. (Version 2016-09-20).}}

\date{November 19, 2016}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

<<setup, include = FALSE>>=
knitr::opts_chunk$set(cache=FALSE,echo=FALSE,message=FALSE,warning=FALSE,comment=NA)##$
@

\section{Introduction}

In this study we take a long-term perspective on the development of social
inequalities in mortality in two Swedish region 1901--2013, in this case
with focus on the adult and elderly population. We investigate the development of
social inequality in mortality during the mortality transition and the
possible changes in the impact of social class in this process, in
particular the hypothesis that modern pattern of social differences in mortality appeared
during the first half of the 20th century.

Several studies have shown how important access to economic and other
resources are for health, but it is not clear how this has developed in
history. \citet{ant67} and \citet{smi83} suggest that social inequality in
health has passed through different phases in history. According to them,
social health differences were
comparatively small during the pre-transitional phase when space was a
strong determinant for the spread of disease. Differences then increased
during the transitional phase when mortality declined and wealthy groups
used their resources to gain better health. Finally, health differences
decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal
differences. \citet{omran82} state, in the third proposition in his theory
of the epidemiological transition, that even if the class differential in
mortality was maintained during the transition, the decline
set in earlier and was faster among privileged groups.

When studies, stimulated by the widely debated Black report \citep{whi98},
began to be published on the late 20th century development, any assumptions
of declining and even disappearing differences could be refuted. Even in
modern wealthy societies social health inequalities are substantial. In
fact, it even seems as if (at least relative) differences are increasing
\citep{kuetal04,frlu07}, contrary to what Antonovsky assumed. Research has
revealed a complex pattern of social class, material conditions,
psycho-social effects, and health-related behaviour. Many contemporary
studies implicitly or explicitly assume a "historical constant" of social
health inequality that often is considered to have been even greater in
earlier times \citep{mar04, liph95}.

It is however time to closely scrutinise the assumption about health and mortality inequalities in the past.
Our knowledge about social health inequalities in history, is
still based on scattered observations and usually not based on thorough
analyses. The social aspects of health began to be observed a long time ago. The 19th century French physician and pioneer in epidemiology Villerm\'e found large differences in mortality between city districts of varying social composition in early 19th century Paris, and
Engels gave evidence of appalling living conditions among the impoverished
in the rapidly expanding industrial environments \citep{ack48}. Despite
their merits as sources for 19th century society and their path-breaking
ideas, these 19th century observations are problematic in many ways. They
lacked adequate sources (no individual data) and methods for analysis. Another issue concerns
the degree to which we can generalise these observations to other
historical contexts.

During recent years, however, increased attention has been paid to the historical
perspective on social health inequalities. The results from different
studies have diverged, some reporting substantial differences but many
finding surprisingly small differences during the first stages of the
transition \citep{bevp11}. Nevertheless, it is certainly not obvious
that historical societies were characterised by the assumed positive
association between socio-economic position and survival. Several traits in
the pre- and early transitional society may explain this circumstance. The
effect of social class is dependent both on patterns of exposure and
resistance against diseases in different populations and sub-populations. In their overview on this topic \citet{bevp11} suggest that consistent hierarchical social health inequalities did not developed until the
first half of the 20th century.

Concerning Sweden, \citet{edli11} found no indication of better health
among wealthy adults living in the Sundsvall region during the
pre-industrial period 1803--1859. This pattern remained for men during the
industrial period 1860--1900, while clearer social health differences
appeared among women. A similar pattern was found for those in old age
\citep{edbr12}. This illustrates the importance of considering gender as
well as age patterns. Results from a study of other parts of northern Sweden,
the Skellefteå  and Umeå regions, 1851--1950 was in accordance with the results
from the Sundsvall region. Higher social position and more resources did not lead
to any health advantages for men, while this was the case for women \citep{segb14}. This is further illustrated with what is known about mortality in younger age groups in the same areas. \citet{edv92, edv04} found weak or
varying social differences in infant mortality in early 19th century northern Sweden, while industrialisation caused large social differences in mortality among children 1--14 years. In their study on socio-economic
mortality differences in a couple of Scanian parishes 1815--1968,
\citet{bedr11} found no indications of any social gradient during most of
the period, suggesting that the increase is a recent phenomenon.

To summarise the present state of knowledge about the development of social
health inequalities (in Sweden but also in other European countries), no
clear social hierarchy has been found before 1900 but substantial
differences towards the late 20th and early 21st century. It appears as if
a more consequent social health divide developed in the period
in-between. This period has not been thoroughly studied on this issue,
mainly because of lack of data to analyse. However, the extension of data at the Demographic Data Base, Umeå University (DDB) makes such studies possible. We are now able to analyse social
differences in mortality with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.

\section{Aims and questions}

The aim of the paper is to investigate the development of social
differences using HISCLASS in mortality for the adult and elderly population during the mortality
transition and its relation to the development of income inequality. The analyses
will be performed on data from the Skellefteå and Umeå regions
1901--2013. The questions are:

\begin{enumerate}
\item Has inequality in mortality between social classes increased among the adult and elderly
  population, primarily benefiting the highest social positions?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any gender differences in the effect of social position?

\item Is social position equally important among the retired population as for those in working age?
\end{enumerate}

\section{Theoretical considerations}

The WHO Commission on Social Determinants of Health framework for the
social determinants of health discusses how socio-economic position impacts
on the health of individuals, placing them within structures that determine
their opportunities and how these influences works
through intermediate determinants related to exposure and vulnerability
\citep[p. 48]{who10}. People are situated within socio-economic and political
contexts, as well as value systems in the society. The socio-economic
position imply different access to resources, power and prestige, all
aspects of crucial importance. The framework takes a holistic approach to
the social determinants, including not only the effect of pure economic and
material resources. The importance of understanding people in their full
context has been emphasised by for example \citet{kri05}. What she calls embodiment refers to the recognition that human beings are
simultaneously social beings and biological organisms.

Social inequality can mean different things, and studies of social
differences 
in health and mortality often differ in what sort of categorisation is used.
This is important to have in mind when comparing results, because the chosen
social variable may measure different aspects of inequality. This relate to an
old subject within social studies, on class and status variables, based on
forerunners 
such as Marx and Weber. In many modern studies, data on income, wealth
and/or education are frequently used. The underlying assumption is that more access
to these resources leads to advantages in life, for example when it comes to survival.
SOCPO \citep{vdpm05} is a system based on different access to what they call social power.
\citet{goldk10} criticise how many economists and epidemiologists operationalize
inequality in their studies, for example by not distinguishing between class
and status aspects. Class refers to relational aspects in society, while many other
variables focus on the distribution of material and/or immaterial resources.
He advocates studies of different aspects, which together will complement each other.

HISCLASS, the classification system used here, is a "\ldots HISCO-based
historical international social class scheme" \citep[p.\ 26]{lema11}. The
different classes in HISCLASS represent distinct categories, based on
whether the work is manual or non--manual, skill level, supervision and
sector and it does not represent a purely hierarchical system, 
but implicitly it reflects large differences in what is in the forefront of
other categorisations, 
as for example access to economic and other resources, status, social power et cetera.
We can thus assume that classes have different opportunities when it comes to matters of health and mortality.

A reasonable assumption is that levels of economic inequality in a society
influences social patterns of health and mortality. Social classes with
more economic resources have better possibilities to stay healthy. We
expect that people from classes with less resources are seriously
disadvantaged in a society with large income inequality. Their poverty
would expose them to larger health risks. The inequality effect could
however influence health in other ways than at the individual economic
level. \citet{wipi09} suggest that income inequality have a contextual
effect as well, meaning that living in an unequal society increases health
risks above what is explained by the individual situation.

How socio-economic position influences lives of people, their health and
their mortality needs to be problematized. When it comes to differential
access to resources, these could be of different character, and not only
material ones. \citet{bou84} distinguishes between economic, 
social and cultural capital. Economic capital allows for much more freedom
to organise lives and to maximise life opportunities. Social capital (here
referring to its individual variant and not as a description on society
level) include different kinds of networks, for example family, friends,
colleagues and other persons acting as valuable contacts and resources. A
rich social network can be crucial when it comes to life opportunities. The
impact of cultural capital on health may seem less obvious. Cultural
capital refers both to the personal skills but also to prestige and
status. This can be demonstrated in taste and choice of life style, which
bring people together with others of similar background and
interests. Achieving a higher cultural capital involve several requirements
on way of living, something that can be relevant for health. Cultural
capital in itself can also bring beneficial psycho-social effects on
individuals, which could be an advantage for health \citep{mar04, wipi09}.

In the historical part of this paper, individuals are grouped
according to their family households, that is, children and married women are
grouped according to the occupation of the head of household, usually the
father or husband. A practical reason for this is that occupation is usually
only provided for the head of household, but there are good research
motives for this as well. We consider the whole family to belong to the
same socio-economic position. But this does not mean that we necessarily
believe that the effects are similar for both sexes. As several researchers
have pointed out, demography has often ignored power relations and
different roles within families \citep{watkins93, mack95, jans07}. By
investigating possible differences in mortality patterns 
according to sex, we may identify gendered patterns within the family, resulting in different effects on family members. Different patterns between the sexes
may be caused by discrimination of the female part in the household. We
know that the society was strongly segregated according to gender and that
women were discriminated and had less space for action. Another possible
cause for different patterns is the gendered expectations on behaviour that
has characterised the relation between sexes in societies throughout
history.
Unfortunately, the same approach cannot be applied on the data from 1960 onward, due to difficulties identifying couples in the modern data. This is further developed in the section on the construction of data for analysis.

\section{Historical background}

The historical context of the present study is a period of major changes in
Sweden and in the studied regions. Swedish society underwent an enormous
development, changing from a rural and sparsely populated society based on
agriculture and with low economic standard to a modern, industrialised and wealthy country. Material conditions improved and welfare provisions for people increased, mainly funded by public finances. Several traits of what was to become the Swedish welfare state had early
predecessors, but became even stronger from the 1930s. These reforms
fundamentally changed living conditions for all citizens but most of all for the
most vulnerable, for example the elderly. Still, levels of income inequality differed during the studied period. The early 20th century was characterised by large income
and wealth differences. This started to decline from the 1930s, but it was
still high into the 1950s. It was only in the
following decades that income inequality became very small in Sweden,
making Sweden one of the most (or even the most) equal societies in the
world around 1980 when inequality levels seem to have been at the lowest. During recent decades, inequality has however increased again \citep{pik14, rowa08, rowa09}.

At the middle of the 19th century, Sweden belonged to the part of the world
with the best survival, being number one in the 1840s, and has continued to belong to the countries with lowest mortality.  Life expectancy was 56 years 1901--10, remaining life
expectancy at age 50 was 24 years. This increased to 72 versus 27 in 1951-59 and 82 versus 33 in 2013. Younger age groups contributed most to the initial and total decline, while improved survival in old age groups adds most to more years during recent decades. The First World War marks an
important breaking point towards declining mortality among young adults, while the period after the Second World War led to large improvements for older adults. \citep{holu76}. If spatial differences determined mortality during the pre-transitional, regional variation and urban
rural differences in mortality levels have diminished, although not disappearing completely.

We do not argue that the Skellefteå and Umeå regions represent the whole of
Sweden, but the availability of data allows us to study the transition
in greater detail here making it a laboratory for studying Swedish demography during a period of fundamental changes. Both regions are part of the county of Västerbotten in the north of Sweden,
along the coast of the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection of
parishes surrounding the town of Skellefteå, founded in 1845 but with a
very small population in the 19th century. The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes. The majority of the population in the region lived in rural villages and hamlets, getting their
livelihood from agricultural production. The
region was vulnerable for harvest failures, one living long in memory was
the famine of 1867--1868 \citep{segb14}. Towards the end of the
century, the region became connected to the Swedish railway system, thus
improving the communications and contacts with other parts of the
country. During the 20th century, industrialisation took place, partly by
the opening of mines in the inlands. This also led to a population increase both in the town and in the rural parts,
and a much more diversified economy.

<<skpop>>=
library(skum)
ske1850 <- with(obs[obs$region == "ske", ],
                sum(birthdate + enter < 1850 &
                    birthdate + exit > 1850))
ske1900 <- with(obs[obs$region == "ske", ],
                sum(birthdate + enter < 1900 &
                    birthdate + exit > 1900))
ske1950 <- with(obs[obs$region == "ske", ],
                sum(birthdate + enter < 1950 &
                    birthdate + exit > 1950))

@

Mortality was fairly low in comparison
with other parts of the country and the fertility transition was late. The
Skellefteå population size as defined in our data-sets was \Sexpr{ske1850} on January 1, 1850,
\Sexpr{ske1900} on January 1,
1900, \Sexpr{ske1950} on January 1, 1950 and 76723 at the end of the century.

\begin{figure}[ht!]
\includegraphics[height=19cm]{myfigs/sweden}
\caption{Skellefteå and Umeå in Sweden.}
\end{figure}

%%\section{Map. Skellefteå and Umeå regions}

<<umpop>>=
ume1900 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1900 &
                    birthdate + exit > 1900))
ume1950 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1950 &
                    birthdate + exit > 1950))
@

The Umeå region in Poplink (until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included. This region had a somewhat different character from that of Skellefteå, even though agriculture was
dominating the economy for a long time. The town of Umeå had a small population but
substantially larger than Skellefteå town during almost the whole studied
period. It was the administrative centre in the county of Västerbotten and
schools and military regiments was placed here. Officials, teachers and the
military were much more common in Umeå. Agriculture dominated in the rural part, while  and there were some foundries, some industries for example in forestry and small-scale production. Umeå had a much more diversified
occupational structure than Skellefteå. The population size as defined in our data-sets was
\Sexpr{ume1900} on January 1, 1900, \Sexpr{ume1950} on January 1, 1950 and 103970 when the century ended.

\section{Data and variables}

The data for the present study comes from two large population databases at
the Demographic Data Base, Umeå University. The early period is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. The population in the two
regions is large enough and socially diverse to enable studies of social
mortality inequalities. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from the regions are accessed for the period 1901--1950s.

The other large data-set comes from the Linnaeus database
\citep{malm10}. It is based on different linked population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from censuses every fifth
year 1960--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare. The construction of these data-sets for analysis is presented below.

Information on individuals are anonymized and the two databases are not
linked between each other. This stops us from following individuals present
in both databases throughout their lives, thus they are treated separately. This
also makes it impossible to add information on individuals in the Linnaeus
database from what we potentially could find in Poplink, for example family
background or occupations earlier in life.

In the data-set analysed here, all individuals 40 years and older and ever being resident in either
of the regions are included. The data file contains the variables on sex,
birth date, death date, first and last date of observation and type of
entrance/exit.

The identification of presence periods as well as the time-dependent variables require some additional comments.

\subsection{Presence periods}

Differences in available information, both when it comes to structure and variables used for attributing social class, from the two data-sets as well as
within the Linnaeus data-set makes necessary diverging approaches when it comes to
the identification of presence periods within our
regions. The Poplink data provides us with exact dates or at least year of
start and exit of the presence of all individuals in our data. This is however not the case
with the Linnaeus database. For the period 1960--2000 we use the presence
in the censuses 1960--1990 together with the yearly population registers
1986--2000 for defining the observation periods as well as their social
class (see below) in our analyses. We have however no exact information on
what happened in between the censuses, for example the exact date of exit
for persons that moved between two censuses. We have solved this by
considering everyone as resident in our regions during the complete
five-year period between censuses if the person is missing in this
census.This overestimates their presence in our regions, but should not
bias the results. We furthermore miss those that both moved in and moved out
or vice versa between censuses. Exit dates are exact if they refer to
death, and in these cases we consider them as constantly present from the
last census to the death data even if the death took place in another part
of Sweden.

For the last period, 2002--2013, we use the yearly population registers
(also known as LISA data from Statistics Sweden), together with information
on deaths from National Board of Health and Welfare. This data allows a
better control of the presence periods in the studied regions.

The presence periods, with start and end date, are reconstructed and
divided according to two time-dependent variables: Social position and
marital status. Thus, at each change in any of
these variables, a new record is created. The time-dependent variables require some additional comments.

\subsection{Social position}

In this paper we analyse mortality differences according to social class
according to the classification scheme HISCLASS. The only type of
socio-economic information available both in Linnaeus and Poplink is
occupations, which can be used to identify distinct social classes or
social status groups. Thus, analysing according to social class is apart
from a scientific choice, also a practical one. Using occupational data is
the only and obvious alternative if we are to make a longitudinal study of
socio-economic mortality differences with consistent classification schemes
(although not without some problems). There are also other problems
connected with using the same classification scheme covering a long time
period, but this will be approached in more detail in the discussion
section. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between sources. The Poplink data usually provides information on the occupation of head of households. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid. Thus, the social class in the census 1990 is considered to be kept until 2000.

The use of different data sets, forces us to apply different approaches towards defining the social position. In the Poplink data, social class is based on the occupational titles given in the parish records. These can be more or less precise. There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers. We consider however their validity to be sufficient for identification of social position. Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}. In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position. This strategy makes the problem of under-recording less problematic.

Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers. Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database. Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period. It is also worth to notice that female labour force participation became the norm, and we could therefore argue that the occupation of the male as signifier of the social class became obsolete. We therefore let all included persons be signified by their individual occupation for this period. Still, we need to keep in mind that results for the different periods are not completely comparable.

Due to the construction of the time-dependent data, some records report unknown social position. This has been updated according to rules in order to optimise available information and minimise missing information. Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census. The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.

The classification scheme used here is HISCLASS \citep{lema11}, a class-based system. The scheme uses HISCO as a basis for assigning categories. HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO. HISCO groups occupations according to tasks and skills.
HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions. This is particularly apparent when analysing the development in separate periods where for example farm workers become more and more rare during the studied century. We have therefore collapsed some categories, ending up with the following six:

\begin{enumerate}
\item \'Elite, HISCLASS 1 and 2, Higher managers and higher professionals.
\item Lower managers, HISCLASS 3.
\item Farmers and fishermen, HISCLASS 8.
\item Lower professionals, HISCLASS 4, 5 and 6, Lower professionals, lower clerical, foremen.
\item Medium skilled workers, HISCLASS 7.
\item Lower skilled workers, HISCLASS 9, 10, 11 and 12, Lower and unskilled workers (including farm workers)
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

In the last studied period, we have brought together the categories even more.

For the Poplink data, we have used the script recoding HISCO into HISCLASS. The procedure for the occupations in the Linnaeus database is a little more complicated. The occupations in the censuses (NYK) must first be transferred to the system later used, SSYK (which is the Swedish version of ISCO88). Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013 can be recoded into HISCLASS. We have used a script provided by Jonas Helgertz, Lund University for this transformation.

\subsection{Marital status}

Marital status is treated as a time-dependent variable in the data file. The statuses used are unmarried, married and dissolution of marriage. The status unmarried is often not explicitly stated, which makes the status unknown in the database Poplink for people for people moving to our parishes.  If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
%'
%'
%' <<getdata, results='asis'>>=
%' library(xtable)
%' library(knitr)
%' load("data/adult.rda")
%' @
%'
%'
%' <<toonerec>>=
%' adult <- rc(adult)
%' ##mad <- adult
%' ##indx <- with(mad, tapply(id, id))
%' ##mad$Enter <- with(mad, tapply(enter, id, min))[indx]
%' ##mad$Exit <- with(mad, tapply(exit, id, max))[indx]
%' ##mad$Event <- with(mad, tapply(event, id, max))[indx]
%'
%' ##mad <- mad[mad$lopnr == 1, ]
%' ##mad$enter <- mad$Enter
%' ##mad$exit <- mad$Exit
%' ##mad$event <- mad$Event
%' ##mad$Enter <- mad$Exit <- mad$Event <- NULL
%'
%' ##mad <- mad[, c("id", "enter", "exit", "event", "birthdate",
%'   ##             "sex", "region",
%'     ##           "socpo20", "socpo30", "socpo40", "socpo50",
%'       ##         "civst20", "civst30", "civst40", "civst50",
%'         ##       "ortnmn")]
%'
%'
%' ## General:
%' source("R/addPer.R")
%' cut <- c(1851, 1876, 1901, 1921, 1940, 1951)
%' xc <- numeric(length(cut) - 1)
%' for (i in 1:(length(cut) - 1)){
%'   xc[i] <- (cut[i] + cut[i + 1]) / 2
%' }
%' ucut <- cut[-(1:2)]
%' uxc <- xc[-(1:2)]
%' mud <- addPer(adult, cuts = cut)
%' source("R/getInter.R")
%'
%' library(xtable)
%' ##xtable(summary(adult[, c("sex", "region",
%'   ##                       "civst40", "socpo40")]))
%' ##save(adult, file = "adult.rda")
%' ##save(mad, file = "mad.rda")
%' @
%'

\section{Models}

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for civil status and rural/urban environment. Since
the central explanatory variable, {\sc Hisclass}, do not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} \citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--59 and 60--99 years of age. The model controls for period, marital
status and whether the individual resides in the urban or rural part. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}.

We analyse both regions together but perform separate analyses for men and
women and for the age groups 40--59 and 60--89. The model in the analyses
include social class, marital status and urban/rural residence. Each decade
during the 20th century gets separate analyses, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period. We present the results for the period 2001--2013
separately. We suggest a heterogeneity index in order to investigate a
possible homogenisation of mortality between social classes.

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. Thus, the main results are
presented \emph{graphically}. Note further that {\tt Hisclass} is a
\emph{time-varying} variable. This causes lack of information in the higher
ages around 1960--1980, because that kind of information is simply missing
to a large extent.

%The effect of *hisclass* on survival prospects is in focus, and the obvious
%tool to use is a proportional hazard (PH) model. The {\sc hisclass} level could
%either be analysed as a time dependent variable, or as a fixed variable at
%a certain point in time. In our case, {\sc hisclass} is defined at the age when
%we start to analyse mortality, age 40 or 50. At these ages, most have
%reached their top position and they are still work active. We thus analyse
%future survival according to the social position in mid-life. It has
%advantages when it comes to drawing causal conclusions of observed
%observations. Changes in social position may be due to circumstances that
%are related to mortality risks, for example changing occupation due to bad
%health. Another advantage of letting the {\sc hisclass} level be fixed is that
%time-dependent are more sensitive to the assumptions in a proportional
%hazard model. In our preliminary analyses of letting {\sc Hisclass} be
%time-dependent, we encounter serious deviations from the PH
%assumption. Marital status is however time-varying.

%As a first crude measure of survival, we use median remaining life at age 40. 
%The reason for showing the median and not the average remaining life is that the latter is impossible to estimate
%nonparametrically without strong assumptions. This depends of course on the
%kind of data we have: Heavily censored and truncated. Our approach can be
%seen as a hybrid between period and cohort analysis.The model is calculated
%for 20-year periods with the variables *hisclass*, marital status and urban
%versus rural residence in the model, and separately for the two regions and
%gender. The reason for presenting the regions separately is that there is
%no data covering the 19th century for the Umeå region. The results
%from the different analyses are presented in a serie of figures. We have
%also made statistical analyses of possible changes in the effect of
%*hisclass*, in order to identify possible appearance of a social
%gradient. These analyses are made separately for the age group 40-59 and
%age group 60 and above.

\subsection{POPLINK data, 1901--1950}

The POPLINK data cover 1901--1950.

\subsection{FOB data, 1961--2000}

The \emph{Fob} data is treated similarly as the POPLINK data. There is
unfortunately no links between the individuals in the two data frames, and
the result of that is a gap in the relevant information between the years
1950 and 1960. Furthermore, the first two decades in the \emph{Fob} data,
1960--1980, will have a lot of missing information on {\sc Hisclass}.

\subsection{LISA data, 2002--2013}

The {\sc Lisa} data is yearly, with {\tt hisclass} information starting in 2002. This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, and 5 + 6.

\subsection{Combining the data sets and tabulate}

<<readtab>>=
load("vbTab.rda")
@

\section{Results}

<<colorc>>=
## Colors for plotting strata:
scol <- c("blue", rep("green", 2), "red")
## Common for all groups:

labb <- levels(vbTab$period)
nper <- length(labb)
source("R/getPHglm.R")
## End 'common'
@

Models are fitted separately for each ten-year time period, from  1901 to
2000, with the exception of the period 1951--1960. We stratify on HISCLASS
and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
(Skellefteå town or outside). The cumulative
hazards at the end of the time period (60 if 40--59, 90 if 60--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

%creating an \emph{heterogeneity index} (a better name is needed:
%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
%Something catchy.).

\subsection{Ages 40--59}

%The distribution over time periods.
<<getske4060>>=
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, women."
scap1w <-  "Cumulative hazards, ages 40--59, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, men."
scap1m <-  "Cumulative hazards, ages 40--59, men."
##
cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard by HISCLASS, ages 40--59."    
cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). Log scale."
scap12log <- "Total hazard by HISCLASS, ages 40--59 (log scale)."

cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, women."
scap2w <-  "Cumulative hazards, ages 60--89, women."
cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, men."
scap2m <-  "Cumulative hazards, ages 60--89, men."
##
cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap22 <- "Total hazard by HISCLASS, ages 60--89."

##
vb4059 <- vbTab[vbTab$age %in% seq(40, 55, by = 5), ]
vb4059$age <- factor(vb4059$age)
@

We start by looking at cumulative hazards by {\sc Hisclass} over time
periods. They are adjusted for {\tt civil status}, and
{\tt urban/rural} residence. We present separate analyses  for women and men. The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.

%\subsubsection{Women}

<<hazcalw>>=
resw <- getPHglm(vb4059[vb4059$sex == "female", ],
            nper, labb)
resm <- getPHglm(vb4059[vb4059$sex == "male", ],
            nper, labb)
@

<<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 10>>=
source("R/plotHaz.glm.R")
plotHaz.glm(resw, main = labb)
@

<<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 10>>=
plotHaz.glm(resm, main = labb)
@

From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
the idea of proportional hazards is not reasonable.The figures do however indicate substantial changes in the social pattern of mortality during the 19th century, something we will return to.

We will use as a summary measure the value of the cumulative hazards at
duration 20 (age 60).

<<cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, scol = scol, logScale = FALSE, ylab = "H(60|40)",
                main = "Women")
moutp <- plotED.glm(resm, scol = scol, logScale = FALSE, ylab = "H(60|40)",
                main = "Men")
par(oldpar)
@

%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--59 per individual by HISCLASS and decade, men.">>=
%%@

Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century. Apart from
the social pattern that will be commented on below, there are a couple of
observations that can be made from the figure. Women had consistently
better survival than men. This is a well-known phenomenon in Swedish
demographic history. Male mortality has almost always been higher than
female in all age groups \citep{wil99, suwi07}. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. The compression that is
visible for later decades is partly explained by the decreasing general
mortality. This effect can be eliminated by showing these numbers on a log
scale, showing the relative levels. Presented in that way, it is not
obvious that differences between social classes have diminished. This
conclusion has been confirmed by separate analyses (not shown). There is no
clear indication of homogenisation of mortality levels between social
classes.

<<cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
oldpar <- par(mfrow = c(2, 1))
wloutp <- plotED.glm(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
main = "Women")
mloutp <- plotED.glm(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
main = "Men")
par(oldpar)
@

The analysis for the last period is based on a separate analysis and is presented separately.

<<lissyoung>>=
liss4059 <- age.window(liss, c(40, 59))
@

<<liswom>>=
fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "female", ])
fitm1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "male", ])
dcol <- c("blue", "darkgreen", "darkgreen", "red")
@

See Figure~\ref{fig:plotliss1} for the cumulative hazards.

<<plotliss1, fig.cap = "Cumulative hazards, ages 40--59, 2002--2013.", fig.height = 4>>=
oldpar <- par(mfrow = c(1, 2))
plot(fitw1, fn = "cum", col = dcol, main = "Women")
plot(fitm1, fn = "cum", col = dcol, main = "Men")
par(oldpar)
@

See Table~\ref{tab:tablisswom1} the PH regression for women.
<<tablisswom1, results = "asis">>=
fitwo1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss4059[liss4059$sex == "female", ])
dr <- drop1(fitwo1, test = "Chisq")
ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--59, 2002--2013.", 
    label = "tab:tablisswom1")
@

See Table~\ref{tab:tablissman1} the PH regression for men.
<<tablissman1, results = "asis">>=
fitma1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss4059[liss4059$sex == "male", ])
dr <- drop1(fitma1, test = "Chisq")
ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--59, 2002--2013.", 
    label = "tab:tablissman1")
@

The central question in this paper is however what classes are most
advantaged when it comes to mortality, and if the social pattern changed
during the 20th century. Notice that we have not analysed the 1950s due to
the fact that we only have data for some parishes at that time,
consequently having no corresponding population for that
decade. Furthermore, the results for the 1960s should be considered with
caution. The way the social class is defined leads to much missing
information for that decade, something that have restricted the size of the
population observed, making the results more shaky. The changing economy
also have consequences for the analysis, particularly when it comes to the
agricultural sector. While farmers were common occupations during the first
half of the 20th century, these occupations has constantly decreased in
numbers thereafter.

The social pattern among women are mixed, but in accordance with previous
analyses from the present research group on other regions in northern
Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
social classes usually have comparatively low mortality, while working
class women have high. This is especially the case during the latest
decades of the century and especially for the period 2002--2013.

A different pattern appears among men. The groups we would expect to have
the lowest mortality, if we assume that access to different resources and
having a high status determines survival, instead have the highest during
much of the century. This is also in accordance with our previous studies
\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
low mortality. This changes however towards the end the 20th century. From
the 1980s, there is a clear advantage when it comes to survival for higher
managers and professionals. This is also what we usually find in analyses
of social health inequalities in present-day Sweden as well as in most
other countries. In the regions we are studying, this is however a quite
recent phenomenon.The advantage for higher classes and disadvantage for low
skilled workers is very clear for the period 2002--2013.

\subsection{Ages 60--89}

%The distribution over time periods.
<<getske6089>>=
vb6089 <- vbTab[vbTab$age %in% seq(60, 85, by = 5), ]
vb6089$age <- factor(vb6089$age)
@

We now turn to the elderly, i.e. those that mostly have left the workforce. We start by looking at cumulative hazards by {\sc Hisclass} over time
periods. In accordance with the analyses for the age group 40--59, the model adjusts for {\tt civil status} and
{\tt urban/rural} residence and with separate analyses for women and men. We do not present any results concerning possible homogenisation of mortality levels, since the main results do not deviate from that of the younger group.

%\subsubsection{Women}

<<hazcalwold>>=
resw <- getPH(vb6089[vb6089$sex == "female", ],
              nper, labb)
resm <- getPH(vb60896089[vb6089$sex == "male", ],
                nper, labb)
@

<<hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 10>>=
##source("R/plotHaz.R")
plotHaz.glm(resw, main = labb)
@

<<hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 10>>=
plotHaz.glm(resm, main = labb)
@

The cumulative hazards for the age group 60--89 in different decades are
presented in Figures~\ref{fig:hazplotwold} and
\ref{fig:hazplotmold}. Corresponding to what we find for the age group
40--59, it is obvious that the assumption of proportional hazards is not
reasonable.

We will use as a summary measure the value of the cumulative hazards at 
duration 30 (age 90), starting from age 60. For the period 2002--2013, a separate analysis is presented.

The general pattern do not differ from the one of the younger age group. The levels are slightly more homogeneous, but we cannot establish that the elderly diverge to any degree from the working-age population. Mortality among women was, as expected, lower than among men. We also find a clear decline in mortality, but this decline did not set in until sometimes after the Second World War. This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases. The results for the 1960s are however uncertain as discussed above, and should not be taken to seriously. The levels seem to be to low. This is something we need to control further.

Concerning the social pattern, we find results fairly similar to the ones for the age group 40--59. For women the results are mixed, even though women from the elite had comparatively low mortality and working class women high. For men we find the same intriguing change among elderly as for the adult population 40--59 years. Mortality was in fact substantially higher among men from the elite and the group of lower managers while it was lowest among workers. This did not change until the last decades of the 20th century. At that time, the highest social classes instead had the best survival while the opposite was the case for workers. The pattern had reversed, thus confirming the results from most studies of this topic concerning the present-day society. The advantage for the higher classes and the disadvantage for the workers are apparent also for this age group during the early 21st century.

<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
##source("R/plotED.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, scol = scol, logScale = FALSE, ylab = "H(90|60)",
                main = "Women")
moutp <- plotED.glm(resm, scol = scol, logScale = FALSE, ylab = "H(90|60)",
                main = "Men")
par(oldpar)
@

%% The compression that is visible for later decades is partly explained by the decereasing general mortality. This effect can be eliminated by showing these numbers on a log scale.

%% <<cumslogwold, fig.cap = "Probability of dying before age 90 for a 60 year old person by HISCLASS and decade, log scale, women (top) and men (bottom).", fig.height = 10>>=
%% oldpar <- par(mfrow = c(2, 1))
%% wloutp <- plotED.ph(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
%% main = "Women")
%% mloutp <- plotED.ph(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
%% main = "Men")
%% par(oldpar)
%% @

%%<<cumslogm, fig.cap = "Expected number of deaths in ages 40--59 per individual by HISCLASS and decade, log scale, men.">>=
%%@

%%<<measm, fig.cap = "Heterogeneity index over time, ages 40--59, men.">>=
%%@

%' For the LISA data:
%' 
%' <<lissold>>=
%' load("liss.rda")
%' liss6089 <- age.window(liss, c(60, 90))
%' @
%' 
%' 
%' <<liswom2>>=
%' fitw2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss6089[liss6089$sex == "female", ])
%' fitm2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss6089[liss6089$sex == "male", ])
%' dcol <- c("blue", "darkgreen", "darkgreen", "red")
%' @
%' 
%' <<plotliss2, fig.cap = "Cumulative hazards, ages 60--89, 2002--2013.", fig.height = 4>>=
%' oldpar <- par(mfrow = c(1, 2))
%' plot(fitw2, fn = "cum", col = dcol, main = "Women")
%' plot(fitm2, fn = "cum", col = dcol, main = "Men")
%' par(oldpar)
%' @
%' 
%' We also present the output from Cox regressions for women and men
%' separately, with {\sc Hisclass} as an 
%' ordinary covariate (no stratification). The non-proportionality issue
%' regarding this covariate is not so severe in the Lisa data.
%' 
%' See Table~\ref{tab:tablisswom2} for the PH regression for women.
%' <<tablisswom2, results = "asis", eval = TRUE>>=
%' fitwo2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
%'                  data = liss6089[liss6089$sex == "female", ])
%' dr <- drop1(fitwo2, test = "Chisq")
%' ltx(fitwo2, dr = dr, caption = "PH regression, women aged 60--89, 2002--2013.", label = "tab:tablisswom2")
%' @
%' 
%' See Table~\ref{tab:tablissman2} for the PH regression for men.
%' <<tablissman2, results = "asis", eval = TRUE>>=
%' fitma2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
%'                  data = liss6089[liss6089$sex == "male", ])
%' dr <- drop1(fitma2, test = "Chisq")
%' ltx(fitma2, dr = dr, caption = "PH regression, men aged 60--89, 2002--2013.", label = "tab:tablissman2")
%' @

%' 
%' \subsection{Cohort modeling}
%' 
%' The models so far have been period-oriented, that is, ten-year periods have
%' been analysed separately. The contributions in different ages then come
%' from different people in separate birth cohorts. An alternative is to
%' analyse different birth cohorts separately, and we do this here. The
%' drawback with the cohort method is of course that the cohorts get shorter
%' and shorter as we move towards younger cohorts, but on the other hand we
%' get a better picture of what happened in real life.
%' 
%' First we need to prepare the cohort data by combining {\tt fob} and {\tt
%'   lisa} data. The central covariate, {\sc Hisclass} can now be kept
%' \emph{time-fixed} at the start of follow-up, the baseline ages 35--40.
%' 
%' <<cohort data, eval = FALSE>>=
%' ##liss$urban <-  liss$FORSAMLING %in% c(248001, 248002, 284004, 284005, 248011, 248201, 248204)
%' ##fo <- fobb[, c("LINNEID", "birthdate", "sex", "enter", "exit", "event", ##"civst", "urban", "hisclass")]
%' ##li <- liss[, c("LINNEID", "birthdate", "sex", "enter", "exit", "event", ##"civst", "urban", "hisclass")]
%' 
%' fobb$hisclass <- cut(fobb$hisclass, c(0, 2.5, 4.5, 5.5, 8), labels = c("1+2", "3+4", "5", "6+7"))
%' foli <- rbind(fobb, liss)
%' foli <- foli[order(foli$id, foli$enter), ]
%' foli <- rc(foli)
%' foli$Enter <- numeric(NROW(foli))
%' foli$Exit <- numeric(NROW(foli))
%' foli$Event <- numeric(NROW(foli))
%' foli$Enter[foli$lopnr == 1] <- foli$enter[foli$lopnr == 1]
%' foli$Exit[foli$lopnr == foli$antrec] <- foli$exit[foli$lopnr == foli$antrec]
%' foli$Event[foli$lopnr == foli$antrec] <- foli$event[foli$lopnr == foli$antrec]
%' indx <- tapply(foli$id, foli$id)
%' Ex <- tapply(foli$Exit, foli$id, max)
%' Ev <- tapply(foli$Event, foli$id, max)
%' foli$Exit <- Ex[indx]
%' foli$Event <- Ev[indx]
%' foli <- foli[foli$lopnr == 1, ]
%' foli$enter <- foli$Enter
%' foli$exit <- foli$Exit
%' foli$event <- foli$Event
%' foli <- foli[!is.na(foli$hisclass), ]
%' foli <- foli[foli$birthdate > 1921 & foli$birthdate < 1961, ]
%' foli$period <- cut(foli$birthdate, c(1921, 1931, 1941, 1951, 1961), labels = c("1921-30", "1931-40", "1941-50", "1951-60"))
%' cols <- c("blue", "green", "green", "red")
%' @
%' 
%' Then, we create four cohorts by birth year: "1921-30", "1931-40", "1941-50", and "1951-60".
%' For men, see Figure~\ref{fig:plotcohm}
%' <<plotcohm, fig.cap = "Birth cohorts, men, life after 40. Blue is 1+2, red is 6+7.",eval=FALSE>>=
%' flm <- foli[foli$sex == "male", ]
%' oldpar <- par(mfrow = c(2, 2))
%' fit1 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1921-30", ])
%' plot(survfit(fit1), fun = "cumhaz", main = "1921-30", col = cols)
%' fit2 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1931-40", ])
%' plot(survfit(fit2), fun = "cumhaz", main = "1931-40", col = cols)
%' fit3 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1941-50", ])
%' plot(survfit(fit3), fun = "cumhaz", main = "1941-50", col = cols)
%' fit4 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1951-60", ])
%' plot(survfit(fit4), fun = "cumhaz", main = "1951-60", col = cols)
%' par(oldpar)
%' @
%' 
%' For women, see Figure~\ref{fig:plotcohw}
%' <<plotcohw, fig.cap = "Birth cohorts, women, life after 40. Blue is 1+2, red is 6+7.",eval=FALSE>>=
%' flw <- foli[foli$sex == "female", ]
%' oldpar <- par(mfrow = c(2, 2))
%' fit1 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1921-30", ])
%' plot(survfit(fit1), fun = "cumhaz", main = "1921-30", col = cols)
%' fit2 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1931-40", ])
%' plot(survfit(fit2), fun = "cumhaz", main = "1931-40", col = cols)
%' fit3 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1941-50", ])
%' plot(survfit(fit3), fun = "cumhaz", main = "1941-50", col = cols)
%' fit4 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1951-60", ])
%' plot(survfit(fit4), fun = "cumhaz", main = "1951-60", col = cols)
%' par(oldpar)
%' @
%' 
%' Another way of analysing mortality is by taking a cohort approach. The
%' cohorts from the decades 1920s until 1950s are followed. The first cohort
%' can be followed until very high age. For the age up to 60, representing the
%' period 1960 to 1989 approximately, the figure do not show any large
%' differences (which can have to do with scale). When they had passed 60, we
%' find clear social differences in the expected direction for men, but less
%' clear for women. The other cohorts follow rather similar patterns as the
%' oldest one.

\section{Discussion}

The authors of this paper have scrutinised different aspects of the social
pattern of mortality in northern Sweden for quite some time, focusing on
different regions, different age groups and partly different social
categories. Already the earliest findings, concerning the social inequality
in a rapidly growing town in an industrial district, surprised and at least
at first confused the author \citep{edv92}. The expected social gradient in
survival, meaning that people from higher social classes always had an
advantage, was in many cases, particularly among adult men, not found
despite the well-documented socio-economic differences in this
society. There were however large differences between men and women and for
different age groups. Further studies have confirmed these
findings \citep{edli11, edbr12}. These earlier studies have ended with an
assumption that more distinct differences to the advantage of higher
classes developed somewhere during and probably quite early in the 20th
century. Until now we have however not been able to investigate the
continuous development on mortality differences among the adult and elderly
population due to no available data. This is the first time it has been
possible, and the results continue to surprise us. The consistent advantage
for higher social classes do not appear until quite late in the regions we
have analysed. The social pattern of female mortality are closer to what is
expected, but a social gradient in male adult mortality seems to be a
recent phenomenon.

The perhaps surprising results makes it reasonable to consider their
validity and possible weaknesses in the study. There are of course
uncertainties involved in all studies of this sort, something that should
not be ignored.  One obstacle concerns the quality of the available
information in the sources. Occupational titles in historical sources are
not always distinct enough to define the correct class. It can be difficult
to identify the skill level in the working class population because many
were often only designated as labourer. The size of the enterprise for free
entrepreneurs is unknown and it may be impossible to distinguish if they
are self-employed or have many employed. The class system and the position
of different classes can also be discussed. Studies covering long period
have the problem that occupations and classes change character - they do
not represent the same thing. Nonetheless, these ambiguities and possible
mis-classifications are not of that sort and magnitude that it would radically change
the basic patterns observed. The social classes identified during the
different periods represent distinct groups with very large differences in
resources, capacities and power. A worker would not end up in the highest
SOCPO level and large-scale entrepreneurs and higher officials would
certainly be assigned the right level.Available data also show that the
differences in income and wealth were large, often enormous.

In this paper we have analysed social inequality in health from the
perspective of social classes using HISCLASS. However, there are
alternative ways to analyse social inequality, and we cannot exclude that
such analyses gives other results or at least would modify the image
presented here. There are other alternatives then HISCLASS, representing
status or class dimensions. SOCPO is another system based on hierarchic
categories of social power, which we have used in a couple of previous
studies. In many modern studies categorisations based on income, wealth
or education are common for studies of inequality. It is possible that
such categorisations are better fit to distinguish between groups in
society. They represent different dimensions of inequality, and we believe
that the topic discussed here would gain from analyses of different aspects
of inequality. One problem is however that reliable data on income and
education are rarely available in historical population databases. We can
however be certain that the highest HISCLASSES level have more education
and the lowest the least. We also know from the 1930 census that the
economic differences were very large at that time and that occupations in
higher HISCLASSES tend to have higher incomes. The groups that may diverge
are those within agriculture and those having a small enterprise where we
can assume that the reported income figures underestimate their economic
standard.

The results are however in accordance with other studies \citep{bevp11},
for example by \citet{bedr11} in a very different setting, a couple of
Scanian parishes. They found the first indications of an emerging social
gradient in the 1950s.The similarities makes us more comfortable about our
results.

\citet{who10} lists three different types of intermediary social
determinants of health, material circumstances, behaviours and biological
factors, and finally psycho-social factors. Most studies have focused on
the material circumstances. Having a better economy can give advantages in
many aspects of relevance for health and mortality, for example good living
conditions, access to food, medical care, better conditions at work et
cetera. Even though these advantages ought to have been at hand in our
studied regions, the final result was not better health for men in the most
privileged groups until the later part of the 20th century. We must
consider other intermediary determinants. Psycho-social factors have been
frequently discussed during the most recent decades, for example by
\citet{mar04}. He argues that being higher in the status ladder leads to
better health, regardless of where in the status ladder we look. This is
also one of the arguments \citet{wipi09} use as an explanation why equal
societies always perform better than unequal ones when it comes to health
and other social conditions. In the case studied here, we do not find any
indication of a status syndrome, since mortality levels was in the opposite
directions to the one we would expect from these theories.

This makes us consider the possible impact of behavioural factors. Our results of the gender aspect might be of interest in this case. When it comes to gender differences, the present study confirms what we have found on the Sundsvall region. There is a clear gendered pattern when it comes to social inequality in mortality. While high HISCLASS did not have any positive impact on survival for men in the early 20th century, we find  that higher social
positions among women led to longer lives. Within the same households, the effect of social
class seems to lead to different results. This opens up for behavioural
aspects based on different gendered expectations. Alcohol has been
suggested as part of the large gender differences in the 19th century
\citep{wil99, edv92}, and a more risk-taking life may be a component of the
male mortality.

Having a high social status could be expressed in attitudes and behaviours
that were harmful for health. Alcohol consumption decreased rapidly during
the period late 19th century to the 1950s, thus possibly making room for an 
increased role of economic resources. But the old pattern in social health
inequalities remained. It is however possible that the relative differences
in alcohol consumption prevailed, although at a lower level. The question 
is however if the lower levels of alcohol consumption had any substantial
negative impact on health. Swedes at that time had very low consumption
levels. Alternative explanations should also be scrutinise. There 
may have been strong and heavy requirements of being successful for the
male \"elite. Many were involved in businesses with high risks. It would be
very damaging to lose the high position and they may have felt that they 
lived under constant threat of economic failure. There are several examples among
the economic \"elite in the Sundsvall region that they had such problems
\citep{edli11}. A possible future study involving analyses of death causes
might throw some light on this issue.

When it comes to the impact of social factors in the working age population versus the elderly, we do not find any fundamental differences. The social class have somewhat smaller effect among the elderly, so there is some difference in levels but not in character.

Our study illustrates the importance of considering contextual factors and
the restrictions of making general statements. We need to look into all
aspects of social classes, not only the material ones but also what this
means for behaviour, norms, way of living. \citet{edli11} suggested that an
explanation for the larger social health inequalities in the present-day
world is that status and higher social position is now expressed through
health and being fit and physical active, while in many historical contexts
higher social status was expressed by low physical activity, obsessive
eating (being fat was impressive), drinking, smoking and sometimes also
taking risks.

Another question is if our results can be generalised. We have studied
environments in northern Sweden without any large metropolis. Perhaps
Sweden and/or this part of Sweden is a special case. The social differences
may have been smaller, the involvement of government both at the national
and local level may have been stronger, thus mitigating socio-economic
disadvantages. These are relevant questions, which hopefully can be
addressed in other and maybe comparative studies.

Finally, we argue that our results have implications for our understanding
on several hot research issues. A central question concerning the mortality
transition is how the enormous increase in survival could be
accomplished. This is still a hot debated question. \citet{mck76} argues that the
improvements up to the middle of the 20th century was first of all made
possible by improved nutrition. This seems to implicate that having more
resources would lead to better survival. If research shows that the most
advantaged groups did not have better health and longer lives, this do not
necessarily speak against McKeown's ideas but it makes the question more
complex. The results are also difficult to fit into \citet{liph95}
hypothesis that social conditions should be seen as a fundamental cause of
disease. Even if we would eliminate disadvantages such as bad living
conditions, lower standard of medical care and other proximate determinants
for the poor, the health disadvantage would remain. In the historical
context we have analysed here, this does not get any support. This further
illustrate the need for situating the analyses in their historical and
geographical contexts.

In recent years, the question about the relation between health and income
inequality been widely discussed. Even though our study do not address this
issue directly, our results are of interest for this question as well. The
development of the social inequality in mortality do not seem to fit that
well into the development of income and wealth inequalities. The early 20th
century, in Sweden as well as in other countries, was characterised by very
large economic differences, still this did in our regions not lead to any
consistent advantage for the highest social classes. After the Second World
War, economic differences diminished, and was at its lowest in Sweden at
about 1980, i.e. at a time when a more expected pattern of a social
hierarchy in survival had appeared. The period thereafter is characterised
by increasing economic differences, and this may coincide with increasing
health differences as well. The long-term relation between economic
inequality and health is worth further studies.

\section{Conclusion}

We started this paper by raising the issue of the development of social
inequalities. When studying this question in history we often get surprised
of its complexity. In a society characterised by large economic inequality, 
we would expect that this would result in corresponding mortality
differences, i.e. higher social classes having an advantage. We find the
appearance of a consistent social pattern in mortality, to the advantage of
the higher classes, but surprisingly late in our history. It is only in the
recent decades that this pattern dominates. Improved survival characterised
however all HISCLASSes. We can however confirm that there are gender
differences in the social pattern. While there is no clear social gradient
among men, the results at least partly indicate advantages for the higher
social classes for women already early in the 20th century. Females had
more to gain from a high social position. Previous studies of the mortality
transition have made similar observations. Among adults, gender has often
been of more significant importance than social class. The reason for this
difference and for the late appearance of the modern social pattern among
men is still an enigma that require further studies, even though we have
suggested the possible importance of behavioural factors and changes in
ideals connected to different social classes. This should stimulate us to
further disentangle the pathways for how social position influence health
and mortality in different societies. Finally, we do not find any clear
differences in the social pattern in the two studied age groups.

%\newpage

%\section{Reproducibility declaration}

%The exact versions of tools used in the analyses are shown here.

%<<reproducibility, echo = TRUE>>=
%sessionInfo()
%@

%\newpage

\section*{Acknowledgements}

This work was undertaken within the Centre for Demographic and Ageing
Research (CEDAR) at Umeå University and with support from the Swedish
Research Council (VR) (grant no. 421-2012-892). We thank Demographic Data
Base for giving us access to data and Jonas Helgertz, Department of
Economic History, Lund University, for providing us with scripts for
creating HISCLASS from Swedish Census data.

\bibliography{mybib}

\end{document}

### 2016-09-22 R/addPer.R
first commit

@@ -1,0 +1,22 @@
+addPer <- function(dat, name = "period", cuts = seq(1901, 1951, by = 10)){
+    n <- length(cuts)
+    start <- cal.window(dat, c(cuts[1], cuts[2]))
+    start[, "period"] <- 1
+    if (n > 2){
+        for (i in (3:n)){
+            tmp <- cal.window(dat, c(cuts[i - 1], cuts[i]))
+            tmp[, "period"] <- i - 1
+            start <- rbind(start, tmp)
+        }
+    }
+    levs <- character(n - 1)
+
+    for (i in 1:(n - 1)){
+        levs[i] <- paste(cuts[i], "-", cuts[i + 1] - 1, sep = "")
+    }
+    start[, "period"] <- factor(start[, "period"], labels = levs)
+    start <- start[order(start$id, start$enter), ]
+    start <- rc(start)
+    start
+}
+

### 2016-09-22 R/ageGroup.R

@@ -1,0 +1,17 @@
+ageGroup <- function(dat, ages){
+    n <- length(ages) - 1
+    ##age <- ages[-(n+1)]
+    if (n > 1){
+        library(eha)
+        dat$age <- numeric(NROW(dat))
+        first <- age.window(dat, ages[1:2])
+        first$age <- ages[1]
+        for (i in 2:n){
+            tmp <- age.window(dat, ages[i:(i+1)])
+            tmp$age <- ages[i]
+            first <- rbind(first, tmp)
+        }
+    }
+    first <- first[order(first$id, first$enter), ]
+    invisible(first)
+}

### 2016-09-22 R/getH.R

@@ -1,0 +1,19 @@
+getH <- function(x, tid){
+    ## 'x' is a vector of coefficients from a stratified (on hisclass, 7 strata) gompertz model.
+    ## Components 1:14 are the baseline parameters (log(scale1), log(shape1), log(scale2), etc)
+    
+    ## 'tid' is the age (or duration!) at which H is evaluated.
+    
+    library(eha)
+
+    x <- x[which(substr(names(x), 1, 3) == "log")] # New 8 Sep 2016!!
+           
+    x <- exp(x[1:14]) # Note!
+    
+    ret <- numeric(7) # NOTE!
+    
+    for (i in 1:7){
+        ret[i] <- Hgompertz(tid, scale = x[2 * i - 1], shape = x[2 * i], param = "canonical")
+    }
+    ret
+}

### 2016-09-22 R/getH.ph.R

@@ -1,0 +1,12 @@
+getH.ph <- function(x){
+    ## 'x' is a fit from a stratified (on hisclass, 7 strata) coxph model.
+    ## covariates civst and urban
+    ## 'tid' is the age (or duration!) at which H is evaluated. Last!!!
+    
+    library(survival)
+    y <- survfit(x, new.data = data.frame(civst = "unmarried", urban = FALSE))
+    hupp <- cumsum(y$strata)
+##    ret <- 1 - y$surv[hupp]
+    ret <- -log(y$surv[hupp]) # Cumulative hazard at the 'end'.
+    ret
+}

### 2016-09-22 R/getPH.R

@@ -1,0 +1,16 @@
+getPH <- function(adu, fo, nper1, nper2, labb, from = 40){
+    ##cat("names(adu) = ", names(adu), "\n")
+    res <- vector(mode = "list", length = nper1 + nper2)
+
+    for (i in 1:nper1){
+        res[[i]] <- coxph(Surv(enter-from, exit-from, event) ~ strata(hisclass) + civst + urban,
+                           data = adu[adu$period == labb[i], ], model = TRUE) 
+    }
+
+    for (i in 1:nper2){
+        i2 <- i + nper1
+        res[[i2]] <- coxph(Surv(enter-from, exit-from, event) ~ strata(hisclass) + civst + urban, 
+                           data = fo[fo$period == labb[i2], ], model = TRUE)
+    }
+    res
+}

### 2016-09-22 R/getPHglm.R

@@ -1,0 +1,34 @@
+getPHglm <- function(dat, nper, labb, from = 40){
+    ## dat$age must be a factor ...
+    ## Note: return value is Cumulative hazards at given age points (straight line between).
+    ##
+    ## 'dat$age' is assumed to be grouped in equidistant intervals!
+    res <- vector(nper, mode = "list")
+    hc <- levels(dat$hisclass)
+    m <- length(hc)
+    k <- length(unique(dat$age))
+    if (k <= 1) error("No. of age intervals must be at least 2!")
+    ageName <- as.numeric(levels(dat$age))
+    ageInt <- ageName[2] - ageName[1]
+    levs <- matrix(0, nrow = m, ncol = k + 1)
+   
+    for (i in 1:nper){
+   
+        for (j in 1:m){
+   
+            fit <- glm(event ~ offset(log(exposure)) +  age +
+                           civst + urban,
+                       data = dat[dat$period == labb[i] &
+                                  dat$hisclass == hc[j], ], family = poisson)
+   
+            lupp <- fit$coef[1:k]
+            lupp[2:k] <- lupp[2:k] + lupp[1]
+            levs[j, ] <-  cumsum(c(0, exp(lupp)) * ageInt) 
+        }
+        colnames(levs) <- c(ageName, ageName[k] + ageInt)
+        rownames(levs) <- hc
+        res[[i]] <- levs
+    }
+    names(res) <- labb
+    res
+}

### 2016-09-22 R/plotED.glm.R

@@ -1,0 +1,54 @@
+plotED.glm <- function(res,
+                       scol = 1:4,
+                       logScale = FALSE,
+                       ylab = "",
+                       main = ""){
+    
+    ## 'res' is a list of matrices of cumulative hazards from a piecewise-contant fit
+    ## We want the last column of each matrix.
+
+    ##nper <- NCOL(res[[1]])
+
+    nper <- length(res)
+    k <- NCOL(res[[1]])
+    m <- NROW(res[[1]])
+    labb <- names(res)
+    
+    outp <- matrix(0, ncol = nper, nrow = m)
+
+    for (i in 1:nper){
+        outp[, i] <- res[[i]][, k]         
+    }
+    
+    if (logScale){
+        outp <- log(outp)
+    }
+    fr <- min(outp, 0)
+    to <- max(outp)
+    xv <- as.numeric(substr(names(res), 1, 4)) + 5 ## NOTE: assuming ten-year periods!
+
+    plot(xv[1:5], outp[1, 1:5], type = "b", lty = 2, col = scol[1], pch = 1,
+         ylim = c(fr, to), main = main,
+         axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
+         xlim = c(min(xv), max(xv) + 10))
+    lines(xv[7:9], outp[1, 7:9], type = "b", lty = 2, col = scol[1], pch = 1)
+    axis(1, at = xv, lab = labb, las = 2)
+    axis(2, las = 2)
+    box()
+    for (i in 2:m){
+        points(xv, outp[i, ], col = scol[i], pch = i)
+        if (i == 2){
+            lines(xv[1:5], outp[i, 1:5], col = scol[i], lty = 2)
+            lines(xv[7:9], outp[i, 7:9], col = scol[i], lty = 2)
+        }else{
+            if (i %in% c(6, 7)){
+                lines(xv[1:5], outp[i, 1:5], col = scol[i], lty = 4)
+                lines(xv[7:9], outp[i, 7:9], col = scol[i], lty = 4)
+            }
+        }
+    }
+    abline(v = xv, lty = 3, col = "magenta")
+    abline(h = 0)
+    legend("topright", legend = 1:4, col = scol, pch = 1:4)
+    invisible(outp)
+}

### 2016-09-22 R/plotED.ph.R

@@ -1,0 +1,53 @@
+plotED.ph <- function(res,
+                   scol = 1:6,
+                   logScale = FALSE,
+                   ylab = "",
+                   main = ""){
+    ## 'res' is a list of results from a fit.
+    source("R/getH.R")
+    outp <- matrix(0, ncol = nper, nrow = 6)
+
+    n <- length(res)
+    for (i in 1:nper){
+        sfi <- survfit(res[[i]], new.data = data.frame(civst = "unmarried", urban = FALSE))
+        x <- cumsum(sfi$strata)
+        
+        ##outp[, i] <- 1 - sfi$surv[x] ### NOTE!!!
+        outp[, i] <- -log(sfi$surv[x])
+    }
+##    for (j in 1:7){
+  ##      outp[j, ] <- outp[j, ] /outp[5, ]
+    ##}
+    
+    if (logScale){
+        outp <- log(outp)
+    }
+    fr <- min(outp, 0)
+    to <- max(outp)
+    xv <- c(seq(1905, 1945, by = 10), seq(1965, 1995, by = 10))
+
+    plot(xv[1:5], outp[1, 1:5], type = "b", lty = 2, col = scol[1], pch = 1,
+         ylim = c(fr, to), main = main,
+         axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
+         xlim = c(min(xv), max(xv) + 10))
+    lines(xv[7:9], outp[1, 7:9], type = "b", lty = 2, col = scol[1], pch = 1)
+    axis(1, at = xv, lab = labb, las = 2)
+    axis(2, las = 2)
+    box()
+    for (i in 2:6){
+        points(xv, outp[i, ], col = scol[i], pch = i)
+        if (i == 2){
+            lines(xv[1:5], outp[i, 1:5], col = scol[i], lty = 2)
+            lines(xv[7:9], outp[i, 7:9], col = scol[i], lty = 2)
+        }else{
+            if (i %in% c(6, 7)){
+                lines(xv[1:5], outp[i, 1:5], col = scol[i], lty = 4)
+                lines(xv[7:9], outp[i, 7:9], col = scol[i], lty = 4)
+            }
+        }
+    }
+    abline(v = xv, lty = 3, col = "magenta")
+    abline(h = 0)
+    legend("topright", legend = 1:6, col = scol, pch = 1:6)
+    invisible(outp)
+}

### 2016-09-22 R/plotHaz.glm.R

@@ -1,0 +1,33 @@
+plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
+                    xlab = "", main = "", layout = c(4, 3)){
+    ## 'res' is list of results from a fit.
+    ## Note, number of levels of hisclass is hardcoded to 4 (four) here!!
+    ## Will be fixed.
+    
+    oldpar <- par(mfrow = layout)
+    on.exit(par(oldpar))
+    n <- length(res)
+
+    if (length(main) == 1){
+        main = rep(main, n)
+    }
+    
+    
+    for (i in 1:n){
+        x <- as.numeric(colnames(res[[i]]))
+        #x <- c(x, x[length(x)] + (x[2] - x[1]) )
+        
+        y <- res[[i]]
+        ymax <- max(y)
+        plot(x, y[1, ], col = scol[1], type = "l", lty = 1,
+              main = names(res)[i], ylim = c(0, ymax),
+             xlab = "Age", ylab = "Cum. Hazards")
+        for (j in 2:4){
+            lines(x, y[j, ], col = scol[j], lty = j)
+        }
+        legend("topleft", legend = 1:4, col = scol, lty = 1:4, cex = 0.9)
+    }
+    abline(h = 0)
+}
+
+    

### 2016-09-22 R/plotHaz.ph.R

@@ -1,0 +1,20 @@
+plotHaz.ph <- function(res, printLegend = FALSE, col = 1:6,
+                    xlab = "", main = "", layout = c(3, 3)){
+    ## 'res' is list of results from a fit.
+    oldpar <- par(mfrow = layout)
+    on.exit(par(oldpar))
+    n <- length(res)
+
+    if (length(main) == 1){
+        main = rep(main, n)
+    }
+    
+    for (i in 1:n){
+        plot(survfit(res[[i]]), fun = "cumhaz", col = scol, 
+             xlab = xlab, main = main[i])
+        legend("topleft", legend = 1:6, col = scol, lty = 1:6, cex = 0.9)
+    }
+    abline(h = 0)
+}
+
+    

### 2016-09-22 oldmort.Rnw
fix file names, more remains
make working

@@
 
-\usepackage[tablesonly]{endfloat}
+%\usepackage[tablesonly]{endfloat}
 \usepackage{endfloat}
@@
 The {\sc Lisa} data is yearly, with {\tt hisclass} information starting in 2002.
-This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, and 5 + 6.
+This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, 5, and 6.
 
@@
 <<readtab>>=
-load("vbTab.rda")
+load("../CreateData/data4/vbTab.rda")
 @
@@
 ## Colors for plotting strata:
-scol <- c("blue", rep("green", 2), "red")
+scol <- c("blue", rep("darkgreen", 2), "red")
 ## Common for all groups:
@@
 scap12 <- "Total hazard by HISCLASS, ages 40--59."    
-cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). Log scale."
-scap12log <- "Total hazard by HISCLASS, ages 40--59 (log scale)."
+cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
+scap12log <- "Total hazard by HISCLASS, ages 40--59, last years."
 
@@
 scap22 <- "Total hazard by HISCLASS, ages 60--89."    
+cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
+scap22log <- "Total hazard by HISCLASS, ages 60--89, last years."
 
@@
 
-<<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 10>>=
+<<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
 source("R/plotHaz.glm.R")
-plotHaz.glm(resw, main = labb)
-@
-
-<<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 10>>=
-plotHaz.glm(resm, main = labb)
+plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+@
+
+<<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
+plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
 @
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, scol = scol, logScale = FALSE, ylab = "H(60|40)",
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(60|40)",
                 main = "Women")
-moutp <- plotED.glm(resm, scol = scol, logScale = FALSE, ylab = "H(60|40)",
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(60|40)",
                 main = "Men")
@@
 
+In Figure~\ref{fig:cumslogw} a "blow-up" of the last decades show clearly the upper hand of the upper classes.
+
 <<cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
 oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
+wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(60|40)",
 main = "Women")
-mloutp <- plotED.glm(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
+mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(60|40)",
 main = "Men")
@@
 <<lissyoung>>=
+load("../CreateData/data4/liss.rda")
 liss4059 <- age.window(liss, c(40, 59))
@@
 
-<<liswom>>=
+<<liswom, cache = TRUE>>=
 fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "female", ])
@@
 See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<tablisswom1, results = "asis">>=
-fitwo1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
+<<tablisswom1, results = "asis", cache = TRUE>>=
+fitwo1 <- coxreg(Surv(enter, exit, event) ~ factor(hisclass) + civst + urban + period, 
                  data = liss4059[liss4059$sex == "female", ])
@@
 See Table~\ref{tab:tablissman1} the PH regression for men.
-<<tablissman1, results = "asis">>=
-fitma1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
+<<tablissman1, results = "asis", cache = TRUE>>=
+fitma1 <- coxreg(Surv(enter, exit, event) ~ factor(hisclass) + civst + urban + period, 
                  data = liss4059[liss4059$sex == "male", ])
@@
 <<hazcalwold>>=
-resw <- getPH(vb6089[vb6089$sex == "female", ],
+resw <- getPHglm(vb6089[vb6089$sex == "female", ],
               nper, labb)
-resm <- getPH(vb60896089[vb6089$sex == "male", ],
+resm <- getPHglm(vb6089[vb6089$sex == "male", ],
                 nper, labb)
@@
 
-<<hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 10>>=
+<<hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 7>>=
 ##source("R/plotHaz.R")
-plotHaz.glm(resw, main = labb)
-@
-
-<<hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 10>>=
-plotHaz.glm(resm, main = labb)
+plotHaz.glm(resw[-(1:6)], main = labb, layout = c(2, 2), col = scol)
+@
+
+<<hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 7>>=
+plotHaz.glm(resm[-(1:6)], main = labb, layout = c(2, 2), col = scol)
 @
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, scol = scol, logScale = FALSE, ylab = "H(90|60)",
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|60)",
                 main = "Women")
-moutp <- plotED.glm(resm, scol = scol, logScale = FALSE, ylab = "H(90|60)",
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|60)",
                 main = "Men")
@@
 
+A blow-up of the last years is shown in Figure~\ref{fig:cumslogwm}.
+
+<<cumslogwm, fig.cap = cap22log, fig.scap = scap22log, fig.height = 10>>=
+oldpar <- par(mfrow = c(2, 1))
+wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|60)",
+main = "Women")
+mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|60)",
+main = "Men")
+par(oldpar)
+@
+
 %% The compression that is visible for later decades is partly explained by the decereasing general mortality. This effect can be eliminated by showing these numbers on a log scale.
@@
 %% oldpar <- par(mfrow = c(2, 1))
-%% wloutp <- plotED.ph(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
+%% wloutp <- plotED.ph(resw, col = scol, logScale = TRUE, ylab = "H(60|40)",
 %% main = "Women")
-%% mloutp <- plotED.ph(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
+%% mloutp <- plotED.ph(resm, col = scol, logScale = TRUE, ylab = "H(60|40)",
 %% main = "Men")

### 2016-09-22 R/plotHaz.glm.R
simplified

@@ -19,13 +19,13 @@ plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
         
         y <- res[[i]]
         ymax <- max(y)
-        plot(x, y[1, ], col = scol[1], type = "l", lty = 1,
+        plot(x, y[1, ], col = col[1], type = "l", lty = 1,
               main = names(res)[i], ylim = c(0, ymax),
-             xlab = "Age", ylab = "Cum. Hazards")
+             xlab = "Age", ylab = "Cum. Hazards", lwd = 2)
         for (j in 2:4){
-            lines(x, y[j, ], col = scol[j], lty = j)
+            lines(x, y[j, ], col = col[j], lty = j, lwd = 2)
         }
-        legend("topleft", legend = 1:4, col = scol, lty = 1:4, cex = 0.9)
+        legend("topleft", legend = 1:4, col = col, lty = 1:4, cex = 1.0)
     }
     abline(h = 0)
 }

### 2016-09-22 R/plotED.glm.R
simplified

@@ -1,5 +1,5 @@
 plotED.glm <- function(res,
-                       scol = 1:4,
+                       col = 1:length(res),
                        logScale = FALSE,
                        ylab = "",
                        main = ""){
@@ -25,30 +25,22 @@ plotED.glm <- function(res,
     }
     fr <- min(outp, 0)
     to <- max(outp)
-    xv <- as.numeric(substr(names(res), 1, 4)) + 5 ## NOTE: assuming ten-year periods!
+    xv <- (as.numeric(substr(names(res), 1, 4)) +
+        as.numeric(substr(names(res), 6, 9))) / 2
 
-    plot(xv[1:5], outp[1, 1:5], type = "b", lty = 2, col = scol[1], pch = 1,
+    plot(xv, outp[1, ], type = "b", lty = 2, col = col[1], pch = 1,
          ylim = c(fr, to), main = main,
          axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
          xlim = c(min(xv), max(xv) + 10))
-    lines(xv[7:9], outp[1, 7:9], type = "b", lty = 2, col = scol[1], pch = 1)
     axis(1, at = xv, lab = labb, las = 2)
     axis(2, las = 2)
     box()
     for (i in 2:m){
-        points(xv, outp[i, ], col = scol[i], pch = i)
-        if (i == 2){
-            lines(xv[1:5], outp[i, 1:5], col = scol[i], lty = 2)
-            lines(xv[7:9], outp[i, 7:9], col = scol[i], lty = 2)
-        }else{
-            if (i %in% c(6, 7)){
-                lines(xv[1:5], outp[i, 1:5], col = scol[i], lty = 4)
-                lines(xv[7:9], outp[i, 7:9], col = scol[i], lty = 4)
-            }
-        }
+        points(xv, outp[i, ], col = col[i], pch = i)
+        lines(xv, outp[i, ], col = col[i], lty = i)
     }
     abline(v = xv, lty = 3, col = "magenta")
     abline(h = 0)
-    legend("topright", legend = 1:4, col = scol, pch = 1:4)
+    legend("topright", legend = 1:m, col = col, pch = 1:m, lty = 1:m)
     invisible(outp)
 }

### 2016-09-25 oldmort.Rnw
add education and wages

@@
 %\usepackage[tablesonly]{endfloat}
-\usepackage{endfloat}
+\usepackage[figuresonly]{endfloat}
 
@@
 
-\section{Introduction}
-
-In this study we take a long-term perspective on the development of social
-inequalities in mortality in two Swedish region 1901--2013, in this case
-with focus on the adult and elderly population. We investigate the development of
-social inequality in mortality during the mortality transition and the
-possible changes in the impact of social class in this process, in
-particular the hypothesis that modern pattern of social differences in mortality appeared
-during the first half of the 20th century.
-
-Several studies have shown how important access to economic and other
-resources are for health, but it is not clear how this has developed in
-history. \citet{ant67} and \citet{smi83} suggest that social inequality in
-health has passed through different phases in history. According to them,
-social health differences were
-comparatively small during the pre-transitional phase when space was a
-strong determinant for the spread of disease. Differences then increased
-during the transitional phase when mortality declined and wealthy groups
-used their resources to gain better health. Finally, health differences
-decreased again in modern low-mortality settings when instead
-health-related behaviour became important, resulting in marginal
-differences. \citet{omran82} state, in the third proposition in his theory
-of the epidemiological transition, that even if the class differential in
-mortality was maintained during the transition, the decline
-set in earlier and was faster among privileged groups.
-
-When studies, stimulated by the widely debated Black report \citep{whi98},
-began to be published on the late 20th century development, any assumptions
-of declining and even disappearing differences could be refuted. Even in
-modern wealthy societies social health inequalities are substantial. In
-fact, it even seems as if (at least relative) differences are increasing
-\citep{kuetal04,frlu07}, contrary to what Antonovsky assumed. Research has
-revealed a complex pattern of social class, material conditions,
-psycho-social effects, and health-related behaviour. Many contemporary
-studies implicitly or explicitly assume a "historical constant" of social
-health inequality that often is considered to have been even greater in
-earlier times \citep{mar04, liph95}.
-
-It is however time to closely scrutinise the assumption about health and mortality inequalities in the past.
-Our knowledge about social health inequalities in history, is
-still based on scattered observations and usually not based on thorough
-analyses.
-The social aspects of health began to be observed a long time ago.
-The 19th century French physician and pioneer in epidemiology Villerm\'e found large differences in mortality between city districts of varying social composition in early 19th century Paris, and
-Engels gave evidence of appalling living conditions among the impoverished
-in the rapidly expanding industrial environments \citep{ack48}. Despite
-their merits as sources for 19th century society and their path-breaking
-ideas, these 19th century observations are problematic in many ways. They
-lacked adequate sources (no individual data) and methods for analysis. Another issue concerns
-the degree to which we can generalise these observations to other
-historical contexts.
-
-During recent years, however, increased attention has been paid to the historical
-perspective on social health inequalities. The results from different
-studies have diverged, some reporting substantial differences but many
-finding surprisingly small differences during the first stages of the
-transition \citep{bevp11}. Nevertheless, it is certainly not obvious
-that historical societies were characterised by the assumed positive
-association between socio-economic position and survival. Several traits in
-the pre- and early transitional society may explain this circumstance. The
-effect of social class is dependent both on patterns of exposure and
-resistance against diseases in different populations and sub-populations.
-In their overview on this topic \citet{bevp11} suggest that consistent hierarchical social health inequalities did not developed until the
-first half of the 20th century.
-
-Concerning Sweden, \citet{edli11} found no indication of better health
-among wealthy adults living in the Sundsvall region during the
-pre-industrial period 1803--1859. This pattern remained for men during the
-industrial period 1860--1900, while clearer social health differences
-appeared among women. A similar pattern was found for those in old age
-\citep{edbr12}. This illustrates the importance of considering gender as
-well as age patterns. Results from a study of other parts of northern Sweden,
-the Skellefteå  and Umeå regions, 1851--1950 was in accordance with the results
-from the Sundsvall region. Higher social position and more resources did not lead
-to any health advantages for men, while this was the case for women \citep{segb14}.
-This is further illustrated with what is known about mortality in younger age groups in the same areas.
-\citet{edv92, edv04} found weak or
-varying social differences in infant mortality in early 19th century northern Sweden, while industrialisation caused large social differences in mortality among children 1--14 years.
-In their study on socio-economic
-mortality differences in a couple of Scanian parishes 1815--1968,
-\citet{bedr11} found no indications of any social gradient during most of
-the period, suggesting that the increase is a recent phenomenon.
-
-To summarise the present state of knowledge about the development of social
-health inequalities (in Sweden but also in other European countries), no
-clear social hierarchy has been found before 1900 but substantial
-differences towards the late 20th and early 21st century. It appears as if
-a more consequent social health divide developed in the period
-in-between. This period has not been thoroughly studied on this issue,
-mainly because of lack of data to analyse. However, the extension of data at the Demographic Data Base, Umeå University (DDB) makes such studies possible. We are now able to analyse social
-differences in mortality with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.
-
-\section{Aims and questions}
-
-The aim of the paper is to investigate the development of social
-differences using HISCLASS in mortality for the adult and elderly population during the mortality
-transition and its relation to the development of income inequality. The analyses
-will be performed on data from the Skellefteå and Umeå regions
-1901--2013. The questions are:
-
-\begin{enumerate}
-\item Has inequality in mortality between social classes increased among the adult and elderly
-  population, primarily benefiting the highest social positions?
-
-%%\item Is the mortality transition initiated by a certain social class?
-
-\item Are there any gender differences in the effect of social position?
-
-\item Is social position equally important among the retired population as for those in working age?
-\end{enumerate}
-
-\section{Theoretical considerations}
-
-The WHO Commission on Social Determinants of Health framework for the
-social determinants of health discusses how socio-economic position impacts
-on the health of individuals, placing them within structures that determine
-their opportunities and how these influences works
-through intermediate determinants related to exposure and vulnerability
-\citep[p. 48]{who10}. People are situated within socio-economic and political
-contexts, as well as value systems in the society. The socio-economic
-position imply different access to resources, power and prestige, all
-aspects of crucial importance. The framework takes a holistic approach to
-the social determinants, including not only the effect of pure economic and
-material resources. The importance of understanding people in their full
-context has been emphasised by for example \citet{kri05}. What she calls embodiment refers to the recognition that human beings are
-simultaneously social beings and biological organisms.
-
-Social inequality can mean different things, and studies of social
-differences 
-in health and mortality often differ in what sort of categorisation is used.
-This is important to have in mind when comparing results, because the chosen
-social variable may measure different aspects of inequality. This relate to an
-old subject within social studies, on class and status variables, based on
-forerunners 
-such as Marx and Weber. In many modern studies, data on income, wealth
-and/or education are frequently used. The underlying assumption is that more access
-to these resources leads to advantages in life, for example when it comes to survival.
-SOCPO \citep{vdpm05} is a system based on different access to what they call social power.
-\citet{goldk10} criticise how many economists and epidemiologists operationalize
-inequality in their studies, for example by not distinguishing between class
-and status aspects. Class refers to relational aspects in society, while many other
-variables focus on the distribution of material and/or immaterial resources.
-He advocates studies of different aspects, which together will complement each other.
-
-HISCLASS, the classification system used here, is a "\ldots HISCO-based
-historical international social class scheme" \citep[p.\ 26]{lema11}. The
-different classes in HISCLASS represent distinct categories, based on
-whether the work is manual or non--manual, skill level, supervision and
-sector and it does not represent a purely hierarchical system, 
-but implicitly it reflects large differences in what is in the forefront of
-other categorisations, 
-as for example access to economic and other resources, status, social power et cetera.
-We can thus assume that classes have different opportunities when it comes to matters of health and mortality.
-
-A reasonable assumption is that levels of economic inequality in a society
-influences social patterns of health and mortality. Social classes with
-more economic resources have better possibilities to stay healthy. We
-expect that people from classes with less resources are seriously
-disadvantaged in a society with large income inequality. Their poverty
-would expose them to larger health risks. The inequality effect could
-however influence health in other ways than at the individual economic
-level. \citet{wipi09} suggest that income inequality have a contextual
-effect as well, meaning that living in an unequal society increases health
-risks above what is explained by the individual situation.
-
-How socio-economic position influences lives of people, their health and
-their mortality needs to be problematized. When it comes to differential
-access to resources, these could be of different character, and not only
-material ones. \citet{bou84} distinguishes between economic, 
-social and cultural capital. Economic capital allows for much more freedom
-to organise lives and to maximise life opportunities. Social capital (here
-referring to its individual variant and not as a description on society
-level) include different kinds of networks, for example family, friends,
-colleagues and other persons acting as valuable contacts and resources. A
-rich social network can be crucial when it comes to life opportunities. The
-impact of cultural capital on health may seem less obvious. Cultural
-capital refers both to the personal skills but also to prestige and
-status. This can be demonstrated in taste and choice of life style, which
-bring people together with others of similar background and
-interests. Achieving a higher cultural capital involve several requirements
-on way of living, something that can be relevant for health. Cultural
-capital in itself can also bring beneficial psycho-social effects on
-individuals, which could be an advantage for health \citep{mar04, wipi09}.
-
-In the historical part of this paper, individuals are grouped
-according to their family households, that is, children and married women are
-grouped according to the occupation of the head of household, usually the
-father or husband. A practical reason for this is that occupation is usually
-only provided for the head of household, but there are good research
-motives for this as well. We consider the whole family to belong to the
-same socio-economic position. But this does not mean that we necessarily
-believe that the effects are similar for both sexes. As several researchers
-have pointed out, demography has often ignored power relations and
-different roles within families \citep{watkins93, mack95, jans07}. By
-investigating possible differences in mortality patterns 
-according to sex, we may identify gendered patterns within the family, resulting in different effects on family members. Different patterns between the sexes
-may be caused by discrimination of the female part in the household. We
-know that the society was strongly segregated according to gender and that
-women were discriminated and had less space for action. Another possible
-cause for different patterns is the gendered expectations on behaviour that
-has characterised the relation between sexes in societies throughout
-history.
-Unfortunately, the same approach cannot be applied on the data from 1960 onward, due to difficulties identifying couples in the modern data.
-This is further developed in the section on the construction of data for analysis.
-
-\section{Historical background}
-
-The historical context of the present study is a period of major changes in
-Sweden and in the studied regions. Swedish society underwent an enormous
-development, changing from a rural and sparsely populated society based on
-agriculture and with low economic standard to a modern, industrialised and wealthy country.
-Material conditions improved and welfare provisions for people increased, mainly funded by public finances.
-Several traits of what was to become the Swedish welfare state had early
-predecessors, but became even stronger from the 1930s. These reforms
-fundamentally changed living conditions for all citizens but most of all for the
-most vulnerable, for example the elderly. Still, levels of income inequality differed during the studied period. The early 20th century was characterised by large income
-and wealth differences. This started to decline from the 1930s, but it was
-still high into the 1950s. It was only in the
-following decades that income inequality became very small in Sweden,
-making Sweden one of the most (or even the most) equal societies in the
-world around 1980 when inequality levels seem to have been at the lowest. During recent decades, inequality has however increased again \citep{pik14, rowa08, rowa09}.
-
-At the middle of the 19th century, Sweden belonged to the part of the world
-with the best survival, being number one in the 1840s, and has continued to belong to the countries with lowest mortality.  Life expectancy was 56 years 1901--10, remaining life
-expectancy at age 50 was 24 years.
-This increased to 72 versus 27 in 1951-59 and 82 versus 33 in 2013.
-Younger age groups contributed most to the initial and total decline, while improved survival in old age groups adds most to more years during recent decades.
-The First World War marks an
-important breaking point towards declining mortality among young adults, while the period after the Second World War led to large improvements for older adults.
-\citep{holu76}.
-If spatial differences determined mortality during the pre-transitional, regional variation and urban
-rural differences in mortality levels have diminished, although not disappearing completely.
-
-We do not argue that the Skellefteå and Umeå regions represent the whole of
-Sweden, but the availability of data allows us to study the transition
-in greater detail here making it a laboratory for studying Swedish demography during a period of fundamental changes. Both regions are part of the county of Västerbotten in the north of Sweden,
-along the coast of the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection of
-parishes surrounding the town of Skellefteå, founded in 1845 but with a
-very small population in the 19th century.
-The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
-The majority of the population in the region lived in rural villages and hamlets, getting their
-livelihood from agricultural production. The
-region was vulnerable for harvest failures, one living long in memory was
-the famine of 1867--1868 \citep{segb14}. Towards the end of the
-century, the region became connected to the Swedish railway system, thus
-improving the communications and contacts with other parts of the
-country. During the 20th century, industrialisation took place, partly by
-the opening of mines in the inlands. This also led to a population increase both in the town and in the rural parts,
-and a much more diversified economy.
-
-<<skpop>>=
-library(skum)
-ske1850 <- with(obs[obs$region == "ske", ],
-                sum(birthdate + enter < 1850 &
-                    birthdate + exit > 1850))
-ske1900 <- with(obs[obs$region == "ske", ],
-                sum(birthdate + enter < 1900 &
-                    birthdate + exit > 1900))
-ske1950 <- with(obs[obs$region == "ske", ],
-                sum(birthdate + enter < 1950 &
-                    birthdate + exit > 1950))
-
-@
-
-Mortality was fairly low in comparison
-with other parts of the country and the fertility transition was late. The
-Skellefteå population size as defined in our data-sets was \Sexpr{ske1850} on January 1, 1850,
-\Sexpr{ske1900} on January 1,
-1900, \Sexpr{ske1950} on January 1, 1950 and 76723 at the end of the century.
-
-\begin{figure}[ht!]
-\includegraphics[height=19cm]{myfigs/sweden}
-\caption{Skellefteå and Umeå in Sweden.}
-\end{figure}
-
-%%\section{Map. Skellefteå and Umeå regions}
-
-<<umpop>>=
-ume1900 <- with(obs[obs$region == "ume", ],
-                sum(birthdate + enter < 1900 &
-                    birthdate + exit > 1900))
-ume1950 <- with(obs[obs$region == "ume", ],
-                sum(birthdate + enter < 1950 &
-                    birthdate + exit > 1950))
-@
-
-The Umeå region in Poplink (until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
-This region had a somewhat different character from that of Skellefteå, even though agriculture was
-dominating the economy for a long time. The town of Umeå had a small population but
-substantially larger than Skellefteå town during almost the whole studied
-period. It was the administrative centre in the county of Västerbotten and
-schools and military regiments was placed here. Officials, teachers and the
-military were much more common in Umeå.
-Agriculture dominated in the rural part, while  and there were some foundries, some industries for example in forestry and small-scale production.
-Umeå had a much more diversified
-occupational structure than Skellefteå. The population size as defined in our data-sets was
-\Sexpr{ume1900} on January 1, 1900, \Sexpr{ume1950} on January 1, 1950 and 103970 when the century ended.
-
-\section{Data and variables}
-
-The data for the present study comes from two large population databases at
-the Demographic Data Base, Umeå University. The early period is covered by
-the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region \citep{wee16}. The population in the two
-regions is large enough and socially diverse to enable studies of social
-mortality inequalities. Poplink is based on linked parish records, allowing
-us to reconstruct life biographies on people as long as they remained in
-the region. The records are linked within but not between the regions. Data
-from the regions are accessed for the period 1901--1950s.
-
-The other large data-set comes from the Linnaeus database
-\citep{malm10}. It is based on different linked population registers from
-1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1960 to 2000 are constructed from censuses every fifth
-year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
-For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
-The construction of these data-sets for analysis is presented below.
-
-Information on individuals are anonymized and the two databases are not
-linked between each other. This stops us from following individuals present
-in both databases throughout their lives, thus they are treated separately. This
-also makes it impossible to add information on individuals in the Linnaeus
-database from what we potentially could find in Poplink, for example family
-background or occupations earlier in life.
-
-In the data-set analysed here, all individuals 40 years and older and ever being resident in either
-of the regions are included. The data file contains the variables on sex,
-birth date, death date, first and last date of observation and type of
-entrance/exit.
-
-The identification of presence periods as well as the time-dependent variables require some additional comments.
-
-\subsection{Presence periods}
-
-Differences in available information, both when it comes to structure and variables used for attributing social class, from the two data-sets as well as
-within the Linnaeus data-set makes necessary diverging approaches when it comes to
-the identification of presence periods within our
-regions. The Poplink data provides us with exact dates or at least year of
-start and exit of the presence of all individuals in our data. This is however not the case
-with the Linnaeus database. For the period 1960--2000 we use the presence
-in the censuses 1960--1990 together with the yearly population registers
-1986--2000 for defining the observation periods as well as their social
-class (see below) in our analyses. We have however no exact information on
-what happened in between the censuses, for example the exact date of exit
-for persons that moved between two censuses. We have solved this by
-considering everyone as resident in our regions during the complete
-five-year period between censuses if the person is missing in this
-census.This overestimates their presence in our regions, but should not
-bias the results. We furthermore miss those that both moved in and moved out
-or vice versa between censuses. Exit dates are exact if they refer to
-death, and in these cases we consider them as constantly present from the
-last census to the death data even if the death took place in another part
-of Sweden.
-
-For the last period, 2002--2013, we use the yearly population registers
-(also known as LISA data from Statistics Sweden), together with information
-on deaths from National Board of Health and Welfare. This data allows a
-better control of the presence periods in the studied regions.
-
-The presence periods, with start and end date, are reconstructed and
-divided according to two time-dependent variables: Social position and
-marital status. Thus, at each change in any of
-these variables, a new record is created. The time-dependent variables require some additional comments.
-
-\subsection{Social position}
-
-In this paper we analyse mortality differences according to social class
-according to the classification scheme HISCLASS. The only type of
-socio-economic information available both in Linnaeus and Poplink is
-occupations, which can be used to identify distinct social classes or
-social status groups. Thus, analysing according to social class is apart
-from a scientific choice, also a practical one. Using occupational data is
-the only and obvious alternative if we are to make a longitudinal study of
-socio-economic mortality differences with consistent classification schemes
-(although not without some problems). There are also other problems
-connected with using the same classification scheme covering a long time
-period, but this will be approached in more detail in the discussion
-section. Social position is treated as a time-dependent variable.
-
-The availability to information on occupations varies over time and between sources.
-The Poplink data usually provides information on the occupation of head of households.
-The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
-The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information.
-In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid.
-Thus, the social class in the census 1990 is considered to be kept until 2000.
-
-The use of different data sets, forces us to apply different approaches towards defining the social position.
-In the Poplink data, social class is based on the occupational titles given in the parish records.
-These can be more or less precise.
-There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
-We consider however their validity to be sufficient for identification of social position.
-Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}.
-In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
-This strategy makes the problem of under-recording less problematic.
-
-Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers.
-Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database.
-Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period.
-It is also worth to notice that female labour force participation became the norm, and we could therefore argue that the occupation of the male as signifier of the social class became obsolete.
-We therefore let all included persons be signified by their individual occupation for this period.
-Still, we need to keep in mind that results for the different periods are not completely comparable.
-
-Due to the construction of the time-dependent data, some records report unknown social position.
-This has been updated according to rules in order to optimise available information and minimise missing information.
-Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
-The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.
-
-The classification scheme used here is HISCLASS \citep{lema11}, a class-based system.
-The scheme uses HISCO as a basis for assigning categories.
-HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO.
-HISCO groups occupations according to tasks and skills.
-HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions.
-This is particularly apparent when analysing the development in separate periods where for example farm workers become more and more rare during the studied century.
-We have therefore collapsed some categories, ending up with the following six:
-
-\begin{enumerate}
-\item \'Elite, HISCLASS 1 and 2, Higher managers and higher professionals.
-\item Lower managers, HISCLASS 3.
-\item Farmers and fishermen, HISCLASS 8.
-\item Lower professionals, HISCLASS 4, 5 and 6, Lower professionals, lower clerical, foremen.
-\item Medium skilled workers, HISCLASS 7.
-\item Lower skilled workers, HISCLASS 9, 10, 11 and 12, Lower and unskilled workers (including farm workers)
-%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
-\end{enumerate}
-
-In the last studied period, we have brought together the categories even more.
-
-For the Poplink data, we have used the script recoding HISCO into HISCLASS.
-The procedure for the occupations in the Linnaeus database is a little more complicated.
-The occupations in the censuses (NYK) must first be transferred to the system later used, SSYK (which is the Swedish version of ISCO88).
-Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013 can be recoded into HISCLASS.
-We have used a script provided by Jonas Helgertz, Lund University for this transformation.
-
-\subsection{Marital status}
-
-Marital status is treated as a time-dependent variable in the data file.
-The statuses used are unmarried, married and dissolution of marriage.
-The status unmarried is often not explicitly stated, which makes the status unknown in the database Poplink for people for people moving to our parishes.
-If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
-%'
-%'
-%' <<getdata, results='asis'>>=
-%' library(xtable)
-%' library(knitr)
-%' load("data/adult.rda")
-%' @
-%'
-%'
-%' <<toonerec>>=
-%' adult <- rc(adult)
-%' ##mad <- adult
-%' ##indx <- with(mad, tapply(id, id))
-%' ##mad$Enter <- with(mad, tapply(enter, id, min))[indx]
-%' ##mad$Exit <- with(mad, tapply(exit, id, max))[indx]
-%' ##mad$Event <- with(mad, tapply(event, id, max))[indx]
-%'
-%' ##mad <- mad[mad$lopnr == 1, ]
-%' ##mad$enter <- mad$Enter
-%' ##mad$exit <- mad$Exit
-%' ##mad$event <- mad$Event
-%' ##mad$Enter <- mad$Exit <- mad$Event <- NULL
-%'
-%' ##mad <- mad[, c("id", "enter", "exit", "event", "birthdate",
-%'   ##             "sex", "region",
-%'     ##           "socpo20", "socpo30", "socpo40", "socpo50",
-%'       ##         "civst20", "civst30", "civst40", "civst50",
-%'         ##       "ortnmn")]
-%'
-%'
-%' ## General:
-%' source("R/addPer.R")
-%' cut <- c(1851, 1876, 1901, 1921, 1940, 1951)
-%' xc <- numeric(length(cut) - 1)
-%' for (i in 1:(length(cut) - 1)){
-%'   xc[i] <- (cut[i] + cut[i + 1]) / 2
-%' }
-%' ucut <- cut[-(1:2)]
-%' uxc <- xc[-(1:2)]
-%' mud <- addPer(adult, cuts = cut)
-%' source("R/getInter.R")
-%'
-%' library(xtable)
-%' ##xtable(summary(adult[, c("sex", "region",
-%'   ##                       "civst40", "socpo40")]))
-%' ##save(adult, file = "adult.rda")
-%' ##save(mad, file = "mad.rda")
-%' @
-%'
-
-\section{Models}
-
-The proportional hazards model in a survival analysis context is used,
-allowing for adjustment for civil status and rural/urban environment. Since
-the central explanatory variable, {\sc Hisclass}, do not follow the
-proportionality property, the analyses are stratified with respect to the
-variable, and the main results are presented graphically.
-
-The analyses are performed in the \R\ environment for statistical computing
-and graphics \citep{cran}, especially using the package {\tt eha} \citep{gb12,eha}
-The proportional hazards model allow us to scrutinise the differences and
-the development in more detail. We divide the analysis into two age groups,
-40--59 and 60--99 years of age. The model controls for period, marital
-status and whether the individual resides in the urban or rural part. There
-are no serious signs of deviation from proportionality assumption, except
-regarding {\sc Hisclass}.
-
-We analyse both regions together but perform separate analyses for men and
-women and for the age groups 40--59 and 60--89. The model in the analyses
-include social class, marital status and urban/rural residence. Each decade
-during the 20th century gets separate analyses, but we then combine results
-in order to show the development of the social patterning of mortality
-during the studied period. We present the results for the period 2001--2013
-separately. We suggest a heterogeneity index in order to investigate a
-possible homogenisation of mortality between social classes.
-
-The important explanatory variable, {\sc Hisclass}, is included in the
-proportional hazards models as a \emph{stratification} variable. The main
-reason for this choice is that the effect of {\sc Hisclass} on survival is
-non-proportional, that is, it varies with age. Thus, the main results are
-presented \emph{graphically}. Note further that {\tt Hisclass} is a
-\emph{time-varying} variable. This causes lack of information in the higher
-ages around 1960--1980, because that kind of information is simply missing
-to a large extent.
-
-%The effect of *hisclass* on survival prospects is in focus, and the obvious
-%tool to use is a proportional hazard (PH) model. The {\sc hisclass} level could
-%either be analysed as a time dependent variable, or as a fixed variable at
-%a certain point in time. In our case, {\sc hisclass} is defined at the age when
-%we start to analyse mortality, age 40 or 50. At these ages, most have
-%reached their top position and they are still work active. We thus analyse
-%future survival according to the social position in mid-life. It has
-%advantages when it comes to drawing causal conclusions of observed
-%observations. Changes in social position may be due to circumstances that
-%are related to mortality risks, for example changing occupation due to bad
-%health. Another advantage of letting the {\sc hisclass} level be fixed is that
-%time-dependent are more sensitive to the assumptions in a proportional
-%hazard model. In our preliminary analyses of letting {\sc Hisclass} be
-%time-dependent, we encounter serious deviations from the PH
-%assumption. Marital status is however time-varying.
-
-%As a first crude measure of survival, we use median remaining life at age 40. 
-%The reason for showing the median and not the average remaining life is that the latter is impossible to estimate
-%nonparametrically without strong assumptions. This depends of course on the
-%kind of data we have: Heavily censored and truncated. Our approach can be
-%seen as a hybrid between period and cohort analysis.The model is calculated
-%for 20-year periods with the variables *hisclass*, marital status and urban
-%versus rural residence in the model, and separately for the two regions and
-%gender. The reason for presenting the regions separately is that there is
-%no data covering the 19th century for the Umeå region. The results
-%from the different analyses are presented in a serie of figures. We have
-%also made statistical analyses of possible changes in the effect of
-%*hisclass*, in order to identify possible appearance of a social
-%gradient. These analyses are made separately for the age group 40-59 and
-%age group 60 and above.
-
-\subsection{POPLINK data, 1901--1950}
-
-The POPLINK data cover 1901--1950.
-
-\subsection{FOB data, 1961--2000}
-
-The \emph{Fob} data is treated similarly as the POPLINK data. There is
-unfortunately no links between the individuals in the two data frames, and
-the result of that is a gap in the relevant information between the years
-1950 and 1960. Furthermore, the first two decades in the \emph{Fob} data,
-1960--1980, will have a lot of missing information on {\sc Hisclass}.
-
-\subsection{LISA data, 2002--2013}
-
-The {\sc Lisa} data is yearly, with {\tt hisclass} information starting in 2002.
-This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, 5, and 6.
+{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}
 
@@
 <<readtab>>=
-load("../CreateData/data4/vbTab.rda")
+load("data/vbTab.rda")
 @
@@
 (Skellefteå town or outside). The cumulative
-hazards at the end of the time period (60 if 40--59, 90 if 60--89) are the
+hazards at the end of the time period (60 if 40--64, 90 if 65--89) are the
 main target in our analysis: These numbers are used in the process of
@@
 
-\subsection{Ages 40--59}
+\subsection{Ages 40--64}
 
@@
 <<getske4060>>=
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, women."
-scap1w <-  "Cumulative hazards, ages 40--59, women."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, men."
-scap1m <-  "Cumulative hazards, ages 40--59, men."
+cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
+scap1w <-  "Cumulative hazards, ages 40--64, women."
+cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
+scap1m <-  "Cumulative hazards, ages 40--64, men."
 ##
 cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard by HISCLASS, ages 40--59."    
+scap12 <- "Total hazard by HISCLASS, ages 40--64."    
 cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap12log <- "Total hazard by HISCLASS, ages 40--59, last years."
+scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."
 
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, women."
-scap2w <-  "Cumulative hazards, ages 60--89, women."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, men."
-scap2m <-  "Cumulative hazards, ages 60--89, men."
+cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women."
+scap2w <-  "Cumulative hazards, ages 65--89, women."
+cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men."
+scap2m <-  "Cumulative hazards, ages 65--89, men."
 ##
 cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap22 <- "Total hazard by HISCLASS, ages 60--89."    
+scap22 <- "Total hazard by HISCLASS, ages 65--89."    
 cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap22log <- "Total hazard by HISCLASS, ages 60--89, last years."
+scap22log <- "Total hazard by HISCLASS, ages 65--89, last years."
 
 ##
-vb4059 <- vbTab[vbTab$age %in% seq(40, 55, by = 5), ]
-vb4059$age <- factor(vb4059$age)
+vb4064 <- vbTab[as.numeric(as.character(vbTab$age)) < 64.5, ]
+vb4064$age <- factor(vb4064$age)
 @
@@
 <<hazcalw>>=
-resw <- getPHglm(vb4059[vb4059$sex == "female", ],
+source("R/getPHglm2.R")
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
             nper, labb)
-resm <- getPHglm(vb4059[vb4059$sex == "male", ],
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
             nper, labb)
@@
 source("R/plotHaz.glm.R")
+
 plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(60|40)",
-                main = "Women")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(60|40)",
-                main = "Men")
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
+                main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
+                main = "Men", exclude = 5:6)
 par(oldpar)
@@
 
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--59 per individual by HISCLASS and decade, men.">>=
+%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
 %%@
@@
 oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(60|40)",
+wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
 main = "Women")
-mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(60|40)",
+mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
 main = "Men")
@@
 
-The analysis for the last period is based on a separate analysis and is presented separately.
+%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
 
 <<lissyoung>>=
-load("../CreateData/data4/liss.rda")
-liss4059 <- age.window(liss, c(40, 59))
+library(eha)
+library(xtable)
+load("data/loTab.rda")
+loTab$period <- factor(loTab$period)
+liss4064 <- loTab[loTab$age < 64.5, ]
 @
 
-<<liswom, cache = TRUE>>=
-fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "female", ])
-fitm1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss4059[liss4059$sex == "male", ])
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<liswom, results = 'asis'>>=
+fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+##cat("Women:\n")
+drw <- drop1(fitw1, test = "Chisq")
+xtable(drw, caption = "Effect of covariates for women 40-65, 1990-2005.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
+source("R/plotWEduc.R")
+plotWEduc(fitw1)
+@
+
+And for men:
+
+<<lismen, results='asis'>>=
+fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+##cat("\nMen:\n")
+drm <- drop1(fitm1, test = "Chisq")
+xtable(drm, caption = "Effect of covariates for men 40-65, 1990-2005.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
 dcol <- c("blue", "darkgreen", "darkgreen", "red")
@@
 
-See Figure~\ref{fig:plotliss1} for the cumulative hazards.
-
-<<plotliss1, fig.cap = "Cumulative hazards, ages 40--59, 2002--2013.", fig.height = 4>>=
-oldpar <- par(mfrow = c(1, 2))
-plot(fitw1, fn = "cum", col = dcol, main = "Women")
-plot(fitm1, fn = "cum", col = dcol, main = "Men")
-par(oldpar)
+<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm1)
 @
 
-See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<tablisswom1, results = "asis", cache = TRUE>>=
-fitwo1 <- coxreg(Surv(enter, exit, event) ~ factor(hisclass) + civst + urban + period, 
-                 data = liss4059[liss4059$sex == "female", ])
-dr <- drop1(fitwo1, test = "Chisq")
-ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--59, 2002--2013.", 
-    label = "tab:tablisswom1")
+<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
+#oldpar <- par(mfrow = c(1, 2))
+#plot(fitw1, fn = "cum", col = dcol, main = "Women")
+#plot(fitm1, fn = "cum", col = dcol, main = "Men")
+#par(oldpar)
+##summary(fitw1)
+##summary(fitm1)
 @
 
-See Table~\ref{tab:tablissman1} the PH regression for men.
-<<tablissman1, results = "asis", cache = TRUE>>=
-fitma1 <- coxreg(Surv(enter, exit, event) ~ factor(hisclass) + civst + urban + period, 
-                 data = liss4059[liss4059$sex == "male", ])
-dr <- drop1(fitma1, test = "Chisq")
-ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--59, 2002--2013.", 
-    label = "tab:tablissman1")
+%See Table~\ref{tab:tablisswom1} the PH regression for women.
+<<tablisswom1, results = "asis">>=
+#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "female", ])
+#dr <- drop1(fitwo1, test = "Chisq")
+#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.", 
+#    label = "tab:tablisswom1")
+@
+
+%See Table~\ref{tab:tablissman1} the PH regression for men.
+<<tablissman1, results = "asis">>=
+#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "male", ])
+#dr <- drop1(fitma1, test = "Chisq")
+#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.", 
+ #   label = "tab:tablissman1")
 @
@@
 
-\subsection{Ages 60--89}
+\subsection{Ages 65--89}
 
@@
 <<getske6089>>=
-vb6089 <- vbTab[vbTab$age %in% seq(60, 85, by = 5), ]
-vb6089$age <- factor(vb6089$age)
+vb6089 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6089$age <- factor(vb6089$age) # NOTE:: No change to '6589'!
 @
@@
 We now turn to the elderly, i.e. those that mostly have left the workforce. We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. In accordance with the analyses for the age group 40--59, the model adjusts for {\tt civil status} and
+periods. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
 {\tt urban/rural} residence and with separate analyses for women and men.
@@
 
-The cumulative hazards for the age group 60--89 in different decades are
+The cumulative hazards for the age group 65--89 in different decades are
 presented in Figures~\ref{fig:hazplotwold} and
 \ref{fig:hazplotmold}. Corresponding to what we find for the age group
-40--59, it is obvious that the assumption of proportional hazards is not
+40--64, it is obvious that the assumption of proportional hazards is not
 reasonable.
@@
 We will use as a summary measure the value of the cumulative hazards at 
-duration 30 (age 90), starting from age 60. For the period 2002--2013, a separate analysis is presented.
+duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
 
@@
 
-Concerning the social pattern, we find results fairly similar to the ones for the age group 40--59.
+Concerning the social pattern, we find results fairly similar to the ones for the age group 40--64.
 For women the results are mixed, even though women from the elite had comparatively low mortality and working class women high.
-For men we find the same intriguing change among elderly as for the adult population 40--59 years.
+For men we find the same intriguing change among elderly as for the adult population 40--64 years.
 Mortality was in fact substantially higher among men from the elite and the group of lower managers while it was lowest among workers.
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|60)",
-                main = "Women")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|60)",
-                main = "Men")
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
+                main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
+                main = "Men", exclude = 5:6)
 par(oldpar)
@@
 oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|60)",
+wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
 main = "Women")
-mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|60)",
+mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
 main = "Men")
@@
 
-%% The compression that is visible for later decades is partly explained by the decereasing general mortality. This effect can be eliminated by showing these numbers on a log scale.
+\subsubsection*{The effect of income and education, 1990--2005}
 
-%% <<cumslogwold, fig.cap = "Probability of dying before age 90 for a 60 year old person by HISCLASS and decade, log scale, women (top) and men (bottom).", fig.height = 10>>=
-%% oldpar <- par(mfrow = c(2, 1))
-%% wloutp <- plotED.ph(resw, col = scol, logScale = TRUE, ylab = "H(60|40)",
-%% main = "Women")
-%% mloutp <- plotED.ph(resm, col = scol, logScale = TRUE, ylab = "H(60|40)",
-%% main = "Men")
-%% par(oldpar)
-%% @
+The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
 
-%%<<cumslogm, fig.cap = "Expected number of deaths in ages 40--59 per individual by HISCLASS and decade, log scale, men.">>=
-%%@
+<<lissyoung2>>=
+library(eha)
+library(xtable)
+##load("data/loTab.rda")
+##loTab$period <- factor(loTab$period)
+liss6589 <- loTab[loTab$age > 64.5, ]
+@
 
-%%<<measm, fig.cap = "Heterogeneity index over time, ages 40--59, men.">>=
-%%@
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<liswom2, results = 'asis'>>=
+fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+##cat("Women:\n")
+drw2 <- drop1(fitw2, test = "Chisq")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
 
-%' For the LISA data:
-%' 
-%' <<lissold>>=
-%' load("liss.rda")
-%' liss6089 <- age.window(liss, c(60, 90))
-%' @
-%' 
-%' 
-%' <<liswom2>>=
-%' fitw2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss6089[liss6089$sex == "female", ])
-%' fitm2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + civst + urban, dist = "gompertz", data = liss6089[liss6089$sex == "male", ])
-%' dcol <- c("blue", "darkgreen", "darkgreen", "red")
-%' @
-%' 
-%' <<plotliss2, fig.cap = "Cumulative hazards, ages 60--89, 2002--2013.", fig.height = 4>>=
-%' oldpar <- par(mfrow = c(1, 2))
-%' plot(fitw2, fn = "cum", col = dcol, main = "Women")
-%' plot(fitm2, fn = "cum", col = dcol, main = "Men")
-%' par(oldpar)
-%' @
-%' 
-%' We also present the output from Cox regressions for women and men
-%' separately, with {\sc Hisclass} as an 
-%' ordinary covariate (no stratification). The non-proportionality issue
-%' regarding this covariate is not so severe in the Lisa data.
-%' 
-%' See Table~\ref{tab:tablisswom2} for the PH regression for women.
-%' <<tablisswom2, results = "asis", eval = TRUE>>=
-%' fitwo2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
-%'                  data = liss6089[liss6089$sex == "female", ])
-%' dr <- drop1(fitwo2, test = "Chisq")
-%' ltx(fitwo2, dr = dr, caption = "PH regression, women aged 60--89, 2002--2013.", label = "tab:tablisswom2")
-%' @
-%' 
-%' See Table~\ref{tab:tablissman2} for the PH regression for men.
-%' <<tablissman2, results = "asis", eval = TRUE>>=
-%' fitma2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
-%'                  data = liss6089[liss6089$sex == "male", ])
-%' dr <- drop1(fitma2, test = "Chisq")
-%' ltx(fitma2, dr = dr, caption = "PH regression, men aged 60--89, 2002--2013.", label = "tab:tablissman2")
-%' @
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
 
-%' 
-%' \subsection{Cohort modeling}
-%' 
-%' The models so far have been period-oriented, that is, ten-year periods have
-%' been analysed separately. The contributions in different ages then come
-%' from different people in separate birth cohorts. An alternative is to
-%' analyse different birth cohorts separately, and we do this here. The
-%' drawback with the cohort method is of course that the cohorts get shorter
-%' and shorter as we move towards younger cohorts, but on the other hand we
-%' get a better picture of what happened in real life.
-%' 
-%' First we need to prepare the cohort data by combining {\tt fob} and {\tt
-%'   lisa} data. The central covariate, {\sc Hisclass} can now be kept
-%' \emph{time-fixed} at the start of follow-up, the baseline ages 35--40.
-%' 
-%' <<cohort data, eval = FALSE>>=
-%' ##liss$urban <-  liss$FORSAMLING %in% c(248001, 248002, 284004, 284005, 248011, 248201, 248204)
-%' ##fo <- fobb[, c("LINNEID", "birthdate", "sex", "enter", "exit", "event", ##"civst", "urban", "hisclass")]
-%' ##li <- liss[, c("LINNEID", "birthdate", "sex", "enter", "exit", "event", ##"civst", "urban", "hisclass")]
-%' 
-%' fobb$hisclass <- cut(fobb$hisclass, c(0, 2.5, 4.5, 5.5, 8), labels = c("1+2", "3+4", "5", "6+7"))
-%' foli <- rbind(fobb, liss)
-%' foli <- foli[order(foli$id, foli$enter), ]
-%' foli <- rc(foli)
-%' foli$Enter <- numeric(NROW(foli))
-%' foli$Exit <- numeric(NROW(foli))
-%' foli$Event <- numeric(NROW(foli))
-%' foli$Enter[foli$lopnr == 1] <- foli$enter[foli$lopnr == 1]
-%' foli$Exit[foli$lopnr == foli$antrec] <- foli$exit[foli$lopnr == foli$antrec]
-%' foli$Event[foli$lopnr == foli$antrec] <- foli$event[foli$lopnr == foli$antrec]
-%' indx <- tapply(foli$id, foli$id)
-%' Ex <- tapply(foli$Exit, foli$id, max)
-%' Ev <- tapply(foli$Event, foli$id, max)
-%' foli$Exit <- Ex[indx]
-%' foli$Event <- Ev[indx]
-%' foli <- foli[foli$lopnr == 1, ]
-%' foli$enter <- foli$Enter
-%' foli$exit <- foli$Exit
-%' foli$event <- foli$Event
-%' foli <- foli[!is.na(foli$hisclass), ]
-%' foli <- foli[foli$birthdate > 1921 & foli$birthdate < 1961, ]
-%' foli$period <- cut(foli$birthdate, c(1921, 1931, 1941, 1951, 1961), labels = c("1921-30", "1931-40", "1941-50", "1951-60"))
-%' cols <- c("blue", "green", "green", "red")
-%' @
-%' 
-%' Then, we create four cohorts by birth year: "1921-30", "1931-40", "1941-50", and "1951-60".
-%' For men, see Figure~\ref{fig:plotcohm}
-%' <<plotcohm, fig.cap = "Birth cohorts, men, life after 40. Blue is 1+2, red is 6+7.",eval=FALSE>>=
-%' flm <- foli[foli$sex == "male", ]
-%' oldpar <- par(mfrow = c(2, 2))
-%' fit1 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1921-30", ])
-%' plot(survfit(fit1), fun = "cumhaz", main = "1921-30", col = cols)
-%' fit2 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1931-40", ])
-%' plot(survfit(fit2), fun = "cumhaz", main = "1931-40", col = cols)
-%' fit3 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1941-50", ])
-%' plot(survfit(fit3), fun = "cumhaz", main = "1941-50", col = cols)
-%' fit4 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flm[flm$period == "1951-60", ])
-%' plot(survfit(fit4), fun = "cumhaz", main = "1951-60", col = cols)
-%' par(oldpar)
-%' @
-%' 
-%' For women, see Figure~\ref{fig:plotcohw}
-%' <<plotcohw, fig.cap = "Birth cohorts, women, life after 40. Blue is 1+2, red is 6+7.",eval=FALSE>>=
-%' flw <- foli[foli$sex == "female", ]
-%' oldpar <- par(mfrow = c(2, 2))
-%' fit1 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1921-30", ])
-%' plot(survfit(fit1), fun = "cumhaz", main = "1921-30", col = cols)
-%' fit2 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1931-40", ])
-%' plot(survfit(fit2), fun = "cumhaz", main = "1931-40", col = cols)
-%' fit3 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1941-50", ])
-%' plot(survfit(fit3), fun = "cumhaz", main = "1941-50", col = cols)
-%' fit4 <- coxph(Surv(enter-40, exit-40, event) ~ strata(hisclass) + civst + urban, data = flw[flw$period == "1951-60", ])
-%' plot(survfit(fit4), fun = "cumhaz", main = "1951-60", col = cols)
-%' par(oldpar)
-%' @
-%' 
-%' Another way of analysing mortality is by taking a cohort approach. The
-%' cohorts from the decades 1920s until 1950s are followed. The first cohort
-%' can be followed until very high age. For the age up to 60, representing the
-%' period 1960 to 1989 approximately, the figure do not show any large
-%' differences (which can have to do with scale). When they had passed 60, we
-%' find clear social differences in the expected direction for men, but less
-%' clear for women. The other cohorts follow rather similar patterns as the
-%' oldest one.
+<<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitw2)
+@
 
-\section{Discussion}
+And for men:
 
-The authors of this paper have scrutinised different aspects of the social
-pattern of mortality in northern Sweden for quite some time, focusing on
-different regions, different age groups and partly different social
-categories. Already the earliest findings, concerning the social inequality
-in a rapidly growing town in an industrial district, surprised and at least
-at first confused the author \citep{edv92}. The expected social gradient in
-survival, meaning that people from higher social classes always had an
-advantage, was in many cases, particularly among adult men, not found
-despite the well-documented socio-economic differences in this
-society. There were however large differences between men and women and for
-different age groups. Further studies have confirmed these
-findings \citep{edli11, edbr12}. These earlier studies have ended with an
-assumption that more distinct differences to the advantage of higher
-classes developed somewhere during and probably quite early in the 20th
-century. Until now we have however not been able to investigate the
-continuous development on mortality differences among the adult and elderly
-population due to no available data. This is the first time it has been
-possible, and the results continue to surprise us. The consistent advantage
-for higher social classes do not appear until quite late in the regions we
-have analysed. The social pattern of female mortality are closer to what is
-expected, but a social gradient in male adult mortality seems to be a
-recent phenomenon.
+<<lismen2, results='asis'>>=
+fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+##cat("\nMen:\n")
+drm2 <- drop1(fitm2, test = "Chisq")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
 
-The perhaps surprising results makes it reasonable to consider their
-validity and possible weaknesses in the study. There are of course
-uncertainties involved in all studies of this sort, something that should
-not be ignored.  One obstacle concerns the quality of the available
-information in the sources. Occupational titles in historical sources are
-not always distinct enough to define the correct class. It can be difficult
-to identify the skill level in the working class population because many
-were often only designated as labourer. The size of the enterprise for free
-entrepreneurs is unknown and it may be impossible to distinguish if they
-are self-employed or have many employed. The class system and the position
-of different classes can also be discussed. Studies covering long period
-have the problem that occupations and classes change character - they do
-not represent the same thing. Nonetheless, these ambiguities and possible
-mis-classifications are not of that sort and magnitude that it would radically change
-the basic patterns observed. The social classes identified during the
-different periods represent distinct groups with very large differences in
-resources, capacities and power. A worker would not end up in the highest
-SOCPO level and large-scale entrepreneurs and higher officials would
-certainly be assigned the right level.Available data also show that the
-differences in income and wealth were large, often enormous.
+<<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm2)
+@
 
-In this paper we have analysed social inequality in health from the
-perspective of social classes using HISCLASS. However, there are
-alternative ways to analyse social inequality, and we cannot exclude that
-such analyses gives other results or at least would modify the image
-presented here. There are other alternatives then HISCLASS, representing
-status or class dimensions. SOCPO is another system based on hierarchic
-categories of social power, which we have used in a couple of previous
-studies. In many modern studies categorisations based on income, wealth
-or education are common for studies of inequality. It is possible that
-such categorisations are better fit to distinguish between groups in
-society. They represent different dimensions of inequality, and we believe
-that the topic discussed here would gain from analyses of different aspects
-of inequality. One problem is however that reliable data on income and
-education are rarely available in historical population databases. We can
-however be certain that the highest HISCLASSES level have more education
-and the lowest the least. We also know from the 1930 census that the
-economic differences were very large at that time and that occupations in
-higher HISCLASSES tend to have higher incomes. The groups that may diverge
-are those within agriculture and those having a small enterprise where we
-can assume that the reported income figures underestimate their economic
-standard.
-
-The results are however in accordance with other studies \citep{bevp11},
-for example by \citet{bedr11} in a very different setting, a couple of
-Scanian parishes. They found the first indications of an emerging social
-gradient in the 1950s.The similarities makes us more comfortable about our
-results.
-
-\citet{who10} lists three different types of intermediary social
-determinants of health, material circumstances, behaviours and biological
-factors, and finally psycho-social factors. Most studies have focused on
-the material circumstances. Having a better economy can give advantages in
-many aspects of relevance for health and mortality, for example good living
-conditions, access to food, medical care, better conditions at work et
-cetera. Even though these advantages ought to have been at hand in our
-studied regions, the final result was not better health for men in the most
-privileged groups until the later part of the 20th century. We must
-consider other intermediary determinants. Psycho-social factors have been
-frequently discussed during the most recent decades, for example by
-\citet{mar04}. He argues that being higher in the status ladder leads to
-better health, regardless of where in the status ladder we look. This is
-also one of the arguments \citet{wipi09} use as an explanation why equal
-societies always perform better than unequal ones when it comes to health
-and other social conditions. In the case studied here, we do not find any
-indication of a status syndrome, since mortality levels was in the opposite
-directions to the one we would expect from these theories.
-
-This makes us consider the possible impact of behavioural factors.
-Our results of the gender aspect might be of interest in this case.
-When it comes to gender differences, the present study confirms what we have found on the Sundsvall region.
-There is a clear gendered pattern when it comes to social inequality in mortality.
-While high HISCLASS did not have any positive impact on survival for men in the early 20th century, we find  that higher social
-positions among women led to longer lives. Within the same households, the effect of social
-class seems to lead to different results. This opens up for behavioural
-aspects based on different gendered expectations. Alcohol has been
-suggested as part of the large gender differences in the 19th century
-\citep{wil99, edv92}, and a more risk-taking life may be a component of the
-male mortality.
-
-Having a high social status could be expressed in attitudes and behaviours
-that were harmful for health. Alcohol consumption decreased rapidly during
-the period late 19th century to the 1950s, thus possibly making room for an 
-increased role of economic resources. But the old pattern in social health
-inequalities remained. It is however possible that the relative differences
-in alcohol consumption prevailed, although at a lower level. The question 
-is however if the lower levels of alcohol consumption had any substantial
-negative impact on health. Swedes at that time had very low consumption
-levels. Alternative explanations should also be scrutinise. There 
-may have been strong and heavy requirements of being successful for the
-male \"elite. Many were involved in businesses with high risks. It would be
-very damaging to lose the high position and they may have felt that they 
-lived under constant threat of economic failure. There are several examples among
-the economic \"elite in the Sundsvall region that they had such problems
-\citep{edli11}. A possible future study involving analyses of death causes
-might throw some light on this issue.
-
-When it comes to the impact of social factors in the working age population versus the elderly, we do not find any fundamental differences.
-The social class have somewhat smaller effect among the elderly, so there is some difference in levels but not in character.
-
-Our study illustrates the importance of considering contextual factors and
-the restrictions of making general statements. We need to look into all
-aspects of social classes, not only the material ones but also what this
-means for behaviour, norms, way of living. \citet{edli11} suggested that an
-explanation for the larger social health inequalities in the present-day
-world is that status and higher social position is now expressed through
-health and being fit and physical active, while in many historical contexts
-higher social status was expressed by low physical activity, obsessive
-eating (being fat was impressive), drinking, smoking and sometimes also
-taking risks.
-
-Another question is if our results can be generalised. We have studied
-environments in northern Sweden without any large metropolis. Perhaps
-Sweden and/or this part of Sweden is a special case. The social differences
-may have been smaller, the involvement of government both at the national
-and local level may have been stronger, thus mitigating socio-economic
-disadvantages. These are relevant questions, which hopefully can be
-addressed in other and maybe comparative studies.
-
-Finally, we argue that our results have implications for our understanding
-on several hot research issues. A central question concerning the mortality
-transition is how the enormous increase in survival could be
-accomplished. This is still a hot debated question. \citet{mck76} argues that the
-improvements up to the middle of the 20th century was first of all made
-possible by improved nutrition. This seems to implicate that having more
-resources would lead to better survival. If research shows that the most
-advantaged groups did not have better health and longer lives, this do not
-necessarily speak against McKeown's ideas but it makes the question more
-complex. The results are also difficult to fit into \citet{liph95}
-hypothesis that social conditions should be seen as a fundamental cause of
-disease. Even if we would eliminate disadvantages such as bad living
-conditions, lower standard of medical care and other proximate determinants
-for the poor, the health disadvantage would remain. In the historical
-context we have analysed here, this does not get any support. This further
-illustrate the need for situating the analyses in their historical and
-geographical contexts.
-
-In recent years, the question about the relation between health and income
-inequality been widely discussed. Even though our study do not address this
-issue directly, our results are of interest for this question as well. The
-development of the social inequality in mortality do not seem to fit that
-well into the development of income and wealth inequalities. The early 20th
-century, in Sweden as well as in other countries, was characterised by very
-large economic differences, still this did in our regions not lead to any
-consistent advantage for the highest social classes. After the Second World
-War, economic differences diminished, and was at its lowest in Sweden at
-about 1980, i.e. at a time when a more expected pattern of a social
-hierarchy in survival had appeared. The period thereafter is characterised
-by increasing economic differences, and this may coincide with increasing
-health differences as well. The long-term relation between economic
-inequality and health is worth further studies.
-
-\section{Conclusion}
-
-We started this paper by raising the issue of the development of social
-inequalities. When studying this question in history we often get surprised
-of its complexity. In a society characterised by large economic inequality, 
-we would expect that this would result in corresponding mortality
-differences, i.e. higher social classes having an advantage. We find the
-appearance of a consistent social pattern in mortality, to the advantage of
-the higher classes, but surprisingly late in our history. It is only in the
-recent decades that this pattern dominates. Improved survival characterised
-however all HISCLASSes. We can however confirm that there are gender
-differences in the social pattern. While there is no clear social gradient
-among men, the results at least partly indicate advantages for the higher
-social classes for women already early in the 20th century. Females had
-more to gain from a high social position. Previous studies of the mortality
-transition have made similar observations. Among adults, gender has often
-been of more significant importance than social class. The reason for this
-difference and for the late appearance of the modern social pattern among
-men is still an enigma that require further studies, even though we have
-suggested the possible importance of behavioural factors and changes in
-ideals connected to different social classes. This should stimulate us to
-further disentangle the pathways for how social position influence health
-and mortality in different societies. Finally, we do not find any clear
-differences in the social pattern in the two studied age groups.
-
-%\newpage
-
-%\section{Reproducibility declaration}
-
-%The exact versions of tools used in the analyses are shown here.
-
-%<<reproducibility, echo = TRUE>>=
-%sessionInfo()
-%@
-
-%\newpage
-
-\section*{Acknowledgements}
-
-This work was undertaken within the Centre for Demographic and Ageing
-Research (CEDAR) at Umeå University and with support from the Swedish
-Research Council (VR) (grant no. 421-2012-892). We thank Demographic Data
-Base for giving us access to data and Jonas Helgertz, Department of
-Economic History, Lund University, for providing us with scripts for
-creating HISCLASS from Swedish Census data.
+{\huge Här ska 'conclusion.tex' in !!!}
 
@@
 \end{document}
+@

### 2016-09-25 R/plotED.glm.R
update

@@ -2,7 +2,8 @@ plotED.glm <- function(res,
                        col = 1:length(res),
                        logScale = FALSE,
                        ylab = "",
-                       main = ""){
+                       main = "",
+                       exclude = 100){
     
     ## 'res' is a list of matrices of cumulative hazards from a piecewise-contant fit
     ## We want the last column of each matrix.
@@ -13,7 +14,9 @@ plotED.glm <- function(res,
     k <- NCOL(res[[1]])
     m <- NROW(res[[1]])
     labb <- names(res)
-    
+
+    out <- 1:nper
+    out <- out[!(out %in% exclude)]
     outp <- matrix(0, ncol = nper, nrow = m)
 
     for (i in 1:nper){
@@ -28,18 +31,20 @@ plotED.glm <- function(res,
     xv <- (as.numeric(substr(names(res), 1, 4)) +
         as.numeric(substr(names(res), 6, 9))) / 2
 
-    plot(xv, outp[1, ], type = "b", lty = 2, col = col[1], pch = 1,
+    plot(xv[out], outp[1, -exclude], type = "b", lty = 2, col = col[1], pch = 1,
          ylim = c(fr, to), main = main,
          axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
          xlim = c(min(xv), max(xv) + 10))
     axis(1, at = xv, lab = labb, las = 2)
     axis(2, las = 2)
     box()
+    ##lines(xv[7:nper], outp[1, 7:nper], col = col[1], pch = 1)
     for (i in 2:m){
-        points(xv, outp[i, ], col = col[i], pch = i)
-        lines(xv, outp[i, ], col = col[i], lty = i)
+        points(xv[out], outp[i, out], col = col[i], pch = i)
+        lines(xv[out], outp[i, out], col = col[i], lty = i)
+        ##lines(xv[7:nper], outp[i, 7:nper], col = col[i], lty = i)
     }
-    abline(v = xv, lty = 3, col = "magenta")
+    abline(v = xv[out], lty = 3, col = "magenta")
     abline(h = 0)
     legend("topright", legend = 1:m, col = col, pch = 1:m, lty = 1:m)
     invisible(outp)

### 2016-09-25 R/getPHglm2.R
first commit

@@ -1,0 +1,47 @@
+getPHglm2 <- function(dat, nper, labb, from = 40){
+    ## dat$age must be a factor ...
+    ## Note: return value is Cumulative hazards at given age points (straight line between).
+    ##
+    ## 'dat$age' is assumed to be grouped in equidistant intervals!
+    res <- vector(nper, mode = "list")
+    hc <- levels(dat$hisclass)
+    m <- length(hc)
+    k <- length(unique(dat$age))
+    if (k <= 1) error("No. of age intervals must be at least 2!")
+    ageName <- as.numeric(levels(dat$age))
+    ageInt <- ageName[2] - ageName[1]
+    levs <- matrix(0, nrow = m, ncol = k + 1)
+   
+    for (i in 1:nper){
+        fit <- glm(event ~ offset(log(exposure)) + civst + urban + age * hisclass, 
+                   data = dat[dat$period == labb[i], ], family = poisson)
+        
+        co <- fit$coefficients
+        ages <- grep("age", names(co))
+        hisces <- grep("hisclass", names(co))
+        interact <- ages[ages %in% hisces]
+        amain <- co[ages[!(ages %in% interact)]]
+        hmain <- co[hisces[!(hisces %in% interact)]]
+        inter <- matrix(co[interact], ncol = k - 1, byrow = TRUE)
+        tmp <- matrix(0, nrow = m, ncol = k)
+        for (j in 1:m){
+            tmp[j, ] <- c(co[1], co[1] + amain)
+        }
+        for (j in 2:m){
+            tmp[j, ] <- tmp[j, ] + hmain[j - 1]
+        }
+        tmp[-1, -1] <- tmp[-1, -1] + inter
+        tmp <- exp(tmp)
+        levs <- cbind(rep(0, m), tmp)
+        for (j in 1:m){
+            levs[j, ] <- cumsum(levs[j, ]) * ageInt
+        }
+        
+
+        colnames(levs) <- c(ageName, ageName[k] + ageInt)
+        rownames(levs) <- hc
+        res[[i]] <- levs
+    }
+    names(res) <- labb
+    res
+}

### 2016-09-25 R/plotWEduc.R
first commit

@@ -1,0 +1,20 @@
+plotWEduc <- function(fit){
+    ## fit from a glm
+    
+    co <- fit$coefficients
+    con <- co[1]
+    educ <- co[substr(names(co), 1, 4) == "educ"]
+    income <- co[substr(names(co), 1, 4) == "inco"]
+
+    educ <- exp(c(con, educ + con))
+    names(educ) <- 1:7
+    income <- exp(c(con, income + con))
+    names(income) = c("lowest", "low", "high", "highest")
+    oldpar <- par(mfrow = c(1, 2))
+    barplot(100 * educ, xlab = "Education", col = "green", las = 1, 
+            ylab = "Mortality per cent")
+    barplot(100 * income, xlab = "Income", col = "blue", las = 2,
+            ylab = "Mortality per cent")
+    par(oldpar)
+    invisible(list(educ = educ, income = income))
+}

### 2016-10-03 oldmort.Rnw
cosmetic changes

@@
     Chicago, Illinois
-    November 17--20, 2016. (Version 2016-09-20).}}
+    November 17--20, 2016. (Version 2016-10-03).}}
 
@@
 drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-65, 1990-2005.", label = "tab:liswom")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-65, 1990-2005.")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)

### 2016-10-06 oldmort.Rnw
add reference

@@
     Chicago, Illinois
-    November 17--20, 2016. (Version 2016-10-03).}}
+    November 17--20, 2016. (Version 2016-10-06).}}
 
@@
 \begin{abstract}
+A long-term perspective on the development of social
+inequalities in mortality in the adult and elderly population is taken. 
+The area under investigation consists of the Skellefteå and Umeå regions in
+the north of Sweden, and the time period is 1851--2013.
+
+The main findings are that the social gradient becomes evident in the later
+alf of the 20th century, with lower classes having the highest mortality.
 
@@
 
-{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}
+%%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}
 
-\subsection{Combining the data sets and tabulate}
+%%\subsection{Combining the data sets and tabulate}
 
@@
 @ 
+\section{Introduction}
 
-\section{Results}
+A long-term perspective on the development of social
+inequalities in mortality in the adult and elderly population is taken. 
+The area under investigation consists of the Skellefteå and Umeå regions in
+the north of Sweden, and the time period is 1851--2013 for Skellefteå and
+1901--2013 for Umeå.
+
+This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
+the ESHD Conference in Leuven, Belgium, 2016. Here we make quick look back
+in time for the Skellefteå region (1851--1900), and we also consider causes
+of death for the later part of the 20th century. The {\sc Lisa} data allow
+us to study the effect of income and education on mortality in a more
+direct way.
+
+\section{Aims and questions}
+
+The aim of the paper is to investigate the development of social 
+differences using HISCLASS in mortality for the adult and elderly
+population during the mortality 
+transition and its relation to the development of income inequality. The
+analyses 
+will be performed on data from the Skellefteå and Umeå regions
+1901--2013. The questions are:
+
+\begin{enumerate}
+\item Has inequality in mortality between social classes increased among the adult and elderly
+  population, primarily benefiting the highest social positions?
+
+%%\item Is the mortality transition initiated by a certain social class?
+
+\item Are there any gender differences in the effect of social position?
+
+\item Is social position equally important among the retired population as for those in working age?
+\end{enumerate}
+
+\section{Previous research}
+
+For a detailed presentation we refer to \citet{segb16}. Her is a short
+summary.
+
+Several studies have shown how important access to economic and other
+resources are for health, but it is not clear how this has developed in
+history. \citet{ant67} and \citet{smi83} suggest that social inequality in
+health has passed through different phases in history. According to them,
+social health differences were
+comparatively small during the pre-transitional phase when space was a
+strong determinant for the spread of disease. Differences then increased
+during the transitional phase when mortality declined and wealthy groups
+used their resources to gain better health. Finally, health differences
+decreased again in modern low-mortality settings when instead
+health-related behaviour became important, resulting in marginal
+differences. \citet{omran82} state, in the third proposition in his theory
+of the epidemiological transition, that even if the class differential in
+mortality was maintained during the transition, the decline
+set in earlier and was faster among privileged groups.
+
+The present state of knowledge about the development of social 
+health inequalities (in Sweden but also in other European countries), no
+clear social hierarchy has been found before 1900 but substantial
+differences towards the late 20th and early 21st century. It appears as if
+a more consequent social health divide developed in the period
+in-between. This period has not been thoroughly studied on this issue,
+mainly because of lack of data to analyse. However, the extension of data
+at the Demographic Data Base, Umeå University (DDB) makes such studies
+possible. We are now able to analyse social 
+differences in mortality with micro-data on a sufficiently large population
+and with substantial social diversity from the 20th century for the
+Skellefteå and Umeå regions \citep{wee16}.
+
+\section{Data and variables}
+
+The data for the present study comes from two large population databases at
+the Demographic Data Base, Umeå University. The early period is covered by
+the database Poplink, the digitisation of historical parish registers for
+the Skellefteå and Umeå region \citep{wee16}. The population in the two
+regions is large enough and socially diverse to enable studies of social
+mortality inequalities. Poplink is based on linked parish records, allowing
+us to reconstruct life biographies on people as long as they remained in
+the region. The records are linked within but not between the regions. Data
+from the regions are accessed for the period 1901--1950s.
+
+The other large data-set comes from the Linnaeus database
+\citep{malm10}. It is based on different linked population registers from
+1960 to 2013 and is used within the ageing program at CEDAR, Umeå
+University. The study period from 1960 to 2000 are constructed from censuses every fifth
+year 1965--1990, with additional information on deaths from National Board of Health and Welfare.
+For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
+The construction of these data-sets for analysis is presented below.
+
+Information on individuals are anonymized and the two databases are not
+linked between each other. This stops us from following individuals present
+in both databases throughout their lives, thus they are treated separately. This
+also makes it impossible to add information on individuals in the Linnaeus
+database from what we potentially could find in Poplink, for example family
+background or occupations earlier in life.
+
+In the data-set analysed here, all individuals 40 years and older and ever being resident in either
+of the regions are included. The data file contains the variables on sex,
+birth date, death date, first and last date of observation and type of
+entrance/exit.
+
+The identification of presence periods as well as the time-dependent variables require some additional comments.
+
+\section{Models}
+
+The proportional hazards model in a survival analysis context is used,
+allowing for adjustment for civil status and rural/urban environment. Since
+the central explanatory variable, {\sc Hisclass}, do not follow the
+proportionality property, the analyses are stratified with respect to the
+variable, and the main results are presented graphically.
+
+The analyses are performed in the \R\ environment for statistical computing
+and graphics \citep{cran}, especially using the package {\tt eha} \citep{gb12,eha}
+The proportional hazards model allow us to scrutinise the differences and
+the development in more detail. We divide the analysis into two age groups,
+40--64 and 65--99 years of age. The model controls for period, marital
+status and whether the individual resides in the urban or rural part. There
+are no serious signs of deviation from proportionality assumption, except
+regarding {\sc Hisclass}.
+
+We analyse both regions together but perform separate analyses for men and
+women and for the age groups 40--64 and 65--89. The model in the analyses
+include social class, marital status and urban/rural residence. Each decade
+during the 20th century gets separate analyses, but we then combine results
+in order to show the development of the social patterning of mortality
+during the studied period. We present the results for the period 2001--2013
+separately. We suggest a heterogeneity index in order to investigate a
+possible homogenisation of mortality between social classes.
+
+The important explanatory variable, {\sc Hisclass}, is included in the
+proportional hazards models as a \emph{stratification} variable. The main
+reason for this choice is that the effect of {\sc Hisclass} on survival is
+non-proportional, that is, it varies with age. Thus, the main results are
+presented \emph{graphically}. Note further that {\tt Hisclass} is a
+\emph{time-varying} variable. This causes lack of information in the higher
+ages around 1960--1980, because that kind of information is simply missing
+to a large extent.
+
+\section{Results, all causes of death}
 
@@
 
-The general pattern do not differ from the one of the younger age group.
-The levels are slightly more homogeneous, but we cannot establish that the elderly diverge to any degree from the working-age population.
-Mortality among women was, as expected, lower than among men.
-We also find a clear decline in mortality, but this decline did not set in until sometimes after the Second World War.
-This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.
-The results for the 1960s are however uncertain as discussed above, and should not be taken to seriously.
-The levels seem to be to low.
-This is something we need to control further.
+The general pattern do not differ from the one of the younger age
+group. The levels are slightly more homogeneous, but we cannot establish
+that the elderly diverge to any degree from the working-age
+population. Mortality among women was, as expected, lower than among
+men. We also find a clear decline in mortality, but this decline did not
+set in until sometimes after the Second World War. This corresponds to the
+Swedish development where we find a turning point at this time with better
+survival among elderly, to a large extent caused by lower mortality in
+cardiovascular diseases. The results for the 1960s are however uncertain as
+discussed above, and should not be taken to seriously. The levels seem to
+be to low. This is something we need to control further.
 
-Concerning the social pattern, we find results fairly similar to the ones for the age group 40--64.
-For women the results are mixed, even though women from the elite had comparatively low mortality and working class women high.
-For men we find the same intriguing change among elderly as for the adult population 40--64 years.
-Mortality was in fact substantially higher among men from the elite and the group of lower managers while it was lowest among workers.
-This did not change until the last decades of the 20th century.
-At that time, the highest social classes instead had the best survival while the opposite was the case for workers.
-The pattern had reversed, thus confirming the results from most studies of this topic concerning the present-day society.
-The advantage for the higher classes and the disadvantage for the workers are apparent also for this age group during the early 21st century.
+Concerning the social pattern, we find results fairly similar to the ones
+for the age group 40--64. For women the results are mixed, even though
+women from the elite had comparatively low mortality and working class
+women high. For men we find the same intriguing change among elderly as for
+the adult population 40--64 years. Mortality was in fact substantially
+higher among men from the elite and the group of lower managers while it
+was lowest among workers. This did not change until the last decades of the
+20th century. At that time, the highest social classes instead had the best
+survival while the opposite was the case for workers. The pattern had
+reversed, thus confirming the results from most studies of this topic
+concerning the present-day society. The advantage for the higher classes
+and the disadvantage for the workers are apparent also for this age group
+during the early 21st century.
 
@@
 
-The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
+The analysis for the last period is based on a separate analysis with
+income and education as explanatory variables.
 
@@
 
-{\huge Här ska 'conclusion.tex' in !!!}
+%%{\huge Här ska 'conclusion.tex' in !!!}
 

### 2016-10-07 oldmort.Rnw
add causes 41

@@
 %\usepackage[tablesonly]{endfloat}
-\usepackage[figuresonly]{endfloat}
+%\usepackage[figuresonly]{endfloat}
+\usepackage{endfloat}
 
@@
 
-For a detailed presentation we refer to \citet{segb16}. Her is a short
+For a detailed presentation we refer to \citet{segb16}. Here is a short
 summary.
@@
 
+\section{Results, causes 41 \ldots}
+
+<<causes41>>=
+load("data/vb.rda")
+x <- substr(vb$ULORSAK, 1, 2)
+vb$event <- vb$event & (!is.na(vb$ULORSAK)) & (x %in% "41")
+vb$exposure <- with(vb, exit - enter)
+vbTab <- aggregate(vb[, c("event", "exposure")], 
+                   vb[, c("sex", "age", "period", "cohort", 
+                          "urban", "civst", "hisclass")], 
+                   FUN = sum)
+vbTab$hisclass <- factor(vbTab$hisclass)
+vbTab$age <- factor(vbTab$age)
+@
+
+<<41colorc>>=
+## Colors for plotting strata:
+scol <- c("blue", rep("darkgreen", 2), "red")
+## Common for all groups:
+
+labb <- levels(vbTab$period)
+nper <- length(labb)
+source("R/getPHglm.R")
+## End 'common'
+@
+
+Models are fitted separately for each ten-year time period, from  1901 to
+2000, with the exception of the period 1951--1960. We stratify on HISCLASS
+and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
+(Skellefteå town or outside). The cumulative
+hazards at the end of the time period (60 if 40--64, 90 if 65--89) are the
+main target in our analysis: These numbers are used in the process of
+graphically illustrating the development over time.
+
+%creating an \emph{heterogeneity index} (a better name is needed:
+%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
+%Something catchy.).
+
+\subsection{Ages 40--64}
+
+%The distribution over time periods.
+<<41getske4060>>=
+cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, causes 41."
+scap1w <-  "Cumulative hazards, ages 40--64, women, causes 41."
+cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, causes 41."
+scap1m <-  "Cumulative hazards, ages 40--64, men, causes 41."
+##
+cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom), causes 41."
+scap12 <- "Total hazard by HISCLASS, ages 40--64, causes 41."    
+cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, causes 41."
+scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, causes 41."
+
+cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, causes 41."
+scap2w <-  "Cumulative hazards, ages 65--89, women, causes 41."
+cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, causes 41."
+scap2m <-  "Cumulative hazards, ages 65--89, men, causes 41."
+##
+cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom), causes 41."
+scap22 <- "Total hazard by HISCLASS, ages 65--89, causes 41."    
+cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, causes 41."
+scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, causes 41."
+
+##
+vb4064 <- vbTab[as.numeric(as.character(vbTab$age)) < 64.5, ]
+vb4064$age <- factor(vb4064$age)
+@
+
+We start by looking at cumulative hazards by {\sc Hisclass} over time
+periods. They are adjusted for {\tt civil status}, and
+{\tt urban/rural} residence.
+We present separate analyses  for women and men.
+The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.
+
+%\subsubsection{Women}
+
+<<41hazcalw>>=
+source("R/getPHglm2.R")
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+            nper, labb)
+@
+
+<<41hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
+source("R/plotHaz.glm.R")
+
+plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+@
+
+<<41hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
+plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+@
+
+From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
+the idea of proportional hazards is not reasonable.The figures do however indicate substantial changes in the social pattern of mortality during the 19th century, something we will return to.
+
+We will use as a summary measure the value of the cumulative hazards at
+duration 20 (age 60).
+
+<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
+                main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
+                main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
+%%@
+
+Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century. Apart from
+the social pattern that will be commented on below, there are a couple of
+observations that can be made from the figure. Women had consistently
+better survival than men. This is a well-known phenomenon in Swedish
+demographic history. Male mortality has almost always been higher than
+female in all age groups \citep{wil99, suwi07}. We also observe a
+substantial decline in mortality for both sexes, resulting in smaller
+absolute differences between social classes. The compression that is
+visible for later decades is partly explained by the decreasing general
+mortality. This effect can be eliminated by showing these numbers on a log
+scale, showing the relative levels. Presented in that way, it is not
+obvious that differences between social classes have diminished. This
+conclusion has been confirmed by separate analyses (not shown). There is no
+clear indication of homogenisation of mortality levels between social
+classes.
+
+In Figure~\ref{fig:cumslogw} a "blow-up" of the last decades show clearly the upper hand of the upper classes.
+
+<<41cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
+oldpar <- par(mfrow = c(2, 1))
+wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
+main = "Women")
+mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
+main = "Men")
+par(oldpar)
+@
+
+%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
+
+<<41lissyoung>>=
+library(eha)
+library(xtable)
+load("data/loTab.rda")
+loTab$period <- factor(loTab$period)
+liss4064 <- loTab[loTab$age < 64.5, ]
+@
+
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<41liswom, results = 'asis'>>=
+fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+##cat("Women:\n")
+drw <- drop1(fitw1, test = "Chisq")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, causes 41.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, causes 41.", fig.height = 4>>=
+source("R/plotWEduc.R")
+plotWEduc(fitw1)
+@
+
+And for men:
+
+<<41lismen, results='asis'>>=
+fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+##cat("\nMen:\n")
+drm <- drop1(fitm1, test = "Chisq")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, causes 41.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, causes 41.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm1)
+@
+
+<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, causes 41.", fig.height = 4>>=
+#oldpar <- par(mfrow = c(1, 2))
+#plot(fitw1, fn = "cum", col = dcol, main = "Women")
+#plot(fitm1, fn = "cum", col = dcol, main = "Men")
+#par(oldpar)
+##summary(fitw1)
+##summary(fitm1)
+@
+
+%See Table~\ref{tab:tablisswom1} the PH regression for women.
+<<41tablisswom1, results = "asis">>=
+#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "female", ])
+#dr <- drop1(fitwo1, test = "Chisq")
+#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.", 
+#    label = "tab:tablisswom1")
+@
+
+%See Table~\ref{tab:tablissman1} the PH regression for men.
+<<41tablissman1, results = "asis">>=
+#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "male", ])
+#dr <- drop1(fitma1, test = "Chisq")
+#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.", 
+ #   label = "tab:tablissman1")
+@
+
+The central question in this paper is however what classes are most
+advantaged when it comes to mortality, and if the social pattern changed
+during the 20th century. Notice that we have not analysed the 1950s due to
+the fact that we only have data for some parishes at that time,
+consequently having no corresponding population for that
+decade. Furthermore, the results for the 1960s should be considered with
+caution. The way the social class is defined leads to much missing
+information for that decade, something that have restricted the size of the
+population observed, making the results more shaky. The changing economy
+also have consequences for the analysis, particularly when it comes to the
+agricultural sector. While farmers were common occupations during the first
+half of the 20th century, these occupations has constantly decreased in
+numbers thereafter.
+
+The social pattern among women are mixed, but in accordance with previous
+analyses from the present research group on other regions in northern
+Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
+social classes usually have comparatively low mortality, while working
+class women have high. This is especially the case during the latest
+decades of the century and especially for the period 2002--2013.
+
+A different pattern appears among men. The groups we would expect to have
+the lowest mortality, if we assume that access to different resources and
+having a high status determines survival, instead have the highest during
+much of the century. This is also in accordance with our previous studies
+\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
+low mortality. This changes however towards the end the 20th century. From
+the 1980s, there is a clear advantage when it comes to survival for higher
+managers and professionals. This is also what we usually find in analyses
+of social health inequalities in present-day Sweden as well as in most
+other countries. In the regions we are studying, this is however a quite
+recent phenomenon.The advantage for higher classes and disadvantage for low
+skilled workers is very clear for the period 2002--2013.
+
+\subsection{Ages 65--89}
+
+%The distribution over time periods.
+<<41getske6089>>=
+vb6089 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6089$age <- factor(vb6089$age) # NOTE:: No change to '6589'!
+@
+
+We now turn to the elderly, i.e. those that mostly have left the workforce. We start by looking at cumulative hazards by {\sc Hisclass} over time
+periods. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
+{\tt urban/rural} residence and with separate analyses for women and men.
+We do not present any results concerning possible homogenisation of mortality levels, since the main results do not deviate from that of the younger group.
+
+%\subsubsection{Women}
+
+<<41hazcalwold>>=
+resw <- getPHglm(vb6089[vb6089$sex == "female", ],
+              nper, labb)
+resm <- getPHglm(vb6089[vb6089$sex == "male", ],
+                nper, labb)
+@
+
+<<41hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 7>>=
+##source("R/plotHaz.R")
+plotHaz.glm(resw[-(1:6)], main = labb, layout = c(2, 2), col = scol)
+@
+
+<<41hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 7>>=
+plotHaz.glm(resm[-(1:6)], main = labb, layout = c(2, 2), col = scol)
+@
+
+The cumulative hazards for the age group 65--89 in different decades are
+presented in Figures~\ref{fig:hazplotwold} and
+\ref{fig:hazplotmold}. Corresponding to what we find for the age group
+40--64, it is obvious that the assumption of proportional hazards is not
+reasonable.
+
+We will use as a summary measure the value of the cumulative hazards at 
+duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
+
+The general pattern do not differ from the one of the younger age
+group. The levels are slightly more homogeneous, but we cannot establish
+that the elderly diverge to any degree from the working-age
+population. Mortality among women was, as expected, lower than among
+men. We also find a clear decline in mortality, but this decline did not
+set in until sometimes after the Second World War. This corresponds to the
+Swedish development where we find a turning point at this time with better
+survival among elderly, to a large extent caused by lower mortality in
+cardiovascular diseases. The results for the 1960s are however uncertain as
+discussed above, and should not be taken to seriously. The levels seem to
+be to low. This is something we need to control further.
+
+Concerning the social pattern, we find results fairly similar to the ones
+for the age group 40--64. For women the results are mixed, even though
+women from the elite had comparatively low mortality and working class
+women high. For men we find the same intriguing change among elderly as for
+the adult population 40--64 years. Mortality was in fact substantially
+higher among men from the elite and the group of lower managers while it
+was lowest among workers. This did not change until the last decades of the
+20th century. At that time, the highest social classes instead had the best
+survival while the opposite was the case for workers. The pattern had
+reversed, thus confirming the results from most studies of this topic
+concerning the present-day society. The advantage for the higher classes
+and the disadvantage for the workers are apparent also for this age group
+during the early 21st century.
+
+<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+##source("R/plotED.R")
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
+                main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
+                main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+A blow-up of the last years is shown in Figure~\ref{fig:cumslogwm}.
+
+<<41cumslogwm, fig.cap = cap22log, fig.scap = scap22log, fig.height = 10>>=
+oldpar <- par(mfrow = c(2, 1))
+wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
+main = "Women")
+mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
+main = "Men")
+par(oldpar)
+@
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period is based on a separate analysis with
+income and education as explanatory variables.
+
+<<41lissyoung2>>=
+library(eha)
+library(xtable)
+##load("data/loTab.rda")
+##loTab$period <- factor(loTab$period)
+liss6589 <- loTab[loTab$age > 64.5, ]
+@
+
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<41liswom2, results = 'asis'>>=
+fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+##cat("Women:\n")
+drw2 <- drop1(fitw2, test = "Chisq")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, causes 41.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, causes 41.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitw2)
+@
+
+And for men:
+
+<<41lismen2, results='asis'>>=
+fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+##cat("\nMen:\n")
+drm2 <- drop1(fitm2, test = "Chisq")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, causes 41.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, causes 41.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm2)
+@
+
 %%{\huge Här ska 'conclusion.tex' in !!!}

### 2016-10-08 oldmort.Rnw
fix errors for cause 41

@@
 load("data/vb.rda")
+vb <- cal.window(vb, c(1961, 2001))
 x <- substr(vb$ULORSAK, 1, 2)
@@
 vb$exposure <- with(vb, exit - enter)
+vb$cohort <- factor(vb$cohort)
+vb$period <- factor(vb$period)
 vbTab <- aggregate(vb[, c("event", "exposure")], 
@@
 
-Models are fitted separately for each ten-year time period, from  1901 to
-2000, with the exception of the period 1951--1960. We stratify on HISCLASS
+Models are fitted separately for each ten-year time period, from  1961 to
+2000. We stratify on HISCLASS
 and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
@@
 
-plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+plotHaz.glm(resw, main = labb, col = scol, layout = c(2, 2))
 @
@@
 <<41hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
-plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+plotHaz.glm(resm, main = labb, col = scol, layout = c(2, 2))
 @
@@
 oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
+wloutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
 main = "Women")
-mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
+mloutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
 main = "Men")
@@
 ##source("R/plotHaz.R")
-plotHaz.glm(resw[-(1:6)], main = labb, layout = c(2, 2), col = scol)
+plotHaz.glm(resw, main = labb, layout = c(2, 2), col = scol)
 @
@@
 <<41hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 7>>=
-plotHaz.glm(resm[-(1:6)], main = labb, layout = c(2, 2), col = scol)
+plotHaz.glm(resm, main = labb, layout = c(2, 2), col = scol)
 @
@@
 oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
+wloutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
 main = "Women")
-mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
+mloutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
 main = "Men")

### 2016-10-27 R/plotED.glm.R
update

@@ -10,6 +10,8 @@ plotED.glm <- function(res,
 
     ##nper <- NCOL(res[[1]])
 
+    ## Assuming 'm == 4', we try:
+    leg = c("1+2", "3", "4+5", "6+7")
     nper <- length(res)
     k <- NCOL(res[[1]])
     m <- NROW(res[[1]])
@@ -46,6 +48,6 @@ plotED.glm <- function(res,
     }
     abline(v = xv[out], lty = 3, col = "magenta")
     abline(h = 0)
-    legend("topright", legend = 1:m, col = col, pch = 1:m, lty = 1:m)
+    legend("topright", legend = leg, col = col, pch = 1:m, lty = 1:m)
     invisible(outp)
 }

### 2016-10-27 R/plotHaz.glm.R
update

@@ -4,6 +4,8 @@ plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
     ## Note, number of levels of hisclass is hardcoded to 4 (four) here!!
     ## Will be fixed.
     
+    leg <- c("1+2", "3+4", "5", "6+7") ## a hack for the time being
+    
     oldpar <- par(mfrow = layout)
     on.exit(par(oldpar))
     n <- length(res)
@@ -25,7 +27,7 @@ plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
         for (j in 2:4){
             lines(x, y[j, ], col = col[j], lty = j, lwd = 2)
         }
-        legend("topleft", legend = 1:4, col = col, lty = 1:4, cex = 1.0)
+        legend("topleft", legend = leg, col = col, lty = 1:4, cex = 1.0)
     }
     abline(h = 0)
 }

### 2016-10-27 oldmort.Rnw
fix spelling error
update

--- 2016-10-06 oldmort.Rnw
@@
 <<readtab>>=
-load("data/vbTab.rda")
+## We want to add hisclass at 40 (hc40) and hisclass at 60 (hc60), so as a shortcut
+## we start over with vb in the creation of the 'new vbTab':
+##load("data/vbTab.rda") 
+load("data/vb.rda")
+vb <- vb[order(vb$id, vb$age), ]
+vb$exposure <- with(vb, exit - enter)
+first <- function(x) x[1]
+hisclass40 <- with(vb, tapply(hisclass, id, first))
+indx <- match(vb$id, names(hisclass40))
+vb$hisclass40 <- hisclass40[indx]
+##
+vb$hisclass60 <- 0
+hisclass60 <- with(vb[vb$age > 59, ], tapply(hisclass, id, first))
+indx <- match(vb$id, names(hisclass60))
+vb$hisclass60 <- hisclass60[indx]
+komp <- is.na(vb$hisclass60)
+vb$hisclass60[komp] <- vb$hisclass40[komp]
+vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+4", "5", "6+7"))
+vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+4", "5", "6+7"))
+## Now to the creation of vbTab:
+vbTab <- aggregate(vb[, c("event", "exposure")],
+                   vb[, c("sex", "age", "period", "cohort", "urban",
+                          "civst", "hisclass40", "hisclass60")],
+                   FUN = sum)
+vbTab$age <- factor(vbTab$age)
 @
+
 \section{Introduction}
@@
 
+\section{The Skellefteå and Umeå regions}
+
+Figure~\ref{fig:mapp} shows the study ares and thier position in Sweden.
+\begin{figure}
+\includegraphics[height=19cm]{myfigs/sweden}
+\caption{The Skellefteå and Umeå regions in Sweden.}
+\label{fig:mapp}
+\end{figure}
+Both regions are part
+of the county of Västerbotten in the north of Sweden, along the coast of
+the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection
+of parishes surrounding the town of Skellefteå, founded in 1845 but with a
+very small population in the 19th century. The data from the period after
+1960 cover the Skellefteå and Malå municipalities, the same area as for the
+earlier period but with the addition of two more parishes. The majority
+of the population in the region lived in rural villages and hamlets, getting
+their livelihood from agricultural production. The region was vulnerable for
+harvest failures, one living long in memory was the famine of 1867–1868
+(Edvinsson and Broström, 2014). Towards the end of the century, the re-
+gion became connected to the Swedish railway system, thus improving the
+communications and contacts with other parts of the country. During the
+20th century, industrialisation took place, partly by the opening of mines in
+the inlands. This also led to a population increase both in the town and in
+the rural parts, and a much more diversified economy.
+Mortality was fairly low in comparison with other parts of the country
+and the fertility transition was late. The Skellefteå population size as defined
+in our datasets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900,
+61\,938 on January 1, 1950 and 76\,723 at the end of the century.
+
+The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and
+rural parish, and in the Linneus database from 1960 onwards of Umeå mu-
+nicipality where another three parishes are included. This region had a
+somewhat different character from that of Skellefteå, even though agricul-
+ture was dominating the economy for a long time. The town of Umeå had a
+small population but substantially larger than Skellefteå town during almost
+the whole studied period. It was the administrative centre in the county of
+Västerbotten and schools and military regiments was placed here. Officials,
+teachers and the military were much more common in Umeå. Agriculture
+dominated in the rural part, while and there were some foundries, some
+industries for example in forestry and small-scale production. Umeå had a
+much more diversified occupational structure than Skellefteå. The popula-
+tion size as defined in our datasets was 18\,970 on January 1, 1900, 32\,900 on
+January 1, 1950 and 103\,970 when the century ended.
+
+Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure time over hisclass  for our selected time periods.
+
+<<hiscskum, fig.cap = "Exposure by hisclass (from the top: 1+2, 3+4, 5, 6+7) and time period", fig.height = 9>>=
+xx <- with(vbTab, tapply(exposure, list(period = period, hisclass = hisclass40), sum))
+yy <- round(prop.table(xx, 1) * 100, 1)
+barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2)
+@
+
 \section{Previous research}
 
-For a detailed presentation we refer to \citet{segb16}. Her is a short
+For a detailed presentation we refer to \citet{segb16}. Here is a short
 summary.
@@
 and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
-(Skellefteå town or outside). The cumulative
-hazards at the end of the time period (60 if 40--64, 90 if 65--89) are the
+(Skellefteå/Umeå town or outside). The cumulative
+hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
 main target in our analysis: These numbers are used in the process of
@@
 ##
-cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
 scap12 <- "Total hazard by HISCLASS, ages 40--64."    
-cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
 scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom)."
 scap22 <- "Total hazard by HISCLASS, ages 65--89."    
-cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
 scap22log <- "Total hazard by HISCLASS, ages 65--89, last years."
@@
 vb4064$age <- factor(vb4064$age)
+vb4064$hisclass <- vb4064$hisclass40
 @
@@
 vb6089$age <- factor(vb6089$age) # NOTE:: No change to '6589'!
+vb6089$hisclass <- vb6089$hisclass60
 @
@@
 
+\section{Results, heart diseases \ldots}
+
+<<causes41>>=
+load("data/vb.rda")
+vb <- cal.window(vb, c(1961, 2001))
+x <- substr(vb$ULORSAK, 1, 2)
+vb$event <- vb$event & (!is.na(x) & (x == "19"))
+vb$exposure <- with(vb, exit - enter)
+vb$cohort <- factor(vb$cohort)
+vb$period <- factor(vb$period)
+vbTab <- aggregate(vb[, c("event", "exposure")], 
+                   vb[, c("sex", "age", "period", "cohort", 
+                          "urban", "civst", "hisclass")], 
+                   FUN = sum)
+vbTab$hisclass <- factor(vbTab$hisclass)
+vbTab$age <- factor(vbTab$age)
+@
+
+<<41colorc>>=
+## Colors for plotting strata:
+scol <- c("blue", rep("darkgreen", 2), "red")
+## Common for all groups:
+
+labb <- levels(vbTab$period)
+nper <- length(labb)
+source("R/getPHglm.R")
+## End 'common'
+@
+
+Models are fitted separately for each ten-year time period, from  1961 to
+2000. We stratify on HISCLASS
+and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
+(Skellefteå town or outside). The cumulative
+hazards at the end of the time period (60 if 40--64, 90 if 65--89) are the
+main target in our analysis: These numbers are used in the process of
+graphically illustrating the development over time.
+
+%creating an \emph{heterogeneity index} (a better name is needed:
+%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
+%Something catchy.).
+
+\subsection{Ages 40--64}
+
+%The distribution over time periods.
+<<41getske4060>>=
+cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, causes 19."
+scap1w <-  "Cumulative hazards, ages 40--64, women, causes 19."
+cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, causes 19."
+scap1m <-  "Cumulative hazards, ages 40--64, men, causes 19."
+##
+cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom), causes 19."
+scap12 <- "Total hazard by HISCLASS, ages 40--64, causes 19."    
+cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, causes 19."
+scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, causes 19."
+
+cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, causes 19."
+scap2w <-  "Cumulative hazards, ages 65--89, women, causes 19."
+cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, causes 19."
+scap2m <-  "Cumulative hazards, ages 65--89, men, causes 19."
+##
+cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom), causes 19."
+scap22 <- "Total hazard by HISCLASS, ages 65--89, causes 19."    
+cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, causes 19."
+scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, causes 19."
+
+##
+vb4064 <- vbTab[as.numeric(as.character(vbTab$age)) < 64.5, ]
+vb4064$age <- factor(vb4064$age)
+@
+
+We start by looking at cumulative hazards by {\sc Hisclass} over time
+periods. They are adjusted for {\tt civil status}, and
+{\tt urban/rural} residence.
+We present separate analyses  for women and men.
+The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.
+
+%\subsubsection{Women}
+
+<<41hazcalw>>=
+source("R/getPHglm2.R")
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+            nper, labb)
+@
+
+<<41hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
+source("R/plotHaz.glm.R")
+
+plotHaz.glm(resw, main = labb, col = scol, layout = c(2, 2))
+@
+
+<<41hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
+plotHaz.glm(resm, main = labb, col = scol, layout = c(2, 2))
+@
+
+From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
+the idea of proportional hazards is not reasonable.The figures do however indicate substantial changes in the social pattern of mortality during the 19th century, something we will return to.
+
+We will use as a summary measure the value of the cumulative hazards at
+duration 20 (age 60).
+
+<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
+                main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
+                main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
+%%@
+
+Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century. Apart from
+the social pattern that will be commented on below, there are a couple of
+observations that can be made from the figure. Women had consistently
+better survival than men. This is a well-known phenomenon in Swedish
+demographic history. Male mortality has almost always been higher than
+female in all age groups \citep{wil99, suwi07}. We also observe a
+substantial decline in mortality for both sexes, resulting in smaller
+absolute differences between social classes. The compression that is
+visible for later decades is partly explained by the decreasing general
+mortality. This effect can be eliminated by showing these numbers on a log
+scale, showing the relative levels. Presented in that way, it is not
+obvious that differences between social classes have diminished. This
+conclusion has been confirmed by separate analyses (not shown). There is no
+clear indication of homogenisation of mortality levels between social
+classes.
+
+In Figure~\ref{fig:cumslogw} a "blow-up" of the last decades show clearly the upper hand of the upper classes.
+
+<<41cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
+oldpar <- par(mfrow = c(2, 1))
+wloutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
+main = "Women")
+mloutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
+main = "Men")
+par(oldpar)
+@
+
+%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
+
+<<41lissyoung>>=
+library(eha)
+library(xtable)
+load("data/loTab.rda")
+loTab$period <- factor(loTab$period)
+liss4064 <- loTab[loTab$age < 64.5, ]
+@
+
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<41liswom, results = 'asis'>>=
+fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+##cat("Women:\n")
+drw <- drop1(fitw1, test = "Chisq")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, causes 19.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, causes 19.", fig.height = 4>>=
+source("R/plotWEduc.R")
+plotWEduc(fitw1)
+@
+
+And for men:
+
+<<41lismen, results='asis'>>=
+fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+##cat("\nMen:\n")
+drm <- drop1(fitm1, test = "Chisq")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, causes 19.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, causes 19.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm1)
+@
+
+<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, causes 19.", fig.height = 4>>=
+#oldpar <- par(mfrow = c(1, 2))
+#plot(fitw1, fn = "cum", col = dcol, main = "Women")
+#plot(fitm1, fn = "cum", col = dcol, main = "Men")
+#par(oldpar)
+##summary(fitw1)
+##summary(fitm1)
+@
+
+%See Table~\ref{tab:tablisswom1} the PH regression for women.
+<<41tablisswom1, results = "asis">>=
+#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "female", ])
+#dr <- drop1(fitwo1, test = "Chisq")
+#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.", 
+#    label = "tab:tablisswom1")
+@
+
+%See Table~\ref{tab:tablissman1} the PH regression for men.
+<<41tablissman1, results = "asis">>=
+#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "male", ])
+#dr <- drop1(fitma1, test = "Chisq")
+#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.", 
+ #   label = "tab:tablissman1")
+@
+
+The central question in this paper is however what classes are most
+advantaged when it comes to mortality, and if the social pattern changed
+during the 20th century. Notice that we have not analysed the 1950s due to
+the fact that we only have data for some parishes at that time,
+consequently having no corresponding population for that
+decade. Furthermore, the results for the 1960s should be considered with
+caution. The way the social class is defined leads to much missing
+information for that decade, something that have restricted the size of the
+population observed, making the results more shaky. The changing economy
+also have consequences for the analysis, particularly when it comes to the
+agricultural sector. While farmers were common occupations during the first
+half of the 20th century, these occupations has constantly decreased in
+numbers thereafter.
+
+The social pattern among women are mixed, but in accordance with previous
+analyses from the present research group on other regions in northern
+Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
+social classes usually have comparatively low mortality, while working
+class women have high. This is especially the case during the latest
+decades of the century and especially for the period 2002--2013.
+
+A different pattern appears among men. The groups we would expect to have
+the lowest mortality, if we assume that access to different resources and
+having a high status determines survival, instead have the highest during
+much of the century. This is also in accordance with our previous studies
+\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
+low mortality. This changes however towards the end the 20th century. From
+the 1980s, there is a clear advantage when it comes to survival for higher
+managers and professionals. This is also what we usually find in analyses
+of social health inequalities in present-day Sweden as well as in most
+other countries. In the regions we are studying, this is however a quite
+recent phenomenon.The advantage for higher classes and disadvantage for low
+skilled workers is very clear for the period 2002--2013.
+
+\subsection{Ages 65--89}
+
+%The distribution over time periods.
+<<41getske6089>>=
+vb6089 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6089$age <- factor(vb6089$age) # NOTE:: No change to '6589'!
+@
+
+We now turn to the elderly, i.e. those that mostly have left the workforce. We start by looking at cumulative hazards by {\sc Hisclass} over time
+periods. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
+{\tt urban/rural} residence and with separate analyses for women and men.
+We do not present any results concerning possible homogenisation of mortality levels, since the main results do not deviate from that of the younger group.
+
+%\subsubsection{Women}
+
+<<41hazcalwold>>=
+resw <- getPHglm(vb6089[vb6089$sex == "female", ],
+              nper, labb)
+resm <- getPHglm(vb6089[vb6089$sex == "male", ],
+                nper, labb)
+@
+
+<<41hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 7>>=
+##source("R/plotHaz.R")
+plotHaz.glm(resw, main = labb, layout = c(2, 2), col = scol)
+@
+
+<<41hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 7>>=
+plotHaz.glm(resm, main = labb, layout = c(2, 2), col = scol)
+@
+
+The cumulative hazards for the age group 65--89 in different decades are
+presented in Figures~\ref{fig:hazplotwold} and
+\ref{fig:hazplotmold}. Corresponding to what we find for the age group
+40--64, it is obvious that the assumption of proportional hazards is not
+reasonable.
+
+We will use as a summary measure the value of the cumulative hazards at 
+duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
+
+The general pattern do not differ from the one of the younger age
+group. The levels are slightly more homogeneous, but we cannot establish
+that the elderly diverge to any degree from the working-age
+population. Mortality among women was, as expected, lower than among
+men. We also find a clear decline in mortality, but this decline did not
+set in until sometimes after the Second World War. This corresponds to the
+Swedish development where we find a turning point at this time with better
+survival among elderly, to a large extent caused by lower mortality in
+cardiovascular diseases. The results for the 1960s are however uncertain as
+discussed above, and should not be taken to seriously. The levels seem to
+be to low. This is something we need to control further.
+
+Concerning the social pattern, we find results fairly similar to the ones
+for the age group 40--64. For women the results are mixed, even though
+women from the elite had comparatively low mortality and working class
+women high. For men we find the same intriguing change among elderly as for
+the adult population 40--64 years. Mortality was in fact substantially
+higher among men from the elite and the group of lower managers while it
+was lowest among workers. This did not change until the last decades of the
+20th century. At that time, the highest social classes instead had the best
+survival while the opposite was the case for workers. The pattern had
+reversed, thus confirming the results from most studies of this topic
+concerning the present-day society. The advantage for the higher classes
+and the disadvantage for the workers are apparent also for this age group
+during the early 21st century.
+
+<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+##source("R/plotED.R")
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
+                main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
+                main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+A blow-up of the last years is shown in Figure~\ref{fig:cumslogwm}.
+
+<<41cumslogwm, fig.cap = cap22log, fig.scap = scap22log, fig.height = 10>>=
+oldpar <- par(mfrow = c(2, 1))
+wloutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
+main = "Women")
+mloutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
+main = "Men")
+par(oldpar)
+@
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period is based on a separate analysis with
+income and education as explanatory variables.
+
+<<41lissyoung2>>=
+library(eha)
+library(xtable)
+##load("data/loTab.rda")
+##loTab$period <- factor(loTab$period)
+liss6589 <- loTab[loTab$age > 64.5, ]
+@
+
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<41liswom2, results = 'asis'>>=
+fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+##cat("Women:\n")
+drw2 <- drop1(fitw2, test = "Chisq")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, causes 19.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, causes 19.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitw2)
+@
+
+And for men:
+
+<<41lismen2, results='asis'>>=
+fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+##cat("\nMen:\n")
+drm2 <- drop1(fitm2, test = "Chisq")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, causes 19.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, causes 19.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm2)
+@
+
 %%{\huge Här ska 'conclusion.tex' in !!!}

### 2016-10-31 oldmort.Rnw
fixing...

@@
 <<readtab>>=
-## We want to add hisclass at 40 (hc40) and hisclass at 60 (hc60), so as a shortcut
-## we start over with vb in the creation of the 'new vbTab':
-##load("data/vbTab.rda") 
+load("data/vbTab.rda") 
 load("data/vb.rda")
-vb <- vb[order(vb$id, vb$age), ]
-vb$exposure <- with(vb, exit - enter)
-first <- function(x) x[1]
-hisclass40 <- with(vb, tapply(hisclass, id, first))
-indx <- match(vb$id, names(hisclass40))
-vb$hisclass40 <- hisclass40[indx]
-##
-vb$hisclass60 <- 0
-hisclass60 <- with(vb[vb$age > 59, ], tapply(hisclass, id, first))
-indx <- match(vb$id, names(hisclass60))
-vb$hisclass60 <- hisclass60[indx]
-komp <- is.na(vb$hisclass60)
-vb$hisclass60[komp] <- vb$hisclass40[komp]
-vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+4", "5", "6+7"))
-vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+4", "5", "6+7"))
-## Now to the creation of vbTab:
-vbTab <- aggregate(vb[, c("event", "exposure")],
-                   vb[, c("sex", "age", "period", "cohort", "urban",
-                          "civst", "hisclass40", "hisclass60")],
-                   FUN = sum)
-vbTab$age <- factor(vbTab$age)
 @
@@
 load("data/vb.rda")
-vb <- cal.window(vb, c(1961, 2001))
-x <- substr(vb$ULORSAK, 1, 2)
-vb$event <- vb$event & (!is.na(x) & (x == "19"))
+##vb <- cal.window(vb, c(1961, 2001))
+##x <- substr(vb$ULORSAK, 1, 2)
+vb$event <- vb$event & vb$heart
 vb$exposure <- with(vb, exit - enter)

### 2016-10-31 R/getCancer.R
adapted to Hadley

@@ -1,0 +1,12 @@
+getCancer <- function(dat){
+    ## 'dat' is 'vb'
+    dat$eb <- dat$exit + dat$birthdate
+    cancer <- with(dat, !is.na(ULORSAK) &
+                       event == 1 &
+                       ((eb < 1951) & ULORSAK == 12 |
+                            (eb >= 1969 & eb < 1997 & ULORSAK %in% 140:239) |
+                            (eb >= 1997 & eb < 2014 & ULORSAK == "C")))
+    dat$cancer <- cancer
+    dat$eb <- NULL
+    dat
+}

### 2016-10-31 R/getHeart.R
adapted to Hadley

@@ -1,0 +1,13 @@
+getHeart <- function(dat){
+    ## 'dat' is 'vb'
+    dat$eb <- dat$exit + dat$birthdate
+    heart <- with(dat, !is.na(ULORSAK) &
+                       event == 1 &
+                       ((eb < 1951) & ULORSAK == 19 |
+                            (eb >= 1969 & eb < 1987 & ULORSAK %in% 390:458) |
+                            (eb >= 1987 & eb < 1997 & ULORSAK %in% 390:459) |
+                            (eb >= 1997 & eb < 2014 & ULORSAK %in% c("I", "J"))))
+    dat$heart <- heart
+    dat$eb <- NULL
+    dat
+}

### 2016-10-31 CreateData/creDat.Rnw
first commit

\documentclass[a4paper,11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\usepackage{epstopdf}
\usepackage{url}

%\usepackage{natbib}

%\bibliographystyle{apalike}
%\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\newcommand{\R}{{\bf R}}

\usepackage[tablesonly]{endfloat}
%\usepackage{endfloat}

\renewcommand{\efloatseparator}{\vfill}

% patch \efloat@iwrite to use \protected@write
%\makeatletter
%\renewcommand\efloat@iwrite[1]{%
%   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
%\makeatother

\author{Göran Broström \\
Centre for Demographic and Ageing Research ({\sc Cedar})\\
Umeå University, Umeå, Sweden}

\title{Creating Data for "Social inequality in mortality among
  adults and elderly in northern Sweden 1850--2013"\footnote{Paper presented
    at the
    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
    Chicago, Illinois
    November 17--20, 2016. Version 2016-09-19.}}

\date{November 19, 2016}

\begin{document}

\maketitle

\begin{abstract}
hej
\end{abstract}

<<setup, include = FALSE>>=
knitr::opts_chunk$set(cache=FALSE,echo=FALSE,message=FALSE,warning=FALSE,comment=NA)##$
@

\section{Introduction}

In this study we take a long-term perspective on the development of social
inequalities in mortality in two Swedish region 1850--2013, in this case
with focus on the adult and elderly population. This is a description of
how we create the tabular form data set (anonymized).

\section{Aims and questions}

The aim of the paper is to investigate the development of social
differences using {\sc Hisclass} in mortality for the adult and elderly
population during the mortality 
transition and its relation to the development of income inequality. The
analyses 
will be performed on data from the Skellefteå and Umeå regions
1850--2013. The questions are:

\begin{enumerate}
\item Has inequality in mortality between social classes increased among
  the adult and elderly 
  population, primarily benefiting the highest social positions?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any gender differences in the effect of social position?

\item Is social position equally important among the retired population as
  for those in working age? 
\end{enumerate}

\section{Theoretical considerations}

The WHO Commission on Social Determinants of Health framework \ldots

\section{Historical background}

The historical context of the present study is \ldots

\section{Data and variables}

The data for the present study comes from two large population databases at
the Demographic Data Base, Umeå University. The early period is covered by
the database {\sc Poplink}, the digitisation of historical parish registers for
the Skellefteå and Umeå region. The population in the two
regions is large enough and socially diverse to enable studies of social
mortality inequalities. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from the regions are accessed for the period 1850--1950s.

The other large data-set comes from the Linnaeus database. It is based on different linked population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from
censuses every fifth 
year 1960--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare. The construction of these data-sets for analysis is presented below.

Information on individuals are anonymized and the two databases are not
linked between each other. This stops us from following individuals present
in both databases throughout their lives.

\subsection{Presence periods}

Differences in available information, both \ldots

\subsection{Social position}

We analyse mortality differences according to social class \ldots

\begin{enumerate}
\item \'Elite, HISCLASS 1, 2, and 3, Higher and lower managers, and higher
  professionals. 
\item HISCALSS 4, 5, 6, and 8. Farmers, fishermen, and lower professionals, lower clerical, foremen.
\item Medium skilled workers, HnISCLASS 7.
\item Lower skilled workers, HISCLASS 9, 10, 11 and 12, Lower and unskilled workers (including farm workers)
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

\subsection{POPLINK data, 1851--1950}

The POPLINK data cover 1901--1950. A covariate {\tt period} is added and used for splitting the data into calendar time periods of length 10 years each.

<<getdata>>=
source("calWin.R")
library(xtable)
library(knitr)
library(eha)
library(skum)
library(dplyr)
library(tidyr)
load("../../../Data/adult.rda")
adult <- calWin(adult, c(1851, 1951))
source("R/addPer.R")
adult <- addPer(adult, cuts = seq(1851, 1951, by = 25))
adult <- rc(adult)
## NEW hisclass, 6 + 7 --> 6:
adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
## Recode to four classes:
hisc <- adult$hisclass
    hisc[is.na(hisc)] <- -1
    x <- hisc
    hisc[x %in% 1:2] <- 1
    hisc[x %in% 3:4] <- 2
    hisc[x == 5] <- 3
    hisc[x >= 5.9] <- 4
    is.na(hisc) <- hisc == -1
    adult$hisclass <- hisc

indx <- match(adult$id, per$id)
adult$dodors <- per$dodors[indx]  ## New 29-10-2016..  and below ...
adult <- adult[, c("id", "birthdate", "sex", "enter", "exit", "event",
                     "period", "urban", "civst", "hisclass", "dodors")]
aduu <- adult
save(aduu, file = "aduu.rda")
@

\subsection{FOB data, 1961--2000}

The \emph{Fob} data is treated similarly as the POPLINK data. There is unfortunately no links between the individuals in the two data frames, and the result of that is a gap in the relevant information between the years 1950 and 1960. Furthermore, the first two decades in the \emph{Fob} data, 1960--1980, will have a lot of missing information on {\sc Hisclass}.

<<ssykonfob>>=
if (file.exists("fobb.rda")){
    load("fobb.rda")
}else{
    load("../../../Data/fobdata.rda")
    ## Cause of death:
    load("../../../Data/socstyr.rda")
    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    source("R/putONssyk.R")
    fobb <- putONssyk(fobdata)
    source("R/ssykTOhisclass.R")
    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
                                       (!is.na(hisclass) & hisclass == -1))
    ##source("R/pushHis.R")
    ##fobb <- pushHis(fobb)
    ##fobb$hisclass <- fobb$nyhisclass
    ## Instead
    fobb <- group_by(fobb, LINNEID) %>% fill_("hisclass")
    fobb$id <- fobb$LINNEID
    fobb <- addPer(fobb, cuts = seq(1961, 2001, by = 10))
    load("../../../Data/sex.rda")
    indx <- match(fobb$LINNEID, sex$LINNEID)
    fobb$KON <- sex$KON[indx]
    fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005, 
                                         248011, 248201, 248204)
    civst <- numeric(NROW(fobb))
    civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
    civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, anka
    fobb$civst <- factor(civst, 
                         labels = c("unmarried", "married", "prev.married"))
    fobb$sex <- fobb$KON
    levels(fobb$sex) <- c("male", "female")
    fobb <- fobb[, c("id", "birthdate", "sex", "enter", "exit", "event",
                     "period", "urban", "civst", "hisclass", "ULORSAK")]
    hisc <- fobb$hisclass
    hisc[is.na(hisc)] <- -1
    x <- hisc
    hisc[x %in% 1:2] <- 1
    hisc[x %in% 3:4] <- 2
    hisc[x == 5] <- 3
    hisc[x >= 5.9] <- 4
    is.na(hisc) <- hisc == -1
    fobb$hisclass <- hisc
    save(fobb, file = "fobb.rda")
}
@

\subsection{LISA data, 2002--2013}

The {\sc Lisa} data is yearly, with {\tt hisclass} information starting in 2002. This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, and 5, 6.

<<lisadata>>=
if (file.exists("liss.rda") & file.exists("lisaold.rda")){
    load("liss.rda")
    ##if (!is.null(liss$nyhisclass)){
      ##  liss$hisclass <- liss$nyhisclass
    ##    liss$nyhisclass <- NULL
    ##}
    load("lisaold.rda")
}else{
    load("../../../Data/lisadata.rda")
    ## Cause of death:
    load("../../../Data/socstyr.rda")
    indx <- match(lisadata$LINNEID, socstyr$LINNEID)
    lisadata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    civst <- lisadata$CIV
    civst[is.na(civst)] <- 0
    civst[civst == 4] <- 3
    civst[civst == 5] <- 2
    civst[civst == 6] <- 3
    is.na(civst) <- civst == 0
    lisadata$civst <- civst
    rm(civst)
    ## Fix KON:
    load("../../../Data/sex.rda")
    indx <- match(lisadata$LINNEID, sex$LINNEID)
    lisadata$sex <- sex$KON[indx]
    levels(lisadata$sex) <- c("male", "female")
    lisaold <- lisadata
    lisadata <- lisadata[lisadata$AR > 1999.5 & !is.na(lisadata$SSYK3), ]
    ## Fix HISCLASS:
    source("R/ssyk3TOhisclass.R")
    lisadata$hisclass <- ssyk3TOhisclass(lisadata$SSYK3)
    ## Push forward:
    ##source("R/pushHis.R")
    ##liss <- pushHis(lisadata)
    ##liss$hisclass <- liss$nyhisclass
    ##liss$nyhisclass <- NULL
    ## Now:
    liss <- group_by(lisadata, LINNEID) %>% fill_("hisclass")
    liss$id <- liss$LINNEID
    lisaold$id <- lisaold$LINNEID

    lisaold <- addPer(lisaold, name = "period", cuts = seq(1986, 2014, by = 2))
    liss <- addPer(liss, name = "period", cuts = seq(2002, 2014, by = 6))
    liss <- liss[!is.na(liss$hisclass), ]
    liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8), 
                         labels = 1:4)
    liss$hisclass <- as.numeric(as.character(liss$hisclass))
    liss <- liss[with(liss, !is.na(civst) &
                            !is.na(hisclass)), ]
    lisaold <- lisaold[with(lisaold, !is.na(civst)), ]
    liss$civst <- factor(liss$civst, 
                         labels = c("unmarried", "married", "prev.married"))
    lisaold$civst <- factor(lisaold$civst, 
                         labels = c("unmarried", "married", "prev.married"))
    liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004, 
                                         284005, 248011, 248201, 248204)
    lisaold$urban <- lisaold$FORSAMLING %in% c(248001, 248002, 284004, 
                                         284005, 248011, 248201, 248204)
    liss$id <- liss$LINNEID
    liss <- liss[, c("id", "birthdate", "sex", "enter", "exit", "event",
                     "period", "urban", "civst", "hisclass", "ULORSAK")]
    save(liss, file = "liss.rda")
    save(lisaold, file = "lisaold.rda")
}
@

\subsection{Combining the data sets and tabulate}

When joining,  we must pay special attention to the {\tt id} variable. There is
overlap between {\tt aduu} on one hand and {\tt fobb} and {\tt liss} on the
other. The solution is to create new id variables.

First we join ({\tt rbind}) the data sets {\tt fobb} and {\tt liss}. They are
linked via the {\tt id}. It is also good to know from which source a record
comes, so we create a new variable with precisely the name {\tt source}.

<<rbindlissfobb>>=
library(stringr)
aduu$ULORSAK <- as.character(aduu$dodors) ## New 13-10-2016
aduu$dodors <- NULL ## Yep...
fobb$source <- "fob"
liss$source <- "lisa"
aduu$source <- "popum"
foli <- rbind(fobb, liss)
##source("R/pushHis2.R")
##foli <- pushHis2(foli)
foli <- group_by(foli, id) %>% fill_("hisclass")
##aduu <- pushHis2(aduu)
aduu <- group_by(aduu, id) %>% fill_("hisclass")
aduu$id <- as.integer(as.factor(aduu$id))
foli$id <- as.integer(as.factor(foli$id)) + max(aduu$id) + 1 #;)
@

<<rbinddata, results = 'asis'>>=

vb <- rbind(aduu, foli)
x <- with(vb, table(period, hisclass, useNA = "ifany"))
y <- prop.table(x, margin = 1) * 100
print(xtable(round(y, 1), digits = 0, caption = "Hisclass by time period, 1901--2013 (per cent)."))
@

Then we group the {\tt birthdate} variable into ten-year intervals ("birth
cohorts"). Since we are interested only in ages 40--89 in the time period
1901--2013, we can remove all born 1810 and earlier and all born 1974 or
later. Finally we apply an "age window" of 40-89.

<<groupbirths, results = 'asis'>>=
source("ageWin.R")
vb <- vb[vb$birthdate > 1761 & vb$birthdate < 1974, ]
vb <- ageWin(vb, c(40, 90))
starts <- seq(1761, 1981, by = 10)
vb$cohort <- cut(vb$birthdate, starts, labels = starts[-length(starts)])
x <- with(vb, table(cohort, hisclass, useNA = "ifany"))
y <- prop.table(x, margin = 1) * 100
print(xtable(round(y, 1), digits = 0, caption = "Hisclass by birth cohort, 1801--1980 (per cent)."))
@

Now we remove all records with missing {\sc Hisclass}, and the we introduce a new variable {\tt age}, grouped in intervals 40--44, 45--49, \ldots 85--89.

<<ageint>>=
vb <- vb[!is.na(vb$hisclass), ] # same as 'na.omit(vb)' (I think)
source("R/ageGroup.R")
vb <- ageGroup(vb, seq(40, 90, by = 5)) ## Oh, oh!
lisaold <- ageGroup(lisaold, seq(40, 90, by = 5))
vb$exposure <- with(vb, exit - enter)

## Here we cut from '../oldmort.Rnw': ######################

vb <- vb[order(vb$id, vb$age), ]
source("../R/getCancer.R")
vb <- getCancer(vb)
source("../R/getHeart.R")
vb <- getHeart(vb)
vb$exposure <- with(vb, exit - enter)
first <- function(x) x[1]
hisclass40 <- with(vb, tapply(hisclass, id, first))
indx <- match(vb$id, names(hisclass40))
vb$hisclass40 <- hisclass40[indx]
##
vb$hisclass60 <- 0
hisclass60 <- with(vb[vb$age > 59, ], tapply(hisclass, id, first))
indx <- match(vb$id, names(hisclass60))
vb$hisclass60 <- hisclass60[indx]
komp <- is.na(vb$hisclass60)
vb$hisclass60[komp] <- vb$hisclass40[komp]
vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+4", "5", "6+7"))
vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+4", "5", "6+7"))

## End cut ####################################################

save(vb, file = "../data/vb.rda")
@

The next, and final, step is aggregation.

<<aggregate>>=
vbTab <- aggregate(vb[, c("event", "exposure")], 
                   vb[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass40", "hisclass60")], 
                   FUN = sum)
## Is this necessary: (See above!)
##vbTab$hisclass40 <- factor(vbTab$hisclass40)
##vbTab$hisclass60 <- factor(vbTab$hisclass60)
vbTab$age <- factor(vbTab$age)
save(vbTab, file = "../data/vbTab.rda")
@

Thus, the one data file to use is {\tt vbTab.rda}.

<<lisaold>>=
lisaold <- age.window(lisaold, c(40, 90))
lisaold$exposure <- with(lisaold, exit - enter)
lisaold$educ <- as.factor(lisaold$SUN2000NIVA_OLD)
lisaold$income <- cut(lisaold$DISPINKPERSF, c(-600, 730, 958, 1270, 500000),
                      labels = c("lowest", "low", "high", "highest"))
loTab <- aggregate(lisaold[, c("event", "exposure")], 
                   lisaold[, c("sex", "age", "period", "urban", 
                               "civst", "income", "educ")],
                   FUN = sum)
save(loTab, file = "../data/loTab.rda")
@

\ldots and of course {\tt loTab.rda}.

\end{document}

### 2016-10-31 CreateData/ageWin.R
first commit

@@ -1,0 +1,36 @@
+ageWin <-
+function (dat, window, surv = c("enter", "exit", "event")) 
+{
+    if (!is.data.frame(dat)) 
+        stop("dat must be a data frame")
+    if (length(surv) != 3) 
+        stop("surv must have length 3")
+    fixed.names <- names(dat)
+    surv.indices <- match(surv, fixed.names)
+    if (length(which(is.na(surv.indices)))) {
+        x <- which(is.na(surv.indices))
+        stop(paste(surv[x], " is not a name in the data frame."))
+    }
+    enter <- dat[[surv.indices[1]]]
+    exit <- dat[[surv.indices[2]]]
+    event <- dat[[surv.indices[3]]]
+    who <- (exit > window[1]) & (enter < window[2])
+    if (sum(who) > 0.5) {
+        enter <- enter[who]
+        exit <- exit[who]
+        event <- event[who]
+        event[exit > window[2]] <- 0
+        exit[exit > window[2]] <- window[2]
+        enter[enter < window[1]] <- window[1]
+        dat <- dat[who, ]
+        dat[[surv.indices[1]]] <- enter
+        dat[[surv.indices[2]]] <- exit
+        dat[[surv.indices[3]]] <- event
+    }
+    else {
+        warning(paste("The age interval", window[1], "-", window[2], 
+            "is empty."))
+        dat <- NULL
+    }
+    dat
+}

### 2016-10-31 CreateData/calWin.R
first commit

@@ -1,0 +1,42 @@
+calWin <-
+function (dat, window, surv = c("enter", "exit", "event", "birthdate")) 
+{
+    if (!is.data.frame(dat)) 
+        stop("dat must be a data frame")
+    if (length(surv) != 4) 
+        stop("surv must have length 4")
+    fixed.names <- names(dat)
+    surv.indices <- match(surv, fixed.names)
+    if (length(which(is.na(surv.indices)))) {
+        x <- which(is.na(surv.indices))
+        stop(paste(surv[x], " is not a name in the fixed data frame."))
+    }
+    enter <- dat[[surv.indices[1]]]
+    exit <- dat[[surv.indices[2]]]
+    event <- dat[[surv.indices[3]]]
+    bdate <- dat[[surv.indices[4]]]
+    who <- ((exit > (window[1] - bdate)) & (enter < (window[2] - 
+        bdate)))
+    if (sum(who) > 0.5) {
+        enter <- enter[who]
+        exit <- exit[who]
+        event <- event[who]
+        bdate <- bdate[who]
+        overShoot <- exit > window[2] - bdate
+        event[overShoot] <- 0
+        exit[overShoot] <- window[2] - bdate[overShoot]
+        underShoot <- enter < window[1] - bdate
+        enter[underShoot] <- window[1] - bdate[underShoot]
+        dat <- dat[who, ]
+        dat[[surv.indices[1]]] <- enter
+        dat[[surv.indices[2]]] <- exit
+        dat[[surv.indices[3]]] <- event
+        dat[[surv.indices[4]]] <- bdate
+    }
+    else {
+        warning(paste("The period", window[1], "-", window[2], 
+            "is empty."))
+        dat <- NULL
+    }
+    dat
+}

### 2016-10-31 CreateData/R/addPer.R
first commit

@@ -1,0 +1,23 @@
+addPer <- function(dat, name = "period", cuts = seq(1901, 1951, by = 10)){
+    n <- length(cuts)
+    source("calWin.R")
+    start <- calWin(dat, c(cuts[1], cuts[2]))
+    start[[name]] <- 1
+    if (n > 2){
+        for (i in (3:n)){
+            tmp <- calWin(dat, c(cuts[i - 1], cuts[i]))
+            tmp[[name]] <- i - 1
+            start <- rbind(start, tmp)
+        }
+    }
+    levs <- character(n - 1)
+
+    for (i in 1:(n - 1)){
+        levs[i] <- paste(cuts[i], "-", cuts[i + 1] - 1, sep = "")
+    }
+    start[[name]] <- factor(start[[name]], labels = levs)
+    start <- start[order(start$id, start$enter), ]
+    start <- rc(start)
+    start
+}
+

### 2016-10-31 CreateData/R/ageGroup.R

@@ -1,0 +1,18 @@
+ageGroup <- function(dat, ages){
+    n <- length(ages) - 1
+    ##age <- ages[-(n+1)]
+    if (n > 1){
+        ##library(eha)
+        source("ageWin.R")
+        dat$age <- numeric(NROW(dat))
+        first <- ageWin(dat, ages[1:2])
+        first$age <- ages[1]
+        for (i in 2:n){
+            tmp <- ageWin(dat, ages[i:(i+1)])
+            tmp$age <- ages[i]
+            first <- rbind(first, tmp)
+        }
+    }
+    first <- first[order(first$id, first$enter), ]
+    invisible(first)
+}

### 2016-10-31 CreateData/R/pH.R

@@ -1,0 +1,14 @@
+pH <- function(dat){
+    ## 'dat' is a data frame with variables 'LINNEID' and 'hisclass'
+    library(tidyr)
+    library(dplyr)
+    ##om <- unique(dat$LINNEID)
+    
+    ##hyhis <- numeric(NROW(dat))
+    ##dat$nyhisclass <- numeric(NROW(dat))
+    x <- group_by(dat, LINNEID) %>% fill_("hisclass")
+    ##for (i in om){
+      ##  dat[dat$LINNEID == i, ] <- fill(dat[dat$LINNEID == i, ], hisclass)
+    ##}
+    x
+}

### 2016-10-31 CreateData/R/pushHis2.R

@@ -1,0 +1,17 @@
+pushHis2 <- function(dat){
+    ## 'dat' is a data frame with variables 'id' and 'hisclass'
+    ## A copy of 'pushHis' with LINNEID <--> id.
+    
+    library(skum)
+    om <- unique(dat$id)
+    
+    hyhis <- numeric(NROW(dat))
+    dat$nyhisclass <- numeric(NROW(dat))
+    for (i in om){
+        who <- dat$id == i
+        ok <- dat$hisclass[who]
+        ok <- pushForward(ok)
+        dat$nyhisclass[who] <- ok
+    }
+    dat
+}

### 2016-10-31 CreateData/R/putONssyk.R

@@ -1,0 +1,59 @@
+putONssyk <- function(dat){
+    ## 'dat' is the processed file 'fobdata'
+    ## This must be run from "Mortality/CreDat"
+    
+    load("../../../Data/yr_ssyk.rda")
+    dat$ssyk <- NA
+
+    ##for (year in seq(60, 75, by = 5)){
+      ##  nam <- paste("yr", year, sep = "")
+    ##    load(paste("data/", nam, ".rda" sep = ""))
+##        nyr <- dat[dat$AR == i, ]
+  ##      indx <- match(nyr$YRKE)
+    ##}
+    ## 1960:
+    load("../../../Data/yr60.rda")
+    who <- dat$AR == 1960 & dat$YRKE %in% yr60$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr60$old)
+    nyr <- yr60$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1965:
+    load("../../../Data/yr65.rda")
+    who <- dat$AR == 1965 & dat$YRKE %in% yr65$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr65$old)
+    nyr <- yr65$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1970:
+    load("../../../Data/yr70.rda")
+    who <- dat$AR == 1970 & dat$YRKE %in% yr70$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr70$old)
+    nyr <- yr70$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1975:
+    load("../../../Data/yr75.rda")
+    who <- dat$AR == 1975 & dat$YRKE %in% yr75$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr75$old)
+    nyr <- yr65$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1980:
+    dat$YRKE[dat$AR == 1980] <- dat$YRKE80[dat$AR == 1980]
+
+    ## 1985:
+    dat$YRKE[dat$AR == 1985] <- dat$YRKE80[dat$AR == 1985]
+
+    is.na(dat$YRKE) <- dat$YRKE %in% c(0, 999)
+    
+    ## Now put on ssyk:
+    
+    indx <- match(dat$YRKE, yr_ssyk$yrke)
+    dat$ssyk <- yr_ssyk$ssyk[indx]
+    dat
+}

### 2016-10-31 CreateData/R/ssyk3TOhisclass.R

@@ -1,0 +1,31 @@
+ssyk3TOhisclass <- function(x){
+    ## 'x' is a vector of ssyk codes.
+    ## as output we want a vector of hisclass codes (7 levels)
+    
+    ## Step 1: Convert shh$ssyk to "three-digit" (except '110'):
+    load("../../../Data/shh.rda")
+    shh$ssyk <- shh$ssyk %/% 10 # removes last digit.
+    ## Remove duplicate ssyk-rows in shh:
+    shh <- shh[!duplicated(shh$ssyk), ]
+    
+    ## Step 2: SSYK3 to 12-levels hisclass
+    indx <- match(x, shh$ssyk)
+    tmp <- shh$hisclass[indx]
+
+    ## Step 2: go from 12 to 7 levels:
+    
+    res <- rep(-1, length(x))
+    
+    tmp[is.na(tmp)] <- -1
+    
+    res[tmp %in% 1:2] <- 1
+    res[tmp == 3] <- 2
+    res[tmp == 8] <- 3
+    res[tmp %in% 4:6] <- 4
+    res[tmp == 7] <- 5
+    res[tmp %in% c(9, 11)] <- 6
+    res[tmp %in% c(10, 12)] <- 7
+    is.na(res) <- res == -1
+    
+    res
+}

### 2016-10-31 CreateData/R/ssykTOhisclass.R

@@ -1,0 +1,28 @@
+ssykTOhisclass <- function(x){
+    ## 'x' is a vector of ssyk codes.
+    ## as output we want a vector of hisclass codes (7 levels)
+    
+    ## Step 1: ssyk to 12-levels hisclass
+    load("../../../Data/shh.rda")
+    indx <- match(x, shh$ssyk)
+    tmp <- shh$hisclass[indx]
+
+    ## Step 2: go from 12 to 7 levels:
+    
+    
+    res <- rep(-1, length(x))
+    
+    tmp[is.na(tmp)] <- -1
+    
+    res[tmp %in% 1:2] <- 1
+    res[tmp == 3] <- 2
+    res[tmp == 8] <- 3
+    res[tmp %in% 4:6] <- 4
+    res[tmp == 7] <- 5
+##    res[tmp %in% c(9, 11)] <- 6
+##    res[tmp %in% c(10, 12)] <- 7 # skip klass 7!
+    res[tmp %in% 9:12] <- 6
+    is.na(res) <- res == -1
+    
+    res
+}

### 2016-11-01 R/getCancer.R
use stringr

@@ -10,3 +10,14 @@ getCancer <- function(dat){
     dat$eb <- NULL
     dat
 }
+## New version: 'ULORSAK' is a character vector:
+    library(stringr)
+    eb <- dat$exit + dat$birthdate
+    ##cancer <- logical(NROW(dat))
+    ulor <- str_trim(dat$ULORSAK)
+    ## First, fix ulor start with "C":
+    fc <- (substr(ulor, 1, 1) == "C") & (eb >= 1997) & (eb < 2014)
+    cancer <- fc
+    ## Then the 'numeric' versions:
+    ##ulor <- as.numeric(ulor)  # NA if letters in it
+    

### 2016-11-01 CreateData/getlisaold.R
first commit

@@ -1,0 +1,43 @@
+getlisaold <- function(){
+    library(eha)
+    library(skum)
+    load("../../Data/lisadata.rda")
+    civst <- lisadata$CIV
+    civst[is.na(civst)] <- 0
+    civst[civst == 4] <- 3
+    civst[civst == 5] <- 2
+    civst[civst == 6] <- 3
+    is.na(civst) <- civst == 0
+    lisadata$civst <- civst
+    rm(civst)
+    ## Fix KON:
+    load("../../Data/sex.rda")
+    indx <- match(lisadata$LINNEID, sex$LINNEID)
+    lisadata$sex <- sex$KON[indx]
+    levels(lisadata$sex) <- c("male", "female")
+    lisaold <- lisadata
+    lisaold$id <- lisaold$LINNEID
+    source("R/addPer.R")
+    cat("addPer:\n")
+    lisaold <- addPer(lisaold, name = "period", cuts = seq(1986, 2014, by = 2))
+    lisaold <- lisaold[with(lisaold, !is.na(civst)), ]
+    lisaold$civst <- factor(lisaold$civst, 
+                         labels = c("unmarried", "married", "prev.married"))
+    lisaold$urban <- lisaold$FORSAMLING %in% c(248001, 248002, 284004, 
+                                               284005, 248011, 248201, 248204)
+    source("R/ageGroup.R")
+    cat("addPer:\n")
+    lisaold <- ageGroup(lisaold, seq(40, 90, by = 5))
+    lisaold <- age.window(lisaold, c(40, 90))
+    lisaold$exposure <- with(lisaold, exit - enter)
+    lisaold$educ <- as.factor(lisaold$SUN2000NIVA_OLD)
+    lisaold$income <- cut(lisaold$DISPINKPERSF, c(-600, 730, 958, 1270, 500000),
+                          labels = c("lowest", "low", "high", "highest"))
+    save(lisaold, file = "lisaold.rda")
+    loTab <- aggregate(lisaold[, c("event", "exposure")], 
+                       lisaold[, c("sex", "age", "period", "urban", 
+                                   "civst", "income", "educ")],
+                       FUN = sum)
+    save(loTab, file = "../Mortality/data/loTab.rda")
+    invisible(loTab)
+}

### 2016-11-01 CreateData/showData4.Rmd
first commit

---
title: "showData 4"
author: "Göran Broström"
date: "September 22, 2016"
output: pdf_document
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
options(stringsAsFactors = FALSE)
```

# Introduction

Now *showData4.Rmd* is updated to include *fob90* and also let the
follow-up for *fob85* be *five years* and the follow-up for *fob90* be *ten
years*. We also add, as an update to *showData3*, guys aged 20 or more to the 
*FoB data*. Note that results (data files) are saved in **data4**!

This is an updated version of *showData2.Rmd*, where we use the newly-found
knowledge that the yearly *lisa* data relates to *December 31* the given
year, while the *FoB* data relate to the date *November 1* the given
year. These facts have the following implications for our plan: For *FoB*
data, we follow individuals from their appearance in a FoB until the next
FoB five years later, even if they do not appear there. The exception is if
they die within five years after the first appearance. So, *fob60* induces
a time interval *from November 1, 1960 to October 31, 1965*, and so on. To
include all 40+ in this interval, we must select all that are *born before
November 1, 1925*, that is, those who have reached the age 35 at November
1, 1960. Since we do not have birth dates but only birth year and quarter,
we decide to use as birthdate the *midpoint of the birth quarter*.

The following FoBs are treated similarly, until we reach the Fob
in 1985. Since the first *lisa* date is December 31, 1986, this "fob
interval" starts at November 1, 1985, and ends at December 31, 1986, that
is, it is 14 months long.

# Fixed data

This time we start with the *inddata* file and treat it as a *fixed-data*
file (no time-varying info). 
We find the birth dates (see above), death dates (for those who are observed die), and the "date last seen".
We also remove individuals that were born 1975 or later because they never get the chance to reach forty before *December 31, 2014*, which is our stop date.

Note that 37 individuals in *inddata* have no information at all, so they are removed.

```{r frominddata}
load("data/inddata.rda")
load("data/sex.rda")
indx <- match(inddata$LINNEID, sex$LINNEID)
inddata$KON <- sex$KON[indx]
per <- inddata[!is.na(inddata$KON) & (inddata$FODELSEAR < 1975), ]
per$birthdate <- per$FODELSEAR + 0.125  +  (per$FODELSEKVARTAL - 1) / 4
```

Now we put on *death date*, *death cause*, and *KOMMUN of death*, from *socstyr*, on *per*:

```{r styronper}
load("data/socstyr.rda")
indx <- match(per$LINNEID, socstyr$LINNEID)
per$DODSDAT <- socstyr$DODSDAT[indx]
fixd <- !is.na(per$DODSDAT) & (per$DODSDAT < 1960010)
per$DODSDAT[fixd] <- per$DODSDAT[fixd] * 100 + 15
per$deathdate <- as.numeric(skum::tillTid(per$DODSDAT))
per$dodsKOMMUN <- socstyr$KOMMUN[indx]
per$dodsALDER <- socstyr$ALDER[indx]
per$ULORSAK <- socstyr$ULORSAK[indx]
```

# The FoBs

Through the *FoB*s we have snapshots of the population on *November 1* the years *1960(5)1990*, giving the dates (decimal form).

```{r days}
ar <- seq(1960, 1990, by = 5)
days <- as.numeric(skum::tillTid(paste(ar, "1101", sep = "")))
days
```

We have to limit the data collection to the *KOMMUN*er that are contained
in our area, that is

```{r kommuner}
ourkom <- c(2406, 2414, 2417, 2418, 2480, 2482)
ourkom <- c(ourkom, 2411, 2412, 2413, 2416, 2415, 2461, 2407, 2402, 2408)
```

Note that this is per year.

## FoB 1960

Persons in the fob and in our KOMMUNer.

```{r fob60}
load("data/fob.rda")
fob60 <- fob[fob$AR == 1960 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1940 or earlier). We
get the birthdates from *inddata*.

```{r mat60}
library(eha)
indx <- match(fob60$LINNEID, per$LINNEID)
fob60$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1940-11-01"))
fob60 <- fob60[fob60$birthdate < lim, ]
indx <- match(fob60$LINNEID, per$LINNEID) # NOTE: Necessary!
fob60$enter <- lim + 20 - fob60$birthdate
dd <- per$deathdate[indx]
fob60$event <- !(is.na(dd)) & (dd <= lim + 25) # NOTE: 40 --> 25!
fob60$exit <- fob60$enter + 5
fob60$exit[fob60$event] <- dd[fob60$event] - fob60$birthdate[fob60$event]
head(fob60)
```

## FoB 1965

Persons in the fob and in our KOMMUNer.

```{r fob65}
fob65 <- fob[fob$AR == 1965 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1930 or earlier). We
get the birthdates from *inddata*.

```{r mat65}
library(eha)
indx <- match(fob65$LINNEID, per$LINNEID)
fob65$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1945-11-01"))
fob65 <- fob65[fob65$birthdate < lim, ]
indx <- match(fob65$LINNEID, per$LINNEID) # NOTE: Necessary!
fob65$enter <- lim + 20 - fob65$birthdate
dd <- per$deathdate[indx]
fob65$event <- !(is.na(dd)) & (dd <= lim + 25) ## NOTE: 40 --> 25!
fob65$exit <- fob65$enter + 5
fob65$exit[fob65$event] <- dd[fob65$event] - fob65$birthdate[fob65$event]
fob65 <- fob65[fob65$exit > fob65$enter, ] # Shouldn't be necessary!
head(fob65)
```

It should be noted that three persons present in the 1965 FoB actually died 1964 or earlier.

## FoB 1970

Persons in the fob and in our KOMMUNer.

```{r fob70}
fob70 <- fob[fob$AR == 1970 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1935 or earlier). We
get the birthdates from *inddata*.

```{r mat70}
library(eha)
indx <- match(fob70$LINNEID, per$LINNEID)
fob70$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1950-11-01"))
fob70 <- fob70[fob70$birthdate <= lim, ]
indx <- match(fob70$LINNEID, per$LINNEID) # NOTE: Necessary!
fob70$enter <- lim + 20 - fob70$birthdate
dd <- per$deathdate[indx]
fob70$event <- !(is.na(dd)) & (dd <= lim + 25)
fob70$exit <- fob70$enter + 5
fob70$exit[fob70$event] <- dd[fob70$event] - fob70$birthdate[fob70$event]
fob70 <- fob70[fob70$exit > fob70$enter, ] # Shouldn't be necessary!
head(fob70)
cat("lim = ", lim, "\n")
```

It should be noted that one person present in the 1970 FoB actually died 1964.

## FoB 1975

Persons in the fob and in our KOMMUNer.

```{r fob75}
fob75 <- fob[fob$AR == 1975 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1940 or earlier). We
get the birthdates from *inddata*.

```{r mat75}
library(eha)
indx <- match(fob75$LINNEID, per$LINNEID)
fob75$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1955-11-01"))
fob75 <- fob75[fob75$birthdate < lim, ]
indx <- match(fob75$LINNEID, per$LINNEID) # NOTE: Necessary!
fob75$enter <- lim + 20 - fob75$birthdate
dd <- per$deathdate[indx]
fob75$event <- !(is.na(dd)) & (dd <= lim + 25)
fob75$exit <- fob75$enter + 5
fob75$exit[fob75$event] <- dd[fob75$event] - fob75$birthdate[fob75$event]
fob75 <- fob75[fob75$exit > fob75$enter, ] # Shouldn't be necessary!
head(fob75)
##cat("lim = ", lim, "\n")
```

It should be noted that one person present in the 1975 FoB actually died 1964 (same person as earlier).

## FoB 1980

Persons in the fob and in our KOMMUNer.

```{r fob80}
fob80 <- fob[fob$AR == 1980 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1945 or earlier). We
get the birthdates from *inddata*.

```{r mat80}
library(eha)
indx <- match(fob80$LINNEID, per$LINNEID)
fob80$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1960-11-01"))
fob80 <- fob80[fob80$birthdate < lim, ]
indx <- match(fob80$LINNEID, per$LINNEID) # NOTE: Necessary!
fob80$enter <- lim + 20 - fob80$birthdate
dd <- per$deathdate[indx]
fob80$event <- !(is.na(dd)) & (dd <= lim + 25)
fob80$exit <- fob80$enter + 5
fob80$exit[fob80$event] <- dd[fob80$event] - fob80$birthdate[fob80$event]
fob80 <- fob80[fob80$exit > fob80$enter, ] # Shouldn't be necessary!
head(fob80)
##cat("lim = ", lim, "\n")
```

It should be noted that one person present in the 1980 FoB actually died 1964 (same person as earlier).

## FoB 1985

This is no longer the last FoB we are using: The follow-up time this time
*is* five years (as before), but five years, from November 1, 1985 to October 31, 1990, when the *lisa* data fills in.

Persons in the fob and in our KOMMUNer.

```{r fob85}
fob85 <- fob[fob$AR == 1985 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1950 or earlier). We
get the birthdates from *inddata*.

```{r mat85}
library(eha)
indx <- match(fob85$LINNEID, per$LINNEID)
fob85$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1965-11-01"))
fob85 <- fob85[fob85$birthdate < lim, ]
indx <- match(fob85$LINNEID, per$LINNEID) # NOTE: Necessary!
fob85$enter <- lim + 20 - fob85$birthdate
##fob85$enter <- as.numeric(toTime("1985-11-01")) - fob85$birthdate
dd <- per$deathdate[indx]
fob85$event <- !(is.na(dd)) & (dd <= lim + 25)
##fob85$event <- !(is.na(dd)) & (dd < 1987) 
fob85$exit <- fob85$enter + 5  ## Not any longer: 1 year and 2 months.
fob85$exit[fob85$event] <- dd[fob85$event] - fob85$birthdate[fob85$event]
fob85 <- fob85[fob85$exit > fob85$enter, ] # Shouldn't be necessary!
head(fob85)
##cat("lim = ", lim, "\n")
```

It should be noted that one person present in the 1985 FoB actually died 1964 (same person as earlier).

## FoB 1990

This is no longer the last FoB we are using: The follow-up time this time
*is* ten years (as before), from November 1, 1990 to October 31, 2000, when
the *lisa* data has been filling in for a long time.

Persons in the fob and in our KOMMUNer.

```{r fob90}
fob90 <- fob[fob$AR == 1990 & 
                 fob$KOMMUN %in% ourkom &
                 fob$LINNEID %in% per$LINNEID, ]
```

They need to be of the right age (born on November 1, 1960 or earlier). We
get the birthdates from *inddata*.

```{r mat90}
library(eha)
indx <- match(fob90$LINNEID, per$LINNEID)
fob90$birthdate <- per$birthdate[indx]
lim <- as.numeric(toTime("1960-11-01"))
fob90 <- fob90[fob90$birthdate < lim, ]
indx <- match(fob90$LINNEID, per$LINNEID) # NOTE: Necessary!
fob90$enter <- lim + 20 - fob90$birthdate
##fob90$enter <- as.numeric(toTime("1985-11-01")) - fob85$birthdate
dd <- per$deathdate[indx]
fob90$event <- !(is.na(dd)) & (dd <= lim + 20)
##fob90$event <- !(is.na(dd)) & (dd < 1987) 
fob90$exit <- fob90$enter + 10  ## Not any longer: 1 year and 2 months.
fob90$exit[fob90$event] <- dd[fob90$event] - fob90$birthdate[fob90$event]
fob90 <- fob90[fob90$exit > fob90$enter, ] # Shouldn't be necessary!
head(fob90)
##cat("lim = ", lim, "\n")
```

It should be noted that one person present in the 1985 FoB actually died 1964 (same person as earlier).

This ends the data collection from the FoBs. Remains to put all together.

## Putting the data from all FoBs in one data frame

We put the event histories described by the FoBs (with FoB90!) into one data frame.

```{r fobgrab}
fobdata <- rbind(fob60, fob65, fob70, fob75, fob80, fob85, fob90)
fobdata <- fobdata[order(fobdata$LINNEID, fobdata$enter), ]
##fobdata <- age.window(fobdata, c(40, 110))
rownames(fobdata) <- 1:NROW(fobdata)
save(fobdata, file = "data4/fobdata.rda")
```

# The Lisa data

The *lisa* data is slightly simpler because it is yearly with cutoff dates 12-31/01-01 each new-year. It starts off at 1987-01-01 (lisa1986) and ends at 2013-12-31 (lisa2013). So the time interval a *lisa* covers is *AR + 1*! But note that the *socstyr* data only cover deaths *before January 1, 2014, so *lisa2013* is in effect useless to us.

```{r lisadat}
load("data/lisa.rda")
lisa <- lisa[lisa$KOMMUN %in% ourkom, ]
indx <- match(lisa$LINNEID, per$LINNEID)
lisa$birthdate <- per$birthdate[indx]
```

Those with *missing birthdate* in *lisa* are those born 1975 or later. They can be removed.

```{r removeyoung}
lisa <- lisa[!is.na(lisa$birthdate), ]
```

So, in *lisa*, all enter in *AR + 1*, at the exact age of *(AR + 1 - birthdate)*, and exit at the exact age *enter + 1* if they do not die in *the year AR + 1*, otherwise at their death date, or at *deathdate - birthdate*.

```{r fixlisaintervals}
lisa$enter <- lisa$AR + 1 - lisa$birthdate
indx <- match(lisa$LINNEID, per$LINNEID) # To be safe ...
lisa$deathdate <- per$deathdate[indx]
lisa$event <- !is.na(lisa$deathdate) & lisa$deathdate > lisa$AR + 1 & lisa$deathdate <= lisa$AR + 2
lisa$exit <- lisa$enter + 1 # First tip.
lisa$exit[lisa$event] <- with(lisa, deathdate[event] - birthdate[event])
```

Now some sorting  and cleaning.

```{r cleanandsort}
##lisadata <- age.window(lisa, c(40, 110))
lisadata <- lisa[order(lisa$LINNEID, lisa$enter), ]
lisadata <- cal.window(lisadata, c(1987, 2014)) # 'Remove lisa2013'
rownames(lisadata) <- 1:NROW(lisadata)
save(lisadata, file = "data4/lisadata.rda")
```

Now, read *checkData* (maybe not).

### 2016-11-01 CreateData/creDat.Rnw
fix errors with causes

@@
 library(tidyr)
+library(stringr)
 load("../../../Data/adult.rda")
@@
 vb <- rbind(aduu, foli)
+vb$period <- as.factor(vb$period)
 x <- with(vb, table(period, hisclass, useNA = "ifany"))

### 2016-11-01 R/getCancer.R
fix errors with cancer causes

@@ -1,23 +1,30 @@
 getCancer <- function(dat){
     ## 'dat' is 'vb'
-    dat$eb <- dat$exit + dat$birthdate
-    cancer <- with(dat, !is.na(ULORSAK) &
-                       event == 1 &
-                       ((eb < 1951) & ULORSAK == 12 |
-                            (eb >= 1969 & eb < 1997 & ULORSAK %in% 140:239) |
-                            (eb >= 1997 & eb < 2014 & ULORSAK == "C")))
-    dat$cancer <- cancer
-    dat$eb <- NULL
-    dat
-}
+    ##dat$eb <- dat$exit + dat$birthdate
+    ##cancer <- with(dat, !is.na(ULORSAK) &
+    ##                   event == 1 &
+    ##                   ((eb < 1951) & ULORSAK == 12 |
+    ##                        (eb >= 1969 & eb < 1997 & ULORSAK %in% 140:239) |
+    ##                        (eb >= 1997 & eb < 2014 & ULORSAK == "C")))
+    ##dat$cancer <- cancer
+    ##dat$eb <- NULL
+    ##dat
+#}
 ## New version: 'ULORSAK' is a character vector:
     library(stringr)
     eb <- dat$exit + dat$birthdate
     ##cancer <- logical(NROW(dat))
+    
     ulor <- str_trim(dat$ULORSAK)
     ## First, fix ulor start with "C":
     fc <- (substr(ulor, 1, 1) == "C") & (eb >= 1997) & (eb < 2014)
     cancer <- fc
     ## Then the 'numeric' versions:
     ##ulor <- as.numeric(ulor)  # NA if letters in it
-    
+    fc <- (ulor %in% as.character(140:239)) & (eb >= 1969) & (eb < 1997)
+    cancer <- cancer | fc
+    fc <- (ulor == "12") & (eb < 1951)
+    cancer <- cancer | fc
+    dat$cancer <- dat$event & cancer
+    dat
+}

### 2016-11-01 R/getHeart.R
fix errors with cardiovascular causes

@@ -1,13 +1,28 @@
 getHeart <- function(dat){
     ## 'dat' is 'vb'
-    dat$eb <- dat$exit + dat$birthdate
-    heart <- with(dat, !is.na(ULORSAK) &
-                       event == 1 &
-                       ((eb < 1951) & ULORSAK == 19 |
-                            (eb >= 1969 & eb < 1987 & ULORSAK %in% 390:458) |
-                            (eb >= 1987 & eb < 1997 & ULORSAK %in% 390:459) |
-                            (eb >= 1997 & eb < 2014 & ULORSAK %in% c("I", "J"))))
-    dat$heart <- heart
-    dat$eb <- NULL
+##    dat$eb <- dat$exit + dat$birthdate
+##    heart <- with(dat, !is.na(ULORSAK) &
+##                       event == 1 &
+##                       ((eb < 1951) & ULORSAK == 19 |
+##                            (eb >= 1969 & eb < 1987 & ULORSAK %in% 390:458) |
+##                            (eb >= 1987 & eb < 1997 & ULORSAK %in% 390:459) |
+##                            (eb >= 1997 & eb < 2014 & ULORSAK %in% c("I", "J"))))
+##    dat$heart <- heart
+##    dat$eb <- NULL
+##    dat
+
+    ## New version (ULORSAK is a character vector)
+    eb <- dat$exit + dat$birthdate
+    library(stringr)
+    ulor <- str_trim(dat$ULORSAK)
+    fc <- (substr(ulor, 1, 1) %in% c("I", "J")) & (eb >= 1997) & (eb < 2014)
+    heart <- fc
+    fc <- (ulor %in% as.character(390:459)) & (eb >= 1987) & (eb < 1997)
+    heart <- heart | fc
+    fc <- (ulor %in% as.character(390:458)) & (eb >= 1969) & (eb < 1987)
+    heart <- heart | fc
+    fc <- (ulor == "19") & (eb < 1951)
+    heart <- heart | fc
+    dat$heart <- dat$event & heart
     dat
 }

### 2016-11-01 oldmort.Rnw
minor fix
run with fixed data

@@
 load("data/vbTab.rda") 
+vbTab$period <- factor(vbTab$period)
 load("data/vb.rda")
+vb$period <- factor(vb$period)
 @
@@
 We will use as a summary measure the value of the cumulative hazards at
-duration 20 (age 60).
+age 65.
 
@@
 <<getske6089>>=
-vb6089 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
-vb6089$age <- factor(vb6089$age) # NOTE:: No change to '6589'!
-vb6089$hisclass <- vb6089$hisclass60
+vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6589$age <- factor(vb6589$age) # NOTE:: No change to '6589'!
+vb6589$hisclass <- vb6589$hisclass60
 @
@@
 <<hazcalwold>>=
-resw <- getPHglm(vb6089[vb6089$sex == "female", ],
+resw <- getPHglm(vb6589[vb6589$sex == "female", ],
               nper, labb)
-resm <- getPHglm(vb6089[vb6089$sex == "male", ],
+resm <- getPHglm(vb6589[vb6589$sex == "male", ],
                 nper, labb)
@@
 
-Models are fitted separately for each ten-year time period, from  1961 to
-2000. We stratify on HISCLASS
+Models are fitted separately for each ten-year time period, from  1981 to
+2013. We stratify on HISCLASS
 and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
 (Skellefteå town or outside). The cumulative
-hazards at the end of the time period (60 if 40--64, 90 if 65--89) are the
+hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
 main target in our analysis: These numbers are used in the process of
@@
 <<41getske4060>>=
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, causes 19."
-scap1w <-  "Cumulative hazards, ages 40--64, women, causes 19."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, causes 19."
-scap1m <-  "Cumulative hazards, ages 40--64, men, causes 19."
+cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cardiovascular death."
+scap1w <-  "Cumulative hazards, ages 40--64, women, cardiovascular death."
+cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cardiovascular death."
+scap1m <-  "Cumulative hazards, ages 40--64, men, cardiovascular death."
 ##
-cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom), causes 19."
-scap12 <- "Total hazard by HISCLASS, ages 40--64, causes 19."    
-cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, causes 19."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, causes 19."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, causes 19."
-scap2w <-  "Cumulative hazards, ages 65--89, women, causes 19."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, causes 19."
-scap2m <-  "Cumulative hazards, ages 65--89, men, causes 19."
+cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
+scap12 <- "Total hazard by HISCLASS, ages 40--64, cardiovascular death."    
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cardiovascular death."
+
+cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cardiovascular death."
+scap2w <-  "Cumulative hazards, ages 65--89, women, cardiovascular death."
+cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cardiovascular death."
+scap2m <-  "Cumulative hazards, ages 65--89, men, cardiovascular death."
 ##
-cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom), causes 19."
-scap22 <- "Total hazard by HISCLASS, ages 65--89, causes 19."    
-cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, causes 19."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, causes 19."
+cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
+scap22 <- "Total hazard by HISCLASS, ages 65--89, cardiovascular death."    
+cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cardiovascular death."
 
@@
 drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, causes 19.", label = "tab:liswom")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:liswom")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 
-<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, causes 19.", fig.height = 4>>=
+<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
 source("R/plotWEduc.R")
@@
 drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, causes 19.")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, causes 19.", fig.height = 4>>=
+<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
-<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, causes 19.", fig.height = 4>>=
+<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
 #oldpar <- par(mfrow = c(1, 2))
@@
 <<41getske6089>>=
-vb6089 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
-vb6089$age <- factor(vb6089$age) # NOTE:: No change to '6589'!
+vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6589$age <- factor(vb6589$age) # NOTE:: Now change to '6589'!
 @
@@
 <<41hazcalwold>>=
-resw <- getPHglm(vb6089[vb6089$sex == "female", ],
+resw <- getPHglm(vb6589[vb6589$sex == "female", ],
               nper, labb)
-resm <- getPHglm(vb6089[vb6089$sex == "male", ],
+resm <- getPHglm(vb6589[vb6589$sex == "male", ],
                 nper, labb)
@@
 drw2 <- drop1(fitw2, test = "Chisq")
-xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, causes 19.", label = "tab:liswom")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:liswom")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 
-<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, causes 19.", fig.height = 4>>=
+<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, causes 19.")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, causes 19.", fig.height = 4>>=
+<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
 #source("R/plotWEduc.R")

### 2016-11-02 R/plotED.glm.R
modify input

@@ -1,6 +1,7 @@
 plotED.glm <- function(res,
-                       col = 1:length(res),
+                       col = 1:NROW(res[[1]]),
                        logScale = FALSE,
+                       relative = FALSE,
                        ylab = "",
                        main = "",
                        exclude = 100){
@@ -11,7 +12,7 @@ plotED.glm <- function(res,
     ##nper <- NCOL(res[[1]])
 
     ## Assuming 'm == 4', we try:
-    leg = c("1+2", "3", "4+5", "6+7")
+    leg = c("1+2", "3+4", "5", "6+7")
     nper <- length(res)
     k <- NCOL(res[[1]])
     m <- NROW(res[[1]])
@@ -28,15 +29,24 @@ plotED.glm <- function(res,
     if (logScale){
         outp <- log(outp)
     }
+    if (relative){
+        outp <- prop.table(outp, 2)
+    }
     fr <- min(outp, 0)
-    to <- max(outp)
+    if (relative){
+        to <- max(outp[, out])
+        fr <- min(outp[, out])
+        #cat("to = ", to, ", fr = ", fr, "\n")
+    }else{    
+        to <- max(outp)
+    }
     xv <- (as.numeric(substr(names(res), 1, 4)) +
         as.numeric(substr(names(res), 6, 9))) / 2
 
-    plot(xv[out], outp[1, -exclude], type = "b", lty = 2, col = col[1], pch = 1,
+    plot(xv[out], outp[1, -exclude], type = "b", lty = 1, col = col[1], pch = 1,
          ylim = c(fr, to), main = main,
          axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
-         xlim = c(min(xv), max(xv) + 10))
+         xlim = c(min(xv), max(xv) + 25))
     axis(1, at = xv, lab = labb, las = 2)
     axis(2, las = 2)
     box()

### 2016-11-02 oldmort.Rnw
fix causes of death, remove some figures

@@
-\documentclass[a4paper,11pt]{article}
+\documentclass[a4paper,11pt,titlepage,twoside]{article}
 
@@
 
-\author{Sören Edvinsson and Göran Broström \\
+\author{Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}} and
+    Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} \\
 Centre for Demographic and Ageing Research ({\sc Cedar})\\
-Umeå University, Umeå, Sweden}
+Umeå University, SE-90187 Umeå, Sweden}
 
@@
     Chicago, Illinois
-    November 17--20, 2016. (Version 2016-10-06).}}
+    November 17--20, 2016}}
 
@@
 \end{abstract}
+
+\tableofcontents
+
+\newpage
 
@@
 
+For the corresponding numbers, with missing values ({\tt NA}) included, see Table~\ref{tab:na}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{r|rrrrr}
+  \hline
+Period & 1+2 & 3+4 & 5 & 6+7 & {\tt NA} \\ 
+  \hline
+1851-1875 & 2 & 32 & 4 & 49 & 14 \\ 
+  1876-1900 & 3 & 34 & 6 & 42 & 15 \\ 
+  1901-1925 & 5 & 34 & 10 & 33 & 18 \\ 
+  1926-1950 & 7 & 26 & 16 & 35 & 16 \\ 
+  1961-1970 & 4 & 18 & 7 & 20 & 50 \\ 
+  1971-1980 & 6 & 25 & 9 & 30 & 31 \\ 
+  1981-1990 & 8 & 28 & 8 & 30 & 26 \\ 
+  1991-2000 & 6 & 24 & 6 & 24 & 39 \\ 
+  2002-2007 & 21 & 35 & 9 & 35 & 0 \\ 
+  2008-2013 & 22 & 35 & 9 & 34 & 0 \\ 
+   \hline
+\end{tabular}
+\caption{Hisclass by time period, 1901--2013 (per cent). {\tt NA}: Missing information.} 
+\label{tab:na}
+\end{table}
+
 \section{Previous research}
@@
 
-Models are fitted separately for each ten-year time period, from  1901 to
-2000, with the exception of the period 1951--1960. We stratify on HISCLASS
+Models are fitted separately for each time period, from  1851 to
+2013, with the exception of the period 1951--1960. We stratify on HISCLASS
 and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
@@
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard by HISCLASS, ages 40--64."    
+cap0 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
+scap0 <- "Total hazard by HISCLASS, ages 40--64." 
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardized by HISCLASS and decade, women (top) and men (bottom)."
+scap12 <- "Total hazard standardized by HISCLASS, ages 40--64."    
 cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
@@
 
-We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. They are adjusted for {\tt civil status}, and
-{\tt urban/rural} residence.
-We present separate analyses  for women and men.
-The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.
+We start by looking at cumulative hazards by {\sc Hisclass} over the first four time
+periods, in order to illustrate the process. They are adjusted for {\tt civil status}, and
+{\tt urban/rural} residence. We present separate analyses  for women and men, see 
+Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.
 
@@
 From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
-the idea of proportional hazards is not reasonable.The figures do however indicate substantial changes in the social pattern of mortality during the 19th century, something we will return to.
-
-We will use as a summary measure the value of the cumulative hazards at
-age 65.
-
-<<cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+the idea of proportional hazards is not reasonable.
+
+The "TMR", that is, the summed hazards over the age interval 40--64 is shown in 
+Figure~\ref{fig:cumsw}.
+
+<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 10>>=
 source("R/plotED.glm.R")
@@
                 main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+Obviously, and not surprisingly, mortality is decreasing over time, and it is possible to see that the upper classes for men are moving from a top position in the nineteenth century to having the lowest mortality around the last turn of centuries.
+Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century.
+Apart from
+the social pattern that will be commented on below, there are a couple of
+observations that can be made from the figure. Women had consistently
+better survival than men. This is a well-known phenomenon in Swedish
+demographic history. Male mortality has almost always been higher than
+female in all age groups \citep{wil99, suwi07}. We also observe a
+substantial decline in mortality for both sexes, resulting in smaller
+absolute differences between social classes. The compression that is
+visible for later decades is partly explained by the decreasing general
+mortality. This effect can be eliminated by showing these numbers on a log
+scale, showing the relative levels. Presented in that way, it is not
+obvious that differences between social classes have diminished. This
+conclusion has been confirmed by separate analyses (not shown). There is no
+clear indication of homogenisation of mortality levels between social
+classes.
+
+However, we are mainly interested in illustrating the change in relative position among the hisclasses over time rather than the ablolute levels, so we \emph{standardize} the levels in each time period to sume to unity.
+It is illustrated in
+Figure~\ref{fig:cumswstand}.
+
+<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
+                    ylab = "TMR(65|40)", main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "TMR(65|40)", main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
+%%@
+
+%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
+
+<<lissyoung>>=
+library(eha)
+library(xtable)
+load("data/loTab.rda")
+loTab$period <- factor(loTab$period)
+liss4064 <- loTab[loTab$age < 64.5, ]
+@
+
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<liswom, results = 'asis'>>=
+fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+##cat("Women:\n")
+drw <- drop1(fitw1, test = "Chisq")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
+source("R/plotWEduc.R")
+plotWEduc(fitw1)
+@
+
+And for men:
+
+<<lismen, results='asis'>>=
+fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+##cat("\nMen:\n")
+drm <- drop1(fitm1, test = "Chisq")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm1)
+@
+
+<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
+#oldpar <- par(mfrow = c(1, 2))
+#plot(fitw1, fn = "cum", col = dcol, main = "Women")
+#plot(fitm1, fn = "cum", col = dcol, main = "Men")
+#par(oldpar)
+##summary(fitw1)
+##summary(fitm1)
+@
+
+%See Table~\ref{tab:tablisswom1} the PH regression for women.
+<<tablisswom1, results = "asis">>=
+#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "female", ])
+#dr <- drop1(fitwo1, test = "Chisq")
+#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.", 
+#    label = "tab:tablisswom1")
+@
+
+%See Table~\ref{tab:tablissman1} the PH regression for men.
+<<tablissman1, results = "asis">>=
+#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "male", ])
+#dr <- drop1(fitma1, test = "Chisq")
+#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.", 
+ #   label = "tab:tablissman1")
+@
+
+The central question in this paper is however what classes are most
+advantaged when it comes to mortality, and if the social pattern changed
+during the 20th century. Notice that we have not analysed the 1950s due to
+the fact that we only have data for some parishes at that time,
+consequently having no corresponding population for that
+decade. Furthermore, the results for the 1960s should be considered with
+caution. The way the social class is defined leads to much missing
+information for that decade, something that have restricted the size of the
+population observed, making the results more shaky. The changing economy
+also have consequences for the analysis, particularly when it comes to the
+agricultural sector. While farmers were common occupations during the first
+half of the 20th century, these occupations has constantly decreased in
+numbers thereafter.
+
+The social pattern among women are mixed, but in accordance with previous
+analyses from the present research group on other regions in northern
+Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
+social classes usually have comparatively low mortality, while working
+class women have high. This is especially the case during the latest
+decades of the century and especially for the period 2002--2013.
+
+A different pattern appears among men. The groups we would expect to have
+the lowest mortality, if we assume that access to different resources and
+having a high status determines survival, instead have the highest during
+much of the century. This is also in accordance with our previous studies
+\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
+low mortality. This changes however towards the end the 20th century. From
+the 1980s, there is a clear advantage when it comes to survival for higher
+managers and professionals. This is also what we usually find in analyses
+of social health inequalities in present-day Sweden as well as in most
+other countries. In the regions we are studying, this is however a quite
+recent phenomenon.The advantage for higher classes and disadvantage for low
+skilled workers is very clear for the period 2002--2013.
+
+\subsection{Ages 65--89}
+
+%The distribution over time periods.
+<<getske6089>>=
+vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6589$age <- factor(vb6589$age) # NOTE:: No change to '6589'!
+vb6589$hisclass <- vb6589$hisclass60
+@
+
+We now turn to the elderly, i.e. those that mostly have left the workforce. 
+We will use as a summary measure the value of the cumulative hazards at 
+duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
+
+The general pattern do not differ from the one of the younger age
+group. The levels are slightly more homogeneous, but we cannot establish
+that the elderly diverge to any degree from the working-age
+population. Mortality among women was, as expected, lower than among
+men. We also find a clear decline in mortality, but this decline did not
+set in until sometimes after the Second World War. This corresponds to the
+Swedish development where we find a turning point at this time with better
+survival among elderly, to a large extent caused by lower mortality in
+cardiovascular diseases. The results for the 1960s are however uncertain as
+discussed above, and should not be taken to seriously. The levels seem to
+be to low. This is something we need to control further.
+
+Concerning the social pattern, we find results fairly similar to the ones
+for the age group 40--64. For women the results are mixed, even though
+women from the elite had comparatively low mortality and working class
+women high. For men we find the same intriguing change among elderly as for
+the adult population 40--64 years. Mortality was in fact substantially
+higher among men from the elite and the group of lower managers while it
+was lowest among workers. This did not change until the last decades of the
+20th century. At that time, the highest social classes instead had the best
+survival while the opposite was the case for workers. The pattern had
+reversed, thus confirming the results from most studies of this topic
+concerning the present-day society. The advantage for the higher classes
+and the disadvantage for the workers are apparent also for this age group
+during the early 21st century.
+
+<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+##source("R/plotED.R")
+resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
+            nper, labb)
+
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "TMR(90|65)", main = "Women", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "TMR(90|65)", main = "Men", exclude = 5:6)
+par(oldpar)
+@
+
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period is based on a separate analysis with
+income and education as explanatory variables.
+
+<<lissyoung2>>=
+library(eha)
+library(xtable)
+##load("data/loTab.rda")
+##loTab$period <- factor(loTab$period)
+liss6589 <- loTab[loTab$age > 64.5, ]
+@
+
+The statistical significance of the covariates are shown in 
+Table~\ref{tab:liswom2}
+<<liswom2, results = 'asis'>>=
+fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+##cat("Women:\n")
+drw2 <- drop1(fitw2, test = "Chisq")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
+
+<<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitw2)
+@
+
+And for men:
+
+<<lismen2, results='asis'>>=
+fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+##cat("\nMen:\n")
+drm2 <- drop1(fitm2, test = "Chisq")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm2)
+@
+
+\section{Results, cardiovascular death}
+
+<<causes41>>=
+cvb <- vb
+cvb$event <- cvb$heart ## Can this be done better?
+cvb <- cvb[!(cvb$period %in% c("1961-1970", "1971-1980")), ]
+cvb$period <- factor(cvb$period)
+
+# load("data/vb.rda")
+# ##vb <- cal.window(vb, c(1961, 2001))
+# ##x <- substr(vb$ULORSAK, 1, 2)
+# vb$event <- vb$event & vb$heart
+# vb$exposure <- with(vb, exit - enter)
+# vb$cohort <- factor(vb$cohort)
+# vb$period <- factor(vb$period)
+cvbTab <- aggregate(cvb[, c("event", "exposure")], 
+                    cvb[, c("sex", "age", "period", "cohort", 
+                           "urban", "civst", "hisclass")], 
+                    FUN = sum)
+ cvbTab$hisclass <- factor(cvbTab$hisclass)
+ cvbTab$age <- factor(cvbTab$age)
+@
+
+<<41colorc>>=
+## Colors for plotting strata:
+scol <- c("blue", rep("darkgreen", 2), "red")
+## Common for all groups:
+
+labb <- levels(cvbTab$period)
+nper <- length(labb)
+source("R/getPHglm.R")
+## End 'common'
+@
+
+Models are fitted separately for each ten-year time period, from  1851 to
+2013. We stratify on HISCLASS
+and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
+(Skellefteå town or outside). The cumulative
+hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
+main target in our analysis: These numbers are used in the process of
+graphically illustrating the development over time.
+
+%creating an \emph{heterogeneity index} (a better name is needed:
+%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
+%Something catchy.).
+
+\subsection{Ages 40--64}
+
+%The distribution over time periods.
+<<41getske4060>>=
+cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cardiovascular death."
+scap1w <-  "Cumulative hazards, ages 40--64, women, cardiovascular death."
+cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cardiovascular death."
+scap1m <-  "Cumulative hazards, ages 40--64, men, cardiovascular death."
+##
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
+scap12 <- "Total hazard by HISCLASS, ages 40--64, cardiovascular death."    
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cardiovascular death."
+
+cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cardiovascular death."
+scap2w <-  "Cumulative hazards, ages 65--89, women, cardiovascular death."
+cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cardiovascular death."
+scap2m <-  "Cumulative hazards, ages 65--89, men, cardiovascular death."
+##
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
+scap22 <- "Total hazard standardized by HISCLASS, ages 65--89, cardiovascular death."    
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cardiovascular death."
+
+##
+cvb4064 <- cvbTab[as.numeric(as.character(cvbTab$age)) < 64.5, ]
+cvb4064$age <- factor(cvb4064$age)
+@
+
+We start by looking at cumulative hazards by {\sc Hisclass} over time
+periods. They are adjusted for {\tt civil status}, and
+{\tt urban/rural} residence.
+We present separate analyses  for women and men.
+The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.
+
+%\subsubsection{Women}
+
+<<41hazcalw>>=
+source("R/getPHglm2.R")
+resw <- getPHglm2(cvb4064[cvb4064$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(cvb4064[cvb4064$sex == "male", ],
+            nper, labb)
+@
+
+We will use as a summary measure the value of the cumulative hazards at
+age 65, given survival to age 40.
+
+<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
+                    ylab = "TMR(65|40)",  main = "Women")
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, 
+                    ylab = "TMR(65|40)", main = "Men")
 par(oldpar)
@@
 
-In Figure~\ref{fig:cumslogw} a "blow-up" of the last decades show clearly the upper hand of the upper classes.
-
-<<cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
-oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
-main = "Women")
-mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(65|40)",
-main = "Men")
-par(oldpar)
-@
-
 %% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
@@
 
-<<lissyoung>>=
+<<41lissyoung>>=
 library(eha)
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
-<<liswom, results = 'asis'>>=
+The statistical significance of the covariates are shown in 
+Table~\ref{tab:41liswom}
+<<41liswom, results = 'asis'>>=
 fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
@@
 drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-
-<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
+The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
+
+<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
 source("R/plotWEduc.R")
@@
 
-<<lismen, results='asis'>>=
+<<41lismen, results='asis'>>=
 fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
@@
 drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
+<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
-<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
+<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
 #oldpar <- par(mfrow = c(1, 2))
@@
 %See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<tablisswom1, results = "asis">>=
+<<41tablisswom1, results = "asis">>=
 #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
@@
 %See Table~\ref{tab:tablissman1} the PH regression for men.
-<<tablissman1, results = "asis">>=
+<<41tablissman1, results = "asis">>=
 #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
@@
 %The distribution over time periods.
-<<getske6089>>=
-vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
-vb6589$age <- factor(vb6589$age) # NOTE:: No change to '6589'!
-vb6589$hisclass <- vb6589$hisclass60
+<<41getske6089>>=
+cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
+cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
 @
@@
 
-<<hazcalwold>>=
-resw <- getPHglm(vb6589[vb6589$sex == "female", ],
+<<41hazcalwold>>=
+resw <- getPHglm(cvb6589[cvb6589$sex == "female", ],
               nper, labb)
-resm <- getPHglm(vb6589[vb6589$sex == "male", ],
+resm <- getPHglm(cvb6589[cvb6589$sex == "male", ],
                 nper, labb)
@@
 
-<<hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 7>>=
-##source("R/plotHaz.R")
-plotHaz.glm(resw[-(1:6)], main = labb, layout = c(2, 2), col = scol)
-@
-
-<<hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 7>>=
-plotHaz.glm(resm[-(1:6)], main = labb, layout = c(2, 2), col = scol)
-@
-
-The cumulative hazards for the age group 65--89 in different decades are
-presented in Figures~\ref{fig:hazplotwold} and
-\ref{fig:hazplotmold}. Corresponding to what we find for the age group
-40--64, it is obvious that the assumption of proportional hazards is not
-reasonable.
-
 We will use as a summary measure the value of the cumulative hazards at 
-duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
+duration 25 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
 
@@
 
-<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
 ##source("R/plotED.R")
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
-                main = "Women", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
-                main = "Men", exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "H(90|65)",  main = "Women", exclude = c(1))
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "H(90|65)", main = "Men", exclude = c(1))
 par(oldpar)
@@
 
-A blow-up of the last years is shown in Figure~\ref{fig:cumslogwm}.
-
-<<cumslogwm, fig.cap = cap22log, fig.scap = scap22log, fig.height = 10>>=
-oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
-main = "Women")
-mloutp <- plotED.glm(resm[-(1:6)], col = scol, logScale = FALSE, ylab = "H(90|65)",
-main = "Men")
-par(oldpar)
-@
-
-\subsubsection*{The effect of income and education, 1990--2005}
+\subsubsection*{The effect of income and education on cardiovascular death, 1990--2005}
 
@@
 
-<<lissyoung2>>=
+<<41lissyoung2>>=
 library(eha)
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
-<<liswom2, results = 'asis'>>=
+The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2}
+<<41liswom2, results = 'asis'>>=
 fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
@@
 drw2 <- drop1(fitw2, test = "Chisq")
-xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 
-<<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
+<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
-<<lismen2, results='asis'>>=
+<<41lismen2, results='asis'>>=
 fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
@@
 drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-<<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
+<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
-\section{Results, heart diseases \ldots}
-
-<<causes41>>=
-load("data/vb.rda")
-##vb <- cal.window(vb, c(1961, 2001))
-##x <- substr(vb$ULORSAK, 1, 2)
-vb$event <- vb$event & vb$heart
-vb$exposure <- with(vb, exit - enter)
-vb$cohort <- factor(vb$cohort)
-vb$period <- factor(vb$period)
-vbTab <- aggregate(vb[, c("event", "exposure")], 
-                   vb[, c("sex", "age", "period", "cohort", 
+\section{Results, cancer death}
+
+<<causes12>>=
+cvb <- vb
+cvb$event <- cvb$cancer ## Can this be done better?
+cvb <- cvb[!(cvb$period %in% c("1961-1970", "1971-1980")), ]
+cvb$period <- factor(cvb$period)
+
+# load("data/vb.rda")
+# ##vb <- cal.window(vb, c(1961, 2001))
+# ##x <- substr(vb$ULORSAK, 1, 2)
+# vb$event <- vb$event & vb$heart
+# vb$exposure <- with(vb, exit - enter)
+# vb$cohort <- factor(vb$cohort)
+# vb$period <- factor(vb$period)
+cvbTab <- aggregate(cvb[, c("event", "exposure")], 
+                    cvb[, c("sex", "age", "period", "cohort", 
                            "urban", "civst", "hisclass")], 
                     FUN = sum)
-vbTab$hisclass <- factor(vbTab$hisclass)
-vbTab$age <- factor(vbTab$age)
-@
-
-<<41colorc>>=
+ cvbTab$hisclass <- factor(cvbTab$hisclass)
+ cvbTab$age <- factor(cvbTab$age)
+@
+
+<<12colorc>>=
 ## Colors for plotting strata:
@@
 
-labb <- levels(vbTab$period)
+labb <- levels(cvbTab$period)
 nper <- length(labb)
@@
 
-Models are fitted separately for each ten-year time period, from  1981 to
+Models are fitted separately for each ten-year time period, from  1851 to
 2013. We stratify on HISCLASS
@@
 %The distribution over time periods.
-<<41getske4060>>=
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cardiovascular death."
-scap1w <-  "Cumulative hazards, ages 40--64, women, cardiovascular death."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cardiovascular death."
-scap1m <-  "Cumulative hazards, ages 40--64, men, cardiovascular death."
+<<12getske4060>>=
+cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cancer death."
+scap1w <-  "Cumulative hazards, ages 40--64, women, cancer death."
+cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cancer death."
+scap1m <-  "Cumulative hazards, ages 40--64, men, cancer death."
 ##
-cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap12 <- "Total hazard by HISCLASS, ages 40--64, cardiovascular death."    
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cardiovascular death."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cardiovascular death."
-scap2w <-  "Cumulative hazards, ages 65--89, women, cardiovascular death."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cardiovascular death."
-scap2m <-  "Cumulative hazards, ages 65--89, men, cardiovascular death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cancer death."
+scap12 <- "Total hazard by HISCLASS, ages 40--64, cancer death."    
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
+scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cancer death."
+
+cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cancer death."
+scap2w <-  "Cumulative hazards, ages 65--89, women, cancer death."
+cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cancer death."
+scap2m <-  "Cumulative hazards, ages 65--89, men, cancer death."
 ##
-cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap22 <- "Total hazard by HISCLASS, ages 65--89, cardiovascular death."    
-cap22log <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cardiovascular death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cancer death."
+scap22 <- "Total hazard standardized by HISCLASS, ages 65--89, cancer death."    
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
+scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cancer death."
 
 ##
-vb4064 <- vbTab[as.numeric(as.character(vbTab$age)) < 64.5, ]
-vb4064$age <- factor(vb4064$age)
+cvb4064 <- cvbTab[as.numeric(as.character(cvbTab$age)) < 64.5, ]
+cvb4064$age <- factor(cvb4064$age)
 @
@@
 
-<<41hazcalw>>=
+<<12hazcalw>>=
 source("R/getPHglm2.R")
-resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+resw <- getPHglm2(cvb4064[cvb4064$sex == "female", ],
             nper, labb)
-resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+resm <- getPHglm2(cvb4064[cvb4064$sex == "male", ],
             nper, labb)
@@
 
-<<41hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
-source("R/plotHaz.glm.R")
-
-plotHaz.glm(resw, main = labb, col = scol, layout = c(2, 2))
-@
-
-<<41hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
-plotHaz.glm(resm, main = labb, col = scol, layout = c(2, 2))
-@
-
-From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
-the idea of proportional hazards is not reasonable.The figures do however indicate substantial changes in the social pattern of mortality during the 19th century, something we will return to.
-
 We will use as a summary measure the value of the cumulative hazards at
-duration 20 (age 60).
-
-<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+age 65, given survival to age 40.
+
+<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
 source("R/plotED.glm.R")
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Women", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Men", exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
+                    ylab = "TMR(65|40)",  main = "Women")
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, 
+                    ylab = "TMR(65|40)", main = "Men")
 par(oldpar)
@@
 
-In Figure~\ref{fig:cumslogw} a "blow-up" of the last decades show clearly the upper hand of the upper classes.
-
-<<41cumslogw, fig.cap = cap12log, fig.scap = scap12log, fig.height = 10>>=
-oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
-main = "Women")
-mloutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
-main = "Men")
-par(oldpar)
-@
-
 %% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
@@
 
-<<41lissyoung>>=
+<<12lissyoung>>=
 library(eha)
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
-<<41liswom, results = 'asis'>>=
+The statistical significance of the covariates are shown in Table~\ref{tab:12liswom}
+<<12liswom, results = 'asis'>>=
 fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
@@
 drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:liswom")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cancer death.", label = "tab:12liswom")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-
-<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
+The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}
+
+<<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
 source("R/plotWEduc.R")
@@
 
-<<41lismen, results='asis'>>=
+<<12lismen, results='asis'>>=
 fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
@@
 drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
+<<12lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cancer death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
-<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
+<<12plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cancer death.", fig.height = 4>>=
 #oldpar <- par(mfrow = c(1, 2))
@@
 %See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<41tablisswom1, results = "asis">>=
+<<12tablisswom1, results = "asis">>=
 #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
@@
 %See Table~\ref{tab:tablissman1} the PH regression for men.
-<<41tablissman1, results = "asis">>=
+<<12tablissman1, results = "asis">>=
 #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
@@
 %The distribution over time periods.
-<<41getske6089>>=
-vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
-vb6589$age <- factor(vb6589$age) # NOTE:: Now change to '6589'!
+<<12getske6089>>=
+cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
+cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
 @
@@
 
-<<41hazcalwold>>=
-resw <- getPHglm(vb6589[vb6589$sex == "female", ],
+<<12hazcalwold>>=
+resw <- getPHglm(cvb6589[cvb6589$sex == "female", ],
               nper, labb)
-resm <- getPHglm(vb6589[vb6589$sex == "male", ],
+resm <- getPHglm(cvb6589[cvb6589$sex == "male", ],
                 nper, labb)
@@
 
-<<41hazplotwold, fig.cap = cap2w, fig.scap = scap2w, fig.height = 7>>=
-##source("R/plotHaz.R")
-plotHaz.glm(resw, main = labb, layout = c(2, 2), col = scol)
-@
-
-<<41hazplotmold, fig.cap = cap2w, fig.scap = scap2m, fig.height = 7>>=
-plotHaz.glm(resm, main = labb, layout = c(2, 2), col = scol)
-@
-
-The cumulative hazards for the age group 65--89 in different decades are
-presented in Figures~\ref{fig:hazplotwold} and
-\ref{fig:hazplotmold}. Corresponding to what we find for the age group
-40--64, it is obvious that the assumption of proportional hazards is not
-reasonable.
-
 We will use as a summary measure the value of the cumulative hazards at 
-duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
+duration 25 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
 
@@
 survival among elderly, to a large extent caused by lower mortality in
-cardiovascular diseases. The results for the 1960s are however uncertain as
+cancer diseases. The results for the 1960s are however uncertain as
 discussed above, and should not be taken to seriously. The levels seem to
@@
 
-<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
 ##source("R/plotED.R")
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
-                main = "Women", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
-                main = "Men", exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "H(90|65)",  main = "Women", exclude = c(1))
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
+                    ylab = "H(90|65)", main = "Men", exclude = c(1))
 par(oldpar)
@@
 
-A blow-up of the last years is shown in Figure~\ref{fig:cumslogwm}.
-
-<<41cumslogwm, fig.cap = cap22log, fig.scap = scap22log, fig.height = 10>>=
-oldpar <- par(mfrow = c(2, 1))
-wloutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(90|65)",
-main = "Women")
-mloutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(90|65)",
-main = "Men")
-par(oldpar)
-@
-
-\subsubsection*{The effect of income and education, 1990--2005}
+\subsubsection*{The effect of income and education on cancer death, 1990--2005}
 
@@
 
-<<41lissyoung2>>=
+<<12lissyoung2>>=
 library(eha)
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
-<<41liswom2, results = 'asis'>>=
+The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2}
+<<12liswom2, results = 'asis'>>=
 fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
@@
 drw2 <- drop1(fitw2, test = "Chisq")
-xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:liswom")
+xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
 ##round(summary(fitw1)$coef[-(1:13), ], 4)
@@
 
-<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
+<<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
-<<41lismen2, results='asis'>>=
+<<12lismen2, results='asis'>>=
 fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
@@
 drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
+<<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
 #source("R/plotWEduc.R")
@@
 
+\section{Discussion}
+
+The authors of this paper have scrutinised different aspects of the social
+pattern of mortality in northern Sweden for quite some time, focusing on
+different regions, different age groups and partly different social
+categories. Already the earliest findings, concerning the social inequality
+in a rapidly growing town in an industrial district, surprised and at least
+at first confused the author \citep{edv92}. The expected social gradient in
+survival, meaning that people from higher social classes always had an
+advantage, was in many cases, particularly among adult men, not found
+despite the well-documented socio-economic differences in this
+society. There were however large differences between men and women and for
+different age groups. Further studies have confirmed these
+findings \citep{edli11, edbr12}. These earlier studies have ended with an
+assumption that more distinct differences to the advantage of higher
+classes developed somewhere during and probably quite early in the 20th
+century. Until now we have however not been able to investigate the
+continuous development on mortality differences among the adult and elderly
+population due to no available data. This is the first time it has been
+possible, and the results continue to surprise us. The consistent advantage
+for higher social classes do not appear until quite late in the regions we
+have analysed. The social pattern of female mortality are closer to what is
+expected, but a social gradient in male adult mortality seems to be a
+recent phenomenon.
+
+The perhaps surprising results makes it reasonable to consider their
+validity and possible weaknesses in the study. There are of course
+uncertainties involved in all studies of this sort, something that should
+not be ignored.  One obstacle concerns the quality of the available
+information in the sources. Occupational titles in historical sources are
+not always distinct enough to define the correct class. It can be difficult
+to identify the skill level in the working class population because many
+were often only designated as labourer. The size of the enterprise for free
+entrepreneurs is unknown and it may be impossible to distinguish if they
+are self-employed or have many employed. The class system and the position
+of different classes can also be discussed. Studies covering long period
+have the problem that occupations and classes change character - they do
+not represent the same thing. Nonetheless, these ambiguities and possible
+mis-classifications are not of that sort and magnitude that it would radically change
+the basic patterns observed. The social classes identified during the
+different periods represent distinct groups with very large differences in
+resources, capacities and power. A worker would not end up in the highest
+SOCPO level and large-scale entrepreneurs and higher officials would
+certainly be assigned the right level.Available data also show that the
+differences in income and wealth were large, often enormous.
+
+In this paper we have analysed social inequality in health from the
+perspective of social classes using HISCLASS. However, there are
+alternative ways to analyse social inequality, and we cannot exclude that
+such analyses gives other results or at least would modify the image
+presented here. There are other alternatives then HISCLASS, representing
+status or class dimensions. SOCPO is another system based on hierarchic
+categories of social power, which we have used in a couple of previous
+studies. In many modern studies categorisations based on income, wealth
+or education are common for studies of inequality. It is possible that
+such categorisations are better fit to distinguish between groups in
+society. They represent different dimensions of inequality, and we believe
+that the topic discussed here would gain from analyses of different aspects
+of inequality. One problem is however that reliable data on income and
+education are rarely available in historical population databases. We can
+however be certain that the highest HISCLASSES level have more education
+and the lowest the least. We also know from the 1930 census that the
+economic differences were very large at that time and that occupations in
+higher HISCLASSES tend to have higher incomes. The groups that may diverge
+are those within agriculture and those having a small enterprise where we
+can assume that the reported income figures underestimate their economic
+standard.
+
+The results are however in accordance with other studies \citep{bevp11},
+for example by \citet{bedr11} in a very different setting, a couple of
+Scanian parishes. They found the first indications of an emerging social
+gradient in the 1950s.The similarities makes us more comfortable about our
+results.
+
+\citet{who10} lists three different types of intermediary social
+determinants of health, material circumstances, behaviours and biological
+factors, and finally psycho-social factors. Most studies have focused on
+the material circumstances. Having a better economy can give advantages in
+many aspects of relevance for health and mortality, for example good living
+conditions, access to food, medical care, better conditions at work et
+cetera. Even though these advantages ought to have been at hand in our
+studied regions, the final result was not better health for men in the most
+privileged groups until the later part of the 20th century. We must
+consider other intermediary determinants. Psycho-social factors have been
+frequently discussed during the most recent decades, for example by
+\citet{mar04}. He argues that being higher in the status ladder leads to
+better health, regardless of where in the status ladder we look. This is
+also one of the arguments \citet{wipi09} use as an explanation why equal
+societies always perform better than unequal ones when it comes to health
+and other social conditions. In the case studied here, we do not find any
+indication of a status syndrome, since mortality levels was in the opposite
+directions to the one we would expect from these theories.
+
+This makes us consider the possible impact of behavioural factors.
+Our results of the gender aspect might be of interest in this case.
+When it comes to gender differences, the present study confirms what we have found on the Sundsvall region.
+There is a clear gendered pattern when it comes to social inequality in mortality.
+While high HISCLASS did not have any positive impact on survival for men in the early 20th century, we find  that higher social
+positions among women led to longer lives. Within the same households, the effect of social
+class seems to lead to different results. This opens up for behavioural
+aspects based on different gendered expectations. Alcohol has been
+suggested as part of the large gender differences in the 19th century
+\citep{wil99, edv92}, and a more risk-taking life may be a component of the
+male mortality.
+
+Having a high social status could be expressed in attitudes and behaviours
+that were harmful for health. Alcohol consumption decreased rapidly during
+the period late 19th century to the 1950s, thus possibly making room for an 
+increased role of economic resources. But the old pattern in social health
+inequalities remained. It is however possible that the relative differences
+in alcohol consumption prevailed, although at a lower level. The question 
+is however if the lower levels of alcohol consumption had any substantial
+negative impact on health. Swedes at that time had very low consumption
+levels. Alternative explanations should also be scrutinise. There 
+may have been strong and heavy requirements of being successful for the
+male \"elite. Many were involved in businesses with high risks. It would be
+very damaging to lose the high position and they may have felt that they 
+lived under constant threat of economic failure. There are several examples among
+the economic \"elite in the Sundsvall region that they had such problems
+\citep{edli11}. A possible future study involving analyses of death causes
+might throw some light on this issue.
+
+When it comes to the impact of social factors in the working age population versus the elderly, we do not find any fundamental differences.
+The social class have somewhat smaller effect among the elderly, so there is some difference in levels but not in character.
+
+Our study illustrates the importance of considering contextual factors and
+the restrictions of making general statements. We need to look into all
+aspects of social classes, not only the material ones but also what this
+means for behaviour, norms, way of living. \citet{edli11} suggested that an
+explanation for the larger social health inequalities in the present-day
+world is that status and higher social position is now expressed through
+health and being fit and physical active, while in many historical contexts
+higher social status was expressed by low physical activity, obsessive
+eating (being fat was impressive), drinking, smoking and sometimes also
+taking risks.
+
+Another question is if our results can be generalised. We have studied
+environments in northern Sweden without any large metropolis. Perhaps
+Sweden and/or this part of Sweden is a special case. The social differences
+may have been smaller, the involvement of government both at the national
+and local level may have been stronger, thus mitigating socio-economic
+disadvantages. These are relevant questions, which hopefully can be
+addressed in other and maybe comparative studies.
+
+Finally, we argue that our results have implications for our understanding
+on several hot research issues. A central question concerning the mortality
+transition is how the enormous increase in survival could be
+accomplished. This is still a hot debated question. \citet{mck76} argues that the
+improvements up to the middle of the 20th century was first of all made
+possible by improved nutrition. This seems to implicate that having more
+resources would lead to better survival. If research shows that the most
+advantaged groups did not have better health and longer lives, this do not
+necessarily speak against McKeown's ideas but it makes the question more
+complex. The results are also difficult to fit into \citet{liph95}
+hypothesis that social conditions should be seen as a fundamental cause of
+disease. Even if we would eliminate disadvantages such as bad living
+conditions, lower standard of medical care and other proximate determinants
+for the poor, the health disadvantage would remain. In the historical
+context we have analysed here, this does not get any support. This further
+illustrate the need for situating the analyses in their historical and
+geographical contexts.
+
+In recent years, the question about the relation between health and income
+inequality been widely discussed. Even though our study do not address this
+issue directly, our results are of interest for this question as well. The
+development of the social inequality in mortality do not seem to fit that
+well into the development of income and wealth inequalities. The early 20th
+century, in Sweden as well as in other countries, was characterised by very
+large economic differences, still this did in our regions not lead to any
+consistent advantage for the highest social classes. After the Second World
+War, economic differences diminished, and was at its lowest in Sweden at
+about 1980, i.e. at a time when a more expected pattern of a social
+hierarchy in survival had appeared. The period thereafter is characterised
+by increasing economic differences, and this may coincide with increasing
+health differences as well. The long-term relation between economic
+inequality and health is worth further studies.
+
+\section{Conclusion}
+
+We started this paper by raising the issue of the development of social
+inequalities. When studying this question in history we often get surprised
+of its complexity. In a society characterised by large economic inequality, 
+we would expect that this would result in corresponding mortality
+differences, i.e. higher social classes having an advantage. We find the
+appearance of a consistent social pattern in mortality, to the advantage of
+the higher classes, but surprisingly late in our history. It is only in the
+recent decades that this pattern dominates. Improved survival characterised
+however all HISCLASSes. We can however confirm that there are gender
+differences in the social pattern. While there is no clear social gradient
+among men, the results at least partly indicate advantages for the higher
+social classes for women already early in the 20th century. Females had
+more to gain from a high social position. Previous studies of the mortality
+transition have made similar observations. Among adults, gender has often
+been of more significant importance than social class. The reason for this
+difference and for the late appearance of the modern social pattern among
+men is still an enigma that require further studies, even though we have
+suggested the possible importance of behavioural factors and changes in
+ideals connected to different social classes. This should stimulate us to
+further disentangle the pathways for how social position influence health
+and mortality in different societies. Finally, we do not find any clear
+differences in the social pattern in the two studied age groups.
+
+\section*{Acknowledgements}
+
+This work was undertaken within the Centre for Demographic and Ageing
+Research (CEDAR) at Umeå University and with support from the Swedish
+Research Council (VR) (grant no. 421-2012-892). We thank Demographic Data
+Base for giving us access to data and Jonas Helgertz, Department of
+Economic History, Lund University, for providing us with scripts for
+creating HISCLASS from Swedish Census data.
+
 \bibliography{mybib}

### 2016-11-02 R/getCancer.R
fix NA

@@ -17,13 +17,13 @@ getCancer <- function(dat){
     
     ulor <- str_trim(dat$ULORSAK)
     ## First, fix ulor start with "C":
-    fc <- (substr(ulor, 1, 1) == "C") & (eb >= 1997) & (eb < 2014)
+    fc <- (!is.na(ulor)) &  (substr(ulor, 1, 1) == "C") & (eb >= 1997) & (eb < 2014)
     cancer <- fc
     ## Then the 'numeric' versions:
     ##ulor <- as.numeric(ulor)  # NA if letters in it
-    fc <- (ulor %in% as.character(140:239)) & (eb >= 1969) & (eb < 1997)
+    fc <- (!is.na(ulor)) & (ulor %in% as.character(140:239)) & (eb >= 1969) & (eb < 1997)
     cancer <- cancer | fc
-    fc <- (ulor == "12") & (eb < 1951)
+    fc <- (!is.na(ulor)) & (ulor == "12") & (eb < 1951)
     cancer <- cancer | fc
     dat$cancer <- dat$event & cancer
     dat

### 2016-11-02 R/getHeart.R
fix NA

@@ -15,13 +15,13 @@ getHeart <- function(dat){
     eb <- dat$exit + dat$birthdate
     library(stringr)
     ulor <- str_trim(dat$ULORSAK)
-    fc <- (substr(ulor, 1, 1) %in% c("I", "J")) & (eb >= 1997) & (eb < 2014)
+    fc <- (!is.na(ulor)) & (substr(ulor, 1, 1) %in% c("I", "J")) & (eb >= 1997) & (eb < 2014)
     heart <- fc
-    fc <- (ulor %in% as.character(390:459)) & (eb >= 1987) & (eb < 1997)
+    fc <- (!is.na(ulor)) & (ulor %in% as.character(390:459)) & (eb >= 1987) & (eb < 1997)
     heart <- heart | fc
-    fc <- (ulor %in% as.character(390:458)) & (eb >= 1969) & (eb < 1987)
+    fc <- (!is.na(ulor)) & (ulor %in% as.character(390:458)) & (eb >= 1969) & (eb < 1987)
     heart <- heart | fc
-    fc <- (ulor == "19") & (eb < 1951)
+    fc <- (!is.na(ulor)) & (ulor == "19") & (eb < 1951)
     heart <- heart | fc
     dat$heart <- dat$event & heart
     dat

### 2016-11-02 oldmort.Rnw
remove plots and add some new

@@
 cvb$event <- cvb$heart ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1961-1970", "1971-1980")), ]
+cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900", "1961-1970", "1971-1980")), ]
 cvb$period <- factor(cvb$period)
@@
 
+The quality of the information about causes of death vary over time as shown in 
+Figure~\ref{fig:qualheart}, where the proportion reported cardiovascular death of all deaths is shown.
+
+<<qualheart, fig.cap = "Proportion cardiovascular deaths of all deaths over time">>=
+x <- with(vb[vb$event == 1, ], table(period, heart))
+x <- prop.table(x, 1)
+x <- x[, 2]
+x <- round(100 * x, 2)
+barplot(x, col = "red", las = 2, ylab = "Per cent")
+@
+
 Models are fitted separately for each ten-year time period, from  1851 to
@@
 cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap22 <- "Total hazard standardized by HISCLASS, ages 65--89, cardiovascular death."    
+scap22 <- "TMR standardized by HISCLASS, ages 65--89, cardiovascular death."    
 cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women", exclude = c(1))
+                    ylab = "TMR(90|65)",  main = "Women")
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)", main = "Men", exclude = c(1))
+                    ylab = "TMR(90|65)", main = "Men")
 par(oldpar)
@@
 cvb$event <- cvb$cancer ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1961-1970", "1971-1980")), ]
+cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900", "1961-1970", "1971-1980", "1981-1990")), ]
 cvb$period <- factor(cvb$period)
@@
 
+The quality of the information about causes of death vary over time as shown in 
+Figure~\ref{fig:qualcancer}, where the proportion reported cancer deaths of all deaths is shown.
+
+<<qualcanncer, fig.cap = "Proportion cancer deaths of all deaths over time.">>=
+x <- with(vb[vb$event == 1, ], table(period, cancer))
+x <- prop.table(x, 1)
+x <- x[, 2]
+x <- round(100 * x, 2)
+barplot(x, col = "darkblue", las = 2, ylab = "Per cent")
+@
+
+We choose to use the time periods 1901--1950 and 1991--2013.
+
 Models are fitted separately for each ten-year time period, from  1851 to
@@
 cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cancer death."
-scap22 <- "Total hazard standardized by HISCLASS, ages 65--89, cancer death."    
+scap22 <- "TMR standardized by HISCLASS, ages 65--89, cancer death."    
 cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women", exclude = c(1))
+                    ylab = "H(90|65)",  main = "Women")
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)", main = "Men", exclude = c(1))
+                    ylab = "H(90|65)", main = "Men")
 par(oldpar)

### 2016-11-07 CreateData/R/putONssyk.R
modify yrken

@@ -40,14 +40,28 @@ putONssyk <- function(dat){
     who <- dat$AR == 1975 & dat$YRKE %in% yr75$old
     nyr <- dat$YRKE[who]
     indx <- match(nyr, yr75$old)
-    nyr <- yr65$new[indx]
+    nyr <- yr75$new[indx]
     dat$YRKE[who] <- nyr
     
-    ## 1980:
-    dat$YRKE[dat$AR == 1980] <- dat$YRKE80[dat$AR == 1980]
+    ## 1980: Vad ska vi ta här?? 'yr80.rda' finns inte!
+    ## och YRKE80 gäller ej 1980!!
+    ## Chansar på att 'yr75.rda' gäller:
+
+    ## 1980 (try):
+    load("../../../Data/yr75.rda")
+    who <- dat$AR == 1980 & dat$YRKE %in% yr75$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr75$old)
+    nyr <- yr75$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ##dat$YRKE[dat$AR == 1980] <- dat$YRKE80[dat$AR == 1980]
 
     ## 1985:
     dat$YRKE[dat$AR == 1985] <- dat$YRKE80[dat$AR == 1985]
+
+    ## 1990:
+    dat$YRKE[dat$AR == 1990] <- dat$YRKE85[dat$AR == 1990]
 
     is.na(dat$YRKE) <- dat$YRKE %in% c(0, 999)
     

### 2016-11-07 CreateData/R/ssyk3TOhisclass.R
modify yrken

@@ -12,19 +12,17 @@ ssyk3TOhisclass <- function(x){
     indx <- match(x, shh$ssyk)
     tmp <- shh$hisclass[indx]
 
-    ## Step 2: go from 12 to 7 levels:
+    ## Step 2: go from 12 to 4 levels: ## NEW!!
     
     res <- rep(-1, length(x))
     
     tmp[is.na(tmp)] <- -1
     
     res[tmp %in% 1:2] <- 1
-    res[tmp == 3] <- 2
-    res[tmp == 8] <- 3
-    res[tmp %in% 4:6] <- 4
-    res[tmp == 7] <- 5
-    res[tmp %in% c(9, 11)] <- 6
-    res[tmp %in% c(10, 12)] <- 7
+    res[tmp %in% c(3, 8)] <- 2
+    res[tmp %in% 4:6] <- 3
+    res[tmp %in% c(7, 9:12)] <- 4
+    
     is.na(res) <- res == -1
     
     res

### 2016-11-07 CreateData/R/ssykTOhisclass.R
modify yrken

@@ -7,7 +7,7 @@ ssykTOhisclass <- function(x){
     indx <- match(x, shh$ssyk)
     tmp <- shh$hisclass[indx]
 
-    ## Step 2: go from 12 to 7 levels:
+    ## Step 2: go from 12 to 4 levels:  ## NEW !!
     
     
     res <- rep(-1, length(x))
@@ -15,13 +15,9 @@ ssykTOhisclass <- function(x){
     tmp[is.na(tmp)] <- -1
     
     res[tmp %in% 1:2] <- 1
-    res[tmp == 3] <- 2
-    res[tmp == 8] <- 3
-    res[tmp %in% 4:6] <- 4
-    res[tmp == 7] <- 5
-##    res[tmp %in% c(9, 11)] <- 6
-##    res[tmp %in% c(10, 12)] <- 7 # skip klass 7!
-    res[tmp %in% 9:12] <- 6
+    res[tmp %in% c(3, 8)] <- 2
+    res[tmp %in% 4:6] <- 3
+    res[tmp %in% c(7, 9, 10, 11, 12)] <- 4
     is.na(res) <- res == -1
     
     res

### 2016-11-07 CreateData/creDat.Rnw
find 18000 deaths

@@
 \begin{abstract}
-hej
+This version is 'old', but all the code is included, so it will be crystal-clear to the reader exacttly what has been done.
+
+Spotted errors must pe reported immediately!
+
 \end{abstract}
@@
 <<setup, include = FALSE>>=
-knitr::opts_chunk$set(cache=FALSE,echo=FALSE,message=FALSE,warning=FALSE,comment=NA)##$
+knitr::opts_chunk$set(cache=FALSE,echo=TRUE,message=TRUE,warning=TRUE,comment=NA)##$
 @
@@
 \begin{enumerate}
-\item \'Elite, HISCLASS 1, 2, and 3, Higher and lower managers, and higher
+\item \'Elite, HISCLASS 1 and 2, Higher and lower managers, and higher
   professionals. 
-\item HISCALSS 4, 5, 6, and 8. Farmers, fishermen, and lower professionals, lower clerical, foremen.
-\item Medium skilled workers, HnISCLASS 7.
-\item Lower skilled workers, HISCLASS 9, 10, 11 and 12, Lower and unskilled workers (including farm workers)
+\item HISCLASS 3  and 8
+\item HISCLASS 4, 5, 6. Farmers, fishermen, and lower professionals, lower clerical, foremen.
+\item Workers, HISCLASS 7, 9, 10, 11, and 12.
 %%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
@@
 source("calWin.R")
+calWin
 library(xtable)
@@
 source("R/addPer.R")
+addPer
 adult <- addPer(adult, cuts = seq(1851, 1951, by = 25))
@@
 ## NEW hisclass, 6 + 7 --> 6:
-adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
+##adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
 ## Recode to four classes:
-hisc <- adult$hisclass
-    hisc[is.na(hisc)] <- -1
-    x <- hisc
-    hisc[x %in% 1:2] <- 1
-    hisc[x %in% 3:4] <- 2
-    hisc[x == 5] <- 3
-    hisc[x >= 5.9] <- 4
+##New Hisclasses:
+ett <- with(adult, !is.na(socBranch) &
+                ((socBranch == "official" & socStatus == "high") |
+                     (socBranch == "business" & socStatus == "high")
+                ))
+tva <- with(adult, !is.na(socBranch) &
+                ((socBranch == "farming" & socStatus == "high") |
+                     (socBranch == "business" & socStatus == "low")
+                ))
+tre <- with(adult, !is.na(socBranch) &
+                    (socBranch == "official" & socStatus == "low")
+               )
+fyra <- with(adult, !is.na(socBranch) &
+                ((socBranch == "farming" & socStatus == "low") |
+                     (socBranch == "worker"))
+                )
+
+hisc <- rep(-1, NROW(adult))
+hisc[ett] <- 1
+hisc[tva] <- 2
+hisc[tre] <- 3
+hisc[fyra] <- 4
 is.na(hisc) <- hisc == -1
 adult$hisclass <- hisc 
+# hisc <- adult$hisclass
+#     hisc[is.na(hisc)] <- -1
+#     x <- hisc
+#     hisc[x %in% 1:2] <- 1
+#     hisc[x %in% 3:4] <- 2
+#     hisc[x == 5] <- 3
+#     hisc[x >= 5.9] <- 4
+#     is.na(hisc) <- hisc == -1
+#     adult$hisclass <- hisc
 
 indx <- match(adult$id, per$id)
+
 adult$dodors <- per$dodors[indx]  ## New 29-10-2016..  and below ...
@@
                      "period", "urban", "civst", "hisclass", "dodors")]
+adult$dataset <- "poplink"
 aduu <- adult
@@
 <<ssykonfob>>=
-if (file.exists("fobb.rda")){
-    load("fobb.rda")
-}else{
-    load("../../../Data/fobdata.rda")
+    ##load("../../../Data/fobdata.rda")
+    load("data4/fobdata.rda")
     ## Cause of death:
@@
     source("R/putONssyk.R")
+    putONssyk
     fobb <- putONssyk(fobdata)
     source("R/ssykTOhisclass.R")
+    ssykTOhisclass
     fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
@@
     load("../../../Data/sex.rda")
+
     indx <- match(fobb$LINNEID, sex$LINNEID)
@@
     civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
-    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, anka
+    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, änka
     fobb$civst <- factor(civst, 
@@
     levels(fobb$sex) <- c("male", "female")
-    fobb <- fobb[, c("id", "birthdate", "sex", "enter", "exit", "event",
+    fobb <- fobb[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
                      "period", "urban", "civst", "hisclass", "ULORSAK")]
-    hisc <- fobb$hisclass
-    hisc[is.na(hisc)] <- -1
-    x <- hisc
-    hisc[x %in% 1:2] <- 1
-    hisc[x %in% 3:4] <- 2
-    hisc[x == 5] <- 3
-    hisc[x >= 5.9] <- 4
-    is.na(hisc) <- hisc == -1
-    fobb$hisclass <- hisc
+    fobb$dataset <- as.character(fobb$AR)
+    ##hisc <- fobb$hisclass
+    ##hisc[is.na(hisc)] <- -1
+    ##x <- hisc
+    ##hisc[x %in% 1:2] <- 1
+    ##hisc[x %in% 3:4] <- 2
+    ##hisc[x == 5] <- 3
+    ##hisc[x >= 5.9] <- 4
+    ##is.na(hisc) <- hisc == -1
+    ##fobb$hisclass <- hisc
     save(fobb, file = "fobb.rda")
-}
 @
@@
 <<lisadata>>=
-if (file.exists("liss.rda") & file.exists("lisaold.rda")){
-    load("liss.rda")
-    ##if (!is.null(liss$nyhisclass)){
-      ##  liss$hisclass <- liss$nyhisclass
-    ##    liss$nyhisclass <- NULL
-    ##}
-    load("lisaold.rda")
-}else{
-    load("../../../Data/lisadata.rda")
+
+    ##load("../../../Data/lisadata.rda")
+    load("data4/lisadata.rda")
     ## Cause of death:
@@
     lisaold <- lisadata
-    lisadata <- lisadata[lisadata$AR > 1999.5 & !is.na(lisadata$SSYK3), ]
+    ##lisadata <- lisadata[lisadata$AR > 1999.5 & !is.na(lisadata$SSYK3), ]
+    lisadata <- lisadata[!is.na(lisadata$SSYK3), ] ##NOTE!
     ## Fix HISCLASS:
     source("R/ssyk3TOhisclass.R")
+    ssyk3TOhisclass
     lisadata$hisclass <- ssyk3TOhisclass(lisadata$SSYK3)
@@
     liss <- addPer(liss, name = "period", cuts = seq(2002, 2014, by = 6))
-    liss <- liss[!is.na(liss$hisclass), ]
-    liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8), 
-                         labels = 1:4)
-    liss$hisclass <- as.numeric(as.character(liss$hisclass))
-    liss <- liss[with(liss, !is.na(civst) &
-                            !is.na(hisclass)), ]
-    lisaold <- lisaold[with(lisaold, !is.na(civst)), ]
+    ##liss <- liss[!is.na(liss$hisclass), ]
+    ##liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8), 
+      ##                   labels = 1:4)
+    ##liss$hisclass <- as.numeric(as.character(liss$hisclass))
+    ##liss <- liss[with(liss, !is.na(civst) &
+      ##                      !is.na(hisclass)), ]
+    ##lisaold <- lisaold[with(lisaold, !is.na(civst)), ]
     liss$civst <- factor(liss$civst, 
@@
     liss$id <- liss$LINNEID
-    liss <- liss[, c("id", "birthdate", "sex", "enter", "exit", "event",
+    liss <- liss[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
                      "period", "urban", "civst", "hisclass", "ULORSAK")]
+    liss$dataset <- paste("lisa", as.character(liss$AR), sep = "")
     save(liss, file = "liss.rda")
     save(lisaold, file = "lisaold.rda")
-}
+
 @
@@
 aduu$dodors <- NULL ## Yep...
-fobb$source <- "fob"
-liss$source <- "lisa"
-aduu$source <- "popum"
+##fobb$source <- "fob"
+##liss$source <- "lisa"
+##aduu$source <- "popum"
 foli <- rbind(fobb, liss)
@@
 aduu$id <- as.integer(as.factor(aduu$id))
+aduu$LINNEID <- NA
+foli$LINNEID <- foli$id ## Saved in time!!;)
 foli$id <- as.integer(as.factor(foli$id)) + max(aduu$id) + 1 #;)
@@
 y <- prop.table(x, margin = 1) * 100
-print(xtable(round(y, 1), digits = 0, caption = "Hisclass by time period, 1901--2013 (per cent)."))
+print(xtable(round(y, 1), digits = 0, 
+             caption = "Hisclass by time period, 1901--2013 (per cent)."))
 @
@@
 
+<<getageWin>>=
+source("ageWin.R")
+ageWin
+@
+
 <<groupbirths, results = 'asis'>>=
-source("ageWin.R")
 vb <- vb[vb$birthdate > 1761 & vb$birthdate < 1974, ]
@@
 <<ageint>>=
-vb <- vb[!is.na(vb$hisclass), ] # same as 'na.omit(vb)' (I think)
+##vb <- vb[!is.na(vb$hisclass), ] # same as 'na.omit(vb)' (I think)
+## Later...
 source("R/ageGroup.R")
+ageGroup
 vb <- ageGroup(vb, seq(40, 90, by = 5)) ## Oh, oh!
@@
 source("../R/getCancer.R")
+getCancer
 vb <- getCancer(vb)
 source("../R/getHeart.R")
+getHeart
 vb <- getHeart(vb)
 vb$exposure <- with(vb, exit - enter)
+vb0 <- vb
+vb <- vb[!is.na(vb$hisclass), ] # same as 'na.omit(vb)' (I think)
 first <- function(x) x[1]
@@
 vb$hisclass60[komp] <- vb$hisclass40[komp]
-vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+4", "5", "6+7"))
-vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+4", "5", "6+7"))
-
+vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
+vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
+vb$hisclass <- factor(vb$hisclass, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
 ## End cut ####################################################
-
+levels(vb$period)[8] <- "1991-2001" ## Changed from "1991-2000", fob90 extended.
+levels(vb0$period)[8] <- "1991-2001"
+save(vb0, file = "../data/vb0.rda")
 save(vb, file = "../data/vb.rda")

### 2016-11-07 CreateData/showData4.Rmd
find 18000 deaths

@@
 author: "Göran Broström"
-date: "September 22, 2016"
+date: "November 5, 2016"
 output: pdf_document
@@
 ```{r frominddata}
-load("data/inddata.rda")
-load("data/sex.rda")
+load("~/Forskning/Data/inddata.rda")
+load("~/Forskning/Data/sex.rda")
 indx <- match(inddata$LINNEID, sex$LINNEID)
@@
 ```{r styronper}
-load("data/socstyr.rda")
+load("~/Forskning/Data/socstyr.rda")
 indx <- match(per$LINNEID, socstyr$LINNEID)
@@
 per$ULORSAK <- socstyr$ULORSAK[indx]
+save(per, file = "data4/per.rda")
 ```
@@
 ```{r fob60}
-load("data/fob.rda")
+load("~/Forskning/Data/fob.rda")
 fob60 <- fob[fob$AR == 1960 & 
@@
 
+**Update 2016-11-05:** Draw it to December 31, 2001, i.e. add 14 months! Reason: 
+No hisclass in LISA before 1 Jan 2002!
+
 Persons in the fob and in our KOMMUNer.
@@
 fob90$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1960-11-01"))
+lim <- as.numeric(toTime("1970-11-01"))
 fob90 <- fob90[fob90$birthdate < lim, ]
@@
 dd <- per$deathdate[indx]
-fob90$event <- !(is.na(dd)) & (dd <= lim + 20)
+fob90$event <- !(is.na(dd)) & (dd <= lim + 31 + 61 / 365)
 ##fob90$event <- !(is.na(dd)) & (dd < 1987) 
-fob90$exit <- fob90$enter + 10  ## Not any longer: 1 year and 2 months.
+fob90$exit <- fob90$enter + 10 + 1 + 61 / 365 ## 10 + 1 + nov, dec: 1 year and 2 months.
 fob90$exit[fob90$event] <- dd[fob90$event] - fob90$birthdate[fob90$event]
@@
 ```{r lisadat}
-load("data/lisa.rda")
+load("~/Forskning/Data/lisa.rda")
 lisa <- lisa[lisa$KOMMUN %in% ourkom, ]

### 2016-11-07 R/getCancer.R
find more cancers

@@ -10,18 +10,22 @@ getCancer <- function(dat){
     ##dat$eb <- NULL
     ##dat
 #}
-## New version: 'ULORSAK' is a character vector:
+## New version: 'ULORSAK' is a factor (character vector):
     library(stringr)
     eb <- dat$exit + dat$birthdate
     ##cancer <- logical(NROW(dat))
     
-    ulor <- str_trim(dat$ULORSAK)
+    ulor <- str_trim(as.character(dat$ULORSAK))
     ## First, fix ulor start with "C":
     fc <- (!is.na(ulor)) &  (substr(ulor, 1, 1) == "C") & (eb >= 1997) & (eb < 2014)
     cancer <- fc
     ## Then the 'numeric' versions:
     ##ulor <- as.numeric(ulor)  # NA if letters in it
-    fc <- (!is.na(ulor)) & (ulor %in% as.character(140:239)) & (eb >= 1969) & (eb < 1997)
+    fc <- (!is.na(ulor)) & (substr(ulor, 1, 3) %in% as.character(140:239)) & 
+        (eb >= 1969) & (eb < 1997)
+    cancer <- cancer | fc
+    fc <- (!is.na(ulor)) & (substr(ulor, 1, 3) %in% as.character(140:239)) &
+        (eb >= 1960) & (eb < 1969)
     cancer <- cancer | fc
     fc <- (!is.na(ulor)) & (ulor == "12") & (eb < 1951)
     cancer <- cancer | fc

### 2016-11-07 R/getHeart.R
find more cardioascular deaths

@@ -14,12 +14,17 @@ getHeart <- function(dat){
     ## New version (ULORSAK is a character vector)
     eb <- dat$exit + dat$birthdate
     library(stringr)
-    ulor <- str_trim(dat$ULORSAK)
+    ulor <- str_trim(as.character(dat$ULORSAK))
     fc <- (!is.na(ulor)) & (substr(ulor, 1, 1) %in% c("I", "J")) & (eb >= 1997) & (eb < 2014)
     heart <- fc
-    fc <- (!is.na(ulor)) & (ulor %in% as.character(390:459)) & (eb >= 1987) & (eb < 1997)
+    fc <- (!is.na(ulor)) & (substr(ulor, 1, 3) %in% as.character(390:459)) & 
+        (eb >= 1987) & (eb < 1997)
     heart <- heart | fc
-    fc <- (!is.na(ulor)) & (ulor %in% as.character(390:458)) & (eb >= 1969) & (eb < 1987)
+    fc <- (!is.na(ulor)) & (substr(ulor, 1, 3) %in% as.character(390:458)) & 
+        (eb >= 1969) & (eb < 1987)
+    heart <- heart | fc
+    fc <- (!is.na(ulor)) & (substr(ulor, 1, 3) %in% as.character(400:468)) & 
+        (eb >= 1960) & (eb < 1969)
     heart <- heart | fc
     fc <- (!is.na(ulor)) & (ulor == "19") & (eb < 1951)
     heart <- heart | fc

### 2016-11-07 R/plotED.glm.R
more points included

@@ -12,7 +12,7 @@ plotED.glm <- function(res,
     ##nper <- NCOL(res[[1]])
 
     ## Assuming 'm == 4', we try:
-    leg = c("1+2", "3+4", "5", "6+7")
+    leg = c("1+2", "3+8", "4+5+6", "7+9+10+11+12")
     nper <- length(res)
     k <- NCOL(res[[1]])
     m <- NROW(res[[1]])
@@ -34,8 +34,8 @@ plotED.glm <- function(res,
     }
     fr <- min(outp, 0)
     if (relative){
-        to <- max(outp[, out])
-        fr <- min(outp[, out])
+        to <- max(outp[, out], na.rm = TRUE)
+        fr <- min(outp[, out], na.rm = TRUE)
         #cat("to = ", to, ", fr = ", fr, "\n")
     }else{    
         to <- max(outp)

### 2016-11-07 oldmort.Rnw
major update

--- 2016-11-02 oldmort.Rnw
@@
-\documentclass[a4paper,11pt,titlepage,twoside]{article}
+\documentclass[a4paper,11pt,titlepage]{article}
 
@@
 A long-term perspective on the development of social
-inequalities in mortality in the adult and elderly population is taken. 
+inequalities in mortality in the adult and elderly population is taken in
+this paper.  
 The area under investigation consists of the Skellefteå and Umeå regions in
-the north of Sweden, and the time period is 1851--2013.
-
-The main findings are that the social gradient becomes evident in the later
-alf of the 20th century, with lower classes having the highest mortality.
+the north of Sweden, and the time period is 1851--2013.The outcome is all
+cause mortality and cause specific mortality for the categories
+cardiovascular diseases and cancers. The analysis of inequality is based on
+Hisclass.
+
+The main findings are that the a consistent social gradient implying better
+survival for classes with more access to economic and other resources, only
+become evident in the later half of the 20th century, with lower classes
+having the highest mortality.The class patterns in mortality differed
+however depending on gender. Women of higher classes were comparatively
+favoured in most of the studied periods, while higher class men were
+disadvantaged when it comes to survival at least until the middle of the
+20th century. The modern pattern of clear socio-economic differences
+appeared surprisingly late. Additional analyses based on education and
+income categories confirmed that there is a large inequality in mortality
+between different groups during the latest decades. The analyses of death
+causes are more sensitive to changes in how they have been recorded, but
+both cardiovascular diseases and cancers conform with the pattern already
+observed when it comes to all-cause mortality.
 
@@
 vbTab$period <- factor(vbTab$period)
+levels(vbTab$period)[8] <- "1991-2001"
 load("data/vb.rda")
 vb$period <- factor(vb$period)
+levels(vb$period)[8] <- "1991-2001"
 @
@@
 
+For a long time, it was generally assumed that socio-economic inequality was high in historical societies, probably higher than in modern societies.
+This can be considered as a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
+Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
+When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
+Results from different studies have diverged, some reporting substantial differences but many have found surprisingly small socio-economic differences during the first stages of the transition.
+The appearance of the consistent pattern of inequality in survival that characterises present-day societies must have developed somewhere on the road from a high-mortality to a low-mortality regime.
+The complete process is however largely unknown.
+
 A long-term perspective on the development of social
-inequalities in mortality in the adult and elderly population is taken. 
+inequalities in mortality in the adult and elderly population is taken in the present paper. 
 The area under investigation consists of the Skellefteå and Umeå regions in
@@
 us to study the effect of income and education on mortality in a more
-direct way.
+direct way. Furthermore, we analyse inequality in mortality based on income and education respectively for the most recent periods, and compare it with the results of the class-based analysis.
 
@@
 differences using HISCLASS in mortality for the adult and elderly
-population during the mortality 
-transition and its relation to the development of income inequality. The
-analyses 
-will be performed on data from the Skellefteå and Umeå regions
-1901--2013. The questions are:
+population during the mortality transition. The
+analyses will be performed on data from the Skellefteå and Umeå regions
+1901--2013. The main questions are:
 
@@
 \item Is social position equally important among the retired population as for those in working age?
+\item Do we find the same pattern of inequality in survival as for all-cause mortality from cardiovscular diseases and cancers respectively?
+\item Are there large differences in survival depending on level of education and income categories?
 \end{enumerate}
@@
 
-Figure~\ref{fig:mapp} shows the study ares and thier position in Sweden.
+Figure~\ref{fig:mapp} shows the study areas and their position in Sweden.
 \begin{figure}
@@
 earlier period but with the addition of two more parishes. The majority
-of the population in the region lived in rural villages and hamlets, getting
+of the 19th century population lived in rural villages and hamlets, getting
 their livelihood from agricultural production. The region was vulnerable for
-harvest failures, one living long in memory was the famine of 1867–1868
-(Edvinsson and Broström, 2014). Towards the end of the century, the re-
-gion became connected to the Swedish railway system, thus improving the
-communications and contacts with other parts of the country. During the
+harvest failures, one living long in memory was the famine of 1867–-1868
+(Edvinsson and Broström, 2014).
+Towards the end of the century, the region became connected to the Swedish railway system, thus improving the communications and contacts with other parts of the country.
+During the
 20th century, industrialisation took place, partly by the opening of mines in
@@
 The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and
-rural parish, and in the Linneus database from 1960 onwards of Umeå mu-
-nicipality where another three parishes are included. This region had a
-somewhat different character from that of Skellefteå, even though agricul-
-ture was dominating the economy for a long time. The town of Umeå had a
-small population but substantially larger than Skellefteå town during almost
+rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
+This region had a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time.
+Umeå town had a
+small population but was substantially larger than Skellefteå town during almost
 the whole studied period. It was the administrative centre in the county of
-Västerbotten and schools and military regiments was placed here. Officials,
+Västerbotten and schools and military regiments was placed here. Consequently, Officials,
 teachers and the military were much more common in Umeå. Agriculture
-dominated in the rural part, while and there were some foundries, some
+dominated the rural part, there were some foundries, some
 industries for example in forestry and small-scale production. Umeå had a
@@
 Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure time over hisclass  for our selected time periods.
-
-<<hiscskum, fig.cap = "Exposure by hisclass (from the top: 1+2, 3+4, 5, 6+7) and time period", fig.height = 9>>=
+The category of workers (6+7) are fairly stable over time, while the lower white collar increases for a while, but diminishes somewhat during the later decades.
+The group of smaller entrepreneurs where farmers also are included, becomes smaller over time.
+This is to a large extent explained by the almost disappearance of farmers in recent time.
+The highest classes (1+2) has increased substantially, although from a very low level in the 19th century.
+The increase of academics in connection to the establishment of the university in Umeå explains much of this increase.
+
+<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, farmers and lower business, lower white collar, workers) and time period", fig.height = 9>>=
 xx <- with(vbTab, tapply(exposure, list(period = period, hisclass = hisclass40), sum))
@@
 
-For the corresponding numbers, with missing values ({\tt NA}) included, see Table~\ref{tab:na}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{r|rrrrr}
-  \hline
-Period & 1+2 & 3+4 & 5 & 6+7 & {\tt NA} \\ 
-  \hline
-1851-1875 & 2 & 32 & 4 & 49 & 14 \\ 
-  1876-1900 & 3 & 34 & 6 & 42 & 15 \\ 
-  1901-1925 & 5 & 34 & 10 & 33 & 18 \\ 
-  1926-1950 & 7 & 26 & 16 & 35 & 16 \\ 
-  1961-1970 & 4 & 18 & 7 & 20 & 50 \\ 
-  1971-1980 & 6 & 25 & 9 & 30 & 31 \\ 
-  1981-1990 & 8 & 28 & 8 & 30 & 26 \\ 
-  1991-2000 & 6 & 24 & 6 & 24 & 39 \\ 
-  2002-2007 & 21 & 35 & 9 & 35 & 0 \\ 
-  2008-2013 & 22 & 35 & 9 & 34 & 0 \\ 
-   \hline
-\end{tabular}
-\caption{Hisclass by time period, 1901--2013 (per cent). {\tt NA}: Missing information.} 
-\label{tab:na}
-\end{table}
+For the corresponding numbers, with missing values ({\tt NA}) included, see Table~\ref{tab:na}.The proportion with missing hisclass was comparatively stable until 1950, about every seventh lacked information allowing for determine class.
+This increased considerably for the beginning of period when the other dataset ---the Linnaeus database--- is used.
+This has to do with difficulties in the determination of class from the censuses, something we discuss below.
+For the most recent period, the information is complete.
+
+<<na, results = "asis">>=
+library(xtable)
+load("data/vb0.rda")
+vb0$hisclass <- factor(vb0$hisclass, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
+x <- with(vb0, table(period, hisclass, useNA = "ifany"))
+x <- round(prop.table(x, 1) * 100)
+xtable(x, legend = "Hisclass by time period, per cent.", label = "tab:na", digits = 0)
+@
 
@@
 
-The present state of knowledge about the development of social 
+According to the present state of knowledge about the development of social 
 health inequalities (in Sweden but also in other European countries), no
-clear social hierarchy has been found before 1900 but substantial
+consisten social hierarchy has been found before 1900 but substantial
 differences towards the late 20th and early 21st century. It appears as if
@@
 University. The study period from 1960 to 2000 are constructed from censuses every fifth
-year 1965--1990, with additional information on deaths from National Board of Health and Welfare.
+year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
 For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
@@
 
-The identification of presence periods as well as the time-dependent variables require some additional comments.
+\subsection{Presence periods}
+
+Differences in available information, both when it comes to structure and variables used for attributing social class, from the two datasets as well as
+within the Linnaeus dataset makes necessary diverging approaches when it comes to
+the identification of presence periods within our
+regions. The Poplink data provides us with exact dates or at least year of
+start and exit of the presence of all individuals in our data. This is however not the case
+with the Linnaeus database.
+For the period 1960--2000 we use the presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare for defining the observation periods as well as their social class (see below) in our analyses.
+We have however no exact information on what happened in between the censuses, for example the exact date of exit for persons that moved between two censuses.
+We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in this census.This overestimates their presence in our regions, but should not bias the results.
+We furthermore miss those that both moved in and moved out or vice versa between censuses.
+Exit dates are exact if they refer to death, and in these cases we consider them as constantly present from the last census to the death data even if the death took place in another part of Sweden.
+
+For the last period, 2002--2013, we use the yearly population registers (also known as LISA data from Statistics Sweden), together with information on deaths from National Board of Health and Welfare.
+This data allows a better control of the presence periods in the studied regions.
+
+\subsection{HISCLASS}
+
+In this paper we analyse mortality differences according to social class according to the classification scheme HISCLASS.
+The only type of socio-economic information available both in Linnaeus and Poplink is occupations, which can be used to identify distinct social classes.
+Thus, analysing according to social class is apart from a scientific choice, also a practical one.
+Using occupational data is the only and obvious alternative if we are to make a longitudinal study of socio-economic mortality differences with consistent classification schemes (although not without some problems).
+There are also other problems connected with using the same classification scheme covering a long time period, but this will be approached in more detail in the discussion section.
+Social position is treated as a time-dependent variable.
+
+The availability to information on occupations varies over time and between sources.
+The Poplink data usually provides information on the occupation of head of households.
+The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
+The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information.
+In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid.
+Thus, the social class in the census 1990 is considered to be kept until 2000.
+
+The use of different data sets, forces us to apply different approaches towards defining the social position.
+In the Poplink data, social class is based on the occupational titles given in the parish records.
+These can be more or less precise.
+There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
+We consider however their validity to be sufficient for identification of social position.
+Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}.
+In this study and according to the research questions in the present paper, we have chosen to categorize wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
+This strategy makes the problem of under-recording less problematic.
+
+Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers.
+Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database.
+Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period.
+It is also worth to notice that female labour force participation became the norm, and we could therefore argue that the occupation of the male as signifier of the social class became obsolete.
+We therefore let all included persons be signified by their individual occupation for this period.
+Still, we need to keep in mind that the results for the different periods are not completely comparable.
+
+Due to the construction of the time-dependent data, some records report unknown social position.
+This has been updated according to rules in order to optimize available information and minimize missing information.
+Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
+The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.
+
+The classification scheme used here is HISCLASS \citep{lema11}, a class-based system.
+The scheme uses HISCO as a basis for assigning categories.
+HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO.
+HISCO groups occupations according to tasks and skills.
+HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions and we have therefore used only the main categories, based on position on the labour market.
+All workers are therefore  This is particularly apparent when analysing the development in separate periods where for example farmers and farm workers become more and more rare during the studied century.
+We have therefore collapsed some categories, ending up with the following four:
+
+\begin{enumerate}
+\item \"elite, HISCLASS 1 and 2, Higher managers and higher professionals.
+\item Lower managers and farmers, HISCLASS 3 and 8.
+\item Lower professionals, HISCLASS 4, 5 and 6. Lower professionals, lower clerical, foremen.
+\item Workers, HISCLASS 7, 9, 10, 11 and 12, Lower and unskilled workers (including farm workers).
+%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
+\end{enumerate}
+
+For the Poplink data, we have used the script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
+The procedure for the occupations in the Linnaeus database is a little more complicated.
+The occupations in the censuses (NYK) must first be transferred to the system later used, SSYK (which is the Swedish version of ISCO88).
+Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013 can be recoded into HISCLASS.
+
+\subsection{Income}
+Income is based on disposable individual income from the LISA database at Statistics Sweden. We have categorized in four categories of the same size.
+
+\subsection{Education}
+Education is here analysed only for the period 1990 onwards.
+Years of education are given in 7 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
+It ranges from basic education (six year Folkskola) at level 1, to high academic degree (PhD and similar) at level 7.
+
+\subsection{Marital status}
+
+Marital status is treated as a time-dependent variable in the data file.
+The statuses used are unmarried, married and dissolution of marriage (without distinguishing between widowed and divorces).
+The status unmarried is often not explicitly stated, which makes the status unknown in the database Poplink for people for people moving to our parishes.
+If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
+
+\subsection{Cause of death}
+In the present study, we analyse mortality in the two most common disease groups in the ages we are investigating -- cardiovascular diseases and cancers.
+Compulsory reporting of cause of death became introduced in 1911 for all Sweden, and not only the urban part which had been the case before that.
+Classification systems have however changed.
+Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
+In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
+
+\subsection{Periods for analysis}
+We have made separate analyses for the different periods.
+For the century 1851-1950 we have divided it into four periods of 25 years each.
+The nineteenth century period represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
+During the years 1901-1950, industrialization became stronger and the urban environments increased in population size.
+The modern welfare state started to develop during this time.
+For the period 1980 onwards, the analyses is made on decades except for the two last periods.
+Sweden at this time is a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
+A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system, not being as generous as before.
+The start of the last period (2008--2013) coincides with the international financial crisis that also had impact on Sweden.
 
@@
 the development in more detail. We divide the analysis into two age groups,
-40--64 and 65--99 years of age. The model controls for period, marital
+40--64 and 65--89 years of age, the first age group representing people that were still mainly in the workforce, and the secon those that mainly were retired. The model controls for period, marital
 status and whether the individual resides in the urban or rural part. There
@@
 non-proportional, that is, it varies with age. Thus, the main results are
-presented \emph{graphically}. Note further that {\tt Hisclass} is a
-\emph{time-varying} variable. This causes lack of information in the higher
-ages around 1960--1980, because that kind of information is simply missing
+presented \emph{graphically}. This causes lack of information in the higher
+ages around 1960--1980 (see Table~\ref{tab:na}), because that kind of information is simply missing
 to a large extent.
@@
 From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
-the idea of proportional hazards is not reasonable.
+the idea of proportional hazards is not reasonable. We therefore use another approach.
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Women", exclude = 5:6)
+                main = "Women")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Men", exclude = 5:6)
+                main = "Men")##, exclude = 5:6)
 par(oldpar)
@@
 visible for later decades is partly explained by the decreasing general
-mortality. This effect can be eliminated by showing these numbers on a log
-scale, showing the relative levels. Presented in that way, it is not
+mortality. It is not
 obvious that differences between social classes have diminished. This
@@
 
-However, we are mainly interested in illustrating the change in relative position among the hisclasses over time rather than the ablolute levels, so we \emph{standardize} the levels in each time period to sume to unity.
+However, we are mainly interested in illustrating the change in relative position among the hisclasses over time rather than the absolute levels, so we \emph{standardize} the levels in each time period.
 It is illustrated in
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
-                    ylab = "TMR(65|40)", main = "Women", exclude = 5:6)
+                    ylab = "TMR(65|40)", main = "Women")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Men", exclude = 5:6)
+                    ylab = "TMR(65|40)", main = "Men")##, exclude = 5:6)
 par(oldpar)
@@
 %% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-\subsubsection*{The effect of income and education, 1990--2005}
-
-The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
-
-<<lissyoung>>=
-library(eha)
-library(xtable)
-load("data/loTab.rda")
-loTab$period <- factor(loTab$period)
-liss4064 <- loTab[loTab$age < 64.5, ]
-@
-
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
-<<liswom, results = 'asis'>>=
-fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-##cat("Women:\n")
-drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-
-<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
-source("R/plotWEduc.R")
-plotWEduc(fitw1)
-@
-
-And for men:
-
-<<lismen, results='asis'>>=
-fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-##cat("\nMen:\n")
-drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm1)
-@
-
-<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
-#oldpar <- par(mfrow = c(1, 2))
-#plot(fitw1, fn = "cum", col = dcol, main = "Women")
-#plot(fitm1, fn = "cum", col = dcol, main = "Men")
-#par(oldpar)
-##summary(fitw1)
-##summary(fitm1)
-@
-
-%See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<tablisswom1, results = "asis">>=
-#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
-#                 data = liss4064[liss4064$sex == "female", ])
-#dr <- drop1(fitwo1, test = "Chisq")
-#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.", 
-#    label = "tab:tablisswom1")
-@
-
-%See Table~\ref{tab:tablissman1} the PH regression for men.
-<<tablissman1, results = "asis">>=
-#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
-#                 data = liss4064[liss4064$sex == "male", ])
-#dr <- drop1(fitma1, test = "Chisq")
-#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.", 
- #   label = "tab:tablissman1")
-@
 
@@
 consequently having no corresponding population for that
-decade. Furthermore, the results for the 1960s should be considered with
-caution. The way the social class is defined leads to much missing
-information for that decade, something that have restricted the size of the
-population observed, making the results more shaky. The changing economy
+decade. Furthermore, we have not presented any results for the 1960s and 1970s due to much missing
+information on Hisclass for these decades, something that have restricted the size of the
+population observed. The changing economy
 also have consequences for the analysis, particularly when it comes to the
@@
 half of the 20th century, these occupations has constantly decreased in
-numbers thereafter.
+numbers thereafter. This means that the Hisclass group 2 has somewhat different character during the last periods in the figures.
 
@@
 
+\subsubsection*{The effect of income and education, 1990--2005}
+
+The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables and with controls of  urban versus rural residence, period and marital status.
+
+<<lissyoung>>=
+library(eha)
+library(xtable)
+load("data/loTab.rda")
+loTab$period <- factor(loTab$period)
+liss4064 <- loTab[loTab$age < 64.5, ]
+@
+
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+<<liswom, results = 'asis'>>=
+fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+##cat("Women:\n")
+drw <- drop1(fitw1, test = "Chisq")
+xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
+##round(summary(fitw1)$coef[-(1:13), ], 4)
+@
+
+The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}
+
+<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
+source("R/plotWEduc.R")
+plotWEduc(fitw1)
+@
+
+And for men:
+
+<<lismen, results='asis'>>=
+fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+##cat("\nMen:\n")
+drm <- drop1(fitm1, test = "Chisq")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
+##round(summary(fitm1)$coef[-(1:13), ], 4)
+dcol <- c("blue", "darkgreen", "darkgreen", "red")
+@
+
+<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitm1)
+@
+
+<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
+#oldpar <- par(mfrow = c(1, 2))
+#plot(fitw1, fn = "cum", col = dcol, main = "Women")
+#plot(fitm1, fn = "cum", col = dcol, main = "Men")
+#par(oldpar)
+##summary(fitw1)
+##summary(fitm1)
+@
+
+%See Table~\ref{tab:tablisswom1} the PH regression for women.
+<<tablisswom1, results = "asis">>=
+#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "female", ])
+#dr <- drop1(fitwo1, test = "Chisq")
+#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.", 
+#    label = "tab:tablisswom1")
+@
+
+%See Table~\ref{tab:tablissman1} the PH regression for men.
+<<tablissman1, results = "asis">>=
+#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA, 
+#                 data = liss4064[liss4064$sex == "male", ])
+#dr <- drop1(fitma1, test = "Chisq")
+#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.", 
+ #   label = "tab:tablissman1")
+@
+Both education and income were strongly significant, as well as marital status as expected.
+Urban or rural residence had however only weak effect.
+Those with the two lowest education level hade the highest mortality, while those with highest level of education had the best survival.
+The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
+The same can be said concerning income.
+More income results in lower mortality.
+This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden.
+Notice however that we cannot definitely establish a causal relationship between income and mortality in particular.
+Having low incomes may be caused by bad health.
+
 \subsection{Ages 65--89}
@@
 vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
-vb6589$age <- factor(vb6589$age) # NOTE:: No change to '6589'!
+vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
 vb6589$hisclass <- vb6589$hisclass60
@@
 We will use as a summary measure the value of the cumulative hazards at 
-duration 30 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
+duration 25 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Women", exclude = 5:6)
+                    ylab = "TMR(90|65)", main = "Women")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Men", exclude = 5:6)
+                    ylab = "TMR(90|65)", main = "Men")##, exclude = 5:6)
 par(oldpar)

### 2016-11-07 R/plotED.glm.R
change y-scale

@@ -34,8 +34,10 @@ plotED.glm <- function(res,
     }
     fr <- min(outp, 0)
     if (relative){
-        to <- max(outp[, out], na.rm = TRUE)
-        fr <- min(outp[, out], na.rm = TRUE)
+        ##to <- max(outp[, out], na.rm = TRUE)
+        ##fr <- min(outp[, out], na.rm = TRUE)
+        fr <- 0
+        to <- 1
         #cat("to = ", to, ", fr = ", fr, "\n")
     }else{    
         to <- max(outp)

### 2016-11-07 oldmort.Rnw
new plots

@@
 \begin{enumerate}
-\item \"elite, HISCLASS 1 and 2, Higher managers and higher professionals.
+\item \'Elite, HISCLASS 1 and 2, Higher managers and higher professionals.
 \item Lower managers and farmers, HISCLASS 3 and 8.
@@
 cvb$event <- cvb$heart ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1961-1970", "1971-1980")), ]
+cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
 cvb$period <- factor(cvb$period)
@@
 
-Models are fitted separately for each ten-year time period, from  1851 to
+Models are fitted separately for each time period, from  1901 to
 2013. We stratify on HISCLASS
@@
 cvb$event <- cvb$cancer ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1961-1970", "1971-1980")), ]
+cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
 cvb$period <- factor(cvb$period)

### 2016-11-08 CreateData/creDat.Rnw
cosmetics

@@
     fobb$id <- fobb$LINNEID
-    fobb <- addPer(fobb, cuts = seq(1961, 2001, by = 10))
+    fobb <- addPer(fobb, cuts = c(1961, 1971, 1981, 1991, 2002))
     load("../../../Data/sex.rda")
@@
 The {\sc Lisa} data is yearly, with {\tt hisclass} information starting in 2002.
-This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+4, and 5, 6.
+This information, however, do not allow us to contemplate seven (or six) levels, so we are using only four classes, grouping together 1+2, 3+8, 4+5+6, and 7+9+10+11+12.
 
@@
     lisaold <- lisadata
+    lisadata <- calWin(lisadata, c(2002, 2014))
+
     ##lisadata <- lisadata[lisadata$AR > 1999.5 & !is.na(lisadata$SSYK3), ]

### 2016-11-08 CreateData/showData4.Rmd
cosmetics

@@
 
-They need to be of the right age (born on November 1, 1930 or earlier). We
+They need to be of the right age (born on November 1, 1945 or earlier). We
 get the birthdates from *inddata*.
@@
 
-They need to be of the right age (born on November 1, 1935 or earlier). We
+They need to be of the right age (born on November 1, 1950 or earlier). We
 get the birthdates from *inddata*.
@@
 
-They need to be of the right age (born on November 1, 1940 or earlier). We
+They need to be of the right age (born on November 1, 1955 or earlier). We
 get the birthdates from *inddata*.
@@
 
-They need to be of the right age (born on November 1, 1945 or earlier). We
+They need to be of the right age (born on November 1, 1960 or earlier). We
 get the birthdates from *inddata*.
@@
 
-They need to be of the right age (born on November 1, 1950 or earlier). We
+They need to be of the right age (born on November 1, 1965 or earlier). We
 get the birthdates from *inddata*.
@@
 This is no longer the last FoB we are using: The follow-up time this time
-*is* ten years (as before), from November 1, 1990 to October 31, 2000, when
-the *lisa* data has been filling in for a long time.
+*is* ten years (as before), from November 1, 1990 to December 31, 2001, when
+the *lisa* data is filling in.
 
@@
 
-They need to be of the right age (born on November 1, 1960 or earlier). We
+They need to be of the right age (born on November 1, 1970 or earlier). We
 get the birthdates from *inddata*.

### 2016-11-08 R/getPHglm2.R
remove dependencies on civst and urban

@@ -13,10 +13,12 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
     levs <- matrix(0, nrow = m, ncol = k + 1)
    
     for (i in 1:nper){
-        fit <- glm(event ~ offset(log(exposure)) + civst + urban + age * hisclass, 
+        fit <- glm(event ~ offset(log(exposure)) + ##urban + ##civst 
+                       age * hisclass, 
                    data = dat[dat$period == labb[i], ], family = poisson)
         
         co <- fit$coefficients
+        ##if(any(is.na(co))) cat("NA in ", i, " period\n")
         ages <- grep("age", names(co))
         hisces <- grep("hisclass", names(co))
         interact <- ages[ages %in% hisces]
@@ -33,6 +35,7 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
         tmp[-1, -1] <- tmp[-1, -1] + inter
         tmp <- exp(tmp)
         levs <- cbind(rep(0, m), tmp)
+        levs[is.na(levs)] <- 0 ## HOPPSAN!!!
         for (j in 1:m){
             levs[j, ] <- cumsum(levs[j, ]) * ageInt
         }

### 2016-11-08 oldmort.Rnw
remove dependencies on civst and urban

@@
 <<setup, include = FALSE>>=
-knitr::opts_chunk$set(cache=FALSE,echo=FALSE,message=FALSE,warning=FALSE,comment=NA)##$
+knitr::opts_chunk$set(cache=FALSE,echo=FALSE,message=FALSE,warning=TRUE,comment=NA)##$
 @
@@
 
-For a long time, it was generally assumed that socio-economic inequality was high in historical societies, probably higher than in modern societies.
+For a long time, it was generally assumed that socio-economic health inequalities were high in historical societies, probably higher than in modern societies.
 This can be considered as a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
@@
 When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
-Results from different studies have diverged, some reporting substantial differences but many have found surprisingly small socio-economic differences during the first stages of the transition.
+Results from different studies have diverged, some reporting substantial differences but many have found surprisingly small socio-economic differences during the first stages of the transition \citep{bevp11, edli11,edbr12}.
 The appearance of the consistent pattern of inequality in survival that characterises present-day societies must have developed somewhere on the road from a high-mortality to a low-mortality regime.
@@
 \item Is social position equally important among the retired population as for those in working age?
-\item Do we find the same pattern of inequality in survival as for all-cause mortality from cardiovscular diseases and cancers respectively?
+\item Is there difference between classes in mortality from cardiovscular diseases and cancers respectively?
 \item Are there large differences in survival depending on level of education and income categories?
@@
 Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure time over hisclass  for our selected time periods.
-The category of workers (6+7) are fairly stable over time, while the lower white collar increases for a while, but diminishes somewhat during the later decades.
-The group of smaller entrepreneurs where farmers also are included, becomes smaller over time.
+The category of workers are fairly stable over time, while the lower white collar group was very small during the 19th century, started to increase during the early 20th century and increasing considerable during the period from the 1960s onwards.
+The group of smaller entrepreneurs where farmers also are included, was large until the middle of the 20th century, but has diminished considerably thereafter.
 This is to a large extent explained by the almost disappearance of farmers in recent time.
-The highest classes (1+2) has increased substantially, although from a very low level in the 19th century.
-The increase of academics in connection to the establishment of the university in Umeå explains much of this increase.
+The highest classes has increased substantially, although from a very low level in the 19th century.
+The increase of academics in connection to the establishment of Umeå university explains much of this increase.
 
@@
 
-For the corresponding numbers, with missing values ({\tt NA}) included, see Table~\ref{tab:na}.The proportion with missing hisclass was comparatively stable until 1950, about every seventh lacked information allowing for determine class.
-This increased considerably for the beginning of period when the other dataset ---the Linnaeus database--- is used.
+For the corresponding numbers, with missing values ({\tt NA}) included, see 
+Table~\ref{tab:nana}.The proportion with missing hisclass was very low, between 4 and 8 \% lacked information allowing for determine class.
+This increased considerably for the beginning of period when the other dataset ---the Linnaeus database--- is used, in particular for the 1960s where every second person lacks hisclass and the 1970s with almost a third have missing information.
 This has to do with difficulties in the determination of class from the censuses, something we discuss below.
@@
 
-<<na, results = "asis">>=
+<<nana, results = 'asis'>>=
 library(xtable)
@@
 x <- round(prop.table(x, 1) * 100)
-xtable(x, legend = "Hisclass by time period, per cent.", label = "tab:na", digits = 0)
+xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
 @
@@
 health inequalities (in Sweden but also in other European countries), no
-consisten social hierarchy has been found before 1900 but substantial
+consistent social hierarchy has been found before 1900 but substantial
 differences towards the late 20th and early 21st century. It appears as if
@@
 The availability to information on occupations varies over time and between sources.
-The Poplink data usually provides information on the occupation of head of households.
+The Poplink data usually provides information only on the occupation of head of households.
 The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
@@
 Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
-The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.
+The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.This, together with the circumstance that many women had no occupation in the census 1960 explans the large proportion where we have not been able to determine class during that decade.
 
@@
 HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions and we have therefore used only the main categories, based on position on the labour market.
-All workers are therefore  This is particularly apparent when analysing the development in separate periods where for example farmers and farm workers become more and more rare during the studied century.
 We have therefore collapsed some categories, ending up with the following four:
@@
 \begin{enumerate}
-\item \'Elite, HISCLASS 1 and 2, Higher managers and higher professionals.
+\item \"elite, HISCLASS 1 and 2. Higher managers and higher professionals.
 \item Lower managers and farmers, HISCLASS 3 and 8.
-\item Lower professionals, HISCLASS 4, 5 and 6. Lower professionals, lower clerical, foremen.
-\item Workers, HISCLASS 7, 9, 10, 11 and 12, Lower and unskilled workers (including farm workers).
+\item Lower white collar, HISCLASS 4, 5 and 6. Lower professionals, lower clerical, foremen.
+\item Workers, HISCLASS 7, 9, 10, 11 and 12. Workers (including farm workers).
 %%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
@@
 In the present study, we analyse mortality in the two most common disease groups in the ages we are investigating -- cardiovascular diseases and cancers.
-Compulsory reporting of cause of death became introduced in 1911 for all Sweden, and not only the urban part which had been the case before that.
+Compulsory reporting of cause of death became introduced in 1911 for all Sweden, and not only the urban part which had been the case from 1860.
 Classification systems have however changed.
@@
 The modern welfare state started to develop during this time.
-For the period 1980 onwards, the analyses is made on decades except for the two last periods.
+For the period 1960 onwards, the analyses is made on decades except for the two last periods.
 Sweden at this time is a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
 A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system, not being as generous as before.
-The start of the last period (2008--2013) coincides with the international financial crisis that also had impact on Sweden.
+The start of the last period (2008--2013) coincides with the international financial crisis that also had strong impact on Swedish economy.
 
@@
 presented \emph{graphically}. This causes lack of information in the higher
-ages around 1960--1980 (see Table~\ref{tab:na}), because that kind of information is simply missing
+ages around 1960--1980 (see Table~\ref{tab:nana}), because that kind of information is simply missing
 to a large extent.
@@
 nper <- length(labb)
-source("R/getPHglm.R")
+#source("R/getPHglm.R")
 ## End 'common'
@@
 
-We start by looking at cumulative hazards by {\sc Hisclass} over the first four time
+We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
 periods, in order to illustrate the process. They are adjusted for {\tt civil status}, and
@@
 
-Obviously, and not surprisingly, mortality is decreasing over time, and it is possible to see that the upper classes for men are moving from a top position in the nineteenth century to having the lowest mortality around the last turn of centuries.
-Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century.
+Figure~\ref{fig:cumsw} shows mortality levels during the different periods.
+Obviously, and not surprisingly, mortality is decreasing over time, and we can already identify some interesting social patterns in mortality which will be commented more on below.
 Apart from
@@
 It is illustrated in
-Figure~\ref{fig:cumswstand}.
+Figure~\ref{fig:cumswstand}. In the following, we only present these relative mortality levels for the different classes for the elderly and according to the two cause of death groups.
 
@@
 
-The central question in this paper is however what classes are most
-advantaged when it comes to mortality, and if the social pattern changed
-during the 20th century. Notice that we have not analysed the 1950s due to
+Notice that we have not analysed the 1950s due to
 the fact that we only have data for some parishes at that time,
 consequently having no corresponding population for that
-decade. Furthermore, we have not presented any results for the 1960s and 1970s due to much missing
-information on Hisclass for these decades, something that have restricted the size of the
-population observed. The changing economy
+decade.
+Furthermore, we need to be cautious when interpreting the results for the 1960s in particular, but also for the 1970s due to much missing information on Hisclass for these decades, something that have restricted the size of the population observed.
+The changing economy
 also have consequences for the analysis, particularly when it comes to the
@@
 
-The social pattern among women are mixed, but in accordance with previous
+The social pattern among women are quite much in accordance with previous
 analyses from the present research group on other regions in northern
-Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
-social classes usually have comparatively low mortality, while working
+Sweden and for other periods \citep{edli11,edbr12,segb16}. The highest
+social class usually have comparatively low mortality, while working
 class women have high. This is especially the case during the latest
-decades of the century and especially for the period 2002--2013.
+decades of the century and until 2013. We once again remind that the results from the 1960s must be interpreted with caution due to the large proportion of missing cases.
 
@@
 the lowest mortality, if we assume that access to different resources and
-having a high status determines survival, instead have the highest during
-much of the century. This is also in accordance with our previous studies
+having a high status determines survival, instead have the highest during the nineteenth and 
+much of the 20th century. This is also in accordance with our previous studies
 \citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
@@
 other countries. In the regions we are studying, this is however a quite
-recent phenomenon.The advantage for higher classes and disadvantage for low
-skilled workers is very clear for the period 2002--2013.
+recent phenomenon.
 
@@
 
-The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables and with controls of  urban versus rural residence, period and marital status.
+A common alternative to a class analysis is using income or education for investigating socio-economic differences.
+Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others disadvantaged.
+For the time being, we have to restrict such a comparative analysis to the last decades.
+
+The analysis for the period 19990--2005, is based on a separate analysis with income and education as explanatory variables and with controls of  urban versus rural residence, period and marital status.
 
@@
 Urban or rural residence had however only weak effect.
-Those with the two lowest education level hade the highest mortality, while those with highest level of education had the best survival.
+Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
 The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
@@
 We will use as a summary measure the value of the cumulative hazards at 
-duration 25 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
-
-The general pattern do not differ from the one of the younger age
-group. The levels are slightly more homogeneous, but we cannot establish
-that the elderly diverge to any degree from the working-age
-population. Mortality among women was, as expected, lower than among
+duration 25 (age 90), starting from age 65.
+
+The general pattern and development of mortality do not differ that much from the younger age group, of course representing a much higher mortality level.
+Mortality among women was, as expected, lower than among
 men. We also find a clear decline in mortality, but this decline did not
-set in until sometimes after the Second World War. This corresponds to the
+set in until sometimes after the Second World War (results not whown here). This corresponds to the
 Swedish development where we find a turning point at this time with better
@@
 cardiovascular diseases. The results for the 1960s are however uncertain as
-discussed above, and should not be taken to seriously. The levels seem to
-be to low. This is something we need to control further.
-
-Concerning the social pattern, we find results fairly similar to the ones
-for the age group 40--64. For women the results are mixed, even though
-women from the elite had comparatively low mortality and working class
-women high. For men we find the same intriguing change among elderly as for
+discussed above, especially for women due to no possibility to determine hisclass for the majority of them and therefore we have not performed any analysis on elderly women in that decade.
+
+Concerning the social pattern, the pattern is a little more difficult to interpret, compared to the younger age group (the new analyses differ somewhat from the ones in \citet{segb16}).
+For women the results are mixed.
+We do not find the same clear advantage for elite women during period until 1950, and neither do we find any large disadvantage for working class elderly women.
+In some periods, the social differences were small, and the elite even had substantially higher mortality during the early 20th century.
+Notice however that this group was rather small at this time, and there is a lot of random fluctuation.
+During the recent decades, the pattern is not very clear, but for the two last periods, the elite have the lowest mortality.
+It is however difficult to identify a clear social gradient in mortality.
+
+For men we find indications of the same intriguing change among elderly as for
 the adult population 40--64 years. Mortality was in fact substantially
-higher among men from the elite and the group of lower managers while it
-was lowest among workers. This did not change until the last decades of the
-20th century. At that time, the highest social classes instead had the best
-survival while the opposite was the case for workers. The pattern had
-reversed, thus confirming the results from most studies of this topic
-concerning the present-day society. The advantage for the higher classes
-and the disadvantage for the workers are apparent also for this age group
-during the early 21st century.
+higher among men from the elite  but the disadvantage is mainly apparent during the 19th century.
+Differences are fairly small during the first half of the 20th century.
+Only during the second half of the 20th century, we find periods with much better survival for the highest social class while retired workers had fairly high mortality.
+But even in recent times, the pattern was not that clear.
+During the last periods, we do not find a social gradient on hisclasse among elderly men (even in this case, the new analyses differ somewhat from the ones in \citet{segb16}).
+This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.
 
@@
 
-The analysis for the last period is based on a separate analysis with
-income and education as explanatory variables.
+The fact that we find rather unclear associations between hisclasses and survival, makes alternative approaches to the question of socio-economic health differences interesting.
+As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
 
@@
 
+These figures shows clear association both with income and with education and for women as well as for men.
+More years of education resulted in lower mortality.
+For women the association with additional levels became however smaller at the higher levels, the effect is more visible at the steps above the basic levels.
+In a similar fashion, increased incomes is associated with improved survival both for men and for women.
+Apparently, education and income is better to catch up socio-economic differences compared to hisclass, at least for this period.
+
 \section{Results, cardiovascular death}
 
+We now turn to a special analysis of mortality for the two dominating causes of death among the adult and elderly population.
+We start with cardiovascular diseases.
+Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group.
+Belonging to working class, having less education and low income increases the mortality risks.
+We would therefore assume that we find a clear social pattern in the present analysis.
+
+First, a general observation is that the analysis of causes of death and their development are more difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ.
+Thus, many deaths during the 19th century were not diagnosed at all, and there was still a substantial part of all death in the early twentieth century where death cause is designated as unknown or dying from old age, despite reporting causes of death had become mandatory.
+Furthermore, we do not present results for the 1960s and 1970s due to much missing data on hisclass, especially for the elderly and for women during these decades.
 <<causes41>>=
@@
 nper <- length(labb)
-source("R/getPHglm.R")
+##source("R/getPHglm.R")
 ## End 'common'
@@
 
-Models are fitted separately for each time period, from  1901 to
+Models are fitted separately for each ten-year time period, from  1851 to
 2013. We stratify on HISCLASS
@@
 periods. They are adjusted for {\tt civil status}, and
-{\tt urban/rural} residence.
-We present separate analyses  for women and men.
-The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.
+{\tt urban/rural} residence. We present separate analyses for women and men.
 
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                     ylab = "TMR(65|40)",  main = "Women")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, 
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                     ylab = "TMR(65|40)", main = "Men")
@@
 
-Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century. Apart from
-the social pattern that will be commented on below, there are a couple of
-observations that can be made from the figure. Women had consistently
-better survival than men. This is a well-known phenomenon in Swedish
-demographic history. Male mortality has almost always been higher than
-female in all age groups \citep{wil99, suwi07}. We also observe a
-substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. The compression that is
-visible for later decades is partly explained by the decreasing general
-mortality. This effect can be eliminated by showing these numbers on a log
-scale, showing the relative levels. Presented in that way, it is not
-obvious that differences between social classes have diminished. This
-conclusion has been confirmed by separate analyses (not shown). There is no
-clear indication of homogenisation of mortality levels between social
-classes.
+Figure~\ref{fig:cumsw} shows mortality levels during different periods from 1851 onwards.
+A couple of observations can be made.
+Women from the highest social class had almost consistently lower mortality from cardiovascular diseases compared to others.This is apparent even for the periods with deficient cause of death data.
+The pattern concerning the other groups are less clear, but if we compare the two largest groups, the workers and the small entrepreneur/farmer group, the differences were rather small during most of the time.
+A new pattern has however developed from 1990 onwards.
+Now working class women have substantially higher cardiovascular compared to others.
+
+For men, we find a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality.
+The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
+Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the other large group, small-scale entrepreneurs/farmers.
+And in the same way as for women at this age, working class me have by far the highest mortality from this cause of death during the last decades.
 
@@
 
-The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
+We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death.
+It is based on a separate analysis with income and education as explanatory variables.
 
@@
 drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 
-The central question in this paper is however what classes are most
-advantaged when it comes to mortality, and if the social pattern changed
-during the 20th century. Notice that we have not analysed the 1950s due to
-the fact that we only have data for some parishes at that time,
-consequently having no corresponding population for that
-decade. Furthermore, the results for the 1960s should be considered with
-caution. The way the social class is defined leads to much missing
-information for that decade, something that have restricted the size of the
-population observed, making the results more shaky. The changing economy
-also have consequences for the analysis, particularly when it comes to the
-agricultural sector. While farmers were common occupations during the first
-half of the 20th century, these occupations has constantly decreased in
-numbers thereafter.
-
-The social pattern among women are mixed, but in accordance with previous
-analyses from the present research group on other regions in northern
-Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
-social classes usually have comparatively low mortality, while working
-class women have high. This is especially the case during the latest
-decades of the century and especially for the period 2002--2013.
-
-A different pattern appears among men. The groups we would expect to have
-the lowest mortality, if we assume that access to different resources and
-having a high status determines survival, instead have the highest during
-much of the century. This is also in accordance with our previous studies
-\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
-low mortality. This changes however towards the end the 20th century. From
-the 1980s, there is a clear advantage when it comes to survival for higher
-managers and professionals. This is also what we usually find in analyses
-of social health inequalities in present-day Sweden as well as in most
-other countries. In the regions we are studying, this is however a quite
-recent phenomenon.The advantage for higher classes and disadvantage for low
-skilled workers is very clear for the period 2002--2013.
+The analysis of cardiovascular mortality accordning to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
+More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system.
 
@@
 
-We now turn to the elderly, i.e. those that mostly have left the workforce. We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
-{\tt urban/rural} residence and with separate analyses for women and men.
-We do not present any results concerning possible homogenisation of mortality levels, since the main results do not deviate from that of the younger group.
+In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years.
+In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
+{\tt urban/rural} residence and with separate analyses for women and men. We use as a summary measure the value of the cumulative hazards at 
+duration 25 (age 90), starting from age 65.
 
@@
 <<41hazcalwold>>=
-resw <- getPHglm(cvb6589[cvb6589$sex == "female", ],
+resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
               nper, labb)
-resm <- getPHglm(cvb6589[cvb6589$sex == "male", ],
+resm <- getPHglm2(cvb6589[cvb6589$sex == "male", ],
                 nper, labb)
@@
 
-We will use as a summary measure the value of the cumulative hazards at 
-duration 25 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
-
-The general pattern do not differ from the one of the younger age
-group. The levels are slightly more homogeneous, but we cannot establish
-that the elderly diverge to any degree from the working-age
-population. Mortality among women was, as expected, lower than among
-men. We also find a clear decline in mortality, but this decline did not
-set in until sometimes after the Second World War. This corresponds to the
-Swedish development where we find a turning point at this time with better
-survival among elderly, to a large extent caused by lower mortality in
-cardiovascular diseases. The results for the 1960s are however uncertain as
-discussed above, and should not be taken to seriously. The levels seem to
-be to low. This is something we need to control further.
-
-Concerning the social pattern, we find results fairly similar to the ones
-for the age group 40--64. For women the results are mixed, even though
-women from the elite had comparatively low mortality and working class
-women high. For men we find the same intriguing change among elderly as for
-the adult population 40--64 years. Mortality was in fact substantially
-higher among men from the elite and the group of lower managers while it
-was lowest among workers. This did not change until the last decades of the
-20th century. At that time, the highest social classes instead had the best
-survival while the opposite was the case for workers. The pattern had
-reversed, thus confirming the results from most studies of this topic
-concerning the present-day society. The advantage for the higher classes
-and the disadvantage for the workers are apparent also for this age group
-during the early 21st century.
+Concerning the social pattern, we find rather mixed results but with some similarities to the ones
+for the age group 40--64.
+The results for the 19th century and the first 20th century period are very shaky and will not be commented on.
+For women mortality in this disease group were much higher among the elite than the others  until only the last decades.
+Differences between the other classes were small, and elderly from a working class background are not disadvantaged.
+The social class pattern has changed during the last periods, but we do not find a consistent class pattern in mortality.
+Also for men, we find high cardiovascular mortality during the period 1926--1950 and the 1980s for the elite, but also for small-scale entrepreneurs/farmers during the first half of the 20th century.
+As for women, we find no clear class pattern during the last periods.The diffuse pattern of class differences in cardiovascular mortality is surprising.
+It is therefore of interest to find out what other aspects of social differentiation tell us.
 
@@
 
-The analysis for the last period is based on a separate analysis with
+The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
 income and education as explanatory variables.
+The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2} for women and Table~\ref{tab:41lismen2} for men.
+It confirms that most of the variables are strongly significant except
 
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2}
 <<41liswom2, results = 'asis'>>=
@@
 
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-
-<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitw2)
-@
-
-And for men:
-
 <<41lismen2, results='asis'>>=
@@
 drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 @
+
+The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+
+<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitw2)
+@
+
+And for men:
 
@@
 
+The pattern is similar to what we find for all-cause mortality.
+The higher educational level, the lower cardiovascular mortality, and more income leads to better survival.
+This association is clear and consistent for women as well as for men.
+A minor difference between men and women is that mortality in the higher levels of education do not lead to lower mortality among women while that is the case among men.
+
 \section{Results, cancer death}
+
+While we assume clear social differences in cardiovascular diseases, this is not necessarily an obvious assumption when it comes to cancers.
+Different forms of cancers have very different origin and different pathways to disease and death.
+The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers.
+In the present analysis we include cancer deaths 1851--2013, but with an unregistered and not analyzed period between 1951--1980.
+Once again we declare that the 19th century information on cause of death is very uncertain and under-registered, and that this problem partly remains into the early 20th century.
 
@@
 
-Models are fitted separately for each ten-year time period, from  1851 to
+Models are fitted separately for each period, from  1851 to
 2013. We stratify on HISCLASS
@@
 (Skellefteå town or outside). The cumulative
-hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
+hazards at the end of the period (65 if 40--64, 90 if 65--89) are the
 main target in our analysis: These numbers are used in the process of
@@
 
-We will use as a summary measure the value of the cumulative hazards at
-age 65, given survival to age 40.
+%%We will use as a summary measure the value of the cumulative hazards at
+%%age 65, given survival to age 40.
 
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2, 
                     ylab = "TMR(65|40)",  main = "Women")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, 
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
                     ylab = "TMR(65|40)", main = "Men")
@@
 
-Figure~\ref{fig:cumsw} shows decadal mortality levels during the 19th century. Apart from
-the social pattern that will be commented on below, there are a couple of
-observations that can be made from the figure. Women had consistently
-better survival than men. This is a well-known phenomenon in Swedish
-demographic history. Male mortality has almost always been higher than
-female in all age groups \citep{wil99, suwi07}. We also observe a
-substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. The compression that is
-visible for later decades is partly explained by the decreasing general
-mortality. This effect can be eliminated by showing these numbers on a log
-scale, showing the relative levels. Presented in that way, it is not
-obvious that differences between social classes have diminished. This
-conclusion has been confirmed by separate analyses (not shown). There is no
-clear indication of homogenisation of mortality levels between social
-classes.
+Figure~\ref{fig:12cumsw} shows mortality levels according to hisclass during the different studied periods.
+The social differences were small until the middle of the 20th century, for women as well as for men.
+We disregard the obviously uncertain results for men during the 19th century.
+For the periods after 1980, the rather small differences remain.
+the results for the 1980s need however to be more thoroughly checked up.
+There is an indication of the elite having lower cancer mortality for both sexes during the last periods, but these differences are small and we do not find any clear social gradient in the risks.
+Thus, for this age group cancer mortality is not something that distinguishes between classes.
 
@@
 
-The analysis for the last period,  is based on a separate analysis with income and education as explanatory variables.
+In the previous section, we concluded that there were no social inequality in cancer mortality according to hisclass.
+The question here is if the lacking socio-economic differentiation holds also when using education and income as categorizing variables.
+The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables.
 
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:12liswom}
+The statistical significance of the covariates are shown in Table~\ref{tab:12liswom} for women and Table~\ref{tab:12lismen} for men. The variables of interest here are all significant.
 <<12liswom, results = 'asis'>>=
@@
 
-The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}
-
-<<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-source("R/plotWEduc.R")
-plotWEduc(fitw1)
-@
-
-And for men:
-
 <<12lismen, results='asis'>>=
@@
 drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.")
+xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.", label = "tab:12lismen")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 @
+
+The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}.
+
+<<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
+source("R/plotWEduc.R")
+plotWEduc(fitw1)
+@
+
+And for men:
 
@@
 
-The central question in this paper is however what classes are most
-advantaged when it comes to mortality, and if the social pattern changed
-during the 20th century. Notice that we have not analysed the 1950s due to
-the fact that we only have data for some parishes at that time,
-consequently having no corresponding population for that
-decade. Furthermore, the results for the 1960s should be considered with
-caution. The way the social class is defined leads to much missing
-information for that decade, something that have restricted the size of the
-population observed, making the results more shaky. The changing economy
-also have consequences for the analysis, particularly when it comes to the
-agricultural sector. While farmers were common occupations during the first
-half of the 20th century, these occupations has constantly decreased in
-numbers thereafter.
-
-The social pattern among women are mixed, but in accordance with previous
-analyses from the present research group on other regions in northern
-Sweden and for other periods \citep{edli11,edbr12,segb16} the highest
-social classes usually have comparatively low mortality, while working
-class women have high. This is especially the case during the latest
-decades of the century and especially for the period 2002--2013.
-
-A different pattern appears among men. The groups we would expect to have
-the lowest mortality, if we assume that access to different resources and
-having a high status determines survival, instead have the highest during
-much of the century. This is also in accordance with our previous studies
-\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
-low mortality. This changes however towards the end the 20th century. From
-the 1980s, there is a clear advantage when it comes to survival for higher
-managers and professionals. This is also what we usually find in analyses
-of social health inequalities in present-day Sweden as well as in most
-other countries. In the regions we are studying, this is however a quite
-recent phenomenon.The advantage for higher classes and disadvantage for low
-skilled workers is very clear for the period 2002--2013.
+When it comes to the association with education and income, we find the similar pattern as those for all-cause mortality and cardiovascular mortality.
+There is a strong effect for both men and women of having more education and income.
+These variables seem to catch up important aspect of socio-economic differences in health and mortality.
 
@@
 
-We now turn to the elderly, i.e. those that mostly have left the workforce. We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
-{\tt urban/rural} residence and with separate analyses for women and men.
-We do not present any results concerning possible homogenisation of mortality levels, since the main results do not deviate from that of the younger group.
+We finally turn to cancer mortality among the elderly.
+In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and {\tt urban/rural} residence and with separate analyses for women and men.
 
@@
 <<12hazcalwold>>=
-resw <- getPHglm(cvb6589[cvb6589$sex == "female", ],
+resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
               nper, labb)
-resm <- getPHglm(cvb6589[cvb6589$sex == "male", ],
+resm <- getPHglm2(cvb6589[cvb6589$sex == "male", ],
                 nper, labb)
@@
 We will use as a summary measure the value of the cumulative hazards at 
-duration 25 (age 90), starting from age 65. For the period 2002--2013, a separate analysis is presented.
-
-The general pattern do not differ from the one of the younger age
-group. The levels are slightly more homogeneous, but we cannot establish
-that the elderly diverge to any degree from the working-age
-population. Mortality among women was, as expected, lower than among
-men. We also find a clear decline in mortality, but this decline did not
-set in until sometimes after the Second World War. This corresponds to the
-Swedish development where we find a turning point at this time with better
-survival among elderly, to a large extent caused by lower mortality in
-cancer diseases. The results for the 1960s are however uncertain as
-discussed above, and should not be taken to seriously. The levels seem to
-be to low. This is something we need to control further.
-
-Concerning the social pattern, we find results fairly similar to the ones
-for the age group 40--64. For women the results are mixed, even though
-women from the elite had comparatively low mortality and working class
-women high. For men we find the same intriguing change among elderly as for
-the adult population 40--64 years. Mortality was in fact substantially
-higher among men from the elite and the group of lower managers while it
-was lowest among workers. This did not change until the last decades of the
-20th century. At that time, the highest social classes instead had the best
-survival while the opposite was the case for workers. The pattern had
-reversed, thus confirming the results from most studies of this topic
-concerning the present-day society. The advantage for the higher classes
-and the disadvantage for the workers are apparent also for this age group
-during the early 21st century.
+duration 25 (age 90), starting from age 65.
+
+The relation between different classes and cancer mortality fluctuates considerable both for women and men in this age group.
+For some periods it has to do with quite few events to analyse, particularly fo the 1960s where hisclass has been difficult to determine for the elderly.
+Women of higher classes sometimes have favorable mortality, but unfavorable at other times.
+It seems however that the higher classes have reached lower cancer levels compared to the working class from 2001 onwards.
+Men from the elite had higher mortality until the 1970s, but a social gradient has appeared thereafter.
+Now the elite with assumed more resources have lower mortality than other groups.
+Elderly of workers origin on the other side have higher cancer mortality during the last decades, sometimes substantially higher than the others.The pattern is however not completely clear.
 
@@
 
-The analysis for the last period is based on a separate analysis with
+Thus, we find indications of social differences in the later decades, particularly between the elite and the workers. The alternative analysis for the last period is based on a separate analysis with
 income and education as explanatory variables.
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2}
+The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men. 
 <<12liswom2, results = 'asis'>>=
@@
 
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-
-<<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitw2)
-@
-
-And for men:
-
 <<12lismen2, results='asis'>>=
@@
 drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
+xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
 ##round(summary(fitm1)$coef[-(1:13), ], 4)
@@
 @
+
+The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
+
+<<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
+#source("R/plotWEduc.R")
+plotWEduc(fitw2)
+@
+
+And for men:
+
+%% <<12lismen2, results='asis'>>=
+%% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
+%% ##cat("\nMen:\n")
+%% drm2 <- drop1(fitm2, test = "Chisq")
+%% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
+%% round(summary(fitm1)$coef[-(1:13), ], 4)
+%% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+%% @
 
@@
 @
+
+The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income.
+We find the same consistent pattern as we have found for all-cause mortality and cardiovascular mortality for this period.
+Higher levels of education and more income is associated with lower cancer mortality.
+We can therefore conclude that both education and income are strongly decisive aspects of social differentiation when it comes to mortality during the last decades.
 

### 2016-11-08 R/plotWEduc.R
in again

@@ -6,14 +6,19 @@ plotWEduc <- function(fit){
     educ <- co[substr(names(co), 1, 4) == "educ"]
     income <- co[substr(names(co), 1, 4) == "inco"]
 
-    educ <- exp(c(con, educ + con))
+    educ <- exp(c(0, educ))
+    ##educ <- exp(c(con, educ + con))
     names(educ) <- 1:7
-    income <- exp(c(con, income + con))
+    income <- exp(c(0, income))
+    ##income <- exp(c(con, income + con))
     names(income) = c("lowest", "low", "high", "highest")
+    base <- co[1:5]
+    base <- c(co[1], co[1] + co[2:5])
+    mort <- 5 * sum(exp(base))
     oldpar <- par(mfrow = c(1, 2))
-    barplot(100 * educ, xlab = "Education", col = "green", las = 1, 
+    barplot(100 * (1 - exp(-educ * mort)), xlab = "Education", col = "green", las = 1, 
             ylab = "Mortality per cent")
-    barplot(100 * income, xlab = "Income", col = "blue", las = 2,
+    barplot(100 * (1 - exp(-income * mort)), xlab = "Income", col = "blue", las = 2,
             ylab = "Mortality per cent")
     par(oldpar)
     invisible(list(educ = educ, income = income))

### 2016-11-08 oldmort.Rnw
in again

@@
 
-<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, farmers and lower business, lower white collar, workers) and time period", fig.height = 9>>=
+<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, farmers and lower business, lower white collar, workers) and time period", fig.height = 4.5>>=
 xx <- with(vbTab, tapply(exposure, list(period = period, hisclass = hisclass40), sum))
 yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2)
+barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
 @
@@
 xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
+@
+
+Regarding the distribution of civil status over time, see Figure~\ref{fig:civstskum}.
+
+<<civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5>>=
+xx <- with(vbTab, tapply(exposure, list(period = period, civst = civst), sum))
+yy <- round(prop.table(xx, 1) * 100, 1)
+barplot(t(yy)[3:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
+@
+
+The distribution of causes of death is shown in Figure~\ref{fig:causeskum}.
+
+<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5>>=
+vb01 <- vb0[vb0$event == 1, ]
+cancer <- tapply(vb01$cancer, vb01$period, sum)
+heart <- tapply(vb01$heart, vb01$period, sum)
+death <- tapply(vb01$period, vb01$period, length)
+other <- death - cancer - heart
+ttt <- rbind(cancer, heart, other)
+ttt <- prop.table(ttt, 2) * 100
+barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = )
+@
+
+Causes of death before 1951.
+
+<<causeof51>>=
+vb051 <- vb0[vb0$period %in% c("1901-1925", "1926-1950"), ]
+vb051$period <- factor(vb051$period)
+with(vb051[vb051$event == 1, ], table(ULORSAK, period, useNA = "ifany"))
 @
@@
 
-The analysis for the period 19990--2005, is based on a separate analysis with income and education as explanatory variables and with controls of  urban versus rural residence, period and marital status.
+The analysis for the period 19990--2005 is based on a separate analysis with income and education as explanatory variables and with controls of  urban versus rural residence, period and marital status.
 
@@
 
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}
+The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
 <<liswom, results = 'asis'>>=
@@
 
-The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}
+The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
 

### 2016-11-08 oldmort.Rnw
soren's latest

@@
 become evident in the later half of the 20th century, with lower classes
-having the highest mortality.The class patterns in mortality differed
+having the highest mortality. The class patterns in mortality differed
 however depending on gender. Women of higher classes were comparatively
@@
 appeared surprisingly late. Additional analyses based on education and
-income categories confirmed that there is a large inequality in mortality
-between different groups during the latest decades. The analyses of death
-causes are more sensitive to changes in how they have been recorded, but
-both cardiovascular diseases and cancers conform with the pattern already
+income categories confirmed large inequality in mortality
+between different groups during the latest decades, but also demonstrated more distinct differences between categories with better survival the more income and the higher edudation level people have.
+Both cardiovascular diseases and cancers conform with the pattern already
 observed when it comes to all-cause mortality.
@@
 This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
-the ESHD Conference in Leuven, Belgium, 2016. Here we make quick look back
+the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
 in time for the Skellefteå region (1851--1900), and we also consider causes
-of death for the later part of the 20th century. The {\sc Lisa} data allow
+of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
 us to study the effect of income and education on mortality in a more
-direct way. Furthermore, we analyse inequality in mortality based on income and education respectively for the most recent periods, and compare it with the results of the class-based analysis.
+direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.
 
@@
 analyses will be performed on data from the Skellefteå and Umeå regions
-1901--2013. The main questions are:
+1851--2013. The main questions are:
 
@@
 \item Is social position equally important among the retired population as for those in working age?
-\item Is there difference between classes in mortality from cardiovscular diseases and cancers respectively?
+\item Is there a difference between classes in mortality from cardiovscular diseases and cancers respectively?
 \item Are there large differences in survival depending on level of education and income categories?
 \end{enumerate}
+
+\section{Previous research}
+
+For a detailed presentation we refer to \citet{segb16}. Here is a short
+summary.
+
+Several studies have shown how important access to economic and other
+resources are for health, but it is not clear how this has developed in
+history. \citet{ant67} and \citet{smi83} suggest that social inequality in
+health has passed through different phases in history. According to them,
+social health differences were
+comparatively small during the pre-transitional phase when space was a
+strong determinant for the spread of disease. Differences then increased
+during the transitional phase when mortality declined and wealthy groups
+used their resources to gain better health. Finally, health differences
+decreased again in modern low-mortality settings when instead
+health-related behaviour became important, resulting in marginal
+differences. \citet{omran82} state, in the third proposition in his theory
+of the epidemiological transition, that even if the class differential in
+mortality was maintained during the transition, the decline
+set in earlier and was faster among privileged groups.
+
+These assumptions has however been questioned by later research. According to the present state of knowledge about the development of social 
+health inequalities (in Sweden but also in other European countries), no
+consistent social hierarchy has been found before 1900 but substantial
+differences towards the late 20th and early 21st century. It appears as if
+a more consequent social health divide developed in the period
+in-between. This period has not been thoroughly studied on this issue,
+mainly because of lack of data to analyse. However, the extension of data
+at the Demographic Data Base, Umeå University (DDB) makes such studies
+possible. We are now able to analyse social 
+differences in mortality with micro-data on a sufficiently large population
+and with substantial social diversity from the 20th century for the
+Skellefteå and Umeå regions \citep{wee16}.
 
@@
 rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
-This region had a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time.
+This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time.
 Umeå town had a
@@
 the whole studied period. It was the administrative centre in the county of
-Västerbotten and schools and military regiments was placed here. Consequently, Officials,
-teachers and the military were much more common in Umeå. Agriculture
+Västerbotten and schools and military regiments were placed here. Consequently, Officials,
+teachers and the military were much more common in Umeå than in Skellefteå. Agriculture
 dominated the rural part, there were some foundries, some
 industries for example in forestry and small-scale production. Umeå had a
-much more diversified occupational structure than Skellefteå. The popula-
-tion size as defined in our datasets was 18\,970 on January 1, 1900, 32\,900 on
+much more diversified occupational structure than Skellefteå. The population size 
+as defined in our datasets was 18\,970 on January 1, 1900, 32\,900 on
 January 1, 1950 and 103\,970 when the century ended.
 
-Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure time over hisclass  for our selected time periods.
-The category of workers are fairly stable over time, while the lower white collar group was very small during the 19th century, started to increase during the early 20th century and increasing considerable during the period from the 1960s onwards.
-The group of smaller entrepreneurs where farmers also are included, was large until the middle of the 20th century, but has diminished considerably thereafter.
-This is to a large extent explained by the almost disappearance of farmers in recent time.
+\section{Data and variables}
+
+The data for the present study comes from two large population databases at
+the DDB, Umeå University. The early period is covered by
+the database Poplink, the digitisation of historical parish registers for
+the Skellefteå and Umeå region \citep{wee16}. The population in the two
+regions is large enough and socially diverse to enable studies of social
+mortality inequalities. Poplink is based on linked parish records, allowing
+us to reconstruct life biographies on people as long as they remained in
+the region. The records are linked within but not between the regions. Data
+from the regions are accessed for the period 1901--1950s.
+
+The other large data-set comes from the Linnaeus database
+\citep{malm10}. It is based on different linked population registers from
+1960 to 2013 and is used within the ageing program at CEDAR, Umeå
+University. The study period from 1960 to 2000 are constructed from censuses every fifth
+year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
+For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
+The construction of these data-sets for analysis is presented below.
+
+Information on individuals are anonymized and the two databases are not
+linked between each other. This stops us from following individuals present
+in both databases throughout their lives, thus they are treated separately. This
+also makes it impossible to add information on individuals in the Linnaeus
+database from what we potentially could find in Poplink, for example family
+background or occupations earlier in life.
+
+In the data-set analysed here, all individuals 40 years and older and ever being resident in either
+of the regions are included. The data file contains the variables on sex,
+birth date, death date, first and last date of observation and type of
+entrance/exit.Total number of deaths analysed are 68.244.
+
+\subsection{Presence periods}
+
+Differences in available information, both when it comes to structure and variables used for attributing social class, from the two datasets as well as
+within the Linnaeus dataset makes necessary diverging approaches when it comes to
+the identification of presence periods within our
+regions. The Poplink data provides us with exact dates or at least year of
+start and exit of the presence of all individuals in our data. This is however not the case
+with the Linnaeus database.
+For the period 1960--2000 we use the presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare for defining the observation periods as well as their social class (see below) in our analyses.
+We have however no exact information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
+We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in this census.
+This overestimates their presence in our regions, but should not bias the results.
+We furthermore miss those that both moved in and moved out or vice versa between censuses.
+Exit dates are exact if they refer to death, and in these cases we consider them as constantly present from the last census to the death data even if the death took place in another part of Sweden.
+
+For the last period, 2002--2013, we use the yearly population registers (also known as LISA data from Statistics Sweden), together with information on deaths from National Board of Health and Welfare.
+This data allows a better control of the presence periods in the studied regions.
+
+\subsection{HISCLASS}
+
+In this paper we analyse mortality differences according to social class according to the classification scheme HISCLASS.
+The only type of socio-economic information available both in Linnaeus and Poplink is occupations, which can be used to identify distinct social classes.
+Thus, analysing according to social class is apart from a scientific choice, also a practical one.
+Using occupational data is the only and obvious alternative if we are to make a longitudinal study of socio-economic mortality differences with consistent classification schemes (although not without some problems).
+There are also other problems connected with using the same classification scheme covering a long time period, but this will be approached in more detail in the discussion section.
+Social position is treated as a time-dependent variable.
+
+The availability to information on occupations varies over time and between sources.
+The Poplink data usually provides information only on the occupation of head of households.
+The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
+The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information.
+In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid.
+Thus, the social class in the census 1990 is considered to be kept until 2000.
+
+The use of different data sets, forces us to apply different approaches towards defining the social position.
+In the Poplink data, social class is based on the occupational titles given in the parish records.
+These can be more or less precise.
+There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
+We consider however their validity to be sufficient for identification of social position.
+Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}.
+In this study and according to the research questions in the present paper, we have chosen to categorize wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
+This strategy makes the problem of under-recording less problematic.
+
+Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers.
+Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database.
+Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period.
+It is also worth to notice that female labour force participation became the norm, and we can therefore argue that the occupation of the male as signifier of the social class became obsolete.
+We let all included persons be signified by their individual occupation for this period.
+Still, we need to keep in mind that the results for the different periods are not completely comparable.
+
+Due to the construction of the time-dependent data, some records report unknown social position.
+This has been updated according to rules in order to optimize available information and minimize missing information.
+Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
+The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.This, together with the circumstance that many women had no occupation in the census 1960 explans the large proportion where we have not been able to determine class during that decade.
+
+The classification scheme used here is HISCLASS \citep{lema11}, a class-based system.
+The scheme uses HISCO as a basis for assigning categories.
+HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO.
+HISCO groups occupations according to tasks and skills.
+HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions and we have therefore used only the main categories, based on position on the labour market.
+We have therefore collapsed categories, ending up with the following four:
+
+\begin{enumerate}
+\item \"elite, HISCLASS 1 and 2. Higher managers and professionals.
+\item Lower managers and farmers, HISCLASS 3 and 8.
+\item Lower white collar, HISCLASS 4, 5 and 6. Lower professionals, lower clerical, foremen.
+\item Workers, HISCLASS 7, 9, 10, 11 and 12. Workers (including farm workers).
+%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
+\end{enumerate}
+
+For the Poplink data, we have used a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
+The procedure for occupations in the Linnaeus database is a little more complicated.
+The occupations in the censuses (NYK) must first be transferred to the system later in use, SSYK (which is the Swedish version of ISCO88).
+Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013, can be recoded into HISCLASS.
+
+Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure time over HISCLASS  for our selected periods.
+The category of workers are fairly stable over time, while the lower white collar group was very small during the 19th century, then starting to increase during the early 20th century and increasing considerably during the period from the 1960s onwards.
+The group of lower managers where farmers also are included, was large until the middle of the 20th century, but has diminished considerably thereafter.
+This is to a large extent explained by farmers almost disappearing as an occupation in recent time.
 The highest classes has increased substantially, although from a very low level in the 19th century.
@@
 For the corresponding numbers, with missing values ({\tt NA}) included, see 
-Table~\ref{tab:nana}.The proportion with missing hisclass was very low, between 4 and 8 \% lacked information allowing for determine class.
-This increased considerably for the beginning of period when the other dataset ---the Linnaeus database--- is used, in particular for the 1960s where every second person lacks hisclass and the 1970s with almost a third have missing information.
+Table~\ref{tab:nana}.The proportion with missing HISCLASS was very low during the periods until 1950, between 4 and 8 \% lacked information allowing determinatin of class.
+This increased considerably for the beginning of period when the other dataset ---the Linnaeus database--- is used, in particular for the 1960s where every second person lacks HISCLASS and the 1970s with almost a third have missing information.
 This has to do with difficulties in the determination of class from the censuses, something we discuss below.
@@
 
-Regarding the distribution of civil status over time, see Figure~\ref{fig:civstskum}.
+\subsection{Marital status}
+
+Marital status is treated as a time-dependent variable in the data file.
+The statuses used are unmarried, married and dissoluted marriages (without distinguishing between widowed and divorces).
+The status unmarried is often not explicitly noted in the source, which makes the status unknown in the database Poplink for people for people moving to our parishes.
+If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
+
+Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.
 
@@
 
+\subsection{Cause of death}
+In the present study, we analyse mortality in the two most common disease groups in the ages we are investigating -- cardiovascular diseases and cancers.
+Compulsory reporting of cause of death became introduced in 1911 for all Sweden, and not only the urban part which had been the case from 1860.
+Classification systems have however changed.
+Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
+In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
+
 The distribution of causes of death is shown in Figure~\ref{fig:causeskum}.
-
-<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5>>=
+Notice the high proportion of other causes in the first two periods.
+This category include cases with unknown cause and unspecific causes which has implications for the cause-specific analyses.
+In the analyse section, we comment more on this.
+The total dominance of the two cause groups we are studying is apparent from the figure.
+
+<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 5.5>>=
 vb01 <- vb0[vb0$event == 1, ]
@@
 
-Causes of death before 1951.
-
-<<causeof51>>=
-vb051 <- vb0[vb0$period %in% c("1901-1925", "1926-1950"), ]
-vb051$period <- factor(vb051$period)
-with(vb051[vb051$event == 1, ], table(ULORSAK, period, useNA = "ifany"))
-@
-
-\section{Previous research}
-
-For a detailed presentation we refer to \citet{segb16}. Here is a short
-summary.
-
-Several studies have shown how important access to economic and other
-resources are for health, but it is not clear how this has developed in
-history. \citet{ant67} and \citet{smi83} suggest that social inequality in
-health has passed through different phases in history. According to them,
-social health differences were
-comparatively small during the pre-transitional phase when space was a
-strong determinant for the spread of disease. Differences then increased
-during the transitional phase when mortality declined and wealthy groups
-used their resources to gain better health. Finally, health differences
-decreased again in modern low-mortality settings when instead
-health-related behaviour became important, resulting in marginal
-differences. \citet{omran82} state, in the third proposition in his theory
-of the epidemiological transition, that even if the class differential in
-mortality was maintained during the transition, the decline
-set in earlier and was faster among privileged groups.
-
-According to the present state of knowledge about the development of social 
-health inequalities (in Sweden but also in other European countries), no
-consistent social hierarchy has been found before 1900 but substantial
-differences towards the late 20th and early 21st century. It appears as if
-a more consequent social health divide developed in the period
-in-between. This period has not been thoroughly studied on this issue,
-mainly because of lack of data to analyse. However, the extension of data
-at the Demographic Data Base, Umeå University (DDB) makes such studies
-possible. We are now able to analyse social 
-differences in mortality with micro-data on a sufficiently large population
-and with substantial social diversity from the 20th century for the
-Skellefteå and Umeå regions \citep{wee16}.
-
-\section{Data and variables}
-
-The data for the present study comes from two large population databases at
-the Demographic Data Base, Umeå University. The early period is covered by
-the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region \citep{wee16}. The population in the two
-regions is large enough and socially diverse to enable studies of social
-mortality inequalities. Poplink is based on linked parish records, allowing
-us to reconstruct life biographies on people as long as they remained in
-the region. The records are linked within but not between the regions. Data
-from the regions are accessed for the period 1901--1950s.
-
-The other large data-set comes from the Linnaeus database
-\citep{malm10}. It is based on different linked population registers from
-1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1960 to 2000 are constructed from censuses every fifth
-year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
-For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
-The construction of these data-sets for analysis is presented below.
-
-Information on individuals are anonymized and the two databases are not
-linked between each other. This stops us from following individuals present
-in both databases throughout their lives, thus they are treated separately. This
-also makes it impossible to add information on individuals in the Linnaeus
-database from what we potentially could find in Poplink, for example family
-background or occupations earlier in life.
-
-In the data-set analysed here, all individuals 40 years and older and ever being resident in either
-of the regions are included. The data file contains the variables on sex,
-birth date, death date, first and last date of observation and type of
-entrance/exit.
-
-\subsection{Presence periods}
-
-Differences in available information, both when it comes to structure and variables used for attributing social class, from the two datasets as well as
-within the Linnaeus dataset makes necessary diverging approaches when it comes to
-the identification of presence periods within our
-regions. The Poplink data provides us with exact dates or at least year of
-start and exit of the presence of all individuals in our data. This is however not the case
-with the Linnaeus database.
-For the period 1960--2000 we use the presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare for defining the observation periods as well as their social class (see below) in our analyses.
-We have however no exact information on what happened in between the censuses, for example the exact date of exit for persons that moved between two censuses.
-We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in this census.This overestimates their presence in our regions, but should not bias the results.
-We furthermore miss those that both moved in and moved out or vice versa between censuses.
-Exit dates are exact if they refer to death, and in these cases we consider them as constantly present from the last census to the death data even if the death took place in another part of Sweden.
-
-For the last period, 2002--2013, we use the yearly population registers (also known as LISA data from Statistics Sweden), together with information on deaths from National Board of Health and Welfare.
-This data allows a better control of the presence periods in the studied regions.
-
-\subsection{HISCLASS}
-
-In this paper we analyse mortality differences according to social class according to the classification scheme HISCLASS.
-The only type of socio-economic information available both in Linnaeus and Poplink is occupations, which can be used to identify distinct social classes.
-Thus, analysing according to social class is apart from a scientific choice, also a practical one.
-Using occupational data is the only and obvious alternative if we are to make a longitudinal study of socio-economic mortality differences with consistent classification schemes (although not without some problems).
-There are also other problems connected with using the same classification scheme covering a long time period, but this will be approached in more detail in the discussion section.
-Social position is treated as a time-dependent variable.
-
-The availability to information on occupations varies over time and between sources.
-The Poplink data usually provides information only on the occupation of head of households.
-The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
-The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information.
-In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid.
-Thus, the social class in the census 1990 is considered to be kept until 2000.
-
-The use of different data sets, forces us to apply different approaches towards defining the social position.
-In the Poplink data, social class is based on the occupational titles given in the parish records.
-These can be more or less precise.
-There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
-We consider however their validity to be sufficient for identification of social position.
-Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}.
-In this study and according to the research questions in the present paper, we have chosen to categorize wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
-This strategy makes the problem of under-recording less problematic.
-
-Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers.
-Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database.
-Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period.
-It is also worth to notice that female labour force participation became the norm, and we could therefore argue that the occupation of the male as signifier of the social class became obsolete.
-We therefore let all included persons be signified by their individual occupation for this period.
-Still, we need to keep in mind that the results for the different periods are not completely comparable.
-
-Due to the construction of the time-dependent data, some records report unknown social position.
-This has been updated according to rules in order to optimize available information and minimize missing information.
-Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
-The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.This, together with the circumstance that many women had no occupation in the census 1960 explans the large proportion where we have not been able to determine class during that decade.
-
-The classification scheme used here is HISCLASS \citep{lema11}, a class-based system.
-The scheme uses HISCO as a basis for assigning categories.
-HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO.
-HISCO groups occupations according to tasks and skills.
-HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions and we have therefore used only the main categories, based on position on the labour market.
-We have therefore collapsed some categories, ending up with the following four:
-
-\begin{enumerate}
-\item \"elite, HISCLASS 1 and 2. Higher managers and higher professionals.
-\item Lower managers and farmers, HISCLASS 3 and 8.
-\item Lower white collar, HISCLASS 4, 5 and 6. Lower professionals, lower clerical, foremen.
-\item Workers, HISCLASS 7, 9, 10, 11 and 12. Workers (including farm workers).
-%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
-\end{enumerate}
-
-For the Poplink data, we have used the script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
-The procedure for the occupations in the Linnaeus database is a little more complicated.
-The occupations in the censuses (NYK) must first be transferred to the system later used, SSYK (which is the Swedish version of ISCO88).
-Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013 can be recoded into HISCLASS.
-
-\subsection{Income}
-Income is based on disposable individual income from the LISA database at Statistics Sweden. We have categorized in four categories of the same size.
-
-\subsection{Education}
-Education is here analysed only for the period 1990 onwards.
-Years of education are given in 7 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
-It ranges from basic education (six year Folkskola) at level 1, to high academic degree (PhD and similar) at level 7.
-
-\subsection{Marital status}
-
-Marital status is treated as a time-dependent variable in the data file.
-The statuses used are unmarried, married and dissolution of marriage (without distinguishing between widowed and divorces).
-The status unmarried is often not explicitly stated, which makes the status unknown in the database Poplink for people for people moving to our parishes.
-If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
-
-\subsection{Cause of death}
-In the present study, we analyse mortality in the two most common disease groups in the ages we are investigating -- cardiovascular diseases and cancers.
-Compulsory reporting of cause of death became introduced in 1911 for all Sweden, and not only the urban part which had been the case from 1860.
-Classification systems have however changed.
-Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
-In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
-
 \subsection{Periods for analysis}
 We have made separate analyses for the different periods.
-For the century 1851-1950 we have divided it into four periods of 25 years each.
-The nineteenth century period represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
-During the years 1901-1950, industrialization became stronger and the urban environments increased in population size.
-The modern welfare state started to develop during this time.
+For the century 1851--1950 we have divided it into four periods of 25 years each.
+The nineteenth century period represents a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
+During the years 1901--1950, industrialization had started and the urban environments increased in population size.
+The modern welfare state began to develop during this time.
 For the period 1960 onwards, the analyses is made on decades except for the two last periods.
@@
 The start of the last period (2008--2013) coincides with the international financial crisis that also had strong impact on Swedish economy.
+
+\subsection{Income}
+Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1990 to 2005.
+
+\subsection{Education}
+Education is here analysed only for the period 1990 onwards.
+Years of education are given in 7 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
+It ranges from basic education (less than nine years schooling) at level 1, to high academic degree (PhD and similar) at level 7.
 
@@
 women and for the age groups 40--64 and 65--89. The model in the analyses
-include social class, marital status and urban/rural residence. Each decade
+include social class, sex and marital status. Each decade
 during the 20th century gets separate analyses, but we then combine results
@@
 2013, with the exception of the period 1951--1960. We stratify on HISCLASS
-and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
-(Skellefteå/Umeå town or outside). The cumulative
+and control for the covariates {\tt sex} and {\tt   civst}. The cumulative
 hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
@@
 We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
-periods, in order to illustrate the process. They are adjusted for {\tt civil status}, and
-{\tt urban/rural} residence. We present separate analyses  for women and men, see 
+periods, in order to illustrate the process. They are adjusted for {\tt civil status}. We present separate analyses  for women and men, see 
 Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.
@@
 the social pattern that will be commented on below, there are a couple of
-observations that can be made from the figure. Women had consistently
+observations to be made from the figure. Women have consistently
 better survival than men. This is a well-known phenomenon in Swedish
@@
 obvious that differences between social classes have diminished. This
-conclusion has been confirmed by separate analyses (not shown). There is no
+conclusion has been confirmed by separate analyses (results not shown). There is no
 clear indication of homogenisation of mortality levels between social
@@
 
-However, we are mainly interested in illustrating the change in relative position among the hisclasses over time rather than the absolute levels, so we \emph{standardize} the levels in each time period.
+However, we are mainly interested in illustrating the change in relative position among the HISCLASSES over time rather than the absolute levels, so we \emph{standardize} the levels in each time period.
 It is illustrated in
-Figure~\ref{fig:cumswstand}. In the following, we only present these relative mortality levels for the different classes for the elderly and according to the two cause of death groups.
+Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these relative mortality levels for the different classes.
 
@@
 decade.
-Furthermore, we need to be cautious when interpreting the results for the 1960s in particular, but also for the 1970s due to much missing information on Hisclass for these decades, something that have restricted the size of the population observed.
+Furthermore, we need to be cautious when interpreting the results for the 1960s in particular, but also for the 1970s due to much missing information on HISCLASS for these decades, something that have restricted the size of the population observed.
 The changing economy
@@
 half of the 20th century, these occupations has constantly decreased in
-numbers thereafter. This means that the Hisclass group 2 has somewhat different character during the last periods in the figures.
-
-The social pattern among women are quite much in accordance with previous
+numbers thereafter. This means that the second HISCLASS group has somewhat different character during the last periods in the figures.
+
+The social pattern of mortality among women is quite much in accordance with previous
 analyses from the present research group on other regions in northern
@@
 \citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
-low mortality. This changes however towards the end the 20th century. From
+low mortality until the 1960s. This changes however towards the end the 20th century. From
 the 1980s, there is a clear advantage when it comes to survival for higher
-managers and professionals. This is also what we usually find in analyses
+managers and professionals. The expected social gradient in mortality has developed. This is also what we usually find in analyses
 of social health inequalities in present-day Sweden as well as in most
@@
 
-A common alternative to a class analysis is using income or education for investigating socio-economic differences.
+A common alternative to a class analysis is to use income or education for investigating socio-economic health differences.
 Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others disadvantaged.
@@
 
-The analysis for the period 19990--2005 is based on a separate analysis with income and education as explanatory variables and with controls of  urban versus rural residence, period and marital status.
+The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables and with controls of period and marital status.
 
@@
 Both education and income were strongly significant, as well as marital status as expected.
-Urban or rural residence had however only weak effect.
 Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
@@
 The same can be said concerning income.
-More income results in lower mortality.
+Higher incomes result in lower mortality.
 This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden.
@@
 We now turn to the elderly, i.e. those that mostly have left the workforce. 
-We will use as a summary measure the value of the cumulative hazards at 
-duration 25 (age 90), starting from age 65.
-
-The general pattern and development of mortality do not differ that much from the younger age group, of course representing a much higher mortality level.
+Corresponding to the previous analysis, we use the value of the cumulative hazards at 
+duration 25 (age 90), starting from age 65 as a summary measure.
+
+The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level.
 Mortality among women was, as expected, lower than among
-men. We also find a clear decline in mortality, but this decline did not
-set in until sometimes after the Second World War (results not whown here). This corresponds to the
+men. We find a clear decline in mortality, but this decline did not
+set in definitely until sometimes after the Second World War (results not whown here), i.e. later than for the age group 40--64 years. This corresponds to the
 Swedish development where we find a turning point at this time with better
@@
 cardiovascular diseases. The results for the 1960s are however uncertain as
-discussed above, especially for women due to no possibility to determine hisclass for the majority of them and therefore we have not performed any analysis on elderly women in that decade.
+discussed above, especially for women due to the difficulty to determine HISCLASS for the majority of them.
 
@@
 In some periods, the social differences were small, and the elite even had substantially higher mortality during the early 20th century.
-Notice however that this group was rather small at this time, and there is a lot of random fluctuation.
+Notice however that this group was rather small at the time.
 During the recent decades, the pattern is not very clear, but for the two last periods, the elite have the lowest mortality.
-It is however difficult to identify a clear social gradient in mortality.
+It is however not possible to identify a clear social gradient in mortality.
 
@@
 Only during the second half of the 20th century, we find periods with much better survival for the highest social class while retired workers had fairly high mortality.
-But even in recent times, the pattern was not that clear.
-During the last periods, we do not find a social gradient on hisclasse among elderly men (even in this case, the new analyses differ somewhat from the ones in \citet{segb16}).
+But even in recent times, the pattern is not completely clear.
+During the last periods, we do not find a social gradient on HISCLASSES among elderly men (even in this case, the new analyses differ somewhat from the ones in \citet{segb16}).
 This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.
@@
 
-The fact that we find rather unclear associations between hisclasses and survival, makes alternative approaches to the question of socio-economic health differences interesting.
+The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting.
 As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
@@
 
-These figures shows clear association both with income and with education and for women as well as for men.
+These figures show clear associations both with income and education and for women as well as for men.
 More years of education resulted in lower mortality.
@@
 Belonging to working class, having less education and low income increases the mortality risks.
-We would therefore assume that we find a clear social pattern in the present analysis.
+We would therefore expect to find a clear social pattern in the present analysis.
 
@@
 Thus, many deaths during the 19th century were not diagnosed at all, and there was still a substantial part of all death in the early twentieth century where death cause is designated as unknown or dying from old age, despite reporting causes of death had become mandatory.
-Furthermore, we do not present results for the 1960s and 1970s due to much missing data on hisclass, especially for the elderly and for women during these decades.
+For the period 1901-1925, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, but at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent.
+Are there many cardiovascular deaths disguised there?
 <<causes41>>=
@@
 
-Models are fitted separately for each ten-year time period, from  1851 to
+As before, models are fitted separately for each ten-year time period, from  1851 to
 2013. We stratify on HISCLASS
-and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
-(Skellefteå town or outside). The cumulative
+and use the covariates {\tt sex} and {\tt   civst}. The cumulative
 hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
@@
 We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. They are adjusted for {\tt civil status}, and
-{\tt urban/rural} residence. We present separate analyses for women and men.
+periods. They are adjusted for {\tt civil status}. We present separate analyses for women and men.
 
@@
 
-Figure~\ref{fig:cumsw} shows mortality levels during different periods from 1851 onwards.
+Figure~\ref{fig:cumsw} shows mortality levels during different periods from 1901 onwards.
 A couple of observations can be made.
 Women from the highest social class had almost consistently lower mortality from cardiovascular diseases compared to others.This is apparent even for the periods with deficient cause of death data.
-The pattern concerning the other groups are less clear, but if we compare the two largest groups, the workers and the small entrepreneur/farmer group, the differences were rather small during most of the time.
+The pattern concerning the other groups are less clear, but if we compare the two largest groups, the workers and the lower manager/farmer group, the differences were rather small during most of the time.
 A new pattern has however developed from 1990 onwards.
@@
 The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
-Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the other large group, small-scale entrepreneurs/farmers.
+Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the other large group, lower manager/farmers.
 And in the same way as for women at this age, working class me have by far the highest mortality from this cause of death during the last decades.
+A social hierarchy in cardiovascular mortality seems to have developed both for men and women.
 
@@
 More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system.
+This is also well in line with our analysis based on HISCLASS above.
 
@@
 In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years.
-In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and
-{\tt urban/rural} residence and with separate analyses for women and men. We use as a summary measure the value of the cumulative hazards at 
+In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and with separate analyses for women and men.
+We use as a summary measure the value of the cumulative hazards at
 duration 25 (age 90), starting from age 65.
@@
 for the age group 40--64.
-The results for the 19th century and the first 20th century period are very shaky and will not be commented on.
-For women mortality in this disease group were much higher among the elite than the others  until only the last decades.
-Differences between the other classes were small, and elderly from a working class background are not disadvantaged.
-The social class pattern has changed during the last periods, but we do not find a consistent class pattern in mortality.
-Also for men, we find high cardiovascular mortality during the period 1926--1950 and the 1980s for the elite, but also for small-scale entrepreneurs/farmers during the first half of the 20th century.
-As for women, we find no clear class pattern during the last periods.The diffuse pattern of class differences in cardiovascular mortality is surprising.
+For women mortality in this disease group were somewhat higher among the elite and the lower manager/farmer group than the others until only the last decades.
+Elderly from a working class background have not been disadvantaged for most of the 20th century.
+The social class pattern has changed somewhat during the last periods and more in line to what is expected.
+The differences are however not that large.
+Cardiovascular disease mortality among elderly women do not present a consistent class pattern.
+Also for men, we find high cardiovascular mortality during the period 1926--1950 and the 1980s for the elite, but also for loer-managers/farmers during the first half of the 20th century.
+As for women, we find no clear class pattern during the last periods and the differences are quite small.The diffuse pattern of class differences in cardiovascular mortality is surprising.
 It is therefore of interest to find out what other aspects of social differentiation tell us.
+
+Notice that the results for the 1960s must be interpreted with caution due to much missing data on HISCLASS among the elderly.
 
@@
 The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers.
-In the present analysis we include cancer deaths 1851--2013, but with an unregistered and not analyzed period between 1951--1980.
-Once again we declare that the 19th century information on cause of death is very uncertain and under-registered, and that this problem partly remains into the early 20th century.
+In the present analysis we include cancer deaths 1901--2013, but with an unregistered and not analyzed period between 1951--1960.
+There is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated above by the large proportion of unknown or vague causes at that time.
+We believe however that cancers were better diagnosed than cardiovascular deaths.
+This is supported by the stability in proportion of cancer deaths between the two periods, 12,5 per cent 1901--1925 and 13,5 per cent for the period 1926--1950.
 
@@
 2013. We stratify on HISCLASS
-and use the covariates {\tt sex}, {\tt   civst}, and {\tt urban}
-(Skellefteå town or outside). The cumulative
+and use the covariates {\tt sex} and {\tt   civst}. The cumulative
 hazards at the end of the period (65 if 40--64, 90 if 65--89) are the
@@
 
-We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. They are adjusted for {\tt civil status}, and
-{\tt urban/rural} residence.
+%%We start by looking at cumulative hazards by {\sc Hisclass} over time
+%%periods. They are adjusted for {\tt civil status}, and
+%%{\tt urban/rural} residence.
 We present separate analyses  for women and men.
-The two highest social classes, those we assume have more access to resources vital for health, are represented by blue lines, while the lower skilled workers have red lines.
+The two highest social %%classes, those we assume have more access to resources vital for health, are represented by blue %%lines, while the lower skilled workers have red lines.
 
@@
 
-Figure~\ref{fig:12cumsw} shows mortality levels according to hisclass during the different studied periods.
+Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods.
 The social differences were small until the middle of the 20th century, for women as well as for men.
-We disregard the obviously uncertain results for men during the 19th century.
 For the periods after 1980, the rather small differences remain.
-the results for the 1980s need however to be more thoroughly checked up.
-There is an indication of the elite having lower cancer mortality for both sexes during the last periods, but these differences are small and we do not find any clear social gradient in the risks.
+There is an indication of the elite having lower cancer mortality for women sexes during the last periods, but these differences are small and we do not find any clear social gradient in the risks.
 Thus, for this age group cancer mortality is not something that distinguishes between classes.
@@
 
-\subsubsection*{The effect of income and education, 1990--2005}
+\subsubsection*{The effect of income and education on cancer mortality, 1990--2005}
 
@@
 
-We finally turn to cancer mortality among the elderly.
-In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and {\tt urban/rural} residence and with separate analyses for women and men.
+We finally turn to cancer mortality among the elderly. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and with separate analyses for women and men.
 
@@
 
-We will use as a summary measure the value of the cumulative hazards at 
-duration 25 (age 90), starting from age 65.
-
-The relation between different classes and cancer mortality fluctuates considerable both for women and men in this age group.
-For some periods it has to do with quite few events to analyse, particularly fo the 1960s where hisclass has been difficult to determine for the elderly.
+We use the value of the cumulative hazards at 
+duration 25 (age 90) as a summary measure, starting from age 65.
+
+The relation between different classes and cancer mortality fluctuates considerably both for women and men in this age group.
+For some periods it has to do with quite few events to analyse, particularly fo the 1960s where HISCLASS has been difficult to determine for the elderly.
 Women of higher classes sometimes have favorable mortality, but unfavorable at other times.

### 2016-11-14 slides.Rnw
first commit

\documentclass[t,a4paper,11pt]{beamer}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{natbib}

%\usepackage[figuresonly]{endfloat}

%\bibliographystyle{apalike}

\usepackage{color}

\newcommand{\emp}[1]{\textcolor{blue}{#1}}

\usetheme{Singapore}% best choice.
  \setbeamercovered{transparent}%

\newcommand{\R}{{\bf R}}
\newcommand{\code}[1]{{\tt #1}}

\title{Social inequality in mortality among adults and elderly in 
  northern  Sweden 1851--2013\footnote{'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
    Chicago,
    November 17--20, 2016.}
}

\date{November 19, 2016}

\author{Göran Broström and Sören Edvinsson}

\begin{document}

\maketitle

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

\begin{frame}{Aims and questions}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased in the adult and elderly 
  population?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any \emp{gender differences} in the effect of social position?

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\end{enumerate}

\end{frame}

\begin{frame}{The Skellefteå and Umeå regions in Sweden}

<<umpop>>=
ume1900 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1900 &
                    birthdate + exit > 1900))
ume1950 <- with(obs[obs$region == "ume", ],
                sum(birthdate + enter < 1950 &
                    birthdate + exit > 1950))
@

\includegraphics[height=8cm]{myfigs/skum17}\includegraphics[height=8cm]{myfigs/sweden}

\end{frame}

\begin{frame}{The Skellefteå region}

\begin{itemize}  
\item Selection of
parishes surrounding the town of Skellefteå. 
\item Town founded in 1843 but with a
very small population during the first decades.
\item The majority of the
population in rural villages and hamlets, getting their
livelihood from agricultural production. 
\item 20th century: \emp{industrialisation}, population \emp{increase in the town},
\emp{diversified economy}. 
\end{itemize}

\end{frame}

\begin{frame}{The Umeå region}

\end{frame}

\begin{frame}[fragile]{Data sources}

\begin{itemize}
\item \emp{Swedish church book} records
\item digitized by the \emp{Demographic Data Base} at Umeå University, 
\begin{itemize}  
   \item \emp{\sc Popum} and 
   \item \emp{\sc Poplink}
\end{itemize}    
\item  \url{http://www.cedar.umu.se/}
\end{itemize}

\end{frame}

\begin{frame}{Software for analysis}

\begin{itemize}
\item  \R : \emp{\url{https://www.R-project.org}}.
  \begin{itemize}
    \item \emp{RStudio}: \url{https://www.rstudio.com}.
  \end{itemize}    
\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
  \begin{itemize}
  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
  \item \emp{rmarkdown} (RStudio Team, 2016).
  \item \emp{knitr} (Xie, 2016).
  \item \emp{eha} (Broström, 2015).
  \item \emp{xtable} (Dahl, 2016).
\end{itemize}            
\end{itemize}

\end{frame}

\begin{frame}{Variables}

\begin{description}

\item[what] is this?

\end{description}

\end{frame}

\begin{frame}{\sc Hisclass}

\begin{enumerate}
\item \'Elite, {\sc Hisclass} \emp{1 and 2}: Higher managers and higher professionals.
\item Lower managers, {\sc Hisclass} \emp{3}.
\item Farmers and fishermen, {\sc Hisclass} \emp{8}.
\item Lower professionals, {\sc Hisclass} \emp{4, 5 and 6}: Lower professionals, lower clerical, foremen.
\item Medium skilled workers, {\sc Hisclass} \emp{7}.
\item Lower skilled workers, {\sc Hisclass} \emp{9, 10, 11 and 12}: Lower and unskilled workers (including farm workers)
%%\item Lower skilled farm workers, {\sc Hisclass} 10 and 12, Lower and unskilled farm workers
\end{enumerate}

\end{frame}

\begin{frame}{{\sc Poplink} data, 1901--1950}

<<getdata, results='asis'>>=
library(xtable)
library(knitr)
load("data/adult.rda")
adult <- cal.window(adult, c(1901, 1951))
source("R/addPer.R")
adult <- addPer(adult, cuts = seq(1901, 1951, by = 10))
adult <- rc(adult)
## NEW hisclass, 6 + 7 --> 6:
adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
@

\end{frame}

\begin{frame}{{\sc Fob} data, 1961--2000}

<<ssykonfob>>=
load("fobb.rda")
@

\end{frame}

\begin{frame}{{\sc Lisa} data, 2002--2013}

<<lisadata>>=
load("liss.rda")
source("R/addPer.R")
liss$id <- liss$LINNEID
liss <- addPer(liss, name = "period", cuts = seq(2002, 2014, by = 6))
liss <- liss[!is.na(liss$hisclass), ]
liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8), 
                     labels = c("1+2", "3+4", "5", "6"))
liss <- liss[with(liss, !is.na(civst) &
                      !is.na(hisclass)), ]
liss$civst <- factor(liss$civst, 
                     labels = c("unmarried", "married", "prev. married"))
liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004, 284005, 
                                     248011, 248201, 248204)
@

\end{frame}

\begin{frame}{Results, ages 40--59}

<<colorc>>=
## Colors for plotting strata:
scol <- c(rep("blue", 2), rep("green", 2), rep("red", 2))
## Common for all groups:
labb1 <- levels(adult$period)
labb2 <- levels(fobb$period)
labb <- c(labb1, labb2)
nper1 <- length(labb1)
nper2 <- length(labb2)
nper <- nper1 + nper2
fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005, 248011, 248201, 248204)
civst <- numeric(NROW(fobb))
civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, anka
fobb$civst <- factor(civst, labels = c("unmarried", "married", "prev. married"))
source("R/getPH.R")
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, women."
scap1w <-  "Cumulative hazards, ages 40--59, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, men."
scap1m <-  "Cumulative hazards, ages 40--59, men."
##
cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard by HISCLASS, ages 40--59."    
cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). Log scale."
scap12log <- "Total hazard by HISCLASS, ages 40--59 (log scale)."

cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, women."
scap2w <-  "Cumulative hazards, ages 60--89, women."
cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, men."
scap2m <-  "Cumulative hazards, ages 60--89, men."
##
cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap22 <- "Total hazard by HISCLASS, ages 60--89."

##
adult4059 <- age.window(adult, c(40, 60))
fobb4059 <- age.window(fobb, c(40, 60))
## End 'common'
@

\end{frame}

\begin{frame}{Ages 40--59, women}

<<hazcalw>>=
resw <- getPH(adult4059[adult4059$sex == "female", ],
                fobb4059[fobb4059$KON == "kvinna", ],
                nper1, nper2, labb)
resm <- getPH(adult4059[adult4059$sex == "male", ],
                fobb4059[fobb4059$KON == "man", ],
                nper1, nper2, labb)
@

<<hazplotw, fig.height = 6>>=
source("R/plotHaz.ph.R")
plotHaz.ph(resw, main = labb, printLegend = FALSE)
@

\end{frame}

\begin{frame}{Ages 40--59, men}

<<hazplotm, fig.height = 6>>=
plotHaz.ph(resm, main = labb, printLegend = FALSE)
@

\end{frame}

\begin{frame}{Total hazards in ages 40--59, women}

<<cumsw, fig.height = 5.5>>=
source("R/plotED.ph.R")
woutp <- plotED.ph(resw, scol = scol, logScale = FALSE, ylab = "H(60|40)")
@

\end{frame}

\begin{frame}{Total hazards in ages 40--59, men}

<<cumsm, fig.height = 5.5>>=
moutp <- plotED.ph(resm, scol = scol, logScale = FALSE, ylab = "H(60|40)")
@

\end{frame}

\begin{frame}{Total hazards in ages 40--59, women: Log scale}

<<cumslogw>>=
wloutp <- plotED.ph(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
main = "Women")
@

\end{frame}

\begin{frame}{Total hazards in ages 40--59, men: Log scale}
<<cumslogm>>=
mloutp <- plotED.ph(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
main = "Men")
@

\end{frame}

\begin{frame}{{\sc Lisa} data, 2002--2013, women 40--59}

<<lissyoung>>=
liss4059 <- age.window(liss, c(40, 59))
fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss4059[liss4059$KON == "kvinna", ])
fitm1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss4059[liss4059$KON == "man", ])
dcol <- c("blue", "darkgreen", "darkgreen", "red")
@

<<plotliss1>>=
plot(fitw1, fn = "cum", col = dcol, main = "Women", lwd = rep(2, 4))
@

\end{frame}

\begin{frame}{{\sc Lisa} data, 2002--2013, men 40--59}

<<plotliss1m>>=  
plot(fitm1, fn = "cum", col = dcol, main = "Men", lwd = rep(2, 4))
@

\end{frame}

\begin{frame}{Cox regression, women 2002--2013, ages 40--59}

\scriptsize

<<tablisswom1, results = "asis">>=
fitwo1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss4059[liss4059$KON == "kvinna", ])
dr <- drop1(fitwo1, test = "Chisq")
ltx(fitwo1, dr = dr)
@

\end{frame}

\begin{frame}{Cox regression, men 2002--2013, ages 40--59}

\scriptsize

<<tablissman1, results = "asis">>=
fitma1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss4059[liss4059$KON == "man", ])
dr <- drop1(fitma1, test = "Chisq")
ltx(fitma1, dr = dr)
@

\end{frame}

\begin{frame}{Results, ages 60--89}

\end{frame}

\begin{frame}{Ages 60--89, women}

<<getske6089>>=
adult6089 <- age.window(adult, c(60, 90))
fobb6089 <- age.window(fobb, c(60, 90))
resw <- getPH(adult6089[adult6089$sex == "female", ],
                fobb6089[fobb6089$KON == "kvinna", ],
                nper1, nper2, labb, from = 60)
resm <- getPH(adult6089[adult6089$sex == "male", ],
                fobb6089[fobb6089$KON == "man", ],
                nper1, nper2, labb, from = 60)
plotHaz.ph(resw, main = labb)

@

\end{frame}

\begin{frame}{Ages 60--89, men}

<<hazplotmold>>=
plotHaz.ph(resm, main = labb)
@

\end{frame}

\begin{frame}{Total hazards in ages 60--89, women}

<<cumswold>>=
woutp <- plotED.ph(resw, scol = scol, logScale = FALSE, ylab = "H(90|60)")  
@

\end{frame}

\begin{frame}{Total hazards in ages 60--89, men}

<<cumsmold>>=
moutp <- plotED.ph(resm, scol = scol, logScale = FALSE, ylab = "H(90|60)")
@

\end{frame}

\begin{frame}{{\sc Lisa} data, 2002--2013, women 60--89}

<<lissold>>=
liss6089 <- age.window(liss, c(60, 90))
fitw2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss6089[liss6089$KON == "kvinna", ])
fitm2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + 
                   civst + urban, dist = "gompertz", 
               data = liss6089[liss6089$KON == "man", ])
dcol <- c("blue", "darkgreen", "red", "red")
plot(fitw2, fn = "cum", col = dcol)

@

\end{frame}

\begin{frame}{Cox regression, women 2002--2013, ages 60--89}

  \scriptsize
<<tablisswom2, results = "asis", eval = TRUE>>=
fitwo2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss6089[liss6089$KON == "kvinna", ])
dr <- drop1(fitwo2, test = "Chisq")
ltx(fitwo2, dr = dr)
@

\end{frame}

\begin{frame}{{\sc Lisa} data, 2002--2013, men 60--89}

<<lissoldmen>>=
plot(fitm2, fn = "cum", col = dcol)
@

\end{frame}

\begin{frame}{Cox regression, men 2002--2013, ages 60--89}

  \scriptsize
<<tablissman2, results = "asis", eval = TRUE>>=
fitma2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
                 data = liss6089[liss6089$KON == "man", ])
dr <- drop1(fitma2, test = "Chisq")
ltx(fitma2, dr = dr)
@

\end{frame}

\begin{frame}{Conclusion}

%\scriptsize

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item There are \emp{gender differences} in the social pattern
\item No difference in the social pattern in the two studied age groups.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\end{itemize}

%We started this paper by raising the issue of the development of social
%inequalities. When studying this question in history we often get surprised
%of its complexity. In a society characterised by large economic inequality, 
%we would expect that this would result in corresponding mortality
%differences, i.e. higher social classes having an advantage. We find the
%appearance of a consistent social pattern in mortality, to the advantage of
%the higher classes, but surprisingly late in our history. It is only in the
%recent decades that this pattern dominates. Improved survival characterised
%however all {\sc Hisclass}es. We can however confirm that there are gender
% differences in the social pattern. While there is no clear social gradient
% among men, the results at least partly indicate advantages for the higher
% social classes for women already early in the 20th century. Females had
% more to gain from a high social position. Previous studies of the mortality
% transition have made similar observations. Among adults, gender has often
% been of more significant importance than social class. The reason for this
% difference and for the late appearance of the modern social pattern among
% men is still an enigma that require further studies, even though we have
% suggested the possible importance of behavioural factors and changes in
% ideals connected to different social classes. This should stimulate us to
% further disentangle the pathways for how social position influence health
% and mortality in different societies. Finally, we do not find any clear
% differences in the social pattern in the two studied age groups.

\end{frame}

\end{document}

### 2016-11-14 R/plotED.glm.R
update

@@ -12,7 +12,7 @@ plotED.glm <- function(res,
     ##nper <- NCOL(res[[1]])
 
     ## Assuming 'm == 4', we try:
-    leg = c("1+2", "3+8", "4+5+6", "7+9+10+11+12")
+    leg = c("elite", "lowMan", "lowWhiteC", "worker")
     nper <- length(res)
     k <- NCOL(res[[1]])
     m <- NROW(res[[1]])

### 2016-11-14 oldmort.Rnw
update

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
 
@@
 income categories confirmed large inequality in mortality
-between different groups during the latest decades, but also demonstrated more distinct differences between categories with better survival the more income and the higher edudation level people have.
+between different groups during the latest decades, but also demonstrated more distinct differences between categories with better survival the more income and the higher education level people have.
 Both cardiovascular diseases and cancers conform with the pattern already
@@
 <<setup, include = FALSE>>=
-knitr::opts_chunk$set(cache=FALSE,echo=FALSE,message=FALSE,warning=TRUE,comment=NA)##$
+knitr::opts_chunk$set(cache=TRUE,echo=FALSE,message=FALSE,warning=TRUE,comment=NA)##$
 @
@@
 vbTab$period <- factor(vbTab$period)
+levels(vbTab$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
+levels(vbTab$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
 levels(vbTab$period)[8] <- "1991-2001"
@@
 vb$period <- factor(vb$period)
+levels(vb$hisclass) <- c("elite", "lowMan", "lowWhiteC", "worker")
+levels(vb$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
+levels(vb$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
 levels(vb$period)[8] <- "1991-2001"
@@
 \item Is social position equally important among the retired population as for those in working age?
-\item Is there a difference between classes in mortality from cardiovscular diseases and cancers respectively?
+\item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
 \item Are there large differences in survival depending on level of education and income categories?
@@
 
-Figure~\ref{fig:mapp} shows the study areas and their position in Sweden.
+Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
+show the study areas and their position in Sweden.
+
+\begin{figure}
+\includegraphics[height=19cm]{myfigs/SKEUME}
+\caption{The Skellefteå and Umeå regions.}
+\label{fig:SKEUME}
+\end{figure}
+
 \begin{figure}
@@
 \end{figure}
+
 Both regions are part
@@
 and the fertility transition was late. The Skellefteå population size as defined
-in our datasets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900,
+in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900,
 61\,938 on January 1, 1950 and 76\,723 at the end of the century.
@@
 much more diversified occupational structure than Skellefteå. The population size 
-as defined in our datasets was 18\,970 on January 1, 1900, 32\,900 on
+as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on
 January 1, 1950 and 103\,970 when the century ended.
@@
 
-Information on individuals are anonymized and the two databases are not
+Information on individuals are anonymised and the two databases are not
 linked between each other. This stops us from following individuals present
@@
 
-Differences in available information, both when it comes to structure and variables used for attributing social class, from the two datasets as well as
-within the Linnaeus dataset makes necessary diverging approaches when it comes to
+Differences in available information, both when it comes to structure and variables used for attributing social class, from the two data sets as well as
+within the Linnaeus data set makes necessary diverging approaches when it comes to
 the identification of presence periods within our
@@
 Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}.
-In this study and according to the research questions in the present paper, we have chosen to categorize wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
+In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
 This strategy makes the problem of under-recording less problematic.
@@
 Due to the construction of the time-dependent data, some records report unknown social position.
-This has been updated according to rules in order to optimize available information and minimize missing information.
+This has been updated according to rules in order to optimise available information and minimise missing information.
 Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
-The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.This, together with the circumstance that many women had no occupation in the census 1960 explans the large proportion where we have not been able to determine class during that decade.
+The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.This, together with the circumstance that many women had no occupation in the census 1960 explains the large proportion where we have not been able to determine class during that decade.
 
@@
 \begin{enumerate}
-\item \"elite, HISCLASS 1 and 2. Higher managers and professionals.
-\item Lower managers and farmers, HISCLASS 3 and 8.
-\item Lower white collar, HISCLASS 4, 5 and 6. Lower professionals, lower clerical, foremen.
-\item Workers, HISCLASS 7, 9, 10, 11 and 12. Workers (including farm workers).
+\item {\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
+\item {\tt lowMan}, HISCLASS 3 and 8. Lower managers and farmers.
+\item {\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
+\item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
 %%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
@@
 
-For the Poplink data, we have used a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
-The procedure for occupations in the Linnaeus database is a little more complicated.
-The occupations in the censuses (NYK) must first be transferred to the system later in use, SSYK (which is the Swedish version of ISCO88).
-Thereafter the SSYK, both those transferred from the censuses and those given in the population registers 2001-2013, can be recoded into HISCLASS.
-
-Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure time over HISCLASS  for our selected periods.
-The category of workers are fairly stable over time, while the lower white collar group was very small during the 19th century, then starting to increase during the early 20th century and increasing considerably during the period from the 1960s onwards.
-The group of lower managers where farmers also are included, was large until the middle of the 20th century, but has diminished considerably thereafter.
-This is to a large extent explained by farmers almost disappearing as an occupation in recent time.
-The highest classes has increased substantially, although from a very low level in the 19th century.
-The increase of academics in connection to the establishment of Umeå university explains much of this increase.
+For the Poplink data, we have used a script re-coding HISCO into HISCLASS,
+provided to us by Jonas Helgertz, Lund University. The procedure for
+occupations in the Linnaeus database is a little more complicated. The
+occupations in the censuses (NYK) must first be transferred to the system
+later in use, SSYK (which is the Swedish version of ISCO88). Thereafter the
+SSYK, both those transferred from the censuses and those given in the
+population registers 2001-2013, can be re-coded into HISCLASS.
+
+Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure
+time over HISCLASS  for our selected periods. The category of workers are
+fairly stable over time, while the lower white collar group was very small
+during the 19th century, then starting to increase during the early 20th
+century and increasing considerably during the period from the 1960s
+onwards. The group of lower managers where farmers also are included, was
+large until the middle of the 20th century, but has diminished considerably
+thereafter. This is to a large extent explained by farmers almost
+disappearing as an occupation in recent time. The highest classes has
+increased substantially, although from a very low level in the 19th
+century. The increase of academics in connection to the establishment of
+Umeå university explains much of this increase.
 
@@
 For the corresponding numbers, with missing values ({\tt NA}) included, see 
-Table~\ref{tab:nana}.The proportion with missing HISCLASS was very low during the periods until 1950, between 4 and 8 \% lacked information allowing determinatin of class.
-This increased considerably for the beginning of period when the other dataset ---the Linnaeus database--- is used, in particular for the 1960s where every second person lacks HISCLASS and the 1970s with almost a third have missing information.
-This has to do with difficulties in the determination of class from the censuses, something we discuss below.
-For the most recent period, the information is complete.
+Table~\ref{tab:nana}.The proportion with missing HISCLASS was very low
+during the periods until 1950, between 4 and 8 \% lacked information
+allowing determination of class. This increased considerably for the
+beginning of period when the other data set---the Linnaeus database---is
+used, in particular for the 1960s where every second person lacks HISCLASS
+and the 1970s with almost a third have missing information. This has to do
+with difficulties in the determination of class from the censuses,
+something we discuss below. For the most recent period, the information is
+complete.
 
@@
 load("data/vb0.rda")
-vb0$hisclass <- factor(vb0$hisclass, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
+vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
 x <- with(vb0, table(period, hisclass, useNA = "ifany"))
@@
 Marital status is treated as a time-dependent variable in the data file.
-The statuses used are unmarried, married and dissoluted marriages (without distinguishing between widowed and divorces).
+The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces).
 The status unmarried is often not explicitly noted in the source, which makes the status unknown in the database Poplink for people for people moving to our parishes.
@@
 \subsection{Cause of death}
-In the present study, we analyse mortality in the two most common disease groups in the ages we are investigating -- cardiovascular diseases and cancers.
-Compulsory reporting of cause of death became introduced in 1911 for all Sweden, and not only the urban part which had been the case from 1860.
-Classification systems have however changed.
-Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
-In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
+In the present study, we analyse mortality in the two most common disease
+groups in the ages we are investigating: Cardiovascular diseases and
+cancers. Compulsory reporting of cause of death became introduced in 1911
+for all Sweden, and not only the urban part which had been the case from
+1860. Classification systems have however changed. Sweden did not join ICD
+until 1952, but the historical causes of death in the Poplink database have
+been coded according to the ICD10 system. In the present study, we have
+identified the codes relevant for cardiovascular diseases and cancers in
+the ICD 8, 9 and 10.
 
@@
 
-<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 5.5>>=
+<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5>>=
 vb01 <- vb0[vb0$event == 1, ]
@@
 ttt <- prop.table(ttt, 2) * 100
-barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = )
+barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = "Per cent")
 @
@@
 The nineteenth century period represents a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
-During the years 1901--1950, industrialization had started and the urban environments increased in population size.
+During the years 1901--1950, industrialisation had started and the urban environments increased in population size.
 The modern welfare state began to develop during this time.
@@
 
+Another practical problem is the size of that data set:
+It contains no less than 2\,432\,871 records of 236\,011 individuals, of
+which 68\,244 are observed to die. 
+This makes it very inconvenient use the nowadays so popular semi-parametric
+technique known as  
+\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
+that allows for reduction of the data  
+to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
+relevant covariates \emph{categorical}, and then create a table with all
+possible combinations of levels of the covariates. For each such
+combination, the total amount of exposure and the number of deaths are
+calculated.  We have two sexes, three civil statuses, ten periods, two
+urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
+ten age groups, and 22 cohorts, in all 422\,400 possible
+combinations. However, in this scheme there are many structural zeroes
+(impossible combinations) and some random zeros, so the actual number of
+combinations we end up with is 24\,326. The data set is thus reduced in
+size by 99 per cent.
+
+The parametric model we use for the baseline hazard function is piecewise
+constant on five-year intervals, so we arrive at the following proportional
+hazards regression model:
+
+\begin{equation}\label{eq:prop}
+h(t; \lambda, \boldsymbol{\beta}, \textbf{x}) = 
+\lambda(t)e^{\boldsymbol{\beta} \mathbf{x}}, \; \; 40 < t \le 90,
+\end{equation}
+where
+
+\begin{equation*}
+\begin{split}  
+\mathbf{x} &= (x_1, \ldots, x_n) \text{\ is a vector of covariates,} \\
+\boldsymbol{\beta} &= (\beta_1, \ldots, \beta_n) \text{\ is the corresponding vector of regression
+parameters.} \\
+\end{split}
+\end{equation*}
+and
+\begin{equation*}
+\lambda(t) = \alpha_i,\; 40 + 5(i - 1) < t \le 40 + 5i, \; i = 1, \ldots, 10. 
+\end{equation*}
+
+Since all components of $\mathbf{x}$ are categorical covariates, the
+estimation of \eqref{eq:prop} can be performed in two ways, first by the
+original data set and a survival analysis routine handling a piecewise
+constant hazard, or second, by the reduced data set (table) and
+\emph{Poisson regression}. The second way is faster than the first by a
+factor of size at least 100, often much more, depending on hardware, and
+the point is that the two ways give exactly the same results. This is
+because the table is a \emph{sufficient statistic} for \eqref{eq:prop} and
+the original data set.
+
 The analyses are performed in the \R\ environment for statistical computing
@@
 the development in more detail. We divide the analysis into two age groups,
-40--64 and 65--89 years of age, the first age group representing people that were still mainly in the workforce, and the secon those that mainly were retired. The model controls for period, marital
+40--64 and 65--89 years of age, the first age group representing people that were still mainly in the workforce, and the second those that mainly were retired. The model controls for period, marital
 status and whether the individual resides in the urban or rural part. There
@@
 reason for this choice is that the effect of {\sc Hisclass} on survival is
-non-proportional, that is, it varies with age. Thus, the main results are
-presented \emph{graphically}. This causes lack of information in the higher
+non-proportional, that is, it varies with age. With the reduced data set
+and Poisson regression, this is simply achieved by introducing an
+interaction term between {\tt age} and {\tt hisclass}. Then, the main
+results are 
+presented \emph{graphically}. There is some  lack of information in the higher
 ages around 1960--1980 (see Table~\ref{tab:nana}), because that kind of information is simply missing
@@
 men. We find a clear decline in mortality, but this decline did not
-set in definitely until sometimes after the Second World War (results not whown here), i.e. later than for the age group 40--64 years. This corresponds to the
+set in definitely until sometimes after the Second World War (results not shown here), i.e. later than for the age group 40--64 years. This corresponds to the
 Swedish development where we find a turning point at this time with better
@@
 
-The analysis of cardiovascular mortality accordning to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
+The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
 More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system.
@@
 Cardiovascular disease mortality among elderly women do not present a consistent class pattern.
-Also for men, we find high cardiovascular mortality during the period 1926--1950 and the 1980s for the elite, but also for loer-managers/farmers during the first half of the 20th century.
+Also for men, we find high cardiovascular mortality during the period 1926--1950 and the 1980s for the elite, but also for lower-managers/farmers during the first half of the 20th century.
 As for women, we find no clear class pattern during the last periods and the differences are quite small.The diffuse pattern of class differences in cardiovascular mortality is surprising.
@@
 The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers.
-In the present analysis we include cancer deaths 1901--2013, but with an unregistered and not analyzed period between 1951--1960.
+In the present analysis we include cancer deaths 1901--2013, but with an unregistered and not analysed period between 1951--1960.
 There is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated above by the large proportion of unknown or vague causes at that time.
@@
 <<12colorc>>=
-## Colors for plotting strata:
+## Colours for plotting strata:
 scol <- c("blue", rep("darkgreen", 2), "red")
@@
 
-%%We start by looking at cumulative hazards by {\sc Hisclass} over time
-%%periods. They are adjusted for {\tt civil status}, and
-%%{\tt urban/rural} residence.
-We present separate analyses  for women and men.
-The two highest social %%classes, those we assume have more access to resources vital for health, are represented by blue %%lines, while the lower skilled workers have red lines.
-
-%\subsubsection{Women}
-
 <<12hazcalw>>=
@@
 In the previous section, we concluded that there were no social inequality in cancer mortality according to hisclass.
-The question here is if the lacking socio-economic differentiation holds also when using education and income as categorizing variables.
+The question here is if the lacking socio-economic differentiation holds also when using education and income as categorising variables.
 The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables.
@@
 The relation between different classes and cancer mortality fluctuates considerably both for women and men in this age group.
-For some periods it has to do with quite few events to analyse, particularly fo the 1960s where HISCLASS has been difficult to determine for the elderly.
-Women of higher classes sometimes have favorable mortality, but unfavorable at other times.
+For some periods it has to do with quite few events to analyse, particularly for the 1960s where HISCLASS has been difficult to determine for the elderly.
+Women of higher classes sometimes have favourable mortality, but unfavourable at other times.
 It seems however that the higher classes have reached lower cancer levels compared to the working class from 2001 onwards.
@@
 findings \citep{edli11, edbr12}. These earlier studies have ended with an
-assumption that more distinct differences to the advantage of higher
+assumption that distinct and consistent differences to the advantage of higher
 classes developed somewhere during and probably quite early in the 20th
@@
 recent phenomenon.
+Furthermore, this transition and the class pattern of mortality differ depending on age group.
+Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly, a somewhat modified result from what we found in the first and preliminary analyses presented in Leuven this year \citep{segb16b}.
+
+We have also added analyses of cause-specific mortality to the previous version.
+Mortality from cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
+What is surprising is however the small class differences among the elderly.
+Cancer is sometimes considered less connected to social class, and we find mixed results in our analyses.
+There are however indications of a social gradient during the last decades in both studied age groups.
+Development of death causes are however problematic since the ways they have been diagnosed and the proportion with missing or vague information has changed over time.
 
@@
 
-The results are however in accordance with other studies \citep{bevp11},
+As a test, we have however included analyses of the associations of
+education and income to mortality for the period 1990--2005. For women as
+well as for men, for both age groups and for all-cause mortality as well as
+cause-specific mortality we find consistent differences with higher
+educational level and more income leading to lower mortality. As already
+mentioned, the class analysis proved that it in most cases are strong
+differences during the last centuries, but the pattern is much clearer in
+the alternative ways of categorisation. This does not mean that a class
+analysis is worthless, rather that the different categorisations measure
+different things. But it certainly point to the importance to also consider
+other aspect of social inequality in the past. We do not know if education
+and income are equally important in the past and that these measures would
+provide us with a completely different picture, but this is something that
+calls for further investigation.
+
+The results of small socio-economic gradient in the past and its late development are however in accordance with other studies \citep{bevp11},
 for example by \citet{bedr11} in a very different setting, a couple of
@@
 might throw some light on this issue.
-
-When it comes to the impact of social factors in the working age population versus the elderly, we do not find any fundamental differences.
-The social class have somewhat smaller effect among the elderly, so there is some difference in levels but not in character.
 
@@
 further disentangle the pathways for how social position influence health
-and mortality in different societies. Finally, we do not find any clear
-differences in the social pattern in the two studied age groups.
+and mortality in different societies.
+Furthermore, we do not find any substantial differences or distinct pattern in the relation between class and the two major cause-of-death groups.
+Finally, we find stronger differences in the social pattern among the working-age population, while the associations were smaller or not so clear among the elderly.
+However, when analysing the association with education and income in the period 1990--2005 we find consistent patterns of higher income and higher education level being associated with lower mortality, indicating that these variables are better to catch socio-economic differences at least in the present-day society.
 

### 2016-11-15 CreateData/creDat.Rnw
keep lisa missing hisclass

@@
 \begin{abstract}
-This version is 'old', but all the code is included, so it will be crystal-clear to the reader exacttly what has been done.
+This version is 'old', but all the code is included, so it will be crystal-clear to the reader exactly what has been done.
 
@@
     ##lisadata <- lisadata[lisadata$AR > 1999.5 & !is.na(lisadata$SSYK3), ]
-    lisadata <- lisadata[!is.na(lisadata$SSYK3), ] ##NOTE!
+
+    q()
+    ## This is new: Do not remove:
+    ##lisadata <- lisadata[!is.na(lisadata$SSYK3), ] ##NOTE!
+
     ## Fix HISCLASS:

### 2016-11-15 R/plotED.glm.R
add text

@@ -37,7 +37,7 @@ plotED.glm <- function(res,
         ##to <- max(outp[, out], na.rm = TRUE)
         ##fr <- min(outp[, out], na.rm = TRUE)
         fr <- 0
-        to <- 1
+        to <- 0.6
         #cat("to = ", to, ", fr = ", fr, "\n")
     }else{    
         to <- max(outp)

### 2016-11-15 R/plotHaz.glm.R
add text

@@ -4,7 +4,7 @@ plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
     ## Note, number of levels of hisclass is hardcoded to 4 (four) here!!
     ## Will be fixed.
     
-    leg <- c("1+2", "3+4", "5", "6+7") ## a hack for the time being
+    leg <- c("elite", "lowMan", "lowWhiteC", "worker") ## a hack for the time being
     
     oldpar <- par(mfrow = layout)
     on.exit(par(oldpar))

### 2016-11-15 slides.Rnw
add stuff

@@
   for those in \emp{working age}?
+\item Is there a difference between classes in mortality from
+  \emp{cardiovascular} diseases and \emp{cancers} respectively? 
+\item Are there large differences in survival depending on level of
+  \emp{education} and \emp{income} categories?
+
 \end{enumerate}
@@
 
+\begin{itemize}
+\item \emp{Town} founded in \emp{1622}
+\begin{itemize}
+\item  Substantially \emp{larger} than Skellefteå
+\item \emp{administrative} centre of the county Västerbotten
+\item \emp{Military} regiments.
+\item \emp{Schools}
+\item \emp{University} from \emp{1963--65}.  
+\end{itemize}  
+\item The rural part similar to the one in the Skellefteå region.
+\end{itemize}
+
 \end{frame}
@@
 \begin{itemize}
-\item \emp{Swedish church book} records
+\item \emp{Swedish church book} records (1851--1950).
+  \begin{itemize}
 \item digitized by the \emp{Demographic Data Base} at Umeå University, 
@@
 \end{itemize}
+\item The \emp{Linnaeus} data base (1960--2013)
+  \begin{itemize}
+\item Census data 1960, 1965, \ldots, 1990
+\item Population registers ({\sc Lisa}, 1986--2013)
+  \begin{itemize}
+    \item \emp{Income}
+    \item \emp{Education}  
+  \end{itemize}    
+ \item Death information from \emp{National Board of Health and Welfare}
+   \begin{itemize}
+\item Death dates from 1 January 1961 to 31 December 2013.
+  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
+   \end{itemize}
+\end{itemize}
+\end{itemize}
 
+\end{frame}
+
+\begin{frame}{The sampling frame}
+
+<<lexis2>>=
+plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40), 
+     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
+     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
+axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(2, las = 2, at = c(0, 40, 65, 90))
+axis(4, las = 2, at = c(0, 40, 65, 90))
+box()
+abline(h = 0)
+lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40), 
+      col = "blue", lwd = 2)
+text(1900, 65, "PopLink", cex = 2, col = "red")
+text(1900, 55, "79593 persons")
+text(1900, 50, "32801 deaths")
+text(1985, 65, "Linnaeus", cex = 2, col = "blue")
+text(1985, 55, "156676 persons")
+text(1985, 50, "42879 deaths")
+@ 
+\end{frame}
+
+\begin{frame}{The Linnaeus data}
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+<<picklexus,fig.width=6,fig.height = 5.9>>=
+
+load("data/vb.rda")
+vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
+x <- vbd$birthdate + vbd$exit
+y <- vbd$exit
+##par(mfrow = c(1, 2))
+plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p", 
+     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
+     axes = FALSE)
+axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
+axis(2, at = c(40, 65, 90))
+axis(4, at = c(40, 65, 90))
+for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
+    abline(v = i, lwd = 2)
+}
+for (i in 2002:2013){
+    abline(v = i, lwd = 2, col = "black")
+}
+abline(h = c(40, 65, 90), lty = 2)
+points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p", 
+     pch = ".", col = "red")
+box()
+@
+
+\end{column}
+
+\begin{column}{0.4\textwidth}
+\scriptsize
+
+\vspace{3\baselineskip}
+
+\begin{tabular}{l|rr}
+      & Number & Deaths \\ \hline
+\textcolor{red}{Women} & 70\,249 & \textcolor{red}{13\,888} \\
+\emp{Men} & 86\,427 & \emp{28\,991} \\ \hline
+Total  &156\,676 & 42\,879
+\end{tabular}
+
+\vspace{\baselineskip}
+
+\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
+\end{column}
+\end{columns}
 \end{frame}
@@
   \item \emp{knitr} (Xie, 2016).
-  \item \emp{eha} (Broström, 2015).
+  \item \emp{eha} (Broström, 2016).
   \item \emp{xtable} (Dahl, 2016).
@@
 
-\item[what] is this?
+\item[sex]
+\item[birth date]
+\item[death date]
+\item[cause of death]
+\item[parish]  
+\item[\sc Hisclass] grouped.
+\item[income] 1990--2005, grouped by quartiles.
+\item[education] 1990--2005. Seven categories.
 
@@
 \begin{enumerate}
-\item \'Elite, {\sc Hisclass} \emp{1 and 2}: Higher managers and higher professionals.
-\item Lower managers, {\sc Hisclass} \emp{3}.
-\item Farmers and fishermen, {\sc Hisclass} \emp{8}.
-\item Lower professionals, {\sc Hisclass} \emp{4, 5 and 6}: Lower professionals, lower clerical, foremen.
-\item Medium skilled workers, {\sc Hisclass} \emp{7}.
-\item Lower skilled workers, {\sc Hisclass} \emp{9, 10, 11 and 12}: Lower and unskilled workers (including farm workers)
-%%\item Lower skilled farm workers, {\sc Hisclass} 10 and 12, Lower and unskilled farm workers
+\item \emp{\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
+\item \emp{\tt lowMan}, HISCLASS 3 and 8. Lower managers and farmers.
+\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
+\item \emp{\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
 \end{enumerate}
@@
 
-\begin{frame}{{\sc Poplink} data, 1901--1950}
+\begin{frame}{Proportional hazards for women?}
 
-<<getdata, results='asis'>>=
-library(xtable)
-library(knitr)
-load("data/adult.rda")
-adult <- cal.window(adult, c(1901, 1951))
-source("R/addPer.R")
-adult <- addPer(adult, cuts = seq(1901, 1951, by = 10))
-adult <- rc(adult)
-## NEW hisclass, 6 + 7 --> 6:
-adult$hisclass[!is.na(adult$hisclass) & (adult$hisclass == 7)] <- 6
-@
+\includegraphics[width=11cm,height=8cm]{figure/hazplotw-1}
 
@@
 
-\begin{frame}{{\sc Fob} data, 1961--2000}
+\begin{frame}{Proportional hazards for men?}
 
-<<ssykonfob>>=
-load("fobb.rda")
-@
+\includegraphics[width=11cm,height=8cm]{figure/hazplotm-1}
 
@@
 
-\begin{frame}{{\sc Lisa} data, 2002--2013}
+\begin{frame}[fragile]{Interactions}
 
-<<lisadata>>=
-load("liss.rda")
-source("R/addPer.R")
-liss$id <- liss$LINNEID
-liss <- addPer(liss, name = "period", cuts = seq(2002, 2014, by = 6))
-liss <- liss[!is.na(liss$hisclass), ]
-liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8), 
-                     labels = c("1+2", "3+4", "5", "6"))
-liss <- liss[with(liss, !is.na(civst) &
-                      !is.na(hisclass)), ]
-liss$civst <- factor(liss$civst, 
-                     labels = c("unmarried", "married", "prev. married"))
-liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004, 284005, 
-                                     248011, 248201, 248204)
-@
+\begin{verbatim}
+fit <- coxph(Surv(enter - 40, exit - 40, event) ~ 
+    period * hisclass * sex + urban, data = vb)
+drop1(fit, test = "Chisq")
+\end{verbatim}
+
+\begin{center}
+\begin{tabular}{lrrrr}
+  \hline
+ & Df & AIC & LRT & Pr($>$Chi) \\ 
+  \hline
+$<$none$>$ &  & 1631901.32 &  &  \\ 
+  urban & 1 & 1631946 & 47.04 & $7.0\times 10^{-12}$ \\ 
+  \emp{period:hisclass:sex} & 27 & 1631957 & 110.11 & \emp{$5.3\times 10^{-12}$} \\ 
+   \hline
+\end{tabular}
+\end{center}
+
+\begin{itemize}
+
+\item This model is described by \emp{80 coefficients}.
+
+\item We show them in \emp{graphs}.
+\end{itemize}
 
@@
 
-\begin{frame}{Results, ages 40--59}
+\begin{frame}{No proportional hazards (wrt \sc Hisclass)!}
 
-<<colorc>>=
-## Colors for plotting strata:
-scol <- c(rep("blue", 2), rep("green", 2), rep("red", 2))
-## Common for all groups:
-labb1 <- levels(adult$period)
-labb2 <- levels(fobb$period)
-labb <- c(labb1, labb2)
-nper1 <- length(labb1)
-nper2 <- length(labb2)
-nper <- nper1 + nper2
-fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005, 248011, 248201, 248204)
-civst <- numeric(NROW(fobb))
-civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
-civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
-civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, anka
-fobb$civst <- factor(civst, labels = c("unmarried", "married", "prev. married"))
-source("R/getPH.R")
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, women."
-scap1w <-  "Cumulative hazards, ages 40--59, women."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--59, men."
-scap1m <-  "Cumulative hazards, ages 40--59, men."
-##
-cap12 <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard by HISCLASS, ages 40--59."    
-cap12log <- "Total hazard of dying before age 60 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). Log scale."
-scap12log <- "Total hazard by HISCLASS, ages 40--59 (log scale)."
+\begin{itemize}
+\item Fit each {\sc Hisclass} \emp{separately}, for each time period.
+\item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
+\end{itemize}
 
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, women."
-scap2w <-  "Cumulative hazards, ages 60--89, women."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 60--89, men."
-scap2m <-  "Cumulative hazards, ages 60--89, men."
-##
-cap22 <- "Total hazard of dying before age 90 for a 60 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap22 <- "Total hazard by HISCLASS, ages 60--89."
+So, we fit separate PH models for each combination of
 
-##
-adult4059 <- age.window(adult, c(40, 60))
-fobb4059 <- age.window(fobb, c(40, 60))
-## End 'common'
+\begin{itemize}
+  \item \emp{\tt sex}
+  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
+  \item \emp{\tt hisclass}
+  \item \emp{\tt age group}: 40--64, 65--89.
+  \item \emp{\tt cause of death}
+\end{itemize}    
+\end{frame}
+
+\begin{frame}{The data table}
+
+<<lexis3, fig.height = 4>>=
+plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40), 
+     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
+     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(40, 90))
+axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(2, las = 2, at = c(0, 40, 65, 90))
+axis(4, las = 2, at = c(0, 40, 65, 90))
+box()
+abline(h = 0)
+lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40), 
+      col = "blue", lwd = 2)
+for (i in c(1851, 1876, 1901, 1926, 1971, 1981, 1991, 2002, 2007)){
+    lines(c(i, i), c(40, 90), lty = 2)
+}
+for (i in c(45, 50, 55, 60, 65, 70, 75, 80, 85)){
+    lines(c(1851, 1951), c(i, i), lty = 2)
+    lines(c(1961, 2014), c(i, i), lty = 2)
+}
 @
+
+\emp{Poisson regression}:
+
+In \emp{each cell}, and for \emp{each} combination of \emp{covariate level}s, calculate 
+\begin{itemize}
+\item  the number of \emp{deaths} and 
+\item the total \emp{exposure time}.
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Women and men, age 40--64}
+
+  \begin{center}
+\includegraphics[height = 7.5cm]{figure/cumsw-1}
+  \end{center}
+\end{frame}
+
+\begin{frame}{Relative differences, ages 40--64 and 65--89}
+
+  %\begin{center}
+\includegraphics[height = 7.5cm]{figure/cumswstand-1}\includegraphics[height=7.5cm]{figure/cumswold-1}
+   %\end{center}
+\end{frame}
+
+\begin{frame}{Cancer mortality, ages 40-64 and 65-89}
+
+\includegraphics[height=7.5cm]{figure/12cumsw-1}\includegraphics[height=7.5cm]{figure/12cumswold-1}
 
@@
 
-\begin{frame}{Ages 40--59, women}
+\begin{frame}{Cardiovascular mortality}
 
-<<hazcalw>>=
-resw <- getPH(adult4059[adult4059$sex == "female", ],
-                fobb4059[fobb4059$KON == "kvinna", ],
-                nper1, nper2, labb)
-resm <- getPH(adult4059[adult4059$sex == "male", ],
-                fobb4059[fobb4059$KON == "man", ],
-                nper1, nper2, labb)
-@
-
-<<hazplotw, fig.height = 6>>=
-source("R/plotHaz.ph.R")
-plotHaz.ph(resw, main = labb, printLegend = FALSE)
-@
+\includegraphics[height=7.5cm]{figure/41cumsw-1}\includegraphics[height=7.5cm]{figure/41cumswold-1}
 
@@
 
-\begin{frame}{Ages 40--59, men}
+\begin{frame}{Education and income, 1990--2005, all deaths}
 
-<<hazplotm, fig.height = 6>>=
-plotHaz.ph(resm, main = labb, printLegend = FALSE)
-@
+\includegraphics[height=3.5cm]{figure/liswomfig-1} Women, 40--64.
+
+\includegraphics[height=3.5cm]{figure/lismanfig-1} Men, 40--64.
 
@@
 
-\begin{frame}{Total hazards in ages 40--59, women}
+\begin{frame}{Education and income, 1990--2005, all deaths}
 
-<<cumsw, fig.height = 5.5>>=
-source("R/plotED.ph.R")
-woutp <- plotED.ph(resw, scol = scol, logScale = FALSE, ylab = "H(60|40)")
-@
+\includegraphics[height=3.5cm]{figure/liswomfig2-1} Women, 65--89.
+
+\includegraphics[height=3.5cm]{figure/lismanfig2-1} Men, 65--89.
 
@@
 
-\begin{frame}{Total hazards in ages 40--59, men}
+\begin{frame}{Education and income, 1990--2005, cancer}
 
-<<cumsm, fig.height = 5.5>>=
-moutp <- plotED.ph(resm, scol = scol, logScale = FALSE, ylab = "H(60|40)")
-@
+\includegraphics[height=3.5cm]{figure/12liswomfig-1} Women, 40--64.
+
+\includegraphics[height=3.5cm]{figure/12lismanfig-1} Men, 40--64.
 
@@
 
-\begin{frame}{Total hazards in ages 40--59, women: Log scale}
+\begin{frame}{Education and income, 1990--2005, cancer}
 
-<<cumslogw>>=
-wloutp <- plotED.ph(resw, scol = scol, logScale = TRUE, ylab = "H(60|40)",
-main = "Women")
-@
+\includegraphics[height=3.5cm]{figure/12liswomfig2-1} Women, 65--89.
+
+\includegraphics[height=3.5cm]{figure/12lismanfig2-1} Men, 65--89.
 
@@
 
-\begin{frame}{Total hazards in ages 40--59, men: Log scale}
-<<cumslogm>>=
-mloutp <- plotED.ph(resm, scol = scol, logScale = TRUE, ylab = "H(60|40)",
-main = "Men")
-@
+\begin{frame}{Education and income, 1990--2005, cardiovascular}
+
+\includegraphics[height=3.5cm]{figure/41liswomfig-1} Women, 40--64.
+
+\includegraphics[height=3.5cm]{figure/41lismanfig-1} Men, 40--64.
 
@@
 
-\begin{frame}{{\sc Lisa} data, 2002--2013, women 40--59}
+\begin{frame}{Education and income, 1990--2005, cardiovascular}
 
-<<lissyoung>>=
-liss4059 <- age.window(liss, c(40, 59))
-fitw1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + 
-                   civst + urban, dist = "gompertz", 
-               data = liss4059[liss4059$KON == "kvinna", ])
-fitm1 <- phreg(Surv(enter-40, exit-40, event) ~ strata(hisclass) + 
-                   civst + urban, dist = "gompertz", 
-               data = liss4059[liss4059$KON == "man", ])
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
+\includegraphics[height=3.5cm]{figure/41liswomfig2-1} Women, 65--89.
 
-<<plotliss1>>=
-plot(fitw1, fn = "cum", col = dcol, main = "Women", lwd = rep(2, 4))
-@
+\includegraphics[height=3.5cm]{figure/41lismanfig2-1} Men, 65--89.
 
@@
 
-\begin{frame}{{\sc Lisa} data, 2002--2013, men 40--59}
-
-<<plotliss1m>>=  
-plot(fitm1, fn = "cum", col = dcol, main = "Men", lwd = rep(2, 4))
-@
-
-\end{frame}
-
-\begin{frame}{Cox regression, women 2002--2013, ages 40--59}
-
-\scriptsize
-
-<<tablisswom1, results = "asis">>=
-fitwo1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
-                 data = liss4059[liss4059$KON == "kvinna", ])
-dr <- drop1(fitwo1, test = "Chisq")
-ltx(fitwo1, dr = dr)
-@
-
-\end{frame}
-
-\begin{frame}{Cox regression, men 2002--2013, ages 40--59}
-
-\scriptsize
-
-<<tablissman1, results = "asis">>=
-fitma1 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
-                 data = liss4059[liss4059$KON == "man", ])
-dr <- drop1(fitma1, test = "Chisq")
-ltx(fitma1, dr = dr)
-@
-
-\end{frame}
-
-\begin{frame}{Results, ages 60--89}
-
-\end{frame}
-
-\begin{frame}{Ages 60--89, women}
-
-<<getske6089>>=
-adult6089 <- age.window(adult, c(60, 90))
-fobb6089 <- age.window(fobb, c(60, 90))
-resw <- getPH(adult6089[adult6089$sex == "female", ],
-                fobb6089[fobb6089$KON == "kvinna", ],
-                nper1, nper2, labb, from = 60)
-resm <- getPH(adult6089[adult6089$sex == "male", ],
-                fobb6089[fobb6089$KON == "man", ],
-                nper1, nper2, labb, from = 60)
-plotHaz.ph(resw, main = labb)
-
-@
-
-\end{frame}
-
-\begin{frame}{Ages 60--89, men}
-
-<<hazplotmold>>=
-plotHaz.ph(resm, main = labb)
-@
-
-\end{frame}
-
-\begin{frame}{Total hazards in ages 60--89, women}
-
-<<cumswold>>=
-woutp <- plotED.ph(resw, scol = scol, logScale = FALSE, ylab = "H(90|60)")  
-@
-
-\end{frame}
-
-\begin{frame}{Total hazards in ages 60--89, men}
-
-<<cumsmold>>=
-moutp <- plotED.ph(resm, scol = scol, logScale = FALSE, ylab = "H(90|60)")
-@
-
-\end{frame}
-
-\begin{frame}{{\sc Lisa} data, 2002--2013, women 60--89}
-
-<<lissold>>=
-liss6089 <- age.window(liss, c(60, 90))
-fitw2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + 
-                   civst + urban, dist = "gompertz", 
-               data = liss6089[liss6089$KON == "kvinna", ])
-fitm2 <- phreg(Surv(enter-60, exit-60, event) ~ strata(hisclass) + 
-                   civst + urban, dist = "gompertz", 
-               data = liss6089[liss6089$KON == "man", ])
-dcol <- c("blue", "darkgreen", "red", "red")
-plot(fitw2, fn = "cum", col = dcol)
-
-@
-
-\end{frame}
-
-\begin{frame}{Cox regression, women 2002--2013, ages 60--89}
-
-  \scriptsize
-<<tablisswom2, results = "asis", eval = TRUE>>=
-fitwo2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
-                 data = liss6089[liss6089$KON == "kvinna", ])
-dr <- drop1(fitwo2, test = "Chisq")
-ltx(fitwo2, dr = dr)
-@
-
-\end{frame}
-
-\begin{frame}{{\sc Lisa} data, 2002--2013, men 60--89}
-
-<<lissoldmen>>=
-plot(fitm2, fn = "cum", col = dcol)
-@
-
-\end{frame}
-
-\begin{frame}{Cox regression, men 2002--2013, ages 60--89}
-
-  \scriptsize
-<<tablissman2, results = "asis", eval = TRUE>>=
-fitma2 <- coxreg(Surv(enter, exit, event) ~ hisclass + civst + urban + period, 
-                 data = liss6089[liss6089$KON == "man", ])
-dr <- drop1(fitma2, test = "Chisq")
-ltx(fitma2, dr = dr)
-@
-
-\end{frame}
+%%% Stop here!
 
@@
 
-%We started this paper by raising the issue of the development of social
-%inequalities. When studying this question in history we often get surprised
-%of its complexity. In a society characterised by large economic inequality, 
-%we would expect that this would result in corresponding mortality
-%differences, i.e. higher social classes having an advantage. We find the
-%appearance of a consistent social pattern in mortality, to the advantage of
-%the higher classes, but surprisingly late in our history. It is only in the
-%recent decades that this pattern dominates. Improved survival characterised
-%however all {\sc Hisclass}es. We can however confirm that there are gender
-% differences in the social pattern. While there is no clear social gradient
-% among men, the results at least partly indicate advantages for the higher
-% social classes for women already early in the 20th century. Females had
-% more to gain from a high social position. Previous studies of the mortality
-% transition have made similar observations. Among adults, gender has often
-% been of more significant importance than social class. The reason for this
-% difference and for the late appearance of the modern social pattern among
-% men is still an enigma that require further studies, even though we have
-% suggested the possible importance of behavioural factors and changes in
-% ideals connected to different social classes. This should stimulate us to
-% further disentangle the pathways for how social position influence health
-% and mortality in different societies. Finally, we do not find any clear
-% differences in the social pattern in the two studied age groups.
-
 \end{frame}

### 2016-11-19 slides.Rnw
fix nice presentation

@@
 
-\begin{frame}{Aims and questions}
+\begin{frame}{Questions}
 
@@
 
-\begin{frame}{The Skellefteå and Umeå regions in Sweden}
+\begin{frame}{Where?}
 
-<<umpop>>=
-ume1900 <- with(obs[obs$region == "ume", ],
-                sum(birthdate + enter < 1900 &
-                    birthdate + exit > 1900))
-ume1950 <- with(obs[obs$region == "ume", ],
-                sum(birthdate + enter < 1950 &
-                    birthdate + exit > 1950))
-@
-
-\includegraphics[height=8cm]{myfigs/skum17}\includegraphics[height=8cm]{myfigs/sweden}
-
+\includegraphics[height=5cm]{myfigs/geo_we.pdf}\includegraphics[height=7cm]{myfigs/sweden.pdf}
 \end{frame}
 
-\begin{frame}{The Skellefteå region}
+\begin{frame}{Here!}
+
+\begin{columns}
+
+\begin{column}{0.5\textwidth}
+
+\includegraphics[height=7cm]{myfigs/skum_latest.pdf}
+\end{column}
+
+\begin{column}{0.5\textwidth}
+
+Skellefteå:
 
 \begin{itemize}
-\item Selection of
-parishes surrounding the town of Skellefteå. 
-\item Town founded in 1843 but with a
-very small population during the first decades.
-\item The majority of the
-population in rural villages and hamlets, getting their
-livelihood from agricultural production. 
-\item 20th century: \emp{industrialisation}, population \emp{increase in the town},
-\emp{diversified economy}. 
+
+\item Small town, heavy industry in the nineteenth century.
+
 \end{itemize}
 
-\end{frame}
-
-\begin{frame}{The Umeå region}
+Umeå:
 
 \begin{itemize}
-\item \emp{Town} founded in \emp{1622}
+
+\item Not so small town, older. Center for administration, education, and military.
+
+\end{itemize}
+
+Rural areas:
+
 \begin{itemize}
-\item  Substantially \emp{larger} than Skellefteå
-\item \emp{administrative} centre of the county Västerbotten
-\item \emp{Military} regiments.
-\item \emp{Schools}
-\item \emp{University} from \emp{1963--65}.  
+
+\item Very similar: Small farms dominating.
+
 \end{itemize}
-\item The rural part similar to the one in the Skellefteå region.
-\end{itemize}
+
+\end{column}
+\end{columns}
+%\includegraphics[height=7cm]{myfigs/sweden.pdf}
 
@@
 \begin{itemize}
-   \item \emp{\sc Popum} and 
+  % \item \emp{\sc Popum} and
    \item \emp{\sc Poplink}
@@
 ##par(mfrow = c(1, 2))
+plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
+     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
+     axes = FALSE)
+axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
+axis(2, at = c(40, 65, 90))
+axis(4, at = c(40, 65, 90))
+for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
+    abline(v = i, lwd = 2)
+}
+for (i in 2002:2013){
+    abline(v = i, lwd = 2, col = "black")
+}
+abline(h = c(40, 65, 90), lty = 2)
+points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
+     pch = ".", col = "red")
+box()
+@
+
+\end{column}
+
+\begin{column}{0.4\textwidth}
+\scriptsize
+
+\vspace{3\baselineskip}
+
+\begin{tabular}{l|rr}
+      & Number & Deaths \\ \hline
+\textcolor{red}{Women} & 70\,249 & \textcolor{red}{13\,888} \\
+\emp{Men} & 86\,427 & \emp{28\,991} \\ \hline
+Total  &156\,676 & 42\,879
+\end{tabular}
+
+\vspace{\baselineskip}
+
+\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}{The Linnaeus data}
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+<<picklexus2,fig.width=6,fig.height = 5.9>>=
+
+load("data/vb.rda")
+vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
+x <- vbd$birthdate + vbd$exit
+y <- vbd$exit
+##par(mfrow = c(1, 2))
 plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
@@
 \end{columns}
-\end{frame}
-
-\begin{frame}{Software for analysis}
-
-\begin{itemize}
-\item  \R : \emp{\url{https://www.R-project.org}}.
-  \begin{itemize}
-    \item \emp{RStudio}: \url{https://www.rstudio.com}.
-  \end{itemize}    
-\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
-  \begin{itemize}
-  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
-  \item \emp{rmarkdown} (RStudio Team, 2016).
-  \item \emp{knitr} (Xie, 2016).
-  \item \emp{eha} (Broström, 2016).
-  \item \emp{xtable} (Dahl, 2016).
-\end{itemize}            
-\end{itemize}
-
 \end{frame}
@@
 
+\begin{frame}{Answers(?)}
+
+\begin{enumerate}
+\item Has \emp{inequality in mortality} between \emp{social classes}
+  increased?
+\begin{itemize}
+\item Not really
+\end{itemize}
+%%\item Is the mortality transition initiated by a certain social class?
+
+\item Are there any \emp{gender differences} in the effect of social position?
+\begin{itemize}
+\item Yes
+\end{itemize}
+
+\item Is social position \emp{equally important} among the \emp{retired} population as
+  for those in \emp{working age}?
+\begin{itemize}
+\item Not really
+\end{itemize}
+
+\item Is there a difference between classes in mortality from
+  \emp{cardiovascular} diseases and \emp{cancers} respectively?
+\begin{itemize}
+\item Yes
+\end{itemize}
+\item Are there large differences in survival depending on level of
+  \emp{education} and \emp{income} categories?
+\begin{itemize}
+\item Yes, definitely.
+\end{itemize}
+
+\end{enumerate}
+
+\end{frame}
+
 \begin{frame}{Conclusion}
@@
 \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
+\item \emp{Education} and \emp{income} are the \emp{most important}
+  variables during late
+  time period.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Software for report writing and analysis}
+
+\begin{itemize}
+\item  \R : \emp{\url{https://www.R-project.org}}.
+  \begin{itemize}
+    \item \emp{RStudio}: \url{https://www.rstudio.com}.
+  \end{itemize}
+\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
+  \begin{itemize}
+  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
+  \item \emp{rmarkdown} (RStudio Team, 2016).
+  \item \emp{knitr} (Xie, 2016).
+  \item \emp{eha} (Broström, 2016).
+  \item \emp{xtable} (Dahl, 2016).
+\end{itemize}
+
+\item \emp{Version control} (reproducibility)
+  \begin{itemize}
+    \item \emp{git}: \url{http://git}
+    \item \emp{gitHub}: \url{http://github.com/goranbrostrom/}
+
+\end{itemize}
 \end{itemize}

### 2017-05-18 icass.Rnw
First commit

--- 2016-11-14 oldmort.Rnw
@@
 \begin{figure}
-\includegraphics[height=19cm]{myfigs/SKEUME}
+\includegraphics[height=19cm]{myfigs/skum17}
 \caption{The Skellefteå and Umeå regions.}
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Women")##, exclude = 5:6)
+                main = "Women 40-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Men")##, exclude = 5:6)
+                main = "Men 40-64")##, exclude = 5:6)
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
-                    ylab = "TMR(65|40)", main = "Women")##, exclude = 5:6)
+                    ylab = "TMR(65|40)", main = "Women 40-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Men")##, exclude = 5:6)
+                    ylab = "TMR(65|40)", main = "Men 40-64")##, exclude = 5:6)
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Women")##, exclude = 5:6)
+                    ylab = "TMR(90|65)", main = "Women 65-89")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Men")##, exclude = 5:6)
+                    ylab = "TMR(90|65)", main = "Men 65-89")##, exclude = 5:6)
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)",  main = "Women")
+                    ylab = "TMR(65|40)",  main = "Women 40-64")
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)", main = "Men")
+                    ylab = "TMR(65|40)", main = "Men 40-64")
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women")
+                    ylab = "H(90|65)",  main = "Women 65-89")
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)", main = "Men")
+                    ylab = "H(90|65)", main = "Men 65-89")
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2, 
-                    ylab = "TMR(65|40)",  main = "Women")
+                    ylab = "TMR(65|40)",  main = "Women 40-64")
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)", main = "Men")
+                    ylab = "TMR(65|40)", main = "Men 40-64")
 par(oldpar)

### 2017-05-18 CreateData/R/putONssyk.R
refresh

@@ -2,7 +2,7 @@ putONssyk <- function(dat){
     ## 'dat' is the processed file 'fobdata'
     ## This must be run from "Mortality/CreDat"
     
-    load("../../../Data/yr_ssyk.rda")
+    load("../../Data/yr_ssyk.rda")
     dat$ssyk <- NA
 
     ##for (year in seq(60, 75, by = 5)){
@@ -12,7 +12,7 @@ putONssyk <- function(dat){
   ##      indx <- match(nyr$YRKE)
     ##}
     ## 1960:
-    load("../../../Data/yr60.rda")
+    load("../../Data/yr60.rda")
     who <- dat$AR == 1960 & dat$YRKE %in% yr60$old
     nyr <- dat$YRKE[who]
     indx <- match(nyr, yr60$old)
@@ -20,7 +20,7 @@ putONssyk <- function(dat){
     dat$YRKE[who] <- nyr
     
     ## 1965:
-    load("../../../Data/yr65.rda")
+    load("../../Data/yr65.rda")
     who <- dat$AR == 1965 & dat$YRKE %in% yr65$old
     nyr <- dat$YRKE[who]
     indx <- match(nyr, yr65$old)
@@ -28,7 +28,7 @@ putONssyk <- function(dat){
     dat$YRKE[who] <- nyr
     
     ## 1970:
-    load("../../../Data/yr70.rda")
+    load("../../Data/yr70.rda")
     who <- dat$AR == 1970 & dat$YRKE %in% yr70$old
     nyr <- dat$YRKE[who]
     indx <- match(nyr, yr70$old)
@@ -36,7 +36,7 @@ putONssyk <- function(dat){
     dat$YRKE[who] <- nyr
     
     ## 1975:
-    load("../../../Data/yr75.rda")
+    load("../../Data/yr75.rda")
     who <- dat$AR == 1975 & dat$YRKE %in% yr75$old
     nyr <- dat$YRKE[who]
     indx <- match(nyr, yr75$old)
@@ -48,7 +48,7 @@ putONssyk <- function(dat){
     ## Chansar på att 'yr75.rda' gäller:
 
     ## 1980 (try):
-    load("../../../Data/yr75.rda")
+    load("../../Data/yr75.rda")
     who <- dat$AR == 1980 & dat$YRKE %in% yr75$old
     nyr <- dat$YRKE[who]
     indx <- match(nyr, yr75$old)

### 2017-05-18 CreateData/R/ssyk3TOhisclass.R
refresh

@@ -1,9 +1,12 @@
 ssyk3TOhisclass <- function(x){
+
+    ## 2017-03.28: Three instead of four hisclasses.
+    
     ## 'x' is a vector of ssyk codes.
     ## as output we want a vector of hisclass codes (7 levels)
     
     ## Step 1: Convert shh$ssyk to "three-digit" (except '110'):
-    load("../../../Data/shh.rda")
+    load("../../Data/shh.rda")
     shh$ssyk <- shh$ssyk %/% 10 # removes last digit.
     ## Remove duplicate ssyk-rows in shh:
     shh <- shh[!duplicated(shh$ssyk), ]
@@ -13,15 +16,16 @@ ssyk3TOhisclass <- function(x){
     tmp <- shh$hisclass[indx]
 
     ## Step 2: go from 12 to 4 levels: ## NEW!!
+    ## Step 2: go from 12 to 3 levels: ## NEWER!!
     
     res <- rep(-1, length(x))
     
     tmp[is.na(tmp)] <- -1
     
     res[tmp %in% 1:2] <- 1
-    res[tmp %in% c(3, 8)] <- 2
-    res[tmp %in% 4:6] <- 3
-    res[tmp %in% c(7, 9:12)] <- 4
+    res[tmp %in% c(3:6, 8)] <- 2
+    ##res[tmp %in% 4:6] <- 3
+    res[tmp %in% c(7, 9:12)] <- 3 # '4' ---> '3'
     
     is.na(res) <- res == -1
     

### 2017-05-18 CreateData/creDat.Rnw
fix typo
refresh

@@
   adults and elderly in northern Sweden 1850--2013"\footnote{Paper presented
-    at the
-    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
-    Chicago, Illinois
-    November 17--20, 2016. Version 2016-09-19.}}
+    at CEDAR, 27 April 2017,
+    Version 2017-03-28.}}
 
-\date{November 19, 2016}
+\date{March 28, 2017}
 
@@
 \begin{abstract}
-This version is 'old', but all the code is included, so it will be crystal-clear to the reader exactly what has been done.
+This version is 'new', but all the code is included, so it will be crystal-clear to the reader exactly what has been done.
 
-Spotted errors must pe reported immediately!
+Spotted errors must be reported immediately!
 
@@
 how we create the tabular form data set (anonymized).
+
+In this version the hisclasses are grouped into three levels instead of four.
 
@@
 \begin{enumerate}
-\item \'Elite, HISCLASS 1 and 2, Higher and lower managers, and higher
+\item {\bf elite}, HISCLASS 1 and 2, Higher and lower managers, and higher
   professionals. 
-\item HISCLASS 3  and 8
-\item HISCLASS 4, 5, 6. Farmers, fishermen, and lower professionals, lower clerical, foremen.
-\item Workers, HISCLASS 7, 9, 10, 11, and 12.
+%%\item HISCLASS 3  and 8
+%%\item HISCLASS 4, 5, 6. Farmers, fishermen, and lower professionals, lower clerical, foremen.
+\item {\bf middle} HISCLASS 3, 4, 5, 6, and 8. Farmers, fishermen, and lower professionals, lower clerical, foremen.
+\item {\bf worker}, HISCLASS 7, 9, 10, 11, and 12.
 %%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
@@
 library(stringr)
-load("../../../Data/adult.rda")
+load("../../Data/adult.rda")
 adult <- calWin(adult, c(1851, 1951))
@@
                 ))
+##tva <- with(adult, !is.na(socBranch) &
+  ##              ((socBranch == "farming" & socStatus == "high") |
+##                     (socBranch == "business" & socStatus == "low")
+##                ))
+##tre <- with(adult, !is.na(socBranch) &
+##                    (socBranch == "official" & socStatus == "low")
+##               )
 tva <- with(adult, !is.na(socBranch) &
                 ((socBranch == "farming" & socStatus == "high") |
-                     (socBranch == "business" & socStatus == "low")
+                 (socBranch == "business" & socStatus == "low") |
+                 (socBranch == "official" & socStatus == "low")
                 ))
-tre <- with(adult, !is.na(socBranch) &
-                    (socBranch == "official" & socStatus == "low")
-               )
-fyra <- with(adult, !is.na(socBranch) &
+
+tre <- with(adult, !is.na(socBranch) & ## Original 'fyra' ---> 'tre'
                 ((socBranch == "farming" & socStatus == "low") |
@@
 hisc[tre] <- 3
-hisc[fyra] <- 4
+##hisc[fyra] <- 4
 is.na(hisc) <- hisc == -1
@@
 <<ssykonfob>>=
-    ##load("../../../Data/fobdata.rda")
+    ##load("../../Data/fobdata.rda")
     load("data4/fobdata.rda")
     ## Cause of death:
-    load("../../../Data/socstyr.rda")
+    load("../../Data/socstyr.rda")
     indx <- match(fobdata$LINNEID, socstyr$LINNEID)
@@
     fobb <- addPer(fobb, cuts = c(1961, 1971, 1981, 1991, 2002))
-    load("../../../Data/sex.rda")
+    load("../../Data/sex.rda")
 
@@
     ## Cause of death:
-    load("../../../Data/socstyr.rda")
+    load("../../Data/socstyr.rda")
     indx <- match(lisadata$LINNEID, socstyr$LINNEID)
@@
     ## Fix KON:
-    load("../../../Data/sex.rda")
+    load("../../Data/sex.rda")
     indx <- match(lisadata$LINNEID, sex$LINNEID)
@@
 
-    q()
     ## This is new: Do not remove:
@@
 vb$hisclass60[komp] <- vb$hisclass40[komp]
-vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
-vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
-vb$hisclass <- factor(vb$hisclass, labels = c("1+2", "3+8", "4+5+6", "7+9+10+11+12"))
+vb$hisclass40 <- factor(vb$hisclass40, labels = c("1+2", "3+4+5+6+8", "7+9+10+11+12"))
+vb$hisclass60 <- factor(vb$hisclass60, labels = c("1+2", "3+4+5+6+8", "7+9+10+11+12"))
+vb$hisclass <- factor(vb$hisclass, labels = c("1+2", "3+4+5+6+8", "7+9+10+11+12"))
 ## End cut ####################################################
@@
 <<aggregate>>=
+vb$poplink <- vb$dataset == "poplink"
 vbTab <- aggregate(vb[, c("event", "exposure")], 
                    vb[, c("sex", "age", "period", "cohort", 
-                          "urban", "civst", "hisclass40", "hisclass60")], 
+                          "urban", "civst", "hisclass40", "hisclass60", "poplink")], 
+                   FUN = sum)
+
+vb0$hisclass[is.na(vb0$hisclass)] <- 0
+vb0$poplink <- vb0$dataset == "poplink"
+vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
+                   vb0[, c("sex", "age", "period", "cohort", 
+                          "urban", "civst", "hisclass", "poplink")], 
                    FUN = sum)
@@
 save(vbTab, file = "../data/vbTab.rda")
+save(vb0Tab, file = "../data/vb0Tab.rda")
 @
 
-Thus, the one data file to use is {\tt vbTab.rda}.
+Thus, the one data file to use is {\tt vbTab.rda} or {\tt vb0Tab.rda}.
 
@@
 save(loTab, file = "../data/loTab.rda")
+save(lisaold, file = "../data/lisaold.rda")
 @

### 2017-05-18 R/plotHaz.glm.R
refresh

@@ -1,10 +1,13 @@
-plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
+plotHaz.glm <- function(res, printLegend = FALSE, col = c("black", "blue", "red"),
                     xlab = "", main = "", layout = c(4, 3)){
     ## 'res' is list of results from a fit.
     ## Note, number of levels of hisclass is hardcoded to 4 (four) here!!
     ## Will be fixed.
     
-    leg <- c("elite", "lowMan", "lowWhiteC", "worker") ## a hack for the time being
+    ##leg <- c("elite", "lowMan", "lowWhiteC", "worker") ## a hack for the time being
+    
+    leg <- rownames(res[[1]])
+    n.leg <- length(leg)
     
     oldpar <- par(mfrow = layout)
     on.exit(par(oldpar))
@@ -24,7 +27,7 @@ plotHaz.glm <- function(res, printLegend = FALSE, col = 1:4,
         plot(x, y[1, ], col = col[1], type = "l", lty = 1,
               main = names(res)[i], ylim = c(0, ymax),
              xlab = "Age", ylab = "Cum. Hazards", lwd = 2)
-        for (j in 2:4){
+        for (j in 2:n.leg){
             lines(x, y[j, ], col = col[j], lty = j, lwd = 2)
         }
         legend("topleft", legend = leg, col = col, lty = 1:4, cex = 1.0)

### 2017-05-18 oldmort.Rnw
refresh

@@
 load("data/vbTab.rda") 
+hisclevels <-  c("elite", "middle", "worker") # NOTE!!
 vbTab$period <- factor(vbTab$period)
-levels(vbTab$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vbTab$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
+levels(vbTab$hisclass40) <- hisclevels
+levels(vbTab$hisclass60) <- hisclevels
 levels(vbTab$period)[8] <- "1991-2001"
@@
 vb$period <- factor(vb$period)
-levels(vb$hisclass) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vb$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vb$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
+levels(vb$hisclass) <- hisclevels
+levels(vb$hisclass40) <- hisclevels
+levels(vb$hisclass60) <- hisclevels
 levels(vb$period)[8] <- "1991-2001"
@@
 yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+who <- length(levels(vbTab$hisclass40))
+barplot(t(yy)[who:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
 @
@@
 load("data/vb0.rda")
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
+vb0$hisclass <- factor(vb0$hisclass, labels = hisclevels)
 x <- with(vb0, table(period, hisclass, useNA = "ifany"))
@@
 ## Colors for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
+scol <- c("black", "blue", "red")
 ## Common for all groups:
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Women")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Men")##, exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "log(1-S(65|40))",
+                main = "Women 40-64")##, exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "log(1-S(65|40))",
+                main = "Men 40-64")##, exclude = 5:6)
 par(oldpar)

### 2017-05-18 R/get_period.R
refresh

@@ -1,0 +1,40 @@
+get_period <- function(period = "1981-1990", hisc = "hisclass40", 
+                       sex = "all", civst = "all", tabl){
+
+    tabl <- tabl[tabl$age != "85", ]
+    tabl <- tabl[tabl$period == period, ]
+    if (sex != "all"){
+        tabl <- tabl[tabl$sex == sex, ]
+    }
+    if (civst != "all"){
+        tabl <- tabl[tabl$civst == civst, ]
+    }
+    tabl$hisc <- tabl[[hisc]]
+    tabl <- aggregate(tabl[, c("event", "exposure")], 
+                      by = tabl[, c("age", "hisc")], FUN = sum)
+    tabl$haz <- tabl$event / tabl$exposure
+    tabl$age <- as.numeric(as.character(tabl$age))
+    tabl
+}
+
+plot_ash <- function(period = "1981-1990", hisc = "hisclass40", 
+                     sex = "all", civst = "all", tabl = vbTab){
+    x <- get_period(period, hisc, sex, civst, tabl)
+    levs <- levels(x$hisc)
+    ##x$haz <- log(x$haz)
+    with(x[x$hisc == levs[1], ], plot(age, haz, main = period, type = "b", col = "black", 
+                                      xlab = "Age", ylim = c(0.0004, 0.15), ylab = "Mortality", log = "y"))
+    with(x[x$hisc == levs[2], ], lines(age, haz, type = "b", col = "darkgreen"))
+    with(x[x$hisc == levs[3], ], lines(age, haz, type = "b", col = "red"))
+    abline(h = 0)
+    legend("topleft", legend = levs, col = c("black", "darkgreen", "red"), lty = 1)
+}
+
+s_show <- function(hisc = "hisclass40", sex = "all", civst = "all", tabl = vbTab, secs = 2){
+    peri <- levels(tabl$period)
+    for (per in peri){
+        cat("per = ", per, "\n")
+        plot_ash(period = per, hisc = hisc, sex = sex, civst = civst, tabl = tabl)
+        Sys.sleep(secs)
+    }
+}

### 2017-05-18 R/mean_pch.R
refresh

@@ -1,0 +1,8 @@
+mean_pch <- function(cuts, levels){
+    ## Mean of a piecewise constant hazards (PCH) distribution
+
+    library(eha)
+    fn <- function(x) ppch(x, cuts, levels, lower.tail = FALSE)
+
+    integrate(fn, lower = 0, upper = Inf)$value
+}

### 2017-05-18 R/plot_surv.R
refresh

@@ -1,0 +1,43 @@
+plot_surv <- function(pop, cols = c("darkgreen", "black", "red", "blue"), sex = "Men"){
+    pp <- levels(pop$period)
+    n <- length(pp)
+    hc <- levels(pop$hisclass)
+    m <- length(hc)
+    res <- matrix(0, nrow = n, ncol = 10)
+    cuts <- seq(5, 45, by = 5)
+    oldpar <- par(mfrow = c(2, n / 2))
+    for (i in 1:n){
+        ##cat("i = ", i, "\n")
+        pim <- pop[pop$period == pp[i], ]
+        for (j in 1:m){
+            fit <- glm(event ~ offset(log(exposure)) + age + urban * civst, 
+                       data = pim[pim$hisclass == hc[j], ], family = poisson)
+            
+            atoms <- c(fit$coef[1], fit$coef[substr(names(fit$coef), 1, 3) == "age"])
+            nl <- length(atoms)                                 
+            atoms[2:nl] <- atoms[2:nl] + atoms[1]
+            if (nl < 10){
+                klipp <- cuts[1:(nl - 1)]
+            }else{
+                klipp <- cuts
+            }
+            
+            levels <- exp(atoms)
+            x <- seq(40, klipp[length(klipp)] + 45, length = 20)
+            y <- ppch(x - 40, cuts = klipp, levels = levels, lower.tail = FALSE)
+            if (j == 1){
+                plot(x, y, type = "l", col = cols[1], ylim = c(0, 1), xlim = c(40, 90),
+                     xlab = "Age", ylab = "", main = paste(sex, pp[i]), axes = FALSE)
+                axis(1)
+                axis(2, at = c(0.0, 0.5, 1.0), las = 1)
+            }else{
+                lines(x, y, col = cols[j])
+            }
+            abline(h = 0.5, lty = 3)
+            abline(h = 0)
+            legend("bottomleft", legend = hc, col = cols, lty = 1)
+        }
+    }
+    par(oldpar)
+    
+}

### 2017-05-18 slides2.Rnw
refresh

\documentclass[t]{beamer}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{natbib}

%\usepackage[figuresonly]{endfloat}

%\bibliographystyle{apalike}

\usepackage{color}

\newcommand{\emp}[1]{\textcolor{blue}{#1}}

\usetheme{Singapore}% best choice.
  \setbeamercovered{transparent}%

\newcommand{\R}{{\bf R}}
\newcommand{\code}[1]{{\tt #1}}

\title{Social inequality in mortality among adults and elderly in
  northern  Sweden 1851--2013\footnote{Seminar April 27, CEDAR, Umeå}
}

\date{April 27, 2017}

\author{Göran Broström and Sören Edvinsson}

\begin{document}

\maketitle

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

\begin{frame}{Questions}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased in the adult and elderly
  population?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any \emp{gender differences} in the effect of social position?

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively?
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?

\end{enumerate}

\end{frame}

%%\begin{frame}{Where?}

%%\includegraphics[height=5cm]{myfigs/geo_we.pdf}\includegraphics[height=7cm]{myfigs/sweden.pdf}
%%\end{frame}

\begin{frame}{Here!}

\begin{columns}

\begin{column}{0.5\textwidth}

\includegraphics[height=7cm]{myfigs/skum17.pdf}

\end{column}

\begin{column}{0.5\textwidth}

Skellefteå:

\begin{itemize}

\item Small town, heavy industry in the nineteenth century.

\end{itemize}

Umeå:

\begin{itemize}

\item Not so small town, older. Center for administration, education, and military.

\end{itemize}

Rural areas:

\begin{itemize}

\item Very similar: Small farms dominating.

\end{itemize}

\end{column}
\end{columns}
%\includegraphics[height=7cm]{myfigs/sweden.pdf}

\end{frame}

\begin{frame}[fragile]{Data sources}

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}
  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{The sampling frame}

<<lexis2>>=
load("data/vb0.rda")
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1985, 70, "Linnaeus", cex = 2, col = "blue")
text(1984, 60, paste(n.rec, "records"))
text(1985.2, 55, paste(n.pl, "persons"))
text(1985.2, 50, paste(n.d, "deaths"))
@
\end{frame}

\begin{frame}[t]{The FOB and Lisa data}

\vspace{-1.5cm}
%\begin{columns}
%\begin{column}{0.6\textwidth}
<<picklexus,fig.width=4.7,fig.height = 4.8, echo=FALSE>>=

##load("data/vb.rda")  ####################################### NOTE!!!!!!!!!!!!!!!! ############################
load("data/vb0.rda")
load("data/vb.rda")

tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")
x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE, ylim = c(28, 90))
axis(1, at = c(seq(1960, 1990, by = 10), 2002, 2014), las = 1)
axis(2, at = c(40, 65, 90), las = 2)
axis(4, at = c(40, 65, 90), las = 2)
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    ##abline(v = i, lwd = 2, col = "red")
    lines(c(i, i), c(40, 90), lwd = 2, col = "red")
}
for (i in seq(1960, 1985, by = 5) + (365 - 61) / 365){
    lines(c(i, i + 5), c(35, 40), col = "red", lty = 2, lwd = 2)
    lines(c(i, i), c(35, 40), col = "red", lty = 2, lwd = 2)
}
lines(c(1990, 1990) + (365 - 61)/365, c(28, 40), lwd = 2, lty = 2, col = "red")
lines(c(1990 + (365 - 61) / 365, 2002), c(28, 40), lwd = 2, lty = 2, col = "red")
for (i in 2002:2014){
    ##abline(v = i, lwd = 2, col = "blue")
    lines(c(i, i), c(40, 90), col = "blue", lwd = 2)
}
for (i in 2002:2013){
    lines(c(i, i), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
    lines(c(i, i+1), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
}
lines(c(2002, 2014), c(40, 40), col = "blue", lwd = 1.5)
lines(c(2002, 2014), c(90, 90), col = "blue", lwd = 1.5)
lines(c(1960, 2002), c(90, 90), lwd = 2, col = "red")
lines(c(1960, 2002), c(40, 40), lwd = 2, col = "red")
##abline(h = c(35, 40, 65, 90), lty = 2)
points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
     pch = ".", col = "red")
box()
@

%\end{column}

%\begin{column}{0.4\textwidth}
%\scriptsize

%\vspace{3\baselineskip}

%\begin{tabular}{l|rr}
%      & Number & Deaths \\ \hline
%\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
%\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
%Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
%\end{tabular}

%\vspace{\baselineskip}

%\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are %\emp{excluded}.
%\end{column}
%\end{columns}
\end{frame}

\begin{frame}{The Linnaeus data with deaths}

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus2,fig.width=6,fig.height = 5.9>>=

##load("data/vb.rda")
tot.m <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "male"]))
tot.f <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "female"]))

vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")

x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE)
axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
axis(2, at = c(40, 65, 90))
axis(4, at = c(40, 65, 90))
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    abline(v = i, lwd = 2)
}
for (i in 2002:2013){
    abline(v = i, lwd = 2, col = "black")
}
abline(h = c(40, 65, 90), lty = 2)
points(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
     pch = ".", col = "blue")
box()
@

\end{column}

\begin{column}{0.4\textwidth}
\scriptsize

\vspace{3\baselineskip}

\begin{tabular}{l|rr}
      & Number & Deaths \\ \hline
\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
\end{tabular}

\vspace{\baselineskip}

\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{included}.
\end{column}
\end{columns}
\end{frame}

\begin{frame}{The Linnaeus data with deaths}

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus3,fig.width=6,fig.height = 5.9>>=

load("data/vb.rda")
tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))

vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")

x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE)
axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
axis(2, at = c(40, 65, 90))
axis(4, at = c(40, 65, 90))
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    abline(v = i, lwd = 2)
}
for (i in 2002:2013){
    abline(v = i, lwd = 2, col = "black")
}
abline(h = c(40, 65, 90), lty = 2)
points(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
     pch = ".", col = "blue")
box()
@

\end{column}

\begin{column}{0.4\textwidth}
\scriptsize

\vspace{3\baselineskip}

\begin{tabular}{l|rr}
      & Number & Deaths \\ \hline
\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
\end{tabular}

\vspace{\baselineskip}

\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
\end{column}
\end{columns}
\end{frame}

\begin{frame}{The sampling frame, II}

<<lexis4>>=
##load("data/vb0.rda")
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 1986, 2014, 2014, 1961), c(40, 65, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1989, 70, "Linnaeus", cex = 2, col = "blue")
text(1988, 60, paste(n.rec, "records"))
text(1989.2, 55, paste(n.pl, "persons"))
text(1989.2, 50, paste(n.d, "deaths"))
@
\end{frame}

\begin{frame}{Variables}

\begin{description}

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[\sc Hisclass] elite, middle, worker.
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, seven categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

\end{description}

\end{frame}

\begin{frame}{\sc Hisclass}

\begin{enumerate}
\item \emp{\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
\item \emp{\tt middle}, HISCLASS 3, 4, 5, 6, and 8. Lower managers, lower white collar, and farmers.
%%\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
\item \emp{\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
\end{enumerate}

\end{frame}

\begin{frame}[fragile]{Preliminary Cox regression}

{\scriptsize

<<prelanal,echo=TRUE>>=
library(eha)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass)
system.time(fit <- coxph(Surv(enter, exit, event) ~
                             hisclass * period + civst + urban + sex, 
                         data = vb0))
system.time(dr <- drop1(fit, test = "Chisq"))
@
}

\emp{It takes time (more than 10 minutes)!}

\end{frame}

\begin{frame}{ANOVA}

<<respre,results='asis',echo=FALSE>>=
library(xtable)
xtable(dr)
@

And with so much data, everything tends to be \emp{statistically} significant.

\end{frame}

\begin{frame}[fragile]{Test of Proportional Hazards}

<<prophaz>>=
system.time(cz <- cox.zph(fit))
@

Another \emp{19 minutes}!
\end{frame}

\begin{frame}[fragile]{Proportional hazards, results}

<<rezph,echo = TRUE>>=
#system.time(cz <- cox.zph(fit))
cz$table[NROW(cz$table), ]
@

As usual, \emp{highly significant} result (PH assumption violated).

\end{frame}

\begin{frame}[fragile]{Sufficient statistics}

Aggregate:

{\scriptsize
<<aggregate,eval = TRUE, echo = TRUE>>=
vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
                   vb0[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass")], 
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
nr <- NROW(vb0Tab)
@
}

\begin{itemize}
\item Gives a table assuming \emp{piecewise constant hazards} on five-year intervals: 40--45, 45--50, \ldots, 85--90 ("age").

\item Number of cells: $2 \times 10 \times 10 \times 20 \times 2 \times 3 \times 4 = $ \emp{96000}

\item Number of \emp{non-empty} cells: \emp{\Sexpr{nr}}.

\end{itemize}

\end{frame}

\begin{frame}[fragile]{PH regression with the table}

\begin{itemize}
\item Piecewise constant hazard model generates the same likelihood as a Poisson model.
\begin{itemize}
\item Does \emp{not} imply that the number of deaths follow a Poisson distribution!
\end{itemize}
\end{itemize}

\small

<<coxpois,echo=TRUE,cache=FALSE>>=
system.time(fitp <- glm(event ~ offset(log(exposure)) + age +
                hisclass * period + civst + urban + sex,
            data = vb0Tab, family = poisson))
system.time(dr <- drop1(fitp, test = "Chisq"))
@

\end{frame}

\begin{frame}{ANOVA with the table}

<<anovatable,results='asis',echo=FALSE>>=
xtable(dr)
@

\end{frame}

\begin{frame}{Age--Period--Cohort data}

\vspace{-1.5cm}

<<moddat,fig.height=4.5,fig.width=5>>=
x <- c(1961, 2014)
y <- c(40, 40)
plot(x, y, type = "l", col = "blue", lwd = 2, xlim = c(1960, 2015), ylim = c(40, 90),
     ylab = "Age", xlab = "Year", axes = FALSE)
axis(1, at = c(1961, 1991, 2014))
axis(2, las = 1)
box()
lines(c(1961, 1961), c(40, 65), col = "blue", lwd = 2)
lines(c(2014, 2014), c(40, 90), col = "blue", lwd = 2)
lines(c(1961, 1986), c(65, 90), col = "blue", lwd = 2)
lines(c(1986, 2014), c(90, 90), col = "blue", lwd = 2)
#
lines(c(1971, 1971), c(40, 75), col = "blue", lty = 2)
lines(c(1981, 1981), c(40, 85), col = "blue", lty = 2)
lines(c(1991, 1991), c(40, 90), col = "blue", lty = 2)
lines(c(2002, 2002), c(40, 90), col = "blue", lty = 2)
lines(c(2008, 2008), c(40, 90), col = "blue", lty = 2)
#
for (age in seq(40, 65, by = 5)){
    lines(c(1961, 2014), c(age, age), col = "blue", lty = 2)
}
for(i in 0:3){
    age <- 70 + 5 * i
    lines(c(1961 + 5 * (i + 1), 2014), c(age, age), col = "blue", lty = 2)
}
#
lines(c(1961, 1961 + 30), c(60, 90), lty = 2, col = "red")
lines(c(1961, 1961 + 40), c(50, 90), lty = 2, col = "red")
lines(c(1961, 1961 + 50), c(40, 90), lty = 2, col = "red")
lines(c(1971, 2014), c(40, 40 + 43), lty = 2, col = "red")
lines(c(1981, 2014), c(40, 40 + 33), lty = 2, col = "red")
lines(c(1991, 2014), c(40, 40 + 23), lty = 2, col = "red")
lines(c(2001, 2014), c(40, 40 + 13), lty = 2, col = "red")
lines(c(2011, 2014), c(40, 40 + 3), lty = 2, col = "red")
@

\end{frame}

\begin{frame}[fragile]{Counting "cells"}

<<cremodern,echo=FALSE>>=
library(eha)
vb0$hisclass[is.na(vb0$hisclass)] <- 4 ## NOTE!!
mode <- cal.window(vb0, c(1961, 2014))
modern <- mode[mode$birthdate > 1896, ]
modern$hisclass <- factor(modern$hisclass)
modern$period <- factor(modern$period)
moTab <- aggregate(modern[, c("event", "exposure")],
                   modern[, c("sex", "age", "period", "cohort",
                              "urban", "civst", "hisclass")], 
                   FUN = sum)
moTab$period <- factor(moTab$period)
moTab$cohort <- factor(moTab$cohort)
moTab$age <- factor(moTab$age)
moTab$hisclass <- factor(moTab$hisclass)
@

\scriptsize

\begin{itemize}
\item There are \emp{115 cells in one figure}
\item There are $2 \times 2 \times 3 \times 4 = $ \emp{48 such figures}.
\item In total: $115 \times 48 = $ \emp{5520 combinations}.
\item Number of \emp{non-empty} combinations: \emp{4072}.
\end{itemize}

<<heada>>=
head(moTab)
@

\end{frame}

\begin{frame}{Period analysis, men}

<<coxper>>=
par(mfrow = c(2, 3))
men <- modern[modern$sex == "male", ]
levels(men$hisclass) <- c("elite", "middle", "worker", "NA")
cols = c("black", "green", "red", "blue")
for (per in levels(men$period)){
    min <- men[men$period == per, ]
   fit <- coxreg(Surv(enter, exit, event) ~ strata(hisclass) + urban + civst, data = min)
   plot(fit, col = cols, main = per, fn = "surv", printLegend = FALSE, xlab = "Age", xlim = c(40, 90))
   legend("bottomleft", legend = levels(min$hisclass), col = cols, lty = 1)
}
@

\end{frame}
\begin{frame}{Proportional hazards for women?}

\includegraphics[width=11cm,height=8cm]{figure/hazplotw-1}

\end{frame}

\begin{frame}{Proportional hazards for men?}

\includegraphics[width=11cm,height=8cm]{figure/hazplotm-1}

\end{frame}

\begin{frame}[fragile]{Interactions}

\begin{verbatim}
fit <- coxph(Surv(enter - 40, exit - 40, event) ~
    period * hisclass * sex + urban, data = vb0)
drop1(fit, test = "Chisq")
\end{verbatim}

\begin{center}
\begin{tabular}{lrrrr}
  \hline
 & Df & AIC & LRT & Pr($>$Chi) \\
  \hline
$<$none$>$ &  & 1631901.32 &  &  \\
  urban & 1 & 1631946 & 47.04 & $7.0\times 10^{-12}$ \\
  \emp{period:hisclass:sex} & 27 & 1631957 & 110.11 & \emp{$5.3\times 10^{-12}$} \\
   \hline
\end{tabular}
\end{center}

\begin{itemize}

\item This model is described by \emp{80 coefficients}.

\item We show them in \emp{graphs}.
\end{itemize}

\end{frame}

\begin{frame}{No proportional hazards (wrt \sc Hisclass)!}

\begin{itemize}
\item Fit each {\sc Hisclass} \emp{separately}, for each time period.
\item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
\end{itemize}

So, we fit separate PH models for each combination of

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
  \item \emp{\tt hisclass}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}
\end{frame}

\begin{frame}{The data table}

<<lexis3, fig.height = 4>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(40, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
for (i in c(1851, 1876, 1901, 1926, 1971, 1981, 1991, 2002, 2007)){
    lines(c(i, i), c(40, 90), lty = 2)
}
for (i in c(45, 50, 55, 60, 65, 70, 75, 80, 85)){
    lines(c(1851, 1951), c(i, i), lty = 2)
    lines(c(1961, 2014), c(i, i), lty = 2)
}
@

\emp{Poisson regression}:

In \emp{each cell}, and for \emp{each} combination of \emp{covariate level}s, calculate
\begin{itemize}
\item  the number of \emp{deaths} and
\item the total \emp{exposure time}.
\end{itemize}
\end{frame}

\begin{frame}{Women and men, age 40--64}

  \begin{center}
\includegraphics[height = 7.5cm]{figure/cumsw-1}
  \end{center}
\end{frame}

\begin{frame}{Relative differences, ages 40--64 and 65--89}

  %\begin{center}
\includegraphics[height = 7.5cm]{figure/cumswstand-1}\includegraphics[height=7.5cm]{figure/cumswold-1}
   %\end{center}
\end{frame}

\begin{frame}{Cancer mortality, ages 40-64 and 65-89}

\includegraphics[height=7.5cm]{figure/12cumsw-1}\includegraphics[height=7.5cm]{figure/12cumswold-1}

\end{frame}

\begin{frame}{Cardiovascular mortality}

\includegraphics[height=7.5cm]{figure/41cumsw-1}\includegraphics[height=7.5cm]{figure/41cumswold-1}

\end{frame}

\begin{frame}{Education and income, 1990--2005, all deaths}

\includegraphics[height=3.5cm]{figure/liswomfig-1} Women, 40--64.

\includegraphics[height=3.5cm]{figure/lismanfig-1} Men, 40--64.

\end{frame}

\begin{frame}{Education and income, 1990--2005, all deaths}

\includegraphics[height=3.5cm]{figure/liswomfig2-1} Women, 65--89.

\includegraphics[height=3.5cm]{figure/lismanfig2-1} Men, 65--89.

\end{frame}

\begin{frame}{Education and income, 1990--2005, cancer}

\includegraphics[height=3.5cm]{figure/12liswomfig-1} Women, 40--64.

\includegraphics[height=3.5cm]{figure/12lismanfig-1} Men, 40--64.

\end{frame}

\begin{frame}{Education and income, 1990--2005, cancer}

\includegraphics[height=3.5cm]{figure/12liswomfig2-1} Women, 65--89.

\includegraphics[height=3.5cm]{figure/12lismanfig2-1} Men, 65--89.

\end{frame}

\begin{frame}{Education and income, 1990--2005, cardiovascular}

\includegraphics[height=3.5cm]{figure/41liswomfig-1} Women, 40--64.

\includegraphics[height=3.5cm]{figure/41lismanfig-1} Men, 40--64.

\end{frame}

\begin{frame}{Education and income, 1990--2005, cardiovascular}

\includegraphics[height=3.5cm]{figure/41liswomfig2-1} Women, 65--89.

\includegraphics[height=3.5cm]{figure/41lismanfig2-1} Men, 65--89.

\end{frame}

%%% Stop here!

\begin{frame}{Answers(?)}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased?
\begin{itemize}
\item Not really
\end{itemize}
%%\item Is the mortality transition initiated by a certain social class?

\item Are there any \emp{gender differences} in the effect of social position?
\begin{itemize}
\item Yes
\end{itemize}

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\begin{itemize}
\item Not really
\end{itemize}

\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively?
\begin{itemize}
\item Yes
\end{itemize}
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?
\begin{itemize}
\item Yes, definitely.
\end{itemize}

\end{enumerate}

\end{frame}

\begin{frame}{Conclusion}

%\scriptsize

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item There are \emp{gender differences} in the social pattern
\item No difference in the social pattern in the two studied age groups.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item \emp{Education} and \emp{income} are the \emp{most important}
  variables during late
  time period.
\end{itemize}

\end{frame}

\begin{frame}{Software for report writing and analysis}

\begin{itemize}
\item  \R : \emp{\url{https://www.R-project.org}}.
  \begin{itemize}
    \item \emp{RStudio}: \url{https://www.rstudio.com}.
  \end{itemize}
\item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
  \begin{itemize}
  \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
  \item \emp{rmarkdown} (RStudio Team, 2016).
  \item \emp{knitr} (Xie, 2016).
  \item \emp{eha} (Broström, 2016).
  \item \emp{xtable} (Dahl, 2016).
\end{itemize}

\item \emp{Version control} (reproducibility)
  \begin{itemize}
    \item \emp{git}: \url{http://git}
    \item \emp{gitHub}: \url{http://github.com/goranbrostrom/}

\end{itemize}
\end{itemize}

\end{frame}

\end{document}

### 2017-05-19 CreateData/creDat.Rnw
from mbp

@@
 <<setup, include = FALSE>>=
-knitr::opts_chunk$set(cache=FALSE,echo=TRUE,message=TRUE,warning=TRUE,comment=NA)##$
+knitr::opts_chunk$set(cache=TRUE,echo=TRUE,message=TRUE,warning=TRUE,comment=NA)##$
 @
@@
     source("R/putONssyk.R")
-    putONssyk
+    ##putONssyk
     fobb <- putONssyk(fobdata)
     source("R/ssykTOhisclass.R")
-    ssykTOhisclass
+    ##ssykTOhisclass
     fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
@@
 <<aggregate>>=
-vb$poplink <- vb$dataset == "poplink"
 vbTab <- aggregate(vb[, c("event", "exposure")], 
                    vb[, c("sex", "age", "period", "cohort", 
-                          "urban", "civst", "hisclass40", "hisclass60", "poplink")], 
-                   FUN = sum)
-
-vb0$hisclass[is.na(vb0$hisclass)] <- 0
-vb0$poplink <- vb0$dataset == "poplink"
-vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
-                   vb0[, c("sex", "age", "period", "cohort", 
-                          "urban", "civst", "hisclass", "poplink")], 
+                          "urban", "civst", "hisclass40", "hisclass60")], 
                    FUN = sum)
@@
 save(vbTab, file = "../data/vbTab.rda")
-save(vb0Tab, file = "../data/vb0Tab.rda")
 @
 
-Thus, the one data file to use is {\tt vbTab.rda} or {\tt vb0Tab.rda}.
+Thus, the one data file to use is {\tt vbTab.rda}.
 
@@
 save(loTab, file = "../data/loTab.rda")
-save(lisaold, file = "../data/lisaold.rda")
 @

### 2017-05-19 CreateData/R/ssykTOhisclass.R

@@ -1,13 +1,17 @@
 ssykTOhisclass <- function(x){
+
+    ## Change 2017-03-28: Three instead of four hisclasses!
+    
     ## 'x' is a vector of ssyk codes.
     ## as output we want a vector of hisclass codes (7 levels)
     
     ## Step 1: ssyk to 12-levels hisclass
-    load("../../../Data/shh.rda")
+    load("../../Data/shh.rda")
     indx <- match(x, shh$ssyk)
     tmp <- shh$hisclass[indx]
 
     ## Step 2: go from 12 to 4 levels:  ## NEW !!
+    ## Step 2: go from 12 to 3 levels:  ## NEWER !!
     
     
     res <- rep(-1, length(x))
@@ -15,9 +19,9 @@ ssykTOhisclass <- function(x){
     tmp[is.na(tmp)] <- -1
     
     res[tmp %in% 1:2] <- 1
-    res[tmp %in% c(3, 8)] <- 2
-    res[tmp %in% 4:6] <- 3
-    res[tmp %in% c(7, 9, 10, 11, 12)] <- 4
+    res[tmp %in% c(3:6, 8)] <- 2
+    ##res[tmp %in% 4:6] <- 3
+    res[tmp %in% c(7, 9, 10, 11, 12)] <- 3 # '4' ---> '3'
     is.na(res) <- res == -1
     
     res

### 2017-05-19 R/plotED.glm.R

@@ -1,6 +1,6 @@
 plotED.glm <- function(res,
                        col = 1:NROW(res[[1]]),
-                       logScale = FALSE,
+                       logScale = TRUE,
                        relative = FALSE,
                        ylab = "",
                        main = "",
@@ -12,7 +12,8 @@ plotED.glm <- function(res,
     ##nper <- NCOL(res[[1]])
 
     ## Assuming 'm == 4', we try:
-    leg = c("elite", "lowMan", "lowWhiteC", "worker")
+    ##leg = c("elite", "lowMan", "lowWhiteC", "worker")
+    leg <- rownames(res[[1]])
     nper <- length(res)
     k <- NCOL(res[[1]])
     m <- NROW(res[[1]])
@@ -27,7 +28,9 @@ plotED.glm <- function(res,
     }
     
     if (logScale){
-        outp <- log(outp)
+        ##outp <- log(outp)
+        outp <- log(1 - exp(-outp))
+        ##outp <- log(outp / (1 - outp))
     }
     if (relative){
         outp <- prop.table(outp, 2)

### 2017-05-19 slides2.Rnw
from mbp

@@
-\documentclass[t]{beamer}
+\documentclass[t,a4paper,11pt]{beamer}
 
@@
 <<lexis2>>=
-load("data/vb0.rda")
 plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
@@
       col = "blue", lwd = 2)
-#
-n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
-text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(n.rec, "records"))
-text(1900, 55, paste(n.pl, "persons"))
-text(1899.5, 50, paste(n.d, "deaths"))
-#
-n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
-n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
-
-text(1985, 70, "Linnaeus", cex = 2, col = "blue")
-text(1984, 60, paste(n.rec, "records"))
-text(1985.2, 55, paste(n.pl, "persons"))
-text(1985.2, 50, paste(n.d, "deaths"))
+text(1900, 65, "PopLink", cex = 2, col = "red")
+text(1900, 55, "79593 persons")
+text(1900, 50, "32801 deaths")
+text(1985, 65, "Linnaeus", cex = 2, col = "blue")
+text(1985, 55, "156676 persons")
+text(1985, 50, "42879 deaths")
 @
@@
 
-\begin{frame}[t]{The FOB and Lisa data}
+\begin{frame}{The Linnaeus data}
 
-\vspace{-1.5cm}
-%\begin{columns}
-%\begin{column}{0.6\textwidth}
-<<picklexus,fig.width=4.7,fig.height = 4.8, echo=FALSE>>=
+\begin{columns}
+\begin{column}{0.6\textwidth}
+<<picklexus,fig.width=6,fig.height = 5.9>>=
 
@@
 load("data/vb0.rda")
-load("data/vb.rda")
+vb <- vb0
 
@@
 tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
-vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
+vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
 d.m <- sum(vbd$sex == "male")
@@
      pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
-     axes = FALSE, ylim = c(28, 90))
-axis(1, at = c(seq(1960, 1990, by = 10), 2002, 2014), las = 1)
-axis(2, at = c(40, 65, 90), las = 2)
-axis(4, at = c(40, 65, 90), las = 2)
+     axes = FALSE)
+axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
+axis(2, at = c(40, 65, 90))
+axis(4, at = c(40, 65, 90))
 for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
-    ##abline(v = i, lwd = 2, col = "red")
-    lines(c(i, i), c(40, 90), lwd = 2, col = "red")
-}
-for (i in seq(1960, 1985, by = 5) + (365 - 61) / 365){
-    lines(c(i, i + 5), c(35, 40), col = "red", lty = 2, lwd = 2)
-    lines(c(i, i), c(35, 40), col = "red", lty = 2, lwd = 2)
-}
-lines(c(1990, 1990) + (365 - 61)/365, c(28, 40), lwd = 2, lty = 2, col = "red")
-lines(c(1990 + (365 - 61) / 365, 2002), c(28, 40), lwd = 2, lty = 2, col = "red")
-for (i in 2002:2014){
-    ##abline(v = i, lwd = 2, col = "blue")
-    lines(c(i, i), c(40, 90), col = "blue", lwd = 2)
+    abline(v = i, lwd = 2)
 }
 for (i in 2002:2013){
-    lines(c(i, i), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
-    lines(c(i, i+1), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
+    abline(v = i, lwd = 2, col = "black")
 }
-lines(c(2002, 2014), c(40, 40), col = "blue", lwd = 1.5)
-lines(c(2002, 2014), c(90, 90), col = "blue", lwd = 1.5)
-lines(c(1960, 2002), c(90, 90), lwd = 2, col = "red")
-lines(c(1960, 2002), c(40, 40), lwd = 2, col = "red")
-##abline(h = c(35, 40, 65, 90), lty = 2)
+abline(h = c(40, 65, 90), lty = 2)
 points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
@@
 
-%\end{column}
+\end{column}
 
-%\begin{column}{0.4\textwidth}
-%\scriptsize
+\begin{column}{0.4\textwidth}
+\scriptsize
 
-%\vspace{3\baselineskip}
+\vspace{3\baselineskip}
 
-%\begin{tabular}{l|rr}
-%      & Number & Deaths \\ \hline
-%\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
-%\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
-%Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
-%\end{tabular}
+\begin{tabular}{l|rr}
+      & Number & Deaths \\ \hline
+\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
+\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
+Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
+\end{tabular}
 
-%\vspace{\baselineskip}
+\vspace{\baselineskip}
 
-%\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are %\emp{excluded}.
-%\end{column}
-%\end{columns}
+\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
+\end{column}
+\end{columns}
 \end{frame}
 
-\begin{frame}{The Linnaeus data with deaths}
+\begin{frame}{The Linnaeus data}
 
@@
 ##load("data/vb.rda")
-tot.m <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "male"]))
-tot.f <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "female"]))
+tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
+tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
 
-vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
+vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
 d.m <- sum(vbd$sex == "male")
@@
 
-\begin{frame}{The Linnaeus data with deaths}
+\begin{frame}{The Linnaeus data}
 
@@
 
-\begin{frame}{The sampling frame, II}
-
-<<lexis4>>=
-##load("data/vb0.rda")
-plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
-     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
-     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
-axis(2, las = 2, at = c(0, 40, 65, 90))
-axis(4, las = 2, at = c(0, 40, 65, 90))
-box()
-abline(h = 0)
-lines(c(1961, 1961, 1986, 2014, 2014, 1961), c(40, 65, 90, 90, 40, 40),
-      col = "blue", lwd = 2)
-#
-n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
-text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(n.rec, "records"))
-text(1900, 55, paste(n.pl, "persons"))
-text(1899.5, 50, paste(n.d, "deaths"))
-#
-n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
-n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
-
-text(1989, 70, "Linnaeus", cex = 2, col = "blue")
-text(1988, 60, paste(n.rec, "records"))
-text(1989.2, 55, paste(n.pl, "persons"))
-text(1989.2, 50, paste(n.d, "deaths"))
-@
-\end{frame}
-
 \begin{frame}{Variables}
@@
 \item[sex]
-\item[birth date] By quarter (15 Feb, May, Aug, Nov)
-\item[death date] By day (\emp{Socialstyrelsen})
-\item[cause of death] from \emp{Socialstyrelsen}
-\item[parish] Urban/Rural
-\item[\sc Hisclass] elite, middle, worker.
+\item[birth date]
+\item[death date]
+\item[cause of death]
+\item[parish]
+\item[\sc Hisclass] grouped.
 \item[income] 1990--2005, grouped by quartiles.
-\item[education] 1990--2005, seven categories.
-\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
-\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).
+\item[education] 1990--2005. Seven categories.
 
@@
 
-\begin{frame}[fragile]{Preliminary Cox regression}
-
-{\scriptsize
-
-<<prelanal,echo=TRUE>>=
-library(eha)
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass)
-system.time(fit <- coxph(Surv(enter, exit, event) ~
-                             hisclass * period + civst + urban + sex, 
-                         data = vb0))
-system.time(dr <- drop1(fit, test = "Chisq"))
-@
-}
-
-\emp{It takes time (more than 10 minutes)!}
-
-\end{frame}
-
-\begin{frame}{ANOVA}
-
-<<respre,results='asis',echo=FALSE>>=
-library(xtable)
-xtable(dr)
-@
-
-And with so much data, everything tends to be \emp{statistically} significant.
-
-\end{frame}
-
-\begin{frame}[fragile]{Test of Proportional Hazards}
-
-<<prophaz>>=
-system.time(cz <- cox.zph(fit))
-@
-
-Another \emp{19 minutes}!
-\end{frame}
-
-\begin{frame}[fragile]{Proportional hazards, results}
-
-<<rezph,echo = TRUE>>=
-#system.time(cz <- cox.zph(fit))
-cz$table[NROW(cz$table), ]
-@
-
-As usual, \emp{highly significant} result (PH assumption violated).
-
-\end{frame}
-
-\begin{frame}[fragile]{Sufficient statistics}
-
-Aggregate:
-
-{\scriptsize
-<<aggregate,eval = TRUE, echo = TRUE>>=
-vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
-                   vb0[, c("sex", "age", "period", "cohort", 
-                          "urban", "civst", "hisclass")], 
-                   FUN = sum)
-vb0Tab$age <- factor(vb0Tab$age)
-nr <- NROW(vb0Tab)
-@
-}
-
-\begin{itemize}
-\item Gives a table assuming \emp{piecewise constant hazards} on five-year intervals: 40--45, 45--50, \ldots, 85--90 ("age").
-
-\item Number of cells: $2 \times 10 \times 10 \times 20 \times 2 \times 3 \times 4 = $ \emp{96000}
-
-\item Number of \emp{non-empty} cells: \emp{\Sexpr{nr}}.
-
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]{PH regression with the table}
-
-\begin{itemize}
-\item Piecewise constant hazard model generates the same likelihood as a Poisson model.
-\begin{itemize}
-\item Does \emp{not} imply that the number of deaths follow a Poisson distribution!
-\end{itemize}
-\end{itemize}
-
-\small
-
-<<coxpois,echo=TRUE,cache=FALSE>>=
-system.time(fitp <- glm(event ~ offset(log(exposure)) + age +
-                hisclass * period + civst + urban + sex,
-            data = vb0Tab, family = poisson))
-system.time(dr <- drop1(fitp, test = "Chisq"))
-@
-
-\end{frame}
-
-\begin{frame}{ANOVA with the table}
-
-<<anovatable,results='asis',echo=FALSE>>=
-xtable(dr)
-@
-
-\end{frame}
-
-\begin{frame}{Age--Period--Cohort data}
-
-\vspace{-1.5cm}
-
-<<moddat,fig.height=4.5,fig.width=5>>=
-x <- c(1961, 2014)
-y <- c(40, 40)
-plot(x, y, type = "l", col = "blue", lwd = 2, xlim = c(1960, 2015), ylim = c(40, 90),
-     ylab = "Age", xlab = "Year", axes = FALSE)
-axis(1, at = c(1961, 1991, 2014))
-axis(2, las = 1)
-box()
-lines(c(1961, 1961), c(40, 65), col = "blue", lwd = 2)
-lines(c(2014, 2014), c(40, 90), col = "blue", lwd = 2)
-lines(c(1961, 1986), c(65, 90), col = "blue", lwd = 2)
-lines(c(1986, 2014), c(90, 90), col = "blue", lwd = 2)
-#
-lines(c(1971, 1971), c(40, 75), col = "blue", lty = 2)
-lines(c(1981, 1981), c(40, 85), col = "blue", lty = 2)
-lines(c(1991, 1991), c(40, 90), col = "blue", lty = 2)
-lines(c(2002, 2002), c(40, 90), col = "blue", lty = 2)
-lines(c(2008, 2008), c(40, 90), col = "blue", lty = 2)
-#
-for (age in seq(40, 65, by = 5)){
-    lines(c(1961, 2014), c(age, age), col = "blue", lty = 2)
-}
-for(i in 0:3){
-    age <- 70 + 5 * i
-    lines(c(1961 + 5 * (i + 1), 2014), c(age, age), col = "blue", lty = 2)
-}
-#
-lines(c(1961, 1961 + 30), c(60, 90), lty = 2, col = "red")
-lines(c(1961, 1961 + 40), c(50, 90), lty = 2, col = "red")
-lines(c(1961, 1961 + 50), c(40, 90), lty = 2, col = "red")
-lines(c(1971, 2014), c(40, 40 + 43), lty = 2, col = "red")
-lines(c(1981, 2014), c(40, 40 + 33), lty = 2, col = "red")
-lines(c(1991, 2014), c(40, 40 + 23), lty = 2, col = "red")
-lines(c(2001, 2014), c(40, 40 + 13), lty = 2, col = "red")
-lines(c(2011, 2014), c(40, 40 + 3), lty = 2, col = "red")
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Counting "cells"}
-
-<<cremodern,echo=FALSE>>=
-library(eha)
-vb0$hisclass[is.na(vb0$hisclass)] <- 4 ## NOTE!!
-mode <- cal.window(vb0, c(1961, 2014))
-modern <- mode[mode$birthdate > 1896, ]
-modern$hisclass <- factor(modern$hisclass)
-modern$period <- factor(modern$period)
-moTab <- aggregate(modern[, c("event", "exposure")],
-                   modern[, c("sex", "age", "period", "cohort",
-                              "urban", "civst", "hisclass")], 
-                   FUN = sum)
-moTab$period <- factor(moTab$period)
-moTab$cohort <- factor(moTab$cohort)
-moTab$age <- factor(moTab$age)
-moTab$hisclass <- factor(moTab$hisclass)
-@
-
-\scriptsize
-
-\begin{itemize}
-\item There are \emp{115 cells in one figure}
-\item There are $2 \times 2 \times 3 \times 4 = $ \emp{48 such figures}.
-\item In total: $115 \times 48 = $ \emp{5520 combinations}.
-\item Number of \emp{non-empty} combinations: \emp{4072}.
-\end{itemize}
-
-<<heada>>=
-head(moTab)
-@
-
-\end{frame}
-
-\begin{frame}{Period analysis, men}
-
-<<coxper>>=
-par(mfrow = c(2, 3))
-men <- modern[modern$sex == "male", ]
-levels(men$hisclass) <- c("elite", "middle", "worker", "NA")
-cols = c("black", "green", "red", "blue")
-for (per in levels(men$period)){
-    min <- men[men$period == per, ]
-   fit <- coxreg(Surv(enter, exit, event) ~ strata(hisclass) + urban + civst, data = min)
-   plot(fit, col = cols, main = per, fn = "surv", printLegend = FALSE, xlab = "Age", xlim = c(40, 90))
-   legend("bottomleft", legend = levels(min$hisclass), col = cols, lty = 1)
-}
-@
-
-\end{frame}
 \begin{frame}{Proportional hazards for women?}
@@
 fit <- coxph(Surv(enter - 40, exit - 40, event) ~
-    period * hisclass * sex + urban, data = vb0)
+    period * hisclass * sex + urban, data = vb)
 drop1(fit, test = "Chisq")

### 2017-05-19 icass.Rnw

@@
 
-\author{Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}} and
-    Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} \\
+\author{Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} and
+Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}}
+     \\
 Centre for Demographic and Ageing Research ({\sc Cedar})\\
@@
     at the
-    session 'Socoeconomic inequalities and mortality I: Europe'. 41st Annual Meeting of the Social Science History Association, 
-    Chicago, Illinois
-    November 17--20, 2016}}
-
-\date{November 19, 2016}
+    session 'Health and Mortality in Arctic Populations'. The Ninth International
+    Congress on Arctic Social Sciences,
+    Umeå,
+    June 8--12, 2017}}
+
+\date{May 18, 2017}
 
@@
 levels(vbTab$period)[8] <- "1991-2001"
-load("data/vb.rda")
-vb$period <- factor(vb$period)
-levels(vb$hisclass) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vb$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vb$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vb$period)[8] <- "1991-2001"
+load("data/vb0.rda")
+vb0$period <- factor(vb0$period)
+vb0$hisclass[is.na(vb0$hisclass)] <- 4
+vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
+vb <- vb0
 @
@@
 library(xtable)
-load("data/vb0.rda")
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
+##load("data/vb0.rda")
+##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
 x <- with(vb0, table(period, hisclass, useNA = "ifany"))

### 2017-05-19 icass2.Rnw

--- 2017-05-19 icass.Rnw
@@
 <<readtab>>=
-load("data/vbTab.rda") 
-vbTab$period <- factor(vbTab$period)
-levels(vbTab$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vbTab$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vbTab$period)[8] <- "1991-2001"
 load("data/vb0.rda")
-vb0$period <- factor(vb0$period)
 vb0$hisclass[is.na(vb0$hisclass)] <- 4
 vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
-vb <- vb0
+plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
+     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
+     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
+axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(2, las = 2, at = c(0, 40, 65, 90))
+axis(4, las = 2, at = c(0, 40, 65, 90))
+box()
+abline(h = 0)
+lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
+      col = "blue", lwd = 2)
+#
+n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
+n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
+n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
+text(1900, 70, "PopLink", cex = 2, col = "red")
+text(1899, 60, paste(n.rec, "records"))
+text(1900, 55, paste(n.pl, "persons"))
+text(1899.5, 50, paste(n.d, "deaths"))
+#
+n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
+n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
+n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
+
+text(1985, 70, "Linnaeus", cex = 2, col = "blue")
+text(1984, 60, paste(n.rec, "records"))
+text(1985.2, 55, paste(n.pl, "persons"))
+text(1985.2, 50, paste(n.d, "deaths"))
 @
@@
 library(xtable)
-##load("data/vb0.rda")
-##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
+load("data/vb0.rda")
+vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
 x <- with(vb0, table(period, hisclass, useNA = "ifany"))

### 2017-05-19 slides3.Rnw

\documentclass[t]{beamer}

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{natbib}

%\usepackage[figuresonly]{endfloat}

%\bibliographystyle{apalike}

\usepackage{color}

\newcommand{\emp}[1]{\textcolor{blue}{#1}}

\usetheme{Singapore}% best choice.
  \setbeamercovered{transparent}%

\newcommand{\R}{{\bf R}}
\newcommand{\code}[1]{{\tt #1}}

\title{Social inequality in mortality among adults and elderly in
  northern  Sweden 1851--2013\footnote{Seminar April 27, CEDAR, Umeå}
}

\date{April 27, 2017}

\author{Göran Broström and Sören Edvinsson}

\begin{document}

\maketitle

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

\begin{frame}{Questions}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased in the adult and elderly
  population?

%%\item Is the mortality transition initiated by a certain social class?

\item Are there any \emp{gender differences} in the effect of social position?

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively?
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?

\end{enumerate}

\end{frame}

\begin{frame}{Outline}

\begin{center}
\begin{enumerate}

\item \emp{Data} management
\item \emp{Modeling} considerations
\item \emp{Analysis}
\item \emp{Results}
\item \emp{Alternative} Results
\end{enumerate}

\end{center}

\end{frame}

\begin{frame}{Here!}

\begin{columns}

\begin{column}{0.5\textwidth}

\includegraphics[height=7cm]{myfigs/skum17.pdf}

\end{column}

\begin{column}{0.5\textwidth}

Skellefteå:

\begin{itemize}

\item Small town, heavy industry in the nineteenth century.

\end{itemize}

Umeå:

\begin{itemize}

\item Not so small town, older. Center for administration, education, and military.

\end{itemize}

Rural areas:

\begin{itemize}

\item Very similar: Small farms dominating.

\end{itemize}

\end{column}
\end{columns}
%\includegraphics[height=7cm]{myfigs/sweden.pdf}

\end{frame}

\begin{frame}[fragile]{Data sources}

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}
  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{The sampling frame}

<<lexis2>>=
load("data/vb0.rda")
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1985, 70, "Linnaeus", cex = 2, col = "blue")
text(1984, 60, paste(n.rec, "records"))
text(1985.2, 55, paste(n.pl, "persons"))
text(1985.2, 50, paste(n.d, "deaths"))
@
\end{frame}

\begin{frame}[t]{The FOB and Lisa data}

\vspace{-1.5cm}
%\begin{columns}
%\begin{column}{0.6\textwidth}
<<picklexus,fig.width=4.7,fig.height = 4.8, echo=FALSE>>=

##load("data/vb.rda")  ####################################### NOTE!!!!!!!!!!!!!!!! ############################
##load("data/vb0.rda")
load("data/vb.rda")

tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")
x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE, ylim = c(28, 90))
axis(1, at = c(seq(1960, 1990, by = 10), 2002, 2014), las = 1)
axis(2, at = c(40, 65, 90), las = 2)
axis(4, at = c(40, 65, 90), las = 2)
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    ##abline(v = i, lwd = 2, col = "red")
    lines(c(i, i), c(40, 90), lwd = 2, col = "red")
}
for (i in seq(1960, 1985, by = 5) + (365 - 61) / 365){
    lines(c(i, i + 5), c(35, 40), col = "red", lty = 2, lwd = 2)
    lines(c(i, i), c(35, 40), col = "red", lty = 2, lwd = 2)
}
lines(c(1990, 1990) + (365 - 61)/365, c(28, 40), lwd = 2, lty = 2, col = "red")
lines(c(1990 + (365 - 61) / 365, 2002), c(28, 40), lwd = 2, lty = 2, col = "red")
for (i in 2002:2014){
    ##abline(v = i, lwd = 2, col = "blue")
    lines(c(i, i), c(40, 90), col = "blue", lwd = 2)
}
for (i in 2002:2013){
    lines(c(i, i), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
    lines(c(i, i+1), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
}
lines(c(2002, 2014), c(40, 40), col = "blue", lwd = 1.5)
lines(c(2002, 2014), c(90, 90), col = "blue", lwd = 1.5)
lines(c(1960, 2002), c(90, 90), lwd = 2, col = "red")
lines(c(1960, 2002), c(40, 40), lwd = 2, col = "red")
##abline(h = c(35, 40, 65, 90), lty = 2)
points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
     pch = ".", col = "red")
box()
@

%\end{column}

%\begin{column}{0.4\textwidth}
%\scriptsize

%\vspace{3\baselineskip}

%\begin{tabular}{l|rr}
%      & Number & Deaths \\ \hline
%\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
%\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
%Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
%\end{tabular}

%\vspace{\baselineskip}

%\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are %\emp{excluded}.
%\end{column}
%\end{columns}
\end{frame}

\begin{frame}{The Linnaeus data with deaths}

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus2,fig.width=6,fig.height = 5.9>>=

##load("data/vb.rda")
tot.m <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "male"]))
tot.f <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "female"]))

vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")

x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE)
axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
axis(2, at = c(40, 65, 90))
axis(4, at = c(40, 65, 90))
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    abline(v = i, lwd = 2)
}
for (i in 2002:2013){
    abline(v = i, lwd = 2, col = "black")
}
abline(h = c(40, 65, 90), lty = 2)
points(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
     pch = ".", col = "blue")
box()
@

\end{column}

\begin{column}{0.4\textwidth}
\scriptsize

\vspace{3\baselineskip}

\begin{tabular}{l|rr}
      & Number & Deaths \\ \hline
\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
\end{tabular}

\vspace{\baselineskip}

\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{included}.
\end{column}
\end{columns}
\end{frame}

\begin{frame}{The Linnaeus data with deaths}

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus3,fig.width=6,fig.height = 5.9>>=

load("data/vb.rda")
tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))

vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")

x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE)
axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
axis(2, at = c(40, 65, 90))
axis(4, at = c(40, 65, 90))
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    abline(v = i, lwd = 2)
}
for (i in 2002:2013){
    abline(v = i, lwd = 2, col = "black")
}
abline(h = c(40, 65, 90), lty = 2)
points(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
     pch = ".", col = "blue")
box()
@

\end{column}

\begin{column}{0.4\textwidth}
\scriptsize

\vspace{3\baselineskip}

\begin{tabular}{l|rr}
      & Number & Deaths \\ \hline
\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
\end{tabular}

\vspace{\baselineskip}

\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
\end{column}
\end{columns}
\end{frame}

\begin{frame}{The sampling frame, II}

<<lexis4>>=
##load("data/vb0.rda")
vb0 <- vb0[vb0$dataset == "poplink" | 
               (vb0$dataset != "poplink" & vb0$birthdate > 1895), ]
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 1986, 2014, 2014, 1961), c(40, 65, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1989, 70, "Linnaeus", cex = 2, col = "blue")
text(1988, 60, paste(n.rec, "records"))
text(1989.2, 55, paste(n.pl, "persons"))
text(1989.2, 50, paste(n.d, "deaths"))
@
\end{frame}

\begin{frame}{Variables}

\begin{description}

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[\sc Hisclass] elite, middle, worker.
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, seven categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

\end{description}

\end{frame}

\begin{frame}{\sc Hisclass}

\begin{enumerate}
\item \emp{\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
\item \emp{\tt middle}, HISCLASS 3, 4, 5, 6, and 8. Lower managers, lower white collar, and farmers.
%%\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
\item \emp{\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
\end{enumerate}

\end{frame}

\begin{frame}[fragile]{Preliminary Cox regression}

{\scriptsize

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{library}\hlstd{(eha)}
\hlkwd{system.time}\hlstd{(fit} \hlkwb{<-} \hlkwd{coxph}\hlstd{(}\hlkwd{Surv}\hlstd{(enter, exit, event)} \hlopt{~}
                              \hlstd{hisclass} \hlopt{*} \hlstd{period} \hlopt{+} \hlstd{civst} \hlopt{+} \hlstd{urban} \hlopt{+} \hlstd{sex,}
                         \hlkwc{data} \hlstd{= vb0))}
\end{alltt}
\begin{verbatim}
   user  system elapsed 
151.360   0.444 151.806 
\end{verbatim}
\begin{alltt}
\hlkwd{system.time}\hlstd{(dr} \hlkwb{<-} \hlkwd{drop1}\hlstd{(fit,} \hlkwc{test} \hlstd{=} \hlstr{"Chisq"}\hlstd{))}
\end{alltt}
\begin{verbatim}
   user  system elapsed 
455.236   0.944 456.188 
\end{verbatim}
\end{kframe}
\end{knitrout}
}

\emp{It takes time (more than 10 minutes)!}

\end{frame}

\begin{frame}{ANOVA}

% latex table generated in R 3.3.3 by xtable 1.8-2 package
% Mon Apr 17 14:43:45 2017
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
 & Df & AIC & LRT & Pr($>$Chi) \\ 
  \hline
$<$none$>$ &  & 2227937.83 &  &  \\ 
  civst & 2 & 2228508.16 & 574.34 & 0.0000 \\ 
  urban & 1 & 2227962.96 & 27.13 & 0.0000 \\ 
  sex & 1 & 2231526.62 & 3590.79 & 0.0000 \\ 
  hisclass:period & 27 & 2228575.58 & 691.75 & 0.0000 \\ 
   \hline
\end{tabular}
\end{table}

And with so much data, everything tends to be \emp{statistically} significant.

\end{frame}

\begin{frame}[fragile]{Test of Proportional Hazards}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
    user   system  elapsed 
1063.928    0.916 1064.028 
\end{verbatim}
\end{kframe}
\end{knitrout}

Another \emp{19 minutes}!
\end{frame}

\begin{frame}[fragile]{Proportional hazards, results}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{#system.time(cz <- cox.zph(fit))}
\hlstd{cz}\hlopt{$}\hlstd{table[}\hlkwd{NROW}\hlstd{(cz}\hlopt{$}\hlstd{table), ]}
\end{alltt}
\begin{verbatim}
    rho   chisq         p
     NA 2064.07    0.0000 
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{itemize}
\item As usual, \emp{highly significant} result (PH assumption violated).
\item Should we worry?
\begin{itemize}
\item Yes
\item No
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}[fragile]{Sufficient statistics}

Aggregate:

{\scriptsize
<<aggregate,eval = TRUE, echo = TRUE>>=
vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
                   vb0[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass")], 
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
nr <- NROW(vb0Tab)
@
}

\begin{itemize}
\item Gives a table assuming \emp{piecewise constant hazards} on five-year intervals: 40--45, 45--50, \ldots, 85--90 ("age").

\item Number of cells: $2 \times 10 \times 10 \times 20 \times 2 \times 3 \times 4 = $ \emp{96000}

\item Number of \emp{non-empty} cells: \emp{\Sexpr{nr}}.

\end{itemize}

\url{http://capa.ddb.umu.se/ds/aggregate.html}

\end{frame}

\begin{frame}[fragile]{Exposure by hisclass, men}

<<exphiswom>>=
xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
who <- length(levels(vb0Tab$hisclass))
barplot(t(yy)[who:1, ], col = c("blue", "red", "green", "black"), las = 2, ylab = "Per cent")
@

\end{frame}

\begin{frame}[fragile]{Exposure by hisclass, women}

<<exphis>>=
xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
who <- length(levels(vb0Tab$hisclass))
barplot(t(yy)[who:1, ], col = c("blue", "red", "green", "black"), las = 2, ylab = "Per cent")
@

\end{frame}

\begin{frame}{Age--Period--Cohort data}

\vspace{-1.5cm}

<<moddat,fig.height=4.5,fig.width=5>>=
x <- c(1961, 2014)
y <- c(40, 40)
plot(x, y, type = "l", col = "blue", lwd = 2, xlim = c(1960, 2015), ylim = c(40, 90),
     ylab = "Age", xlab = "Year", axes = FALSE)
axis(1, at = c(1961, 1991, 2014))
axis(2, las = 1)
box()
lines(c(1961, 1961), c(40, 65), col = "blue", lwd = 2)
lines(c(2014, 2014), c(40, 90), col = "blue", lwd = 2)
lines(c(1961, 1986), c(65, 90), col = "blue", lwd = 2)
lines(c(1986, 2014), c(90, 90), col = "blue", lwd = 2)
#
lines(c(1971, 1971), c(40, 75), col = "blue", lty = 2)
lines(c(1981, 1981), c(40, 85), col = "blue", lty = 2)
lines(c(1991, 1991), c(40, 90), col = "blue", lty = 2)
lines(c(2002, 2002), c(40, 90), col = "blue", lty = 2)
lines(c(2008, 2008), c(40, 90), col = "blue", lty = 2)
#
for (age in seq(40, 65, by = 5)){
    lines(c(1961, 2014), c(age, age), col = "blue", lty = 2)
}
for(i in 0:3){
    age <- 70 + 5 * i
    lines(c(1961 + 5 * (i + 1), 2014), c(age, age), col = "blue", lty = 2)
}
#
lines(c(1961, 1961 + 30), c(60, 90), lty = 2, col = "red")
lines(c(1961, 1961 + 40), c(50, 90), lty = 2, col = "red")
lines(c(1961, 1961 + 50), c(40, 90), lty = 2, col = "red")
lines(c(1971, 2014), c(40, 40 + 43), lty = 2, col = "red")
lines(c(1981, 2014), c(40, 40 + 33), lty = 2, col = "red")
lines(c(1991, 2014), c(40, 40 + 23), lty = 2, col = "red")
lines(c(2001, 2014), c(40, 40 + 13), lty = 2, col = "red")
lines(c(2011, 2014), c(40, 40 + 3), lty = 2, col = "red")
@

\end{frame}

\begin{frame}[fragile]{Counting "cells"}

<<cremodern,echo=FALSE>>=
library(eha)
##vb0$hisclass[is.na(vb0$hisclass)] <- 4 ## NOTE!!
##mode <- cal.window(vb0, c(1961, 2014))
##modern <- mode[mode$birthdate > 1896, ]
##modern$hisclass <- factor(modern$hisclass)
##modern$period <- factor(modern$period)
##moTab <- aggregate(modern[, c("event", "exposure")],
  ##                 modern[, c("sex", "age", "period", "cohort",
    ##                          "urban", "civst", "hisclass")], 
      ##             FUN = sum)
moTab <- vb0Tab[vb0Tab$period %in% levels(vb0Tab$period)[7:10], ]
moTab$period <- factor(moTab$period)
moTab$cohort <- factor(moTab$cohort)
moTab$age <- factor(moTab$age)
moTab$hisclass <- factor(moTab$hisclass)
moTab$hisclass <- relevel(moTab$hisclass, ref = "middle")
moTab$civst <- relevel(moTab$civst, ref = "married")
##
olTab <- vb0Tab[vb0Tab$period %in% levels(vb0Tab$period)[2:4], ]
olTab$period <- factor(olTab$period)
olTab$cohort <- factor(olTab$cohort)
olTab$age <- factor(olTab$age)
olTab$hisclass <- factor(olTab$hisclass)
olTab$hisclass <- relevel(olTab$hisclass, ref = "middle")
olTab$civst <- relevel(olTab$civst, ref = "married")

@

\begin{itemize}
\item There are \emp{115 cells in one figure}
\item There are $2 \times 2 \times 3 \times 4 = $ \emp{48 such figures}.
\item In total: $115 \times 48 = $ \emp{5520 combinations}.
\item Number of \emp{non-empty} combinations: \emp{4072}.
\end{itemize}

{\scriptsize
<<heada>>=
head(moTab)
@
}

An \emp{anonymous} data set!
\end{frame}

\begin{frame}[fragile]{PH regression with the table}

\begin{itemize}
\item Piecewise constant hazard model generates the same likelihood as a Poisson model.
\begin{itemize}
\item Does \emp{not} imply that the number of deaths follow a Poisson distribution!
\end{itemize}
\end{itemize}

\small

<<coxpois,echo=TRUE,cache=FALSE>>=
system.time(fitp <- glm(event ~ offset(log(exposure)) + age +
                hisclass * period + civst + urban + sex,
            data = vb0Tab, family = poisson))
system.time(dr <- drop1(fitp, test = "Chisq"))
@

\end{frame}

\begin{frame}{ANOVA with the table}

<<anovatable,results='asis',echo=FALSE>>=
xtable(dr)
@

\end{frame}

\begin{frame}[fragile]{The 'baseline hazard function'}

{\scriptsize
<<extab,echo=FALSE,results='asis'>>=
xtable(summary(fitp)$coefficients[1:10, 1:2])
@
}

\begin{itemize}
\item \emp{(Intercept)} is log hazard in 40--45. \emp{Reference} interval.
\item \emp{age45} is the \emp{difference} in log hazard between 45--50 and 40--45,
\item \emp{age50} is the \emp{difference} in log hazard between 50--55 and 40--45,
\item Etcetera \ldots
\end{itemize}

\end{frame}

\begin{frame}[fragile]{Baseline hazards}

<<fixthem,fig.height = 3.4>>=
alpha <- summary(fitp)$coefficients[1:10, 1]
alpha[2:10] <- alpha[2:10] + alpha[1]
oldpar <- par(mfrow = c(1, 2))
plot(seq(40, 85, by = 5), alpha, col = "red", xlab = "Age", main = "log(hazard)", ylab = "", axes = FALSE)
axis(1)
axis(2, las = 1)
axis(1)
box()
x <- seq(40, 85, by = 5)
plot(x, exp(alpha), col = "blue", type = "b", xlab = "Age", main = "hazard", ylab = "", axes = FALSE)
axis(1)
axis(2, las = 1)
box()
abline(h = 0)
par(oldpar)
##ff <- lm(alpha ~ x)
##plot(ff)
@

Why not try:

\begin{equation*}
\log\{h(t)\} = \alpha + \beta t, \quad 40 < t \le 90,
\end{equation*}

a \emp{Gompertz} distribution?

\end{frame}

%%\begin{frame}{Period analysis, men}

%%<<coxper,eval=FALSE>>=
%%oldpar <- par(mfrow = c(2, 3))
%%men <- modern[modern$sex == "male", ]
%%##levels(men$hisclass) <- c("elite", "middle", "worker", "NA")
%%cols = c("black", "darkgreen", "red", "blue")
%%for (per in levels(men$period)){
%%    min <- men[men$period == per, ]
%%   fit <- coxreg(Surv(enter, exit, event) ~ strata(hisclass) + urban * civst, data = min)
%%   plot(fit, col = cols, main = paste("Men", per), fn = "surv", printLegend = FALSE, xlab = "Age", xlim = c(40, 90))
%%   legend("bottomleft", legend = levels(min$hisclass), col = cols, lty = 1)
%%   abline(h = 0.5, lty = 3)
%%}
%%par(oldpar)
%%@

%%\end{frame}

\begin{frame}[fragile]{Survival analysis, men 1961-2013}

<<coxper2>>=
source("R/plot_surv.R")
##oldpar <- par(mfrow = c(2, 3))
men <- moTab[moTab$sex == "male", ]
cols <- c("darkgreen", "black", "red", "blue")
##men$hisclass <- relevel(men$hisclass, ref = "middle")
plot_surv(men, sex = "Men")
@

\end{frame}

\begin{frame}[fragile]{Median life after 40, men 1961-2013}

<<coxpertab>>=
source("R/mean_pch.R")
source("R/median_pch.R")
source("R/plot_median.R")

##pop <- vb0Tab[!(vb0Tab$period %in% oldperiod) & 
  ##                vb0Tab$sex == "male" &
    ##              vb0Tab$civst != "unmarried" &
      ##            !vb0Tab$urban, ]
##pop$hisclass <- relevel(pop$hisclass, ref = "middle")
##pop$civst <- relevel(pop$civst, ref = "married")
##pop$period <- factor(pop$period)
##pop$age <- factor(pop$age)
plot_median(men)
@

\end{frame}

\begin{frame}{Survival analysis, women 1961-2013}

<<coxperwom>>=
par(mfrow = c(2, 3))
women <- moTab[moTab$sex == "female", ]
##levels(men$hisclass) <- c("elite", "middle", "worker", "NA")
##cols = c("black", "green", "red", "blue")
plot_surv(women, sex = "Women")
##for (per in levels(women$period)){
##    min <- women[women$period == per, ]
##   fit <- coxreg(Surv(enter, exit, event) ~ strata(hisclass) + urban * civst, data = min)
##   plot(fit, col = cols, main = paste("Women,", per), fn = "surv", printLegend = FALSE, xlab = "Age", xlim = c(40, 90))
##   legend("bottomleft", legend = levels(min$hisclass), col = cols, lty = 1)
##}
@

\end{frame}

\begin{frame}[fragile]{Median life after 40, women 1961-2013}
<<coxpertabwom,warning = FALSE>>=
source("R/mean_pch.R")
source("R/median_pch.R")
##oldperiod <- levels(vb0Tab$period)[1:4]
##pop <- modern[modern$sex == "female", ]
##pop <- vb0Tab[!(vb0Tab$period %in% oldperiod) & 
  ##                vb0Tab$sex == "male" &
    ##              vb0Tab$civst != "unmarried" &
      ##            !vb0Tab$urban, ]
##pop$hisclass <- relevel(pop$hisclass, ref = "middle")
##pop$civst <- relevel(pop$civst, ref = "married")
##pop$period <- factor(pop$period)
##pop$age <- factor(pop$age)
plot_median(women)
@

\end{frame}

\begin{frame}[fragile]{Median life after 40, men 1851-1950}

<<medaf40olmen>>=
men <- olTab[olTab$sex == "male", ]
plot_median(men)
@

\end{frame}

\begin{frame}[fragile]{Median life after 40, women 1851-1950}

<<medaf40olwomen,warning=FALSE>>=
women <- olTab[olTab$sex == "female", ]
plot_median(women)
@

\end{frame}

%\begin{frame}[frame]{Income, men, 1990-2005}

%<<incomemen>>=
%source("R/plot_income.R")
%load("data/loTab.rda")
%loTab$period <- factor(loTab$period)
%men <- loTab[loTab$sex == "male",]
%plot_income(men)
%@

%\end{frame}

% \begin{frame}{Conclusion}

% \begin{itemize}
% \item Most striking: Very bad prospects for \emp{unemployed}, both men and women, all periods.
% \item Apart from unemployed: Equal opportunities for men 1960--80, for women 1960--90, but after that a clear separation with the \emp{elite on top}.
% \end{itemize}

% \vspace{2\baselineskip}

% After this slide: \emp{Old stuff!}

% \end{frame}

\begin{frame}{Proportional hazards for women?}

\includegraphics[width=11cm,height=8cm]{figure/hazplotw-1}

\end{frame}

\begin{frame}{Proportional hazards for men?}

\includegraphics[width=11cm,height=8cm]{figure/hazplotm-1}

\end{frame}

% \begin{frame}[fragile]{Interactions}

% \begin{verbatim}
% fit <- coxph(Surv(enter - 40, exit - 40, event) ~
%     period * hisclass * sex + urban, data = vb0)
% drop1(fit, test = "Chisq")
% \end{verbatim}

% \begin{center}
% \begin{tabular}{lrrrr}
%   \hline
%  & Df & AIC & LRT & Pr($>$Chi) \\
%   \hline
% $<$none$>$ &  & 1631901.32 &  &  \\
%   urban & 1 & 1631946 & 47.04 & $7.0\times 10^{-12}$ \\
%   \emp{period:hisclass:sex} & 27 & 1631957 & 110.11 & \emp{$5.3\times 10^{-12}$} \\
%    \hline
% \end{tabular}
% \end{center}

% \begin{itemize}

% \item This model is described by \emp{80 coefficients}.

% \item We show them in \emp{graphs}.
% \end{itemize}

% \end{frame}

\begin{frame}{No proportional hazards (wrt \sc Hisclass)!}

\begin{itemize}
\item Fit each {\sc Hisclass} \emp{separately}, for each time period.
\item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
\end{itemize}

So, we fit separate PH models for each combination of

\begin{itemize}
  \item \emp{\tt sex}
  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
  \item \emp{\tt hisclass}
  \item \emp{\tt age group}: 40--64, 65--89.
  \item \emp{\tt cause of death}
\end{itemize}
\end{frame}

% \begin{frame}{The data table}

% <<lexis3, fig.height = 4>>=
% plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
%      type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
%      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(40, 90))
% axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
% axis(2, las = 2, at = c(0, 40, 65, 90))
% axis(4, las = 2, at = c(0, 40, 65, 90))
% box()
% abline(h = 0)
% lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
%       col = "blue", lwd = 2)
% for (i in c(1851, 1876, 1901, 1926, 1971, 1981, 1991, 2002, 2007)){
%     lines(c(i, i), c(40, 90), lty = 2)
% }
% for (i in c(45, 50, 55, 60, 65, 70, 75, 80, 85)){
%     lines(c(1851, 1951), c(i, i), lty = 2)
%     lines(c(1961, 2014), c(i, i), lty = 2)
% }
% @

% \emp{Poisson regression}:

% In \emp{each cell}, and for \emp{each} combination of \emp{covariate level}s, calculate
% \begin{itemize}
% \item  the number of \emp{deaths} and
% \item the total \emp{exposure time}.
% \end{itemize}
% \end{frame}

\begin{frame}{Women and men, age 40--64}

  \begin{center}
\includegraphics[height = 7.5cm]{figure/cumsw-1}
  \end{center}
\end{frame}

\begin{frame}{Relative differences, ages 40--64 and 65--89}

  %\begin{center}
\includegraphics[height = 7.5cm]{figure/cumswstand-1}\includegraphics[height=7.5cm]{figure/cumswold-1}
   %\end{center}
\end{frame}

\begin{frame}{Cancer mortality, ages 40-64 and 65-89}

\includegraphics[height=7.5cm]{figure/12cumsw-1}\includegraphics[height=7.5cm]{figure/12cumswold-1}

\end{frame}

\begin{frame}{Cardiovascular mortality}

\includegraphics[height=7.5cm]{figure/41cumsw-1}\includegraphics[height=7.5cm]{figure/41cumswold-1}

\end{frame}

\begin{frame}[fragile]{Education 1990--2013}

<<educ>>=
load("data/lisaold.rda")
educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
educ$period <- factor(educ$period)
educ <- educ[!is.na(educ$educ), ]
levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
educ$educ <- relevel(educ$educ, ref = "middle")
educ$civst <- relevel(educ$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban, 
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban, 
             data = educ[educ$sex == "female", ])
cols <- c("darkgreen", "red", "black")
@

<<ploteduc>>=
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",  
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)

plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",  
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)

par(oldpar)
@

\end{frame}

\begin{frame}{Income 1986--2005}

<<femeduc>>=
income <- lisaold[!is.na(lisaold$income), ]
income$period <- factor(income$period)
##income <- income[!is.na(income$income), ]
##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
income$income <- relevel(income$income, ref = "low")
income$civst <- relevel(income$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
             data = income[income$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
             data = income[income$sex == "female", ])
cols <- c("darkgreen", "red", "blue", "black")

@

<<plotincome>>=
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",  
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)

plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",  
     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)

par(oldpar)
@

\end{frame}

\begin{frame}[fragile]{Income by education, 1990--2005, ages 40--75!}

<<incomeandeduc>>=
both <- income[!is.na(income$educ), ]
levels(both$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
both$educ <- relevel(both$educ, ref = "middle")
bmal <- both[both$sex == "male", ]
bfem <- both[both$sex == "female", ]
fitme1 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
                data = bmal[bmal$educ == "low", ])
fitme2 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
                data = bmal[bmal$educ == "middle", ])
fitme3 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
                data = bmal[bmal$educ == "high", ])
##
fitfe1 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
                data = bfem[bfem$educ == "low", ])
fitfe2 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
                data = bfem[bfem$educ == "middle", ])
fitfe3 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban, 
                data = bfem[bfem$educ == "high", ])

@

<<plotedin>>=
oldpar <- par(mfrow = c(2, 3))
ylim = c(0.5, 1)
plot(survfit(fitme1, start.time = 40), col = cols, main = "Men, low education", xlab = "Age", ylim = ylim)
legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.0)

plot(survfit(fitme2, start.time = 40), col = cols, main = "Men, average education", xlab = "Age", ylim = ylim)
plot(survfit(fitme3, start.time = 40), col = cols, main = "Men, high education", xlab = "Age", ylim = ylim)
##
plot(survfit(fitfe1, start.time = 40), col = cols, main = "Women, low education", xlab = "Age", ylim = ylim)
legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.0)

plot(survfit(fitfe2, start.time = 40), col = cols, main = "Women, average education", xlab = "Age", ylim = ylim)
plot(survfit(fitfe3, start.time = 40), col = cols, main = "Women, high education", xlab = "Age", ylim = ylim)
par(oldpar)
@

\end{frame}

%%\begin{frame}{Education and income, 1990--2005, all deaths}

%%\includegraphics[height=3.5cm]{figure/liswomfig-1} Women, 40--64.

%%\includegraphics[height=3.5cm]{figure/lismanfig-1} Men, 40--64.

%%\end{frame}

%%\begin{frame}{Education and income, 1990--2005, all deaths}

%%\includegraphics[height=3.5cm]{figure/liswomfig2-1} Women, 65--89.

%%\includegraphics[height=3.5cm]{figure/lismanfig2-1} Men, 65--89.

%%\end{frame}

% \begin{frame}{Education and income, 1990--2005, cancer}

% \includegraphics[height=3.5cm]{figure/12liswomfig-1} Women, 40--64.

% \includegraphics[height=3.5cm]{figure/12lismanfig-1} Men, 40--64.

% \end{frame}

% \begin{frame}{Education and income, 1990--2005, cancer}

% \includegraphics[height=3.5cm]{figure/12liswomfig2-1} Women, 65--89.

% \includegraphics[height=3.5cm]{figure/12lismanfig2-1} Men, 65--89.

% \end{frame}

% \begin{frame}{Education and income, 1990--2005, cardiovascular}

% \includegraphics[height=3.5cm]{figure/41liswomfig-1} Women, 40--64.

% \includegraphics[height=3.5cm]{figure/41lismanfig-1} Men, 40--64.

% \end{frame}

% \begin{frame}{Education and income, 1990--2005, cardiovascular}

% \includegraphics[height=3.5cm]{figure/41liswomfig2-1} Women, 65--89.

% \includegraphics[height=3.5cm]{figure/41lismanfig2-1} Men, 65--89.

% \end{frame}

%%% Stop here!

\begin{frame}{Answers(?)}

\begin{enumerate}
\item Has \emp{inequality in mortality} between \emp{social classes}
  increased?
\begin{itemize}
\item Not really
\end{itemize}
%%\item Is the mortality transition initiated by a certain social class?

\item Are there any \emp{gender differences} in the effect of social position?
\begin{itemize}
\item Yes
\end{itemize}

\item Is social position \emp{equally important} among the \emp{retired} population as
  for those in \emp{working age}?
\begin{itemize}
\item Not really
\end{itemize}

\item Is there a difference between classes in mortality from
  \emp{cardiovascular} diseases and \emp{cancers} respectively?
\begin{itemize}
\item Yes
\end{itemize}
\item Are there large differences in survival depending on level of
  \emp{education} and \emp{income} categories?
\begin{itemize}
\item Yes, definitely.
\end{itemize}

\end{enumerate}

\end{frame}

\begin{frame}{Conclusion}

%\scriptsize

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item There are \emp{gender differences} in the social pattern
\item No difference in the social pattern in the two studied age groups.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item \emp{Education} and \emp{income} are the \emp{most important}
  variables during late
  time period.
\end{itemize}

\end{frame}

% \begin{frame}{Software for report writing and analysis}

% \begin{itemize}
% \item  \R : \emp{\url{https://www.R-project.org}}.
%   \begin{itemize}
%     \item \emp{RStudio}: \url{https://www.rstudio.com}.
%   \end{itemize}
% \item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
%   \begin{itemize}
%   \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
%   \item \emp{rmarkdown} (RStudio Team, 2016).
%   \item \emp{knitr} (Xie, 2016).
%   \item \emp{eha} (Broström, 2016).
%   \item \emp{xtable} (Dahl, 2016).
% \end{itemize}

% \item \emp{Version control} (reproducibility)
%   \begin{itemize}
%     \item \emp{git}: \url{http://git}
%     \item \emp{gitHub}: \url{http://github.com/goranbrostrom/}

% \end{itemize}
% \end{itemize}

% \end{frame}

\end{document}

### 2017-05-19 icass.Rnw
data from capa

@@
 \maketitle
+
+<<setup, include = FALSE>>=
+library(knitr)
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+library(knitr)
+library(eha)
+library(xtable)
+library(skum)
+@
+
+<<readalldata>>=
+## Load remote data:
+con <- url("http://capa.ddb.umu.se/data/vb0.rda")
+##load("data/vb0.rda")
+load(con)
+close(con)
+vb0$hisclass[is.na(vb0$hisclass)] <- 4
+vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
+##
+con <- url("http://capa.ddb.umu.se/data/vb.rda")
+##load("data/vb.rda")
+load(con)
+close(con)
+##
+con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+##load("data/lisaold.rda")
+load(con)
+close(con)
+##
+con <- url("http://capa.ddb.umu.se/data/loTab.rda")
+load(con)
+close(con)
+loTab$period <- factor(loTab$period)
+##
+vb0Tab <- aggregate(vb0[, c("event", "exposure")],
+                   vb0[, c("sex", "age", "period", "cohort",
+                          "urban", "civst", "hisclass")],
+                   FUN = sum)
+vb0Tab$age <- factor(vb0Tab$age)
+@
 
@@
 
-<<setup, include = FALSE>>=
-knitr::opts_chunk$set(cache=TRUE,echo=FALSE,message=FALSE,warning=TRUE,comment=NA)##$
-@
-
 %%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}
@@
 %%\subsection{Combining the data sets and tabulate}
-
-<<readtab>>=
-load("data/vbTab.rda") 
-vbTab$period <- factor(vbTab$period)
-levels(vbTab$hisclass40) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vbTab$hisclass60) <- c("elite", "lowMan", "lowWhiteC", "worker")
-levels(vbTab$period)[8] <- "1991-2001"
-load("data/vb0.rda")
-vb0$period <- factor(vb0$period)
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
-vb <- vb0
-@
 
@@
 \item {\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
-\item {\tt lowMan}, HISCLASS 3 and 8. Lower managers and farmers.
-\item {\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
+\item {\tt middle}, HISCLASS 3, 4, 5, 6 and 8. Lower managers and farmers.
 \item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
@@
 
-<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, farmers and lower business, lower white collar, workers) and time period", fig.height = 4.5>>=
-xx <- with(vbTab, tapply(exposure, list(period = period, hisclass = hisclass40), sum))
+<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
+xx <- with(vb0Tab, tapply(exposure, list(period = period, hisclass = hisclass), sum))
 yy <- round(prop.table(xx, 1) * 100, 1)
@@
 <<civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5>>=
-xx <- with(vbTab, tapply(exposure, list(period = period, civst = civst), sum))
+xx <- with(vb0Tab, tapply(exposure, list(period = period, civst = civst), sum))
 yy <- round(prop.table(xx, 1) * 100, 1)
@@
 
-labb <- levels(vbTab$period)
+labb <- levels(vb0Tab$period)
 nper <- length(labb)
@@
 ##
-vb4064 <- vbTab[as.numeric(as.character(vbTab$age)) < 64.5, ]
+vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
 vb4064$age <- factor(vb4064$age)
-vb4064$hisclass <- vb4064$hisclass40
+##vb4064$hisclass <- vb4064$hisclass40
 @
@@
 source("R/getPHglm2.R")
+##nper <- 4
+nper <- length(labb)
 resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
@@
 library(xtable)
-load("data/loTab.rda")
-loTab$period <- factor(loTab$period)
 liss4064 <- loTab[loTab$age < 64.5, ]
@@
 <<getske6089>>=
-vb6589 <- vbTab[as.numeric(as.character(vbTab$age)) >= 64.5, ]
+vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
 vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
-vb6589$hisclass <- vb6589$hisclass60
+##vb6589$hisclass <- vb6589$hisclass60
 @
@@
 library(xtable)
-load("data/loTab.rda")
+##load("data/loTab.rda")
 loTab$period <- factor(loTab$period)
@@
 library(xtable)
-load("data/loTab.rda")
-loTab$period <- factor(loTab$period)
+##load("data/loTab.rda")
+##loTab$period <- factor(loTab$period)
 liss4064 <- loTab[loTab$age < 64.5, ]

### 2017-05-19 slides3.Rnw
touch

@@
 library(skum)
+## Load remote data:
+con <- url("http://capa.ddb.umu.se/data/vb0.rda")
+##load("data/vb0.rda")
+load(con)
+close(con)
+vb0$hisclass[is.na(vb0$hisclass)] <- 4
+vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
+##
+con <- url("http://capa.ddb.umu.se/data/vb.rda")
+##load("data/vb.rda")
+load(con)
+close(con)
+##
+con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+##load("data/lisaold.rda")
+load(con)
+close(con)
+##
+vb0Tab <- aggregate(vb0[, c("event", "exposure")],
+                   vb0[, c("sex", "age", "period", "cohort",
+                          "urban", "civst", "hisclass")],
+                   FUN = sum)
+vb0Tab$age <- factor(vb0Tab$age)
+
 @
@@
 <<lexis2>>=
-load("data/vb0.rda")
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
 plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
@@
 ##load("data/vb0.rda")
-load("data/vb.rda")
-
 tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
@@
 
-load("data/vb.rda")
+##load("data/vb.rda")
 tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
@@
 <<aggregate,eval = TRUE, echo = TRUE>>=
-vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
-                   vb0[, c("sex", "age", "period", "cohort", 
-                          "urban", "civst", "hisclass")], 
-                   FUN = sum)
-vb0Tab$age <- factor(vb0Tab$age)
 nr <- NROW(vb0Tab)
@@
 <<educ>>=
-load("data/lisaold.rda")
+
 educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]

### 2017-05-22 icass.Rnw
Update introduction

@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{plainnat}
+
 %\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
@@
 
-For a long time, it was generally assumed that socio-economic health inequalities were high in historical societies, probably higher than in modern societies.
+One of the most central aspects of survival on individual level is social class and access to economic and other resources.
+Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and they seem to be increasing as well \citep{kuetal04, frlu07}.
+It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}.
+For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies.
 This can be considered as a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
 Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
-When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
-Results from different studies have diverged, some reporting substantial differences but many have found surprisingly small socio-economic differences during the first stages of the transition \citep{bevp11, edli11,edbr12}.
-The appearance of the consistent pattern of inequality in survival that characterises present-day societies must have developed somewhere on the road from a high-mortality to a low-mortality regime.
-The complete process is however largely unknown.
-
-A long-term perspective on the development of social
-inequalities in mortality in the adult and elderly population is taken in the present paper.
-The area under investigation consists of the Skellefteå and Umeå regions in
-the north of Sweden, and the time period is 1851--2013 for Skellefteå and
-1901--2013 for Umeå.
-
-This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
-the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
-in time for the Skellefteå region (1851--1900), and we also consider causes
-of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
-us to study the effect of income and education on mortality in a more
-direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.
+Not only should we expect to find a direct effect of the lack of resources on health, we might even find an independent effect of inequality as \citet{Wipi} suggest.
+They argue that unequal societies perform less well when it comes to health than equal ones.
+Even if it not a necessary implication of this, we would still assume that poorer groups were also those most disadvantaged of large inequality in itself.
+Some recent studies have however indicated that the assumed pattern of distinct socio-economic health differences was not always apparent in pre- and early-transitional societies and some suggest that the modern pattern might be a fairly recent phenomenon \citep{bevp11}.
+Solid empirical evidence is about the process is however lacking and the available studies focusing on the issue are still few.
+There is a need for more reliable studies from different geographical and historical settings in order to better understand the role of socio-economic conditions for health over time.
+
+% This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
+% the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
+% in time for the Skellefteå region (1851--1900), and we also consider causes
+% of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
+% us to study the effect of income and education on mortality in a more
+% direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.
 
@@
 
-The aim of the paper is to investigate the development of social
-differences using HISCLASS in mortality for the adult and elderly
-population during the mortality transition. The
-analyses will be performed on data from the Skellefteå and Umeå regions
-1851--2013. The main questions are:
+In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
+The areas under investigation consist of the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.The main questions are:
 
@@
 \item Has inequality in mortality between social classes increased among the adult and elderly
-  population, primarily benefiting the highest social positions?
+  population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
 
@@
 \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
-\item Are there large differences in survival depending on level of education and income categories?
+\item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 onwards?
 \end{enumerate}
@@
 set in earlier and was faster among privileged groups.
+When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
+Results from different studies have diverged, some reporting substantial differences but many have found surprisingly small socio-economic differences during the first stages of the transition \citep{bevp11, edli11,edbr12}.
 
@@
 \begin{figure}
-\includegraphics[height=19cm]{myfigs/skum17}
+\includegraphics[height=19cm]{myfigs/skum}
 \caption{The Skellefteå and Umeå regions.}

### 2017-05-22 icass.Rnw
Update region description

@@
 
-\date{May 18, 2017}
+\date{May 22, 2017}
 
@@
 
-For a detailed presentation we refer to \citet{segb16}. Here is a short
-summary.
-
-Several studies have shown how important access to economic and other
-resources are for health, but it is not clear how this has developed in
-history. \citet{ant67} and \citet{smi83} suggest that social inequality in
-health has passed through different phases in history. According to them,
-social health differences were
-comparatively small during the pre-transitional phase when space was a
-strong determinant for the spread of disease. Differences then increased
-during the transitional phase when mortality declined and wealthy groups
-used their resources to gain better health. Finally, health differences
-decreased again in modern low-mortality settings when instead
-health-related behaviour became important, resulting in marginal
-differences. \citet{omran82} state, in the third proposition in his theory
-of the epidemiological transition, that even if the class differential in
-mortality was maintained during the transition, the decline
-set in earlier and was faster among privileged groups.
+Several studies have shown how important access to economic and other resources are for health, but it is not clear how this has developed in history.
+\citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different phases in history.
+According to them, social health differences were
+comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease.
+Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health.
+Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences.
+\citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.
 When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
@@
 
-Both regions are part
-of the county of Västerbotten in the north of Sweden, along the coast of
-the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection
-of parishes surrounding the town of Skellefteå, founded in 1845 but with a
-very small population in the 19th century. The data from the period after
-1960 cover the Skellefteå and Malå municipalities, the same area as for the
-earlier period but with the addition of two more parishes. The majority
-of the 19th century population lived in rural villages and hamlets, getting
-their livelihood from agricultural production. The region was vulnerable for
-harvest failures, one living long in memory was the famine of 1867–-1868
-(Edvinsson and Broström, 2014).
+The Skellefteå and Umeå regions are part of county of Västerbotten in the north of Sweden.
+This was a remote part of the country, where communications with the rest of Sweden were restricted until the late 20th century.
+During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
+Communications during winter used the winter roads.
+The first ships during the year came to Umeå in May?
+Towards the end of the 19th century, railways were built up to our regions, thereby facilitating contacts with the rest of Sweden.
+But of course, they were not totally outside what happened in other parts of the country.
+The inlands were however very remote.
+But our regions are situated mostly close to the coast.
+During the 19th century, these parts became integrated in the same epidemiological pattern as in the rest of Sweden.
+Diseases reached to the places that earlier were saved.
+Both regions are part of the county of Västerbotten in the north of Sweden, along the coast of the Gulf of Bothnia.
+The Skellefteå region in Poplink consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population in the 19th century.
+The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
+The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production.
+The region was vulnerable to harvest failures, one living long in memory was the famine of 1867–-1868 (Edvinsson and Broström, 2014).
 Towards the end of the century, the region became connected to the Swedish railway system, thus improving the communications and contacts with other parts of the country.
-During the
-20th century, industrialisation took place, partly by the opening of mines in
-the inlands. This also led to a population increase both in the town and in
-the rural parts, and a much more diversified economy.
-Mortality was fairly low in comparison with other parts of the country
-and the fertility transition was late. The Skellefteå population size as defined
-in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900,
-61\,938 on January 1, 1950 and 76\,723 at the end of the century.
-
-The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and
-rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
+During the 20th century, industrialisation took place, partly by the opening of mines in the inlands.
+This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.
+
+Mortality was fairly low in comparison with other parts of the country and the fertility transition was late.
+The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.
+
+The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
 This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time.
-Umeå town had a
-small population but was substantially larger than Skellefteå town during almost
-the whole studied period. It was the administrative centre in the county of
-Västerbotten and schools and military regiments were placed here. Consequently, Officials,
-teachers and the military were much more common in Umeå than in Skellefteå. Agriculture
-dominated the rural part, there were some foundries, some
-industries for example in forestry and small-scale production. Umeå had a
-much more diversified occupational structure than Skellefteå. The population size
-as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on
-January 1, 1950 and 103\,970 when the century ended.
+Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period.
+It was the administrative centre in the county of Västerbotten and schools and military regiments were placed here.
+Consequently, officials, teachers and the military were much more common in Umeå than in Skellefteå.
+Agriculture dominated the rural part, there were some foundries, some industries for example in forestry and small-scale production.
+Umeå had a much more diversified occupational structure than Skellefteå.
+The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.
 

### 2017-05-23 icass.Rnw
Update data and variables

--- 2017-05-22 icass.Rnw
@@
 The area under investigation consists of the Skellefteå and Umeå regions in
-the north of Sweden, and the time period is 1851--2013.The outcome is all
+the north of Sweden, and the time period is 1851--2013. The outcome is all
 cause mortality and cause specific mortality for the categories
@@
 
-The main findings are that the a consistent social gradient implying better
+The main findings are that a consistent social gradient implying better
 survival for classes with more access to economic and other resources, only
@@
 income categories confirmed large inequality in mortality
-between different groups during the latest decades, but also demonstrated more distinct differences between categories with better survival the more income and the higher education level people have.
+between different groups during the most recent decades, but also demonstrated more distinct differences between categories with better survival the more income and the higher education level people have.
 Both cardiovascular diseases and cancers conform with the pattern already
@@
 One of the most central aspects of survival on individual level is social class and access to economic and other resources.
-Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and they seem to be increasing as well \citep{kuetal04, frlu07}.
+Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and the impact seems to be increasing as well \citep{kuetal04, frlu07}.
 It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}.
@@
 Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
-Not only should we expect to find a direct effect of the lack of resources on health, we might even find an independent effect of inequality as \citet{Wipi} suggest.
+Not only should we expect to find a direct effect of the lack of resources on health, we might even find an independent effect of inequality as \citet{wipi09} suggest.
 They argue that unequal societies perform less well when it comes to health than equal ones.
-Even if it not a necessary implication of this, we would still assume that poorer groups were also those most disadvantaged of large inequality in itself.
+Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.
 Some recent studies have however indicated that the assumed pattern of distinct socio-economic health differences was not always apparent in pre- and early-transitional societies and some suggest that the modern pattern might be a fairly recent phenomenon \citep{bevp11}.
-Solid empirical evidence is about the process is however lacking and the available studies focusing on the issue are still few.
-There is a need for more reliable studies from different geographical and historical settings in order to better understand the role of socio-economic conditions for health over time.
+Solid empirical evidence about the process is however lacking and studies focusing on the issue are still few.
+There is a need for more reliable studies from different geographical and historical settings as well as according to age and gender in order to better understand the role of socio-economic conditions for health and survival over time.
 
@@
 In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
-The areas under investigation consist of the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.The main questions are:
+The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
+The main questions are:
 
@@
 
-For a detailed presentation we refer to \citet{segb16}. Here is a short
-summary.
-
-Several studies have shown how important access to economic and other
-resources are for health, but it is not clear how this has developed in
-history. \citet{ant67} and \citet{smi83} suggest that social inequality in
-health has passed through different phases in history. According to them,
-social health differences were
-comparatively small during the pre-transitional phase when space was a
-strong determinant for the spread of disease. Differences then increased
-during the transitional phase when mortality declined and wealthy groups
-used their resources to gain better health. Finally, health differences
-decreased again in modern low-mortality settings when instead
-health-related behaviour became important, resulting in marginal
-differences. \citet{omran82} state, in the third proposition in his theory
-of the epidemiological transition, that even if the class differential in
-mortality was maintained during the transition, the decline
-set in earlier and was faster among privileged groups.
+Several studies have shown how important access to economic and other resources are for health, but it is not clear how this has developed in history.
+\citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases.
+According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease.
+Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health.
+Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences.
+\citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.
 When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
-Results from different studies have diverged, some reporting substantial differences but many have found surprisingly small socio-economic differences during the first stages of the transition \citep{bevp11, edli11,edbr12}.
-
-These assumptions has however been questioned by later research. According to the present state of knowledge about the development of social
-health inequalities (in Sweden but also in other European countries), no
-consistent social hierarchy has been found before 1900 but substantial
-differences towards the late 20th and early 21st century. It appears as if
-a more consequent social health divide developed in the period
-in-between. This period has not been thoroughly studied on this issue,
-mainly because of lack of data to analyse. However, the extension of data
-at the Demographic Data Base, Umeå University (DDB) makes such studies
-possible. We are now able to analyse social
-differences in mortality with micro-data on a sufficiently large population
-and with substantial social diversity from the 20th century for the
-Skellefteå and Umeå regions \citep{wee16}.
+When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small socio-economic differences during the first stages of the transition \citep{bevp11, bedr11, edli11,edbr12}.
+The Swedish results show no consistent social hierarchy before 1900 but substantial differences towards the late 20th and early 21st century.
+It appears as if a more consequent social health divide developed in the period
+in-between. 
+\citet{wipi09} argue that income inequality has an independent effect on mortality, separated from that of the actual differential levels of economic resources.
+They find an positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world.
+This has initiated a vital scholarly debate and the topic has been extensively studied (see for example \citet{suka04, wado00}.
+How the association looked like in previous periods has however not been studied.
+This is of particular interest since the levels of inequality were very different from what we find during most of the recent decades.
+Data are uncertain for the 19th century, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
+From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach the lowest point at around 1980.
+Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}.
+The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.
 
@@
 
-Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
-show the study areas and their position in Sweden.
+% Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
+% show the study areas and their position in Sweden.
 
@@
 
-Both regions are part
-of the county of Västerbotten in the north of Sweden, along the coast of
-the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection
-of parishes surrounding the town of Skellefteå, founded in 1845 but with a
-very small population in the 19th century. The data from the period after
-1960 cover the Skellefteå and Malå municipalities, the same area as for the
-earlier period but with the addition of two more parishes. The majority
-of the 19th century population lived in rural villages and hamlets, getting
-their livelihood from agricultural production. The region was vulnerable for
-harvest failures, one living long in memory was the famine of 1867–-1868
-(Edvinsson and Broström, 2014).
+The Skellefteå and Umeå regions (Figures~\ref{fig:SKEUME} and~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of the Gulf of Bothnia.
+This was a remote part of the country, where communications with the rest of Sweden were restricted until the late 20th century, dominated by agriculture and characterise by a vulnerable economy.
+Several severe famines struck the regions during the 1800s, making living difficult for the population and increasing mortality \citep{segb14}.
+During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
+Communications during winter used the winter roads.
+The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}.
+Towards the end of the 19th century, railways construction reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures.
+The inland was particularly remote.
+During the 19th century, our regions parts became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
+Diseases reached to the places that earlier were spared.
+The Skellefteå region in Poplink consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
+The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
+The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production.
+The region was vulnerable to harvest failures, one living long in memory was the famine of 1867–-1868 \citep{segb14}.
 Towards the end of the century, the region became connected to the Swedish railway system, thus improving the communications and contacts with other parts of the country.
-During the
-20th century, industrialisation took place, partly by the opening of mines in
-the inlands. This also led to a population increase both in the town and in
-the rural parts, and a much more diversified economy.
-Mortality was fairly low in comparison with other parts of the country
-and the fertility transition was late. The Skellefteå population size as defined
-in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900,
-61\,938 on January 1, 1950 and 76\,723 at the end of the century.
-
-The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and
-rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
+During the 20th century, industrialisation took place, partly by the opening of mines in the inlands.
+This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.
+
+Mortality was fairly low in comparison with other parts of the country and the fertility transition was late.
+The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.
+
+The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
 This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time.
-Umeå town had a
-small population but was substantially larger than Skellefteå town during almost
-the whole studied period. It was the administrative centre in the county of
-Västerbotten and schools and military regiments were placed here. Consequently, Officials,
-teachers and the military were much more common in Umeå than in Skellefteå. Agriculture
-dominated the rural part, there were some foundries, some
-industries for example in forestry and small-scale production. Umeå had a
-much more diversified occupational structure than Skellefteå. The population size
-as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on
-January 1, 1950 and 103\,970 when the century ended.
+Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period.
+It was the administrative centre in the county of Västerbotten and schools and military regiments were placed here.
+Consequently, officials, teachers and the military were much more common in Umeå than in Skellefteå.
+Agriculture dominated the rural part, there were some foundries, some industries for example in forestry and small-scale production.
+Umeå had a much more diversified occupational structure than Skellefteå.
+The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.
 
@@
 
-The data for the present study comes from two large population databases at
-the DDB, Umeå University. The early period is covered by
+The extension in time of data at the Demographic Data Base, Umeå University (DDB) makes it possible to investigate the long-term development of social inequality in mortality by including the period between 1900 and 1950s when almost no micro-data has been available for research.
+The databases provide us with with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.The data for the present study comes from two large population databases at the DDB, Umeå University.
+The early period is covered by
 the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region \citep{wee16}. The population in the two
-regions is large enough and socially diverse to enable studies of social
-mortality inequalities. Poplink is based on linked parish records, allowing
+the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked parish records, allowing
 us to reconstruct life biographies on people as long as they remained in
@@
 Information on individuals are anonymised and the two databases are not
-linked between each other. This stops us from following individuals present
+linked. This stops us from following individuals present
 in both databases throughout their lives, thus they are treated separately. This
@@
 birth date, death date, first and last date of observation and type of
-entrance/exit.Total number of deaths analysed are 68.244.
+entrance/exit.Total number of persons are 84.599 leading to 35.001 deaths in Poplink and 197.864 persons leading to 66297 deaths in the Linnaeus database.
+(SKA VI HA FIGUREN MED THE SAMPLING FRAME HÄR?)
 
@@
 
-Differences in available information, both when it comes to structure and variables used for attributing social class, from the two data sets as well as
-within the Linnaeus data set makes necessary diverging approaches when it comes to
-the identification of presence periods within our
+Differences in available information in the two datasets as well as within the Linnaeus data makes it necessary to apply diverging approaches when it comes to identification of presence periods within our
 regions. The Poplink data provides us with exact dates or at least year of
-start and exit of the presence of all individuals in our data. This is however not the case
-with the Linnaeus database.
-For the period 1960--2000 we use the presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare for defining the observation periods as well as their social class (see below) in our analyses.
+start and exit of the presence of all individuals in our data, allowing us to have full and continuous control of the de jure population.
+This is however not the case with the Linnaeus database.
+For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare for defining the observation periods as well as their social class (see below) in our analyses.
 We have however no exact information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
-We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in this census.
-This overestimates their presence in our regions, but should not bias the results.
+We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in the later census.
+This overestimates their presence in our regions, but should not bias the results when it comes to socioeconomic differences.
 We furthermore miss those that both moved in and moved out or vice versa between censuses.
@@
 
-\subsection{HISCLASS}
-
-In this paper we analyse mortality differences according to social class according to the classification scheme HISCLASS.
-The only type of socio-economic information available both in Linnaeus and Poplink is occupations, which can be used to identify distinct social classes.
-Thus, analysing according to social class is apart from a scientific choice, also a practical one.
-Using occupational data is the only and obvious alternative if we are to make a longitudinal study of socio-economic mortality differences with consistent classification schemes (although not without some problems).
-There are also other problems connected with using the same classification scheme covering a long time period, but this will be approached in more detail in the discussion section.
+\subsection{Social class}
+
+In this paper we analyse mortality differences according to social class based on the classification scheme HISCLASS \citep{lema11}.
+The only type of socio-economic information available both in Linnaeus and Poplink is occupations.
+Thus, analysing according to social class is apart from being a scientific choice, also a practical one.
+Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes.
+The obvious problems connected with using the same classification scheme covering a long time period will be addresed in the discussion section.
 Social position is treated as a time-dependent variable.
@@
 The availability to information on occupations varies over time and between sources.
-The Poplink data usually provides information only on the occupation of head of households.
+Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system.
 The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
@@
 Thus, the social class in the census 1990 is considered to be kept until 2000.
+We consider that the social position is kept after retirement (in Sweden usually at age 65).
 
@@
 In the Poplink data, social class is based on the occupational titles given in the parish records.
-These can be more or less precise.
-There are some problems with titles that are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
+Some titles are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
 We consider however their validity to be sufficient for identification of social position.
@@
 
-Female labour is however much better covered in the Linnaeus database, both in the censuses and the population registers.
+Female labour is much better covered in the Linnaeus database, both in the censuses and the population registers.
 Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database.
-Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became very common in Sweden during this period.
-It is also worth to notice that female labour force participation became the norm, and we can therefore argue that the occupation of the male as signifier of the social class became obsolete.
-We let all included persons be signified by their individual occupation for this period.
-Still, we need to keep in mind that the results for the different periods are not completely comparable.
+Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became common and female labour force participation developed as the norm in Sweden during this period.
+The occupation of the husband as signifier of social class became if not obsolete at least less relevant.
+All included persons are signified by their individual occupation for this period.
+Still, the results for the different periods are not completely comparable and makes the interpretation of the development for women more shaky.
 
 Due to the construction of the time-dependent data, some records report unknown social position.
-This has been updated according to rules in order to optimise available information and minimise missing information.
+This has been updated according to rules in order to optimise available information.
 Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
-The fact that Linnaeus and Poplink are separate databases stops us from updating this information from Poplink.This, together with the circumstance that many women had no occupation in the census 1960 explains the large proportion where we have not been able to determine class during that decade.
-
-The classification scheme used here is HISCLASS \citep{lema11}, a class-based system.
-The scheme uses HISCO as a basis for assigning categories.
-HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations, representing a historical version of ISCO.
-HISCO groups occupations according to tasks and skills.
-HISCLASS have 12 categories, but some categories become quite small in number even when analysing these comparatively large regions and we have therefore used only the main categories, based on position on the labour market.
-We have therefore collapsed categories, ending up with the following four:
+This, together with the circumstance that many women had no occupation in the census 1960 explains the large proportion where we have not been able to determine class during that decade.
+As a consequence, we leave out the 1960s and 1970s in some analyses.
+
+The classification scheme used here is  based on HISCLASS \citep{lema11}, a class-based system.
+The scheme uses HISCO for assigning categories.
+HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations.
+HISCO categorize occupations according to tasks and skills.
+We have merged the original 12 classes in HISCLASS, based on position on the labour market, ending up with the following three and an additional fourth category with unknown class four:
 
@@
 occupations in the Linnaeus database is a little more complicated. The
-occupations in the censuses (NYK) must first be transferred to the system
-later in use, SSYK (which is the Swedish version of ISCO88). Thereafter the
+occupations in the censuses (NYK) are first transferred to the system
+later in use, SSYK (the Swedish version of ISCO88). Thereafter the
 SSYK, both those transferred from the censuses and those given in the
@@
 Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure
-time over HISCLASS  for our selected periods. The category of workers are
-fairly stable over time, while the lower white collar group was very small
-during the 19th century, then starting to increase during the early 20th
-century and increasing considerably during the period from the 1960s
-onwards. The group of lower managers where farmers also are included, was
-large until the middle of the 20th century, but has diminished considerably
-thereafter. This is to a large extent explained by farmers almost
-disappearing as an occupation in recent time. The highest classes has
+time over class  for our selected periods. The category of workers is
+fairly stable over time. The highest class has
 increased substantially, although from a very low level in the 19th
@@
 Umeå university explains much of this increase.
+Within the middle group, there has been a transfer from a majority of farmer during the 19th century to lower white collar from the middle of the 20th century.
 
@@
 For the corresponding numbers, with missing values ({\tt NA}) included, see
-Table~\ref{tab:nana}.The proportion with missing HISCLASS was very low
+Table~\ref{tab:nana}.The proportion with missing class was very low
 during the periods until 1950, between 4 and 8 \% lacked information
-allowing determination of class. This increased considerably for the
-beginning of period when the other data set---the Linnaeus database---is
-used, in particular for the 1960s where every second person lacks HISCLASS
-and the 1970s with almost a third have missing information. This has to do
-with difficulties in the determination of class from the censuses,
-something we discuss below. For the most recent period, the information is
-complete.
+allowing determination of class. This increased considerably during the first two decades of data brom the Linnaeus database. During the 1960s, every second person lacks class category
+and almost a third have missing information during the 1970s. For the most recent period, the information is fairly complete.
 
@@
 The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces).
-The status unmarried is often not explicitly noted in the source, which makes the status unknown in the database Poplink for people for people moving to our parishes.
-If no explicit status is given in the sources and partner is missing, the status has been set to unmarried in our data file.
+The status unmarried is often not explicitly noted in the source, which makes the status unknown in the database Poplink for people for people moving into our parishes.
+If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.
 
@@
 In the present study, we analyse mortality in the two most common disease
-groups in the ages we are investigating: Cardiovascular diseases and
-cancers. Compulsory reporting of cause of death became introduced in 1911
-for all Sweden, and not only the urban part which had been the case from
-1860. Classification systems have however changed. Sweden did not join ICD
-until 1952, but the historical causes of death in the Poplink database have
-been coded according to the ICD10 system. In the present study, we have
-identified the codes relevant for cardiovascular diseases and cancers in
-the ICD 8, 9 and 10.
+groups in the ages we are investigating -- cardiovascular diseases and
+cancers.
+National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year.
+Classification systems have however changed.
+Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
+In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
 
@@
 Notice the high proportion of other causes in the first two periods.
-This category include cases with unknown cause and unspecific causes which has implications for the cause-specific analyses.
-In the analyse section, we comment more on this.
-The total dominance of the two cause groups we are studying is apparent from the figure.
+This category include cases with unknown or unspecific causes which has implications for the cause-specific analyses (see later discussion).
+The total dominance of our two disease groups is apparent.
 
@@
 We have made separate analyses for the different periods.
-For the century 1851--1950 we have divided it into four periods of 25 years each.
-The nineteenth century period represents a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
+The century 1851--1950 is divided into four periods of 25 years each.
+The nineteenth century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
 During the years 1901--1950, industrialisation had started and the urban environments increased in population size.
 The modern welfare state began to develop during this time.
-For the period 1960 onwards, the analyses is made on decades except for the two last periods.
-Sweden at this time is a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
-A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system, not being as generous as before.
-The start of the last period (2008--2013) coincides with the international financial crisis that also had strong impact on Swedish economy.
+For the period 1960 onwards, there are separate analyses on decades except for the two last periods.
+Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
+A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to not being as generous as before.
+The start of the last period (2008--2013) coincides with the international financial crisis.
 
@@
 Education is here analysed only for the period 1990 onwards.
-Years of education are given in 7 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
-It ranges from basic education (less than nine years schooling) at level 1, to high academic degree (PhD and similar) at level 7.
+Years of education are given in 4 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
+It ranges from basic education (less than nine years schooling) at the lowest level, to high academic degree (PhD and similar) at the highest level.
 
@@
 \section{Discussion}
+
+The WHO Commission on Social Determinants of Health framework for the social determinants of health discusses how socio-economic position impacts on health.
+The influences are not direct but works through intermediate determinants related to exposure and vulnerability \citep{who10}.
+People are placed within socio-economic and political contexts, as well as value systems in the society.
+The socio-economic position imply different access to resources, power and prestige, all aspects of crucial importance.
+The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
+The importance of understanding people in their context has also been emphasized by for example Krieger \citet{kri05}.
+What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
 
@@
 Research (CEDAR) at Umeå University and with support from the Swedish
-Research Council (VR) (grant no. 421-2012-892). We thank Demographic Data
+Research Council (VR) (grant no. 421-2012-892) and the project Old age ... financed by the Swedish Research Council. We thank Demographic Data
 Base for giving us access to data and Jonas Helgertz, Department of

### 2017-05-24 icass.Rnw
Updating text in introduction and more

@@
 
-\date{May 22, 2017}
+\date{May 24, 2017}
 
@@
 Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
-Not only should we expect to find a direct effect of the lack of resources on health, we might even find an independent effect of inequality as \citet{wipi09} suggest.
+\citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases.
+According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease.
+Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health.
+Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences.
+\citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.
+
+Recent studies, investigating social inequality in health and mortality with micro-data, has however questioned the assumed pattern.
+These studies suggest that the modern pattern might be a fairly recent development, although the pattern differs depending on age, gender or historical context.
+When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small socio-economic differences or differences in the opposite direction than the expected during the first stages of the transition \citep{bevp11, bedr11, edli11,edbr12}, sharply contrasting against what we know from present-day society.
+
+Another interesting aspect of the developent of social inequality in mortality, is its relation to economic inequality.
+\citet{wipi09} argue that income inequality has an independent effect on mortality, separated from the direct effect of the actual different access to economic resources.
 They argue that unequal societies perform less well when it comes to health than equal ones.
-Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.
-Some recent studies have however indicated that the assumed pattern of distinct socio-economic health differences was not always apparent in pre- and early-transitional societies and some suggest that the modern pattern might be a fairly recent phenomenon \citep{bevp11}.
+Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world.
+This has initiated a vital scholarly debate and the topic has been extensively studied (see for example \citet{suka04, wado00}.
+How the association looked like in previous periods is unknown.
+This is of particular interest since the levels of inequality were very different from what we find during most of the recent decades.
+Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.Data are uncertain for the 19th century, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
+From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach the lowest point at around 1980.
+Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}.
+The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.
+
+It thus appears as if a more consistent social health divide appeared at some point during the mortality transition, at least in Sweden.
 Solid empirical evidence about the process is however lacking and studies focusing on the issue are still few.
@@
 
-\section{Previous research}
-
-Several studies have shown how important access to economic and other resources are for health, but it is not clear how this has developed in history.
-\citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases.
-According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease.
-Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health.
-Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences.
-\citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.
-When in recent years new studies have appeared, investigating social inequality in health and mortality with micro-data, the assumed pattern has been questioned.
-When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small socio-economic differences during the first stages of the transition \citep{bevp11, bedr11, edli11,edbr12}.
-The Swedish results show no consistent social hierarchy before 1900 but substantial differences towards the late 20th and early 21st century.
-It appears as if a more consequent social health divide developed in the period
-in-between. 
-\citet{wipi09} argue that income inequality has an independent effect on mortality, separated from that of the actual differential levels of economic resources.
-They find an positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world.
-This has initiated a vital scholarly debate and the topic has been extensively studied (see for example \citet{suka04, wado00}.
-How the association looked like in previous periods has however not been studied.
-This is of particular interest since the levels of inequality were very different from what we find during most of the recent decades.
-Data are uncertain for the 19th century, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
-From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach the lowest point at around 1980.
-Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}.
-The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.
-
 \section{The Skellefteå and Umeå regions}
@@
 
-\begin{figure}
-\includegraphics[height=19cm]{myfigs/skum}
-\caption{The Skellefteå and Umeå regions.}
-\label{fig:SKEUME}
-\end{figure}
+% \begin{figure}
+% \includegraphics[height=19cm]{myfigs/skum}
+% \caption{The Skellefteå and Umeå regions.}
+% \label{fig:SKEUME}
+% \end{figure}
 
@@
 
-The Skellefteå and Umeå regions (Figures~\ref{fig:SKEUME} and~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of the Gulf of Bothnia.
+The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of the Gulf of Bothnia.
 This was a remote part of the country, where communications with the rest of Sweden were restricted until the late 20th century, dominated by agriculture and characterise by a vulnerable economy.
-Several severe famines struck the regions during the 1800s, making living difficult for the population and increasing mortality \citep{segb14}.
+The county was vulnerable to harvest failures and several severe famines struck the regions during the 1800s.
+One living long in memory was the famine of 1867–-1868 \citep{segb14}.
 During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
@@
 The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}.
-Towards the end of the 19th century, railways construction reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures.
+Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures.
 The inland was particularly remote.
-During the 19th century, our regions parts became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
-Diseases reached to the places that earlier were spared.
-The Skellefteå region in Poplink consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
+During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
+Diseases reached to places that earlier were spared.
+
+The Skellefteå region consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
 The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
 The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production.
-The region was vulnerable to harvest failures, one living long in memory was the famine of 1867–-1868 \citep{segb14}.
-Towards the end of the century, the region became connected to the Swedish railway system, thus improving the communications and contacts with other parts of the country.
 During the 20th century, industrialisation took place, partly by the opening of mines in the inlands.
@@
 
-The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included.
-This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time.
+The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality where another three parishes are included.
+This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy as well for a long time.
 Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period.
@@
 
-The extension in time of data at the Demographic Data Base, Umeå University (DDB) makes it possible to investigate the long-term development of social inequality in mortality by including the period between 1900 and 1950s when almost no micro-data has been available for research.
-The databases provide us with with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.The data for the present study comes from two large population databases at the DDB, Umeå University.
+The extension in time of data at the Demographic Data Base, Umeå University (DDB) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
+The databases provide us with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.The data for the present study comes from two large population databases at the DDB, Umeå University.
 The early period is covered by
@@
 the region. The records are linked within but not between the regions. Data
-from the regions are accessed for the period 1901--1950s.
-
-The other large data-set comes from the Linnaeus database
-\citep{malm10}. It is based on different linked population registers from
+from Poplink are accessed for the period 1851--1950s.
+
+The other large data-set is extracted from the Linnaeus database
+\citep{malm10}. Linnaeus is based on different linked population registers from
 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
@@
 For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
-The construction of these data-sets for analysis is presented below.
 
@@
 
-In the data-set analysed here, all individuals 40 years and older and ever being resident in either
+In the data-set analysed here, all individuals 40 years and older ever being resident in either
 of the regions are included. The data file contains the variables on sex,
@@
 This is however not the case with the Linnaeus database.
-For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare for defining the observation periods as well as their social class (see below) in our analyses.
+For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare to identify the observation periods in our analyses.
 We have however no exact information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
@@
 For the last period, 2002--2013, we use the yearly population registers (also known as LISA data from Statistics Sweden), together with information on deaths from National Board of Health and Welfare.
-This data allows a better control of the presence periods in the studied regions.
+This data allows good control of the presence periods in the studied regions.
 
@@
 Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes.
-The obvious problems connected with using the same classification scheme covering a long time period will be addresed in the discussion section.
 Social position is treated as a time-dependent variable.
@@
 The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
-The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1990 to 2001 where we lack information.
-In the censuses, we have solved this by extending the periods for which we consider the defined social class be valid.
+The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information.
+In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid.
 Thus, the social class in the census 1990 is considered to be kept until 2000.
-We consider that the social position is kept after retirement (in Sweden usually at age 65).
+This strategy is also used for other cases of missing social class.
+Furthermore, we consider the social position after retirement (in Sweden usually at age 65) to be the same as the last during working age.
 
 The use of different data sets, forces us to apply different approaches towards defining the social position.
-In the Poplink data, social class is based on the occupational titles given in the parish records.
-Some titles are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
-We consider however their validity to be sufficient for identification of social position.
-Another problem is that we in Poplink usually get the occupation only for the head of household, thus under-estimating the female working-force participation and the occupations of adult children residing with their parents \citep{vik10}.
+In the Poplink data, some titles are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
+We consider however the validity of the information to be sufficient for identification of social position.
+Another problem is that Poplink usually only provides the occupation only for the head of household, thus under-estimating female working-force participation as well as that of adult children residing with their parents \citep{vik10}.
 In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
@@
 
-Female labour is much better covered in the Linnaeus database, both in the censuses and the population registers.
-Another difference is that it is much more difficult to define households in the same way as in Poplink from the Linnaeus database.
+Female labour is much better covered in the Linnaeus database.
+A complication is however that it is  difficult to define households in the same way as in Poplink.
 Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became common and female labour force participation developed as the norm in Sweden during this period.
-The occupation of the husband as signifier of social class became if not obsolete at least less relevant.
+The occupation of the husband as signifier of social class became, if not obsolete at least, less relevant.
 All included persons are signified by their individual occupation for this period.
@@
 
-Due to the construction of the time-dependent data, some records report unknown social position.
-This has been updated according to rules in order to optimise available information.
-Those that still lack information on social class are not included in the analysis, something that have consequences for elderly living in the 1960s in particular, since many were already retired and thus without occupation when we first observe them in the 1960 census.
-This, together with the circumstance that many women had no occupation in the census 1960 explains the large proportion where we have not been able to determine class during that decade.
-As a consequence, we leave out the 1960s and 1970s in some analyses.
-
-The classification scheme used here is  based on HISCLASS \citep{lema11}, a class-based system.
+The classification scheme we use is  based on HISCLASS \citep{lema11}, a class-based system.
 The scheme uses HISCO for assigning categories.
-HISCO \citet{lema02} (Historical International Standard Classifications of Occupations) is a classification system of historical occupations.
-HISCO categorize occupations according to tasks and skills.
-We have merged the original 12 classes in HISCLASS, based on position on the labour market, ending up with the following three and an additional fourth category with unknown class four:
+HISCO (Historical International Standard Classifications of Occupations) \citet{lema02}  is a classification system of historical occupations, categorizing occupations according to tasks and skills.
+We have merged the original 12 classes in HISCLASS, based on position on the labour market, ending up with the following three and an additional fourth category with unknown class:
 
@@
 
-For the corresponding numbers, with missing values ({\tt NA}) included, see
-Table~\ref{tab:nana}.The proportion with missing class was very low
+Those that we first observe in the first included censuses often lack stated occupations.
+This is especially the case for women and elderly being retired.
+The proportion with missing values ({\tt NA}) included is presented in Table~\ref{tab:nana}.
+It was very low
 during the periods until 1950, between 4 and 8 \% lacked information
 allowing determination of class. This increased considerably during the first two decades of data brom the Linnaeus database. During the 1960s, every second person lacks class category
-and almost a third have missing information during the 1970s. For the most recent period, the information is fairly complete.
+and almost a third have missing information during the 1970s.
+As a consequence, we only report the results for men 40--64 for the 1960s and 1970s.
+For the most recent periods, the information is fairly complete.
 
@@
 
-Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.
+Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.The figure indicate an under-registration of those married and and over-registration of previously married during the 1960s.
+(WHY?)
 
@@
 \subsection{Education}
-Education is here analysed only for the period 1990 onwards.
+Education is here analysed only for the period 1990 to 2005.
 Years of education are given in 4 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
 It ranges from basic education (less than nine years schooling) at the lowest level, to high academic degree (PhD and similar) at the highest level.
+Which are the new categories?)
 
@@
 which 68\,244 are observed to die.
-This makes it very inconvenient use the nowadays so popular semi-parametric
+This makes it very inconvenient to use the nowadays so popular semi-parametric
 technique known as
@@
 
-\subsubsection*{The effect of income and education, 1990--2005}
-
-We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death.
+% \subsubsection*{The effect of income and education, 1990--2005}
+% 
+% We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death.
 It is based on a separate analysis with income and education as explanatory variables.
-
-<<41lissyoung>>=
-library(eha)
-library(xtable)
-##load("data/loTab.rda")
-loTab$period <- factor(loTab$period)
-liss4064 <- loTab[loTab$age < 64.5, ]
-@
-
-The statistical significance of the covariates are shown in
-Table~\ref{tab:41liswom}
-<<41liswom, results = 'asis'>>=
-fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-##cat("Women:\n")
-drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
-
-<<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
-source("R/plotWEduc.R")
-plotWEduc(fitw1)
-@
-
-And for men:
-
-<<41lismen, results='asis'>>=
-fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-##cat("\nMen:\n")
-drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-<<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm1)
-@
-
-<<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
-#oldpar <- par(mfrow = c(1, 2))
-#plot(fitw1, fn = "cum", col = dcol, main = "Women")
-#plot(fitm1, fn = "cum", col = dcol, main = "Men")
-#par(oldpar)
-##summary(fitw1)
-##summary(fitm1)
-@
-
-%See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<41tablisswom1, results = "asis">>=
-#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-#                 data = liss4064[liss4064$sex == "female", ])
-#dr <- drop1(fitwo1, test = "Chisq")
-#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-#    label = "tab:tablisswom1")
-@
-
-%See Table~\ref{tab:tablissman1} the PH regression for men.
-<<41tablissman1, results = "asis">>=
-#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-#                 data = liss4064[liss4064$sex == "male", ])
-#dr <- drop1(fitma1, test = "Chisq")
-#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
- #   label = "tab:tablissman1")
-@
-
-The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
+% 
+% <<41lissyoung>>=
+% library(eha)
+% library(xtable)
+% ##load("data/loTab.rda")
+% loTab$period <- factor(loTab$period)
+% liss4064 <- loTab[loTab$age < 64.5, ]
+% @
+% 
+% 
+% The statistical significance of the covariates are shown in
+% Table~\ref{tab:41liswom}
+% <<41liswom, results = 'asis'>>=
+% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+% ##cat("Women:\n")
+% drw <- drop1(fitw1, test = "Chisq")
+% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
+% ##round(summary(fitw1)$coef[-(1:13), ], 4)
+% @
+% 
+% The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
+% 
+% <<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
+% source("R/plotWEduc.R")
+% plotWEduc(fitw1)
+% @
+% 
+% 
+% And for men:
+% 
+% <<41lismen, results='asis'>>=
+% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+% ##cat("\nMen:\n")
+% drm <- drop1(fitm1, test = "Chisq")
+% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
+% ##round(summary(fitm1)$coef[-(1:13), ], 4)
+% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% @
+% 
+% <<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitm1)
+% @
+% 
+% 
+% <<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
+% #oldpar <- par(mfrow = c(1, 2))
+% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
+% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
+% #par(oldpar)
+% ##summary(fitw1)
+% ##summary(fitm1)
+% @
+% 
+% %See Table~\ref{tab:tablisswom1} the PH regression for women.
+% <<41tablisswom1, results = "asis">>=
+% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
+% #                 data = liss4064[liss4064$sex == "female", ])
+% #dr <- drop1(fitwo1, test = "Chisq")
+% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
+% #    label = "tab:tablisswom1")
+% @
+% 
+% %See Table~\ref{tab:tablissman1} the PH regression for men.
+% <<41tablissman1, results = "asis">>=
+% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
+% #                 data = liss4064[liss4064$sex == "male", ])
+% #dr <- drop1(fitma1, test = "Chisq")
+% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
+%  #   label = "tab:tablissman1")
+% @
+% 
+% The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
 More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system.
@@
 
-\subsubsection*{The effect of income and education on cardiovascular death, 1990--2005}
-
-The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
-income and education as explanatory variables.
+% \subsubsection*{The effect of income and education on cardiovascular death, 1990--2005}
+% 
+% The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
+% income and education as explanatory variables.
 The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2} for women and Table~\ref{tab:41lismen2} for men.
 It confirms that most of the variables are strongly significant except
-
-<<41lissyoung2>>=
-library(eha)
-library(xtable)
-##load("data/loTab.rda")
-##loTab$period <- factor(loTab$period)
-liss6589 <- loTab[loTab$age > 64.5, ]
-@
-
-<<41liswom2, results = 'asis'>>=
-fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-##cat("Women:\n")
-drw2 <- drop1(fitw2, test = "Chisq")
-xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-<<41lismen2, results='asis'>>=
-fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-##cat("\nMen:\n")
-drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-
-<<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitw2)
-@
-
-And for men:
-
-<<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm2)
-@
-
-The pattern is similar to what we find for all-cause mortality.
+% 
+% <<41lissyoung2>>=
+% library(eha)
+% library(xtable)
+% ##load("data/loTab.rda")
+% ##loTab$period <- factor(loTab$period)
+% liss6589 <- loTab[loTab$age > 64.5, ]
+% @
+% 
+% 
+% 
+% <<41liswom2, results = 'asis'>>=
+% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+% ##cat("Women:\n")
+% drw2 <- drop1(fitw2, test = "Chisq")
+% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
+% ##round(summary(fitw1)$coef[-(1:13), ], 4)
+% @
+% 
+% 
+% <<41lismen2, results='asis'>>=
+% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+% ##cat("\nMen:\n")
+% drm2 <- drop1(fitm2, test = "Chisq")
+% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
+% ##round(summary(fitm1)$coef[-(1:13), ], 4)
+% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% @
+% 
+% The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
+% 
+% <<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitw2)
+% @
+% 
+% 
+% And for men:
+% 
+% 
+% 
+% <<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitm2)
+% @
+% 
+% The pattern is similar to what we find for all-cause mortality.
 The higher educational level, the lower cardiovascular mortality, and more income leads to better survival.
@@
 
-\subsubsection*{The effect of income and education on cancer mortality, 1990--2005}
-
-In the previous section, we concluded that there were no social inequality in cancer mortality according to hisclass.
+% \subsubsection*{The effect of income and education on cancer mortality, 1990--2005}
+% 
+% In the previous section, we concluded that there were no social inequality in cancer mortality according to hisclass.
 The question here is if the lacking socio-economic differentiation holds also when using education and income as categorising variables.
 The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables.
-
-<<12lissyoung>>=
-library(eha)
-library(xtable)
-##load("data/loTab.rda")
-##loTab$period <- factor(loTab$period)
-liss4064 <- loTab[loTab$age < 64.5, ]
-@
-
-The statistical significance of the covariates are shown in Table~\ref{tab:12liswom} for women and Table~\ref{tab:12lismen} for men. The variables of interest here are all significant.
-<<12liswom, results = 'asis'>>=
-fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-##cat("Women:\n")
-drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cancer death.", label = "tab:12liswom")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-<<12lismen, results='asis'>>=
-fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-##cat("\nMen:\n")
-drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.", label = "tab:12lismen")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}.
-
-<<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-source("R/plotWEduc.R")
-plotWEduc(fitw1)
-@
-
-And for men:
-
-<<12lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm1)
-@
-
-<<12plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cancer death.", fig.height = 4>>=
-#oldpar <- par(mfrow = c(1, 2))
-#plot(fitw1, fn = "cum", col = dcol, main = "Women")
-#plot(fitm1, fn = "cum", col = dcol, main = "Men")
-#par(oldpar)
-##summary(fitw1)
-##summary(fitm1)
-@
-
-%See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<12tablisswom1, results = "asis">>=
-#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-#                 data = liss4064[liss4064$sex == "female", ])
-#dr <- drop1(fitwo1, test = "Chisq")
-#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-#    label = "tab:tablisswom1")
-@
-
-%See Table~\ref{tab:tablissman1} the PH regression for men.
-<<12tablissman1, results = "asis">>=
-#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-#                 data = liss4064[liss4064$sex == "male", ])
-#dr <- drop1(fitma1, test = "Chisq")
-#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
- #   label = "tab:tablissman1")
-@
-
-When it comes to the association with education and income, we find the similar pattern as those for all-cause mortality and cardiovascular mortality.
+% 
+% <<12lissyoung>>=
+% library(eha)
+% library(xtable)
+% ##load("data/loTab.rda")
+% ##loTab$period <- factor(loTab$period)
+% liss4064 <- loTab[loTab$age < 64.5, ]
+% @
+% 
+% 
+% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom} for women and Table~\ref{tab:12lismen} for men. The variables of interest here are all significant.
+% <<12liswom, results = 'asis'>>=
+% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+% ##cat("Women:\n")
+% drw <- drop1(fitw1, test = "Chisq")
+% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cancer death.", label = "tab:12liswom")
+% ##round(summary(fitw1)$coef[-(1:13), ], 4)
+% @
+% 
+% <<12lismen, results='asis'>>=
+% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+% ##cat("\nMen:\n")
+% drm <- drop1(fitm1, test = "Chisq")
+% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.", label = "tab:12lismen")
+% ##round(summary(fitm1)$coef[-(1:13), ], 4)
+% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% @
+% 
+% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}.
+% 
+% <<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
+% source("R/plotWEduc.R")
+% plotWEduc(fitw1)
+% @
+% 
+% 
+% And for men:
+% 
+% 
+% <<12lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cancer death.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitm1)
+% @
+% 
+% 
+% <<12plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cancer death.", fig.height = 4>>=
+% #oldpar <- par(mfrow = c(1, 2))
+% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
+% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
+% #par(oldpar)
+% ##summary(fitw1)
+% ##summary(fitm1)
+% @
+% 
+% %See Table~\ref{tab:tablisswom1} the PH regression for women.
+% <<12tablisswom1, results = "asis">>=
+% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
+% #                 data = liss4064[liss4064$sex == "female", ])
+% #dr <- drop1(fitwo1, test = "Chisq")
+% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
+% #    label = "tab:tablisswom1")
+% @
+% 
+% %See Table~\ref{tab:tablissman1} the PH regression for men.
+% <<12tablissman1, results = "asis">>=
+% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
+% #                 data = liss4064[liss4064$sex == "male", ])
+% #dr <- drop1(fitma1, test = "Chisq")
+% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
+%  #   label = "tab:tablissman1")
+% @
+% 
+% When it comes to the association with education and income, we find the similar pattern as those for all-cause mortality and cardiovascular mortality.
 There is a strong effect for both men and women of having more education and income.
@@
 
-\subsubsection*{The effect of income and education on cancer death, 1990--2005}
-
-Thus, we find indications of social differences in the later decades, particularly between the elite and the workers. The alternative analysis for the last period is based on a separate analysis with
-income and education as explanatory variables.
-
-<<12lissyoung2>>=
-library(eha)
-library(xtable)
-##load("data/loTab.rda")
-##loTab$period <- factor(loTab$period)
-liss6589 <- loTab[loTab$age > 64.5, ]
-@
-
-The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men.
-<<12liswom2, results = 'asis'>>=
-fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-##cat("Women:\n")
-drw2 <- drop1(fitw2, test = "Chisq")
-xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-<<12lismen2, results='asis'>>=
-fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-##cat("\nMen:\n")
-drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
-
-<<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitw2)
-@
-
-And for men:
-
-%% <<12lismen2, results='asis'>>=
-%% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
-%% ##cat("\nMen:\n")
-%% drm2 <- drop1(fitm2, test = "Chisq")
-%% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
-%% round(summary(fitm1)$coef[-(1:13), ], 4)
-%% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-%% @
-
-<<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm2)
-@
-
-The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income.
+% \subsubsection*{The effect of income and education on cancer death, 1990--2005}
+% 
+% Thus, we find indications of social differences in the later decades, particularly between the elite and the workers.
+The alternative analysis for the last period is based on a separate analysis with
+% income and education as explanatory variables.
+% 
+% <<12lissyoung2>>=
+% library(eha)
+% library(xtable)
+% ##load("data/loTab.rda")
+% ##loTab$period <- factor(loTab$period)
+% liss6589 <- loTab[loTab$age > 64.5, ]
+% @
+% 
+% 
+% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men.
+% <<12liswom2, results = 'asis'>>=
+% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+% ##cat("Women:\n")
+% drw2 <- drop1(fitw2, test = "Chisq")
+% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
+% ##round(summary(fitw1)$coef[-(1:13), ], 4)
+% @
+% 
+% <<12lismen2, results='asis'>>=
+% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+% ##cat("\nMen:\n")
+% drm2 <- drop1(fitm2, test = "Chisq")
+% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
+% ##round(summary(fitm1)$coef[-(1:13), ], 4)
+% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% @
+% 
+% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
+% 
+% <<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitw2)
+% @
+% 
+% 
+% And for men:
+% 
+% %% <<12lismen2, results='asis'>>=
+% %% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
+% %% ##cat("\nMen:\n")
+% %% drm2 <- drop1(fitm2, test = "Chisq")
+% %% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
+% %% round(summary(fitm1)$coef[-(1:13), ], 4)
+% %% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% %% @
+% 
+% <<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitm2)
+% @
+% 
+% The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income.
 We find the same consistent pattern as we have found for all-cause mortality and cardiovascular mortality for this period.
@@
 The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
-The importance of understanding people in their context has also been emphasized by for example Krieger \citet{kri05}.
+The importance of understanding people in their context has also been emphasized by for example \citet{kri05}.
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.

### 2017-05-24 icass.Rnw
Removing some figures

@@
 
-We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
-periods, in order to illustrate the process. They are adjusted for {\tt civil status}. We present separate analyses  for women and men, see
-Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.
+% We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
+% periods, in order to illustrate the process. They are adjusted for {\tt civil status}. We present separate analyses  for women and men, see
+% Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.
 
@@
 
-<<hazcalw>>=
-source("R/getPHglm2.R")
-##nper <- 4
-nper <- length(labb)
-resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
-            nper, labb)
-@
-
-<<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
-source("R/plotHaz.glm.R")
-
-plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
-@
-
-<<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
-plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
-@
-
-From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
-the idea of proportional hazards is not reasonable. We therefore use another approach.
+% <<hazcalw>>=
+% source("R/getPHglm2.R")
+% ##nper <- 4
+% nper <- length(labb)
+% resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+%             nper, labb)
+% resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+%             nper, labb)
+% @
+% 
+% 
+% <<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
+% source("R/plotHaz.glm.R")
+% 
+% plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+% @
+% 
+% <<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
+% plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
+% @
+% 
+% From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
+% the idea of proportional hazards is not reasonable. We therefore use another approach.
 

### 2017-05-26 icass.Rnw
reorganising ...

@@
 century. The increase of academics in connection to the establishment of
-Umeå university explains much of this increase.
+Umeå University explains much of this increase.
 Within the middle group, there has been a transfer from a majority of farmer during the 19th century to lower white collar from the middle of the 20th century.
 
-<<hiscskum, fig.cap = "Exposure by hisclass (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
-xx <- with(vb0Tab, tapply(exposure, list(period = period, hisclass = hisclass), sum))
+<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
+xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+yy <- round(prop.table(xx, 1) * 100, 1)
+barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+@
+
+<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
+xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
 yy <- round(prop.table(xx, 1) * 100, 1)
@@
 
-<<nana, results = 'asis'>>=
+<<nanam, results = 'asis'>>=
 library(xtable)
@@
 ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-x <- with(vb0, table(period, hisclass, useNA = "ifany"))
+x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
+x <- round(prop.table(x, 1) * 100)
+xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
+@
+
+<<nanaw, results = 'asis'>>=
+##library(xtable)
+##load("data/vb0.rda")
+##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
+x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
 x <- round(prop.table(x, 1) * 100)
@@
 
-\section{Results, all causes of death}
+\section{The effect of Hisclass on All causes of death}
 
@@
 
-% <<hazcalw>>=
-% source("R/getPHglm2.R")
-% ##nper <- 4
-% nper <- length(labb)
-% resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
-%             nper, labb)
-% resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
-%             nper, labb)
-% @
+<<hazcalw>>=
+source("R/getPHglm2.R")
+##nper <- 4
+nper <- length(labb)
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+            nper, labb)
+@
+
 % 
@@
 
-\subsubsection*{The effect of income and education, 1990--2005}
-
-A common alternative to a class analysis is to use income or education for investigating socio-economic health differences.
-Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others disadvantaged.
-For the time being, we have to restrict such a comparative analysis to the last decades.
-
-The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables and with controls of period and marital status.
-
-<<lissyoung>>=
-library(eha)
-library(xtable)
-liss4064 <- loTab[loTab$age < 64.5, ]
-@
-
-The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
-<<liswom, results = 'asis'>>=
-fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-##cat("Women:\n")
-drw <- drop1(fitw1, test = "Chisq")
-xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
-
-<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
-source("R/plotWEduc.R")
-plotWEduc(fitw1)
-@
-
-And for men:
-
-<<lismen, results='asis'>>=
-fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-##cat("\nMen:\n")
-drm <- drop1(fitm1, test = "Chisq")
-xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm1)
-@
-
-<<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
-#oldpar <- par(mfrow = c(1, 2))
-#plot(fitw1, fn = "cum", col = dcol, main = "Women")
-#plot(fitm1, fn = "cum", col = dcol, main = "Men")
-#par(oldpar)
-##summary(fitw1)
-##summary(fitm1)
-@
-
-%See Table~\ref{tab:tablisswom1} the PH regression for women.
-<<tablisswom1, results = "asis">>=
-#fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-#                 data = liss4064[liss4064$sex == "female", ])
-#dr <- drop1(fitwo1, test = "Chisq")
-#ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-#    label = "tab:tablisswom1")
-@
-
-%See Table~\ref{tab:tablissman1} the PH regression for men.
-<<tablissman1, results = "asis">>=
-#fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-#                 data = liss4064[liss4064$sex == "male", ])
-#dr <- drop1(fitma1, test = "Chisq")
-#ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
- #   label = "tab:tablissman1")
-@
-Both education and income were strongly significant, as well as marital status as expected.
-Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
+% \subsubsection{The effect of income and education, 1986--2013}
+% 
+% A common alternative to a class analysis is to use income or education for investigating socio-economic health differences. 
+% Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
+% disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
+% 
+% The analysis for the period 1990--2005 is based on a separate analysis with 
+% income and education as explanatory variables and with controls of period 
+% and marital status.
+% 
+% <<lissyoung>>=
+% library(eha)
+% library(xtable)
+% liss4064 <- loTab[loTab$age < 64.5, ]
+% con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+% ##load("data/lisaold.rda")
+% load(con)
+% close(con)
+% @
+% 
+% Both education and income had a strong, statistically significant, effect on survival, as expected. 
+% 
+% \paragraph{Education, 1990--2013}
+% 
+% <<educ>>=
+% 
+% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
+% educ$period <- factor(educ$period)
+% educ <- educ[!is.na(educ$educ), ]
+% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
+% ##educ$educ <- relevel(educ$educ, ref = "middle")
+% educ$civst <- relevel(educ$civst, ref = "married")
+% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
+%              data = educ[educ$sex == "male", ])
+% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
+%              data = educ[educ$sex == "female", ])
+% cols <- c("red", "darkgreen", "black")
+% @
+% 
+% <<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
+% oldpar <- par(mfrow = c(1, 2))
+% 
+% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+% 
+% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+% 
+% par(oldpar)
+% @
+% 
+% Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
 The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
-The same can be said concerning income.
-Higher incomes result in lower mortality.
-This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden.
-Notice however that we cannot definitely establish a causal relationship between income and mortality in particular.
-Having low incomes may be caused by bad health.
+% 
+% \paragraph{Income, 1986--2005}
+% 
+% <<femeduc>>=
+% income <- lisaold[!is.na(lisaold$income), ]
+% income$period <- factor(income$period)
+% ##income <- income[!is.na(income$income), ]
+% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
+% ##income$income <- relevel(income$income, ref = "low")
+% income$civst <- relevel(income$civst, ref = "married")
+% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
+%              data = income[income$sex == "male", ])
+% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
+%              data = income[income$sex == "female", ])
+% cols <- c("darkgreen", "red", "blue", "black")
+% 
+% @
+% 
+% <<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
+% oldpar <- par(mfrow = c(1, 2))
+% 
+% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+% 
+% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+% 
+% par(oldpar)
+% @
+% 
+% Higher incomes result in lower mortality. This is also what could be expected, 
+% and conforms to what we know of social health differences in present-day Sweden. 
+% Notice however that we cannot establish a causal relationship between 
+% income and mortality. Bad health, for instance, may be a confounder.
+% 
+% %%The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
+% %%<<liswom, results = 'asis'>>=
+% %%fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
+% %%##cat("Women:\n")
+% %%drw <- drop1(fitw1, test = "Chisq")
+% %%xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
+% %##round(summary(fitw1)$coef[-(1:13), ], 4)
+% %@
+% 
+% %%The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
+% 
+% %%<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
+% %%source("R/plotWEduc.R")
+% %%plotWEduc(fitw1)
+% %%@
+% 
+% 
+% %%And for men:
+% 
+% %%<<lismen, results='asis'>>=
+% %%fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
+% %%##cat("\nMen:\n")
+% %%drm <- drop1(fitm1, test = "Chisq")
+% %%xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
+% %%##round(summary(fitm1)$coef[-(1:13), ], 4)
+% %%dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% %%@
+% 
+% %%<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
+% %%#source("R/plotWEduc.R")
+% %%plotWEduc(fitm1)
+% %%@
+% 
+% % 
+% % <<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
+% % #oldpar <- par(mfrow = c(1, 2))
+% % #plot(fitw1, fn = "cum", col = dcol, main = "Women")
+% % #plot(fitm1, fn = "cum", col = dcol, main = "Men")
+% % #par(oldpar)
+% % ##summary(fitw1)
+% % ##summary(fitm1)
+% % @
+% % 
+% % %See Table~\ref{tab:tablisswom1} the PH regression for women.
+% % <<tablisswom1, results = "asis">>=
+% % #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
+% % #                 data = liss4064[liss4064$sex == "female", ])
+% % #dr <- drop1(fitwo1, test = "Chisq")
+% % #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
+% % #    label = "tab:tablisswom1")
+% % @
+% % 
+% % %See Table~\ref{tab:tablissman1} the PH regression for men.
+% % <<tablissman1, results = "asis">>=
+% % #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
+% % #                 data = liss4064[liss4064$sex == "male", ])
+% % #dr <- drop1(fitma1, test = "Chisq")
+% % #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
+% %  #   label = "tab:tablissman1")
+% % @
+%
 
@@
 
-\subsubsection*{The effect of income and education, 1990--2005}
-
-The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting.
+% \subsubsection{The effect of income and education, 1990--2005}
+% 
+% The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting.
 As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
-
-<<lissyoung2>>=
-library(eha)
-library(xtable)
-##load("data/loTab.rda")
-##loTab$period <- factor(loTab$period)
-liss6589 <- loTab[loTab$age > 64.5, ]
-@
-
-The statistical significance of the covariates are shown in
-Table~\ref{tab:liswom2}
-<<liswom2, results = 'asis'>>=
-fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-##cat("Women:\n")
-drw2 <- drop1(fitw2, test = "Chisq")
-xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
-##round(summary(fitw1)$coef[-(1:13), ], 4)
-@
-
-The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
-
-<<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitw2)
-@
-
-And for men:
-
-<<lismen2, results='asis'>>=
-fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-##cat("\nMen:\n")
-drm2 <- drop1(fitm2, test = "Chisq")
-xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
-##round(summary(fitm1)$coef[-(1:13), ], 4)
-dcol <- c("blue", "darkgreen", "darkgreen", "red")
-@
-
-<<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
-#source("R/plotWEduc.R")
-plotWEduc(fitm2)
-@
-
-These figures show clear associations both with income and education and for women as well as for men.
+% 
+% <<lissyoung2>>=
+% library(eha)
+% library(xtable)
+% ##load("data/loTab.rda")
+% ##loTab$period <- factor(loTab$period)
+% liss6589 <- loTab[loTab$age > 64.5, ]
+% @
+% 
+% 
+% The statistical significance of the covariates are shown in
+% Table~\ref{tab:liswom2}
+% <<liswom2, results = 'asis'>>=
+% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
+% ##cat("Women:\n")
+% drw2 <- drop1(fitw2, test = "Chisq")
+% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
+% ##round(summary(fitw1)$coef[-(1:13), ], 4)
+% @
+% 
+% The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
+% 
+% <<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitw2)
+% @
+% 
+% 
+% And for men:
+% 
+% <<lismen2, results='asis'>>=
+% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
+% ##cat("\nMen:\n")
+% drm2 <- drop1(fitm2, test = "Chisq")
+% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
+% ##round(summary(fitm1)$coef[-(1:13), ], 4)
+% dcol <- c("blue", "darkgreen", "darkgreen", "red")
+% @
+% 
+% <<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
+% #source("R/plotWEduc.R")
+% plotWEduc(fitm2)
+% @
+% 
+% These figures show clear associations both with income and education and for women as well as for men.
 More years of education resulted in lower mortality.
@@
 
-\section{Results, cardiovascular death}
+\section{The effect of Hisclass on cardiovascular death}
 
@@
 
-% \subsubsection*{The effect of income and education, 1990--2005}
+% \subsubsection{The effect of income and education, 1990--2005}
 % 
@@
 
-% \subsubsection*{The effect of income and education on cardiovascular death, 1990--2005}
+% \subsubsection{The effect of income and education on cardiovascular death, 1990--2005}
 % 
@@
 
-\section{Results, cancer death}
+\section{The effect of Hisclass on cancer death}
 
@@
 
-% \subsubsection*{The effect of income and education on cancer mortality, 1990--2005}
+% \subsubsection{The effect of income and education on cancer mortality, 1990--2005}
 % 
@@
 
-% \subsubsection*{The effect of income and education on cancer death, 1990--2005}
+% \subsubsection{The effect of income and education on cancer death, 1990--2005}
 % 
@@
 %%{\huge Här ska 'conclusion.tex' in !!!}
+
+\section{The effect of Education on All causes of death, 1990--2013}
+
+A common alternative to a class analysis is to use income or education for investigating socio-economic health differences. 
+Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
+disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
+
+The analysis for the period 1990--2005 is based on a separate analysis with 
+income and education as explanatory variables and with controls of period 
+and marital status.
+
+<<lissyoung3>>=
+library(eha)
+library(xtable)
+liss4064 <- loTab[loTab$age < 64.5, ]
+con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+##load("data/lisaold.rda")
+load(con)
+close(con)
+@
+
+Both education and income had a strong, statistically significant, effect on survival, as expected.
+
+<<educ>>=
+
+educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
+educ$period <- factor(educ$period)
+educ <- educ[!is.na(educ$educ), ]
+levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
+##educ$educ <- relevel(educ$educ, ref = "middle")
+educ$civst <- relevel(educ$civst, ref = "married")
+fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
+             data = educ[educ$sex == "male", ])
+fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
+             data = educ[educ$sex == "female", ])
+cols <- c("red", "darkgreen", "black")
+@
+
+<<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
+oldpar <- par(mfrow = c(1, 2))
+
+plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
+     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+axis(1)
+axis(2, at = c(0, 0.5, 1.0), las = 1)
+abline(h = 0.5, lty = 3)
+box()
+legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+
+plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
+     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+axis(1)
+axis(2, at = c(0, 0.5, 1.0), las = 1)
+abline(h = 0.5, lty = 3)
+box()
+legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+
+par(oldpar)
+@
+
+Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
+The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
+
+\section{The effect of Education on Cardiovascular death}
+
+\section{The effect of Education on Cancer death}
+
+\section{The effect of Income on All causes of death, 1986--2005}
+
+<<femeduc>>=
+income <- lisaold[!is.na(lisaold$income), ]
+income$period <- factor(income$period)
+##income <- income[!is.na(income$income), ]
+##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
+##income$income <- relevel(income$income, ref = "low")
+income$civst <- relevel(income$civst, ref = "married")
+fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
+             data = income[income$sex == "male", ])
+fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
+             data = income[income$sex == "female", ])
+cols <- c("darkgreen", "red", "blue", "black")
+
+@
+
+<<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
+oldpar <- par(mfrow = c(1, 2))
+
+plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
+     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+axis(1)
+axis(2, at = c(0, 0.5, 1.0), las = 1)
+abline(h = 0.5, lty = 3)
+box()
+legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+
+plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
+     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+axis(1)
+axis(2, at = c(0, 0.5, 1.0), las = 1)
+abline(h = 0.5, lty = 3)
+box()
+legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+
+par(oldpar)
+@
+
+Higher incomes result in lower mortality. This is also what could be expected, 
+and conforms to what we know of social health differences in present-day Sweden. 
+Notice however that we cannot establish a causal relationship between 
+income and mortality. Bad health, for instance, may be a confounder.
+
+\section{The effect of Income on Cardiovascular death}
+
+\section{The effect of Income on Cancer death}
 
@@
 
-\section*{Acknowledgements}
+\section{Acknowledgements}
 

### 2017-05-28 icass.Rnw
reorganising and adding cause-specific to education and income

@@
 %\usepackage[figuresonly]{endfloat}
-\usepackage{endfloat}
-
-\renewcommand{\efloatseparator}{\vfill}
+%\usepackage[tablesonly]{endfloat}
+
+%\renewcommand{\efloatseparator}{\vfill}
 
 % patch \efloat@iwrite to use \protected@write
-\makeatletter
-\renewcommand\efloat@iwrite[1]{%
-   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
+%\makeatletter
+%\renewcommand\efloat@iwrite[1]{%
+%   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
+%\makeatother
 
@@
 
-\date{May 24, 2017}
+\date{May 28, 2017}
 
@@
 
-Another interesting aspect of the developent of social inequality in mortality, is its relation to economic inequality.
+Another interesting aspect of the development of social inequality in mortality, is its relation to economic inequality.
 \citet{wipi09} argue that income inequality has an independent effect on mortality, separated from the direct effect of the actual different access to economic resources.
@@
 \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
-\item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 onwards?
+\item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
 \end{enumerate}
@@
 The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production.
-During the 20th century, industrialisation took place, partly by the opening of mines in the inlands.
+During the 20th century, industrialisation took place, partly by the opening of mines in the in-lands.
 This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.
@@
 
-The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality where another three parishes are included.
+The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 on-wards of Umeå municipality where another three parishes are included.
 This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy as well for a long time.
@@
 birth date, death date, first and last date of observation and type of
-entrance/exit.Total number of persons are 84.599 leading to 35.001 deaths in Poplink and 197.864 persons leading to 66297 deaths in the Linnaeus database.
-(SKA VI HA FIGUREN MED THE SAMPLING FRAME HÄR?)
+entrance/exit.Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
+
+<<lexis2,fig.cap="The sampling frame.", fig.height = 5>>=
+plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
+     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
+     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
+axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(2, las = 2, at = c(0, 40, 65, 90))
+axis(4, las = 2, at = c(0, 40, 65, 90))
+box()
+abline(h = 0)
+lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
+      col = "blue", lwd = 2)
+#
+n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
+n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
+n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
+text(1900, 70, "PopLink", cex = 2, col = "red")
+text(1899, 60, paste(n.rec, "records"))
+text(1900, 55, paste(n.pl, "persons"))
+text(1899.5, 50, paste(n.d, "deaths"))
+#
+n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
+n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
+n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
+
+text(1985, 70, "Linnaeus", cex = 2, col = "blue")
+text(1984, 60, paste(n.rec, "records"))
+text(1985.2, 55, paste(n.pl, "persons"))
+text(1985.2, 50, paste(n.d, "deaths"))
+@
 
@@
 The scheme uses HISCO for assigning categories.
-HISCO (Historical International Standard Classifications of Occupations) \citet{lema02}  is a classification system of historical occupations, categorizing occupations according to tasks and skills.
+HISCO (Historical International Standard Classifications of Occupations) \citet{lema02}  is a classification system of historical occupations, categorising occupations according to tasks and skills.
 We have merged the original 12 classes in HISCLASS, based on position on the labour market, ending up with the following three and an additional fourth category with unknown class:
@@
 
-Figure~\ref{fig:hiscskum} shows the distribution (per cent) of exposure
+Figures~\ref{fig:hiscskumm} and \ref{fig:hiscumw} show the distribution (per cent) of exposure
 time over class  for our selected periods. The category of workers is
@@
 
-<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
+<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
 xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period", fig.height = 4.5>>=
+<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
 xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-Those that we first observe in the first included censuses often lack stated occupations.
-This is especially the case for women and elderly being retired.
-The proportion with missing values ({\tt NA}) included is presented in Table~\ref{tab:nana}.
-It was very low
+Those that we first observe in the first included censuses often lack stated 
+occupations. This is especially the case for women and elderly being retired. 
+The proportion with missing values ({\tt NA}) is shown at the bottom of the 
+figures. It was very low
 during the periods until 1950, between 4 and 8 \% lacked information
-allowing determination of class. This increased considerably during the first two decades of data brom the Linnaeus database. During the 1960s, every second person lacks class category
-and almost a third have missing information during the 1970s.
-As a consequence, we only report the results for men 40--64 for the 1960s and 1970s.
-For the most recent periods, the information is fairly complete.
-
-<<nanam, results = 'asis'>>=
-library(xtable)
-##load("data/vb0.rda")
-##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
-x <- round(prop.table(x, 1) * 100)
-xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
-@
-
-<<nanaw, results = 'asis'>>=
-##library(xtable)
-##load("data/vb0.rda")
-##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
-x <- round(prop.table(x, 1) * 100)
-xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
-@
+allowing determination of class. This increased considerably during the first 
+two decades of data from the Linnaeus database. During the 1960s, every second 
+person lacks class category
+and almost a third have missing information during the 1970s. As a consequence, 
+we only report the results for men 40--64 for the 1960s and 1970s. For the most 
+recent periods, the information is fairly complete.
+
+% <<nanam, results = 'asis'>>=
+% library(xtable)
+% ##load("data/vb0.rda")
+% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
+% x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
+% x <- round(prop.table(x, 1) * 100)
+% xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
+% @
+% 
+% <<nanaw, results = 'asis'>>=
+% ##library(xtable)
+% ##load("data/vb0.rda")
+% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
+% x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
+% x <- round(prop.table(x, 1) * 100)
+% xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
+% @
 
@@
 The modern welfare state began to develop during this time.
-For the period 1960 onwards, there are separate analyses on decades except for the two last periods.
+For the period 1960 on-wards, there are separate analyses on decades except for the two last periods.
 Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
@@
 
-The parametric model we use for the baseline hazard function is piecewise
+The parametric model we use for the baseline hazard function is piece-wise
 constant on five-year intervals, so we arrive at the following proportional
@@
 The analyses are performed in the \R\ environment for statistical computing
-and graphics \citep{cran}, especially using the package {\tt eha} \citep{gb12,eha}
+and graphics \citep{cran}, especially using the package {\tt eha} 
+\citep{gb12,eha}
 The proportional hazards model allow us to scrutinise the differences and
 the development in more detail. We divide the analysis into two age groups,
-40--64 and 65--89 years of age, the first age group representing people that were still mainly in the workforce, and the second those that mainly were retired. The model controls for period, marital
+40--64 and 65--89 years of age, the first age group representing people that
+were still mainly in the workforce, and the second those that mainly were 
+retired. The model controls for period, marital
 status and whether the individual resides in the urban or rural part. There
@@
 presented \emph{graphically}. There is some  lack of information in the higher
-ages around 1960--1980 (see Table~\ref{tab:nana}), because that kind of information is simply missing
-to a large extent.
-
-\section{The effect of Hisclass on All causes of death}
+ages around 1960--1980 (see Table~\ref{tab:nana}), because that kind of 
+information is simply missing to a large extent.
+
+\section{The effect of Hisclass, 1851--1950, 1961--2013}
+
+\subsection{All causes of death}
 
@@
 
-\subsection{Ages 40--64}
+\subsubsection{Ages 40--64}
 
@@
 scap0 <- "Total hazard by HISCLASS, ages 40--64."
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardized by HISCLASS and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard standardized by HISCLASS, ages 40--64."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom)."
+scap12 <- "Total hazard standardised by HISCLASS, ages 40--64."
 cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom)."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). Standardised."
 scap22 <- "Total hazard by HISCLASS, ages 65--89."
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "H(65|40)",
-                main = "Women 40-64")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "H(65|40)",
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
+                main = "Women 40-64", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
                 main = "Men 40-64")##, exclude = 5:6)
@@
 Figure~\ref{fig:cumsw} shows mortality levels during the different periods. 
-Obviously, and not surprisingly, mortality is decreasing over time, and we can already identify some interesting social patterns in mortality which will be commented more on below.
-Apart from
+Obviously, and not surprisingly, mortality is decreasing over time, and we can 
+already identify some interesting social patterns in mortality which will be 
+commented more on below. Apart from
 the social pattern that will be commented on below, there are a couple of
@@
 
-However, we are mainly interested in illustrating the change in relative position among the HISCLASSES over time rather than the absolute levels, so we \emph{standardize} the levels in each time period.
-It is illustrated in
-Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these relative mortality levels for the different classes.
+However, we are mainly interested in illustrating the change in relative 
+position among the HISCLASSES over time rather than the absolute levels, so we 
+\emph{standardise} the levels in each time period. It is illustrated in
+Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
+relative mortality levels for the different classes.
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Women 40-64")##, exclude = 5:6)
+                    ylab = "TMR(65|40)", main = "Women 40-64", exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 
-\subsection{Ages 65--89}
+\subsubsection{Ages 65--89}
 
@@
 This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.
+See Figure~\ref{fig:cumswold}.
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Women 65-89")##, exclude = 5:6)
+                    ylab = "TMR(90|65)", main = "Women 65-89", exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 
-\section{The effect of Hisclass on cardiovascular death}
+\subsection{Cardiovascular death}
 
@@
 <<causes41>>=
-cvb <- vb
+cvb <- vb0
 cvb$event <- cvb$heart ## Can this be done better?
@@
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
 scap12 <- "Total hazard by HISCLASS, ages 40--64, cardiovascular death."
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap22 <- "Total hazard standardized by HISCLASS, ages 65--89, cardiovascular death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
+scap22 <- "Total hazard standardised by HISCLASS, ages 65--89, cardiovascular death."
 cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
                     ylab = "TMR(65|40)",  main = "Women 40-64")
@@
 
-Figure~\ref{fig:cumsw} shows mortality levels during different periods from 1901 onwards.
+Figure~\ref{fig:cumsw} shows mortality levels during different periods from 1901 on-wards.
 A couple of observations can be made.
@@
 The pattern concerning the other groups are less clear, but if we compare the two largest groups, the workers and the lower manager/farmer group, the differences were rather small during most of the time.
-A new pattern has however developed from 1990 onwards.
+A new pattern has however developed from 1990 on-wards.
 Now working class women have substantially higher cardiovascular compared to others.
@@
 For men, we find a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality.
-The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
+The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 on-wards.
 Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the other large group, lower manager/farmers.
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women 65-89")
+                    ylab = "H(90|65)",  main = "Women 65-89", exclude = 3:4)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 
-\section{The effect of Hisclass on cancer death}
+\subsection{Cancer death}
 
@@
 <<causes12>>=
-cvb <- vb
+cvb <- vb0
 cvb$event <- cvb$cancer ## Can this be done better?
@@
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cancer death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cancer death."
 scap12 <- "Total hazard by HISCLASS, ages 40--64, cancer death."
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardized by HISCLASS and decade, women (top) and men (bottom), cancer death."
-scap22 <- "Total hazard standardized by HISCLASS, ages 65--89, cancer death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cancer death."
+scap22 <- "Total hazard standardised by HISCLASS, ages 65--89, cancer death."
 cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
@@
 Women of higher classes sometimes have favourable mortality, but unfavourable at other times.
-It seems however that the higher classes have reached lower cancer levels compared to the working class from 2001 onwards.
+It seems however that the higher classes have reached lower cancer levels compared to the working class from 2001 on-wards.
 Men from the elite had higher mortality until the 1970s, but a social gradient has appeared thereafter.
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women 65-89")
+                    ylab = "TMR(90|65)",  main = "Women 65-89", exclude = 3:4)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)", main = "Men 65-89")
+                    ylab = "TMR(90|65)", main = "Men 65-89")
 par(oldpar)
@@
 
-\section{The effect of Education on All causes of death, 1990--2013}
+\section{The effect of Education, 1990--2013}
+
+\subsection{All causes of death}
 
@@
 income and education as explanatory variables and with controls of period 
-and marital status.
+and marital status. See Figure~\ref{fig:ploteduc}.
 
@@
 plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+     col = cols, lwd = 1.5, lty = c(1, 2, 5), axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
 axis(1)
@@
 box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
 
 plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+     col = cols, lty = c(1, 2, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
 axis(1)
@@
 box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
 
@@
 
-\section{The effect of Education on Cardiovascular death}
-
-\section{The effect of Education on Cancer death}
-
-\section{The effect of Income on All causes of death, 1986--2005}
+\subsection{Cardiovascular death}
+
+<<cardeduc>>=
+ulcard <- unique(vb0$ULORSAK[vb0$heart])
+educ$heart <- FALSE
+educ$heart <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
+fitm <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
+             data = educ[educ$sex == "male", ])
+fitf <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
+             data = educ[educ$sex == "female", ])
+@
+
+Illustrating differences for cause-specific mortality is conveniently done by comparing 
+cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
+cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
+mortality.
+
+<<ploteducheart, fig.cap="Cumulative hazards by education and sex, cardiovascular death, 1990--2013.">>=
+oldpar <- par(mfrow = c(1, 2))
+
+plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
+     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
+abline(h = 0)
+plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
+     col = cols, lty = c(1:2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1:2, 5), lwd = 1.5)
+abline(h = 0)
+par(oldpar)
+@
+
+\subsection{Cancer death}
+
+<<canceduc>>=
+ulcard <- unique(vb0$ULORSAK[vb0$cancer])
+educ$cancer <- FALSE
+educ$cancer <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
+fitm <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
+             data = educ[educ$sex == "male", ])
+fitf <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
+             data = educ[educ$sex == "female", ])
+@
+
+Although mortality in cancer is lower than mortality in cardiovascular diseases,
+the pattern is the same: The more education, the less mortality. See 
+Figure~\ref{fig:ploteduccancer}.
+
+<<ploteduccancer, fig.cap="Cumulative hazards by education and sex, cancer death, 1990--2013.">>=
+oldpar <- par(mfrow = c(1, 2))
+
+plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
+     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
+abline(h = 0)
+plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
+     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
+abline(h = 0)
+par(oldpar)
+@
+
+\section{The effect of Income, 1986--2005}
+
+\subsection{All causes of death}
 
@@
 plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
 axis(1)
@@
 box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
 
 plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
 axis(1)
@@
 box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
 
@@
 
-\section{The effect of Income on Cardiovascular death}
-
-\section{The effect of Income on Cancer death}
+\subsection{Cardiovascular death}
+
+<<heartincome>>=
+ulcard <- unique(vb0$ULORSAK[vb0$heart])
+income$heart <- FALSE
+income$heart <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
+fitm <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
+             data = income[income$sex == "male", ])
+fitf <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
+             data = income[income$sex == "female", ])
+
+@
+
+The overall pattern is the same for cardiovascular mortality, high income pays off.
+<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1990--2013.">>=
+oldpar <- par(mfrow = c(1, 2))
+
+plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
+     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
+abline(h = 0)
+plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
+     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
+abline(h = 0)
+par(oldpar)
+@
+
+\subsection{Cancer death}
+
+<<cancerincome>>=
+ulcard <- unique(vb0$ULORSAK[vb0$cancer])
+income$cancer <- FALSE
+income$cancer <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
+fitm <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
+             data = income[income$sex == "male", ])
+fitf <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
+             data = income[income$sex == "female", ])
+
+@
+
+The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.
+
+<<plotinccancer, fig.cap="Cumulative hazards by income and sex, cancer death, 1990--2013.">>=
+oldpar <- par(mfrow = c(1, 2))
+
+plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
+     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
+abline(h = 0)
+plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
+     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
+#axis(1)
+#axis(2, at = c(0, 0.5, 1.0), las = 1)
+#abline(h = 0.5, lty = 3)
+#box()
+legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
+abline(h = 0)
+par(oldpar)
+@
 
@@
 The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
-The importance of understanding people in their context has also been emphasized by for example \citet{kri05}.
+The importance of understanding people in their context has also been emphasised by for example \citet{kri05}.
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.

### 2017-05-29 icass.Rnw
Revisions until section 6

@@
 
-\date{May 28, 2017}
+\date{May 29, 2017}
 
@@
 
-One of the most central aspects of survival on individual level is social class and access to economic and other resources.
-Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and the impact seems to be increasing as well \citep{kuetal04, frlu07}.
+One of the central aspects of survival is social class and access to economic and other resources.
+Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}.
 It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}.
 For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies.
-This can be considered as a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
+This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
 Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
@@
 
-Recent studies, investigating social inequality in health and mortality with micro-data, has however questioned the assumed pattern.
-These studies suggest that the modern pattern might be a fairly recent development, although the pattern differs depending on age, gender or historical context.
-When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small socio-economic differences or differences in the opposite direction than the expected during the first stages of the transition \citep{bevp11, bedr11, edli11,edbr12}, sharply contrasting against what we know from present-day society.
-
-Another interesting aspect of the development of social inequality in mortality, is its relation to economic inequality.
-\citet{wipi09} argue that income inequality has an independent effect on mortality, separated from the direct effect of the actual different access to economic resources.
+Recent studies, investigating social inequality in health and mortality with micro-data, has however questioned how generalizable the assumed pattern is \citep{bevp11}.
+Instead, some studies suggest that the modern pattern might be a fairly recent development, although the pattern differs depending on age, gender or historical context.
+When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small or even reversed socio-economic differences during the first stages of the transition \citep{bedr11, edli11,edbr12}, sharply contrasting against what we know from present-day society.
+
+Another question of the development of social inequality in mortality, concerns its relation to economic inequality.
+\citet{wipi09} argue that income inequality has an independent effect on mortality, separate from the direct effect of actual different access to economic resources.
 They argue that unequal societies perform less well when it comes to health than equal ones.
 Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world.
-This has initiated a vital scholarly debate and the topic has been extensively studied (see for example \citet{suka04, wado00}.
+This has initiated a vital scholarly debate and the topic has been extensively studied \citep{suka04, wado00}.
 How the association looked like in previous periods is unknown.
-This is of particular interest since the levels of inequality were very different from what we find during most of the recent decades.
-Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.Data are uncertain for the 19th century, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
-From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach the lowest point at around 1980.
+This is of particular interest since the levels of economic inequality were very different from those of the recent decades.
+Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.
+The conditions during the 19th century are uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
+From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980.
 Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}.
@@
 The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of the Gulf of Bothnia.
-This was a remote part of the country, where communications with the rest of Sweden were restricted until the late 20th century, dominated by agriculture and characterise by a vulnerable economy.
-The county was vulnerable to harvest failures and several severe famines struck the regions during the 1800s.
+This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century.
+The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s.
 One living long in memory was the famine of 1867–-1868 \citep{segb14}.
 During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
-Communications during winter used the winter roads.
 The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}.
 Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures.
-The inland was particularly remote.
+The inland was particularly remote and isolated.
 During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
@@
 
-The Skellefteå region consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
+The Skellefteå region until the 1950 in our dataset consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
 The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
@@
 
-The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 on-wards of Umeå municipality where another three parishes are included.
-This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy as well for a long time.
+The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 on-wards of Umeå municipality when another three parishes are included.
+This region has a somewhat different character from that of Skellefteå.
 Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period.
-It was the administrative centre in the county of Västerbotten and schools and military regiments were placed here.
-Consequently, officials, teachers and the military were much more common in Umeå than in Skellefteå.
-Agriculture dominated the rural part, there were some foundries, some industries for example in forestry and small-scale production.
-Umeå had a much more diversified occupational structure than Skellefteå.
+It was the administrative centre in the county of Västerbotten.
+Schools and military regiments were placed here and the economy was more diversified compared to Skellefteå.
+Agriculture dominated the rural part, but there were also some foundries as well as industries for example in forestry and small-scale production.
 The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.
@@
 
-The extension in time of data at the Demographic Data Base, Umeå University (DDB) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
-The databases provide us with micro-data on a sufficiently large population and with substantial social diversity from the 20th century for the Skellefteå and Umeå regions \citep{wee16}.The data for the present study comes from two large population databases at the DDB, Umeå University.
-The early period is covered by
+The recent extension in time of data at the Demographic Data Base, Umeå University (DDB) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
+The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}.
+The data for the present study comes from two large population databases at the DDB, Umeå University.
+The early period until 1950 is covered by
 the database Poplink, the digitisation of historical parish registers for
@@
 the region. The records are linked within but not between the regions. Data
-from Poplink are accessed for the period 1851--1950s.
+from Poplink are accessed for the period 1851--1950.
 
@@
 birth date, death date, first and last date of observation and type of
-entrance/exit.Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
+entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
 
@@
 
-Differences in available information in the two datasets as well as within the Linnaeus data makes it necessary to apply diverging approaches when it comes to identification of presence periods within our
+Differences in available information in the two datasets as well as within the Linnaeus data makes it necessary to apply different approaches when it comes to identification of presence periods within our
 regions. The Poplink data provides us with exact dates or at least year of
@@
 For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare to identify the observation periods in our analyses.
-We have however no exact information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
+There is however no full information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
 We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in the later census.
@@
 In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid.
-Thus, the social class in the census 1990 is considered to be kept until 2000.
+Thus, the social class in the census 1990 is considered to be kept until 2001.
 This strategy is also used for other cases of missing social class.
@@
 
-The use of different data sets, forces us to apply different approaches towards defining the social position.
-In the Poplink data, some titles are not precise enough to identify the size of the enterprise, or the division between skills, semi-skilled and unskilled workers.
-We consider however the validity of the information to be sufficient for identification of social position.
-Another problem is that Poplink usually only provides the occupation only for the head of household, thus under-estimating female working-force participation as well as that of adult children residing with their parents \citep{vik10}.
+The use of different data sets, forces us to apply different approaches towards defining the social position in some respects.
+A problem is that Poplink usually only provides the occupation for the head of household, thus under-estimating female working-force participation as well as that of adult children residing with their parents \citep{vik10}.
 In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
@@
 Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became common and female labour force participation developed as the norm in Sweden during this period.
-The occupation of the husband as signifier of social class became, if not obsolete at least, less relevant.
+The occupation of the husband as signifier of social class became, if not obsolete, at least less relevant.
 All included persons are signified by their individual occupation for this period.
@@
 
-Figures~\ref{fig:hiscskumm} and \ref{fig:hiscumw} show the distribution (per cent) of exposure
-time over class  for our selected periods. The category of workers is
+Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution (per cent) of exposure
+time according to class for men and women respectively. The category of workers is
 fairly stable over time. The highest class has
@@
 Umeå University explains much of this increase.
-Within the middle group, there has been a transfer from a majority of farmer during the 19th century to lower white collar from the middle of the 20th century.
+What is not illustrated in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century.
+Furthermore we observe an increase during the 1960s and partly during later decades as well in those where social class has not been possible to define.
+When first observed in the earliest included censuses, occupation is often missing.
+For men the majority are however in the age group 65 and older, i.e.
+those that were retired.
+Until 1950, there were however very few with missing occupation.
+For men in working-active age, the proportion missing does not cause any problems for our analyses.
 
@@
 
-Those that we first observe in the first included censuses often lack stated 
-occupations. This is especially the case for women and elderly being retired. 
-The proportion with missing values ({\tt NA}) is shown at the bottom of the 
-figures. It was very low
-during the periods until 1950, between 4 and 8 \% lacked information
-allowing determination of class. This increased considerably during the first 
-two decades of data from the Linnaeus database. During the 1960s, every second 
-person lacks class category
-and almost a third have missing information during the 1970s. As a consequence, 
-we only report the results for men 40--64 for the 1960s and 1970s. For the most 
-recent periods, the information is fairly complete.
+The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}.
+Corresponding to what we found for men, the proportion without social position is very low until the middle of the 20th century.
+This is a reflection of how womens social position is defined for that period.
+Missing social class is however a real problem during the 1960s and 1970s in particular.
+Around 80 \% lack information for the 1960s and 50 \% for the 1970s.
+Since social class for this period is defined according to own occupation, the consequence of many women being outside the workforce, the proportion missing becomes very high.
+We have chosen not to report results for women for these two decades.
+For the later decades, the results are fairly complete and ought to be representative.
 
@@
 The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces).
-The status unmarried is often not explicitly noted in the source, which makes the status unknown in the database Poplink for people for people moving into our parishes.
+The status unmarried is often not explicitly noted in the source, which makes the marital status unknown in the database Poplink for people for people moving into our parishes.
 If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.
@@
 Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.The figure indicate an under-registration of those married and and over-registration of previously married during the 1960s.
-(WHY?)
 
@@
 The century 1851--1950 is divided into four periods of 25 years each.
-The nineteenth century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
-During the years 1901--1950, industrialisation had started and the urban environments increased in population size.
+The 19th century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
+During the years 1901--1950, industrialisation started and the urban environments increased in population size.
 The modern welfare state began to develop during this time.
@@
 Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
-A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to not being as generous as before.
+A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to being less generous than before.
 The start of the last period (2008--2013) coincides with the international financial crisis.
@@
 Education is here analysed only for the period 1990 to 2005.
-Years of education are given in 4 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
-It ranges from basic education (less than nine years schooling) at the lowest level, to high academic degree (PhD and similar) at the highest level.
-Which are the new categories?)
+Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
+It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.
 
@@
 presented \emph{graphically}. There is some  lack of information in the higher
-ages around 1960--1980 (see Table~\ref{tab:nana}), because that kind of 
+ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
 information is simply missing to a large extent.
@@
 Research (CEDAR) at Umeå University and with support from the Swedish
-Research Council (VR) (grant no. 421-2012-892) and the project Old age ... financed by the Swedish Research Council. We thank Demographic Data
+Research Council (VR) (grant no.
+421-2012-892) and the project Old age, health and inequality: mortality from a life course perspective in Sweden 1986-2009, financed by the Swedish Research Council (grant no 421-2012-892.
+We thank Demographic Data
 Base for giving us access to data and Jonas Helgertz, Department of

### 2017-05-29 icass.Rnw
Revisions until 6.2.1

@@
 \subsection{Income}
-Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1990 to 2005.
+Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to 2005.
 
 \subsection{Education}
-Education is here analysed only for the period 1990 to 2005.
+Education is here analysed only for the period 1990 to 2013.
 Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
@@
 Models are fitted separately for each time period, from  1851 to
-2013, with the exception of the period 1951--1960. We stratify on HISCLASS
-and control for the covariates {\tt sex} and {\tt   civst}. The cumulative
+2013, with the exception of the period 1951--1960 and separately for sex. We stratify on HISCLASS
+and control for the covariates {\tt   civst} and urban/rural residence. The cumulative
 hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
@@
 
-The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
-Figure~\ref{fig:cumsw}.
+%The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
+%Figure~\ref{fig:cumsw}.
 
@@
 
-Figure~\ref{fig:cumsw} shows mortality levels during the different periods. 
-Obviously, and not surprisingly, mortality is decreasing over time, and we can 
-already identify some interesting social patterns in mortality which will be 
-commented more on below. Apart from
-the social pattern that will be commented on below, there are a couple of
-observations to be made from the figure. Women have consistently
-better survival than men. This is a well-known phenomenon in Swedish
+Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval 40--64 during the different periods.
+Obviously, and not surprisingly, mortality is decreasing over time for both sexes.
+Women have consistently better survival than men.
+This is a well-known phenomenon in Swedish
 demographic history. Male mortality has almost always been higher than
@@
 visible for later decades is partly explained by the decreasing general
-mortality. It is not
-obvious that differences between social classes have diminished. This
-conclusion has been confirmed by separate analyses (results not shown). There is no
-clear indication of homogenisation of mortality levels between social
-classes.
-
-However, we are mainly interested in illustrating the change in relative 
-position among the HISCLASSES over time rather than the absolute levels, so we 
-\emph{standardise} the levels in each time period. It is illustrated in
-Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
+mortality and is not necessarily representing a homogenisation of relative risks.
+
+By \emph{standardising e} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
 relative mortality levels for the different classes.
@@
 
-Notice that we have not analysed the 1950s due to
-the fact that we only have data for some parishes at that time,
-consequently having no corresponding population for that
-decade.
-Furthermore, we need to be cautious when interpreting the results for the 1960s in particular, but also for the 1970s due to much missing information on HISCLASS for these decades, something that have restricted the size of the population observed.
-The changing economy
-also have consequences for the analysis, particularly when it comes to the
-agricultural sector. While farmers were common occupations during the first
-half of the 20th century, these occupations has constantly decreased in
-numbers thereafter. This means that the second HISCLASS group has somewhat different character during the last periods in the figures.
-
-The social pattern of mortality among women is quite much in accordance with previous
+Notice that we have not analysed the 1950s due to the fact that we only have data for some parishes at that time, consequently having no corresponding population for that decade.
+Furthermore, we have excluded the results for women during the 1960s and 1970 of reasons discussed above.
+The changing economy also have consequences for the analysis, particularly when it comes to the agricultural sector.
+While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter.
+This means that the middle class has somewhat different character during the last periods in the figures.
+
+The most striking result of the analysis is the different patterns for men and women. The social pattern of mortality among women is quite much in accordance with previous
 analyses from the present research group on other regions in northern
-Sweden and for other periods \citep{edli11,edbr12,segb16}. The highest
+Sweden and for other periods \citep{edli11,edbr12}. The highest
 social class usually have comparatively low mortality, while working
-class women have high. This is especially the case during the latest
-decades of the century and until 2013. We once again remind that the results from the 1960s must be interpreted with caution due to the large proportion of missing cases.
-
-A different pattern appears among men. The groups we would expect to have
-the lowest mortality, if we assume that access to different resources and
-having a high status determines survival, instead have the highest during the nineteenth and
-much of the 20th century. This is also in accordance with our previous studies
-\citep{edli11, edbr12, segb16}. Workers on the other hand have comparatively
-low mortality until the 1960s. This changes however towards the end the 20th century. From
-the 1980s, there is a clear advantage when it comes to survival for higher
-managers and professionals. The expected social gradient in mortality has developed. This is also what we usually find in analyses
-of social health inequalities in present-day Sweden as well as in most
+class women have high.
+There is however no substantial change in the relative risks between the classes.
+Only the group with unknown social class is distinguished by moch higher mortality from the 1980s onwards.
+This can however have to do with the differences how social position for women is defined before and after 1950.
+After 1950 it is based on the own occupation.
+Being without occupation might include women that are more frail both when it comes to health and socially.
+
+A different pattern appears among men, being the most surprising finding in our analyses. The groups we would expect to have the lowest mortality, if we assume that access to different resources and
+having a high status determines survival, instead have by far the highest hazards of dying during the nineteenth and much of the 20th century. This is also in accordance with our previous studies
+\citep{edli11, edbr12}.
+Workers on the other hand have comparatively low mortality until the 1960s, at the same level as the middle class.
+The relative positions change radically towards the end the 20th century.
+From the 1980s, there is a clear advantage when it comes to survival for higher
+managers and professionals.
+The relative differences are however more compressed than for the period until 1950.
+The expected social gradient in mortality has developed.
+This is also what we usually find in analyses of social health inequalities in present-day Sweden as well as in most
 other countries. In the regions we are studying, this is however a quite
 recent phenomenon.
+Corresponding to what we find for women, those without known occupation have much higher mortality than the others from the 1960s onwards.
+The assumption made for women of unknowns social class is probably even more relevant for men.
+They probably represent groups in society that are more vulnerable when it comes to mortality.
+
+A further observation can be made from Figure~\ref{fig:cumsw}.
+The gender difference in death hazards is very big for the élite group from 1851--1950, men having about three times higher risks than women in the same class.
+In the other classes, the difference was much smaller.
+It was men in high position that experienced the largest risks.
 
@@
 
-We now turn to the elderly, i.e. those that mostly have left the workforce.
+We now turn to the elderly, i.e. those that in most cases have left the workforce.
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
@@
 
-The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level.
-Mortality among women was, as expected, lower than among
-men. We find a clear decline in mortality, but this decline did not
-set in definitely until sometimes after the Second World War (results not shown here), i.e. later than for the age group 40--64 years. This corresponds to the
-Swedish development where we find a turning point at this time with better
-survival among elderly, to a large extent caused by lower mortality in
-cardiovascular diseases. The results for the 1960s are however uncertain as
-discussed above, especially for women due to the difficulty to determine HISCLASS for the majority of them.
-
-Concerning the social pattern, the pattern is a little more difficult to interpret, compared to the younger age group (the new analyses differ somewhat from the ones in \citet{segb16}).
+The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level earlier in history.
+Mortality among women was, as expected, lower than among men.
+We find a clear decline in mortality, but the fasted decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e.
+later than for the age group 40--64 years.
+This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.
+
+Concerning the social pattern, the pattern is less clear compared to what we found for the younger age group (see Figure~\ref{fig:cumswold}).
 For women the results are mixed.
-We do not find the same clear advantage for elite women during period until 1950, and neither do we find any large disadvantage for working class elderly women.
-In some periods, the social differences were small, and the elite even had substantially higher mortality during the early 20th century.
+Once again, too much missing data on social class leaves out the results for the 1960s and 1970s.
+We do not find the same clear advantage for élite women during period until 1950, but neither do we find any large disadvantage for working class elderly women at that time.
+In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century.
 Notice however that this group was rather small at the time.
-During the recent decades, the pattern is not very clear, but for the two last periods, the elite have the lowest mortality.
-It is however not possible to identify a clear social gradient in mortality.
+The social differences continued to be small during the last decades of the 20th century.
+Only during the two last periods (after the year 2000) is it possible to establish the expected social gradient in mortality.
 
 For men we find indications of the same intriguing change among elderly as for
-the adult population 40--64 years. Mortality was in fact substantially
-higher among men from the elite  but the disadvantage is mainly apparent during the 19th century.
+the adult male opulation 40--64 years. Mortality was in fact substantially
+higher among élite men  but their disadvantage was primarily apparent during the 19th century.
 Differences are fairly small during the first half of the 20th century.
-Only during the second half of the 20th century, we find periods with much better survival for the highest social class while retired workers had fairly high mortality.
-But even in recent times, the pattern is not completely clear.
-During the last periods, we do not find a social gradient on HISCLASSES among elderly men (even in this case, the new analyses differ somewhat from the ones in \citet{segb16}).
+Not until the second half of the 20th century is the survival much better for the highest social class while workers in the age of retirement had high mortality in comparison.
+).
 This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.
-See Figure~\ref{fig:cumswold}.
 
@@
 
-We now turn to a special analysis of mortality for the two dominating causes of death among the adult and elderly population.
+In this section, the focus is on analyses of mortality for the two dominating causes of death among the adult and elderly population.
 We start with cardiovascular diseases.
@@
 First, a general observation is that the analysis of causes of death and their development are more difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ.
-Thus, many deaths during the 19th century were not diagnosed at all, and there was still a substantial part of all death in the early twentieth century where death cause is designated as unknown or dying from old age, despite reporting causes of death had become mandatory.
-For the period 1901-1925, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, but at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent.
-Are there many cardiovascular deaths disguised there?
+Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers.
+There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory.
+For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases.
+However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent.
+It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and the increase therefore is explained by improvements in how deaths were diagnosed.
 <<causes41>>=
@@
 
-As before, models are fitted separately for each ten-year time period, from  1851 to
-2013. We stratify on HISCLASS
-and use the covariates {\tt sex} and {\tt   civst}. The cumulative
+As before, models are fitted separately for sex and each ten-year time period, from  1901 to
+2013 and . We stratify on HISCLASS
+and use the covariates urban/rural and {\tt   civst}. The cumulative
 hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
@@
 Furthermore, this transition and the class pattern of mortality differ depending on age group.
-Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly, a somewhat modified result from what we found in the first and preliminary analyses presented in Leuven this year \citep{segb16b}.
+Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly.
 

### 2017-05-29 icass.Rnw
add ages 40-65 in missing hisclass tables

@@
 
+<<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+yy <- round(prop.table(xx, 1) * 100, 1)
+barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+@
+
+<<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+yy <- round(prop.table(xx, 1) * 100, 1)
+barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+@
+
 The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}.
@@
 For the later decades, the results are fairly complete and ought to be representative.
+
+The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
+\ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.
 

### 2017-05-30 icass.Rnw
Update until Education

@@
 
-By \emph{standardising e} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
+By \emph{standardising} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
 relative mortality levels for the different classes.
@@
 
-We start by looking at cumulative hazards by {\sc Hisclass} over time
-periods. They are adjusted for {\tt civil status}. We present separate analyses for women and men.
+Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40.
+They are adjusted for {\tt marital status}.
+We present separate analyses for women and men.
 
@@
 
-We will use as a summary measure the value of the cumulative hazards at
-age 65, given survival to age 40.
+% We will use as a summary measure the value of the cumulative hazards at
+% age 65, given survival to age 40.
 
@@
 
-Figure~\ref{fig:cumsw} shows mortality levels during different periods from 1901 on-wards.
-A couple of observations can be made.
-Women from the highest social class had almost consistently lower mortality from cardiovascular diseases compared to others.This is apparent even for the periods with deficient cause of death data.
-The pattern concerning the other groups are less clear, but if we compare the two largest groups, the workers and the lower manager/farmer group, the differences were rather small during most of the time.
-A new pattern has however developed from 1990 on-wards.
+Figure~\ref{fig:41cumsw} shows mortality levels during different periods from 1901 on-wards.
+Women from the highest social class had consistently lower mortality from cardiovascular diseases compared to the others.
+We also observe increasing social differences from the first half of the century to the last decades.
+A new pattern has developed from 1980 onwards.
 Now working class women have substantially higher cardiovascular compared to others.
-
-For men, we find a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality.
-The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 on-wards.
+We can however not exclude that this is an artefact of different diagnostics for deaths in different groups.
+Working class women may have had less precise diagnoses in the early 20th century.
+
+For men, there is a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality.
+The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
 Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the other large group, lower manager/farmers.
-And in the same way as for women at this age, working class me have by far the highest mortality from this cause of death during the last decades.
+And in the same way as for women at this age, working class men have by far the highest mortality from this cause of death during the last decades.
 A social hierarchy in cardiovascular mortality seems to have developed both for men and women.
+
+A word of caution is however needed when it comes to these results.
+We can not exclude that the smaller differences in cardiovascular diseases for women in the early periods are an artefact of how deaths were diagnose.
+Working class men and women may have had less precise diagnoses in the early 20th century.
 
@@
 
-Concerning the social pattern, we find rather mixed results but with some similarities to the ones
-for the age group 40--64.
-For women mortality in this disease group were somewhat higher among the elite and the lower manager/farmer group than the others until only the last decades.
-Elderly from a working class background have not been disadvantaged for most of the 20th century.
+Concerning the social pattern, we find somewhat different results from that of the younger age group of women.
+Cardiovascular mortality among elderly women do not present a consistent class pattern.
+Female cardiovascular mortality was somewhat higher among the elite and the middle group than among working class women until the last decades.
+Elderly from a working class background were not disadvantaged for most of the 20th century.
 The social class pattern has changed somewhat during the last periods and more in line to what is expected.
 The differences are however not that large.
-Cardiovascular disease mortality among elderly women do not present a consistent class pattern.
-Also for men, we find high cardiovascular mortality during the period 1926--1950 and the 1980s for the elite, but also for lower-managers/farmers during the first half of the 20th century.
-As for women, we find no clear class pattern during the last periods and the differences are quite small.The diffuse pattern of class differences in cardiovascular mortality is surprising.
-It is therefore of interest to find out what other aspects of social differentiation tell us.
-
-Notice that the results for the 1960s must be interpreted with caution due to much missing data on HISCLASS among the elderly.
+
+Even if results for elderly men point in different directions for some periods, the general trend resembles that of working-active men.
+The highest class had high cardiovascular mortality during the first half of the 20th century, while the expected social pattern has developed only during the most recent periods.
 
@@
 The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers.
-In the present analysis we include cancer deaths 1901--2013, but with an unregistered and not analysed period between 1951--1960.
-There is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated above by the large proportion of unknown or vague causes at that time.
+In our analyses, there is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated by the already mentioned large proportion of unknown or vague causes at that time.
 We believe however that cancers were better diagnosed than cardiovascular deaths.
-This is supported by the stability in proportion of cancer deaths between the two periods, 12,5 per cent 1901--1925 and 13,5 per cent for the period 1926--1950.
+This is supported by the stability in proportion of cancer deaths between the two first periods, 12,5 per cent 1901--1925 and 13,5 per cent for the period 1926--1950.
 
@@
 
-Models are fitted separately for each period, from  1851 to
-2013. We stratify on HISCLASS
-and use the covariates {\tt sex} and {\tt   civst}. The cumulative
+We use similar types of analyses as in the earlier sections. The cumulative
 hazards at the end of the period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of
+main target in our analysis. These numbers are used in the process of
 graphically illustrating the development over time.
@@
 
-Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods.
-The social differences were small until the middle of the 20th century, for women as well as for men.
-For the periods after 1980, the rather small differences remain.
-There is an indication of the elite having lower cancer mortality for women sexes during the last periods, but these differences are small and we do not find any clear social gradient in the risks.
-Thus, for this age group cancer mortality is not something that distinguishes between classes.
+Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods for the age group 40--64.
+Cancer mortality among the élite women was in fact lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods.
+Consequently, we find a more consistent advantage of higher class in cancers compared to cardiovascular diseases for women.
+We do however not find the same pattern for men.
+It is difficult to identify any general features of 20th century male cancer mortality when it comes to social differences.
 
@@
 
-We finally turn to cancer mortality among the elderly. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and with separate analyses for women and men.
+In our analyses of class patterns in mortality, we finally turn to cancer mortality among the elderly.
 
@@
 
-We use the value of the cumulative hazards at
-duration 25 (age 90) as a summary measure, starting from age 65.
-
 The relation between different classes and cancer mortality fluctuates considerably both for women and men in this age group.
-For some periods it has to do with quite few events to analyse, particularly for the 1960s where HISCLASS has been difficult to determine for the elderly.
-Women of higher classes sometimes have favourable mortality, but unfavourable at other times.
-It seems however that the higher classes have reached lower cancer levels compared to the working class from 2001 on-wards.
-Men from the elite had higher mortality until the 1970s, but a social gradient has appeared thereafter.
-Now the elite with assumed more resources have lower mortality than other groups.
-Elderly of workers origin on the other side have higher cancer mortality during the last decades, sometimes substantially higher than the others.The pattern is however not completely clear.
+For some periods it has to do with quite few events to analyse.
+Women of the highest class have favourable mortality during several periods, consistent with the advantages we found for women in the younger age group, but on the other hand had high levels at other times.
+The social differences are however quite small during the most recent periods.
+
+For men we find a similar lack of social pattern in cancer mortality as we found for men in working-active ages.
+There is however a tendency during the last decades of higher mortality among working class men.
 

### 2017-05-30 icass.Rnw
Revising discussion. Adding references

@@
 
-\title{Social inequality in mortality among
+\title{Long--term perspectives of social inequality in mortality among
   adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
@@
 
-\date{May 29, 2017}
+\date{May 30, 2017}
 
@@
 
-The classification scheme we use is  based on HISCLASS \citep{lema11}, a class-based system.
-The scheme uses HISCO for assigning categories.
-HISCO (Historical International Standard Classifications of Occupations) \citet{lema02}  is a classification system of historical occupations, categorising occupations according to tasks and skills.
-We have merged the original 12 classes in HISCLASS, based on position on the labour market, ending up with the following three and an additional fourth category with unknown class:
+HISCLASS, the classification system used here, is a ”. . . HISCO-based
+historical international social class scheme” \citep{lema11, lema02}.
+The different classes in HISCLASS represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector.
+Implicitly it reflects large differences in what is in the forefront of other
+categorisations, as for example access to economic and other resources, status, social power et cetera.
+We have merged the original 12 classes in HISCLASS, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:
 
 \begin{enumerate}
-\item {\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
-\item {\tt middle}, HISCLASS 3, 4, 5, 6 and 8. Lower managers and farmers.
-\item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
+\item {\tt elite}, HISCLASS 1 and 2. Higher managers, professionals and higher white collar.
+\item {\tt middle}, HISCLASS 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
+\item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers including farm workers.
 %%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
@@
 
-\subsection{Cardiovascular death}
+\subsection{Cardiovascular mortality}
 
@@
 
-\subsection{Cancer death}
+\subsection{Cancer mortality}
 
@@
 A common alternative to a class analysis is to use income or education for investigating socio-economic health differences.
-Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
-disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
+\citet{goldk10} criticize how many economists and epidemiologists operationalize inequality in their studies, for example by not distinguishing between class and status aspects.
+Class refers to relational aspects in society, while many other variables focus on the distribution of material and/or immaterial resources.
+He advocates studies of different aspects, which together complement each other.
+Comparing different approaches are valuable in order to understand why some are advantaged and others disadvantaged and what sort of factors determines health differences.
+In this and the following section, we introduce analyses of education and income.
+In that way, we can compare the class analyses with these alternative systems.
+For the time being, we have to restrict such a comparative analysis to the last decades.
 
@@
 
-The WHO Commission on Social Determinants of Health framework for the social determinants of health discusses how socio-economic position impacts on health.
-The influences are not direct but works through intermediate determinants related to exposure and vulnerability \citep{who10}.
-People are placed within socio-economic and political contexts, as well as value systems in the society.
-The socio-economic position imply different access to resources, power and prestige, all aspects of crucial importance.
-The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
-The importance of understanding people in their context has also been emphasised by for example \citet{kri05}.
-What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-
 The authors of this paper have scrutinised different aspects of the social
-pattern of mortality in northern Sweden for quite some time, focusing on
-different regions, different age groups and partly different social
-categories. Already the earliest findings, concerning the social inequality
-in a rapidly growing town in an industrial district, surprised and at least
-at first confused the author \citep{edv92}. The expected social gradient in
+pattern of mortality in historical northern Sweden for some time, focusing on
+different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}.
+When it comes to infant and child mortality pre-- and early--transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}.
+Studies based on microdata on the adult population in history have been rare, mainly due to lack of adequate data.
+The results from such studies show other patterns than that of infant and child mortality.
+The expected social gradient in
 survival, meaning that people from higher social classes always had an
 advantage, was in many cases, particularly among adult men, not found
-despite the well-documented socio-economic differences in this
-society. There were however large differences between men and women and for
-different age groups. Further studies have confirmed these
-findings \citep{edli11, edbr12}. These earlier studies have ended with an
-assumption that distinct and consistent differences to the advantage of higher
-classes developed somewhere during and probably quite early in the 20th
-century. Until now we have however not been able to investigate the
-continuous development on mortality differences among the adult and elderly
-population due to no available data. This is the first time it has been
-possible, and the results continue to surprise us. The consistent advantage
-for higher social classes do not appear until quite late in the regions we
-have analysed. The social pattern of female mortality are closer to what is
-expected, but a social gradient in male adult mortality seems to be a
-recent phenomenon.
+despite the well-documented socio-economic differences \citep{edli11, edbr12}.
+There were however large differences between men and women and for different age groups.
+These earlier studies have suggested that distinct and consistent differences to the advantage of higher classes developed somewhere during and probably quite early in the 20th century \citep{bevp11}.
+Until now we have however not been able to investigate the continuous development on mortality differences among the adult and elderly
+population due to no available data.
+This is the first time it has been possible, and the results continue to surprise us.
+The consistent advantage for higher social classes do not appear until quite late in the regions we have analysed.
+The social pattern of female mortality are in line to what is
+expected, but a social gradient in male adult mortality seems to be a recent phenomenon.
 Furthermore, this transition and the class pattern of mortality differ depending on age group.
@@
 
+The results of small socio-economic gradient in the past and its late development are however in accordance with other studies \citep{bevp11}, for example by \citet{bedr11} in a very different setting, a couple of Scanian parishes, although with somewhat different social categories.
+They found the first indications of an emerging social gradient in the 1950s.
+
 We have also added analyses of cause-specific mortality to the previous version.
 Mortality from cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
-What is surprising is however the small class differences among the elderly.
 Cancer is sometimes considered less connected to social class, and we find mixed results in our analyses.
-There are however indications of a social gradient during the last decades in both studied age groups.
-Development of death causes are however problematic since the ways they have been diagnosed and the proportion with missing or vague information has changed over time.
-
-The perhaps surprising results makes it reasonable to consider their
-validity and possible weaknesses in the study. There are of course
-uncertainties involved in all studies of this sort, something that should
-not be ignored.  One obstacle concerns the quality of the available
-information in the sources. Occupational titles in historical sources are
-not always distinct enough to define the correct class. It can be difficult
-to identify the skill level in the working class population because many
+
+The surprising results makes it reasonable to consider the
+validity and possible weaknesses in the study.
+One obstacle concerns the quality of available information.
+Occupational titles in historical sources are not always distinct enough to define the correct class.
+It can be difficult to identify the skill level in the working class population because many
 were often only designated as labourer. The size of the enterprise for free
 entrepreneurs is unknown and it may be impossible to distinguish if they
-are self-employed or have many employed. The class system and the position
+are self-employed or have employees. The class system and the position
 of different classes can also be discussed. Studies covering long period
 have the problem that occupations and classes change character - they do
-not represent the same thing. Nonetheless, these ambiguities and possible
+not represent the same thing. Nevertheless, these ambiguities and possible
 mis-classifications are not of that sort and magnitude that it would radically change
@@
 different periods represent distinct groups with very large differences in
-resources, capacities and power. A worker would not end up in the highest
-SOCPO level and large-scale entrepreneurs and higher officials would
-certainly be assigned the right level.Available data also show that the
-differences in income and wealth were large, often enormous.
+resources, capacities and power. Available data, for example from the 1930 census confirm that class differences in income and wealth were large, often enormous.
 
@@
 such analyses gives other results or at least would modify the image
-presented here. There are other alternatives then HISCLASS, representing
-status or class dimensions. SOCPO is another system based on hierarchic
-categories of social power, which we have used in a couple of previous
-studies. In many modern studies categorisations based on income, wealth
-or education are common for studies of inequality. It is possible that
-such categorisations are better fit to distinguish between groups in
-society. They represent different dimensions of inequality, and we believe
-that the topic discussed here would gain from analyses of different aspects
-of inequality. One problem is however that reliable data on income and
-education are rarely available in historical population databases. We can
-however be certain that the highest HISCLASSES level have more education
-and the lowest the least. We also know from the 1930 census that the
-economic differences were very large at that time and that occupations in
-higher HISCLASSES tend to have higher incomes. The groups that may diverge
-are those within agriculture and those having a small enterprise where we
-can assume that the reported income figures underestimate their economic
-standard.
-
-As a test, we have however included analyses of the associations of
-education and income to mortality for the period 1990--2005. For women as
+presented here. In many modern studies categorisations based on income, wealth
+or education are common for studies of inequality. They represent different dimensions of 
+inequality. One problem is however that reliable data on income and
+education are rarely available in historical population databases. 
+As a test, we have included analyses of the associations of
+education and income to mortality for recent decades. For women as
 well as for men, for both age groups and for all-cause mortality as well as
 cause-specific mortality we find consistent differences with higher
-educational level and more income leading to lower mortality. As already
-mentioned, the class analysis proved that it in most cases are strong
-differences during the last centuries, but the pattern is much clearer in
-the alternative ways of categorisation. This does not mean that a class
-analysis is worthless, rather that the different categorisations measure
-different things. But it certainly point to the importance to also consider
-other aspect of social inequality in the past. We do not know if education
-and income are equally important in the past and that these measures would
-provide us with a completely different picture, but this is something that
-calls for further investigation.
-
-The results of small socio-economic gradient in the past and its late development are however in accordance with other studies \citep{bevp11},
-for example by \citet{bedr11} in a very different setting, a couple of
-Scanian parishes. They found the first indications of an emerging social
-gradient in the 1950s.The similarities makes us more comfortable about our
-results.
-
-\citet{who10} lists three different types of intermediary social
-determinants of health, material circumstances, behaviours and biological
-factors, and finally psycho-social factors. Most studies have focused on
-the material circumstances. Having a better economy can give advantages in
-many aspects of relevance for health and mortality, for example good living
-conditions, access to food, medical care, better conditions at work et
-cetera. Even though these advantages ought to have been at hand in our
-studied regions, the final result was not better health for men in the most
-privileged groups until the later part of the 20th century. We must
-consider other intermediary determinants. Psycho-social factors have been
-frequently discussed during the most recent decades, for example by
-\citet{mar04}. He argues that being higher in the status ladder leads to
-better health, regardless of where in the status ladder we look. This is
-also one of the arguments \citet{wipi09} use as an explanation why equal
-societies always perform better than unequal ones when it comes to health
-and other social conditions. In the case studied here, we do not find any
-indication of a status syndrome, since mortality levels was in the opposite
-directions to the one we would expect from these theories.
+educational level and more income leading to lower mortality, thus basically confirming what we find in the class analysis. We do not know if education and income are equally important in the past.
+
+The WHO Commission on Social Determinants of Health framework for the social determinants of health provides a model on how socio-economic position influence health.
+Social class works through intermediate determinants related to exposure and vulnerability \citep{who10}.
+People are situated within socio-economic and political contexts, as well as prevailing value systems in society.
+Socio-economic position imply different access to resources, power and prestige, all being aspects of crucial importance.
+The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
+The importance of understanding people in their context is also emphasised by for example \citet{kri05}.
+What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
+The WHO framework
+\citep{who10} lists three different types of intermediary social determinants of health; material circumstances, behaviours, biological factors, and finally psycho-social factors.
+Most studies focus on material circumstances.
+Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions.
+Even though these advantages were at hand in our studied regions, the final result was not better health for men in the most privileged groups until the late 20th century.
+We must therefore consider other explanations.
+Psycho-social factors have been frequently discussed during the most recent decades, for example by \citet{mar04}, who argues that a higher status leads to better health, regardless of where in the status ladder we make our observations.
+Psychosocial factors are also suggested by \citet{wipi09} as an explanation why equal societies always perform better than unequal ones when it comes to health and other social conditions.
+In the case studied here, we do not find any indication of a status syndrome, since mortality levels was in the opposite directions to the one we would expect from these theories.
 
@@
 Our results of the gender aspect might be of interest in this case.
-When it comes to gender differences, the present study confirms what we have found on the Sundsvall region.
+When it comes to gender differences, the present study confirms what we previously have found in the Sundsvall region \citep{edli11}.
 There is a clear gendered pattern when it comes to social inequality in mortality.
-While high HISCLASS did not have any positive impact on survival for men in the early 20th century, we find  that higher social
-positions among women led to longer lives. Within the same households, the effect of social
-class seems to lead to different results. This opens up for behavioural
-aspects based on different gendered expectations. Alcohol has been
-suggested as part of the large gender differences in the 19th century
-\citep{wil99, edv92}, and a more risk-taking life may be a component of the
-male mortality.
-
-Having a high social status could be expressed in attitudes and behaviours
-that were harmful for health. Alcohol consumption decreased rapidly during
-the period late 19th century to the 1950s, thus possibly making room for an
-increased role of economic resources. But the old pattern in social health
-inequalities remained. It is however possible that the relative differences
-in alcohol consumption prevailed, although at a lower level. The question
-is however if the lower levels of alcohol consumption had any substantial
-negative impact on health. Swedes at that time had very low consumption
-levels. Alternative explanations should also be scrutinise. There
-may have been strong and heavy requirements of being successful for the
-male \"elite. Many were involved in businesses with high risks. It would be
-very damaging to lose the high position and they may have felt that they
-lived under constant threat of economic failure. There are several examples among
-the economic \"elite in the Sundsvall region that they had such problems
-\citep{edli11}. A possible future study involving analyses of death causes
-might throw some light on this issue.
+Within the same households, the effect of social class seems to lead to different results, high social position being beneficial for women but not for men.
+We might consider the role of behaviours based on different gendered expectations.
+Having a high social status could be expressed in attitudes and behaviours that were harmful for health (this is discussed in more detail in \citet{edli11}.
+Alcohol has been suggested as part of the large gender differences in the 19th century \citep{wil99, edv92}, and a more risk-taking life may be a component of male mortality \citep{cour00}.
+There are however some problems with attributing most of the male excess mortality to heavy drinking.
+Alcohol consumption decreased substantially during the period late 19th century to the 1950s, but the old pattern in social health inequalities remained.
+Swedes at that time had very low consumption levels.
+Alternative explanations should also be considered.
+There
+were strong and heavy requirements of being successful for the male \"elite. Many were involved in businesses with high risks and they may have felt a constant threat of economic failure.
 
@@
 aspects of social classes, not only the material ones but also what this
-means for behaviour, norms, way of living. \citet{edli11} suggested that an
-explanation for the larger social health inequalities in the present-day
-world is that status and higher social position is now expressed through
-health and being fit and physical active, while in many historical contexts
-higher social status was expressed by low physical activity, obsessive
-eating (being fat was impressive), drinking, smoking and sometimes also
-taking risks.
-
-Another question is if our results can be generalised. We have studied
+means for behaviour, norms, way of living. \citet{edli11} suggest that an
+explanation for the larger social health inequalities in present-day
+society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive
+eating, drinking, smoking and sometimes also taking risks. Different historical contexts promote different ways to spend economic resources.
+
+One of the major findings in this study relate to gender.
+As several researchers have pointed out, demography has often ignored power relations and different roles within families \citep{watkins93, mack95, jans07}.
+By investigating possible differences in mortality patterns according to sex, we have identified gendered patterns within the family.
+Different patterns between the sexes may be caused by discrimination on the female part in the household.
+We know that the society was
+strongly segregated according to gender and that women were discriminated
+and had less space for action.
+In our case, we do not find this had negative consequences for women, indicating that the relation between sexessociety was characterised by distinct gendered expectations on behaviour.
+
+Another question is if our results can be generalised to other historical contexts. We have studied
 environments in northern Sweden without any large metropolis. Perhaps
@@
 and local level may have been stronger, thus mitigating socio-economic
-disadvantages. These are relevant questions, which hopefully can be
+disadvantages. In comparison with many other countries, the level of mortality has been fairly low. These are relevant questions, which hopefully can be
 addressed in other and maybe comparative studies.
 
-Finally, we argue that our results have implications for our understanding
-on several hot research issues. A central question concerning the mortality
-transition is how the enormous increase in survival could be
-accomplished. This is still a hot debated question. \citet{mck76} argues that the
-improvements up to the middle of the 20th century was first of all made
-possible by improved nutrition. This seems to implicate that having more
-resources would lead to better survival. If research shows that the most
+Finally, we argue that our results have implications for the understanding
+on several other hot research topics. A central and hot debated question concerning the mortality
+transition is how the enormous increase in survival was accomplished. \citet{mck76} argues that the
+improvements up to the middle of the 20th century was primarily caused by improved nutrition. This implicate that having more resources would lead to better survival. If research shows that the most
 advantaged groups did not have better health and longer lives, this do not
@@
 hypothesis that social conditions should be seen as a fundamental cause of
-disease. Even if we would eliminate disadvantages such as bad living
-conditions, lower standard of medical care and other proximate determinants
-for the poor, the health disadvantage would remain. In the historical
-context we have analysed here, this does not get any support. This further
-illustrate the need for situating the analyses in their historical and
-geographical contexts.
+disease and not only expressions of different disadvantages. In the historical
+context analysed here, this idea is not supported.
 
 In recent years, the question about the relation between health and income
-inequality been widely discussed. Even though our study do not address this
+inequality been widely discussed \citep{wipi09}. Even though our study do not address this
 issue directly, our results are of interest for this question as well. The
@@
 
-\section{Conclusion}
-
-We started this paper by raising the issue of the development of social
-inequalities. When studying this question in history we often get surprised
-of its complexity. In a society characterised by large economic inequality,
-we would expect that this would result in corresponding mortality
-differences, i.e. higher social classes having an advantage. We find the
-appearance of a consistent social pattern in mortality, to the advantage of
-the higher classes, but surprisingly late in our history. It is only in the
-recent decades that this pattern dominates. Improved survival characterised
-however all HISCLASSes. We can however confirm that there are gender
-differences in the social pattern. While there is no clear social gradient
-among men, the results at least partly indicate advantages for the higher
-social classes for women already early in the 20th century. Females had
-more to gain from a high social position. Previous studies of the mortality
-transition have made similar observations. Among adults, gender has often
-been of more significant importance than social class. The reason for this
-difference and for the late appearance of the modern social pattern among
-men is still an enigma that require further studies, even though we have
-suggested the possible importance of behavioural factors and changes in
-ideals connected to different social classes. This should stimulate us to
-further disentangle the pathways for how social position influence health
-and mortality in different societies.
+% \section{Conclusion}
+% 
+% We started this paper by raising the issue of the development of social
+% inequalities. When studying this question in history we often get surprised
+% of its complexity. In a society characterised by large economic inequality,
+% we would expect that this would result in corresponding mortality
+% differences, i.e. higher social classes having an advantage. We find the
+% appearance of a consistent social pattern in mortality, to the advantage of
+% the higher classes, but surprisingly late in our history. It is only in the
+% recent decades that this pattern dominates. Improved survival characterised
+% however all HISCLASSes. We can however confirm that there are gender
+% differences in the social pattern. While there is no clear social gradient
+% among men, the results at least partly indicate advantages for the higher
+% social classes for women already early in the 20th century. Females had
+% more to gain from a high social position. Previous studies of the mortality
+% transition have made similar observations. Among adults, gender has often
+% been of more significant importance than social class. The reason for this
+% difference and for the late appearance of the modern social pattern among
+% men is still an enigma that require further studies, even though we have
+% suggested the possible importance of behavioural factors and changes in
+% ideals connected to different social classes. This should stimulate us to
+% further disentangle the pathways for how social position influence health
+% and mortality in different societies.
 Furthermore, we do not find any substantial differences or distinct pattern in the relation between class and the two major cause-of-death groups.

### 2017-05-30 icass.Rnw
Minor revision in abstract

@@
 20th century. The modern pattern of clear socio-economic differences
-appeared surprisingly late. Additional analyses based on education and
+appeared surprisingly late. Additional and alternative analyses based on education and
 income categories confirmed large inequality in mortality
-between different groups during the most recent decades, but also demonstrated more distinct differences between categories with better survival the more income and the higher education level people have.
-Both cardiovascular diseases and cancers conform with the pattern already
-observed when it comes to all-cause mortality.
+between different groups during the most recent decades.
+There were no clear signs of qualitative differences depending on whether it was the working-active or the retired population that was analysed, but the differences according to social class was larger in the first-mentioned group.
+Both cardiovascular diseases and cancers conform with the pattern already observed when it comes to all-cause mortality.
 

### 2017-05-31 icass.Rnw
Revision of diverse sections. Adding references

@@
 
-\date{May 30, 2017}
+\date{May 31, 2017}
 
@@
 inequalities in mortality in the adult and elderly population is taken in
-this paper.
-The area under investigation consists of the Skellefteå and Umeå regions in
+this paper. The area under investigation consists of the Skellefteå and Umeå regions in
 the north of Sweden, and the time period is 1851--2013. The outcome is all
@@
 cardiovascular diseases and cancers. The analysis of inequality is based on
-Hisclass.
+three social classes, reorganized from Hisclass.
 
@@
 One of the central aspects of survival is social class and access to economic and other resources.
-Even in present welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}.
+Even in present-day welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}.
 It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}.
@@
 Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.
-The conditions during the 19th century are uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
+The situation during the 19th century is uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
 From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980.
@@
 The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
-The main questions are:
-
-\begin{enumerate}
-\item Has inequality in mortality between social classes increased among the adult and elderly
-  population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
-
-%%\item Is the mortality transition initiated by a certain social class?
-
-\item Are there any gender differences in the effect of social position?
-
-\item Is social position equally important among the retired population as for those in working age?
-\item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
-\item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
-\end{enumerate}
+The main question we investigate is if inequality in mortality between social classes increased among the adult and elderly population, primarily benefiting the highest social positions and how this is connected to the development of economic inequality in society.
+This issue is then analysed according to gender, age-group (working age vs retired) and cause of death (cardiovascular and cancer).
+Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period 1986--2013.
+
+% ? :
+% 
+% \begin{enumerate}
+% \item Has inequality in mortality between social classes increased among the adult and elderly
+%   population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
+% 
+% %%\item Is the mortality transition initiated by a certain social class?
+% 
+% \item Are there any gender differences in the effect of social position?
+% 
+% \item Is social position equally important among the retired population as for those in working age?
+% \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
+% \item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
+% \end{enumerate}
+% 
+%
 
@@
 
-The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of the Gulf of Bothnia.
+The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of Gulf of Bothnia.
 This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century.
@@
 
-Mortality was fairly low in comparison with other parts of the country and the fertility transition was late.
+Mortality was fairly low in comparison with other parts of the country and the fertility transition was late \citep{coa86}.
 The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.
 
-The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 on-wards of Umeå municipality when another three parishes are included.
+The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality when another three parishes are included.
 This region has a somewhat different character from that of Skellefteå.
@@
 
-The recent extension in time of data at the Demographic Data Base, Umeå University (DDB) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
+The recent extension in time of data at the Demographic Data Base (DDB), Umeå University (www.cedar.umu.se) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
 The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}.
-The data for the present study comes from two large population databases at the DDB, Umeå University.
+The data for the present study comes from two large population databases at the DDB.
 The early period until 1950 is covered by
@@
 The other large data-set is extracted from the Linnaeus database
-\citep{malm10}. Linnaeus is based on different linked population registers from
+\citep{malm10}. Linnaeus is based on different linked national population registers from
 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
@@
 year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
-For the period 2002--2013 we use the information from the yearly population registers (LISA) together with death information from National Board of Health and Welfare.
+For the period 2002--2013 we use the information from the yearly population registers from Statistics Sweden (LISA) together with death information from National Board of Health and Welfare.
 
@@
 In the data-set analysed here, all individuals 40 years and older ever being resident in either
-of the regions are included. The data file contains the variables on sex,
-birth date, death date, first and last date of observation and type of
+of the regions are included. The data file contains the variables on social class (HISCLASS) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
 entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
@@
 
-For the last period, 2002--2013, we use the yearly population registers (also known as LISA data from Statistics Sweden), together with information on deaths from National Board of Health and Welfare.
+For the last period, 2002--2013, we use the yearly population registers (LISA data), together with information on deaths from National Board of Health and Welfare.
 This data allows good control of the presence periods in the studied regions.
@@
 This strategy is also used for other cases of missing social class.
-Furthermore, we consider the social position after retirement (in Sweden usually at age 65) to be the same as the last during working age.
+Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65.
+For the Linnaeus data, we use a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
 
@@
 
-For the Poplink data, we have used a script re-coding HISCO into HISCLASS,
-provided to us by Jonas Helgertz, Lund University. The procedure for
-occupations in the Linnaeus database is a little more complicated. The
-occupations in the censuses (NYK) are first transferred to the system
-later in use, SSYK (the Swedish version of ISCO88). Thereafter the
-SSYK, both those transferred from the censuses and those given in the
-population registers 2001-2013, can be re-coded into HISCLASS.
+% The procedure for
+% occupations in the Linnaeus database is a little more complicated. The
+% occupations in the censuses (NYK) are first transferred to the system
+% later in use, SSYK (the Swedish version of ISCO88). Thereafter the
+% SSYK, both those transferred from the censuses and those given in the
+% population registers 2001-2013, can be re-coded into HISCLASS.
 
@@
 time according to class for men and women respectively. The category of workers is
-fairly stable over time. The highest class has
-increased substantially, although from a very low level in the 19th
-century. The increase of academics in connection to the establishment of
+fairly stable over time. The highest class has increased substantially, although from a very low level in the 19th century. The increase of academics in connection to the establishment of
 Umeå University explains much of this increase.
-What is not illustrated in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century.
-Furthermore we observe an increase during the 1960s and partly during later decades as well in those where social class has not been possible to define.
+What is hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century.
+Furthermore we observe an increase during the 1960s and partly during later decades as well in those with undefined.
 When first observed in the earliest included censuses, occupation is often missing.
-For men the majority are however in the age group 65 and older, i.e.
-those that were retired.
+For men the proportion is somewhat higher in the age group 65 and above, i.e.
+those that were retired, but still low enough to be used in our analyses.
 Until 1950, there were however very few with missing occupation.
@@
 
-<<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
-xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-<<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
-xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-@
+% <<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+% xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+% yy <- round(prop.table(xx, 1) * 100, 1)
+% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+% @
+% 
+% <<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+% xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+% yy <- round(prop.table(xx, 1) * 100, 1)
+% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+% @
 
@@
 We have chosen not to report results for women for these two decades.
-For the later decades, the results are fairly complete and ought to be representative.
-
-The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
-\ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.
+For the later decades, the results are fairly complete and should be representative.
+
+% The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
+% \ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.
 
@@
 The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces).
-The status unmarried is often not explicitly noted in the source, which makes the marital status unknown in the database Poplink for people for people moving into our parishes.
 If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.
 
-Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.The figure indicate an under-registration of those married and and over-registration of previously married during the 1960s.
+Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.
+The figure indicates an under-registration of those married and and over-registration of previously married during the 1960s.
 
@@
 @
+
+\subsection{Urban vs rural residence}
+In 19th and early 20th century Sweden (as in other countries), mortality was much higher in urban environments.
+The unsanitary and overcrowded living conditions as well as the risks for importing infections made the health risks higher in towns and cities.
+In the present paper, we control for this aspect by distinguishing between urban and rural residence, based on in which parish people lived.
 
@@
 in order to show the development of the social patterning of mortality
-during the studied period. We present the results for the period 2001--2013
-separately. We suggest a heterogeneity index in order to investigate a
-possible homogenisation of mortality between social classes.
+during the studied period.
+
+% We present the results for the period 2001--2013
+% separately. We suggest a heterogeneity index in order to investigate a
+% possible homogenisation of mortality between social classes.
 
@@
 interaction term between {\tt age} and {\tt hisclass}. Then, the main
-results are
-presented \emph{graphically}. There is some  lack of information in the higher
-ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
-information is simply missing to a large extent.
+results are presented \emph{graphically}.
+
+% There is some  lack of information in the higher
+% ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
+% information is simply missing to a large extent.
 
@@
 While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter.
-This means that the middle class has somewhat different character during the last periods in the figures.
+This means that the middle class has somewhat different character during the last periods.
 
@@
 The assumption made for women of unknowns social class is probably even more relevant for men.
-They probably represent groups in society that are more vulnerable when it comes to mortality.
+They probably represent groups in society that are particularly vulnerable.
 
@@
 Mortality among women was, as expected, lower than among men.
-We find a clear decline in mortality, but the fasted decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e.
+We find a clear decline in mortality, but the fastest decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e.
 later than for the age group 40--64 years.
@@
 In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century.
-Notice however that this group was rather small at the time.
 The social differences continued to be small during the last decades of the 20th century.
@@
 For men we find indications of the same intriguing change among elderly as for
-the adult male opulation 40--64 years. Mortality was in fact substantially
+the adult male population 40--64 years. Mortality was in fact substantially
 higher among élite men  but their disadvantage was primarily apparent during the 19th century.
@@
 
-In this section, the focus is on analyses of mortality for the two dominating causes of death among the adult and elderly population.
+In this section, the focus is on analyses of mortality from the two dominant causes of death among the adult and elderly population.
 We start with cardiovascular diseases.
@@
 
-First, a general observation is that the analysis of causes of death and their development are more difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ.
+First, a general observation is that the analysis of causes of death and its development is difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ.
 Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers.
@@
 However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent.
-It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and the increase therefore is explained by improvements in how deaths were diagnosed.
+It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and that the increase partly is explained by improvements in how deaths were diagnosed.
 <<causes41>>=
@@
 As before, models are fitted separately for sex and each ten-year time period, from  1901 to
-2013 and . We stratify on HISCLASS
-and use the covariates urban/rural and {\tt   civst}. The cumulative
+2013. We stratify on HISCLASS and use the covariates urban/rural and {\tt   civst}. The cumulative
 hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of
-graphically illustrating the development over time.
+main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.
 
@@
 Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40.
-They are adjusted for {\tt marital status}.
+They are adjusted for {\tt marital status} and urban/rural residence.
 We present separate analyses for women and men.
@@
 A new pattern has developed from 1980 onwards.
-Now working class women have substantially higher cardiovascular compared to others.
+Now working class women have substantially higher cardiovascular mortality compared to others.
 We can however not exclude that this is an artefact of different diagnostics for deaths in different groups.
@@
 The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
-Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the other large group, lower manager/farmers.
-And in the same way as for women at this age, working class men have by far the highest mortality from this cause of death during the last decades.
+Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the middle category consisting of lower manager/farmers.
+And in the same way as for women at this age, working class men have the highest mortality from this cause of death during the last decades.
 A social hierarchy in cardiovascular mortality seems to have developed both for men and women.
 
-A word of caution is however needed when it comes to these results.
-We can not exclude that the smaller differences in cardiovascular diseases for women in the early periods are an artefact of how deaths were diagnose.
+% A word of caution is however needed when it comes to these results.
+We can not exclude that the smaller differences in cardiovascular diseases for women in the early periods are an artefact of how deaths were diagnosed.
 Working class men and women may have had less precise diagnoses in the early 20th century.
@@
 In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years.
-In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and with separate analyses for women and men.
+In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and urban/rural residence and with separate analyses for women and men.
 We use as a summary measure the value of the cumulative hazards at
@@
 The social class pattern has changed somewhat during the last periods and more in line to what is expected.
-The differences are however not that large.
+The differences are however rather small.
 
@@
 Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods for the age group 40--64.
-Cancer mortality among the élite women was in fact lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods.
+Cancer mortality among the élite women was lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods.
 Consequently, we find a more consistent advantage of higher class in cancers compared to cardiovascular diseases for women.
@@
 In this and the following section, we introduce analyses of education and income.
-In that way, we can compare the class analyses with these alternative systems.
+In that way, we can compare the class analyses with these alternative categorizations.
 For the time being, we have to restrict such a comparative analysis to the last decades.
 
-The analysis for the period 1990--2005 is based on a separate analysis with 
-income and education as explanatory variables and with controls of period 
+The analysis for the period 1990--2005 for the whole population 40 years and older is based on separate analyses with income and education as explanatory variables and with controls of period 
 and marital status. See Figure~\ref{fig:ploteduc}.
@@
 
-Both education and income had a strong, statistically significant, effect on survival, as expected.
+Education had a strong, statistically significant, effect on survival, as expected.
 
@@
 They found the first indications of an emerging social gradient in the 1950s.
-
-We have also added analyses of cause-specific mortality to the previous version.
-Mortality from cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
+Similar results have also been found in Estonia \citep{Jaadla17}.
+
+When it comes to specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
 Cancer is sometimes considered less connected to social class, and we find mixed results in our analyses.
@@
 society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive
-eating, drinking, smoking and sometimes also taking risks. Different historical contexts promote different ways to spend economic resources.
+eating, drinking, smoking and sometimes also taking risks.
+Different historical contexts promote different ways to spend economic resources.This is also in line with what \citet{rasp06} and \citet{Jaadla17} suggest.
+If this interpretation is correct, the hypothesis of \citet{ant67}on an increasing role of health related behaviour in the modern low-mortality régime,while economic inequality determined the social mortality pattern during the early phases of the mortality transition.
+Instead, health-related seems to be at least equally strong earlier in history but having a completely different expression.
 

### 2017-05-31 icass.Rnw
Updating education and income

@@
 were still mainly in the workforce, and the second those that mainly were 
-retired. The model controls for period, marital
+retired. The model controls for marital
 status and whether the individual resides in the urban or rural part. There
 are no serious signs of deviation from proportionality assumption, except
-regarding {\sc Hisclass}.
-
-We analyse both regions together but perform separate analyses for men and
-women and for the age groups 40--64 and 65--89. The model in the analyses
-include social class, sex and marital status. Each decade
-during the 20th century gets separate analyses, but we then combine results
+regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
 in order to show the development of the social patterning of mortality
 during the studied period.
+
+% We analyse both regions together but perform separate analyses for men and
+% women and for the age groups 40--64 and 65--89. The model in the analyses
+% include social class, sex and marital status.
 
@@
 
-\subsection{All causes of death}
+%\subsection{All causes of death}
 
@@
 
-The analysis for the period 1990--2005 for the whole population 40 years and older is based on separate analyses with income and education as explanatory variables and with controls of period 
-and marital status. See Figure~\ref{fig:ploteduc}.
+In this section we analyse differences according to education level for the period 1990--2013 for the whole population 40 years and older.
+Education is a good indicator of both human and cultural capital, but aer also strongly associated with income and social class.
+We present analyses on all cause mortality (survival curves) as well as on the cause of death groups of cardiovascular diseases and cancers (cumulative hazards).
+The pattern is clear in total mortality as well as cause specific mortality, and this is valid for both men and women, see Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart} and~\ref{fig:ploteduccancer}.
+Education had a strong, statistically significant, effect on survival, as expected.
+Those with the lowest education levels hade the highest mortality, while those with highest level of education had by far the best survival.
+The differences must be considered large, and it follows a continuous improvement according to increasing levels.
 
@@
 @
-
-Education had a strong, statistically significant, effect on survival, as expected.
 
@@
 
-Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
-The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
-
-\subsection{Cardiovascular death}
+%\subsection{Cardiovascular death}
 
@@
 
-Illustrating differences for cause-specific mortality is conveniently done by comparing 
-cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
-cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
-mortality.
+% Illustrating differences for cause-specific mortality is conveniently done by comparing 
+% cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
+% cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
+% mortality.
 
@@
 
-\subsection{Cancer death}
+%\subsection{Cancer death}
 
@@
 
-Although mortality in cancer is lower than mortality in cardiovascular diseases,
-the pattern is the same: The more education, the less mortality. See 
-Figure~\ref{fig:ploteduccancer}.
+% Although mortality in cancer is lower than mortality in cardiovascular diseases,
+% the pattern is the same: The more education, the less mortality. See 
+% Figure~\ref{fig:ploteduccancer}.
 
@@
 
-Higher incomes result in lower mortality. This is also what could be expected, 
-and conforms to what we know of social health differences in present-day Sweden. 
-Notice however that we cannot establish a causal relationship between 
-income and mortality. Bad health, for instance, may be a confounder.
-
-\subsection{Cardiovascular death}
+Finally, we present analyses of different income levels for the age group 40 and above and the period 1986--2005.
+Disposable individual income must be considered as a robust measure of economic condition, and increases the possibility to allocate these economic resources towards health promoting living conditions and behaviours.
+We categorize the income according to quartiles.
+As expected, we find that higher incomes result in lower mortality, both when it comes to total mortality as to deaths from cardiovascular diseases, see Figures~\ref{fig:plotincome} and~\ref{fig:plotincheart}.
+This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden.
+We also find indications of a similar difference for cancers but less clear, particularly for male cancer mortality, see Figure~\ref{fig:plotinccancer}.
+Notice however that we cannot establish a causal relationship between income and mortality.
+Bad health may be a confounder, causing both low incomes and higher death risks.
+
+%\subsection{Cardiovascular death}
 
@@
 
-The overall pattern is the same for cardiovascular mortality, high income pays off.
+%The overall pattern is the same for cardiovascular mortality, high income pays off.
 <<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1990--2013.">>=
@@
 
-The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.
+% The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.
 
@@
 and had less space for action.
-In our case, we do not find this had negative consequences for women, indicating that the relation between sexessociety was characterised by distinct gendered expectations on behaviour.
+In our case, we do not find this to have negative consequences for women, indicating that the relation between sexes in the society was characterised by distinct gendered expectations on behaviour.
 

### 2017-06-01 R/getPHglm2.R
add civst and urban

@@ -13,7 +13,7 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
     levs <- matrix(0, nrow = m, ncol = k + 1)
    
     for (i in 1:nper){
-        fit <- glm(event ~ offset(log(exposure)) + ##urban + ##civst 
+        fit <- glm(event ~ offset(log(exposure)) + urban + civst +
                        age * hisclass, 
                    data = dat[dat$period == labb[i], ], family = poisson)
         

### 2017-06-01 icass.Rnw
add tables, natbib with apalike

@@
 
+\usepackage{float}
+
 \usepackage{epstopdf}
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{apalike}
+%\bibliographystyle{plainnat}
 
@@
 
-%\usepackage[tablesonly]{endfloat}
+\usepackage[tablesonly]{endfloat}
 %\usepackage[figuresonly]{endfloat}
@@
 
-%\renewcommand{\efloatseparator}{\vfill}
+\renewcommand{\efloatseparator}{\vfill}
 
 % patch \efloat@iwrite to use \protected@write
-%\makeatletter
-%\renewcommand\efloat@iwrite[1]{%
-%   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-%\makeatother
+\makeatletter
+\renewcommand\efloat@iwrite[1]{%
+   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
+\makeatother
 
@@
 
-\begin{document}
-
-\maketitle
-
 <<setup, include = FALSE>>=
 library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
 library(knitr)
@@
 
+\begin{document}
+
+\maketitle
+
 \begin{abstract}
@@
 
-<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 10>>=
+<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8>>=
 source("R/plotED.glm.R")
@@
 
-Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval 40--64 during the different periods.
-Obviously, and not surprisingly, mortality is decreasing over time for both sexes.
-Women have consistently better survival than men.
-This is a well-known phenomenon in Swedish
+From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
+The most interesting feature is the three-way interaction with sex, period, and hisclass.
+
+<<cumswtab,results='asis',cap="Table">>=
+fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period, 
+            data = vb4064, family = poisson)
+dr <- drop1(fitt, test = "Chisq")
+##knitr::kable(dr)
+xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, 40-65.")
+@
+
+Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval 
+40--64 during the different periods. Obviously, and not surprisingly, mortality 
+is decreasing over time for both sexes. Women have consistently better survival 
+than men. This is a well-known phenomenon in Swedish
 demographic history. Male mortality has almost always been higher than
@@
 
-<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
 oldpar <- par(mfrow = c(2, 1))
@@
 
-<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
 ##source("R/plotED.R")
@@
 par(oldpar)
+@
+
+See Table~\ref{tab:cumswoldtab}.
+
+<<cumswoldtab,results='asis'>>=
+fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period, 
+            data = vb6589, family = poisson)
+dr <- drop1(fitt, test = "Chisq")
+##knitr::kable(dr)
+xtable(dr,label = "tab:cumswoldtab", caption = "Anova, all causes of death, 65-89.")
 @
@@
 
-<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
 source("R/plotED.glm.R")
@@
 %%@
+<<41cumswtab,results='asis'>>=
+fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass + 
+                sex * period + hisclass * period, 
+            data = cvb4064, family = poisson)
+dr <- drop1(fitt, test = "Chisq")
+##knitr::kable(dr)
+xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64.")
+@
+
+See Table~\ref{tab:41cumswtab}.
 
@@
 
-<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
 ##source("R/plotED.R")
@@
 @
+
+<<41cumswoldtab,results='asis'>>=
+fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * period *hisclass,
+                ##sex * period + hisclass * period, 
+            data = cvb6589, family = poisson)
+dr <- drop1(fitt, test = "Chisq")
+##knitr::kable(dr)
+xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
+@
+
+See Table~\ref{tab:41cumswoldtab}.
 
@@
 
-<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 10>>=
+<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
 source("R/plotED.glm.R")
@@
 
+<<12cumswtab,results='asis'>>=
+fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex *hisclass +
+                sex * period + hisclass * period, 
+            data = cvb4064, family = poisson)
+dr <- drop1(fitt, test = "Chisq")
+##knitr::kable(dr)
+xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
+@
+
+See Table~\ref{tab:12cumswtab}.
+
 Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods for the age group 40--64.
@@
 
-<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 10>>=
+<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
 ##source("R/plotED.R")
@@
 
+<<12cumswoldtab,results='asis'>>=
+fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex +
+                sex * period + hisclass * period, 
+            data = cvb6589, family = poisson)
+dr <- drop1(fitt, test = "Chisq")
+##knitr::kable(dr)
+xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
+@
+
+See Table~\ref{tab:12cumswoldtab}.
 % \subsubsection{The effect of income and education on cancer death, 1990--2005}
@@
 @
+
+<<tableeduc,results='asis'>>=
+fit <- coxreg(Surv(enter, exit, event) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
+              hazards = FALSE)
+dr <- drop1(fit, test = "Chisq")
+ltx(fit, dr = dr, caption = "Effect of education on mortality in ages 40-90, 1990-2013.", 
+    label = "tab:tableeduc", digits = 4)
+@
+
+See Table~\ref{tab:tableeduc}.
 
@@
 
+<<tableeducheart,results='asis'>>=
+fit <- coxreg(Surv(enter, exit, heart) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
+              hazards = FALSE)
+dr <- drop1(fit, test = "Chisq")
+ltx(fit, dr = dr, caption = "Effect of education on cardiovascular mortality in ages 40-90, 1990-2013.", 
+    label = "tab:tableeducheart", digits = 4)
+@
+
+See Table~\ref{tab:tableeduc}.
+
 %\subsection{Cancer death}
@@
 
+<<tableeduccancer,results='asis'>>=
+fit <- coxreg(Surv(enter, exit, cancer) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
+              hazards = FALSE)
+dr <- drop1(fit, test = "Chisq")
+ltx(fit, dr = dr, caption = "Effect of education on cancer mortality in ages 40-89, 1990-2013.", 
+    label = "tab:tableeduccancer", digits = 4)
+@
+
+See Table~\ref{tab:tableeduc}.
+
 \section{The effect of Income, 1986--2005}
@@
 
+<<tableinc>>=
+fit <- coxreg(Surv(enter, exit, event) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
+              hazards = FALSE)
+dr <- drop1(fit, test = "Chisq")
+@
+
+<<tableincpr,results='asis'>>=
+ltx(fit, dr = dr, caption = "Effect of income on mortality in ages 40-89, 1986-2005.", 
+    label = "tab:tableincpr", digits = 4)
+@
+
+See Table~\ref{tab:tableincpr}.
+
 Finally, we present analyses of different income levels for the age group 40 and above and the period 1986--2005.
@@
 %The overall pattern is the same for cardiovascular mortality, high income pays off.
-<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1990--2013.">>=
+<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005.">>=
 oldpar <- par(mfrow = c(1, 2))
@@
 
+<<tableincheart>>=
+fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
+              hazards = FALSE)
+dr <- drop1(fit, test = "Chisq")
+@
+
+<<tableincprheart,results='asis'>>=
+ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.", 
+    label = "tab:tableincpr", digits = 4)
+@
+
+See Table~\ref{tab:tableincpr}.
+
 %\subsection{Cancer death}
@@
 @
+
+<<tableinccancer>>=
+fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
+              hazards = FALSE)
+dr <- drop1(fit, test = "Chisq")
+@
+
+<<tableincprcancer,results='asis'>>=
+ltx(fit, dr = dr, caption = "Effect of income on cancer mortality in ages 40-89, 1986-2005.", 
+    label = "tab:tableincprcancer", digits = 4)
+@
+
+See Table~\ref{tab:tableincpr}.
 

### 2017-06-01 icass.Rnw
Adding comments on tables

@@
 
-\bibliographystyle{apalike}
-%\bibliographystyle{plainnat}
+%\bibliographystyle{apalike}
+\bibliographystyle{plainnat}
 
@@
 retired. The model controls for marital
-status and whether the individual resides in the urban or rural part. There
+status and whether the individual resides in the urban or rural parts of the regions. There
 are no serious signs of deviation from proportionality assumption, except
@@
 results are presented \emph{graphically}.
+Tables with results from the different models are presented at the end of the article, but are not commented on in detail.
+The comments on the results are primarily based on the figures and focuses on the central variable in our study -- {\sc Hisclass}.
+As can be seen from the tables, most variables in the model are significant at 5 per cent level.
 
@@
 
-Models are fitted separately for each time period, from  1851 to
-2013, with the exception of the period 1951--1960 and separately for sex. We stratify on HISCLASS
-and control for the covariates {\tt   civst} and urban/rural residence. The cumulative
+The cumulative
 hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
@@
 Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40.
-They are adjusted for {\tt marital status} and urban/rural residence.
-We present separate analyses for women and men.
 
@@
 
-In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years.
-In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and urban/rural residence and with separate analyses for women and men.
-We use as a summary measure the value of the cumulative hazards at
+In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years. We use as a summary measure the value of the cumulative hazards at
 duration 25 (age 90), starting from age 65.

### 2017-06-01 icass.Rnw
bibliography and cosmetic changes

@@
 
-%\bibliographystyle{apalike}
-\bibliographystyle{plainnat}
+\bibliographystyle{apalike}
+%\bibliographystyle{plainnat}
 
@@
 
-\title{Long--term perspectives of social inequality in mortality among
+\title{Long-term perspectives of social inequality in mortality among
   adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
@@
 cardiovascular diseases and cancers. The analysis of inequality is based on
-three social classes, reorganized from Hisclass.
+three social classes, reorganized from {\sc Hisclass}.
 
@@
 This issue is then analysed according to gender, age-group (working age vs retired) and cause of death (cardiovascular and cancer).
-Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period 1986--2013.
+Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period
+1986--2013.
 
@@
 
-The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of Gulf of Bothnia.
+The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf of Bothnia.
 This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century.
@@
 In the data-set analysed here, all individuals 40 years and older ever being resident in either
-of the regions are included. The data file contains the variables on social class (HISCLASS) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
+of the regions are included. The data file contains the variables on social class ({\sc Hisclass}) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
 entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
@@
 
-In this paper we analyse mortality differences according to social class based on the classification scheme HISCLASS \citep{lema11}.
+In this paper we analyse mortality differences according to social class based on the classification scheme {\sc Hisclass} \citep{lema11}.
 The only type of socio-economic information available both in Linnaeus and Poplink is occupations.
@@
 Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65.
-For the Linnaeus data, we use a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
+For the Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided to us by Jonas Helgertz, Lund University.
 
@@
 
-HISCLASS, the classification system used here, is a ”. . . HISCO-based
-historical international social class scheme” \citep{lema11, lema02}.
-The different classes in HISCLASS represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector.
+{\sc Hisclass}, the classification system used here, is a "\ldots HISCO-based
+historical international social class scheme" \citep{lema11, lema02}.
+The different classes in {\sc Hisclass} represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector.
 Implicitly it reflects large differences in what is in the forefront of other
 categorisations, as for example access to economic and other resources, status, social power et cetera.
-We have merged the original 12 classes in HISCLASS, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:
+We have merged the original 12 classes in {\sc Hisclass}, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:
 
 \begin{enumerate}
-\item {\tt elite}, HISCLASS 1 and 2. Higher managers, professionals and higher white collar.
-\item {\tt middle}, HISCLASS 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
-\item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers including farm workers.
+\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals and higher white collar.
+\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
+\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including farm workers.
 %%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
@@
 In the present study, we analyse mortality in the two most common disease
-groups in the ages we are investigating -- cardiovascular diseases and
+groups in the ages we are investigating---cardiovascular diseases and
 cancers.
@@
 
-The important explanatory variable, {\sc Hisclass}, is included in the
+The important explanatory variable, {\tt hisclass}, is included in the
 proportional hazards models as a \emph{stratification} variable. The main
-reason for this choice is that the effect of {\sc Hisclass} on survival is
+reason for this choice is that the effect of {\tt hisclass} on survival is
 non-proportional, that is, it varies with age. With the reduced data set
@@
 Tables with results from the different models are presented at the end of the article, but are not commented on in detail.
-The comments on the results are primarily based on the figures and focuses on the central variable in our study -- {\sc Hisclass}.
+The comments on the results are primarily based on the figures and focuses on the central variable in our study---{\tt hisclass}.
 As can be seen from the tables, most variables in the model are significant at 5 per cent level.
@@
 
-\section{The effect of Hisclass, 1851--1950, 1961--2013}
+\section{The effect of {\sc Hisclass}, 1851--1950, 1961--2013}
 
@@
 From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with sex, period, and hisclass.
+The most interesting feature is the three-way interaction with {\tt sex}, {\tt period}, and {\tt hisclass}.
 
@@
 As before, models are fitted separately for sex and each ten-year time period, from  1901 to
-2013. We stratify on HISCLASS and use the covariates urban/rural and {\tt   civst}. The cumulative
-hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
+2013. We stratify on {\tt hisclass} and include the covariates {\tt urban} (TRUE or FALSE) and 
+{\tt civst} (marital status). The cumulative
+hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
 main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.
@@
 We believe however that cancers were better diagnosed than cardiovascular deaths.
-This is supported by the stability in proportion of cancer deaths between the two first periods, 12,5 per cent 1901--1925 and 13,5 per cent for the period 1926--1950.
+This is supported by the stability in proportion of cancer deaths between the two first periods, 12.5 per cent 1901--1925 and 13.5 per cent for the period 1926--1950.
 
@@
 
-Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods for the age group 40--64.
+Figure~\ref{fig:12cumsw} shows relative mortality levels according to {\sc Hisclass} during the different studied periods for the age group 40--64.
 Cancer mortality among the élite women was lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods.
@@
 
-See Table~\ref{tab:tableeduc}.
+See Table~\ref{tab:tableeducheart}.
 
@@
 
-See Table~\ref{tab:tableeduc}.
+See Table~\ref{tab:tableeduccancer}.
 
@@
 ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.", 
-    label = "tab:tableincpr", digits = 4)
-@
-
-See Table~\ref{tab:tableincpr}.
+    label = "tab:tableincprheart", digits = 4)
+@
+
+See Table~\ref{tab:tableincprheart} for the effect of income on cardiovascular death.
 
@@
 
-See Table~\ref{tab:tableincpr}.
+See Table~\ref{tab:tableincprcancer} for the income effect on cancer death.
 
@@
 pattern of mortality in historical northern Sweden for some time, focusing on
-different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}.
-When it comes to infant and child mortality pre-- and early--transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}.
+different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}. When it comes to infant and child mortality pre---and 
+early---transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}.
 Studies based on microdata on the adult population in history have been rare, mainly due to lack of adequate data.
@@
 In this paper we have analysed social inequality in health from the
-perspective of social classes using HISCLASS. However, there are
+perspective of social classes using {\sc Hisclass}. However, there are
 alternative ways to analyse social inequality, and we cannot exclude that
@@
 Research (CEDAR) at Umeå University and with support from the Swedish
-Research Council (VR) (grant no.
-421-2012-892) and the project Old age, health and inequality: mortality from a life course perspective in Sweden 1986-2009, financed by the Swedish Research Council (grant no 421-2012-892.
-We thank Demographic Data
+Research Council (VR) (grant No. 421-2012-892) and the project 
+\emph{Old age, health, and inequality: Mortality from a life course perspective in Sweden 1986--2009}, 
+ financed by the Swedish Research Council (grant No. 421-2012-892). We thank The Demographic Data
 Base for giving us access to data and Jonas Helgertz, Department of
 Economic History, Lund University, for providing us with scripts for
-creating HISCLASS from Swedish Census data.
+creating {\sc Hisclass} from Swedish Census data.
 

### 2017-06-02 icass.Rnw
new acknowledgement

@@
 
-This work was undertaken within the Centre for Demographic and Ageing
-Research (CEDAR) at Umeå University and with support from the Swedish
-Research Council (VR) (grant No. 421-2012-892) and the project 
-\emph{Old age, health, and inequality: Mortality from a life course perspective in Sweden 1986--2009}, 
- financed by the Swedish Research Council (grant No. 421-2012-892). We thank The Demographic Data
-Base for giving us access to data and Jonas Helgertz, Department of
-Economic History, Lund University, for providing us with scripts for
-creating {\sc Hisclass} from Swedish Census data.
+This work was undertaken within the research program \emph{Ageing and Living Conditions} at the Centre for Demographic and Ageing Research (CEDAR) at Umeå University, financed by the Swedish Research Council (grant No 349-2008-6592) and the project \emph{Old age, health and inequality: Mortality from a life course perspective in Sweden 1986--2009}, financed by the Swedish Research Council (grant No 421-2012-892).
+We thank the Demographic Data Base, Umeå University, for giving us access to data and Jonas Helgertz, Department of Economic History, Lund University, for providing us with scripts for creating {\sc Hisclass} from Swedish Census data.
 

### 2017-06-07 CreateData/creDat.Rnw
koll

@@
 \title{Creating Data for "Social inequality in mortality among
-  adults and elderly in northern Sweden 1850--2013"\footnote{Paper presented
-    at CEDAR, 27 April 2017,
-    Version 2017-03-28.}}
+  adults and elderly in northern Sweden 1850--2013"\footnote{Version 2017-05-27.}}
 
-\date{March 28, 2017}
+\date{\today}
 

### 2017-06-07 Umea.Rnw
first commit
fix

--- 2017-05-19 slides3.Rnw
@@
 \title{Social inequality in mortality among adults and elderly in
-  northern  Sweden 1851--2013\footnote{Seminar April 27, CEDAR, Umeå}
+  northern  Sweden 1851--2013\footnote{Seminar June 10, 2017, ICASS, Umeå}
 }
 
-\date{April 27, 2017}
+\date{\today}
 
-\author{Göran Broström and Sören Edvinsson}
+\author{Sören Edvinsson and Göran Broström}
 
@@
 library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/ume-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+@
+
+<<environ>>=
 library(knitr)
@@
 
+\begin{frame}{Top incomes}
+
+\includegraphics[height=8cm]{myfigs/Top_incomes.jpg}
+
+\end{frame}
+
 \begin{frame}{Outline}
@@
 
-\includegraphics[height=7cm]{myfigs/skum17.pdf}
+\includegraphics[height=7cm]{myfigs/sweden.pdf}
 
@@
 
-\begin{frame}[t]{The FOB and Lisa data}
-
-\vspace{-1.5cm}
-%\begin{columns}
-%\begin{column}{0.6\textwidth}
-<<picklexus,fig.width=4.7,fig.height = 4.8, echo=FALSE>>=
-
-##load("data/vb.rda")  ####################################### NOTE!!!!!!!!!!!!!!!! ############################
-##load("data/vb0.rda")
-tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
-tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
-vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
-d.m <- sum(vbd$sex == "male")
-d.f <- sum(vbd$sex == "female")
-x <- vbd$birthdate + vbd$exit
-y <- vbd$exit
-##par(mfrow = c(1, 2))
-plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
-     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
-     axes = FALSE, ylim = c(28, 90))
-axis(1, at = c(seq(1960, 1990, by = 10), 2002, 2014), las = 1)
-axis(2, at = c(40, 65, 90), las = 2)
-axis(4, at = c(40, 65, 90), las = 2)
-for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
-    ##abline(v = i, lwd = 2, col = "red")
-    lines(c(i, i), c(40, 90), lwd = 2, col = "red")
-}
-for (i in seq(1960, 1985, by = 5) + (365 - 61) / 365){
-    lines(c(i, i + 5), c(35, 40), col = "red", lty = 2, lwd = 2)
-    lines(c(i, i), c(35, 40), col = "red", lty = 2, lwd = 2)
-}
-lines(c(1990, 1990) + (365 - 61)/365, c(28, 40), lwd = 2, lty = 2, col = "red")
-lines(c(1990 + (365 - 61) / 365, 2002), c(28, 40), lwd = 2, lty = 2, col = "red")
-for (i in 2002:2014){
-    ##abline(v = i, lwd = 2, col = "blue")
-    lines(c(i, i), c(40, 90), col = "blue", lwd = 2)
-}
-for (i in 2002:2013){
-    lines(c(i, i), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
-    lines(c(i, i+1), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
-}
-lines(c(2002, 2014), c(40, 40), col = "blue", lwd = 1.5)
-lines(c(2002, 2014), c(90, 90), col = "blue", lwd = 1.5)
-lines(c(1960, 2002), c(90, 90), lwd = 2, col = "red")
-lines(c(1960, 2002), c(40, 40), lwd = 2, col = "red")
-##abline(h = c(35, 40, 65, 90), lty = 2)
-points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
-     pch = ".", col = "red")
-box()
-@
-
-%\end{column}
-
-%\begin{column}{0.4\textwidth}
-%\scriptsize
-
-%\vspace{3\baselineskip}
-
-%\begin{tabular}{l|rr}
-%      & Number & Deaths \\ \hline
-%\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
-%\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
-%Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
-%\end{tabular}
-
-%\vspace{\baselineskip}
-
-%\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are %\emp{excluded}.
-%\end{column}
-%\end{columns}
-\end{frame}
-
-\begin{frame}{The Linnaeus data with deaths}
-
-\begin{columns}
-\begin{column}{0.6\textwidth}
-<<picklexus2,fig.width=6,fig.height = 5.9>>=
-
-##load("data/vb.rda")
-tot.m <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "male"]))
-tot.f <- length(unique(vb0$id[vb0$dataset != "poplink" & vb0$sex == "female"]))
-
-vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
-d.m <- sum(vbd$sex == "male")
-d.f <- sum(vbd$sex == "female")
-
-x <- vbd$birthdate + vbd$exit
-y <- vbd$exit
-##par(mfrow = c(1, 2))
-plot(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p",
-     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
-     axes = FALSE)
-axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
-axis(2, at = c(40, 65, 90))
-axis(4, at = c(40, 65, 90))
-for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
-    abline(v = i, lwd = 2)
-}
-for (i in 2002:2013){
-    abline(v = i, lwd = 2, col = "black")
-}
-abline(h = c(40, 65, 90), lty = 2)
-points(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
-     pch = ".", col = "blue")
-box()
-@
-
-\end{column}
-
-\begin{column}{0.4\textwidth}
-\scriptsize
-
-\vspace{3\baselineskip}
-
-\begin{tabular}{l|rr}
-      & Number & Deaths \\ \hline
-\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
-\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
-Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
-\end{tabular}
-
-\vspace{\baselineskip}
-
-\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{included}.
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}{The Linnaeus data with deaths}
-
-\begin{columns}
-\begin{column}{0.6\textwidth}
-<<picklexus3,fig.width=6,fig.height = 5.9>>=
-
-##load("data/vb.rda")
-tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
-tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
-
-vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
-d.m <- sum(vbd$sex == "male")
-d.f <- sum(vbd$sex == "female")
-
-x <- vbd$birthdate + vbd$exit
-y <- vbd$exit
-##par(mfrow = c(1, 2))
-plot(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "p",
-     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
-     axes = FALSE)
-axis(1, at = c(seq(1960, 1990, by = 5), 2002, 2007, 2014), las = 2)
-axis(2, at = c(40, 65, 90))
-axis(4, at = c(40, 65, 90))
-for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
-    abline(v = i, lwd = 2)
-}
-for (i in 2002:2013){
-    abline(v = i, lwd = 2, col = "black")
-}
-abline(h = c(40, 65, 90), lty = 2)
-points(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "p",
-     pch = ".", col = "blue")
-box()
-@
-
-\end{column}
-
-\begin{column}{0.4\textwidth}
-\scriptsize
-
-\vspace{3\baselineskip}
-
-\begin{tabular}{l|rr}
-      & Number & Deaths \\ \hline
-\textcolor{red}{Women} & \Sexpr{tot.f} & \textcolor{red}{\Sexpr{d.f}} \\
-\emp{Men} & \Sexpr{tot.m} & \emp{\Sexpr{d.m}} \\ \hline
-Total  & \Sexpr{tot.f + tot.m} & \Sexpr{d.f + d.m}
-\end{tabular}
-
-\vspace{\baselineskip}
-
-\emp{\underline{Note}}: Persons with \emp{missing {\sc Hisclass}} (occupation) are \emp{excluded}.
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}{The sampling frame, II}
-
-<<lexis4>>=
-##load("data/vb0.rda")
-vb0 <- vb0[vb0$dataset == "poplink" |
-               (vb0$dataset != "poplink" & vb0$birthdate > 1895), ]
-plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
-     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
-     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
-axis(2, las = 2, at = c(0, 40, 65, 90))
-axis(4, las = 2, at = c(0, 40, 65, 90))
-box()
-abline(h = 0)
-lines(c(1961, 1961, 1986, 2014, 2014, 1961), c(40, 65, 90, 90, 40, 40),
-      col = "blue", lwd = 2)
-#
-n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
-text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(n.rec, "records"))
-text(1900, 55, paste(n.pl, "persons"))
-text(1899.5, 50, paste(n.d, "deaths"))
-#
-n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
-n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
-
-text(1989, 70, "Linnaeus", cex = 2, col = "blue")
-text(1988, 60, paste(n.rec, "records"))
-text(1989.2, 55, paste(n.pl, "persons"))
-text(1989.2, 50, paste(n.d, "deaths"))
-@
-\end{frame}
-
 \begin{frame}{Variables}
@@
 
-\begin{frame}[fragile]{Preliminary Cox regression}
-
-{\scriptsize
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlkwd{library}\hlstd{(eha)}
-\hlkwd{system.time}\hlstd{(fit} \hlkwb{<-} \hlkwd{coxph}\hlstd{(}\hlkwd{Surv}\hlstd{(enter, exit, event)} \hlopt{~}
-                              \hlstd{hisclass} \hlopt{*} \hlstd{period} \hlopt{+} \hlstd{civst} \hlopt{+} \hlstd{urban} \hlopt{+} \hlstd{sex,}
-                         \hlkwc{data} \hlstd{= vb0))}
-\end{alltt}
-\begin{verbatim}
-   user  system elapsed
-151.360   0.444 151.806
-\end{verbatim}
-\begin{alltt}
-\hlkwd{system.time}\hlstd{(dr} \hlkwb{<-} \hlkwd{drop1}\hlstd{(fit,} \hlkwc{test} \hlstd{=} \hlstr{"Chisq"}\hlstd{))}
-\end{alltt}
-\begin{verbatim}
-   user  system elapsed
-455.236   0.944 456.188
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-}
-
-\emp{It takes time (more than 10 minutes)!}
-
-\end{frame}
-
-\begin{frame}{ANOVA}
-
-% latex table generated in R 3.3.3 by xtable 1.8-2 package
-% Mon Apr 17 14:43:45 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{lrrrr}
-  \hline
- & Df & AIC & LRT & Pr($>$Chi) \\
-  \hline
-$<$none$>$ &  & 2227937.83 &  &  \\
-  civst & 2 & 2228508.16 & 574.34 & 0.0000 \\
-  urban & 1 & 2227962.96 & 27.13 & 0.0000 \\
-  sex & 1 & 2231526.62 & 3590.79 & 0.0000 \\
-  hisclass:period & 27 & 2228575.58 & 691.75 & 0.0000 \\
-   \hline
-\end{tabular}
-\end{table}
-
-And with so much data, everything tends to be \emp{statistically} significant.
-
-\end{frame}
-
-\begin{frame}[fragile]{Test of Proportional Hazards}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-    user   system  elapsed
-1063.928    0.916 1064.028
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-Another \emp{19 minutes}!
-\end{frame}
-
-\begin{frame}[fragile]{Proportional hazards, results}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{#system.time(cz <- cox.zph(fit))}
-\hlstd{cz}\hlopt{$}\hlstd{table[}\hlkwd{NROW}\hlstd{(cz}\hlopt{$}\hlstd{table), ]}
-\end{alltt}
-\begin{verbatim}
-    rho   chisq         p
-     NA 2064.07    0.0000
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\begin{itemize}
-\item As usual, \emp{highly significant} result (PH assumption violated).
-\item Should we worry?
-\begin{itemize}
-\item Yes
-\item No
-\end{itemize}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]{Sufficient statistics}
-
-Aggregate:
-
-{\scriptsize
-<<aggregate,eval = TRUE, echo = TRUE>>=
-nr <- NROW(vb0Tab)
-@
-}
-
-\begin{itemize}
-\item Gives a table assuming \emp{piecewise constant hazards} on five-year intervals: 40--45, 45--50, \ldots, 85--90 ("age").
-
-\item Number of cells: $2 \times 10 \times 10 \times 20 \times 2 \times 3 \times 4 = $ \emp{96000}
-
-\item Number of \emp{non-empty} cells: \emp{\Sexpr{nr}}.
-
-\end{itemize}
-
-\url{http://capa.ddb.umu.se/ds/aggregate.html}
-
-\end{frame}
-
-\begin{frame}[fragile]{Exposure by hisclass, men}
-
-<<exphiswom>>=
-xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-who <- length(levels(vb0Tab$hisclass))
-barplot(t(yy)[who:1, ], col = c("blue", "red", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Exposure by hisclass, women}
-
-<<exphis>>=
-xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-who <- length(levels(vb0Tab$hisclass))
-barplot(t(yy)[who:1, ], col = c("blue", "red", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-\end{frame}
-
-\begin{frame}{Age--Period--Cohort data}
-
-\vspace{-1.5cm}
-
-<<moddat,fig.height=4.5,fig.width=5>>=
-x <- c(1961, 2014)
-y <- c(40, 40)
-plot(x, y, type = "l", col = "blue", lwd = 2, xlim = c(1960, 2015), ylim = c(40, 90),
-     ylab = "Age", xlab = "Year", axes = FALSE)
-axis(1, at = c(1961, 1991, 2014))
-axis(2, las = 1)
-box()
-lines(c(1961, 1961), c(40, 65), col = "blue", lwd = 2)
-lines(c(2014, 2014), c(40, 90), col = "blue", lwd = 2)
-lines(c(1961, 1986), c(65, 90), col = "blue", lwd = 2)
-lines(c(1986, 2014), c(90, 90), col = "blue", lwd = 2)
-#
-lines(c(1971, 1971), c(40, 75), col = "blue", lty = 2)
-lines(c(1981, 1981), c(40, 85), col = "blue", lty = 2)
-lines(c(1991, 1991), c(40, 90), col = "blue", lty = 2)
-lines(c(2002, 2002), c(40, 90), col = "blue", lty = 2)
-lines(c(2008, 2008), c(40, 90), col = "blue", lty = 2)
-#
-for (age in seq(40, 65, by = 5)){
-    lines(c(1961, 2014), c(age, age), col = "blue", lty = 2)
-}
-for(i in 0:3){
-    age <- 70 + 5 * i
-    lines(c(1961 + 5 * (i + 1), 2014), c(age, age), col = "blue", lty = 2)
-}
-#
-lines(c(1961, 1961 + 30), c(60, 90), lty = 2, col = "red")
-lines(c(1961, 1961 + 40), c(50, 90), lty = 2, col = "red")
-lines(c(1961, 1961 + 50), c(40, 90), lty = 2, col = "red")
-lines(c(1971, 2014), c(40, 40 + 43), lty = 2, col = "red")
-lines(c(1981, 2014), c(40, 40 + 33), lty = 2, col = "red")
-lines(c(1991, 2014), c(40, 40 + 23), lty = 2, col = "red")
-lines(c(2001, 2014), c(40, 40 + 13), lty = 2, col = "red")
-lines(c(2011, 2014), c(40, 40 + 3), lty = 2, col = "red")
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Counting "cells"}
-
-<<cremodern,echo=FALSE>>=
-library(eha)
-##vb0$hisclass[is.na(vb0$hisclass)] <- 4 ## NOTE!!
-##mode <- cal.window(vb0, c(1961, 2014))
-##modern <- mode[mode$birthdate > 1896, ]
-##modern$hisclass <- factor(modern$hisclass)
-##modern$period <- factor(modern$period)
-##moTab <- aggregate(modern[, c("event", "exposure")],
-  ##                 modern[, c("sex", "age", "period", "cohort",
-    ##                          "urban", "civst", "hisclass")],
-      ##             FUN = sum)
-moTab <- vb0Tab[vb0Tab$period %in% levels(vb0Tab$period)[7:10], ]
-moTab$period <- factor(moTab$period)
-moTab$cohort <- factor(moTab$cohort)
-moTab$age <- factor(moTab$age)
-moTab$hisclass <- factor(moTab$hisclass)
-moTab$hisclass <- relevel(moTab$hisclass, ref = "middle")
-moTab$civst <- relevel(moTab$civst, ref = "married")
-##
-olTab <- vb0Tab[vb0Tab$period %in% levels(vb0Tab$period)[2:4], ]
-olTab$period <- factor(olTab$period)
-olTab$cohort <- factor(olTab$cohort)
-olTab$age <- factor(olTab$age)
-olTab$hisclass <- factor(olTab$hisclass)
-olTab$hisclass <- relevel(olTab$hisclass, ref = "middle")
-olTab$civst <- relevel(olTab$civst, ref = "married")
-
-@
-
-\begin{itemize}
-\item There are \emp{115 cells in one figure}
-\item There are $2 \times 2 \times 3 \times 4 = $ \emp{48 such figures}.
-\item In total: $115 \times 48 = $ \emp{5520 combinations}.
-\item Number of \emp{non-empty} combinations: \emp{4072}.
-\end{itemize}
-
-{\scriptsize
-<<heada>>=
-head(moTab)
-@
-}
-
-An \emp{anonymous} data set!
-\end{frame}
-
-\begin{frame}[fragile]{PH regression with the table}
-
-\begin{itemize}
-\item Piecewise constant hazard model generates the same likelihood as a Poisson model.
-\begin{itemize}
-\item Does \emp{not} imply that the number of deaths follow a Poisson distribution!
-\end{itemize}
-\end{itemize}
-
-\small
-
-<<coxpois,echo=TRUE,cache=FALSE>>=
-system.time(fitp <- glm(event ~ offset(log(exposure)) + age +
-                hisclass * period + civst + urban + sex,
-            data = vb0Tab, family = poisson))
-system.time(dr <- drop1(fitp, test = "Chisq"))
-@
-
-\end{frame}
-
-\begin{frame}{ANOVA with the table}
-
-<<anovatable,results='asis',echo=FALSE>>=
-xtable(dr)
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{The 'baseline hazard function'}
-
-{\scriptsize
-<<extab,echo=FALSE,results='asis'>>=
-xtable(summary(fitp)$coefficients[1:10, 1:2])
-@
-}
-
-\begin{itemize}
-\item \emp{(Intercept)} is log hazard in 40--45. \emp{Reference} interval.
-\item \emp{age45} is the \emp{difference} in log hazard between 45--50 and 40--45,
-\item \emp{age50} is the \emp{difference} in log hazard between 50--55 and 40--45,
-\item Etcetera \ldots
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}[fragile]{Baseline hazards}
-
-<<fixthem,fig.height = 3.4>>=
-alpha <- summary(fitp)$coefficients[1:10, 1]
-alpha[2:10] <- alpha[2:10] + alpha[1]
-oldpar <- par(mfrow = c(1, 2))
-plot(seq(40, 85, by = 5), alpha, col = "red", xlab = "Age", main = "log(hazard)", ylab = "", axes = FALSE)
-axis(1)
-axis(2, las = 1)
-axis(1)
-box()
-x <- seq(40, 85, by = 5)
-plot(x, exp(alpha), col = "blue", type = "b", xlab = "Age", main = "hazard", ylab = "", axes = FALSE)
-axis(1)
-axis(2, las = 1)
-box()
-abline(h = 0)
-par(oldpar)
-##ff <- lm(alpha ~ x)
-##plot(ff)
-@
-
-Why not try:
-
-\begin{equation*}
-\log\{h(t)\} = \alpha + \beta t, \quad 40 < t \le 90,
-\end{equation*}
-
-a \emp{Gompertz} distribution?
-
-\end{frame}
-
-%%\begin{frame}{Period analysis, men}
-
-%%<<coxper,eval=FALSE>>=
-%%oldpar <- par(mfrow = c(2, 3))
-%%men <- modern[modern$sex == "male", ]
-%%##levels(men$hisclass) <- c("elite", "middle", "worker", "NA")
-%%cols = c("black", "darkgreen", "red", "blue")
-%%for (per in levels(men$period)){
-%%    min <- men[men$period == per, ]
-%%   fit <- coxreg(Surv(enter, exit, event) ~ strata(hisclass) + urban * civst, data = min)
-%%   plot(fit, col = cols, main = paste("Men", per), fn = "surv", printLegend = FALSE, xlab = "Age", xlim = c(40, 90))
-%%   legend("bottomleft", legend = levels(min$hisclass), col = cols, lty = 1)
-%%   abline(h = 0.5, lty = 3)
-%%}
-%%par(oldpar)
-%%@
-
-%%\end{frame}
-
-\begin{frame}[fragile]{Survival analysis, men 1961-2013}
-
-<<coxper2>>=
-source("R/plot_surv.R")
-##oldpar <- par(mfrow = c(2, 3))
-men <- moTab[moTab$sex == "male", ]
-cols <- c("darkgreen", "black", "red", "blue")
-##men$hisclass <- relevel(men$hisclass, ref = "middle")
-plot_surv(men, sex = "Men")
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Median life after 40, men 1961-2013}
-
-<<coxpertab>>=
-source("R/mean_pch.R")
-source("R/median_pch.R")
-source("R/plot_median.R")
-
-##pop <- vb0Tab[!(vb0Tab$period %in% oldperiod) &
-  ##                vb0Tab$sex == "male" &
-    ##              vb0Tab$civst != "unmarried" &
-      ##            !vb0Tab$urban, ]
-##pop$hisclass <- relevel(pop$hisclass, ref = "middle")
-##pop$civst <- relevel(pop$civst, ref = "married")
-##pop$period <- factor(pop$period)
-##pop$age <- factor(pop$age)
-plot_median(men)
-@
-
-\end{frame}
-
-\begin{frame}{Survival analysis, women 1961-2013}
-
-<<coxperwom>>=
-par(mfrow = c(2, 3))
-women <- moTab[moTab$sex == "female", ]
-##levels(men$hisclass) <- c("elite", "middle", "worker", "NA")
-##cols = c("black", "green", "red", "blue")
-plot_surv(women, sex = "Women")
-##for (per in levels(women$period)){
-##    min <- women[women$period == per, ]
-##   fit <- coxreg(Surv(enter, exit, event) ~ strata(hisclass) + urban * civst, data = min)
-##   plot(fit, col = cols, main = paste("Women,", per), fn = "surv", printLegend = FALSE, xlab = "Age", xlim = c(40, 90))
-##   legend("bottomleft", legend = levels(min$hisclass), col = cols, lty = 1)
-##}
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Median life after 40, women 1961-2013}
-<<coxpertabwom,warning = FALSE>>=
-source("R/mean_pch.R")
-source("R/median_pch.R")
-##oldperiod <- levels(vb0Tab$period)[1:4]
-##pop <- modern[modern$sex == "female", ]
-##pop <- vb0Tab[!(vb0Tab$period %in% oldperiod) &
-  ##                vb0Tab$sex == "male" &
-    ##              vb0Tab$civst != "unmarried" &
-      ##            !vb0Tab$urban, ]
-##pop$hisclass <- relevel(pop$hisclass, ref = "middle")
-##pop$civst <- relevel(pop$civst, ref = "married")
-##pop$period <- factor(pop$period)
-##pop$age <- factor(pop$age)
-plot_median(women)
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Median life after 40, men 1851-1950}
-
-<<medaf40olmen>>=
-men <- olTab[olTab$sex == "male", ]
-plot_median(men)
-@
-
-\end{frame}
-
-\begin{frame}[fragile]{Median life after 40, women 1851-1950}
-
-<<medaf40olwomen,warning=FALSE>>=
-women <- olTab[olTab$sex == "female", ]
-plot_median(women)
-@
-
-\end{frame}
-
-%\begin{frame}[frame]{Income, men, 1990-2005}
-
-%<<incomemen>>=
-%source("R/plot_income.R")
-%load("data/loTab.rda")
-%loTab$period <- factor(loTab$period)
-%men <- loTab[loTab$sex == "male",]
-%plot_income(men)
-%@
-
-%\end{frame}
-
-% \begin{frame}{Conclusion}
-
-% \begin{itemize}
-% \item Most striking: Very bad prospects for \emp{unemployed}, both men and women, all periods.
-% \item Apart from unemployed: Equal opportunities for men 1960--80, for women 1960--90, but after that a clear separation with the \emp{elite on top}.
-% \end{itemize}
-
-% \vspace{2\baselineskip}
-
-% After this slide: \emp{Old stuff!}
-
-% \end{frame}
-
-\begin{frame}{Proportional hazards for women?}
-
-\includegraphics[width=11cm,height=8cm]{figure/hazplotw-1}
-
-\end{frame}
-
-\begin{frame}{Proportional hazards for men?}
-
-\includegraphics[width=11cm,height=8cm]{figure/hazplotm-1}
-
-\end{frame}
-
-% \begin{frame}[fragile]{Interactions}
-
-% \begin{verbatim}
-% fit <- coxph(Surv(enter - 40, exit - 40, event) ~
-%     period * hisclass * sex + urban, data = vb0)
-% drop1(fit, test = "Chisq")
-% \end{verbatim}
-
-% \begin{center}
-% \begin{tabular}{lrrrr}
-%   \hline
-%  & Df & AIC & LRT & Pr($>$Chi) \\
-%   \hline
-% $<$none$>$ &  & 1631901.32 &  &  \\
-%   urban & 1 & 1631946 & 47.04 & $7.0\times 10^{-12}$ \\
-%   \emp{period:hisclass:sex} & 27 & 1631957 & 110.11 & \emp{$5.3\times 10^{-12}$} \\
-%    \hline
-% \end{tabular}
-% \end{center}
-
-% \begin{itemize}
-
-% \item This model is described by \emp{80 coefficients}.
-
-% \item We show them in \emp{graphs}.
-% \end{itemize}
-
-% \end{frame}
-
 \begin{frame}{No proportional hazards (wrt \sc Hisclass)!}
@@
 
-% \begin{frame}{The data table}
-
-% <<lexis3, fig.height = 4>>=
-% plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
-%      type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
-%      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(40, 90))
-% axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
-% axis(2, las = 2, at = c(0, 40, 65, 90))
-% axis(4, las = 2, at = c(0, 40, 65, 90))
-% box()
-% abline(h = 0)
-% lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
-%       col = "blue", lwd = 2)
-% for (i in c(1851, 1876, 1901, 1926, 1971, 1981, 1991, 2002, 2007)){
-%     lines(c(i, i), c(40, 90), lty = 2)
-% }
-% for (i in c(45, 50, 55, 60, 65, 70, 75, 80, 85)){
-%     lines(c(1851, 1951), c(i, i), lty = 2)
-%     lines(c(1961, 2014), c(i, i), lty = 2)
-% }
-% @
-
-% \emp{Poisson regression}:
-
-% In \emp{each cell}, and for \emp{each} combination of \emp{covariate level}s, calculate
-% \begin{itemize}
-% \item  the number of \emp{deaths} and
-% \item the total \emp{exposure time}.
-% \end{itemize}
-% \end{frame}
-
 \begin{frame}{Women and men, age 40--64}
 
-  \begin{center}
-\includegraphics[height = 7.5cm]{figure/cumsw-1}
-  \end{center}
+%  \begin{center}
+%\includegraphics[height = 7.5cm]{figure/cumsw-1}
+%  \end{center}
+
+<<hazcalw>>=
+source("R/getPHglm2.R")
+##nper <- 4
+vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
+vb4064$age <- factor(vb4064$age)
+labb <- levels(vb0Tab$period)
+nper <- length(labb)
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+            nper, labb)
+@
+
+<<cumsw>>=
+source("R/plotED.glm.R")
+scol <- c("blue", "darkgreen", "red", "black")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
+                main = "Women 40-64", exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
+                main = "Men 40-64")##, exclude = 5:6)
+par(oldpar)
+@
+
 \end{frame}
@@
   %\begin{center}
-\includegraphics[height = 7.5cm]{figure/cumswstand-1}\includegraphics[height=7.5cm]{figure/cumswold-1}
+\includegraphics[height = 7.5cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm]{figs/slide-cumswold-1}
    %\end{center}
-\end{frame}
-
-\begin{frame}{Cancer mortality, ages 40-64 and 65-89}
-
-\includegraphics[height=7.5cm]{figure/12cumsw-1}\includegraphics[height=7.5cm]{figure/12cumswold-1}
-
-\end{frame}
-
-\begin{frame}{Cardiovascular mortality}
-
-\includegraphics[height=7.5cm]{figure/41cumsw-1}\includegraphics[height=7.5cm]{figure/41cumswold-1}
-
 \end{frame}
@@
 
-\begin{frame}[fragile]{Income by education, 1990--2005, ages 40--75!}
-
-<<incomeandeduc>>=
-both <- income[!is.na(income$educ), ]
-levels(both$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-both$educ <- relevel(both$educ, ref = "middle")
-bmal <- both[both$sex == "male", ]
-bfem <- both[both$sex == "female", ]
-fitme1 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-                data = bmal[bmal$educ == "low", ])
-fitme2 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-                data = bmal[bmal$educ == "middle", ])
-fitme3 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-                data = bmal[bmal$educ == "high", ])
-##
-fitfe1 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-                data = bfem[bfem$educ == "low", ])
-fitfe2 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-                data = bfem[bfem$educ == "middle", ])
-fitfe3 <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-                data = bfem[bfem$educ == "high", ])
-
-@
-
-<<plotedin>>=
-oldpar <- par(mfrow = c(2, 3))
-ylim = c(0.5, 1)
-plot(survfit(fitme1, start.time = 40), col = cols, main = "Men, low education", xlab = "Age", ylim = ylim)
-legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.0)
-
-plot(survfit(fitme2, start.time = 40), col = cols, main = "Men, average education", xlab = "Age", ylim = ylim)
-plot(survfit(fitme3, start.time = 40), col = cols, main = "Men, high education", xlab = "Age", ylim = ylim)
-##
-plot(survfit(fitfe1, start.time = 40), col = cols, main = "Women, low education", xlab = "Age", ylim = ylim)
-legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.0)
-
-plot(survfit(fitfe2, start.time = 40), col = cols, main = "Women, average education", xlab = "Age", ylim = ylim)
-plot(survfit(fitfe3, start.time = 40), col = cols, main = "Women, high education", xlab = "Age", ylim = ylim)
-par(oldpar)
-@
-
-\end{frame}
-
-%%\begin{frame}{Education and income, 1990--2005, all deaths}
-
-%%\includegraphics[height=3.5cm]{figure/liswomfig-1} Women, 40--64.
-
-%%\includegraphics[height=3.5cm]{figure/lismanfig-1} Men, 40--64.
-
-%%\end{frame}
-
-%%\begin{frame}{Education and income, 1990--2005, all deaths}
-
-%%\includegraphics[height=3.5cm]{figure/liswomfig2-1} Women, 65--89.
-
-%%\includegraphics[height=3.5cm]{figure/lismanfig2-1} Men, 65--89.
-
-%%\end{frame}
-
-% \begin{frame}{Education and income, 1990--2005, cancer}
-
-% \includegraphics[height=3.5cm]{figure/12liswomfig-1} Women, 40--64.
-
-% \includegraphics[height=3.5cm]{figure/12lismanfig-1} Men, 40--64.
-
-% \end{frame}
-
-% \begin{frame}{Education and income, 1990--2005, cancer}
-
-% \includegraphics[height=3.5cm]{figure/12liswomfig2-1} Women, 65--89.
-
-% \includegraphics[height=3.5cm]{figure/12lismanfig2-1} Men, 65--89.
-
-% \end{frame}
-
-% \begin{frame}{Education and income, 1990--2005, cardiovascular}
-
-% \includegraphics[height=3.5cm]{figure/41liswomfig-1} Women, 40--64.
-
-% \includegraphics[height=3.5cm]{figure/41lismanfig-1} Men, 40--64.
-
-% \end{frame}
-
-% \begin{frame}{Education and income, 1990--2005, cardiovascular}
-
-% \includegraphics[height=3.5cm]{figure/41liswomfig2-1} Women, 65--89.
-
-% \includegraphics[height=3.5cm]{figure/41lismanfig2-1} Men, 65--89.
-
-% \end{frame}
-
 %%% Stop here!
-
-\begin{frame}{Answers(?)}
-
-\begin{enumerate}
-\item Has \emp{inequality in mortality} between \emp{social classes}
-  increased?
-\begin{itemize}
-\item Not really
-\end{itemize}
-%%\item Is the mortality transition initiated by a certain social class?
-
-\item Are there any \emp{gender differences} in the effect of social position?
-\begin{itemize}
-\item Yes
-\end{itemize}
-
-\item Is social position \emp{equally important} among the \emp{retired} population as
-  for those in \emp{working age}?
-\begin{itemize}
-\item Not really
-\end{itemize}
-
-\item Is there a difference between classes in mortality from
-  \emp{cardiovascular} diseases and \emp{cancers} respectively?
-\begin{itemize}
-\item Yes
-\end{itemize}
-\item Are there large differences in survival depending on level of
-  \emp{education} and \emp{income} categories?
-\begin{itemize}
-\item Yes, definitely.
-\end{itemize}
-
-\end{enumerate}
-
-\end{frame}
 
@@
 
-% \begin{frame}{Software for report writing and analysis}
-
-% \begin{itemize}
-% \item  \R : \emp{\url{https://www.R-project.org}}.
-%   \begin{itemize}
-%     \item \emp{RStudio}: \url{https://www.rstudio.com}.
-%   \end{itemize}
-% \item \R\ \emp{packages} (\emp{\url{https://cran.R-project.org}}):
-%   \begin{itemize}
-%   \item \emp{skum} (Broström, 2016) data from \emp{DDB}.
-%   \item \emp{rmarkdown} (RStudio Team, 2016).
-%   \item \emp{knitr} (Xie, 2016).
-%   \item \emp{eha} (Broström, 2016).
-%   \item \emp{xtable} (Dahl, 2016).
-% \end{itemize}
-
-% \item \emp{Version control} (reproducibility)
-%   \begin{itemize}
-%     \item \emp{git}: \url{http://git}
-%     \item \emp{gitHub}: \url{http://github.com/goranbrostrom/}
-
-% \end{itemize}
-% \end{itemize}
-
-% \end{frame}
-
 \end{document}

### 2017-06-07 Umea.Rnw
Revised slides

@@
 
-\title{Social inequality in mortality among adults and elderly in
-  northern  Sweden 1851--2013\footnote{Seminar June 10, 2017, ICASS, Umeå}
+\title{Long--term perspectives of social inequality in mortality among adults and elderly in
+  northern  Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 10, 2017, Umeå}
 }
@@
 
-\author{Sören Edvinsson and Göran Broström}
+\author{Sören Edvinsson and Göran Broström \\
+Centre for Ageing and Demographic Research, Umeå University}
 
@@
 
-\begin{frame}{Questions}
+\begin{frame}{Aims}
 
-\begin{enumerate}
-\item Has \emp{inequality in mortality} between \emp{social classes}
-  increased in the adult and elderly
-  population?
+To investigate the association between social class and mortality among adults and elderly from the 19th century to the present and how it is connected to socio-economic inequality in society.
+This is analysed according to 
+\begin{itemize}
+\item gender
+\item age group (working age vs retired)
+\item cause of death (cardiovascular, cancer)
+\item (education 1990--2013)
+\item (income 1986-2005)
+
+% \item Has \emp{inequality in mortality} between \emp{social classes}
+%   increased in the adult and elderly
+%   population?
 
@@
 
-\item Are there any \emp{gender differences} in the effect of social position?
+% \item Are there any \emp{gender differences} in the effect of social position?
+% 
+% \item Is social position \emp{equally important} among the \emp{retired} population as
+%   for those in \emp{working age}?
+% \item Is there a difference between classes in mortality from
+%   \emp{cardiovascular} diseases and \emp{cancers} respectively?
+% \item Are there large differences in survival depending on level of
+%   \emp{education} and \emp{income} categories?
 
-\item Is social position \emp{equally important} among the \emp{retired} population as
-  for those in \emp{working age}?
-\item Is there a difference between classes in mortality from
-  \emp{cardiovascular} diseases and \emp{cancers} respectively?
-\item Are there large differences in survival depending on level of
-  \emp{education} and \emp{income} categories?
-
-\end{enumerate}
+\end{itemize}
 
@@
 
-\begin{frame}{Top incomes}
+\begin{frame}{Main results}
+
+\begin{itemize}
+\item A consistent mortality difference according to social class appears surprisingly late in our regions and has no clear connection to the development of income inequality.
+\item However, there is a strong gender difference in the development of social inequality.
+\item We argue that health-related behavior is an important aspect in historical as well as present-day mortality, and that such behaviours are based on strong gender and class expectations.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Economic inequality}
 
@@
 \item \emp{Results}
-\item \emp{Alternative} Results
+\item \emp{Implications} of the results
 \end{enumerate}
@@
 
-\begin{frame}{Here!}
+\begin{frame}{Two regions in Västerbotten}
 
@@
 
+Remote area in northern Sweden, dominated by agriculture until well into the 20th century. 19th century communications were rudimentary.
+
 Skellefteå:
-
 \begin{itemize}
 
-\item Small town, heavy industry in the nineteenth century.
+\item Small town, industrialization of the region during the 20th century.
 
@@
 
-\item Not so small town, older. Center for administration, education, and military.
+\item Center for administration, education, and military. University from 1965
 
@@
 
-Rural areas:
-
-\begin{itemize}
-
-\item Very similar: Small farms dominating.
-
-\end{itemize}
+% Rural areas:
+% 
+% \begin{itemize}
+% 
+% \item Very similar: Small farms dominating.
+% 
+% \end{itemize}
 
@@
 \item[parish] Urban/Rural
-\item[\sc Hisclass] elite, middle, worker.
+\item[Social class] elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
 \item[income] 1990--2005, grouped by quartiles.
-\item[education] 1990--2005, seven categories.
+\item[education] 1990--2005, three categories.
 \item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
-\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).
+%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).
 
@@
 
-\begin{frame}{\sc Hisclass}
+\begin{frame}{\sc Social class}
 
@@
 
-\begin{frame}{No proportional hazards (wrt \sc Hisclass)!}
+\begin{frame}{No proportional hazards (wrt class)!}
 
 \begin{itemize}
-\item Fit each {\sc Hisclass} \emp{separately}, for each time period.
+\item Fit each social class \emp{separately}, for each time period.
 \item Use the \emp{cumulative hazard} at 65 (or 90) for comparison.
@@
   \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
-  \item \emp{\tt hisclass}
+  \item \emp{\tt class}
   \item \emp{\tt age group}: 40--64, 65--89.
@@
   %\begin{center}
-\includegraphics[height = 7.5cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm]{figs/slide-cumswold-1}
+\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswold-1}
    %\end{center}
@@
 
-\begin{frame}{Conclusion}
+\begin{frame}{Summarizing the results}
 
@@
 \item \emp{Improved} survival characterised all {\sc Hisclass}es.
-\item There are \emp{gender differences} in the social pattern
-\item No difference in the social pattern in the two studied age groups.
+\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
+\item Distinct \emp{gender} differences in the development.
 \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
-\item \emp{Education} and \emp{income} are the \emp{most important}
-  variables during late
-  time period.
+\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
+\item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
+\item In the long run, no clear connection between income inequality and social differences in mortality.
 \end{itemize}
@@
 
+\begin{frame}{Some implications}
+
+%\scriptsize
+
+\begin{itemize}
+\item The development of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
+\item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
+\item The role of \emp{living standard} for mortality decline (McKeown).
+\item \emp{Gender} and mortality (Watkins, Mckinnon, Janssens).
+\item Social conditions as \emp{fundamental causes of death} (Link/Phelan).
+
+\end{itemize}
+
+\end{frame}
+
+%\begin{frame}{Conclusion}
+
+%\scriptsize
+
+% \begin{itemize}
+% \item \emp{Improved} survival characterised all {\sc Hisclass}es.
+% \item There are \emp{gender differences} in the social pattern
+% \item No difference in the social pattern in the two studied age groups.
+% \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
+% \item \emp{Educational} and \emp{income} income differences during the last decades conform with what we found for class differences.
+% \end{itemize}
+
+%\end{frame}
+
 \end{document}

### 2017-06-08 Umea.Rnw
Small revisions

@@
 \item[parish] Urban/Rural
-\item[Social class] elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
+\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
 \item[income] 1990--2005, grouped by quartiles.
 \item[education] 1990--2005, three categories.
-\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
+\item[period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
 %\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).
@@
 \begin{itemize}
-\item \emp{Improved} survival characterised all {\sc Hisclass}es.
+\item \emp{Improved} survival characterised all social classes.
 \item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
@@
 \begin{itemize}
-\item The development of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
+\item The \emp{development} of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
 \item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
+\item \emp{Status syndrome} (Marmot)
 \item The role of \emp{living standard} for mortality decline (McKeown).

### 2017-06-08 abo.Rnw
first commit (from copy of 'Umeå.Rnw')

--- 2017-06-08 Umea.Rnw
@@
 \title{Long--term perspectives of social inequality in mortality among adults and elderly in
-  northern  Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 10, 2017, Umeå}
+  northern  Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}
 }
@@
 \author{Sören Edvinsson and Göran Broström \\
-Centre for Ageing and Demographic Research, Umeå University}
+Centre for Ageing and Demographic Research \\
+Umeå University, Umeå, Sweden}
 

### 2017-06-09 Umea.Rnw
Last corrections

@@
 \title{Long--term perspectives of social inequality in mortality among adults and elderly in
-  northern  Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 10, 2017, Umeå}
+  northern  Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 11, 2017, Umeå}
 }
 
-\date{\today}
+\date{June 11, 2017}
 
@@
 \item age group (working age vs retired)
-\item cause of death (cardiovascular, cancer)
-\item (education 1990--2013)
-\item (income 1986-2005)
+\item *cause of death (cardiovascular, cancer)
+\item *education 1990--2013
+\item *income 1986-2005
 
@@
 
-\begin{frame}{Economic inequality}
-
-\includegraphics[height=8cm]{myfigs/Top_incomes.jpg}
-
-\end{frame}
-
 \begin{frame}{Outline}
@@
 \end{center}
+
+\end{frame}
+
+\begin{frame}{Economic inequality}
+
+\includegraphics[height=8cm]{myfigs/Top_incomes.jpg}
 
@@
 
-\item[sex]
-\item[birth date] By quarter (15 Feb, May, Aug, Nov)
-\item[death date] By day (\emp{Socialstyrelsen})
-\item[cause of death] from \emp{Socialstyrelsen}
-\item[parish] Urban/Rural
+\item[Sex]
+\item[Marital status]
+\item[Birth date] By quarter (15 Feb, May, Aug, Nov)
+\item[Death date] By day (\emp{Socialstyrelsen})
+\item[Cause of death] from \emp{Socialstyrelsen}
+\item[Parish] Urban/Rural
 \item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
-\item[income] 1990--2005, grouped by quartiles.
-\item[education] 1990--2005, three categories.
-\item[period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
+\item[Income] 1990--2005, grouped by quartiles.
+\item[Education] 1990--2005, three categories.
+\item[Period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
 %\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).
@@
   \item \emp{\tt age group}: 40--64, 65--89.
-  \item \emp{\tt cause of death}
+  \item \emp{\tt *cause of death}
 \end{itemize}

### 2017-06-12 abo.Rnw
Small revision
Åbo version

@@
 \item age group (working age vs retired)
-\item cause of death (cardiovascular, cancer)
-\item (education 1990--2013)
-\item (income 1986-2005)
+\item *cause of death (cardiovascular, cancer)
+\item *education 1990--2013)
+\item *income 1986-2005)
 
@@
 
-\begin{frame}{Economic inequality}
+\begin{frame}{Economic inequality (figure from Roine and Waldenström 2008)}
 
@@
 
-\begin{frame}{Outline}
-
-\begin{center}
-\begin{enumerate}
-
-\item \emp{Data} management
-\item \emp{Modeling} considerations
-\item \emp{Analysis}
-\item \emp{Results}
-\item \emp{Implications} of the results
-\end{enumerate}
-
-\end{center}
-
-\end{frame}
+% \begin{frame}{Outline}
+% 
+% \begin{center}
+% \begin{enumerate}
+% 
+% \item \emp{Data} management
+% \item \emp{Modeling} considerations
+% \item \emp{Analysis}
+% \item \emp{Results}
+% \item \emp{Implications} of the results
+% \end{enumerate}
+% 
+% \end{center}
+% 
+% \end{frame}
 
@@
 
-Remote area in northern Sweden, dominated by agriculture until well into the 20th century. 19th century communications were rudimentary.
+Remote area in northern Sweden, dominated by agriculture until well into the 20th century.
 
@@
 
-\item Small town, industrialization of the region during the 20th century.
+\item Small town with rural hinterland, industrialization of the region during the 20th century.
 
@@
 \begin{frame}{The sampling frame}
+{Two separate datasets}
 
@@
 
-\item[sex]
-\item[birth date] By quarter (15 Feb, May, Aug, Nov)
-\item[death date] By day (\emp{Socialstyrelsen})
-\item[cause of death] from \emp{Socialstyrelsen}
-\item[parish] Urban/Rural
+\item[Sex]
+\item[Marital status]
+\item[Birth date] By quarter (15 Feb, May, Aug, Nov)
+\item[Death date] By day (\emp{Socialstyrelsen}, \emp{Poplink})
+\item[*Death cause] from \emp{Socialstyrelsen} and \emp{Poplink}. Adapted to ICD10
+\item[Parish] Urban/Rural
 \item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
-\item[income] 1990--2005, grouped by quartiles.
-\item[education] 1990--2005, three categories.
-\item[period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
+\item[*Income] 1986--2005, grouped by quartiles.
+\item[*Education] 1990--2013, three categories.
+\item[Period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
+\item[Age group] Separate analyses for working active (40-65) vs retired (65-89) population
 %\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).
@@
 
-\begin{frame}[fragile]{Education 1990--2013}
+% \begin{frame}[fragile]{Education 1990--2013}
+% 
+% <<educ>>=
+% 
+% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
+% educ$period <- factor(educ$period)
+% educ <- educ[!is.na(educ$educ), ]
+% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
+% educ$educ <- relevel(educ$educ, ref = "middle")
+% educ$civst <- relevel(educ$civst, ref = "married")
+% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
+%              data = educ[educ$sex == "male", ])
+% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
+%              data = educ[educ$sex == "female", ])
+% cols <- c("darkgreen", "red", "black")
+% @
+% 
+% <<ploteduc>>=
+% oldpar <- par(mfrow = c(1, 2))
+% 
+% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+% 
+% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
+% 
+% par(oldpar)
+% @
+% 
+% \end{frame}
 
-<<educ>>=
-
-educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
-educ$period <- factor(educ$period)
-educ <- educ[!is.na(educ$educ), ]
-levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-educ$educ <- relevel(educ$educ, ref = "middle")
-educ$civst <- relevel(educ$civst, ref = "married")
-fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-cols <- c("darkgreen", "red", "black")
-@
-
-<<ploteduc>>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
-
-plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
-
-par(oldpar)
-@
-
-\end{frame}
-
-\begin{frame}{Income 1986--2005}
-
-<<femeduc>>=
-income <- lisaold[!is.na(lisaold$income), ]
-income$period <- factor(income$period)
-##income <- income[!is.na(income$income), ]
-##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-income$income <- relevel(income$income, ref = "low")
-income$civst <- relevel(income$civst, ref = "married")
-fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-cols <- c("darkgreen", "red", "blue", "black")
-
-@
-
-<<plotincome>>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
-
-plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
-
-par(oldpar)
-@
-
-\end{frame}
+% \begin{frame}{Income 1986--2005}
+% 
+% <<femeduc>>=
+% income <- lisaold[!is.na(lisaold$income), ]
+% income$period <- factor(income$period)
+% ##income <- income[!is.na(income$income), ]
+% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
+% income$income <- relevel(income$income, ref = "low")
+% income$civst <- relevel(income$civst, ref = "married")
+% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
+%              data = income[income$sex == "male", ])
+% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
+%              data = income[income$sex == "female", ])
+% cols <- c("darkgreen", "red", "blue", "black")
+% 
+% @
+% 
+% <<plotincome>>=
+% oldpar <- par(mfrow = c(1, 2))
+% 
+% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+% 
+% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
+%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
+% axis(1)
+% axis(2, at = c(0, 0.5, 1.0), las = 1)
+% abline(h = 0.5, lty = 3)
+% box()
+% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
+% 
+% par(oldpar)
+% @
+% 
+% \end{frame}
 
@@
 \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
-\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
-\item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
-\item In the long run, no clear connection between income inequality and social differences in mortality.
+% \item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
+% \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
+\item In the long run, no clear connection in the development of income inequality and social differences in mortality.
+\item This suggests an important role of gender and class related expectations on health behaviour
 \end{itemize}

### 2017-06-12 abo.Rnw
Correcting title in final version

@@
 
-\title{Long--term perspectives of social inequality in mortality among adults and elderly in
-  northern  Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}
+\title{Increasing the mortality gap? Social inequality in mortality among elderly and adults in
+  Northern  Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}
 }

### 2017-06-28 data.Rmd
first commit: Swedish

---
title: "Data issues"
author: "Göran Broström and Sören Edvinsson"
date: "28 juni 2017"
output:
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: true
    number_sections: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inledning

**Sören:** Jag har försökt identifiera vilka skillnader som finns mellan våra 
analyser baserade på Poplink respektive Linné. Om du kommer på några fler kan du 
meddela mig det. Vi kan behöva fundera över vad dessa skillnader innebär för 
analysen, och om det finns några risker för säkerheten i våra resultat. Om det 
skulle vara så, kan man behöva överväga sätt att hantera det, alternativt 
diskutera dessa begränsningar i artikeln. Jag tror inte att det ska ha någon 
betydelse, men om något går att göra bättre med befintliga data och utan stora 
extrainsatser kan man förstås överväga förändringar. (Idealt skulle vi göra helt 
nytt dataset från Poplink där vi använder HISCO-koder direkt och låta SCB länka 
till data från 1960 och framåt). Hur som helst bör vi fundera på konsekvenserna 
av skillnader och också vad som är nödvändigt att redogöra för i artikeln.

**Göran:** *Så länge vi noggrant och korrekt beskriver skillnaderna så ser jag 
inga risker för säkerheten i våra resultat.*

# Definition av socialklass

**Sören:** Definition av social klass: I Poplink är social klass hushållsbaserat 
(bygger på hushållsföreståndarens, alltså i allmänhet mannens, yrke) medan Linné 
är individbaserat. Orsaken är dels källmässiga, att vi i praktiken inte kommer 
åt alla individers yrken i Poplink och att det finns problem att definiera 
hushåll i Linné (samboförhållanden osv), samt att arbetsmarknadsförhållanden 
förändrats så att kvinnors arbetslivsdeltagande har kraftigt ökat. Det kan vara 
motiverat att betrakta kvinnor som självständiga på det sättet. Om vi skulle 
göra det hushållsbaserat, skulle vi också behöva hantera en 
klassificeringsproblematik, är det mannens yrke som ska styra, eller kan det 
lika gärna vara kvinnans, alternativt att man tar det som betraktas högst, eller 
att man gör någon sammanjämkning. De skilda principerna gör att alla i samma 
hushåll tillhör samma sociala klass i Poplink, medan de kan tillhöra olika i 
Linné. Det kommer inte att ha någon betydelse för analysen av sociala skillnader 
bland män, men det kan ha vissa effekter för kvinnor. Vi kan säkert ändå 
argumentera att resultaten inte borde påverkas. Män och kvinnor kommer tämligen 
ofta att tillhöra samma sociala klass (skulle kanske behövas referens för 
detta). Det som gör skillnad är att många kvinnor kommer att bli odefinierade 
socialt, men detta finner vi framförallt under 1960- och 70-talet, och vi väljer 
därför att inte presentera analyser för dessa decennier.

**Göran:** *Här är naturligtvis svagheten i vår artikel: Även sent, när vi har 
kvinnans yrke i Linnédatabasen så speglar det inte nödvändigtvis den sociala 
tillhörigheten, eftersom hon kan vara gift med en man i helt annan socialklass 
baserat på yrke. Jag tror att det kan vara vanligt i början på Linnéperioden 
åtminstone. Vi borde verkligen ha länkar mellan sambos/gifta. Jag tror för 
övrigt att Martin Dribe och Maria Stanfors har studerat "poer couples" och kan 
ha relevant information i ämnet.*

#   Närvarodefinition

**Sören:** I Poplink analyseras dödligheten efter närvaron i våra regioner, en 
närvaro som vi kan följa kontinuerligt. Dödsfallen sker därför också i dessa 
regioner. I Linnédatat 1960-2000 identifierar vi närvaron utifrån tvärsnitten i 
Fobbarna vart femte år 1960-1990. Vi betraktar individerna som under observation 
fram till följande Fob, och om de också då är närvarande fram till nästa Fob 
osv, alternativt fram till dödsdatum om det sker mellan två Fobbar. Dödsfallen 
inom intervall mellan två Fobbar (eller Fobben 1990 fram till 2000) kan inträffa 
vart som helst i Sverige. Det blir således en viss skillnad mot vad vi gör för 
Poplink. Dödsfall behöver inte inträffa i regionen, och vi kommer att 
(marginellt) överskatta närvaron i regionerna. För de två sista perioderna 
(2002-2007, 2008-2013) följer vi däremot individerna kontinuerligt (kan vi väl 
säga, vi har uppgift om närvaro under året) och vi analyserar dödsfallen i våra 
regioner (visst stämmer det?). I teorin skulle vi också kunna skapa ”exakta” 
närvaroperioder för hela tiden 1986-2013 (el 1985-2013) från LISA-data, men det 
skulle kräva en del ytterligare organisering av data.

**Göran:** *Här är jag däremot inte så orolig: Vad vi beskriver är utvecklingen 
för de som var närvarande i vår region som fyrtioåringar. Möjligen skulle vi 
inte censurera de som inte är närvarande i "nästa" Fob utan att ha dött på 
vägen. Ja, det vore kanske det mest konsekventa. Tyvärr kan vi inte göra 
motsvarande för Poplinkperioden, så vi släpper det.*

# Församlingsunderlaget

**Sören:** Poplink innehåller data från ett antal församlingar i vad som är 
DDB:s Skellefteå- och Umeå-regioner. Från Linnédatabasen använder vi de som är 
boende i Skellefteå, Malå och Umeå kommuner. I huvudsak sammanfaller det 
geografiska området över tid, men det tillkommer några fler församlingar i 
analyserna efter 1960. I princip skulle man kunna identifiera exakt 
församlingarna i Poplink också i Linné, men det tillför knappast något av värde, 
och skulle inte förändra resultaten.

**Göran:** *Vi har ju också den stora geografiska skillnaden före och efter 
förra sekelskiftet, då Umeå tillkommer. Vi borde egentligen hålla Umeå utanför
det hela, alternativt ha en separat artikel om denna Västerbottens pärla.*

# Undersökta tidsperioder

**Sören:** Skellefteå-regionen analyseras 1851-2013, medan Umeå-regionen är 
begränsad till tiden 1901-2013. Konsekvensen är att vi troligen undersöker 
en mindre socialt diversifierad miljö under 1800-talet. Tidigare gjorde vi 
separata analyser för de två regionerna, men resultaten var väl inte olika i 
något avgörande avseende?

**Göran:** *Se ovan!*

# Skapandet av social klass

**Sören:** I Linné baseras social klass på en omkodning av Fobbarnas och LISA:s 
yrken till HISCO och därifrån till HISCLASS vilket sedan förts samman till de 
tre grupper vi har valt. För Poplink-tiden är underlaget baserat på DDB:s 
traditionella yrkeskodning men redan grupperat i större sociala kategorier för 
att skapa närvaroperioder enligt klass. Dessa kategorier har vi identifierat 
enligt kategorierna i HISCLASS. Detta har i huvudsak varit oproblematiskt, men 
det finns vissa placeringar som skulle krävt att vi hade mer exakt 
yrkesinformation för att få systemet att helt motsvara HISCLASS. Vi hanterar 
detta rätt väl genom att bara använda tre kategorier, något som således har en 
praktisk komponent men som vi också kan motivera utifrån teoretiska skäl.

**Göran:** *Hänger samman med "Definition av socialklass". Jag känner en viss 
oro här, men som jag påpekade inledningsvis: Redovisa ärligt och läsaren får
själv dra sina slutsatser (med viss hjälp av författarna ...).*

### 2017-07-01 article.Rmd
first commit

--- 2017-06-01 icass.Rnw
@@
-\documentclass[a4paper,11pt,titlepage]{article}
+---
+title: "Long-term perspectives of social inequality in mortality among
+  adults and elderly in northern Sweden 1851--2013"
+author: "Sören Edvinsson and Göran Broström"
+date: "28 juni 2017"
+output:
+  bookdown::html_document2:
+    toc: yes
+    toc_depth: 2
+    toc_float: true
+    number_sections: true
+bibliography: ["mybib.bib"]
+biblio-style: "apalike"
+link-citations: true
+---
 
-\usepackage[utf8]{inputenc}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
-\usepackage{float}
+# Introduction
 
-\usepackage{epstopdf}
-\usepackage{url}
-
-\usepackage{natbib}
-
-%\bibliographystyle{apalike}
-\bibliographystyle{plainnat}
-
-%\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\newcommand{\R}{{\bf R}}
-
-\usepackage[tablesonly]{endfloat}
-%\usepackage[figuresonly]{endfloat}
-%\usepackage[tablesonly]{endfloat}
-
-\renewcommand{\efloatseparator}{\vfill}
-
-% patch \efloat@iwrite to use \protected@write
-\makeatletter
-\renewcommand\efloat@iwrite[1]{%
-   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-
-\author{Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} and
-Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}}
-     \\
-Centre for Demographic and Ageing Research ({\sc Cedar})\\
-Umeå University, SE-90187 Umeå, Sweden}
-
-\title{Long--term perspectives of social inequality in mortality among
-  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
-    at the
-    session 'Health and Mortality in Arctic Populations'. The Ninth International
-    Congress on Arctic Social Sciences,
-    Umeå,
-    June 8--12, 2017}}
-
-\date{May 31, 2017}
-
-<<setup, include = FALSE>>=
-library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
-library(knitr)
-library(eha)
-library(xtable)
-library(skum)
-@
-
-<<readalldata>>=
-## Load remote data:
-con <- url("http://capa.ddb.umu.se/data/vb0.rda")
-##load("data/vb0.rda")
-load(con)
-close(con)
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
-##
-con <- url("http://capa.ddb.umu.se/data/vb.rda")
-##load("data/vb.rda")
-load(con)
-close(con)
-##
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-##load("data/lisaold.rda")
-load(con)
-close(con)
-##
-con <- url("http://capa.ddb.umu.se/data/loTab.rda")
-load(con)
-close(con)
-loTab$period <- factor(loTab$period)
-##
-vb0Tab <- aggregate(vb0[, c("event", "exposure")],
-                   vb0[, c("sex", "age", "period", "cohort",
-                          "urban", "civst", "hisclass")],
-                   FUN = sum)
-vb0Tab$age <- factor(vb0Tab$age)
-@
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-A long-term perspective on the development of social
-inequalities in mortality in the adult and elderly population is taken in
-this paper. The area under investigation consists of the Skellefteå and Umeå regions in
-the north of Sweden, and the time period is 1851--2013. The outcome is all
-cause mortality and cause specific mortality for the categories
-cardiovascular diseases and cancers. The analysis of inequality is based on
-three social classes, reorganized from Hisclass.
-
-The main findings are that a consistent social gradient implying better
-survival for classes with more access to economic and other resources, only
-become evident in the later half of the 20th century, with lower classes
-having the highest mortality. The class patterns in mortality differed
-however depending on gender. Women of higher classes were comparatively
-favoured in most of the studied periods, while higher class men were
-disadvantaged when it comes to survival at least until the middle of the
-20th century. The modern pattern of clear socio-economic differences
-appeared surprisingly late. Additional and alternative analyses based on education and
-income categories confirmed large inequality in mortality
-between different groups during the most recent decades.
-There were no clear signs of qualitative differences depending on whether it was the working-active or the retired population that was analysed, but the differences according to social class was larger in the first-mentioned group.
-Both cardiovascular diseases and cancers conform with the pattern already observed when it comes to all-cause mortality.
-
-\end{abstract}
-
-\tableofcontents
-
-\newpage
-
-%%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}
-
-%%\subsection{Combining the data sets and tabulate}
-
-\section{Introduction}
-
-One of the central aspects of survival is social class and access to economic and other resources.
-Even in present-day welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}.
-It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}.
-For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies.
-This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
-Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
-\citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases.
-According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease.
-Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health.
-Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences.
-\citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.
-
-Recent studies, investigating social inequality in health and mortality with micro-data, has however questioned how generalizable the assumed pattern is \citep{bevp11}.
-Instead, some studies suggest that the modern pattern might be a fairly recent development, although the pattern differs depending on age, gender or historical context.
-When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small or even reversed socio-economic differences during the first stages of the transition \citep{bedr11, edli11,edbr12}, sharply contrasting against what we know from present-day society.
-
-Another question of the development of social inequality in mortality, concerns its relation to economic inequality.
-\citet{wipi09} argue that income inequality has an independent effect on mortality, separate from the direct effect of actual different access to economic resources.
-They argue that unequal societies perform less well when it comes to health than equal ones.
-Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world.
-This has initiated a vital scholarly debate and the topic has been extensively studied \citep{suka04, wado00}.
-How the association looked like in previous periods is unknown.
-This is of particular interest since the levels of economic inequality were very different from those of the recent decades.
-Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.
-The situation during the 19th century is uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
-From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980.
-Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}.
-The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.
-
-It thus appears as if a more consistent social health divide appeared at some point during the mortality transition, at least in Sweden.
-Solid empirical evidence about the process is however lacking and studies focusing on the issue are still few.
-There is a need for more reliable studies from different geographical and historical settings as well as according to age and gender in order to better understand the role of socio-economic conditions for health and survival over time.
-
-% This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
-% the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
-% in time for the Skellefteå region (1851--1900), and we also consider causes
-% of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
-% us to study the effect of income and education on mortality in a more
-% direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.
-
-\section{Aims and questions}
-
-In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
-The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
-The main question we investigate is if inequality in mortality between social classes increased among the adult and elderly population, primarily benefiting the highest social positions and how this is connected to the development of economic inequality in society.
-This issue is then analysed according to gender, age-group (working age vs retired) and cause of death (cardiovascular and cancer).
-Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period 1986--2013.
-
-% ? :
-% 
-% \begin{enumerate}
-% \item Has inequality in mortality between social classes increased among the adult and elderly
-%   population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
-% 
-% %%\item Is the mortality transition initiated by a certain social class?
-% 
-% \item Are there any gender differences in the effect of social position?
-% 
-% \item Is social position equally important among the retired population as for those in working age?
-% \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
-% \item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
-% \end{enumerate}
-% 
-%
-
-\section{The Skellefteå and Umeå regions}
-
-% Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
-% show the study areas and their position in Sweden.
-
-% \begin{figure}
-% \includegraphics[height=19cm]{myfigs/skum}
-% \caption{The Skellefteå and Umeå regions.}
-% \label{fig:SKEUME}
-% \end{figure}
-
-\begin{figure}
-\includegraphics[height=19cm]{myfigs/sweden}
-\caption{The Skellefteå and Umeå regions in Sweden.}
-\label{fig:mapp}
-\end{figure}
-
-The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of Gulf of Bothnia.
-This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century.
-The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s.
-One living long in memory was the famine of 1867–-1868 \citep{segb14}.
-During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
-The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}.
-Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures.
-The inland was particularly remote and isolated.
-During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
-Diseases reached to places that earlier were spared.
-
-The Skellefteå region until the 1950 in our dataset consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
-The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
-The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production.
-During the 20th century, industrialisation took place, partly by the opening of mines in the in-lands.
-This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.
-
-Mortality was fairly low in comparison with other parts of the country and the fertility transition was late \citep{coa86}.
-The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.
-
-The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality when another three parishes are included.
-This region has a somewhat different character from that of Skellefteå.
-Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period.
-It was the administrative centre in the county of Västerbotten.
-Schools and military regiments were placed here and the economy was more diversified compared to Skellefteå.
-Agriculture dominated the rural part, but there were also some foundries as well as industries for example in forestry and small-scale production.
-The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.
-
-\section{Data and variables}
-
-The recent extension in time of data at the Demographic Data Base (DDB), Umeå University (www.cedar.umu.se) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
-The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}.
-The data for the present study comes from two large population databases at the DDB.
-The early period until 1950 is covered by
-the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked parish records, allowing
-us to reconstruct life biographies on people as long as they remained in
-the region. The records are linked within but not between the regions. Data
-from Poplink are accessed for the period 1851--1950.
-
-The other large data-set is extracted from the Linnaeus database
-\citep{malm10}. Linnaeus is based on different linked national population registers from
-1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1960 to 2000 are constructed from censuses every fifth
-year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
-For the period 2002--2013 we use the information from the yearly population registers from Statistics Sweden (LISA) together with death information from National Board of Health and Welfare.
-
-Information on individuals are anonymised and the two databases are not
-linked. This stops us from following individuals present
-in both databases throughout their lives, thus they are treated separately. This
-also makes it impossible to add information on individuals in the Linnaeus
-database from what we potentially could find in Poplink, for example family
-background or occupations earlier in life.
-
-In the data-set analysed here, all individuals 40 years and older ever being resident in either
-of the regions are included. The data file contains the variables on social class (HISCLASS) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
-entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
-
-<<lexis2,fig.cap="The sampling frame.", fig.height = 5>>=
-plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
-     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
-     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
-axis(2, las = 2, at = c(0, 40, 65, 90))
-axis(4, las = 2, at = c(0, 40, 65, 90))
-box()
-abline(h = 0)
-lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
-      col = "blue", lwd = 2)
-#
-n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
-text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(n.rec, "records"))
-text(1900, 55, paste(n.pl, "persons"))
-text(1899.5, 50, paste(n.d, "deaths"))
-#
-n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
-n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
-
-text(1985, 70, "Linnaeus", cex = 2, col = "blue")
-text(1984, 60, paste(n.rec, "records"))
-text(1985.2, 55, paste(n.pl, "persons"))
-text(1985.2, 50, paste(n.d, "deaths"))
-@
-
-\subsection{Presence periods}
-
-Differences in available information in the two datasets as well as within the Linnaeus data makes it necessary to apply different approaches when it comes to identification of presence periods within our
-regions. The Poplink data provides us with exact dates or at least year of
-start and exit of the presence of all individuals in our data, allowing us to have full and continuous control of the de jure population.
-This is however not the case with the Linnaeus database.
-For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare to identify the observation periods in our analyses.
-There is however no full information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
-We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in the later census.
-This overestimates their presence in our regions, but should not bias the results when it comes to socioeconomic differences.
-We furthermore miss those that both moved in and moved out or vice versa between censuses.
-Exit dates are exact if they refer to death, and in these cases we consider them as constantly present from the last census to the death data even if the death took place in another part of Sweden.
-
-For the last period, 2002--2013, we use the yearly population registers (LISA data), together with information on deaths from National Board of Health and Welfare.
-This data allows good control of the presence periods in the studied regions.
-
-\subsection{Social class}
-
-In this paper we analyse mortality differences according to social class based on the classification scheme HISCLASS \citep{lema11}.
-The only type of socio-economic information available both in Linnaeus and Poplink is occupations.
-Thus, analysing according to social class is apart from being a scientific choice, also a practical one.
-Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes.
-Social position is treated as a time-dependent variable.
-
-The availability to information on occupations varies over time and between sources.
-Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system.
-The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
-The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information.
-In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid.
-Thus, the social class in the census 1990 is considered to be kept until 2001.
-This strategy is also used for other cases of missing social class.
-Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65.
-For the Linnaeus data, we use a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.
-
-The use of different data sets, forces us to apply different approaches towards defining the social position in some respects.
-A problem is that Poplink usually only provides the occupation for the head of household, thus under-estimating female working-force participation as well as that of adult children residing with their parents \citep{vik10}.
-In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
-This strategy makes the problem of under-recording less problematic.
-
-Female labour is much better covered in the Linnaeus database.
-A complication is however that it is  difficult to define households in the same way as in Poplink.
-Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became common and female labour force participation developed as the norm in Sweden during this period.
-The occupation of the husband as signifier of social class became, if not obsolete, at least less relevant.
-All included persons are signified by their individual occupation for this period.
-Still, the results for the different periods are not completely comparable and makes the interpretation of the development for women more shaky.
-
-HISCLASS, the classification system used here, is a ”. . . HISCO-based
-historical international social class scheme” \citep{lema11, lema02}.
-The different classes in HISCLASS represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector.
-Implicitly it reflects large differences in what is in the forefront of other
-categorisations, as for example access to economic and other resources, status, social power et cetera.
-We have merged the original 12 classes in HISCLASS, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:
-
-\begin{enumerate}
-\item {\tt elite}, HISCLASS 1 and 2. Higher managers, professionals and higher white collar.
-\item {\tt middle}, HISCLASS 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
-\item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers including farm workers.
-%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
-\end{enumerate}
-
-% The procedure for
-% occupations in the Linnaeus database is a little more complicated. The
-% occupations in the censuses (NYK) are first transferred to the system
-% later in use, SSYK (the Swedish version of ISCO88). Thereafter the
-% SSYK, both those transferred from the censuses and those given in the
-% population registers 2001-2013, can be re-coded into HISCLASS.
-
-Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution (per cent) of exposure
-time according to class for men and women respectively. The category of workers is
-fairly stable over time. The highest class has increased substantially, although from a very low level in the 19th century. The increase of academics in connection to the establishment of
-Umeå University explains much of this increase.
-What is hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century.
-Furthermore we observe an increase during the 1960s and partly during later decades as well in those with undefined.
-When first observed in the earliest included censuses, occupation is often missing.
-For men the proportion is somewhat higher in the age group 65 and above, i.e.
-those that were retired, but still low enough to be used in our analyses.
-Until 1950, there were however very few with missing occupation.
-For men in working-active age, the proportion missing does not cause any problems for our analyses.
-
-<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
-xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
-xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-% <<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
-% xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-% yy <- round(prop.table(xx, 1) * 100, 1)
-% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-% @
-% 
-% <<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
-% xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-% yy <- round(prop.table(xx, 1) * 100, 1)
-% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-% @
-
-The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}.
-Corresponding to what we found for men, the proportion without social position is very low until the middle of the 20th century.
-This is a reflection of how womens social position is defined for that period.
-Missing social class is however a real problem during the 1960s and 1970s in particular.
-Around 80 \% lack information for the 1960s and 50 \% for the 1970s.
-Since social class for this period is defined according to own occupation, the consequence of many women being outside the workforce, the proportion missing becomes very high.
-We have chosen not to report results for women for these two decades.
-For the later decades, the results are fairly complete and should be representative.
-
-% The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
-% \ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.
-
-% <<nanam, results = 'asis'>>=
-% library(xtable)
-% ##load("data/vb0.rda")
-% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-% x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
-% x <- round(prop.table(x, 1) * 100)
-% xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
-% @
-% 
-% <<nanaw, results = 'asis'>>=
-% ##library(xtable)
-% ##load("data/vb0.rda")
-% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-% x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
-% x <- round(prop.table(x, 1) * 100)
-% xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
-% @
-
-\subsection{Marital status}
-
-Marital status is treated as a time-dependent variable in the data file.
-The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces).
-If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.
-
-Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.
-The figure indicates an under-registration of those married and and over-registration of previously married during the 1960s.
-
-<<civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5>>=
-xx <- with(vb0Tab, tapply(exposure, list(period = period, civst = civst), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[3:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
-@
-
-\subsection{Cause of death}
-In the present study, we analyse mortality in the two most common disease
-groups in the ages we are investigating -- cardiovascular diseases and
-cancers.
-National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year.
-Classification systems have however changed.
-Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
-In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
-
-The distribution of causes of death is shown in Figure~\ref{fig:causeskum}.
-Notice the high proportion of other causes in the first two periods.
-This category include cases with unknown or unspecific causes which has implications for the cause-specific analyses (see later discussion).
-The total dominance of our two disease groups is apparent.
-
-<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5>>=
-vb01 <- vb0[vb0$event == 1, ]
-cancer <- tapply(vb01$cancer, vb01$period, sum)
-heart <- tapply(vb01$heart, vb01$period, sum)
-death <- tapply(vb01$period, vb01$period, length)
-other <- death - cancer - heart
-ttt <- rbind(cancer, heart, other)
-ttt <- prop.table(ttt, 2) * 100
-barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = "Per cent")
-@
-
-\subsection{Urban vs rural residence}
-In 19th and early 20th century Sweden (as in other countries), mortality was much higher in urban environments.
-The unsanitary and overcrowded living conditions as well as the risks for importing infections made the health risks higher in towns and cities.
-In the present paper, we control for this aspect by distinguishing between urban and rural residence, based on in which parish people lived.
-
-\subsection{Periods for analysis}
-We have made separate analyses for the different periods.
-The century 1851--1950 is divided into four periods of 25 years each.
-The 19th century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
-During the years 1901--1950, industrialisation started and the urban environments increased in population size.
-The modern welfare state began to develop during this time.
-For the period 1960 on-wards, there are separate analyses on decades except for the two last periods.
-Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
-A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to being less generous than before.
-The start of the last period (2008--2013) coincides with the international financial crisis.
-
-\subsection{Income}
-Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to 2005.
-
-\subsection{Education}
-Education is here analysed only for the period 1990 to 2013.
-Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
-It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.
-
-\section{Models}
-
-The proportional hazards model in a survival analysis context is used,
-allowing for adjustment for civil status and rural/urban environment. Since
-the central explanatory variable, {\sc Hisclass}, do not follow the
-proportionality property, the analyses are stratified with respect to the
-variable, and the main results are presented graphically.
-
-Another practical problem is the size of that data set:
-It contains no less than 2\,432\,871 records of 236\,011 individuals, of
-which 68\,244 are observed to die.
-This makes it very inconvenient to use the nowadays so popular semi-parametric
-technique known as
-\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
-that allows for reduction of the data
-to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
-relevant covariates \emph{categorical}, and then create a table with all
-possible combinations of levels of the covariates. For each such
-combination, the total amount of exposure and the number of deaths are
-calculated.  We have two sexes, three civil statuses, ten periods, two
-urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
-ten age groups, and 22 cohorts, in all 422\,400 possible
-combinations. However, in this scheme there are many structural zeroes
-(impossible combinations) and some random zeros, so the actual number of
-combinations we end up with is 24\,326. The data set is thus reduced in
-size by 99 per cent.
-
-The parametric model we use for the baseline hazard function is piece-wise
-constant on five-year intervals, so we arrive at the following proportional
-hazards regression model:
-
-\begin{equation}\label{eq:prop}
-h(t; \lambda, \boldsymbol{\beta}, \textbf{x}) =
-\lambda(t)e^{\boldsymbol{\beta} \mathbf{x}}, \; \; 40 < t \le 90,
-\end{equation}
-where
-
-\begin{equation*}
-\begin{split}
-\mathbf{x} &= (x_1, \ldots, x_n) \text{\ is a vector of covariates,} \\
-\boldsymbol{\beta} &= (\beta_1, \ldots, \beta_n) \text{\ is the corresponding vector of regression
-parameters.} \\
-\end{split}
-\end{equation*}
-and
-\begin{equation*}
-\lambda(t) = \alpha_i,\; 40 + 5(i - 1) < t \le 40 + 5i, \; i = 1, \ldots, 10.
-\end{equation*}
-
-Since all components of $\mathbf{x}$ are categorical covariates, the
-estimation of \eqref{eq:prop} can be performed in two ways, first by the
-original data set and a survival analysis routine handling a piecewise
-constant hazard, or second, by the reduced data set (table) and
-\emph{Poisson regression}. The second way is faster than the first by a
-factor of size at least 100, often much more, depending on hardware, and
-the point is that the two ways give exactly the same results. This is
-because the table is a \emph{sufficient statistic} for \eqref{eq:prop} and
-the original data set.
-
-The analyses are performed in the \R\ environment for statistical computing
-and graphics \citep{cran}, especially using the package {\tt eha} 
-\citep{gb12,eha}
-The proportional hazards model allow us to scrutinise the differences and
-the development in more detail. We divide the analysis into two age groups,
-40--64 and 65--89 years of age, the first age group representing people that
-were still mainly in the workforce, and the second those that mainly were 
-retired. The model controls for marital
-status and whether the individual resides in the urban or rural parts of the regions. There
-are no serious signs of deviation from proportionality assumption, except
-regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
-in order to show the development of the social patterning of mortality
-during the studied period.
-
-% We analyse both regions together but perform separate analyses for men and
-% women and for the age groups 40--64 and 65--89. The model in the analyses
-% include social class, sex and marital status.
-
-% We present the results for the period 2001--2013
-% separately. We suggest a heterogeneity index in order to investigate a
-% possible homogenisation of mortality between social classes.
-
-The important explanatory variable, {\sc Hisclass}, is included in the
-proportional hazards models as a \emph{stratification} variable. The main
-reason for this choice is that the effect of {\sc Hisclass} on survival is
-non-proportional, that is, it varies with age. With the reduced data set
-and Poisson regression, this is simply achieved by introducing an
-interaction term between {\tt age} and {\tt hisclass}. Then, the main
-results are presented \emph{graphically}.
-Tables with results from the different models are presented at the end of the article, but are not commented on in detail.
-The comments on the results are primarily based on the figures and focuses on the central variable in our study -- {\sc Hisclass}.
-As can be seen from the tables, most variables in the model are significant at 5 per cent level.
-
-% There is some  lack of information in the higher
-% ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
-% information is simply missing to a large extent.
-
-\section{The effect of Hisclass, 1851--1950, 1961--2013}
-
-\subsection{All causes of death}
-
-<<colorc>>=
-## Colors for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
-## Common for all groups:
-
-labb <- levels(vb0Tab$period)
-nper <- length(labb)
-#source("R/getPHglm.R")
-## End 'common'
-@
-
-The cumulative
-hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of
-graphically illustrating the development over time.
-
-%creating an \emph{heterogeneity index} (a better name is needed:
-%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
-%Something catchy.).
-
-\subsubsection{Ages 40--64}
-
-%The distribution over time periods.
-<<getske4060>>=
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
-scap1w <-  "Cumulative hazards, ages 40--64, women."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
-scap1m <-  "Cumulative hazards, ages 40--64, men."
-##
-cap0 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap0 <- "Total hazard by HISCLASS, ages 40--64."
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard standardised by HISCLASS, ages 40--64."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women."
-scap2w <-  "Cumulative hazards, ages 65--89, women."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men."
-scap2m <-  "Cumulative hazards, ages 65--89, men."
-##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). Standardised."
-scap22 <- "Total hazard by HISCLASS, ages 65--89."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years."
-
-##
-vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
-vb4064$age <- factor(vb4064$age)
-##vb4064$hisclass <- vb4064$hisclass40
-@
-
-% We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
-% periods, in order to illustrate the process. They are adjusted for {\tt civil status}. We present separate analyses  for women and men, see
-% Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.
-
-%\subsubsection{Women}
-
-<<hazcalw>>=
-source("R/getPHglm2.R")
-##nper <- 4
-nper <- length(labb)
-resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
-            nper, labb)
-@
-
-% 
-% 
-% <<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
-% source("R/plotHaz.glm.R")
-% 
-% plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
-% @
-% 
-% <<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
-% plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
-% @
-% 
-% From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
-% the idea of proportional hazards is not reasonable. We therefore use another approach.
-
-%The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
-%Figure~\ref{fig:cumsw}.
-
-<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8>>=
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Women 40-64", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Men 40-64")##, exclude = 5:6)
-par(oldpar)
-@
-
-From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with sex, period, and hisclass.
-
-<<cumswtab,results='asis',cap="Table">>=
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period, 
-            data = vb4064, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, 40-65.")
-@
-
-Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval 
-40--64 during the different periods. Obviously, and not surprisingly, mortality 
-is decreasing over time for both sexes. Women have consistently better survival 
-than men. This is a well-known phenomenon in Swedish
-demographic history. Male mortality has almost always been higher than
-female in all age groups \citep{wil99, suwi07}. We also observe a
-substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. The compression that is
-visible for later decades is partly explained by the decreasing general
-mortality and is not necessarily representing a homogenisation of relative risks.
-
-By \emph{standardising} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these 
-relative mortality levels for the different classes.
-
-<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Women 40-64", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Men 40-64")##, exclude = 5:6)
-par(oldpar)
-@
-
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
-%%@
-
-%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-Notice that we have not analysed the 1950s due to the fact that we only have data for some parishes at that time, consequently having no corresponding population for that decade.
-Furthermore, we have excluded the results for women during the 1960s and 1970 of reasons discussed above.
-The changing economy also have consequences for the analysis, particularly when it comes to the agricultural sector.
-While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter.
-This means that the middle class has somewhat different character during the last periods.
-
-The most striking result of the analysis is the different patterns for men and women. The social pattern of mortality among women is quite much in accordance with previous
-analyses from the present research group on other regions in northern
-Sweden and for other periods \citep{edli11,edbr12}. The highest
-social class usually have comparatively low mortality, while working
-class women have high.
-There is however no substantial change in the relative risks between the classes.
-Only the group with unknown social class is distinguished by moch higher mortality from the 1980s onwards.
-This can however have to do with the differences how social position for women is defined before and after 1950.
-After 1950 it is based on the own occupation.
-Being without occupation might include women that are more frail both when it comes to health and socially.
-
-A different pattern appears among men, being the most surprising finding in our analyses. The groups we would expect to have the lowest mortality, if we assume that access to different resources and
-having a high status determines survival, instead have by far the highest hazards of dying during the nineteenth and much of the 20th century. This is also in accordance with our previous studies
-\citep{edli11, edbr12}.
-Workers on the other hand have comparatively low mortality until the 1960s, at the same level as the middle class.
-The relative positions change radically towards the end the 20th century.
-From the 1980s, there is a clear advantage when it comes to survival for higher
-managers and professionals.
-The relative differences are however more compressed than for the period until 1950.
-The expected social gradient in mortality has developed.
-This is also what we usually find in analyses of social health inequalities in present-day Sweden as well as in most
-other countries. In the regions we are studying, this is however a quite
-recent phenomenon.
-Corresponding to what we find for women, those without known occupation have much higher mortality than the others from the 1960s onwards.
-The assumption made for women of unknowns social class is probably even more relevant for men.
-They probably represent groups in society that are particularly vulnerable.
-
-A further observation can be made from Figure~\ref{fig:cumsw}.
-The gender difference in death hazards is very big for the élite group from 1851--1950, men having about three times higher risks than women in the same class.
-In the other classes, the difference was much smaller.
-It was men in high position that experienced the largest risks.
-
-% \subsubsection{The effect of income and education, 1986--2013}
-% 
-% A common alternative to a class analysis is to use income or education for investigating socio-economic health differences. 
-% Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
-% disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
-% 
-% The analysis for the period 1990--2005 is based on a separate analysis with 
-% income and education as explanatory variables and with controls of period 
-% and marital status.
-% 
-% <<lissyoung>>=
-% library(eha)
-% library(xtable)
-% liss4064 <- loTab[loTab$age < 64.5, ]
-% con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-% ##load("data/lisaold.rda")
-% load(con)
-% close(con)
-% @
-% 
-% Both education and income had a strong, statistically significant, effect on survival, as expected. 
-% 
-% \paragraph{Education, 1990--2013}
-% 
-% <<educ>>=
-% 
-% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
-% educ$period <- factor(educ$period)
-% educ <- educ[!is.na(educ$educ), ]
-% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-% ##educ$educ <- relevel(educ$educ, ref = "middle")
-% educ$civst <- relevel(educ$civst, ref = "married")
-% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-%              data = educ[educ$sex == "male", ])
-% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-%              data = educ[educ$sex == "female", ])
-% cols <- c("red", "darkgreen", "black")
-% @
-% 
-% <<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
-% oldpar <- par(mfrow = c(1, 2))
-% 
-% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
-% 
-% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
-% 
-% par(oldpar)
-% @
-% 
-% Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
-The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
-% 
-% \paragraph{Income, 1986--2005}
-% 
-% <<femeduc>>=
-% income <- lisaold[!is.na(lisaold$income), ]
-% income$period <- factor(income$period)
-% ##income <- income[!is.na(income$income), ]
-% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-% ##income$income <- relevel(income$income, ref = "low")
-% income$civst <- relevel(income$civst, ref = "married")
-% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-%              data = income[income$sex == "male", ])
-% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-%              data = income[income$sex == "female", ])
-% cols <- c("darkgreen", "red", "blue", "black")
-% 
-% @
-% 
-% <<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
-% oldpar <- par(mfrow = c(1, 2))
-% 
-% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
-% 
-% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
-% 
-% par(oldpar)
-% @
-% 
-% Higher incomes result in lower mortality. This is also what could be expected, 
-% and conforms to what we know of social health differences in present-day Sweden. 
-% Notice however that we cannot establish a causal relationship between 
-% income and mortality. Bad health, for instance, may be a confounder.
-% 
-% %%The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
-% %%<<liswom, results = 'asis'>>=
-% %%fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-% %%##cat("Women:\n")
-% %%drw <- drop1(fitw1, test = "Chisq")
-% %%xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
-% %##round(summary(fitw1)$coef[-(1:13), ], 4)
-% %@
-% 
-% %%The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
-% 
-% %%<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
-% %%source("R/plotWEduc.R")
-% %%plotWEduc(fitw1)
-% %%@
-% 
-% 
-% %%And for men:
-% 
-% %%<<lismen, results='asis'>>=
-% %%fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-% %%##cat("\nMen:\n")
-% %%drm <- drop1(fitm1, test = "Chisq")
-% %%xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
-% %%##round(summary(fitm1)$coef[-(1:13), ], 4)
-% %%dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% %%@
-% 
-% %%<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
-% %%#source("R/plotWEduc.R")
-% %%plotWEduc(fitm1)
-% %%@
-% 
-% % 
-% % <<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
-% % #oldpar <- par(mfrow = c(1, 2))
-% % #plot(fitw1, fn = "cum", col = dcol, main = "Women")
-% % #plot(fitm1, fn = "cum", col = dcol, main = "Men")
-% % #par(oldpar)
-% % ##summary(fitw1)
-% % ##summary(fitm1)
-% % @
-% % 
-% % %See Table~\ref{tab:tablisswom1} the PH regression for women.
-% % <<tablisswom1, results = "asis">>=
-% % #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% % #                 data = liss4064[liss4064$sex == "female", ])
-% % #dr <- drop1(fitwo1, test = "Chisq")
-% % #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-% % #    label = "tab:tablisswom1")
-% % @
-% % 
-% % %See Table~\ref{tab:tablissman1} the PH regression for men.
-% % <<tablissman1, results = "asis">>=
-% % #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% % #                 data = liss4064[liss4064$sex == "male", ])
-% % #dr <- drop1(fitma1, test = "Chisq")
-% % #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
-% %  #   label = "tab:tablissman1")
-% % @
-%
-
-\subsubsection{Ages 65--89}
-
-%The distribution over time periods.
-<<getske6089>>=
-vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
-vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
-##vb6589$hisclass <- vb6589$hisclass60
-@
-
-We now turn to the elderly, i.e. those that in most cases have left the workforce.
-Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure.
-
-The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level earlier in history.
-Mortality among women was, as expected, lower than among men.
-We find a clear decline in mortality, but the fastest decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e.
-later than for the age group 40--64 years.
-This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.
-
-Concerning the social pattern, the pattern is less clear compared to what we found for the younger age group (see Figure~\ref{fig:cumswold}).
-For women the results are mixed.
-Once again, too much missing data on social class leaves out the results for the 1960s and 1970s.
-We do not find the same clear advantage for élite women during period until 1950, but neither do we find any large disadvantage for working class elderly women at that time.
-In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century.
-The social differences continued to be small during the last decades of the 20th century.
-Only during the two last periods (after the year 2000) is it possible to establish the expected social gradient in mortality.
-
-For men we find indications of the same intriguing change among elderly as for
-the adult male population 40--64 years. Mortality was in fact substantially
-higher among élite men  but their disadvantage was primarily apparent during the 19th century.
-Differences are fairly small during the first half of the 20th century.
-Not until the second half of the 20th century is the survival much better for the highest social class while workers in the age of retirement had high mortality in comparison.
-).
-This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.
-
-<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
-##source("R/plotED.R")
-resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
-            nper, labb)
-
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Women 65-89", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Men 65-89")##, exclude = 5:6)
-par(oldpar)
-@
-
-See Table~\ref{tab:cumswoldtab}.
-
-<<cumswoldtab,results='asis'>>=
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period, 
-            data = vb6589, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:cumswoldtab", caption = "Anova, all causes of death, 65-89.")
-@
-
-% \subsubsection{The effect of income and education, 1990--2005}
-% 
-% The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting.
-As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
-% 
-% <<lissyoung2>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss6589 <- loTab[loTab$age > 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in
-% Table~\ref{tab:liswom2}
-% <<liswom2, results = 'asis'>>=
-% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-% ##cat("Women:\n")
-% drw2 <- drop1(fitw2, test = "Chisq")
-% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
-% 
-% <<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitw2)
-% @
-% 
-% 
-% And for men:
-% 
-% <<lismen2, results='asis'>>=
-% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm2 <- drop1(fitm2, test = "Chisq")
-% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% <<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm2)
-% @
-% 
-% These figures show clear associations both with income and education and for women as well as for men.
-More years of education resulted in lower mortality.
-For women the association with additional levels became however smaller at the higher levels, the effect is more visible at the steps above the basic levels.
-In a similar fashion, increased incomes is associated with improved survival both for men and for women.
-Apparently, education and income is better to catch up socio-economic differences compared to hisclass, at least for this period.
-
-\subsection{Cardiovascular mortality}
-
-In this section, the focus is on analyses of mortality from the two dominant causes of death among the adult and elderly population.
-We start with cardiovascular diseases.
-Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group.
-Belonging to working class, having less education and low income increases the mortality risks.
-We would therefore expect to find a clear social pattern in the present analysis.
-
-First, a general observation is that the analysis of causes of death and its development is difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ.
-Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers.
-There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory.
-For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases.
-However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent.
-It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and that the increase partly is explained by improvements in how deaths were diagnosed.
-<<causes41>>=
-cvb <- vb0
-cvb$event <- cvb$heart ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
-cvb$period <- factor(cvb$period)
-
-# load("data/vb.rda")
-# ##vb <- cal.window(vb, c(1961, 2001))
-# ##x <- substr(vb$ULORSAK, 1, 2)
-# vb$event <- vb$event & vb$heart
-# vb$exposure <- with(vb, exit - enter)
-# vb$cohort <- factor(vb$cohort)
-# vb$period <- factor(vb$period)
-cvbTab <- aggregate(cvb[, c("event", "exposure")],
-                    cvb[, c("sex", "age", "period", "cohort",
-                           "urban", "civst", "hisclass")],
-                    FUN = sum)
- cvbTab$hisclass <- factor(cvbTab$hisclass)
- cvbTab$age <- factor(cvbTab$age)
-@
-
-<<41colorc>>=
-## Colors for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
-## Common for all groups:
-
-labb <- levels(cvbTab$period)
-nper <- length(labb)
-##source("R/getPHglm.R")
-## End 'common'
-@
-
-As before, models are fitted separately for sex and each ten-year time period, from  1901 to
-2013. We stratify on HISCLASS and use the covariates urban/rural and {\tt   civst}. The cumulative
-hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.
-
-%creating an \emph{heterogeneity index} (a better name is needed:
-%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
-%Something catchy.).
-
-\subsubsection{Ages 40--64}
-
-%The distribution over time periods.
-<<41getske4060>>=
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cardiovascular death."
-scap1w <-  "Cumulative hazards, ages 40--64, women, cardiovascular death."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cardiovascular death."
-scap1m <-  "Cumulative hazards, ages 40--64, men, cardiovascular death."
-##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap12 <- "Total hazard by HISCLASS, ages 40--64, cardiovascular death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cardiovascular death."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cardiovascular death."
-scap2w <-  "Cumulative hazards, ages 65--89, women, cardiovascular death."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cardiovascular death."
-scap2m <-  "Cumulative hazards, ages 65--89, men, cardiovascular death."
-##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap22 <- "Total hazard standardised by HISCLASS, ages 65--89, cardiovascular death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cardiovascular death."
-
-##
-cvb4064 <- cvbTab[as.numeric(as.character(cvbTab$age)) < 64.5, ]
-cvb4064$age <- factor(cvb4064$age)
-@
-
-Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40.
-
-%\subsubsection{Women}
-
-<<41hazcalw>>=
-source("R/getPHglm2.R")
-resw <- getPHglm2(cvb4064[cvb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(cvb4064[cvb4064$sex == "male", ],
-            nper, labb)
-@
-
-% We will use as a summary measure the value of the cumulative hazards at
-% age 65, given survival to age 40.
-
-<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
-                    ylab = "TMR(65|40)",  main = "Women 40-64")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)", main = "Men 40-64")
-par(oldpar)
-@
-
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
-%%@
-<<41cumswtab,results='asis'>>=
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass + 
-                sex * period + hisclass * period, 
-            data = cvb4064, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64.")
-@
-
-See Table~\ref{tab:41cumswtab}.
-
-Figure~\ref{fig:41cumsw} shows mortality levels during different periods from 1901 on-wards.
-Women from the highest social class had consistently lower mortality from cardiovascular diseases compared to the others.
-We also observe increasing social differences from the first half of the century to the last decades.
-A new pattern has developed from 1980 onwards.
-Now working class women have substantially higher cardiovascular mortality compared to others.
-We can however not exclude that this is an artefact of different diagnostics for deaths in different groups.
-Working class women may have had less precise diagnoses in the early 20th century.
-
-For men, there is a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality.
-The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
-Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the middle category consisting of lower manager/farmers.
-And in the same way as for women at this age, working class men have the highest mortality from this cause of death during the last decades.
-A social hierarchy in cardiovascular mortality seems to have developed both for men and women.
-
-% A word of caution is however needed when it comes to these results.
-We can not exclude that the smaller differences in cardiovascular diseases for women in the early periods are an artefact of how deaths were diagnosed.
-Working class men and women may have had less precise diagnoses in the early 20th century.
-
-%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-% \subsubsection{The effect of income and education, 1990--2005}
-% 
-% We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death.
-It is based on a separate analysis with income and education as explanatory variables.
-% 
-% <<41lissyoung>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% loTab$period <- factor(loTab$period)
-% liss4064 <- loTab[loTab$age < 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in
-% Table~\ref{tab:41liswom}
-% <<41liswom, results = 'asis'>>=
-% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-% ##cat("Women:\n")
-% drw <- drop1(fitw1, test = "Chisq")
-% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
-% 
-% <<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% source("R/plotWEduc.R")
-% plotWEduc(fitw1)
-% @
-% 
-% 
-% And for men:
-% 
-% <<41lismen, results='asis'>>=
-% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm <- drop1(fitm1, test = "Chisq")
-% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% <<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm1)
-% @
-% 
-% 
-% <<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
-% #oldpar <- par(mfrow = c(1, 2))
-% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
-% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
-% #par(oldpar)
-% ##summary(fitw1)
-% ##summary(fitm1)
-% @
-% 
-% %See Table~\ref{tab:tablisswom1} the PH regression for women.
-% <<41tablisswom1, results = "asis">>=
-% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "female", ])
-% #dr <- drop1(fitwo1, test = "Chisq")
-% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-% #    label = "tab:tablisswom1")
-% @
-% 
-% %See Table~\ref{tab:tablissman1} the PH regression for men.
-% <<41tablissman1, results = "asis">>=
-% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "male", ])
-% #dr <- drop1(fitma1, test = "Chisq")
-% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
-%  #   label = "tab:tablissman1")
-% @
-% 
-% The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
-More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system.
-This is also well in line with our analysis based on HISCLASS above.
-
-\subsubsection{Ages 65--89}
-
-%The distribution over time periods.
-<<41getske6089>>=
-cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
-cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
-@
-
-In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years. We use as a summary measure the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65.
-
-%\subsubsection{Women}
-
-<<41hazcalwold>>=
-resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
-              nper, labb)
-resm <- getPHglm2(cvb6589[cvb6589$sex == "male", ],
-                nper, labb)
-@
-
-Concerning the social pattern, we find somewhat different results from that of the younger age group of women.
-Cardiovascular mortality among elderly women do not present a consistent class pattern.
-Female cardiovascular mortality was somewhat higher among the elite and the middle group than among working class women until the last decades.
-Elderly from a working class background were not disadvantaged for most of the 20th century.
-The social class pattern has changed somewhat during the last periods and more in line to what is expected.
-The differences are however rather small.
-
-Even if results for elderly men point in different directions for some periods, the general trend resembles that of working-active men.
-The highest class had high cardiovascular mortality during the first half of the 20th century, while the expected social pattern has developed only during the most recent periods.
-
-<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
-##source("R/plotED.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women 65-89", exclude = 3:4)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)", main = "Men 65-89")
-par(oldpar)
-@
-
-<<41cumswoldtab,results='asis'>>=
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * period *hisclass,
-                ##sex * period + hisclass * period, 
-            data = cvb6589, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
-@
-
-See Table~\ref{tab:41cumswoldtab}.
-
-% \subsubsection{The effect of income and education on cardiovascular death, 1990--2005}
-% 
-% The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
-% income and education as explanatory variables.
-The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2} for women and Table~\ref{tab:41lismen2} for men.
-It confirms that most of the variables are strongly significant except
-% 
-% <<41lissyoung2>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss6589 <- loTab[loTab$age > 64.5, ]
-% @
-% 
-% 
-% 
-% <<41liswom2, results = 'asis'>>=
-% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-% ##cat("Women:\n")
-% drw2 <- drop1(fitw2, test = "Chisq")
-% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% 
-% <<41lismen2, results='asis'>>=
-% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm2 <- drop1(fitm2, test = "Chisq")
-% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-% 
-% <<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitw2)
-% @
-% 
-% 
-% And for men:
-% 
-% 
-% 
-% <<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm2)
-% @
-% 
-% The pattern is similar to what we find for all-cause mortality.
-The higher educational level, the lower cardiovascular mortality, and more income leads to better survival.
-This association is clear and consistent for women as well as for men.
-A minor difference between men and women is that mortality in the higher levels of education do not lead to lower mortality among women while that is the case among men.
-
-\subsection{Cancer mortality}
-
-While we assume clear social differences in cardiovascular diseases, this is not necessarily an obvious assumption when it comes to cancers.
-Different forms of cancers have very different origin and different pathways to disease and death.
-The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers.
-In our analyses, there is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated by the already mentioned large proportion of unknown or vague causes at that time.
-We believe however that cancers were better diagnosed than cardiovascular deaths.
-This is supported by the stability in proportion of cancer deaths between the two first periods, 12,5 per cent 1901--1925 and 13,5 per cent for the period 1926--1950.
-
-<<causes12>>=
-cvb <- vb0
-cvb$event <- cvb$cancer ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
-cvb$period <- factor(cvb$period)
-
-# load("data/vb.rda")
-# ##vb <- cal.window(vb, c(1961, 2001))
-# ##x <- substr(vb$ULORSAK, 1, 2)
-# vb$event <- vb$event & vb$heart
-# vb$exposure <- with(vb, exit - enter)
-# vb$cohort <- factor(vb$cohort)
-# vb$period <- factor(vb$period)
-cvbTab <- aggregate(cvb[, c("event", "exposure")],
-                    cvb[, c("sex", "age", "period", "cohort",
-                           "urban", "civst", "hisclass")],
-                    FUN = sum)
- cvbTab$hisclass <- factor(cvbTab$hisclass)
- cvbTab$age <- factor(cvbTab$age)
-@
-
-<<12colorc>>=
-## Colours for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
-## Common for all groups:
-
-labb <- levels(cvbTab$period)
-nper <- length(labb)
-source("R/getPHglm.R")
-## End 'common'
-@
-
-We use similar types of analyses as in the earlier sections. The cumulative
-hazards at the end of the period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis. These numbers are used in the process of
-graphically illustrating the development over time.
-
-%creating an \emph{heterogeneity index} (a better name is needed:
-%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
-%Something catchy.).
-
-\subsubsection{Ages 40--64}
-
-%The distribution over time periods.
-<<12getske4060>>=
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cancer death."
-scap1w <-  "Cumulative hazards, ages 40--64, women, cancer death."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cancer death."
-scap1m <-  "Cumulative hazards, ages 40--64, men, cancer death."
-##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cancer death."
-scap12 <- "Total hazard by HISCLASS, ages 40--64, cancer death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cancer death."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cancer death."
-scap2w <-  "Cumulative hazards, ages 65--89, women, cancer death."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cancer death."
-scap2m <-  "Cumulative hazards, ages 65--89, men, cancer death."
-##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cancer death."
-scap22 <- "Total hazard standardised by HISCLASS, ages 65--89, cancer death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cancer death."
-
-##
-cvb4064 <- cvbTab[as.numeric(as.character(cvbTab$age)) < 64.5, ]
-cvb4064$age <- factor(cvb4064$age)
-@
-
-<<12hazcalw>>=
-source("R/getPHglm2.R")
-resw <- getPHglm2(cvb4064[cvb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(cvb4064[cvb4064$sex == "male", ],
-            nper, labb)
-@
-
-%%We will use as a summary measure the value of the cumulative hazards at
-%%age 65, given survival to age 40.
-
-<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
-                    ylab = "TMR(65|40)",  main = "Women 40-64")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)", main = "Men 40-64")
-par(oldpar)
-@
-
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
-%%@
-
-<<12cumswtab,results='asis'>>=
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex *hisclass +
-                sex * period + hisclass * period, 
-            data = cvb4064, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
-@
-
-See Table~\ref{tab:12cumswtab}.
-
-Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods for the age group 40--64.
-Cancer mortality among the élite women was lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods.
-Consequently, we find a more consistent advantage of higher class in cancers compared to cardiovascular diseases for women.
-We do however not find the same pattern for men.
-It is difficult to identify any general features of 20th century male cancer mortality when it comes to social differences.
-
-%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-% \subsubsection{The effect of income and education on cancer mortality, 1990--2005}
-% 
-% In the previous section, we concluded that there were no social inequality in cancer mortality according to hisclass.
-The question here is if the lacking socio-economic differentiation holds also when using education and income as categorising variables.
-The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables.
-% 
-% <<12lissyoung>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss4064 <- loTab[loTab$age < 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom} for women and Table~\ref{tab:12lismen} for men. The variables of interest here are all significant.
-% <<12liswom, results = 'asis'>>=
-% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-% ##cat("Women:\n")
-% drw <- drop1(fitw1, test = "Chisq")
-% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cancer death.", label = "tab:12liswom")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% <<12lismen, results='asis'>>=
-% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm <- drop1(fitm1, test = "Chisq")
-% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.", label = "tab:12lismen")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}.
-% 
-% <<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-% source("R/plotWEduc.R")
-% plotWEduc(fitw1)
-% @
-% 
-% 
-% And for men:
-% 
-% 
-% <<12lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm1)
-% @
-% 
-% 
-% <<12plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cancer death.", fig.height = 4>>=
-% #oldpar <- par(mfrow = c(1, 2))
-% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
-% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
-% #par(oldpar)
-% ##summary(fitw1)
-% ##summary(fitm1)
-% @
-% 
-% %See Table~\ref{tab:tablisswom1} the PH regression for women.
-% <<12tablisswom1, results = "asis">>=
-% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "female", ])
-% #dr <- drop1(fitwo1, test = "Chisq")
-% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-% #    label = "tab:tablisswom1")
-% @
-% 
-% %See Table~\ref{tab:tablissman1} the PH regression for men.
-% <<12tablissman1, results = "asis">>=
-% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "male", ])
-% #dr <- drop1(fitma1, test = "Chisq")
-% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
-%  #   label = "tab:tablissman1")
-% @
-% 
-% When it comes to the association with education and income, we find the similar pattern as those for all-cause mortality and cardiovascular mortality.
-There is a strong effect for both men and women of having more education and income.
-These variables seem to catch up important aspect of socio-economic differences in health and mortality.
-
-\subsubsection{Ages 65--89}
-
-%The distribution over time periods.
-<<12getske6089>>=
-cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
-cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
-@
-
-In our analyses of class patterns in mortality, we finally turn to cancer mortality among the elderly.
-
-%\subsubsection{Women}
-
-<<12hazcalwold>>=
-resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
-              nper, labb)
-resm <- getPHglm2(cvb6589[cvb6589$sex == "male", ],
-                nper, labb)
-@
-
-The relation between different classes and cancer mortality fluctuates considerably both for women and men in this age group.
-For some periods it has to do with quite few events to analyse.
-Women of the highest class have favourable mortality during several periods, consistent with the advantages we found for women in the younger age group, but on the other hand had high levels at other times.
-The social differences are however quite small during the most recent periods.
-
-For men we find a similar lack of social pattern in cancer mortality as we found for men in working-active ages.
-There is however a tendency during the last decades of higher mortality among working class men.
-
-<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
-##source("R/plotED.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)",  main = "Women 65-89", exclude = 3:4)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Men 65-89")
-par(oldpar)
-@
-
-<<12cumswoldtab,results='asis'>>=
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex +
-                sex * period + hisclass * period, 
-            data = cvb6589, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
-@
-
-See Table~\ref{tab:12cumswoldtab}.
-% \subsubsection{The effect of income and education on cancer death, 1990--2005}
-% 
-% Thus, we find indications of social differences in the later decades, particularly between the elite and the workers.
-The alternative analysis for the last period is based on a separate analysis with
-% income and education as explanatory variables.
-% 
-% <<12lissyoung2>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss6589 <- loTab[loTab$age > 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men.
-% <<12liswom2, results = 'asis'>>=
-% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-% ##cat("Women:\n")
-% drw2 <- drop1(fitw2, test = "Chisq")
-% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% <<12lismen2, results='asis'>>=
-% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm2 <- drop1(fitm2, test = "Chisq")
-% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
-% 
-% <<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitw2)
-% @
-% 
-% 
-% And for men:
-% 
-% %% <<12lismen2, results='asis'>>=
-% %% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
-% %% ##cat("\nMen:\n")
-% %% drm2 <- drop1(fitm2, test = "Chisq")
-% %% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
-% %% round(summary(fitm1)$coef[-(1:13), ], 4)
-% %% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% %% @
-% 
-% <<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm2)
-% @
-% 
-% The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income.
-We find the same consistent pattern as we have found for all-cause mortality and cardiovascular mortality for this period.
-Higher levels of education and more income is associated with lower cancer mortality.
-We can therefore conclude that both education and income are strongly decisive aspects of social differentiation when it comes to mortality during the last decades.
-
-%%{\huge Här ska 'conclusion.tex' in !!!}
-
-\section{The effect of Education, 1990--2013}
-
-%\subsection{All causes of death}
-
-A common alternative to a class analysis is to use income or education for investigating socio-economic health differences.
-\citet{goldk10} criticize how many economists and epidemiologists operationalize inequality in their studies, for example by not distinguishing between class and status aspects.
-Class refers to relational aspects in society, while many other variables focus on the distribution of material and/or immaterial resources.
-He advocates studies of different aspects, which together complement each other.
-Comparing different approaches are valuable in order to understand why some are advantaged and others disadvantaged and what sort of factors determines health differences.
-In this and the following section, we introduce analyses of education and income.
-In that way, we can compare the class analyses with these alternative categorizations.
-For the time being, we have to restrict such a comparative analysis to the last decades.
-
-In this section we analyse differences according to education level for the period 1990--2013 for the whole population 40 years and older.
-Education is a good indicator of both human and cultural capital, but aer also strongly associated with income and social class.
-We present analyses on all cause mortality (survival curves) as well as on the cause of death groups of cardiovascular diseases and cancers (cumulative hazards).
-The pattern is clear in total mortality as well as cause specific mortality, and this is valid for both men and women, see Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart} and~\ref{fig:ploteduccancer}.
-Education had a strong, statistically significant, effect on survival, as expected.
-Those with the lowest education levels hade the highest mortality, while those with highest level of education had by far the best survival.
-The differences must be considered large, and it follows a continuous improvement according to increasing levels.
-
-<<lissyoung3>>=
-library(eha)
-library(xtable)
-liss4064 <- loTab[loTab$age < 64.5, ]
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-##load("data/lisaold.rda")
-load(con)
-close(con)
-@
-
-<<educ>>=
-
-educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
-educ$period <- factor(educ$period)
-educ <- educ[!is.na(educ$educ), ]
-levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-##educ$educ <- relevel(educ$educ, ref = "middle")
-educ$civst <- relevel(educ$civst, ref = "married")
-fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-cols <- c("red", "darkgreen", "black")
-@
-
-<<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lwd = 1.5, lty = c(1, 2, 5), axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-
-plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-
-par(oldpar)
-@
-
-<<tableeduc,results='asis'>>=
-fit <- coxreg(Surv(enter, exit, event) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-ltx(fit, dr = dr, caption = "Effect of education on mortality in ages 40-90, 1990-2013.", 
-    label = "tab:tableeduc", digits = 4)
-@
-
-See Table~\ref{tab:tableeduc}.
-
-%\subsection{Cardiovascular death}
-
-<<cardeduc>>=
-ulcard <- unique(vb0$ULORSAK[vb0$heart])
-educ$heart <- FALSE
-educ$heart <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
-fitm <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-@
-
-% Illustrating differences for cause-specific mortality is conveniently done by comparing 
-% cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
-% cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
-% mortality.
-
-<<ploteducheart, fig.cap="Cumulative hazards by education and sex, cardiovascular death, 1990--2013.">>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1:2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1:2, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-@
-
-<<tableeducheart,results='asis'>>=
-fit <- coxreg(Surv(enter, exit, heart) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-ltx(fit, dr = dr, caption = "Effect of education on cardiovascular mortality in ages 40-90, 1990-2013.", 
-    label = "tab:tableeducheart", digits = 4)
-@
-
-See Table~\ref{tab:tableeduc}.
-
-%\subsection{Cancer death}
-
-<<canceduc>>=
-ulcard <- unique(vb0$ULORSAK[vb0$cancer])
-educ$cancer <- FALSE
-educ$cancer <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
-fitm <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-@
-
-% Although mortality in cancer is lower than mortality in cardiovascular diseases,
-% the pattern is the same: The more education, the less mortality. See 
-% Figure~\ref{fig:ploteduccancer}.
-
-<<ploteduccancer, fig.cap="Cumulative hazards by education and sex, cancer death, 1990--2013.">>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-@
-
-<<tableeduccancer,results='asis'>>=
-fit <- coxreg(Surv(enter, exit, cancer) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ], 
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-ltx(fit, dr = dr, caption = "Effect of education on cancer mortality in ages 40-89, 1990-2013.", 
-    label = "tab:tableeduccancer", digits = 4)
-@
-
-See Table~\ref{tab:tableeduc}.
-
-\section{The effect of Income, 1986--2005}
-
-%\subsection{All causes of death}
-
-<<femeduc>>=
-income <- lisaold[!is.na(lisaold$income), ]
-income$period <- factor(income$period)
-##income <- income[!is.na(income$income), ]
-##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-##income$income <- relevel(income$income, ref = "low")
-income$civst <- relevel(income$civst, ref = "married")
-fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-cols <- c("darkgreen", "red", "blue", "black")
-
-@
-
-<<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-
-plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-
-par(oldpar)
-@
-
-<<tableinc>>=
-fit <- coxreg(Surv(enter, exit, event) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-@
-
-<<tableincpr,results='asis'>>=
-ltx(fit, dr = dr, caption = "Effect of income on mortality in ages 40-89, 1986-2005.", 
-    label = "tab:tableincpr", digits = 4)
-@
-
-See Table~\ref{tab:tableincpr}.
-
-Finally, we present analyses of different income levels for the age group 40 and above and the period 1986--2005.
-Disposable individual income must be considered as a robust measure of economic condition, and increases the possibility to allocate these economic resources towards health promoting living conditions and behaviours.
-We categorize the income according to quartiles.
-As expected, we find that higher incomes result in lower mortality, both when it comes to total mortality as to deaths from cardiovascular diseases, see Figures~\ref{fig:plotincome} and~\ref{fig:plotincheart}.
-This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden.
-We also find indications of a similar difference for cancers but less clear, particularly for male cancer mortality, see Figure~\ref{fig:plotinccancer}.
-Notice however that we cannot establish a causal relationship between income and mortality.
-Bad health may be a confounder, causing both low incomes and higher death risks.
-
-%\subsection{Cardiovascular death}
-
-<<heartincome>>=
-ulcard <- unique(vb0$ULORSAK[vb0$heart])
-income$heart <- FALSE
-income$heart <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
-fitm <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-
-@
-
-%The overall pattern is the same for cardiovascular mortality, high income pays off.
-<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005.">>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-@
-
-<<tableincheart>>=
-fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-@
-
-<<tableincprheart,results='asis'>>=
-ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.", 
-    label = "tab:tableincpr", digits = 4)
-@
-
-See Table~\ref{tab:tableincpr}.
-
-%\subsection{Cancer death}
-
-<<cancerincome>>=
-ulcard <- unique(vb0$ULORSAK[vb0$cancer])
-income$cancer <- FALSE
-income$cancer <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
-fitm <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-
-@
-
-% The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.
-
-<<plotinccancer, fig.cap="Cumulative hazards by income and sex, cancer death, 1990--2013.">>=
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-@
-
-<<tableinccancer>>=
-fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-@
-
-<<tableincprcancer,results='asis'>>=
-ltx(fit, dr = dr, caption = "Effect of income on cancer mortality in ages 40-89, 1986-2005.", 
-    label = "tab:tableincprcancer", digits = 4)
-@
-
-See Table~\ref{tab:tableincpr}.
-
-\section{Discussion}
-
-The authors of this paper have scrutinised different aspects of the social
-pattern of mortality in historical northern Sweden for some time, focusing on
-different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}.
-When it comes to infant and child mortality pre-- and early--transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}.
-Studies based on microdata on the adult population in history have been rare, mainly due to lack of adequate data.
-The results from such studies show other patterns than that of infant and child mortality.
-The expected social gradient in
-survival, meaning that people from higher social classes always had an
-advantage, was in many cases, particularly among adult men, not found
-despite the well-documented socio-economic differences \citep{edli11, edbr12}.
-There were however large differences between men and women and for different age groups.
-These earlier studies have suggested that distinct and consistent differences to the advantage of higher classes developed somewhere during and probably quite early in the 20th century \citep{bevp11}.
-Until now we have however not been able to investigate the continuous development on mortality differences among the adult and elderly
-population due to no available data.
-This is the first time it has been possible, and the results continue to surprise us.
-The consistent advantage for higher social classes do not appear until quite late in the regions we have analysed.
-The social pattern of female mortality are in line to what is
-expected, but a social gradient in male adult mortality seems to be a recent phenomenon.
-Furthermore, this transition and the class pattern of mortality differ depending on age group.
-Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly.
-
-The results of small socio-economic gradient in the past and its late development are however in accordance with other studies \citep{bevp11}, for example by \citet{bedr11} in a very different setting, a couple of Scanian parishes, although with somewhat different social categories.
-They found the first indications of an emerging social gradient in the 1950s.
-Similar results have also been found in Estonia \citep{Jaadla17}.
-
-When it comes to specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
-Cancer is sometimes considered less connected to social class, and we find mixed results in our analyses.
-
-The surprising results makes it reasonable to consider the
-validity and possible weaknesses in the study.
-One obstacle concerns the quality of available information.
-Occupational titles in historical sources are not always distinct enough to define the correct class.
-It can be difficult to identify the skill level in the working class population because many
-were often only designated as labourer. The size of the enterprise for free
-entrepreneurs is unknown and it may be impossible to distinguish if they
-are self-employed or have employees. The class system and the position
-of different classes can also be discussed. Studies covering long period
-have the problem that occupations and classes change character - they do
-not represent the same thing. Nevertheless, these ambiguities and possible
-mis-classifications are not of that sort and magnitude that it would radically change
-the basic patterns observed. The social classes identified during the
-different periods represent distinct groups with very large differences in
-resources, capacities and power. Available data, for example from the 1930 census confirm that class differences in income and wealth were large, often enormous.
-
-In this paper we have analysed social inequality in health from the
-perspective of social classes using HISCLASS. However, there are
-alternative ways to analyse social inequality, and we cannot exclude that
-such analyses gives other results or at least would modify the image
-presented here. In many modern studies categorisations based on income, wealth
-or education are common for studies of inequality. They represent different dimensions of 
-inequality. One problem is however that reliable data on income and
-education are rarely available in historical population databases. 
-As a test, we have included analyses of the associations of
-education and income to mortality for recent decades. For women as
-well as for men, for both age groups and for all-cause mortality as well as
-cause-specific mortality we find consistent differences with higher
-educational level and more income leading to lower mortality, thus basically confirming what we find in the class analysis. We do not know if education and income are equally important in the past.
-
-The WHO Commission on Social Determinants of Health framework for the social determinants of health provides a model on how socio-economic position influence health.
-Social class works through intermediate determinants related to exposure and vulnerability \citep{who10}.
-People are situated within socio-economic and political contexts, as well as prevailing value systems in society.
-Socio-economic position imply different access to resources, power and prestige, all being aspects of crucial importance.
-The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
-The importance of understanding people in their context is also emphasised by for example \citet{kri05}.
-What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-The WHO framework
-\citep{who10} lists three different types of intermediary social determinants of health; material circumstances, behaviours, biological factors, and finally psycho-social factors.
-Most studies focus on material circumstances.
-Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions.
-Even though these advantages were at hand in our studied regions, the final result was not better health for men in the most privileged groups until the late 20th century.
-We must therefore consider other explanations.
-Psycho-social factors have been frequently discussed during the most recent decades, for example by \citet{mar04}, who argues that a higher status leads to better health, regardless of where in the status ladder we make our observations.
-Psychosocial factors are also suggested by \citet{wipi09} as an explanation why equal societies always perform better than unequal ones when it comes to health and other social conditions.
-In the case studied here, we do not find any indication of a status syndrome, since mortality levels was in the opposite directions to the one we would expect from these theories.
-
-This makes us consider the possible impact of behavioural factors.
-Our results of the gender aspect might be of interest in this case.
-When it comes to gender differences, the present study confirms what we previously have found in the Sundsvall region \citep{edli11}.
-There is a clear gendered pattern when it comes to social inequality in mortality.
-Within the same households, the effect of social class seems to lead to different results, high social position being beneficial for women but not for men.
-We might consider the role of behaviours based on different gendered expectations.
-Having a high social status could be expressed in attitudes and behaviours that were harmful for health (this is discussed in more detail in \citet{edli11}.
-Alcohol has been suggested as part of the large gender differences in the 19th century \citep{wil99, edv92}, and a more risk-taking life may be a component of male mortality \citep{cour00}.
-There are however some problems with attributing most of the male excess mortality to heavy drinking.
-Alcohol consumption decreased substantially during the period late 19th century to the 1950s, but the old pattern in social health inequalities remained.
-Swedes at that time had very low consumption levels.
-Alternative explanations should also be considered.
-There
-were strong and heavy requirements of being successful for the male \"elite. Many were involved in businesses with high risks and they may have felt a constant threat of economic failure.
-
-Our study illustrates the importance of considering contextual factors and
-the restrictions of making general statements. We need to look into all
-aspects of social classes, not only the material ones but also what this
-means for behaviour, norms, way of living. \citet{edli11} suggest that an
-explanation for the larger social health inequalities in present-day
-society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive
-eating, drinking, smoking and sometimes also taking risks.
-Different historical contexts promote different ways to spend economic resources.This is also in line with what \citet{rasp06} and \citet{Jaadla17} suggest.
-If this interpretation is correct, the hypothesis of \citet{ant67}on an increasing role of health related behaviour in the modern low-mortality régime,while economic inequality determined the social mortality pattern during the early phases of the mortality transition.
-Instead, health-related seems to be at least equally strong earlier in history but having a completely different expression.
-
-One of the major findings in this study relate to gender.
-As several researchers have pointed out, demography has often ignored power relations and different roles within families \citep{watkins93, mack95, jans07}.
-By investigating possible differences in mortality patterns according to sex, we have identified gendered patterns within the family.
-Different patterns between the sexes may be caused by discrimination on the female part in the household.
-We know that the society was
-strongly segregated according to gender and that women were discriminated
-and had less space for action.
-In our case, we do not find this to have negative consequences for women, indicating that the relation between sexes in the society was characterised by distinct gendered expectations on behaviour.
-
-Another question is if our results can be generalised to other historical contexts. We have studied
-environments in northern Sweden without any large metropolis. Perhaps
-Sweden and/or this part of Sweden is a special case. The social differences
-may have been smaller, the involvement of government both at the national
-and local level may have been stronger, thus mitigating socio-economic
-disadvantages. In comparison with many other countries, the level of mortality has been fairly low. These are relevant questions, which hopefully can be
-addressed in other and maybe comparative studies.
-
-Finally, we argue that our results have implications for the understanding
-on several other hot research topics. A central and hot debated question concerning the mortality
-transition is how the enormous increase in survival was accomplished. \citet{mck76} argues that the
-improvements up to the middle of the 20th century was primarily caused by improved nutrition. This implicate that having more resources would lead to better survival. If research shows that the most
-advantaged groups did not have better health and longer lives, this do not
-necessarily speak against McKeown's ideas but it makes the question more
-complex. The results are also difficult to fit into \citet{liph95}
-hypothesis that social conditions should be seen as a fundamental cause of
-disease and not only expressions of different disadvantages. In the historical
-context analysed here, this idea is not supported.
-
-In recent years, the question about the relation between health and income
-inequality been widely discussed \citep{wipi09}. Even though our study do not address this
-issue directly, our results are of interest for this question as well. The
-development of the social inequality in mortality do not seem to fit that
-well into the development of income and wealth inequalities. The early 20th
-century, in Sweden as well as in other countries, was characterised by very
-large economic differences, still this did in our regions not lead to any
-consistent advantage for the highest social classes. After the Second World
-War, economic differences diminished, and was at its lowest in Sweden at
-about 1980, i.e. at a time when a more expected pattern of a social
-hierarchy in survival had appeared. The period thereafter is characterised
-by increasing economic differences, and this may coincide with increasing
-health differences as well. The long-term relation between economic
-inequality and health is worth further studies.
-
-% \section{Conclusion}
-% 
-% We started this paper by raising the issue of the development of social
-% inequalities. When studying this question in history we often get surprised
-% of its complexity. In a society characterised by large economic inequality,
-% we would expect that this would result in corresponding mortality
-% differences, i.e. higher social classes having an advantage. We find the
-% appearance of a consistent social pattern in mortality, to the advantage of
-% the higher classes, but surprisingly late in our history. It is only in the
-% recent decades that this pattern dominates. Improved survival characterised
-% however all HISCLASSes. We can however confirm that there are gender
-% differences in the social pattern. While there is no clear social gradient
-% among men, the results at least partly indicate advantages for the higher
-% social classes for women already early in the 20th century. Females had
-% more to gain from a high social position. Previous studies of the mortality
-% transition have made similar observations. Among adults, gender has often
-% been of more significant importance than social class. The reason for this
-% difference and for the late appearance of the modern social pattern among
-% men is still an enigma that require further studies, even though we have
-% suggested the possible importance of behavioural factors and changes in
-% ideals connected to different social classes. This should stimulate us to
-% further disentangle the pathways for how social position influence health
-% and mortality in different societies.
-Furthermore, we do not find any substantial differences or distinct pattern in the relation between class and the two major cause-of-death groups.
-Finally, we find stronger differences in the social pattern among the working-age population, while the associations were smaller or not so clear among the elderly.
-However, when analysing the association with education and income in the period 1990--2005 we find consistent patterns of higher income and higher education level being associated with lower mortality, indicating that these variables are better to catch socio-economic differences at least in the present-day society.
-
-\section{Acknowledgements}
-
-This work was undertaken within the Centre for Demographic and Ageing
-Research (CEDAR) at Umeå University and with support from the Swedish
-Research Council (VR) (grant no.
-421-2012-892) and the project Old age, health and inequality: mortality from a life course perspective in Sweden 1986-2009, financed by the Swedish Research Council (grant no 421-2012-892.
-We thank Demographic Data
-Base for giving us access to data and Jonas Helgertz, Department of
-Economic History, Lund University, for providing us with scripts for
-creating HISCLASS from Swedish Census data.
-
-\bibliography{mybib}
-
-\end{document}
-@
+One of the central aspects of survival is social class and access to economic 
+and other resources. Even in present-day welfare societies, social position is a 
+strong determinant when it comes to health and mortality and the impact even 
+seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that 
+"... social conditions have been, are and will continue to be irreducible 
+determinants of health outcomes and therefore deserves their appellation as 
+'fundamental causes' of disease and death" \citep{liph95}. For a long time, 
+the general view has been that socio-economic health inequalities were high in 
+historical societies, probably higher than in modern societies. This is a 
+reasonable assumption since these societies were in most cases characterised by 
+very large differences between groups and classes. Knowing that access to 
+resources provides advantages in all ways, the health advantage of higher 
+classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social 
+inequality in health has passed through different historical phases. According 
+to them, social health differences were comparatively small during the 
+pre-transitional phase when space was a strong determinant for the spread of 
+disease. Differences then increased during the transitional phase when mortality 
+declined and wealthy groups used their resources to gain better health. Finally, 
+health differences decreased again in modern low-mortality settings when instead 
+health-related behaviour became important, resulting in marginal differences. 
+\citet{omran82} state, in the third proposition in his theory of the 
+epidemiological transition, that even if the class differential in mortality was 
+maintained during the transition, the decline set in earlier and was faster 
+among privileged groups.

### 2017-07-02 article.Rmd
convert halfway from Rnw to Rmd

--- 2017-06-02 icass.Rnw
@@
-\documentclass[a4paper,11pt,titlepage]{article}
-
-\usepackage[utf8]{inputenc}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\usepackage{float}
-
-\usepackage{epstopdf}
-\usepackage{url}
-
-\usepackage{natbib}
-
-\bibliographystyle{apalike}
-%\bibliographystyle{plainnat}
-
-%\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\newcommand{\R}{{\bf R}}
-
-\usepackage[tablesonly]{endfloat}
-%\usepackage[figuresonly]{endfloat}
-%\usepackage[tablesonly]{endfloat}
-
-\renewcommand{\efloatseparator}{\vfill}
-
-% patch \efloat@iwrite to use \protected@write
-\makeatletter
-\renewcommand\efloat@iwrite[1]{%
-   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-
-\author{Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} and
-Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}}
-     \\
-Centre for Demographic and Ageing Research ({\sc Cedar})\\
-Umeå University, SE-90187 Umeå, Sweden}
-
-\title{Long-term perspectives of social inequality in mortality among
-  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
-    at the
-    session 'Health and Mortality in Arctic Populations'. The Ninth International
-    Congress on Arctic Social Sciences,
-    Umeå,
-    June 8--12, 2017}}
-
-\date{May 31, 2017}
-
-<<setup, include = FALSE>>=
-library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
-library(knitr)
-library(eha)
-library(xtable)
-library(skum)
-@
-
-<<readalldata>>=
-## Load remote data:
-con <- url("http://capa.ddb.umu.se/data/vb0.rda")
-##load("data/vb0.rda")
-load(con)
-close(con)
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
-##
-con <- url("http://capa.ddb.umu.se/data/vb.rda")
-##load("data/vb.rda")
-load(con)
-close(con)
-##
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-##load("data/lisaold.rda")
-load(con)
-close(con)
-##
-con <- url("http://capa.ddb.umu.se/data/loTab.rda")
-load(con)
-close(con)
-loTab$period <- factor(loTab$period)
-##
-vb0Tab <- aggregate(vb0[, c("event", "exposure")],
-                   vb0[, c("sex", "age", "period", "cohort",
-                          "urban", "civst", "hisclass")],
-                   FUN = sum)
-vb0Tab$age <- factor(vb0Tab$age)
-@
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-A long-term perspective on the development of social
-inequalities in mortality in the adult and elderly population is taken in
-this paper. The area under investigation consists of the Skellefteå and Umeå regions in
-the north of Sweden, and the time period is 1851--2013. The outcome is all
-cause mortality and cause specific mortality for the categories
-cardiovascular diseases and cancers. The analysis of inequality is based on
-three social classes, reorganized from {\sc Hisclass}.
-
-The main findings are that a consistent social gradient implying better
-survival for classes with more access to economic and other resources, only
-become evident in the later half of the 20th century, with lower classes
-having the highest mortality. The class patterns in mortality differed
-however depending on gender. Women of higher classes were comparatively
-favoured in most of the studied periods, while higher class men were
-disadvantaged when it comes to survival at least until the middle of the
-20th century. The modern pattern of clear socio-economic differences
-appeared surprisingly late. Additional and alternative analyses based on education and
-income categories confirmed large inequality in mortality
-between different groups during the most recent decades.
-There were no clear signs of qualitative differences depending on whether it was the working-active or the retired population that was analysed, but the differences according to social class was larger in the first-mentioned group.
-Both cardiovascular diseases and cancers conform with the pattern already observed when it comes to all-cause mortality.
-
-\end{abstract}
-
-\tableofcontents
-
-\newpage
-
-%%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}
-
-%%\subsection{Combining the data sets and tabulate}
-
-\section{Introduction}
-
-One of the central aspects of survival is social class and access to economic and other resources.
-Even in present-day welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}.
-It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}.
-For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies.
-This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes.
-Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious.
-\citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases.
-According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease.
-Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health.
-Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences.
-\citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.
-
-Recent studies, investigating social inequality in health and mortality with micro-data, has however questioned how generalizable the assumed pattern is \citep{bevp11}.
-Instead, some studies suggest that the modern pattern might be a fairly recent development, although the pattern differs depending on age, gender or historical context.
-When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small or even reversed socio-economic differences during the first stages of the transition \citep{bedr11, edli11,edbr12}, sharply contrasting against what we know from present-day society.
-
-Another question of the development of social inequality in mortality, concerns its relation to economic inequality.
-\citet{wipi09} argue that income inequality has an independent effect on mortality, separate from the direct effect of actual different access to economic resources.
-They argue that unequal societies perform less well when it comes to health than equal ones.
-Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world.
-This has initiated a vital scholarly debate and the topic has been extensively studied \citep{suka04, wado00}.
-How the association looked like in previous periods is unknown.
-This is of particular interest since the levels of economic inequality were very different from those of the recent decades.
-Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself.
-The situation during the 19th century is uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden.
-From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980.
-Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}.
-The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.
-
-It thus appears as if a more consistent social health divide appeared at some point during the mortality transition, at least in Sweden.
-Solid empirical evidence about the process is however lacking and studies focusing on the issue are still few.
-There is a need for more reliable studies from different geographical and historical settings as well as according to age and gender in order to better understand the role of socio-economic conditions for health and survival over time.
-
-% This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
-% the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
-% in time for the Skellefteå region (1851--1900), and we also consider causes
-% of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
-% us to study the effect of income and education on mortality in a more
-% direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.
-
-\section{Aims and questions}
-
-In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
-The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
-The main question we investigate is if inequality in mortality between social classes increased among the adult and elderly population, primarily benefiting the highest social positions and how this is connected to the development of economic inequality in society.
-This issue is then analysed according to gender, age-group (working age vs retired) and cause of death (cardiovascular and cancer).
-Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period
-1986--2013.
-
-% ? :
-% 
-% \begin{enumerate}
-% \item Has inequality in mortality between social classes increased among the adult and elderly
-%   population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
-% 
-% %%\item Is the mortality transition initiated by a certain social class?
-% 
-% \item Are there any gender differences in the effect of social position?
-% 
-% \item Is social position equally important among the retired population as for those in working age?
-% \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
-% \item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
-% \end{enumerate}
-% 
-%
-
-\section{The Skellefteå and Umeå regions}
-
-% Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
-% show the study areas and their position in Sweden.
-
-% \begin{figure}
-% \includegraphics[height=19cm]{myfigs/skum}
-% \caption{The Skellefteå and Umeå regions.}
-% \label{fig:SKEUME}
-% \end{figure}
+---
+title: "Long-term perspectives of social inequality in mortality among
+  adults and elderly in northern Sweden 1851--2013"
+author: "Sören Edvinsson and Göran Broström"
+date: "28 juni 2017"
+output:
+  bookdown::html_document2:
+    toc: yes
+    toc_depth: 2
+    toc_float: true
+    number_sections: true
+bibliography: ["mybib.bib"]
+biblio-style: "apalike"
+link-citations: true
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
+# Introduction
+
+One of the central aspects of survival is social class and access to economic 
+and other resources. Even in present-day welfare societies, social position is a 
+strong determinant when it comes to health and mortality and the impact even 
+seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that 
+"... social conditions have been, are and will continue to be irreducible 
+determinants of health outcomes and therefore deserves their appellation as 
+'fundamental causes' of disease and death" \citep{liph95}. For a long time, 
+the general view has been that socio-economic health inequalities were high in 
+historical societies, probably higher than in modern societies. This is a 
+reasonable assumption since these societies were in most cases characterised by 
+very large differences between groups and classes. Knowing that access to 
+resources provides advantages in all ways, the health advantage of higher 
+classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social 
+inequality in health has passed through different historical phases. According 
+to them, social health differences were comparatively small during the 
+pre-transitional phase when space was a strong determinant for the spread of 
+disease. Differences then increased during the transitional phase when mortality 
+declined and wealthy groups used their resources to gain better health. Finally, 
+health differences decreased again in modern low-mortality settings when instead 
+health-related behaviour became important, resulting in marginal differences. 
+\citet{omran82} state, in the third proposition in his theory of the 
+epidemiological transition, that even if the class differential in mortality was 
+maintained during the transition, the decline set in earlier and was faster 
+among privileged groups.
+
+Recent studies, investigating social inequality in health and mortality
+with micro-data, has however questioned how generalizable the assumed
+pattern is \citep{bevp11}. Instead, some studies suggest that the modern
+pattern might be a fairly recent development, although the pattern differs
+depending on age, gender or historical context. When it comes to the adult
+and elderly population results from different studies have diverged, some
+reporting substantial differences but many, particularly those based on
+Swedish data, have found surprisingly small or even reversed socio-economic
+differences during the first stages of the transition \citep{bedr11,
+edli11,edbr12}, sharply contrasting against what we know from present-day
+society.
+
+Another question of the development of social inequality in mortality,
+concerns its relation to economic inequality. \citet{wipi09} argue that
+income inequality has an independent effect on mortality, separate from the
+direct effect of actual different access to economic resources. They argue
+that unequal societies perform less well when it comes to health than equal
+ones. Wilkinson and Pickett find a positive association between inequality
+and health (as well as other social conditions) at country level in the
+present-day economically developed world. This has initiated a vital
+scholarly debate and the topic has been extensively studied \citep{suka04,
+wado00}. How the association looked like in previous periods is
+unknown. This is of particular interest since the levels of economic
+inequality were very different from those of the recent decades. Even if it
+not a necessary implication, we would still assume that poorer groups were
+those most disadvantaged of large inequality in itself. The situation
+during the 19th century is uncertain, but at least during the late 19th and
+early 20th century both income and wealth distribution was strongly skewed
+in Sweden. From a couple of decades into the 20th century, income and
+wealth inequality continuously diminished to reach a low point at around
+1980. Thereafter, economic inequality has increased substantially until the
+present, although far from being as high as a century ago \citep{rowa08,
+rowa09}. The Swedish development resembles that of other European and North
+American countries in its basic features \citep{rowa14}.
+
+It thus appears as if a more consistent social health divide appeared at
+some point during the mortality transition, at least in Sweden. Solid
+empirical evidence about the process is however lacking and studies
+focusing on the issue are still few. There is a need for more reliable
+studies from different geographical and historical settings as well as
+according to age and gender in order to better understand the role of
+socio-economic conditions for health and survival over time.
+
+# Aims and questions
+
+In the present paper, we take a long-term perspective on the development of
+social inequalities in mortality in the adult and elderly population. The
+areas under investigation are the Skellefteå and Umeå regions in the north
+of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013
+for Umeå. The main question we investigate is if inequality in mortality
+between social classes increased among the adult and elderly population,
+primarily benefiting the highest social positions and how this is connected
+to the development of economic inequality in society. This issue is then
+analysed according to gender, age-group (working age vs retired) and cause
+of death (cardiovascular and cancer). Furthermore, we compare our results
+witn analyses of alternative inequality indicators, education and income
+for the period 1986--2013.
+
+# The Skellefteå and Umeå regions
 
@@
 
-The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf of Bothnia.
-This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century.
-The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s.
-One living long in memory was the famine of 1867–-1868 \citep{segb14}.
-During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
-The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}.
-Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures.
-The inland was particularly remote and isolated.
-During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
-Diseases reached to places that earlier were spared.
-
-The Skellefteå region until the 1950 in our dataset consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
-The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes.
-The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production.
-During the 20th century, industrialisation took place, partly by the opening of mines in the in-lands.
-This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.
-
-Mortality was fairly low in comparison with other parts of the country and the fertility transition was late \citep{coa86}.
-The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.
-
-The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality when another three parishes are included.
-This region has a somewhat different character from that of Skellefteå.
-Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period.
-It was the administrative centre in the county of Västerbotten.
-Schools and military regiments were placed here and the economy was more diversified compared to Skellefteå.
-Agriculture dominated the rural part, but there were also some foundries as well as industries for example in forestry and small-scale production.
-The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.
-
-\section{Data and variables}
-
-The recent extension in time of data at the Demographic Data Base (DDB), Umeå University (www.cedar.umu.se) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013.
-The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}.
-The data for the present study comes from two large population databases at the DDB.
-The early period until 1950 is covered by
+The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the
+county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf
+of Bothnia. This was a remote part of the country, where communication with
+the rest of Sweden was difficult until the late 19th century. The economy
+was dominated by agriculture, making it vulnerable to harvest failures, and
+several severe famines affected the regions during the 1800s. One living
+long in memory was the famine of 1867–-1868 \citep{segb14}. During the long
+winters, sea communication was hindered due to the Gulf of Bothnia being
+frozen. The first ships during the year came to this area in May or even as
+late as June in some cases \citep{fah56}. Towards the end of the 19th
+century, the Swedish railway system reached this part of Sweden, thereby
+facilitating contacts with the rest of Sweden, improving the economy and
+mitigating the effects of harvest failures. The inland was particularly
+remote and isolated. During the 19th century, our regions became
+increasingly integrated in the same epidemiological pattern as the rest of
+Sweden. Diseases reached to places that earlier were spared.
+
+The Skellefteå region until the 1950 in our dataset consists of a selection
+of parishes surrounding the town of Skellefteå, founded in 1845 but with a
+very small population during the 19th century. The data from the period
+after 1960 cover the Skellefteå and Malå municipalities, the same area as
+for the earlier period but with the addition of two more parishes. The
+majority of the 19th century population lived in rural villages and
+hamlets, getting their livelihood from agricultural production. During the
+20th century, industrialisation took place, partly by the opening of mines
+in the in-lands. This also led to a population increase both in the town
+and in the rural parts, and a much more diversified economy.
+
+Mortality was fairly low in comparison with other parts of the country and
+the fertility transition was late \citep{coa86}. The Skellefteå population
+size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on
+January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the
+century.
+
+The Umeå region in this study consists of Umeå urban and rural parish until
+1950, and from 1960 onwards of Umeå municipality when another three
+parishes are included. This region has a somewhat different character from
+that of Skellefteå. Umeå town had a small population but was substantially
+larger than Skellefteå town during almost the whole studied period. It was
+the administrative centre in the county of Västerbotten. Schools and
+military regiments were placed here and the economy was more diversified
+compared to Skellefteå. Agriculture dominated the rural part, but there
+were also some foundries as well as industries for example in forestry and
+small-scale production. The population size as defined in our data sets was
+18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when
+the century ended.
+
+# Data and variables
+
+The recent extension in time of data at the Demographic Data Base (DDB),
+Umeå University (www.cedar.umu.se) by including the period between 1900 and
+1950s makes it possible to investigate the long-term development of social
+inequality in mortality, in the present case from 1851 to 2013. The
+databases provide us with micro-data on a sufficiently large population and
+with substantial social diversity for the Skellefteå and Umeå regions
+\citep{wee16}. The data for the present study comes from two large
+population databases at the DDB. The early period until 1950 is covered by 
 the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked parish records, allowing
-us to reconstruct life biographies on people as long as they remained in
-the region. The records are linked within but not between the regions. Data
-from Poplink are accessed for the period 1851--1950.
+the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked
+parish records, allowing us to reconstruct life biographies on people as
+long as they remained in the region. The records are linked within but not
+between the regions. Data from Poplink are accessed for the period 1851--1950.
 
 The other large data-set is extracted from the Linnaeus database
-\citep{malm10}. Linnaeus is based on different linked national population registers from
-1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1960 to 2000 are constructed from censuses every fifth
-year 1960--1990, with additional information on deaths from National Board of Health and Welfare.
-For the period 2002--2013 we use the information from the yearly population registers from Statistics Sweden (LISA) together with death information from National Board of Health and Welfare.
+\citep{malm10}. Linnaeus is based on different linked national population
+registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
+University. The study period from 1960 to 2000 are constructed from
+censuses every fifth year 1960--1990, with additional information on deaths
+from National Board of Health and Welfare. For the period 2002--2013 we use
+the information from the yearly population registers from Statistics Sweden
+(LISA) together with death information from National Board of Health and
+Welfare.
 
@@
 
-In the data-set analysed here, all individuals 40 years and older ever being resident in either
-of the regions are included. The data file contains the variables on social class ({\sc Hisclass}) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
-entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.
+In the data-set analysed here, all individuals 40 years and older ever
+being resident in either of the regions are included. The data file
+contains the variables on social class ({\sc Hisclass}) sex, urban/rural
+residence, birth date, death date, first and last date of observation and
+type of entrance/exit. Total number of persons are 84\,599 leading to
+35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in
+the Linnaeus database. See Figure~\ref{fig:lexis2}.
 
@@
 
-\subsection{Presence periods}
-
-Differences in available information in the two datasets as well as within the Linnaeus data makes it necessary to apply different approaches when it comes to identification of presence periods within our
+## Presence periods
+
+Differences in available information in the two datasets as well as within
+the Linnaeus data makes it necessary to apply different approaches when it
+comes to identification of presence periods within our 
 regions. The Poplink data provides us with exact dates or at least year of
-start and exit of the presence of all individuals in our data, allowing us to have full and continuous control of the de jure population.
-This is however not the case with the Linnaeus database.
-For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare to identify the observation periods in our analyses.
-There is however no full information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses.
-We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in the later census.
-This overestimates their presence in our regions, but should not bias the results when it comes to socioeconomic differences.
-We furthermore miss those that both moved in and moved out or vice versa between censuses.
-Exit dates are exact if they refer to death, and in these cases we consider them as constantly present from the last census to the death data even if the death took place in another part of Sweden.
-
-For the last period, 2002--2013, we use the yearly population registers (LISA data), together with information on deaths from National Board of Health and Welfare.
-This data allows good control of the presence periods in the studied regions.
-
-\subsection{Social class}
-
-In this paper we analyse mortality differences according to social class based on the classification scheme {\sc Hisclass} \citep{lema11}.
-The only type of socio-economic information available both in Linnaeus and Poplink is occupations.
-Thus, analysing according to social class is apart from being a scientific choice, also a practical one.
-Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes.
-Social position is treated as a time-dependent variable.
-
-The availability to information on occupations varies over time and between sources.
-Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system.
-The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering).
-The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information.
-In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid.
-Thus, the social class in the census 1990 is considered to be kept until 2001.
-This strategy is also used for other cases of missing social class.
-Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65.
-For the Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided to us by Jonas Helgertz, Lund University.
-
-The use of different data sets, forces us to apply different approaches towards defining the social position in some respects.
-A problem is that Poplink usually only provides the occupation for the head of household, thus under-estimating female working-force participation as well as that of adult children residing with their parents \citep{vik10}.
-In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position.
-This strategy makes the problem of under-recording less problematic.
-
-Female labour is much better covered in the Linnaeus database.
-A complication is however that it is  difficult to define households in the same way as in Poplink.
-Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became common and female labour force participation developed as the norm in Sweden during this period.
-The occupation of the husband as signifier of social class became, if not obsolete, at least less relevant.
-All included persons are signified by their individual occupation for this period.
-Still, the results for the different periods are not completely comparable and makes the interpretation of the development for women more shaky.
-
-{\sc Hisclass}, the classification system used here, is a "\ldots HISCO-based
-historical international social class scheme" \citep{lema11, lema02}.
-The different classes in {\sc Hisclass} represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector.
-Implicitly it reflects large differences in what is in the forefront of other
-categorisations, as for example access to economic and other resources, status, social power et cetera.
-We have merged the original 12 classes in {\sc Hisclass}, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:
+start and exit of the presence of all individuals in our data, allowing us
+to have full and continuous control of the de jure population. This is
+however not the case with the Linnaeus database. For the period 1960--2000
+we use presence in the censuses 1960--1990 together with the yearly
+population registers 1986--2000 and information on deaths from the National
+Board of Health and Welfare to identify the observation periods in our
+analyses. There is however no full information on what happened in-between
+the censuses, for example the exact date of exit for persons that moved
+between two censuses. We have solved this by considering everyone as
+resident in our regions during the complete five-year period between
+censuses if the person is missing in the later census. This overestimates
+their presence in our regions, but should not bias the results when it
+comes to socioeconomic differences. We furthermore miss those that both
+moved in and moved out or vice versa between censuses. Exit dates are exact
+if they refer to death, and in these cases we consider them as constantly
+present from the last census to the death data even if the death took place
+in another part of Sweden.
+
+For the last period, 2002--2013, we use the yearly population registers
+(LISA data), together with information on deaths from National Board of
+Health and Welfare. This data allows good control of the presence periods
+in the studied regions.
+
+## Social class
+
+In this paper we analyse mortality differences according to social class
+based on the classification scheme {\sc Hisclass} \citep{lema11}. The only
+type of socio-economic information available both in Linnaeus and Poplink
+is occupations. Thus, analysing according to social class is apart from
+being a scientific choice, also a practical one. Using occupational data is
+the only and obvious alternative for making a longitudinal study of
+socio-economic mortality differences with consistent classification
+schemes. Social position is treated as a time-dependent variable.
+
+The availability to information on occupations varies over time and between
+sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the
+original DDB coding system. The censuses 1960--1990 include occupational
+codes based on the system used in Sweden at that time, NYK (Nordisk
+Yrkesklassificering). The yearly population registers (LISA) do not report
+occupations until 2001, thus there is a window of 11 years from 1991 to
+2001 where we lack information. In the censuses, we have solved this by
+extending the periods for which we consider the observed social class to be
+valid. Thus, the social class in the census 1990 is considered to be kept
+until 2001. This strategy is also used for other cases of missing social
+class. Furthermore, we consider the last social position before retirement
+to be kept until death or last observation for those above 65. For the
+Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided
+to us by Jonas Helgertz, Lund University.
+
+The use of different data sets, forces us to apply different approaches
+towards defining the social position in some respects. A problem is that
+Poplink usually only provides the occupation for the head of household,
+thus under-estimating female working-force participation as well as that of
+adult children residing with their parents \citep{vik10}. In this study and
+according to the research questions in the present paper, we have chosen to
+categorise wife according to the position of the head of household, usually
+the husband/father, assuming that the family shares the same socio-economic
+position. This strategy makes the problem of under-recording less
+problematic.
+
+Female labour is much better covered in the Linnaeus database. A
+complication is however that it is  difficult to define households in the
+same way as in Poplink. Extra-marital cohabitation, connected to the Second
+Demographic Transition \citep{van87, van03}, became common and female
+labour force participation developed as the norm in Sweden during this
+period. The occupation of the husband as signifier of social class became,
+if not obsolete, at least less relevant. All included persons are signified
+by their individual occupation for this period. Still, the results for the
+different periods are not completely comparable and makes the
+interpretation of the development for women more shaky.
+
+{\sc Hisclass}, the classification system used here, is a "\ldots
+HISCO-based historical international social class scheme" \citep{lema11,
+lema02}. The different classes in {\sc Hisclass} represent distinct
+categories, based on whether the work is manual or non–manual, skill level,
+supervision and sector. Implicitly it reflects large differences in what is
+in the forefront of other categorisations, as for example access to
+economic and other resources, status, social power et cetera. We have
+merged the original 12 classes in {\sc Hisclass}, based on position in the
+labour market, ending up with the following three and an additional fourth
+category with unknown class:
 
 \begin{enumerate}
-\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals and higher white collar.
-\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
-\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including farm workers.
-%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
+\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals
+and higher white collar. 
+\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers,
+farmers and lower white collar. 
+\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including
+farm workers. 
 \end{enumerate}
 
-% The procedure for
-% occupations in the Linnaeus database is a little more complicated. The
-% occupations in the censuses (NYK) are first transferred to the system
-% later in use, SSYK (the Swedish version of ISCO88). Thereafter the
-% SSYK, both those transferred from the censuses and those given in the
-% population registers 2001-2013, can be re-coded into HISCLASS.
-
-Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution (per cent) of exposure
-time according to class for men and women respectively. The category of workers is
-fairly stable over time. The highest class has increased substantially, although from a very low level in the 19th century. The increase of academics in connection to the establishment of
-Umeå University explains much of this increase.
-What is hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century.
-Furthermore we observe an increase during the 1960s and partly during later decades as well in those with undefined.
-When first observed in the earliest included censuses, occupation is often missing.
-For men the proportion is somewhat higher in the age group 65 and above, i.e.
-those that were retired, but still low enough to be used in our analyses.
-Until 1950, there were however very few with missing occupation.
-For men in working-active age, the proportion missing does not cause any problems for our analyses.
+Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution
+(per cent) of exposure time according to class for men and women
+respectively. The category of workers is fairly stable over time. The
+highest class has increased substantially, although from a very low level
+in the 19th century. The increase of academics in connection to the
+establishment of Umeå University explains much of this increase. What is
+hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle
+group from a majority of farmers during the 19th century to lower white
+collar from the middle of the 20th century. Furthermore we observe an
+increase during the 1960s and partly during later decades as well in those
+with undefined. When first observed in the earliest included censuses,
+occupation is often missing. For men the proportion is somewhat higher in
+the age group 65 and above, i.e. those that were retired, but still low
+enough to be used in our analyses. Until 1950, there were however very few
+with missing occupation. For men in working-active age, the proportion
+missing does not cause any problems for our analyses.
 
@@
 
-The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}.
-Corresponding to what we found for men, the proportion without social position is very low until the middle of the 20th century.
-This is a reflection of how womens social position is defined for that period.
-Missing social class is however a real problem during the 1960s and 1970s in particular.
-Around 80 \% lack information for the 1960s and 50 \% for the 1970s.
-Since social class for this period is defined according to own occupation, the consequence of many women being outside the workforce, the proportion missing becomes very high.
-We have chosen not to report results for women for these two decades.
-For the later decades, the results are fairly complete and should be representative.
-
-% The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
-% \ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.
-
-% <<nanam, results = 'asis'>>=
-% library(xtable)
-% ##load("data/vb0.rda")
-% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-% x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
-% x <- round(prop.table(x, 1) * 100)
-% xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
-% @
-% 
-% <<nanaw, results = 'asis'>>=
-% ##library(xtable)
-% ##load("data/vb0.rda")
-% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
-% x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
-% x <- round(prop.table(x, 1) * 100)
-% xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
-% @
+The proportion with missing values among women is however a larger problem,
+see Figure~\ref{fig:hiscskumw}. Corresponding to what we found for men, the
+proportion without social position is very low until the middle of the 20th
+century. This is a reflection of how womens social position is defined for
+that period. Missing social class is however a real problem during the
+1960s and 1970s in particular. Around 80 \% lack information for the 1960s
+and 50 \% for the 1970s. Since social class for this period is defined
+according to own occupation, the consequence of many women being outside
+the workforce, the proportion missing becomes very high. We have chosen not
+to report results for women for these two decades. For the later decades,
+the results are fairly complete and should be representative.
 
@@
 
-Marital status is treated as a time-dependent variable in the data file.
-The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces).
-If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.
-
-Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}.
-The figure indicates an under-registration of those married and and over-registration of previously married during the 1960s.
+Marital status is treated as a time-dependent variable in the data
+file. The statuses used are unmarried, married and dissolved marriages
+(without distinguishing between widowed and divorces). If no explicit
+status is given in the sources and partner is missing, the status has been
+set to unmarried.
+
+Regarding the distribution of marital status over time, see
+Figure~\ref{fig:civstskum}. The figure indicates an under-registration of
+those married and and over-registration of previously married during the
+1960s.
 
@@
 
-\subsection{Cause of death}
+## Cause of death
+
 In the present study, we analyse mortality in the two most common disease
 groups in the ages we are investigating---cardiovascular diseases and 
-cancers.
-National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year.
-Classification systems have however changed.
-Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system.
-In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.
-
-The distribution of causes of death is shown in Figure~\ref{fig:causeskum}.
-Notice the high proportion of other causes in the first two periods.
-This category include cases with unknown or unspecific causes which has implications for the cause-specific analyses (see later discussion).
-The total dominance of our two disease groups is apparent.
+cancers. National compulsory reporting of cause of death was introduced in
+1911, thus our analysis of the two disease groups starts from that
+year. Classification systems have however changed. Sweden did not join ICD
+until 1952, but the historical causes of death in the Poplink database have
+been coded according to the ICD10 system. In the present study, we have
+identified the codes relevant for cardiovascular diseases and cancers in
+the ICD 8, 9 and 10.
+
+The distribution of causes of death is shown in
+Figure~\ref{fig:causeskum}. Notice the high proportion of other causes in
+the first two periods. This category include cases with unknown or
+unspecific causes which has implications for the cause-specific analyses
+(see later discussion). The total dominance of our two disease groups is
+apparent.
 
@@
 
-\subsection{Urban vs rural residence}
-In 19th and early 20th century Sweden (as in other countries), mortality was much higher in urban environments.
-The unsanitary and overcrowded living conditions as well as the risks for importing infections made the health risks higher in towns and cities.
-In the present paper, we control for this aspect by distinguishing between urban and rural residence, based on in which parish people lived.
-
-\subsection{Periods for analysis}
-We have made separate analyses for the different periods.
-The century 1851--1950 is divided into four periods of 25 years each.
-The 19th century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions.
-During the years 1901--1950, industrialisation started and the urban environments increased in population size.
-The modern welfare state began to develop during this time.
-For the period 1960 on-wards, there are separate analyses on decades except for the two last periods.
-Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s.
-A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to being less generous than before.
-The start of the last period (2008--2013) coincides with the international financial crisis.
-
-\subsection{Income}
-Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to 2005.
-
-\subsection{Education}
-Education is here analysed only for the period 1990 to 2013.
-Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden.
-It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.
-
-\section{Models}
+## Urban vs rural residence
+
+In 19th and early 20th century Sweden (as in other countries), mortality
+was much higher in urban environments. The unsanitary and overcrowded
+living conditions as well as the risks for importing infections made the
+health risks higher in towns and cities. In the present paper, we control
+for this aspect by distinguishing between urban and rural residence, based
+on in which parish people lived.
+
+## Periods for analysis
+
+We have made separate analyses for the different periods. The century
+1851--1950 is divided into four periods of 25 years each. The 19th century
+periods represent a mainly pre-industrial society dominated by agriculture
+and with restricted welfare provisions. During the years 1901--1950,
+industrialisation started and the urban environments increased in
+population size. The modern welfare state began to develop during this
+time. For the period 1960 on-wards, there are separate analyses on decades
+except for the two last periods. Sweden had developed into a wealthy
+welfare society with low income inequality in international comparison,
+particularly until the 1990s. A financial crisis struck Sweden in the early
+1990s, which among other things increased unemployment and changed parts of
+the welfare system to being less generous than before. The start of the
+last period (2008--2013) coincides with the international financial
+crisis.
+
+## Income
+
+Income is based on disposable individual income from the LISA database at
+Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to
+2005.
+
+## Education
+
+Education is here analysed only for the period 1990 to 2013. Years of
+education are given in 3 categories and is based on level of education
+according to categories in the LISA database at Statistics Sweden. It
+ranges from basic education (nine years of schooling or less) at the lowest
+level, secondary education (high school or similar) for the middle group
+and to academic degree at the highest level.
+
+# Models
 
@@
 Another practical problem is the size of that data set:
-It contains no less than 2\,432\,871 records of 236\,011 individuals, of
-which 68\,244 are observed to die.
+It contains no less than 2432871 records of 236011 individuals, of
+which 68244 are observed to die.
 This makes it very inconvenient to use the nowadays so popular semi-parametric
@@
 
-% We analyse both regions together but perform separate analyses for men and
-% women and for the age groups 40--64 and 65--89. The model in the analyses
-% include social class, sex and marital status.
-
-% We present the results for the period 2001--2013
-% separately. We suggest a heterogeneity index in order to investigate a
-% possible homogenisation of mortality between social classes.
-
 The important explanatory variable, {\tt hisclass}, is included in the
@@
 interaction term between {\tt age} and {\tt hisclass}. Then, the main
-results are presented \emph{graphically}.
-Tables with results from the different models are presented at the end of the article, but are not commented on in detail.
-The comments on the results are primarily based on the figures and focuses on the central variable in our study---{\tt hisclass}.
-As can be seen from the tables, most variables in the model are significant at 5 per cent level.
-
-% There is some  lack of information in the higher
-% ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
-% information is simply missing to a large extent.
-
-\section{The effect of {\sc Hisclass}, 1851--1950, 1961--2013}
-
-\subsection{All causes of death}
+results are presented \emph{graphically}. Tables with results from the
+different models are presented at the end of the article, but are not
+commented on in detail. The comments on the results are primarily based on
+the figures and focuses on the central variable in our study---{\tt
+hisclass}. As can be seen from the tables, most variables in the model are
+significant at 5 per cent level.
+
+# The effect of {\sc Hisclass}, 1851--1950, 1961--2013
+
+## All causes of death
 
@@
 
-%creating an \emph{heterogeneity index} (a better name is needed:
-%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
-%Something catchy.).
-
-\subsubsection{Ages 40--64}
-
-%The distribution over time periods.
+### Ages 40--64
+
 <<getske4060>>=
@@
 
-% We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
-% periods, in order to illustrate the process. They are adjusted for {\tt civil status}. We present separate analyses  for women and men, see
-% Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.
-
-%\subsubsection{Women}
-
 <<hazcalw>>=
@@
 @
-
-% 
-% 
-% <<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
-% source("R/plotHaz.glm.R")
-% 
-% plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
-% @
-% 
-% <<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
-% plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
-% @
-% 
-% From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
-% the idea of proportional hazards is not reasonable. We therefore use another approach.
-
-%The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
-%Figure~\ref{fig:cumsw}.
 
@@
 
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
-%%@
-
-%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-Notice that we have not analysed the 1950s due to the fact that we only have data for some parishes at that time, consequently having no corresponding population for that decade.
-Furthermore, we have excluded the results for women during the 1960s and 1970 of reasons discussed above.
-The changing economy also have consequences for the analysis, particularly when it comes to the agricultural sector.
-While farmers were common occupations during the first half of the 20th century, these occupations has constantly decreased in numbers thereafter.
-This means that the middle class has somewhat different character during the last periods.
-
-The most striking result of the analysis is the different patterns for men and women. The social pattern of mortality among women is quite much in accordance with previous
-analyses from the present research group on other regions in northern
-Sweden and for other periods \citep{edli11,edbr12}. The highest
-social class usually have comparatively low mortality, while working
-class women have high.
-There is however no substantial change in the relative risks between the classes.
-Only the group with unknown social class is distinguished by moch higher mortality from the 1980s onwards.
-This can however have to do with the differences how social position for women is defined before and after 1950.
-After 1950 it is based on the own occupation.
-Being without occupation might include women that are more frail both when it comes to health and socially.
-
-A different pattern appears among men, being the most surprising finding in our analyses. The groups we would expect to have the lowest mortality, if we assume that access to different resources and
-having a high status determines survival, instead have by far the highest hazards of dying during the nineteenth and much of the 20th century. This is also in accordance with our previous studies
-\citep{edli11, edbr12}.
-Workers on the other hand have comparatively low mortality until the 1960s, at the same level as the middle class.
-The relative positions change radically towards the end the 20th century.
-From the 1980s, there is a clear advantage when it comes to survival for higher
-managers and professionals.
-The relative differences are however more compressed than for the period until 1950.
-The expected social gradient in mortality has developed.
-This is also what we usually find in analyses of social health inequalities in present-day Sweden as well as in most
-other countries. In the regions we are studying, this is however a quite
-recent phenomenon.
-Corresponding to what we find for women, those without known occupation have much higher mortality than the others from the 1960s onwards.
-The assumption made for women of unknowns social class is probably even more relevant for men.
-They probably represent groups in society that are particularly vulnerable.
-
-A further observation can be made from Figure~\ref{fig:cumsw}.
-The gender difference in death hazards is very big for the élite group from 1851--1950, men having about three times higher risks than women in the same class.
-In the other classes, the difference was much smaller.
-It was men in high position that experienced the largest risks.
-
-% \subsubsection{The effect of income and education, 1986--2013}
-% 
-% A common alternative to a class analysis is to use income or education for investigating socio-economic health differences. 
-% Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
-% disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
-% 
-% The analysis for the period 1990--2005 is based on a separate analysis with 
-% income and education as explanatory variables and with controls of period 
-% and marital status.
-% 
-% <<lissyoung>>=
-% library(eha)
-% library(xtable)
-% liss4064 <- loTab[loTab$age < 64.5, ]
-% con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-% ##load("data/lisaold.rda")
-% load(con)
-% close(con)
-% @
-% 
-% Both education and income had a strong, statistically significant, effect on survival, as expected. 
-% 
-% \paragraph{Education, 1990--2013}
-% 
-% <<educ>>=
-% 
-% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
-% educ$period <- factor(educ$period)
-% educ <- educ[!is.na(educ$educ), ]
-% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-% ##educ$educ <- relevel(educ$educ, ref = "middle")
-% educ$civst <- relevel(educ$civst, ref = "married")
-% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-%              data = educ[educ$sex == "male", ])
-% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-%              data = educ[educ$sex == "female", ])
-% cols <- c("red", "darkgreen", "black")
-% @
-% 
-% <<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
-% oldpar <- par(mfrow = c(1, 2))
-% 
-% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
-% 
-% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
-% 
-% par(oldpar)
-% @
-% 
-% Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival.
-The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.
-% 
-% \paragraph{Income, 1986--2005}
-% 
-% <<femeduc>>=
-% income <- lisaold[!is.na(lisaold$income), ]
-% income$period <- factor(income$period)
-% ##income <- income[!is.na(income$income), ]
-% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-% ##income$income <- relevel(income$income, ref = "low")
-% income$civst <- relevel(income$civst, ref = "married")
-% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-%              data = income[income$sex == "male", ])
-% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-%              data = income[income$sex == "female", ])
-% cols <- c("darkgreen", "red", "blue", "black")
-% 
-% @
-% 
-% <<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
-% oldpar <- par(mfrow = c(1, 2))
-% 
-% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
-% 
-% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-% axis(1)
-% axis(2, at = c(0, 0.5, 1.0), las = 1)
-% abline(h = 0.5, lty = 3)
-% box()
-% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
-% 
-% par(oldpar)
-% @
-% 
-% Higher incomes result in lower mortality. This is also what could be expected, 
-% and conforms to what we know of social health differences in present-day Sweden. 
-% Notice however that we cannot establish a causal relationship between 
-% income and mortality. Bad health, for instance, may be a confounder.
-% 
-% %%The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
-% %%<<liswom, results = 'asis'>>=
-% %%fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-% %%##cat("Women:\n")
-% %%drw <- drop1(fitw1, test = "Chisq")
-% %%xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
-% %##round(summary(fitw1)$coef[-(1:13), ], 4)
-% %@
-% 
-% %%The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
-% 
-% %%<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
-% %%source("R/plotWEduc.R")
-% %%plotWEduc(fitw1)
-% %%@
-% 
-% 
-% %%And for men:
-% 
-% %%<<lismen, results='asis'>>=
-% %%fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-% %%##cat("\nMen:\n")
-% %%drm <- drop1(fitm1, test = "Chisq")
-% %%xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
-% %%##round(summary(fitm1)$coef[-(1:13), ], 4)
-% %%dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% %%@
-% 
-% %%<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
-% %%#source("R/plotWEduc.R")
-% %%plotWEduc(fitm1)
-% %%@
-% 
-% % 
-% % <<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
-% % #oldpar <- par(mfrow = c(1, 2))
-% % #plot(fitw1, fn = "cum", col = dcol, main = "Women")
-% % #plot(fitm1, fn = "cum", col = dcol, main = "Men")
-% % #par(oldpar)
-% % ##summary(fitw1)
-% % ##summary(fitm1)
-% % @
-% % 
-% % %See Table~\ref{tab:tablisswom1} the PH regression for women.
-% % <<tablisswom1, results = "asis">>=
-% % #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% % #                 data = liss4064[liss4064$sex == "female", ])
-% % #dr <- drop1(fitwo1, test = "Chisq")
-% % #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-% % #    label = "tab:tablisswom1")
-% % @
-% % 
-% % %See Table~\ref{tab:tablissman1} the PH regression for men.
-% % <<tablissman1, results = "asis">>=
-% % #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% % #                 data = liss4064[liss4064$sex == "male", ])
-% % #dr <- drop1(fitma1, test = "Chisq")
-% % #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
-% %  #   label = "tab:tablissman1")
-% % @
-%
-
-\subsubsection{Ages 65--89}
-
-%The distribution over time periods.
+Notice that we have not analysed the 1950s due to the fact that we only
+have data for some parishes at that time, consequently having no
+corresponding population for that decade. Furthermore, we have excluded the
+results for women during the 1960s and 1970 of reasons discussed above. The
+changing economy also have consequences for the analysis, particularly when
+it comes to the agricultural sector. While farmers were common occupations
+during the first half of the 20th century, these occupations has constantly
+decreased in numbers thereafter. This means that the middle class has
+somewhat different character during the last periods.
+
+The most striking result of the analysis is the different patterns for men
+and women. The social pattern of mortality among women is quite much in
+accordance with previous analyses from the present research group on other
+regions in northern Sweden and for other periods \citep{edli11,edbr12}. The
+highest social class usually have comparatively low mortality, while
+working class women have high. There is however no substantial change in
+the relative risks between the classes. Only the group with unknown social
+class is distinguished by moch higher mortality from the 1980s
+onwards. This can however have to do with the differences how social
+position for women is defined before and after 1950. After 1950 it is based
+on the own occupation. Being without occupation might include women that
+are more frail both when it comes to health and socially.
+
+A different pattern appears among men, being the most surprising finding in
+our analyses. The groups we would expect to have the lowest mortality, if
+we assume that access to different resources and having a high status
+determines survival, instead have by far the highest hazards of dying
+during the nineteenth and much of the 20th century. This is also in
+accordance with our previous studies \citep{edli11, edbr12}. Workers on the
+other hand have comparatively low mortality until the 1960s, at the same
+level as the middle class. The relative positions change radically towards
+the end the 20th century. From the 1980s, there is a clear advantage when
+it comes to survival for higher managers and professionals. The relative
+differences are however more compressed than for the period until 1950. The
+expected social gradient in mortality has developed. This is also what we
+usually find in analyses of social health inequalities in present-day
+Sweden as well as in most other countries. In the regions we are studying,
+this is however a quite recent phenomenon. Corresponding to what we find
+for women, those without known occupation have much higher mortality than
+the others from the 1960s onwards. The assumption made for women of
+unknowns social class is probably even more relevant for men. They probably
+represent groups in society that are particularly vulnerable.
+
+A further observation can be made from Figure~\ref{fig:cumsw}. The gender
+difference in death hazards is very big for the élite group from
+1851--1950, men having about three times higher risks than women in the
+same class. In the other classes, the difference was much smaller. It was
+men in high position that experienced the largest risks.
+
+### Ages 65--89
+
 <<getske6089>>=
@@
 
-The general pattern and development of mortality (not shown here) do not differ that much from the younger age group, of course representing a much higher mortality level earlier in history.
-Mortality among women was, as expected, lower than among men.
-We find a clear decline in mortality, but the fastest decline did not set in definitely until sometimes after the Second World War (results not shown here), i.e.
-later than for the age group 40--64 years.
-This corresponds to the Swedish development where we find a turning point at this time with better survival among elderly, to a large extent caused by lower mortality in cardiovascular diseases.
-
-Concerning the social pattern, the pattern is less clear compared to what we found for the younger age group (see Figure~\ref{fig:cumswold}).
-For women the results are mixed.
-Once again, too much missing data on social class leaves out the results for the 1960s and 1970s.
-We do not find the same clear advantage for élite women during period until 1950, but neither do we find any large disadvantage for working class elderly women at that time.
-In some periods, the social differences were small, and the élite even had substantially higher mortality during the early 20th century.
-The social differences continued to be small during the last decades of the 20th century.
-Only during the two last periods (after the year 2000) is it possible to establish the expected social gradient in mortality.
-
-For men we find indications of the same intriguing change among elderly as for
-the adult male population 40--64 years. Mortality was in fact substantially
-higher among élite men  but their disadvantage was primarily apparent during the 19th century.
-Differences are fairly small during the first half of the 20th century.
-Not until the second half of the 20th century is the survival much better for the highest social class while workers in the age of retirement had high mortality in comparison.
-).
-This is quite unexpected, since other studies have shown clear socio-economic differences in Sweden during recent decades.
+The general pattern and development of mortality (not shown here) do not
+differ that much from the younger age group, of course representing a much
+higher mortality level earlier in history. Mortality among women was, as
+expected, lower than among men. We find a clear decline in mortality, but
+the fastest decline did not set in definitely until sometimes after the
+Second World War (results not shown here), i.e. later than for the age
+group 40--64 years. This corresponds to the Swedish development where we
+find a turning point at this time with better survival among elderly, to a
+large extent caused by lower mortality in cardiovascular diseases.
+
+Concerning the social pattern, the pattern is less clear compared to what
+we found for the younger age group (see Figure~\ref{fig:cumswold}). For
+women the results are mixed. Once again, too much missing data on social
+class leaves out the results for the 1960s and 1970s. We do not find the
+same clear advantage for élite women during period until 1950, but neither
+do we find any large disadvantage for working class elderly women at that
+time. In some periods, the social differences were small, and the élite
+even had substantially higher mortality during the early 20th century. The
+social differences continued to be small during the last decades of the
+20th century. Only during the two last periods (after the year 2000) is it
+possible to establish the expected social gradient in mortality.
+
+For men we find indications of the same intriguing change among elderly as
+for the adult male population 40--64 years. Mortality was in fact
+substantially higher among élite men  but their disadvantage was primarily
+apparent during the 19th century. Differences are fairly small during the
+first half of the 20th century. Not until the second half of the 20th
+century is the survival much better for the highest social class while
+workers in the age of retirement had high mortality in comparison. ). This
+is quite unexpected, since other studies have shown clear socio-economic
+differences in Sweden during recent decades.
 
@@
 
-% \subsubsection{The effect of income and education, 1990--2005}
-% 
-% The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting.
-As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
-% 
-% <<lissyoung2>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss6589 <- loTab[loTab$age > 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in
-% Table~\ref{tab:liswom2}
-% <<liswom2, results = 'asis'>>=
-% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-% ##cat("Women:\n")
-% drw2 <- drop1(fitw2, test = "Chisq")
-% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
-% 
-% <<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitw2)
-% @
-% 
-% 
-% And for men:
-% 
-% <<lismen2, results='asis'>>=
-% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm2 <- drop1(fitm2, test = "Chisq")
-% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% <<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm2)
-% @
-% 
-% These figures show clear associations both with income and education and for women as well as for men.
-More years of education resulted in lower mortality.
-For women the association with additional levels became however smaller at the higher levels, the effect is more visible at the steps above the basic levels.
-In a similar fashion, increased incomes is associated with improved survival both for men and for women.
-Apparently, education and income is better to catch up socio-economic differences compared to hisclass, at least for this period.
-
-\subsection{Cardiovascular mortality}
-
-In this section, the focus is on analyses of mortality from the two dominant causes of death among the adult and elderly population.
-We start with cardiovascular diseases.
-Many studies have shown that socio-economic conditions are strongly associated with mortality in this disease group.
-Belonging to working class, having less education and low income increases the mortality risks.
-We would therefore expect to find a clear social pattern in the present analysis.
-
-First, a general observation is that the analysis of causes of death and its development is difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ.
-Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers.
-There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory.
-For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases.
-However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent.
-It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and that the increase partly is explained by improvements in how deaths were diagnosed.
+## Cardiovascular mortality
+
+In this section, the focus is on analyses of mortality from the two
+dominant causes of death among the adult and elderly population. We start
+with cardiovascular diseases. Many studies have shown that socio-economic
+conditions are strongly associated with mortality in this disease
+group. Belonging to working class, having less education and low income
+increases the mortality risks. We would therefore expect to find a clear
+social pattern in the present analysis.
+
+First, a general observation is that the analysis of causes of death and
+its development is difficult to interpret, partly because the number of
+cases are fewer but also that the quality and coverage of reported death
+causes differ. Thus, many deaths during the early 20th got quite unspecific
+causes reported in the death registers. There was still a substantial part
+of all deaths where cause was diagnosed as unknown or dying from old age,
+despite the fact that reporting causes of death had become mandatory. For
+the period 1901--1925 to 1926--1950, the proportion of deaths from
+cardiovascular diseases increased from 25 to 43 per cent for the age group
+40 and above, thus indicating an increase of these diseases. However, at
+the same time unknown cause and death of old age decreased from 28 percent
+to 18 per cent. It is quite reasonable to suspect that many cardiovascular
+deaths were disguised in unspecific categories and that the increase partly
+is explained by improvements in how deaths were diagnosed.
+
 <<causes41>>=
@@
 hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.
-
-%creating an \emph{heterogeneity index} (a better name is needed:
-%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
-%Something catchy.).
-
-\subsubsection{Ages 40--64}
-
-%The distribution over time periods.
+main target in our analysis: These numbers are used in the process of
+graphically illustrating the development over time.
+
+### Ages 40--64
+
 <<41getske4060>>=
@@
 
-Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40.
-
-%\subsubsection{Women}
+Performing the same types of analyses as for all-cause mortality, we start
+by looking at cumulative hazards by {\sc Hisclass} at age 65, given
+survival to age 40.
 
@@
 @
-
-% We will use as a summary measure the value of the cumulative hazards at
-% age 65, given survival to age 40.
 
@@
 
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
-%%@
 <<41cumswtab,results='asis'>>=
@@
 
-Figure~\ref{fig:41cumsw} shows mortality levels during different periods from 1901 on-wards.
-Women from the highest social class had consistently lower mortality from cardiovascular diseases compared to the others.
-We also observe increasing social differences from the first half of the century to the last decades.
-A new pattern has developed from 1980 onwards.
-Now working class women have substantially higher cardiovascular mortality compared to others.
-We can however not exclude that this is an artefact of different diagnostics for deaths in different groups.
-Working class women may have had less precise diagnoses in the early 20th century.
-
-For men, there is a different pattern, a pattern however that to a large extent is consistent with what we observed concerning all-cause mortality.
-The elite group had much higher cardiovascular diseases up until the middle of the 20th century, but instead have the lowest mortality from 1980 onwards.
-Working class men had a fairly low cardiovascular mortality, but in the same way as among women, the levels were fairly similar to the one of the middle category consisting of lower manager/farmers.
-And in the same way as for women at this age, working class men have the highest mortality from this cause of death during the last decades.
-A social hierarchy in cardiovascular mortality seems to have developed both for men and women.
-
-% A word of caution is however needed when it comes to these results.
-We can not exclude that the smaller differences in cardiovascular diseases for women in the early periods are an artefact of how deaths were diagnosed.
-Working class men and women may have had less precise diagnoses in the early 20th century.
-
-%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-% \subsubsection{The effect of income and education, 1990--2005}
-% 
-% We have also performed the analysis on the alternative approaches on socio-economic differences for the causes of death.
-It is based on a separate analysis with income and education as explanatory variables.
-% 
-% <<41lissyoung>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% loTab$period <- factor(loTab$period)
-% liss4064 <- loTab[loTab$age < 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in
-% Table~\ref{tab:41liswom}
-% <<41liswom, results = 'asis'>>=
-% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-% ##cat("Women:\n")
-% drw <- drop1(fitw1, test = "Chisq")
-% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cardiovascular death.", label = "tab:41liswom")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:41liswomfig}
-% 
-% <<41liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% source("R/plotWEduc.R")
-% plotWEduc(fitw1)
-% @
-% 
-% 
-% And for men:
-% 
-% <<41lismen, results='asis'>>=
-% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm <- drop1(fitm1, test = "Chisq")
-% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cardiovascular death.", label = "tab:41lismen")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% <<41lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm1)
-% @
-% 
-% 
-% <<41plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cardiovascular death.", fig.height = 4>>=
-% #oldpar <- par(mfrow = c(1, 2))
-% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
-% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
-% #par(oldpar)
-% ##summary(fitw1)
-% ##summary(fitm1)
-% @
-% 
-% %See Table~\ref{tab:tablisswom1} the PH regression for women.
-% <<41tablisswom1, results = "asis">>=
-% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "female", ])
-% #dr <- drop1(fitwo1, test = "Chisq")
-% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-% #    label = "tab:tablisswom1")
-% @
-% 
-% %See Table~\ref{tab:tablissman1} the PH regression for men.
-% <<41tablissman1, results = "asis">>=
-% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "male", ])
-% #dr <- drop1(fitma1, test = "Chisq")
-% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
-%  #   label = "tab:tablissman1")
-% @
-% 
-% The analysis of cardiovascular mortality according to education and income conforms very well to what we find for all-cause mortality, and we will therefore not delve into detail on this issue.
-More education as well as higher income is advantageous both for women and men for avoiding deaths related to the circulatory system.
-This is also well in line with our analysis based on HISCLASS above.
-
-\subsubsection{Ages 65--89}
-
-%The distribution over time periods.
+Figure~\ref{fig:41cumsw} shows mortality levels during different periods
+from 1901 on-wards. Women from the highest social class had consistently
+lower mortality from cardiovascular diseases compared to the others. We
+also observe increasing social differences from the first half of the
+century to the last decades. A new pattern has developed from 1980
+onwards. Now working class women have substantially higher cardiovascular
+mortality compared to others. We can however not exclude that this is an
+artefact of different diagnostics for deaths in different groups. Working
+class women may have had less precise diagnoses in the early 20th century.
+
+For men, there is a different pattern, a pattern however that to a large
+extent is consistent with what we observed concerning all-cause
+mortality. The elite group had much higher cardiovascular diseases up until
+the middle of the 20th century, but instead have the lowest mortality from
+1980 onwards. Working class men had a fairly low cardiovascular mortality,
+but in the same way as among women, the levels were fairly similar to the
+one of the middle category consisting of lower manager/farmers. And in the
+same way as for women at this age, working class men have the highest
+mortality from this cause of death during the last decades. A social
+hierarchy in cardiovascular mortality seems to have developed both for men
+and women.
+
+### The effect of income and education, 1990--2005
+
+### Ages 65--89
+
 <<41getske6089>>=
@@
 
-In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years. We use as a summary measure the value of the cumulative hazards at
+In this section we focus on cardiovascular mortality among those that
+mostly have left the workforce, the age group 65--89 years. We use as a
+summary measure the value of the cumulative hazards at 
 duration 25 (age 90), starting from age 65.
-
-%\subsubsection{Women}
 
@@
 
-Concerning the social pattern, we find somewhat different results from that of the younger age group of women.
-Cardiovascular mortality among elderly women do not present a consistent class pattern.
-Female cardiovascular mortality was somewhat higher among the elite and the middle group than among working class women until the last decades.
-Elderly from a working class background were not disadvantaged for most of the 20th century.
-The social class pattern has changed somewhat during the last periods and more in line to what is expected.
-The differences are however rather small.
-
-Even if results for elderly men point in different directions for some periods, the general trend resembles that of working-active men.
-The highest class had high cardiovascular mortality during the first half of the 20th century, while the expected social pattern has developed only during the most recent periods.
+Concerning the social pattern, we find somewhat different results from that
+of the younger age group of women. Cardiovascular mortality among elderly
+women do not present a consistent class pattern. Female cardiovascular
+mortality was somewhat higher among the elite and the middle group than
+among working class women until the last decades. Elderly from a working
+class background were not disadvantaged for most of the 20th century. The
+social class pattern has changed somewhat during the last periods and more
+in line to what is expected. The differences are however rather small.
+
+Even if results for elderly men point in different directions for some
+periods, the general trend resembles that of working-active men. The
+highest class had high cardiovascular mortality during the first half of
+the 20th century, while the expected social pattern has developed only
+during the most recent periods.
 
@@
 
-% \subsubsection{The effect of income and education on cardiovascular death, 1990--2005}
-% 
-% The analysis of cardiovascular mortality for the period 1990--2005 is in this section based on a separate analysis with
-% income and education as explanatory variables.
-The statistical significance of the covariates are shown in Table~\ref{tab:41liswom2} for women and Table~\ref{tab:41lismen2} for men.
-It confirms that most of the variables are strongly significant except
-% 
-% <<41lissyoung2>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss6589 <- loTab[loTab$age > 64.5, ]
-% @
-% 
-% 
-% 
-% <<41liswom2, results = 'asis'>>=
-% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-% ##cat("Women:\n")
-% drw2 <- drop1(fitw2, test = "Chisq")
-% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cardiovascular death.", label = "tab:41liswom2")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% 
-% <<41lismen2, results='asis'>>=
-% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm2 <- drop1(fitm2, test = "Chisq")
-% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cardiovascular death.", label = "tab:41lismen2")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:liswomfig}
-% 
-% <<41liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitw2)
-% @
-% 
-% 
-% And for men:
-% 
-% 
-% 
-% <<41lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cardiovascular death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm2)
-% @
-% 
-% The pattern is similar to what we find for all-cause mortality.
-The higher educational level, the lower cardiovascular mortality, and more income leads to better survival.
-This association is clear and consistent for women as well as for men.
-A minor difference between men and women is that mortality in the higher levels of education do not lead to lower mortality among women while that is the case among men.
-
-\subsection{Cancer mortality}
-
-While we assume clear social differences in cardiovascular diseases, this is not necessarily an obvious assumption when it comes to cancers.
-Different forms of cancers have very different origin and different pathways to disease and death.
-The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers.
-In our analyses, there is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated by the already mentioned large proportion of unknown or vague causes at that time.
-We believe however that cancers were better diagnosed than cardiovascular deaths.
-This is supported by the stability in proportion of cancer deaths between the two first periods, 12.5 per cent 1901--1925 and 13.5 per cent for the period 1926--1950.
+## Cancer mortality
+
+While we assume clear social differences in cardiovascular diseases, this
+is not necessarily an obvious assumption when it comes to
+cancers. Different forms of cancers have very different origin and
+different pathways to disease and death. The social pattern of higher
+mortality in classes with less power and accordingly less life chances are
+considered to be less clear when it comes to cancers. In our analyses,
+there is a risk that cancer mortality is under-estimated during the first
+half of the 20th century, illustrated by the already mentioned large
+proportion of unknown or vague causes at that time. We believe however that
+cancers were better diagnosed than cardiovascular deaths. This is supported
+by the stability in proportion of cancer deaths between the two first
+periods, 12.5 per cent 1901--1925 and 13.5 per cent for the period
+1926--1950.
 
@@
 
-%creating an \emph{heterogeneity index} (a better name is needed:
-%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
-%Something catchy.).
-
-\subsubsection{Ages 40--64}
-
-%The distribution over time periods.
+### Ages 40--64
+
 <<12getske4060>>=
@@
 
-%%We will use as a summary measure the value of the cumulative hazards at
-%%age 65, given survival to age 40.
-
 <<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
@@
 
-%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
-%%@
-
 <<12cumswtab,results='asis'>>=
@@
 
-Figure~\ref{fig:12cumsw} shows relative mortality levels according to {\sc Hisclass} during the different studied periods for the age group 40--64.
-Cancer mortality among the élite women was lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods.
-Consequently, we find a more consistent advantage of higher class in cancers compared to cardiovascular diseases for women.
-We do however not find the same pattern for men.
-It is difficult to identify any general features of 20th century male cancer mortality when it comes to social differences.
-
-%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH
-
-% \subsubsection{The effect of income and education on cancer mortality, 1990--2005}
-% 
-% In the previous section, we concluded that there were no social inequality in cancer mortality according to hisclass.
-The question here is if the lacking socio-economic differentiation holds also when using education and income as categorising variables.
-The analysis for the period 1990--2005 is based on a separate analysis with income and education as explanatory variables.
-% 
-% <<12lissyoung>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss4064 <- loTab[loTab$age < 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom} for women and Table~\ref{tab:12lismen} for men. The variables of interest here are all significant.
-% <<12liswom, results = 'asis'>>=
-% fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
-% ##cat("Women:\n")
-% drw <- drop1(fitw1, test = "Chisq")
-% xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005, cancer death.", label = "tab:12liswom")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% <<12lismen, results='asis'>>=
-% fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm <- drop1(fitm1, test = "Chisq")
-% xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005, cancer death.", label = "tab:12lismen")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig}.
-% 
-% <<12liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-% source("R/plotWEduc.R")
-% plotWEduc(fitw1)
-% @
-% 
-% 
-% And for men:
-% 
-% 
-% <<12lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005, cancer death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm1)
-% @
-% 
-% 
-% <<12plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013, cancer death.", fig.height = 4>>=
-% #oldpar <- par(mfrow = c(1, 2))
-% #plot(fitw1, fn = "cum", col = dcol, main = "Women")
-% #plot(fitm1, fn = "cum", col = dcol, main = "Men")
-% #par(oldpar)
-% ##summary(fitw1)
-% ##summary(fitm1)
-% @
-% 
-% %See Table~\ref{tab:tablisswom1} the PH regression for women.
-% <<12tablisswom1, results = "asis">>=
-% #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "female", ])
-% #dr <- drop1(fitwo1, test = "Chisq")
-% #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
-% #    label = "tab:tablisswom1")
-% @
-% 
-% %See Table~\ref{tab:tablissman1} the PH regression for men.
-% <<12tablissman1, results = "asis">>=
-% #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
-% #                 data = liss4064[liss4064$sex == "male", ])
-% #dr <- drop1(fitma1, test = "Chisq")
-% #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
-%  #   label = "tab:tablissman1")
-% @
-% 
-% When it comes to the association with education and income, we find the similar pattern as those for all-cause mortality and cardiovascular mortality.
-There is a strong effect for both men and women of having more education and income.
-These variables seem to catch up important aspect of socio-economic differences in health and mortality.
-
-\subsubsection{Ages 65--89}
-
-%The distribution over time periods.
+Figure~\ref{fig:12cumsw} shows relative mortality levels according to {\sc
+Hisclass} during the different studied periods for the age group
+40--64. Cancer mortality among the élite women was lower than that for the
+other classes during the complete studied period, while working class women
+had the highest levels during almost all periods. Consequently, we find a
+more consistent advantage of higher class in cancers compared to
+cardiovascular diseases for women. We do however not find the same pattern
+for men. It is difficult to identify any general features of 20th century
+male cancer mortality when it comes to social differences.
+
+### The effect of income and education on cancer mortality, 1990--2005
+
+### Ages 65--89
+
 <<12getske6089>>=
@@
 
-In our analyses of class patterns in mortality, we finally turn to cancer mortality among the elderly.
-
-%\subsubsection{Women}
+In our analyses of class patterns in mortality, we finally turn to cancer
+mortality among the elderly.
 
@@
 
-The relation between different classes and cancer mortality fluctuates considerably both for women and men in this age group.
-For some periods it has to do with quite few events to analyse.
-Women of the highest class have favourable mortality during several periods, consistent with the advantages we found for women in the younger age group, but on the other hand had high levels at other times.
-The social differences are however quite small during the most recent periods.
-
-For men we find a similar lack of social pattern in cancer mortality as we found for men in working-active ages.
-There is however a tendency during the last decades of higher mortality among working class men.
+The relation between different classes and cancer mortality fluctuates
+considerably both for women and men in this age group. For some periods it
+has to do with quite few events to analyse. Women of the highest class have
+favourable mortality during several periods, consistent with the advantages
+we found for women in the younger age group, but on the other hand had high
+levels at other times. The social differences are however quite small
+during the most recent periods.
+
+For men we find a similar lack of social pattern in cancer mortality as we
+found for men in working-active ages. There is however a tendency during
+the last decades of higher mortality among working class men.
 
@@
 See Table~\ref{tab:12cumswoldtab}.
-% \subsubsection{The effect of income and education on cancer death, 1990--2005}
-% 
-% Thus, we find indications of social differences in the later decades, particularly between the elite and the workers.
-The alternative analysis for the last period is based on a separate analysis with
-% income and education as explanatory variables.
-% 
-% <<12lissyoung2>>=
-% library(eha)
-% library(xtable)
-% ##load("data/loTab.rda")
-% ##loTab$period <- factor(loTab$period)
-% liss6589 <- loTab[loTab$age > 64.5, ]
-% @
-% 
-% 
-% The statistical significance of the covariates are shown in Table~\ref{tab:12liswom2} for women and Table~\ref{tab:12lismen2} for men.
-% <<12liswom2, results = 'asis'>>=
-% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
-% ##cat("Women:\n")
-% drw2 <- drop1(fitw2, test = "Chisq")
-% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005, cancer death.", label = "tab:12liswom2")
-% ##round(summary(fitw1)$coef[-(1:13), ], 4)
-% @
-% 
-% <<12lismen2, results='asis'>>=
-% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
-% ##cat("\nMen:\n")
-% drm2 <- drop1(fitm2, test = "Chisq")
-% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.", label = "tab:12lismen2")
-% ##round(summary(fitm1)$coef[-(1:13), ], 4)
-% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% @
-% 
-% The estimated effect sizes are shown in Figure \ref{fig:12liswomfig2}
-% 
-% <<12liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitw2)
-% @
-% 
-% 
-% And for men:
-% 
-% %% <<12lismen2, results='asis'>>=
-% %% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = ##poisson, data = liss6589[liss6589$sex == "male", ])
-% %% ##cat("\nMen:\n")
-% %% drm2 <- drop1(fitm2, test = "Chisq")
-% %% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005, cancer death.")
-% %% round(summary(fitm1)$coef[-(1:13), ], 4)
-% %% dcol <- c("blue", "darkgreen", "darkgreen", "red")
-% %% @
-% 
-% <<12lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005, cancer death.", fig.height = 4>>=
-% #source("R/plotWEduc.R")
-% plotWEduc(fitm2)
-% @
-% 
-% The social inequality in cancer mortality is confirmed by the approach of analysing the effect of education and income.
-We find the same consistent pattern as we have found for all-cause mortality and cardiovascular mortality for this period.
-Higher levels of education and more income is associated with lower cancer mortality.
-We can therefore conclude that both education and income are strongly decisive aspects of social differentiation when it comes to mortality during the last decades.
-
-%%{\huge Här ska 'conclusion.tex' in !!!}
-
-\section{The effect of Education, 1990--2013}
-
-%\subsection{All causes of death}
-
-A common alternative to a class analysis is to use income or education for investigating socio-economic health differences.
-\citet{goldk10} criticize how many economists and epidemiologists operationalize inequality in their studies, for example by not distinguishing between class and status aspects.
-Class refers to relational aspects in society, while many other variables focus on the distribution of material and/or immaterial resources.
-He advocates studies of different aspects, which together complement each other.
-Comparing different approaches are valuable in order to understand why some are advantaged and others disadvantaged and what sort of factors determines health differences.
-In this and the following section, we introduce analyses of education and income.
-In that way, we can compare the class analyses with these alternative categorizations.
-For the time being, we have to restrict such a comparative analysis to the last decades.
-
-In this section we analyse differences according to education level for the period 1990--2013 for the whole population 40 years and older.
-Education is a good indicator of both human and cultural capital, but aer also strongly associated with income and social class.
-We present analyses on all cause mortality (survival curves) as well as on the cause of death groups of cardiovascular diseases and cancers (cumulative hazards).
-The pattern is clear in total mortality as well as cause specific mortality, and this is valid for both men and women, see Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart} and~\ref{fig:ploteduccancer}.
-Education had a strong, statistically significant, effect on survival, as expected.
-Those with the lowest education levels hade the highest mortality, while those with highest level of education had by far the best survival.
-The differences must be considered large, and it follows a continuous improvement according to increasing levels.
+
+# The effect of Education, 1990--2013
+
+A common alternative to a class analysis is to use income or education for
+investigating socio-economic health differences. \citet{goldk10} criticize
+how many economists and epidemiologists operationalize inequality in their
+studies, for example by not distinguishing between class and status
+aspects. Class refers to relational aspects in society, while many other
+variables focus on the distribution of material and/or immaterial
+resources. He advocates studies of different aspects, which together
+complement each other. Comparing different approaches are valuable in order
+to understand why some are advantaged and others disadvantaged and what
+sort of factors determines health differences. In this and the following
+section, we introduce analyses of education and income. In that way, we can
+compare the class analyses with these alternative categorizations. For the
+time being, we have to restrict such a comparative analysis to the last
+decades.
+
+In this section we analyse differences according to education level for the
+period 1990--2013 for the whole population 40 years and older. Education is
+a good indicator of both human and cultural capital, but aer also strongly
+associated with income and social class. We present analyses on all cause
+mortality (survival curves) as well as on the cause of death groups of
+cardiovascular diseases and cancers (cumulative hazards). The pattern is
+clear in total mortality as well as cause specific mortality, and this is
+valid for both men and women, see
+Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart}
+and~\ref{fig:ploteduccancer}. Education had a strong, statistically
+significant, effect on survival, as expected. Those with the lowest
+education levels hade the highest mortality, while those with highest level
+of education had by far the best survival. The differences must be
+considered large, and it follows a continuous improvement according to
+increasing levels.
 
@@
 <<educ>>=
-
 educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
@@
 
-%\subsection{Cardiovascular death}
-
 <<cardeduc>>=
@@
 @
-
-% Illustrating differences for cause-specific mortality is conveniently done by comparing 
-% cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
-% cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
-% mortality.
 
@@
 
-%\subsection{Cancer death}
-
 <<canceduc>>=
@@
 @
-
-% Although mortality in cancer is lower than mortality in cardiovascular diseases,
-% the pattern is the same: The more education, the less mortality. See 
-% Figure~\ref{fig:ploteduccancer}.
 
@@
 
-\section{The effect of Income, 1986--2005}
-
-%\subsection{All causes of death}
+# The effect of Income, 1986--2005
 
@@
 
-Finally, we present analyses of different income levels for the age group 40 and above and the period 1986--2005.
-Disposable individual income must be considered as a robust measure of economic condition, and increases the possibility to allocate these economic resources towards health promoting living conditions and behaviours.
-We categorize the income according to quartiles.
-As expected, we find that higher incomes result in lower mortality, both when it comes to total mortality as to deaths from cardiovascular diseases, see Figures~\ref{fig:plotincome} and~\ref{fig:plotincheart}.
-This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden.
-We also find indications of a similar difference for cancers but less clear, particularly for male cancer mortality, see Figure~\ref{fig:plotinccancer}.
-Notice however that we cannot establish a causal relationship between income and mortality.
-Bad health may be a confounder, causing both low incomes and higher death risks.
-
-%\subsection{Cardiovascular death}
+Finally, we present analyses of different income levels for the age group
+40 and above and the period 1986--2005. Disposable individual income must
+be considered as a robust measure of economic condition, and increases the
+possibility to allocate these economic resources towards health promoting
+living conditions and behaviours. We categorize the income according to
+quartiles. As expected, we find that higher incomes result in lower
+mortality, both when it comes to total mortality as to deaths from
+cardiovascular diseases, see Figures~\ref{fig:plotincome}
+and~\ref{fig:plotincheart}. This is also what could be expected, and
+conforms to what we know of social health differences in present-day
+Sweden. We also find indications of a similar difference for cancers but
+less clear, particularly for male cancer mortality, see
+Figure~\ref{fig:plotinccancer}. Notice however that we cannot establish a
+causal relationship between income and mortality. Bad health may be a
+confounder, causing both low incomes and higher death risks.
 
@@
 
-%The overall pattern is the same for cardiovascular mortality, high income pays off.
 <<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005.">>=
@@
 
-%\subsection{Cancer death}
-
 <<cancerincome>>=
@@
 @
-
-% The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.
 
@@
 The authors of this paper have scrutinised different aspects of the social
-pattern of mortality in historical northern Sweden for some time, focusing on
-different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}. When it comes to infant and child mortality pre---and 
-early---transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}.
-Studies based on microdata on the adult population in history have been rare, mainly due to lack of adequate data.
-The results from such studies show other patterns than that of infant and child mortality.
-The expected social gradient in
-survival, meaning that people from higher social classes always had an
-advantage, was in many cases, particularly among adult men, not found
-despite the well-documented socio-economic differences \citep{edli11, edbr12}.
-There were however large differences between men and women and for different age groups.
-These earlier studies have suggested that distinct and consistent differences to the advantage of higher classes developed somewhere during and probably quite early in the 20th century \citep{bevp11}.
-Until now we have however not been able to investigate the continuous development on mortality differences among the adult and elderly
-population due to no available data.
-This is the first time it has been possible, and the results continue to surprise us.
-The consistent advantage for higher social classes do not appear until quite late in the regions we have analysed.
-The social pattern of female mortality are in line to what is
-expected, but a social gradient in male adult mortality seems to be a recent phenomenon.
-Furthermore, this transition and the class pattern of mortality differ depending on age group.
-Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly.
-
-The results of small socio-economic gradient in the past and its late development are however in accordance with other studies \citep{bevp11}, for example by \citet{bedr11} in a very different setting, a couple of Scanian parishes, although with somewhat different social categories.
-They found the first indications of an emerging social gradient in the 1950s.
-Similar results have also been found in Estonia \citep{Jaadla17}.
-
-When it comes to specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
-Cancer is sometimes considered less connected to social class, and we find mixed results in our analyses.
-
-The surprising results makes it reasonable to consider the
-validity and possible weaknesses in the study.
-One obstacle concerns the quality of available information.
-Occupational titles in historical sources are not always distinct enough to define the correct class.
-It can be difficult to identify the skill level in the working class population because many
-were often only designated as labourer. The size of the enterprise for free
+pattern of mortality in historical northern Sweden for some time, focusing
+on different regions, different age groups and different time periods
+\citep{edv92, edv04, edli11, edbr12}. When it comes to infant and child
+mortality pre---and early---transitional time, the expected socio-economic
+differences was often but not always confirmed \citep{edv92, edv04},
+something that has been found in other studies as well \citep{bredro02,
+bube01, mol17}. Studies based on microdata on the adult population in
+history have been rare, mainly due to lack of adequate data. The results
+from such studies show other patterns than that of infant and child
+mortality. The expected social gradient in survival, meaning that people
+from higher social classes always had an advantage, was in many cases,
+particularly among adult men, not found despite the well-documented
+socio-economic differences \citep{edli11, edbr12}. There were however large
+differences between men and women and for different age groups. These
+earlier studies have suggested that distinct and consistent differences to
+the advantage of higher classes developed somewhere during and probably
+quite early in the 20th century \citep{bevp11}. Until now we have however
+not been able to investigate the continuous development on mortality
+differences among the adult and elderly population due to no available
+data. This is the first time it has been possible, and the results continue
+to surprise us. The consistent advantage for higher social classes do not
+appear until quite late in the regions we have analysed. The social pattern
+of female mortality are in line to what is expected, but a social gradient
+in male adult mortality seems to be a recent phenomenon. Furthermore, this
+transition and the class pattern of mortality differ depending on age
+group. Inequality in mortality appear clearer for the population at working
+age, while the differences are smaller among the elderly.
+
+The results of small socio-economic gradient in the past and its late
+development are however in accordance with other studies \citep{bevp11},
+for example by \citet{bedr11} in a very different setting, a couple of
+Scanian parishes, although with somewhat different social categories. They
+found the first indications of an emerging social gradient in the
+1950s. Similar results have also been found in Estonia \citep{Jaadla17}.
+
+When it comes to specific causes of death, the analysis shows that
+cardiovascular diseases contributes to a large proportion of all deaths
+during the 20th century, and we are not surprised that it has a fairly
+similar pattern as for all-cause mortality. Cancer is sometimes considered
+less connected to social class, and we find mixed results in our analyses.
+
+The surprising results makes it reasonable to consider the validity and
+possible weaknesses in the study. One obstacle concerns the quality of
+available information. Occupational titles in historical sources are not
+always distinct enough to define the correct class. It can be difficult to
+identify the skill level in the working class population because many were
+often only designated as labourer. The size of the enterprise for free 
 entrepreneurs is unknown and it may be impossible to distinguish if they
@@
 not represent the same thing. Nevertheless, these ambiguities and possible
-mis-classifications are not of that sort and magnitude that it would radically change
-the basic patterns observed. The social classes identified during the
-different periods represent distinct groups with very large differences in
-resources, capacities and power. Available data, for example from the 1930 census confirm that class differences in income and wealth were large, often enormous.
+mis-classifications are not of that sort and magnitude that it would
+radically change the basic patterns observed. The social classes identified
+during the different periods represent distinct groups with very large
+differences in resources, capacities and power. Available data, for example
+from the 1930 census confirm that class differences in income and wealth
+were large, often enormous.
 
@@
 such analyses gives other results or at least would modify the image
-presented here. In many modern studies categorisations based on income, wealth
-or education are common for studies of inequality. They represent different dimensions of 
-inequality. One problem is however that reliable data on income and
-education are rarely available in historical population databases. 
-As a test, we have included analyses of the associations of
+presented here. In many modern studies categorisations based on income,
+wealth or education are common for studies of inequality. They represent
+different dimensions of inequality. One problem is however that reliable
+data on income and education are rarely available in historical population
+databases.  As a test, we have included analyses of the associations of 
 education and income to mortality for recent decades. For women as
@@
 cause-specific mortality we find consistent differences with higher
-educational level and more income leading to lower mortality, thus basically confirming what we find in the class analysis. We do not know if education and income are equally important in the past.
-
-The WHO Commission on Social Determinants of Health framework for the social determinants of health provides a model on how socio-economic position influence health.
-Social class works through intermediate determinants related to exposure and vulnerability \citep{who10}.
-People are situated within socio-economic and political contexts, as well as prevailing value systems in society.
-Socio-economic position imply different access to resources, power and prestige, all being aspects of crucial importance.
-The framework takes a holistic approach to the social determinants, including not only the effect of pure economic and material resources.
-The importance of understanding people in their context is also emphasised by for example \citet{kri05}.
-What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-The WHO framework
-\citep{who10} lists three different types of intermediary social determinants of health; material circumstances, behaviours, biological factors, and finally psycho-social factors.
-Most studies focus on material circumstances.
-Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions.
-Even though these advantages were at hand in our studied regions, the final result was not better health for men in the most privileged groups until the late 20th century.
-We must therefore consider other explanations.
-Psycho-social factors have been frequently discussed during the most recent decades, for example by \citet{mar04}, who argues that a higher status leads to better health, regardless of where in the status ladder we make our observations.
-Psychosocial factors are also suggested by \citet{wipi09} as an explanation why equal societies always perform better than unequal ones when it comes to health and other social conditions.
-In the case studied here, we do not find any indication of a status syndrome, since mortality levels was in the opposite directions to the one we would expect from these theories.
-
-This makes us consider the possible impact of behavioural factors.
-Our results of the gender aspect might be of interest in this case.
-When it comes to gender differences, the present study confirms what we previously have found in the Sundsvall region \citep{edli11}.
-There is a clear gendered pattern when it comes to social inequality in mortality.
-Within the same households, the effect of social class seems to lead to different results, high social position being beneficial for women but not for men.
-We might consider the role of behaviours based on different gendered expectations.
-Having a high social status could be expressed in attitudes and behaviours that were harmful for health (this is discussed in more detail in \citet{edli11}.
-Alcohol has been suggested as part of the large gender differences in the 19th century \citep{wil99, edv92}, and a more risk-taking life may be a component of male mortality \citep{cour00}.
-There are however some problems with attributing most of the male excess mortality to heavy drinking.
-Alcohol consumption decreased substantially during the period late 19th century to the 1950s, but the old pattern in social health inequalities remained.
-Swedes at that time had very low consumption levels.
-Alternative explanations should also be considered.
-There
-were strong and heavy requirements of being successful for the male \"elite. Many were involved in businesses with high risks and they may have felt a constant threat of economic failure.
+educational level and more income leading to lower mortality, thus
+basically confirming what we find in the class analysis. We do not know if
+education and income are equally important in the past.
+
+The WHO Commission on Social Determinants of Health framework for the
+social determinants of health provides a model on how socio-economic
+position influence health. Social class works through intermediate
+determinants related to exposure and vulnerability \citep{who10}. People
+are situated within socio-economic and political contexts, as well as
+prevailing value systems in society. Socio-economic position imply
+different access to resources, power and prestige, all being aspects of
+crucial importance. The framework takes a holistic approach to the social
+determinants, including not only the effect of pure economic and material
+resources. The importance of understanding people in their context is also
+emphasised by for example \citet{kri05}. What she calls embodiment refers
+to the recognition that human beings are simultaneously social beings and
+biological organisms. The WHO framework  \citep{who10} lists three
+different types of intermediary social determinants of health; material
+circumstances, behaviours, biological factors, and finally psycho-social
+factors. Most studies focus on material circumstances. Having a good
+economy gives advantages in many aspects of relevance for health and
+mortality, for example good living conditions, access to food, medical care
+and better working conditions. Even though these advantages were at hand in
+our studied regions, the final result was not better health for men in the
+most privileged groups until the late 20th century. We must therefore
+consider other explanations. Psycho-social factors have been frequently
+discussed during the most recent decades, for example by \citet{mar04}, who
+argues that a higher status leads to better health, regardless of where in
+the status ladder we make our observations. Psychosocial factors are also
+suggested by \citet{wipi09} as an explanation why equal societies always
+perform better than unequal ones when it comes to health and other social
+conditions. In the case studied here, we do not find any indication of a
+status syndrome, since mortality levels was in the opposite directions to
+the one we would expect from these theories.
+
+This makes us consider the possible impact of behavioural factors. Our
+results of the gender aspect might be of interest in this case. When it
+comes to gender differences, the present study confirms what we previously
+have found in the Sundsvall region \citep{edli11}. There is a clear
+gendered pattern when it comes to social inequality in mortality.  Within
+the same households, the effect of social class seems to lead to different
+results, high social position being beneficial for women but not for
+men. We might consider the role of behaviours based on different gendered
+expectations. Having a high social status could be expressed in attitudes
+and behaviours that were harmful for health (this is discussed in more
+detail in \citet{edli11}. Alcohol has been suggested as part of the large
+gender differences in the 19th century \citep{wil99, edv92}, and a more
+risk-taking life may be a component of male mortality \citep{cour00}. There
+are however some problems with attributing most of the male excess
+mortality to heavy drinking. Alcohol consumption decreased substantially
+during the period late 19th century to the 1950s, but the old pattern in
+social health inequalities remained. Swedes at that time had very low
+consumption levels. Alternative explanations should also be
+considered. There were strong and heavy requirements of being successful
+for the male \"elite. Many were involved in businesses with high risks and
+they may have felt a constant threat of economic failure.
 
@@
 explanation for the larger social health inequalities in present-day
-society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive
-eating, drinking, smoking and sometimes also taking risks.
-Different historical contexts promote different ways to spend economic resources.This is also in line with what \citet{rasp06} and \citet{Jaadla17} suggest.
-If this interpretation is correct, the hypothesis of \citet{ant67}on an increasing role of health related behaviour in the modern low-mortality régime,while economic inequality determined the social mortality pattern during the early phases of the mortality transition.
-Instead, health-related seems to be at least equally strong earlier in history but having a completely different expression.
-
-One of the major findings in this study relate to gender.
-As several researchers have pointed out, demography has often ignored power relations and different roles within families \citep{watkins93, mack95, jans07}.
-By investigating possible differences in mortality patterns according to sex, we have identified gendered patterns within the family.
-Different patterns between the sexes may be caused by discrimination on the female part in the household.
-We know that the society was
-strongly segregated according to gender and that women were discriminated
-and had less space for action.
-In our case, we do not find this to have negative consequences for women, indicating that the relation between sexes in the society was characterised by distinct gendered expectations on behaviour.
-
-Another question is if our results can be generalised to other historical contexts. We have studied
-environments in northern Sweden without any large metropolis. Perhaps
-Sweden and/or this part of Sweden is a special case. The social differences
-may have been smaller, the involvement of government both at the national
+society is that high status is expressed through health, being fit and
+physically active, while in many historical contexts higher social status
+(for men) was expressed in low physical activity, obsessive eating,
+drinking, smoking and sometimes also taking risks.  Different historical
+contexts promote different ways to spend economic resources.This is also in
+line with what \citet{rasp06} and \citet{Jaadla17} suggest. If this
+interpretation is correct, the hypothesis of \citet{ant67}on an increasing
+role of health related behaviour in the modern low-mortality régime,while
+economic inequality determined the social mortality pattern during the
+early phases of the mortality transition. Instead, health-related seems to
+be at least equally strong earlier in history but having a completely
+different expression.
+
+One of the major findings in this study relate to gender. As several
+researchers have pointed out, demography has often ignored power relations
+and different roles within families \citep{watkins93, mack95, jans07}. By
+investigating possible differences in mortality patterns according to sex,
+we have identified gendered patterns within the family. Different patterns
+between the sexes may be caused by discrimination on the female part in the
+household. We know that the society was strongly segregated according to
+gender and that women were discriminated and had less space for action. In
+our case, we do not find this to have negative consequences for women,
+indicating that the relation between sexes in the society was characterised
+by distinct gendered expectations on behaviour.
+
+Another question is if our results can be generalised to other historical
+contexts. We have studied environments in northern Sweden without any large
+metropolis. Perhaps Sweden and/or this part of Sweden is a special
+case. The social differences may have been smaller, the involvement of
+government both at the national 
 and local level may have been stronger, thus mitigating socio-economic
-disadvantages. In comparison with many other countries, the level of mortality has been fairly low. These are relevant questions, which hopefully can be
-addressed in other and maybe comparative studies.
+disadvantages. In comparison with many other countries, the level of
+mortality has been fairly low. These are relevant questions, which
+hopefully can be addressed in other and maybe comparative studies.
 
 Finally, we argue that our results have implications for the understanding
-on several other hot research topics. A central and hot debated question concerning the mortality
-transition is how the enormous increase in survival was accomplished. \citet{mck76} argues that the
-improvements up to the middle of the 20th century was primarily caused by improved nutrition. This implicate that having more resources would lead to better survival. If research shows that the most
-advantaged groups did not have better health and longer lives, this do not
-necessarily speak against McKeown's ideas but it makes the question more
-complex. The results are also difficult to fit into \citet{liph95}
-hypothesis that social conditions should be seen as a fundamental cause of
-disease and not only expressions of different disadvantages. In the historical
-context analysed here, this idea is not supported.
+on several other hot research topics. A central and hot debated question
+concerning the mortality transition is how the enormous increase in
+survival was accomplished. \citet{mck76} argues that the improvements up to
+the middle of the 20th century was primarily caused by improved
+nutrition. This implicate that having more resources would lead to better
+survival. If research shows that the most advantaged groups did not have
+better health and longer lives, this do not necessarily speak against
+McKeown's ideas but it makes the question more complex. The results are
+also difficult to fit into \citet{liph95} hypothesis that social conditions
+should be seen as a fundamental cause of disease and not only expressions
+of different disadvantages. In the historical context analysed here, this
+idea is not supported.
 
 In recent years, the question about the relation between health and income
-inequality been widely discussed \citep{wipi09}. Even though our study do not address this
-issue directly, our results are of interest for this question as well. The
+inequality been widely discussed \citep{wipi09}. Even though our study do
+not address this issue directly, our results are of interest for this
+question as well. The 
 development of the social inequality in mortality do not seem to fit that
@@
 
-% \section{Conclusion}
-% 
-% We started this paper by raising the issue of the development of social
-% inequalities. When studying this question in history we often get surprised
-% of its complexity. In a society characterised by large economic inequality,
-% we would expect that this would result in corresponding mortality
-% differences, i.e. higher social classes having an advantage. We find the
-% appearance of a consistent social pattern in mortality, to the advantage of
-% the higher classes, but surprisingly late in our history. It is only in the
-% recent decades that this pattern dominates. Improved survival characterised
-% however all HISCLASSes. We can however confirm that there are gender
-% differences in the social pattern. While there is no clear social gradient
-% among men, the results at least partly indicate advantages for the higher
-% social classes for women already early in the 20th century. Females had
-% more to gain from a high social position. Previous studies of the mortality
-% transition have made similar observations. Among adults, gender has often
-% been of more significant importance than social class. The reason for this
-% difference and for the late appearance of the modern social pattern among
-% men is still an enigma that require further studies, even though we have
-% suggested the possible importance of behavioural factors and changes in
-% ideals connected to different social classes. This should stimulate us to
-% further disentangle the pathways for how social position influence health
-% and mortality in different societies.
-Furthermore, we do not find any substantial differences or distinct pattern in the relation between class and the two major cause-of-death groups.
-Finally, we find stronger differences in the social pattern among the working-age population, while the associations were smaller or not so clear among the elderly.
-However, when analysing the association with education and income in the period 1990--2005 we find consistent patterns of higher income and higher education level being associated with lower mortality, indicating that these variables are better to catch socio-economic differences at least in the present-day society.
-
-\section{Acknowledgements}
-
-This work was undertaken within the research program \emph{Ageing and Living Conditions} at the Centre for Demographic and Ageing Research (CEDAR) at Umeå University, financed by the Swedish Research Council (grant No 349-2008-6592) and the project \emph{Old age, health and inequality: Mortality from a life course perspective in Sweden 1986--2009}, financed by the Swedish Research Council (grant No 421-2012-892).
-We thank the Demographic Data Base, Umeå University, for giving us access to data and Jonas Helgertz, Department of Economic History, Lund University, for providing us with scripts for creating {\sc Hisclass} from Swedish Census data.
+# Acknowledgements}
+
+This work was undertaken within the research program \emph{Ageing and
+Living Conditions} at the Centre for Demographic and Ageing Research
+(CEDAR) at Umeå University, financed by the Swedish Research Council (grant
+No 349-2008-6592) and the project \emph{Old age, health and inequality:
+Mortality from a life course perspective in Sweden 1986--2009}, financed by
+the Swedish Research Council (grant No 421-2012-892). We thank the
+Demographic Data Base, Umeå University, for giving us access to data and
+Jonas Helgertz, Department of Economic History, Lund University, for
+providing us with scripts for creating {\sc Hisclass} from Swedish Census
+data.
 
@@
 \end{document}
-@

### 2017-07-03 R/getPHglm2.R
add debug 'cat'

@@ -27,6 +27,7 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
         inter <- matrix(co[interact], ncol = k - 1, byrow = TRUE)
         tmp <- matrix(0, nrow = m, ncol = k)
         for (j in 1:m){
+            cat("j = ", j, "\n")
             tmp[j, ] <- c(co[1], co[1] + amain)
         }
         for (j in 2:m){

### 2017-07-03 article.Rmd
Rmarkdownifying .Rnw more

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+library(knitr)
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
+library(knitr)
+library(eha)
+library(xtable)
+library(skum)
+```
+
+```{r readalldata}
+## Load remote data:
+##con <- url("http://capa.ddb.umu.se/data/vb0.rda")
+load("data/vb0.rda")
+##load(con)
+##close(con)
+vb0$hisclass[is.na(vb0$hisclass)] <- 4
+vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
+##
+##con <- url("http://capa.ddb.umu.se/data/vb.rda")
+load("data/vb.rda")
+##load(con)
+##close(con)
+##
+##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+load("data/lisaold.rda")
+##load(con)
+##close(con)
+##
+##con <- url("http://capa.ddb.umu.se/data/loTab.rda")
+##load(con)
+##close(con)
+load("data/loTab.rda")
+loTab$period <- factor(loTab$period)
+##
+vb0Tab <- aggregate(vb0[, c("event", "exposure")],
+                   vb0[, c("sex", "age", "period", "cohort",
+                          "urban", "civst", "hisclass")],
+                   FUN = sum)
+vb0Tab$age <- factor(vb0Tab$age)
 ```
@@
 
-<<lexis2,fig.cap="The sampling frame.", fig.height = 5>>=
+```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
 plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
@@
 text(1985.2, 50, paste(n.d, "deaths"))
-@
+```
 
@@
 
-<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+```{r hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
 xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+```
+
+```{r hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
 xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-@
-
-% <<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+```
+
+% ```{r hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
 % xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 % barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-% @
+% ```
 %
-% <<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
+% ```{r hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
 % xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 % barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-% @
+% ```
 
@@
 
-<<civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5>>=
+```{r civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5}
 xx <- with(vb0Tab, tapply(exposure, list(period = period, civst = civst), sum))
@@
 barplot(t(yy)[3:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
-@
+```
 
@@
 
-<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5>>=
+```{r causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5}
 vb01 <- vb0[vb0$event == 1, ]
@@
 barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = "Per cent")
-@
+```
 
@@
 
-<<colorc>>=
+```{r colorc}
 ## Colors for plotting strata:
@@
 ## End 'common'
-@
+```
 
@@
 
-<<getske4060>>=
+```{r getske4060}
 cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
@@
 ##vb4064$hisclass <- vb4064$hisclass40
-@
-
-<<hazcalw>>=
+```
+
+```{r hazcalw}
 source("R/getPHglm2.R")
@@
             nper, labb)
-@
-
-<<cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8>>=
+```
+
+```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8}
 source("R/plotED.glm.R")
@@
 par(oldpar)
-@
+```
 
@@
 
-<<cumswtab,results='asis',cap="Table">>=
+```{r cumswtab,results='asis',cap="Table"}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
@@
 xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, 40-65.")
-@
+```
 
@@
 
-<<cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
+```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8}
 oldpar <- par(mfrow = c(2, 1))
@@
 
-<<getske6089>>=
+```{r getske6089}
 vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
@@
 ##vb6589$hisclass <- vb6589$hisclass60
-@
+```
 
@@
 
-<<cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
+```{r cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8}
 ##source("R/plotED.R")
@@
 
-<<cumswoldtab,results='asis'>>=
+```{r cumswoldtab,results='asis'}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
@@
 xtable(dr,label = "tab:cumswoldtab", caption = "Anova, all causes of death, 65-89.")
-@
+```
 
@@
 
-<<causes41>>=
+```{r causes41}
 cvb <- vb0
@@
  cvbTab$age <- factor(cvbTab$age)
-@
-
-<<41colorc>>=
+```
+
+```{r 41colorc}
 ## Colors for plotting strata:
@@
 
-<<41getske4060>>=
+```{r 41getske4060}
 cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cardiovascular death."
@@
 cvb4064$age <- factor(cvb4064$age)
-@
+```
 
@@
 
-<<41hazcalw>>=
+```{r 41hazcalw}
 source("R/getPHglm2.R")
@@
             nper, labb)
-@
-
-<<41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
+```
+
+```{r 41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8}
 source("R/plotED.glm.R")
@@
 par(oldpar)
-@
-
-<<41cumswtab,results='asis'>>=
+```
+
+```{r 41cumswtab,results='asis'}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass +
@@
 xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64.")
-@
+```
 
@@
 
-<<41getske6089>>=
+```{r 41getske6089}
 cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
 cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
-@
+```
 
@@
 
-<<41hazcalwold>>=
+```{r 41hazcalwold}
 resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
@@
                 nper, labb)
-@
+```
 
@@
 
-<<41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
+```{r 41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8}
 ##source("R/plotED.R")
@@
 par(oldpar)
-@
-
-<<41cumswoldtab,results='asis'>>=
+```
+
+```{r 41cumswoldtab,results='asis'}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * period *hisclass,
@@
 xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
-@
+```
 
@@
 
-<<causes12>>=
+```{r causes12}
 cvb <- vb0
@@
  cvbTab$age <- factor(cvbTab$age)
-@
-
-<<12colorc>>=
+```
+
+```{r 12colorc}
 ## Colours for plotting strata:
@@
 
-<<12getske4060>>=
+```{r 12getske4060}
 cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cancer death."
@@
 cvb4064$age <- factor(cvb4064$age)
-@
-
-<<12hazcalw>>=
+```
+
+```{r 12hazcalw}
 source("R/getPHglm2.R")
@@
             nper, labb)
-@
-
-<<12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8>>=
+```
+
+```{r 12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8}
 source("R/plotED.glm.R")
@@
 par(oldpar)
-@
-
-<<12cumswtab,results='asis'>>=
+```
+
+```{r 12cumswtab,results='asis'}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex *hisclass +
@@
 xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
-@
+```
 
@@
 
-<<12getske6089>>=
+```{r 12getske6089}
 cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
 cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
-@
+```
 
@@
 
-<<12hazcalwold>>=
+```{r 12hazcalwold}
 resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
@@
 
-<<12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8>>=
+```{r 12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8}
 ##source("R/plotED.R")
@@
 par(oldpar)
-@
-
-<<12cumswoldtab,results='asis'>>=
+```
+
+```{r 12cumswoldtab,results='asis'}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex +
@@
 xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
-@
+```
 
@@
 
-<<lissyoung3>>=
+```{r lissyoung3}
 library(eha)
@@
 close(con)
-@
-
-<<educ>>=
+```
+
+```{r educ}
 educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
@@
 cols <- c("red", "darkgreen", "black")
-@
-
-<<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
+```
+
+```{r ploteduc, fig.cap="Survival by education and sex, 1990--2013."}
 oldpar <- par(mfrow = c(1, 2))
@@
 par(oldpar)
-@
-
-<<tableeduc,results='asis'>>=
+```
+
+```{r tableeduc,results='asis'}
 fit <- coxreg(Surv(enter, exit, event) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
@@
     label = "tab:tableeduc", digits = 4)
-@
+```
 
@@
 
-<<cardeduc>>=
+```{r cardeduc}
 ulcard <- unique(vb0$ULORSAK[vb0$heart])
@@
              data = educ[educ$sex == "female", ])
-@
-
-<<ploteducheart, fig.cap="Cumulative hazards by education and sex, cardiovascular death, 1990--2013.">>=
+```
+
+```{r ploteducheart, fig.cap="Cumulative hazards by education and sex, cardiovascular death, 1990--2013."}
 oldpar <- par(mfrow = c(1, 2))
@@
 par(oldpar)
-@
-
-<<tableeducheart,results='asis'>>=
+```
+
+```{r tableeducheart,results='asis'}
 fit <- coxreg(Surv(enter, exit, heart) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
@@
     label = "tab:tableeducheart", digits = 4)
-@
+```
 
@@
 
-<<canceduc>>=
+```{r canceduc}
 ulcard <- unique(vb0$ULORSAK[vb0$cancer])
@@
              data = educ[educ$sex == "female", ])
-@
-
-<<ploteduccancer, fig.cap="Cumulative hazards by education and sex, cancer death, 1990--2013.">>=
+```
+
+```{r ploteduccancer, fig.cap="Cumulative hazards by education and sex, cancer death, 1990--2013."}
 oldpar <- par(mfrow = c(1, 2))
@@
 par(oldpar)
-@
-
-<<tableeduccancer,results='asis'>>=
+```
+
+```{r tableeduccancer,results='asis'}
 fit <- coxreg(Surv(enter, exit, cancer) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
@@
     label = "tab:tableeduccancer", digits = 4)
-@
+```
 
@@
 
-<<femeduc>>=
+```{r femeduc}
 income <- lisaold[!is.na(lisaold$income), ]
@@
 
-@
-
-<<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
+```
+
+```{r plotincome, fig.cap="Survival by income and sex, 1986--2005."}
 oldpar <- par(mfrow = c(1, 2))
@@
 par(oldpar)
-@
-
-<<tableinc>>=
+```
+
+```{r tableinc}
 fit <- coxreg(Surv(enter, exit, event) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
@@
 dr <- drop1(fit, test = "Chisq")
-@
-
-<<tableincpr,results='asis'>>=
+```
+
+```{r tableincpr,results='asis'}
 ltx(fit, dr = dr, caption = "Effect of income on mortality in ages 40-89, 1986-2005.",
     label = "tab:tableincpr", digits = 4)
-@
+```
 
@@
 
-<<heartincome>>=
+```{r heartincome}
 ulcard <- unique(vb0$ULORSAK[vb0$heart])
@@
 
-@
-
-<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005.">>=
+```
+
+```{r plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005."}
 oldpar <- par(mfrow = c(1, 2))
@@
 par(oldpar)
-@
-
-<<tableincheart>>=
+```
+
+```{r tableincheart}
 fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
@@
 dr <- drop1(fit, test = "Chisq")
-@
-
-<<tableincprheart,results='asis'>>=
+```
+
+```{r tableincprheart,results='asis'}
 ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.",
     label = "tab:tableincprheart", digits = 4)
-@
+```
 
@@
 
-<<cancerincome>>=
+```{r cancerincome}
 ulcard <- unique(vb0$ULORSAK[vb0$cancer])
@@
 
-@
-
-<<plotinccancer, fig.cap="Cumulative hazards by income and sex, cancer death, 1990--2013.">>=
+```
+
+```{r plotinccancer, fig.cap="Cumulative hazards by income and sex, cancer death, 1990--2013."}
 oldpar <- par(mfrow = c(1, 2))
@@
 par(oldpar)
-@
-
-<<tableinccancer>>=
+```
+
+```{r tableinccancer}
 fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
@@
 dr <- drop1(fit, test = "Chisq")
-@
-
-<<tableincprcancer,results='asis'>>=
+```
+
+```{r tableincprcancer,results='asis'}
 ltx(fit, dr = dr, caption = "Effect of income on cancer mortality in ages 40-89, 1986-2005.",
     label = "tab:tableincprcancer", digits = 4)
-@
+```
 

### 2017-07-03 icass.Rnw
add Yihui's float suggestions in preamble
nothing done

@@
 
-\usepackage{float}
+%\usepackage{float}
 
@@
 \renewcommand{\efloatseparator}{\vfill}
+
+\renewcommand{\textfraction}{0.05}
+\renewcommand{\topfraction}{0.8}
+\renewcommand{\bottomfraction}{0.8}
+\renewcommand{\floatpagefraction}{0.75}
 
@@
 ## Load remote data:
-con <- url("http://capa.ddb.umu.se/data/vb0.rda")
-##load("data/vb0.rda")
-load(con)
-close(con)
+##con <- url("http://capa.ddb.umu.se/data/vb0.rda")
+load("data/vb0.rda")
+##load(con)
+##close(con)
 vb0$hisclass[is.na(vb0$hisclass)] <- 4
@@
 ##
-con <- url("http://capa.ddb.umu.se/data/vb.rda")
-##load("data/vb.rda")
-load(con)
-close(con)
+##con <- url("http://capa.ddb.umu.se/data/vb.rda")
+load("data/vb.rda")
+##load(con)
+##close(con)
 ##
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-##load("data/lisaold.rda")
-load(con)
-close(con)
+##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+load("data/lisaold.rda")
+##load(con)
+##close(con)
 ##
-con <- url("http://capa.ddb.umu.se/data/loTab.rda")
-load(con)
-close(con)
+##con <- url("http://capa.ddb.umu.se/data/loTab.rda")
+##load(con)
+##close(con)
+load("data/loTab.rda")
 loTab$period <- factor(loTab$period)
@@
 liss4064 <- loTab[loTab$age < 64.5, ]
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
 ##load("data/lisaold.rda")
-load(con)
-close(con)
+##load(con)
+##close(con)
 @

### 2017-07-03 R/getPHglm2.R
remove debug stuff

@@ -27,7 +27,7 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
         inter <- matrix(co[interact], ncol = k - 1, byrow = TRUE)
         tmp <- matrix(0, nrow = m, ncol = k)
         for (j in 1:m){
-            cat("j = ", j, "\n")
+            ##cat("j = ", j, "\n")
             tmp[j, ] <- c(co[1], co[1] + amain)
         }
         for (j in 2:m){

### 2017-07-03 article.Rmd
more translation from Rnw

@@
 library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
+opts_chunk$set(engine='R',fig.path='figs/art-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 library(knitr)
@@
 ```{r readalldata}
+source("R/get.file.R")
+vb0 <- get.file("vb0")
+##if (file.exists("data/vb0.rda")){
+##    load("data/vb0.rda")
+##}else{
 ## Load remote data:
-##con <- url("http://capa.ddb.umu.se/data/vb0.rda")
-load("data/vb0.rda")
-##load(con)
-##close(con)
+##   con <- url("http://capa.ddb.umu.se/data/vb0.rda")
+##   load(con)
+##   close(con)
+##   save(vb0, file = "data/vb0.rda")
+##}
 vb0$hisclass[is.na(vb0$hisclass)] <- 4
@@
 ##
+
+vb <- get.file("vb")
+##if (file.exists("data/vb.rda"))
 ##con <- url("http://capa.ddb.umu.se/data/vb.rda")
-load("data/vb.rda")
+##load("data/vb.rda")
 ##load(con)
@@
 ##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-load("data/lisaold.rda")
+lisaold <- get.file("lisaold")
+##load("data/lisaold.rda")
 ##load(con)
@@
 ##close(con)
-load("data/loTab.rda")
+loTab <- get.file("loTab")
+##load("data/loTab.rda")
 loTab$period <- factor(loTab$period)
@@
 strong determinant when it comes to health and mortality and the impact even
-seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that
+seems to be increasing [@kuetal04, @frlu07]. It has been suggested that
 "... social conditions have been, are and will continue to be irreducible
 determinants of health outcomes and therefore deserves their appellation as
-'fundamental causes' of disease and death" \citep{liph95}. For a long time,
+'fundamental causes' of disease and death" [@liph95]. For a long time,
 the general view has been that socio-economic health inequalities were high in
@@
 resources provides advantages in all ways, the health advantage of higher
-classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social
+classes ought to be obvious. @ant67 and @smi83 suggest that social
 inequality in health has passed through different historical phases. According
@@
 health-related behaviour became important, resulting in marginal differences.
-\citet{omran82} state, in the third proposition in his theory of the
+@omran82 states, in the third proposition in his theory of the
 epidemiological transition, that even if the class differential in mortality was
@@
 with micro-data, has however questioned how generalizable the assumed
-pattern is \citep{bevp11}. Instead, some studies suggest that the modern
+pattern is [@bevp11]. Instead, some studies suggest that the modern
 pattern might be a fairly recent development, although the pattern differs
@@
 Swedish data, have found surprisingly small or even reversed socio-economic
-differences during the first stages of the transition \citep{bedr11,
-edli11,edbr12}, sharply contrasting against what we know from present-day
+differences during the first stages of the transition [@bedr11,
+@edli11,@edbr12], sharply contrasting against what we know from present-day
 society.
@@
 Another question of the development of social inequality in mortality,
-concerns its relation to economic inequality. \citet{wipi09} argue that
+concerns its relation to economic inequality. @wipi09 argue that
 income inequality has an independent effect on mortality, separate from the
@@
 present-day economically developed world. This has initiated a vital
-scholarly debate and the topic has been extensively studied \citep{suka04,
-wado00}. How the association looked like in previous periods is
+scholarly debate and the topic has been extensively studied [@suka04,
+@wado00]. How the association looked like in previous periods is
 unknown. This is of particular interest since the levels of economic
@@
 1980. Thereafter, economic inequality has increased substantially until the
-present, although far from being as high as a century ago \citep{rowa08,
-rowa09}. The Swedish development resembles that of other European and North
-American countries in its basic features \citep{rowa14}.
+present, although far from being as high as a century ago [@rowa08,
+@rowa09]. The Swedish development resembles that of other European and North
+American countries in its basic features [@rowa14].
 
@@
 
-The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the
-county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf
+```{r mapp,fig.cap = "The regions."}
+include_graphics("myfigs/sweden.png")
+```
+
+The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
+county of Västerbotten in the north of Sweden along the coast of Gulf
 of Bothnia. This was a remote part of the country, where communication with
@@
 several severe famines affected the regions during the 1800s. One living
-long in memory was the famine of 1867–-1868 \citep{segb14}. During the long
+long in memory was the famine of 1867–-1868 [@segb14]. During the long
 winters, sea communication was hindered due to the Gulf of Bothnia being
 frozen. The first ships during the year came to this area in May or even as
-late as June in some cases \citep{fah56}. Towards the end of the 19th
+late as June in some cases [@fah56]. Towards the end of the 19th
 century, the Swedish railway system reached this part of Sweden, thereby
@@
 Mortality was fairly low in comparison with other parts of the country and
-the fertility transition was late \citep{coa86}. The Skellefteå population
+the fertility transition was late [@coa86]. The Skellefteå population
 size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on
@@
 with substantial social diversity for the Skellefteå and Umeå regions
-\citep{wee16}. The data for the present study comes from two large
+[@wee16]. The data for the present study comes from two large
 population databases at the DDB. The early period until 1950 is covered by
 the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked
+the Skellefteå and Umeå region [@wee16]. Poplink is based on linked
 parish records, allowing us to reconstruct life biographies on people as
@@
 The other large data-set is extracted from the Linnaeus database
-\citep{malm10}. Linnaeus is based on different linked national population
+[@malm10]. Linnaeus is based on different linked national population
 registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
@@
 being resident in either of the regions are included. The data file
-contains the variables on social class ({\sc Hisclass}) sex, urban/rural
+contains the variables on social class (Hisclass) sex, urban/rural
 residence, birth date, death date, first and last date of observation and
@@
 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in
-the Linnaeus database. See Figure~\ref{fig:lexis2}.
+the Linnaeus database. See Figure @\ref(fig:lexis2).
 
@@
 In this paper we analyse mortality differences according to social class
-based on the classification scheme {\sc Hisclass} \citep{lema11}. The only
+based on the classification scheme Hisclass [@lema11]. The only
 type of socio-economic information available both in Linnaeus and Poplink
@@
 The availability to information on occupations varies over time and between
-sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the
+sources. Occupations in Poplink are coded into HISCO [@lema02] and the
 original DDB coding system. The censuses 1960--1990 include occupational
@@
 to be kept until death or last observation for those above 65. For the
-Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided
+Linnaeus data, we use a script recoding HISCO into Hisclass, provided
 to us by Jonas Helgertz, Lund University.
@@
 thus under-estimating female working-force participation as well as that of
-adult children residing with their parents \citep{vik10}. In this study and
+adult children residing with their parents [@vik10]. In this study and
 according to the research questions in the present paper, we have chosen to
@@
 same way as in Poplink. Extra-marital cohabitation, connected to the Second
-Demographic Transition \citep{van87, van03}, became common and female
+Demographic Transition [@van87, @van03], became common and female
 labour force participation developed as the norm in Sweden during this
@@
 
-{\sc Hisclass}, the classification system used here, is a "\ldots
-HISCO-based historical international social class scheme" \citep{lema11,
-lema02}. The different classes in {\sc Hisclass} represent distinct
+Hisclass, the classification system used here, is a "\ldots
+HISCO-based historical international social class scheme" [@lema11,
+@lema02]. The different classes in Hisclass represent distinct
 categories, based on whether the work is manual or non–manual, skill level,
@@
 economic and other resources, status, social power et cetera. We have
-merged the original 12 classes in {\sc Hisclass}, based on position in the
+merged the original 12 classes in Hisclass, based on position in the
 labour market, ending up with the following three and an additional fourth
@@
 \begin{enumerate}
-\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals
+\item {\tt elite}, Hisclass 1 and 2. Higher managers, professionals
 and higher white collar.
-\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers,
+\item {\tt middle}, Hisclass 3, 4, 5, 6 and 8. Lower managers,
 farmers and lower white collar.
-\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including
+\item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
 farm workers.
@@
 
-Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution
+Figures @\ref(fig:hiscskumm) and @\ref{fig:hiscskumw} show the distribution
 (per cent) of exposure time according to class for men and women
@@
 establishment of Umeå University explains much of this increase. What is
-hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle
+hidden in Figure @\ref(fig:hiscskumm) is the transfer within the middle
 group from a majority of farmers during the 19th century to lower white
@@
 The proportion with missing values among women is however a larger problem,
-see Figure~\ref{fig:hiscskumw}. Corresponding to what we found for men, the
+see Figure @\ref(fig:hiscskumw). Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th
@@
 Regarding the distribution of marital status over time, see
-Figure~\ref{fig:civstskum}. The figure indicates an under-registration of
+Figure \@ref(fig:civstskum). The figure indicates an under-registration of
 those married and and over-registration of previously married during the
@@
 The distribution of causes of death is shown in
-Figure~\ref{fig:causeskum}. Notice the high proportion of other causes in
+Figure \@ref(fig:causeskum). Notice the high proportion of other causes in
 the first two periods. This category include cases with unknown or
@@
 allowing for adjustment for civil status and rural/urban environment. Since
-the central explanatory variable, {\sc Hisclass}, do not follow the
+the central explanatory variable, Hisclass, do not follow the
 proportionality property, the analyses are stratified with respect to the
@@
 technique known as
-\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
+\emph{Cox regression} [@cox72,@gb02]. Instead we use a parametric model
 that allows for reduction of the data
-to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
+to \emph{sufficient statistics} [@fisher22]. The idea is to keep all
 relevant covariates \emph{categorical}, and then create a table with all
@@
 The analyses are performed in the \R\ environment for statistical computing
-and graphics \citep{cran}, especially using the package {\tt eha}
-\citep{gb12,eha}
+and graphics [@cran], especially using the package {\tt eha}
+[@gb12,@eha]
 The proportional hazards model allow us to scrutinise the differences and
@@
 are no serious signs of deviation from proportionality assumption, except
-regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
+regarding Hisclass. Each period are analysed separately, but we then combine results
 in order to show the development of the social patterning of mortality
@@
 
-# The effect of {\sc Hisclass}, 1851--1950, 1961--2013
+# The effect of Hisclass, 1851--1950, 1961--2013
 
@@
 
-Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval
+Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
 40--64 during the different periods. Obviously, and not surprisingly, mortality
@@
 demographic history. Male mortality has almost always been higher than
-female in all age groups \citep{wil99, suwi07}. We also observe a
+female in all age groups [@wil99, @suwi07]. We also observe a
 substantial decline in mortality for both sexes, resulting in smaller
@@
 
-By \emph{standardising} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these
+By \emph{standardising} the levels in each time period for each class, we 
+illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
+the paper, we only present these
 relative mortality levels for the different classes.
@@
 accordance with previous analyses from the present research group on other
-regions in northern Sweden and for other periods \citep{edli11,edbr12}. The
+regions in northern Sweden and for other periods [@edli11,@edbr12]. The
 highest social class usually have comparatively low mortality, while
@@
 during the nineteenth and much of the 20th century. This is also in
-accordance with our previous studies \citep{edli11, edbr12}. Workers on the
+accordance with our previous studies [@edli11, @edbr12]. Workers on the
 other hand have comparatively low mortality until the 1960s, at the same
@@
 
-A further observation can be made from Figure~\ref{fig:cumsw}. The gender
+A further observation can be made from Figure \@ref(fig:cumsw). The gender
 difference in death hazards is very big for the élite group from
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we found for the younger age group (see Figure~\ref{fig:cumswold}). For
+we found for the younger age group (see Figure \@ref(fig:cumswold). For
 women the results are mixed. Once again, too much missing data on social
@@
 Performing the same types of analyses as for all-cause mortality, we start
-by looking at cumulative hazards by {\sc Hisclass} at age 65, given
+by looking at cumulative hazards by Hisclass at age 65, given
 survival to age 40.
@@
 
-Figure~\ref{fig:41cumsw} shows mortality levels during different periods
+Figure \@ref(fig:41cumsw) shows mortality levels during different periods
 from 1901 on-wards. Women from the highest social class had consistently
@@
 
-See Table~\ref{tab:12cumswtab}.
-
-Figure~\ref{fig:12cumsw} shows relative mortality levels according to {\sc
-Hisclass} during the different studied periods for the age group
+See Table \@ref(tab:12cumswtab).
+
+Figure \@ref(fig:12cumsw) shows relative mortality levels according to
+Hisclass during the different studied periods for the age group
 40--64. Cancer mortality among the élite women was lower than that for the
@@
 A common alternative to a class analysis is to use income or education for
-investigating socio-economic health differences. \citet{goldk10} criticize
+investigating socio-economic health differences. @goldk10 criticize
 how many economists and epidemiologists operationalize inequality in their
@@
 on different regions, different age groups and different time periods
-\citep{edv92, edv04, edli11, edbr12}. When it comes to infant and child
+[@edv92, @edv04, @edli11, @edbr12]. When it comes to infant and child
 mortality pre---and early---transitional time, the expected socio-economic
-differences was often but not always confirmed \citep{edv92, edv04},
-something that has been found in other studies as well \citep{bredro02,
-bube01, mol17}. Studies based on microdata on the adult population in
+differences was often but not always confirmed [@edv92, @edv04],
+something that has been found in other studies as well [@bredro02,
+@bube01, @mol17]. Studies based on microdata on the adult population in
 history have been rare, mainly due to lack of adequate data. The results
@@
 particularly among adult men, not found despite the well-documented
-socio-economic differences \citep{edli11, edbr12}. There were however large
+socio-economic differences [@edli11, @edbr12]. There were however large
 differences between men and women and for different age groups. These
@@
 the advantage of higher classes developed somewhere during and probably
-quite early in the 20th century \citep{bevp11}. Until now we have however
+quite early in the 20th century [@bevp11]. Until now we have however
 not been able to investigate the continuous development on mortality
@@
 The results of small socio-economic gradient in the past and its late
-development are however in accordance with other studies \citep{bevp11},
-for example by \citet{bedr11} in a very different setting, a couple of
+development are however in accordance with other studies [@bevp11],
+for example by @bedr11 in a very different setting, a couple of
 Scanian parishes, although with somewhat different social categories. They
 found the first indications of an emerging social gradient in the
-1950s. Similar results have also been found in Estonia \citep{Jaadla17}.
+1950s. Similar results have also been found in Estonia [@jaadla17].
 
@@
 In this paper we have analysed social inequality in health from the
-perspective of social classes using {\sc Hisclass}. However, there are
+perspective of social classes using Hisclass. However, there are
 alternative ways to analyse social inequality, and we cannot exclude that
@@
 position influence health. Social class works through intermediate
-determinants related to exposure and vulnerability \citep{who10}. People
+determinants related to exposure and vulnerability [@who10]. People
 are situated within socio-economic and political contexts, as well as
@@
 resources. The importance of understanding people in their context is also
-emphasised by for example \citet{kri05}. What she calls embodiment refers
+emphasised by for example @kri05. What she calls embodiment refers
 to the recognition that human beings are simultaneously social beings and
-biological organisms. The WHO framework  \citep{who10} lists three
+biological organisms. The WHO framework  [@who10] lists three
 different types of intermediary social determinants of health; material
@@
 consider other explanations. Psycho-social factors have been frequently
-discussed during the most recent decades, for example by \citet{mar04}, who
+discussed during the most recent decades, for example by @mar04, who
 argues that a higher status leads to better health, regardless of where in
 the status ladder we make our observations. Psychosocial factors are also
-suggested by \citet{wipi09} as an explanation why equal societies always
+suggested by @wipi09 as an explanation why equal societies always
 perform better than unequal ones when it comes to health and other social
@@
 comes to gender differences, the present study confirms what we previously
-have found in the Sundsvall region \citep{edli11}. There is a clear
+have found in the Sundsvall region [@edli11]. There is a clear
 gendered pattern when it comes to social inequality in mortality.  Within
@@
 and behaviours that were harmful for health (this is discussed in more
-detail in \citet{edli11}. Alcohol has been suggested as part of the large
-gender differences in the 19th century \citep{wil99, edv92}, and a more
-risk-taking life may be a component of male mortality \citep{cour00}. There
+detail in @edli11. Alcohol has been suggested as part of the large
+gender differences in the 19th century [@wil99, @edv92], and a more
+risk-taking life may be a component of male mortality [@cour00]. There
 are however some problems with attributing most of the male excess
@@
 aspects of social classes, not only the material ones but also what this
-means for behaviour, norms, way of living. \citet{edli11} suggest that an
+means for behaviour, norms, way of living. @edli11 suggest that an
 explanation for the larger social health inequalities in present-day
@@
 contexts promote different ways to spend economic resources.This is also in
-line with what \citet{rasp06} and \citet{Jaadla17} suggest. If this
-interpretation is correct, the hypothesis of \citet{ant67}on an increasing
+line with what @rasp06 and @jaadla17 suggest. If this
+interpretation is correct, the hypothesis of @ant67 on an increasing
 role of health related behaviour in the modern low-mortality régime,while
@@
 researchers have pointed out, demography has often ignored power relations
-and different roles within families \citep{watkins93, mack95, jans07}. By
+and different roles within families [@watkins93, @mack95, @jans07]. By
 investigating possible differences in mortality patterns according to sex,
@@
 concerning the mortality transition is how the enormous increase in
-survival was accomplished. \citet{mck76} argues that the improvements up to
+survival was accomplished. @mck76 argues that the improvements up to
 the middle of the 20th century was primarily caused by improved
@@
 McKeown's ideas but it makes the question more complex. The results are
-also difficult to fit into \citet{liph95} hypothesis that social conditions
+also difficult to fit into @liph95 hypothesis that social conditions
 should be seen as a fundamental cause of disease and not only expressions
@@
 In recent years, the question about the relation between health and income
-inequality been widely discussed \citep{wipi09}. Even though our study do
+inequality been widely discussed [@wipi09]. Even though our study do
 not address this issue directly, our results are of interest for this
@@
 
-# Acknowledgements}
+# Acknowledgements
 
@@
 Jonas Helgertz, Department of Economic History, Lund University, for
-providing us with scripts for creating {\sc Hisclass} from Swedish Census
+providing us with scripts for creating Hisclass from Swedish Census
 data.
 
-\bibliography{mybib}
-
-\end{document}
+# References

### 2017-07-03 R/get.file.R
first commit

@@ -1,0 +1,12 @@
+get.file <- function(name, link = "http://capa.ddb.umu.se/data/"){
+    filename = paste("data/", name, ".rda", sep = "")
+    if (file.exists(filename)){
+        load(filename)
+    }else{
+        con <- url(paste(link, name, ".rda", sep = ""))
+        load(con)
+        close(con)
+        save(list = name, file = filename)
+    }
+    get(name)
+}

### 2017-07-03 article.Rmd
closer to Rmd

@@
 ---
-title: "Long-term perspectives of social inequality in mortality among
-  adults and elderly in northern Sweden 1851--2013"
+title: Long-term perspectives of social inequality in mortality among adults and elderly
+  in northern Sweden 1851--2013
 author: "Sören Edvinsson and Göran Broström"
@@
 output:
-  bookdown::html_document2:
+  bookdown::pdf_document2:
+    number_sections: yes
     toc: yes
     toc_depth: 2
-    toc_float: true
-    number_sections: true
-bibliography: ["mybib.bib"]
-biblio-style: "apalike"
-link-citations: true
+  bookdown::html_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
+    toc_float: yes
+link-citations: yes
+bibliography: mybib.bib
+biblio-style: apalike
 ---
@@
 
-\begin{figure}
-\includegraphics[height=19cm]{myfigs/sweden}
-\caption{The Skellefteå and Umeå regions in Sweden.}
-\label{fig:mapp}
-\end{figure}
-
-```{r mapp,fig.cap = "The regions."}
+```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in"}
 include_graphics("myfigs/sweden.png")
 ```
-
 The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
@@
 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in
-the Linnaeus database. See Figure @\ref(fig:lexis2).
+the Linnaeus database. See Figure \@ref(fig:lexis2).
 
@@
 
-Figures @\ref(fig:hiscskumm) and @\ref{fig:hiscskumw} show the distribution
+Figures \@ref(fig:hiscskumm) and \@ref{fig:hiscskumw} show the distribution
 (per cent) of exposure time according to class for men and women
@@
 establishment of Umeå University explains much of this increase. What is
-hidden in Figure @\ref(fig:hiscskumm) is the transfer within the middle
+hidden in Figure \@ref(fig:hiscskumm) is the transfer within the middle
 group from a majority of farmers during the 19th century to lower white
@@
 The proportion with missing values among women is however a larger problem,
-see Figure @\ref(fig:hiscskumw). Corresponding to what we found for men, the
+see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th
@@
 
-The analyses are performed in the \R\ environment for statistical computing
-and graphics [@cran], especially using the package {\tt eha}
-[@gb12,@eha]
+The analyses are performed in the **R** environment for statistical computing
+and graphics [@cran], especially using the package **eha**
+[@gb12,@eha].
 The proportional hazards model allow us to scrutinise the differences and
@@
 
-The important explanatory variable, {\tt hisclass}, is included in the
-proportional hazards models as a \emph{stratification} variable. The main
-reason for this choice is that the effect of {\tt hisclass} on survival is
+The important explanatory variable, **hisclass**, is included in the
+proportional hazards models as a *stratification* variable. The main
+reason for this choice is that the effect of **hisclass** on survival is
 non-proportional, that is, it varies with age. With the reduced data set
 and Poisson regression, this is simply achieved by introducing an
-interaction term between {\tt age} and {\tt hisclass}. Then, the main
-results are presented \emph{graphically}. Tables with results from the
+interaction term between **age** and **hisclass**. Then, the main
+results are presented graphically. Tables with results from the
 different models are presented at the end of the article, but are not
 commented on in detail. The comments on the results are primarily based on
-the figures and focuses on the central variable in our study---{\tt
-hisclass}. As can be seen from the tables, most variables in the model are
+the figures and focuses on the central variable in our study---**hisclass**. 
+As can be seen from the tables, most variables in the model are
 significant at 5 per cent level.
@@
 From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with {\tt sex}, {\tt period}, and {\tt hisclass}.
+The most interesting feature is the three-way interaction with **sex**, **period**, and **hisclass**.
 
@@
 
-By \emph{standardising} the levels in each time period for each class, we 
+By standardising the levels in each time period for each class, we 
 illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
@@
 As before, models are fitted separately for sex and each ten-year time period, from  1901 to
-2013. We stratify on {\tt hisclass} and include the covariates {\tt urban} (TRUE or FALSE) and
-{\tt civst} (marital status). The cumulative
+2013. We stratify on **hisclass** and include the covariates **urban** (TRUE or FALSE) and
+**civst** (marital status). The cumulative
 hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
@@
 
-See Table~\ref{tab:41cumswtab}.
+See Table \@ref(tab:41cumswtab).
 
@@
 
-See Table~\ref{tab:41cumswoldtab}.
+See Table \@ref(tab:41cumswoldtab).
 
@@
 
-See Table~\ref{tab:12cumswoldtab}.
+See Table \@ref(tab:12cumswoldtab).
 
@@
 
-See Table~\ref{tab:tableeduc}.
+See Table ref{tab:tableeduc}.
 
@@
 
-This work was undertaken within the research program \emph{Ageing and
-Living Conditions} at the Centre for Demographic and Ageing Research
+This work was undertaken within the research program *Ageing and
+Living Conditions* at the Centre for Demographic and Ageing Research
 (CEDAR) at Umeå University, financed by the Swedish Research Council (grant
-No 349-2008-6592) and the project \emph{Old age, health and inequality:
-Mortality from a life course perspective in Sweden 1986--2009}, financed by
+No 349-2008-6592) and the project *Old age, health and inequality:
+Mortality from a life course perspective in Sweden 1986--2009*, financed by
 the Swedish Research Council (grant No 421-2012-892). We thank the

### 2017-07-04 article.Rmd
no reading of 'vb.rda' now

@@
 output:
-  bookdown::pdf_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
   bookdown::html_document2:
@@
     toc_float: yes
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
 link-citations: yes
@@
 vb0 <- get.file("vb0")
-##if (file.exists("data/vb0.rda")){
-##    load("data/vb0.rda")
-##}else{
-## Load remote data:
-##   con <- url("http://capa.ddb.umu.se/data/vb0.rda")
-##   load(con)
-##   close(con)
-##   save(vb0, file = "data/vb0.rda")
-##}
+
 vb0$hisclass[is.na(vb0$hisclass)] <- 4
@@
 
-vb <- get.file("vb")
-##if (file.exists("data/vb.rda"))
-##con <- url("http://capa.ddb.umu.se/data/vb.rda")
-##load("data/vb.rda")
-##load(con)
-##close(con)
-##
-##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
+## We don't need "vb", do we?
+## vb <- get.file("vb")
+
 lisaold <- get.file("lisaold")
-##load("data/lisaold.rda")
-##load(con)
-##close(con)
-##
-##con <- url("http://capa.ddb.umu.se/data/loTab.rda")
-##load(con)
-##close(con)
+
 loTab <- get.file("loTab")
-##load("data/loTab.rda")
+
 loTab$period <- factor(loTab$period)

### 2017-07-24 article.Rmd
First update article version

@@
 ---
-title: Long-term perspectives of social inequality in mortality among adults and elderly
-  in northern Sweden 1851--2013
+title: Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013
 author: "Sören Edvinsson and Göran Broström"
-date: "28 juni 2017"
+date: "24 JulY 2017"
 output:
@@
 strong determinant when it comes to health and mortality and the impact even
-seems to be increasing [@kuetal04, @frlu07]. It has been suggested that
+seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand2010]. It has been suggested that
 "... social conditions have been, are and will continue to be irreducible

### 2017-07-24 article.Rmd
Deleting sections income and education

@@
 analysed according to gender, age-group (working age vs retired) and cause
-of death (cardiovascular and cancer). Furthermore, we compare our results
-witn analyses of alternative inequality indicators, education and income
-for the period 1986--2013.
+of death (cardiovascular and cancer).
 
@@
 
-## Income
-
-Income is based on disposable individual income from the LISA database at
-Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to
-2005.
-
-## Education
-
-Education is here analysed only for the period 1990 to 2013. Years of
-education are given in 3 categories and is based on level of education
-according to categories in the LISA database at Statistics Sweden. It
-ranges from basic education (nine years of schooling or less) at the lowest
-level, secondary education (high school or similar) for the middle group
-and to academic degree at the highest level.
-
 # Models
@@
 decades.
-
-In this section we analyse differences according to education level for the
-period 1990--2013 for the whole population 40 years and older. Education is
-a good indicator of both human and cultural capital, but aer also strongly
-associated with income and social class. We present analyses on all cause
-mortality (survival curves) as well as on the cause of death groups of
-cardiovascular diseases and cancers (cumulative hazards). The pattern is
-clear in total mortality as well as cause specific mortality, and this is
-valid for both men and women, see
-Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart}
-and~\ref{fig:ploteduccancer}. Education had a strong, statistically
-significant, effect on survival, as expected. Those with the lowest
-education levels hade the highest mortality, while those with highest level
-of education had by far the best survival. The differences must be
-considered large, and it follows a continuous improvement according to
-increasing levels.
-
-```{r lissyoung3}
-library(eha)
-library(xtable)
-liss4064 <- loTab[loTab$age < 64.5, ]
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-##load("data/lisaold.rda")
-load(con)
-close(con)
-```
-
-```{r educ}
-educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
-educ$period <- factor(educ$period)
-educ <- educ[!is.na(educ$educ), ]
-levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-##educ$educ <- relevel(educ$educ, ref = "middle")
-educ$civst <- relevel(educ$civst, ref = "married")
-fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-cols <- c("red", "darkgreen", "black")
-```
-
-```{r ploteduc, fig.cap="Survival by education and sex, 1990--2013."}
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lwd = 1.5, lty = c(1, 2, 5), axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-
-plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-
-par(oldpar)
-```
-
-```{r tableeduc,results='asis'}
-fit <- coxreg(Surv(enter, exit, event) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-ltx(fit, dr = dr, caption = "Effect of education on mortality in ages 40-90, 1990-2013.",
-    label = "tab:tableeduc", digits = 4)
-```
-
-See Table ref{tab:tableeduc}.
-
-```{r cardeduc}
-ulcard <- unique(vb0$ULORSAK[vb0$heart])
-educ$heart <- FALSE
-educ$heart <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
-fitm <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-```
-
-```{r ploteducheart, fig.cap="Cumulative hazards by education and sex, cardiovascular death, 1990--2013."}
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1:2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1:2, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-```
-
-```{r tableeducheart,results='asis'}
-fit <- coxreg(Surv(enter, exit, heart) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-ltx(fit, dr = dr, caption = "Effect of education on cardiovascular mortality in ages 40-90, 1990-2013.",
-    label = "tab:tableeducheart", digits = 4)
-```
-
-See Table~\ref{tab:tableeducheart}.
-
-```{r canceduc}
-ulcard <- unique(vb0$ULORSAK[vb0$cancer])
-educ$cancer <- FALSE
-educ$cancer <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
-fitm <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
-             data = educ[educ$sex == "female", ])
-```
-
-```{r ploteduccancer, fig.cap="Cumulative hazards by education and sex, cancer death, 1990--2013."}
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-```
-
-```{r tableeduccancer,results='asis'}
-fit <- coxreg(Surv(enter, exit, cancer) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-ltx(fit, dr = dr, caption = "Effect of education on cancer mortality in ages 40-89, 1990-2013.",
-    label = "tab:tableeduccancer", digits = 4)
-```
-
-See Table~\ref{tab:tableeduccancer}.
-
-# The effect of Income, 1986--2005
-
-```{r femeduc}
-income <- lisaold[!is.na(lisaold$income), ]
-income$period <- factor(income$period)
-##income <- income[!is.na(income$income), ]
-##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
-##income$income <- relevel(income$income, ref = "low")
-income$civst <- relevel(income$civst, ref = "married")
-fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-cols <- c("darkgreen", "red", "blue", "black")
-
-```
-
-```{r plotincome, fig.cap="Survival by income and sex, 1986--2005."}
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-
-plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
-axis(1)
-axis(2, at = c(0, 0.5, 1.0), las = 1)
-abline(h = 0.5, lty = 3)
-box()
-legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-
-par(oldpar)
-```
-
-```{r tableinc}
-fit <- coxreg(Surv(enter, exit, event) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-```
-
-```{r tableincpr,results='asis'}
-ltx(fit, dr = dr, caption = "Effect of income on mortality in ages 40-89, 1986-2005.",
-    label = "tab:tableincpr", digits = 4)
-```
-
-See Table~\ref{tab:tableincpr}.
-
-Finally, we present analyses of different income levels for the age group
-40 and above and the period 1986--2005. Disposable individual income must
-be considered as a robust measure of economic condition, and increases the
-possibility to allocate these economic resources towards health promoting
-living conditions and behaviours. We categorize the income according to
-quartiles. As expected, we find that higher incomes result in lower
-mortality, both when it comes to total mortality as to deaths from
-cardiovascular diseases, see Figures~\ref{fig:plotincome}
-and~\ref{fig:plotincheart}. This is also what could be expected, and
-conforms to what we know of social health differences in present-day
-Sweden. We also find indications of a similar difference for cancers but
-less clear, particularly for male cancer mortality, see
-Figure~\ref{fig:plotinccancer}. Notice however that we cannot establish a
-causal relationship between income and mortality. Bad health may be a
-confounder, causing both low incomes and higher death risks.
-
-```{r heartincome}
-ulcard <- unique(vb0$ULORSAK[vb0$heart])
-income$heart <- FALSE
-income$heart <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
-fitm <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-
-```
-
-```{r plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005."}
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-```
-
-```{r tableincheart}
-fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-```
-
-```{r tableincprheart,results='asis'}
-ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.",
-    label = "tab:tableincprheart", digits = 4)
-```
-
-See Table~\ref{tab:tableincprheart} for the effect of income on cardiovascular death.
-
-```{r cancerincome}
-ulcard <- unique(vb0$ULORSAK[vb0$cancer])
-income$cancer <- FALSE
-income$cancer <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
-fitm <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
-             data = income[income$sex == "male", ])
-fitf <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
-             data = income[income$sex == "female", ])
-
-```
-
-```{r plotinccancer, fig.cap="Cumulative hazards by income and sex, cancer death, 1990--2013."}
-oldpar <- par(mfrow = c(1, 2))
-
-plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
-     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
-#axis(1)
-#axis(2, at = c(0, 0.5, 1.0), las = 1)
-#abline(h = 0.5, lty = 3)
-#box()
-legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
-abline(h = 0)
-par(oldpar)
-```
-
-```{r tableinccancer}
-fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
-              hazards = FALSE)
-dr <- drop1(fit, test = "Chisq")
-```
-
-```{r tableincprcancer,results='asis'}
-ltx(fit, dr = dr, caption = "Effect of income on cancer mortality in ages 40-89, 1986-2005.",
-    label = "tab:tableincprcancer", digits = 4)
-```
-
-See Table~\ref{tab:tableincprcancer} for the income effect on cancer death.
 

### 2017-07-24 article.Rmd
Fix conflicting sizes, add tables (by replacing 'xtable' by 'kable')

@@
 ---
-title: Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013
+title: "Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013"
 author: "Sören Edvinsson and Göran Broström"
-date: "24 JulY 2017"
+date: "24 July 2017"
 output:
@@
 strong determinant when it comes to health and mortality and the impact even
-seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand2010]. It has been suggested that
+seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand10]. It has been suggested that
 "... social conditions have been, are and will continue to be irreducible
@@
 
-In the data-set analysed here, all individuals 40 years and older ever
-being resident in either of the regions are included. The data file
-contains the variables on social class (Hisclass) sex, urban/rural
-residence, birth date, death date, first and last date of observation and
-type of entrance/exit. Total number of persons are 84\,599 leading to
-35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in
-the Linnaeus database. See Figure \@ref(fig:lexis2).
-
 ```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
@@
 #
-n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
+pn.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
+pn.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
+pn.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
+pn.d <- as.integer(pn.d)
 text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(n.rec, "records"))
-text(1900, 55, paste(n.pl, "persons"))
-text(1899.5, 50, paste(n.d, "deaths"))
+text(1899, 60, paste(pn.rec, "records"))
+text(1900, 55, paste(pn.pl, "persons"))
+text(1899.5, 50, paste(pn.d, "deaths"))
 #
@@
 n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
+n.d <- as.integer(n.d)
 
@@
 ```
+
+In the data-set analysed here, all individuals 40 years and older ever
+being resident in either of the regions are included. The data file
+contains the variables on social class (Hisclass) sex, urban/rural
+residence, birth date, death date, first and last date of observation and
+type of entrance/exit. Total number of persons are `r pn.pl` leading to
+`r pn.d` deaths in Poplink and `r n.pl` persons leading 
+to `r n.d` deaths in the Linnaeus database. See 
+Figure \@ref(fig:lexis2).
 
@@
 
-Hisclass, the classification system used here, is a "\ldots
+Hisclass, the classification system used here, is a "...
 HISCO-based historical international social class scheme" [@lema11,
@@
 
-Figures \@ref(fig:hiscskumm) and \@ref{fig:hiscskumw} show the distribution
+Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
 (per cent) of exposure time according to class for men and women
@@
 
-% ```{r hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
-% xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-% yy <- round(prop.table(xx, 1) * 100, 1)
-% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-% ```
-%
-% ```{r hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
-% xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-% yy <- round(prop.table(xx, 1) * 100, 1)
-% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
-% ```
-
 The proportion with missing values among women is however a larger problem,
@@
 
-\subsection{Marital status}
+## Marital status
 
@@
 
-```{r civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5}
+```{r civstskum, fig.cap = "Exposure by civil status (from the top: never married, married and previously married) and time period", fig.height = 4.5}
 xx <- with(vb0Tab, tapply(exposure, list(period = period, civst = civst), sum))
@@
 Another practical problem is the size of that data set:
-It contains no less than 2432871 records of 236011 individuals, of
-which 68244 are observed to die.
+It contains no less than `r n.rec + pn.rec` records of `r n.pl + pn.pl` individuals, of
+which `r n.d + pn.d` are observed to die.
 This makes it very inconvenient to use the nowadays so popular semi-parametric
@@
 dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64.")
+knitr::kable(dr, caption = "Anova, cardiovascular deaths, 40-64.")
+#print(xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64."), type = "html")
 ```
@@
 dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
+knitr::kable(dr, caption = "Anova, cardiovascular deaths, 65-90.")
+##xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
 ```
@@
 dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
+knitr::kable(dr, caption = "Anova, cancer deaths, 40-64.")
+##xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
 ```
@@
 dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
+knitr::kable(dr, caption = "Anova, cancer deaths, 65-90.")
+##xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
 ```

### 2017-07-26 article.Rmd
Update Data and variables

@@
 ---
-title: "Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013"
-author: "Sören Edvinsson and Göran Broström"
+title: "'Is high social class always beneficial for survival? Long-term perspectives"
+  of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013'
+author: "SÃƒÆ’Ã‚Â¶ren Edvinsson and GÃƒÆ’Ã‚Â¶ran BrostrÃƒÆ’Ã‚Â¶m"
 date: "24 July 2017"
 output:
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::html_document2:
@@
     toc_float: yes
-  bookdown::pdf_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
 link-citations: yes
@@
 
+In the present paper, we take a long-term perspective on the development of
+social inequalities in mortality in the adult and elderly population, and how this is connected
+to the development of economic inequality in society.
+The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
+This issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
+The results are then discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the hypotheses and concepts in health research.
+
+On the basis of these results, we argue that a consistent hierarchical social gradient in mortality appears surprisingly late, but that there were large gender differences in trends and levels.
+We suggest that more attention should be paid on health-related behaviour also in earlier phases of the mortality transition.
+
 One of the central aspects of survival is social class and access to economic
@@
 'fundamental causes' of disease and death" [@liph95]. For a long time,
-the general view has been that socio-economic health inequalities were high in
+the general view has been that socio-economic health and mortality inequalities were high in
 historical societies, probably higher than in modern societies. This is a
@@
 very large differences between groups and classes. Knowing that access to
-resources provides advantages in all ways, the health advantage of higher
+resources provides advantages in all aspects of life, the health advantage of higher
 classes ought to be obvious. @ant67 and @smi83 suggest that social
-inequality in health has passed through different historical phases. According
-to them, social health differences were comparatively small during the
+inequality in mortality has passed through different historical phases. According
+to them, differences were comparatively small during the
 pre-transitional phase when space was a strong determinant for the spread of
 disease. Differences then increased during the transitional phase when mortality
-declined and wealthy groups used their resources to gain better health. Finally,
-health differences decreased again in modern low-mortality settings when instead
-health-related behaviour became important, resulting in marginal differences.
-@omran82 states, in the third proposition in his theory of the
+declined and wealthy groups used their economic and human resources to gain better health. Finally,
+mortality differences decreased again resulting in marginal levels of inequality in modern low-mortality societies when instead
+health-related behaviour became the decisive determinant for health and survival.
+@omran82 comes to a similar conclusion states, in the third proposition in his theory of the
 epidemiological transition, that even if the class differential in mortality was
 maintained during the transition, the decline set in earlier and was faster
-among privileged groups.
+among privileged groups. These assumption if partly supported by data from England and Wales [@pamuk85].
 
@@
 pattern is [@bevp11]. Instead, some studies suggest that the modern
-pattern might be a fairly recent development, although the pattern differs
+pattern might be a fairly recent development [@bedr11], although the pattern differs
 depending on age, gender or historical context. When it comes to the adult
@@
 Swedish data, have found surprisingly small or even reversed socio-economic
-differences during the first stages of the transition [@bedr11,
-@edli11,@edbr12], sharply contrasting against what we know from present-day
-society.
-
-Another question of the development of social inequality in mortality,
+differences during the 19th and early 20th century [@bedr11,
+@edli11,@edbr12], sharply contrasting from the situation in present-day society.
+
+It thus appears as if a more consistent social health divide appeared at
+some point during the mortality transition, at least in Sweden. Solid
+empirical evidence about the process is however lacking and studies
+focusing on the issue are still few. There is a need for more reliable
+studies from different geographical and historical settings in order to 
+better understand the role of socio-economic conditions for health and survival over time. 
+Furthermore, previous studies indicate substantial differences in the development according to age and gender.
+
+Another aspect of the development of social inequality in mortality,
 concerns its relation to economic inequality. @wipi09 argue that
 income inequality has an independent effect on mortality, separate from the
-direct effect of actual different access to economic resources. They argue
+direct effect of actual access to economic resources. They argue
 that unequal societies perform less well when it comes to health than equal
@@
 inequality were very different from those of the recent decades. Even if it
-not a necessary implication, we would still assume that poorer groups were
-those most disadvantaged of large inequality in itself. The situation
-during the 19th century is uncertain, but at least during the late 19th and
-early 20th century both income and wealth distribution was strongly skewed
-in Sweden. From a couple of decades into the 20th century, income and
+not a necessary implication, it is reasonable to assume that poorer groups were
+those most disadvantaged of large inequality in itself.
+During the last decades, the association between trends of inequalities in mortality and income respectively are weak or non-existant according to @hoetal16.
+
+When it comes to Sweden both income and wealth distribution was strongly skewed from the 1870's to the early 20th century. Starting during the inter-war period, income and
 wealth inequality continuously diminished to reach a low point at around
@@
 American countries in its basic features [@rowa14].
-
-It thus appears as if a more consistent social health divide appeared at
-some point during the mortality transition, at least in Sweden. Solid
-empirical evidence about the process is however lacking and studies
-focusing on the issue are still few. There is a need for more reliable
-studies from different geographical and historical settings as well as
-according to age and gender in order to better understand the role of
-socio-economic conditions for health and survival over time.
-
-# Aims and questions
-
-In the present paper, we take a long-term perspective on the development of
-social inequalities in mortality in the adult and elderly population. The
-areas under investigation are the Skellefteå and Umeå regions in the north
-of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013
-for Umeå. The main question we investigate is if inequality in mortality
-between social classes increased among the adult and elderly population,
-primarily benefiting the highest social positions and how this is connected
-to the development of economic inequality in society. This issue is then
-analysed according to gender, age-group (working age vs retired) and cause
-of death (cardiovascular and cancer).
 
@@
 increasingly integrated in the same epidemiological pattern as the rest of
-Sweden. Diseases reached to places that earlier were spared.
+Sweden. Diseases reached to places that were spared in earlier times.
 
@@
 resident in our regions during the complete five-year period between
-censuses if the person is missing in the later census. This overestimates
+censuses if the person is missing in the later census (the last period covering ten years). This overestimates
 their presence in our regions, but should not bias the results when it
@@
 (LISA data), together with information on deaths from National Board of
-Health and Welfare. This data allows good control of the presence periods
+Health and Welfare. This data allows for full control of the presence periods
 in the studied regions.
@@
 ## Social class
+
+The choice of variable for analysing social differences vary between studies. A common alternative to a class analysis is to use income or education for
+investigating socio-economic health differences. @goldk10 criticize
+how many economists and epidemiologists operationalize inequality in their
+studies, for example by not distinguishing between class and status
+aspects. Class refers to relational aspects in society, while many other
+variables focus on the distribution of material and/or immaterial
+resources. He advocates studies of different aspects, which together
+complement each other. Comparing different approaches are valuable in order
+to understand why some are advantaged and others disadvantaged and what
+sort of factors determines health differences.
 
@@
 based on the classification scheme Hisclass [@lema11]. The only
-type of socio-economic information available both in Linnaeus and Poplink
+type of socio-economic information available both in the Linnaeus and Poplink database
 is occupations. Thus, analysing according to social class is apart from
@@
 The availability to information on occupations varies over time and between
-sources. Occupations in Poplink are coded into HISCO [@lema02] and the
-original DDB coding system. The censuses 1960--1990 include occupational
+sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. The censuses 1960--1990 include occupational
 codes based on the system used in Sweden at that time, NYK (Nordisk
-Yrkesklassificering). The yearly population registers (LISA) do not report
-occupations until 2001, thus there is a window of 11 years from 1991 to
-2001 where we lack information. In the censuses, we have solved this by
-extending the periods for which we consider the observed social class to be
-valid. Thus, the social class in the census 1990 is considered to be kept
-until 2001. This strategy is also used for other cases of missing social
+Yrkesklassificering).
+We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our treatment of presence periods.
+For the period 1991--2000 the social class in the census 1990 is considered to be kept
+until 2001. This strategy is also used for cases of missing social
 class. Furthermore, we consider the last social position before retirement
@@
 The use of different data sets, forces us to apply different approaches
-towards defining the social position in some respects. A problem is that
+towards defining the social position. A problem is that
 Poplink usually only provides the occupation for the head of household,
@@
 according to the research questions in the present paper, we have chosen to
-categorise wife according to the position of the head of household, usually
+categorise wives according to position of the head of household, usually
 the husband/father, assuming that the family shares the same socio-economic
@@
 different periods are not completely comparable and makes the
-interpretation of the development for women more shaky.
+interpretation of the development for women more uncertain.
+Notice however that there is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way if we use social class on household level also for the Linneaus data.
 
@@
 in the forefront of other categorisations, as for example access to
-economic and other resources, status, social power et cetera. We have
+economic and other resources, status, social power et cetera.
+Longitudinal analyses of social classes for such a long period covered in this study involves several problems, related to changes in occupational structure and the meaning and contents of occupations and social positions.
+We have
 merged the original 12 classes in Hisclass, based on position in the
@@
 establishment of Umeå University explains much of this increase. What is
-hidden in Figure \@ref(fig:hiscskumm) is the transfer within the middle
+invisible in Figure \@ref(fig:hiscskumm) is the transfer within the middle
 group from a majority of farmers during the 19th century to lower white
@@
 increase during the 1960s and partly during later decades as well in those
-with undefined. When first observed in the earliest included censuses,
+with undefined class. When first observed in the earliest included censuses,
 occupation is often missing. For men the proportion is somewhat higher in
@@
 to report results for women for these two decades. For the later decades,
-the results are fairly complete and should be representative.
+the results are fairly complete.
 
@@
 Figure \@ref(fig:civstskum). The figure indicates an under-registration of
-those married and and over-registration of previously married during the
+those married and over-registration of unmarried during the
 1960s.
@@
 See Table \@ref(tab:12cumswoldtab).
-
-# The effect of Education, 1990--2013
-
-A common alternative to a class analysis is to use income or education for
-investigating socio-economic health differences. @goldk10 criticize
-how many economists and epidemiologists operationalize inequality in their
-studies, for example by not distinguishing between class and status
-aspects. Class refers to relational aspects in society, while many other
-variables focus on the distribution of material and/or immaterial
-resources. He advocates studies of different aspects, which together
-complement each other. Comparing different approaches are valuable in order
-to understand why some are advantaged and others disadvantaged and what
-sort of factors determines health differences. In this and the following
-section, we introduce analyses of education and income. In that way, we can
-compare the class analyses with these alternative categorizations. For the
-time being, we have to restrict such a comparative analysis to the last
-decades.
 

### 2017-07-27 article.Rmd
Revisions of variables and added references

@@
 ---
-title: "'Is high social class always beneficial for survival? Long-term perspectives"
-  of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013'
-author: "SÃƒÆ’Ã‚Â¶ren Edvinsson and GÃƒÆ’Ã‚Â¶ran BrostrÃƒÆ’Ã‚Â¶m"
+title: Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013
+author: Sören Edvinsson and Göran Broström
 date: "24 July 2017"
@@
 In the present paper, we take a long-term perspective on the development of
-social inequalities in mortality in the adult and elderly population, and how this is connected
-to the development of economic inequality in society.
+social inequalities in mortality in the adult and elderly population.
 The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
 This issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
-The results are then discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the hypotheses and concepts in health research.
+Furthermore we put this in he context of how the inequality in mortality is associated
+with the development of economic inequality in society.The results are then discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
 
@@
 In 19th and early 20th century Sweden (as in other countries), mortality
-was much higher in urban environments. The unsanitary and overcrowded
+was much higher in urban environments [@hani94, @edv92]. The unsanitary and overcrowded
 living conditions as well as the risks for importing infections made the
@@
 for this aspect by distinguishing between urban and rural residence, based
-on in which parish people lived.
+on parish of residence.
 
@@
 periods represent a mainly pre-industrial society dominated by agriculture
-and with restricted welfare provisions. During the years 1901--1950,
+and with only basic welfare provisions. During the years 1901--1950,
 industrialisation started and the urban environments increased in
-population size. The modern welfare state began to develop during this
-time. For the period 1960 on-wards, there are separate analyses on decades
+population size. The modern Swedish welfare state began to develop during this
+time. For the period 1960 onwards, there are separate analyses on decades
 except for the two last periods. Sweden had developed into a wealthy
-welfare society with low income inequality in international comparison,
+welfare society with low income inequality,
 particularly until the 1990s. A financial crisis struck Sweden in the early
 1990s, which among other things increased unemployment and changed parts of
-the welfare system to being less generous than before. The start of the
+the welfare system towards being less generous than before. The start of the
 last period (2008--2013) coincides with the international financial
@@
 The proportional hazards model in a survival analysis context is used,
-allowing for adjustment for civil status and rural/urban environment. Since
-the central explanatory variable, Hisclass, do not follow the
+allowing for adjustment for marital status and rural/urban environment. Since
+the central explanatory variable, social class (Hisclass), do not follow the
 proportionality property, the analyses are stratified with respect to the
@@
 combination, the total amount of exposure and the number of deaths are
-calculated.  We have two sexes, three civil statuses, ten periods, two
-urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
+calculated.  We have two sexes, three marital statuses, ten periods, two
+urban/rural groups, four social classes at age 40,  four social classes at age 60,
 ten age groups, and 22 cohorts, in all 422\,400 possible
@@
 are no serious signs of deviation from proportionality assumption, except
-regarding Hisclass. Each period are analysed separately, but we then combine results
+regarding docial class. Each period are analysed separately, but we then combine results
 in order to show the development of the social patterning of mortality
@@
 
-The important explanatory variable, **hisclass**, is included in the
+The important explanatory variable, social class (**hisclass**), is included in the
 proportional hazards models as a *stratification* variable. The main
-reason for this choice is that the effect of **hisclass** on survival is
+reason for this choice is that the effect of social class (**hisclass**) on survival is
 non-proportional, that is, it varies with age. With the reduced data set
 and Poisson regression, this is simply achieved by introducing an
-interaction term between **age** and **hisclass**. Then, the main
+interaction term between **age** and social class (**hisclass**). Then, the main
 results are presented graphically. Tables with results from the
@@
 commented on in detail. The comments on the results are primarily based on
-the figures and focuses on the central variable in our study---**hisclass**. 
-As can be seen from the tables, most variables in the model are
+the figures and focuses on the central variable in our study--- social class (**hisclass**). 
+As can be seen from the tables, most variables in the models are
 significant at 5 per cent level.
 
-# The effect of Hisclass, 1851--1950, 1961--2013
+# The effect of social class, 1851--1950, 1961--2013
 
@@
 From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with **sex**, **period**, and **hisclass**.
+The most interesting feature is the three-way interaction with **sex**, **period**, and social class (**hisclass**).
 
@@
 substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. The compression that is
+absolute differences between social classes. We must however distinguish between differences in absolute risks compared to relative risks [@hoetal16). The compression that is
 visible for later decades is partly explained by the decreasing general
@@
 the paper, we only present these
-relative mortality levels for the different classes.
+relative mortality levels for the different classes since they are the main focus in this study.
 
@@
 Notice that we have not analysed the 1950s due to the fact that we only
-have data for some parishes at that time, consequently having no
-corresponding population for that decade. Furthermore, we have excluded the
-results for women during the 1960s and 1970 of reasons discussed above. The
+have data for some parishes at that time. Furthermore, we have excluded the
+results for women during the 1960s and 1970 out of the reasons discussed above. The
 changing economy also have consequences for the analysis, particularly when
@@
 the relative risks between the classes. Only the group with unknown social
-class is distinguished by moch higher mortality from the 1980s
+class is distinguished by much higher mortality from the 1980s
 onwards. This can however have to do with the differences how social
@@
 on the own occupation. Being without occupation might include women that
-are more frail both when it comes to health and socially.
+are more frail both when it comes to health and social characteristics.
 

### 2017-07-29 article.Rmd
Revising results and references

@@
 ---
-title: Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013
+title: "Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013""
 author: Sören Edvinsson and Göran Broström
@@
 working class women have high. There is however no substantial change in
-the relative risks between the classes. Only the group with unknown social
+the relative risks between classes. Only the group with unknown social
 class is distinguished by much higher mortality from the 1980s
-onwards. This can however have to do with the differences how social
-position for women is defined before and after 1950. After 1950 it is based
-on the own occupation. Being without occupation might include women that
+onwards. This can however have to do with how we have defined social
+position for women before and after 1950. After 1950 it is based
+on their own occupation. Being without occupation might include women that
 are more frail both when it comes to health and social characteristics.
@@
 our analyses. The groups we would expect to have the lowest mortality, if
-we assume that access to different resources and having a high status
+we assume that access to different resources and belonging to a high social class
 determines survival, instead have by far the highest hazards of dying
-during the nineteenth and much of the 20th century. This is also in
-accordance with our previous studies [@edli11, @edbr12]. Workers on the
+during the nineteenth and much of the 20th century. Workers on the
 other hand have comparatively low mortality until the 1960s, at the same
@@
 unknowns social class is probably even more relevant for men. They probably
-represent groups in society that are particularly vulnerable.
+represent groups in society that are particularly vulnerable, having difficulties getting into the laour force.
 
@@
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure.
+duration 25 (age 90), starting from age 65 as a summary measure. In Table~\ref{tab:cumswoldtab} the results of the model is presented.
 
@@
 class leaves out the results for the 1960s and 1970s. We do not find the
-same clear advantage for élite women during period until 1950, but neither
+same clear advantage for élite women during the period before 1950, but neither
 do we find any large disadvantage for working class elderly women at that
@@
 for the adult male population 40--64 years. Mortality was in fact
-substantially higher among élite men  but their disadvantage was primarily
+substantially higher among élite men  but their disadvantage was most 
 apparent during the 19th century. Differences are fairly small during the
@@
 century is the survival much better for the highest social class while
-workers in the age of retirement had high mortality in comparison. ). This
-is quite unexpected, since other studies have shown clear socio-economic
-differences in Sweden during recent decades.
+workers in the age of retirement had high mortality in comparison.
 
@@
 
-In this section, the focus is on analyses of mortality from the two
-dominant causes of death among the adult and elderly population. We start
+In this and the following section, the focus is on analyses of mortality from the two
+dominant causes of death groups among the adult and elderly population. We start
 with cardiovascular diseases. Many studies have shown that socio-economic
@@
 group. Belonging to working class, having less education and low income
-increases the mortality risks. We would therefore expect to find a clear
+increases mortality risks. We would therefore expect to find a clear
 social pattern in the present analysis.
 
-First, a general observation is that the analysis of causes of death and
+A general observation is that the results on causes of death and
 its development is difficult to interpret, partly because the number of
 cases are fewer but also that the quality and coverage of reported death
-causes differ. Thus, many deaths during the early 20th got quite unspecific
-causes reported in the death registers. There was still a substantial part
+causes differ in time. Thus, many deaths during the early 20th got quite vague diagnoses reported in the death registers. There was still a substantial part
 of all deaths where cause was diagnosed as unknown or dying from old age,
@@
 to 18 per cent. It is quite reasonable to suspect that many cardiovascular
-deaths were disguised in unspecific categories and that the increase partly
-is explained by improvements in how deaths were diagnosed.
+deaths were disguised in unspecific categories and that the increase is partly
+explained by improvements in how deaths were diagnosed, something that also seems to have been the case in Scotland at this time [@reidetal15].
 
@@
 As before, models are fitted separately for sex and each ten-year time period, from  1901 to
-2013. We stratify on **hisclass** and include the covariates **urban** (TRUE or FALSE) and
+2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
 **civst** (marital status). The cumulative
@@
 by looking at cumulative hazards by Hisclass at age 65, given
-survival to age 40.
+survival to age 40. The result of the model can be seen in Table \@ref(tab:41cumswtab).
 
@@
 but in the same way as among women, the levels were fairly similar to the
-one of the middle category consisting of lower manager/farmers. And in the
-same way as for women at this age, working class men have the highest
-mortality from this cause of death during the last decades. A social
-hierarchy in cardiovascular mortality seems to have developed both for men
+one of the middle category consisting of white collar, lower managers and farmers. In the last decades and similarly as for women, working class men have the highest
+mortality from this cause of death group. A social
+hierarchy in cardiovascular mortality developed both for men
 and women.
-
-### The effect of income and education, 1990--2005
 
@@
 summary measure the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65.
+duration 25 (age 90), starting from age 65. The results from the model is presented in Table \@ref(tab:41cumswoldtab).
 
@@
 
-Concerning the social pattern, we find somewhat different results from that
-of the younger age group of women. Cardiovascular mortality among elderly
+Concerning the social pattern, we find somewhat different results for elderly women compared to the women in working-active age. Cardiovascular mortality among elderly
 women do not present a consistent class pattern. Female cardiovascular
@@
 class background were not disadvantaged for most of the 20th century. The
-social class pattern has changed somewhat during the last periods and more
+social class pattern has changed somewhat during the last periods becoming more
 in line to what is expected. The differences are however rather small.
@@
 highest class had high cardiovascular mortality during the first half of
-the 20th century, while the expected social pattern has developed only
+the 20th century, while the expected social pattern developed and became apparent only
 during the most recent periods.
@@
 is not necessarily an obvious assumption when it comes to
-cancers. Different forms of cancers have very different origin and
+cancers. Different forms of cancers have different origin and
 different pathways to disease and death. The social pattern of higher
@@
 
-See Table \@ref(tab:12cumswtab).
+For the results of the model, see Table \@ref(tab:12cumswtab).
 
@@
 Hisclass during the different studied periods for the age group
-40--64. Cancer mortality among the élite women was lower than that for the
+40--64. Cancer mortality among the elite women was lower than that for the
 other classes during the complete studied period, while working class women
@@
 more consistent advantage of higher class in cancers compared to
-cardiovascular diseases for women. We do however not find the same pattern
+cardiovascular diseases for women, thus confirming what we have expected. We do however not find the same pattern
 for men. It is difficult to identify any general features of 20th century
 male cancer mortality when it comes to social differences.
-
-### The effect of income and education on cancer mortality, 1990--2005
 
@@
 In our analyses of class patterns in mortality, we finally turn to cancer
-mortality among the elderly.
+mortality among the elderly. The model is presented in Table \@ref(tab:12cumswoldtab).
 

### 2017-07-29 article.Rmd
Small corection in title

@@
 ---
-title: "Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013""
+title: "Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013"
 author: Sören Edvinsson and Göran Broström

### 2017-07-29 article.Rmd
trivial change

@@
 ---
-title: "Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013"
-author: Sören Edvinsson and Göran Broström
+title: Is high social class always beneficial for survival? Long-term perspectives
+  of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013
+author: "Sören Edvinsson and Göran Broström"
 date: "24 July 2017"
 output:
-  bookdown::pdf_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
   bookdown::html_document2:

### 2017-07-29 article.Rmd
fix title

@@
 author: "Sören Edvinsson and Göran Broström"
-date: "24 July 2017"
+date: '`r Sys.Date()`'
 output:

### 2017-07-30 article.Rmd
add word_output

--- 2017-07-29 article.Rmd
@@
 author: "Sören Edvinsson and Göran Broström"
-date: "24 July 2017"
+date: '`r Sys.Date()`'
 output:
+  bookdown::word_document2:
+    toc: yes
+    toc_depth: 2
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::html_document2:

### 2017-07-30 article.Rmd
miktex works

--- 2017-07-29 article.Rmd
@@
 author: "Sören Edvinsson and Göran Broström"
-date: "24 July 2017"
+date: '`r Sys.Date()`'
 output:
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::html_document2:
@@
     toc_float: yes
-  bookdown::pdf_document2:
-    number_sections: yes
+  bookdown::word_document2:
     toc: yes

### 2017-08-21 article.Rmd
encoding issue?

--- 2017-07-29 article.Rmd
@@
 ---
-title: Is high social class always beneficial for survival? Long-term perspectives
-  of social inequality in mortality among adults and elderly in Northern Sweden 1851--2013
+title: Is high social class always beneficial for survival? Social inequality in mortality
+  among adults and elderly in Northern Sweden 1851--2013
 author: "Sören Edvinsson and Göran Broström"
-date: '`r Sys.Date()`'
+date: "`r Sys.Date()`"
+documentclass: article
+citation-package: natbib
+bibliography: "mybib.bib"
+biblio-style: "apalike"
 output:
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::html_document2:
@@
     toc_float: yes
-  bookdown::pdf_document2:
-    number_sections: yes
+  bookdown::word_document2:
     toc: yes
     toc_depth: 2
-link-citations: yes
-bibliography: mybib.bib
-biblio-style: apalike
+    toc_float: yes
 ---
@@
 determinants of health outcomes and therefore deserves their appellation as
-'fundamental causes' of disease and death" [@liph95]. For a long time,
+'fundamental causes' of disease and death" [@liph95].
+The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in the theory of @liph98.
+For a long time,
 the general view has been that socio-economic health and mortality inequalities were high in
@@
 maintained during the transition, the decline set in earlier and was faster
-among privileged groups. These assumption if partly supported by data from England and Wales [@pamuk85].
+among privileged groups. These assumption if partly supported by data from England and Wales [@pamuk85, @wowi95].
 
@@
 ```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in"}
-include_graphics("myfigs/sweden.png")
+knitr::include_graphics("myfigs/sweden.png")
 ```
@@
 [@edv92, @edv04, @edli11, @edbr12]. When it comes to infant and child
-mortality pre---and early---transitional time, the expected socio-economic
+mortality during pre---and early---transitional time, the expected socio-economic
 differences was often but not always confirmed [@edv92, @edv04],
@@
 particularly among adult men, not found despite the well-documented
-socio-economic differences [@edli11, @edbr12]. There were however large
-differences between men and women and for different age groups. These
-earlier studies have suggested that distinct and consistent differences to
-the advantage of higher classes developed somewhere during and probably
-quite early in the 20th century [@bevp11]. Until now we have however
-not been able to investigate the continuous development on mortality
-differences among the adult and elderly population due to no available
-data. This is the first time it has been possible, and the results continue
-to surprise us. The consistent advantage for higher social classes do not
-appear until quite late in the regions we have analysed. The social pattern
-of female mortality are in line to what is expected, but a social gradient
-in male adult mortality seems to be a recent phenomenon. Furthermore, this
-transition and the class pattern of mortality differ depending on age
-group. Inequality in mortality appear clearer for the population at working
-age, while the differences are smaller among the elderly.
-
-The results of small socio-economic gradient in the past and its late
-development are however in accordance with other studies [@bevp11],
-for example by @bedr11 in a very different setting, a couple of
-Scanian parishes, although with somewhat different social categories. They
-found the first indications of an emerging social gradient in the
-1950s. Similar results have also been found in Estonia [@jaadla17].
-
-When it comes to specific causes of death, the analysis shows that
-cardiovascular diseases contributes to a large proportion of all deaths
-during the 20th century, and we are not surprised that it has a fairly
-similar pattern as for all-cause mortality. Cancer is sometimes considered
-less connected to social class, and we find mixed results in our analyses.
-
-The surprising results makes it reasonable to consider the validity and
-possible weaknesses in the study. One obstacle concerns the quality of
-available information. Occupational titles in historical sources are not
-always distinct enough to define the correct class. It can be difficult to
-identify the skill level in the working class population because many were
-often only designated as labourer. The size of the enterprise for free
-entrepreneurs is unknown and it may be impossible to distinguish if they
-are self-employed or have employees. The class system and the position
-of different classes can also be discussed. Studies covering long period
-have the problem that occupations and classes change character - they do
-not represent the same thing. Nevertheless, these ambiguities and possible
-mis-classifications are not of that sort and magnitude that it would
-radically change the basic patterns observed. The social classes identified
-during the different periods represent distinct groups with very large
-differences in resources, capacities and power. Available data, for example
-from the 1930 census confirm that class differences in income and wealth
-were large, often enormous.
+socio-economic differences in society [@edli11, @edbr12].
+Until recently it has not been possible to investigate the development on mortality differences among the adult and elderly population.
+The present study is unique in its time perspective by investigating the development from the 19th century until the present.
+The results probably surprise most readers, and partly surprised us as well.
+The consistent advantage for higher social classes do not appear until quite late in the investigated regions.
+The results of small socio-economic gradient in the past and its late development are however in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
+They find the first indications of an emerging social gradient in the 1950s.
+Similar results have also been found in Estonia [@jaadla17].
+A study of mortality and social class in Sweden 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total and cardiovascular mortality [@vano89].
+Apparently high social class is not always beneficial for survival.
+
+There were however large differences in the pattern between men and women as well as between different age groups.
+The social pattern of female mortality are in line with the expected, but a social gradient in male adult mortality is a recent phenomenon.
+Furthermore, inequality in mortality appears clearer for the working—age population, while the differences are smaller among the elderly.
+Concerning specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
+Cancer is often considered less connected to social class, and we find mixed results in our analyses.
+
+The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
+Some of the limitations have been addressed in the presentation of the variables---The fact that social class was defined according to household until 1950 but individually defined from 1960 (mainly affecting the social class of women), different definitions of households during the same two periods, some differences in parishes included at different periods.
+These differences are however not of the magnitude that the main conclusions would be affected.
+Most women would end up in the same category as their husbands.
+
+One obstacle concerns the quality of available information.
+Occupational titles in historical sources are not always distinct enough to define the correct class.
+Studies covering long periods face the problem that occupations and classes change character---they do not represent the same thing.
+Nevertheless, these ambiguities and possible misclassifications should not radically change the basic patterns observed.
+The social classes identified during the different periods represent distinct groups with very large differences in resources, capacities and power.
 
 In this paper we have analysed social inequality in health from the
-perspective of social classes using Hisclass. However, there are
+perspective of social classes. However, there are
 alternative ways to analyse social inequality, and we cannot exclude that
-such analyses gives other results or at least would modify the image
-presented here. In many modern studies categorisations based on income,
+such analyses give other results or at least would modify the conclusions suggested here. In many modern studies categorisations based on income,
 wealth or education are common for studies of inequality. They represent
@@
 data on income and education are rarely available in historical population
-databases.  As a test, we have included analyses of the associations of
-education and income to mortality for recent decades. For women as
-well as for men, for both age groups and for all-cause mortality as well as
-cause-specific mortality we find consistent differences with higher
-educational level and more income leading to lower mortality, thus
-basically confirming what we find in the class analysis. We do not know if
-education and income are equally important in the past.
-
-The WHO Commission on Social Determinants of Health framework for the
-social determinants of health provides a model on how socio-economic
-position influence health. Social class works through intermediate
-determinants related to exposure and vulnerability [@who10]. People
-are situated within socio-economic and political contexts, as well as
-prevailing value systems in society. Socio-economic position imply
-different access to resources, power and prestige, all being aspects of
-crucial importance. The framework takes a holistic approach to the social
-determinants, including not only the effect of pure economic and material
-resources. The importance of understanding people in their context is also
-emphasised by for example @kri05. What she calls embodiment refers
-to the recognition that human beings are simultaneously social beings and
-biological organisms. The WHO framework  [@who10] lists three
-different types of intermediary social determinants of health; material
-circumstances, behaviours, biological factors, and finally psycho-social
-factors. Most studies focus on material circumstances. Having a good
-economy gives advantages in many aspects of relevance for health and
-mortality, for example good living conditions, access to food, medical care
-and better working conditions. Even though these advantages were at hand in
-our studied regions, the final result was not better health for men in the
-most privileged groups until the late 20th century. We must therefore
-consider other explanations. Psycho-social factors have been frequently
-discussed during the most recent decades, for example by @mar04, who
-argues that a higher status leads to better health, regardless of where in
-the status ladder we make our observations. Psychosocial factors are also
-suggested by @wipi09 as an explanation why equal societies always
-perform better than unequal ones when it comes to health and other social
-conditions. In the case studied here, we do not find any indication of a
-status syndrome, since mortality levels was in the opposite directions to
-the one we would expect from these theories.
-
-This makes us consider the possible impact of behavioural factors. Our
-results of the gender aspect might be of interest in this case. When it
-comes to gender differences, the present study confirms what we previously
-have found in the Sundsvall region [@edli11]. There is a clear
-gendered pattern when it comes to social inequality in mortality.  Within
-the same households, the effect of social class seems to lead to different
-results, high social position being beneficial for women but not for
-men. We might consider the role of behaviours based on different gendered
-expectations. Having a high social status could be expressed in attitudes
-and behaviours that were harmful for health (this is discussed in more
-detail in @edli11. Alcohol has been suggested as part of the large
-gender differences in the 19th century [@wil99, @edv92], and a more
-risk-taking life may be a component of male mortality [@cour00]. There
-are however some problems with attributing most of the male excess
-mortality to heavy drinking. Alcohol consumption decreased substantially
-during the period late 19th century to the 1950s, but the old pattern in
-social health inequalities remained. Swedes at that time had very low
-consumption levels. Alternative explanations should also be
-considered. There were strong and heavy requirements of being successful
-for the male \"elite. Many were involved in businesses with high risks and
-they may have felt a constant threat of economic failure.
-
-Our study illustrates the importance of considering contextual factors and
-the restrictions of making general statements. We need to look into all
-aspects of social classes, not only the material ones but also what this
-means for behaviour, norms, way of living. @edli11 suggest that an
-explanation for the larger social health inequalities in present-day
-society is that high status is expressed through health, being fit and
-physically active, while in many historical contexts higher social status
-(for men) was expressed in low physical activity, obsessive eating,
-drinking, smoking and sometimes also taking risks.  Different historical
-contexts promote different ways to spend economic resources.This is also in
-line with what @rasp06 and @jaadla17 suggest. If this
-interpretation is correct, the hypothesis of @ant67 on an increasing
-role of health related behaviour in the modern low-mortality régime,while
-economic inequality determined the social mortality pattern during the
-early phases of the mortality transition. Instead, health-related seems to
-be at least equally strong earlier in history but having a completely
-different expression.
-
-One of the major findings in this study relate to gender. As several
-researchers have pointed out, demography has often ignored power relations
-and different roles within families [@watkins93, @mack95, @jans07]. By
-investigating possible differences in mortality patterns according to sex,
-we have identified gendered patterns within the family. Different patterns
-between the sexes may be caused by discrimination on the female part in the
-household. We know that the society was strongly segregated according to
-gender and that women were discriminated and had less space for action. In
-our case, we do not find this to have negative consequences for women,
-indicating that the relation between sexes in the society was characterised
-by distinct gendered expectations on behaviour.
-
-Another question is if our results can be generalised to other historical
-contexts. We have studied environments in northern Sweden without any large
-metropolis. Perhaps Sweden and/or this part of Sweden is a special
-case. The social differences may have been smaller, the involvement of
-government both at the national
-and local level may have been stronger, thus mitigating socio-economic
-disadvantages. In comparison with many other countries, the level of
-mortality has been fairly low. These are relevant questions, which
-hopefully can be addressed in other and maybe comparative studies.
-
-Finally, we argue that our results have implications for the understanding
-on several other hot research topics. A central and hot debated question
-concerning the mortality transition is how the enormous increase in
-survival was accomplished. @mck76 argues that the improvements up to
-the middle of the 20th century was primarily caused by improved
-nutrition. This implicate that having more resources would lead to better
-survival. If research shows that the most advantaged groups did not have
-better health and longer lives, this do not necessarily speak against
-McKeown's ideas but it makes the question more complex. The results are
-also difficult to fit into @liph95 hypothesis that social conditions
-should be seen as a fundamental cause of disease and not only expressions
-of different disadvantages. In the historical context analysed here, this
-idea is not supported.
-
-In recent years, the question about the relation between health and income
-inequality been widely discussed [@wipi09]. Even though our study do
-not address this issue directly, our results are of interest for this
-question as well. The
-development of the social inequality in mortality do not seem to fit that
-well into the development of income and wealth inequalities. The early 20th
-century, in Sweden as well as in other countries, was characterised by very
-large economic differences, still this did in our regions not lead to any
-consistent advantage for the highest social classes. After the Second World
-War, economic differences diminished, and was at its lowest in Sweden at
-about 1980, i.e. at a time when a more expected pattern of a social
-hierarchy in survival had appeared. The period thereafter is characterised
-by increasing economic differences, and this may coincide with increasing
-health differences as well. The long-term relation between economic
-inequality and health is worth further studies.
+databases.
+
+How does the results correspond to research on other health measures?
+In a study of southern Sweden, @oberg14 find large social differences in body height, a measure that is assumed to sum up health during life.
+This does not contradict what we find here.
+Height is mainly determined by conditions in early years, and earlier studies in Sweden and in the areas studied here [@bredro02, @bube01, @mol17, @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
+
+The results of our study seem to contradict many of the dominating theories in present—day health research.
+In an overview of explanations for the persistence of health inequalities (although primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital et cetera.
+Most of these explanations are however based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
+One of these leading theories is the hypothesis of @liph95 that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to the social position.
+Others emphasize psycho-social factors, for example @mar04, who argues that a higher status leads to better health regardless of where in the status ladder individuals are situated.
+Psychosocial factors are also suggested by @wipi09 as an explanation why equal societies always perform better than unequal ones when it comes to health and other social conditions.
+In the long-term perspective provided here, neither of these theories fits the observation in earlier history.
+Their validity are mainly apparent for the most recent decades.
+To be fair, however, our study do not explicitly test for the possible impact of different material or psycho-social mechanisms.
+They could still be valid and influence the social pattern of mortality.
+What we can establish is that our study show that there is no necessary eternal social gradient favouring higher social classes.
+The lacking social hierarchy may be caused by other mechanisms having stronger impact on the pattern.
+We might need to consider the possibility of socially countervailing mechanisms [@lufr05].
+
+The historical perspective shows the complexity of how health and mortality is socially constructed and illustrates the need to contextualize our analyses.
+The dominant theories are not necessarily wrong but they may need to be further developed and refined.
+The WHO Commission on Social Determinants of Health framework provides a holistic model on how socio-economic position influences health.
+Social class works through intermediate determinants (material circumstances, behaviours, biological factors and psycho-social factors) related to exposure and vulnerability [@who10].
+People are situated within socio-economic and political contexts, as well as prevailing societal value systems.
+The importance of understanding people in their context is also emphasized by @kri05.
+What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
+Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, the final result was not better health for men in the most privileged groups until the late 20th century.
+We must therefore consider other explanations.
+
+This makes us consider the possible impact of behavioural factors.
+Our results of the gender aspect is important in this respect.
+As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93, @mack95, @jans07].
+The gender difference in mortality may be caused by discrimination on the female part in the household.
+We know that historical Sweden was strongly gender segregated, discriminating women and giving them less space for action.
+In our study as well as in a previous study [@edli11], we do not find this to have negative consequences for women.
+Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men.
+Instead, we might consider the role of behaviours based on distinct gendered and class expectations.
+Having a high social status could be expressed in attitudes, norms and way of living that could be harmful for health.
+@edli11 suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and sometimes also taking risks.
+Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99, @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
+Different historical contexts promote different ways to spend economic resources.
+This is also in line with what @rasp06 and @jaadla17 suggest.
+
+In the present-day high-income world, smoking and drinking contributes to a substantial part of the socio-economic health differences [@mapemy14].
+Our study suggests a strong impact also in historical Sweden.
+If this interpretation is correct, the hypothesis of @ant67 on an increasing role of health related behaviour in the modern low-mortality régime, while economic inequality determined the social mortality pattern during the early phases of the mortality transition can be questioned.
+Instead, health-related can be equally strong earlier in history but with a completely different expression.
+To corroborate this claim, we would need to have evidence on the social differentiation of consumption patterns and behaviours in history.
+Unfortunately, such data are rarely available before the 1960’s.
+At best, we have circumstantial and anecdotal evidence.
+Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
+@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained the higher tobacco consumption among men in these groups.
+The social pattern among women was however reversed, something that is in line with the present study.
+There is no solid information on smoking in different social classes before the 1960, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians.
+There was also an association between higher incomes and consumption [@loco94].
+The same lack of statistical evidence characterizes alcohol consumption in different social classes before 1960s.
+Access to alcohol was however easier for the wealthy during the period of strong alcohol restriction 1915-1955.
+Overweight and obesity was also probably higher in higher social classes.
+
+An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
+Even though our study do not address the hypothesis raised by [@wipi09] directly, our results are of interest for this question as well.
+The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
+After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
+at a time when a more expected pattern of a social hierarchy in survival appeared.
+The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
+The long-term relation between economic inequality and health require further studies.
+
+A relevant question is if our results can be generalised to other historical contexts.
+We have studied environments in northern Sweden without any large metropolis.
+Perhaps Sweden and/or this part of Sweden is a special case.
+The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages.
+The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century.
+In comparison with many other countries, the level of mortality has been fairly low.
+These are relevant questions, which hopefully can be addressed in other and maybe comparative studies.
 

### 2017-08-21 article.Rmd
why?

--- 2017-07-30 article.Rmd
@@
     toc_float: yes
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::word_document2:

### 2017-08-21 article.Rmd
Adding conclusion

--- 2017-08-21 article.Rmd
@@
 socio-economic differences in society [@edli11, @edbr12].
-Until recently it has not been possible to investigate the development on mortality differences among the adult and elderly population.
-The present study is unique in its time perspective by investigating the development from the 19th century until the present.
-The results probably surprise most readers, and partly surprised us as well.
-The consistent advantage for higher social classes do not appear until quite late in the investigated regions.
+The results presented here show that the consistent advantage for higher social classes do not appear until quite late in the investigated regions.
 The results of small socio-economic gradient in the past and its late development are however in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
@@
 A study of mortality and social class in Sweden 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total and cardiovascular mortality [@vano89].
-Apparently high social class is not always beneficial for survival.
-
-There were however large differences in the pattern between men and women as well as between different age groups.
-The social pattern of female mortality are in line with the expected, but a social gradient in male adult mortality is a recent phenomenon.
-Furthermore, inequality in mortality appears clearer for the working—age population, while the differences are smaller among the elderly.
-Concerning specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
-Cancer is often considered less connected to social class, and we find mixed results in our analyses.
 
@@
 
+\section{Conclusion}
+In this article we set up to analyse social differences in mortality from a long-term perspective.Until recently it has not been possible to investigate the development on mortality differences among the adult and elderly population.
+Still, the prevailing idea in many parts of the scholarly society is that strong socio--economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to the large class differences.
+The present study is unique in its time perspective by investigating the development from the 19th century until the present.
+The results probably surprise most readers, and partly surprised us as well.
+The consistent advantage of belonging to the elite have not always existed and the modern pattern of social divide in health appeared much later than expected in the regions we have studied.
+Apparently high social class is not always beneficial for survival.
+
+There were however large differences in the pattern between men and women as well as between different age groups.
+The social pattern of female mortality are in line with the expected, but a social gradient in male adult mortality is a recent phenomenon.
+Furthermore, inequality in mortality appears clearer for the working—age population, while the differences are smaller among the elderly.
+Concerning specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
+Cancer is often considered less connected to social class, and we find mixed results in our analyses.
+
+To explain the observed differences is difficult and we can only provide tentative hypotheses.
+We have discussed the possible role of behavioural influences on mortality in different historical context, suggesting that such aspects are not only important in the modern epidemiological regime but might have been decisive in earlier phases of the epidemiologic transition as well.
+The results indicate that we should consider the role of gender and class expectations of behaviour.
+Unfortunately, the evidence for behavioural differences in history are usually circumstantial and anecdotal, at least when it comes to social class.
+Studies of these aspects and their role for the health and mortality in different times and places require further development.
+
 An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not address the hypothesis raised by [@wipi09] directly, our results are of interest for this question as well.
+Even though our study do not directly address the hypothesis raised by [@wipi09], our results are of interest for this question as well.
 The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
@@
 
-A relevant question is if our results can be generalised to other historical contexts.
+An important question is if our results can be generalised to other historical contexts.
 We have studied environments in northern Sweden without any large metropolis.
@@
 In comparison with many other countries, the level of mortality has been fairly low.
-These are relevant questions, which hopefully can be addressed in other and maybe comparative studies.
+This is a relevant question, which hopefully can be addressed in other and maybe comparative studies.
 

### 2017-08-24 article.Rmd
cape town

@@
 author: "Sören Edvinsson and Göran Broström"
-date: "`r Sys.Date()`"
-documentclass: article
-citation-package: natbib
-bibliography: "mybib.bib"
-biblio-style: "apalike"
+date: '`r Sys.Date()`'
 output:
@@
     toc_float: yes
+bibliography: mybib.bib
+documentclass: article
+citation-package: natbib
+biblio-style: apalike
 ---

### 2017-08-28 article.Rmd
Revisions until causes of death analyses

@@
 ---
-title: Is high social class always beneficial for survival? Social inequality in mortality
-  among adults and elderly in Northern Sweden 1851--2013
+title: Is high social class always beneficial for survival? A study of adults and elderly in Northern Sweden 1851--2013
 author: "Sören Edvinsson and Göran Broström"
@@
 
-In the present paper, we take a long-term perspective on the development of
-social inequalities in mortality in the adult and elderly population.
-The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå.
-This issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
+A major theme in demographic and epidemiological studies, is the seemingly persistent effect of social class on mortality.
+In the present study, we approach this issue by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
+Our main focus is on the relative differences.
+The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden for the period 1851--2013 for Skellefteå and 1901--2013 for Umeå.
+The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
 Furthermore we put this in he context of how the inequality in mortality is associated
@@
 
-On the basis of these results, we argue that a consistent hierarchical social gradient in mortality appears surprisingly late, but that there were large gender differences in trends and levels.
-We suggest that more attention should be paid on health-related behaviour also in earlier phases of the mortality transition.
+On the basis of these results, we argue that high social class is not necessarily favourable for survival.
+This is based on the finding that a consistent hierarchical social gradient in mortality appears surprisingly late in the studied regions, but that there were large gender differences in trends and levels.
+We suggest that health-related behaviour was important not only in present-day societies, but were decisive also in earlier phases of the mortality transition.
+The results implicate that the association between social class and health are more complex than is assumed in many of the dominant theories in demography and epidemiology.
 
@@
 resources provides advantages in all aspects of life, the health advantage of higher
-classes ought to be obvious. @ant67 and @smi83 suggest that social
+classes ought to be obvious. @ant67 suggests that social
 inequality in mortality has passed through different historical phases. According
-to them, differences were comparatively small during the
+to him, differences were comparatively small during the
 pre-transitional phase when space was a strong determinant for the spread of
@@
 health-related behaviour became the decisive determinant for health and survival.
-@omran82 comes to a similar conclusion states, in the third proposition in his theory of the
-epidemiological transition, that even if the class differential in mortality was
+@omran82 comes to a similar conclusion in his theory of the epidemiologic transition. He states in the third proposition in the theory, that even if the class differentials in mortality were
 maintained during the transition, the decline set in earlier and was faster
-among privileged groups. These assumption if partly supported by data from England and Wales [@pamuk85, @wowi95].
+among privileged groups. These assumptions are primarily based on studies of England and Wales [@ant67, @wowi95], for example @pamuk85.
 
 Recent studies, investigating social inequality in health and mortality
-with micro-data, has however questioned how generalizable the assumed
+with micro-data, has however questioned how general the assumed
 pattern is [@bevp11]. Instead, some studies suggest that the modern
@@
 differences during the 19th and early 20th century [@bedr11,
-@edli11,@edbr12], sharply contrasting from the situation in present-day society.
-
-It thus appears as if a more consistent social health divide appeared at
+@edli11,@edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
 some point during the mortality transition, at least in Sweden. Solid
@@
 better understand the role of socio-economic conditions for health and survival over time.
-Furthermore, previous studies indicate substantial differences in the development according to age and gender.
 
@@
 direct effect of actual access to economic resources. They argue
-that unequal societies perform less well when it comes to health than equal
-ones. Wilkinson and Pickett find a positive association between inequality
-and health (as well as other social conditions) at country level in the
-present-day economically developed world. This has initiated a vital
-scholarly debate and the topic has been extensively studied [@suka04,
-@wado00]. How the association looked like in previous periods is
-unknown. This is of particular interest since the levels of economic
+that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the present-day economically developed world.
+This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04,
+@wado00].
+During the last decades, the association between trends of inequalities in mortality and income respectively are weak or non-existant according to @hoetal16.
+How the association looked like in previous periods is unknown.
+This is of particular interest since the levels of economic
 inequality were very different from those of the recent decades. Even if it
 not a necessary implication, it is reasonable to assume that poorer groups were
-those most disadvantaged of large inequality in itself.
-During the last decades, the association between trends of inequalities in mortality and income respectively are weak or non-existant according to @hoetal16.
+those most disadvantaged of large inequalities.
 
@@
 was dominated by agriculture, making it vulnerable to harvest failures, and
-several severe famines affected the regions during the 1800s. One living
-long in memory was the famine of 1867–-1868 [@segb14]. During the long
-winters, sea communication was hindered due to the Gulf of Bothnia being
-frozen. The first ships during the year came to this area in May or even as
+several severe famines frequently affected the regions during the 1800s, for example after the harvest failure in 1867 [@segb14].
+During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
+The first ships during the year came to this area in May or even as
 late as June in some cases [@fah56]. Towards the end of the 19th
-century, the Swedish railway system reached this part of Sweden, thereby
-facilitating contacts with the rest of Sweden, improving the economy and
-mitigating the effects of harvest failures. The inland was particularly
+century, the Swedish railway system reached this part of Sweden. Thereby
+contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. The inland was particularly
 remote and isolated. During the 19th century, our regions became
 increasingly integrated in the same epidemiological pattern as the rest of
-Sweden. Diseases reached to places that were spared in earlier times.
+Sweden.
 
@@
 1950s makes it possible to investigate the long-term development of social
-inequality in mortality, in the present case from 1851 to 2013. The
-databases provide us with micro-data on a sufficiently large population and
+inequality in mortality in a part of Sweden. The available
+databases  The data for the present study comes from two large
+population databases at the DDB, whom provide us with micro-data on a sufficiently large population and
 with substantial social diversity for the Skellefteå and Umeå regions
-[@wee16]. The data for the present study comes from two large
-population databases at the DDB. The early period until 1950 is covered by
+[@wee16]. The early period until 1950 is covered by
 the database Poplink, the digitisation of historical parish registers for
@@
 also makes it impossible to add information on individuals in the Linnaeus
-database from what we potentially could find in Poplink, for example family
-background or occupations earlier in life.
+database from what we potentially could extract from Poplink, for example family
+background or previous social class.
 
@@
 being resident in either of the regions are included. The data file
-contains the variables on social class (Hisclass) sex, urban/rural
+contains the variables social class, sex, urban/rural
 residence, birth date, death date, first and last date of observation and
@@
 the Linnaeus data makes it necessary to apply different approaches when it
-comes to identification of presence periods within our
-regions. The Poplink data provides us with exact dates or at least year of
-start and exit of the presence of all individuals in our data, allowing us
+comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of the presence of all individuals in our data, allowing us
 to have full and continuous control of the de jure population. This is
 however not the case with the Linnaeus database. For the period 1960--2000
-we use presence in the censuses 1960--1990 together with the yearly
-population registers 1986--2000 and information on deaths from the National
+we use presence in the censuses 1960--1990 and information on deaths from the National
 Board of Health and Welfare to identify the observation periods in our
 analyses. There is however no full information on what happened in-between
-the censuses, for example the exact date of exit for persons that moved
+the censuses, for example on the exact date of exit for persons having moved
 between two censuses. We have solved this by considering everyone as
 resident in our regions during the complete five-year period between
-censuses if the person is missing in the later census (the last period covering ten years). This overestimates
-their presence in our regions, but should not bias the results when it
+censuses if the person is missing in the later census (the last period covering ten years). This overestimates their presence, but should not bias the results when it
 comes to socioeconomic differences. We furthermore miss those that both
@@
 In this paper we analyse mortality differences according to social class
-based on the classification scheme Hisclass [@lema11]. The only
+based on the classification scheme Hisclass in modified form [@lema11]. The only
 type of socio-economic information available both in the Linnaeus and Poplink database
@@
 thus under-estimating female working-force participation as well as that of
-adult children residing with their parents [@vik10]. In this study and
-according to the research questions in the present paper, we have chosen to
-categorise wives according to position of the head of household, usually
+adult children residing with their parents [@vik10]. For the Poplink period (until 1950), we have chosen to categorise wives according to position of the head of household, usually
 the husband/father, assuming that the family shares the same socio-economic
@@
 different periods are not completely comparable and makes the
-interpretation of the development for women more uncertain.
-Notice however that there is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way if we use social class on household level also for the Linneaus data.
+interpretation of the development for women uncertain for some periods.
+Notice however that there is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data.
 
@@
 proportion without social position is very low until the middle of the 20th
-century. This is a reflection of how womens social position is defined for
-that period. Missing social class is however a real problem during the
-1960s and 1970s in particular. Around 80 \% lack information for the 1960s
-and 50 \% for the 1970s. Since social class for this period is defined
+century, reflecting the way womens social position is defined.
+Missing social class is however a real problem during particularly the 1960s and 1970s.
+Around 80 \% lack information for the 1960s and 50 \% for the 1970s.
+Since social class for this period is defined
 according to own occupation, the consequence of many women being outside
@@
 cancers. National compulsory reporting of cause of death was introduced in
-1911, thus our analysis of the two disease groups starts from that
-year. Classification systems have however changed. Sweden did not join ICD
+1911, but our analysis of the two disease groups starts from 1901 since deaths were fairly frequently reported even before 1911. Classification systems have however changed. Sweden did not join ICD
 until 1952, but the historical causes of death in the Poplink database have
-been coded according to the ICD10 system. In the present study, we have
-identified the codes relevant for cardiovascular diseases and cancers in
+been coded according to the ICD10 system. In the present study, we identify the codes relevant for cardiovascular diseases and cancers in
 the ICD 8, 9 and 10.
@@
 last period (2008--2013) coincides with the international financial
-crisis.
+crisis. During the last decades income inequality has increased considerably.
 
@@
 and graphics [@cran], especially using the package **eha**
-[@gb12,@eha].
-The proportional hazards model allow us to scrutinise the differences and
+[@gb12,@eha]. The proportional hazards model allows us to scrutinise the differences and
 the development in more detail. We divide the analysis into two age groups,
@@
 were still mainly in the workforce, and the second those that mainly were
-retired. The model controls for marital
-status and whether the individual resides in the urban or rural parts of the regions. There
-are no serious signs of deviation from proportionality assumption, except
-regarding docial class. Each period are analysed separately, but we then combine results
+retired.
+The model controls for marital status and whether the individual resides in the urban or rural parts of the regions.
+There are no serious signs of deviation from proportionality assumption, except
+regarding social class. Each period are analysed separately, but we then combine results
 in order to show the development of the social patterning of mortality

### 2017-08-29 article.Rmd

@@
 position for women before and after 1950. After 1950 it is based
-on their own occupation. Being without occupation might include women that
-are more frail both when it comes to health and social characteristics.
-
-A different pattern appears among men, being the most surprising finding in
+on their own occupation. Being without occupation probably include women that
+are more frail when it comes to health and vulnerable concerning social characteristics.
+
+A completely different pattern appears among men, being the most surprising finding in
 our analyses. The groups we would expect to have the lowest mortality, if
-we assume that access to different resources and belonging to a high social class
+we assume that access to important resources and belonging to a high social class
 determines survival, instead have by far the highest hazards of dying
@@
 unknowns social class is probably even more relevant for men. They probably
-represent groups in society that are particularly vulnerable, having difficulties getting into the laour force.
+represent groups in society that are particularly vulnerable, having difficulties getting into the labour force.
 
 A further observation can be made from Figure \@ref(fig:cumsw). The gender
-difference in death hazards is very big for the élite group from
+difference in death hazards is extremely large for the elite group from
 1851--1950, men having about three times higher risks than women in the
-same class. In the other classes, the difference was much smaller. It was
+same class (and in the same households). In the other classes, the difference was much smaller. It was
 men in high position that experienced the largest risks.
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we found for the younger age group (see Figure \@ref(fig:cumswold). For
+we find for the younger age group (see Figure \@ref(fig:cumswold). For
 women the results are mixed. Once again, too much missing data on social
-class leaves out the results for the 1960s and 1970s. We do not find the
-same clear advantage for élite women during the period before 1950, but neither
+class force us to leave out the results for the 1960s and 1970s. We do not find the
+same clear advantage for elite women during the period before 1950, but neither
 do we find any large disadvantage for working class elderly women at that
-time. In some periods, the social differences were small, and the élite
+time. In some periods, the social differences were small, and the elite
 even had substantially higher mortality during the early 20th century. The
-social differences continued to be small during the last decades of the
+social differences continued to be small until the last decades of the
 20th century. Only during the two last periods (after the year 2000) is it
@@
 for the adult male population 40--64 years. Mortality was in fact
-substantially higher among élite men  but their disadvantage was most 
+substantially higher among elite men but their disadvantage was mainly 
 apparent during the 19th century. Differences are fairly small during the
@@
 In this and the following section, the focus is on analyses of mortality from the two
-dominant causes of death groups among the adult and elderly population. We start
+dominant groups of death causes in the adult and elderly population. We start
 with cardiovascular diseases. Many studies have shown that socio-economic
@@
 hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of
+main target in our analysis. These numbers are used in the process of
 graphically illustrating the development over time.

### 2017-08-29 CreateData/creDat2.Rnw

--- 2017-06-07 CreateData/creDat.Rnw
@@
 \title{Creating Data for "Social inequality in mortality among
-  adults and elderly in northern Sweden 1850--2013"\footnote{Version 2017-05-27.}}
+  adults and elderly in northern Sweden 1801--2013"\footnote{Version 2017-08-27, with a longer time period.}}
 
@@
 
-\subsection{POPLINK data, 1851--1950}
+\subsection{POPLINK data, 1801--1950}
 
-The POPLINK data cover 1901--1950. A covariate {\tt period} is added and used for splitting the data into calendar time periods of length 10 years each.
+In this version, we start at 1801-01-01. The POPLINK data cover 1801--1950. A covariate {\tt period} is added and used for splitting the data into calendar time periods of length 10 years each.
 
@@
 load("../../Data/adult.rda")
-adult <- calWin(adult, c(1851, 1951))
+adult <- calWin(adult, c(1801, 1951))
 source("R/addPer.R")
 addPer
-adult <- addPer(adult, cuts = seq(1851, 1951, by = 25))
+adult <- addPer(adult, cuts = seq(1801, 1951, by = 25))
 adult <- rc(adult)
@@
 
+Now let us put on emph{partner} on {\bf lisaold}.
+
+<<partner>>=
+load("../../Data/partner.rda")
+##females <- lisaold[lisaold$sex == "female", ]
+##par.fem <- partner[partner$LINNEID1 %in% females$LINNEID | partner$LINNEID2 %in% females$LINNEID2, ]
+partner$id1 <- pmin(partner$LINNEID1, partner$LINNEID2)
+partner$id2 <- pmax(partner$LINNEID1, partner$LINNEID2)
+partner <- unique(partner[, c("LINNEIDAR", "id1", "id2")])
+@
+
 \end{document}

### 2017-08-29 article.Rmd
Revisions in text and references

--- 2017-08-29 article.Rmd
@@
 The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
-Furthermore we put this in he context of how the inequality in mortality is associated
-with the development of economic inequality in society.The results are then discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
-
-On the basis of these results, we argue that high social class is not necessarily favourable for survival.
-This is based on the finding that a consistent hierarchical social gradient in mortality appears surprisingly late in the studied regions, but that there were large gender differences in trends and levels.
+Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.The results are then discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
+
+On the basis of the results we present, we argue that high social class is not necessarily favourable for survival.
+Mortality risks in different contexts must be understood in the intersection between class and gender.
 We suggest that health-related behaviour was important not only in present-day societies, but were decisive also in earlier phases of the mortality transition.
-The results implicate that the association between social class and health are more complex than is assumed in many of the dominant theories in demography and epidemiology.
+The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
 
@@
 maintained during the transition, the decline set in earlier and was faster
-among privileged groups. These assumptions are primarily based on studies of England and Wales [@ant67, @wowi95], for example @pamuk85.
+among privileged groups.
 
@@
 is not necessarily an obvious assumption when it comes to
-cancers. Different forms of cancers have different origin and
+cancer. Different forms of cancers have different origin and
 different pathways to disease and death. The social pattern of higher
@@
 Hisclass during the different studied periods for the age group
-40--64. Cancer mortality among the elite women was lower than that for the
+40--64. Cancer mortality among elite women was lower than that of the
 other classes during the complete studied period, while working class women
@@
 more consistent advantage of higher class in cancers compared to
-cardiovascular diseases for women, thus confirming what we have expected. We do however not find the same pattern
-for men. It is difficult to identify any general features of 20th century
+cardiovascular diseases for women, consistent with the idea of a social hierarchy in health.
+What is less expected, is that this we find a stronger association between social class and survival in cancer than cardiovascular diseases.
+We do however not find the same pattern
+for men. It is difficult to identify any general trends and features of 20th century
 male cancer mortality when it comes to social differences.
@@
 
-See Table \@ref(tab:12cumswoldtab).
-
 \section{Discussion}
@@
 [@edv92, @edv04, @edli11, @edbr12]. When it comes to infant and child
-mortality during pre---and early---transitional time, the expected socio-economic
+mortality during pre- and early-transitional time, the expected socio-economic
 differences was often but not always confirmed [@edv92, @edv04],
 something that has been found in other studies as well [@bredro02,
-@bube01, @mol17]. Studies based on microdata on the adult population in
+@bube01, @mol17].
+
+Studies based on microdata on the adult population in
 history have been rare, mainly due to lack of adequate data. The results
-from such studies show other patterns than that of infant and child
+from such studies show other patterns than that for infant and child
 mortality. The expected social gradient in survival, meaning that people
@@
 The results presented here show that the consistent advantage for higher social classes do not appear until quite late in the investigated regions.
-The results of small socio-economic gradient in the past and its late development are however in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
+
+The results showing a small or even reversed socio-economic gradient in the past and its late development are however in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
 They find the first indications of an emerging social gradient in the 1950s.
@@
 A study of mortality and social class in Sweden 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total and cardiovascular mortality [@vano89].
+These findings are however contrary towards what usually is assumed and what other studies have shown.
+Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67, @wowi95 @pamuk85].
+We know less about how general these results are, but apparently the Swedish context has been different from the English and Welsh.
 
 The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
-Some of the limitations have been addressed in the presentation of the variables---The fact that social class was defined according to household until 1950 but individually defined from 1960 (mainly affecting the social class of women), different definitions of households during the same two periods, some differences in parishes included at different periods.
-These differences are however not of the magnitude that the main conclusions would be affected.
-Most women would end up in the same category as their husbands.
+Some of the limitations have been addressed in the presentation of the variables: The fact that social class was defined according to household until 1950 but individually defined from 1960 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
+These differences are however not of the magnitude that the main conclusions are affected.
+Most women would end up in the same category as their husbands and the migrations between census dates or census date and death were not frequent.
 
@@
 alternative ways to analyse social inequality, and we cannot exclude that
-such analyses give other results or at least would modify the conclusions suggested here. In many modern studies categorisations based on income,
-wealth or education are common for studies of inequality. They represent
-different dimensions of inequality. One problem is however that reliable
-data on income and education are rarely available in historical population
+such analyses give other results or at least would modify the conclusions suggested here.
+In many modern studies categorisations based on income, wealth or education are common for studies of inequality.
+They represent different dimensions of inequality.
+It would be interesting to compare the results from this study with alternative measures, but reliable
+data on income and education are unfortunately rarely available in historical population
 databases.
@@
 How does the results correspond to research on other health measures?
-In a study of southern Sweden, @oberg14 find large social differences in body height, a measure that is assumed to sum up health during life.
+In a study of southern Sweden, @oberg14 finds large social differences in body height, a measure that is assumed to sum up health during the life span.
 This does not contradict what we find here.
-Height is mainly determined by conditions in early years, and earlier studies in Sweden and in the areas studied here [@bredro02, @bube01, @mol17, @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
-
-The results of our study seem to contradict many of the dominating theories in present—day health research.
-In an overview of explanations for the persistence of health inequalities (although primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital et cetera.
+Height is mainly determined by conditions in early years, and are not reflecting inequalities in adulthood.
+Earlier studies in Sweden and in the areas studied here [@bredro02, @bube01, @mol17, @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
+
+Our study seems to be in conflict with many of the dominating theories in present—day health research.
+In an overview of explanations for the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital et cetera.
 Most of these explanations are however based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
@@
 To be fair, however, our study do not explicitly test for the possible impact of different material or psycho-social mechanisms.
-They could still be valid and influence the social pattern of mortality.
-What we can establish is that our study show that there is no necessary eternal social gradient favouring higher social classes.
+These factors could still be valid and influence the social pattern of mortality.
+What our study establishes is that there there is no necessary eternal social gradient favouring higher social classes.
 The lacking social hierarchy may be caused by other mechanisms having stronger impact on the pattern.
-We might need to consider the possibility of socially countervailing mechanisms [@lufr05].
+We might have to consider the possibility of socially countervailing mechanisms [@lufr05].
 
@@
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, the final result was not better health for men in the most privileged groups until the late 20th century.
+Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
 We must therefore consider other explanations.
@@
 We know that historical Sweden was strongly gender segregated, discriminating women and giving them less space for action.
-In our study as well as in a previous study [@edli11], we do not find this to have negative consequences for women.
+In our study as well as in a previous study [@edli11], we do not find this to have negative health consequences for women.
 Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men.
 Instead, we might consider the role of behaviours based on distinct gendered and class expectations.
-Having a high social status could be expressed in attitudes, norms and way of living that could be harmful for health.
+Having a high social status is sometimes expressed in attitudes, norms and way of living potentially harmful for health.
 @edli11 suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and sometimes also taking risks.
 Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99, @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
-Different historical contexts promote different ways to spend economic resources.
-This is also in line with what @rasp06 and @jaadla17 suggest.
+Different historical contexts promote different ways to spend economic resources (compare @rasp06 and @jaadla17.
 
@@
 If this interpretation is correct, the hypothesis of @ant67 on an increasing role of health related behaviour in the modern low-mortality régime, while economic inequality determined the social mortality pattern during the early phases of the mortality transition can be questioned.
-Instead, health-related can be equally strong earlier in history but with a completely different expression.
+Instead, health-related can be equally strong earlier in history but expressed in a completely other way.
 To corroborate this claim, we would need to have evidence on the social differentiation of consumption patterns and behaviours in history.
@@
 There was also an association between higher incomes and consumption [@loco94].
-The same lack of statistical evidence characterizes alcohol consumption in different social classes before 1960s.
-Access to alcohol was however easier for the wealthy during the period of strong alcohol restriction 1915-1955.
+The same lack of statistical evidence characterizes alcohol consumption in different social classes before 1960s with some exceptions.
+Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with highest incomes.
+There was also an over-representation of alcohol-related deaths in higher social classes in the 1960 [@noro98].
+@noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers.
 Overweight and obesity was also probably higher in higher social classes.

### 2017-08-29 CreateData/creDat2.Rnw
from 1801

@@
 print(xtable(round(y, 1), digits = 0, 
-             caption = "Hisclass by time period, 1901--2013 (per cent)."))
+             caption = "Hisclass by time period, 1801--2013 (per cent)."))
 @
@@
 <<groupbirths, results = 'asis'>>=
-vb <- vb[vb$birthdate > 1761 & vb$birthdate < 1974, ]
+vb <- vb[vb$birthdate > 1711 & vb$birthdate < 1974, ]
 vb <- ageWin(vb, c(40, 90))
-starts <- seq(1761, 1981, by = 10)
+starts <- seq(1711, 1981, by = 10)
 vb$cohort <- cut(vb$birthdate, starts, labels = starts[-length(starts)])
@@
 levels(vb0$period)[8] <- "1991-2001"
-save(vb0, file = "../data/vb0.rda")
-save(vb, file = "../data/vb.rda")
+save(vb0, file = "../data2/vb0.rda")
+save(vb, file = "../data2/vb.rda")
 @
@@
 vbTab$age <- factor(vbTab$age)
-save(vbTab, file = "../data/vbTab.rda")
-save(vb0Tab, file = "../data/vb0Tab.rda")
+save(vbTab, file = "../data2/vbTab.rda")
+save(vb0Tab, file = "../data2/vb0Tab.rda")
 @
@@
                    FUN = sum)
-save(loTab, file = "../data/loTab.rda")
-save(lisaold, file = "../data/lisaold.rda")
+save(loTab, file = "../data2/loTab.rda")
+save(lisaold, file = "../data2/lisaold.rda")
 @
@@
 
-Now let us put on emph{partner} on {\bf lisaold}.
+Now let us put on \emph{partner} on {\bf lisaold}. First we need to fix the file {\bf partner} so that we can distinguish 
+females and males. We also remove homosexual couples: We want to check a female's status compared to her male partner's.
 
@@
 load("../../Data/partner.rda")
-##females <- lisaold[lisaold$sex == "female", ]
-##par.fem <- partner[partner$LINNEID1 %in% females$LINNEID | partner$LINNEID2 %in% females$LINNEID2, ]
 partner$id1 <- pmin(partner$LINNEID1, partner$LINNEID2)
@@
 partner <- unique(partner[, c("LINNEIDAR", "id1", "id2")])
+indx <- match(partner$id1, lisaold$LINNEID)
+partner$sex1 <- lisaold$sex[indx]
+indx <- match(partner$id2, lisaold$LINNEID)
+partner$sex2 <- lisaold$sex[indx]
+partner <- partner[!(is.na(partner$sex1) | is.na(partner$sex2)), ]
+partner <- partner[partner$sex1 != partner$sex2, ]
+dim(partner)
+partner$female <- ifelse(partner$sex1 == "female", partner$id1, partner$id2)
+partner$male <- ifelse(partner$sex1 == "male", partner$id1, partner$id2)
+partner <- data.frame(AR = partner$LINNEIDAR,
+                      female = partner$female,
+                      male = partner$male)
 @
 
+Next, put on partners on the females of {\bf lisaold}:
+
+<<putonpartner>>=
+key <- paste(lisaold$LINNEID, lisaold$AR)
+partner$key <- paste(partner$female, partner$AR)
+indx <- match(key, partner$key)
+lisaold$partner <- partner$male[indx]
+@
+
+Next, put on partner's \emph{education} and \emph{income}:
+
+<<putinceduc>>=
+males <- lisaold[lisaold$LINNEID %in% lisaold$partner, ]
+females <- lisaold[!is.na(lisaold$partner), ]
+indx <- match(females$partner, males$LINNEID)
+females$m.educ <- males$educ[indx]
+females$m.income <- males$income[indx]
+@
+
+Tabulating female's income and partner's income (\emph{m.income}):
+
+<<tabminc>>=
+x <- with(females, table(income, m.income))
+x
+@
+
+Tabulating female's education and partner's education (\emph{m.educ}):
+
+<<tabmeduc>>=
+x <- with(females, table(educ, m.educ))
+x
+@
+
+No comments!
+
 \end{document}

### 2017-08-29 R/get.file2.R

@@ -1,0 +1,12 @@
+get.file2 <- function(name, link = "http://capa.ddb.umu.se/data2/"){
+    filename = paste("data2/", name, ".rda", sep = "")
+    if (file.exists(filename)){
+        load(filename)
+    }else{
+        con <- url(paste(link, name, ".rda", sep = ""))
+        load(con)
+        close(con)
+        save(list = name, file = filename)
+    }
+    get(name)
+}

### 2017-08-30 article.Rmd
Adding references and making small revisions

@@
 
-In the present-day high-income world, smoking and drinking contributes to a substantial part of the socio-economic health differences [@mapemy14].
+In the present-day high-income world including Sweden, smoking and drinking contributes to a substantial part of the socio-economic health differences [@mapemy14, @syla16].
 Our study suggests a strong impact also in historical Sweden.
@@
 Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
+Overweight and obesity was probably higher in higher social classes.
 @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained the higher tobacco consumption among men in these groups.
@@
 There was also an over-representation of alcohol-related deaths in higher social classes in the 1960 [@noro98].
-@noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers.
-Overweight and obesity was also probably higher in higher social classes.
+@noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with a more skewed consumption distribution, indicating more high-consumers, in this group.
+A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
+Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being inconsistent in 1961 and reversed in 1991 with highest mortality in the lowest class [@cropr10].
 
 \section{Conclusion}
-In this article we set up to analyse social differences in mortality from a long-term perspective.Until recently it has not been possible to investigate the development on mortality differences among the adult and elderly population.
-Still, the prevailing idea in many parts of the scholarly society is that strong socio--economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to the large class differences.
+In this article we set our task to analyse social differences in mortality from a long-term perspective.
+Until recently it has not been possible to investigate the development on mortality differences among the adult and elderly population.
+Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to the large class differences.
 The present study is unique in its time perspective by investigating the development from the 19th century until the present.
@@
 There were however large differences in the pattern between men and women as well as between different age groups.
+Class as well as gender must be considered.
 The social pattern of female mortality are in line with the expected, but a social gradient in male adult mortality is a recent phenomenon.

### 2017-08-30 article.Rmd
Small changes

@@
 Recent studies, investigating social inequality in health and mortality
-with micro-data, has however questioned how general the assumed
+with micro-data, have however questioned how general the assumed
 pattern is [@bevp11]. Instead, some studies suggest that the modern
@@
 Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
-Overweight and obesity was probably higher in higher social classes.
+Overweight and obesity was probably higher and physical exercise lower in higher social classes.
 @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained the higher tobacco consumption among men in these groups.

### 2017-08-30 article2.Rmd
from 1801
check consistency (bad)

--- 2017-08-29 article.Rmd
@@
 Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
+Overweight, obesity and physical inactivity were probably more common in higher social classes.
 @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained the higher tobacco consumption among men in these groups.

### 2017-08-31 article.Rmd
Revisions at different places

@@
 The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
-Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.The results are then discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
+Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
+The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
 
@@
 'fundamental causes' of disease and death" [@liph95].
-The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in the theory of @liph98.
-For a long time,
-the general view has been that socio-economic health and mortality inequalities were high in
+The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in this theory [@liph98].
+For a long time, the general view has been that socio-economic health and mortality inequalities were high in
 historical societies, probably higher than in modern societies. This is a
 reasonable assumption since these societies were in most cases characterised by
-very large differences between groups and classes. Knowing that access to
+very large socioeconomic differences. Knowing that access to
 resources provides advantages in all aspects of life, the health advantage of higher
@@
 differences during the 19th and early 20th century [@bedr11,
-@edli11,@edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
+@edli11, @edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
 some point during the mortality transition, at least in Sweden. Solid
@@
 20th century, industrialisation took place, partly by the opening of mines
-in the in-lands. This also led to a population increase both in the town
+in the inlands. This also led to a population increase both in the town
 and in the rural parts, and a much more diversified economy.
@@
 
-The Umeå region in this study consists of Umeå urban and rural parish until
+The Umeå region consists of Umeå urban and rural parish until
 1950, and from 1960 onwards of Umeå municipality when another three
@@
 1950s makes it possible to investigate the long-term development of social
-inequality in mortality in a part of Sweden. The available
-databases  The data for the present study comes from two large
+inequality in mortality in a part of Sweden. The data for the present study comes from two large
 population databases at the DDB, whom provide us with micro-data on a sufficiently large population and
@@
 Yrkesklassificering).
-We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our treatment of presence periods.
+We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods.
 For the period 1991--2000 the social class in the census 1990 is considered to be kept
@@
 Longitudinal analyses of social classes for such a long period covered in this study involves several problems, related to changes in occupational structure and the meaning and contents of occupations and social positions.
-We have
-merged the original 12 classes in Hisclass, based on position in the
-labour market, ending up with the following three and an additional fourth
+We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth
 category with unknown class:
@@
 \begin{enumerate}
-\item {\tt elite}, Hisclass 1 and 2. Higher managers, professionals
-and higher white collar.
+\item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
 \item {\tt middle}, Hisclass 3, 4, 5, 6 and 8. Lower managers,
-farmers and lower white collar.
+farmers and lower professionals, clerical and sales personell.
 \item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
@@
 establishment of Umeå University explains much of this increase. What is
-invisible in Figure \@ref(fig:hiscskumm) is the transfer within the middle
+not apparent from Figure \@ref(fig:hiscskumm) is the transfer within the middle
 group from a majority of farmers during the 19th century to lower white
@@
 In the present study, we analyse mortality in the two most common disease
-groups in the ages we are investigating---cardiovascular diseases and
+groups in the ages we are investigating; cardiovascular diseases and
 cancers. National compulsory reporting of cause of death was introduced in
@@
 until 1952, but the historical causes of death in the Poplink database have
-been coded according to the ICD10 system. In the present study, we identify the codes relevant for cardiovascular diseases and cancers in
+been coded according to an adapted form of the ICD10 system. In the present study, we identify the codes relevant for cardiovascular diseases and cancers in
 the ICD 8, 9 and 10.
@@
 health risks higher in towns and cities. In the present paper, we control
-for this aspect by distinguishing between urban and rural residence, based
+for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
 on parish of residence.
@@
 industrialisation started and the urban environments increased in
-population size. The modern Swedish welfare state began to develop during this
-time. For the period 1960 onwards, there are separate analyses on decades
-except for the two last periods. Sweden had developed into a wealthy
+population size.
+The modern Swedish welfare state began to develop.
+For the period 1960 onwards, there are separate analyses on decades except for the two last periods.
+Sweden had developed into a wealthy
 welfare society with low income inequality,
@@
 1990s, which among other things increased unemployment and changed parts of
-the welfare system towards being less generous than before. The start of the
+the welfare system towards being less generous. The start of the
 last period (2008--2013) coincides with the international financial
@@
 It contains no less than `r n.rec + pn.rec` records of `r n.pl + pn.pl` individuals, of
-which `r n.d + pn.d` are observed to die.
+which `r n.d + pn.d` are observed to death.
 This makes it very inconvenient to use the nowadays so popular semi-parametric
 technique known as
-\emph{Cox regression} [@cox72,@gb02]. Instead we use a parametric model
+\emph{Cox regression} [@cox72, @gb02]. Instead we use a parametric model
 that allows for reduction of the data
@@
 it comes to the agricultural sector. While farmers were common occupations
-during the first half of the 20th century, these occupations has constantly
+up until the first half of the 20th century, these occupations has constantly
 decreased in numbers thereafter. This means that the middle class has
@@
 highest social class usually have comparatively low mortality, while
-working class women have high. There is however no substantial change in
-the relative risks between classes. Only the group with unknown social
+working class women have high, but about the same level as the middle category. There is however no substantial change in the relative risks between classes. Only the group with unknown social
 class is distinguished by much higher mortality from the 1980s
@@
 on their own occupation. Being without occupation probably include women that
-are more frail when it comes to health and vulnerable concerning social characteristics.
+are more frail when it comes to health or are vulnerable due to social characteristics.
 
@@
 the fastest decline did not set in definitely until sometimes after the
-Second World War (results not shown here), i.e. later than for the age
+Second World War, i.e. later than for the age
 group 40--64 years. This corresponds to the Swedish development where we
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group (see Figure \@ref(fig:cumswold). For
+we find for the younger age group (see Figure \@ref(fig:cumswold)). For
 women the results are mixed. Once again, too much missing data on social
@@
 extent is consistent with what we observed concerning all-cause
-mortality. The elite group had much higher cardiovascular diseases up until
+mortality. The elite group had much higher cardiovascular mortality up until
 the middle of the 20th century, but instead have the lowest mortality from
@@
 has to do with quite few events to analyse. Women of the highest class have
-favourable mortality during several periods, consistent with the advantages
+favourable mortality levels during several periods, consistent with the advantages
 we found for women in the younger age group, but on the other hand had high
@@
 pattern of mortality in historical northern Sweden for some time, focusing
-on different regions, different age groups and different time periods
-[@edv92, @edv04, @edli11, @edbr12]. When it comes to infant and child
-mortality during pre- and early-transitional time, the expected socio-economic
+on different regions, age groups and time periods [@edv92, @edv04, @edli11, @edbr12]. When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic
 differences was often but not always confirmed [@edv92, @edv04],
-something that has been found in other studies as well [@bredro02,
+something that has been found in other Swedish studies as well [@bredro02,
 @bube01, @mol17].
@@
 Occupational titles in historical sources are not always distinct enough to define the correct class.
-Studies covering long periods face the problem that occupations and classes change character---they do not represent the same thing.
+Studies covering long periods face the problem that occupations and classes change character --- they do not represent the same thing.
+We have chosen to work with a few broad groups.
 Nevertheless, these ambiguities and possible misclassifications should not radically change the basic patterns observed.
-The social classes identified during the different periods represent distinct groups with very large differences in resources, capacities and power.
+The social classes identified during the different periods represent distinct groups when it comes to resources, capacities and power.
 
@@
 Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
-We must therefore consider other explanations.
-
-This makes us consider the possible impact of behavioural factors.
+It is possible that they had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
+We must however consider other and potentially more important explanations as well.
+
+This makes us consider the possible impact of behavioural factors, something that could represent mechanisms related to stressful lives.
 Our results of the gender aspect is important in this respect.
@@
 Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
-Overweight and obesity was probably higher and physical exercise lower in higher social classes.
+
+Overweight, obesity and physical inactivity were probably more common in higher social classes.
 @vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained the higher tobacco consumption among men in these groups.
@@
 There was also an association between higher incomes and consumption [@loco94].
-The same lack of statistical evidence characterizes alcohol consumption in different social classes before 1960s with some exceptions.
+
+The same lack of statistical evidence characterizes knowledge about alcohol consumption in different social classes before 1960s, but with some exceptions.
 Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with highest incomes.
 There was also an over-representation of alcohol-related deaths in higher social classes in the 1960 [@noro98].
-@noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with a more skewed consumption distribution, indicating more high-consumers, in this group.
+@noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with a more skewed consumption distribution, indicating more high-consumers in this category.
 A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
 Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being inconsistent in 1961 and reversed in 1991 with highest mortality in the lowest class [@cropr10].
+If these indications of changes in health-related behaviour are significant enough to explain our findings here can be discussed.
+We believe that they certainly are at least an important part of the story.
 
@@
 In this article we set our task to analyse social differences in mortality from a long-term perspective.
-Until recently it has not been possible to investigate the development on mortality differences among the adult and elderly population.
+Until recently it has not been possible to investigate the development on mortality differences among adult and elderly populations.
 Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to the large class differences.
@@
 Unfortunately, the evidence for behavioural differences in history are usually circumstantial and anecdotal, at least when it comes to social class.
-Studies of these aspects and their role for the health and mortality in different times and places require further development.
+Studies of these aspects and their role for health and mortality in different times and places require further development.
 
@@
 In comparison with many other countries, the level of mortality has been fairly low.
-This is a relevant question, which hopefully can be addressed in other and maybe comparative studies.
+This is a relevant question, which hopefully can be addressed in other studies, maybe with a comparative approach.
 

### 2017-09-01 CreateData/creDat2.Rnw

@@
 cohorts"). Since we are interested only in ages 40--89 in the time period
-1901--2013, we can remove all born 1810 and earlier and all born 1974 or
+1801--2013, we can remove all born 1710 and earlier and all born 1974 or
 later. Finally we apply an "age window" of 40-89.
@@
 ## End cut ####################################################
-levels(vb$period)[8] <- "1991-2001" ## Changed from "1991-2000", fob90 extended.
-levels(vb0$period)[8] <- "1991-2001"
+##levels(vb$period)[8] <- "1991-2001" ## Changed from "1991-2000", fob90 extended.
+##levels(vb0$period)[8] <- "1991-2001"
 save(vb0, file = "../data2/vb0.rda")

### 2017-09-04 CreateData/R/ageGroup.R
remove 'CreateData'

@@ -1,18 +1,0 @@
-ageGroup <- function(dat, ages){
-    n <- length(ages) - 1
-    ##age <- ages[-(n+1)]
-    if (n > 1){
-        ##library(eha)
-        source("ageWin.R")
-        dat$age <- numeric(NROW(dat))
-        first <- ageWin(dat, ages[1:2])
-        first$age <- ages[1]
-        for (i in 2:n){
-            tmp <- ageWin(dat, ages[i:(i+1)])
-            tmp$age <- ages[i]
-            first <- rbind(first, tmp)
-        }
-    }
-    first <- first[order(first$id, first$enter), ]
-    invisible(first)
-}

### 2017-09-04 CreateData/R/pushHis.R

@@ -1,15 +1,0 @@
-pushHis <- function(dat){
-    ## 'dat' is a data frame with variables 'LINNEID' and 'hisclass'
-    library(skum)
-    om <- unique(dat$LINNEID)
-    
-    hyhis <- numeric(NROW(dat))
-    dat$nyhisclass <- numeric(NROW(dat))
-    for (i in om){
-        who <- dat$LINNEID == i
-        ok <- dat$hisclass[who]
-        ok <- pushForward(ok)
-        dat$nyhisclass[who] <- ok
-    }
-    dat
-}

### 2017-09-04 CreateData/R/pushHis2.R

@@ -1,17 +1,0 @@
-pushHis2 <- function(dat){
-    ## 'dat' is a data frame with variables 'id' and 'hisclass'
-    ## A copy of 'pushHis' with LINNEID <--> id.
-    
-    library(skum)
-    om <- unique(dat$id)
-    
-    hyhis <- numeric(NROW(dat))
-    dat$nyhisclass <- numeric(NROW(dat))
-    for (i in om){
-        who <- dat$id == i
-        ok <- dat$hisclass[who]
-        ok <- pushForward(ok)
-        dat$nyhisclass[who] <- ok
-    }
-    dat
-}

### 2017-09-04 CreateData/R/putONssyk.R

@@ -1,73 +1,0 @@
-putONssyk <- function(dat){
-    ## 'dat' is the processed file 'fobdata'
-    ## This must be run from "Mortality/CreDat"
-    
-    load("../../Data/yr_ssyk.rda")
-    dat$ssyk <- NA
-
-    ##for (year in seq(60, 75, by = 5)){
-      ##  nam <- paste("yr", year, sep = "")
-    ##    load(paste("data/", nam, ".rda" sep = ""))
-##        nyr <- dat[dat$AR == i, ]
-  ##      indx <- match(nyr$YRKE)
-    ##}
-    ## 1960:
-    load("../../Data/yr60.rda")
-    who <- dat$AR == 1960 & dat$YRKE %in% yr60$old
-    nyr <- dat$YRKE[who]
-    indx <- match(nyr, yr60$old)
-    nyr <- yr60$new[indx]
-    dat$YRKE[who] <- nyr
-    
-    ## 1965:
-    load("../../Data/yr65.rda")
-    who <- dat$AR == 1965 & dat$YRKE %in% yr65$old
-    nyr <- dat$YRKE[who]
-    indx <- match(nyr, yr65$old)
-    nyr <- yr65$new[indx]
-    dat$YRKE[who] <- nyr
-    
-    ## 1970:
-    load("../../Data/yr70.rda")
-    who <- dat$AR == 1970 & dat$YRKE %in% yr70$old
-    nyr <- dat$YRKE[who]
-    indx <- match(nyr, yr70$old)
-    nyr <- yr70$new[indx]
-    dat$YRKE[who] <- nyr
-    
-    ## 1975:
-    load("../../Data/yr75.rda")
-    who <- dat$AR == 1975 & dat$YRKE %in% yr75$old
-    nyr <- dat$YRKE[who]
-    indx <- match(nyr, yr75$old)
-    nyr <- yr75$new[indx]
-    dat$YRKE[who] <- nyr
-    
-    ## 1980: Vad ska vi ta här?? 'yr80.rda' finns inte!
-    ## och YRKE80 gäller ej 1980!!
-    ## Chansar på att 'yr75.rda' gäller:
-
-    ## 1980 (try):
-    load("../../Data/yr75.rda")
-    who <- dat$AR == 1980 & dat$YRKE %in% yr75$old
-    nyr <- dat$YRKE[who]
[…]

### 2017-09-04 CreateData/R/ssyk3TOhisclass.R

@@ -1,33 +1,0 @@
-ssyk3TOhisclass <- function(x){
-
-    ## 2017-03.28: Three instead of four hisclasses.
-    
-    ## 'x' is a vector of ssyk codes.
-    ## as output we want a vector of hisclass codes (7 levels)
-    
-    ## Step 1: Convert shh$ssyk to "three-digit" (except '110'):
-    load("../../Data/shh.rda")
-    shh$ssyk <- shh$ssyk %/% 10 # removes last digit.
-    ## Remove duplicate ssyk-rows in shh:
-    shh <- shh[!duplicated(shh$ssyk), ]
-    
-    ## Step 2: SSYK3 to 12-levels hisclass
-    indx <- match(x, shh$ssyk)
-    tmp <- shh$hisclass[indx]
-
-    ## Step 2: go from 12 to 4 levels: ## NEW!!
-    ## Step 2: go from 12 to 3 levels: ## NEWER!!
-    
-    res <- rep(-1, length(x))
-    
-    tmp[is.na(tmp)] <- -1
-    
-    res[tmp %in% 1:2] <- 1
-    res[tmp %in% c(3:6, 8)] <- 2
-    ##res[tmp %in% 4:6] <- 3
-    res[tmp %in% c(7, 9:12)] <- 3 # '4' ---> '3'
-    
-    is.na(res) <- res == -1
-    
-    res
-}

### 2017-09-04 article2.Rmd
first commit of new model: tabular data

@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art2-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 library(knitr)
@@
 source("R/get.file.R")
-vb0 <- get.file("vb0")
-
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
+skel <- get.file("skel")
 ##
-
-## We don't need "vb", do we?
-## vb <- get.file("vb")
-
-lisaold <- get.file("lisaold")
-
-loTab <- get.file("loTab")
-
-loTab$period <- factor(loTab$period)
+skume <- get.file("skume")
 ##
-vb0Tab <- aggregate(vb0[, c("event", "exposure")],
-                   vb0[, c("sex", "age", "period", "cohort",
-                          "urban", "civst", "hisclass")],
-                   FUN = sum)
-vb0Tab$age <- factor(vb0Tab$age)
+foli <- get.file("foli")
 ```
@@
 ```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
-plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
-     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
+plot(c(1801, 1801, 1951, 1951, 1801), c(40, 90, 90, 40, 40),
+     type = "l", xlim = c(1800, 2014), lwd = 2, col = "red",
      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(1, at = c(1801, 1851, 1901, 1951, 1961, 1981, 2014), las = 2)
 axis(2, las = 2, at = c(0, 40, 65, 90))
@@
 #
-pn.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-pn.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-pn.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
-pn.d <- as.integer(pn.d)
-text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(pn.rec, "records"))
-text(1900, 55, paste(pn.pl, "persons"))
-text(1899.5, 50, paste(pn.d, "deaths"))
+pn.exp <- round(sum(skel$exposure) + sum(skume$exposure))
+pn.d <-  as.integer(sum(skel$event) + sum(skume$event))
 #
-n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
-n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
-n.d <- as.integer(n.d)
+text(1875, 70, "PopLink", cex = 2, col = "red")
+##text(1899, 60, paste(pn.rec, "records"))
+text(1875, 55, paste(pn.exp, "p-years"))
+text(1874.5, 50, paste(pn.d, "deaths"))
+#
+n.exp <- round(sum(foli$exposure))
+n.d <-  as.integer(sum(foli$event))
 
 text(1985, 70, "Linnaeus", cex = 2, col = "blue")
-text(1984, 60, paste(n.rec, "records"))
-text(1985.2, 55, paste(n.pl, "persons"))
+##text(1984, 60, paste(n.rec, "records"))
+text(1985.2, 55, paste(n.exp, "p-years"))
 text(1985.2, 50, paste(n.d, "deaths"))
@@
 residence, birth date, death date, first and last date of observation and
-type of entrance/exit. Total number of persons are `r pn.pl` leading to
-`r pn.d` deaths in Poplink and `r n.pl` persons leading 
+type of entrance/exit. Total number of person years is `r pn.exp` leading to
+`r pn.d` deaths in Poplink and `r n.exp` person years leading 
 to `r n.d` deaths in the Linnaeus database. See 
@@
 ```{r hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
-xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
+xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xxx <- rbind(xx, xy, xz)
+yy <- round(prop.table(xxx, 1) * 100, 1)
 barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@@
 ```{r hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
-xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
+xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xxx <- rbind(xx, xy, xz)
+yy <- round(prop.table(xxx, 1) * 100, 1)
 barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@@
 Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
+
 Overweight, obesity and physical inactivity were probably more common in higher social classes.

### 2017-09-05 R/getPHglm2.R
new versions forced by extending the time period

@@ -13,7 +13,8 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
     levs <- matrix(0, nrow = m, ncol = k + 1)
    
     for (i in 1:nper){
-        fit <- glm(event ~ offset(log(exposure)) + urban + civst +
+        fit <- glm(event ~ offset(log(exposure)) + ##urban + 
+                       civst +
                        age * hisclass, 
                    data = dat[dat$period == labb[i], ], family = poisson)
         

### 2017-09-05 R/plotED.glm.R

@@ -51,7 +51,7 @@ plotED.glm <- function(res,
     plot(xv[out], outp[1, -exclude], type = "b", lty = 1, col = col[1], pch = 1,
          ylim = c(fr, to), main = main,
          axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
-         xlim = c(min(xv), max(xv) + 25))
+         xlim = c(min(xv), max(xv) + 33))
     axis(1, at = xv, lab = labb, las = 2)
     axis(2, las = 2)
     box()

### 2017-09-05 article2.Rmd

@@
 ---
-title: Is high social class always beneficial for survival? A study of adults and elderly in Northern Sweden 1851--2013
+title: Is high social class always beneficial for survival? A study of adults and
+  elderly in Northern Sweden 1801--2013
 author: "Sören Edvinsson and Göran Broström"
-date: '`r Sys.Date()`'
+date: "`r Sys.Date()`"
 output:
@@
 foli <- get.file("foli")
+skel$urban <- FALSE
+vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
 ```
@@
 Our main focus is on the relative differences.
-The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden for the period 1851--2013 for Skellefteå and 1901--2013 for Umeå.
+The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
 The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
@@
 the fertility transition was late [@coa86]. The Skellefteå population
-size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on
-January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the
-century.
+size as defined in our data sets (all ages) was 6142 on January 1, 1801, 17355 on January 1, 1851, 43212 on
+January 1, 1901, 62136 on December 31, 1950 and 76723 at the end of the
+20th century.
 
@@
 were also some foundries as well as industries for example in forestry and
-small-scale production. The population size as defined in our data sets was
-18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when
-the century ended.
+small-scale production. The population size as defined in our data sets (all ages) was
+19138 on January 1, 1901, 33393 on December 31, 1950 and 103970 when
+the 20th century ended.
 
@@
 residence, birth date, death date, first and last date of observation and
-type of entrance/exit. Total number of person years is `r pn.exp` leading to
-`r pn.d` deaths in Poplink and `r n.exp` person years leading 
+type of entrance/exit. Total number of person years is `r as.integer(pn.exp)` leading to
+`r pn.d` deaths in Poplink and `r as.integer(n.exp)` person years leading 
 to `r n.d` deaths in the Linnaeus database. See 
@@
 ```{r civstskum, fig.cap = "Exposure by civil status (from the top: never married, married and previously married) and time period", fig.height = 4.5}
-xx <- with(vb0Tab, tapply(exposure, list(period = period, civst = civst), sum))
-yy <- round(prop.table(xx, 1) * 100, 1)
+xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))
+xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
+xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
+xxx <- rbind(xx, xy, xz)
+yy <- round(prop.table(xxx, 1) * 100, 1)
 barplot(t(yy)[3:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
@@
 ```{r causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5}
-vb01 <- vb0[vb0$event == 1, ]
-cancer <- tapply(vb01$cancer, vb01$period, sum)
-heart <- tapply(vb01$heart, vb01$period, sum)
-death <- tapply(vb01$period, vb01$period, length)
+##vb01 <- vb0[vb0$event == 1, ]
+cancer <- c(tapply(skume$cancer, skume$period, sum), tapply(foli$cancer, foli$period, sum))
+heart <- c(tapply(skume$heart, skume$period, sum), tapply(foli$heart, foli$period, sum))
+death <- c(tapply(skume$event, skume$period, sum), tapply(foli$event, foli$period, sum))
 other <- death - cancer - heart
@@
 ttt <- prop.table(ttt, 2) * 100
-barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = "Per cent")
+barplot(ttt[3:1, ], col = 3:1, las = 2, ylab = "Per cent")
 ```
@@
 industrialisation started and the urban environments increased in
-population size.
-The modern Swedish welfare state began to develop.
-For the period 1960 onwards, there are separate analyses on decades except for the two last periods.
-Sweden had developed into a wealthy
+population size. The modern Swedish welfare state began to develop. For the 
+period 1960 onwards, there are separate analyses on decades except for the two 
+last periods. Sweden had developed into a wealthy
 welfare society with low income inequality,
@@
 Another practical problem is the size of that data set:
-It contains no less than `r n.rec + pn.rec` records of `r n.pl + pn.pl` individuals, of
-which `r n.d + pn.d` are observed to death.
+It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
+ `r n.d + pn.d` deaths.
 This makes it very inconvenient to use the nowadays so popular semi-parametric
@@
 Since all components of $\mathbf{x}$ are categorical covariates, the
-estimation of \eqref{eq:prop} can be performed in two ways, first by the
+estimation of \@ref(eq:prop) can be performed in two ways, first by the
 original data set and a survival analysis routine handling a piecewise
@@
 the point is that the two ways give exactly the same results. This is
-because the table is a \emph{sufficient statistic} for \eqref{eq:prop} and
+because the table is a \emph{sufficient statistic} for \@ref(eq:prop) and
 the original data set.
@@
 and graphics [@cran], especially using the package **eha**
-[@gb12,@eha]. The proportional hazards model allows us to scrutinise the differences and
+[@gb12;@eha]. The proportional hazards model allows us to scrutinise the differences and
 the development in more detail. We divide the analysis into two age groups,
@@
 were still mainly in the workforce, and the second those that mainly were
-retired.
-The model controls for marital status and whether the individual resides in the urban or rural parts of the regions.
-There are no serious signs of deviation from proportionality assumption, except
-regarding social class. Each period are analysed separately, but we then combine results
+retired. The model controls for marital status and whether the individual resides 
+in the urban or rural parts of the regions. Each period are analysed separately, 
+but we then combine results
 in order to show the development of the social patterning of mortality
@@
 interaction term between **age** and social class (**hisclass**). Then, the main
-results are presented graphically. Tables with results from the
-different models are presented at the end of the article, but are not
-commented on in detail. The comments on the results are primarily based on
+results are presented graphically.  The comments on the results are based on
 the figures and focuses on the central variable in our study--- social class (**hisclass**).
-As can be seen from the tables, most variables in the models are
-significant at 5 per cent level.
-
-# The effect of social class, 1851--1950, 1961--2013
+
+# The effect of social class, 1801--1950, 1961--2013
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Women 40-64", exclude = 5:6)
+                main = "Women 40-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
@@
 
-From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
+From Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
 The most interesting feature is the three-way interaction with **sex**, **period**, and social class (**hisclass**).
 
-```{r cumswtab,results='asis',cap="Table"}
+```{r cumswtab,cap="Table",results="asis"}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
@@
 ##knitr::kable(dr)
-xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, 40-65.")
+print(xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, ages 40--65."), 
+      type = "latex", comment = FALSE)
 ```
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Women 40-64", exclude = 5:6)
+                    ylab = "TMR(65|40)", main = "Women 40-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 Notice that we have not analysed the 1950s due to the fact that we only
-have data for some parishes at that time. Furthermore, we have excluded the
-results for women during the 1960s and 1970 out of the reasons discussed above. The
+have data for some parishes at that time. Furthermore, the
+results for women during the 1960s and 1970 are highly misleading for the reasons discussed above. The
 changing economy also have consequences for the analysis, particularly when
@@
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure. In Table~\ref{tab:cumswoldtab} the results of the model is presented.
+duration 25 (age 90), starting from age 65 as a summary measure.
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Women 65-89", exclude = 5:6)
+                    ylab = "TMR(90|65)", main = "Women 65-89")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 
-See Table~\ref{tab:cumswoldtab}.
-
-```{r cumswoldtab,results='asis'}
+```{r cumswoldtab,results='asis',eval=FALSE}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
@@
 ```{r causes41}
-cvb <- vb0
-cvb$event <- cvb$heart ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
-cvb$period <- factor(cvb$period)
-
-# load("data/vb.rda")
-# ##vb <- cal.window(vb, c(1961, 2001))
-# ##x <- substr(vb$ULORSAK, 1, 2)
-# vb$event <- vb$event & vb$heart
-# vb$exposure <- with(vb, exit - enter)
-# vb$cohort <- factor(vb$cohort)
-# vb$period <- factor(vb$period)
-cvbTab <- aggregate(cvb[, c("event", "exposure")],
-                    cvb[, c("sex", "age", "period", "cohort",
-                           "urban", "civst", "hisclass")],
-                    FUN = sum)
- cvbTab$hisclass <- factor(cvbTab$hisclass)
- cvbTab$age <- factor(cvbTab$age)
+cvbTab <- rbind(skume[, names(foli)], foli)
+cvbTab$event <- cvbTab$heart
 ```
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, exclude = 3:4,
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 3:4,
                     ylab = "TMR(65|40)",  main = "Women 40-64")
@@
 
-```{r 41cumswtab,results='asis'}
+```{r 41cumswtab,results='asis',eval = FALSE}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass +
@@
 ```
-
-See Table \@ref(tab:41cumswtab).
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women 65-89", exclude = 3:4)
+                    ylab = "H(90|65)",  main = "Women 65-89")##, exclude = 3:4)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 
-```{r 41cumswoldtab,results='asis'}
+```{r 41cumswoldtab,results='asis',eval=FALSE}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * period *hisclass,
@@
 ```
-
-See Table \@ref(tab:41cumswoldtab).
 
@@
 ```{r causes12}
-cvb <- vb0
-cvb$event <- cvb$cancer ## Can this be done better?
-cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
-cvb$period <- factor(cvb$period)
-
-# load("data/vb.rda")
-# ##vb <- cal.window(vb, c(1961, 2001))
-# ##x <- substr(vb$ULORSAK, 1, 2)
-# vb$event <- vb$event & vb$heart
-# vb$exposure <- with(vb, exit - enter)
-# vb$cohort <- factor(vb$cohort)
-# vb$period <- factor(vb$period)
-cvbTab <- aggregate(cvb[, c("event", "exposure")],
-                    cvb[, c("sex", "age", "period", "cohort",
-                           "urban", "civst", "hisclass")],
-                    FUN = sum)
- cvbTab$hisclass <- factor(cvbTab$hisclass)
- cvbTab$age <- factor(cvbTab$age)
+cvbTab <- rbind(skume[, names(foli)], foli)
+cvbTab$event <- cvbTab$cancer
 ```
@@
 
-```{r 12cumswtab,results='asis'}
+```{r 12cumswtab,results='asis',eval=FALSE}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex *hisclass +
@@
 ```
-
-For the results of the model, see Table \@ref(tab:12cumswtab).
 
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)",  main = "Women 65-89", exclude = 3:4)
+                    ylab = "TMR(90|65)",  main = "Women 65-89")##, exclude = 3:4)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
@@
 
-```{r 12cumswoldtab,results='asis'}
+```{r 12cumswoldtab,results='asis',eval=FALSE}
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex +

### 2017-09-07 article2.Rmd
Minor changes in introduction

@@
 A major theme in demographic and epidemiological studies, is the seemingly persistent effect of social class on mortality.
-In the present study, we approach this issue by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
-Our main focus is on the relative differences.
-The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
+In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
 The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
 Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
-The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
-
-On the basis of the results we present, we argue that high social class is not necessarily favourable for survival.
+The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
+
+On the basis of the results we present, we argue that high social class is not necessarily always favourable for survival.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
+
+## Social class and mortality
 

### 2017-09-08 article2.Rmd
Fortsatta revideringar och referens

@@
 health-related behaviour became the decisive determinant for health and survival.
-@omran82 comes to a similar conclusion in his theory of the epidemiologic transition. He states in the third proposition in the theory, that even if the class differentials in mortality were
+@omran82 comes to a similar conclusion in his theory of the epidemiologic transition. He states in the third proposition of the theory that even if the class differentials in mortality were
 maintained during the transition, the decline set in earlier and was faster
@@
 with micro-data, have however questioned how general the assumed
-pattern is [@bevp11]. Instead, some studies suggest that the modern
-pattern might be a fairly recent development [@bedr11], although the pattern differs
-depending on age, gender or historical context. When it comes to the adult
-and elderly population results from different studies have diverged, some
-reporting substantial differences but many, particularly those based on
-Swedish data, have found surprisingly small or even reversed socio-economic
-differences during the 19th and early 20th century [@bedr11,
-@edli11, @edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
-some point during the mortality transition, at least in Sweden. Solid
-empirical evidence about the process is however lacking and studies
+pattern is [@bevp11]. Solid empirical evidence about the process is however lacking and studies
 focusing on the issue are still few. There is a need for more reliable
@@
 county of Västerbotten in the north of Sweden along the coast of Gulf
-of Bothnia. This was a remote part of the country, where communication with
+of Bothnia. This was a remote part of the country where communication with
 the rest of Sweden was difficult until the late 19th century. The economy
 was dominated by agriculture, making it vulnerable to harvest failures, and
-several severe famines frequently affected the regions during the 1800s, for example after the harvest failure in 1867 [@segb14].
+several severe famines frequently occurred in the regions during the 1800s, for example after the harvest failure in 1867 [@segb14].
 During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
@@
 
-The Skellefteå region until the 1950 in our dataset consists of a selection
+In our dataset, the Skellefteå region before 1950 consists of a selection
 of parishes surrounding the town of Skellefteå, founded in 1845 but with a
@@
 hamlets, getting their livelihood from agricultural production. During the
-20th century, industrialisation took place, partly by the opening of mines
-in the inlands. This also led to a population increase both in the town
-and in the rural parts, and a much more diversified economy.
+20th century, industrialisation took place. This also led to a population increase both in the town
+and in the rural parts, resulting in a much more diversified economy.
 
@@
 the fertility transition was late [@coa86]. The Skellefteå population
-size as defined in our data sets (all ages) was 6142 on January 1, 1801, 17355 on January 1, 1851, 43212 on
-January 1, 1901, 62136 on December 31, 1950 and 76723 at the end of the
+size as defined in our data sets (all ages) was 6142 on January 1, 1801, 17355 on January 1, 1851, 43212 on January 1, 1901, 62136 on December 31, 1950 and 76723 at the end of the
 20th century.
 
-The Umeå region consists of Umeå urban and rural parish until
-1950, and from 1960 onwards of Umeå municipality when another three
+The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1960 onwards of Umeå municipality when another three
 parishes are included. This region has a somewhat different character from
@@
 population databases at the DDB, whom provide us with micro-data on a sufficiently large population and
-with substantial social diversity for the Skellefteå and Umeå regions
-[@wee16]. The early period until 1950 is covered by
-the database Poplink, the digitisation of historical parish registers for
+with substantial social diversity for the Skellefteå and Umeå regions. The early period until 1950 is covered by the database Poplink, the digitisation of historical parish registers for
 the Skellefteå and Umeå region [@wee16]. Poplink is based on linked
@@
 long as they remained in the region. The records are linked within but not
-between the regions. Data from Poplink are accessed for the period 1851--1950.
+between the regions. Data from Poplink are accessed for the period 1801--1950.
 
@@
 the information from the yearly population registers from Statistics Sweden
-(LISA) together with death information from National Board of Health and
+(LISA) together with information on mortality from National Board of Health and
 Welfare.
 
-Information on individuals are anonymised and the two databases are not
-linked. This stops us from following individuals present
-in both databases throughout their lives, thus they are treated separately. This
-also makes it impossible to add information on individuals in the Linnaeus
-database from what we potentially could extract from Poplink, for example family
+Individuals are anonymised and the two databases are not
+linked, thus they are treated separately.
+This prevent us from following individuals between the two databases throughout their lives.
+It also makes it impossible to add information on individuals in the Linnaeus database from what we potentially could extract from Poplink, for example family
 background or previous social class.
@@
 comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of the presence of all individuals in our data, allowing us
-to have full and continuous control of the de jure population. This is
+to have full and continuous control of the \emph{de jure} population. This is
 however not the case with the Linnaeus database. For the period 1960--2000
@@
 analyses. There is however no full information on what happened in-between
-the censuses, for example on the exact date of exit for persons having moved
+the censuses, for example the exact date of exit for persons having moved
 between two censuses. We have solved this by considering everyone as
 resident in our regions during the complete five-year period between
-censuses if the person is missing in the later census (the last period covering ten years). This overestimates their presence, but should not bias the results when it
+censuses if the person is missing in the later census (the last period covering ten years). This marginally overestimates presence, but should not bias the results when it
 comes to socioeconomic differences. We furthermore miss those that both
 moved in and moved out or vice versa between censuses. Exit dates are exact
-if they refer to death, and in these cases we consider them as constantly
+if they refer to death. In these cases we consider them as constantly
 present from the last census to the death data even if the death took place
@@
 
-The choice of variable for analysing social differences vary between studies. A common alternative to a class analysis is to use income or education for
-investigating socio-economic health differences. @goldk10 criticize
-how many economists and epidemiologists operationalize inequality in their
-studies, for example by not distinguishing between class and status
-aspects. Class refers to relational aspects in society, while many other
-variables focus on the distribution of material and/or immaterial
-resources. He advocates studies of different aspects, which together
-complement each other. Comparing different approaches are valuable in order
-to understand why some are advantaged and others disadvantaged and what
-sort of factors determines health differences.
-
-In this paper we analyse mortality differences according to social class
-based on the classification scheme Hisclass in modified form [@lema11]. The only
+In this paper we analyse mortality differences according to social class, treated as a time-dependent variable, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
 type of socio-economic information available both in the Linnaeus and Poplink database
@@
 The availability to information on occupations varies over time and between
-sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. The censuses 1960--1990 include occupational
-codes based on the system used in Sweden at that time, NYK (Nordisk
+sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
+Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
 Yrkesklassificering).
@@
 For the period 1991--2000 the social class in the census 1990 is considered to be kept
-until 2001. This strategy is also used for cases of missing social
-class. Furthermore, we consider the last social position before retirement
-to be kept until death or last observation for those above 65. For the
-Linnaeus data, we use a script recoding HISCO into Hisclass, provided
-to us by Jonas Helgertz, Lund University.
+until 2001. Furthermore, we consider the last social position before retirement
+to be kept until death or last observation for those above 65.
 
@@
 the husband/father, assuming that the family shares the same socio-economic
-position. This strategy makes the problem of under-recording less
-problematic.
+position.
 
@@
 same way as in Poplink. Extra-marital cohabitation, connected to the Second
-Demographic Transition [@van87, @van03], became common and female
+Demographic Transition [@van87], became common and female
 labour force participation developed as the norm in Sweden during this
@@
 interpretation of the development for women uncertain for some periods.
-Notice however that there is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data.
+Notice however that there probably is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].
 
@@
 the age group 65 and above, i.e. those that were retired, but still low
-enough to be used in our analyses. Until 1950, there were however very few
+enough to be used in our analyses (results not shown). Until 1950, there were however very few
 with missing occupation. For men in working-active age, the proportion
@@
 proportion without social position is very low until the middle of the 20th
-century, reflecting the way womens social position is defined.
+century, reflecting the way womens' social position is defined.
 Missing social class is however a real problem during particularly the 1960s and 1970s.
@@
 status is given in the sources and partner is missing, the status has been
-set to unmarried.
+set to unmarried. Notice that it has not been possible to identify cohabitation.
 
@@
 
-```{r civstskum, fig.cap = "Exposure by civil status (from the top: never married, married and previously married) and time period", fig.height = 4.5}
+```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.height = 4.5}
 xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))
@@
 for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
-on parish of residence.
+on parish or place of residence.
 
@@
 We have made separate analyses for the different periods. The century
-1851--1950 is divided into four periods of 25 years each. The 19th century
+1801--1950 is divided into six periods of 25 years each. The 19th century
 periods represent a mainly pre-industrial society dominated by agriculture
@@
 decreased in numbers thereafter. This means that the middle class has
-somewhat different character during the last periods.
+a different character during the last periods.
 
@@
 accordance with previous analyses from the present research group on other
-regions in northern Sweden and for other periods [@edli11,@edbr12]. The
+regions in northern Sweden and for other periods [@edli11, @edbr12]. The
 highest social class usually have comparatively low mortality, while
-working class women have high, but about the same level as the middle category. There is however no substantial change in the relative risks between classes. Only the group with unknown social
-class is distinguished by much higher mortality from the 1980s
+working class women have high, but almost constantly about the same level as the middle category.
+There is however no substantial change in the relative risks between classes.
+Only the group with unknown social class is distinguished by much higher mortality from the 1980s
 onwards. This can however have to do with how we have defined social
-position for women before and after 1950. After 1950 it is based
+position for women before and after 1950. After 1960 it is based
 on their own occupation. Being without occupation probably include women that
@@
 our analyses. The groups we would expect to have the lowest mortality, if
-we assume that access to important resources and belonging to a high social class
+we assume that access to vital resources and belonging to a high social class
 determines survival, instead have by far the highest hazards of dying
@@
 the others from the 1960s onwards. The assumption made for women of
-unknowns social class is probably even more relevant for men. They probably
-represent groups in society that are particularly vulnerable, having difficulties getting into the labour force.
+unknown social class is probably even more relevant for men. They probably
+represent groups in society that are particularly vulnerable.
 
@@
 1851--1950, men having about three times higher risks than women in the
-same class (and in the same households). In the other classes, the difference was much smaller. It was
-men in high position that experienced the largest risks.
+same class (basically in the same households). In the other classes, the difference was much smaller. It was men in high position that experienced the largest risks.
 
@@
 
+Instead, some studies suggest that the modern
+pattern might be a fairly recent development [@bedr11], although the pattern differs
+depending on age, gender or historical context. When it comes to the adult
+and elderly population results from different studies have diverged, some
+reporting substantial differences but many, particularly those based on
+Swedish data, have found surprisingly small or even reversed socio-economic
+differences during the 19th and early 20th century [@bedr11,
+@edli11, @edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
+some point during the mortality transition, at least in Sweden.
+
 The surprising results make it reasonable to consider the validity and possible weaknesses in the study.

### 2017-09-08 article2.Rmd

@@
 ```{r getske4060}
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
+cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
 scap1w <-  "Cumulative hazards, ages 40--64, women."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
+cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
 scap1m <-  "Cumulative hazards, ages 40--64, men."
 ##
-cap0 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
-scap0 <- "Total hazard by HISCLASS, ages 40--64."
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard standardised by HISCLASS, ages 40--64."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women."
+cap0 <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom)."
+scap0 <- "Total hazard by social class, ages 40--64."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom)."
+scap12 <- "Total hazard standardised by social class, ages 40--64."
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades."
+scap12log <- "Total hazard by social class, ages 40--64, last years."
+
+cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women."
 scap2w <-  "Cumulative hazards, ages 65--89, women."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men."
+cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men."
 scap2m <-  "Cumulative hazards, ages 65--89, men."
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). Standardised."
-scap22 <- "Total hazard by HISCLASS, ages 65--89."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). Standardised."
+scap22 <- "Total hazard by social class, ages 65--89."
 cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years."
+scap22log <- "Total hazard by social class, ages 65--89, last years."
 
@@
 group 40--64 years. This corresponds to the Swedish development where we
-find a turning point at this time with better survival among elderly, to a
-large extent caused by lower mortality in cardiovascular diseases.
+find a turning point at this time with better survival among elderly.
 
@@
 do we find any large disadvantage for working class elderly women at that
-time. In some periods, the social differences were small, and the elite
-even had substantially higher mortality during the early 20th century. The
+time. In some periods, the social differences were small, particularly between the worker and the middle group, and the elite even had substantially higher mortality during the early 20th century. The
 social differences continued to be small until the last decades of the
@@
 century is the survival much better for the highest social class while
-workers in the age of retirement had high mortality in comparison.
+workers in the age of retirement had high mortality in comparison. The difference between the worker and middle group have in most periods been small.
 
@@
 A general observation is that the results on causes of death and
-its development is difficult to interpret, partly because the number of
+its development can be difficult to interpret, partly because the number of
 cases are fewer but also that the quality and coverage of reported death
@@
 40 and above, thus indicating an increase of these diseases. However, at
-the same time unknown cause and death of old age decreased from 28 percent
+the same time unknown cause and death in old age decreased from 28 percent
 to 18 per cent. It is quite reasonable to suspect that many cardiovascular
@@
 ```{r 41getske4060}
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cardiovascular death."
+cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women, cardiovascular death."
 scap1w <-  "Cumulative hazards, ages 40--64, women, cardiovascular death."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cardiovascular death."
+cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men, cardiovascular death."
 scap1m <-  "Cumulative hazards, ages 40--64, men, cardiovascular death."
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap12 <- "Total hazard by HISCLASS, ages 40--64, cardiovascular death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cardiovascular death."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cardiovascular death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom), cardiovascular death."
+scap12 <- "Total hazard by social class, ages 40--64, cardiovascular death."
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+scap12log <- "Total hazard by social class, ages 40--64, last years, cardiovascular death."
+
+cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women, cardiovascular death."
 scap2w <-  "Cumulative hazards, ages 65--89, women, cardiovascular death."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cardiovascular death."
+cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men, cardiovascular death."
 scap2m <-  "Cumulative hazards, ages 65--89, men, cardiovascular death."
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cardiovascular death."
-scap22 <- "Total hazard standardised by HISCLASS, ages 65--89, cardiovascular death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cardiovascular death."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cardiovascular death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and decade, women (top) and men (bottom), cardiovascular death."
+scap22 <- "Total hazard standardised by social class, ages 65--89, cardiovascular death."
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+scap22log <- "Total hazard by social class, ages 65--89, last years, cardiovascular death."
 
@@
 ```
-
-Performing the same types of analyses as for all-cause mortality, we start
-by looking at cumulative hazards by Hisclass at age 65, given
-survival to age 40. The result of the model can be seen in Table \@ref(tab:41cumswtab).
 
@@
 Figure \@ref(fig:41cumsw) shows mortality levels during different periods
-from 1901 on-wards. Women from the highest social class had consistently
+from 1901 on-wards and the result of the model can be seen in Table \@ref(tab:41cumswtab). Women from the highest social class had consistently
 lower mortality from cardiovascular diseases compared to the others. We
@@
 In this section we focus on cardiovascular mortality among those that
-mostly have left the workforce, the age group 65--89 years. We use as a
-summary measure the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65. The results from the model is presented in Table \@ref(tab:41cumswoldtab).
+mostly have left the workforce, the age group 65--89 years.
+Much of the decline in mortality among elderly from the middle of the 20th century can be explained by lower death rates in this disease group.
+The results from the model is presented in Table \@ref(tab:41cumswoldtab).
 
@@
 ```{r 12getske4060}
-cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women, cancer death."
+cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women, cancer death."
 scap1w <-  "Cumulative hazards, ages 40--64, women, cancer death."
-cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men, cancer death."
+cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men, cancer death."
 scap1m <-  "Cumulative hazards, ages 40--64, men, cancer death."
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cancer death."
-scap12 <- "Total hazard by HISCLASS, ages 40--64, cancer death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
-scap12log <- "Total hazard by HISCLASS, ages 40--64, last years, cancer death."
-
-cap2w <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, women, cancer death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom), cancer death."
+scap12 <- "Total hazard by social class, ages 40--64, cancer death."
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades, cancer death."
+scap12log <- "Total hazard by social class, ages 40--64, last years, cancer death."
+
+cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women, cancer death."
 scap2w <-  "Cumulative hazards, ages 65--89, women, cancer death."
-cap2m <-  "Cumulative hazards for HISCLASS by time period, ages 65--89, men, cancer death."
+cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men, cancer death."
 scap2m <-  "Cumulative hazards, ages 65--89, men, cancer death."
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by HISCLASS and decade, women (top) and men (bottom), cancer death."
-scap22 <- "Total hazard standardised by HISCLASS, ages 65--89, cancer death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades, cancer death."
-scap22log <- "Total hazard by HISCLASS, ages 65--89, last years, cancer death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and decade, women (top) and men (bottom), cancer death."
+scap22 <- "Total hazard standardised by social class, ages 65--89, cancer death."
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). The last decades, cancer death."
+scap22log <- "Total hazard by social class, ages 65--89, last years, cancer death."
 
@@
 cardiovascular diseases for women, consistent with the idea of a social hierarchy in health.
-What is less expected, is that this we find a stronger association between social class and survival in cancer than cardiovascular diseases.
+What is less expected, is that we find a stronger association between social class and survival in cancer than in cardiovascular diseases.
 We do however not find the same pattern
@@
 particularly among adult men, not found despite the well-documented
-socio-economic differences in society [@edli11, @edbr12].
+socio-economic differences in society, for example in the Sundsvall region [@edli11, @edbr12].
 The results presented here show that the consistent advantage for higher social classes do not appear until quite late in the investigated regions.
@@
 Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67, @wowi95 @pamuk85].
-We know less about how general these results are, but apparently the Swedish context has been different from the English and Welsh.
+We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh.
 
@@
 In a study of southern Sweden, @oberg14 finds large social differences in body height, a measure that is assumed to sum up health during the life span.
-This does not contradict what we find here.
+This does not necessarily contradict what we find here.
 Height is mainly determined by conditions in early years, and are not reflecting inequalities in adulthood.
@@
 Psychosocial factors are also suggested by @wipi09 as an explanation why equal societies always perform better than unequal ones when it comes to health and other social conditions.
-In the long-term perspective provided here, neither of these theories fits the observation in earlier history.
+In the long-term perspective provided here, neither of these theories fit the historical observations.
 Their validity are mainly apparent for the most recent decades.
@@
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy gives advantages in many aspects of relevance for health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
+Having a good economy gives advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
 It is possible that they had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
-We must however consider other and potentially more important explanations as well.
-
-This makes us consider the possible impact of behavioural factors, something that could represent mechanisms related to stressful lives.
+
+We must however consider other and potentially more important explanations as well, for example the possible impact of behavioural factors, something maybe reflecting mechanisms related to stressful lives.
 Our results of the gender aspect is important in this respect.
@@
 Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being inconsistent in 1961 and reversed in 1991 with highest mortality in the lowest class [@cropr10].
-If these indications of changes in health-related behaviour are significant enough to explain our findings here can be discussed.
+If these indications of changes in health-related behaviour are significant enough to explain the present findings can be discussed.
 We believe that they certainly are at least an important part of the story.
@@
 In this article we set our task to analyse social differences in mortality from a long-term perspective.
-Until recently it has not been possible to investigate the development on mortality differences among adult and elderly populations.
+Until recently it has not been possible to investigate the really long-term development on mortality differences among adult and elderly populations.
 Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to the large class differences.
@@
 The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
-The long-term relation between economic inequality and health require further studies.
+The long-term relation between economic inequality and health is still a puzzle requiring further studies.
 

### 2017-09-09 article2.Rmd
so what?

--- 2017-09-07 article2.Rmd
@@
 A major theme in demographic and epidemiological studies, is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
-The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
+In the present study, we approach this issue by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population.
+Our main focus is on the relative differences.
+The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
 The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
 Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
-The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
-
-On the basis of the results we present, we argue that high social class is not necessarily always favourable for survival.
+The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, and their implications on some of the most influential hypotheses and concepts in health research.
+
+On the basis of the results we present, we argue that high social class is not necessarily favourable for survival.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
-
-## Social class and mortality
 
@@
 technique known as
-\emph{Cox regression} [@cox72, @gb02]. Instead we use a parametric model
+\emph{Cox regression} [@cox72]. Instead we use a parametric model
 that allows for reduction of the data

### 2017-09-09 R/get.file.R
need internet connection

@@ -1,12 +1,12 @@
 get.file <- function(name, link = "http://capa.ddb.umu.se/data/"){
     filename = paste("data/", name, ".rda", sep = "")
-    if (file.exists(filename)){
-        load(filename)
-    }else{
+    ##if (file.exists(filename)){
+      ##  load(filename)
+    ##}else{
         con <- url(paste(link, name, ".rda", sep = ""))
         load(con)
         close(con)
-        save(list = name, file = filename)
-    }
+        ##save(list = name, file = filename)
+    ##}
     get(name)
 }

### 2017-09-09 article2.Rmd
edit new periods

--- 2017-09-08 article2.Rmd
@@
 biblio-style: apalike
+fontsize: 12pt
 ---
@@
 
-A major theme in demographic and epidemiological studies, is the seemingly persistent effect of social class on mortality.
+A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
 In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
@@
 
-```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in"}
+```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in", out.height = "4.1in", fig.align="center"}
 knitr::include_graphics("myfigs/sweden.png")
@@
      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1801, 1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1991, 2014), las = 2)
 axis(2, las = 2, at = c(0, 40, 65, 90))
@@
 abline(h = 0)
-lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
+lines(c(1976, 1976, 2014, 2014, 1976), c(40, 90, 90, 40, 40),
       col = "blue", lwd = 2)
@@
 
-text(1985, 70, "Linnaeus", cex = 2, col = "blue")
+text(1995, 70, "Linnaeus", cex = 1.5, col = "blue")
 ##text(1984, 60, paste(n.rec, "records"))
-text(1985.2, 55, paste(n.exp, "p-years"))
-text(1985.2, 50, paste(n.d, "deaths"))
+text(1995.2, 55, paste(n.exp, "p-years"), cex = 0.8)
+text(1995.2, 50, paste(n.d, "deaths"))
 ```
@@
 
-# The effect of social class, 1801--1950, 1961--2013
+# The effect of social class, 1801--1950, 1976--2013
 
@@
 A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
-Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being inconsistent in 1961 and reversed in 1991 with highest mortality in the lowest class [@cropr10].
+Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with highest mortality in the lowest class [@cropr10].
 If these indications of changes in health-related behaviour are significant enough to explain the present findings can be discussed.

### 2017-09-09 article.Rmd
nothing...

@@
     toc_depth: 2
-    toc_float: yes
 bibliography: mybib.bib
@@
 vb0Tab$age <- factor(vb0Tab$age)
+rm(vb0)
 ```

### 2017-09-14 article2.Rmd
Nonsense

--- 2017-09-08 article2.Rmd
@@
   elderly in Northern Sweden 1801--2013
-author: "Sören Edvinsson and Göran Broström"
-date: "`r Sys.Date()`"
+author: Sören Edvinsson and Göran
+  Broström
+date: '`r Sys.Date()`'
 output:
-  bookdown::pdf_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
   bookdown::html_document2:

### 2017-09-15 article2.Rmd
Change of title

@@
 ---
-title: Is high social class always beneficial for survival? A study of adults and
+title: Is high social class always beneficial for survival? Adults and
   elderly in Northern Sweden 1801--2013

### 2017-09-19 article2.Rmd
Revisions and references

@@
 A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+In the present study, we challenge common notions of this by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
 The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
+The main issue is then analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age -- the retired population -- versus the population in working age.
+We analyse both total and mortality and mortality from the major causes of death (cardiovascular diseases and cancers).
 Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
@@
 to him, differences were comparatively small during the
-pre-transitional phase when space was a strong determinant for the spread of
-disease. Differences then increased during the transitional phase when mortality
+pre-transitional phase. This period was characterised by space being a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality
 declined and wealthy groups used their economic and human resources to gain better health. Finally,
@@
 pattern is [@bevp11]. Solid empirical evidence about the process is however lacking and studies
-focusing on the issue are still few. There is a need for more reliable
+focusing on the issue are still few. There is a need for additional reliable
 studies from different geographical and historical settings in order to 
 better understand the role of socio-economic conditions for health and survival over time.
+
+What about life course aspects of the impact of social position for health and mortality? There are
+identified three main hypotheses concerning social health inequalities in adulthood versus old age [@hof08].
+Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage).
+Diminishing
+differences may be a consequence of the circumstance that biological factors play a stronger role than social factors later in life. The status maintenance hypothesis basically assume 
+continuity in the determinants for social health inequalities from adulthood to old age.
+The third is the cumulative (dis)advantage hypothesis [@dan03], implying that advantages and disadvantages tend to persist and ackumulate during life, constantly rewarding some while disfavouring others.
+This would lead to larger differences in old age.
 
@@
 This is of particular interest since the levels of economic
-inequality were very different from those of the recent decades. Even if it
-not a necessary implication, it is reasonable to assume that poorer groups were
-those most disadvantaged of large inequalities.
+inequality were very different from those of the recent decades. Even if not a necessary implication, it is reasonable to assume that poorer groups were those most disadvantaged of large inequalities.
 
@@
 century, the Swedish railway system reached this part of Sweden. Thereby
-contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. The inland was particularly
-remote and isolated. During the 19th century, our regions became
+contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became
 increasingly integrated in the same epidemiological pattern as the rest of
@@
 very small population during the 19th century. The data from the period
-after 1960 cover the Skellefteå and Malå municipalities, the same area as
+after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
 for the earlier period but with the addition of two more parishes. The
@@
 
-The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1960 onwards of Umeå municipality when another three
+The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1976 onwards of Umeå municipality when another three
 parishes are included. This region has a somewhat different character from
@@
 registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1960 to 2000 are constructed from
-censuses every fifth year 1960--1990, with additional information on deaths
-from National Board of Health and Welfare. For the period 2002--2013 we use
+University. The study period from 1976 to 2000 are constructed from
+censuses every fifth year 1976--1990, with additional information on deaths
+from National Board of Health and Welfare (we have left out the period 1960--1975 due to more missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis).
+For the period 2002--2013 we use
 the information from the yearly population registers from Statistics Sweden
@@
 to have full and continuous control of the \emph{de jure} population. This is
-however not the case with the Linnaeus database. For the period 1960--2000
-we use presence in the censuses 1960--1990 and information on deaths from the National
+however not the case with the Linnaeus database. For the period 1976--2000
+we use presence in the censuses 1976--1990 and information on deaths from the National
 Board of Health and Welfare to identify the observation periods in our
@@
 
-In this paper we analyse mortality differences according to social class, treated as a time-dependent variable, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
+In this paper we analyse mortality differences according to social class, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
 type of socio-economic information available both in the Linnaeus and Poplink database
@@
 socio-economic mortality differences with consistent classification
-schemes. Social position is treated as a time-dependent variable.
+schemes.
+Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, i.e.
+from age 65 until death or last observation.
 
@@
 sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
-Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
+Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
 Yrkesklassificering).
 We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods.
-For the period 1991--2000 the social class in the census 1990 is considered to be kept
-until 2001. Furthermore, we consider the last social position before retirement
-to be kept until death or last observation for those above 65.
+For the period 1991--2000 the social class in the census 1990 is considered to be kept until 2001.
 
@@
 complication is however that it is  difficult to define households in the
-same way as in Poplink. Extra-marital cohabitation, connected to the Second
-Demographic Transition [@van87], became common and female
+same way as in Poplink. Extra-marital cohabitation became common and female
 labour force participation developed as the norm in Sweden during this
@@
 group from a majority of farmers during the 19th century to lower white
-collar from the middle of the 20th century. Furthermore we observe an
-increase during the 1960s and partly during later decades as well in those
-with undefined class. When first observed in the earliest included censuses,
-occupation is often missing. For men the proportion is somewhat higher in
-the age group 65 and above, i.e. those that were retired, but still low
-enough to be used in our analyses (results not shown). Until 1950, there were however very few
-with missing occupation. For men in working-active age, the proportion
-missing does not cause any problems for our analyses.
+collar from the middle of the 20th century.
+Furthermore we observe a rather high percentage of missing social class in the early 19th century.
+For men the proportion is somewhat higher in the age group 65 and above, i.e.
+those that were retired, but still low enough to be used in our analyses (results not shown).
+Until 1950, there were however very few with missing occupation.
 
@@
 century, reflecting the way womens' social position is defined.
-Missing social class is however a real problem during particularly the 1960s and 1970s.
-Around 80 \% lack information for the 1960s and 50 \% for the 1970s.
-Since social class for this period is defined
+Missing social class is however quite high during particularly the first two periods with data from the Linnaeus database -- from 1976.
+Since female social class for this period is defined
 according to own occupation, the consequence of many women being outside
-the workforce, the proportion missing becomes very high. We have chosen not
-to report results for women for these two decades. For the later decades,
+the workforce is that the proportion missing becomes very high. For the later periods,
 the results are fairly complete.
@@
 Regarding the distribution of marital status over time, see
-Figure \@ref(fig:civstskum). The figure indicates an under-registration of
-those married and over-registration of unmarried during the
-1960s.
+Figure \@ref(fig:civstskum).
 
@@
 living conditions as well as the risks for importing infections made the
-health risks higher in towns and cities. In the present paper, we control
+health risks higher in towns and cities. In the present paper, the variable **urban** controls
 for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
@@
 population size. The modern Swedish welfare state began to develop. For the 
-period 1960 onwards, there are separate analyses on decades except for the two 
+period 1976 onwards, there are separate analyses on decades except for the two 
 last periods. Sweden had developed into a wealthy
@@
 
-The cumulative
-hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis: These numbers are used in the process of
+The models are fitted separately for sex and each period, from  1801 to
+2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
+**civst** (marital status). The cumulative
+hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
+main target in our analysis. These numbers are used in the process of
 graphically illustrating the development over time.
@@
 Notice that we have not analysed the 1950s due to the fact that we only
-have data for some parishes at that time. Furthermore, the
-results for women during the 1960s and 1970 are highly misleading for the reasons discussed above. The
-changing economy also have consequences for the analysis, particularly when
+have data for some parishes at that time. The
+changing economy have some consequences for the analysis, particularly when
 it comes to the agricultural sector. While farmers were common occupations
@@
 onwards. This can however have to do with how we have defined social
-position for women before and after 1950. After 1960 it is based
+position for women before and after 1950. After 1975 it is based
 on their own occupation. Being without occupation probably include women that
@@
 during the nineteenth and much of the 20th century. Workers on the
-other hand have comparatively low mortality until the 1960s, at the same
-level as the middle class. The relative positions change radically towards
+other hand have comparatively low mortality until 1950, at the same
+level as the middle class.
+The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we still find better survival among workers compared to the elite.
+
+The relative positions change radically towards
 the end the 20th century. From the 1980s, there is a clear advantage when
@@
 for women, those without known occupation have much higher mortality than
-the others from the 1960s onwards. The assumption made for women of
+the others from the late 1970s onwards. The assumption made for women of
 unknown social class is probably even more relevant for men. They probably
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group (see Figure \@ref(fig:cumswold)). For
-women the results are mixed. Once again, too much missing data on social
-class force us to leave out the results for the 1960s and 1970s. We do not find the
+we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For
+women the results are mixed. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
@@
 In this and the following section, the focus is on analyses of mortality from the two
-dominant groups of death causes in the adult and elderly population. We start
+dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality. We restrict the analyses to the period 1901 on wards. We start
 with cardiovascular diseases. Many studies have shown that socio-economic
@@
 
-As before, models are fitted separately for sex and each ten-year time period, from  1901 to
-2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
-**civst** (marital status). The cumulative
-hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis. These numbers are used in the process of
-graphically illustrating the development over time.
-
 ### Ages 40--64
@@
 Figure \@ref(fig:41cumsw) shows mortality levels during different periods
-from 1901 on-wards and the result of the model can be seen in Table \@ref(tab:41cumswtab). Women from the highest social class had consistently
+from 1901 on-wards. Women from the highest social class had consistently
 lower mortality from cardiovascular diseases compared to the others. We
@@
 
-For men, there is a different pattern, a pattern however that to a large
-extent is consistent with what we observed concerning all-cause
+For men, there is a different pattern being to a large
+extent consistent with what we observed concerning all-cause
 mortality. The elite group had much higher cardiovascular mortality up until
@@
 Much of the decline in mortality among elderly from the middle of the 20th century can be explained by lower death rates in this disease group.
-The results from the model is presented in Table \@ref(tab:41cumswoldtab).
 
@@
 
-We use similar types of analyses as in the earlier sections. The cumulative
-hazards at the end of the period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis. These numbers are used in the process of
-graphically illustrating the development over time.
-
 ### Ages 40--64
@@
 
-In our analyses of class patterns in mortality, we finally turn to cancer
-mortality among the elderly. The model is presented in Table \@ref(tab:12cumswoldtab).
-
 ```{r 12hazcalwold}
@@
 
-The relation between different classes and cancer mortality fluctuates
+In our analyses of class patterns in mortality, we finally turn to cancer
+mortality among the elderly. The relation between different classes and cancer mortality fluctuates
 considerably both for women and men in this age group. For some periods it
@@
 
-Studies based on microdata on the adult population in
+Studies based on microdata on the adult and elderly population in
 history have been rare, mainly due to lack of adequate data. The results
 from such studies show other patterns than that for infant and child
-mortality. The expected social gradient in survival, meaning that people
+mortality. Some report substantial differences but many, particularly those based on
+Swedish data, have found surprisingly small or even reversed socio-economic
+differences during the 19th and early 20th century, sharply contrasting from the situation in present-day society. The expected social gradient in survival, meaning that people
 from higher social classes always had an advantage, was in many cases,
@@
 socio-economic differences in society, for example in the Sundsvall region [@edli11, @edbr12].
-The results presented here show that the consistent advantage for higher social classes do not appear until quite late in the investigated regions.
-
-The results showing a small or even reversed socio-economic gradient in the past and its late development are however in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
-They find the first indications of an emerging social gradient in the 1950s.
+The results presented here show that the consistent advantage for higher social classes do not appear until surprisingly late in the investigated regions.
+The basic pattern is the same in the working-age and the retired population, but the differences are smaller in old age in accordance with the status levelling hypothesis.
+
+The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
+They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as in the 1950s.
 Similar results have also been found in Estonia [@jaadla17].
@@
 Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67, @wowi95 @pamuk85].
-We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh.
-
-Instead, some studies suggest that the modern
-pattern might be a fairly recent development [@bedr11], although the pattern differs
-depending on age, gender or historical context. When it comes to the adult
-and elderly population results from different studies have diverged, some
-reporting substantial differences but many, particularly those based on
-Swedish data, have found surprisingly small or even reversed socio-economic
-differences during the 19th and early 20th century [@bedr11,
-@edli11, @edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
-some point during the mortality transition, at least in Sweden.
+We know less about how general these results are, but apparently the Swedish context we are studying differed from the English and Welsh.
 
 The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
-Some of the limitations have been addressed in the presentation of the variables: The fact that social class was defined according to household until 1950 but individually defined from 1960 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
+Some of the limitations have been addressed in the presentation of the variables: The fact that social class was defined according to household until 1950 but individually defined from 1976 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
 These differences are however not of the magnitude that the main conclusions are affected.
@@
 Occupational titles in historical sources are not always distinct enough to define the correct class.
-Studies covering long periods face the problem that occupations and classes change character --- they do not represent the same thing.
+Studies covering long periods face the problem that occupations and classes change character -- they do not represent the same thing.
 We have chosen to work with a few broad groups.
@@
 This does not necessarily contradict what we find here.
-Height is mainly determined by conditions in early years, and are not reflecting inequalities in adulthood.
+Height is mainly determined by conditions in early childhood, and are not reflecting inequalities in adulthood.
 Earlier studies in Sweden and in the areas studied here [@bredro02, @bube01, @mol17, @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
+
+An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
+Even though our study do not directly address the hypothesis raised by [@wipi09], our results are of interest for this question as well.
+The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
+After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
+at a time when a more expected pattern of a social hierarchy in survival appeared.
+The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
+The long-term relation between economic inequality and health is still a puzzle requiring further studies.
 
@@
 
-An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not directly address the hypothesis raised by [@wipi09], our results are of interest for this question as well.
-The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
-After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
-at a time when a more expected pattern of a social hierarchy in survival appeared.
-The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
-The long-term relation between economic inequality and health is still a puzzle requiring further studies.
-
 An important question is if our results can be generalised to other historical contexts.

### 2017-09-19 article2.Rmd
Changing title

@@
 ---
-title: Is high social class always beneficial for survival? Adults and
-  elderly in Northern Sweden 1801--2013
+title: Life course and long-term perspectives of social inequality in 
+       mortality among elderly and adults:  Northern Sweden                                     1801–2013
+
 author: Sören Edvinsson and Göran

### 2017-09-19 article2.Rmd
add &ouml;

--- 2017-09-19 article2.Rmd
@@
 ---
-title: Is high social class always beneficial for survival? Adults and
-  elderly in Northern Sweden 1801--2013
-author: Sören Edvinsson and Göran
-  Broström
+title: Life course and long-term perspectives of social inequality in mortality among
+  elderly and adults in Northern Sweden      1801–2013
+author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
 date: '`r Sys.Date()`'
 output:
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::html_document2:
@@
     toc_float: yes
-  bookdown::pdf_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
   bookdown::word_document2:
@@
 documentclass: article
+fontsize: 12pt
 citation-package: natbib
 biblio-style: apalike
-fontsize: 12pt
 ---

### 2017-09-19 article2.Rmd
fix refs, add cohort

--- 2017-09-19 article2.Rmd
@@
 ---
-title: Is high social class always beneficial for survival? Adults and
-  elderly in Northern Sweden 1801--2013
-author: Sören Edvinsson and Göran
-  Broström
+title: Life course and long-term perspectives of social inequality in mortality among
+  elderly and adults in Northern Sweden      1801–2013
+author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
 date: '`r Sys.Date()`'
 output:
+  bookdown::pdf_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
+    keep_tex: true
+    citation_package: natbib
   bookdown::html_document2:
@@
 documentclass: article
-citation-package: natbib
+fontsize: 12pt
+citation_package: natbib
 biblio-style: apalike
-fontsize: 12pt
 ---
@@
 
-\section{Discussion}
+# A cohort approach, all causes of death
+
+```{r cohortapproach}
+cohs <- levels(vb0Tab$cohort)
+fit <- glm(event ~ offset(log(exposure)) + hisclass, data = vb0Tab[vb0Tab$cohort == cohs[1], ],
+           family = poisson)
+summary(fit)
+```
+
+# Discussion
 
@@
 
-\section{Conclusion}
+# Conclusion
+
 In this article we set our task to analyse social differences in mortality from a long-term perspective.

### 2017-09-20 article2.Rmd
fix citations, English; remove life course aspects

--- 2017-09-19 article2.Rmd
@@
 ---
-title: Life course and long-term perspectives of social inequality in 
-       mortality among elderly and adults:  Northern Sweden                                     1801–2013
-
-author: Sören Edvinsson and Göran
-  Broström
+title: Life course and long-term perspectives of social inequality in mortality among
+  elderly and adults in Northern Sweden      1801–2013
+author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
 date: '`r Sys.Date()`'
 output:
+  bookdown::pdf_document2:
+    citation_package: natbib
+    keep_tex: yes
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
   bookdown::html_document2:
@@
 A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge common notions of this by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
-The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is then analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age -- the retired population -- versus the population in working age.
+In the present study, we challenge common notions of this by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
+The main issue is then analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age---the retired population---versus the population in working age.
 We analyse both total and mortality and mortality from the major causes of death (cardiovascular diseases and cancers).
@@
 Mortality risks in different contexts must be understood in the intersection between class and gender.
-We suggest that health-related behaviour was important not only in present-day societies, but were decisive also in earlier phases of the mortality transition.
+We suggest that health-related behaviour was important not only in present-day societies, but was decisive also in earlier phases of the mortality transition.
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
@@
 strong determinant when it comes to health and mortality and the impact even
-seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand10]. It has been suggested that
+seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. It has been suggested that
 "... social conditions have been, are and will continue to be irreducible
@@
 The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in this theory [@liph98].
-For a long time, the general view has been that socio-economic health and mortality inequalities were high in
-historical societies, probably higher than in modern societies. This is a
+For a long time, the general view has been that socio-economic health and mortality inequalities were large in
+historical societies, probably larger than in modern societies. This is a
 reasonable assumption since these societies were in most cases characterised by
@@
 Recent studies, investigating social inequality in health and mortality
-with micro-data, have however questioned how general the assumed
-pattern is [@bevp11]. Solid empirical evidence about the process is however lacking and studies
+with micro-data, have however questioned the generality of the assumed
+pattern [@bevp11]. Solid empirical evidence about the process is however lacking and studies
 focusing on the issue are still few. There is a need for additional reliable
@@
 
+<!--
 What about life course aspects of the impact of social position for health and mortality? There are
@@
 This would lead to larger differences in old age.
+-->
 
@@
 that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the present-day economically developed world.
-This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04,
-@wado00].
-During the last decades, the association between trends of inequalities in mortality and income respectively are weak or non-existant according to @hoetal16.
-How the association looked like in previous periods is unknown.
+This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04; @wado00].
+During the last decades, the association between trends of inequalities in mortality and income respectively is weak or non-existant according to @hoetal16.
+What the association looked like in previous periods is unknown.
 This is of particular interest since the levels of economic
@@
 
-When it comes to Sweden both income and wealth distribution was strongly skewed from the 1870's to the early 20th century. Starting during the inter-war period, income and
+When it comes to Sweden both income and wealth distributions were strongly skewed from the 1870's to the early 20th century. Starting during the inter-war period, income and
 wealth inequality continuously diminished to reach a low point at around
 1980. Thereafter, economic inequality has increased substantially until the
-present, although far from being as high as a century ago [@rowa08,
-@rowa09]. The Swedish development resembles that of other European and North
+present, although far from being as high as a century ago [@rowa08; @rowa09]. 
+The Swedish development resembles that of other European and North
 American countries in its basic features [@rowa14].
@@
 was dominated by agriculture, making it vulnerable to harvest failures, and
-several severe famines frequently occurred in the regions during the 1800s, for example after the harvest failure in 1867 [@segb14].
+several severe famines occurred in the regions during the 1800s, for example after the harvest failure in 1867 [@segb14].
 During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
@@
 majority of the 19th century population lived in rural villages and
-hamlets, getting their livelihood from agricultural production. During the
+hamlets, getting its livelihood from agricultural production. During the
 20th century, industrialisation took place. This also led to a population increase both in the town
@@
 The recent extension in time of data at the Demographic Data Base (DDB),
-Umeå University (www.cedar.umu.se) by including the period between 1900 and
-1950s makes it possible to investigate the long-term development of social
-inequality in mortality in a part of Sweden. The data for the present study comes from two large
-population databases at the DDB, whom provide us with micro-data on a sufficiently large population and
+Umeå University (http://www.cedar.umu.se) by including the period between 1900 and
+1950 makes it possible to investigate the long-term development of social
+inequality in mortality in a part of Sweden. The data for the present study come from two large
+population databases at the DDB, whom provide us with micro-data on a large population and
 with substantial social diversity for the Skellefteå and Umeå regions. The early period until 1950 is covered by the database Poplink, the digitisation of historical parish registers for
@@
 
-The other large data-set is extracted from the Linnaeus database
-[@malm10]. Linnaeus is based on different linked national population
+The other large data set is extracted from the Linnaeus database
+[@malm10], which is based on different linked national population
 registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1976 to 2000 are constructed from
-censuses every fifth year 1976--1990, with additional information on deaths
+University. The study period from 1976 to 2000 is constructed from
+censuses every fifth year 1975--1990, with additional information on deaths
 from National Board of Health and Welfare (we have left out the period 1960--1975 due to more missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis).
@@
 
-In the data-set analysed here, all individuals 40 years and older ever
+In the data set analysed here, all individuals 40 years and older ever
 being resident in either of the regions are included. The data file
@@
 `r pn.d` deaths in Poplink and `r as.integer(n.exp)` person years leading 
-to `r n.d` deaths in the Linnaeus database. See 
+to `r n.d` deaths in the Linnaeus database, see 
 Figure \@ref(fig:lexis2).
@@
 however not the case with the Linnaeus database. For the period 1976--2000
-we use presence in the censuses 1976--1990 and information on deaths from the National
+we use presence in the censuses 1975--1990 and information on deaths from the National
 Board of Health and Welfare to identify the observation periods in our
@@
 
-In this paper we analyse mortality differences according to social class, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
-type of socio-economic information available both in the Linnaeus and Poplink database
+Mortality differences are analysed according to social class, based on a modified form 
+of the classification scheme Hisclass [@lema11]. The only
+type of socio-economic information available both in the Linnaeus and the Poplink databases
 is occupations. Thus, analysing according to social class is apart from
@@
 schemes.
-Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, i.e.
-from age 65 until death or last observation.
-
-The availability to information on occupations varies over time and between
+Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, that is, from age 65 until death or last observation.
+
+The availability of information on occupations varies over time and between
 sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
-Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
+Linnaeus data, HISCO is recoded into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
 Yrkesklassificering).
@@
 
-The use of different data sets, forces us to apply different approaches
+The use of different data sets forces us to apply different approaches
 towards defining the social position. A problem is that
-Poplink usually only provides the occupation for the head of household,
-thus under-estimating female working-force participation as well as that of
+Poplink usually only provides the occupation of the head of household,
+thus underestimating female working-force participation as well as that of
 adult children residing with their parents [@vik10]. For the Poplink period (until 1950), we have chosen to categorise wives according to position of the head of household, usually
@@
 Hisclass, the classification system used here, is a "...
-HISCO-based historical international social class scheme" [@lema11,
-@lema02]. The different classes in Hisclass represent distinct
+HISCO-based historical international social class scheme" [@lema11; @lema02]. 
+The different classes in Hisclass represent distinct
 categories, based on whether the work is manual or non–manual, skill level,
@@
 Furthermore we observe a rather high percentage of missing social class in the early 19th century.
-For men the proportion is somewhat higher in the age group 65 and above, i.e.
-those that were retired, but still low enough to be used in our analyses (results not shown).
+For men the proportion is somewhat higher in the age group 65 and above, those that were retired, but still low enough to be used in our analyses (results not shown).
 Until 1950, there were however very few with missing occupation.
@@
 century, reflecting the way womens' social position is defined.
-Missing social class is however quite high during particularly the first two periods with data from the Linnaeus database -- from 1976.
+The frequency of missing social class is however quite high during particularly the first two periods with data from the Linnaeus database---from 1976.
 Since female social class for this period is defined
@@
 cancers. National compulsory reporting of cause of death was introduced in
-1911, but our analysis of the two disease groups starts from 1901 since deaths were fairly frequently reported even before 1911. Classification systems have however changed. Sweden did not join ICD
+1911, but our analysis of the two disease groups starts from 1901 since cause of death was fairly frequently reported even before 1911.
+Classification systems have however changed.
+Sweden did not join ICD
 until 1952, but the historical causes of death in the Poplink database have
@@
 Figure \@ref(fig:causeskum). Notice the high proportion of other causes in
-the first two periods. This category include cases with unknown or
+the first two periods. This category includes cases with unknown or
 unspecific causes which has implications for the cause-specific analyses
@@
 In 19th and early 20th century Sweden (as in other countries), mortality
-was much higher in urban environments [@hani94, @edv92]. The unsanitary and overcrowded
+was much higher in urban environments [@hani94; @edv92]. The unsanitary and overcrowded
 living conditions as well as the risks for importing infections made the
@@
 
-We have made separate analyses for the different periods. The century
-1801--1950 is divided into six periods of 25 years each. The 19th century
+We have made separate analyses for the different time periods. The period
+1801--1950 is divided into six subperiods of length 25 years each. The 19th century
 periods represent a mainly pre-industrial society dominated by agriculture
@@
 The proportional hazards model in a survival analysis context is used,
-allowing for adjustment for marital status and rural/urban environment. Since
-the central explanatory variable, social class (Hisclass), do not follow the
-proportionality property, the analyses are stratified with respect to the
+allowing for adjustment for marital status and rural/urban environment.
+The central explanatory variable, social class (Hisclass), does not follow the
+proportionality property, therefore the analyses are stratified with respect to the
 variable, and the main results are presented graphically.
 
-Another practical problem is the size of that data set:
+Another practical problem is the size of the data set:
 It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
@@
 calculated.  We have two sexes, three marital statuses, ten periods, two
-urban/rural groups, four social classes at age 40,  four social classes at age 60,
-ten age groups, and 22 cohorts, in all 422\,400 possible
+urban/rural groups, four social classes,
+ten age groups, and 22 cohorts, in all 105600 possible
 combinations. However, in this scheme there are many structural zeroes
 (impossible combinations) and some random zeros, so the actual number of
-combinations we end up with is 24\,326. The data set is thus reduced in
+combinations we end up with is 14084. The data set is thus reduced in
 size by 99 per cent.
@@
 \emph{Poisson regression}. The second way is faster than the first by a
-factor of size at least 100, often much more, depending on hardware, and
+factor of size at least 100, often much more, and
 the point is that the two ways give exactly the same results. This is
@@
 and graphics [@cran], especially using the package **eha**
-[@gb12;@eha]. The proportional hazards model allows us to scrutinise the differences and
-the development in more detail. We divide the analysis into two age groups,
+[@gb12;@eha]. In the the analysis, data are divided into two age groups,
 40--64 and 65--89 years of age, the first age group representing people that
@@
 results are presented graphically.  The comments on the results are based on
-the figures and focuses on the central variable in our study--- social class (**hisclass**).
+the figures and focuses on the central variable in our study, **hisclass**.
 
@@
 The models are fitted separately for sex and each period, from  1801 to
-2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
+2013. We stratify on **hisclass** and include the covariates **urban** (TRUE or FALSE) and
 **civst** (marital status). The cumulative
-hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
+hazards at the end of the age period (65 if 40--64, 90 if 65--89), conditional on survival 
+to the start of the age period, are the
 main target in our analysis. These numbers are used in the process of
@@
 Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
-40--64 during the different periods. Obviously, and not surprisingly, mortality
+40--64 during the different periods. Obviously, but not surprisingly, mortality
 is decreasing over time for both sexes. Women have consistently better survival
@@
 demographic history. Male mortality has almost always been higher than
-female in all age groups [@wil99, @suwi07]. We also observe a
+female in all age groups [@wil99; @suwi07]. We also observe a
 substantial decline in mortality for both sexes, resulting in smaller
@@
 accordance with previous analyses from the present research group on other
-regions in northern Sweden and for other periods [@edli11, @edbr12]. The
+regions in northern Sweden and for other periods [@edli11; @edbr12]. The
 highest social class usually have comparatively low mortality, while
@@
 
-We now turn to the elderly, i.e. those that in most cases have left the workforce.
+We now turn to the elderly, those that in most cases have left the workforce.
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
@@
 the fastest decline did not set in definitely until sometimes after the
-Second World War, i.e. later than for the age
+Second World War, that is, later than for the age
 group 40--64 years. This corresponds to the Swedish development where we
@@
 In this and the following section, the focus is on analyses of mortality from the two
-dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality. We restrict the analyses to the period 1901 on wards. We start
+dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality. We restrict the analyses to the period 1901 onwards. We start
 with cardiovascular diseases. Many studies have shown that socio-economic
@@
 Figure \@ref(fig:41cumsw) shows mortality levels during different periods
-from 1901 on-wards. Women from the highest social class had consistently
+from 1901 onwards. Women from the highest social class had consistently
 lower mortality from cardiovascular diseases compared to the others. We
@@
 
-In this section we focus on cardiovascular mortality among those that
+Here we focus on cardiovascular mortality among those that
 mostly have left the workforce, the age group 65--89 years.
@@
 Concerning the social pattern, we find somewhat different results for elderly women compared to the women in working-active age. Cardiovascular mortality among elderly
-women do not present a consistent class pattern. Female cardiovascular
+women does not present a consistent class pattern. Female cardiovascular
 mortality was somewhat higher among the elite and the middle group than
@@
 social class pattern has changed somewhat during the last periods becoming more
-in line to what is expected. The differences are however rather small.
+in line with what to expect. The differences are however rather small.
 
@@
 considered to be less clear when it comes to cancers. In our analyses,
-there is a risk that cancer mortality is under-estimated during the first
+there is a risk that cancer mortality is underestimated during the first
 half of the 20th century, illustrated by the already mentioned large
@@
 
-\section{Discussion}
+# A cohort approach, all causes of death, men 40--65
+
+```{r cohortapproach, fig.cap = "Cohort mortality, men aged 40-65."}
+vv <- vb0Tab[vb0Tab$cohort %in% seq(1751, 1951, by = 10) &
+               vb0Tab$age %in% seq(40, 60, by = 5), ]
+vv$age <- factor(vv$age)
+vv$cohort <- factor(vv$cohort)
+cohs <- levels(vv$cohort)
+vv$age <- factor(vv$age)
+n <- length(cohs)
+coe <- matrix(0, nrow = 4, ncol = n)
+for (i in 1:n){
+   fit <- glm(event ~ offset(log(exposure)) + hisclass + age + urban + civst, 
+              data = vv[vv$cohort == cohs[i] & vv$sex == "male", ],
+              family = poisson)
+   x <- coefficients(fit)[1:4]
+   x[2:4] <- x[1] + x[2:4]
+   coe[, i] <-  exp(x) / sum(exp(x))
+}
+x <- as.numeric(as.character(cohs))
+plot(x[-n], coe[1, -n], type = "b", col = scol[1], ylim = c(0, 0.7), xlab = "Birth cohort", ylab = "Standardized rel. risks")
+lines(x[-n], coe[2, -n], type = "b", col = scol[2])
+lines(x[-n], coe[3, -n], type = "b", col = scol[3])
+lines(x[-n], coe[4, -n], type = "b", col = scol[4])
+text(1775, 0.65, "elite", col = scol[1])
+text(1775, 0.55, "middle", col = scol[2])
+text(1825, 0.65, "worker", col = scol[3])
+text(1825, 0.55, "NA", col = scol[4])
+```
+
+In Figure \@ref(fig:cohortapproach) a crude cohort analysis for men is exposed (total mortality).
+It is "crude" in the sense that proportional hazards are assumed without checking it (in contrast to the period analyses).
+More work needs to be put into the cohort analysis.
+However, this crude result is well in line with what we have seen in the period analyses.
+
+# Discussion
 
@@
 pattern of mortality in historical northern Sweden for some time, focusing
-on different regions, age groups and time periods [@edv92, @edv04, @edli11, @edbr12]. When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic
-differences was often but not always confirmed [@edv92, @edv04],
-something that has been found in other Swedish studies as well [@bredro02,
-@bube01, @mol17].
+on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12]. When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic
+differences were often but not always confirmed [@edv92; @edv04],
+something that has been found in other Swedish studies as well 
+[@bredro02; @bube01; @mol17].
 
@@
 particularly among adult men, not found despite the well-documented
-socio-economic differences in society, for example in the Sundsvall region [@edli11, @edbr12].
+socio-economic differences in society, for example in the Sundsvall region [@edli11; @edbr12].
 The results presented here show that the consistent advantage for higher social classes do not appear until surprisingly late in the investigated regions.
@@
 These findings are however contrary towards what usually is assumed and what other studies have shown.
-Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67, @wowi95 @pamuk85].
+Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85].
 We know less about how general these results are, but apparently the Swedish context we are studying differed from the English and Welsh.
@@
 Occupational titles in historical sources are not always distinct enough to define the correct class.
-Studies covering long periods face the problem that occupations and classes change character -- they do not represent the same thing.
+Studies covering long periods face the problem that occupations and classes change character---they do not represent the same thing.
 We have chosen to work with a few broad groups.
@@
 
-In this paper we have analysed social inequality in health from the
+We have analysed social inequality in health from the
 perspective of social classes. However, there are
@@
 Height is mainly determined by conditions in early childhood, and are not reflecting inequalities in adulthood.
-Earlier studies in Sweden and in the areas studied here [@bredro02, @bube01, @mol17, @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
+Earlier studies in Sweden and in the areas studied here [@bredro02; @bube01; @mol17; @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
 
 An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not directly address the hypothesis raised by [@wipi09], our results are of interest for this question as well.
+Even though our study do not directly address the hypothesis raised by @wipi09, our results are of interest for this question as well.
 The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
-After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
-at a time when a more expected pattern of a social hierarchy in survival appeared.
+After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, that is, at a time when a more expected pattern of a social hierarchy in survival appeared.
 The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
@@
 
-Our study seems to be in conflict with many of the dominating theories in present—day health research.
+Our study seems to be in conflict with many of the dominating theories in present-day health research.
 In an overview of explanations for the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital et cetera.
@@
 Our results of the gender aspect is important in this respect.
-As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93, @mack95, @jans07].
+As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07].
 The gender difference in mortality may be caused by discrimination on the female part in the household.
@@
 @edli11 suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and sometimes also taking risks.
-Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99, @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
-Different historical contexts promote different ways to spend economic resources (compare @rasp06 and @jaadla17.
-
-In the present-day high-income world including Sweden, smoking and drinking contributes to a substantial part of the socio-economic health differences [@mapemy14, @syla16].
+Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
+Different historical contexts promote different ways to spend economic resources [@rasp06; @jaadla17].
+
+In the present-day high-income world including Sweden, smoking and drinking contributes to a substantial part of the socio-economic health differences [@mapemy14; @syla16].
 Our study suggests a strong impact also in historical Sweden.
 If this interpretation is correct, the hypothesis of @ant67 on an increasing role of health related behaviour in the modern low-mortality régime, while economic inequality determined the social mortality pattern during the early phases of the mortality transition can be questioned.
-Instead, health-related can be equally strong earlier in history but expressed in a completely other way.
+Instead, health-related inequality can be equally strong earlier in history but expressed in a completely other way.
 To corroborate this claim, we would need to have evidence on the social differentiation of consumption patterns and behaviours in history.
@@
 Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with highest incomes.
-There was also an over-representation of alcohol-related deaths in higher social classes in the 1960 [@noro98].
-@noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with a more skewed consumption distribution, indicating more high-consumers in this category.
+There was also an over-representation of alcohol-related deaths in higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with a more skewed consumption distribution, indicating more high-consumers in this category.
 A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
@@
 The social pattern of female mortality are in line with the expected, but a social gradient in male adult mortality is a recent phenomenon.
-Furthermore, inequality in mortality appears clearer for the working—age population, while the differences are smaller among the elderly.
+Furthermore, inequality in mortality appears clearer for the working-age population, while the differences are smaller among the elderly.
 Concerning specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.

### 2017-09-21 article2.Rmd
Fina revisions of the IUSSP version

@@
 The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is then analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age---the retired population---versus the population in working age.
+The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age---the retired population---versus the population in working age.
 We analyse both total and mortality and mortality from the major causes of death (cardiovascular diseases and cancers).
@@
 The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in this theory [@liph98].
-For a long time, the general view has been that socio-economic health and mortality inequalities were large in
-historical societies, probably larger than in modern societies. This is a
+For a long time, the general view has been that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
+This is a
 reasonable assumption since these societies were in most cases characterised by
@@
 resources provides advantages in all aspects of life, the health advantage of higher
-classes ought to be obvious. @ant67 suggests that social
-inequality in mortality has passed through different historical phases. According
-to him, differences were comparatively small during the
+classes ought to be obvious.
+
+@ant67 suggests that social inequality in mortality has passed through different historical phases. According to him, differences were comparatively small during the
 pre-transitional phase. This period was characterised by space being a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality
@@
 
-<!--
-What about life course aspects of the impact of social position for health and mortality? There are
-identified three main hypotheses concerning social health inequalities in adulthood versus old age [@hof08].
-Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage).
+What about life course aspects of the impact of social position for health and mortality ?
+Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage) (on this issue, see [@hof08]).
 Diminishing
-differences may be a consequence of the circumstance that biological factors play a stronger role than social factors later in life. The status maintenance hypothesis basically assume 
+differences may be a consequence of the circumstance that biological factors becomes increasingly important during the ageing process and in old age, leaving less impact for social factors.
+The status maintenance hypothesis basically assume
 continuity in the determinants for social health inequalities from adulthood to old age.
-The third is the cumulative (dis)advantage hypothesis [@dan03], implying that advantages and disadvantages tend to persist and ackumulate during life, constantly rewarding some while disfavouring others.
-This would lead to larger differences in old age.
--->
+The cumulative (dis)advantage hypothesis [@dan03], imply that advantages and disadvantages tend to persist and ackumulate during life in a negative spiral rewarding some while disfavouring others.
+This leads to larger differences in old age.
 
@@
 inequality in mortality in a part of Sweden. The data for the present study come from two large
-population databases at the DDB, whom provide us with micro-data on a large population and
-with substantial social diversity for the Skellefteå and Umeå regions. The early period until 1950 is covered by the database Poplink, the digitisation of historical parish registers for
-the Skellefteå and Umeå region [@wee16]. Poplink is based on linked
+population databases at the DDB, whom provide us with micro-data for the Skellefteå and Umeå regions in nortehrn Sweden.
+Our dataset constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
+The early period until 1950 is covered by the database Poplink, the digitisation of historical parish registers for
+the two regions [@wee16]. Poplink is based on linked
 parish records, allowing us to reconstruct life biographies on people as
@@
 to have full and continuous control of the \emph{de jure} population. This is
-however not the case with the Linnaeus database. For the period 1976--2000
+however not the case with the Linnaeus database. For the period 1976--2001
 we use presence in the censuses 1975--1990 and information on deaths from the National
@@
 resident in our regions during the complete five-year period between
-censuses if the person is missing in the later census (the last period covering ten years). This marginally overestimates presence, but should not bias the results when it
+censuses if the person is missing in the later census (the last period covering eleven years). This marginally overestimates presence, but should not bias the results when it
 comes to socioeconomic differences. We furthermore miss those that both
@@
 The use of different data sets forces us to apply different approaches
-towards defining the social position. A problem is that
-Poplink usually only provides the occupation of the head of household,
-thus underestimating female working-force participation as well as that of
+towards defining the social position. Poplink usually only provides occupation for the head of household, thus underestimating female working-force participation as well as that of
 adult children residing with their parents [@vik10]. For the Poplink period (until 1950), we have chosen to categorise wives according to position of the head of household, usually
@@
 by their individual occupation for this period. Still, the results for the
-different periods are not completely comparable and makes the
-interpretation of the development for women uncertain for some periods.
-Notice however that there probably is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].
-
-Hisclass, the classification system used here, is a "...
+different periods are not completely comparable.
+Notice however that we can expect a strong social homogamy within the married couple, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].
+
+Hisclass, the classification system used as a basis for our categorisation, is a "...
 HISCO-based historical international social class scheme" [@lema11; @lema02]. 
@@
 Longitudinal analyses of social classes for such a long period covered in this study involves several problems, related to changes in occupational structure and the meaning and contents of occupations and social positions.
-We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth
-category with unknown class:
+We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth category of unknown class:
 
@@
 For men the proportion is somewhat higher in the age group 65 and above, those that were retired, but still low enough to be used in our analyses (results not shown).
-Until 1950, there were however very few with missing occupation.
-
-```{r hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
+If we disregard the very first period, there were however very few with missing occupation until 1950.
+
+```{r hiscskumm, fig.cap = "Exposure by social class, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
 xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
+```{r hiscskumw, fig.cap = "Exposure by social class, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
 xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. We must however distinguish between differences in absolute risks compared to relative risks [@hoetal16). The compression that is
+absolute differences between social classes. We must however distinguish between differences in absolute risks compared to relative risks [@hoetal16]. The compression that is
 visible for later decades is partly explained by the decreasing general
@@
 up until the first half of the 20th century, these occupations has constantly
-decreased in numbers thereafter. This means that the middle class has
+decreased in numbers thereafter. This means that the middle category has
 a different character during the last periods.
@@
 
-For men, there is a different pattern being to a large
+For men, we find a different pattern, however it being to a large
 extent consistent with what we observed concerning all-cause
@@
 Figure \@ref(fig:12cumsw) shows relative mortality levels according to
-Hisclass during the different studied periods for the age group
+social class during the different studied periods for the age group
 40--64. Cancer mortality among elite women was lower than that of the
@@
 
-In Figure \@ref(fig:cohortapproach) a crude cohort analysis for men is exposed (total mortality).
+In Figure \@ref(fig:cohortapproach) a crude cohort analysis (ten year birth cohorts) for men in working age is exposed (total mortality), i.e.
+the group where we found the largest shift in socio-economic pattern in mortality.
 It is "crude" in the sense that proportional hazards are assumed without checking it (in contrast to the period analyses).
@@
 However, this crude result is well in line with what we have seen in the period analyses.
+Mortality is highest in the elite for almost all cohorts until the late 19th century.
+Only from the cohort born in the 20th century do we find disappearing differences, ending with a consistent hierarchical social pattern for the cohorts born a couple of decades into that century.
 

### 2017-09-22 article2.Rmd
trivia

--- 2017-09-19 article2.Rmd
@@
 ---
-title: Life course and long-term perspectives of social inequality in mortality among
-  elderly and adults in Northern Sweden      1801–2013
+title: Is high social class always beneficial for survival? A study of adults and
+  elderly in Northern Sweden 1801--2013
 author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
-date: '`r Sys.Date()`'
+date: "`r Sys.Date()`"
 output:
-  bookdown::pdf_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
-    keep_tex: true
-    citation_package: natbib
   bookdown::html_document2:
@@
 documentclass: article
-fontsize: 12pt
-citation_package: natbib
+citation-package: natbib
 biblio-style: apalike
@@
 A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge common notions of this by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
 The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is then analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age -- the retired population -- versus the population in working age.
-We analyse both total and mortality and mortality from the major causes of death (cardiovascular diseases and cancers).
+The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers).
 Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
@@
 to him, differences were comparatively small during the
-pre-transitional phase. This period was characterised by space being a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality
+pre-transitional phase when space was a strong determinant for the spread of
+disease. Differences then increased during the transitional phase when mortality
 declined and wealthy groups used their economic and human resources to gain better health. Finally,
@@
 pattern is [@bevp11]. Solid empirical evidence about the process is however lacking and studies
-focusing on the issue are still few. There is a need for additional reliable
+focusing on the issue are still few. There is a need for more reliable
 studies from different geographical and historical settings in order to 
 better understand the role of socio-economic conditions for health and survival over time.
-
-What about life course aspects of the impact of social position for health and mortality? There are
-identified three main hypotheses concerning social health inequalities in adulthood versus old age [@hof08].
-Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage).
-Diminishing
-differences may be a consequence of the circumstance that biological factors play a stronger role than social factors later in life. The status maintenance hypothesis basically assume 
-continuity in the determinants for social health inequalities from adulthood to old age.
-The third is the cumulative (dis)advantage hypothesis [@dan03], implying that advantages and disadvantages tend to persist and ackumulate during life, constantly rewarding some while disfavouring others.
-This would lead to larger differences in old age.
 
@@
 This is of particular interest since the levels of economic
-inequality were very different from those of the recent decades. Even if not a necessary implication, it is reasonable to assume that poorer groups were those most disadvantaged of large inequalities.
+inequality were very different from those of the recent decades. Even if it
+not a necessary implication, it is reasonable to assume that poorer groups were
+those most disadvantaged of large inequalities.
 
@@
 
-```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in", out.height = "4.1in", fig.align="center"}
+```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in"}
 knitr::include_graphics("myfigs/sweden.png")
@@
 century, the Swedish railway system reached this part of Sweden. Thereby
-contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became
+contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. The inland was particularly
+remote and isolated. During the 19th century, our regions became
 increasingly integrated in the same epidemiological pattern as the rest of
@@
 very small population during the 19th century. The data from the period
-after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
+after 1960 cover the Skellefteå and Malå municipalities, the same area as
 for the earlier period but with the addition of two more parishes. The
@@
 
-The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1976 onwards of Umeå municipality when another three
+The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1960 onwards of Umeå municipality when another three
 parishes are included. This region has a somewhat different character from
@@
 registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1976 to 2000 are constructed from
-censuses every fifth year 1976--1990, with additional information on deaths
-from National Board of Health and Welfare (we have left out the period 1960--1975 due to more missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis).
-For the period 2002--2013 we use
+University. The study period from 1960 to 2000 are constructed from
+censuses every fifth year 1960--1990, with additional information on deaths
+from National Board of Health and Welfare. For the period 2002--2013 we use
 the information from the yearly population registers from Statistics Sweden
@@
      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1991, 2014), las = 2)
+axis(1, at = c(1801, 1851, 1901, 1951, 1961, 1981, 2014), las = 2)
 axis(2, las = 2, at = c(0, 40, 65, 90))
@@
 abline(h = 0)
-lines(c(1976, 1976, 2014, 2014, 1976), c(40, 90, 90, 40, 40),
+lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
       col = "blue", lwd = 2)
@@
 
-text(1995, 70, "Linnaeus", cex = 1.5, col = "blue")
+text(1985, 70, "Linnaeus", cex = 2, col = "blue")
 ##text(1984, 60, paste(n.rec, "records"))
-text(1995.2, 55, paste(n.exp, "p-years"), cex = 0.8)
-text(1995.2, 50, paste(n.d, "deaths"))
+text(1985.2, 55, paste(n.exp, "p-years"))
+text(1985.2, 50, paste(n.d, "deaths"))
 ```
@@
 to have full and continuous control of the \emph{de jure} population. This is
-however not the case with the Linnaeus database. For the period 1976--2000
-we use presence in the censuses 1976--1990 and information on deaths from the National
+however not the case with the Linnaeus database. For the period 1960--2000
+we use presence in the censuses 1960--1990 and information on deaths from the National
 Board of Health and Welfare to identify the observation periods in our
@@
 
-In this paper we analyse mortality differences according to social class, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
+In this paper we analyse mortality differences according to social class, treated as a time-dependent variable, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
 type of socio-economic information available both in the Linnaeus and Poplink database
@@
 socio-economic mortality differences with consistent classification
-schemes.
-Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, i.e.
-from age 65 until death or last observation.
+schemes. Social position is treated as a time-dependent variable.
 
@@
 sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
-Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
+Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
 Yrkesklassificering).
 We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods.
-For the period 1991--2000 the social class in the census 1990 is considered to be kept until 2001.
+For the period 1991--2000 the social class in the census 1990 is considered to be kept
+until 2001. Furthermore, we consider the last social position before retirement
+to be kept until death or last observation for those above 65.
 
@@
 complication is however that it is  difficult to define households in the
-same way as in Poplink. Extra-marital cohabitation became common and female
+same way as in Poplink. Extra-marital cohabitation, connected to the Second
+Demographic Transition [@van87], became common and female
 labour force participation developed as the norm in Sweden during this
@@
 group from a majority of farmers during the 19th century to lower white
-collar from the middle of the 20th century.
-Furthermore we observe a rather high percentage of missing social class in the early 19th century.
-For men the proportion is somewhat higher in the age group 65 and above, i.e.
-those that were retired, but still low enough to be used in our analyses (results not shown).
-Until 1950, there were however very few with missing occupation.
+collar from the middle of the 20th century. Furthermore we observe an
+increase during the 1960s and partly during later decades as well in those
+with undefined class. When first observed in the earliest included censuses,
+occupation is often missing. For men the proportion is somewhat higher in
+the age group 65 and above, i.e. those that were retired, but still low
+enough to be used in our analyses (results not shown). Until 1950, there were however very few
+with missing occupation. For men in working-active age, the proportion
+missing does not cause any problems for our analyses.
 
@@
 century, reflecting the way womens' social position is defined.
-Missing social class is however quite high during particularly the first two periods with data from the Linnaeus database -- from 1976.
-Since female social class for this period is defined
+Missing social class is however a real problem during particularly the 1960s and 1970s.
+Around 80 \% lack information for the 1960s and 50 \% for the 1970s.
+Since social class for this period is defined
 according to own occupation, the consequence of many women being outside
-the workforce is that the proportion missing becomes very high. For the later periods,
+the workforce, the proportion missing becomes very high. We have chosen not
+to report results for women for these two decades. For the later decades,
 the results are fairly complete.
@@
 Regarding the distribution of marital status over time, see
-Figure \@ref(fig:civstskum).
+Figure \@ref(fig:civstskum). The figure indicates an under-registration of
+those married and over-registration of unmarried during the
+1960s.
 
@@
 living conditions as well as the risks for importing infections made the
-health risks higher in towns and cities. In the present paper, the variable **urban** controls
+health risks higher in towns and cities. In the present paper, we control
 for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
@@
 population size. The modern Swedish welfare state began to develop. For the 
-period 1976 onwards, there are separate analyses on decades except for the two 
+period 1960 onwards, there are separate analyses on decades except for the two 
 last periods. Sweden had developed into a wealthy
@@
 
-# The effect of social class, 1801--1950, 1976--2013
+# The effect of social class, 1801--1950, 1961--2013
 
@@
 
-The models are fitted separately for sex and each period, from  1801 to
-2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
-**civst** (marital status). The cumulative
-hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
-main target in our analysis. These numbers are used in the process of
+The cumulative
+hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
+main target in our analysis: These numbers are used in the process of
 graphically illustrating the development over time.
@@
 Notice that we have not analysed the 1950s due to the fact that we only
-have data for some parishes at that time. The
-changing economy have some consequences for the analysis, particularly when
+have data for some parishes at that time. Furthermore, the
+results for women during the 1960s and 1970 are highly misleading for the reasons discussed above. The
+changing economy also have consequences for the analysis, particularly when
 it comes to the agricultural sector. While farmers were common occupations
@@
 onwards. This can however have to do with how we have defined social
-position for women before and after 1950. After 1975 it is based
+position for women before and after 1950. After 1960 it is based
 on their own occupation. Being without occupation probably include women that
@@
 during the nineteenth and much of the 20th century. Workers on the
-other hand have comparatively low mortality until 1950, at the same
-level as the middle class.
-The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we still find better survival among workers compared to the elite.
-
-The relative positions change radically towards
+other hand have comparatively low mortality until the 1960s, at the same
+level as the middle class. The relative positions change radically towards
 the end the 20th century. From the 1980s, there is a clear advantage when
@@
 for women, those without known occupation have much higher mortality than
-the others from the late 1970s onwards. The assumption made for women of
+the others from the 1960s onwards. The assumption made for women of
 unknown social class is probably even more relevant for men. They probably
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For
-women the results are mixed. We do not find the
+we find for the younger age group (see Figure \@ref(fig:cumswold)). For
+women the results are mixed. Once again, too much missing data on social
+class force us to leave out the results for the 1960s and 1970s. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
@@
 In this and the following section, the focus is on analyses of mortality from the two
-dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality. We restrict the analyses to the period 1901 on wards. We start
+dominant groups of death causes in the adult and elderly population. We start
 with cardiovascular diseases. Many studies have shown that socio-economic
@@
 
+As before, models are fitted separately for sex and each ten-year time period, from  1901 to
+2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
+**civst** (marital status). The cumulative
+hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
+main target in our analysis. These numbers are used in the process of
+graphically illustrating the development over time.
+
 ### Ages 40--64
@@
 
-For men, there is a different pattern being to a large
-extent consistent with what we observed concerning all-cause
+For men, there is a different pattern, a pattern however that to a large
+extent is consistent with what we observed concerning all-cause
 mortality. The elite group had much higher cardiovascular mortality up until
@@
 Much of the decline in mortality among elderly from the middle of the 20th century can be explained by lower death rates in this disease group.
+The results from the model is presented in Table \@ref(tab:41cumswoldtab).
 
@@
 
+We use similar types of analyses as in the earlier sections. The cumulative
+hazards at the end of the period (65 if 40--64, 90 if 65--89) are the
+main target in our analysis. These numbers are used in the process of
+graphically illustrating the development over time.
+
 ### Ages 40--64
@@
 
+In our analyses of class patterns in mortality, we finally turn to cancer
+mortality among the elderly. The model is presented in Table \@ref(tab:12cumswoldtab).
+
 ```{r 12hazcalwold}
@@
 
-In our analyses of class patterns in mortality, we finally turn to cancer
-mortality among the elderly. The relation between different classes and cancer mortality fluctuates
+The relation between different classes and cancer mortality fluctuates
 considerably both for women and men in this age group. For some periods it
@@
 
-# A cohort approach, all causes of death
-
-```{r cohortapproach}
-cohs <- levels(vb0Tab$cohort)
-fit <- glm(event ~ offset(log(exposure)) + hisclass, data = vb0Tab[vb0Tab$cohort == cohs[1], ],
-           family = poisson)
-summary(fit)
-```
-
-# Discussion
+\section{Discussion}
 
@@
 
-Studies based on microdata on the adult and elderly population in
+Studies based on microdata on the adult population in
 history have been rare, mainly due to lack of adequate data. The results
 from such studies show other patterns than that for infant and child
-mortality. Some report substantial differences but many, particularly those based on
-Swedish data, have found surprisingly small or even reversed socio-economic
-differences during the 19th and early 20th century, sharply contrasting from the situation in present-day society. The expected social gradient in survival, meaning that people
+mortality. The expected social gradient in survival, meaning that people
 from higher social classes always had an advantage, was in many cases,
@@
 socio-economic differences in society, for example in the Sundsvall region [@edli11, @edbr12].
-The results presented here show that the consistent advantage for higher social classes do not appear until surprisingly late in the investigated regions.
-The basic pattern is the same in the working-age and the retired population, but the differences are smaller in old age in accordance with the status levelling hypothesis.
-
-The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
-They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as in the 1950s.
+The results presented here show that the consistent advantage for higher social classes do not appear until quite late in the investigated regions.
+
+The results showing a small or even reversed socio-economic gradient in the past and its late development are however in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
+They find the first indications of an emerging social gradient in the 1950s.
 Similar results have also been found in Estonia [@jaadla17].
@@
 Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67, @wowi95 @pamuk85].
-We know less about how general these results are, but apparently the Swedish context we are studying differed from the English and Welsh.
+We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh.
+
+Instead, some studies suggest that the modern
+pattern might be a fairly recent development [@bedr11], although the pattern differs
+depending on age, gender or historical context. When it comes to the adult
+and elderly population results from different studies have diverged, some
+reporting substantial differences but many, particularly those based on
+Swedish data, have found surprisingly small or even reversed socio-economic
+differences during the 19th and early 20th century [@bedr11,
+@edli11, @edbr12], sharply contrasting from the situation in present-day society. It thus appears as if a more consistent social health divide in adult age appeared at
+some point during the mortality transition, at least in Sweden.
 
 The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
-Some of the limitations have been addressed in the presentation of the variables: The fact that social class was defined according to household until 1950 but individually defined from 1976 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
+Some of the limitations have been addressed in the presentation of the variables: The fact that social class was defined according to household until 1950 but individually defined from 1960 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
 These differences are however not of the magnitude that the main conclusions are affected.
@@
 Occupational titles in historical sources are not always distinct enough to define the correct class.
-Studies covering long periods face the problem that occupations and classes change character -- they do not represent the same thing.
+Studies covering long periods face the problem that occupations and classes change character --- they do not represent the same thing.
 We have chosen to work with a few broad groups.
@@
 This does not necessarily contradict what we find here.
-Height is mainly determined by conditions in early childhood, and are not reflecting inequalities in adulthood.
+Height is mainly determined by conditions in early years, and are not reflecting inequalities in adulthood.
 Earlier studies in Sweden and in the areas studied here [@bredro02, @bube01, @mol17, @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
-
-An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not directly address the hypothesis raised by [@wipi09], our results are of interest for this question as well.
-The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
-After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
-at a time when a more expected pattern of a social hierarchy in survival appeared.
-The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
-The long-term relation between economic inequality and health is still a puzzle requiring further studies.
 
@@
 A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
-Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with highest mortality in the lowest class [@cropr10].
+Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being inconsistent in 1961 and reversed in 1991 with highest mortality in the lowest class [@cropr10].
 If these indications of changes in health-related behaviour are significant enough to explain the present findings can be discussed.
@@
 
-# Conclusion
-
+\section{Conclusion}
 In this article we set our task to analyse social differences in mortality from a long-term perspective.
@@
 Studies of these aspects and their role for health and mortality in different times and places require further development.
+
+An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
+Even though our study do not directly address the hypothesis raised by [@wipi09], our results are of interest for this question as well.
+The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
+After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
+at a time when a more expected pattern of a social hierarchy in survival appeared.
+The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
+The long-term relation between economic inequality and health is still a puzzle requiring further studies.
 

### 2017-09-22 article2.Rmd

--- 2017-09-21 article2.Rmd
@@
     toc_depth: 2
-  bookdown::html_document2:
+    toc_float: yes
+  bookdown::pdf_document2:
     number_sections: yes
@@
     toc_depth: 2
-    toc_float: yes
   bookdown::word_document2:

### 2017-09-22 article2.Rmd
fix yaml

@@
     toc_depth: 2
-    toc_float: yes
-  bookdown::pdf_document2:
+  bookdown::html_document2:
     number_sections: yes
@@
     toc_depth: 2
+    toc_float: yes
   bookdown::word_document2:

### 2017-10-05 cape.Rnw
first commit

--- 2017-06-12 abo.Rnw
@@
 
-\title{Increasing the mortality gap? Social inequality in mortality among elderly and adults in
-  Northern  Sweden 1851--2013\footnote{Session "Health, morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, Turku, Finland}
+\title{Increasing the mortality gap? Social inequality in mortality among 
+elderly and adults in Northern  Sweden 1851--2013\footnote{Session "Health, 
+morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, 
+Turku, Finland}
 }
@@
 library(skum)
-## Load remote data:
-con <- url("http://capa.ddb.umu.se/data/vb0.rda")
-##load("data/vb0.rda")
-load(con)
-close(con)
-vb0$hisclass[is.na(vb0$hisclass)] <- 4
-vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
+@
+
+<<readalldata>>=
+source("R/get.file.R")
+skel <- get.file("skel")
 ##
-con <- url("http://capa.ddb.umu.se/data/vb.rda")
-##load("data/vb.rda")
-load(con)
-close(con)
+skume <- get.file("skume")
 ##
-con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
-##load("data/lisaold.rda")
-load(con)
-close(con)
-##
-vb0Tab <- aggregate(vb0[, c("event", "exposure")],
-                   vb0[, c("sex", "age", "period", "cohort",
-                          "urban", "civst", "hisclass")],
-                   FUN = sum)
+foli <- get.file("foli")
+skel$urban <- FALSE
+vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
 vb0Tab$age <- factor(vb0Tab$age)
-
 @
@@
 <<lexis2>>=
-plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
-     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
+plot(c(1801, 1801, 1951, 1951, 1801), c(40, 90, 90, 40, 40),
+     type = "l", xlim = c(1800, 2014), lwd = 2, col = "red",
      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
-axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
+axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1991, 2014), las = 2)
 axis(2, las = 2, at = c(0, 40, 65, 90))
@@
 abline(h = 0)
-lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
+lines(c(1976, 1976, 2014, 2014, 1976), c(40, 90, 90, 40, 40),
       col = "blue", lwd = 2)
 #
-n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
-n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
-text(1900, 70, "PopLink", cex = 2, col = "red")
-text(1899, 60, paste(n.rec, "records"))
-text(1900, 55, paste(n.pl, "persons"))
-text(1899.5, 50, paste(n.d, "deaths"))
+pn.exp <- round(sum(skel$exposure) + sum(skume$exposure))
+pn.d <-  as.integer(sum(skel$event) + sum(skume$event))
 #
-n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
-n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
-n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))
+text(1875, 70, "PopLink", cex = 2, col = "red")
+##text(1899, 60, paste(pn.rec, "records"))
+text(1875, 55, paste(pn.exp, "p-years"))
+text(1874.5, 50, paste(pn.d, "deaths"))
+#
+n.exp <- round(sum(foli$exposure))
+n.d <-  as.integer(sum(foli$event))
 
-text(1985, 70, "Linnaeus", cex = 2, col = "blue")
-text(1984, 60, paste(n.rec, "records"))
-text(1985.2, 55, paste(n.pl, "persons"))
-text(1985.2, 50, paste(n.d, "deaths"))
+text(1995, 70, "Linnaeus", cex = 1.5, col = "blue")
+##text(1984, 60, paste(n.rec, "records"))
+text(1995.2, 55, paste(n.exp, "p-years"), cex = 0.8)
+text(1995.2, 50, paste(n.d, "deaths"))
 @
@@
   %\begin{center}
-\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswold-1}
+\includegraphics[height=7.5cm,width=5.4cm]{figs/art2-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/art2-cumswold-1}
    %\end{center}

### 2017-10-06 cape.Rnw
improve a figure

@@
 library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/ume-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/ume-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 @
 
-<<environ>>=
+<<environ>>=n
 library(knitr)
@@
 source("R/plotED.glm.R")
-scol <- c("blue", "darkgreen", "red", "black")
+scol <- c("blue", "darkgreen", "darkgreen", "red")
 oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Women 40-64", exclude = 5:6)
+                main = "Women 40-64")## exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
                 main = "Men 40-64")##, exclude = 5:6)
+par(oldpar)
+@
+
+\end{frame}
+
+\begin{frame}{Women and men, age 40--64, 1976--2013}
+
+<<cumswlate>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
+                main = "Women 40-64", exclude = 1:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
+                main = "Men 40-64", exclude = 1:6)
 par(oldpar)
@@
 \includegraphics[height=7.5cm,width=5.4cm]{figs/art2-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/art2-cumswold-1}
+   %\end{center}
+\end{frame}
+
+\begin{frame}{Relative differences, ages 40--64}
+
+  %\begin{center}
+\includegraphics[height=7.5cm,width=10.8cm]{figs/art2-cumswstand-1}
+   %\end{center}
+\end{frame}
+
+\begin{frame}{Relative differences, ages 65--89}
+
+  %\begin{center}
+\includegraphics[height=7.5cm,width=10.8cm]{figs/art2-cumswold-1}
    %\end{center}

### 2017-10-06 R/plotED.glm.R

@@ -4,7 +4,7 @@ plotED.glm <- function(res,
                        relative = FALSE,
                        ylab = "",
                        main = "",
-                       exclude = 100){
+                       exclude = numeric(0)){
     
     ## 'res' is a list of matrices of cumulative hazards from a piecewise-contant fit
     ## We want the last column of each matrix.
@@ -13,6 +13,8 @@ plotED.glm <- function(res,
 
     ## Assuming 'm == 4', we try:
     ##leg = c("elite", "lowMan", "lowWhiteC", "worker")
+    chs <- !(1:length(res) %in% exclude)
+    res <- res[chs]
     leg <- rownames(res[[1]])
     nper <- length(res)
     k <- NCOL(res[[1]])
@@ -20,7 +22,7 @@ plotED.glm <- function(res,
     labb <- names(res)
 
     out <- 1:nper
-    out <- out[!(out %in% exclude)]
+    ##out <- out[!(out %in% exclude)]
     outp <- matrix(0, ncol = nper, nrow = m)
 
     for (i in 1:nper){
@@ -48,10 +50,12 @@ plotED.glm <- function(res,
     xv <- (as.numeric(substr(names(res), 1, 4)) +
         as.numeric(substr(names(res), 6, 9))) / 2
 
-    plot(xv[out], outp[1, -exclude], type = "b", lty = 1, col = col[1], pch = 1,
+    ##xv <- xv[-exclude]
+    ran <- max(xv) - min(xv)
+    plot(xv[out], outp[1, ], type = "b", lty = 1, col = col[1], pch = 1,
          ylim = c(fr, to), main = main,
          axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
-         xlim = c(min(xv), max(xv) + 33))
+         xlim = c(min(xv), max(xv) + 33 * ran / 200))
     axis(1, at = xv, lab = labb, las = 2)
     axis(2, las = 2)
     box()

### 2017-10-08 R/pchtab.R
check log scale

@@ -1,0 +1,16 @@
+pchtab <- function(formula, data){
+    ## Assume 'events' as name for response, 'exposure' as name for 
+    ## exposure, and 'age' as name for factor time.
+    fit <- glm(formula, data = data, family = poisson)    
+    coeff <- coefficients(fit)
+    on <- substr(names(coeff), 1, 3) %in% c("(In", "age")
+    bas <- coeff[on]
+    f_age <- as.numeric(substr(names(bas)[2], 4, 5)) - 5
+    names(bas)[1] <- paste("age", f_age, sep = "")
+    bas[2:length(bas)] <- bas[2:length(bas)] + bas[1]
+    bas <- exp(bas)
+    cuts <- substr(names(bas), 4, 5)
+    list(levels = c(0, bas), 
+         cuts = as.numeric(cuts))
+    
+}

### 2017-10-08 R/plotED.glm.R

@@ -4,6 +4,7 @@ plotED.glm <- function(res,
                        relative = FALSE,
                        ylab = "",
                        main = "",
+                       prob = TRUE,
                        exclude = numeric(0)){
     
     ## 'res' is a list of matrices of cumulative hazards from a piecewise-contant fit
@@ -29,9 +30,11 @@ plotED.glm <- function(res,
         outp[, i] <- res[[i]][, k]         
     }
     
-    if (logScale){
+    if (!logScale){
+        ## logScale <--> cum. Hazard
+        ## !logScale <--> P(die )
         ##outp <- log(outp)
-        outp <- log(1 - exp(-outp))
+        outp <- 1 - exp(-outp)
         ##outp <- log(outp / (1 - outp))
     }
     if (relative){
@@ -42,7 +45,7 @@ plotED.glm <- function(res,
         ##to <- max(outp[, out], na.rm = TRUE)
         ##fr <- min(outp[, out], na.rm = TRUE)
         fr <- 0
-        to <- 0.6
+        to <- max(outp)
         #cat("to = ", to, ", fr = ", fr, "\n")
     }else{    
         to <- max(outp)

### 2017-10-08 article2.Rmd
improve a figure
fix cape slides and figures

@@
 vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
+vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
 ```
@@
 ##
-cap0 <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom)."
-scap0 <- "Total hazard by social class, ages 40--64."
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom)."
-scap12 <- "Total hazard standardised by social class, ages 40--64."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades."
-scap12log <- "Total hazard by social class, ages 40--64, last years."
+cap0 <- "Probability of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom)."
+scap0 <- "Probability by social class, ages 40--64."
+cap12 <- "Probability of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom)."
+scap12 <- "Probability standardised by social class, ages 40--64."
+cap12log <- "Probability of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades."
+scap12log <- "Probability by social class, ages 40--64, last years."
 
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). Standardised."
-scap22 <- "Total hazard by social class, ages 65--89."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
-scap22log <- "Total hazard by social class, ages 65--89, last years."
+cap22 <- "Probability of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). Standardised."
+scap22 <- "Probability by social class, ages 65--89."
+cap22log <- "Probability of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
+scap22log <- "Probability by social class, ages 65--89, last years."
 
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Women 40-64")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Men 40-64")##, exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Women 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Men 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Women 40-64")##, exclude = 5:6)
+                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.1, 0.55), main = "Women 40-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(65|40)", main = "Men 40-64")##, exclude = 5:6)
+                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.1, 0.55), main = "Men 40-64")##, exclude = 5:6)
 par(oldpar)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Women 65-89")##, exclude = 5:6)
+                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.1, 0.55), main = "Women 65-89")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Men 65-89")##, exclude = 5:6)
+                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.1, 0.55), main = "Men 65-89")##, exclude = 5:6)
 par(oldpar)
@@
 ## Colors for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
+##scol <- c("blue", rep("darkgreen", 2), "red")
 ## Common for all groups:

### 2017-10-08 cape.Rnw
fix cape slides and figures

--- 2017-10-05 cape.Rnw
@@
 vb0Tab$age <- factor(vb0Tab$age)
+vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
 @
@@
 
-\begin{frame}{Women and men, age 40--64}
+\begin{frame}{Married, rural areas, ages 40--64}
 
@@
 source("R/plotED.glm.R")
-scol <- c("blue", "darkgreen", "red", "black")
+scol <- c("blue", "darkgreen", "darkgreen", "red")
 oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Women 40-64", exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "TMR(65|40)",
-                main = "Men 40-64")##, exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Women", ylim = c(0, 0.6))## exclude = 5:6
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Men", ylim = c(0, 0.6))##, exclude = 5:6)
+par(oldpar)
+@
+
+\end{frame}
+
+\begin{frame}{Married, rural areas, ages 40--64, 1976--2013}
+
+<<cumswlate>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Women", exclude = 1:6, ylim = c(0, 0.3))
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Men", exclude = 1:6, ylim = c(0, 0.3))
+par(oldpar)
+@
+
+\end{frame}
+
+\begin{frame}{Married, rural areas, ages 65--89}
+
+%  \begin{center}
+%\includegraphics[height = 7.5cm]{figure/cumsw-1}
+%  \end{center}
+
+<<hazcalw2>>=
+source("R/getPHglm2.R")
+##nper <- 4
+vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) > 64.5, ]
+vb6589$age <- factor(vb6589$age)
+labb <- levels(vb0Tab$period)
+nper <- length(labb)
+resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
+            nper, labb)
+@
+
+<<cumsw2>>=
+source("R/plotED.glm.R")
+scol <- c("blue", "darkgreen", "darkgreen", "red")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Women", ylim = c(0.4, 1.0))## exclude = 5:6
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Men", ylim = c(0.4, 1.0))##, exclude = 5:6)
+par(oldpar)
+@
+
+\end{frame}
+
+\begin{frame}{Married, rural areas, ages 65--89, 1976--2013}
+
+<<cumswlate2>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Women", exclude = 1:6, ylim = c(0.4, 1.0))
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Men", exclude = 1:6, ylim = c(0.4, 1.0))
 par(oldpar)
@@
 \end{frame}
+
+% \begin{frame}{Relative differences, ages 40--64}
+% 
+%   %\begin{center}
+% \includegraphics[height=7.5cm,width=10.8cm]{figs/art2-cumswstand-1}
+%    %\end{center}
+% \end{frame}
+% 
+% \begin{frame}{Relative differences, ages 65--89}
+% 
+%   %\begin{center}
+% \includegraphics[height=7.5cm,width=10.8cm]{figs/art2-cumswold-1}
+%    %\end{center}
+% \end{frame}
+%
 

### 2017-10-08 R/plotED.glm.R

@@ -4,6 +4,7 @@ plotED.glm <- function(res,
                        relative = FALSE,
                        ylab = "",
                        main = "",
+                       ylim = c(0, 1),
                        prob = TRUE,
                        exclude = numeric(0)){
     
@@ -56,7 +57,7 @@ plotED.glm <- function(res,
     ##xv <- xv[-exclude]
     ran <- max(xv) - min(xv)
     plot(xv[out], outp[1, ], type = "b", lty = 1, col = col[1], pch = 1,
-         ylim = c(fr, to), main = main,
+         ylim = ylim, main = main,
          axes = FALSE, ylab = ylab, cex.axis = 0.4, xlab = "", 
          xlim = c(min(xv), max(xv) + 33 * ran / 200))
     axis(1, at = xv, lab = labb, las = 2)

### 2017-10-09 cape.Rnw
Changing title

@@
 
-\title{Increasing the mortality gap? Social inequality in mortality among 
-elderly and adults in Northern  Sweden 1851--2013\footnote{Session "Health, 
-morbidity, and mortality" at the Nordic Demographic Symposium 14--16 June 2017, 
-Turku, Finland}
+\title{Life course and long-term perspectives of social inequality in mortality among 
+elderly and adults in Northern  Sweden 1801--2013\footnote{Session "Social inequalities in health at older ages: life course perspectives" at the XXVIII International Population Conference 29 October--4 November 2017, Cape Town, South Africa}
 }

### 2017-10-09 cape.Rnw
Revising slides

@@
 \item *cause of death (cardiovascular, cancer)
-\item *education 1990--2013)
-\item *income 1986-2005)
+%\item *education 1990--2013)
+%\item *income 1986-2005)
 
@@
 \item A consistent mortality difference according to social class appears surprisingly late in our regions and has no clear connection to the development of income inequality.
+\item The association with social class is stronger in the working-age population compared with the elderly, retired population.
 \item However, there is a strong gender difference in the development of social inequality.
@@
 \end{itemize}
-\item The \emp{Linnaeus} data base (1960--2013)
+\item The \emp{Linnaeus} data base (1976--2013)
   \begin{itemize}
@@
 \item Population registers ({\sc Lisa}, 1986--2013)
-  \begin{itemize}
-    \item \emp{Income}
-    \item \emp{Education}
-  \end{itemize}
+%  \begin{itemize}
+%    \item \emp{Income}
+%    \item \emp{Education}
+%  \end{itemize}
  \item Death information from \emp{National Board of Health and Welfare}
    \begin{itemize}
-\item Death dates from 1 January 1961 to 31 December 2013.
+\item Death dates from 1 January 1976 to 31 December 2013.
   \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
@@
 \item[Parish] Urban/Rural
-\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
-\item[*Income] 1986--2005, grouped by quartiles.
-\item[*Education] 1990--2013, three categories.
-\item[Period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
+\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1976--2013
+%\item[*Income] 1986--2005, grouped by quartiles.
+%\item[*Education] 1990--2013, three categories.
+\item[Period]  1851--1875, \ldots,  2008--2013 (13 periods). 
 \item[Age group] Separate analyses for working active (40-65) vs retired (65-89) population
@@
 \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
-% \item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
+\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
 % \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
@@
 
+\begin{frame}
+
+Thank you for your attention
+
+\end{frame}
+
 %\begin{frame}{Conclusion}

### 2017-10-09 article2.Rmd
fix some figures

@@
 yy <- round(prop.table(xxx, 1) * 100, 1)
-barplot(t(yy)[3:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
+barplot(t(yy)[3:1, ], col = c("red", "blue", "green"), las = 2, ylab = "Per cent")
 ```

### 2017-10-09 cape.Rnw
fix some figures

@@
 
+\begin{frame}{The sampling frame}
+
+<<lexis2>>=
+@
+
+\end{frame}
+
+\begin{frame}{Exposure by social class, women}
+
+\begin{center}
+\includegraphics[height = 5cm]{figs/art2-hiscskumw-1}
+
+\small From the top: Elite, middle, workers, missing.
+\end{center}
+\end{frame}
+
+\begin{frame}{Exposure by social class, men}
+
+\begin{center}
+\includegraphics[height = 5cm]{figs/art2-hiscskumm-1}
+
+\small From the top: Elite, middle, workers, missing.
+\end{center}
+\end{frame}
+
+\begin{frame}{Exposure by marital status}
+
+\begin{center}
+\includegraphics[height = 5cm]{figs/art2-civstskum-1.pdf}
+
+\small From the top: Never married, married, previously married.
+\end{center}
+
+\end{frame}
+
+\begin{frame}[fragile]{Cohort mortality, men 40--65}
+
+<<cohortapproach>>=
+vv <- vb0Tab[vb0Tab$cohort %in% seq(1751, 1951, by = 10) &
+               vb0Tab$age %in% seq(40, 60, by = 5), ]
+vv$age <- factor(vv$age)
+vv$cohort <- factor(vv$cohort)
+cohs <- levels(vv$cohort)
+vv$age <- factor(vv$age)
+n <- length(cohs)
+coe <- matrix(0, nrow = 4, ncol = n)
+for (i in 1:n){
+   fit <- glm(event ~ offset(log(exposure)) + hisclass + age + urban + civst, 
+              data = vv[vv$cohort == cohs[i] & vv$sex == "male", ],
+              family = poisson)
+   x <- coefficients(fit)[1:4]
+   x[2:4] <- x[1] + x[2:4]
+   coe[, i] <-  exp(x) / sum(exp(x))
+}
+x <- as.numeric(as.character(cohs))
+par(lwd = 2)
+plot(x[-n], coe[1, -n], type = "b", col = scol[1], ylim = c(0, 0.7), xlab = "Birth cohort", ylab = "Standardized rel. risks")
+lines(x[-n], coe[2, -n], type = "b", col = scol[2], pch = 2, lty = 2)
+lines(x[-n], coe[3, -n], type = "b", col = scol[3], pch = 3, lty = 3)
+lines(x[-n], coe[4, -n], type = "b", col = scol[4], pch = 4, lty = 4)
+legend("topleft", legend = c("elite", "middle", "worker", "NA"), pch = 1:4, col = scol, lty = 1:4)
+@
+
+\end{frame}
+
 \end{document}

### 2017-10-12 cape.Rnw
add slide with deaths by period and sex

@@
 
+\begin{frame}{Crude statistics}
+
+<<crude>>=
+dat <- with(vb0Tab, aggregate(list(event = event, exposure = exposure), 
+                              by = list(period = period, sex = sex), 
+                              FUN = sum))
+males <- dat[dat$sex == "male", c("period", "event", "exposure")]
+females <- dat[dat$sex == "female", c("period", "event", "exposure")]
+oldpar <- par(mfrow = c(1, 2))
+x <- c(1812.5, 1837.5, 1862.5, 1888.5, 1912.5, 1937.5, 1987.5, 1982.5, 1987.5, 
+       1992.5, 1997.5, 2004.5, 2010.5)
+plot(x, males$event, type = "b", col = "blue", main = "Number of deaths by period",
+     ylim = c(0, 7500), ylab = "No. of deaths", xlab = "Time period", lty = 1, axes = FALSE)
+axis(1, at = c(1801, 1826, 1851, 1876, 1901, 1926, 1951, 1976, 2014), las = 2)
+axis(2, las = 1)
+box()
+lines(x, females$event, type = "b", col = "red", lty = 2)
+legend("bottomright", legend = c("Males", "Females"), lty = c(1, 2), col = c("blue", "red"))
+plot(x, males$event / males$exposure * 1000, type = "b", col = "blue", 
+     main = "Crude death rates, ages 40-89", ylim = c(0, 30), xlim = c(1800, 2014),
+     xlab = "Time period", ylab = "Deaths per 1000", axes = FALSE)
+axis(1, at = c(1801, 1826, 1851, 1876, 1901, 1926, 1951, 1976, 2014), las = 2)
+axis(2, las = 1)
+box()
+lines(x, females$event / females$exposure * 1000, type = "b", col = "red")
+legend("bottomright", legend = c("Males", "Females"), lty = c(1, 2), col = c("blue", "red"))
+@
+
+\end{frame}
 \end{document}

### 2017-10-17 cape.Rnw
A couple of corrections
Moving slides

@@
 
-\begin{frame}{Married, rural areas, ages 40--64}
+\begin{frame}{Probabilities of dying, ages 40--64, 1801-2013}
 
@@
 
-\begin{frame}{Married, rural areas, ages 40--64, 1976--2013}
-
-<<cumswlate>>=
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
-                main = "Women", exclude = 1:6, ylim = c(0, 0.3))
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
-                main = "Men", exclude = 1:6, ylim = c(0, 0.3))
-par(oldpar)
-@
-
-\end{frame}
-
-\begin{frame}{Married, rural areas, ages 65--89}
-
-%  \begin{center}
-%\includegraphics[height = 7.5cm]{figure/cumsw-1}
-%  \end{center}
-
-<<hazcalw2>>=
-source("R/getPHglm2.R")
-##nper <- 4
-vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) > 64.5, ]
-vb6589$age <- factor(vb6589$age)
-labb <- levels(vb0Tab$period)
-nper <- length(labb)
-resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
-            nper, labb)
-@
-
-<<cumsw2>>=
-source("R/plotED.glm.R")
-scol <- c("blue", "darkgreen", "darkgreen", "red")
-oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
-                main = "Women", ylim = c(0.4, 1.0))## exclude = 5:6
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
-                main = "Men", ylim = c(0.4, 1.0))##, exclude = 5:6)
-par(oldpar)
-@
-
-\end{frame}
-
-\begin{frame}{Married, rural areas, ages 65--89, 1976--2013}
-
-<<cumswlate2>>=
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
-                main = "Women", exclude = 1:6, ylim = c(0.4, 1.0))
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
-                main = "Men", exclude = 1:6, ylim = c(0.4, 1.0))
-par(oldpar)
-@
-
-\end{frame}
-
-\begin{frame}{Relative differences, ages 40--64 and 65--89}
+\begin{frame}{Relative differences in probabilities of dying, ages 40--64 and 65--89}
 
@@
 \item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
-\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
+\item \emp{Small} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
 % \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
@@
 
-Thank you for your attention
+\begin{center}
+
+\Large
+
+Thank you for your attention!
+
+\end{center}
 
@@
 %\end{frame}
+
+\begin{frame}{Probabilities of dying, ages 40--64, 1976--2013}
+
+<<cumswlate>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Women", exclude = 1:6, ylim = c(0, 0.3))
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+                main = "Men", exclude = 1:6, ylim = c(0, 0.3))
+par(oldpar)
+@
+
+\end{frame}
+
+\begin{frame}{Probabilities of dying, ages 65--89, 1801-2013}
+
+%  \begin{center}
+%\includegraphics[height = 7.5cm]{figure/cumsw-1}
+%  \end{center}
+
+<<hazcalw2>>=
+source("R/getPHglm2.R")
+##nper <- 4
+vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) > 64.5, ]
+vb6589$age <- factor(vb6589$age)
+labb <- levels(vb0Tab$period)
+nper <- length(labb)
+resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
+            nper, labb)
+@
+
+<<cumsw2>>=
+source("R/plotED.glm.R")
+scol <- c("blue", "darkgreen", "darkgreen", "red")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Women", ylim = c(0.4, 1.0))## exclude = 5:6
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Men", ylim = c(0.4, 1.0))##, exclude = 5:6)
+par(oldpar)
+@
+
+\end{frame}
+
+\begin{frame}{Probabilities of dying, ages 65--89, 1976--2013}
+
+<<cumswlate2>>=
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1), cex = 0.7, lwd = 1.5)
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Women", exclude = 1:6, ylim = c(0.4, 1.0))
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 90 | alive at 65)",
+                main = "Men", exclude = 1:6, ylim = c(0.4, 1.0))
+par(oldpar)
+@
+
+\end{frame}
 

### 2017-10-19 article3.Rmd
first commit (copy of 'article2.Rmd')

--- 2017-10-09 article2.Rmd
@@
 ---
-title: Life course and long-term perspectives of social inequality in mortality among
-  elderly and adults in Northern Sweden      1801–2013
+title: Life course and long-term perspectives of social inequality in mortality among elderly and adults in Northern Sweden 1801--2013
 author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art2-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 library(knitr)

### 2017-10-24 cape.Rnw
Correcting affiliation
Adding and correcting slides

@@
 \author{Sören Edvinsson and Göran Broström \\
-Centre for Ageing and Demographic Research \\
+Centre for Demographic and Ageing Research \\
 Umeå University, Umeå, Sweden}
@@
 
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]{Two time perspectives}
+
+\begin{itemize}
+\item Development over \emp{time}.
+  \begin{itemize}
+\item Explicit or implicit assumption of an \epm{eternal} social gradient in history in many theories (Omran, Link and Phelan, Marmot, McKeown), some implying even larger in history.
+\item Antonovsky – divergence followed by convergence
+\end{itemize}
+\item Development over the \emp/life course}
+  \begin{itemize}
+\item Convergence, divergence or maintenance.
+\item Antonovsky – large differences in adulthood, convergence in old age
 \end{itemize}
@@
 \begin{itemize}
-\item \emp{Swedish church book} records (1851--1950).
+\item \emp{Swedish church book} records (1801--1950).
   \begin{itemize}
@@
 \item Distinct \emp{gender} differences in the development.
-\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
-\item \emp{Small} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
+\item \emp{Large} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
 % \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
@@
 
-\Large
+\Huge
 

### 2017-10-24 cape.Rnw
fix small errors

@@
   \begin{itemize}
-\item Explicit or implicit assumption of an \epm{eternal} social gradient in history in many theories (Omran, Link and Phelan, Marmot, McKeown), some implying even larger in history.
+\item Explicit or implicit assumption of an \emp{eternal} social gradient in history in many theories (Omran, Link and Phelan, Marmot, McKeown), some implying even larger in history.
 \item Antonovsky – divergence followed by convergence
 \end{itemize}
-\item Development over the \emp/life course}
+\item Development over the \emp{life course}
   \begin{itemize}
@@
 \item Antonovsky – large differences in adulthood, convergence in old age
+\end{itemize}
 \end{itemize}
@@
 
-Thank you for your attention!
+\vspace{1\baselineskip}
+\textcolor{red}{\fbox{Thank you for your attention!}}
 
+\includegraphics[height=5cm]{myfigs/Top_incomes}
 \end{center}

### 2017-10-24 apc.Rmd
first commit, beginning of apc model

---
title: "apc.Rnw"
author: "Göran Broström"
date: "October 24, 2017"
output: html_document
---

## Read data

```{r setup2, include=FALSE}
library(knitr)
opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
```

```{r readalldata}
source("R/get.file.R")
skel <- get.file("skel")
##
skume <- get.file("skume")
##
foli <- get.file("foli")
skel$urban <- FALSE
vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
```

## Age-period-cohort data

```{r datai}
library(Epi)
dat <- vb0Tab[vb0Tab$sex == "male", c("age", "period", "cohort", "event", "exposure")]
dat$age <- as.numeric(dat$age)
dat$A <- dat$age + 2.5
dat$Per <- as.numeric(dat$period)
dat$P <- 25 * dat$Per + 1787.5
dat$P[dat$period == "1976-1980"] <- 1977.5
dat$P[dat$period == "1981-1985"] <- 1982.5
dat$P[dat$period == "1986-1990"] <- 1987.5
dat$P[dat$period == "1991-1995"] <- 1992.5
dat$P[dat$period == "1996-2001"] <- 1999
dat$P[dat$period == "2002-2007"] <- 2005
dat$P[dat$period == "2008-2013"] <- 2011
dat$D <-dat$event
dat$Y <- dat$exposure
summary(dat)
```

## APC model fit

```{r}
ex.H <- apc.fit(dat, npar = 7, model = "ns", dr.extr = "Holford",  parm = "ACP", scale = 100)
ex.W <- apc.fit(dat, npar = 7, model = "ns", dr.extr = "weighted", parm = "ACP", scale = 100)
ex.S <- apc.fit(dat, npar = 7, model = "ns", parm = "AC-P", scale = 100)
## Show the estimated drifts
ex.H[["Drift"]]
ex.W[["Drift"]]
ex.S[["Drift"]]
```

Then we plot the results

```{r}
# Plot the effects
fp <- apc.plot( ex.H )
apc.lines( ex.W, frame.par=fp, col="red" )
apc.lines( ex.S, frame.par=fp, col="blue" )
```

### 2017-10-24 article3.Rmd

@@
 ---
-title: Life course and long-term perspectives of social inequality in mortality among elderly and adults in Northern Sweden 1801--2013
-author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
+title: Is high social class always beneficial for survival? A study of adults and elderly in Northern Sweden 1851--2013
+author: "Sören Edvinsson and Göran Broström"
 date: '`r Sys.Date()`'

### 2017-10-24 cape.Rnw
Corrections

@@
   \begin{itemize}
-\item Explicit or implicit assumption of an \emp{eternal} social gradient in history in many theories (Omran, Link and Phelan, Marmot, McKeown), some implying even larger in history.
+\item Explicit or implicit assumption of an \emp{eternal} social gradient (Omran, Link and Phelan, Marmot, McKeown), some implying even larger differences in history.
 \item Antonovsky – divergence followed by convergence
@@
 \item Convergence, divergence or maintenance.
-\item Antonovsky – large differences in adulthood, convergence in old age
+\item Antonovsky – \emp{Convergence} from adulthood to old age
 \end{itemize}
@@
 
-Remote area in northern Sweden, dominated by agriculture until well into the 20th century.
+Remote areas in northern Sweden, dominated by agriculture until well into the 20th century.
 
@@
 
-\item[Sex]
-\item[Marital status]
+\item[Sex] Male/female
+\item[Marital status] Unmarried/married/previously married
 \item[Birth date] By quarter (15 Feb, May, Aug, Nov)
@@
 \item[Parish] Urban/Rural
-\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1976--2013
+\item[Social class] Elite, middle, worker. Household definition 1801--1950. Individual definition 1976--2013
 %\item[*Income] 1986--2005, grouped by quartiles.
 %\item[*Education] 1990--2013, three categories.
-\item[Period]  1851--1875, \ldots,  2008--2013 (13 periods). 
+\item[Period]  1801--1825, \ldots,  2008--2013 (13 periods). 
 \item[Age group] Separate analyses for working active (40-65) vs retired (65-89) population
@@
   \item \emp{\tt sex}
-  \item \emp{\tt period}: 1851--75, 1876--1900, 1901--10, etc.
+  \item \emp{\tt period}: 1801--25, 1826--1850, ... 1976--1980, etc.
   \item \emp{\tt class}
@@
 \item Distinct \emp{gender} differences in the development.
-\item \emp{Large} differences between those in working age compared to the retired population, indications of \emp{status convergence}.
+\item \emp{Large} differences between those in working age and the retired population, indications of \emp{status convergence}.
 % \item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
-\item In the long run, no clear connection in the development of income inequality and social differences in mortality.
-\item This suggests an important role of gender and class related expectations on health behaviour
+\item No clear connection in the development of income inequality and social differences in mortality.
+\item The results suggest an important role of gender and class related expectations on health behaviour
 \end{itemize}
@@
 \begin{itemize}
-\item The \emp{development} of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
+\item The role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
 \item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).

### 2017-10-24 article3.Rmd
Change of title

@@
 ---
-title: Is high social class always beneficial for survival? A study of adults and elderly in Northern Sweden 1851--2013
+title: Is high social class always beneficial for survival? A study of Northern Sweden 1801--2013
 author: "Sören Edvinsson and Göran Broström"

### 2017-10-25 article3.Rmd
First revisions of article

@@
 
-On the basis of the results we present, we argue that high social class is not necessarily always favourable for survival.
+On the basis of the results, we argue that high social class is not necessarily always favourable for survival.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 One of the central aspects of survival is social class and access to economic
-and other resources. Even in present-day welfare societies, social position is a
+and other resources. In present-day welfare societies, social position is a
 strong determinant when it comes to health and mortality and the impact even
-seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. It has been suggested that
+seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph95 suggest that
 "... social conditions have been, are and will continue to be irreducible
 determinants of health outcomes and therefore deserves their appellation as
-'fundamental causes' of disease and death" [@liph95].
-The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in this theory [@liph98].
-For a long time, the general view has been that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
+'fundamental causes' of disease and death".
+The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph98].
+The reigning view has been that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
 This is a
-reasonable assumption since these societies were in most cases characterised by
-very large socioeconomic differences. Knowing that access to
-resources provides advantages in all aspects of life, the health advantage of higher
+reasonable assumption since these societies were often characterised by
+enourmous socioeconomic differences. Knowing that access to different kinds of resources
+provide advantages in all aspects of life, the health advantage of higher
 classes ought to be obvious.
-
-@ant67 suggests that social inequality in mortality has passed through different historical phases. According to him, differences were comparatively small during the
-pre-transitional phase. This period was characterised by space being a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality
-declined and wealthy groups used their economic and human resources to gain better health. Finally,
-mortality differences decreased again resulting in marginal levels of inequality in modern low-mortality societies when instead
-health-related behaviour became the decisive determinant for health and survival.
-@omran82 comes to a similar conclusion in his theory of the epidemiologic transition. He states in the third proposition of the theory that even if the class differentials in mortality were
-maintained during the transition, the decline set in earlier and was faster
-among privileged groups.
 
@@
 20th century, industrialisation took place. This also led to a population increase both in the town
-and in the rural parts, resulting in a much more diversified economy.
+and in the rural parts, resulting in a more diversified economy.
 
@@
 the fertility transition was late [@coa86]. The Skellefteå population
-size as defined in our data sets (all ages) was 6142 on January 1, 1801, 17355 on January 1, 1851, 43212 on January 1, 1901, 62136 on December 31, 1950 and 76723 at the end of the
-20th century.
+size as defined in our data sets (all ages) was 6142 on January 1, 1801, 43212 on January 1, 1901, and 76723 at the end of the 20th century.
 
@@
 that of Skellefteå. Umeå town had a small population but was substantially
-larger than Skellefteå town during almost the whole studied period. It was
-the administrative centre in the county of Västerbotten. Schools and
-military regiments were placed here and the economy was more diversified
-compared to Skellefteå. Agriculture dominated the rural part, but there
-were also some foundries as well as industries for example in forestry and
-small-scale production. The population size as defined in our data sets (all ages) was
-19138 on January 1, 1901, 33393 on December 31, 1950 and 103970 when
-the 20th century ended.
+larger than Skellefteå town during almost the whole studied period and was
+the administrative, educational and military centre in the county of Västerbotten.
+During the later part of the 20th century, the establishment of a university in Umeå led to a rapid population increase.
+Agriculture dominated the rural part, but there were also some foundries as well as forestry industries.
+Consequently the economy was more diversified compared to Skellefteå.The population size as defined in our data sets (all ages) was 19138 on January 1, 1901 and 103970 when the 20th century ended.
 
@@
 1950 makes it possible to investigate the long-term development of social
-inequality in mortality in a part of Sweden. The data for the present study come from two large
-population databases at the DDB, whom provide us with micro-data for the Skellefteå and Umeå regions in nortehrn Sweden.
+inequality in mortality in this part of Sweden. The data for the present study come from two large
+population databases at the DDB, whom provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
 Our dataset constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
-The early period until 1950 is covered by the database Poplink, the digitisation of historical parish registers for
-the two regions [@wee16]. Poplink is based on linked
-parish records, allowing us to reconstruct life biographies on people as
-long as they remained in the region. The records are linked within but not
+The period 1801--1950 is covered by the database Poplink [@wee16].
+Poplink is based on linked parish records, allowing us to reconstruct life biographies on people as long as they remained in the region.
+The records are linked within but not
 between the regions. Data from Poplink are accessed for the period 1801--1950.
@@
 registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. The study period from 1976 to 2000 is constructed from
+University. Due to data issues, we choose to use Linnaeus data from 1976. The study period from 1976 to 2000 is constructed from
 censuses every fifth year 1975--1990, with additional information on deaths
@@
 Individuals are anonymised and the two databases are not
-linked, thus they are treated separately.
-This prevent us from following individuals between the two databases throughout their lives.
+linked, thus they are treated as separate units.
+This prevents us from following individuals between the two databases throughout their lives.
 It also makes it impossible to add information on individuals in the Linnaeus database from what we potentially could extract from Poplink, for example family
@@
 the Linnaeus data makes it necessary to apply different approaches when it
-comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of the presence of all individuals in our data, allowing us
+comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of presence, allowing us
 to have full and continuous control of the \emph{de jure} population. This is
@@
 Mortality differences are analysed according to social class, based on a modified form 
-of the classification scheme Hisclass [@lema11]. The only
-type of socio-economic information available both in the Linnaeus and the Poplink databases
-is occupations. Thus, analysing according to social class is apart from
-being a scientific choice, also a practical one. Using occupational data is
-the only and obvious alternative for making a longitudinal study of
-socio-economic mortality differences with consistent classification
-schemes.
+of the classification scheme Hisclass [@lema11]. The only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification
+schemes in the two databases, is occupations. Thus, analysing according to social class is apart from
+being a scientific choice, also a practical one.
 Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, that is, from age 65 until death or last observation.
@@
 The availability of information on occupations varies over time and between
-sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
-Linnaeus data, HISCO is recoded into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
+sources.
+Occupations in Poplink are coded according to the original DDB coding system that then has been adapted to Hisclass.
+For the Linnaeus data, HISCO is recoded into Hisclass.
+The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
 Yrkesklassificering).
@@
 towards defining the social position. Poplink usually only provides occupation for the head of household, thus underestimating female working-force participation as well as that of
-adult children residing with their parents [@vik10]. For the Poplink period (until 1950), we have chosen to categorise wives according to position of the head of household, usually
+adult children residing with their parents [@vik10]. For the Poplink period, we have chosen to categorise wives according to position of the head of household, usually
 the husband/father, assuming that the family shares the same socio-economic
@@
 different periods are not completely comparable.
-Notice however that we can expect a strong social homogamy within the married couple, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].
+Notice however that there was a strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].
 
@@
 categories, based on whether the work is manual or non–manual, skill level,
-supervision and sector. Implicitly it reflects large differences in what is
-in the forefront of other categorisations, as for example access to
+supervision and sector. Implicitly it reflects large differences in access to
 economic and other resources, status, social power et cetera.
@@
 according to own occupation, the consequence of many women being outside
-the workforce is that the proportion missing becomes very high. For the later periods,
+the workforce is that the proportion missing increases. For the later periods,
 the results are fairly complete.
@@
 
-Marital status is treated as a time-dependent variable in the data
+Marital status, see Figure \@ref(fig:civstskum), is treated as a time-dependent variable in the data
 file. The statuses used are unmarried, married and dissolved marriages
@@
 set to unmarried. Notice that it has not been possible to identify cohabitation.
-
-Regarding the distribution of marital status over time, see
-Figure \@ref(fig:civstskum).
 

### 2017-10-26 article3.Rmd
Further revisions

@@
 population size. The modern Swedish welfare state began to develop. For the 
-period 1976 onwards, there are separate analyses on decades except for the two 
-last periods. Sweden had developed into a wealthy
-welfare society with low income inequality,
+period 1976 onwards, there are separate analyses on 5- or 6-year periods. Sweden had developed into a wealthy welfare society with low income inequality,
 particularly until the 1990s. A financial crisis struck Sweden in the early
@@
 and graphics [@cran], especially using the package **eha**
-[@gb12;@eha]. In the the analysis, data are divided into two age groups,
-40--64 and 65--89 years of age, the first age group representing people that
-were still mainly in the workforce, and the second those that mainly were
-retired. The model controls for marital status and whether the individual resides 
-in the urban or rural parts of the regions. Each period are analysed separately, 
-but we then combine results
-in order to show the development of the social patterning of mortality
-during the studied period.
-
-The important explanatory variable, social class (**hisclass**), is included in the
+[@gb12;@eha]. In the analysis, data are divided into two age groups,
+40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period are analysed separately, 
+but we then combine results in order to show the development of the social patterning of mortality during the studied period.
+
+The important explanatory variable, social class, is included in the
 proportional hazards models as a *stratification* variable. The main
-reason for this choice is that the effect of social class (**hisclass**) on survival is
+reason for this choice is that the effect of social class on survival is
 non-proportional, that is, it varies with age. With the reduced data set
 and Poisson regression, this is simply achieved by introducing an
-interaction term between **age** and social class (**hisclass**). Then, the main
+interaction term between **age** and social class. Then, the main
 results are presented graphically.  The comments on the results are based on
-the figures and focuses on the central variable in our study, **hisclass**.
+the figures and focuses on the central variable in our study, social class.
 
@@
 The models are fitted separately for sex and each period, from  1801 to
-2013. We stratify on **hisclass** and include the covariates **urban** (TRUE or FALSE) and
+2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
 **civst** (marital status). The cumulative
@@
 From Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with **sex**, **period**, and social class (**hisclass**).
+The most interesting feature is the three-way interaction with **sex**, **period**, and social class.
 
@@
 than men. This is a well-known phenomenon in Swedish
-demographic history. Male mortality has almost always been higher than
-female in all age groups [@wil99; @suwi07]. We also observe a
+demographic history [@wil99; @suwi07]. We also observe a
 substantial decline in mortality for both sexes, resulting in smaller
@@
 
-Notice that we have not analysed the 1950s due to the fact that we only
-have data for some parishes at that time. The
-changing economy have some consequences for the analysis, particularly when
+The changing economy have some consequences for the analysis, particularly when
 it comes to the agricultural sector. While farmers were common occupations
@@
 regions in northern Sweden and for other periods [@edli11; @edbr12]. The
-highest social class usually have comparatively low mortality, while
+highest social class usually have comparatively low mortality (except for the first period when however the statistical power was weak), while
 working class women have high, but almost constantly about the same level as the middle category.
@@
 A completely different pattern appears among men, being the most surprising finding in
-our analyses. The groups we would expect to have the lowest mortality, if
+our analyses. The groups we expect to have the lowest mortality, if
 we assume that access to vital resources and belonging to a high social class
@@
 level as the middle class.
-The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we still find better survival among workers compared to the elite.
+The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we find better survival among workers compared to the elite.
 
 The relative positions change radically towards
-the end the 20th century. From the 1980s, there is a clear advantage when
-it comes to survival for higher managers and professionals. The relative
-differences are however more compressed than for the period until 1950. The
+the end of the 20th century. From the 1980s, there is a clear advantage when
+it comes to survival for teh elite. The relative
+differences are however more compressed than for the periods until 1950. The
 expected social gradient in mortality has developed. This is also what we
@@
 1851--1950, men having about three times higher risks than women in the
-same class (basically in the same households). In the other classes, the difference was much smaller. It was men in high position that experienced the largest risks.
+same class (basically living in the same households). In the other classes, the difference was much smaller. It was men in high position that experienced the largest risks.
 
@@
 expected, lower than among men. We find a clear decline in mortality, but
-the fastest decline did not set in definitely until sometimes after the
+the fastest decline did not set in definitely until some time after the
 Second World War, that is, later than for the age
@@
 In this and the following section, the focus is on analyses of mortality from the two
-dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality. We restrict the analyses to the period 1901 onwards. We start
-with cardiovascular diseases. Many studies have shown that socio-economic
-conditions are strongly associated with mortality in this disease
-group. Belonging to working class, having less education and low income
+dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality.
+We restrict the analyses to the period 1901 onwards.
+Many studies have shown that socio-economic conditions are strongly associated with mortality in cardiovascular diseases.
+Belonging to working class, having less education and low income
 increases mortality risks. We would therefore expect to find a clear
@@
 A general observation is that the results on causes of death and
-its development can be difficult to interpret, partly because the number of
+its development are less clear, partly because the number of
 cases are fewer but also that the quality and coverage of reported death
@@
 the same time unknown cause and death in old age decreased from 28 percent
-to 18 per cent. It is quite reasonable to suspect that many cardiovascular
+to 18 per cent. It is reasonable to suspect that many cardiovascular
 deaths were disguised in unspecific categories and that the increase is partly
@@
 also observe increasing social differences from the first half of the
-century to the last decades. A new pattern has developed from 1980
+century to the last decades. A new pattern has developed from 1981
 onwards. Now working class women have substantially higher cardiovascular
@@
 
-For men, we find a different pattern, however it being to a large
-extent consistent with what we observed concerning all-cause
-mortality. The elite group had much higher cardiovascular mortality up until
-the middle of the 20th century, but instead have the lowest mortality from
+For men, we find a different pattern, however being basically consistent with what we observed concerning all-cause mortality.
+The elite group had much higher cardiovascular mortality up until the middle of the 20th century, but instead have the lowest mortality from
 1980 onwards. Working class men had a fairly low cardiovascular mortality,
@@
 
-# A cohort approach, all causes of death, men 40--65
-
-```{r cohortapproach, fig.cap = "Cohort mortality, men aged 40-65."}
-vv <- vb0Tab[vb0Tab$cohort %in% seq(1751, 1951, by = 10) &
-               vb0Tab$age %in% seq(40, 60, by = 5), ]
-vv$age <- factor(vv$age)
-vv$cohort <- factor(vv$cohort)
-cohs <- levels(vv$cohort)
-vv$age <- factor(vv$age)
-n <- length(cohs)
-coe <- matrix(0, nrow = 4, ncol = n)
-for (i in 1:n){
-   fit <- glm(event ~ offset(log(exposure)) + hisclass + age + urban + civst, 
-              data = vv[vv$cohort == cohs[i] & vv$sex == "male", ],
-              family = poisson)
-   x <- coefficients(fit)[1:4]
-   x[2:4] <- x[1] + x[2:4]
-   coe[, i] <-  exp(x) / sum(exp(x))
-}
-x <- as.numeric(as.character(cohs))
-plot(x[-n], coe[1, -n], type = "b", col = scol[1], ylim = c(0, 0.7), xlab = "Birth cohort", ylab = "Standardized rel. risks")
-lines(x[-n], coe[2, -n], type = "b", col = scol[2])
-lines(x[-n], coe[3, -n], type = "b", col = scol[3])
-lines(x[-n], coe[4, -n], type = "b", col = scol[4])
-text(1775, 0.65, "elite", col = scol[1])
-text(1775, 0.55, "middle", col = scol[2])
-text(1825, 0.65, "worker", col = scol[3])
-text(1825, 0.55, "NA", col = scol[4])
-```
-
-In Figure \@ref(fig:cohortapproach) a crude cohort analysis (ten year birth cohorts) for men in working age is exposed (total mortality), i.e.
-the group where we found the largest shift in socio-economic pattern in mortality.
-It is "crude" in the sense that proportional hazards are assumed without checking it (in contrast to the period analyses).
-More work needs to be put into the cohort analysis.
-However, this crude result is well in line with what we have seen in the period analyses.
-Mortality is highest in the elite for almost all cohorts until the late 19th century.
-Only from the cohort born in the 20th century do we find disappearing differences, ending with a consistent hierarchical social pattern for the cohorts born a couple of decades into that century.
-
 # Discussion
@@
 The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
-Some of the limitations have been addressed in the presentation of the variables: The fact that social class was defined according to household until 1950 but individually defined from 1976 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
+Some of the limitations have been addressed in the presentation of the variables: the fact that social class was defined according to household until 1950 but individually defined from 1976 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
 These differences are however not of the magnitude that the main conclusions are affected.
-Most women would end up in the same category as their husbands and the migrations between census dates or census date and death were not frequent.
+Most women end up in the same category as their husbands and the migrations between census dates or census date and death were not frequent.
 
@@
 perspective of social classes. However, there are
-alternative ways to analyse social inequality, and we cannot exclude that
-such analyses give other results or at least would modify the conclusions suggested here.
+alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude that such analyses give other results or modify the conclusions suggested here.
 In many modern studies categorisations based on income, wealth or education are common for studies of inequality.
-They represent different dimensions of inequality.
-It would be interesting to compare the results from this study with alternative measures, but reliable
-data on income and education are unfortunately rarely available in historical population
-databases.
-
-How does the results correspond to research on other health measures?
-In a study of southern Sweden, @oberg14 finds large social differences in body height, a measure that is assumed to sum up health during the life span.
-This does not necessarily contradict what we find here.
-Height is mainly determined by conditions in early childhood, and are not reflecting inequalities in adulthood.
-Earlier studies in Sweden and in the areas studied here [@bredro02; @bube01; @mol17; @edv04] indicate that social disadvantage of being in lower social position were strong in childhood.
+It would be interesting to compare the results from this study with alternative measures, but reliable data on income and education are unfortunately rarely available in historical population databases stretching back as far as in this article.
 
 An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not directly address the hypothesis raised by @wipi09, our results are of interest for this question as well.
-The early 20th century, in Sweden as well as in other countries, was characterised by very large economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
-After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, that is, at a time when a more expected pattern of a social hierarchy in survival appeared.
+Even though our study do not directly address the hypothesis raised by @wipi09, our results are of relevance for this question as well.
+The early 20th century, in Sweden as well as in other countries, was characterised by enourmous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
+After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
+at a time when the expected pattern of a social hierarchy in survival appeared.
 The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.

### 2017-10-26 article3.Rmd
Spelling correction

@@
 reasonable assumption since these societies were often characterised by
-enourmous socioeconomic differences. Knowing that access to different kinds of resources
+enormous socioeconomic differences. Knowing that access to different kinds of resources
 provide advantages in all aspects of life, the health advantage of higher
@@
 Even though our study do not directly address the hypothesis raised by @wipi09, our results are of relevance for this question as well.
-The early 20th century, in Sweden as well as in other countries, was characterised by enourmous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
+The early 20th century, in Sweden as well as in other countries, was characterised by enormous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
 After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.

### 2017-10-26 article3.Rmd
Revision of discussion

@@
 Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85].
-We know less about how general these results are, but apparently the Swedish context we are studying differed from the English and Welsh.
+We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh.
 
@@
 Others emphasize psycho-social factors, for example @mar04, who argues that a higher status leads to better health regardless of where in the status ladder individuals are situated.
-Psychosocial factors are also suggested by @wipi09 as an explanation why equal societies always perform better than unequal ones when it comes to health and other social conditions.
+Psychosocial factors are also suggested by @wipi09 as an explanation why equal societies perform better than unequal ones when it comes to health and other social conditions.
 In the long-term perspective provided here, neither of these theories fit the historical observations.
@@
 To be fair, however, our study do not explicitly test for the possible impact of different material or psycho-social mechanisms.
-These factors could still be valid and influence the social pattern of mortality.
+These factors could still be (and probably are) valid and influence the social pattern of mortality.
 What our study establishes is that there there is no necessary eternal social gradient favouring higher social classes.
@@
 
-We must however consider other and potentially more important explanations as well, for example the possible impact of behavioural factors, something maybe reflecting mechanisms related to stressful lives.
+We must however consider other and potentially more important explanations as well.
+It is possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
+Such stress can also be expressed in unhealthful ways of living.
+These behaviours can however also have to do with certain expectations on being a man of high status.
 Our results of the gender aspect is important in this respect.
@@
 The gender difference in mortality may be caused by discrimination on the female part in the household.
-We know that historical Sweden was strongly gender segregated, discriminating women and giving them less space for action.
-In our study as well as in a previous study [@edli11], we do not find this to have negative health consequences for women.
+Historical Sweden was strongly gender segregated, discriminating women and giving them less space for action.
+In our study, we do not find the lack of power to have negative health consequences for women.
 Within the same households, the effect of social class leads to different results, high social position being beneficial for women but not for men.
@@
 Having a high social status is sometimes expressed in attitudes, norms and way of living potentially harmful for health.
-@edli11 suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and sometimes also taking risks.
+@edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and more inclined to take risks.
 Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
@@
 
-In the present-day high-income world including Sweden, smoking and drinking contributes to a substantial part of the socio-economic health differences [@mapemy14; @syla16].
-Our study suggests a strong impact also in historical Sweden.
-If this interpretation is correct, the hypothesis of @ant67 on an increasing role of health related behaviour in the modern low-mortality régime, while economic inequality determined the social mortality pattern during the early phases of the mortality transition can be questioned.
-Instead, health-related inequality can be equally strong earlier in history but expressed in a completely other way.
-To corroborate this claim, we would need to have evidence on the social differentiation of consumption patterns and behaviours in history.
+In the present-day high-income world including Sweden, smoking and drinking contributes to a substantial part of socio-economic health differences [@mapemy14; @syla16].
+Our study suggests a strong impact in historical Sweden as well.
+If this interpretation is correct, the hypothesis of @ant67, on an increasing role of health related behaviour in the modern low-mortality regime while economic inequality determined the social mortality pattern during the early phases of the mortality transition, can be questioned.
+Instead, health-related inequality can be equally strong earlier in history but expressed in a completely different way.
+To corroborate this claim, we need to have evidence on the social differentiation of consumption patterns and behaviours in history.
 Unfortunately, such data are rarely available before the 1960’s.
 At best, we have circumstantial and anecdotal evidence.
-Such evidence can however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
+Such evidence do however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
 
 Overweight, obesity and physical inactivity were probably more common in higher social classes.
-@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained the higher tobacco consumption among men in these groups.
-The social pattern among women was however reversed, something that is in line with the present study.
-There is no solid information on smoking in different social classes before the 1960, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians.
-There was also an association between higher incomes and consumption [@loco94].
-
-The same lack of statistical evidence characterizes knowledge about alcohol consumption in different social classes before 1960s, but with some exceptions.
+@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained by the higher tobacco consumption among men in these groups.
+The social pattern among women was however reversed in line with the present study.
+There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians.
+There was also an association between higher incomes and tobacco consumption [@loco94].
+
+The same lack of statistical evidence characterizes knowledge about alcohol consumption in different social classes in Sweden before 1960, but with some exceptions.
 Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with highest incomes.
-There was also an over-representation of alcohol-related deaths in higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with a more skewed consumption distribution, indicating more high-consumers in this category.
+There was also an over-representation of alcohol-related deaths in higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with an increasingly skewed consumption distribution, indicating more high-consumers in this category.
 A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
@@
 Until recently it has not been possible to investigate the really long-term development on mortality differences among adult and elderly populations.
-Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to the large class differences.
-The present study is unique in its time perspective by investigating the development from the 19th century until the present.
+Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to large class differences.
+The present study is unique in its time perspective by investigating the development from the early 19th century until the present.
 The results probably surprise most readers, and partly surprised us as well.
@@
 An important question is if our results can be generalised to other historical contexts.
-We have studied environments in northern Sweden without any large metropolis.
-Perhaps Sweden and/or this part of Sweden is a special case.
-The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages.
+We have studied environments in northern Sweden without any large metropolitan areas.Perhaps Sweden or this part of Sweden is a special case.
+The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century.
 The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century.
 In comparison with many other countries, the level of mortality has been fairly low.
-This is a relevant question, which hopefully can be addressed in other studies, maybe with a comparative approach.
+This is a relevant question, which hopefully can be addressed in other studies, preferrably with a comparative approach.
 

### 2017-10-26 article3.Rmd
Small revision

@@
 Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to large class differences.
-The present study is unique in its time perspective by investigating the development from the early 19th century until the present.
+The present study is unique in its time perspective by investigating the development from the early 19th century until the present, utilizing micro-data of high quality on a population of considerable size and social diversification.
 The results probably surprise most readers, and partly surprised us as well.

### 2017-10-26 article3.Rmd
Figure titles, corrections

@@
 ##
-cap0 <- "Probability of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom)."
+cap0 <- "Probability of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom)."
 scap0 <- "Probability by social class, ages 40--64."
-cap12 <- "Probability of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom)."
+cap12 <- "Probability of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom)."
 scap12 <- "Probability standardised by social class, ages 40--64."
-cap12log <- "Probability of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades."
+cap12log <- "Probability of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades."
 scap12log <- "Probability by social class, ages 40--64, last years."
@@
 ##
-cap22 <- "Probability of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). Standardised."
+cap22 <- "Probability of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). Standardised."
 scap22 <- "Probability by social class, ages 65--89."
-cap22log <- "Probability of dying before age 90 for a 65 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
+cap22log <- "Probability of dying before age 90 for a 65 year old person by HISCLASS and period, women (top) and men (bottom). The last decades."
 scap22log <- "Probability by social class, ages 65--89, last years."
@@
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom), cardiovascular death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom), cardiovascular death."
 scap12 <- "Total hazard by social class, ages 40--64, cardiovascular death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades, cardiovascular death."
 scap12log <- "Total hazard by social class, ages 40--64, last years, cardiovascular death."
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and decade, women (top) and men (bottom), cardiovascular death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and period, women (top) and men (bottom), cardiovascular death."
 scap22 <- "Total hazard standardised by social class, ages 65--89, cardiovascular death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). The last decades, cardiovascular death."
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). The last decades, cardiovascular death."
 scap22log <- "Total hazard by social class, ages 65--89, last years, cardiovascular death."
@@
 ##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and decade, women (top) and men (bottom), cancer death."
+cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom), cancer death."
 scap12 <- "Total hazard by social class, ages 40--64, cancer death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and decade, women (top) and men (bottom). The last decades, cancer death."
+cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades, cancer death."
 scap12log <- "Total hazard by social class, ages 40--64, last years, cancer death."
@@
 ##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and decade, women (top) and men (bottom), cancer death."
+cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and period, women (top) and men (bottom), cancer death."
 scap22 <- "Total hazard standardised by social class, ages 65--89, cancer death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and decade, women (top) and men (bottom). The last decades, cancer death."
+cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). The last decades, cancer death."
 scap22log <- "Total hazard by social class, ages 65--89, last years, cancer death."

### 2017-10-26 article3.Rmd
Correcting author names

@@
 ---
-title: Is high social class always beneficial for survival? A study of Northern Sweden 1801--2013
+title: Is high social class always beneficial for survival? A study of Northern Sweden
+  1801--2013
 author: "Sören Edvinsson and Göran Broström"
@@
 output:
+  bookdown::word_document2:
+    toc: yes
+    toc_depth: 2
   bookdown::pdf_document2:
@@
     toc_float: yes
-  bookdown::word_document2:
-    toc: yes
-    toc_depth: 2
 bibliography: mybib.bib

### 2017-10-27 cape.Rnw
merge Social class and exposure to one slide
merge Social class and exposure to split slide

@@
 \newcommand{\emp}[1]{\textcolor{blue}{#1}}
+
+\newcommand{\empg}[1]{\textcolor{green}{#1}}
+
+\newcommand{\empr}[1]{\textcolor{red}{#1}}
 
@@
 
-\begin{frame}{\sc Social class}
+\begin{frame}{Social class and relative exposure}
 
-\begin{enumerate}
-\item \emp{\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
-\item \emp{\tt middle}, HISCLASS 3, 4, 5, 6, and 8. Lower managers, lower white collar, and farmers.
+\begin{center}
+\small {Women \hspace{4cm} Men}
+\includegraphics[height = 3cm]{figs/art2-hiscskumw-1}
+\includegraphics[height = 3cm]{figs/art2-hiscskumm-1}
+\end{center}
+\footnotesize
+\begin{description}
+\item[\bf \textcolor{black}{elite}] {\sc Hisclass} 1 and 2. \emph{Higher managers and professionals.}
+\item[\empg{\bf middle}] {\sc Hisclass} \empg{3, 4, 5, 6, and 8}. \emph{Lower managers, lower white collar, and farmers.}
 %%\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
-\item \emp{\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
-\end{enumerate}
+\item[\emp{\bf worker}] {\sc Hisclass} \emp{7, 9, 10, 11 and 12}. \emph{Workers of different skills including farm workers.}
+\item[\empr{\bf NA}] {\sc Hisclass} is \empr{missing}.
+\end{description}
 
 \end{frame}
+
+\begin{frame}{Social class and relative exposure}
+
+\scriptsize
+\begin{columns}[t]
+\begin{column}{.48\textwidth}
+\begin{center}
+Social classes
+
+\begin{description}
+\item[\bf \textcolor{black}{elite}] {\sc Hisclass} 1 and 2. \emph{Higher managers and professionals.}
+\item[\empg{\bf middle}] {\sc Hisclass} \empg{3, 4, 5, 6, and 8}. \emph{Lower managers, lower white collar, and farmers.}
+%%\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
+\item[\emp{\bf worker}] {\sc Hisclass} \emp{7, 9, 10, 11 and 12}. \emph{Workers of different skills including farm workers.}
+\item[\empr{\bf NA}] {\sc Hisclass} is \empr{missing}.
+\end{description}
+
+\end{center}
+\end{column}
+\begin{column}[t]{.48\textwidth}
+
+\includegraphics[height = 3cm]{figs/art2-hiscskumw-1}
+\hspace{2cm} Exposure, women
+
+\includegraphics[height = 3cm]{figs/art2-hiscskumm-1}
+Exposure, men
+\end{column}
+
+\end{columns}
+
+\end{frame}
+
+%\begin{frame}{\sc Social class}
+
+%\begin{enumerate}
+%\item \emp{\tt elite}, HISCLASS 1 and 2. Higher managers and professionals.
+%\item \emp{\tt middle}, HISCLASS 3, 4, 5, 6, and 8. Lower managers, lower white collar, and farmers.
+%%\item \emp{\tt lowWhiteC}, HISCLASS 4, 5 and 6. Lower white collar.
+%\item \emp{\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers of different skills including farm workers.
+%\end{enumerate}
+
+%\end{frame}
 
@@
 
-\includegraphics[height=5cm]{myfigs/Top_incomes}
+\includegraphics[height=5cm]{myfigs/skord05.jpg}
 \end{center}
@@
 
-\small From the top: Elite, middle, workers, missing.
 \end{center}
+
 \end{frame}
+
+%% Combined slide:
 

### 2017-11-01 article3.Rmd
Revising presence periods

@@
 we use presence in the censuses 1975--1990 and information on deaths from the National
-Board of Health and Welfare to identify the observation periods in our
-analyses. There is however no full information on what happened in-between
-the censuses, for example the exact date of exit for persons having moved
-between two censuses. We have solved this by considering everyone as
-resident in our regions during the complete five-year period between
-censuses if the person is missing in the later census (the last period covering eleven years). This marginally overestimates presence, but should not bias the results when it
-comes to socioeconomic differences. We furthermore miss those that both
-moved in and moved out or vice versa between censuses. Exit dates are exact
-if they refer to death. In these cases we consider them as constantly
-present from the last census to the death data even if the death took place
-in another part of Sweden.
+Board of Health and Welfare.
+The population in the Skellefteå and Umeå regions are those residing there according to the censuses.
+Each census constitute the baseline for which persons are followed up during the five years until the next census (1975 census have a follow-up from 1 January 1976 until the end of 1980).
+The expection is the census 1990 where the follow-up is until 2001.
+Thus, the deaths do not necessarily take place in our region for the Linnaeus analyses, although this is usually the case.
+
+##There is however no full information on what happened in-between
+##the censuses, for example the exact date of exit for persons having moved
+##between two censuses. We have solved this by considering everyone as
+##resident in our regions during the complete five-year period between
+##censuses if the person is missing in the later census (the last period covering eleven years). ##This marginally overestimates presence, but should not bias the results when it
+##comes to socioeconomic differences. We furthermore miss those that both
+##moved in and moved out or vice versa between censuses. Exit dates are exact
+##if they refer to death. In these cases we consider them as constantly
+##present from the last census to the death data even if the death took place
+##in another part of Sweden.
 

### 2017-11-10 article3.Rmd
Revisions until 3.1

@@
 The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age---the retired population---versus the population in working age.
-We analyse both total and mortality and mortality from the major causes of death (cardiovascular diseases and cancers).
 Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
@@
 On the basis of the results, we argue that high social class is not necessarily always favourable for survival.
+Social conditions and social position have impact, but not always in the expected direction.
+In our case this is apparent for men during a large part of the studied period.
+Only at a surprisingly late time appears a male mortality class reversal.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph98].
-The reigning view has been that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
+The reigning view is that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
 This is a
@@
 
-What about life course aspects of the impact of social position for health and mortality ?
-Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage) (on this issue, see [@hof08]).
+Social conditions affect health and mortality differently during the life course.
+There are different hypotheses about the impact in adulthood and old age respectively.
+Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage) [@hof08].
 Diminishing
@@
 During the last decades, the association between trends of inequalities in mortality and income respectively is weak or non-existant according to @hoetal16.
-What the association looked like in previous periods is unknown.
-This is of particular interest since the levels of economic
-inequality were very different from those of the recent decades. Even if not a necessary implication, it is reasonable to assume that poorer groups were those most disadvantaged of large inequalities.
+How the association looked like in previous periods is unknown.
+This is of particular interest since the levels of economic inequality differed fundamentally from those of the recent decades.
+Even if not a necessary implication, it is reasonable to assume that poorer groups were those most disadvantaged of large inequalities.
 
@@
 ```
+Sweden developed during the studied period from a poor agricultural society with low urbanization to a rich welfare state.
+If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country.
 The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
@@
 During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
-The first ships during the year came to this area in May or even as
-late as June in some cases [@fah56]. Towards the end of the 19th
+The first ships during the year came to this area in May or even as late as June in some cases [@fah56].
+Towards the end of the 19th
 century, the Swedish railway system reached this part of Sweden. Thereby
@@
 inequality in mortality in this part of Sweden. The data for the present study come from two large
-population databases at the DDB, whom provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
+population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
 Our dataset constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
@@
 registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
-University. Due to data issues, we choose to use Linnaeus data from 1976. The study period from 1976 to 2000 is constructed from
+University.
+Due to data issues (much missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis).
+The study period from 1976 to 2000 is constructed from
 censuses every fifth year 1975--1990, with additional information on deaths
-from National Board of Health and Welfare (we have left out the period 1960--1975 due to more missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis).
-For the period 2002--2013 we use
+from National Board of Health and Welfare . For the period 2002--2013 we use
 the information from the yearly population registers from Statistics Sweden

### 2017-11-10 article3.Rmd
Revisions until discussion

@@
 
-##There is however no full information on what happened in-between
-##the censuses, for example the exact date of exit for persons having moved
-##between two censuses. We have solved this by considering everyone as
-##resident in our regions during the complete five-year period between
-##censuses if the person is missing in the later census (the last period covering eleven years). ##This marginally overestimates presence, but should not bias the results when it
-##comes to socioeconomic differences. We furthermore miss those that both
-##moved in and moved out or vice versa between censuses. Exit dates are exact
-##if they refer to death. In these cases we consider them as constantly
-##present from the last census to the death data even if the death took place
-##in another part of Sweden.
-
 For the last period, 2002--2013, we use the yearly population registers
 (LISA data), together with information on deaths from National Board of
-Health and Welfare. This data allows for full control of the presence periods
+Health and Welfare. This allows for full control of the presence periods
 in the studied regions.
@@
 being a scientific choice, also a practical one.
-Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, that is, from age 65 until death or last observation.
+Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population, i.e.
+from age 65 until death or last observation.
 
@@
 
-## Cause of death
-
-In the present study, we analyse mortality in the two most common disease
-groups in the ages we are investigating; cardiovascular diseases and
-cancers. National compulsory reporting of cause of death was introduced in
-1911, but our analysis of the two disease groups starts from 1901 since cause of death was fairly frequently reported even before 1911.
-Classification systems have however changed.
-Sweden did not join ICD
-until 1952, but the historical causes of death in the Poplink database have
-been coded according to an adapted form of the ICD10 system. In the present study, we identify the codes relevant for cardiovascular diseases and cancers in
-the ICD 8, 9 and 10.
-
-The distribution of causes of death is shown in
-Figure \@ref(fig:causeskum). Notice the high proportion of other causes in
-the first two periods. This category includes cases with unknown or
-unspecific causes which has implications for the cause-specific analyses
-(see later discussion). The total dominance of our two disease groups is
-apparent.
-
-```{r causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5}
-##vb01 <- vb0[vb0$event == 1, ]
-cancer <- c(tapply(skume$cancer, skume$period, sum), tapply(foli$cancer, foli$period, sum))
-heart <- c(tapply(skume$heart, skume$period, sum), tapply(foli$heart, foli$period, sum))
-death <- c(tapply(skume$event, skume$period, sum), tapply(foli$event, foli$period, sum))
-other <- death - cancer - heart
-ttt <- rbind(cancer, heart, other)
-ttt <- prop.table(ttt, 2) * 100
-barplot(ttt[3:1, ], col = 3:1, las = 2, ylab = "Per cent")
-```
-
 ## Urban vs rural residence
@@
 health risks higher in towns and cities. In the present paper, the variable **urban** controls
-for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
-on parish or place of residence.
+for this aspect by distinguishing between urban and rural residence for the periods after 1900, based on parish or place of residence.
 
@@
 allowing for adjustment for marital status and rural/urban environment.
-The central explanatory variable, social class (Hisclass), does not follow the
+The central explanatory variable, social class, does not follow the
 proportionality property, therefore the analyses are stratified with respect to the
@@
 
-# The effect of social class, 1801--1950, 1976--2013
+The models are fitted separately for sex and each period, from  1801 to
+2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
+**civst** (marital status). The cumulative
+hazards at the end of the age period (65 if 40--64, 90 if 65--89), conditional on survival 
+to the start of the age period, are the
+main target in our analysis. These numbers are used in the process of
+graphically illustrating the development over time.
 
-## All causes of death
+# The effect of social class on all--cause mortality, 1801--1950, 1976--2013
 
@@
 ```
-
-The models are fitted separately for sex and each period, from  1801 to
-2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
-**civst** (marital status). The cumulative
-hazards at the end of the age period (65 if 40--64, 90 if 65--89), conditional on survival 
-to the start of the age period, are the
-main target in our analysis. These numbers are used in the process of
-graphically illustrating the development over time.
 
@@
 The most interesting feature is the three-way interaction with **sex**, **period**, and social class.
+This indicates that fundamental changes has taken place in the relation between the variables during our study period.
 
@@
 and women. The social pattern of mortality among women is quite much in
-accordance with previous analyses from the present research group on other
-regions in northern Sweden and for other periods [@edli11; @edbr12]. The
-highest social class usually have comparatively low mortality (except for the first period when however the statistical power was weak), while
+accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12].
+The highest social class usually have comparatively low mortality (except for the first period when however the statistical power was weak), while
 working class women have high, but almost constantly about the same level as the middle category.
-There is however no substantial change in the relative risks between classes.
-Only the group with unknown social class is distinguished by much higher mortality from the 1980s
-onwards. This can however have to do with how we have defined social
-position for women before and after 1950. After 1975 it is based
+There is however no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide between workers and the middle category appears.
+A consisten social gradient has developed.
+The group with unknown social class is distinguished by much higher mortality from the 1976 onwards.
+This can however have to do with how we have defined social position for women before and after 1950.
+After 1975 it is based
 on their own occupation. Being without occupation probably include women that
@@
 the end of the 20th century. From the 1980s, there is a clear advantage when
-it comes to survival for teh elite. The relative
-differences are however more compressed than for the periods until 1950. The
+it comes to survival for the elite. The relative
+differences are however more compressed compared to the periods until 1950. The
 expected social gradient in mortality has developed. This is also what we
 usually find in analyses of social health inequalities in present-day
-Sweden as well as in most other countries. In the regions we are studying,
+Sweden as well as in most other countries, results that has formed the basis for many of the contemporary epidemiologic and demographic theories. In the regions we are studying,
 this is however a quite recent phenomenon. Corresponding to what we find
@@
 the others from the late 1970s onwards. The assumption made for women of
-unknown social class is probably even more relevant for men. They probably
+unknown social class is probably even more relevant for men. These men probably
 represent groups in society that are particularly vulnerable.
@@
 group 40--64 years. This corresponds to the Swedish development where we
-find a turning point at this time with better survival among elderly.
+find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.
 
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For
-women the results are mixed. We do not find the
+we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For women the results are mixed. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
 do we find any large disadvantage for working class elderly women at that
-time. In some periods, the social differences were small, particularly between the worker and the middle group, and the elite even had substantially higher mortality during the early 20th century. The
-social differences continued to be small until the last decades of the
+time.
+In most periods, the social differences were small, particularly between the worker and the middle group, and the elite even had somewhat higher mortality during the early 20th century.
+The social differences continued to be small until the last decades of the
 20th century. Only during the two last periods (after the year 2000) is it
@@
 For men we find indications of the same intriguing change among elderly as
-for the adult male population 40--64 years. Mortality was in fact
-substantially higher among elite men but their disadvantage was mainly 
-apparent during the 19th century. Differences are fairly small during the
-first half of the 20th century. Not until the second half of the 20th
-century is the survival much better for the highest social class while
-workers in the age of retirement had high mortality in comparison. The difference between the worker and middle group have in most periods been small.
+for the adult male population 40--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
+the 20th century. Not until the second half of the 20th
+century is the survival better for the highest social class while
+workers in the age of retirement had a little higher mortality. The difference between the worker and middle group have in most periods been small.
 
@@
 xtable(dr,label = "tab:cumswoldtab", caption = "Anova, all causes of death, 65-89.")
-```
-
-## Cardiovascular mortality
-
-In this and the following section, the focus is on analyses of mortality from the two
-dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality.
-We restrict the analyses to the period 1901 onwards.
-Many studies have shown that socio-economic conditions are strongly associated with mortality in cardiovascular diseases.
-Belonging to working class, having less education and low income
-increases mortality risks. We would therefore expect to find a clear
-social pattern in the present analysis.
-
-A general observation is that the results on causes of death and
-its development are less clear, partly because the number of
-cases are fewer but also that the quality and coverage of reported death
-causes differ in time. Thus, many deaths during the early 20th got quite vague diagnoses reported in the death registers. There was still a substantial part
-of all deaths where cause was diagnosed as unknown or dying from old age,
-despite the fact that reporting causes of death had become mandatory. For
-the period 1901--1925 to 1926--1950, the proportion of deaths from
-cardiovascular diseases increased from 25 to 43 per cent for the age group
-40 and above, thus indicating an increase of these diseases. However, at
-the same time unknown cause and death in old age decreased from 28 percent
-to 18 per cent. It is reasonable to suspect that many cardiovascular
-deaths were disguised in unspecific categories and that the increase is partly
-explained by improvements in how deaths were diagnosed, something that also seems to have been the case in Scotland at this time [@reidetal15].
-
-```{r causes41}
-cvbTab <- rbind(skume[, names(foli)], foli)
-cvbTab$event <- cvbTab$heart
-```
-
-```{r 41colorc}
-## Colors for plotting strata:
-##scol <- c("blue", rep("darkgreen", 2), "red")
-## Common for all groups:
-
-labb <- levels(cvbTab$period)
-nper <- length(labb)
-##source("R/getPHglm.R")
-## End 'common'
-```
-
-### Ages 40--64
-
-```{r 41getske4060}
-cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women, cardiovascular death."
-scap1w <-  "Cumulative hazards, ages 40--64, women, cardiovascular death."
-cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men, cardiovascular death."
-scap1m <-  "Cumulative hazards, ages 40--64, men, cardiovascular death."
-##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom), cardiovascular death."
-scap12 <- "Total hazard by social class, ages 40--64, cardiovascular death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades, cardiovascular death."
-scap12log <- "Total hazard by social class, ages 40--64, last years, cardiovascular death."
-
-cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women, cardiovascular death."
-scap2w <-  "Cumulative hazards, ages 65--89, women, cardiovascular death."
-cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men, cardiovascular death."
-scap2m <-  "Cumulative hazards, ages 65--89, men, cardiovascular death."
-##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and period, women (top) and men (bottom), cardiovascular death."
-scap22 <- "Total hazard standardised by social class, ages 65--89, cardiovascular death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). The last decades, cardiovascular death."
-scap22log <- "Total hazard by social class, ages 65--89, last years, cardiovascular death."
-
-##
-cvb4064 <- cvbTab[as.numeric(as.character(cvbTab$age)) < 64.5, ]
-cvb4064$age <- factor(cvb4064$age)
-```
-
-```{r 41hazcalw}
-source("R/getPHglm2.R")
-resw <- getPHglm2(cvb4064[cvb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(cvb4064[cvb4064$sex == "male", ],
-            nper, labb)
-```
-
-```{r 41cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8}
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 3:4,
-                    ylab = "TMR(65|40)",  main = "Women 40-64")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)", main = "Men 40-64")
-par(oldpar)
-```
-
-```{r 41cumswtab,results='asis',eval = FALSE}
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass +
-                sex * period + hisclass * period,
-            data = cvb4064, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-knitr::kable(dr, caption = "Anova, cardiovascular deaths, 40-64.")
-#print(xtable(dr,label = "tab:41cumswtab", caption = "Anova, cardiovascular deaths, 40-64."), type = "html")
-```
-
-Figure \@ref(fig:41cumsw) shows mortality levels during different periods
-from 1901 onwards. Women from the highest social class had consistently
-lower mortality from cardiovascular diseases compared to the others. We
-also observe increasing social differences from the first half of the
-century to the last decades. A new pattern has developed from 1981
-onwards. Now working class women have substantially higher cardiovascular
-mortality compared to others. We can however not exclude that this is an
-artefact of different diagnostics for deaths in different groups. Working
-class women may have had less precise diagnoses in the early 20th century.
-
-For men, we find a different pattern, however being basically consistent with what we observed concerning all-cause mortality.
-The elite group had much higher cardiovascular mortality up until the middle of the 20th century, but instead have the lowest mortality from
-1980 onwards. Working class men had a fairly low cardiovascular mortality,
-but in the same way as among women, the levels were fairly similar to the
-one of the middle category consisting of white collar, lower managers and farmers. In the last decades and similarly as for women, working class men have the highest
-mortality from this cause of death group. A social
-hierarchy in cardiovascular mortality developed both for men
-and women.
-
-### Ages 65--89
-
-```{r 41getske6089}
-cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
-cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
-```
-
-Here we focus on cardiovascular mortality among those that
-mostly have left the workforce, the age group 65--89 years.
-Much of the decline in mortality among elderly from the middle of the 20th century can be explained by lower death rates in this disease group.
-
-```{r 41hazcalwold}
-resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
-              nper, labb)
-resm <- getPHglm2(cvb6589[cvb6589$sex == "male", ],
-                nper, labb)
-```
-
-Concerning the social pattern, we find somewhat different results for elderly women compared to the women in working-active age. Cardiovascular mortality among elderly
-women does not present a consistent class pattern. Female cardiovascular
-mortality was somewhat higher among the elite and the middle group than
-among working class women until the last decades. Elderly from a working
-class background were not disadvantaged for most of the 20th century. The
-social class pattern has changed somewhat during the last periods becoming more
-in line with what to expect. The differences are however rather small.
-
-Even if results for elderly men point in different directions for some
-periods, the general trend resembles that of working-active men. The
-highest class had high cardiovascular mortality during the first half of
-the 20th century, while the expected social pattern developed and became apparent only
-during the most recent periods.
-
-```{r 41cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8}
-##source("R/plotED.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)",  main = "Women 65-89")##, exclude = 3:4)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "H(90|65)", main = "Men 65-89")
-par(oldpar)
-```
-
-```{r 41cumswoldtab,results='asis',eval=FALSE}
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * period *hisclass,
-                ##sex * period + hisclass * period,
-            data = cvb6589, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-knitr::kable(dr, caption = "Anova, cardiovascular deaths, 65-90.")
-##xtable(dr,label = "tab:41cumswoldtab", caption = "Anova, cardiovascular deaths, 65-90.")
-```
-
-## Cancer mortality
-
-While we assume clear social differences in cardiovascular diseases, this
-is not necessarily an obvious assumption when it comes to
-cancer. Different forms of cancers have different origin and
-different pathways to disease and death. The social pattern of higher
-mortality in classes with less power and accordingly less life chances are
-considered to be less clear when it comes to cancers. In our analyses,
-there is a risk that cancer mortality is underestimated during the first
-half of the 20th century, illustrated by the already mentioned large
-proportion of unknown or vague causes at that time. We believe however that
-cancers were better diagnosed than cardiovascular deaths. This is supported
-by the stability in proportion of cancer deaths between the two first
-periods, 12.5 per cent 1901--1925 and 13.5 per cent for the period
-1926--1950.
-
-```{r causes12}
-cvbTab <- rbind(skume[, names(foli)], foli)
-cvbTab$event <- cvbTab$cancer
-```
-
-```{r 12colorc}
-## Colours for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
-## Common for all groups:
-
-labb <- levels(cvbTab$period)
-nper <- length(labb)
-source("R/getPHglm.R")
-## End 'common'
-```
-
-### Ages 40--64
-
-```{r 12getske4060}
-cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women, cancer death."
-scap1w <-  "Cumulative hazards, ages 40--64, women, cancer death."
-cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men, cancer death."
-scap1m <-  "Cumulative hazards, ages 40--64, men, cancer death."
-##
-cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom), cancer death."
-scap12 <- "Total hazard by social class, ages 40--64, cancer death."
-cap12log <- "Total hazard of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades, cancer death."
-scap12log <- "Total hazard by social class, ages 40--64, last years, cancer death."
-
-cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women, cancer death."
-scap2w <-  "Cumulative hazards, ages 65--89, women, cancer death."
-cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men, cancer death."
-scap2m <-  "Cumulative hazards, ages 65--89, men, cancer death."
-##
-cap22 <- "Total hazard of dying before age 90 for a 65 year old person standardised by social class and period, women (top) and men (bottom), cancer death."
-scap22 <- "Total hazard standardised by social class, ages 65--89, cancer death."
-cap22log <- "Total hazard of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). The last decades, cancer death."
-scap22log <- "Total hazard by social class, ages 65--89, last years, cancer death."
-
-##
-cvb4064 <- cvbTab[as.numeric(as.character(cvbTab$age)) < 64.5, ]
-cvb4064$age <- factor(cvb4064$age)
-```
-
-```{r 12hazcalw}
-source("R/getPHglm2.R")
-resw <- getPHglm2(cvb4064[cvb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(cvb4064[cvb4064$sex == "male", ],
-            nper, labb)
-```
-
-```{r 12cumsw, fig.cap = cap12, fig.scap = scap12, fig.height = 8}
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 3:4,
-                    ylab = "TMR(65|40)",  main = "Women 40-64")
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE, ##exclude = 1:2,
-                    ylab = "TMR(65|40)", main = "Men 40-64")
-par(oldpar)
-```
-
-```{r 12cumswtab,results='asis',eval=FALSE}
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex *hisclass +
-                sex * period + hisclass * period,
-            data = cvb4064, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-knitr::kable(dr, caption = "Anova, cancer deaths, 40-64.")
-##xtable(dr,label = "tab:12cumswtab", caption = "Anova, cancer deaths, 40-64.")
-```
-
-Figure \@ref(fig:12cumsw) shows relative mortality levels according to
-social class during the different studied periods for the age group
-40--64. Cancer mortality among elite women was lower than that of the
-other classes during the complete studied period, while working class women
-had the highest levels during almost all periods. Consequently, we find a
-more consistent advantage of higher class in cancers compared to
-cardiovascular diseases for women, consistent with the idea of a social hierarchy in health.
-What is less expected, is that we find a stronger association between social class and survival in cancer than in cardiovascular diseases.
-We do however not find the same pattern
-for men. It is difficult to identify any general trends and features of 20th century
-male cancer mortality when it comes to social differences.
-
-### Ages 65--89
-
-```{r 12getske6089}
-cvb6589 <- cvbTab[as.numeric(as.character(cvbTab$age)) >= 64.5, ]
-cvb6589$age <- factor(cvb6589$age) # NOTE:: Now change to '6589'!
-```
-
-```{r 12hazcalwold}
-resw <- getPHglm2(cvb6589[cvb6589$sex == "female", ],
-              nper, labb)
-resm <- getPHglm2(cvb6589[cvb6589$sex == "male", ],
-                nper, labb)
-```
-
-In our analyses of class patterns in mortality, we finally turn to cancer
-mortality among the elderly. The relation between different classes and cancer mortality fluctuates
-considerably both for women and men in this age group. For some periods it
-has to do with quite few events to analyse. Women of the highest class have
-favourable mortality levels during several periods, consistent with the advantages
-we found for women in the younger age group, but on the other hand had high
-levels at other times. The social differences are however quite small
-during the most recent periods.
-
-For men we find a similar lack of social pattern in cancer mortality as we
-found for men in working-active ages. There is however a tendency during
-the last decades of higher mortality among working class men.
-
-```{r 12cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8}
-##source("R/plotED.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)",  main = "Women 65-89")##, exclude = 3:4)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "TMR(90|65)", main = "Men 65-89")
-par(oldpar)
-```
-
-```{r 12cumswoldtab,results='asis',eval=FALSE}
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex +
-                sex * period + hisclass * period,
-            data = cvb6589, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-knitr::kable(dr, caption = "Anova, cancer deaths, 65-90.")
-##xtable(dr,label = "tab:12cumswoldtab", caption = "Anova, cancer deaths, 65-90.")
 ```

### 2017-11-10 article3.Rmd
Revision of discussion

@@
 output:
-  bookdown::word_document2:
-    toc: yes
-    toc_depth: 2
   bookdown::pdf_document2:
@@
     toc_float: yes
+  bookdown::word_document2:
+    toc: yes
+    toc_depth: 2
 bibliography: mybib.bib
@@
 
-The authors of this paper have scrutinised different aspects of the social
+Previous studies have scrutinised different aspects of the social
 pattern of mortality in historical northern Sweden for some time, focusing
-on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12]. When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic
-differences were often but not always confirmed [@edv92; @edv04],
-something that has been found in other Swedish studies as well 
-[@bredro02; @bube01; @mol17].
+on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12].
+When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economicdifferences were often but not always confirmed [@edv92; @edv04],
+something that has been found in other Swedish studies as well [@bredro02; @bube01; @mol17].
 
@@
 socio-economic differences in society, for example in the Sundsvall region [@edli11; @edbr12].
+This raises the question when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century.
 The results presented here show that the consistent advantage for higher social classes do not appear until surprisingly late in the investigated regions.
@@
 A study of mortality and social class in Sweden 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total and cardiovascular mortality [@vano89].
+A recent study based on census data 1880--1910 linked to death registers of Eriksson and Dribe on the social pattern of mortality shows that the unexpected pattern found in the present study is not unique, it seems to have characterised all parts of the country.
 These findings are however contrary towards what usually is assumed and what other studies have shown.
@@
 These differences are however not of the magnitude that the main conclusions are affected.
-Most women end up in the same category as their husbands and the migrations between census dates or census date and death were not frequent.
+Most women end up in the same category as their husbands and using a time-dependent definition or a definition based on baseline with a short follow-up can only marginally change the results, especially since most people in our age groups remained in the same region during follow-up.
 
@@
 We have chosen to work with a few broad groups.
-Nevertheless, these ambiguities and possible misclassifications should not radically change the basic patterns observed.
+Nevertheless, these ambiguities and possible misclassifications cannot radically change the basic patterns observed.
 The social classes identified during the different periods represent distinct groups when it comes to resources, capacities and power.
 
-We have analysed social inequality in health from the
-perspective of social classes. However, there are
-alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude that such analyses give other results or modify the conclusions suggested here.
+We have analysed social inequality in health from the perspective of social classes.
+However, there are alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude that such analyses give other results or modify the conclusions suggested here.
 In many modern studies categorisations based on income, wealth or education are common for studies of inequality.
 It would be interesting to compare the results from this study with alternative measures, but reliable data on income and education are unfortunately rarely available in historical population databases stretching back as far as in this article.
+In the future it will however be possible to utilize a database on the 1930 census which can be linked to the regions studied here.
+That census include information on income and wealth as well as educational level.
 

### 2017-11-11 article3.Rmd
Revisions of discussion

@@
 An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not directly address the hypothesis raised by @wipi09, our results are of relevance for this question as well.
-The early 20th century, in Sweden as well as in other countries, was characterised by enormous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
+Even though our study do not directly address the hypothesis raised by @wipi09 and can thus not falsify it, our results are of relevance for this question as well.
+The early 20th century, Sweden as well as other countries was characterised by enormous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
 After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
@@
 The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
+We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower but that tye follow the general development of inequality.
 The long-term relation between economic inequality and health is still a puzzle requiring further studies.
 
-Our study seems to be in conflict with many of the dominating theories in present-day health research.
+We believe that our results have implications on several issues concerning our understanding of the demographic and epidemiologic development.
+One of the central theories has beenforwarded by @mck76, who argue that improved nutrition has been the main determinant for the increase in life expectancy during the last centuries.
+This ought to imply that people with less resources are those most vulnerable in the earlier phases of the mortality transition.
+The McKeown theses has been refuted by several scholars [@szreter88, @Colgrove02].
+Still, it has had a central role within epidemiology and social medicine.
+
+Our study seems thus to be in conflict with many of the dominating theories in present-day health research.
 In an overview of explanations for the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital et cetera.
 Most of these explanations are however based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
+
 One of these leading theories is the hypothesis of @liph95 that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to the social position.
-Others emphasize psycho-social factors, for example @mar04, who argues that a higher status leads to better health regardless of where in the status ladder individuals are situated.
+The critique againt The arguments of McKeown forced them to address the question:  "If Colgrove is right about the McKeown theses, social epidemiologic is left with a gaping hole int its explanatory reportoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02].
+This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantee better survival chances.Others emphasize psycho-social factors, for example @mar04, who argues that a higher status leads to better health regardless of where in the status ladder individuals are situated.
 Psychosocial factors are also suggested by @wipi09 as an explanation why equal societies perform better than unequal ones when it comes to health and other social conditions.
@@
 What our study establishes is that there there is no necessary eternal social gradient favouring higher social classes.
-The lacking social hierarchy may be caused by other mechanisms having stronger impact on the pattern.
+They do however point to the important role of social position.
+The lacking social hierarchy may be caused by other socially determined mechanisms having stronger impact on the pattern.
 We might have to consider the possibility of socially countervailing mechanisms [@lufr05].
@@
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy gives advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
-It is possible that they had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
+Having a good economy gives advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which will we embodied in individuals, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
 
-We must however consider other and potentially more important explanations as well.
+We must however consider other and potentially more important explanations as well and that the embodied body can be materialized in different ways.
 It is possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.

### 2017-11-11 article3.Rmd
Small revisions

@@
 Having a high social status is sometimes expressed in attitudes, norms and way of living potentially harmful for health.
-@edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and more inclined to take risks.
+@edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society now is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and more inclined to take risks.
 Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
@@
 There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians.
-There was also an association between higher incomes and tobacco consumption [@loco94].
+There was also a clear association between higher incomes and tobacco consumption [@loco94].
 

### 2017-11-12 article3.Rmd
Revising discussion and conclusion

@@
 The results presented here show that the consistent advantage for higher social classes do not appear until surprisingly late in the investigated regions.
-The basic pattern is the same in the working-age and the retired population, but the differences are smaller in old age in accordance with the status levelling hypothesis.
+The basic pattern is primarily observed in the working-age, while the differences are smaller or even non-existent in old age in accordance with the status levelling hypothesis.
 
@@
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy gives advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which will we embodied in individuals, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
+Having a good economy gives advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which become embodied in individuals, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
+The embodyment took another direction.
 
-We must however consider other and potentially more important explanations as well and that the embodied body can be materialized in different ways.
-It is possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
-Such stress can also be expressed in unhealthful ways of living.
-These behaviours can however also have to do with certain expectations on being a man of high status.
+We must thus consider other and potentially more important explanations than the pure advantage of being in high social position, reflecting that the embodiment can be materialized in different ways.
+In the following we discuss three possible pathways -- the life course perspective, social stress and class and gender based life styles.
+Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence health later in life.
+Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden.
+We are controlling for the type of environment during the studied age groups, but we do not have full information about previous residences.
+There is a possibility that the elite group to a larger extent had spent their childhood in the more harmful urban environments, and that this could influence their health and survival in adulthood.
+It seems however improbable that differences in residential backgrounds can explain most of the observed pattern.
+The second pathway we discuss relates to health risks of belonging to a certain social class.
+Usually we assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article.
+It is however possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
+Business often failed which meant a substantial loss of prestige and social network.
+Suicides certainly appeared among the elite, but it is difficult to verify the extent of social stress in this and other social classes.
+
+Social stress can also be expressed in unhealthful life styles, the third pathway we discuss.
+Harmful behaviours can be connected to certain expectations on being a man of high status.
 Our results of the gender aspect is important in this respect.
@@
 The consistent advantage of belonging to the elite have not always existed and the modern pattern of social divide in health appeared much later than expected in the regions we have studied.
+A male mortality class reversal has taken place.
 Apparently high social class is not always beneficial for survival.
@@
 Furthermore, inequality in mortality appears clearer for the working-age population, while the differences are smaller among the elderly.
-Concerning specific causes of death, the analysis shows that cardiovascular diseases contributes to a large proportion of all deaths during the 20th century, and we are not surprised that it has a fairly similar pattern as for all-cause mortality.
-Cancer is often considered less connected to social class, and we find mixed results in our analyses.
 
@@
 Studies of these aspects and their role for health and mortality in different times and places require further development.
+A possible way to proceed is to analyse the development of causes of death, and how they are distributed in different social classes.
+In particular, we should focus on causes that indicate risk behaviours.
 
 An important question is if our results can be generalised to other historical contexts.
-We have studied environments in northern Sweden without any large metropolitan areas.Perhaps Sweden or this part of Sweden is a special case.
+We have studied environments in northern Sweden without any large metropolitan areas.
+Perhaps Sweden or this part of Sweden is a special case.
 The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century.

### 2017-11-12 article3.Rmd
Small revision
edits

@@
 output:
-  bookdown::pdf_document2:
-    citation_package: natbib
-    keep_tex: yes
-    number_sections: yes
+  bookdown::word_document2:
     toc: yes
@@
     toc_float: yes
-  bookdown::word_document2:
+  bookdown::pdf_document2:
+    citation_package: natbib
+    keep_tex: yes
+    number_sections: yes
     toc: yes
@@
 The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age---the retired population---versus the population in working age.
+The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevail in old age---the retired population---as in working age.
 Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
@@
 Mortality risks in different contexts must be understood in the intersection between class and gender.
-We suggest that health-related behaviour was important not only in present-day societies, but was decisive also in earlier phases of the mortality transition.
+We suggest that health-related behaviour is important not only in present-day societies, but was decisive also in earlier phases of the mortality transition.
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
@@
 
-# The effect of social class on all--cause mortality, 1801--1950, 1976--2013
+# The effect of social class on mortality, 1801--1950, 1976--2013
 
@@
 
-### Ages 40--64
+## Ages 40--64
 
@@
 From Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with **sex**, **period**, and social class.
+The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
 This indicates that fundamental changes has taken place in the relation between the variables during our study period.
@@
 ##knitr::kable(dr)
-print(xtable(dr,label = "tab:cumswtab", caption = "Anova, all causes of death, ages 40--65."), 
+print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: Anova, ages 40--64."), 
       type = "latex", comment = FALSE)
@@
 
-### Ages 65--89
+## Ages 65--89
 
@@
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure.
+duration 25 (age 90), starting from age 65 as a summary measure. From Table \@ref(tab:cumswtabold) it is obvious that the statistical significance still is strong.
+
+```{r cumswtabold,cap="Table",results="asis"}
+fittold <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
+            data = vb6589, family = poisson)
+drold <- drop1(fittold, test = "Chisq")
+##knitr::kable(dr)
+print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: Anova, ages 65--89."), 
+      type = "latex", comment = FALSE)
+```
 
@@
 
+An important question is if our results can be generalised to other historical contexts.
+We have studied environments in northern Sweden without any large metropolitan areas.
+Perhaps Sweden or this part of Sweden is a special case.
+The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century.
+The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century.
+In comparison with many other countries, the level of mortality has been fairly low.
+The question of generalisability is relevant, which hopefully can be addressed in other studies, preferrably with a comparative approach.
+
 To explain the observed differences is difficult and we can only provide tentative hypotheses.
@@
 In particular, we should focus on causes that indicate risk behaviours.
-
-An important question is if our results can be generalised to other historical contexts.
-We have studied environments in northern Sweden without any large metropolitan areas.
-Perhaps Sweden or this part of Sweden is a special case.
-The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century.
-The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century.
-In comparison with many other countries, the level of mortality has been fairly low.
-This is a relevant question, which hopefully can be addressed in other studies, preferrably with a comparative approach.
 

### 2017-11-15 article3.Rmd
Beginning language corrections

--- 2017-11-12 article3.Rmd
@@
 In the present study, we challenge common notions of this by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
-The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age---the retired population---versus the population in working age.
-Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
+The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
+The main issue is analysed according to gender and age group (working age vs retired), testing whether the same patterns prevails in old age---the retired population---versus the population in working age.
+Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
 The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
 
-On the basis of the results, we argue that high social class is not necessarily always favourable for survival.
+Basen on the results, we argue that high social class is not necessarily always favourable for survival.
 Social conditions and social position have impact, but not always in the expected direction.

### 2017-11-16 article3.Rmd
Small correction

--- 2017-11-12 article3.Rmd
@@
 In the present study, we challenge common notions of this by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
-The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
+The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
 The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevail in old age---the retired population---as in working age.

### 2017-11-16 article3.Rmd
Language corrections part 1

@@
 The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is analysed according to gender and age-group (working age vs retired), testing if the same patterns prevail in old age---the retired population---as in working age.
-Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
+The main issue is analysed according to gender and age group (working age vs retired), testing whether the same patterns prevail in old age---the retired population---as in working age.
+Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
 The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
 
-On the basis of the results, we argue that high social class is not necessarily always favourable for survival.
+Based on the results, we argue that high social class is not necessarily always favourable for survival.
 Social conditions and social position have impact, but not always in the expected direction.
-In our case this is apparent for men during a large part of the studied period.
-Only at a surprisingly late time appears a male mortality class reversal.
+In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date there appear a male mortality class reversal.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
-We suggest that health-related behaviour is important not only in present-day societies, but was decisive also in earlier phases of the mortality transition.
+We suggest that health-related behaviour is not only important in present-day societies, but was also decisive in earlier phases of the mortality transition.
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
@@
 and other resources. In present-day welfare societies, social position is a
-strong determinant when it comes to health and mortality and the impact even
-seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph95 suggest that
+strong determinant when it comes to health and mortality and its impact even
+seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
 "... social conditions have been, are and will continue to be irreducible
-determinants of health outcomes and therefore deserves their appellation as
+determinants of health outcomes and therefore deserve their appellation of
 'fundamental causes' of disease and death".
-The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph98].
+The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph95; @liph98].
 The reigning view is that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
 This is a
-reasonable assumption since these societies were often characterised by
+reasonable assumption since these societies were often characterized by
 enormous socioeconomic differences. Knowing that access to different kinds of resources
-provide advantages in all aspects of life, the health advantage of higher
+provides advantages in all aspects of life, the health advantage of higher
 classes ought to be obvious.
 
-Recent studies, investigating social inequality in health and mortality
-with micro-data, have however questioned the generality of the assumed
-pattern [@bevp11]. Solid empirical evidence about the process is however lacking and studies
+However, recent studies investigating social inequality in health and mortality
+with micro-data have questioned the generality of the assumed
+pattern [@bevp11]. Solid empirical evidence about the process is lacking however and studies
 focusing on the issue are still few. There is a need for additional reliable
 studies from different geographical and historical settings in order to 
-better understand the role of socio-economic conditions for health and survival over time.
+better understand the role of socio-economic conditions in health and survival over time.
 
@@
 There are different hypotheses about the impact in adulthood and old age respectively.
-Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage) [@hof08].
-Diminishing
-differences may be a consequence of the circumstance that biological factors becomes increasingly important during the ageing process and in old age, leaving less impact for social factors.
-The status maintenance hypothesis basically assume
-continuity in the determinants for social health inequalities from adulthood to old age.
-The cumulative (dis)advantage hypothesis [@dan03], imply that advantages and disadvantages tend to persist and ackumulate during life in a negative spiral rewarding some while disfavouring others.
+The differences either converge in old age (status levelling), are constant (status maintenance), or diverge (cumulative advantage) [@hof08].
+Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors.
+The status maintenance hypothesis basically assumes
+continuity in the determinants of social health inequalities from adulthood to old age.
+The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages tend to persist and accumulate during life in a negative spiral rewarding some while disfavouring others.
 This leads to larger differences in old age.
 
-Another aspect of the development of social inequality in mortality,
-concerns its relation to economic inequality. @wipi09 argue that
-income inequality has an independent effect on mortality, separate from the
-direct effect of actual access to economic resources. They argue
-that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the present-day economically developed world.
+Another aspect of the development of social inequality in mortality concerns its relation to economic inequality.
+@wipi09 argue that income inequality has an independent effect on mortality, separate from the direct effect of actual access to economic resources.
+They argue
+that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the todays economically developed world.
 This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04; @wado00].
-During the last decades, the association between trends of inequalities in mortality and income respectively is weak or non-existant according to @hoetal16.
-How the association looked like in previous periods is unknown.
-This is of particular interest since the levels of economic inequality differed fundamentally from those of the recent decades.
-Even if not a necessary implication, it is reasonable to assume that poorer groups were those most disadvantaged of large inequalities.
+In recent decades, the respective association between trends of inequalities in mortality and income is weak or non-existent according to @hoetal16.
+What the association looked like in previous periods is unknown;  this is of particular interest as the levels of economic inequality differed fundamentally from those of recent decades.
+Even if not a necessary implication, it is reasonable to assume that poorer groups were the most disadvantaged due to large levels of inequality.
 
-When it comes to Sweden both income and wealth distributions were strongly skewed from the 1870's to the early 20th century. Starting during the inter-war period, income and
+When it comes to Sweden both income and wealth distributions were strongly skewed from the 1870s to the early 20th century. Starting during the inter-war period, income and
 wealth inequality continuously diminished to reach a low point at around
-1980. Thereafter, economic inequality has increased substantially until the
-present, although far from being as high as a century ago [@rowa08; @rowa09]. 
+1980. Since then, economic inequality has increased substantially until the
+present, although it is far from as high as it was a century ago [@rowa08; @rowa09]. 
 The Swedish development resembles that of other European and North
@@
 ```
-Sweden developed during the studied period from a poor agricultural society with low urbanization to a rich welfare state.
+During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state.
 If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country.
 The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
-county of Västerbotten in the north of Sweden along the coast of Gulf
-of Bothnia. This was a remote part of the country where communication with
+county of Västerbotten in northern of Sweden along the coast of the Gulf
+of Bothnia. This is a remote part of the country where communication with
 the rest of Sweden was difficult until the late 19th century. The economy
-was dominated by agriculture, making it vulnerable to harvest failures, and
-several severe famines occurred in the regions during the 1800s, for example after the harvest failure in 1867 [@segb14].
+was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14].
 During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
-The first ships during the year came to this area in May or even as late as June in some cases [@fah56].
-Towards the end of the 19th
-century, the Swedish railway system reached this part of Sweden. Thereby
-contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became
-increasingly integrated in the same epidemiological pattern as the rest of
-Sweden.
+The first ships of the year came to this area in May or even as late as June in some cases [@fah56].
+Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden and, improving the economy and making it possible to mitigate the effects of harvest failures.
+During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
 
-In our dataset, the Skellefteå region before 1950 consists of a selection
-of parishes surrounding the town of Skellefteå, founded in 1845 but with a
-very small population during the 19th century. The data from the period
-after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
+In our dataset, before 1950 the Skellefteå region consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century.
+The data from the period after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
 for the earlier period but with the addition of two more parishes. The
 majority of the 19th century population lived in rural villages and
-hamlets, getting its livelihood from agricultural production. During the
-20th century, industrialisation took place. This also led to a population increase both in the town
-and in the rural parts, resulting in a more diversified economy.
+hamlets, making its livelihood from agricultural production. During the
+20th century, industrialization took place. This also led to a population increase both in both the town
+and the rural parts, resulting in a more diversified economy.
 
-Mortality was fairly low in comparison with other parts of the country and
+Mortality was fairly low compared with other parts of the country and
 the fertility transition was late [@coa86]. The Skellefteå population
-size as defined in our data sets (all ages) was 6142 on January 1, 1801, 43212 on January 1, 1901, and 76723 at the end of the 20th century.
+size as defined in our data sets (all ages) was 6,142 on January 1, 1801, 43,212 on January 1, 1901, and 76,723 at the end of the 20th century.
 
-The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1976 onwards of Umeå municipality when another three
-parishes are included. This region has a somewhat different character from
-that of Skellefteå. Umeå town had a small population but was substantially
-larger than Skellefteå town during almost the whole studied period and was
-the administrative, educational and military centre in the county of Västerbotten.
-During the later part of the 20th century, the establishment of a university in Umeå led to a rapid population increase.
+The Umeå region in the dataset consists of Umeå urban and rural parishes 1901--1950, and from 1976 onwards of Umeå municipality, withn another three parishes included.
+This region has a somewhat different character to that of Skellefteå.
+Umeå town had a small population but was substantially
+larger than Skellefteå town for almost the whole studied period and was
+the administrative, educational and military centre of the county of Västerbotten.
+During the latter part of the 20th century, the establishment of a university in Umeå led to a rapid population increase.
 Agriculture dominated the rural part, but there were also some foundries as well as forestry industries.
-Consequently the economy was more diversified compared to Skellefteå.The population size as defined in our data sets (all ages) was 19138 on January 1, 1901 and 103970 when the 20th century ended.
+Consequently the economy was more diversified than that of Skellefteå.The population size as defined in our data sets (all ages) was 19,138 on January 1, 1901 and 103,970 when the 20th century ended.
 
@@
 The recent extension in time of data at the Demographic Data Base (DDB),
-Umeå University (http://www.cedar.umu.se) by including the period between 1900 and
+Umeå University (http://www.cedar.umu.se) by including the period 1900 to
 1950 makes it possible to investigate the long-term development of social
@@
 population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
-Our dataset constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
+Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
 The period 1801--1950 is covered by the database Poplink [@wee16].
 Poplink is based on linked parish records, allowing us to reconstruct life biographies on people as long as they remained in the region.
-The records are linked within but not
-between the regions. Data from Poplink are accessed for the period 1801--1950.
+The records are linked within, but not
+between, the regions. Data from Poplink are accessed for the period 1801--1950.
 
@@
 [@malm10], which is based on different linked national population
-registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
+registers from 1960 to 2013 and is used within the ageing programme at CEDAR, Umeå
 University.
-Due to data issues (much missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis).
+Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis).
 The study period from 1976 to 2000 is constructed from
-censuses every fifth year 1975--1990, with additional information on deaths
-from National Board of Health and Welfare . For the period 2002--2013 we use
+censuses every fifth year during the period 1975--1990, with additional information on deaths
+from the National Board of Health and Welfare. For the period 2002--2013 we use
 the information from the yearly population registers from Statistics Sweden
-(LISA) together with information on mortality from National Board of Health and
+(LISA) together with information on mortality from the National Board of Health and
 Welfare.
 
-Individuals are anonymised and the two databases are not
-linked, thus they are treated as separate units.
+Individuals are anonymized and as the two databases are not
+linked they are treated as separate units.
 This prevents us from following individuals between the two databases throughout their lives.
-It also makes it impossible to add information on individuals in the Linnaeus database from what we potentially could extract from Poplink, for example family
+It also makes it impossible to add information on individuals in the Linnaeus database from what we could potentially extract from Poplink, for example family
 background or previous social class.
@@
 
-In the data set analysed here, all individuals 40 years and older ever
-being resident in either of the regions are included. The data file
-contains the variables social class, sex, urban/rural
+In the data set analysed here, all individuals aged 40 years and older ever
+having resided in either of the regions are included. The data file
+contains the variables social class, gender, urban/rural
 residence, birth date, death date, first and last date of observation and
-type of entrance/exit. Total number of person years is `r as.integer(pn.exp)` leading to
+type of entrance/exit. The total number of person years is `r as.integer(pn.exp)` leading to
 `r pn.d` deaths in Poplink and `r as.integer(n.exp)` person years leading 
-to `r n.d` deaths in the Linnaeus database, see 
+to `r n.d` deaths in the Linnaeus database: see 
 Figure \@ref(fig:lexis2).

### 2017-11-16 article3.Rmd
Language corrections, part 2

@@
 
-Differences in available information in the two datasets as well as within
-the Linnaeus data makes it necessary to apply different approaches when it
-comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of presence, allowing us
-to have full and continuous control of the \emph{de jure} population. This is
-however not the case with the Linnaeus database. For the period 1976--2001
+Differences in available information in the two datasets as well as in
+the Linnaeus data make it necessary to apply different approaches when it
+comes to the identification of presence periods. The Poplink data provide us with exact dates, or at least year of start and exit of presence, allowing us
+to have full and continuous control over the \emph{de jure} population. This is
+not the case with the Linnaeus database, however. For the period 1976--2001
 we use presence in the censuses 1975--1990 and information on deaths from the National
 Board of Health and Welfare.
-The population in the Skellefteå and Umeå regions are those residing there according to the censuses.
-Each census constitute the baseline for which persons are followed up during the five years until the next census (1975 census have a follow-up from 1 January 1976 until the end of 1980).
-The expection is the census 1990 where the follow-up is until 2001.
+The populations in the Skellefteå and Umeå regions are those residing there according to the censuses.
+Each census constitute the baseline for which persons are followed up during the five years until the next census (e.g.
+the 1975 census has a follow-up from 1 January 1976 until the end of 1980).
+The expection of thisis the 1990 census for which the follow-up is until 2001.
 Thus, the deaths do not necessarily take place in our region for the Linnaeus analyses, although this is usually the case.
@@
 For the last period, 2002--2013, we use the yearly population registers
-(LISA data), together with information on deaths from National Board of
-Health and Welfare. This allows for full control of the presence periods
+(LISA data) together with information on deaths from the National Board of
+Health and Welfare. This allows for full control over the presence periods
 in the studied regions.
@@
 Mortality differences are analysed according to social class, based on a modified form 
-of the classification scheme Hisclass [@lema11]. The only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification
-schemes in the two databases, is occupations. Thus, analysing according to social class is apart from
-being a scientific choice, also a practical one.
-Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population, i.e.
+of the classification scheme Hisclass [@lema11]. The only, and obvious, alternative for conducting a longitudinal study of socio-economic mortality differences with consistent classification
+schemes in the two databases is occupations. Thus, analysing according to social class, apart from
+being a scientific choice, is also a practical one.
+Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population; i.e.
 from age 65 until death or last observation.
@@
 sources.
-Occupations in Poplink are coded according to the original DDB coding system that then has been adapted to Hisclass.
+In Poplink, occupations are coded according to the original DDB coding system that has then been adapted to Hisclass.
 For the Linnaeus data, HISCO is recoded into Hisclass.
-The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
+The censuses include occupational codes based on the system used in Sweden at the time, NYK (Nordisk
 Yrkesklassificering).
 We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods.
-For the period 1991--2000 the social class in the census 1990 is considered to be kept until 2001.
+For the period 1991--2000 the social class in the 1990 census is considered to be maintained until 2001.
 
 The use of different data sets forces us to apply different approaches
-towards defining the social position. Poplink usually only provides occupation for the head of household, thus underestimating female working-force participation as well as that of
-adult children residing with their parents [@vik10]. For the Poplink period, we have chosen to categorise wives according to position of the head of household, usually
+to defining the social position. Poplink usually only provides the occupation of the head of household, thus underestimating female labour force participation as well as that of
+adult children residing with their parents [@vik10]. For the Poplink period, we have chosen to categorize wives according to the position of the head of household, usually
 the husband/father, assuming that the family shares the same socio-economic
@@
 Female labour is much better covered in the Linnaeus database. A
-complication is however that it is  difficult to define households in the
-same way as in Poplink. Extra-marital cohabitation became common and female
-labour force participation developed as the norm in Sweden during this
-period. The occupation of the husband as signifier of social class became,
+complication, however, is that it is  difficult to define households in the
+same way as in Poplink. Extramarital cohabitation became common and female
+labour force participation developed as the norm, in Sweden during this
+period. The husband's occupation, as signifier of social class became,
 if not obsolete, at least less relevant. All included persons are signified
@@
 different periods are not completely comparable.
-Notice however that there was a strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].
+Noti, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].
 
-Hisclass, the classification system used as a basis for our categorisation, is a "...
+Hisclass, the classification system used as a basis for our categorization, is a "...
 HISCO-based historical international social class scheme" [@lema11; @lema02]. 
@@
 categories, based on whether the work is manual or non–manual, skill level,
-supervision and sector. Implicitly it reflects large differences in access to
-economic and other resources, status, social power et cetera.
-Longitudinal analyses of social classes for such a long period covered in this study involves several problems, related to changes in occupational structure and the meaning and contents of occupations and social positions.
+supervision, and sector. Implicitly it reflects large differences in access to
+economic and other resources, status, social power, etc.
+Longitudinal analyses of social classes for such a long period covered in this study present several problems, related to changes in occupational structure and the meaning and content of occupations and social positions.
 We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth category of unknown class:
@@
 Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
-(per cent) of exposure time according to class for men and women
+(per cent) of exposure time according to class for men and women,
 respectively. The category of workers is fairly stable over time. The
-highest class has increased substantially, although from a very low level
-in the 19th century. The increase of academics in connection to the
+highest class has increased substantially, albeit from a very low level
+in the 19th century. The increase in academics in connection to the
 establishment of Umeå University explains much of this increase. What is
-not apparent from Figure \@ref(fig:hiscskumm) is the transfer within the middle
-group from a majority of farmers during the 19th century to lower white
-collar from the middle of the 20th century.
+not apparent in Figure \@ref(fig:hiscskumm) is the transfer within the middle
+group from a majority of farmers during the 19th century to lower white-collar from the middle of the 20th century.
 Furthermore we observe a rather high percentage of missing social class in the early 19th century.
-For men the proportion is somewhat higher in the age group 65 and above, those that were retired, but still low enough to be used in our analyses (results not shown).
-If we disregard the very first period, there were however very few with missing occupation until 1950.
+For men the proportion is somewhat higher in the age group 65 and above, those who were retired, but still low enough to be used in our analyses (results not shown).
+If we disregard the very first period, however, there were very few with occupation missing until 1950.
 
@@
 
-The proportion with missing values among women is however a larger problem,
+The proportion with missing values among women is a larger problem, however;
 see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th
-century, reflecting the way womens' social position is defined.
-The frequency of missing social class is however quite high during particularly the first two periods with data from the Linnaeus database---from 1976.
+century, reflecting the way women's social position is defined.
+However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976.
 Since female social class for this period is defined
 according to own occupation, the consequence of many women being outside
-the workforce is that the proportion missing increases. For the later periods,
+the workforce is that the proportion of missing increases. For the later periods,
 the results are fairly complete.
@@
 
-Marital status, see Figure \@ref(fig:civstskum), is treated as a time-dependent variable in the data
-file. The statuses used are unmarried, married and dissolved marriages
-(without distinguishing between widowed and divorces). If no explicit
-status is given in the sources and partner is missing, the status has been
-set to unmarried. Notice that it has not been possible to identify cohabitation.
+Marital status (see Figure \@ref(fig:civstskum)), is treated as a time-dependent variable in the data
+file. The statuses used are unmarried, married and dissolved marriage
+(without distinguishing between widowed and divorced). If no explicit
+status is given in the sources and the partner is missing, the status has been
+set to unmarried. Note that it has not been possible to identify cohabitation.
 
@@
 was much higher in urban environments [@hani94; @edv92]. The unsanitary and overcrowded
-living conditions as well as the risks for importing infections made the
+living conditions, as well as the risk of importing infections, made the
 health risks higher in towns and cities. In the present paper, the variable **urban** controls
@@
 
-## Periods for analysis
+## Periods of analysis
 
-We have made separate analyses for the different time periods. The period
-1801--1950 is divided into six subperiods of length 25 years each. The 19th century
+We have done separate analyses for the different time periods. The period
+1801--1950 is divided into six subperiods of 25 years each. The 19th century
 periods represent a mainly pre-industrial society dominated by agriculture
 and with only basic welfare provisions. During the years 1901--1950,
-industrialisation started and the urban environments increased in
+industrialization started and the urban environments increased in
 population size. The modern Swedish welfare state began to develop. For the 
-period 1976 onwards, there are separate analyses on 5- or 6-year periods. Sweden had developed into a wealthy welfare society with low income inequality,
-particularly until the 1990s. A financial crisis struck Sweden in the early
-1990s, which among other things increased unemployment and changed parts of
+period 1976 onwards, there are separate analyses of five- or six-year periods.
+Sweden had developed into a wealthy welfare society with low income inequality.
+Then in the early 1990s, a financial crisis struck Sweden, which among other things increased unemployment and changed parts of
 the welfare system towards being less generous. The start of the
 last period (2008--2013) coincides with the international financial
-crisis. During the last decades income inequality has increased considerably.
+crisis. In the most recent decades income inequality has increased considerably.
 
@@
 allowing for adjustment for marital status and rural/urban environment.
-The central explanatory variable, social class, does not follow the
-proportionality property, therefore the analyses are stratified with respect to the
+As the central explanatory variable, social class, does not follow the
+proportionality property, the analyses are stratified with respect to the
 variable, and the main results are presented graphically.
@@
 It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
- `r n.d + pn.d` deaths.
-This makes it very inconvenient to use the nowadays so popular semi-parametric
-technique known as
-\emph{Cox regression} [@cox72]. Instead we use a parametric model
-that allows for reduction of the data
-to \emph{sufficient statistics} [@fisher22]. The idea is to keep all
-relevant covariates \emph{categorical}, and then create a table with all
+ `r n.d + pn.d` deaths. This makes it highly inconvenient to use the currently so popular semi-parametric technique known as \emph{Cox regression} [@cox72]. Instead we use a parametric model
+that allows for a reduction of the data to \emph{sufficient statistics} [@fisher22]. The idea is to keep all relevant covariates as \emph{categorical} and then create a table with all
 possible combinations of levels of the covariates. For each such
@@
 urban/rural groups, four social classes,
-ten age groups, and 22 cohorts, in all 105600 possible
+ten age groups, and 22 cohorts; in all, 105,600 possible
 combinations. However, in this scheme there are many structural zeroes
 (impossible combinations) and some random zeros, so the actual number of
-combinations we end up with is 14084. The data set is thus reduced in
+combinations we end up with is 14,084. The data set is thus reduced in
 size by 99 per cent.
 
-The parametric model we use for the baseline hazard function is piece-wise
-constant on five-year intervals, so we arrive at the following proportional
+The parametric model we use for the baseline hazard function is piecewise
+constant at five-year intervals, so we arrive at the following proportional
 hazards regression model:
@@
 Since all components of $\mathbf{x}$ are categorical covariates, the
-estimation of \@ref(eq:prop) can be performed in two ways, first by the
+estimation of \@ref(eq:prop) can be performed in two ways; based on the
 original data set and a survival analysis routine handling a piecewise
-constant hazard, or second, by the reduced data set (table) and
+constant hazard;  or based on the reduced data set (table) and
 \emph{Poisson regression}. The second way is faster than the first by a
-factor of size at least 100, often much more, and
+size factor of at least 100, often much more, and
 the point is that the two ways give exactly the same results. This is
@@
 [@gb12;@eha]. In the analysis, data are divided into two age groups,
-40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period are analysed separately, 
+40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
 but we then combine results in order to show the development of the social patterning of mortality during the studied period.
@@
 reason for this choice is that the effect of social class on survival is
-non-proportional, that is, it varies with age. With the reduced data set
+non-proportional; that is, it varies with age. With the reduced data set
 and Poisson regression, this is simply achieved by introducing an
@@
 results are presented graphically.  The comments on the results are based on
-the figures and focuses on the central variable in our study, social class.
+the figures and focus on the central variable in our study, social class.
 
@@
 ##
-cap0 <- "Probability of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom)."
+cap0 <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
 scap0 <- "Probability by social class, ages 40--64."
-cap12 <- "Probability of dying before age 65 for a 40 year old person standardised by social class and period, women (top) and men (bottom)."
-scap12 <- "Probability standardised by social class, ages 40--64."
-cap12log <- "Probability of dying before age 65 for a 40 year old person by social class and period, women (top) and men (bottom). The last decades."
+cap12 <- "Probability of dying before age 65 for a 40-year-old standardized by social class and period, women (top) and men (bottom)."
+scap12 <- "Probability standardized by social class, ages 40--64."
+cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
 scap12log <- "Probability by social class, ages 40--64, last years."
@@
 ##
-cap22 <- "Probability of dying before age 90 for a 65 year old person by social class and period, women (top) and men (bottom). Standardised."
+cap22 <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). Standardised."
 scap22 <- "Probability by social class, ages 65--89."
-cap22log <- "Probability of dying before age 90 for a 65 year old person by HISCLASS and period, women (top) and men (bottom). The last decades."
+cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
 scap22log <- "Probability by social class, ages 65--89, last years."
@@
 
-From Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
+Looking at Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
 The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
-This indicates that fundamental changes has taken place in the relation between the variables during our study period.
+This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
 
@@
 40--64 during the different periods. Obviously, but not surprisingly, mortality
-is decreasing over time for both sexes. Women have consistently better survival
+decreases over time for both sexes. Women have consistently better survival
 than men. This is a well-known phenomenon in Swedish
@@
 substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. We must however distinguish between differences in absolute risks compared to relative risks [@hoetal16]. The compression that is
+absolute differences between social classes. We must, however, distinguish between differences in absolute risks compared to relative risks [@hoetal16]. The compression that is
 visible for later decades is partly explained by the decreasing general
-mortality and is not necessarily representing a homogenisation of relative risks.
+mortality and does not necessarily represent a homogenization of relative risks.
 
-By standardising the levels in each time period for each class, we 
+By standardizing the levels in each time period for each class, we 
 illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
@@
 
-The changing economy have some consequences for the analysis, particularly when
-it comes to the agricultural sector. While farmers were common occupations
-up until the first half of the 20th century, these occupations has constantly
-decreased in numbers thereafter. This means that the middle category has
-a different character during the last periods.
+The changing economy has some consequences for the analysis, particularly when
+it comes to the agricultural sector. While farming occupations were common
+up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then. This means that the middle category has
+a different character in the last periods.
 
 The most striking result of the analysis is the different patterns for men
-and women. The social pattern of mortality among women is quite much in
+and women. The social pattern of mortality among women is a great deal in
 accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12].
-The highest social class usually have comparatively low mortality (except for the first period when however the statistical power was weak), while
-working class women have high, but almost constantly about the same level as the middle category.
-There is however no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide between workers and the middle category appears.
-A consisten social gradient has developed.
-The group with unknown social class is distinguished by much higher mortality from the 1976 onwards.
-This can however have to do with how we have defined social position for women before and after 1950.
-After 1975 it is based
-on their own occupation. Being without occupation probably include women that
-are more frail when it comes to health or are vulnerable due to social characteristics.
+The highest social class usually has comparatively low mortality (except for the first period when the statistical power was weak), while
+working-class women have high mortality, but almost constantly about the same level as the middle category.
+However, there is no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide appears between workers and the middle category.
+A consistent social gradient has developed; the group with unknown social class is distinguished by much higher mortality from the 1976 onwards.
+This might have to do with how we have defined social position for women before and after 1950.
+After 1975 it is based on their own occupation; lacking an occupation likely includes women who are more frail when it comes to health or are vulnerable due to social characteristics.
 
-A completely different pattern appears among men, being the most surprising finding in
+A completely different pattern appears among men, which is the most surprising finding in
 our analyses. The groups we expect to have the lowest mortality, if
@@
 determines survival, instead have by far the highest hazards of dying
-during the nineteenth and much of the 20th century. Workers on the
-other hand have comparatively low mortality until 1950, at the same
+during the 19th century and much of the 20th. Workers, on the
+other hand, have comparatively low mortality until 1950, at the same
 level as the middle class.
-The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we find better survival among workers compared to the elite.
+The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.
 
-The relative positions change radically towards
-the end of the 20th century. From the 1980s, there is a clear advantage when
-it comes to survival for the elite. The relative
-differences are however more compressed compared to the periods until 1950. The
-expected social gradient in mortality has developed. This is also what we
-usually find in analyses of social health inequalities in present-day
-Sweden as well as in most other countries, results that has formed the basis for many of the contemporary epidemiologic and demographic theories. In the regions we are studying,
-this is however a quite recent phenomenon. Corresponding to what we find
-for women, those without known occupation have much higher mortality than
+The relative positions change radically towards the end of the 20th century. From the 1980s, there is a clear advantage when it comes to survival for the elite. However, the relative
+differences are more compressed compared to the periods until 1950.
+A reversal of the class pattern in mortality has taken place, and the expected social gradient in mortality has developed.
+This is also what we usually find in analyses of social health inequalities in present-day
+Sweden as well as in most other countries, results that have formed the basis for many of the contemporary epidemiologic and demographic theories. In the regions we are studying,
+this is a quite recent phenomenon. Corresponding to what we find
+for women, those without a known occupation have much higher mortality than
 the others from the late 1970s onwards. The assumption made for women of
-unknown social class is probably even more relevant for men. These men probably
+unknown social class is probably even more relevant for men, who likely
 represent groups in society that are particularly vulnerable.
@@
 1851--1950, men having about three times higher risks than women in the
-same class (basically living in the same households). In the other classes, the difference was much smaller. It was men in high position that experienced the largest risks.
+same class (basically living in the same households). In the other classes, the difference was much smaller. It was men in high positions who experienced the largest risks.
 
@@
 
-We now turn to the elderly, those that in most cases have left the workforce.
+We now turn to the elderly, those who in most cases have left the workforce.
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure. From Table \@ref(tab:cumswtabold) it is obvious that the statistical significance still is strong.
+duration 25 (age 90), starting from age 65 as a summary measure. Looking at Table \@ref(tab:cumswtabold) it is obvious that the statistical significance is still strong.
 
@@
 The general pattern and development of mortality (not shown here) do not
-differ that much from the younger age group, of course representing a much
-higher mortality level earlier in history. Mortality among women was, as
-expected, lower than among men. We find a clear decline in mortality, but
-the fastest decline did not set in definitely until some time after the
-Second World War, that is, later than for the age
-group 40--64 years. This corresponds to the Swedish development where we
+differ much from the younger age group, of course representing a much
+higher mortality level earlier in history. Mortality among women, as
+expected, was lower than among men. We find a clear decline in mortality, but
+the fastest decline did not set in definitely until sometime after the
+Second World War; that is, later than for the age
+group 40--64 years. This corresponds to the Swedish development whereby we
 find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For women the results are mixed. We do not find the
+we find for the younger age group in line with the status-levelling hypothesis (see Figure \@ref(fig:cumswold)). For women the results are mixed. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
-do we find any large disadvantage for working class elderly women at that
+do we find any large disadvantage for working-class elderly women at that
 time.
-In most periods, the social differences were small, particularly between the worker and the middle group, and the elite even had somewhat higher mortality during the early 20th century.
+In most periods the social differences were small, particularly between the workers and the middle group, and the elite even had somewhat higher mortality during the early 20th century.
 The social differences continued to be small until the last decades of the
-20th century. Only during the two last periods (after the year 2000) is it
+20th century. It is only during the two last periods (after the year 2000) that it is
 possible to establish the expected social gradient in mortality.
@@
 For men we find indications of the same intriguing change among elderly as
-for the adult male population 40--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
-the 20th century. Not until the second half of the 20th
-century is the survival better for the highest social class while
-workers in the age of retirement had a little higher mortality. The difference between the worker and middle group have in most periods been small.
+for the adult male population aged 40--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
+the 20th century. It is not until the second half of the 20th
+century that the survival is better for the highest social class while
+workers of retirement age have slightly higher mortality. The difference between the worker and middle group has been small in most periods.
 
@@
 
-Previous studies have scrutinised different aspects of the social
+Previous studies have scrutinized different aspects of the social
 pattern of mortality in historical northern Sweden for some time, focusing
 on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12].
-When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economicdifferences were often but not always confirmed [@edv92; @edv04],
+When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic differences have often, but not always,  been confirmed [@edv92; @edv04],
 something that has been found in other Swedish studies as well [@bredro02; @bube01; @mol17].

### 2017-11-16 article3.Rmd
Language corrections part 3

@@
 
-Studies based on microdata on the adult and elderly population in
+Studies based on micro-data on adult and elderly populations in
 history have been rare, mainly due to lack of adequate data. The results
 from such studies show other patterns than that for infant and child
-mortality. Some report substantial differences but many, particularly those based on
-Swedish data, have found surprisingly small or even reversed socio-economic
-differences during the 19th and early 20th century, sharply contrasting from the situation in present-day society. The expected social gradient in survival, meaning that people
-from higher social classes always had an advantage, was in many cases,
+mortality. While some report substantial differences, many -- particularly those based on
+Swedish data -- have found surprisingly small or even reversed socio-economic
+differences during the 19th and early 20th centuries, sharply contrasting from the situation in present-day society. The expected social gradient in survival -- meaning that people
+from higher social classes always had an advantage -- was in many cases,
 particularly among adult men, not found despite the well-documented
 socio-economic differences in society, for example in the Sundsvall region [@edli11; @edbr12].
-This raises the question when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century.
-The results presented here show that the consistent advantage for higher social classes do not appear until surprisingly late in the investigated regions.
-The basic pattern is primarily observed in the working-age, while the differences are smaller or even non-existent in old age in accordance with the status levelling hypothesis.
+This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century.
+The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions.
+The basic pattern is primarily observed in the working-age population, while the differences are smaller or even non-existent in old age in accordance with the status-levelling hypothesis.
 
-The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, although with somewhat different social categories.
-They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as in the 1950s.
+The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories.
+They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s.
 Similar results have also been found in Estonia [@jaadla17].
-A study of mortality and social class in Sweden 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total and cardiovascular mortality [@vano89].
-A recent study based on census data 1880--1910 linked to death registers of Eriksson and Dribe on the social pattern of mortality shows that the unexpected pattern found in the present study is not unique, it seems to have characterised all parts of the country.
-These findings are however contrary towards what usually is assumed and what other studies have shown.
-Those studies usually referred to when discussing the historical large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85].
-We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh.
+A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89].
+A recent study Eriksson and Dribe on the social pattern of mortality, based on 1880---1910 census data linked to death registers shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country.
+These findings are contrary to what is usually assumed and what other studies have shown.
+The studies usually referred to when discussing the historically large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85].
+We know less about how general these results are, but apparently the Swedish context differed from the English and Welsh ones.
 
 The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
-Some of the limitations have been addressed in the presentation of the variables: the fact that social class was defined according to household until 1950 but individually defined from 1976 (affecting the social class of women), different definitions of presence during the same two periods and some differences in parishes included at different periods.
-These differences are however not of the magnitude that the main conclusions are affected.
-Most women end up in the same category as their husbands and using a time-dependent definition or a definition based on baseline with a short follow-up can only marginally change the results, especially since most people in our age groups remained in the same region during follow-up.
+Some of the limitations have been addressed in the presentation of the variables: the fact that social class of women was defined according to household until 1950 but was defined individually from 1976, different definitions of presence during the same two periods and some differences in parishes included at different periods.
+However, these differences are neither of the magnitude nor of the character that the main conclusions are affected.
+Most women end up in the same category as their husbands and using a time-dependent definition or a definition based on baseline with a short follow-up can only marginally change the results, especially since most people in our age groups remained in the same region at follow-up.
 
 One obstacle concerns the quality of available information.
-Occupational titles in historical sources are not always distinct enough to define the correct class.
+Occupational titles in historical sources are not always sufficiently distinct to define the correct class.
 Studies covering long periods face the problem that occupations and classes change character---they do not represent the same thing.
@@
 Nevertheless, these ambiguities and possible misclassifications cannot radically change the basic patterns observed.
-The social classes identified during the different periods represent distinct groups when it comes to resources, capacities and power.
+The social classes identified during the different periods represent distinct groups when it comes to resources, capacities, and power.
 
 We have analysed social inequality in health from the perspective of social classes.
-However, there are alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude that such analyses give other results or modify the conclusions suggested here.
-In many modern studies categorisations based on income, wealth or education are common for studies of inequality.
+However, there are alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude the possibility that such analyses might give other results or modify the conclusions suggested here.
+In many modern studies categorizations based on income, wealth or education are common when examining inequality.
 It would be interesting to compare the results from this study with alternative measures, but reliable data on income and education are unfortunately rarely available in historical population databases stretching back as far as in this article.
-In the future it will however be possible to utilize a database on the 1930 census which can be linked to the regions studied here.
-That census include information on income and wealth as well as educational level.
+In the future, however, it will be possible to utilize a database on the 1930 census that can be linked to the regions studied here.
+That census includes information on income and wealth as well as educational level.
 
 An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
-Even though our study do not directly address the hypothesis raised by @wipi09 and can thus not falsify it, our results are of relevance for this question as well.
-The early 20th century, Sweden as well as other countries was characterised by enormous economic differences, still this did in our regions not lead to any consistent advantage for the highest social class.
-After the Second World War, economic differences diminished, and was at its lowest in Sweden at about 1980, i.e.
+Even though our study does not directly address the hypothesis posed by @wipi09 and thus cannot falsify it, our results are relevant for this question as well.
+In the early 20th century, like other countries, Sweden was characterised by enormous economic differences; still, in our regions this did not lead to any consistent advantage for the highest social class.
+After the Second World War, economic differences began diminishing, and were at their lowest in Sweden around 1980; i.e.
 at a time when the expected pattern of a social hierarchy in survival appeared.
 The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
-We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower but that tye follow the general development of inequality.
-The long-term relation between economic inequality and health is still a puzzle requiring further studies.
+We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower than in most other parts of Sweden but that it has followed the general development of first diminishing and later increasing inequality.
+The long-term relationship between economic inequality and health is still a puzzle requiring further studies.
 
-We believe that our results have implications on several issues concerning our understanding of the demographic and epidemiologic development.
-One of the central theories has beenforwarded by @mck76, who argue that improved nutrition has been the main determinant for the increase in life expectancy during the last centuries.
-This ought to imply that people with less resources are those most vulnerable in the earlier phases of the mortality transition.
-The McKeown theses has been refuted by several scholars [@szreter88, @Colgrove02].
-Still, it has had a central role within epidemiology and social medicine.
+We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development.
+One of the central theories has been presented by @mck76, who argues that improved nutrition has been the main determinant of the increase in life expectancy in recent centuries.
+This ought to imply that people with less adequate resources are the most vulnerable in the earlier phases of the mortality transition.
+While the McKeown thesis has been refuted by several scholars [@szreter88, @Colgrove02], it has played a central role within epidemiology and social medicine.
 
 Our study seems thus to be in conflict with many of the dominating theories in present-day health research.
-In an overview of explanations for the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital et cetera.
-Most of these explanations are however based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
+In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc.
+However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
 

### 2017-11-17 article3.Rmd
Language corrections to the end

@@
 
-One of these leading theories is the hypothesis of @liph95 that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to the social position.
-The critique againt The arguments of McKeown forced them to address the question:  "If Colgrove is right about the McKeown theses, social epidemiologic is left with a gaping hole int its explanatory reportoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02].
-This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantee better survival chances.Others emphasize psycho-social factors, for example @mar04, who argues that a higher status leads to better health regardless of where in the status ladder individuals are situated.
-Psychosocial factors are also suggested by @wipi09 as an explanation why equal societies perform better than unequal ones when it comes to health and other social conditions.
-In the long-term perspective provided here, neither of these theories fit the historical observations.
-Their validity are mainly apparent for the most recent decades.
-To be fair, however, our study do not explicitly test for the possible impact of different material or psycho-social mechanisms.
+One of these leading theories is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to social position.
+The criticism of the arguments by McKeown forced them to address following question:  "If Colgrove is right about the McKeown thesis, social epidemiology is left with a gaping hole in its explanatory repertoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02].
+This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantees better survival chances.Others emphasize psycho-social factors, for example @mar04, who argues that higher status leads to better health regardless of where on the status ladder the individual is situated.
+Psychosocial factors are also suggested by @wipi09 as an explanation for why equal societies perform better than unequal ones when it comes to health and other social conditions.
+In the long-term perspective provided here, neither of these theories fits the historical observations; their validity is mainly apparent for the most recent decades.
+To be fair, however, our study does not explicitly test for the possible impact of different material or psycho-social mechanisms.
 These factors could still be (and probably are) valid and influence the social pattern of mortality.
-What our study establishes is that there there is no necessary eternal social gradient favouring higher social classes.
-They do however point to the important role of social position.
-The lacking social hierarchy may be caused by other socially determined mechanisms having stronger impact on the pattern.
+What our study establishes is that there is no necessary eternal social gradient favouring higher social classes.
+Our results do however point to the important role of social position.
+The lacking social hierarchy may be caused by other socially determined mechanisms having a stronger impact on the pattern.
 We might have to consider the possibility of socially countervailing mechanisms [@lufr05].
 
-The historical perspective shows the complexity of how health and mortality is socially constructed and illustrates the need to contextualize our analyses.
-The dominant theories are not necessarily wrong but they may need to be further developed and refined.
+The historical perspective shows the complexity of how health and mortality are socially constructed, and illustrates the need to contextualize our analyses.
+The dominant theories are not necessarily wrong but may need to be further developed and refined.
 The WHO Commission on Social Determinants of Health framework provides a holistic model on how socio-economic position influences health.
@@
 What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy gives advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which become embodied in individuals, but even though these advantages were at hand in our studied regions, it did not result in better health for men in the most privileged groups until the late 20th century.
-The embodyment took another direction.
+Having a good economy offers advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which become embodied in individuals; but even though these advantages were at hand in our studied regions, this did not result in better health for men in the most privileged groups until the late 20th century.
+The embodiment took another direction.
 
-We must thus consider other and potentially more important explanations than the pure advantage of being in high social position, reflecting that the embodiment can be materialized in different ways.
-In the following we discuss three possible pathways -- the life course perspective, social stress and class and gender based life styles.
-Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence health later in life.
+We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways.
+In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles.
+Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one'shealth later in life.
 Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden.
-We are controlling for the type of environment during the studied age groups, but we do not have full information about previous residences.
-There is a possibility that the elite group to a larger extent had spent their childhood in the more harmful urban environments, and that this could influence their health and survival in adulthood.
-It seems however improbable that differences in residential backgrounds can explain most of the observed pattern.
-The second pathway we discuss relates to health risks of belonging to a certain social class.
-Usually we assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article.
-It is however possible that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantagous position.
+We control for the type of environment people live in, but we do not have full information about previous residences.
+There is a possibility that the elite group to a greater extent had spent their childhood in the more harmful urban environments, and that this might have influenced their health and survival in adulthood.
+It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern.
+The second pathway we discuss relates to the health risks of belonging to a certain social class.
+We usually assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article.
+It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position.
 Business often failed which meant a substantial loss of prestige and social network.
-Suicides certainly appeared among the elite, but it is difficult to verify the extent of social stress in this and other social classes.
+Suicides certainly occurred among the elite, but it is difficult to verify the extent of social stress in this and other social classes.
 
-Social stress can also be expressed in unhealthful life styles, the third pathway we discuss.
-Harmful behaviours can be connected to certain expectations on being a man of high status.
-Our results of the gender aspect is important in this respect.
+Social stress can also be expressed in unhealthy lifestyles, the third pathway we discuss.
+Harmful behaviours can be connected to certain expectationsassociated with being a man of high status.
+Our results regarding the gender aspect are important in this respect.
 As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07].
-The gender difference in mortality may be caused by discrimination on the female part in the household.
-Historical Sweden was strongly gender segregated, discriminating women and giving them less space for action.
+The gender difference in mortality may be caused by discrimination of the female party in the household.
+Historical Sweden was strongly gender-segregated, discriminating against women and giving them less space for action.
 In our study, we do not find the lack of power to have negative health consequences for women.
@@
 Instead, we might consider the role of behaviours based on distinct gendered and class expectations.
-Having a high social status is sometimes expressed in attitudes, norms and way of living potentially harmful for health.
-@edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society now is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking, smoking and more inclined to take risks.
-Alcohol has been suggested as part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking life may be an important component of male mortality [@cour00].
+Having a high social status is sometimes expressed in attitudes, norms and way of living that are potentially harmful to one's health.
+@edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.
+Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00].
 Different historical contexts promote different ways to spend economic resources [@rasp06; @jaadla17].
 
-In the present-day high-income world including Sweden, smoking and drinking contributes to a substantial part of socio-economic health differences [@mapemy14; @syla16].
+In the today's high-income world including Sweden, smoking and drinking contribute to a substantial part of socio-economic health differences [@mapemy14; @syla16].
 Our study suggests a strong impact in historical Sweden as well.
-If this interpretation is correct, the hypothesis of @ant67, on an increasing role of health related behaviour in the modern low-mortality regime while economic inequality determined the social mortality pattern during the early phases of the mortality transition, can be questioned.
-Instead, health-related inequality can be equally strong earlier in history but expressed in a completely different way.
-To corroborate this claim, we need to have evidence on the social differentiation of consumption patterns and behaviours in history.
-Unfortunately, such data are rarely available before the 1960’s.
+If this interpretation is correct, the hypothesis by @ant67, on an increasing role of health-related behaviour in the modern low-mortality regime while economic inequality determined the social mortality pattern during the early phases of the mortality transition, can be questioned.
+Instead, health-related inequality might be equally strong earlier in history but expressed in a completely different way.
+To corroborate this claim, we need to find evidence on the social differentiation of consumption patterns and behaviours in history.
+Unfortunately, such data are rarely available before the 1960s.
 At best, we have circumstantial and anecdotal evidence.
-Such evidence do however support the idea of less health-beneficial ways of living among elite men (@edli11 ).
+However, such evidence does support the idea of less health-beneficial lifestyles among elite men (@edli11 ).
 
 Overweight, obesity and physical inactivity were probably more common in higher social classes.
-@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, something they suggest can be explained by the higher tobacco consumption among men in these groups.
-The social pattern among women was however reversed in line with the present study.
-There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, illustrated for example by the fact that smoking initially was more prevalent among physicians.
+@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, which they suggest can be explained by the higher tobacco consumption among men in these groups.
+The social pattern among women was the reverse, however, in line with the present study.
+There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating a higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, for example illustrated by the fact that smoking was initially more prevalent among physicians.
 There was also a clear association between higher incomes and tobacco consumption [@loco94].
@@
 The same lack of statistical evidence characterizes knowledge about alcohol consumption in different social classes in Sweden before 1960, but with some exceptions.
-Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with highest incomes.
-There was also an over-representation of alcohol-related deaths in higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there has been an equalization and even reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with an increasingly skewed consumption distribution, indicating more high-consumers in this category.
-A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, maybe reflecting a change in consumption patterns.
-Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with highest mortality in the lowest class [@cropr10].
-If these indications of changes in health-related behaviour are significant enough to explain the present findings can be discussed.
-We believe that they certainly are at least an important part of the story.
+Access to alcohol was much easier for the wealthy during the period of strong alcohol restriction in 1915-1955 and the purchase of alcoholic beverages was substantially higher among those with the highest incomes.
+There was also an overrepresentation of alcohol-related deaths in the higher social classes in the 1960s: @noro98 suggest that during the last decades of the 20th century, there was an equalization and even a reversal of alcohol consumption between social strata, with higher levels among manual labourers but also with an increasingly skewed consumption distribution, indicating more high consumers in this category.
+A similar reversal has been observed in England and Wales during the 20th century when it comes to cirrhosis mortality, possibly reflecting a change in consumption patterns.
+Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with the highest mortality in the lowest class [@cropr10].
+It could be discussed whether these indications of changes in health-related behaviour are sufficiently significant to explain the present findings; we believe they are certainly at least an important part of the story.
 
@@
 
-In this article we set our task to analyse social differences in mortality from a long-term perspective.
-Until recently it has not been possible to investigate the really long-term development on mortality differences among adult and elderly populations.
+In this article we set our task as analysing social differences in mortality from a long-term perspective.
+Until recently it has not been possible to investigate the truly long-term development of mortality differences among adult and elderly populations.
 Still, the prevailing idea in many parts of the scholarly society is that strong socio-economic differences in health and mortality have always existed and were possibly even stronger in historical societies due to large class differences.
-The present study is unique in its time perspective by investigating the development from the early 19th century until the present, utilizing micro-data of high quality on a population of considerable size and social diversification.
-The results probably surprise most readers, and partly surprised us as well.
-The consistent advantage of belonging to the elite have not always existed and the modern pattern of social divide in health appeared much later than expected in the regions we have studied.
+The present study is unique in its time perspective, investigating the development from the early 19th century until the present, utilizing micro-data of high quality on a population of considerable size and social diversification.
+The resultswill likely surprise most readers; they partly surprised us as well.
+The consistent advantage of belonging to the elite has not always existed, and the modern pattern of social divide in health appeared much later than expected in the regions we studied.
 A male mortality class reversal has taken place.
@@
 
-There were however large differences in the pattern between men and women as well as between different age groups.
+However, there were large differences in the pattern between men and women as well as between different age groups.
 Class as well as gender must be considered.
-The social pattern of female mortality are in line with the expected, but a social gradient in male adult mortality is a recent phenomenon.
+The social pattern of female mortality is in line with what was expected, but a social gradient in male adult mortality is a recent phenomenon.
+It is an interesting paradox hat both the class and gender aspects in the analysis show that those with more power and resources were in fact less successful when it came to survival.
 Furthermore, inequality in mortality appears clearer for the working-age population, while the differences are smaller among the elderly.
 
-An important question is if our results can be generalised to other historical contexts.
-We have studied environments in northern Sweden without any large metropolitan areas.
+An important question is whether our results can be generalized to other historical contexts.
+We have studied environments in northern Sweden with no large metropolitan areas.
 Perhaps Sweden or this part of Sweden is a special case.
-The involvement of government both at the national and local level may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century.
+The involvement of government at both the national and local levels may have been stronger, thus mitigating socio-economic disadvantages, but this is doubtful concerning the 19th century.
 The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century.
-In comparison with many other countries, the level of mortality has been fairly low.
-The question of generalisability is relevant, which hopefully can be addressed in other studies, preferrably with a comparative approach.
+Compared with many other countries, the level of mortality has been fairly low.
+The question of generalizability is relevant, which can hopefully be addressed in other studies, preferably with a comparative approach.
 
-To explain the observed differences is difficult and we can only provide tentative hypotheses.
-We have discussed the possible role of behavioural influences on mortality in different historical context, suggesting that such aspects are not only important in the modern epidemiological regime but might have been decisive in earlier phases of the epidemiologic transition as well.
+It is difficult to explain the observed differences, and we can only offer tentative hypotheses.
+We have discussed the possible role of behavioural influences on mortality in different historical contexts, suggesting that such aspects are not only important in the modern epidemiological regime but might have been decisive in earlier phases of the epidemiologic transition as well.
 The results indicate that we should consider the role of gender and class expectations of behaviour.
-Unfortunately, the evidence for behavioural differences in history are usually circumstantial and anecdotal, at least when it comes to social class.
-Studies of these aspects and their role for health and mortality in different times and places require further development.
+Unfortunately, the evidence for behavioural differences in history is usually circumstantial and anecdotal, at least when it comes to social class.
+Studies of these aspects and their role in health and mortality in different times and places require further development.
 A possible way to proceed is to analyse the development of causes of death, and how they are distributed in different social classes.
@@
 
-This work was undertaken within the research program *Ageing and
+This work was undertaken within the research programme *Ageing and
 Living Conditions* at the Centre for Demographic and Ageing Research
@@
 Demographic Data Base, Umeå University, for giving us access to data and
-Jonas Helgertz, Department of Economic History, Lund University, for
-providing us with scripts for creating Hisclass from Swedish Census
+Jonas Helgertz at the Department of Economic History, Lund University, for
+providing us with scripts for creating Hisclass from Swedish census
 data.

### 2017-11-22 article3.Rmd
Making last corrections, and adding a reference

@@
 output:
-  bookdown::word_document2:
+  bookdown::pdf_document2:
+    citation_package: natbib
+    keep_tex: yes
+    number_sections: yes
     toc: yes
@@
     toc_float: yes
-  bookdown::pdf_document2:
-    citation_package: natbib
-    keep_tex: yes
-    number_sections: yes
+  bookdown::word_document2:
     toc: yes
@@
 Based on the results, we argue that high social class is not necessarily always favourable for survival.
-Social conditions and social position have impact, but not always in the expected direction.
-In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date there appear a male mortality class reversal.
+Social conditions and social position certainly have impact, but not always in the expected direction.
+In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date there appear a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 other hand, have comparatively low mortality until 1950, at the same
-level as the middle class.
+level as the middle category.
 The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.
@@
 The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions.
+Instead, the elite category experienced the highest death risks.
 The basic pattern is primarily observed in the working-age population, while the differences are smaller or even non-existent in old age in accordance with the status-levelling hypothesis.
@@
 They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s.
-Similar results have also been found in Estonia [@jaadla17].
+Compared to previous studies, the present one shows a much clearer survival disadvantage of the elite, and not only the lack of social gradient.
+There are in fact some studies indicating a similar pattern.
 A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89].
-A recent study Eriksson and Dribe on the social pattern of mortality, based on 1880---1910 census data linked to death registers shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country.
+A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study in not unique, but rather seems to have characterised all parts of the country.
 These findings are contrary to what is usually assumed and what other studies have shown.
@@
 This ought to imply that people with less adequate resources are the most vulnerable in the earlier phases of the mortality transition.
-While the McKeown thesis has been refuted by several scholars [@szreter88, @Colgrove02], it has played a central role within epidemiology and social medicine.
+The results presented here prove that the most wealthy and privileged groups were not able to avoid the most common health risks in historical societies.
+This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition.
+While the McKeown thesis has been refuted by several scholars [@szreter88, @Colgrove02], it has played and is still playing a central role within epidemiology and social medicine.
 
@@
 In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc.
-However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
+However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality and that the different mechanisms all contribute to the advantage of higher social position.
 
@@
 The criticism of the arguments by McKeown forced them to address following question:  "If Colgrove is right about the McKeown thesis, social epidemiology is left with a gaping hole in its explanatory repertoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02].
-This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantees better survival chances.Others emphasize psycho-social factors, for example @mar04, who argues that higher status leads to better health regardless of where on the status ladder the individual is situated.
+This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantees better survival chances.
+
+Others emphasize psycho-social factors, for example @mar04, who argues that higher status leads to better health regardless of where on the status ladder the individual is situated.
 Psychosocial factors are also suggested by @wipi09 as an explanation for why equal societies perform better than unequal ones when it comes to health and other social conditions.
@@
 In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles.
-Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one'shealth later in life.
+Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one's health later in life.
 Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden.
@@
 Social stress can also be expressed in unhealthy lifestyles, the third pathway we discuss.
-Harmful behaviours can be connected to certain expectationsassociated with being a man of high status.
+Harmful behaviours can be connected to certain expectations associated with being a man of high status.
 Our results regarding the gender aspect are important in this respect.
@@
 Having a high social status is sometimes expressed in attitudes, norms and way of living that are potentially harmful to one's health.
-@edli11, finding similar results in another environment, suggest that an explanation for the larger social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.
-Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00].
-Different historical contexts promote different ways to spend economic resources [@rasp06; @jaadla17].
+@edli11, finding similar results in another environment, suggest that an explanation for the large social health inequalities in present-day society is that high status is expressed through health, being fit and physically active, while in many historical contexts higher social status (for men) was expressed in low physical activity, obsessive eating, drinking and smoking and being more inclined to take risks.
+Alcohol has been suggested to be part of the large gender differences in the 19th century [@wil99; @edv92], and a more risk-taking lifestyle may be an important component of male mortality [@cour00; @rasp06; @jaadla17].
+Different historical contexts promote different ways to spend economic resources.
 
@@
 Our study suggests a strong impact in historical Sweden as well.
-If this interpretation is correct, the hypothesis by @ant67, on an increasing role of health-related behaviour in the modern low-mortality regime while economic inequality determined the social mortality pattern during the early phases of the mortality transition, can be questioned.
+If this interpretation is correct, the hypothesis by @ant67, on an increasing role of health-related behaviour in the modern low-mortality regime while economic and material conditions determined the social mortality pattern during the early phases of the mortality transition, can be questioned.
 Instead, health-related inequality might be equally strong earlier in history but expressed in a completely different way.
@@
 The present study is unique in its time perspective, investigating the development from the early 19th century until the present, utilizing micro-data of high quality on a population of considerable size and social diversification.
-The resultswill likely surprise most readers; they partly surprised us as well.
+The results will likely surprise most readers; they partly surprised us as well.
 The consistent advantage of belonging to the elite has not always existed, and the modern pattern of social divide in health appeared much later than expected in the regions we studied.
@@
 The social pattern of female mortality is in line with what was expected, but a social gradient in male adult mortality is a recent phenomenon.
-It is an interesting paradox hat both the class and gender aspects in the analysis show that those with more power and resources were in fact less successful when it came to survival.
+It is an interesting paradox that both the class and gender aspects in the analysis show that those with more power and resources were in fact less successful when it came to survival.
 Furthermore, inequality in mortality appears clearer for the working-age population, while the differences are smaller among the elderly.

### 2017-11-22 article3.Rmd
endfloat

@@
     citation_package: natbib
-    keep_tex: yes
     number_sections: yes
@@
     toc_depth: 2
+    keep_tex: true
   bookdown::html_document2:
@@
 biblio-style: apalike
+header-includes:
+  - \usepackage{endfloat}
 ---
@@
 
-```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in", out.height = "4.1in", fig.align="center"}
+```{r mapp,fig.cap = "The Skelleftea and Umea regions in Sweden.", out.width="3in", out.height = "4.1in", fig.align="center"}
 knitr::include_graphics("myfigs/sweden.png")

### 2017-11-23 article3.Rmd
fixes

--- 2017-11-22 article3.Rmd
@@
 ---
-title: Is high social class always beneficial for survival? A study of Northern Sweden
-  1801--2013
+title: Is high social class always beneficial for survival? Northern Sweden 1801--2013
 author: "Sören Edvinsson and Göran Broström"
@@
     citation_package: natbib
-    keep_tex: yes
     number_sections: yes
-    toc: yes
+    toc: no
     toc_depth: 2
+    keep_tex: true
   bookdown::html_document2:
@@
     toc_depth: 2
+titlepage: yes
 bibliography: mybib.bib
 documentclass: article
+classoption: titlepage
 fontsize: 12pt
@@
 biblio-style: apalike
+header-includes:
+  - \usepackage{endfloat}
+  - \usepackage[utf8]{inputenc}
+abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class.
+The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality.
+The results show that high social class is not always favourable for survival.
+Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
+We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
+The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
 ---
@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=6.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 library(knitr)
@@
 
-```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in", out.height = "4.1in", fig.align="center"}
+```{r mapp,fig.cap = "The Skelleftea and Umea regions in Sweden.", out.width="3in", fig.height = 16.1, fig.align="center"}
 knitr::include_graphics("myfigs/sweden.png")
@@
 
-```{r hiscskumm, fig.cap = "Exposure by social class, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
+```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '15cm'}
 xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r hiscskumw, fig.cap = "Exposure by social class, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5}
+```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '15cm', fig.scap="Exposure by social class and time period, women."}
 xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.height = 4.5}
+```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exosure by marital status and time period.',out.height = '24.5cm'}
 xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))

### 2017-11-23 article3.Rmd
formalia

@@
 ---
-title: Is high social class always beneficial for survival? Northern Sweden 1801--2013
+title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
 author: "Sören Edvinsson and Göran Broström"
@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=6.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 library(knitr)
@@
 ```{r mapp,fig.cap = "The Skelleftea and Umea regions in Sweden.", out.width="3in", fig.height = 16.1, fig.align="center"}
-knitr::include_graphics("myfigs/sweden.png")
+knitr::include_graphics("myfigs/sweden.pdf")
 ```
@@
 cap0 <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
-scap0 <- "Probability by social class, ages 40--64."
-cap12 <- "Probability of dying before age 65 for a 40-year-old standardized by social class and period, women (top) and men (bottom)."
-scap12 <- "Probability standardized by social class, ages 40--64."
+scap0 <- "Probability of dying before age 65 for a 40-year old by social class."
+cap12 <- "Relative probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
+scap12 <- "Relative probability of dying before age 65 for a 40-year-old by social class."
 cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
@@
 cap22 <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). Standardised."
-scap22 <- "Probability by social class, ages 65--89."
+scap22 <- "Relative probability of dying before age 90 for a 65-year-old by social class."
 cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
@@
 
-```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8}
+```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
 source("R/plotED.glm.R")
@@
 
-```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8}
+```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm"}
 oldpar <- par(mfrow = c(2, 1))
@@
 
-```{r cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8}
+```{r cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.R")

### 2017-11-27 article3.Rmd
affiliation, million

@@
 title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
-author: "Sören Edvinsson and Göran Broström"
-date: '`r Sys.Date()`'
+author: Sören Edvinsson\footnote{Corresponding author} and Göran Broström
+date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
 output:
@@
 ##text(1899, 60, paste(pn.rec, "records"))
-text(1875, 55, paste(pn.exp, "p-years"))
-text(1874.5, 50, paste(pn.d, "deaths"))
+text(1875, 55, paste(round(pn.exp / 1000000, 1), "million p-years"))
+text(1874.5, 50, paste(round(pn.d / 1000, 1), "thousand deaths"))
 #
@@
 ##text(1984, 60, paste(n.rec, "records"))
-text(1995.2, 55, paste(n.exp, "p-years"), cex = 0.8)
-text(1995.2, 50, paste(n.d, "deaths"))
+text(1995.2, 55, paste(round(n.exp / 1000000, 1), "million p-years"), cex = 0.8)
+text(1995.2, 50, paste(round(n.d / 1000, 1), "thousand deaths"), cex = 0.7)
 ```

### 2017-11-27 abo.Rnw
a cosmetic change
reproduce

@@
 library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/ume-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/ume-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 @

### 2017-11-27 article3.Rmd
reproduce

@@
 title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
-author: Sören Edvinsson\footnote{Corresponding author} and Göran Broström
+author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
 date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'

### 2017-11-27 icass.Rnw
reproduce

@@
 library(knitr)
-opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
+opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
 library(knitr)

### 2017-11-28 Umea.Rnw
kolling
kolling...

@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 
@@
 
-\begin{frame}{Aims}
+\begin{frame}[fragile]{Aims}
 
@@
 \item age group (working age vs retired)
-\item *cause of death (cardiovascular, cancer)
-\item *education 1990--2013
-\item *income 1986-2005
+\item cause of death (cardiovascular, cancer)
+\item education 1990--2013
+\item income 1986--2005
 
@@
 
-\begin{frame}{The sampling frame}
+\begin{frame}[fragile]{The sampling frame}
 
@@
 
-\begin{frame}{Women and men, age 40--64}
+\begin{frame}[fragile]{Women and men, age 40--64}
 
@@
 
-\begin{frame}{Relative differences, ages 40--64 and 65--89}
+\begin{frame}[fragile]{Relative differences, ages 40--64 and 65--89}
 
   %\begin{center}
-\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswold-1}
+\includegraphics[height=7.5cm,width=5.4cm]{figs/art2-cumswstand-1}\includegraphics[height=7.5cm,width=5.4cm]{figs/slide-cumswold-1}
    %\end{center}
@@
 
-\begin{frame}{Income 1986--2005}
+\begin{frame}[fragile]{Income 1986--2005}
 

### 2017-11-28 slides.Rnw
kolling...

@@
 load("data/vb.rda")
-vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
+vbd <- vb[vb$event == 1, c("birthdate", "exit", "sex")]
 x <- vbd$birthdate + vbd$exit

### 2017-11-28 article2.Rmd
manipulating!

@@
 ---
-title: Life course and long-term perspectives of social inequality in mortality among
-  elderly and adults in Northern Sweden      1801–2013
+title: "Life course and long-term perspectives of social inequality in mortality among
+  elderly and adults in Northern Sweden 1801–2013"
 author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"

### 2017-11-28 article3.Rmd
manipulating!

@@
 vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
+vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
+                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
+                    FUN = sum)
 ```
@@
 ##
+
 cap0 <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
 scap0 <- "Probability of dying before age 65 for a 40-year old by social class."
+cap1 <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
+scap1 <- "Probability of dying before age 90 for a 65-year old by social class."
 cap12 <- "Relative probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
@@
 source("R/getPHglm2.R")
+source("R/getPHglm.R")
 ##nper <- 4
 nper <- length(labb)
-resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+resw <- getPHglm(vb4064[vb4064$sex == "female", ],
             nper, labb)
-resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+resm <- getPHglm(vb4064[vb4064$sex == "male", ],
             nper, labb)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.1, 0.55), main = "Women 40-64")##, exclude = 5:6)
+                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.1, 0.55), main = "Men 40-64")##, exclude = 5:6)
+                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
 par(oldpar)
@@
 
-```{r cumswold, fig.cap = cap22, fig.scap = scap22, fig.height = 8, out.height = "14cm"}
+```{r cumswold, fig.cap = cap1, fig.scap = scap1, fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.R")
-resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
+source("R/getPHglm.R")
+resw <- getPHglm(vb6589[vb6589$sex == "female", ],
             nper, labb)
-resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
+resm <- getPHglm(vb6589[vb6589$sex == "male", ],
             nper, labb)
@@
 oldpar <- par(mfrow = c(2, 1))
+woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = FALSE,
+                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.4, 1.0), main = "Women 65-89")##, exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = FALSE,
+                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.4, 1.0), main = "Men 65-89")##, exclude = 5:6)
+par(oldpar)
+```
+
+```{r cumsw2, fig.cap = cap22, fig.scap = scap22, fig.height = 8, out.height = "14cm"}
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
 woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
-                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.1, 0.55), main = "Women 65-89")##, exclude = 5:6)
+                    ylab = "P(dead at 90 | alive at 65)",
+                main = "Women 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.1, 0.55), main = "Men 65-89")##, exclude = 5:6)
+                    ylab = "P(dead at 90 | alive at 65)",
+                main = "Men 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
 par(oldpar)

### 2017-11-28 R/getPHglm2.R

@@ -13,7 +13,7 @@ getPHglm2 <- function(dat, nper, labb, from = 40){
     levs <- matrix(0, nrow = m, ncol = k + 1)
    
     for (i in 1:nper){
-        fit <- glm(event ~ offset(log(exposure)) + ##urban + 
+        fit <- glm(event ~ offset(log(exposure)) + urban + 
                        civst +
                        age * hisclass, 
                    data = dat[dat$period == labb[i], ], family = poisson)

### 2017-11-28 R/getPHglm.R
unknown change

@@ -1,6 +1,6 @@
-getPHglm <- function(dat, nper, labb, from = 40){
+getRR <- function(dat, nper, labb, from = 40){
     ## dat$age must be a factor ...
-    ## Note: return value is Cumulative hazards at given age points (straight line between).
+    ## Note: return value is relative risks from a "Cox regression"
     ##
     ## 'dat$age' is assumed to be grouped in equidistant intervals!
     res <- vector(nper, mode = "list")

### 2017-11-28 article4.Rmd
first commit

--- 2017-11-28 article3.Rmd
@@
                     FUN = sum)
+vb0Tab$hisclass <- relevel(vb0Tab$hisclass, ref = "middle")
 ```
@@
 ```{r hazcalw}
-source("R/getPHglm2.R")
-source("R/getPHglm.R")
+source("R/preg.R")
 ##nper <- 4
 nper <- length(labb)
-resw <- getPHglm(vb4064[vb4064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm(vb4064[vb4064$sex == "male", ],
-            nper, labb)
+resw <- matrix(0, nrow = 4, ncol = nper)
+resm <- matrix(0, nrow = 4, ncol = nper)
+for (i in 1:nper){
+  resw[, i] <- preg("+ civst", data = vb4064[vb4064$sex == "female" &
+                                               vb4064$period == labb[i], ])
+    resm[, i] <- preg("+ civst", data = vb4064[vb4064$sex == "male" &
+                                               vb4064$period == labb[i], ])
+}
+##resw <- getPHglm(vb4064[vb4064$sex == "female", ],
+##            nper, labb)
+##resm <- getPHglm(vb4064[vb4064$sex == "male", ],
+##            nper, labb)
 ```

### 2017-11-28 R/preg.R

@@ -1,0 +1,11 @@
+preg <- function(formula = "", data){
+  ## formula = "event ~ offset(log(exposure) + hisclass + age...)"
+  ## Argument: formula is a text string equal to ... above
+  f1 <- "event ~ offset(log(exposure)) + hisclass + age"
+  form <- as.formula(paste(f1, formula))
+  fit <- glm(form, data = data, family = poisson)
+  print(summary(fit)$coefficients)
+  y <- exp(c(0, fit$coefficients[2:4]))
+  names(y) <- c("middle", "elite", "worker", "NA")
+  y
+}

### 2017-11-29 article3.Rmd
editorial

@@
 
-Marital status (see Figure \@ref(fig:civstskum)), is treated as a time-dependent variable in the data
-file. The statuses used are unmarried, married and dissolved marriage
+Marital status (see Figure \@ref(fig:civstskum)) is a time-dependent variable. 
+The statuses used are unmarried, married and dissolved marriage
 (without distinguishing between widowed and divorced). If no explicit
@@
 living conditions, as well as the risk of importing infections, made the
-health risks higher in towns and cities. In the present paper, the variable **urban** controls
+health risks higher in towns and cities. In the present paper, the variable **urban** (values *TRUE* or *FALSE*) controls
 for this aspect by distinguishing between urban and rural residence for the periods after 1900, based on parish or place of residence.
@@
 
-cap0 <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
-scap0 <- "Probability of dying before age 65 for a 40-year old by social class."
-cap1 <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
-scap1 <- "Probability of dying before age 90 for a 65-year old by social class."
-cap12 <- "Relative probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
-scap12 <- "Relative probability of dying before age 65 for a 40-year-old by social class."
+cap0 <- "Cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
+scap0 <- "Cumulative hazard of dying before age 65 for a 40-year old by social class."
+cap1 <- "Cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
+scap1 <- "Cumulative hazard of dying before age 90 for a 65-year old by social class."
+cap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
+scap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class."
 cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
@@
 ##
-cap22 <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). Standardised."
-scap22 <- "Relative probability of dying before age 90 for a 65-year-old by social class."
+cap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
+scap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class."
 cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
@@
 nper <- length(labb)
-resw <- getPHglm(vb4064[vb4064$sex == "female", ],
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
             nper, labb)
-resm <- getPHglm(vb4064[vb4064$sex == "male", ],
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
             nper, labb)
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
-                main = "Women 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
-                main = "Men 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
+                main = "Women 40-64", ylim = c(0, 1))##, exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
+                main = "Men 40-64", ylim = c(0, 1))##, exclude = 5:6)
 par(oldpar)
@@
 ```{r cumswtab,cap="Table",results="asis"}
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
+vb4064$social_class <- vb4064$hisclass
+vb4064$marital_status <- vb4064$civst
+vb4064$urban_or_rural <- vb4064$urban
+vb4064$age_group <- vb4064$age
+fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
             data = vb4064, family = poisson)
@@
 ##knitr::kable(dr)
-print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: Anova, ages 40--64."), 
+print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
       type = "latex", comment = FALSE)
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "P(dead at 65 | alive at 40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
+                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
+                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
 par(oldpar)
@@
 ```{r cumswtabold,cap="Table",results="asis"}
-fittold <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
+vb6589$marital_status <- vb6589$civst
+vb6589$social_class <- vb6589$hisclass
+vb6589$urban_or_rural <- vb6589$urban
+vb6589$age_group <- vb6589$age
+fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
             data = vb6589, family = poisson)
@@
 ##knitr::kable(dr)
-print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: Anova, ages 65--89."), 
+print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--89."), 
       type = "latex", comment = FALSE)
@@
 
-The general pattern and development of mortality (not shown here) do not
+The general pattern and development of mortality (Figure \@ref(fig:cumswold)) do not
 differ much from the younger age group, of course representing a much
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status-levelling hypothesis (see Figure \@ref(fig:cumswold)). For women the results are mixed. We do not find the
+we find for the younger age group in line with the status-levelling hypothesis (see Figure \@ref(fig:cumsw2)). For women the results are mixed. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
@@
 source("R/getPHglm.R")
-resw <- getPHglm(vb6589[vb6589$sex == "female", ],
+resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
             nper, labb)
-resm <- getPHglm(vb6589[vb6589$sex == "male", ],
+resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
             nper, labb)
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = FALSE,
-                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.4, 1.0), main = "Women 65-89")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = FALSE,
-                    ylab = "P(dead at 90 | alive at 65)", ylim = c(0.4, 1.0), main = "Men 65-89")##, exclude = 5:6)
+woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = FALSE,
+                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Women 65-89")##, exclude = 5:6)
+moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = FALSE,
+                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Men 65-89")##, exclude = 5:6)
 par(oldpar)
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, relative = TRUE, 
-                    ylab = "P(dead at 90 | alive at 65)",
+woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE, 
+                    ylab = "H(90|65)",
                 main = "Women 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, relative = TRUE,
-                    ylab = "P(dead at 90 | alive at 65)",
+moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
+                    ylab = "H(90|65)",
                 main = "Men 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
@@
 ##knitr::kable(dr)
-xtable(dr,label = "tab:cumswoldtab", caption = "Anova, all causes of death, 65-89.")
+xtable(dr,label = "tab:cumswoldtab", caption = "ANOVA, all causes of death, 65-89.")
 ```

### 2017-11-29 article4.Rmd
editorial

@@
 the results are fairly complete.
-
+np
 ## Marital status
@@
 ## Colors for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
+scol <- c("blue", "darkgreen", "black", "red")
 ## Common for all groups:
@@
 source("R/preg.R")
+source("R/getPHglm2.R")
 ##nper <- 4
 nper <- length(labb)
-resw <- matrix(0, nrow = 4, ncol = nper)
-resm <- matrix(0, nrow = 4, ncol = nper)
+reswh <- matrix(0, nrow = 4, ncol = nper)
+resmh <- matrix(0, nrow = 4, ncol = nper)
 for (i in 1:nper){
-  resw[, i] <- preg("+ civst", data = vb4064[vb4064$sex == "female" &
+  reswh[, i] <- preg("+ civst + urban", data = vb4064[vb4064$sex == "female" &
                                                vb4064$period == labb[i], ])
-    resm[, i] <- preg("+ civst", data = vb4064[vb4064$sex == "male" &
-                                               vb4064$period == labb[i], ])
+    resmh[, i] <- preg("+ civst + urban", data = vb4064[vb4064$sex == "male" &
+                                               vb4064$period == labb[i], ], debug = FALSE)
 }
-##resw <- getPHglm(vb4064[vb4064$sex == "female", ],
-##            nper, labb)
-##resm <- getPHglm(vb4064[vb4064$sex == "male", ],
-##            nper, labb)
+resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+            nper, labb)
+resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+            nper, labb)
 ```
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
-                main = "Women 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
-                main = "Men 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
+##woutp <- plotED.glm(resw, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+  ##              main = "Women 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
+##moutp <- plotED.glm(resm, col = scol, logScale = FALSE, ylab = "P(dead at 65 | alive at 40)",
+  ##              main = "Men 40-64", ylim = c(0, 0.6))##, exclude = 5:6)
+xx <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1977.5, 1982.5, 1987.5, 1992.5, 1998.5, 2004.5, 2010.5)
+plot(xx, reswh[1, ], col = scol[1], type = "b", pch = 1, lty = 3, ylim = c(0.0, 4), main = "Women 40-64")
+for(i in 2:4){
+   lines(xx, reswh[i, ], col = scol[i], type = "b", lty = (i == 2) + (i != 2) * 3, pch = i)
+}
+plot(xx, resmh[1, ], col = scol[1], type = "b", pch = 1, lty = 3, ylim = c(0.0, 4), main = "Men 40-64")
+for(i in 2:4){
+   lines(xx, resmh[i, ], col = scol[i], type = "b", lty = (i == 2) + (i != 2) * 3, pch = i)
+}
 par(oldpar)

### 2017-11-29 R/getPHglm.R

@@ -1,4 +1,4 @@
-getRR <- function(dat, nper, labb, from = 40){
+getPHglm <- function(dat, nper, labb, from = 40){
     ## dat$age must be a factor ...
     ## Note: return value is relative risks from a "Cox regression"
     ##

### 2017-11-29 R/preg.R

@@ -1,11 +1,12 @@
-preg <- function(formula = "", data){
+preg <- function(formula = "", data, debug = FALSE){
   ## formula = "event ~ offset(log(exposure) + hisclass + age...)"
   ## Argument: formula is a text string equal to ... above
   f1 <- "event ~ offset(log(exposure)) + hisclass + age"
   form <- as.formula(paste(f1, formula))
   fit <- glm(form, data = data, family = poisson)
-  print(summary(fit)$coefficients)
+  if (debug) print(summary(fit)$coefficients)
   y <- exp(c(0, fit$coefficients[2:4]))
-  names(y) <- c("middle", "elite", "worker", "NA")
+  y <- exp(c(fit$coefficients[2], 0, fit$coefficients[3:4]))
+  names(y) <- c("elite", "middle", "worker", "NA")
   y
 }

### 2017-12-01 article3.Rmd
cosmetics

@@
 residence, birth date, death date, first and last date of observation and
-type of entrance/exit. The total number of person years is `r as.integer(pn.exp)` leading to
-`r pn.d` deaths in Poplink and `r as.integer(n.exp)` person years leading 
-to `r n.d` deaths in the Linnaeus database: see 
+type of entrance/exit. The total number of person years is `r round(pn.exp / 1000000, 2)` millions leading to
+`r round(pn.d /1000, 2)` thousand deaths in Poplink and `r round(n.exp / 1000000, 2)` million person years leading 
+to `r round(n.d / 1000, 2)` thousand deaths in the Linnaeus database: see 
 Figure \@ref(fig:lexis2).
@@
 time.
-In most periods the social differences were small, particularly between the workers and the middle group, and the elite even had somewhat higher mortality during the early 20th century.
+In most periods the social differences were small, particularly between the workers and the middle group, and the elite women even had somewhat higher mortality during some periods until the 190--1925 period.
 The social differences continued to be small until the last decades of the

### 2017-12-06 pdr.Rmd
first commit

--- 2017-12-01 article3.Rmd
@@
 title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
-author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
-date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
+author: X and Y
 output:
@@
 header-includes:
-  - \usepackage{endfloat}
+  - \usepackage[nolists]{endfloat}
   - \usepackage[utf8]{inputenc}

### 2017-12-06 article3.Rmd
Some smaller revisions

@@
 A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge common notions of this by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
 The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
@@
 Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
-The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
+The results are discussed in relation to the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
 
@@
 One of the central aspects of survival is social class and access to economic
-and other resources. In present-day welfare societies, social position is a
+and other resources. In present-day welfare societies, social position, either according to social class, education or income, is a
 strong determinant when it comes to health and mortality and its impact even
@@
 reasonable assumption since these societies were often characterized by
-enormous socioeconomic differences. Knowing that access to different kinds of resources
-provides advantages in all aspects of life, the health advantage of higher
-classes ought to be obvious.
+enormous class differences.
+Knowing that access to different kinds of resources, such as economic, social and cultural capital, provides advantages in all aspects of life, the health advantage of higher classes ought to be obvious.
 
@@
 studies from different geographical and historical settings in order to 
-better understand the role of socio-economic conditions in health and survival over time.
+better understand the role of class and socio-economic conditions in health and survival over time.
 
-Social conditions affect health and mortality differently during the life course.
+Social position affects health and mortality differently during the life course.
 There are different hypotheses about the impact in adulthood and old age respectively.
@@
 If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country.
-The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
-county of Västerbotten in northern of Sweden along the coast of the Gulf
+The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
 of Bothnia. This is a remote part of the country where communication with
@@
 The records are linked within, but not
-between, the regions. Data from Poplink are accessed for the period 1801--1950.
+between, the regions.
 
@@
 different periods are not completely comparable.
-Noti, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].
+Notice, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].
 

### 2017-12-06 article3.Rmd
Final revision before submission

@@
 In Poplink, occupations are coded according to the original DDB coding system that has then been adapted to Hisclass.
-For the Linnaeus data, HISCO is recoded into Hisclass.
-The censuses include occupational codes based on the system used in Sweden at the time, NYK (Nordisk
-Yrkesklassificering).
+For the Linnaeus data, the occupations are available in the coding scheme HISCO [@lema02] who then are recoded into Hisclass.
 We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods.
@@
 
-The use of different data sets forces us to apply different approaches
-to defining the social position. Poplink usually only provides the occupation of the head of household, thus underestimating female labour force participation as well as that of
-adult children residing with their parents [@vik10]. For the Poplink period, we have chosen to categorize wives according to the position of the head of household, usually
+The two different data sets represent different ways of reporting occupations.
+Poplink usually only provides the occupation of the head of household, thus underestimating female labour force participation as well as that of adult children residing with their parents [@vik10].
+For the Poplink period, we have chosen to categorize wives according to the position of the head of household, usually
 the husband/father, assuming that the family shares the same socio-economic
-position.
+position.This is a reasonable approach for this period.
 
-Female labour is much better covered in the Linnaeus database. A
-complication, however, is that it is  difficult to define households in the
-same way as in Poplink. Extramarital cohabitation became common and female
+Female labour is much better covered in the Linnaeus database, and it is difficult to define households in the same way as in Poplink. Extramarital cohabitation became common and female
 labour force participation developed as the norm, in Sweden during this
@@
 if not obsolete, at least less relevant. All included persons are signified
-by their individual occupation for this period. Still, the results for the
-different periods are not completely comparable.
+by their individual occupation for this period.
+Although the results for the different periods are not completely comparable, the difference in approach reflects an actual change in how social class is structured.
 Notice, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].
@@
 supervision, and sector. Implicitly it reflects large differences in access to
-economic and other resources, status, social power, etc.
-Longitudinal analyses of social classes for such a long period covered in this study present several problems, related to changes in occupational structure and the meaning and content of occupations and social positions.
-We have chosen to work with a broad definition of classes, merging the original 12 classes in Hisclass into the following three and an additional fourth category of unknown class:
+economic resources, status, social power, etc.
+We have chosen to work with a broad definition of classes, merging the original 12 social classes in Hisclass into three, representing different levels of control and access to resources vital for life chances, and an additional fourth category of unknown class:
 
@@
 in the 19th century. The increase in academics in connection to the
-establishment of Umeå University explains much of this increase. What is
+establishment of Umeå University explains much of this increase in later time. What is
 not apparent in Figure \@ref(fig:hiscskumm) is the transfer within the middle
@@
 from such studies show other patterns than that for infant and child
-mortality. While some report substantial differences, many -- particularly those based on
-Swedish data -- have found surprisingly small or even reversed socio-economic
-differences during the 19th and early 20th centuries, sharply contrasting from the situation in present-day society. The expected social gradient in survival -- meaning that people
-from higher social classes always had an advantage -- was in many cases,
-particularly among adult men, not found despite the well-documented
-socio-economic differences in society, for example in the Sundsvall region [@edli11; @edbr12].
+mortality.
+While some report substantial differences, others have found surprisingly small or even reversed class differences, for example in the 19th century Sundsvall region [@edli11; @edbr12].
 This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century.
@@
 The results indicate that we should consider the role of gender and class expectations of behaviour.
-Unfortunately, the evidence for behavioural differences in history is usually circumstantial and anecdotal, at least when it comes to social class.
 Studies of these aspects and their role in health and mortality in different times and places require further development.

### 2017-12-07 pdr.Rmd
remove abstract and acknowledgements

@@
   - \usepackage[utf8]{inputenc}
-abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class.
-The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality.
-The results show that high social class is not always favourable for survival.
-Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
-We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
-The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
 ---
@@
 
-# Acknowledgements
-
-This work was undertaken within the research programme *Ageing and
-Living Conditions* at the Centre for Demographic and Ageing Research
-(CEDAR) at Umeå University, financed by the Swedish Research Council (grant
-No 349-2008-6592) and the project *Old age, health and inequality:
-Mortality from a life course perspective in Sweden 1986--2009*, financed by
-the Swedish Research Council (grant No 421-2012-892). We thank the
-Demographic Data Base, Umeå University, for giving us access to data and
-Jonas Helgertz at the Department of Economic History, Lund University, for
-providing us with scripts for creating Hisclass from Swedish census
-data.
-
 # References

### 2018-01-13 article3.Rmd
fix for html version

@@
 output:
+  bookdown::html_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
+    toc_float: yes
   bookdown::pdf_document2:
@@
     keep_tex: true
-  bookdown::html_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
-    toc_float: yes
   bookdown::word_document2:
@@
 library(skum)
+options(show.signif.stars = FALSE)
 ```
@@
 
-```{r mapp,fig.cap = "The Skelleftea and Umea regions in Sweden.", out.width="3in", fig.height = 16.1, fig.align="center"}
-knitr::include_graphics("myfigs/sweden.pdf")
+(ref:cities) The Skellefteå and Umeå regions in Sweden.
+
+```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height='50%'}
+knitr::include_graphics("myfigs/sweden.png")
 ```
@@
 
-```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '15cm'}
+```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
 xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '15cm', fig.scap="Exposure by social class and time period, women."}
+```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '50%', fig.scap="Exposure by social class and time period, women.",fig.height=4}
 xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exosure by marital status and time period.',out.height = '24.5cm'}
+```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = '50%',fig.height=4}
 xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))
@@
 \begin{split}
-\mathbf{x} &= (x_1, \ldots, x_n) \text{\ is a vector of covariates,} \\
-\boldsymbol{\beta} &= (\beta_1, \ldots, \beta_n) \text{\ is the corresponding vector of regression
+\mathbf{x} &= (x_1, \ldots, x_n) \text{ is a vector of covariates,} \\
+\boldsymbol{\beta} &= (\beta_1, \ldots, \beta_n) \text{ is the corresponding vector of regression
 parameters.} \\
@@
 Since all components of $\mathbf{x}$ are categorical covariates, the
-estimation of \@ref(eq:prop) can be performed in two ways; based on the
+estimation of the model can be performed in two ways; based on the
 original data set and a survival analysis routine handling a piecewise
@@
 the point is that the two ways give exactly the same results. This is
-because the table is a \emph{sufficient statistic} for \@ref(eq:prop) and
+because the table is a \emph{sufficient statistic} for the model and
 the original data set.
@@
 
-# The effect of social class on mortality, 1801--1950, 1976--2013
+# The effect of social class on mortality
 
@@
 
-Looking at Table \@ref(tab:cumswtab) it is obvious that the statistical significance is strong.
-The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
-This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
+Looking at the table below, it is obvious that the statistical significance is strong.
 
-```{r cumswtab,cap="Table",results="asis"}
+```{r cumswtab,cap="Table"}
 vb4064$social_class <- vb4064$hisclass
@@
 dr <- drop1(fitt, test = "Chisq")
+dr
 ##knitr::kable(dr)
-print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
-      type = "latex", comment = FALSE)
+##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
+  ##    type = "latex", comment = FALSE)
 ```
+
+The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
+This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
 
@@
 Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure. Looking at Table \@ref(tab:cumswtabold) it is obvious that the statistical significance is still strong.
+duration 25 (age 90), starting from age 65 as a summary measure. Looking at the table below, it is obvious that the statistical significance is still strong.
 
-```{r cumswtabold,cap="Table",results="asis"}
+```{r cumswtabold,cap="Table"}
 vb6589$marital_status <- vb6589$civst
@@
 ##knitr::kable(dr)
-print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--89."), 
-      type = "latex", comment = FALSE)
+drold
+##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--89."), 
+  ##    type = "latex", comment = FALSE)
 ```
@@
 This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition.
-While the McKeown thesis has been refuted by several scholars [@szreter88, @Colgrove02], it has played and is still playing a central role within epidemiology and social medicine.
+While the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it has played and is still playing a central role within epidemiology and social medicine.
 

### 2018-01-20 ols.Rmd
first commit

---
title: "Sampling deaths and OLS"
author: "Göran Broström"
date: "January 19, 2018"
output: html_document
---

## Introduction

We check the effect of assuming normally distributed remaining life after age 60.

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(engine='R',fig.path='figs/ols-',tidy=FALSE,fig.height=4,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
options(show.signif.stars = FALSE)
```

## Data

Data come from Skellefteå and Umeå. The selection is all persons born between January 1, 1841 and December 31, 1850 and present in 
the data at age 60.
They are followed from age 60 to death or right censoring (at most until December 31, 1950).

```{r readalldata}
source("R/get.file.R")
skume <- get.file("skume")
## Not needed(?):
##vb0Tab <- skume
##vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
##vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
##                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
##                    FUN = sum)
## End: Not needed(?)
ols <- obs[, c("id", "sex", "birthdate", "region", "civst", "hisclass", "sluttyp", "enter", "exit")]
ols$event <- ols$sluttyp == 2
ols$sluttyp <- NULL
ols <- ols[ols$birthdate >= 1841 & ols$birthdate < 1851, ]
ols <- age.window(ols, c(60, 100))
ids <- ols$id[ols$enter < 60.01]
ols <- ols[ols$id %in% ids, ]
indx <- tapply(ols$id, ols$id)
out <- tapply(ols$exit, ols$id, max)
ols$out <- out[indx]
ols <- ols[order(ols$id, -ols$exit, -ols$event),]
##sum(ols$event)
ols <- ols[!duplicated(ols$id), ]
##sum(ols$event)
ols$exit <- ols$out
ols$out <- NULL
summary(ols)
```

We want to observe time since age 60 (exact!), so we subtract 60 from the ages `enter` and `exit`.

```{r}
ols$enter <- ols$enter - 60
ols$exit <- ols$exit - 60
zero <- ols$enter >= ols$exit
ols$enter[zero] <- ols$exit[zero] - 0.001 # Fix zero-length intervals (no big deal)
```

In the following, we are interested in the *sex difference* in expected life after 60 (just as an illustration). The categorical variable `region` is included as an example of a confounder.

## Standard survival analyses

### Ordinary *Cox regression*

We first try to get a picture of what really is happening in mortality relative to `sex` and `region` (the figure is truncated at duration 30, i.e., age 90).

```{r standards}
fit <- coxph(Surv(enter, exit, event) ~ strata(region, sex), data = ols)
plot(survfit(fit), fun = "cumhaz", col = 1:4, lty = 1:4, xlab = "Years after 60.", ylab = "Cumulative hazards", xlim = c(0, 30))
legend("topleft", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
abline(h = 0)
```

It seems as if males have the same mortality experience in Umeå and Skellefteå, while women from Umeå are better off than those from Skellefteå. Of course, women are much better off than men, reardless of region. So, there is a tendency of *interaction* between the effects on mortality of `sex` and `region`.

### AFT regression

Then an *accelerated failure time (AFT)* model ("linear regression on the log scale, with censoring").

```{r afts}
fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = ols, dist = "lognormal", param = "lifeExp")
meanlog <- fit$coefficients["log(scale)"]
sdlog <- 1 / exp(fit$coefficients["log(shape)"])
fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
summary(fit)
```

The lognormal baseline plot looks like this:

```{r plotlns}
plot(fit, fn = "haz", xlab = "Years after 60.")
```

Not very convincing: The mortality is *decreasing* after age 70 (forced by the lognormal assumption)!

We try the *Gompertz* distribution instead:

```{r aftgomps}
fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = ols, dist = "gompertz", param = "lifeExp")

summary(fit.g)
```

The Gompertz baseline plot looks like this:

```{r plotgomps}
plot(fit.g, fn = "haz", xlab = "Years after 60.")
```

## Sampling deaths

We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes.

```{r smallols}
ids <- ols$id[ols$event == 1]
sols <- ols[ols$id %in% ids, ]
```

### Ordinary *Cox regression*

```{r standard}
fit <- coxreg(Surv(enter, exit, event) ~ sex + region, data = sols)
summary(fit)
```

### AFT regression

Then an *accelerated failure time (AFT)* model ("linear regression on the log scale, with censoring").

```{r aft}
fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "lognormal", param = "lifeExp")
meanlog <- fit$coefficients["log(scale)"]
sdlog <- 1 / exp(fit$coefficients["log(shape)"])
fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
summary(fit)
```

The lognormal baseline plot looks like this:

```{r plotln}
plot(fit, fn = "haz", xlab = "Years after 60.")
```

Not very convincing: The mortality is *decreasing* after age 70 (forced by the lognormal assumption)!

We try the *Gompertz* distribution instead:

```{r aftgomp}
fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "gompertz", param = "lifeExp")

summary(fit.g)
```

The Gompertz baseline plot looks like this:

```{r plotgomp}
plot(fit.g, fn = "haz", xlab = "Years after 60.")
```

## OLS regression

Finally, let us perform ordinary OLS regression.

```{r ordOLS}
ools <- sols[sols$event == 1, ]
summary(fit <- lm(exit ~ sex + region, data = ools)) 
```

### 2018-01-20 ols.Rmd
revision

@@
 date: "January 19, 2018"
-output: html_document
+output:
+  bookdown::html_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
+    toc_float: yes
+  bookdown::pdf_document2:
+    citation_package: natbib
+    number_sections: yes
+    toc: no
+    toc_depth: 2
+    keep_tex: true
+
 ---
 
-## Introduction
+# Introduction
 
@@
 
-## Data
+# Data
 
@@
 ```{r readalldata}
-source("R/get.file.R")
-skume <- get.file("skume")
+##source("R/get.file.R")
+##skume <- get.file("skume")
 ## Not needed(?):
@@
 ## End: Not needed(?)
-ols <- obs[, c("id", "sex", "birthdate", "region", "civst", "hisclass", "sluttyp", "enter", "exit")]
+ols <- obs[, c("id", "sex", "birthdate", "region", "civst", "sluttyp", "enter", "exit")]
 ols$event <- ols$sluttyp == 2
@@
 
-In the following, we are interested in the *sex difference* in expected life after 60 (just as an illustration). The categorical variable `region` is included as an example of a confounder.
+In the following, we are interested in the *sex difference* in expected remaining life after 60 (*e60*) (just as an illustration).
+The categorical variable `region` is included as an example of a confounder.
 
-## Standard survival analyses
+# Standard survival analyses
 
-### Ordinary *Cox regression*
+## Ordinary *Cox regression*
 
@@
 
-### AFT regression
+We cannot easily calculate *e60* from this figure (or table above), but the estimated median remaining life at 60 (*m60*) is easy to illustrate graphically.
+For that porpose we need to plot the estimated *survival functions*.
+```{r }
+plot(survfit(fit), col = 1:4, lty = 1:4, xlab = "Years after 60.", 
+     ylab = "Remaining fraction", xlim = c(0, 30), axes = FALSE)
+axis(1, at = c(0, 9, 13, 20, 30))
+axis(2, at = c(0, 0.5, 1), las = 1)
+box()
+legend("bottomleft", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
+abline(h = 0)
+abline(h = 0.5, lty = 2)
+```
+
+We can also make an exact calculation, which results in
+
+```{r exactki}
+x <- survfit(fit)
+x
+```
+
+## AFT regression
 
@@
 ```
+
+The *e60* for men in Skellefteå is 9.69, and for Skellefteå women it is $1.329 \times 9.69 = 12.88$, a difference of 3.19 years.
 
@@
 
+The *e60* for men in Skellefteå is now 11.2, and for Skellefteå women it is $1.075 \times 11.2 = 12.04$, a difference of only 0.84 years.
+
 The Gompertz baseline plot looks like this:
@@
 
-## Sampling deaths
+A lot more reasonable.
+
+# Sampling deaths
 
@@
 
-### Ordinary *Cox regression*
+## Ordinary *Cox regression*
 
 ```{r standard}
-fit <- coxreg(Surv(enter, exit, event) ~ sex + region, data = sols)
-summary(fit)
+fit <- coxph(Surv(enter, exit, event) ~ sex + region, data = sols)
+survfit(fit)
 ```
 
-### AFT regression
+## AFT regression
 
@@
 
-## OLS regression
+# OLS regression
 

### 2018-01-21 ols.Rmd
change

@@
 
-We check the effect of assuming normally distributed remaining life after age 60.
+We check the effect of studying mortality of the dead and whether it is reasonable to assume normally distributed remaining life after age 60.
 
@@
 
-# Standard survival analyses
+# Survival analysis with OLS
+
+## Full data
+
+*OLS* means assuming *normally distributed* remaining life after 60, and we can use standard linear regression.
+In the case of right censoring, we need to fall back on maximum likelihood for the estimation.
+The function *survreg* in the **R** package **survival** facilitates this.
+
+```{r survrcens}
+fit <- survreg(Surv(exit, event) ~ sex + region, data = ols, dist = "gaussian")
+x <- summary(fit)
+x
+```
+
+So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` and for women it is `r round(x$coef[2], 2)` more years.
+
+# Sampling deaths
+
+We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes.
+
+```{r smallols}
+ids <- ols$id[ols$event == 1]
+sols <- ols[ols$id %in% ids, ]
+```
 
 ## Ordinary *Cox regression*
+
+```{r standard}
+fit <- coxph(Surv(enter, exit, event) ~ sex + region, data = sols)
+survfit(fit)
+```
+
+## AFT regression
+
+Then an *accelerated failure time (AFT)* model ("linear regression on the log scale, with censoring").
+
+```{r aft}
+fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "lognormal", param = "lifeExp")
+meanlog <- fit$coefficients["log(scale)"]
+sdlog <- 1 / exp(fit$coefficients["log(shape)"])
+fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
+summary(fit)
+```
+
+The lognormal baseline plot looks like this:
+
+```{r plotln}
+plot(fit, fn = "haz", xlab = "Years after 60.")
+```
+
+Not very convincing: The mortality is *decreasing* after age 70 (forced by the lognormal assumption)!
+
+We try the *Gompertz* distribution instead:
+
+```{r aftgomp}
+fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "gompertz", param = "lifeExp")
+
+summary(fit.g)
+```
+
+The Gompertz baseline plot looks like this:
+
+```{r plotgomp}
+plot(fit.g, fn = "haz", xlab = "Years after 60.")
+```
+
+# OLS regression
+
+Finally, let us perform ordinary OLS regression.
+
+```{r ordOLS}
+ools <- sols[sols$event == 1, ]
+summary(fit <- lm(exit ~ sex + region, data = ools)) 
+```
 
@@
 A lot more reasonable.
-
-# Sampling deaths
-
-We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes.
-
-```{r smallols}
-ids <- ols$id[ols$event == 1]
-sols <- ols[ols$id %in% ids, ]
-```
-
-## Ordinary *Cox regression*
-
-```{r standard}
-fit <- coxph(Surv(enter, exit, event) ~ sex + region, data = sols)
-survfit(fit)
-```
-
-## AFT regression
-
-Then an *accelerated failure time (AFT)* model ("linear regression on the log scale, with censoring").
-
-```{r aft}
-fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "lognormal", param = "lifeExp")
-meanlog <- fit$coefficients["log(scale)"]
-sdlog <- 1 / exp(fit$coefficients["log(shape)"])
-fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
-summary(fit)
-```
-
-The lognormal baseline plot looks like this:
-
-```{r plotln}
-plot(fit, fn = "haz", xlab = "Years after 60.")
-```
-
-Not very convincing: The mortality is *decreasing* after age 70 (forced by the lognormal assumption)!
-
-We try the *Gompertz* distribution instead:
-
-```{r aftgomp}
-fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "gompertz", param = "lifeExp")
-
-summary(fit.g)
-```
-
-The Gompertz baseline plot looks like this:
-
-```{r plotgomp}
-plot(fit.g, fn = "haz", xlab = "Years after 60.")
-```
-
-# OLS regression
-
-Finally, let us perform ordinary OLS regression.
-
-```{r ordOLS}
-ools <- sols[sols$event == 1, ]
-summary(fit <- lm(exit ~ sex + region, data = ools)) 
-```

### 2018-01-21 ols.Rmd
go on

@@
 
-So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` and for women it is `r round(x$coef[2], 2)` more years.
+So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` years and for women it is `r round(x$coef[2], 2)` more years.
 
-# Sampling deaths
+## Sampling the dead
 
-We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes.
+We repeat the analyses with the data subset consisting of all those who were observerd to die in the parishes. This implies that 13 percent censored individuals are thrown away.
 
-```{r smallols}
+```{r smallols, echo = FALSE}
 ids <- ols$id[ols$event == 1]
@@
 
-## Ordinary *Cox regression*
-
-```{r standard}
-fit <- coxph(Surv(enter, exit, event) ~ sex + region, data = sols)
-survfit(fit)
+```{r survrnocens}
+fit2 <- survreg(Surv(exit, event) ~ sex + region, data = sols, dist = "gaussian")
+x2 <- summary(fit2)
+x2
 ```
 
-## AFT regression
+So the *e60* for men from Skellefteå is now `r round(x2$coef[1], 2)` years and for women it is `r round(x2$coef[2], 2)` more years.
 
-Then an *accelerated failure time (AFT)* model ("linear regression on the log scale, with censoring").
+## Conclusion
 
-```{r aft}
-fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "lognormal", param = "lifeExp")
-meanlog <- fit$coefficients["log(scale)"]
-sdlog <- 1 / exp(fit$coefficients["log(shape)"])
-fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
-summary(fit)
-```
+Sampling of deaths leads to overestimation of mortality (underestimation of *e60*): This is a trivial (well-known) consequence.
+More seriously, it distorts the balance between categories (`sex`, `region`) in unpredictable ways.
 
-The lognormal baseline plot looks like this:
+# AFT regression
 
-```{r plotln}
-plot(fit, fn = "haz", xlab = "Years after 60.")
-```
+The OLS model is an example of what is called  an *accelerated failure time (AFT)* model in survival analysis, a competitor to the *proportional hazards (PH)* model. 
+A common assumption about old age mortality is that a *Gompertz* distribution fits well.
 
-Not very convincing: The mortality is *decreasing* after age 70 (forced by the lognormal assumption)!
+So we go back to the full data situation and fit a Gompertz AFT model. The *survreg* function does not include the Gompertz distribution, but the *aftreg* function in the package **eha** does.
 
-We try the *Gompertz* distribution instead:
-
-```{r aftgomp}
-fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = sols, dist = "gompertz", param = "lifeExp")
-
+```{r aft, echo = FALSE}
+fit.g <- aftreg(Surv(exit, event) ~ sex, data = ols, dist = "gompertz", param = "lifeExp")
 summary(fit.g)
@@
 
-The Gompertz baseline plot looks like this:
-
-```{r plotgomp}
+```{r plotln}
 plot(fit.g, fn = "haz", xlab = "Years after 60.")
 ```
-
-# OLS regression
-
-Finally, let us perform ordinary OLS regression.
-
-```{r ordOLS}
-ools <- sols[sols$event == 1, ]
-summary(fit <- lm(exit ~ sex + region, data = ools)) 
-```
-
-We first try to get a picture of what really is happening in mortality relative to `sex` and `region` (the figure is truncated at duration 30, i.e., age 90).
 

### 2018-01-21 ols.Rmd
aft

@@
 ols$enter[zero] <- ols$exit[zero] - 0.001 # Fix zero-length intervals (no big deal)
+##ols$event[ols$exit > 30] <- 1
 ```
@@
 ```{r aft, echo = FALSE}
-fit.g <- aftreg(Surv(exit, event) ~ sex, data = ols, dist = "gompertz", param = "lifeExp")
+fit.g <- aftreg(Surv(exit, event) ~ sex + region, data = ols, dist = "gompertz", param = "lifeExp")
 summary(fit.g)
+em <- fit.g$baselineMean
+ef <- fit.g$baselineMean * exp(fit.g$coef[1])
 ```
+So the *e60* for Skellefteå men is `r round(em, 2)` years, and for the women we get 
+`r round(ef, 2)` years, a diff of `r round(ef - em, 2)` years, that is, *men* are better off.
+
+Compare with the result of the OLS regression:
 

### 2018-01-21 ols.Rmd
add three references

@@
 ---
-title: "Sampling deaths and OLS"
+title: "On mortality of the dead and OLS"
 author: "Göran Broström"
-date: "January 19, 2018"
+date: '`r Sys.Date()`'
 output:
@@
     keep_tex: true
-
+bibliography: mybib.bib
 ---
@@
 
-We check the effect of studying mortality of the dead and whether it is reasonable to assume normally distributed remaining life after age 60.
+We ask if studying mortality by sampling the dead is a good idea and whether it is reasonable to assume normally distributed remaining life after age 60. We answer *No* and *Maybe*.
 
@@
 the data at age 60.
-They are followed from age 60 to death or right censoring (at most until December 31, 1950).
+They are followed from age 60 to death or right censoring (at most until December 31, 1950). The first six rows of the data frame:
 
@@
 ols$out <- NULL
-summary(ols)
+head(ols)
 ```
@@
 
-# Survival analysis with OLS
+# OLS regression
+
+OLS regression in survival analysis is a special case of the AFT regression model, see the next section.
 
@@
 
-*OLS* means assuming *normally distributed* remaining life after 60, and we can use standard linear regression.
+*OLS* means assuming *normally distributed* remaining life after 60, and we can use standard linear regression, given that we are willing to assume an *accelerated failure time (AFT)* model for our data.
 In the case of right censoring, we need to fall back on maximum likelihood for the estimation.
-The function *survreg* in the **R** package **survival** facilitates this.
+The function *survreg* in the **R** [@cran] package **survival** facilitates this.
 
@@
 
-So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` years and for women it is `r round(x$coef[2], 2)` more years.
+So the *e60* for men from Skellefteå is `r round(x$coef[1], 2)` years and for women it is `r round(x$coef[2], 2)` more years. The difference is statistically significant.
 
@@
 
-So the *e60* for men from Skellefteå is now `r round(x2$coef[1], 2)` years and for women it is `r round(x2$coef[2], 2)` more years.
+So the *e60* for men from Skellefteå is now only `r round(x2$coef[1], 2)` years and for women it is `r round(x2$coef[2], 2)` more years. The difference is statistically significant, 
+and larger than in the case with full data.
 
@@
 Sampling of deaths leads to overestimation of mortality (underestimation of *e60*): This is a trivial (well-known) consequence.
-More seriously, it distorts the balance between categories (`sex`, `region`) in unpredictable ways.
+More seriously, it distorts the balance between the effects of categories (`sex`, `region`) on mortality in unpredictable ways.
 
@@
 
-So we go back to the full data situation and fit a Gompertz AFT model. The *survreg* function does not include the Gompertz distribution, but the *aftreg* function in the package **eha** does.
+So we go back to the full data situation and fit a Gompertz AFT model. The *survreg* function does not include the Gompertz distribution, but the *aftreg* function in the package **eha** [@eha] does.
 
@@
 So the *e60* for Skellefteå men is `r round(em, 2)` years, and for the women we get 
-`r round(ef, 2)` years, a diff of `r round(ef - em, 2)` years, that is, *men* are better off.
+`r round(ef, 2)` years, a difference of `r round(ef - em, 2)` years, that is, *men* are better off. This difference is *not* statistically significant.
 
-Compare with the result of the OLS regression:
+## Is the assumption of normally distributed residual life reasonable?
+
+Yes, as long as condition on survival past the mode of the survival distribution, according to @kann96. So what is the mode in the life distribution in our data?
+
+```{r mode, echo = FALSE, cache = TRUE}
+oj <- obs
+oj$event <- oj$sluttyp == 2
+oj <- age.window(oj, c(0, 100))
+as <- oj$enter >= oj$exit
+oj$enter[as] <- oj$exit[as] - 0.0001
+ff <- phreg(Surv(enter, exit, event) ~ 1, data = oj, dist = "pch", cuts = 1:99)
+par(las = 1)
+plot(ff, fn = "den", xlab = "Age", main = "", col = "blue")
+```
+
+Well, it is closer to 80 than to 60, so the normal distribution assumption may be in doubt. We can compare the baseline (men in Skellefteå) distributions from 
+the OLS fit and nonparametric estimation.
 
 ```{r plotln}
-plot(fit.g, fn = "haz", xlab = "Years after 60.")
+fit.cox <- coxreg(Surv(exit, event) ~ 1, data = ols[ols$sex == "male" & ols$region == "ske", ])
+par(las = 1)
+plot(fit.cox, fn = "cum", xlab = "Years after 60.", main = "Baseline Cumulative hazards  functions", ylim = c(0, 5))
+x <- seq(0, 37, length = 1000)
+y <- -log(pnorm(x, mean = 16.474, sd = 8.96, lower.tail = FALSE))
+lines(x, y, col = "red")
+legend("topleft", legend = c("Nonparametric", "OLS regression"), lty = c(1, 1), col = c("black", "red"))
+abline(h = 0)
 ```
+
+Looks quite ok, even though the OLS curve heavily overestimates the ages above 90. No simple parametric model will catch the heavy slow-down of the mortality increase in the high ages.
+
+# PH regression
+
+Maybe the PH model fits better than the AFT? Compare max log likelihoods.
+
+```{r phbetter}
+fit.ph <- phreg(Surv(exit, event) ~ sex + region, data = ols, dist = "gompertz")
+fit.ph
+```
+
+No, the difference is very small: `r round(fit.ph$loglik[2] - fit.g$loglik[2], 3)` in favor of the PH model. Not much to argue about.
+
+# A look at data
+
+These are the mortality patterns for the four groups we are contemplating. Worth noting is that male mortality above age 90 is so low (compared to women).
 
@@
 fit <- coxph(Surv(enter, exit, event) ~ strata(region, sex), data = ols)
-plot(survfit(fit), fun = "cumhaz", col = 1:4, lty = 1:4, xlab = "Years after 60.", ylab = "Cumulative hazards", xlim = c(0, 30))
+par(las = 1)
+plot(survfit(fit), fun = "cumhaz", col = 1:4, lty = 1:4, xlab = "Years after 60.", ylab = "Cum. hazards", axes = FALSE)
+axis(1, at = c(0, 9, 13, 20, 30, 40))
+axis(2, at = c(0, log(2), 3, 5), labels = c("0", "log(2)", "3", "5"))
+box()
 legend("topleft", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
 abline(h = 0)
+abline(h = log(2), lty = 2)
 ```
@@
 It seems as if males have the same mortality experience in Umeå and Skellefteå, while women from Umeå are better off than those from Skellefteå.
-Of course, women are much better off than men, reardless of region.
+Of course, women are much better off than men (before age 90!), reardless of region.
 So, there is a tendency of *interaction* between the effects on mortality of `sex` and `region`.
 
-We cannot easily calculate *e60* from this figure (or table above), but the estimated median remaining life at 60 (*m60*) is easy to illustrate graphically.
-For that porpose we need to plot the estimated *survival functions*.
+We cannot easily calculate *e60* from this figure (or table above), but the estimated *median remaining life at 60 (m60)* is easy to illustrate: It is the years-after-60 at which the horizontal line at $\log(2)$ intersects the curves: 9 years for men and 13 for women.
+Maybe easier to see in the plot of the estimated *survival functions* below.
 ```{r }
 plot(survfit(fit), col = 1:4, lty = 1:4, xlab = "Years after 60.", 
-     ylab = "Remaining fraction", xlim = c(0, 30), axes = FALSE)
+     ylab = "Remaining fraction", axes = FALSE)
 axis(1, at = c(0, 9, 13, 20, 30))
@@
 box()
-legend("bottomleft", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
+legend("topright", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
 abline(h = 0)
@@
 
-## AFT regression
+# Conclusion
 
-Then an *accelerated failure time (AFT)* model ("linear regression on the log scale, with censoring").
+We can live with the assumption of normally distributed residual life after age 60, but we must be aware of the fact that applying OLS implies rather strong distributional assumptions.
+It is a misconception to believe that least squares methods are "distribution free".
+Sampling of the dead, on the other hand, is problematic and should be avoided without measures that are specifically designed to take care of that situation (what ever they are).
 
-```{r afts}
-fit <- aftreg(Surv(enter, exit, event) ~ sex + region, data = ols, dist = "lognormal", param = "lifeExp")
-meanlog <- fit$coefficients["log(scale)"]
-sdlog <- 1 / exp(fit$coefficients["log(shape)"])
-fit$baselineMean <- exp(meanlog + sdlog^2 / 2)
-summary(fit)
-```
-
-The *e60* for men in Skellefteå is 9.69, and for Skellefteå women it is $1.329 \times 9.69 = 12.88$, a difference of 3.19 years.
-
-The lognormal baseline plot looks like this:
-
-```{r plotlns}
-plot(fit, fn = "haz", xlab = "Years after 60.")
-```
-
-Not very convincing: The mortality is *decreasing* after age 70 (forced by the lognormal assumption)!
-
-We try the *Gompertz* distribution instead:
-
-```{r aftgomps}
-fit.g <- aftreg(Surv(enter, exit, event) ~ sex + region, data = ols, dist = "gompertz", param = "lifeExp")
-
-summary(fit.g)
-```
-
-The *e60* for men in Skellefteå is now 11.2, and for Skellefteå women it is $1.075 \times 11.2 = 12.04$, a difference of only 0.84 years.
-
-The Gompertz baseline plot looks like this:
-
-```{r plotgomps}
-plot(fit.g, fn = "haz", xlab = "Years after 60.")
-```
-
-A lot more reasonable.
+# References

### 2018-01-22 ols.Rmd
add plot for the dead

@@
 
+## Full data
+
 These are the mortality patterns for the four groups we are contemplating. Worth noting is that male mortality above age 90 is so low (compared to women).
 
-```{r standards}
+```{r standards, fig.cap = "Full data."}
 fit <- coxph(Surv(enter, exit, event) ~ strata(region, sex), data = ols)
@@
 
+## Mortality of the dead
+
+Median life becomes about two years shorter for the "dead". Furthermore, mortality continue to rise after age 90, in contrast to
+what happens with full data. I suspect many too late censored in the high ages, a problem with the data registration. I will return with 
+a study of mortality after age 90 in a coming post.
+
+```{r nonstandards, fig.cap = "Sampling of the dead."}
+sfit <- coxph(Surv(enter, exit, event) ~ strata(region, sex), data = sols)
+par(las = 1)
+plot(survfit(sfit), fun = "cumhaz", col = 1:4, lty = 1:4, xlab = "Years after 60.", ylab = "Cum. hazards", axes = FALSE)
+axis(1, at = c(0, 7, 11, 20, 30, 40))
+axis(2, at = c(0, log(2), 3, 5, 7), labels = c("0", "log(2)", "3", "5", "7"))
+box()
+legend("topleft", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
+abline(h = 0)
+abline(h = log(2), lty = 2)
+```
+
+```{r }
+plot(survfit(sfit), col = 1:4, lty = 1:4, xlab = "Years after 60, the dead.", 
+     ylab = "Remaining fraction", axes = FALSE)
+axis(1, at = c(0, 7, 11, 20, 30))
+axis(2, at = c(0, 0.5, 1), las = 1)
+box()
+legend("topright", legend = levels(strata(ols$region, ols$sex)), col = 1:4, lty = 1:4)
+abline(h = 0)
+abline(h = 0.5, lty = 2)
+```
+
+We can also here make an exact calculation, which results in
+
+```{r exactkis}
+x <- survfit(sfit)
+x
+```
 # Conclusion

### 2018-01-30 particle3.Rmd
first commit

--- 2018-01-13 article3.Rmd
@@
 source("R/get.file.R")
-skel <- get.file("skel")
+##skel <- get.file("skel")
+load("data/skel.rda")
 ##
-skume <- get.file("skume")
+##skume <- get.file("skume")
 ##
+load("data/skume.rda")
 foli <- get.file("foli")
@@
                     FUN = sum)
+vb0Tab <- vb0Tab[!(vb0Tab$age %in% c(30, 35)), ]
 ```
@@
 ##
+cap0m <- "Median age at death for a 40-year-old by social class and period, women (top) and men (bottom)."
+scap0m <- "Median age at death for a 40-year old by social class."
 
@@
 ##
-vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
-vb4064$age <- factor(vb4064$age)
-##vb4064$hisclass <- vb4064$hisclass40
+vb40 <- vb0Tab
+vb40$age <- factor(vb40$age)
+##vb40$hisclass <- vb40$hisclass40
 ```
@@
 ```{r hazcalw}
-source("R/getPHglm2.R")
+source("R/pgetPHglm2.R")
 source("R/getPHglm.R")
@@
 nper <- length(labb)
-resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+resw <- pgetPHglm2(vb40[vb40$sex == "female", ],
             nper, labb)
-resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+resm <- pgetPHglm2(vb40[vb40$sex == "male", ],
             nper, labb)
@@
 
-```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
+```{r meli, fig.cap = cap0m, fig.height = 8, out.height = "14cm"}
+oldpar <- par(mfrow = c(2, 1))
+plot(1:13, resw[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(40, 75), main = "Women", ylab = "Age", xlab = "", xlim = c(1, 15.5))
+axis(1, at = 1:13, labels = labb, las = 2)
+axis(2, at = c(40, 50, 60, 70), las = 1)
+abline(h = c(50, 60, 70), lty = 3)
+box()
+lines(1:13, resw[2, ], type = "b", lty = 2, pch = 2, col = scol[2])
+lines(1:13, resw[3, ], type = "b", lty = 3, pch = 3, col = scol[3])
+lines(1:13, resw[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
+abline(h = 40)
+legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
+## Men:
+plot(1:13, resm[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(40, 75), main = "Men", ylab = "Age", xlab = "", xlim = c(1, 15.5))
+axis(1, at = 1:13, labels = labb, las = 2)
+axis(2, at = c(40, 50, 60, 70), las = 1)
+box()
+lines(1:13, resm[2, ], type = "b", lty = 2, pch = 2, col = scol[2])
+lines(1:13, resm[3, ], type = "b", lty = 3, pch = 3, col = scol[3])
+lines(1:13, resm[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
+abline(h = 40)
+abline(h = c(50, 60, 70), lty = 3)
+legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
+par(oldpar)
+source("R/plotED.glm.R") # From below (not eval)
+```
+
+```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm",eval = FALSE}
 source("R/plotED.glm.R")
@@
 ```{r cumswtab,cap="Table"}
-vb4064$social_class <- vb4064$hisclass
-vb4064$marital_status <- vb4064$civst
-vb4064$urban_or_rural <- vb4064$urban
-vb4064$age_group <- vb4064$age
+vb40$social_class <- vb40$hisclass
+vb40$marital_status <- vb40$civst
+vb40$urban_or_rural <- vb40$urban
+vb40$age_group <- vb40$age
 fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
-            data = vb4064, family = poisson)
+            data = vb40, family = poisson)
 dr <- drop1(fitt, test = "Chisq")
@@
 
-```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm"}
+```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm",eval = FALSE}
 oldpar <- par(mfrow = c(2, 1))
@@
 ##source("R/plotED.R")
-source("R/getPHglm.R")
+source("R/getPHglm2.R")
 resw <- getPHglm2(vb6589[vb6589$sex == "female", ],

### 2018-01-30 slides3.Rnw
first commit

@@
 
-\begin{document}
-
-\maketitle
-
 <<setup, include = FALSE>>=
@@
 opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+@
+
+<<getgoing>>=
 library(knitr)
@@
 
+\begin{document}
+
+\maketitle
+
 \begin{frame}{Questions}
@@
 
-\includegraphics[height=7cm]{myfigs/skum17.pdf}
-
+%\includegraphics[height=7cm]{myfigs/skum17.pdf}
+\includegraphics[height=7cm]{myfigs/skum.pdf}
 \end{column}
@@
    \end{itemize}
-\end{itemize}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{The sampling frame}
-
-<<lexis2>>=
+\end{itemize}sampling fra
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}[fragile]{The sampling frame}
+
+<<lexis2,echo = FALSE>>=
 plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
@@
 @
+
 \end{frame}
@@
 
-\begin{frame}{The Linnaeus data with deaths}
+\begin{frame}[fragile]{The Linnaeus data with deaths}
 
@@
 \begin{column}{0.6\textwidth}
-<<picklexus2,fig.width=6,fig.height = 5.9>>=
+
+<<picklexus2,fig.width=6,fig.height = 5.9,echo=FALSE>>=
 
@@
 \begin{column}{0.6\textwidth}
-<<picklexus3,fig.width=6,fig.height = 5.9>>=
-
+<<picklexus3,fig.width=6,fig.height = 5.9,echo=FALSE>>=
 ##load("data/vb.rda")
-tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
-tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
-
-vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
+##tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
+##tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
+tot.m <- length(unique(vb$id[vb$from != "popum" & vb$sex == "male"]))
+tot.f <- length(unique(vb$id[vb$from != "popum" & vb$sex == "female"]))
+
+vbd <- vb[vb$event == 1 & vb$from != "popum", c("birthdate", "exit", "sex")]
 d.m <- sum(vbd$sex == "male")
@@
 
-<<anovatable,results='asis',echo=FALSE>>=
-xtable(dr)
+<<anovatable,echo=FALSE>>=
+knitr::kable(dr)
 @
@@
 {\scriptsize
-<<extab,echo=FALSE,results='asis'>>=
-xtable(summary(fitp)$coefficients[1:10, 1:2])
+<<extab,echo=FALSE>>=
+knitr::kable(summary(fitp)$coefficients[1:10, 1:2])
 @
@@
 
+\end{document}
 \begin{frame}{Cancer mortality, ages 40-64 and 65-89}

### 2018-01-30 R/mpch.R

@@ -1,0 +1,6 @@
+mpch <- function(cuts, levels){
+  n <- length(levels)
+  if (length(cuts) != (n-1)) stop("length(cuts) = ", length(cuts), " != ", n, "\n")
+  res <- integrate(ppch, 0, Inf, cuts = cuts, levels = levels, lower.tail = FALSE)
+  res$value
+}

### 2018-01-30 R/pgetPHglm2.R

@@ -1,0 +1,74 @@
+pgetPHglm2 <- function(dat, nper, labb, from = 40, median = TRUE){
+    ## For calculating remaining life:
+    ## dat$age must be a factor ...
+    ## Note: return value is Cumulative hazards at given age points (straight line between).
+    ##
+    ## 'dat$age' is assumed to be grouped in equidistant intervals!
+    source("R/mpch.R")
+    res <- vector(nper, mode = "list")
+    hc <- levels(dat$hisclass)
+    m <- length(hc)
+    k <- length(unique(dat$age))
+    if (k <= 1) error("No. of age intervals must be at least 2!")
+    ageName <- as.numeric(levels(dat$age))
+    ageInt <- ageName[2] - ageName[1]
+    levs <- matrix(0, nrow = m, ncol = k + 1)
+   
+    for (i in 1:nper){
+        fit <- glm(event ~ offset(log(exposure)) + urban + 
+                       civst +
+                       age * hisclass, 
+                   data = dat[dat$period == labb[i], ], family = poisson)
+        
+        co <- fit$coefficients
+        ##if(any(is.na(co))) cat("NA in ", i, " period\n")
+        ages <- grep("age", names(co))
+        hisces <- grep("hisclass", names(co))
+        interact <- ages[ages %in% hisces]
+        amain <- co[ages[!(ages %in% interact)]]
+        hmain <- co[hisces[!(hisces %in% interact)]]
+        inter <- matrix(co[interact], ncol = k - 1, byrow = TRUE)
+        tmp <- matrix(0, nrow = m, ncol = k)
+        for (j in 1:m){
+            ##cat("j = ", j, "\n")
+            tmp[j, ] <- c(co[1], co[1] + amain)
+        }
+        for (j in 2:m){
+            tmp[j, ] <- tmp[j, ] + hmain[j - 1]
+        }
+        tmp[-1, -1] <- tmp[-1, -1] + inter
+        tmp <- exp(tmp)
+        levs <- cbind(rep(0, m), tmp)
+        levs[is.na(levs)] <- 0 ## HOPPSAN!!!
+        for (j in 1:m){
+            levs[j, ] <- cumsum(levs[j, ]) * ageInt
+        }
+        
+
+        colnames(levs) <- c(ageName, ageName[k] + ageInt)
+        rownames(levs) <- hc
+        res[[i]] <- levs
+    }
[…]

### 2018-01-31 particle3.Rmd
fixes

@@
 
-## Ages 40--64
+## Median remaining life at age 40
 
@@
 scap0m <- "Median age at death for a 40-year old by social class."
+cap0md <- "Differences in median age at death for a 40-year-old by social class and period (ref: middle class), women (top) and men (bottom)."
+scap0md <- "Differences in median age at death for a 40-year old by social class."
 
@@
 resw <- pgetPHglm2(vb40[vb40$sex == "female", ],
-            nper, labb)
+            nper, labb)$medi
 resm <- pgetPHglm2(vb40[vb40$sex == "male", ],
-            nper, labb)
+            nper, labb)$medi
 ```
@@
 oldpar <- par(mfrow = c(2, 1))
-plot(1:13, resw[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(40, 75), main = "Women", ylab = "Age", xlab = "", xlim = c(1, 15.5))
+plot(1:13, resw[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(60, 95), main = "Women", ylab = "Age", xlab = "", xlim = c(1, 15.5))
 axis(1, at = 1:13, labels = labb, las = 2)
-axis(2, at = c(40, 50, 60, 70), las = 1)
-abline(h = c(50, 60, 70), lty = 3)
+axis(2, at = c(60, 70, 80, 90), las = 1)
+abline(h = c(60), lty = 3)
+abline(h = c(70, 80, 90), lty = 3)
 box()
@@
 lines(1:13, resw[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
-abline(h = 40)
+abline(h = 60)
 legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
 ## Men:
-plot(1:13, resm[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(40, 75), main = "Men", ylab = "Age", xlab = "", xlim = c(1, 15.5))
+plot(1:13, resm[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(60, 95), main = "Men", ylab = "Age", xlab = "", xlim = c(1, 15.5))
 axis(1, at = 1:13, labels = labb, las = 2)
-axis(2, at = c(40, 50, 60, 70), las = 1)
+axis(2, at = c(60, 70, 80, 90), las = 1)
 box()
@@
 lines(1:13, resm[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
-abline(h = 40)
-abline(h = c(50, 60, 70), lty = 3)
+abline(h = 60)
+abline(h = c(70, 80, 90), lty = 3)
 legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
@@
 
+In Figure \@ref(fig:melidiff) the difference in median life (Reference: middle class) is shown.
+
+```{r melidiff, fig.cap = cap0md, fig.height = 8, out.height = "14cm"}
+oldpar <- par(mfrow = c(2, 1))
+for (m in c(1, 3, 4)) resw[m, ] <- resw[m, ] - resw[2, ]
+resw[2, ] <- 0
+for (m in c(1, 3, 4)) resm[m, ] <- resm[m, ] - resm[2, ]
+resm[2, ] <- 0
+plot(1:13, resw[1, ], type = "b", col = scol[1], axes = FALSE, main = "Women", ylab = "Years", xlab = "", xlim = c(1, 15.5), ylim = c(-8, 8))
+axis(1, at = 1:13, labels = labb, las = 2)
+axis(2, at = c(-5, 0, 5), las = 1)
+abline(h = c(-5, 5), lty = 3)
+box()
+lines(1:13, resw[2, ], type = "b", lty = 2, pch = 2, col = scol[2])
+lines(1:13, resw[3, ], type = "b", lty = 3, pch = 3, col = scol[3])
+lines(1:13, resw[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
+abline(h = 60)
+legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
+## Men:
+plot(1:13, resm[1, ], type = "b", col = scol[1], axes = FALSE, main = "Men", ylab = "Years", xlab = "", xlim = c(1, 15.5), ylim = c(-8, 8))
+axis(1, at = 1:13, labels = labb, las = 2)
+axis(2, at = c(-5, 0, 5), las = 1)
+box()
+lines(1:13, resm[2, ], type = "b", lty = 2, pch = 2, col = scol[2])
+lines(1:13, resm[3, ], type = "b", lty = 3, pch = 3, col = scol[3])
+lines(1:13, resm[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
+abline(h = c(-5, 5), lty = 3)
+legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
+par(oldpar)
+source("R/plotED.glm.R") # From below (not eval)
+```
 ```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm",eval = FALSE}

### 2018-01-31 R/pgetPHglm2.R

@@ -40,9 +40,9 @@ pgetPHglm2 <- function(dat, nper, labb, from = 40, median = TRUE){
         tmp <- exp(tmp)
         levs <- cbind(rep(0, m), tmp)
         levs[is.na(levs)] <- 0 ## HOPPSAN!!!
-        for (j in 1:m){
-            levs[j, ] <- cumsum(levs[j, ]) * ageInt
-        }
+        ##for (j in 1:m){
+          ##  levs[j, ] <- cumsum(levs[j, ]) * ageInt
+        ##}
         
 
         colnames(levs) <- c(ageName, ageName[k] + ageInt)
@@ -54,7 +54,7 @@ pgetPHglm2 <- function(dat, nper, labb, from = 40, median = TRUE){
     cuts <- as.numeric(colnames(res[[1]]))
     cuts <- cuts[-c(1, length(cuts))]
     medi <- matrix(0, ncol = nper, nrow = 4)
-    cat("cuts = ", cuts, "\n")
+    ##cat("cuts = ", cuts, "\n")
     ##cat("levs = ", levs, "\n")
     for (i in 1:nper){
       mat <- res[[i]]
@@ -70,5 +70,5 @@ pgetPHglm2 <- function(dat, nper, labb, from = 40, median = TRUE){
     ##res
     colnames(medi) <- names(res)
     rownames(medi) <- c("elite", "middle", "worker", "NA")
-    medi + 40
+    list(medi = medi + 40, mat = mat)
 }

### 2018-01-31 R/pgetmedian.R
first commit

@@ -1,0 +1,74 @@
+pgetPHglm2 <- function(dat, nper, labb, from = 40, median = TRUE){
+    ## For calculating remaining life:
+    ## dat$age must be a factor ...
+    ## Note: return value is Cumulative hazards at given age points (straight line between).
+    ##
+    ## 'dat$age' is assumed to be grouped in equidistant intervals!
+    source("R/mpch.R")
+    res <- vector(nper, mode = "list")
+    hc <- levels(dat$hisclass)
+    m <- length(hc)
+    k <- length(unique(dat$age))
+    if (k <= 1) error("No. of age intervals must be at least 2!")
+    ageName <- as.numeric(levels(dat$age))
+    ageInt <- ageName[2] - ageName[1]
+    levs <- matrix(0, nrow = m, ncol = k + 1)
+   
+    for (i in 1:nper){
+        fit <- glm(event ~ offset(log(exposure)) + urban + 
+                       civst +
+                       age * hisclass, 
+                   data = dat[dat$period == labb[i], ], family = poisson)
+        
+        co <- fit$coefficients
+        ##if(any(is.na(co))) cat("NA in ", i, " period\n")
+        ages <- grep("age", names(co))
+        hisces <- grep("hisclass", names(co))
+        interact <- ages[ages %in% hisces]
+        amain <- co[ages[!(ages %in% interact)]]
+        hmain <- co[hisces[!(hisces %in% interact)]]
+        inter <- matrix(co[interact], ncol = k - 1, byrow = TRUE)
+        tmp <- matrix(0, nrow = m, ncol = k)
+        for (j in 1:m){
+            ##cat("j = ", j, "\n")
+            tmp[j, ] <- c(co[1], co[1] + amain)
+        }
+        for (j in 2:m){
+            tmp[j, ] <- tmp[j, ] + hmain[j - 1]
+        }
+        tmp[-1, -1] <- tmp[-1, -1] + inter
+        tmp <- exp(tmp)
+        levs <- cbind(rep(0, m), tmp)
+        levs[is.na(levs)] <- 0 ## HOPPSAN!!!
+        for (j in 1:m){
+            levs[j, ] <- cumsum(levs[j, ]) * ageInt
+        }
+        
+
+        colnames(levs) <- c(ageName, ageName[k] + ageInt)
+        rownames(levs) <- hc
+        res[[i]] <- levs
+    }
[…]

### 2018-02-04 article3b.Rmd
introduce median life

--- 2018-01-13 article3.Rmd
@@
 \item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
-\item {\tt middle}, Hisclass 3, 4, 5, 6 and 8. Lower managers,
-farmers and lower professionals, clerical and sales personell.
+\item {\tt middle}, Hisclass 3, 4, 5, and 6. Lower managers,
+farm hands and lower professionals, clerical and sales personell.
 \item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
 farm workers.
+\item {\tt farmer}, Hisclass 8. Farmers. 
 \end{enumerate}
@@
 
-```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
+```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
 xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 yy <- round(prop.table(xxx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
 ```
@@
 yy <- round(prop.table(xxx, 1) * 100, 1)
-barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+barplot(t(yy)[5:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
 ```
@@
 [@gb12;@eha]. In the analysis, data are divided into two age groups,
-40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
+30--64 and 65--99 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
 but we then combine results in order to show the development of the social patterning of mortality during the studied period.
@@
 2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
-**civst** (marital status). The cumulative
-hazards at the end of the age period (65 if 40--64, 90 if 65--89), conditional on survival 
-to the start of the age period, are the
-main target in our analysis. These numbers are used in the process of
+**civst** (marital status). The *cumulative hazards* at the end of the age period *30--64* 
+conditional on survival to age 30,  and  *remaining expected life at 65* are the
+main targets in our analysis. These numbers are used in the process of
 graphically illustrating the development over time.
@@
 ## Colors for plotting strata:
-scol <- c("blue", rep("darkgreen", 2), "red")
+scol <- c("black", "green", "blue", "darkgreen", "red")
 ## Common for all groups:
@@
 
-## Ages 40--64
+## Ages 30--64
 
-```{r getske4060}
-cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
-scap1w <-  "Cumulative hazards, ages 40--64, women."
-cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
-scap1m <-  "Cumulative hazards, ages 40--64, men."
+```{r getske3064}
+cap1w <-  "Cumulative hazards for social class by time period, ages 30--64, women."
+scap1w <-  "Cumulative hazards, ages 30--64, women."
+cap1m <-  "Cumulative hazards for social class by time period, ages 30--64, men."
+scap1m <-  "Cumulative hazards, ages 30--64, men."
 ##
 
-cap0 <- "Cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
-scap0 <- "Cumulative hazard of dying before age 65 for a 40-year old by social class."
-cap1 <- "Cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
-scap1 <- "Cumulative hazard of dying before age 90 for a 65-year old by social class."
+cap0 <- "Cumulative hazard of dying before age 65 for a 30-year-old by social class and period, women (top) and men (bottom)."
+scap0 <- "Cumulative hazard of dying before age 65 for a 30-year old by social class."
+cap1 <- "Remaining years of life for a 65-year-old by social class and period, women (top) and men (bottom)."
+scap1 <- "Remaining life for a 65-year old by social class."
 cap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
@@
 
-cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women."
-scap2w <-  "Cumulative hazards, ages 65--89, women."
-cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men."
-scap2m <-  "Cumulative hazards, ages 65--89, men."
+cap2w <-  "Cumulative hazards for social class by time period, ages 65--99, women."
+scap2w <-  "Cumulative hazards, ages 65--99, women."
+cap2m <-  "Cumulative hazards for social class by time period, ages 65--99, men."
+scap2m <-  "Cumulative hazards, ages 65--99, men."
 ##
@@
 cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
-scap22log <- "Probability by social class, ages 65--89, last years."
+scap22log <- "Probability by social class, ages 65--99, last years."
 
 ##
-vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
-vb4064$age <- factor(vb4064$age)
-##vb4064$hisclass <- vb4064$hisclass40
+vb3064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
+vb3064$age <- factor(vb3064$age)
+##vb3064$hisclass <- vb3064$hisclass40
 ```
@@
 nper <- length(labb)
-resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
+resw <- getPHglm2(vb3064[vb3064$sex == "female", ],
             nper, labb)
-resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
+resm <- getPHglm2(vb3064[vb3064$sex == "male", ],
             nper, labb)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
-                main = "Women 40-64", ylim = c(0, 1))##, exclude = 5:6)
+                main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
-                main = "Men 40-64", ylim = c(0, 1))##, exclude = 5:6)
+                main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
 par(oldpar)
@@
 ```{r cumswtab,cap="Table"}
-vb4064$social_class <- vb4064$hisclass
-vb4064$marital_status <- vb4064$civst
-vb4064$urban_or_rural <- vb4064$urban
-vb4064$age_group <- vb4064$age
+vb3064$social_class <- vb3064$hisclass
+vb3064$marital_status <- vb3064$civst
+vb3064$urban_or_rural <- vb3064$urban
+vb3064$age_group <- vb3064$age
 fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
-            data = vb4064, family = poisson)
+            data = vb3064, family = poisson)
 dr <- drop1(fitt, test = "Chisq")
@@
 woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
+                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 30-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
+                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 30-64")##, exclude = 5:6)
 par(oldpar)
@@
 
-## Ages 65--89
+## Ages 65--99
 
-```{r getske6089}
-vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
-vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
-##vb6589$hisclass <- vb6589$hisclass60
+```{r rawolddat,echo=FALSE}
+rawskel <- get.file("rawskel")
+rawskel$urban <- FALSE
+rawskel <- rawskel[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
+##
+rawskume <- get.file("rawskume")
+rawskume <- rawskume[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
+##
+rawfoli <- get.file("rawfoli")
+rawfoli <- rawfoli[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
+raw <- rbind(rawskel, rawskume, rawfoli)
+raw <- age.window(raw, c(65, 100))
 ```
 
+<!--
+```{r getske6099}
+vb6599 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
+vb6599$age <- factor(vb6599$age) # NOTE:: Change to '6599'!
+##vb6599$hisclass <- vb6599$hisclass60
+```
+-->
 We now turn to the elderly, those who in most cases have left the workforce.
@@
 ```{r cumswtabold,cap="Table"}
-vb6589$marital_status <- vb6589$civst
-vb6589$social_class <- vb6589$hisclass
-vb6589$urban_or_rural <- vb6589$urban
-vb6589$age_group <- vb6589$age
+vb6599$marital_status <- vb6599$civst
+vb6599$social_class <- vb6599$hisclass
+vb6599$urban_or_rural <- vb6599$urban
+vb6599$age_group <- vb6599$age
 fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
-            data = vb6589, family = poisson)
+            data = vb6599, family = poisson)
 drold <- drop1(fittold, test = "Chisq")
@@
 drold
-##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--89."), 
+##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--99."), 
   ##    type = "latex", comment = FALSE)
@@
 ##source("R/plotED.R")
-source("R/getPHglm.R")
-resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
+source("R/pgetPHglm2.R")
+resw <- pgetPHglm2(vb6599[vb6599$sex == "female", ],
             nper, labb)
-resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
+resm <- pgetPHglm2(vb6599[vb6599$sex == "male", ],
             nper, labb)
@@
 woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = FALSE,
-                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Women 65-89")##, exclude = 5:6)
+                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Women 65-99")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = FALSE,
-                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Men 65-89")##, exclude = 5:6)
+                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Men 65-99")##, exclude = 5:6)
 par(oldpar)
@@
                     ylab = "H(90|65)",
-                main = "Women 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
+                main = "Women 65-99", ylim = c(0, 0.6))##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                     ylab = "H(90|65)",
-                main = "Men 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
+                main = "Men 65-99", ylim = c(0, 0.6))##, exclude = 5:6)
 par(oldpar)
@@
 fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
-            data = vb6589, family = poisson)
+            data = vb6599, family = poisson)
 dr <- drop1(fitt, test = "Chisq")
 ##knitr::kable(dr)
-xtable(dr,label = "tab:cumswoldtab", caption = "ANOVA, all causes of death, 65-89.")
+xtable(dr,label = "tab:cumswoldtab", caption = "ANOVA, all causes of death, 65-99.")
 ```

### 2018-02-04 scb.Rmd
introduce median life

---
title: "Statistics Sweden"
author: "Göran Broström"
date: "February 2, 2018"
output: 
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Studying ages 20--99, Sweden 1968--2014, for trend in expected life at 20 and median remaining life at 20. Using data from Statistics Sweden, stored in the **R** package **Sweden** .

# Get relevant data

```{r getdata}
library(Sweden)
fdeaths <- getDeaths(sex = "females", ages = 20:99)
fpop <- getPop(sex = "females", years = c(1968:2014), ages = 20:99)
mdeaths <- getDeaths(sex = "males", ages = 20:99)
mpop <- getPop(sex = "males", years = c(1968:2014), ages = 20:99)
```

Then we calculate yearly age-specific death rates:

```{r calcrates}
frates <- fdeaths / fpop
mrates <- mdeaths / mpop
```

And then we calculate $e_{20}$ and $m_{20}$ for the years 1968--2014.

# Statistics for men, age 20

```{r statmen, echo = FALSE}
library(eha)
source("R/mpch.R")
nYears <- NCOL(mrates)
me20 <- numeric(nYears)
mm20 <- numeric(nYears)
cts <- (21:99)
for (i in seq_len(nYears)){
    me20[i] <- mpch(cuts = cts - 20, levels = mrates[, i]) + 20
    mm20[i] <-  qpch(0.5, cuts = cts - 20, levels = mrates[, i]) + 20
}
plot(1968:2014, mm20, type = "b", main = "Men", ylim = c(70, 90), xlab = "Year", ylab = "Age", col = "red", las = 1)
lines(1968:2014, me20, col = "blue", type = "b")
text(1995, 82, "e.20", col = "red")
text(1995, 73, "m.20", col = "blue")
abline(h = c(75, 80, 85), lty = 3, col = "green")
abline(v = c(1970, 1980, 1990, 2000, 2010), lty = 3, col = "green")
mdiff <- mean(mm20 - me20)
```

We can also see that the difference between expected life and median life is fairly constant over time with a mean value of `r round(mdiff, 2)` years.

# Statistics for women, age 20

```{r statwomen, echo = FALSE}
library(eha)
source("R/mpch.R")
nYears <- NCOL(frates)
fe20 <- numeric(nYears)
fm20 <- numeric(nYears)
cts <- (21:99)
for (i in seq_len(nYears)){
    fe20[i] <- mpch(cuts = cts - 20, levels = frates[, i]) + 20
    fm20[i] <-  qpch(0.5, cuts = cts - 20, levels = frates[, i]) + 20
}
plot(1968:2014, fm20, type = "b", main = "Women", ylim = c(70, 90), xlab = "Year", ylab = "Age", col = "red", las = 1)
lines(1968:2014, fe20, col = "blue", type = "b")
text(1995, 86, "e.20", col = "red")
text(1995, 79, "m.20", col = "blue")
abline(h = c(75, 80, 85), lty = 3, col = "green")
abline(v = c(1970, 1980, 1990, 2000, 2010), lty = 3, col = "green")
fdiff <- mean(fm20 - fe20)
```

We can also see that the difference between expected life and median life is fairly constant over time with a mean value of `r round(fdiff, 2)` years.

### 2018-02-04 R/pgetmedian.R

@@ -1,74 +1,39 @@
-pgetPHglm2 <- function(dat, nper, labb, from = 40, median = TRUE){
+pgetmedian <- function(dat, nper, labb, from = 65, median = TRUE){
     ## For calculating remaining life:
     ## dat$age must be a factor ...
     ## Note: return value is Cumulative hazards at given age points (straight line between).
     ##
     ## 'dat$age' is assumed to be grouped in equidistant intervals!
     source("R/mpch.R")
-    res <- vector(nper, mode = "list")
+    
+    labb <- levels(dat$period)
+    nper <- length(labb)
     hc <- levels(dat$hisclass)
     m <- length(hc)
+    dat$age <- factor(dat$age)
     k <- length(unique(dat$age))
     if (k <= 1) error("No. of age intervals must be at least 2!")
-    ageName <- as.numeric(levels(dat$age))
-    ageInt <- ageName[2] - ageName[1]
-    levs <- matrix(0, nrow = m, ncol = k + 1)
-   
-    for (i in 1:nper){
-        fit <- glm(event ~ offset(log(exposure)) + urban + 
-                       civst +
-                       age * hisclass, 
-                   data = dat[dat$period == labb[i], ], family = poisson)
-        
-        co <- fit$coefficients
-        ##if(any(is.na(co))) cat("NA in ", i, " period\n")
-        ages <- grep("age", names(co))
-        hisces <- grep("hisclass", names(co))
-        interact <- ages[ages %in% hisces]
-        amain <- co[ages[!(ages %in% interact)]]
-        hmain <- co[hisces[!(hisces %in% interact)]]
-        inter <- matrix(co[interact], ncol = k - 1, byrow = TRUE)
-        tmp <- matrix(0, nrow = m, ncol = k)
-        for (j in 1:m){
-            ##cat("j = ", j, "\n")
-            tmp[j, ] <- c(co[1], co[1] + amain)
+    cts <- as.numeric(levels(dat$age))[-1]
+    medi <- matrix(0, ncol = nper, nrow = m)
+    for (j in 10:nper){
+        cat("j = ", j, "\n")
+        for (i in 1:m){
+            cat("   i = ", i, "\n")
+            fit <- glm(event ~ offset(log(exposure)) + age + urban + civst,
[…]

### 2018-02-04 article3b.Rmd
median life at 65 complete

@@
 ```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
-plot(c(1801, 1801, 1951, 1951, 1801), c(40, 90, 90, 40, 40),
+plot(c(1801, 1801, 1951, 1951, 1801), c(30, 100, 100, 30, 30),
      type = "l", xlim = c(1800, 2014), lwd = 2, col = "red",
-     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
+     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 100))
 axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1991, 2014), las = 2)
-axis(2, las = 2, at = c(0, 40, 65, 90))
-axis(4, las = 2, at = c(0, 40, 65, 90))
+axis(2, las = 2, at = c(0, 30, 65, 100))
+axis(4, las = 2, at = c(0, 30, 65, 100))
 box()
 abline(h = 0)
-lines(c(1976, 1976, 2014, 2014, 1976), c(40, 90, 90, 40, 40),
+lines(c(1976, 1976, 2014, 2014, 1976), c(30, 100, 100, 30, 30),
       col = "blue", lwd = 2)
@@
 
-In the data set analysed here, all individuals aged 40 years and older ever
+In the data set analysed here, all individuals aged 30 years and older ever
 having resided in either of the regions are included. The data file
@@
 yy <- round(prop.table(xxx, 1) * 100, 1)
-barplot(t(yy)[5:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
+barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
 ```
@@
 h(t; \lambda, \boldsymbol{\beta}, \textbf{x}) =
-\lambda(t)e^{\boldsymbol{\beta} \mathbf{x}}, \; \; 40 < t \le 90,
+\lambda(t)e^{\boldsymbol{\beta} \mathbf{x}}, \; \; 30 < t \le 100,
 \end{equation}
@@
 \begin{equation*}
-\lambda(t) = \alpha_i,\; 40 + 5(i - 1) < t \le 40 + 5i, \; i = 1, \ldots, 10.
+\lambda(t) = \alpha_i,\; 30 + 5(i - 1) < t \le 30 + 5i, \; i = 1, \ldots, 10.
 \end{equation*}
@@
 scap1 <- "Remaining life for a 65-year old by social class."
-cap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
-scap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class."
-cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
-scap12log <- "Probability by social class, ages 40--64, last years."
+cap12 <- "Relative cumulative hazard of dying before age 65 for a 30-year-old by social class and period, women (top) and men (bottom)."
+scap12 <- "Relative cumulative hazard of dying before age 65 for a 30-year-old by social class."
+cap12log <- "Probability of dying before age 65 for a 30-year-old by social class and period, women (top) and men (bottom). The last decades."
+scap12log <- "Probability by social class, ages 30--64, last years."
 
@@
 ##
-cap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
-scap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class."
-cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
+cap22 <- "Relative cumulative hazard of dying before age 100 for a 65-year-old by social class and period, women (top) and men (bottom)."
+scap22 <- "Relative cumulative hazard of dying before age 100 for a 65-year-old by social class."
+cap22log <- "Probability of dying before age 100 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
 scap22log <- "Probability by social class, ages 65--99, last years."
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
+woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
                 main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
+moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
                 main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
@@
 ##knitr::kable(dr)
-##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
+##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."), 
   ##    type = "latex", comment = FALSE)
@@
 Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
-40--64 during the different periods. Obviously, but not surprisingly, mortality
+30--64 during the different periods. Obviously, but not surprisingly, mortality
 decreases over time for both sexes. Women have consistently better survival
@@
 woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 30-64")##, exclude = 5:6)
+                    ylab = "H(65|30)", ylim = c(0.0, 0.6), main = "Women 30-64")##, exclude = 5:6)
 moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 30-64")##, exclude = 5:6)
+                    ylab = "H(65|30)", ylim = c(0.0, 0.6), main = "Men 30-64")##, exclude = 5:6)
 par(oldpar)
@@
 
-<!--
 ```{r getske6099}
@@
 ```
--->
+
 We now turn to the elderly, those who in most cases have left the workforce.
-Corresponding to the previous analysis, we use the value of the cumulative hazards at
-duration 25 (age 90), starting from age 65 as a summary measure. Looking at the table below, it is obvious that the statistical significance is still strong.
+We use the *median remaining life at 65* (reported as median age at death) as the summary measure. Looking at the table below, it is obvious that the statistical significance is still strong.
 
@@
 
-The general pattern and development of mortality (Figure \@ref(fig:cumswold)) do not
+The general pattern and development of mortality (Figures \@ref(fig:plotrawwom) and \@ref(fig:plotrawm)) do not
 differ much from the younger age group, of course representing a much
@@
 Second World War; that is, later than for the age
-group 40--64 years. This corresponds to the Swedish development whereby we
+group 30--64 years. This corresponds to the Swedish development whereby we
 find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status-levelling hypothesis (see Figure \@ref(fig:cumsw2)). For women the results are mixed. We do not find the
+we find for the younger age group in line with the status-levelling hypothesis. For women the results are mixed. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
@@
 time.
-In most periods the social differences were small, particularly between the workers and the middle group, and the elite women even had somewhat higher mortality during some periods until the 190--1925 period.
+In most periods the social differences were small, particularly between the workers and the middle group, and the elite women even had somewhat higher mortality during some periods until the 1901--1925 period.
 The social differences continued to be small until the last decades of the
@@
 For men we find indications of the same intriguing change among elderly as
-for the adult male population aged 40--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
+for the adult male population aged 30--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
 the 20th century. It is not until the second half of the 20th
@@
 
-```{r cumswold, fig.cap = cap1, fig.scap = scap1, fig.height = 8, out.height = "14cm"}
-##source("R/plotED.R")
-source("R/pgetPHglm2.R")
-resw <- pgetPHglm2(vb6599[vb6599$sex == "female", ],
-            nper, labb)
-resm <- pgetPHglm2(vb6599[vb6599$sex == "male", ],
-            nper, labb)
-
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = FALSE,
-                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Women 65-99")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = FALSE,
-                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Men 65-99")##, exclude = 5:6)
-par(oldpar)
+```{r rawreswomen}
+hc <- levels(raw$hisclass)
+n.hc <- length(hc)
+peri <- levels(raw$period)
+n.peri <- length(peri)
+med <- matrix(0, nrow = n.hc, ncol = n.peri)
+for (i in 1:n.hc){
+    fit <- coxph(Surv(enter, exit, event) ~ civst + urban + strata(period), data = raw[raw$sex == "female" & 
+                                                                              raw$hisclass == hc[i], ])
+    med[i, ] <- summary(survfit(fit, newdata = data.frame(civst = "married", urban = FALSE)))$table[, "median"]
+}
 ```
 
-```{r cumsw2, fig.cap = cap22, fig.scap = scap22, fig.height = 8, out.height = "14cm"}
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE, 
-                    ylab = "H(90|65)",
-                main = "Women 65-99", ylim = c(0, 0.6))##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(90|65)",
-                main = "Men 65-99", ylim = c(0, 0.6))##, exclude = 5:6)
-par(oldpar)
+```{r plotrawwom, fig.cap="Median age at death for a 65 year old married woman on the countryside by SES and period.", fig.height = 5, out.height = "10cm"}
+x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
+plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Median age at death | 65", ylim = c(70, 90), main = "Women", axes = FALSE)
+axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+box()
+abline(h = c(75, 80, 85), lty = 3)
+for (i in 2:5){
+    lines(x, med[i, -7], col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("bottomright", legend = hc, lty = 1:5, pch = 1:5, col = scol)
 ```
 
-```{r cumswoldtab,results='asis',eval=FALSE}
-fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
-            data = vb6599, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
-##knitr::kable(dr)
-xtable(dr,label = "tab:cumswoldtab", caption = "ANOVA, all causes of death, 65-99.")
+```{r rawresmen}
+hc <- levels(raw$hisclass)
+n.hc <- length(hc)
+peri <- levels(raw$period)
+n.peri <- length(peri)
+med <- matrix(0, nrow = n.hc, ncol = n.peri)
+for (i in 1:n.hc){
+    fit <- coxph(Surv(enter, exit, event) ~ civst + urban + strata(period), data = raw[raw$sex == "male" & 
+                                                                              raw$hisclass == hc[i], ])
+    med[i, ] <- summary(survfit(fit, newdata = data.frame(civst = "married", urban = FALSE)))$table[, "median"]
+}
+```
+
+```{r plotrawm, fig.cap="Median age at death for a 65 year old married man on the countryside by SES and period.", fig.height = 5}
+x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
+plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Median age at death | 65", ylim = c(70, 90), main = "Men", axes = FALSE)
+axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+abline(h = c(75, 80, 85), lty = 3)
+box()
+for (i in 2:5){
+    lines(x, med[i, -7], col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("bottomright", legend = hc, lty = 1:5, pch = 1:5, col = scol)
 ```

### 2018-02-05 article3b.Rmd
fix small error, some remain

@@
 output:
-  bookdown::html_document2:
-    number_sections: yes
-    toc: yes
-    toc_depth: 2
-    toc_float: yes
   bookdown::pdf_document2:
@@
     keep_tex: true
+  bookdown::html_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
+    toc_float: yes
   bookdown::word_document2:
@@
 titlepage: yes
-bibliography: mybib.bib
+bibliography: mybib
 documentclass: article
@@
 x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
-plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Median age at death | 65", ylim = c(70, 90), main = "Women", axes = FALSE)
+plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Women", axes = FALSE)
 axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
@@
 x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
-plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Median age at death | 65", ylim = c(70, 90), main = "Men", axes = FALSE)
+plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Men", axes = FALSE)
 axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)

### 2018-02-05 article3b.Rmd

@@
 titlepage: yes
-bibliography: mybib
+bibliography: mybib.bib
 documentclass: article

### 2018-02-05 article3b.Rmd
nonsense

--- 2018-02-05 article3b.Rmd
@@
 header-includes:
-  - \usepackage{endfloat}
   - \usepackage[utf8]{inputenc}

### 2018-02-05 article3b.Rmd
fix some text details

@@
 titlepage: yes
-bibliography: mybib
+bibliography: "mybib.bib"
 documentclass: article
@@
 citation_package: natbib
-biblio-style: apalike
+biblio-style: "apalike"
 header-includes:
@@
 economic resources, status, social power, etc.
-We have chosen to work with a broad definition of classes, merging the original 12 social classes in Hisclass into three, representing different levels of control and access to resources vital for life chances, and an additional fourth category of unknown class:
+We have chosen to work with a broad definition of classes, merging the original 12 social classes in Hisclass into four, representing different levels of control and access to resources vital for life chances, and an additional fifth category of unknown class:
 
@@
 \item {\tt farmer}, Hisclass 8. Farmers. 
+\item {\tt NA}, Not Available, unknown. 
 \end{enumerate}
@@
 in the 19th century. The increase in academics in connection to the
-establishment of Umeå University explains much of this increase in later time. What is
-not apparent in Figure \@ref(fig:hiscskumm) is the transfer within the middle
-group from a majority of farmers during the 19th century to lower white-collar from the middle of the 20th century.
-Furthermore we observe a rather high percentage of missing social class in the early 19th century.
-For men the proportion is somewhat higher in the age group 65 and above, those who were retired, but still low enough to be used in our analyses (results not shown).
-If we disregard the very first period, however, there were very few with occupation missing until 1950.
+establishment of Umeå University explains much of this increase in later time. Furthermore, 
+we observe a rather high percentage of missing social class in the early 19th century. 
+For men the proportion is somewhat higher in the age group 65 and above, those who were retired, 
+but still low enough to be used in our analyses (results not shown). If we disregard the very first period, 
+however, there were very few with occupation missing until 1950.
 
@@
 
-Looking at the table below, it is obvious that the statistical significance is strong.
+Looking at Table \@ref(tab:cumswtab), it is obvious that the statistical significance is strong.
 
-```{r cumswtab,cap="Table"}
+```{r cumswtab,message=FALSE}
 vb3064$social_class <- vb3064$hisclass
@@
 dr <- drop1(fitt, test = "Chisq")
-dr
-##knitr::kable(dr)
+##dr
+knitr::kable(dr, caption = "Single term deletions, ages 30-64.")
 ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."), 
@@
 1851--1950, men having about three times higher risks than women in the
-same class (basically living in the same households). In the other classes, the difference was much smaller. It was men in high positions who experienced the largest risks.
+same class (basically living in the same households). In the other classes, 
+the difference was much smaller. It was men in high positions who experienced the largest risks.
 
@@
 We now turn to the elderly, those who in most cases have left the workforce.
-We use the *median remaining life at 65* (reported as median age at death) as the summary measure. Looking at the table below, it is obvious that the statistical significance is still strong.
+We use the *median remaining life at 65* (reported as median age at death) as the summary measure.
+Looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance is still strong.
 
@@
 drold <- drop1(fittold, test = "Chisq")
-##knitr::kable(dr)
-drold
+knitr::kable(dr, caption = "Single term deletions, ages 65-99.")
+##drold
 ##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--99."), 
@@
 
-```{r plotrawwom, fig.cap="Median age at death for a 65 year old married woman on the countryside by SES and period.", fig.height = 5, out.height = "10cm"}
+```{r plotrawwom, fig.cap="Median age at death for a 65 year old married woman on the countryside by SES and period.", fig.height = 4, out.height = "10cm"}
 x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
@@
 }
-legend("bottomright", legend = hc, lty = 1:5, pch = 1:5, col = scol)
+legend("bottomright", legend = hc, lty = 1:5, pch = 1:5, col = scol, cex = 0.6)
 ```
@@
 
-```{r plotrawm, fig.cap="Median age at death for a 65 year old married man on the countryside by SES and period.", fig.height = 5}
+```{r plotrawm, fig.cap="Median age at death for a 65 year old married man on the countryside by SES and period.", fig.height = 4}
 x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)

### 2018-02-05 R/plotED.glm.R
smaller font in legend

@@ -71,6 +71,6 @@ plotED.glm <- function(res,
     }
     abline(v = xv[out], lty = 3, col = "magenta")
     abline(h = 0)
-    legend("topright", legend = leg, col = col, pch = 1:m, lty = 1:m)
+    legend("topright", legend = leg, col = col, pch = 1:m, lty = 1:m, cex = 0.7)
     invisible(outp)
 }

### 2018-02-06 article3b.Rmd
fix out.height

@@
   - \usepackage[utf8]{inputenc}
+  - \usepackage{graphicx}
 abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class.
@@
 
-```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height='50%'}
+```{r mapp,fig.cap = "(ref:cities)", fig.height=6,fig.align='center', out.height="5in"}
 knitr::include_graphics("myfigs/sweden.png")
@@
 
-```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
+```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4}
 xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '50%', fig.scap="Exposure by social class and time period, women.",fig.height=4}
+```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
 xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 
-```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = '50%',fig.height=4}
+```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4}
 xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))

### 2018-02-06 article3.Rmd
Some revisions

@@
 A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population.
 The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
@@
 
-Based on the results, we argue that high social class is not necessarily always favourable for survival.
+Based on the results, we argue that high social class is not necessarily favourable for survival.
 Social conditions and social position certainly have impact, but not always in the expected direction.
-In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date there appear a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
+In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 and other resources. In present-day welfare societies, social position, either according to social class, education or income, is a
-strong determinant when it comes to health and mortality and its impact even
-seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
+strong determinant when it comes to health and mortality and its impact are even
+increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
 "... social conditions have been, are and will continue to be irreducible
@@
 continuity in the determinants of social health inequalities from adulthood to old age.
-The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages tend to persist and accumulate during life in a negative spiral rewarding some while disfavouring others.
+The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral rewarding some while disfavouring others.
 This leads to larger differences in old age.
@@
 @wipi09 argue that income inequality has an independent effect on mortality, separate from the direct effect of actual access to economic resources.
-They argue
+They find
 that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the todays economically developed world.
@@
 If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country.
-The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
-of Bothnia. This is a remote part of the country where communication with
+The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf of Bothnia.
+Communication with
 the rest of Sweden was difficult until the late 19th century. The economy
@@
 This region has a somewhat different character to that of Skellefteå.
-Umeå town had a small population but was substantially
-larger than Skellefteå town for almost the whole studied period and was
+Umeå town had a small population, although substantially
+larger than Skellefteå town, for almost the whole studied period and was
 the administrative, educational and military centre of the county of Västerbotten.
@@
 It would be interesting to compare the results from this study with alternative measures, but reliable data on income and education are unfortunately rarely available in historical population databases stretching back as far as in this article.
-In the future, however, it will be possible to utilize a database on the 1930 census that can be linked to the regions studied here.
-That census includes information on income and wealth as well as educational level.
 
@@
 People are situated within socio-economic and political contexts, as well as prevailing societal value systems.
-The importance of understanding people in their context is also emphasized by @kri05.
-What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy offers advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which become embodied in individuals; but even though these advantages were at hand in our studied regions, this did not result in better health for men in the most privileged groups until the late 20th century.
-The embodiment took another direction.
 
-We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways.
+We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position.
 In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles.
@@
 The second pathway we discuss relates to the health risks of belonging to a certain social class.
-We usually assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article.
+We usually assume that health risks are higher in social classes with less access to resources.
 It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position.

### 2018-02-07 pure3b.Rmd

--- 2018-02-05 article3b.Rmd
@@
 ---
-title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
+title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013 (cohort version)"
 author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
@@
 source("R/get.file.R")
-skel <- get.file("skel")
+skel <- get.file("rawskel")
+skel$urban <- FALSE
 ##
-skume <- get.file("skume")
+skume <- get.file("rawskume")
 ##
-foli <- get.file("foli")
-skel$urban <- FALSE
-vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
-vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
-vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
-                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
-                    FUN = sum)
+su <- rbind(skel[, c("id", "birthdate", "sex", "period", "civst", "hisclass", "enter", "exit", "event")], 
+            skume[, c("id", "birthdate", "sex", "period", "civst", "hisclass", "enter", "exit", "event")])
+su$id <- paste("s", su$id, sep = "_")
+su <- rc(su)
+Exit <- tapply(su$exit, su$id, max)
+Event <- tapply(su$event, su$id, max)
+idx <- tapply(su$id, su$id)
+su$Exit <- Exit[idx]
+su$Event <- Event[idx]
+su <- su[su$lopnr == 1, ]
+su$exit <- su$Exit
+su$event <- su$Event
+
+foli <- get.file("rawfoli")
+
+foli <- foli[, c("id", "birthdate", "sex", "period", "civst", "hisclass", "enter", "exit", "event")]
+foli <- rc(foli)
+foli$id <- paste("f", foli$id, sep = "_")
+Exit <- tapply(foli$exit, foli$id, max)
+Event <- tapply(foli$event, foli$id, max)
+idx <- tapply(foli$id, foli$id)
+foli$Exit <- Exit[idx]
+foli$Event <- Event[idx]
+foli <- foli[foli$lopnr == 1, ]
+foli$exit <- foli$Exit
+foli$event <- foli$Event
+
+vb0 <- rbind(su, foli)
+rownames(vb0) <- 1:NROW(vb0)
+
+##vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
+##vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
+##                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
+##                    FUN = sum)
 ```

### 2018-02-07 article3b.Rmd
fix time periods

@@
 skel$urban <- FALSE
-vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
+su <- rbind(skel, skume[, names(skel)])
+
+vb0Tab <- rbind(su, foli[, names(su)])
+
 vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
@@
                     FUN = sum)
+vb0Tab <- vb0Tab[!(vb0Tab$period %in% c("1961-1970", "1971-1975")), ]
+vb0Tab$period <- factor(vb0Tab$period)
 ```
@@
 rawskume <- rawskume[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
+raw <- rbind(rawskel, rawskume)
+raw$id <- paste("pop", raw$id, sep = "_")
 ##
@@
 rawfoli <- rawfoli[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
-raw <- rbind(rawskel, rawskume, rawfoli)
+raw <- rbind(raw, rawfoli)
+raw <- raw[!(raw$period %in% c("1961-1970", "1971-1975")), ]
+raw$period <- factor(raw$period)
 raw <- age.window(raw, c(65, 100))
@@
     med[i, ] <- summary(survfit(fit, newdata = data.frame(civst = "married", urban = FALSE)))$table[, "median"]
+}
+for (i in 1:n.hc){
+    fit <- coxph(Surv(enter, exit, event) ~ strata(period), data = raw[raw$sex == "male" & 
+                                                                              raw$hisclass == hc[i], ])
+    med[i, ] <- summary(survfit(fit))$table[, "median"]
 }

### 2018-02-08 article3b.Rmd
Some changes

@@
 strong determinant when it comes to health and mortality and its impact even
-seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
+are increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
 "... social conditions have been, are and will continue to be irreducible
@@
 During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state.
-If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country.
+If Sweden is peripheral from a European continental perspective, the regions studied here are remote from the central parts of the country.
 The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
-of Bothnia. This is a remote part of the country where communication with
+of Bothnia, where communication with
 the rest of Sweden was difficult until the late 19th century. The economy
@@
 This region has a somewhat different character to that of Skellefteå.
-Umeå town had a small population but was substantially
-larger than Skellefteå town for almost the whole studied period and was
-the administrative, educational and military centre of the county of Västerbotten.
-During the latter part of the 20th century, the establishment of a university in Umeå led to a rapid population increase.
+Umeå town had for a long time a small population, although substantially, and was the administrative, educational and military centre of the county of Västerbotten.
+During the latter part of the 20th century, the establishment of Umeå University led to a rapid population increase.
 Agriculture dominated the rural part, but there were also some foundries as well as forestry industries.
-Consequently the economy was more diversified than that of Skellefteå.The population size as defined in our data sets (all ages) was 19,138 on January 1, 1901 and 103,970 when the 20th century ended.
+Consequently the economy was more diversified than that of Skellefteå.
+The population size as defined in our data sets (all ages) was 19,138 on January 1, 1901 and 103,970 when the 20th century ended.
 
@@
 [@malm10], which is based on different linked national population
-registers from 1960 to 2013 and is used within the ageing programme at CEDAR, Umeå
+registers from 1960 to 2013 (censuses, LISA from Statistics Sweden and cause of death registers) and is used within the ageing programme at CEDAR, Umeå
 University.
-Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis).
+Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data from 1976 onwards.
 The study period from 1976 to 2000 is constructed from
@@
 the 1975 census has a follow-up from 1 January 1976 until the end of 1980).
-The expection of thisis the 1990 census for which the follow-up is until 2001.
+The exception of this is the 1990 census for which the follow-up is until 2001.
 Thus, the deaths do not necessarily take place in our region for the Linnaeus analyses, although this is usually the case.
@@
 the husband/father, assuming that the family shares the same socio-economic
-position.This is a reasonable approach for this period.
+position. We consider this as a reasonable approach for this period.
 
@@
 Although the results for the different periods are not completely comparable, the difference in approach reflects an actual change in how social class is structured.
-Notice, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].
 
@@
 (per cent) of exposure time according to class for men and women,
-respectively. The category of workers is fairly stable over time. The
-highest class has increased substantially, albeit from a very low level
+respectively.
+The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women.
+The category of workers is fairly stable over time.
+The highest class has increased substantially, albeit from a very low level
 in the 19th century. The increase in academics in connection to the

### 2018-02-10 pure3b.Rmd
introduce inverse probabiity weighting

@@
 
+## Inverse probability weighting
+
+```{r ipw}
+head(vb0)
+da <- vb0[vb0$sex == "male" & vb0$period == "1876-1900", c("id", "civst", "hisclass", "enter", "exit", "event")]
+gg <- as.numeric(da$hisclass)
+lfit1 <- glm(gg == 1 ~ civst, data = da, family = "binomial")
+p1 <- predict(lfit1, type = "response")
+lfit2 <- glm(gg == 2 ~ civst, data = da, family = "binomial")
+p2 <- predict(lfit2, type = "response")
+lfit3 <- glm(gg == 3 ~ civst, data = da, family = "binomial")
+p3 <- predict(lfit3, type = "response")
+lfit4 <- glm(gg == 4 ~ civst, data = da, family = "binomial")
+p4 <- predict(lfit4, type = "response")
+lfit5 <- glm(gg == 5 ~ civst, data = da, family = "binomial")
+p5 <- predict(lfit5, type = "response")
+tmp <- p1
+tmp[gg==2] <- p2[gg==2]
+tmp[gg==3] <- p3[gg==3]
+tmp[gg==4] <- p4[gg==4]
+tmp[gg==5] <- p5[gg==5]
+wgt <- 1 / tmp
+```
+
 # The effect of social class on mortality

### 2018-02-11 article3c.Rmd
introduce marginal modeling

--- 2018-02-10 article3b.Rmd
@@
 We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
-The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
+The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
+In this version of article3, we are adopting a marginal approach, in contrast to earlier versions, where the approach was conditional."
 ---
@@
 combination, the total amount of exposure and the number of deaths are
-calculated.  We have two sexes, three marital statuses, ten periods, two
-urban/rural groups, four social classes,
-ten age groups, and 22 cohorts; in all, 105,600 possible
+calculated.  We have 2 sexes, 3 marital statuses, 12 periods, 2
+urban/rural groups, 5 social classes,
+14 age groups, and 22 cohorts; in all, 221,760 possible
 combinations. However, in this scheme there are many structural zeroes
 (impossible combinations) and some random zeros, so the actual number of
-combinations we end up with is 14,084. The data set is thus reduced in
+combinations we end up with is 17,181. The data set is thus reduced in
 size by 99 per cent.
+
+## The conditional model
 
@@
 
-# The effect of social class on mortality
+## The marginal model
+
+A alternative, and preferable, approach is the marginal model, where the disturbace effect of confounders is 
+balanced out through weighting. In our case , `civil status` and `urban/rural` are confounders, that is, they 
+are correlated to both our outcome, mortality, and our explanatory variable, social class. The latter association 
+manifests itself through the facts that the marriage frequency varies social class and certain social classes are 
+more common in the rural area than others (we find farmers on the countryside).
 
@@
 ```
+
+### Association between civil status and social class over time
+
+In order to simplify the illustration, we look at `civil status` in two groups, `ever married` and `unmarried`, in Figure \@ref(fig:assesciv).
+
+```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4}
+orr <- aggregate(x = vb0Tab$exposure, by = vb0Tab[, c("period", "civst", "hisclass")], FUN = sum)
+
+levper <- levels(orr$period)
+nper <- length(levper)
+levhis <- levels(orr$hisclass)
+nhis <- length(levhis)
+out <- matrix(0, nrow = nhis, ncol = nper)
+for (per in 1:nper){
+    for (his in 1:nhis){
+        oo <- orr[orr$period == levper[per] & orr$hisclass == levhis[his], ]
+        nam <- sum(oo$x)
+        out[his, per] <- sum(oo$x[oo$civst != "unmarried"]) / nam
+    }
+}
+out <- 100 * out
+xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1978, 1982, 1988, 1995, 2005, 2011)
+plot(xli, out[1, ], type = "b", xlab = "", ylab = "Fraction ever married", 
+     axes = FALSE, col = scol[1], lty = 1, pch = 1, ylim = c(0,100))
+axis(1, at = xli, labels = levper, las = 2, cex = 0.7)
+axis(2, las = 1)
+box()
+abline(h = 0)
+for (r in 2:nhis){
+    lines(xli, out[r, ], col = scol[r], lty = r, pch = r, type = "b")
+}
+legend("bottomright", legend = levhis, lty = 1:5, col = scol, pch = 1:5)
+```
+
+### Association between urban/rural and social class over time
+
+```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4}
+orr <- aggregate(x = vb0Tab$exposure, by = vb0Tab[, c("period", "urban", "hisclass")], FUN = sum)
+
+levper <- levels(orr$period)
+nper <- length(levper)
+levhis <- levels(orr$hisclass)
+nhis <- length(levhis)
+out <- matrix(0, nrow = nhis, ncol = nper)
+for (per in 1:nper){
+    for (his in 1:nhis){
+        oo <- orr[orr$period == levper[per] & orr$hisclass == levhis[his], ]
+        nam <- sum(oo$x)
+        out[his, per] <- sum(oo$x[oo$urban == TRUE]) / nam
+    }
+}
+out <- 100 * out
+xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1978, 1982, 1988, 1995, 2005, 2011)
+plot(xli, out[1, ], type = "b", xlab = "", ylab = "Fraction urban", 
+     axes = FALSE, col = scol[1], lty = 1, pch = 1, ylim = c(0,100))
+axis(1, at = xli, labels = levper, las = 2, cex = 0.7)
+axis(2, las = 1)
+box()
+abline(h = 0)
+for (r in 2:nhis){
+    lines(xli, out[r, ], col = scol[r], lty = r, pch = r, type = "b")
+}
+legend("bottomleft", legend = levhis, lty = 1:5, col = scol, pch = 1:5)
+```
+
+### Inverse frequency weighting
+
+In order to eliminate the disturbing effect on the association between hisclass and mortality by
+the confounders civil status and urban/rural, weights that eliminates the associations between hisclass 
+and the confounders are introduced. This is done period by period.
+
+As an illustration, take the period `1901-1925` with males. There are six combinations of civil status and urban/rural, and for 
+each we calculate the total exposure time. The total distribution by confounders is
+
+```{r exweight}
+exm <- vb0Tab[vb0Tab$period == "1901-1925" & vb0Tab$sex == "male", c("age", "civst", "urban", "exposure", "event", "hisclass")]
+all <- aggregate(exm["exposure"], by = exm[c("civst", "urban")], FUN = sum)
+all
+```
+
+and for the elite we have
+
+```{r forelite}
+elite <- exm[exm$hisclass == "elite", c("age", "civst", "urban", "exposure", "event", "hisclass")]
+el <- aggregate(elite["exposure"], by = elite[, c("civst", "urban")], FUN = sum)
+el
+```
+
+The weights are now 'all$exposure / el$exposure`, that is
+
+```{r wei}
+all$wei <- all$exposure / el$exposure
+all
+```
+
+Put these weights on the data frame `elite`:
+
+```{r putonel}
+all$id <- paste(all$civst, all$urban, sep = "_")
+elite$id <- paste(elite$civst, elite$urban, sep = "_")
+indx <- match(elite$id, all$id)
+elite$wght <- all$wei[indx]
+```
+
+Now repeat this for the other four hisclasses:
+
+```{r therest}
+middle <- exm[exm$hisclass == "middle", c("age", "civst", "urban", "exposure", "event", "hisclass")]
+el <- aggregate(middle["exposure"], by = middle[, c("civst", "urban")], FUN = sum)
+all$wei <- all$exposure / el$exposure
+middle$id <- paste(middle$civst, middle$urban, sep = "_")
+indx <- match(middle$id, all$id)
+middle$wght <- all$wei[indx]
+el
+##
+worker <- exm[exm$hisclass == "worker", c("age", "civst", "urban", "exposure", "event", "hisclass")]
+el <- aggregate(worker["exposure"], by = worker[, c("civst", "urban")], FUN = sum)
+all$wei <- all$exposure / el$exposure
+worker$id <- paste(worker$civst, worker$urban, sep = "_")
+indx <- match(worker$id, all$id)
+worker$wght <- all$wei[indx]
+el
+##
+farmer <- exm[exm$hisclass == "farmer", c("age", "civst", "urban", "exposure", "event", "hisclass")]
+el <- aggregate(farmer["exposure"], by = farmer[, c("civst", "urban")], FUN = sum)
+all$wei <- all$exposure / el$exposure
+farmer$id <- paste(farmer$civst, farmer$urban, sep = "_")
+indx <- match(farmer$id, all$id)
+farmer$wght <- all$wei[indx]
+el
+##
+non <- exm[exm$hisclass == "NA", c("age", "civst", "urban", "exposure", "event", "hisclass")]
+el <- aggregate(non["exposure"], by = non[, c("civst", "urban")], FUN = sum)
+all$wei <- all$exposure / el$exposure
+non$id <- paste(non$civst, non$urban, sep = "_")
+indx <- match(non$id, all$id)
+non$wght <- all$wei[indx]
+el
+```
+
+And put together:
+
+```{r puttog}
+per1901 <- rbind(elite, middle, worker, farmer, non)
+per1901$age <- factor(per1901$age)
+```
+
+Let us fit with and without weights. Without weights:
+
+```{r testfit, fig.cap = "Survival by hisclass, men 1901-1925, non-weighted analysis.", fig.height = 4}
+fit1 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "elite", ], family = poisson)
+co1 <- fit1$coefficients
+co1[2:length(co1)] <- co1[2:length(co1)] + co1[1]
+co1 <- exp(co1)
+##
+fit2 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "middle", ], family = "poisson")
+co2 <- fit2$coefficients
+co2[2:length(co2)] <- co2[2:length(co2)] + co2[1]
+co2 <- exp(co2)
+##
+fit3 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "worker", ], family = "poisson")
+co3 <- fit3$coefficients
+co3[2:length(co3)] <- co3[2:length(co3)] + co3[1]
+co3 <- exp(co3)
+##
+fit4 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "farmer", ], family = "poisson")
+co4 <- fit4$coefficients
+co4[2:length(co4)] <- co4[2:length(co4)] + co4[1]
+co4 <- exp(co4)
+##
+fit5 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "NA", ], family = "poisson")
+co5 <- fit5$coefficients
+co5[2:length(co5)] <- co5[2:length(co5)] + co5[1]
+co5 <- exp(co5)
+###
+cuts <- seq(5, 55, by = 5)
+x <- seq(0, 60, length = 1000)
+s1 <- ppch(x, cuts = cuts, levels = co1[1:12], lower.tail = FALSE)
+plot(x, s1, type = "l", col = scol[1], axes = FALSE, xlab = "age", ylab = "Surviving fraction after 30.")
+axis(1, at = seq(0, 60, by = 10), labels = seq(30, 90, by = 10))
+box()
+axis(2, las = 1)
+abline(h = 0)
+abline(v = 0)
+s2 <- ppch(x, cuts = cuts, levels = co2[1:12], lower.tail = FALSE)
+lines(x, s2, col = scol[2], lty = 2)
+s3 <- ppch(x, cuts = cuts, levels = co3[1:12], lower.tail = FALSE)
+lines(x, s3, col = scol[3], lty = 3)
+s4 <- ppch(x, cuts = cuts, levels = co4[1:12], lower.tail = FALSE)
+lines(x, s4, col = scol[4], lty = 4)
+s5 <- ppch(x, cuts = cuts, levels = co5[1:12], lower.tail = FALSE)
+lines(x, s5, col = scol[5], lty = 5)
+##
+```
+
+With weights, see Figure \@ref(fig:testfit2):
+
+```{r testfit2, fig.cap = "Survival by hisclass, men 1901-1925, weighted analysis.", fig.height = 4}
+fit1 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "elite", ], family = poisson, weights = wght)
+co1 <- fit1$coefficients
+co1[2:length(co1)] <- co1[2:length(co1)] + co1[1]
+co1 <- exp(co1)
+##
+fit2 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "middle", ], family = "poisson", weights = wght)
+co2 <- fit2$coefficients
+co2[2:length(co2)] <- co2[2:length(co2)] + co2[1]
+co2 <- exp(co2)
+##
+fit3 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "worker", ], family = "poisson", weights = wght)
+co3 <- fit3$coefficients
+co3[2:length(co3)] <- co3[2:length(co3)] + co3[1]
+co3 <- exp(co3)
+##
+fit4 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "farmer", ], family = "poisson", weights = wght)
+co4 <- fit4$coefficients
+co4[2:length(co4)] <- co4[2:length(co4)] + co4[1]
+co4 <- exp(co4)
+##
+fit5 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "NA", ], family = "poisson", weights = wght)
+co5 <- fit5$coefficients
+co5[2:length(co5)] <- co5[2:length(co5)] + co5[1]
+co5 <- exp(co5)
+###
+cuts <- seq(5, 55, by = 5)
+x <- seq(0, 60, length = 1000)
+s1 <- ppch(x, cuts = cuts, levels = co1[1:12], lower.tail = FALSE)
+plot(x, s1, type = "l", col = scol[1], axes = FALSE, xlab = "Age", ylab = "Surviving fraction after 30")
+axis(1, at = seq(0, 60, by = 10), labels = seq(30, 90, by = 10))
+axis(2, las = 1)
+box()
+abline(h = 0)
+abline(v = 0)
+s2 <- ppch(x, cuts = cuts, levels = co2[1:12], lower.tail = FALSE)
+lines(x, s2, col = scol[2], lty = 2)
+s3 <- ppch(x, cuts = cuts, levels = co3[1:12], lower.tail = FALSE)
+lines(x, s3, col = scol[3], lty = 3)
+s4 <- ppch(x, cuts = cuts, levels = co4[1:12], lower.tail = FALSE)
+lines(x, s4, col = scol[4], lty = 4)
+s5 <- ppch(x, cuts = cuts, levels = co5[1:12], lower.tail = FALSE)
+lines(x, s5, col = scol[5], lty = 5)
+##
+```
+
+# The effect of social class on mortality
 

### 2018-02-13 article3c.Rmd
introduce new weights

@@
 exm <- vb0Tab[vb0Tab$period == "1901-1925" & vb0Tab$sex == "male", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-all <- aggregate(exm["exposure"], by = exm[c("civst", "urban")], FUN = sum)
+all <- aggregate(exm["exposure"], by = exm[c("civst", "urban", "age")], FUN = sum)
 all
@@
 elite <- exm[exm$hisclass == "elite", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(elite["exposure"], by = elite[, c("civst", "urban")], FUN = sum)
+el <- aggregate(elite["exposure"], by = elite[, c("civst", "urban", "age")], FUN = sum)
 el
@@
 
-The weights are now 'all$exposure / el$exposure`, that is
+The weights are now 'all["exposure"]" / el["exposure"]`, that is
 
 ```{r wei}
-all$wei <- all$exposure / el$exposure
+wei <-  all$exposure / el$exposure
+all$wei <- wei 
+all$Expos <- all$exposure * wei
+all$Eve <- all$event * wei
 all
@@
 per1901$age <- factor(per1901$age)
-```
-
-Let us fit with and without weights. Without weights:
+#fit <- lm(exposure ~ civst * urban * age, data = per1901)
+#pred <- predict(fit, type = "response")
+#per1901$wght2 <- pred / per1901$exposure
+```
+
+Let us fit with and without weights. Without weights, see Figure \@ref(fig:testfit).
 

### 2018-02-13 R/ipw.R
first commit

@@ -1,0 +1,52 @@
+ipw <- function(dat){
+    ## Creates 'inverse probability weights'
+    ## Confounders: 'urban' and 'civst', 2 x 3 levels.
+    ## Independently for: 'period', 'sex', and 'age' (5-year intervals.)
+    
+    ## We add together "married" and "prev.married":
+    
+    levels(dat$civst) <- c("married", "unmarried", "married")
+    dat$civst <- factor(dat$civst)
+    
+    ## We aggregate over 'cohort':
+    dat <- aggregate(dat[c("event", "exposure")], by = dat[c("sex", "age", "period", "civst", "hisclass", "urban")], FUN = sum)
+    
+    dat <- dat[order(dat$sex, dat$period, dat$age), ]
+    rownames(dat) <- 1:nrow(dat)
+    dat$age <- factor(dat$age)
+    dat$wght <- NA
+    periods <- levels(dat$period)
+    n.per <- length(periods)
+    ages <- levels(dat$age)
+    n.ages <- length(ages)
+    
+    for (per in 1:n.per){
+        for (age in 1:n.ages){
+            for (sx in c("male", "female")){
+                ##cat("sex = ", sx, "\n")
+                select <- dat$period == periods[per] & dat$age == ages[age] & dat$sex == sx
+                ct <- dat[select, c("hisclass", "urban", "civst", "exposure", "event")]
+                ct$id <- paste(ct$civst, ct$urban, sep = "_")
+                all <- aggregate(ct[c("event", "exposure")], by = ct["hisclass"], FUN = sum)
+                indx <- match(ct$hisclass, all$hisclass)
+                totexpo <- all$exposure[indx]
+                wght <- totexpo / ct$exposure
+                ct$event <- wght * ct$event
+                ct$exposure <- wght * ct$exposure
+                out <- aggregate(ct[c("event", "exposure")], by = ct["hisclass"], FUN = sum)
+                out$sex <- sx
+                out$period <- periods[per]
+                out$age <- ages[age]
+                out$rate <- out$event / out$exposure
+                ##return(out)
+                if (per == 1 & age == 1 & sx == "male"){
+                    res <- out
+                }else{
+                    res <- rbind(res, out)
+                }
+                ##
+            }
+        }
+    }
+    res
+}

### 2018-02-13 article3b.Rmd
Some small revisions

@@
 University.
-Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data from 1976 onwards.
+Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we we restrict the analysis to the period 1976 onwards.
 The study period from 1976 to 2000 is constructed from
@@
 
-For the last period, 2002--2013, we use the yearly population registers
+For the last period, 2002--2013, we use the yearly population registers from Statistics Sweden
 (LISA data) together with information on deaths from the National Board of
@@
 For men the proportion is somewhat higher in the age group 65 and above, those who were retired, 
-but still low enough to be used in our analyses (results not shown). If we disregard the very first period, 
-however, there were very few with occupation missing until 1950.
+but still low enough to be used in our analyses (results not shown). If we disregard the very first period, however, there were very few with occupation missing until 1950.
 

### 2018-02-13 article3b.Rmd
Further revisions

@@
 By standardizing the levels in each time period for each class, we 
-illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
-the paper, we only present these
-relative mortality levels for the different classes since they are the main focus in this study.
+illustrate the relative risks in Figure \@ref(fig:cumswstand).
 
@@
 People are situated within socio-economic and political contexts, as well as prevailing societal value systems.
-The importance of understanding people in their context is also emphasized by @kri05.
-What she calls embodiment refers to the recognition that human beings are simultaneously social beings and biological organisms.
-Having a good economy offers advantages in many aspects related to health and mortality, for example good living conditions, access to food, medical care and better working conditions which become embodied in individuals; but even though these advantages were at hand in our studied regions, this did not result in better health for men in the most privileged groups until the late 20th century.
-The embodiment took another direction.
 
@@
 It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern.
-The second pathway we discuss relates to the health risks of belonging to a certain social class.
-We usually assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article.
+The second pathway relates to the health risks of belonging to a certain social class.
+We usually assume that health risks are higher in social classes with less access to resources.
 It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position.

### 2018-02-13 article3b.Rmd
Further revisions

@@
 
-Looking at Table \@ref(tab:cumswtab), it is obvious that the statistical significance is strong.
-
 ```{r cumswtab,message=FALSE}
@@
 
+Looking at Table \@ref(tab:cumswtab), it is obvious that the statistical significance is strong.
 The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
@@
 it comes to the agricultural sector. While farming occupations were common
-up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then. This means that the middle category has
-a different character in the last periods.
+up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then, and represent a very small minority at the enbd, particularly among women.
 
@@
 accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12].
-The highest social class usually has comparatively low mortality (except for the first period when the statistical power was weak), while
-working-class women have high mortality, but almost constantly about the same level as the middle category.
-However, there is no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide appears between workers and the middle category.
-A consistent social gradient has developed; the group with unknown social class is distinguished by much higher mortality from the 1976 onwards.
+The highest social class usually has comparatively low mortality (except for the first period when the statistical power was weak), while working-class women have high mortality.
+For women in the farming category, the mortality risks was at about the same levels as those of working-class women until the 1950s, while those in the middle category had somewhat lower mortality.
+However, there is no substantial change during the studied period in the relative risk pattern between classes except for the results of the marginal farming category after 1976.
+Female mortality is characterised by a consistent social gradient.
+The group with unknown social class is distinguished by much higher mortality from the 1976 onwards.
 This might have to do with how we have defined social position for women before and after 1950.
@@
 A completely different pattern appears among men, which is the most surprising finding in
-our analyses. The groups we expect to have the lowest mortality, if
+our analyses. The group we expect to have the lowest mortality, if
 we assume that access to vital resources and belonging to a high social class
-determines survival, instead have by far the highest hazards of dying
+determines survival, instead has by far the highest hazards of dying
 during the 19th century and much of the 20th. Workers, on the
-other hand, have comparatively low mortality until 1950, at the same
-level as the middle category.
+other hand, have comparatively low mortality until 1950, while mortality in the middle category represents an intermediate level.
+Farmers have similar levels as workers until the middle of the 20th century.
 The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.
@@
 A reversal of the class pattern in mortality has taken place, and the expected social gradient in mortality has developed.
-This is also what we usually find in analyses of social health inequalities in present-day
+This latter pattern is also what we usually find in analyses of social health inequalities in present-day
 Sweden as well as in most other countries, results that have formed the basis for many of the contemporary epidemiologic and demographic theories. In the regions we are studying,

### 2018-02-13 article3b.Rmd
Some corrections

@@
 
-The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories.
+The results showing a small or even reversed socio-economic gradient in the past and its late development are in accordance with other studies [@bevp11], for example by @bedr11 in a different setting, a couple of Scanian parishes in southern Sweden, albeit with somewhat different social categories.
 They find the first indications of an emerging social gradient to be a fairly recent phenomenon, in their case as late as the 1950s.
@@
 This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition.
-While the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it has played and is still playing a central role within epidemiology and social medicine.
+While the strong version of the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it continues to play a central role within epidemiology and social medicine.
 
@@
 
-We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways.
+We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position.
 In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles.
@@
 
-Social stress can also be expressed in unhealthy lifestyles, the third pathway we discuss.
+Social stress can also be expressed in unhealthy lifestyles -- the third pathway.
 Harmful behaviours can be connected to certain expectations associated with being a man of high status.
@@
 Overweight, obesity and physical inactivity were probably more common in higher social classes.
-@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers, which they suggest can be explained by the higher tobacco consumption among men in these groups.
+@vano89 find a higher total as well as cardiovascular mortality among self-employed men and non-manual male workers in Sweden during the 1960s, which they suggest can be explained by the higher tobacco consumption among men in these groups.
 The social pattern among women was the reverse, however, in line with the present study.
@@
 Mortality from this disease was substantially higher in the managerial and professional classes in 1921, being reversed in 1991 with the highest mortality in the lowest class [@cropr10].
-It could be discussed whether these indications of changes in health-related behaviour are sufficiently significant to explain the present findings; we believe they are certainly at least an important part of the story.
+Whether these indications of changes in health-related behaviour are sufficiently significant to explain the present findings can be discussed; we believe they certainly are at least an important part of the story.
 
@@
 The level of income inequality was certainly lower in our regions than in other parts of the country but it was still substantially higher in 1930 than towards the end of the 20th century.
-Compared with many other countries, the level of mortality has been fairly low.
+Compared with many other countries, the mortality level has been fairly low.
 The question of generalizability is relevant, which can hopefully be addressed in other studies, preferably with a comparative approach.

### 2018-02-14 article3b.Rmd
Further revisions

@@
 We now turn to the elderly, those who in most cases have left the workforce.
-We use the *median remaining life at 65* (reported as median age at death) as the summary measure.
+We use the *median remaining life at 65* (reported as median age at death) as the summary measure, here restricted to married people from the countryside.
 Looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance is still strong.
@@
 
-The general pattern and development of mortality (Figures \@ref(fig:plotrawwom) and \@ref(fig:plotrawm)) do not
-differ much from the younger age group, of course representing a much
+The general pattern and development of mortality (Figures \@ref(fig:plotrawwom) and \@ref(fig:plotrawm)) do not differ much from the younger age group, of course representing a much
 higher mortality level earlier in history. Mortality among women, as
-expected, was lower than among men. We find a clear decline in mortality, but
-the fastest decline did not set in definitely until sometime after the
+expected, was lower than among men. We find a clear increase in survival, but
+the fastest increase did not set in definitely until sometime after the
 Second World War; that is, later than for the age
@@
 Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status-levelling hypothesis. For women the results are mixed. We do not find the
-same clear advantage for elite women during the period before 1950, but neither
-do we find any large disadvantage for working-class elderly women at that
+we find for the younger age group in line with the status-levelling hypothesis.
+For women the results are mixed.
+We do not find the same clear advantage for elite women during the period before 1950, but neither do we find any large disadvantage for working-class elderly women at that
 time.
-In most periods the social differences were small, particularly between the workers and the middle group, and the elite women even had somewhat higher mortality during some periods until the 1901--1925 period.
+In most periods the social differences were small, particularly between the workers, farmers and the middle category, and the elite women even had somewhat lower mean age at death during some periods in the 19th century.
 The social differences continued to be small until the last decades of the
-20th century. It is only during the two last periods (after the year 2000) that it is
-possible to establish the expected social gradient in mortality.
+20th century. It is only from around 1980 that it is possible to establish the expected social gradient in mortality.
 
 For men we find indications of the same intriguing change among elderly as
-for the adult male population aged 30--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
+for the adult male population aged 30--64 years, but at a much lower level and less consistent.
+Mean age at death was somewhat lower among elite men during the 19th century, a pattern that prevailed until the middle of the
 the 20th century. It is not until the second half of the 20th
 century that the survival is better for the highest social class while
-workers of retirement age have slightly higher mortality. The difference between the worker and middle group has been small in most periods.
+workers of retirement age have slightly lower age at death. The difference between the worker, farmers and the middle group has been small in most periods.
 
@@
 While some report substantial differences, others have found surprisingly small or even reversed class differences, for example in the 19th century Sundsvall region [@edli11; @edbr12].
-This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century.
+This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century [@bevp11].
 The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions.

### 2018-02-16 R/ipw.R
change weight scedule: constant over age

@@ -21,22 +21,23 @@ ipw <- function(dat){
     n.ages <- length(ages)
     
     for (per in 1:n.per){
-        for (age in 1:n.ages){
+        ##for (age in 1:n.ages){
             for (sx in c("male", "female")){
                 ##cat("sex = ", sx, "\n")
-                select <- dat$period == periods[per] & dat$age == ages[age] & dat$sex == sx
-                ct <- dat[select, c("hisclass", "urban", "civst", "exposure", "event")]
-                ct$id <- paste(ct$civst, ct$urban, sep = "_")
+                select <- dat$period == periods[per] & dat$sex == sx
+                ct <- dat[select, c("hisclass", "urban", "civst", "exposure", "event", "age")]
+                ct$id <- paste(ct$civst, ct$urban, sep = "_") # A 'stickspår'?
                 all <- aggregate(ct[c("event", "exposure")], by = ct["hisclass"], FUN = sum)
                 indx <- match(ct$hisclass, all$hisclass)
                 totexpo <- all$exposure[indx]
                 wght <- totexpo / ct$exposure
+                ##wght <- 1
                 ct$event <- wght * ct$event
                 ct$exposure <- wght * ct$exposure
-                out <- aggregate(ct[c("event", "exposure")], by = ct["hisclass"], FUN = sum)
+                out <- aggregate(ct[c("event", "exposure")], by = ct[c("hisclass", "age")], FUN = sum)
                 out$sex <- sx
                 out$period <- periods[per]
-                out$age <- ages[age]
+                ##out$age <- ages[age]
                 out$rate <- out$event / out$exposure
                 ##return(out)
                 if (per == 1 & age == 1 & sx == "male"){
@@ -46,7 +47,7 @@ ipw <- function(dat){
                 }
                 ##
             }
-        }
+        #}
     }
     res
 }

### 2018-02-16 article3c.Rmd
trivia

@@
 
-### Inverse frequency weighting
+### Inverse exposure weighting
 
@@
 
+## All ages, survival curves
+
+```{r inipw}
+source("R/ipw.R")
+vb1Tab <- aggregate(vb0Tab[c("event", "exposure")], by = vb0Tab[c("sex", "age", "period", "civst", "hisclass", "urban")], FUN = sum, drop = TRUE)
+##problem <- vb1Tab$exposure <= 0 
+##vb1Tab$exposure[problem] <- 1
+##vb1Tab$exposure[vb1Tab$exposure < 1] <- 1
+##vb1Tab <- vb1Tab[vb1Tab$exposure > 1, ] # Removes extreme cases blowing up weights.
+cu <- ipw(vb1Tab)
+##sex <- "male"
+##period <- "1801-1825"
+##hisclass <- "farmer"
+##first <- cu[cu$sex == sex & cu$period == period & cu$hisclass == hisclass, c("rate", "event", "exposure")]
+##plot(first$rate, type = "b")
+cu$period <- factor(cu$period)
+cu$age <- as.numeric(cu$age)
+pers <- levels(cu$period)
+n.per <- length(pers)
+n <- n.per
+elite <- numeric(n)
+for (i in 1:n){
+    dat <- cu[cu$sex == "male" & cu$hisclass == "elite" & cu$period == pers[i], "rate"]
+    elite[i] <- sum(dat[1:7]) * 5
+}   
+##
+middle <- numeric(n)
+for (i in 1:n){
+    dat <- cu[cu$sex == "male" & cu$hisclass == "middle" & cu$period == pers[i], "rate"]
+    middle[i] <- sum(dat[1:7]) * 5
+}   
+kk <- middle
+##
+worker <- numeric(n)
+for (i in 1:n){
+    dat <- cu[cu$sex == "male" & cu$hisclass == "worker" & cu$period == pers[i], "rate"]
+    worker[i] <- sum(dat[1:7]) * 5
+}   
+##
+farmer<- numeric(n)
+for (i in 1:n){
+    dat <- cu[cu$sex == "male" & cu$hisclass == "farmer" & cu$period == pers[i], "rate"]
+    farmer[i] <- sum(dat[1:7]) * 5
+}   
+##
+no <- numeric(n)
+for (i in 1:n){
+    dat <- cu[cu$sex == "male" & cu$hisclass == "NA" & cu$period == pers[i], "rate"]
+    no[i] <- sum(dat[1:7]) * 5
+}
+
+plot(c(1812, 1838, 1862, 1888, 1912, 1938, 1978, 1982, 1988, 1995, 2005, 2010), elite / kk, type = "b", col = "black", lty = 1, pch = 1, ylim = c(0, 4))
+lines(c(1812, 1838, 1862, 1888, 1912, 1938, 1978, 1982, 1988, 1995, 2005, 2010), middle / kk, type = "b", col = "green", lty = 2, pch = 2)
+lines(c(1812, 1838, 1862, 1888, 1912, 1938, 1978, 1982, 1988, 1995, 2005, 2010), worker / kk, type = "b", col = "blue", lty = 3, pch = 3)
+lines(c(1812, 1838, 1862, 1888, 1912, 1938, 1978, 1982, 1988, 1995, 2005, 2010), farmer / kk, type = "b", col = "darkgreen", lty = 4, pch = 4)
+lines(c(1812, 1838, 1862, 1888, 1912, 1938, 1978, 1982, 1988, 1995, 2005, 2010), no / kk, type = "b", col = "red", lty = 5, pch = 5)
+```
+
 ## Ages 30--64

### 2018-02-22 article3.Rmd

@@
 During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state.
-If Sweden is remote from a European continental perspective, the regions studied here are remote from the central parts of the country.
+If Sweden is peripheral from a European continental perspective, the regions studied here are remote from the central parts of the country.
 The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf of Bothnia.
@@
 the administrative, educational and military centre of the county of Västerbotten.
-During the latter part of the 20th century, the establishment of a university in Umeå led to a rapid population increase.
+During the latter part of the 20th century, the establishment of Umeå University led to a rapid population increase.
 Agriculture dominated the rural part, but there were also some foundries as well as forestry industries.
@@
 1950 makes it possible to investigate the long-term development of social
-inequality in mortality in this part of Sweden. The data for the present study come from two large
-population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
+inequality in mortality in a part of Sweden. The data for the present study come from two large
+population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions.
 Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
@@
 [@malm10], which is based on different linked national population
-registers from 1960 to 2013 and is used within the ageing programme at CEDAR, Umeå
+registers from 1960 to 2013 (censuses, LISA and data from National Board of Health and Welfare) and is used within the ageing programme at CEDAR, Umeå
 University.
-Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis).
+Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis), we restrict the analysis to the period 1976 onwards.
 The study period from 1976 to 2000 is constructed from
-censuses every fifth year during the period 1975--1990, with additional information on deaths
+censuses every fifth year 1975--1990, with additional information on deaths
 from the National Board of Health and Welfare. For the period 2002--2013 we use
@@
 establishment of Umeå University explains much of this increase in later time. What is
-not apparent in Figure \@ref(fig:hiscskumm) is the transfer within the middle
+apparent in Figure \@ref(fig:hiscskumm) is the transition from a social distribution of  fer within the middle
 group from a majority of farmers during the 19th century to lower white-collar from the middle of the 20th century.

### 2018-02-28 article3c.Rmd
marginal efforts

@@
 are correlated to both our outcome, mortality, and our explanatory variable, social class. The latter association 
-manifests itself through the facts that the marriage frequency varies social class and certain social classes are 
+manifests itself through the facts that the marriage frequency varies by social class and certain social classes are 
 more common in the rural area than others (we find farmers on the countryside).
@@
 exm <- vb0Tab[vb0Tab$period == "1901-1925" & vb0Tab$sex == "male", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-all <- aggregate(exm["exposure"], by = exm[c("civst", "urban", "age")], FUN = sum)
+all <- aggregate(exm[, c("exposure", "event")], by = exm[, c("civst", "urban")], FUN = sum)
 all
@@
 elite <- exm[exm$hisclass == "elite", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(elite["exposure"], by = elite[, c("civst", "urban", "age")], FUN = sum)
+el <- aggregate(elite[, c("exposure", "event")], by = elite[, c("civst", "urban")], FUN = sum)
 el

### 2018-02-28 article3b.Rmd

@@
 ##
-foli <- get.file("foli")
+foli <- get.file("foli2")
+##load("../Data/foli2.rda")
 skel$urban <- FALSE
@@
 ##
-rawfoli <- get.file("rawfoli")
+rawfoli <- get.file("rawfoli2")
 rawfoli <- rawfoli[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
 raw <- rbind(raw, rawfoli)
-raw <- raw[!(raw$period %in% c("1961-1970", "1971-1975")), ]
+##raw <- raw[!(raw$period %in% c("1961-1970", "1971-1975")), ]
 raw$period <- factor(raw$period)
@@
 ```{r plotrawwom, fig.cap="Median age at death for a 65 year old married woman on the countryside by SES and period.", fig.height = 4, out.height = "10cm"}
-x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
-plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Women", axes = FALSE)
-axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
+x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1963, 1969, 1975, 1981, 1987, 1993, 1999, 2006, 2012)
+plot(x, med[1, ], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Women", axes = FALSE)
+axis(1, at = x, labels = peri, las = 2, cex.axis = 0.5)
 axis(2, las = 1)
@@
 for (i in 2:5){
-    lines(x, med[i, -7], col = scol[i], type = "b", lty = i, pch = i)
+    lines(x, med[i, ], col = scol[i], type = "b", lty = i, pch = i)
 }
@@
 ```{r plotrawm, fig.cap="Median age at death for a 65 year old married man on the countryside by SES and period.", fig.height = 4}
-x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
-plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Men", axes = FALSE)
-axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
+##x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
+plot(x, med[1, ], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Men", axes = FALSE)
+axis(1, at = x, labels = peri, las = 2, cex.axis = 0.5)
 axis(2, las = 1)

### 2018-02-28 CreData/fromPoplink.Rmd
first commit

---
title: "From Poplink"
author: "Göran Broström"
date: "28 February 2018"
output:
  pdf_document: default
  html_document: default
---

## Introduction

In this version we start at January 1, 1801, and we **do not** split the *poplink* data 
into two data sets (no causes of death, yet). End at December 31, 1950

```{r setup}
knitr::opts_chunk$set(cache=FALSE,echo=TRUE,message=TRUE,warning=TRUE,comment=NA)
library(xtable)
library(knitr)
library(eha)
library(skum)
##library(dplyr)
##library(tidyr)
##library(stringr)
```

## PopLink data

The POPLINK data cover 1801--1950. The resulting output consists of two tables,
**skel** and **skume** (I avoid the natural name **skum** in order to avoid a
mix-up with the **R** package by that name).
A covariate **period** is added and used for 
splitting the data into calendar time periods of length 10 years each.

### Get data

```{r getdata}
library(skum)
##load("../../Data/adult.rda")
##adult <- cal.window(adult, c(1801, 1951))
adult <- obs
indx <- match(adult$id, per$id)
adult$mid <- per$mid[indx]
adult$fid <- per$fid[indx]
adult$event <- !is.na(adult$sluttyp) & adult$sluttyp == 2
```

## Put on hisclass and civil status

```{r hisclass}
##New Hisclasses:
ett <- with(adult, !is.na(socBranch) &
              (hisclass == 1))
               ## ((socBranch == "official" & socStatus == "high") |
                 ##    (socBranch == "business" & socStatus == "high")
                ##))

tva <- with(adult, !is.na(socBranch) &
              (hisclass %in% c(2, 4)))
                ##((socBranch == "farming" & socStatus == "high") |
                ## (socBranch == "business" & socStatus == "low") |
                ## (socBranch == "official" & socStatus == "low")
                ##))

tre <- with(adult, !is.na(socBranch) &
              (hisclass %in% 5:7)) ## Original 'fyra' ---> 'tre'
              ##  ((socBranch == "farming" & socStatus == "low") |
                ##     (socBranch == "worker"))
                ##)
fyra <- with(adult, !is.na(socBranch) &
               (hisclass == 3))

hisc <- rep(-1, NROW(adult))
hisc[ett] <- 1
hisc[tva] <- 2
hisc[tre] <- 3
hisc[fyra] <- 4
is.na(hisc) <- hisc == -1
adult$hisclass <- hisc
```

Continue ...

```{r urbanetc}
adult$urban <- with(adult, !is.na(ortnmn) & (ortnmn %in% c("UMEÅ", "SKELLEFTEÅ STAD")))
levels(adult$civst) <- c("unmarried", "married", "prev.married", "prev.married", "prev.married")
```

## Propagate

```{r propagate}
##if (file.exists("adult.rda")){
  ##load("adult.rda")
if (TRUE){
source("R/fixCovar2.R")  
    adult$hisclass40 <- with(adult, fixCovar2(id, hisclass, enter, exit, 40, 20))
    adult$hisclass65 <- with(adult, fixCovar2(id, hisclass, enter, exit, 65, 45))
    adult$civst40 <- with(adult, fixCovar2(id, civst, enter, exit, 40, 20))
    adult$civst65 <- with(adult, fixCovar2(id, civst, enter, exit, 65, 45))
    adult$urban40 <- with(adult, fixCovar2(id, civst, enter, exit, 40, 20))
    adult$urban65 <- with(adult, fixCovar2(id, civst, enter, exit, 65, 45))
}else{
  adult <- adult[order(adult$id, adult$enter), ]
  adult <- group_by(adult, id) %>% fill_("hisclass")
  adult$hisclass[is.na(adult$hisclass)] <- 4
  adult$hisclass <- factor(adult$hisclass, labels = c("elite", "middle", "worker", "NA"))
  ##
  adult <- group_by(adult, id) %>%  fill_("civst")
  adult$civst[is.na(adult$civst)] <- "unmarried"
  save(adult, file = "adult.rda")
}
```

## The Lexis cut

Our data covers the years 1801--1950 and the ages 40--99.

```{r lexis}
##adult <- cal.window(adult, c(1801, 1951))
adult <- age.window(adult, c(40, 100))
```

## Split and period

```{r splitper}
source("R/addPer.R")
##skel <- cal.window(adult, c(1801, 1901))
##skel <- addPer(skel, cuts = seq(1801, 1901, by = 25))
skume <- cal.window(adult, c(1801, 1951))
skume <- addPer(skume, cuts = seq(1801, 1951, by = 25))
```

### Trim skume

```{r skumefilt}
skume <- skume[, c("id", "region", "birthdate", "sex", "enter", "exit", "event",
                    "period", "urban", "urban40", "urban65", "civst", "civst40", "civst65", 
                    "hisclass", "hisclass40", "hisclass65")]
skume <- skume[order(skume$id, skume$exit), ]
skume <- rc(skume)
indx <- tapply(skume$id, skume$id)
Exit <- with(skume, tapply(exit, id, max))
Enter <- with(skume, tapply(enter, id, min))
Event <- with(skume, tapply(event, id, max))
skume$Enter <- Enter[indx]
skume$Exit <- Exit[indx]
skume$Event <- Event[indx]
skume <- skume[!duplicated(skume$id), ]
skume$enter <- skume$Enter
skume$exit <- skume$Exit
skume$event <- skume$Event
skume$id <- paste("pop", skume$id, sep = "_")
skume$Exit <- NULL
skume$Enter <- NULL
skume$Event <- NULL
```

## Save

```{r save}
##save(skel, file = "skel.rda")
save(skume, file = "skume.rda")
```

### 2018-02-28 article3d.Rmd
first commit

--- 2018-02-28 article3c.Rmd
@@
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
-In this version of article3, we are adopting a marginal approach, in contrast to earlier versions, where the approach was conditional."
+In this version of article3, we are adopting a marginal approach, in contrast to earlier versions, where the approach was conditional.
+In addition to article3c, we are going back to individual, raw, data."
 ---
@@
 source("R/get.file.R")
-skel <- get.file("skel")
-##
-skume <- get.file("skume")
-##
-foli <- get.file("foli")
+skel <- get.file("rawskel")
+skel$region <- "ske"
 skel$urban <- FALSE
-su <- rbind(skel, skume[, names(skel)])
-
-vb0Tab <- rbind(su, foli[, names(su)])
-
-vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
-vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
-                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
-                    FUN = sum)
-vb0Tab <- vb0Tab[!(vb0Tab$period %in% c("1961-1970", "1971-1975")), ]
-vb0Tab$period <- factor(vb0Tab$period)
+vars <- c("id", "birthdate", "sex", "period", "urban", "civst", "hisclass", "enter", "exit", "event")
+skel <- skel[, vars]
+##
+skume <- get.file("rawskume")
+skume <- skume[, vars]
+##
+vb <- rbind(skel, skume)
+vb <- vb[order(vb$id, vb$exit), ]
+vb$id <- paste("vb", vb$id, sep = "_")
+##
+foli <- get.file("rawfoli")
+foli <- foli[, vars]
 ```

### 2018-02-28 CreData/R/addPer.R

@@ -1,0 +1,23 @@
+addPer <- function(dat, name = "period", cuts = seq(1901, 1951), by = 10){
+    n <- length(cuts)
+    start <- cal.window(dat, c(cuts[1], cuts[2]))
+    start[[name]] <- 1
+
+    if (n > 2){
+        for (i in (3:n)){
+            tmp <- cal.window(dat, c(cuts[i - 1], cuts[i]))
+            tmp[[name]] <- i - 1
+            start <- rbind(start, tmp)
+        }
+    }
+    levs <- character(n - 1)
+
+    for (i in 1:(n - 1)){
+        levs[i] <- paste(cuts[i], "-", cuts[i + 1] - 1, sep = "")
+    }
+    start[[name]] <- factor(start[[name]], labels = levs)
+    start <- start[order(start$id, start$enter), ]
+    start <- rc(start)
+    start
+}
+

### 2018-02-28 CreData/R/fixCovar2.R

@@ -1,0 +1,45 @@
+fixCovar2 <- function(id, covar, enter, exit, age, lookBack = 0){
+    ## The difference to the function 'fixCovar' is the argument
+    ## 'lookBack', which means that we search for the value in an
+    ## interval (age - lookBack, age]. The last non-NA value (if any)
+    ## is chosen, otherwise NA.
+
+    ## The assumption here is NOT that covar is positive numeric, with
+    ## the value 0 (zero) a code for missing.
+
+    ord <- order(id, enter)
+    id <- id[ord]
+    covar <- covar[ord]
+    enter <- enter[ord]
+    exit <- exit[ord]
+
+    isF <- is.factor(covar)
+    if (isF){
+        ##print(table(covar, useNA = "ifany"))
+        covar <- factor(covar) # remove unused levels
+        levs <- levels(covar)
+        ##cat("levs = ", levs, "\n")
+        covar <- as.numeric(covar)
+    }
+
+    out <- enter > age | exit < age - lookBack
+    is.na(covar) <- out | is.na(covar)
+
+    getval <- function(x){ # Get the last non-NA value
+        if (length(x) <= 1 | all(is.na(x))){
+            val <- x[1]
+        }else{
+            y <- x[!is.na(x)]
+            val <- y[length(y)]
+        }
+        val
+    }
+
+    indx <- tapply(id, id)
+
+    ret <- tapply(covar, id, getval)[indx]
+    if (isF){
+        ret <- factor(ret, levels = 1:length(levs), labels = levs)
+    }
+    ret
+}

### 2018-02-28 CreData/fromPoplink.Rmd
fix hisclass

@@
     adult$hisclass40 <- with(adult, fixCovar2(id, hisclass, enter, exit, 40, 20))
+    adult$hisclass40[is.na(adult$hisclass40)] <- 5
+    adult$hisclass40 <- factor(adult$hisclass40, labels = c("elite", "middle", "worker", "farmer", "none"))
     adult$hisclass65 <- with(adult, fixCovar2(id, hisclass, enter, exit, 65, 45))
+    adult$hisclass65[is.na(adult$hisclass65)] <- 5
+    adult$hisclass65 <- factor(adult$hisclass65, labels = c("elite", "middle", "worker", "farmer", "none"))
+
     adult$civst40 <- with(adult, fixCovar2(id, civst, enter, exit, 40, 20))
     adult$civst65 <- with(adult, fixCovar2(id, civst, enter, exit, 65, 45))
+
     adult$urban40 <- with(adult, fixCovar2(id, civst, enter, exit, 40, 20))
     adult$urban65 <- with(adult, fixCovar2(id, civst, enter, exit, 65, 45))
+
 }else{

### 2018-03-01 CreData/fromFobLisa.Rmd
first commit

--- 2016-11-01 CreateData/showData4.Rmd
@@
 ---
-title: "showData 4"
+title: "fromFobLisa, Version 2"
 author: "Göran Broström"
-date: "September 22, 2016"
+date: "March 1, 2018"
 output: pdf_document
@@
 
-Now *showData4.Rmd* is updated to include *fob90* and also let the
+Now *fromFobLisa.Rmd* is updated to include *fob90* and also let the
 follow-up for *fob85* be *five years* and the follow-up for *fob90* be *ten
-years*. We also add, as an update to *showData3*, guys aged 20 or more to the 
-*FoB data*. Note that results (data files) are saved in **data4**!
+years*.
 
-This is an updated version of *showData2.Rmd*, where we use the newly-found
+This is an updated version of *showData4.Rmd*, where we used the newly-found
 knowledge that the yearly *lisa* data relates to *December 31* the given
@@
 they die within five years after the first appearance. So, *fob60* induces
-a time interval *from November 1, 1960 to October 31, 1965*, and so on. To
-include all 40+ in this interval, we must select all that are *born before
-November 1, 1925*, that is, those who have reached the age 35 at November
-1, 1960. Since we do not have birth dates but only birth year and quarter,
+a time interval *from November 1, 1960 to October 31, 1965*, and so on. 
+Since we do not have birth dates but only birth year and quarter,
 we decide to use as birthdate the *midpoint of the birth quarter*. 
+WE work with all ages 0-110, and only as the final step, `age.window` is called.
 
@@
 
-This time we start with the *inddata* file and treat it as a *fixed-data*
+We start with the *inddata* file and treat it as a *fixed-data*
 file (no time-varying info). 
 We find the birth dates (see above), death dates (for those who are observed die), and the "date last seen".
-We also remove individuals that were born 1975 or later because they never get the chance to reach forty before *December 31, 2014*, which is our stop date.
 
@@
 ```{r frominddata}
-load("data/inddata.rda")
-load("data/sex.rda")
+load("../../Data/inddata.rda")
+load("../../Data/sex.rda")
 indx <- match(inddata$LINNEID, sex$LINNEID)
 inddata$KON <- sex$KON[indx]
-per <- inddata[!is.na(inddata$KON) & (inddata$FODELSEAR < 1975), ]
+per <- inddata[!is.na(inddata$KON), ]
 per$birthdate <- per$FODELSEAR + 0.125  +  (per$FODELSEKVARTAL - 1) / 4
@@
 ```{r styronper}
-load("data/socstyr.rda")
+load("../../Data/socstyr.rda")
 indx <- match(per$LINNEID, socstyr$LINNEID)
@@
 ```{r fob60}
-load("data/fob.rda")
+load("../../Data/fob.rda")
 fob60 <- fob[fob$AR == 1960 & 
@@
 rownames(fobdata) <- 1:NROW(fobdata)
-save(fobdata, file = "data4/fobdata.rda")
+save(fobdata, file = "fobdata.rda")
 ```
@@
 ```{r lisadat}
-load("data/lisa.rda")
+load("../../Data/lisa.rda")
 lisa <- lisa[lisa$KOMMUN %in% ourkom, ]
@@
 rownames(lisadata) <- 1:NROW(lisadata)
-save(lisadata, file = "data4/lisadata.rda")
+save(lisadata, file = "lisadata.rda")
 ```

### 2018-03-02 CreData/fromFobLisa.Rmd
continue

@@
 ourkom <- c(ourkom, 2411, 2412, 2413, 2416, 2415, 2461, 2407, 2402, 2408)
+fob <- fob[fob$KOMMUN %in% ourkom, ]
 ```
@@
 fob60$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1940-11-01"))
-fob60 <- fob60[fob60$birthdate < lim, ]
+fob60$fobdate <- as.numeric(toTime("1960-11-01"))
+##lim <- as.numeric(toTime("1940-11-01"))
+##fob60 <- fob60[fob60$birthdate < lim, ]
 indx <- match(fob60$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob60$enter <- lim + 20 - fob60$birthdate
+fob60$enter <- fob60$fobdate - fob60$birthdate
 dd <- per$deathdate[indx]
-fob60$event <- !(is.na(dd)) & (dd <= lim + 25) # NOTE: 40 --> 25!
+##fob60$event <- !(is.na(dd)) & (dd <= lim + 25) # NOTE: 40 --> 25!
+fob60$event <- !(is.na(dd)) & (dd <= fob60$fobdate +5)
 fob60$exit <- fob60$enter + 5

### 2018-03-03 CreData/R/extractFob.R
first commit

@@ -1,0 +1,26 @@
+extractFob <- function(year, followUp, data){
+    ## data = fob
+    out <- fob[fob$AR == year & 
+                     fob$KOMMUN %in% ourkom &
+                     fob$LINNEID %in% per$LINNEID, ]
+    ```
+    
+    They need to be of the right age (born on November 1, 1940 or earlier). We
+    get the birthdates from *inddata*.
+    
+    ```{r mat60}
+    library(eha)
+    indx <- match(fob60$LINNEID, per$LINNEID)
+    fob60$birthdate <- per$birthdate[indx]
+    fob60$fobdate <- as.numeric(toTime("1960-11-01"))
+    ##lim <- as.numeric(toTime("1940-11-01"))
+    ##fob60 <- fob60[fob60$birthdate < lim, ]
+    indx <- match(fob60$LINNEID, per$LINNEID) # NOTE: Necessary!
+    fob60$enter <- fob60$fobdate - fob60$birthdate
+    dd <- per$deathdate[indx]
+    ##fob60$event <- !(is.na(dd)) & (dd <= lim + 25) # NOTE: 40 --> 25!
+    fob60$event <- !(is.na(dd)) & (dd <= fob60$fobdate +5)
+    fob60$exit <- fob60$enter + 5
+    fob60$exit[fob60$event] <- dd[fob60$event] - fob60$birthdate[fob60$event]
+    head(fob60)
+}

### 2018-03-03 CreData/fromFobLisa.Rmd
continuing

@@
 date: "March 1, 2018"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
 fontsize: 11pt
@@
 ```{r days}
+load("../../Data/fob.rda")
 ar <- seq(1960, 1990, by = 5)
@@
 
-Note that this is per year.
+Get `birthdate` from `per`.
 
-## FoB 1960
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob60}
-load("../../Data/fob.rda")
-fob60 <- fob[fob$AR == 1960 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
+```{r getbirthdateanddathdate}
+indx <- match(fob$LINNEID, per$LINNEID)
+fob$birthdate <- per$birthdate[indx]
+fob$deathdate <- per$deathdate[indx]
 ```
 
-They need to be of the right age (born on November 1, 1940 or earlier). We
-get the birthdates from *inddata*.
+Create individual Fobs (1960, 1965, etc):
 
-```{r mat60}
-library(eha)
-indx <- match(fob60$LINNEID, per$LINNEID)
-fob60$birthdate <- per$birthdate[indx]
-fob60$fobdate <- as.numeric(toTime("1960-11-01"))
-##lim <- as.numeric(toTime("1940-11-01"))
-##fob60 <- fob60[fob60$birthdate < lim, ]
-indx <- match(fob60$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob60$enter <- fob60$fobdate - fob60$birthdate
-dd <- per$deathdate[indx]
-##fob60$event <- !(is.na(dd)) & (dd <= lim + 25) # NOTE: 40 --> 25!
-fob60$event <- !(is.na(dd)) & (dd <= fob60$fobdate +5)
-fob60$exit <- fob60$enter + 5
-fob60$exit[fob60$event] <- dd[fob60$event] - fob60$birthdate[fob60$event]
-head(fob60)
-```
-
-## FoB 1965
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob65}
-fob65 <- fob[fob$AR == 1965 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
-```
-
-They need to be of the right age (born on November 1, 1930 or earlier). We
-get the birthdates from *inddata*.
-
-```{r mat65}
-library(eha)
-indx <- match(fob65$LINNEID, per$LINNEID)
-fob65$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1945-11-01"))
-fob65 <- fob65[fob65$birthdate < lim, ]
-indx <- match(fob65$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob65$enter <- lim + 20 - fob65$birthdate
-dd <- per$deathdate[indx]
-fob65$event <- !(is.na(dd)) & (dd <= lim + 25) ## NOTE: 40 --> 25!
-fob65$exit <- fob65$enter + 5
-fob65$exit[fob65$event] <- dd[fob65$event] - fob65$birthdate[fob65$event]
-fob65 <- fob65[fob65$exit > fob65$enter, ] # Shouldn't be necessary!
-head(fob65)
-```
-
-It should be noted that three persons present in the 1965 FoB actually died 1964 or earlier.
-
-## FoB 1970
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob70}
-fob70 <- fob[fob$AR == 1970 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
-```
-
-They need to be of the right age (born on November 1, 1935 or earlier). We
-get the birthdates from *inddata*.
-
-```{r mat70}
-library(eha)
-indx <- match(fob70$LINNEID, per$LINNEID)
-fob70$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1950-11-01"))
-fob70 <- fob70[fob70$birthdate <= lim, ]
-indx <- match(fob70$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob70$enter <- lim + 20 - fob70$birthdate
-dd <- per$deathdate[indx]
-fob70$event <- !(is.na(dd)) & (dd <= lim + 25)
-fob70$exit <- fob70$enter + 5
-fob70$exit[fob70$event] <- dd[fob70$event] - fob70$birthdate[fob70$event]
-fob70 <- fob70[fob70$exit > fob70$enter, ] # Shouldn't be necessary!
-head(fob70)
-cat("lim = ", lim, "\n")
-```
-
-It should be noted that one person present in the 1970 FoB actually died 1964.
-
-## FoB 1975
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob75}
-fob75 <- fob[fob$AR == 1975 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
-```
-
-They need to be of the right age (born on November 1, 1940 or earlier). We
-get the birthdates from *inddata*.
-
-```{r mat75}
-library(eha)
-indx <- match(fob75$LINNEID, per$LINNEID)
-fob75$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1955-11-01"))
-fob75 <- fob75[fob75$birthdate < lim, ]
-indx <- match(fob75$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob75$enter <- lim + 20 - fob75$birthdate
-dd <- per$deathdate[indx]
-fob75$event <- !(is.na(dd)) & (dd <= lim + 25)
-fob75$exit <- fob75$enter + 5
-fob75$exit[fob75$event] <- dd[fob75$event] - fob75$birthdate[fob75$event]
-fob75 <- fob75[fob75$exit > fob75$enter, ] # Shouldn't be necessary!
-head(fob75)
-##cat("lim = ", lim, "\n")
-```
-
-It should be noted that one person present in the 1975 FoB actually died 1964 (same person as earlier).
-
-## FoB 1980
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob80}
-fob80 <- fob[fob$AR == 1980 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
-```
-
-They need to be of the right age (born on November 1, 1945 or earlier). We
-get the birthdates from *inddata*.
-
-```{r mat80}
-library(eha)
-indx <- match(fob80$LINNEID, per$LINNEID)
-fob80$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1960-11-01"))
-fob80 <- fob80[fob80$birthdate < lim, ]
-indx <- match(fob80$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob80$enter <- lim + 20 - fob80$birthdate
-dd <- per$deathdate[indx]
-fob80$event <- !(is.na(dd)) & (dd <= lim + 25)
-fob80$exit <- fob80$enter + 5
-fob80$exit[fob80$event] <- dd[fob80$event] - fob80$birthdate[fob80$event]
-fob80 <- fob80[fob80$exit > fob80$enter, ] # Shouldn't be necessary!
-head(fob80)
-##cat("lim = ", lim, "\n")
-```
-
-It should be noted that one person present in the 1980 FoB actually died 1964 (same person as earlier).
-
-## FoB 1985
-
-This is no longer the last FoB we are using: The follow-up time this time
-*is* five years (as before), but five years, from November 1, 1985 to October 31, 1990, when the *lisa* data fills in.
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob85}
-fob85 <- fob[fob$AR == 1985 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
-```
-
-They need to be of the right age (born on November 1, 1950 or earlier). We
-get the birthdates from *inddata*.
-
-```{r mat85}
-library(eha)
-indx <- match(fob85$LINNEID, per$LINNEID)
-fob85$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1965-11-01"))
-fob85 <- fob85[fob85$birthdate < lim, ]
-indx <- match(fob85$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob85$enter <- lim + 20 - fob85$birthdate
-##fob85$enter <- as.numeric(toTime("1985-11-01")) - fob85$birthdate
-dd <- per$deathdate[indx]
-fob85$event <- !(is.na(dd)) & (dd <= lim + 25)
-##fob85$event <- !(is.na(dd)) & (dd < 1987) 
-fob85$exit <- fob85$enter + 5  ## Not any longer: 1 year and 2 months.
-fob85$exit[fob85$event] <- dd[fob85$event] - fob85$birthdate[fob85$event]
-fob85 <- fob85[fob85$exit > fob85$enter, ] # Shouldn't be necessary!
-head(fob85)
-##cat("lim = ", lim, "\n")
-```
-
-It should be noted that one person present in the 1985 FoB actually died 1964 (same person as earlier).
-
-## FoB 1990
-
-This is no longer the last FoB we are using: The follow-up time this time
-*is* ten years (as before), from November 1, 1990 to October 31, 2000, when
-the *lisa* data has been filling in for a long time.
-
-Persons in the fob and in our KOMMUNer.
-
-```{r fob90}
-fob90 <- fob[fob$AR == 1990 & 
-                 fob$KOMMUN %in% ourkom &
-                 fob$LINNEID %in% per$LINNEID, ]
-```
-
-They need to be of the right age (born on November 1, 1960 or earlier). We
-get the birthdates from *inddata*.
-
-```{r mat90}
-library(eha)
-indx <- match(fob90$LINNEID, per$LINNEID)
-fob90$birthdate <- per$birthdate[indx]
-lim <- as.numeric(toTime("1960-11-01"))
-fob90 <- fob90[fob90$birthdate < lim, ]
-indx <- match(fob90$LINNEID, per$LINNEID) # NOTE: Necessary!
-fob90$enter <- lim + 20 - fob90$birthdate
-##fob90$enter <- as.numeric(toTime("1985-11-01")) - fob85$birthdate
-dd <- per$deathdate[indx]
-fob90$event <- !(is.na(dd)) & (dd <= lim + 20)
-##fob90$event <- !(is.na(dd)) & (dd < 1987) 
-fob90$exit <- fob90$enter + 10  ## Not any longer: 1 year and 2 months.
-fob90$exit[fob90$event] <- dd[fob90$event] - fob90$birthdate[fob90$event]
-fob90 <- fob90[fob90$exit > fob90$enter, ] # Shouldn't be necessary!
-head(fob90)
-##cat("lim = ", lim, "\n")
-```
-
-It should be noted that one person present in the 1985 FoB actually died 1964 (same person as earlier).
-
-This ends the data collection from the FoBs. Remains to put all together.
-
-## Putting the data from all FoBs in one data frame
-
-We put the event histories described by the FoBs (with FoB90!) into one data frame.
-
-```{r fobgrab}
-fobdata <- rbind(fob60, fob65, fob70, fob75, fob80, fob85, fob90)
+```{r rawfob}
+source("R/extractFob.R")
+n <- length(ar)
+fobbar <- vector(mode = "list", length = n)
+names(fobbar) <- paste("fob", (ar %% 100), sep = "")
+for (i in 1:(n-1)){
+    fobbar[[i]] <- extractFob(ar[i], followUp = 5, data = fob)
+}
+fobbar[[n]] <- extractFob(ar[n], followUp = 11 + 61 / 365, data = fob)
+fobdata <- do.call(rbind, fobbar)
 fobdata <- fobdata[order(fobdata$LINNEID, fobdata$enter), ]
@@
 rownames(fobdata) <- 1:NROW(fobdata)
-save(fobdata, file = "fobdata.rda")
+save(fobdata, file = "fobdata2.rda")
 ```

### 2018-03-03 CreData/R/extractFob.R

@@ -1,26 +1,15 @@
 extractFob <- function(year, followUp, data){
-    ## data = fob
-    out <- fob[fob$AR == year & 
-                     fob$KOMMUN %in% ourkom &
-                     fob$LINNEID %in% per$LINNEID, ]
-    ```
+    ## data = fob (or lisa?)
     
-    They need to be of the right age (born on November 1, 1940 or earlier). We
-    get the birthdates from *inddata*.
+    out <- data[data$AR == year, ]
     
-    ```{r mat60}
     library(eha)
-    indx <- match(fob60$LINNEID, per$LINNEID)
-    fob60$birthdate <- per$birthdate[indx]
-    fob60$fobdate <- as.numeric(toTime("1960-11-01"))
-    ##lim <- as.numeric(toTime("1940-11-01"))
-    ##fob60 <- fob60[fob60$birthdate < lim, ]
-    indx <- match(fob60$LINNEID, per$LINNEID) # NOTE: Necessary!
-    fob60$enter <- fob60$fobdate - fob60$birthdate
-    dd <- per$deathdate[indx]
-    ##fob60$event <- !(is.na(dd)) & (dd <= lim + 25) # NOTE: 40 --> 25!
-    fob60$event <- !(is.na(dd)) & (dd <= fob60$fobdate +5)
-    fob60$exit <- fob60$enter + 5
-    fob60$exit[fob60$event] <- dd[fob60$event] - fob60$birthdate[fob60$event]
-    head(fob60)
+    
+    out$fobdate <- as.numeric(toTime(paste(year, "11", "01", sep = "-")))
+    
+    out$enter <- out$fobdate - out$birthdate
+    out$event <- !(is.na(out$deathdate)) & (out$deathdate <= out$fobdate + followUp)
+    out$exit <- out$enter + followUp
+    out$exit[out$event] <- out$deathdate[out$event] - out$birthdate[out$event]
+    out
 }

### 2018-03-03 CreData/R/extractLisa.R
first commit

@@ -1,0 +1,15 @@
+extractLisa <- function(year, followUp, data){
+    ## data = lisa
+    
+    out <- data[data$AR == year, ]
+    
+    library(eha)
+    
+    out$fobdate <- as.numeric(toTime(paste(year + 1, "01", "01", sep = "-")))
+    
+    out$enter <- out$fobdate - out$birthdate
+    out$event <- !(is.na(out$deathdate)) & (out$deathdate <= out$fobdate + followUp)
+    out$exit <- out$enter + followUp
+    out$exit[out$event] <- out$deathdate[out$event] - out$birthdate[out$event]
+    out
+}

### 2018-03-04 CreData/fromFobLisa.Rmd
almost first commit

@@
 output:
+  pdf_document: default
   html_document:
@@
 options(stringsAsFactors = FALSE)
+library(stringr)
+library(tidyr)
+library(dplyr)
 ```
@@
 inddata$KON <- sex$KON[indx]
-per <- inddata[!is.na(inddata$KON), ]
+per <- inddata[!is.na(inddata$KON) & !is.na(inddata$FODELSEAR), ]
 per$birthdate <- per$FODELSEAR + 0.125  +  (per$FODELSEKVARTAL - 1) / 4
@@
 load("../../Data/fob.rda")
+fob <- fob[fob$LINNEID %in% per$LINNEID, ]
 ar <- seq(1960, 1990, by = 5)
@@
 
+# Fob gets hisclass and the rest
+
+WE continue as in "Mort2/CreateData2/fromFobLisa.Rmd":
+
+```{r syskonfob}
+    load("../../Data/socstyr.rda")
+    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
+    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
+    source("R/putONssyk.R")
+    ##putONssyk
+    fobb <- putONssyk(fobdata)
+    source("R/ssykTOhisclass.R")
+    ##ssykTOhisclass
+    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
+    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
+                                       (!is.na(hisclass) & hisclass == -1))
+    source("R/addPer.R")
+    ##fobb <- pushHis(fobb)
+    ##fobb$hisclass <- fobb$nyhisclass
+    ## Instead
+    ##fobb <- group_by(fobb, LINNEID) %>% fill_("hisclass") # Later!!summary(lisa)
+    fobb$id <- fobb$LINNEID
+    ##fobb <- addPer(fobb, cuts = c(1961, 1976, 1986, 1996, 2002)) # Wait!
+    load("../../Data/sex.rda")
+
+    indx <- match(fobb$LINNEID, sex$LINNEID)
+    fobb$KON <- sex$KON[indx]
+    fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005,
+                                         248011, 248201, 248204)
+    civst <- numeric(NROW(fobb))
+    civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
+    civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
+    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, änka
+    fobb$civst <- factor(civst,
+                         labels = c("unmarried", "married", "prev.married"))
+    fobb$sex <- fobb$KON
+    levels(fobb$sex) <- c("male", "female")
+    fobb <- fobb[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
+                     "urban", "civst", "hisclass", "ULORSAK")]
+```
+
 # The Lisa data
@@
 
-Those with *missing birthdate* in *lisa* are those born 1975 or later. They can be removed.
+Those with *missing birthdate* in *lisa* are those born 1975 or later (What??). They can be removed.
 
@@
 rownames(lisadata) <- 1:NROW(lisadata)
-save(lisadata, file = "lisadata.rda")
+save(lisadata, file = "lisadata2.rda")
+```
+
+## Lisa gets hisclass and the rest
+
+Note, this is just cut from "Mort2/CreateData2/fromFobLisa.Rmd". May need
+some editing ...
+
+```{r gethislisa}
+    load("../../Data/socstyr.rda")
+    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
+    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
+    source("R/putONssyk.R")
+    ##putONssyk
+    fobb <- putONssyk(fobdata)
+    source("R/ssykTOhisclass.R")
+    ##ssykTOhisclass
+    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
+    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
+                                       (!is.na(hisclass) & hisclass == -1))
+    source("R/addPer.R")
+    ##fobb <- pushHis(fobb)
+    ##fobb$hisclass <- fobb$nyhisclass
+    ## Instead
+    ##fobb <- group_by(fobb, LINNEID) %>% fill_("hisclass") # Later!
+    fobb$id <- fobb$LINNEID
+    fobb <- addPer(fobb, cuts = c(1961, 1971, 1981, 1991, 2002))
+    load("../../Data/sex.rda")
+
+    indx <- match(fobb$LINNEID, sex$LINNEID)
+    fobb$KON <- sex$KON[indx]
+    fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005,
+                                         248011, 248201, 248204)
+    civst <- numeric(NROW(fobb))
+    civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
+    civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
+    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, änka
+    fobb$civst <- factor(civst,
+                         labels = c("unmarried", "married", "prev.married"))
+    fobb$sex <- fobb$KON
+    levels(fobb$sex) <- c("male", "female")
+    fobb <- fobb[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
+                     "period", "urban", "civst", "hisclass", "ULORSAK")]
+   fobb$dataset <- as.character(fobb$AR)
+```
+
+## Lisa
+
+```{r lisadata}
+    ##load("../../../Data/lisadata.rda")
+    ##load("data4/lisadata.rda")
+    ## Cause of death:
+    load("../../Data/socstyr.rda")
+    indx <- match(lisadata$LINNEID, socstyr$LINNEID)
+    lisadata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
+    civst <- lisadata$CIV
+    civst[is.na(civst)] <- 0
+    civst[civst == 4] <- 3
+    civst[civst == 5] <- 2
+    civst[civst == 6] <- 3
+    is.na(civst) <- civst == 0
+    lisadata$civst <- civst
+    rm(civst)
+    ## Fix KON:
+    load("../../Data/sex.rda")
+    indx <- match(lisadata$LINNEID, sex$LINNEID)
+    lisadata$sex <- sex$KON[indx]
+    levels(lisadata$sex) <- c("male", "female")
+    lisaold <- lisadata
+    lisadata <- cal.window(lisadata, c(2002, 2014))
+
+    ##lisadata <- lisadata[lisadata$AR > 1999.5 & !is.na(lisadata$SSYK3), ]
+
+    ## This is new: Do not remove:
+    ##lisadata <- lisadata[!is.na(lisadata$SSYK3), ] ##NOTE!
+
+    ## Fix HISCLASS:
+    source("R/ssyk3TOhisclass.R")
+    ##ssyk3TOhisclass
+    lisadata$hisclass <- ssyk3TOhisclass(lisadata$SSYK3)
+    ## Push forward:
+    source("R/pushHis.R")
+    ## liss <- pushHis(lisadata) # Later!
+    liss <- lisadata
+    ##liss$hisclass <- liss$nyhisclass
+    ##liss$nyhisclass <- NULL
+    ## Now:
+    ##liss <- group_by(lisadata, LINNEID) %>% fill_("hisclass") # Later!!
+    liss$id <- liss$LINNEID
+    lisaold$id <- lisaold$LINNEID
+
+    ## lisaold <- addPer(lisaold, name = "period", cuts = seq(1986, 2014, by = 2)) ## Later! 
+    ## liss <- addPer(liss, name = "period", cuts = seq(2002, 2014, by = 6)) # Later!!
+    ##liss <- liss[!is.na(liss$hisclass), ]
+    ##liss$hisclass <- cut(liss$hisclass, c(0, 2.5, 4.5, 5.5, 8),
+      ##                   labels = 1:4)
+    ##liss$hisclass <- as.numeric(as.character(liss$hisclass))
+    ##liss <- liss[with(liss, !is.na(civst) &
+      ##                      !is.na(hisclass)), ]
+    ##lisaold <- lisaold[with(lisaold, !is.na(civst)), ]
+    liss$civst <- factor(liss$civst,
+                         labels = c("unmarried", "married", "prev.married"))
+    lisaold$civst <- factor(lisaold$civst,
+                         labels = c("unmarried", "married", "prev.married"))
+    liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004,
+                                         284005, 248011, 248201, 248204)
+    lisaold$urban <- lisaold$FORSAMLING %in% c(248001, 248002, 284004,
+                                         284005, 248011, 248201, 248204)
+    liss$id <- liss$LINNEID
+    liss <- liss[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
+                     "urban", "civst", "hisclass", "ULORSAK")]
+    liss$dataset <- paste("lisa", as.character(liss$AR), sep = "")
+    save(liss, file = "liss2.rda")
+    save(lisaold, file = "lisaold.rda")
 ```

### 2018-03-04 CreData/R/addPer.R

@@ -17,7 +17,7 @@ addPer <- function(dat, name = "period", cuts = seq(1901, 1951), by = 10){
     }
     start[[name]] <- factor(start[[name]], labels = levs)
     start <- start[order(start$id, start$enter), ]
-    start <- rc(start)
+    ##start <- rc(start)
     start
 }
 

### 2018-03-04 CreData/R/pushForward.R

@@ -1,0 +1,9 @@
+pushForward <- function(x){
+    n <- length(x)
+    if (n > 1){
+        for (i in 2:n){
+            if (is.na(x[i])) x[i] <- x[i-1]
+        }
+    }
+    x
+}

### 2018-03-04 CreData/R/pushHis.R

@@ -1,0 +1,15 @@
+pushHis <- function(dat){
+    ## 'dat' is a data frame with variables 'LINNEID' and 'hisclass'
+    library(skum)
+    om <- unique(dat$LINNEID)
+    
+    hyhis <- numeric(NROW(dat))
+    dat$nyhisclass <- numeric(NROW(dat))
+    for (i in om){
+        who <- dat$LINNEID == i
+        ok <- dat$hisclass[who]
+        ok <- pushForward(ok)
+        dat$nyhisclass[who] <- ok
+    }
+    dat
+}

### 2018-03-04 CreData/R/putONssyk.R

@@ -1,0 +1,73 @@
+putONssyk <- function(dat){
+    ## 'dat' is the processed file 'fobdata'
+    ## This must be run from "Mortality/CreDat"
+    
+    load("../../Data/yr_ssyk.rda")
+    dat$ssyk <- NA
+
+    ##for (year in seq(60, 75, by = 5)){
+      ##  nam <- paste("yr", year, sep = "")
+    ##    load(paste("data/", nam, ".rda" sep = ""))
+##        nyr <- dat[dat$AR == i, ]
+  ##      indx <- match(nyr$YRKE)
+    ##}
+    ## 1960:
+    load("../../Data/yr60.rda")
+    who <- dat$AR == 1960 & dat$YRKE %in% yr60$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr60$old)
+    nyr <- yr60$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1965:
+    load("../../Data/yr65.rda")
+    who <- dat$AR == 1965 & dat$YRKE %in% yr65$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr65$old)
+    nyr <- yr65$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1970:
+    load("../../Data/yr70.rda")
+    who <- dat$AR == 1970 & dat$YRKE %in% yr70$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr70$old)
+    nyr <- yr70$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1975:
+    load("../../Data/yr75.rda")
+    who <- dat$AR == 1975 & dat$YRKE %in% yr75$old
+    nyr <- dat$YRKE[who]
+    indx <- match(nyr, yr75$old)
+    nyr <- yr75$new[indx]
+    dat$YRKE[who] <- nyr
+    
+    ## 1980: Vad ska vi ta här?? 'yr80.rda' finns inte!
+    ## och YRKE80 gäller ej 1980!!
+    ## Chansar på att 'yr75.rda' gäller:
+
+    ## 1980 (try):
+    load("../../Data/yr75.rda")
+    who <- dat$AR == 1980 & dat$YRKE %in% yr75$old
+    nyr <- dat$YRKE[who]
[…]

### 2018-03-04 CreData/R/ssyk3TOhisclass.R

@@ -1,0 +1,34 @@
+ssyk3TOhisclass <- function(x){
+
+    ## 2017-03.28: Three instead of four hisclasses.
+    ## 2018-02-03: Four instead of three hisclasses. 8 == farmer
+  
+    ## 'x' is a vector of ssyk codes.
+    ## as output we want a vector of hisclass codes (7 levels)
+    
+    ## Step 1: Convert shh$ssyk to "three-digit" (except '110'):
+    load("../../Data/shh.rda")
+    shh$ssyk <- shh$ssyk %/% 10 # removes last digit.
+    ## Remove duplicate ssyk-rows in shh:
+    shh <- shh[!duplicated(shh$ssyk), ]
+    
+    ## Step 2: SSYK3 to 12-levels hisclass
+    indx <- match(x, shh$ssyk)
+    tmp <- shh$hisclass[indx]
+
+    ## Step 2: go from 12 to 4 levels: ## NEW!!
+    ## Step 2: go from 12 to 3 levels: ## NEWER!!
+    
+    res <- rep(-1, length(x))
+    
+    tmp[is.na(tmp)] <- -1
+    
+    res[tmp %in% 1:2] <- 1
+    res[tmp %in% c(3:6)] <- 2
+    ##res[tmp %in% 4:6] <- 3
+    res[tmp %in% c(7, 9:12)] <- 3 # '4' ---> '3'
+    res[tmp == 8] <- 4
+    is.na(res) <- res == -1
+    
+    res
+}

### 2018-03-04 CreData/R/ssykTOhisclass.R

@@ -1,0 +1,30 @@
+ssykTOhisclass <- function(x){
+
+    ## Change 2017-03-28: Three instead of four hisclasses!
+    ## Change 2018-02-03: Four instead of three hisclasses! 8 == farmers  
+    
+    ## 'x' is a vector of ssyk codes.
+    ## as output we want a vector of hisclass codes (7 levels)
+    
+    ## Step 1: ssyk to 12-levels hisclass
+    load("../../Data/shh.rda")
+    indx <- match(x, shh$ssyk)
+    tmp <- shh$hisclass[indx]
+
+    ## Step 2: go from 12 to 4 levels:  ## NEW !!
+    ## Step 2: go from 12 to 3 levels:  ## NEWER !!
+    
+    
+    res <- rep(-1, length(x))
+    
+    tmp[is.na(tmp)] <- -1
+    
+    res[tmp %in% 1:2] <- 1
+    res[tmp %in% c(3:6)] <- 2
+    ##res[tmp %in% 4:6] <- 3
+    res[tmp %in% c(7, 9, 10, 11, 12)] <- 3 # '4' ---> '3'
+    res[tmp == 8] <- 4
+    is.na(res) <- res == -1
+    
+    res
+}

### 2018-03-04 CreData/fromFobLisa.Rmd
last minute

@@
 load("../../Data/lisa.rda")
-lisa <- lisa[lisa$KOMMUN %in% ourkom, ]
+lisa <- lisa[(lisa$KOMMUN %in% ourkom) & (lisa$LINNEID %in% per$LINNEID), ]
 indx <- match(lisa$LINNEID, per$LINNEID)
+
 lisa$birthdate <- per$birthdate[indx]
-```
 
-Those with *missing birthdate* in *lisa* are those born 1975 or later (What??). They can be removed.
-
-```{r removeyoung}
-lisa <- lisa[!is.na(lisa$birthdate), ]
 ```
@@
 rownames(lisadata) <- 1:NROW(lisadata)
-save(lisadata, file = "lisadata2.rda")
+##save(lisadata, file = "lisadata2.rda")
 ```
@@
 ## Lisa gets hisclass and the rest
-
-Note, this is just cut from "Mort2/CreateData2/fromFobLisa.Rmd". May need
-some editing ...
-
-```{r gethislisa}
-    load("../../Data/socstyr.rda")
-    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
-    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
-    source("R/putONssyk.R")
-    ##putONssyk
-    fobb <- putONssyk(fobdata)
-    source("R/ssykTOhisclass.R")
-    ##ssykTOhisclass
-    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
-    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
-                                       (!is.na(hisclass) & hisclass == -1))
-    source("R/addPer.R")
-    ##fobb <- pushHis(fobb)
-    ##fobb$hisclass <- fobb$nyhisclass
-    ## Instead
-    ##fobb <- group_by(fobb, LINNEID) %>% fill_("hisclass") # Later!
-    fobb$id <- fobb$LINNEID
-    fobb <- addPer(fobb, cuts = c(1961, 1971, 1981, 1991, 2002))
-    load("../../Data/sex.rda")
-
-    indx <- match(fobb$LINNEID, sex$LINNEID)
-    fobb$KON <- sex$KON[indx]
-    fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005,
-                                         248011, 248201, 248204)
-    civst <- numeric(NROW(fobb))
-    civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
-    civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
-    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, änka
-    fobb$civst <- factor(civst,
-                         labels = c("unmarried", "married", "prev.married"))
-    fobb$sex <- fobb$KON
-    levels(fobb$sex) <- c("male", "female")
-    fobb <- fobb[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
-                     "period", "urban", "civst", "hisclass", "ULORSAK")]
-   fobb$dataset <- as.character(fobb$AR)
-```
-
-## Lisa
 
@@
     liss$id <- liss$LINNEID
-    lisaold$id <- lisaold$LINNEID
+    ##lisaold$id <- lisaold$LINNEID
 
@@
                          labels = c("unmarried", "married", "prev.married"))
-    lisaold$civst <- factor(lisaold$civst,
-                         labels = c("unmarried", "married", "prev.married"))
+    ##lisaold$civst <- factor(lisaold$civst,
+      ##                   labels = c("unmarried", "married", "prev.married"))
     liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004,
                                          284005, 248011, 248201, 248204)
-    lisaold$urban <- lisaold$FORSAMLING %in% c(248001, 248002, 284004,
-                                         284005, 248011, 248201, 248204)
+    ##lisaold$urban <- lisaold$FORSAMLING %in% c(248001, 248002, 284004,
+      ##                                   284005, 248011, 248201, 248204)
     liss$id <- liss$LINNEID
@@
                      "urban", "civst", "hisclass", "ULORSAK")]
-    liss$dataset <- paste("lisa", as.character(liss$AR), sep = "")
+    ##liss$dataset <- paste("lisa", as.character(liss$AR), sep = "")
     save(liss, file = "liss2.rda")
-    save(lisaold, file = "lisaold.rda")
+    ##save(lisaold, file = "lisaold.rda")
 ```
 
-Now, read *checkData* (maybe not).
+# Combine Fob and Lisa
+
+```{r combfoblisa}
+source("R/fixCovar2.R")
+foli <- rbind(fobb, liss)
+foli <- foli[order(foli$id, foli$enter), ]
+source("R/pushForward.R")
+foli$hisclass <- unlist(tapply(foli$hisclass, foli$id, pushForward))
+foli$hisclass[is.na(foli$hisclass)] <- 5
+foli$hisclass <- factor(foli$hisclass, labels = c("elite", "middle", "worker", "farmer", "none"))
+foli$hisclass40 <- fixCovar2(foli$id, foli$hisclass, foli$enter, foli$exit, age = 40, lookBack = 20)
+foli$hisclass65 <- fixCovar2(foli$id, foli$hisclass, foli$enter, foli$exit, age = 65, lookBack = 45)
+save(foli, file = "foli.rda")
+```

### 2018-03-04 CreData/fromPoplink.Rmd
continue
considerations

@@
 author: "Göran Broström"
-date: "28 February 2018"
+date: "4 March, 2018"
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
 ---
@@
 if (TRUE){
+    adult <- adult[order(adult$id, adult$enter), ]
+    adult <- group_by(adult, id) %>% fill_("hisclass")
+    source("R/pus")
     source("R/fixCovar2.R")  
@@
 }else{
-  adult <- adult[order(adult$id, adult$enter), ]
-  adult <- group_by(adult, id) %>% fill_("hisclass")
   adult$hisclass[is.na(adult$hisclass)] <- 4

### 2018-03-05 CreData/fromFobLisa.Rmd
finishing data creation

--- 2018-03-03 CreData/fromFobLisa.Rmd
@@
 fobdata <- fobdata[order(fobdata$LINNEID, fobdata$enter), ]
-##fobdata <- age.window(fobdata, c(40, 110))
+##
 rownames(fobdata) <- 1:NROW(fobdata)
-save(fobdata, file = "fobdata2.rda")
+```
+
+# Fob gets hisclass and the rest
+
+We continue as in "Mort2/CreateData2/fromFobLisa.Rmd":
+
+```{r syskonfob}
+    load("../../Data/socstyr.rda")
+    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
+    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
+    source("R/putONssyk.R")
+    ##putONssyk
+    fobb <- putONssyk(fobdata)
+    source("R/ssykTOhisclass.R")
+    ##ssykTOhisclass
+    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
+    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
+                                       (!is.na(hisclass) & hisclass == -1))
+    fobb$id <- fobb$LINNEID
+    load("../../Data/sex.rda")
+
+    indx <- match(fobb$LINNEID, sex$LINNEID)
+    fobb$KON <- sex$KON[indx]
+    fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005,
+                                         248011, 248201, 248204)
+    civst <- numeric(NROW(fobb))
+    civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
+    civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
+    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, änka
+    fobb$civst <- factor(civst,
+                         labels = c("unmarried", "married", "prev.married"))
+    fobb$sex <- fobb$KON
+    levels(fobb$sex) <- c("male", "female")
+    fobb <- fobb[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
+                     "urban", "civst", "hisclass", "ULORSAK")]
 ```
@@
 load("../../Data/lisa.rda")
-lisa <- lisa[lisa$KOMMUN %in% ourkom, ]
+lisa <- lisa[(lisa$KOMMUN %in% ourkom) & (lisa$LINNEID %in% per$LINNEID), ]
 indx <- match(lisa$LINNEID, per$LINNEID)
+
 lisa$birthdate <- per$birthdate[indx]
-```
 
-Those with *missing birthdate* in *lisa* are those born 1975 or later. They can be removed.
-
-```{r removeyoung}
-lisa <- lisa[!is.na(lisa$birthdate), ]
 ```
@@
 lisa$deathdate <- per$deathdate[indx]
-lisa$event <- !is.na(lisa$deathdate) & lisa$deathdate > lisa$AR + 1 & lisa$deathdate <= lisa$AR + 2
+lisa$event <- !is.na(lisa$deathdate) &
+    lisa$deathdate > lisa$AR + 1 & 
+    lisa$deathdate <= lisa$AR + 2
 lisa$exit <- lisa$enter + 1 # First tip.
@@
 ```{r cleanandsort}
-##lisadata <- age.window(lisa, c(40, 110))
 lisadata <- lisa[order(lisa$LINNEID, lisa$enter), ]
@@
 rownames(lisadata) <- 1:NROW(lisadata)
-save(lisadata, file = "lisadata.rda")
 ```
 
-Now, read *checkData* (maybe not).
+## Lisa gets hisclass and the rest
+
+```{r lisadata}
+    ## Cause of death:
+    load("../../Data/socstyr.rda")
+    indx <- match(lisadata$LINNEID, socstyr$LINNEID)
+    lisadata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
+    ## Civil status:
+    civst <- lisadata$CIV
+    civst[is.na(civst)] <- 0
+    civst[civst == 4] <- 3
+    civst[civst == 5] <- 2
+    civst[civst == 6] <- 3
+    is.na(civst) <- civst == 0
+    lisadata$civst <- civst
+    rm(civst)
+    ## Fix KON:
+    load("../../Data/sex.rda")
+    indx <- match(lisadata$LINNEID, sex$LINNEID)
+    lisadata$sex <- sex$KON[indx]
+    levels(lisadata$sex) <- c("male", "female")
+    lisaold <- lisadata
+    lisadata <- cal.window(lisadata, c(2002, 2014))
+
+    ## Fix HISCLASS:
+    source("R/ssyk3TOhisclass.R")
+    ##ssyk3TOhisclass
+    lisadata$hisclass <- ssyk3TOhisclass(lisadata$SSYK3)
+    liss <- lisadata
+
+    liss$id <- liss$LINNEID
+    liss$civst <- factor(liss$civst,
+                         labels = c("unmarried", "married", "prev.married"))
+
+    liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004,
+                                         284005, 248011, 248201, 248204)
+
+    liss$id <- liss$LINNEID
+    liss <- liss[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
+                     "urban", "civst", "hisclass", "ULORSAK")]
+
+```
+
+# Combine Fob and Lisa
+
+```{r combfoblisa}
+
+foli <- rbind(fobb, liss)
+foli <- foli[order(foli$id, foli$enter), ]
+source("R/pushForward.R")
+foli$hisclass <- unlist(tapply(foli$hisclass, foli$id, pushForward))
+foli$hisclass[is.na(foli$hisclass)] <- 5
+foli$hisclass <- factor(foli$hisclass, 
+                        labels = c("elite", "middle", "worker", "farmer", "none"))
+zz <- levels(foli$civst)
+foli$civst <- unlist(tapply(as.integer(foli$civst), foli$id, pushForward))
+foli$civst <- factor(foli$civst, labels = zz)
+
+foli <- age.window(foli, c(30, 100))
+save(foli, file = "../data/foli.rda")
+```

### 2018-03-05 CreData/fromPoplink.Rmd
finishing data creation

--- 2018-02-28 CreData/fromPoplink.Rmd
@@
 author: "Göran Broström"
-date: "28 February 2018"
+date: "5 March, 2018"
 output:
@@
 In this version we start at January 1, 1801, and we **do not** split the *poplink* data 
-into two data sets (no causes of death, yet). End at December 31, 1950
+into two data sets (no causes of death, yet). Instead, `skume` refers to Beginning at January 1, 1801 and End at December 31, 1950.
 
@@
 library(skum)
-##load("../../Data/adult.rda")
-##adult <- cal.window(adult, c(1801, 1951))
+
 adult <- obs
@@
 ```{r hisclass}
-##New Hisclasses:
+## Hisclasses:
 ett <- with(adult, !is.na(socBranch) &
               (hisclass == 1))
-               ## ((socBranch == "official" & socStatus == "high") |
-                 ##    (socBranch == "business" & socStatus == "high")
-                ##))
 
@@
               (hisclass %in% c(2, 4)))
-                ##((socBranch == "farming" & socStatus == "high") |
-                ## (socBranch == "business" & socStatus == "low") |
-                ## (socBranch == "official" & socStatus == "low")
-                ##))
 
@@
               (hisclass %in% 5:7)) ## Original 'fyra' ---> 'tre'
-              ##  ((socBranch == "farming" & socStatus == "low") |
-                ##     (socBranch == "worker"))
-                ##)
+
 fyra <- with(adult, !is.na(socBranch) &
@@
 ```{r propagate}
-##if (file.exists("adult.rda")){
-  ##load("adult.rda")
-if (TRUE){
-source("R/fixCovar2.R")  
-    adult$hisclass40 <- with(adult, fixCovar2(id, hisclass, enter, exit, 40, 20))
-    adult$hisclass65 <- with(adult, fixCovar2(id, hisclass, enter, exit, 65, 45))
-    adult$civst40 <- with(adult, fixCovar2(id, civst, enter, exit, 40, 20))
-    adult$civst65 <- with(adult, fixCovar2(id, civst, enter, exit, 65, 45))
-    adult$urban40 <- with(adult, fixCovar2(id, civst, enter, exit, 40, 20))
-    adult$urban65 <- with(adult, fixCovar2(id, civst, enter, exit, 65, 45))
-}else{
+    source("R/pushForward.R")
     adult <- adult[order(adult$id, adult$enter), ]
-  adult <- group_by(adult, id) %>% fill_("hisclass")
-  adult$hisclass[is.na(adult$hisclass)] <- 4
-  adult$hisclass <- factor(adult$hisclass, labels = c("elite", "middle", "worker", "NA"))
-  ##
-  adult <- group_by(adult, id) %>%  fill_("civst")
-  adult$civst[is.na(adult$civst)] <- "unmarried"
-  save(adult, file = "adult.rda")
-}
+    ##adult <- group_by(adult, id) %>% fill_("hisclass")
+    adult$hisclass <- unlist(tapply(adult$hisclass, adult$id, pushForward))
+    adult$hisclass[is.na(adult$hisclass)] <- 5
+    adult$hisclass <- factor(adult$hisclass, 
+                             labels = c("elite", "middle", "worker", "farmer", "none"))
 ```
@@
 ```{r lexis}
-##adult <- cal.window(adult, c(1801, 1951))
-adult <- age.window(adult, c(40, 100))
-```
-
-## Split and period
-
-```{r splitper}
-source("R/addPer.R")
-##skel <- cal.window(adult, c(1801, 1901))
-##skel <- addPer(skel, cuts = seq(1801, 1901, by = 25))
-skume <- cal.window(adult, c(1801, 1951))
-skume <- addPer(skume, cuts = seq(1801, 1951, by = 25))
+skume <- age.window(adult, c(30, 100))
+skume <- cal.window(skume, c(1801, 1951))
 ```
@@
 ```{r skumefilt}
-skume <- skume[, c("id", "region", "birthdate", "sex", "enter", "exit", "event",
-                    "period", "urban", "urban40", "urban65", "civst", "civst40", "civst65", 
-                    "hisclass", "hisclass40", "hisclass65")]
-skume <- skume[order(skume$id, skume$exit), ]
-skume <- rc(skume)
-indx <- tapply(skume$id, skume$id)
-Exit <- with(skume, tapply(exit, id, max))
-Enter <- with(skume, tapply(enter, id, min))
-Event <- with(skume, tapply(event, id, max))
-skume$Enter <- Enter[indx]
-skume$Exit <- Exit[indx]
-skume$Event <- Event[indx]
-skume <- skume[!duplicated(skume$id), ]
-skume$enter <- skume$Enter
-skume$exit <- skume$Exit
-skume$event <- skume$Event
-skume$id <- paste("pop", skume$id, sep = "_")
-skume$Exit <- NULL
-skume$Enter <- NULL
-skume$Event <- NULL
+skume <- skume[, c("id", "birthdate", "sex", "enter", "exit", "event",
+                    "urban", "civst",  "hisclass")]
+skume <- skume[order(skume$id, skume$enter), ]
+summary(skume)
 ```
@@
 ```{r save}
-##save(skel, file = "skel.rda")
-save(skume, file = "skume.rda")
+save(skume, file = "../data/skume.rda")
 ```

### 2018-03-05 CreData/R/addPer.R

@@ -1,4 +1,4 @@
-addPer <- function(dat, name = "period", cuts = seq(1901, 1951), by = 10){
+addPer <- function(dat, name = "period", cuts = seq(1801, 1951, by = 25)){
     n <- length(cuts)
     start <- cal.window(dat, c(cuts[1], cuts[2]))
     start[[name]] <- 1

### 2018-03-05 CreData/R/pushForward.R

@@ -1,8 +1,14 @@
-pushForward <- function(x){
+pushForward <- function(x, fill = TRUE){
     n <- length(x)
     if (n > 1){
         for (i in 2:n){
             if (is.na(x[i])) x[i] <- x[i-1]
+        }
+        if (fill & is.na(x[1])){
+            if (!is.na(x[n])){
+                m <- which(!is.na(x))[1]
+                x[1:(m-1)] <- x[m]
+            }
         }
     }
     x

### 2018-03-05 article3d.Rmd
article on oneSpell data

@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
 library(knitr)
@@
 library(skum)
+source("R/oneSpell.R")
 options(show.signif.stars = FALSE)
@@
 ```{r readalldata}
-source("R/get.file.R")
-skel <- get.file("rawskel")
-skel$region <- "ske"
-skel$urban <- FALSE
-vars <- c("id", "birthdate", "sex", "period", "urban", "civst", "hisclass", "enter", "exit", "event")
-skel <- skel[, vars]
-##
-skume <- get.file("rawskume")
-skume <- skume[, vars]
-##
-vb <- rbind(skel, skume)
-vb <- vb[order(vb$id, vb$exit), ]
-vb$id <- paste("vb", vb$id, sep = "_")
-##
-foli <- get.file("rawfoli")
-foli <- foli[, vars]
+load("data/skume.rda")
+skume <- skume[order(skume$id, skume$enter), ]
+skume$id <- paste("pop", skume$id, sep = "_")
+##
+skume1 <- age.window(skume, c(40, 65))
+skume1 <- oneSpell(skume1)
+##
+skume2 <- age.window(skume, c(65, 100))
+skume2 <- oneSpell(skume2)
+##################################################
+load("data/foli.rda")
+foli <- foli[order(foli$id, foli$enter), ]
+foli$AR <- foli$ULORSAK <- NULL
+##
+foli1 <- age.window(foli, c(40, 65))
+foli1 <- oneSpell(foli1)
+##
+foli2 <- age.window(foli, c(65, 100))
+foli2 <- oneSpell(foli2)
+##################################################
+source("CreData/R/addPer.R")
+skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
+skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
+##
+foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
+foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
+##################################################
+vb1 <- rbind(skume1, foli1)
+vb1$period <- factor(vb1$period)
+##
+vb2 <- rbind(skume2, foli2)
+vb2$period <- factor(vb2$period)
+vb <- rbind(vb1, vb2)
+vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
 ```
@@
      xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 100))
-axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1991, 2014), las = 2)
+axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1986, 1996, 2005, 2014), las = 2)
 axis(2, las = 2, at = c(0, 30, 65, 100))
@@
 #
-pn.exp <- round(sum(skel$exposure) + sum(skume$exposure))
-pn.d <-  as.integer(sum(skel$event) + sum(skume$event))
+pn.exp <- round(sum(skume1$exit - skume1$enter) + sum(skume2$exit - skume2$enter))
+pn.d <-  as.integer(sum(skume1$event) + sum(skume2$event))
 #
@@
 #
-n.exp <- round(sum(foli$exposure))
-n.d <-  as.integer(sum(foli$event))
+n.exp <- round(sum(foli1$exit - foli1$enter) + sum(foli2$exit -foli2$enter))
+n.d <-  as.integer(sum(foli1$event) + sum(foli2$event))
 
@@
 
-```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4}
-xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-xxx <- rbind(xx, xy, xz)
-yy <- round(prop.table(xxx, 1) * 100, 1)
+```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4}
+xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xxx <- rbind(xx, xy, xz)
+yy <- round(prop.table(xx, 1) * 100, 1)
 barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
@@
 ```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
-xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-xxx <- rbind(xx, xy, xz)
-yy <- round(prop.table(xxx, 1) * 100, 1)
+xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xxx <- rbind(xx, xy, xz)
+yy <- round(prop.table(xx, 1) * 100, 1)
 barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
@@
 ```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4}
-xx <- with(skel, tapply(exposure, list(period = period, civst = civst), sum))
-xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
-xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
-xxx <- rbind(xx, xy, xz)
-yy <- round(prop.table(xxx, 1) * 100, 1)
+xx <- with(vb, tapply(exposure, list(period = period, civst = civst), sum))
+##xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
+##xz <- with(foli, tapply(exposure, list(period = period, civst = civst), sum))
+##xxx <- rbind(xx, xy, xz)
+yy <- round(prop.table(xx, 1) * 100, 1)
 barplot(t(yy)[3:1, ], col = c("red", "blue", "green"), las = 2, ylab = "Per cent")
@@
 
-labb <- levels(vb0Tab$period)
+labb <- levels(vb$period)
 nper <- length(labb)
@@
 ```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4}
-orr <- aggregate(x = vb0Tab$exposure, by = vb0Tab[, c("period", "civst", "hisclass")], FUN = sum)
+orr <- aggregate(x = vb$exposure, by = vb[, c("period", "civst", "hisclass")], FUN = sum)
 
@@
 out <- 100 * out
-xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1978, 1982, 1988, 1995, 2005, 2011)
+xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1981, 1991, 2000, 2009)
 plot(xli, out[1, ], type = "b", xlab = "", ylab = "Fraction ever married", 
@@
 ```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4}
-orr <- aggregate(x = vb0Tab$exposure, by = vb0Tab[, c("period", "urban", "hisclass")], FUN = sum)
+orr <- aggregate(x = vb$exposure, by = vb[, c("period", "urban", "hisclass")], FUN = sum)
 
@@
 out <- 100 * out
-xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1978, 1982, 1988, 1995, 2005, 2011)
+xli <-  c(1812, 1837, 1862, 1887, 1912, 1932, 1981, 1991, 2000, 2009)
 plot(xli, out[1, ], type = "b", xlab = "", ylab = "Fraction urban", 
@@
 
-### Inverse exposure weighting
+### Inverse probability weighting
 
@@
 
-As an illustration, take the period `1901-1925` with males. There are six combinations of civil status and urban/rural, and for 
+As an illustration, take the period `2005-2013` with males, 65--99 years of age. There are six combinations of civil status and urban/rural, and for 
 each we calculate the total exposure time. The total distribution by confounders is
 
-```{r exweight}
-exm <- vb0Tab[vb0Tab$period == "1901-1925" & vb0Tab$sex == "male", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-all <- aggregate(exm[, c("exposure", "event")], by = exm[, c("civst", "urban")], FUN = sum)
-all
-```
-
-and for the elite we have
-
-```{r forelite}
-elite <- exm[exm$hisclass == "elite", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(elite[, c("exposure", "event")], by = elite[, c("civst", "urban")], FUN = sum)
-el
-```
-
-The weights are now 'all["exposure"]" / el["exposure"]`, that is
-
-```{r wei}
-wei <-  all$exposure / el$exposure
-all$wei <- wei 
-all$Expos <- all$exposure * wei
-all$Eve <- all$event * wei
-all
-```
-
-Put these weights on the data frame `elite`:
-
-```{r putonel}
-all$id <- paste(all$civst, all$urban, sep = "_")
-elite$id <- paste(elite$civst, elite$urban, sep = "_")
-indx <- match(elite$id, all$id)
-elite$wght <- all$wei[indx]
-```
-
-Now repeat this for the other four hisclasses:
-
-```{r therest}
-middle <- exm[exm$hisclass == "middle", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(middle["exposure"], by = middle[, c("civst", "urban")], FUN = sum)
-all$wei <- all$exposure / el$exposure
-middle$id <- paste(middle$civst, middle$urban, sep = "_")
-indx <- match(middle$id, all$id)
-middle$wght <- all$wei[indx]
-el
-##
-worker <- exm[exm$hisclass == "worker", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(worker["exposure"], by = worker[, c("civst", "urban")], FUN = sum)
-all$wei <- all$exposure / el$exposure
-worker$id <- paste(worker$civst, worker$urban, sep = "_")
-indx <- match(worker$id, all$id)
-worker$wght <- all$wei[indx]
-el
-##
-farmer <- exm[exm$hisclass == "farmer", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(farmer["exposure"], by = farmer[, c("civst", "urban")], FUN = sum)
-all$wei <- all$exposure / el$exposure
-farmer$id <- paste(farmer$civst, farmer$urban, sep = "_")
-indx <- match(farmer$id, all$id)
-farmer$wght <- all$wei[indx]
-el
-##
-non <- exm[exm$hisclass == "NA", c("age", "civst", "urban", "exposure", "event", "hisclass")]
-el <- aggregate(non["exposure"], by = non[, c("civst", "urban")], FUN = sum)
-all$wei <- all$exposure / el$exposure
-non$id <- paste(non$civst, non$urban, sep = "_")
-indx <- match(non$id, all$id)
-non$wght <- all$wei[indx]
-el
-```
-
-And put together:
-
-```{r puttog}
-per1901 <- rbind(elite, middle, worker, farmer, non)
-per1901$age <- factor(per1901$age)
-#fit <- lm(exposure ~ civst * urban * age, data = per1901)
-#pred <- predict(fit, type = "response")
-#per1901$wght2 <- pred / per1901$exposure
-```
-
-Let us fit with and without weights. Without weights, see Figure \@ref(fig:testfit).
+```{r exweight, echo = TRUE}
+vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
+fit1 <- glm(I(hisclass == "elite") ~ urban * civst, data = vb2_2005_men, family = binomial)
+fit2 <- glm(I(hisclass == "middle") ~ urban * civst, data = vb2_2005_men, family = binomial)
+fit3 <- glm(I(hisclass == "worker") ~ urban * civst, data = vb2_2005_men, family = binomial)
+fit4 <- glm(I(hisclass == "farmer") ~ urban * civst, data = vb2_2005_men, family = binomial)
+fit5 <- glm(I(hisclass == "none") ~ urban * civst, data = vb2_2005_men, family = binomial)
+wght <- numeric(NROW(vb2_2005_men))
+##
+gg <- vb2_2005_men$hisclass == "elite"
+wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+##
+gg <- vb2_2005_men$hisclass == "middle"
+wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+##
+gg <- vb2_2005_men$hisclass == "worker"
+wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+##
+gg <- vb2_2005_men$hisclass == "farmer"
+wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+##
+gg <- vb2_2005_men$hisclass == "none"
+wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+##
+vb2_2005_men$wght <- wght
+```
+
+Let us fit with and without weights, see Figure \@ref(fig:testfit).
 
 ```{r testfit, fig.cap = "Survival by hisclass, men 1901-1925, non-weighted analysis.", fig.height = 4}
-fit1 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "elite", ], family = poisson)
-co1 <- fit1$coefficients
-co1[2:length(co1)] <- co1[2:length(co1)] + co1[1]
-co1 <- exp(co1)
-##
-fit2 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "middle", ], family = "poisson")
-co2 <- fit2$coefficients
-co2[2:length(co2)] <- co2[2:length(co2)] + co2[1]
-co2 <- exp(co2)
-##
-fit3 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "worker", ], family = "poisson")
-co3 <- fit3$coefficients
-co3[2:length(co3)] <- co3[2:length(co3)] + co3[1]
-co3 <- exp(co3)
-##
-fit4 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "farmer", ], family = "poisson")
-co4 <- fit4$coefficients
-co4[2:length(co4)] <- co4[2:length(co4)] + co4[1]
-co4 <- exp(co4)
-##
-fit5 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "NA", ], family = "poisson")
-co5 <- fit5$coefficients
-co5[2:length(co5)] <- co5[2:length(co5)] + co5[1]
-co5 <- exp(co5)
-###
-cuts <- seq(5, 55, by = 5)
-x <- seq(0, 60, length = 1000)
-s1 <- ppch(x, cuts = cuts, levels = co1[1:12], lower.tail = FALSE)
-plot(x, s1, type = "l", col = scol[1], axes = FALSE, xlab = "age", ylab = "Surviving fraction after 30.")
-axis(1, at = seq(0, 60, by = 10), labels = seq(30, 90, by = 10))
-box()
-axis(2, las = 1)
+fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = vb2_2005_men)
+fitw <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = vb2_2005_men)
+oldpar <- par(mfrow = c(1, 2))
+plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), xlab = "Age", ylab = "Survival", main = "Unweighted")
 abline(h = 0)
-abline(v = 0)
-s2 <- ppch(x, cuts = cuts, levels = co2[1:12], lower.tail = FALSE)
-lines(x, s2, col = scol[2], lty = 2)
-s3 <- ppch(x, cuts = cuts, levels = co3[1:12], lower.tail = FALSE)
-lines(x, s3, col = scol[3], lty = 3)
-s4 <- ppch(x, cuts = cuts, levels = co4[1:12], lower.tail = FALSE)
-lines(x, s4, col = scol[4], lty = 4)
-s5 <- ppch(x, cuts = cuts, levels = co5[1:12], lower.tail = FALSE)
-lines(x, s5, col = scol[5], lty = 5)
-##
-```
-
-With weights, see Figure \@ref(fig:testfit2):
-
-```{r testfit2, fig.cap = "Survival by hisclass, men 1901-1925, weighted analysis.", fig.height = 4}
-fit1 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "elite", ], family = poisson, weights = wght)
-co1 <- fit1$coefficients
-co1[2:length(co1)] <- co1[2:length(co1)] + co1[1]
-co1 <- exp(co1)
-##
-fit2 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "middle", ], family = "poisson", weights = wght)
-co2 <- fit2$coefficients
-co2[2:length(co2)] <- co2[2:length(co2)] + co2[1]
-co2 <- exp(co2)
-##
-fit3 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "worker", ], family = "poisson", weights = wght)
-co3 <- fit3$coefficients
-co3[2:length(co3)] <- co3[2:length(co3)] + co3[1]
-co3 <- exp(co3)
-##
-fit4 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "farmer", ], family = "poisson", weights = wght)
-co4 <- fit4$coefficients
-co4[2:length(co4)] <- co4[2:length(co4)] + co4[1]
-co4 <- exp(co4)
-##
-fit5 <- glm(event ~ offset(log(exposure)) + age, data = per1901[per1901$hisclass == "NA", ], family = "poisson", weights = wght)
-co5 <- fit5$coefficients
-co5[2:length(co5)] <- co5[2:length(co5)] + co5[1]
-co5 <- exp(co5)
-###
-cuts <- seq(5, 55, by = 5)
-x <- seq(0, 60, length = 1000)
-s1 <- ppch(x, cuts = cuts, levels = co1[1:12], lower.tail = FALSE)
-plot(x, s1, type = "l", col = scol[1], axes = FALSE, xlab = "Age", ylab = "Surviving fraction after 30")
-axis(1, at = seq(0, 60, by = 10), labels = seq(30, 90, by = 10))
-axis(2, las = 1)
-box()
+abline(h = 0.5, lty = 3)
+med <- summary(survfit(fit))$table[, "median"]
+lines(c(med["none"], med["none"]), c(0, 0.5), col = "red", lty = 3)
+lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
+plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), xlab = "Age", ylab = "Survival", main = "Weighted")
 abline(h = 0)
-abline(v = 0)
-s2 <- ppch(x, cuts = cuts, levels = co2[1:12], lower.tail = FALSE)
-lines(x, s2, col = scol[2], lty = 2)
-s3 <- ppch(x, cuts = cuts, levels = co3[1:12], lower.tail = FALSE)
-lines(x, s3, col = scol[3], lty = 3)
-s4 <- ppch(x, cuts = cuts, levels = co4[1:12], lower.tail = FALSE)
-lines(x, s4, col = scol[4], lty = 4)
-s5 <- ppch(x, cuts = cuts, levels = co5[1:12], lower.tail = FALSE)
-lines(x, s5, col = scol[5], lty = 5)
-##
+abline(h = 0.5, lty = 3)
+medw <- summary(survfit(fitw))$table[, "median"]
+lines(c(medw["none"], medw["none"]), c(0, 0.5), col = "red", lty = 3)
+lines(c(medw["elite"], medw["elite"]), c(0, 0.5), col = "black", lty = 3)
+medi <- cbind(med, medw)
+colnames(medi) <- c("Unweighted", "weighted")
+##
+```
+
+The estimated medians are
+
+```{r estmed}
+knitr::kable(round(medi, 1))
 ```

### 2018-03-05 R/oneSpell.R
first commit

@@ -1,0 +1,13 @@
+oneSpell <- function(dat){
+    ## 'dat' is a survival data frame, with
+    ## id, enter, exit, event (+ some covariates)
+    ##
+    ## Output is a dataframe with one record per id.
+    indx <- tapply(dat$id, dat$id)
+    dat$enter <- tapply(dat$enter, dat$id, min)[indx]
+    dat$exit <- tapply(dat$exit, dat$id, max)[indx]
+    dat$event <- tapply(dat$event, dat$id, max)[indx]
+    dat <- dat[!duplicated(dat$id), ]
+    dat <- dat[dat$exit > dat$enter, ] # Clean...no big deal!
+    dat
+}

### 2018-03-06 article3d.Rmd
weights complete

@@
 load("data/foli.rda")
+foli$civst[is.na(foli$civst)] <- "unmarried"
 foli <- foli[order(foli$id, foli$enter), ]
@@
 
-As an illustration, take the period `2005-2013` with males, 65--99 years of age. There are six combinations of civil status and urban/rural, and for 
-each we calculate the total exposure time. The total distribution by confounders is
-
-```{r exweight, echo = TRUE}
+As an illustration, take the period `2005-2013` with males, 65--99 years of age.
+
+```{r exweight}
 vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
@@
 gg <- vb2_2005_men$hisclass == "middle"
-wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+wght[gg] <- 1 / predict(fit2, type = "response")[gg]
 ##
 gg <- vb2_2005_men$hisclass == "worker"
-wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+wght[gg] <- 1 / predict(fit3, type = "response")[gg]
 ##
 gg <- vb2_2005_men$hisclass == "farmer"
-wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+wght[gg] <- 1 / predict(fit4, type = "response")[gg]
 ##
 gg <- vb2_2005_men$hisclass == "none"
-wght[gg] <- 1 / predict(fit1, type = "response")[gg]
+wght[gg] <- 1 / predict(fit5, type = "response")[gg]
 ##
@@
 fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = vb2_2005_men)
+med <- summary(survfit(fit))$table[, "median"]
 fitw <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = vb2_2005_men)
+medw <- summary(survfit(fitw))$table[, "median"]
 oldpar <- par(mfrow = c(1, 2))
-plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), xlab = "Age", ylab = "Survival", main = "Unweighted")
+plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), 
+     xlab = "Age", ylab = "Survival", main = "Unweighted", axes = FALSE)
+tics <- c(65, 70, 75, round(med["none"], 1), round(med["elite"], 1), 90, 95, 100)
+axis(1, at = tics)
+axis(2, at = c(0, 0.5, 1), las = 1)
+box()
 abline(h = 0)
 abline(h = 0.5, lty = 3)
-med <- summary(survfit(fit))$table[, "median"]
+
 lines(c(med["none"], med["none"]), c(0, 0.5), col = "red", lty = 3)
 lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
-plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), xlab = "Age", ylab = "Survival", main = "Weighted")
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), 
+     xlab = "Age", ylab = "Survival", main = "Weighted", axes = FALSE)
+tics <- c(65, 70, 75, round(medw["none"], 1), round(medw["elite"], 1), 90, 95, 100)
+axis(1, at = tics)
+axis(2, at = c(0, 0.5, 1), las = 1)
+box()
 abline(h = 0)
 abline(h = 0.5, lty = 3)
-medw <- summary(survfit(fitw))$table[, "median"]
+
 lines(c(medw["none"], medw["none"]), c(0, 0.5), col = "red", lty = 3)
 lines(c(medw["elite"], medw["elite"]), c(0, 0.5), col = "black", lty = 3)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
 medi <- cbind(med, medw)
@@
 
+For the age interval 40--64 we get
+
+```{r forty64}
+source("R/getWeights.R")
+da <- vb1[vb1$sex == "male" & vb1$period == "2005-2013", ]
+da$wght <- getWeights(da)
+##
+fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = da)
+med <- summary(survfit(fit))$table[, "median"]
+fitw <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = da)
+medw <- summary(survfit(fitw))$table[, "median"]
+oldpar <- par(mfrow = c(1, 2))
+plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65), 
+     xlab = "Age", ylab = "Survival", main = "Unweighted", axes = FALSE)
+tics <- c(40, 45, 50, 55, 60, 65)
+axis(1, at = tics)
+axis(2, at = c(0, 0.5, 1), las = 1)
+box()
+abline(h = 0)
+abline(h = 0.5, lty = 3)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+##lines(c(med["none"], med["none"]), c(0, 0.5), col = "red", lty = 3)
+##lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
+plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65), 
+     xlab = "Age", ylab = "Survival", main = "Weighted", axes = FALSE)
+tics <- c(40, 45, 50, 55, 60, 65)
+axis(1, at = tics)
+axis(2, at = c(0, 0.5, 1), las = 1)
+box()
+abline(h = 0)
+abline(h = 0.5, lty = 3)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+```
+
+We cannot use the "median remaining life" as a summary comparison measure, for obvious reasons.
+Instead, the *area under the survival curves* will be used.
+This area can be interpreted as the *expected years of life before 65(100)*.
+Here we get
+
+```{r explife40}
+round(summary(survfit(fitw), rmean = 65)$table[, "*rmean"], 2)
+```
+
+This is now repeated for each period, and for men and women separately.
+
 # The effect of social class on mortality
@@
 
-```{r inipw}
+```{r inipw, eval = FALSE}
 source("R/ipw.R")
@@
 ##
-vb3064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
-vb3064$age <- factor(vb3064$age)
+##vb3064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
+##vb3064$age <- factor(vb3064$age)
 ##vb3064$hisclass <- vb3064$hisclass40
@@
 ```{r hazcalw}
-source("R/getPHglm2.R")
-source("R/getPHglm.R")
-##nper <- 4
 nper <- length(labb)
-resw <- getPHglm2(vb3064[vb3064$sex == "female", ],
-            nper, labb)
-resm <- getPHglm2(vb3064[vb3064$sex == "male", ],
-            nper, labb)
+his <- levels(vb1$hisclass)
+nhis <- length(his)
+##
+resm <- matrix(0, nrow = nhis, ncol = nper)
+for (i in 1:nper){
+    dat <- vb1[vb1$sex == "male" & vb1$period == labb[i], ]
+    dat$wght <- getWeights(dat)
+    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    resm[, i] <- summary(survfit(fit), rmean = 65)$table[, "*rmean"]
+}
+##
+resw <- matrix(0, nrow = nhis, ncol = nper)
+for (i in 1:nper){
+    dat <- vb1[vb1$sex == "female" & vb1$period == labb[i], ]
+    dat$wght <- getWeights(dat)
+    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    resw[, i] <- summary(survfit(fit), rmean = 65)$table[, "*rmean"]
+}
+
 ```
@@
 oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
-                main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
-                main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
+midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
+plot(midper, resw[1, ], col = scol[1], type = "b", ylim = c(58, 65), xlab = "Period", 
+     ylab = "Age", axes = FALSE, main = "Women, 40-64")
+
+axis(1, at = midper, labels = labb)
+axis(2, las = 1)
+box()
+for (i in 2:5){
+    lines(midper, resw[i, ], col = scol[i], type = "b")
+}
+plot(midper, resm[1, ], col = scol[1], type = "b", ylim = c(58, 65), xlab = "Period", 
+     ylab = "Age", axes = FALSE, main = "Men, 40-64")
+axis(1, at = midper, labels = labb)
+axis(2, las = 1)
+box()
+for (i in 2:5){
+    lines(midper, resm[i, ], col = scol[i], type = "b")
+}
+##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
+##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
 par(oldpar)
@@
 
-```{r cumswtab,message=FALSE}
-vb3064$social_class <- vb3064$hisclass
-vb3064$marital_status <- vb3064$civst
-vb3064$urban_or_rural <- vb3064$urban
-vb3064$age_group <- vb3064$age
-fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
-            data = vb3064, family = poisson)
+```{r cumswtab, message=FALSE, cache = TRUE, eval = FALSE}
+fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
+
+# vb3064$social_class <- vb3064$hisclass
+# vb3064$marital_status <- vb3064$civst
+# vb3064$urban_or_rural <- vb3064$urban
+# vb3064$age_group <- vb3064$age
+# fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
+#             data = vb3064, family = poisson)
 dr <- drop1(fitt, test = "Chisq")
-##dr
+# ##dr
 knitr::kable(dr, caption = "Single term deletions, ages 30-64.")
-##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."), 
+# ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."), 
   ##    type = "latex", comment = FALSE)
@@
 
+<!--
 By standardizing the levels in each time period for each class, we 
@@
 
-```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm"}
+```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm", eval = FALSE}
 oldpar <- par(mfrow = c(2, 1))
@@
 ```
+
+-->
 
@@
 
-```{r rawolddat,echo=FALSE}
-rawskel <- get.file("rawskel")
-rawskel$urban <- FALSE
-rawskel <- rawskel[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
-##
-rawskume <- get.file("rawskume")
-rawskume <- rawskume[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
-raw <- rbind(rawskel, rawskume)
-raw$id <- paste("pop", raw$id, sep = "_")
-##
-rawfoli <- get.file("rawfoli")
-rawfoli <- rawfoli[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
-raw <- rbind(raw, rawfoli)
-raw <- raw[!(raw$period %in% c("1961-1970", "1971-1975")), ]
-raw$period <- factor(raw$period)
-raw <- age.window(raw, c(65, 100))
-```
-
-```{r getske6099}
-vb6599 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
-vb6599$age <- factor(vb6599$age) # NOTE:: Change to '6599'!
-##vb6599$hisclass <- vb6599$hisclass60
-```
-
 We now turn to the elderly, those who in most cases have left the workforce.
@@
 
-```{r cumswtabold,cap="Table"}
-vb6599$marital_status <- vb6599$civst
-vb6599$social_class <- vb6599$hisclass
-vb6599$urban_or_rural <- vb6599$urban
-vb6599$age_group <- vb6599$age
-fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
-            data = vb6599, family = poisson)
+```{r cumswtabold, cap="Table", cache = TRUE, eval = FALSE}
+# vb6599$marital_status <- vb6599$civst
+# vb6599$social_class <- vb6599$hisclass
+# vb6599$urban_or_rural <- vb6599$urban
+# vb6599$age_group <- vb6599$age
+# fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
+#             data = vb6599, family = poisson)
+
+fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
 drold <- drop1(fittold, test = "Chisq")
-knitr::kable(dr, caption = "Single term deletions, ages 65-99.")
+knitr::kable(drold, caption = "Single term deletions, ages 65-99.")
 ##drold
@@
   ##    type = "latex", comment = FALSE)
+```
+
+```{r hazcalw2}
+nper <- length(labb)
+his <- levels(vb1$hisclass)
+nhis <- length(his)
+##
+resm <- matrix(0, nrow = nhis, ncol = nper)
+for (i in 1:nper){
+    dat <- vb2[vb2$sex == "male" & vb2$period == labb[i], ]
+    dat$wght <- getWeights(dat)
+    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+}
+##
+resw <- matrix(0, nrow = nhis, ncol = nper)
+for (i in 1:nper){
+    dat <- vb2[vb2$sex == "female" & vb2$period == labb[i], ]
+    dat$wght <- getWeights(dat)
+    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+}
+
+```
+
+```{r cumsw2, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
+##source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
+mm <- resw[3, ]
+plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
+     ylab = "Excess years", axes = FALSE, main = "Women, 65-99")
+
+axis(1, at = midper, labels = labb)
+axis(2, las = 1)
+box()
+for (i in 2:5){
+    lines(midper, resw[i, ] - mm, col = scol[i], type = "b")
+}
+
+mm <- resm[3, ]
+plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
+     ylab = "Excess years", axes = FALSE, main = "Men, 65-99")
+axis(1, at = midper, labels = labb)
+axis(2, las = 1)
+box()
+for (i in 2:5){
+    lines(midper, resm[i, ] - mm, col = scol[i], type = "b")
+}
+##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
+##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
+par(oldpar)
 ```
@@
 workers of retirement age have slightly higher mortality. The difference between the worker and middle group has been small in most periods.
-
-```{r rawreswomen}
-hc <- levels(raw$hisclass)
-n.hc <- length(hc)
-peri <- levels(raw$period)
-n.peri <- length(peri)
-med <- matrix(0, nrow = n.hc, ncol = n.peri)
-for (i in 1:n.hc){
-    fit <- coxph(Surv(enter, exit, event) ~ civst + urban + strata(period), data = raw[raw$sex == "female" & 
-                                                                              raw$hisclass == hc[i], ])
-    med[i, ] <- summary(survfit(fit, newdata = data.frame(civst = "married", urban = FALSE)))$table[, "median"]
-}
-```
-
-```{r plotrawwom, fig.cap="Median age at death for a 65 year old married woman on the countryside by SES and period.", fig.height = 4, out.height = "10cm"}
-x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
-plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Women", axes = FALSE)
-axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
-axis(2, las = 1)
-box()
-abline(h = c(75, 80, 85), lty = 3)
-for (i in 2:5){
-    lines(x, med[i, -7], col = scol[i], type = "b", lty = i, pch = i)
-}
-legend("bottomright", legend = hc, lty = 1:5, pch = 1:5, col = scol, cex = 0.6)
-```
-
-```{r rawresmen}
-hc <- levels(raw$hisclass)
-n.hc <- length(hc)
-peri <- levels(raw$period)
-n.peri <- length(peri)
-med <- matrix(0, nrow = n.hc, ncol = n.peri)
-for (i in 1:n.hc){
-    fit <- coxph(Surv(enter, exit, event) ~ civst + urban + strata(period), data = raw[raw$sex == "male" & 
-                                                                              raw$hisclass == hc[i], ])
-    med[i, ] <- summary(survfit(fit, newdata = data.frame(civst = "married", urban = FALSE)))$table[, "median"]
-}
-for (i in 1:n.hc){
-    fit <- coxph(Surv(enter, exit, event) ~ strata(period), data = raw[raw$sex == "male" & 
-                                                                              raw$hisclass == hc[i], ])
-    med[i, ] <- summary(survfit(fit))$table[, "median"]
-}
-```
-
-```{r plotrawm, fig.cap="Median age at death for a 65 year old married man on the countryside by SES and period.", fig.height = 4}
-x <- c(1812, 1837, 1862, 1887, 1912, 1937, 1975, 1985, 1995, 2004, 2011)
-plot(x, med[1, -7], col = scol[1], type = "b", lty = 1, pch = 1, xlab = "Period", ylab = "Age", ylim = c(70, 90), main = "Men", axes = FALSE)
-axis(1, at = x, labels = peri[-7], las = 2, cex.axis = 0.5)
-axis(2, las = 1)
-abline(h = c(75, 80, 85), lty = 3)
-box()
-for (i in 2:5){
-    lines(x, med[i, -7], col = scol[i], type = "b", lty = i, pch = i)
-}
-legend("bottomright", legend = hc, lty = 1:5, pch = 1:5, col = scol, cex = 0.6)
-```
 

### 2018-03-06 R/getWeights.R
first commit

@@ -1,0 +1,15 @@
+getWeights <- function(dat){
+    options(na.action = "na.exclude")
+    ## "Inverse probability weights" are created.
+    hlevs <- levels(dat$hisclass)
+    n <- length(hlevs)
+    wght <- numeric(NROW(dat))
+    ##
+    for (i in seq_along(hlevs)){
+        fit <- glm(I(hisclass == hlevs[i]) ~ urban * civst, data = dat, family = poisson)
+        gg <- dat$hisclass == hlevs[i]
+        wght[gg] <- 1 / predict(fit, type = "response")[gg]
+    }
+    ##
+    wght
+}

### 2018-03-06 article3d.Rmd
clean figures

@@
 
-The estimated medians are
+The estimated medians are (see Table \@ref(tab:estmed).
 
 ```{r estmed}
-knitr::kable(round(medi, 1))
+knitr::kable(round(medi, 1), caption = "Estimated medians, men 65-99, 2005-2013.")
 ```
@@
 
-```{r forty64}
+```{r forty64, fig.cap = "Survival curves, men 40-65, 2005-2013.", fig.height = 4}
 source("R/getWeights.R")
@@
 
-# The effect of social class on mortality
-
 ## All ages, survival curves
@@
 
-```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
+```{r cumsw, fig.cap = "Expected years of survival until the age of 65, given survival to age 40, by period and hisclass.", fig.scap = scap0, fig.height = 8, out.height = "14cm"}
 source("R/plotED.glm.R")
@@
 plot(midper, resw[1, ], col = scol[1], type = "b", ylim = c(58, 65), xlab = "Period", 
-     ylab = "Age", axes = FALSE, main = "Women, 40-64")
-
-axis(1, at = midper, labels = labb)
+     ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
+
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
 axis(2, las = 1)
+abline(h = 65)
 box()
 for (i in 2:5){
-    lines(midper, resw[i, ], col = scol[i], type = "b")
-}
+    lines(midper, resw[i, ], col = scol[i], type = "b", pch = i, lty = i)
+}
+legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
 plot(midper, resm[1, ], col = scol[1], type = "b", ylim = c(58, 65), xlab = "Period", 
-     ylab = "Age", axes = FALSE, main = "Men, 40-64")
-axis(1, at = midper, labels = labb)
+     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
 axis(2, las = 1)
+abline(h = 65)
 box()
 for (i in 2:5){
-    lines(midper, resm[i, ], col = scol[i], type = "b")
-}
+    lines(midper, resm[i, ], col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",

### 2018-03-07 CreData/fromFobLisa.Rmd
trivia

@@
 library(stringr)
-library(tidyr)
+##library(tidyr)
 library(dplyr)

### 2018-03-07 article3d.Rmd
trivia

@@
 }
-legend("bottomleft", legend = levhis, lty = 1:5, col = scol, pch = 1:5)
+legend("bottomleft", legend = levhis, lty = 1:5, col = scol, pch = 1:5, cex = 0.8)
 ```
@@
 
-```{r exweight}
+```{r exweight,echo=TRUE}
 vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
@@
 
-```{r testfit, fig.cap = "Survival by hisclass, men 1901-1925, non-weighted analysis.", fig.height = 4}
+```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4}
 fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = vb2_2005_men)
@@
      xlab = "Age", ylab = "Survival", main = "Unweighted", axes = FALSE)
-tics <- c(65, 70, 75, round(med["none"], 1), round(med["elite"], 1), 90, 95, 100)
-axis(1, at = tics)
+tics <- c(65, round(med["none"], 1), round(med["elite"], 1), 100)
+axis(1, at = tics, las = 2)
 axis(2, at = c(0, 0.5, 1), las = 1)
@@
 lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
-legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.8)
 plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), 
      xlab = "Age", ylab = "Survival", main = "Weighted", axes = FALSE)
-tics <- c(65, 70, 75, round(medw["none"], 1), round(medw["elite"], 1), 90, 95, 100)
-axis(1, at = tics)
+tics <- c(65, round(medw["none"], 1), round(medw["elite"], 1), 100)
+axis(1, at = tics, las = 2)
 axis(2, at = c(0, 0.5, 1), las = 1)
@@
 lines(c(medw["elite"], medw["elite"]), c(0, 0.5), col = "black", lty = 3)
-legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.8)
 medi <- cbind(med, medw)
@@
 abline(h = 0.5, lty = 3)
-legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.7)
 ##lines(c(med["none"], med["none"]), c(0, 0.5), col = "red", lty = 3)
@@
 abline(h = 0.5, lty = 3)
-legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5)
+legend("bottomleft", legend = levels(vb2$hisclass), col = scol, lty = 1:5, cex = 0.7)
 ```
@@
 
-```{r cumsw2, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
+```{r cumsw2, fig.cap = "Excess years of life (compared to 'worker') before 100 for a 65-year-old, by hisclass and period (experimental!).", fig.scap = scap0, fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")
@@
 mm <- resw[3, ]
-plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
+plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-4, 5), xlab = "Period", 
      ylab = "Excess years", axes = FALSE, main = "Women, 65-99")
 
-axis(1, at = midper, labels = labb)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
 axis(2, las = 1)
@@
 for (i in 2:5){
-    lines(midper, resw[i, ] - mm, col = scol[i], type = "b")
-}
-
+    lines(midper, resw[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
+}
+
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
 mm <- resm[3, ]
-plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
+plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-4, 5), xlab = "Period", 
      ylab = "Excess years", axes = FALSE, main = "Men, 65-99")
-axis(1, at = midper, labels = labb)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
 axis(2, las = 1)
@@
 for (i in 2:5){
-    lines(midper, resm[i, ] - mm, col = scol[i], type = "b")
-}
+    lines(midper, resm[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",

### 2018-03-10 article3d.Rmd
small fixes

@@
 The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
-In this version of article3, we are adopting a marginal approach, in contrast to earlier versions, where the approach was conditional.
-In addition to article3c, we are going back to individual, raw, data."
+We are adopting a marginal approach to the problem of adjust for confounders by impementing the inverse probability weighting technique."
 ---
@@
 ##
-skume1 <- age.window(skume, c(40, 65))
+skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
 skume1 <- oneSpell(skume1)
@@
 
-# Models
-
-The proportional hazards model in a survival analysis context is used,
-allowing for adjustment for marital status and rural/urban environment.
-As the central explanatory variable, social class, does not follow the
-proportionality property, the analyses are stratified with respect to the
-variable, and the main results are presented graphically.
-
-Another practical problem is the size of the data set:
-It contains no less than `r as.integer(round(n.exp + pn.exp))` person years and
- `r n.d + pn.d` deaths. This makes it highly inconvenient to use the currently so popular semi-parametric technique known as \emph{Cox regression} [@cox72]. Instead we use a parametric model
-that allows for a reduction of the data to \emph{sufficient statistics} [@fisher22]. The idea is to keep all relevant covariates as \emph{categorical} and then create a table with all
-possible combinations of levels of the covariates. For each such
-combination, the total amount of exposure and the number of deaths are
-calculated.  We have 2 sexes, 3 marital statuses, 12 periods, 2
-urban/rural groups, 5 social classes,
-14 age groups, and 22 cohorts; in all, 221,760 possible
-combinations. However, in this scheme there are many structural zeroes
-(impossible combinations) and some random zeros, so the actual number of
-combinations we end up with is 17,181. The data set is thus reduced in
-size by 99 per cent.
-
+# Model
+
+The comparison of mortality in the five social classes, period by period, is done
+by comparing estimated survival curves, adjusted for the distribution of the confounding variables over the social classes.
+The adjustment carried out by inverse probability weighting (**reference**) in the following way: First, for each social class, 
+the probability that an individual in the class belongs to that class is estimated by logistic regression with the 
+confounders as explanatory variables. Second, these probabilities are inverted and used as weights in the estimation of the survival 
+curves for each social class.
+
+<!--
 ## The conditional model
@@
 
-## The marginal model
-
-A alternative, and preferable, approach is the marginal model, where the disturbace effect of confounders is 
+-->
+
+The disturbace effect of confounders is 
 balanced out through weighting. In our case , `civil status` and `urban/rural` are confounders, that is, they 
@@
 
-### Association between civil status and social class over time
+## Association between civil status and social class over time
 
@@
 
-### Association between urban/rural and social class over time
+## Association between urban/rural and social class over time
 
@@
 
-### Inverse probability weighting
+## Inverse probability weighting
 
@@
 
-```{r exweight,echo=TRUE}
+```{r exweight,echo=TRUE,echo = FALSE}
 vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
@@
 ```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4}
+par(cex.axis = 0.8)
 fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = vb2_2005_men)
@@
 tics <- c(65, round(med["none"], 1), round(med["elite"], 1), 100)
-axis(1, at = tics, las = 2)
+axis(1, at = tics, las = 1, cex.axis = 0.8)
 axis(2, at = c(0, 0.5, 1), las = 1)
@@
 tics <- c(65, round(medw["none"], 1), round(medw["elite"], 1), 100)
-axis(1, at = tics, las = 2)
+axis(1, at = tics, las = 1, cex.axis = 0.8)
 axis(2, at = c(0, 0.5, 1), las = 1)
@@
 
-The estimated medians are (see Table \@ref(tab:estmed).
+The social class differences are smaller in the weighted analysis, meaning that some of the raw differences are due to different distributions of the confounders  between the social classes.
+
+For simple comparison over time, we need  simple summary measures, for instance the medians, shown in Figure \@ref(fig:testfit) and Table \@ref(tab:estmed).
 
@@
 
-For the age interval 40--64 we get
-
-```{r forty64, fig.cap = "Survival curves, men 40-65, 2005-2013.", fig.height = 4}
+However, for the age interval 40--64 the median is not a good measure, because the probability to survive age 65, given survival to age 40, is much larger than 0.5, see Figure \@ref(fig:forty64).
+
+```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4}
 source("R/getWeights.R")
@@
 ##
-fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = da)
-med <- summary(survfit(fit))$table[, "median"]
-fitw <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = da)
-medw <- summary(survfit(fitw))$table[, "median"]
+fit0 <- coxph(Surv(enter, exit, event) ~ strata(hisclass), data = da)
+med0 <- summary(survfit(fit))$table[, "median"]
+fitw0 <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = da)
+medw0 <- summary(survfit(fitw))$table[, "median"]
 oldpar <- par(mfrow = c(1, 2))
-plot(survfit(fit), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65), 
-     xlab = "Age", ylab = "Survival", main = "Unweighted", axes = FALSE)
+plot(survfit(fitw0), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65), 
+     xlab = "Age", ylab = "Survival", main = "40-65", axes = FALSE)
 tics <- c(40, 45, 50, 55, 60, 65)
@@
 ##lines(c(med["elite"], med["elite"]), c(0, 0.5), col = "black", lty = 3)
-plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(40, 65), 
-     xlab = "Age", ylab = "Survival", main = "Weighted", axes = FALSE)
-tics <- c(40, 45, 50, 55, 60, 65)
+plot(survfit(fitw), mark.time = FALSE, col = scol, lty = 1:5, xlim = c(65, 100), 
+     xlab = "Age", ylab = "Survival", main = "65-100", axes = FALSE)
+tics <- c(65, 70, 75, 80, 85, 90, 95, 100)
 axis(1, at = tics)
@@
 
-We cannot use the "median remaining life" as a summary comparison measure, for obvious reasons.
+We cannot use the "median remaining life" as a summary comparison measure here, for obvious reasons.
 Instead, the *area under the survival curves* will be used.
-This area can be interpreted as the *expected years of life before 65(100)*.
+This area can be interpreted as the *expected years of life before 65(100)* for a person aged 40(65).
 Here we get
@@
 ```{r explife40}
-round(summary(survfit(fitw), rmean = 65)$table[, "*rmean"], 2)
+avew <- round(summary(survfit(fitw), rmean = 100)$table[, "*rmean"], 2)
+avew0 <- round(summary(survfit(fitw0), rmean = 65)$table[, "*rmean"], 2)
+aver <- cbind(avew0, avew)
+colnames(aver) <- c("Men, 40-64", "Men, 65-99")
+knitr::kable(round(aver, 1), caption = "Expected years of life before 65(100) for a 40(65) year old man, by social class." )
 ```
@@
 This is now repeated for each period, and for men and women separately.
-
-## All ages, survival curves
 
@@
 
-## Ages 30--64
-
-```{r getske3064}
+# Results
+
+In the graphical presentation of the results, the *worker* social class is used as reference, and all curves show the difference in years to the expected years in the reference.
+
+## Ages 40--64
+
+```{r getske3064, eval = FALSE}
 cap1w <-  "Cumulative hazards for social class by time period, ages 30--64, women."
@@
 
-```{r cumsw, fig.cap = "Expected years of survival until the age of 65, given survival to age 40, by period and hisclass.", fig.scap = scap0, fig.height = 8, out.height = "14cm"}
+```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass.
+Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm"}
 source("R/plotED.glm.R")
@@
 midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
-plot(midper, resw[1, ], col = scol[1], type = "b", ylim = c(58, 65), xlab = "Period", 
+refclw <- resw[3, ]
+plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2030), 
      ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
@@
 for (i in 2:5){
-    lines(midper, resw[i, ], col = scol[i], type = "b", pch = i, lty = i)
+    lines(midper, resw[i, ] - refclw, col = scol[i], type = "b", pch = i, lty = i)
 }
 legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
-plot(midper, resm[1, ], col = scol[1], type = "b", ylim = c(58, 65), xlab = "Period", 
-     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1)
+refclm <- resm[3, ]
+plot(midper, resm[1, ] - refclm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
+     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2030))
 axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
@@
 for (i in 2:5){
-    lines(midper, resm[i, ], col = scol[i], type = "b", lty = i, pch = i)
+    lines(midper, resm[i, ] - refclm, col = scol[i], type = "b", lty = i, pch = i)
 }
@@
 
-Looking at Table \@ref(tab:cumswtab), it is obvious that the statistical significance is strong.
-
-```{r cumswtab, message=FALSE, cache = TRUE, eval = FALSE}
+Obviously, differences are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class, the exception being those who have no notation of social class.
+
+Hovever, looking at Table \@ref(tab:cumswtab), it is obvious that the *statistical significance* is strong.
+
+```{r cumswtab, message=FALSE, cache = TRUE, eval = TRUE}
 fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
@@
 
-<!--
-By standardizing the levels in each time period for each class, we 
-illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
-the paper, we only present these
-relative mortality levels for the different classes since they are the main focus in this study.
-
-```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm", eval = FALSE}
-oldpar <- par(mfrow = c(2, 1))
-woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(65|30)", ylim = c(0.0, 0.6), main = "Women 30-64")##, exclude = 5:6)
-moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
-                    ylab = "H(65|30)", ylim = c(0.0, 0.6), main = "Men 30-64")##, exclude = 5:6)
-par(oldpar)
-```
-
--->
-
 The changing economy has some consequences for the analysis, particularly when
@@
 
-```{r cumswtabold, cap="Table", cache = TRUE, eval = FALSE}
+```{r cumswtabold, cap="Table", cache = TRUE, eval = TRUE}
 # vb6599$marital_status <- vb6599$civst
@@
 
-```{r cumsw2, fig.cap = "Excess years of life (compared to 'worker') before 100 for a 65-year-old, by hisclass and period (experimental!).", fig.scap = scap0, fig.height = 8, out.height = "14cm"}
+```{r cumsw2, fig.cap = "Excess years of life (compared to 'worker') before 100 for a 65-year-old, by hisclass and period (experimental!).", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")

### 2018-03-11 article3b.Rmd
random

@@
 While some report substantial differences, others have found surprisingly small or even reversed class differences, for example in the 19th century Sundsvall region [@edli11; @edbr12].
-This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century [@bevp11].
+This raises the question of when the modern social gradient appeared, if this was something tls g*
+hat developed in the early parts of the 20th century [@bevp11].
 The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions.
@@
 
-In this article we set our task as analysing social differences in mortality from a long-term perspective.
+We set our task as analysing social differences in mortality from a long-term perspective.
 Until recently it has not been possible to investigate the truly long-term development of mortality differences among adult and elderly populations.

### 2018-03-14 marginal.Rmd

--- 2018-03-10 article3d.Rmd
@@
 We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
-The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.
-We are adopting a marginal approach to the problem of adjust for confounders by impementing the inverse probability weighting technique."
+The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
 ---
@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warnings = FALSE)##$
 library(knitr)
@@
 ```{r readalldata}
-load("data/skume.rda")
+source("R/get.file.R")
+skume <- get.file("skume")
+##load("data/skume.rda")
 skume <- skume[order(skume$id, skume$enter), ]
@@
 ##################################################
-load("data/foli.rda")
+foli <- get.file("foli")
+##load("data/foli.rda")
 foli$civst[is.na(foli$civst)] <- "unmarried"
@@
 \item {\tt farmer}, Hisclass 8. Farmers. 
-\item {\tt NA}, Not Available, unknown. 
+\item {\tt none}, not given, unknown. 
 \end{enumerate}
@@
 
-Marital status (see Figure \@ref(fig:civstskum)) is a time-dependent variable. 
+Marital status (see Figure \@ref(fig:civstskum)) is a time-fixed covariate.
 The statuses used are unmarried, married and dissolved marriage
@@
 axis(2, las = 1)
+axis(4, las = 1)
 abline(h = 65)
@@
 axis(2, las = 1)
+axis(4, las = 1,)
 abline(h = 65)
@@
 Obviously, differences are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class, the exception being those who have no notation of social class.
-
-Hovever, looking at Table \@ref(tab:cumswtab), it is obvious that the *statistical significance* is strong.
+In Sweden of the year 2014, the probability of surviving age 65 for a forty-year-old man is 91 per cent  (a period measure).
+The corresponding figure for a forty-year-old woman is 94 per cent.
+More interesting is to look at the late differences in the age group 65--100.
+
+Hovever, from Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong.
 
@@
 mm <- resw[3, ]
-plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-4, 5), xlab = "Period", 
+plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "Period", 
      ylab = "Excess years", axes = FALSE, main = "Women, 65-99")
@@
 axis(2, las = 1)
+axis(4, las = 1)
 box()
@@
 mm <- resm[3, ]
-plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-4, 5), xlab = "Period", 
+plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "Period", 
      ylab = "Excess years", axes = FALSE, main = "Men, 65-99")

### 2018-03-15 CreData/fromFobLisa.Rmd
fix Lexis diagram

@@
 we decide to use as birthdate the *midpoint of the birth quarter*. 
-WE work with all ages 0-110, and only as the final step, `age.window` is called.
+We work with all ages 0-110, and only as the final step, `age.window` is called.
 
@@
 per$DODSDAT <- socstyr$DODSDAT[indx]
-fixd <- !is.na(per$DODSDAT) & (per$DODSDAT < 1960010)
+fixd <- !is.na(per$DODSDAT) & (per$DODSDAT < 1960010) # Only month and year
 per$DODSDAT[fixd] <- per$DODSDAT[fixd] * 100 + 15

### 2018-03-15 marginal.Rmd
fix Lexis diagram

@@
 ```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
-plot(c(1801, 1801, 1951, 1951, 1801), c(30, 100, 100, 30, 30),
+plot(c(1801, 1801, 1951, 1951, 1801), c(40, 100, 100, 40, 40),
      type = "l", xlim = c(1800, 2014), lwd = 2, col = "red",
@@
 axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1986, 1996, 2005, 2014), las = 2)
-axis(2, las = 2, at = c(0, 30, 65, 100))
-axis(4, las = 2, at = c(0, 30, 65, 100))
+axis(2, las = 2, at = c(0, 40, 65, 100))
+axis(4, las = 2, at = c(0, 40, 65, 100))
 box()
 abline(h = 0)
-lines(c(1976, 1976, 2014, 2014, 1976), c(30, 100, 100, 30, 30),
+lines(c(1976, 1976, 2014, 2014, 1976), c(40, 100, 100, 40, 40),
       col = "blue", lwd = 2)

### 2018-03-15 marginal.Rmd
fix Lexis, turn off warnings

@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warnings = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
 library(knitr)

### 2018-03-25 article3b.Rmd
Small revisions

@@
 The results show that high social class is not always favourable for survival.
-Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
+Men in the elite category, particularly among those in working age, had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
 We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
@@
 Social conditions and social position certainly have impact, but not always in the expected direction.
-In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
+In our case this is apparent for men during a large part of the studied period, particularly among those in working age; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 During the studied period, Sweden developed from a poor agricultural society with low urbanization to a rich welfare state.
-If Sweden is peripheral from a European continental perspective, the regions studied here are remote from the central parts of the country.
+The regions studied here were for a long time remote from the central parts of the country.
 The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the county of Västerbotten in northern of Sweden along the coast of the Gulf
@@
 This region has a somewhat different character to that of Skellefteå.
-Umeå town had for a long time a small population, although substantially, and was the administrative, educational and military centre of the county of Västerbotten.
+Umeå town had for a long time a small population, and was the administrative, educational and military centre of the county of Västerbotten.
 During the latter part of the 20th century, the establishment of Umeå University led to a rapid population increase.
@@
 
-In the data set analysed here, all individuals aged 30 years and older ever
+In the data set analysed here, all individuals aged 40 years and older ever
 having resided in either of the regions are included. The data file
@@
 being a scientific choice, is also a practical one.
-Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population; i.e.
+Social position during working age is defined at around the age of 40 but defined from the last occupation in working age for the elderly and retired population; i.e.
 from age 65 until death or last observation.
@@
 population size. The modern Swedish welfare state began to develop. For the 
-period 1976 onwards, there are separate analyses of five- or six-year periods.
+period 1976 onwards, there are separate analyses of ten-year periods.
 Sweden had developed into a wealthy welfare society with low income inequality.

### 2018-03-25 marginal.Rmd
Some revision of the text

@@
 The results show that high social class is not always favourable for survival.
-Men in the elite category had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
+Men in the elite category, particularly in working age, had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
 We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
@@
 Social conditions and social position certainly have impact, but not always in the expected direction.
-In our case this is apparent for men during a large part of the studied period; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
+In our case this is apparent for men, during a large part of the studied period, particularly for men in working age; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 being a scientific choice, is also a practical one.
-Social position is treated as a time-dependent variable during working age but defined from the last occupation in working age for the elderly and retired population; i.e.
+Social position during working age is defined at around age 40 but defined from the last occupation in working age for the elderly and retired population; i.e.
 from age 65 until death or last observation.
@@
 
-The disturbace effect of confounders is 
+The disturbance effect of confounders is 
 balanced out through weighting. In our case , `civil status` and `urban/rural` are confounders, that is, they 
@@
 Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
-30--64 during the different periods. Obviously, but not surprisingly, mortality
+40--64 during the different periods. Obviously, but not surprisingly, mortality
 decreases over time for both sexes. Women have consistently better survival

### 2018-03-29 article3c.Rmd
fix description of presence periods

@@
 Each census constitute the baseline for which persons are followed up during the five years until the next census (e.g.
-the 1975 census has a follow-up from 1 January 1976 until the end of 1980).
-The exception of this is the 1990 census for which the follow-up is until 2001.
+the 1975 census has a follow-up from 1 November 1975 until 31 October 1980).
+The exception of this is the 1990 census for which the follow-up is until 31 December 2001.
 Thus, the deaths do not necessarily take place in our region for the Linnaeus analyses, although this is usually the case.
 
-For the last period, 2002--2013, we use the yearly population registers
+For the last period, from 1 January 2002 until 31 December 2013, we use the yearly population registers
 (LISA data) together with information on deaths from the National Board of
-Health and Welfare. This allows for full control over the presence periods
-in the studied regions.
+Health and Welfare.
+The LISA data depict the situation at the end of the given year, so the follow-up period is from 1 January to 31 December the next year.
+Deaths may still occur outside the region, but less frequently than in the Fob case.
 

### 2018-03-29 marginal.Rmd
add timing (temporarily)
fix description of presence periods

@@
 title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
-author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström
-date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
+author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson@umu.se} and Göran Broström
+date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden `r Sys.time()`'
 output:
@@
 
+```{r time1, echo = FALSE}
+time1 <- Sys.time() # Time the document rendering
+```
+
 ```{r setup, include=FALSE}
@@
 
+```{r time2, echo = FALSE}
+time2 <- Sys.time()
+time2 - time1
+```
+
 # Acknowledgements

### 2018-04-04 marginal.Rmd
fix random error in Model

@@
 by comparing estimated survival curves, adjusted for the distribution of the confounding variables over the social classes.
-The adjustment carried out by inverse probability weighting (**reference**) in the following way: First, for each social class, 
+The adjustment carried out by inverse probability weighting [@hoth52;@hero06] in the following way: First, for each social class, 
 the probability that an individual in the class belongs to that class is estimated by logistic regression with the 
@@
 curves for each social class.
+
+The social classes are compared by the \emph{expected years of life
+before 65(100)} for a person aged 40(65). This measure is calculated
+as the area under the survival curve for each class.
 
@@
 -->
-
+<!--
 The disturbance effect of confounders is 
@@
 ```
+-->
 

### 2018-04-07 marginal.Rmd
eval

@@
 ```{r hazcalw}
+source("R/getWeights.R")
+labb <- levels(vb$period)
 nper <- length(labb)

### 2018-04-07 marginal.Rmd
fixes

--- 2018-04-04 marginal.Rmd
@@
 more common in the rural area than others (we find farmers on the countryside).
+-->
 
@@
 
+<!--
 ## Association between civil status and social class over time
@@
 resm <- matrix(0, nrow = nhis, ncol = nper)
+pdm <-  matrix(0, nrow = nhis, ncol = nper)
 for (i in 1:nper){
@@
     fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resm[, i] <- summary(survfit(fit), rmean = 65)$table[, "*rmean"]
+    x <- survfit(fit)
+    resm[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
+    pdm[, i] <- 1 - x$surv[cumsum(x$strata)]
 }
@@
 resw <- matrix(0, nrow = nhis, ncol = nper)
+pdw <-  matrix(0, nrow = nhis, ncol = nper)
 for (i in 1:nper){
@@
     fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resw[, i] <- summary(survfit(fit), rmean = 65)$table[, "*rmean"]
+    x <- survfit(fit)
+    resw[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
+    pdw[, i] <- 1 - x$surv[cumsum(x$strata)]
 }
@@
 refclw <- resw[3, ]
-plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2030), 
+plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2015), 
      ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
@@
 }
-legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
+legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
 refclm <- resm[3, ]
 plot(midper, resm[1, ] - refclm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", 
-     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2030))
+     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2015))
 axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
 axis(2, las = 1)
-axis(4, las = 1,)
+axis(4, las = 1)
 abline(h = 65)
@@
 }
-legend("bottomright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
+legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
@@
 More interesting is to look at the late differences in the age group 65--100.
+
+A more relevant measure of comparison is perhaps the probability to die before age 65 for a forty-year-old, see Figure \@ref(fig:probsw).
+
+```{r probsw, fig.cap = "Relative probability of dying before the age of 65, given survival to age 40, by period and hisclass.
+Reference is worker.", fig.scap = "Relative probability of dying before 65.", fig.height = 8, out.height = "14cm"}
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
+refclw <- pdw[3, ]
+plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(1/3, 3), 
+     xlab = "Period", xlim = c(1800, 2015), log = "y",
+     ylab = "Relative probability", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
+
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+axis(4, las = 1)
+abline(h = 0)
+box()
+for (i in 2:5){
+    lines(midper, pdw[i, ] / refclw, col = scol[i], type = "b", pch = i, lty = i)
+}
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+refclm <- pdm[3, ]
+plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(1/3, 3), xlab = "Period", 
+     ylab = "Relative probability", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
+     xlim = c(1800, 2015), log = "y")
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+axis(4, las = 1)
+abline(h = 0)
+box()
+for (i in 2:5){
+    lines(midper, pdm[i, ] / refclm, col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
+##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
+par(oldpar)
+```
 
@@
 
-Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
+Figure \@ref(fig:cumsw) shows the differences in expected years of survival in the age interval
 40--64 during the different periods. Obviously, but not surprisingly, mortality
@@
 substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. We must, however, distinguish between differences in absolute risks compared to relative risks [@hoetal16]. The compression that is
+absolute differences between social classes. We must, however, distinguish between differences in absolute risks and relative risks. The compression that is
 visible for later decades is partly explained by the decreasing general
-mortality and does not necessarily represent a homogenization of relative risks.
+mortality and does not imply decreasing relative risks, see Figure \@ref(fig:probsw).
 
@@
 
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
 mm <- resm[3, ]
@@
 }
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7)
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",

### 2018-04-08 marginal.Rmd
fix finish

@@
 
-# Model
+# Models and methods
 
@@
 
-The social classes are compared by the \emph{expected years of life
-before 65(100)} for a person aged 40(65). This measure is calculated
-as the area under the survival curve for each class.
+The important explanatory variable, social class, is included in the
+proportional hazards models as a *stratification* variable. The main
+reason for this choice is that the effect of social class on survival is
+non-proportional; that is, it varies with age. The
+results are presented graphically.  The comments on the results are based on
+the figures and focus on the central variable in our study, social class. 
+The models are fitted separately for sex and time period, from  1801 to
+2013.
+
+The social classes are compared by the *expected years of life
+before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40, respectively.
+The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older.
+
+The analyses are performed in the **R** environment for statistical computing
+and graphics [@cran], especially using the package **eha**
+[@gb12;@eha]. In the analysis, data are divided into two age groups,
+40--64 and 65--99 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
+but we then combine results in order to show the development of the social patterning of mortality during the studied period.
 
@@
 the original data set.
-
-The analyses are performed in the **R** environment for statistical computing
-and graphics [@cran], especially using the package **eha**
-[@gb12;@eha]. In the analysis, data are divided into two age groups,
-30--64 and 65--99 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
-but we then combine results in order to show the development of the social patterning of mortality during the studied period.
-
-The important explanatory variable, social class, is included in the
-proportional hazards models as a *stratification* variable. The main
-reason for this choice is that the effect of social class on survival is
-non-proportional; that is, it varies with age. With the reduced data set
-and Poisson regression, this is simply achieved by introducing an
-interaction term between **age** and social class. Then, the main
-results are presented graphically.  The comments on the results are based on
-the figures and focus on the central variable in our study, social class.
-
-The models are fitted separately for sex and each period, from  1801 to
-2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
-**civst** (marital status). The *cumulative hazards* at the end of the age period *30--64* 
-conditional on survival to age 30,  and  *remaining expected life at 65* are the
-main targets in our analysis. These numbers are used in the process of
-graphically illustrating the development over time.
 
@@
 
+<!--
 ```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass.
@@
 
-Obviously, differences are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class, the exception being those who have no notation of social class.
+-->
+Differences in mortality in the age group 40--64 are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class.
 In Sweden of the year 2014, the probability of surviving age 65 for a forty-year-old man is 91 per cent  (a period measure).
 The corresponding figure for a forty-year-old woman is 94 per cent.
-More interesting is to look at the late differences in the age group 65--100.
-
-A more relevant measure of comparison is perhaps the probability to die before age 65 for a forty-year-old, see Figure \@ref(fig:probsw).
-
-```{r probsw, fig.cap = "Relative probability of dying before the age of 65, given survival to age 40, by period and hisclass.
-Reference is worker.", fig.scap = "Relative probability of dying before 65.", fig.height = 8, out.height = "14cm"}
+
+A more relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.
+Figure \@ref(fig:probsw) shows the development of the relative risks over time, with worker as the reference category.
+
+```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and hisclass.
+Reference is worker.", fig.scap = "Relative risk of dying before 65.", fig.height = 8, out.height = "14cm"}
 source("R/plotED.glm.R")
@@
      xlab = "Period", xlim = c(1800, 2015), log = "y",
-     ylab = "Relative probability", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
+     ylab = "Relative risk", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
 
@@
 plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(1/3, 3), xlab = "Period", 
-     ylab = "Relative probability", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
+     ylab = "Relative risk", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
      xlim = c(1800, 2015), log = "y")
@@
 
-Hovever, from Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong.
+From Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong.
+The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
+This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
 
@@
 
-The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
-This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
-
-Figure \@ref(fig:cumsw) shows the differences in expected years of survival in the age interval
-40--64 during the different periods. Obviously, but not surprisingly, mortality
-decreases over time for both sexes. Women have consistently better survival
+Figure \@ref(fig:probsw) does not show that mortality
+decreases over time for both sexes, and that women have consistently better survival
 than men. This is a well-known phenomenon in Swedish
-demographic history [@wil99; @suwi07]. We also observe a
-substantial decline in mortality for both sexes, resulting in smaller
-absolute differences between social classes. We must, however, distinguish between differences in absolute risks and relative risks. The compression that is
-visible for later decades is partly explained by the decreasing general
-mortality and does not imply decreasing relative risks, see Figure \@ref(fig:probsw).
-
-The changing economy has some consequences for the analysis, particularly when
-it comes to the agricultural sector. While farming occupations were common
-up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then. This means that the middle category has
-a different character in the last periods.
+demographic history [@wil99; @suwi07] as well as in the rest of the world. The
+substantial decline in mortality for both sexes results in smaller
+absolute differences between social classes in the later time periods.
 
 The most striking result of the analysis is the different patterns for men
-and women. The social pattern of mortality among women is a great deal in
-accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12].
-The highest social class usually has comparatively low mortality (except for the first period when the statistical power was weak), while
+and women, especially regarding the elite class. The social pattern of mortality among women is a great deal in
+accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12]. The highest social class usually has a comparatively low mortality, while
 working-class women have high mortality, but almost constantly about the same level as the middle category.
-However, there is no substantial change in the relative risks between classes except for the periods 1976 onwards when a divide appears between workers and the middle category.
-A consistent social gradient has developed; the group with unknown social class is distinguished by much higher mortality from the 1976 onwards.
+However, there is no substantial change in the relative risks between classes except for the periods after 1976.
+A consistent social gradient has developed; the group with unknown social class is distinguished by much higher mortality from 1976 and onwards.
 This might have to do with how we have defined social position for women before and after 1950.
-After 1975 it is based on their own occupation; lacking an occupation likely includes women who are more frail when it comes to health or are vulnerable due to social characteristics.
-
-A completely different pattern appears among men, which is the most surprising finding in
-our analyses. The groups we expect to have the lowest mortality, if
-we assume that access to vital resources and belonging to a high social class
-determines survival, instead have by far the highest hazards of dying
+After it is based on their own occupation; lacking an occupation likely includes women who are more frail when it comes to health or are vulnerable due to social characteristics.
+
+A completely different pattern appears among men. The groups we expect to have the lowest mortality, if
+assuming that access to vital resources and belonging to a high social class
+determines survival, instead have by far the highest risks of dying before 65
 during the 19th century and much of the 20th. Workers, on the
@@
 level as the middle category.
-The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead, we find better survival among workers compared to the elite.
-
-The relative positions change radically towards the end of the 20th century. From the 1980s, there is a clear advantage when it comes to survival for the elite. However, the relative
-differences are more compressed compared to the periods until 1950.
+
+The relative positions change radically towards the end of the 20th century.
+From the 1980s, there is a clear advantage when it comes to survival for the elite.
 A reversal of the class pattern in mortality has taken place, and the expected social gradient in mortality has developed.
@@
 unknown social class is probably even more relevant for men, who likely
-represent groups in society that are particularly vulnerable.
-
+represent groups in society that are particularly vulnerable. The reason for being without notation of occupation may also vary over time.
+
+<!--
 A further observation can be made from Figure \@ref(fig:cumsw). The gender
@@
 the difference was much smaller. It was men in high positions who experienced the largest risks.
+-->
+The changing economy has some consequences for the analysis, particularly when
+it comes to the agricultural sector. While farming occupations were common
+up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then.
+This means that the middle category has a different character in the last periods.
 
@@
 We now turn to the elderly, those who in most cases have left the workforce.
-We use the *median remaining life at 65* (reported as median age at death) as the summary measure.
-Looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance is still strong.
+We use the *expected remaining years of life at 65 up to age 100*  as the summary measure, since the relative risk is a not so good measure when risks are high.
+First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is still strong.
 
@@
 
-```{r cumsw2, fig.cap = "Excess years of life (compared to 'worker') before 100 for a 65-year-old, by hisclass and period (experimental!).", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2, fig.cap = "Excess years of life (compared to worker) before 100 for a 65-year-old, by hisclass and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")
@@
 
-The general pattern and development of mortality (Figures \@ref(fig:plotrawwom) and \@ref(fig:plotrawm)) do not
-differ much from the younger age group, of course representing a much
+The general pattern and development of mortality do not
+differ much from the younger age group, but on a much
 higher mortality level earlier in history. Mortality among women, as
@@
 Second World War; that is, later than for the age
-group 30--64 years. This corresponds to the Swedish development whereby we
+group 40--64 years. This corresponds to the Swedish development whereby we
 find a turning point at this time with better survival among elderly, first for women and a couple of decades later for men.
 
-Concerning the social pattern, the pattern is less clear compared to what
-we find for the younger age group in line with the status-levelling hypothesis. For women the results are mixed. We do not find the
+Concerning the differences in the social pattern (Figure \@ref(fig:cumsw2)), it is less clear compared to what
+we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the
 same clear advantage for elite women during the period before 1950, but neither
@@
 The social differences continued to be small until the last decades of the
-20th century. It is only during the two last periods (after the year 2000) that it is
+20th century. It is only during the four last periods (after the year 1975) that it is
 possible to establish the expected social gradient in mortality.
@@
 For men we find indications of the same intriguing change among elderly as
-for the adult male population aged 30--64 years, but at a much lower level. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
+for the adult male population aged 40--64 years. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
 the 20th century. It is not until the second half of the 20th
@@
 at a time when the expected pattern of a social hierarchy in survival appeared.
-The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences as well.
+The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences.
 We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower than in most other parts of Sweden but that it has followed the general development of first diminishing and later increasing inequality.
@@
 We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways.
-In the following we discuss three possible pathways -- the life course perspective, social stress and class- and gender-based lifestyles.
+In the following we discuss three possible pathways---the life course perspective, social stress and class- and gender-based lifestyles.
 Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one's health later in life.
@@
 time2 <- Sys.time()
-time2 - time1
+##time2 - time1
 ```
@@
 (CEDAR) at Umeå University, financed by the Swedish Research Council (grant
-No 349-2008-6592) and the project *Old age, health and inequality:
-Mortality from a life course perspective in Sweden 1986--2009*, financed by
-the Swedish Research Council (grant No 421-2012-892). We thank the
+No 349-2008-6592) and Edvinsson was economically supported by the Swedish Research Council via the project *Old age, health and inequality:
+Mortality from a life course perspective in Sweden 1986--2009* (grant No 421-2012-892). We thank the
 Demographic Data Base, Umeå University, for giving us access to data and

### 2018-04-11 marginal.Rmd
add absolute mortality

@@
 A more relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.
-Figure \@ref(fig:probsw) shows the development of the relative risks over time, with worker as the reference category.
+
+Figure \@ref(fig:probswabs) shows the development of the relative risks over time.
+
+```{r probswabs, fig.cap = "Absolute risk of dying before the age of 65, given survival to age 40, by period and hisclass.", fig.scap = "Absolute risk of dying before 65.", fig.height = 8, out.height = "14cm"}
+source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
+refclw <- 1 ## pdw[3, ]
+plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(0.0, 0.55), 
+     xlab = "Period", xlim = c(1800, 2015), ##log = "y",
+     ylab = "P(die before 65 | alive at 40", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
+
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+axis(4, las = 1)
+abline(h = 0)
+box()
+for (i in 2:5){
+    lines(midper, pdw[i, ] / refclw, col = scol[i], type = "b", pch = i, lty = i)
+}
+legend("bottomleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+refclm <- 1 ##pdm[3, ]
+plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(0.0, 0.55), xlab = "Period", 
+     ylab = "P(die before 65 | alive at 40)", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
+     xlim = c(1800, 2015))#, log = "y")
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+axis(4, las = 1)
+abline(h = 0)
+box()
+for (i in 2:5){
+    lines(midper, pdm[i, ] / refclm, col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("bottomleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
+##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
+par(oldpar)
+```
+
+Figure \@ref(fig:probsw) shows the development of the *relative* risks over time, with worker as the reference category.
 
@@
 ```
+
+The absolute years remaining are shown in Figure \@ref(fig:cumsw2abs).
+
+```{r cumsw2abs, fig.cap = "Remaining years of life before 100 for a 65-year-old, by hisclass and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+##source("R/plotED.glm.R")
+oldpar <- par(mfrow = c(2, 1))
+midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
+mm <- 0 ##resw[3, ]
+plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(75, 88), xlab = "Period", 
+     ylab = "Average death age", axes = FALSE, main = "Women, 65-99")
+
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+axis(4, las = 1)
+box()
+for (i in 2:5){
+    lines(midper, resw[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
+}
+
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+mm <- 0 ##resm[3, ]
+plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(73, 85), xlab = "Period", 
+     ylab = "Average death age", axes = FALSE, main = "Men, 65-99")
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(2, las = 1)
+axis(4, las = 1)
+box()
+for (i in 2:5){
+    lines(midper, resm[i, ] - mm, col = scol[i], type = "b", lty = i, pch = i)
+}
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
+##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
+  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
+par(oldpar)
+```
+
+The excess years of remaining life compared to worker is shown in Figure \@ref(fig:cumsw2).
 

### 2018-04-16 marginal.Rmd
Small corrections until results

@@
 Social conditions and social position certainly have impact, but not always in the expected direction.
-In our case this is apparent for men, during a large part of the studied period, particularly for men in working age; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
+In our case this is apparent for men during a large part of the studied period, particularly for men in working age; only at a surprisingly late date appears a male mortality class reversal, changing the relation to a substantial advantage of being in a higher social position.
 Mortality risks in different contexts must be understood in the intersection between class and gender.
@@
 and other resources. In present-day welfare societies, social position, either according to social class, education or income, is a
-strong determinant when it comes to health and mortality and its impact even
-are increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
+strong determinant when it comes to health and mortality and its impact are even
+increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
 "... social conditions have been, are and will continue to be irreducible
@@
 The LISA data depict the situation at the end of the given year, so the follow-up period is from 1 January to 31 December the next year.
-Deaths may still occur outside the region, but less frequently than in the Fob case.
+Deaths may still occur outside the region, but less frequently than when exposure is restricted to presence in the censuses.
 
@@
 labour force participation developed as the norm, in Sweden during this
-period. The husband's occupation, as signifier of social class became,
+period. The husband's occupation as signifier of social class became,
 if not obsolete, at least less relevant. All included persons are signified
@@
 Then in the early 1990s, a financial crisis struck Sweden, which among other things increased unemployment and changed parts of
-the welfare system towards being less generous. The start of the
-last period (2008--2013) coincides with the international financial
-crisis. In the most recent decades income inequality has increased considerably.
+the welfare system towards being less generous. In the most recent decades income inequality has increased considerably.
 
@@
 before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40, respectively.
-The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older.
+The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older, at least recent periods.
 

### 2018-04-16 marginal.Rmd
Revisions of results 40--64

@@
 
-In the graphical presentation of the results, the *worker* social class is used as reference, and all curves show the difference in years to the expected years in the reference.
-
 ## Ages 40--64
@@
 The corresponding figure for a forty-year-old woman is 94 per cent.
-
 A more relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.
 
-Figure \@ref(fig:probswabs) shows the development of the relative risks over time.
+From Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong in our statistical model of the probability of dying.
+The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
+This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
 
@@
 
-Figure \@ref(fig:probsw) shows the development of the *relative* risks over time, with worker as the reference category.
-
 ```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and hisclass.
@@
 ```
-
-From Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong.
-The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.
-This indicates that fundamental changes have taken place in the relationship between the variables during our study period.
 
@@
 
-Figure \@ref(fig:probsw) does not show that mortality
-decreases over time for both sexes, and that women have consistently better survival
+Figure \@ref(fig:probswabs) shows the absolute risks of dying for in the age interval 40--64 during the different periods. Obviously, but not surprisingly, mortality
+decreases over time for both sexes. Women have consistently better survival
 than men. This is a well-known phenomenon in Swedish
@@
 absolute differences between social classes in the later time periods.
-
-The most striking result of the analysis is the different patterns for men
+We must, however, distinguish between differences in absolute risks compared to relative risks [@hoetal16].
+The compression that is visible for later decades is partly explained by the decreasing general mortality and does not necessarily represent a homogenisation of relative risks.
+
+By standardizing the levels in each time period for each class, we illustrate the relative risks in Figure \@ref(fig:probsw).
+
+The most striking and surprising result of the analysis is the different patterns for men
 and women, especially regarding the elite class. The social pattern of mortality among women is a great deal in
 accordance with previous analyses on other regions in northern Sweden and for other periods [@edli11; @edbr12]. The highest social class usually has a comparatively low mortality, while
-working-class women have high mortality, but almost constantly about the same level as the middle category.
-However, there is no substantial change in the relative risks between classes except for the periods after 1976.
-A consistent social gradient has developed; the group with unknown social class is distinguished by much higher mortality from 1976 and onwards.
+working-class women have high mortality, but during most periods at about the same level as the farmer and only a little higher than the middle category.
+However, there is no substantial change in the relative risks between classes except for the periods after 1976 (considering the very few female farmers during the latest decades).
+The group with unknown social class is distinguished by much higher mortality from 1976 and onwards.
 This might have to do with how we have defined social position for women before and after 1950.
@@
 other hand, have comparatively low mortality until 1950, at the same
-level as the middle category.
+level as the farmers and somewhat below the middle category. Consequently, the gender
+difference was extremely large for the elite group. Men had much higher mortality risks compared to women, basically living in the same households. In the other classes, 
+the difference was much smaller. It was men in high positions who were the most vulnerable.
 
@@
 -->
-The changing economy has some consequences for the analysis, particularly when
-it comes to the agricultural sector. While farming occupations were common
-up until the first half of the 20th century, they have constantly decreased both in numbers as proportionally since then.
-This means that the middle category has a different character in the last periods.
 

### 2018-04-16 marginal.Rmd
Small changes

@@
 
-```{r probswabs, fig.cap = "Absolute risk of dying before the age of 65, given survival to age 40, by period and hisclass.", fig.scap = "Absolute risk of dying before 65.", fig.height = 8, out.height = "14cm"}
+```{r probswabs, fig.cap = "Absolute risk of dying before the age of 65, given survival to age 40, by period and social class.", fig.scap = "Absolute risk of dying before 65.", fig.height = 8, out.height = "14cm"}
 source("R/plotED.glm.R")
@@
 
-```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and hisclass.
+```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and social class.
 Reference is worker.", fig.scap = "Relative risk of dying before 65.", fig.height = 8, out.height = "14cm"}
@@
 
-```{r cumsw2abs, fig.cap = "Remaining years of life before 100 for a 65-year-old, by hisclass and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2abs, fig.cap = "Remaining years of life before 100 for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")
@@
 
-```{r cumsw2, fig.cap = "Excess years of life (compared to worker) before 100 for a 65-year-old, by hisclass and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2, fig.cap = "Excess years of life (compared to worker) before 100 for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")

### 2018-04-17 marginal.Rmd
Revisions of results for the 65+ group

@@
 We now turn to the elderly, those who in most cases have left the workforce.
-We use the *expected remaining years of life at 65 up to age 100*  as the summary measure, since the relative risk is a not so good measure when risks are high.
+We use the *expected remaining years of life at 65 up to age 100* (reported as medan age at death) as the summary measure, since the relative risk is a not so good measure when risks are high.
 First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is still strong.
@@
 
-The absolute years remaining are shown in Figure \@ref(fig:cumsw2abs).
-
-```{r cumsw2abs, fig.cap = "Remaining years of life before 100 for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2abs, fig.cap = "Median age at death for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")
@@
 
-The excess years of remaining life compared to worker is shown in Figure \@ref(fig:cumsw2).
-
 ```{r cumsw2, fig.cap = "Excess years of life (compared to worker) before 100 for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
@@
 
-The general pattern and development of mortality do not
-differ much from the younger age group, but on a much
-higher mortality level earlier in history. Mortality among women, as
-expected, was lower than among men. We find a clear decline in mortality, but
+The general pattern and development of mortality resembles that from the younger age group, according to the median age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs).
+Median death age among women, as expected, was higher than among men.
+We find a clear increase in life expectancy, but
 the fastest decline did not set in definitely until sometime after the
@@
 
-Concerning the differences in the social pattern (Figure \@ref(fig:cumsw2)), it is less clear compared to what
-we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the
-same clear advantage for elite women during the period before 1950, but neither
+The excess years of remaining life in different social classes compared to workers are shown in Figure \@ref(fig:cumsw2).
+Concerning the differences in the social pattern, it is less consisten compared to what
+we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the same clear advantage for elite women during the period before 1950, but neither
 do we find any large disadvantage for working-class elderly women at that
 time.
-In most periods the social differences were small, particularly between the workers and the middle group, and the elite women even had somewhat higher mortality during some periods until the 1901--1925 period.
-The social differences continued to be small until the last decades of the
-20th century. It is only during the four last periods (after the year 1975) that it is
+In most periods the social differences were small between working class women and those in the farming and middle catetories.
+The social differences continued to be small until the last decades of the 20th century.
+It is only during the four last periods (after the year 1975) that it is
 possible to establish the expected social gradient in mortality.

### 2018-04-17 marginal.Rmd
Small changes

@@
 time.
-In most periods the social differences were small between working class women and those in the farming and middle catetories.
+In most periods the social differences were small between working class women and those in the farming and middle categories (only differences of a couple of months in median age at death).
 The social differences continued to be small until the last decades of the 20th century.
 It is only during the four last periods (after the year 1975) that it is
-possible to establish the expected social gradient in mortality.
+possible to establish the expected and consistent social gradient in mortality.
 

### 2018-04-17 marginal.Rmd
Revision of discussion

@@
 The results presented here show that the consistent advantage for higher social classes does not appear until surprisingly late in the investigated regions.
-Instead, the elite category experienced the highest death risks.
+Instead, men in the elite category experienced the highest death risks.
 The basic pattern is primarily observed in the working-age population, while the differences are smaller or even non-existent in old age in accordance with the status-levelling hypothesis.
@@
 However, these differences are neither of the magnitude nor of the character that the main conclusions are affected.
-Most women end up in the same category as their husbands and using a time-dependent definition or a definition based on baseline with a short follow-up can only marginally change the results, especially since most people in our age groups remained in the same region at follow-up.
+Most women end up in the same category as their husbands and using presence based strictly on living in the regions or a definition based on baseline with a short follow-up can only marginally change the results, especially since most people in our age groups remained in the same region at follow-up.
 
@@
 
-We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development.
+In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc.
+However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality and that the different mechanisms all contribute to the advantage of higher social position.
+
+We believe our results have implications on several issues concerning our understanding of the demographic and epidemiologic development and seems to be in conflict with many of the dominating theories in present-day health research.
 One of the central theories has been presented by @mck76, who argues that improved nutrition has been the main determinant of the increase in life expectancy in recent centuries.
@@
 
-Our study seems thus to be in conflict with many of the dominating theories in present-day health research.
-In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc.
-However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality and that the different mechanisms all contribute to the advantage of higher social position.
-
-One of these leading theories is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to social position.
+One of these leading theories in social determinants of health is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to social position.
 The criticism of the arguments by McKeown forced them to address following question:  "If Colgrove is right about the McKeown thesis, social epidemiology is left with a gaping hole in its explanatory repertoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02].

### 2018-04-20 marginal.Rmd
Revision of results in the elderly category

@@
 
-The general pattern and development of mortality resembles that from the younger age group, according to the median age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs).
+The general pattern and development of mortality resembles that from the younger age group (notice that the comparisons are made with different methods in the two age groups, and the figures are therefore only comparable for the identification of patterns in the development), according to the median age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs).
 Median death age among women, as expected, was higher than among men.
@@
 The excess years of remaining life in different social classes compared to workers are shown in Figure \@ref(fig:cumsw2).
-Concerning the differences in the social pattern, it is less consisten compared to what
-we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the same clear advantage for elite women during the period before 1950, but neither
+Concerning the differences in the social pattern, it is less consistent compared to what
+we find for the younger age group in the earlier time periods. For women the results are mixed. We do not find the same clear advantage for elite women during the period before 1925, but neither
 do we find any large disadvantage for working-class elderly women at that
@@
 For men we find indications of the same intriguing change among elderly as
-for the adult male population aged 40--64 years. Mortality was somewhat higher among elite men during the 19th century, a pattern that prevailed until the middle of the
-the 20th century. It is not until the second half of the 20th
-century that the survival is better for the highest social class while
-workers of retirement age have slightly higher mortality. The difference between the worker and middle group has been small in most periods.
+for the adult male population aged 40--64 years, but with a more mixed social pattern in this age group until the last periods.
+There was no clear social gradient during the 19th century, a pattern that prevailed until the middle of the the 20th century.
+Similar to what we found for men in working age, it is not until the second half of the 20th
+century that a distinct social gradient appears, characterised by better survival for higher social classes.
+The difference between the worker and the farmer group, that towards the end of the 20th century has become marginal in numbers has been small in most periods.
 

### 2018-04-23 marginal.Rmd
change median to mean; correct small errors

@@
 author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson@umu.se} and Göran Broström
-date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden `r Sys.time()`'
+date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
 output:
@@
 There are different hypotheses about the impact in adulthood and old age respectively.
-The differences either converge in old age (status levelling), are constant (status maintenance), or diverge (cumulative advantage) [@hof08].
+The differences either decrease in old age (status levelling), are constant (status maintenance), or increase (cumulative advantage) [@hof08].
 Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors.
@@
 
-The Umeå region in the dataset consists of Umeå urban and rural parishes 1901--1950, and from 1976 onwards of Umeå municipality, withn another three parishes included.
+The Umeå region in the dataset consists of Umeå urban and rural parishes 1901--1950, and from 1976 onwards of Umeå municipality, with another three parishes included.
 This region has a somewhat different character to that of Skellefteå.
@@
 
-The two different data sets represent different ways of reporting occupations.
+The two data sets represent different ways of reporting occupations.
 Poplink usually only provides the occupation of the head of household, thus underestimating female labour force participation as well as that of adult children residing with their parents [@vik10].
@@
 by comparing estimated survival curves, adjusted for the distribution of the confounding variables over the social classes.
-The adjustment carried out by inverse probability weighting [@hoth52;@hero06] in the following way: First, for each social class, 
+The adjustment is carried out by inverse probability weighting [@hoth52;@hero06] in the following way: First, for each social class, 
 the probability that an individual in the class belongs to that class is estimated by logistic regression with the 
@@
 The corresponding figure for a forty-year-old woman is 94 per cent.
-A more relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.
+A relevant measure of comparison is the probability (risk) of dying before age 65 for a forty-year-old, remembering that the absolute risks are very small in the later time periods.
 
@@
 absolute differences between social classes in the later time periods.
-We must, however, distinguish between differences in absolute risks compared to relative risks [@hoetal16].
-The compression that is visible for later decades is partly explained by the decreasing general mortality and does not necessarily represent a homogenisation of relative risks.
-
-By standardizing the levels in each time period for each class, we illustrate the relative risks in Figure \@ref(fig:probsw).
+We must, however, distinguish between differences in absolute risks and in relative risks.
+The compression that is visible for later decades is partly explained by the decreasing general mortality and does not necessarily represent a homogenisation of relative risks [@hoetal16].
+The relative risks are shown in Figure \@ref(fig:probsw).
 
@@
 We now turn to the elderly, those who in most cases have left the workforce.
-We use the *expected remaining years of life at 65 up to age 100* (reported as medan age at death) as the summary measure, since the relative risk is a not so good measure when risks are high.
+We use the *expected remaining years of life at 65 up to age 100* (reported as mean age at death) as the summary measure, since the relative risk is a not so good measure when risks are high.
 First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is still strong.
@@
 
-```{r cumsw2abs, fig.cap = "Median age at death for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2abs, fig.cap = "Mean age at death for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
 ##source("R/plotED.glm.R")
@@
 
-The general pattern and development of mortality resembles that from the younger age group (notice that the comparisons are made with different methods in the two age groups, and the figures are therefore only comparable for the identification of patterns in the development), according to the median age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs).
-Median death age among women, as expected, was higher than among men.
+The general pattern and development of mortality resembles that from the younger age group (notice that the comparisons are made with different methods in the two age groups, and the figures are therefore only comparable for the identification of patterns in the development), according to the mean age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs).
+Mean death age among women, as expected, was higher than among men.
 We find a clear increase in life expectancy, but
@@
 A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89].
-A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study in not unique, but rather seems to have characterised all parts of the country.
+A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country.
 These findings are contrary to what is usually assumed and what other studies have shown.
@@
 Occupational titles in historical sources are not always sufficiently distinct to define the correct class.
-Studies covering long periods face the problem that occupations and classes change character---they do not represent the same thing.
+Studies covering long periods face the problem that occupations and classes change character---they do not represent the same thing over time.
 We have chosen to work with a few broad groups.
@@
 We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways.
-In the following we discuss three possible pathways---the life course perspective, social stress and class- and gender-based lifestyles.
+In the following we discuss three possible pathways---the life course perspective, social stress, and class- and gender-based lifestyles.
 Concerning the life course perspective, exposure to diseases and other negative experiences in childhood can influence one's health later in life.

### 2018-04-24 marginal.Rmd
Some final corrections and additions.

@@
 continuity in the determinants of social health inequalities from adulthood to old age.
-The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral rewarding some while disfavouring others.
+The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral, rewarding some while disfavouring others.
 This leads to larger differences in old age.
@@
 The first ships of the year came to this area in May or even as late as June in some cases [@fah56].
-Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden and, improving the economy and making it possible to mitigate the effects of harvest failures.
+Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden, improving the economy and making it possible to mitigate the effects of harvest failures.
 During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
@@
 University.
-Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data from 1976 onwards.
+Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data only from 1976 onwards.
 The study period from 1976 to 2000 is constructed from
@@
 
-```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
+```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
 xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 than men. This is a well-known phenomenon in Swedish
-demographic history [@wil99; @suwi07] as well as in the rest of the world. The
+demographic history [@wil99; @suwi07] as well as in other parts of the world. The
 substantial decline in mortality for both sexes results in smaller
@@
 
+Earlier versions have been presented at 11th European Social Science History Conference, Valencia 2016, 2nd Conference of the European Society of Historical Demography, Leuven 2016, 41st Annual Meeting of the Social Science History Association, Chicago 2016, The Ninth International Congress of Arctic Social Sciences ICASS IX, Umeå 2017, 20th Nordic Demographic Symposium, Åbo 2017 and XXVII International Population Conference, Cape Town 2017.
+We are grateful for comments and suggestions at these conferences as well as comments from our colleagues ag CEDAR, Umeå University.
+
 # References

### 2018-04-24 marginal.Rmd
Small correction
Another small revision

@@
 
-Earlier versions have been presented at 11th European Social Science History Conference, Valencia 2016, 2nd Conference of the European Society of Historical Demography, Leuven 2016, 41st Annual Meeting of the Social Science History Association, Chicago 2016, The Ninth International Congress of Arctic Social Sciences ICASS IX, Umeå 2017, 20th Nordic Demographic Symposium, Åbo 2017 and XXVII International Population Conference, Cape Town 2017.
-We are grateful for comments and suggestions at these conferences as well as comments from our colleagues ag CEDAR, Umeå University.
-
 # References

### 2018-04-25 marginal.Rmd
add Keywords

@@
 We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
-The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology."
+The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.\\par
+
+\\textbf{Keywords:} Class reversal in mortality; Cox regression; Gender differences; Inverse probability weighting; Mortality transition; Northern Sweden; Social inequality; 19th to 21 century
+"
 ---
@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
 library(knitr)

### 2018-04-27 marginal.Rmd
endfloat, anonymous

@@
 title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
-author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson@umu.se} and Göran Broström
-date: 'Centre for Demographic and Ageing Research \newline Umeå University, SE-90187 Umeå, Sweden'
+author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com} and Y
+date: "`r Sys.Date()"
 output:
@@
 titlepage: yes
-bibliography: "mybib.bib"
+bibliography: "mybib"
 documentclass: article
@@
   - \usepackage{graphicx}
+  - \usepackage{endfloat}
 abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class.

### 2018-05-08 marginal.Rmd
remove Acknowledgements

@@
 author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com} and Y
-date: "`r Sys.Date()"
+date: "`r Sys.Date()`"
 output:
@@
 
-Marital status (see Figure \@ref(fig:civstskum)) is a time-fixed covariate.
+Marital status <!-- (see Figure \@ref(fig:civstskum)) --> is a time-fixed covariate.
 The statuses used are unmarried, married and dissolved marriage
@@
 set to unmarried. Note that it has not been possible to identify cohabitation.
+
+The proportion married varies between 60 and 70 percent over time, never married between 15 and 25 percent, and previously married between 10 and 15 per cent.
+
+<!--
 
@@
 
+<!--
 # Acknowledgements
@@
 data.
+-->
 

### 2018-09-05 marginal.Rmd
Small change
Some cutting in the first pages

@@
 A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality.
-In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population with a special focus on the relative differences.
+In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population.
 The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå.
-The main issue is analysed according to gender and age group (working age vs retired), testing whether the same patterns prevail in old age---the retired population---as in working age.
+The main issue is analysed according to gender and age group (working age vs retired).
 Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society.
@@
 The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph95; @liph98].
-The reigning view is that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies.
-This is a
+The idea that socio-economic health inequalities were probably larger in historical societies is a
 reasonable assumption since these societies were often characterized by
@@
 
-Social position affects health and mortality differently during the life course.
-There are different hypotheses about the impact in adulthood and old age respectively.
+Social position affects health and mortality differently during the life course and from adulthood into old age.
 The differences either decrease in old age (status levelling), are constant (status maintenance), or increase (cumulative advantage) [@hof08].
@@
 The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages persist and accumulate during life in a negative spiral, rewarding some while disfavouring others.
-This leads to larger differences in old age.
 
@@
 They find
-that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the todays economically developed world.
+that unequal societies perform less well when it comes to health (as well as other social conditions) than equal ones in the todays economically developed world.
 This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04; @wado00].
@@
 Even if not a necessary implication, it is reasonable to assume that poorer groups were the most disadvantaged due to large levels of inequality.
-
-When it comes to Sweden both income and wealth distributions were strongly skewed from the 1870s to the early 20th century. Starting during the inter-war period, income and
-wealth inequality continuously diminished to reach a low point at around
-1980. Since then, economic inequality has increased substantially until the
-present, although it is far from as high as it was a century ago [@rowa08; @rowa09].
-The Swedish development resembles that of other European and North
-American countries in its basic features [@rowa14].
 
@@
 was dominated by agriculture, making it vulnerable to harvest failures; several severe famines occurred in the regions during the 1800s, for example after the harvest failure of 1867 [@segb14].
-During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen.
-The first ships of the year came to this area in May or even as late as June in some cases [@fah56].
-Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden, improving the economy and making it possible to mitigate the effects of harvest failures.
-During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
+During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen, in some cases as late as June [@fah56].
+Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, facilitating contact with the rest of Sweden, improving the economy and making it possible to mitigate the effects of harvest failures.The regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden.
 
@@
 hamlets, making its livelihood from agricultural production. During the
-20th century, industrialization took place. This also led to a population increase both in both the town
-and the rural parts, resulting in a more diversified economy.
-
-Mortality was fairly low compared with other parts of the country and
-the fertility transition was late [@coa86]. The Skellefteå population
+20th century, industrialization took place. This also led to a population increase both in the town
+and the rural parts, resulting in a more diversified economy. The Skellefteå population
 size as defined in our data sets (all ages) was 6,142 on January 1, 1801, 43,212 on January 1, 1901, and 76,723 at the end of the 20th century.
@@
 The Umeå region in the dataset consists of Umeå urban and rural parishes 1901--1950, and from 1976 onwards of Umeå municipality, with another three parishes included.
-This region has a somewhat different character to that of Skellefteå.
 Umeå town had for a long time a small population, and was the administrative, educational and military centre of the county of Västerbotten.
 During the latter part of the 20th century, the establishment of Umeå University led to a rapid population increase.
-Agriculture dominated the rural part, but there were also some foundries as well as forestry industries.
+Agriculture dominated the rural part.
 Consequently the economy was more diversified than that of Skellefteå.
@@
 Even though our study does not directly address the hypothesis posed by @wipi09 and thus cannot falsify it, our results are relevant for this question as well.
-In the early 20th century, like other countries, Sweden was characterised by enormous economic differences; still, in our regions this did not lead to any consistent advantage for the highest social class.
+In the 19th and early 20th century, like other countries, Sweden was characterised by enormous economic differences; still, in our regions this did not lead to any consistent advantage for the highest social class.
 After the Second World War, economic differences began diminishing, and were at their lowest in Sweden around 1980; i.e.
 at a time when the expected pattern of a social hierarchy in survival appeared.
-The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences.
+The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences [@rowa08; @rowa09].
 We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower than in most other parts of Sweden but that it has followed the general development of first diminishing and later increasing inequality.

### 2018-09-06 compare.Rmd
first commit

---
title: "Compare foli and new"
author: "Göran Broström"
date: "September 6, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(eha)
```

## Intro

I compare the old data file **foli.rmd** with the new version **per_60.rds**. The latter is slightly more strict with right censoring ("last seen").

## The foli file

```{r getfoli}
load("data/foli.rda")
summary(foli[, c("enter", "exit", "event", "sex", "birthdate")])
```

## The per_60 file

```{r getper}
per60 <- readRDS("../Long_Life/mydata/per_60.rds")
per60$sex <- factor(per60$KON, labels = c("male", "female"))
summary(per60[, c("enter", "exit", "event", "sex", "birthdate")])
```

## Comparison

We see that **foli** is age-cut 30-100, while **p60** only has a lower cut, at age 60. In order to make them comparable, we cut them into the largest common age interval, 6--100:

```{r cutcomp}
ageint <- c(60, 100)
per60 <- age.window(per60, ageint)[, c("LINNEID", "enter", "exit", "event", "sex", "birthdate")]
foli <- age.window(foli, ageint)[, c("id", "enter", "exit", "event", "sex", "birthdate")]
summary(foli)
summary(per60)
```

### 2018-09-07 marginal.Rmd
Further revisions until Result part

@@
 
-The recent extension in time of data at the Demographic Data Base (DDB),
-Umeå University (http://www.cedar.umu.se) by including the period 1900 to
-1950 makes it possible to investigate the long-term development of social
-inequality in mortality in this part of Sweden. The data for the present study come from two large
-population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
+The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
 Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
@@
 registers from 1960 to 2013 (censuses, LISA from Statistics Sweden and cause of death registers) and is used within the ageing programme at CEDAR, Umeå
-University.
-Due to data issues (a great deal of missing data in some variables, for example social class, and additional work required to define the exact geographical areas of analysis) we choose to use Linnaeus data only from 1976 onwards.
-The study period from 1976 to 2000 is constructed from
-censuses every fifth year during the period 1975--1990, with additional information on deaths
-from the National Board of Health and Welfare. For the period 2002--2013 we use
-the information from the yearly population registers from Statistics Sweden
-(LISA) together with information on mortality from the National Board of Health and
-Welfare.
+University. Due to data issues we choose to use Linnaeus data only from 1976 onwards.
 
@@
 Mortality differences are analysed according to social class, based on a modified form
-of the classification scheme Hisclass [@lema11]. The only, and obvious, alternative for conducting a longitudinal study of socio-economic mortality differences with consistent classification
-schemes in the two databases is occupations. Thus, analysing according to social class, apart from
-being a scientific choice, is also a practical one.
+of the classification scheme Hisclass [@lema11].
 Social position during working age is defined at around age 40 but defined from the last occupation in working age for the elderly and retired population; i.e.
@@
 before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40, respectively.
-The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older, at least recent periods.
+The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older, at least during recent periods.
 

### 2018-09-11 marginal.Rmd
Making revisions to the end

@@
 We use the *expected remaining years of life at 65 up to age 100* (reported as mean age at death) as the summary measure, since the relative risk is a not so good measure when risks are high.
-First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is still strong.
+First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is strong.
 
@@
 
-The general pattern and development of mortality resembles that from the younger age group (notice that the comparisons are made with different methods in the two age groups, and the figures are therefore only comparable for the identification of patterns in the development), according to the mean age at death for a 65-year-old shown in Figure \@ref(fig:cumsw2abs).
+The general pattern and development of mortality resembles that from the younger age group, see  Figure \@ref(fig:cumsw2abs).
 Mean death age among women, as expected, was higher than among men.
@@
 on different regions, age groups and time periods [@edv92; @edv04; @edli11; @edbr12].
-When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic differences have often, but not always,  been confirmed [@edv92; @edv04],
-something that has been found in other Swedish studies as well [@bredro02; @bube01; @mol17].
+When it comes to infant and child mortality during pre- and early-transitional time, the expected socio-economic differences have often, but not always, been confirmed [@edv92; @edv04; @bredro02; @bube01; @mol17].
 
@@
 from such studies show other patterns than that for infant and child
-mortality.
-While some report substantial differences, others have found surprisingly small or even reversed class differences, for example in the 19th century Sundsvall region [@edli11; @edbr12].
+mortality, reporting surprisingly small or even reversed class differences, for example in the 19th century Sundsvall region [@edli11; @edbr12].
 This raises the question of when the modern social gradient appeared, if this was something that developed in the early parts of the 20th century.
@@
 
-We have analysed social inequality in health from the perspective of social classes.
-However, there are alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude the possibility that such analyses might give other results or modify the conclusions suggested here.
-In many modern studies categorizations based on income, wealth or education are common when examining inequality.
-It would be interesting to compare the results from this study with alternative measures, but reliable data on income and education are unfortunately rarely available in historical population databases stretching back as far as in this article.
-
 An intriguing aspect of this study is the lack of association between level of economic inequality and social differences in mortality.
@@
 In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc.
-However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality and that the different mechanisms all contribute to the advantage of higher social position.
+However, most of these explanations are based on the assumption that higher social classes are always advantaged when it comes to health and mortality.
 
@@
 The results presented here prove that the most wealthy and privileged groups were not able to avoid the most common health risks in historical societies.
-This should make us cautious of accepting nutritional improvements as the most important explanation for the mortality transition.
 While the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it has played and is still playing a central role within epidemiology and social medicine.
@@
 Others emphasize psycho-social factors, for example @mar04, who argues that higher status leads to better health regardless of where on the status ladder the individual is situated.
-Psychosocial factors are also suggested by @wipi09 as an explanation for why equal societies perform better than unequal ones when it comes to health and other social conditions.
+Psychosocial factors are also suggested by @wipi09 as an explanation for why equal societies perform better than unequal ones.
 In the long-term perspective provided here, neither of these theories fits the historical observations; their validity is mainly apparent for the most recent decades.
@@
 Up until the early 20th century, being exposed to an urban environment had far higher health consequences than  living in rural Sweden.
-We control for the type of environment people live in, but we do not have full information about previous residences.
 There is a possibility that the elite group to a greater extent had spent their childhood in the more harmful urban environments, and that this might have influenced their health and survival in adulthood.
 It seems improbable, however, that differences in residential backgrounds could explain most of the observed pattern.
-The second pathway we discuss relates to the health risks of belonging to a certain social class.
+The second pathway relates to the health risks of belonging to a certain social class.
 We usually assume that health risks are higher in social classes with less access to resources, as discussed earlier in this article.
 It is possible, however, that men in high positions had a stressful life, for example entrepreneurs facing the risk of losing an advantageous position.
-Business often failed which meant a substantial loss of prestige and social network.
+Business could fail which meant a substantial loss of prestige and social network.
 Suicides certainly occurred among the elite, but it is difficult to verify the extent of social stress in this and other social classes.
@@
 As several researchers have pointed out, demographic research has often ignored power relations and different roles within families [@watkins93; @mack95; @jans07].
-The gender difference in mortality may be caused by discrimination of the female party in the household.
 Historical Sweden was strongly gender-segregated, discriminating against women and giving them less space for action.
@@
 In the today's high-income world including Sweden, smoking and drinking contribute to a substantial part of socio-economic health differences [@mapemy14; @syla16].
-Our study suggests a strong impact in historical Sweden as well.
+Our study suggests a possible impact in historical Sweden as well.
 If this interpretation is correct, the hypothesis by @ant67, on an increasing role of health-related behaviour in the modern low-mortality regime while economic and material conditions determined the social mortality pattern during the early phases of the mortality transition, can be questioned.
@@
 The social pattern among women was the reverse, however, in line with the present study.
-There is no solid information on smoking in different social classes before the 1960s, but there are international studies indicating a higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, for example illustrated by the fact that smoking was initially more prevalent among physicians.
+There is no solid information on smoking in different social classes before the 1960s in Sweden, but there are international studies indicating a higher prevalence and consumption of tobacco in higher classes during the early phases of the tobacco epidemic, for example illustrated by the fact that smoking was initially more prevalent among physicians.
 There was also a clear association between higher incomes and tobacco consumption [@loco94].
@@
 In particular, we should focus on causes that indicate risk behaviours.
+
+Finally, we have analysed social inequality in health from the perspective of social classes.
+However, there are alternative ways to analyse social inequality, representing other dimensions, and we cannot exclude the possibility that such analyses might give other results or modify the conclusions suggested here.
+In many modern studies categorizations based on income, wealth or education are common when examining inequality.
+It would be interesting to compare the results from this study with alternative measures, but reliable data on income and education are unfortunately rarely available in historical population databases stretching back as far as in this article.
 

### 2018-09-16 marginal.Rmd
Final (almost) revisions

@@
 title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
-author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com} and Y
-date: "`r Sys.Date()`"
+author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com}
+  and Y
+date: '`r Sys.Date()`'
 output:
-  bookdown::pdf_document2:
-    citation_package: natbib
-    number_sections: yes
-    toc: no
-    toc_depth: 2
-    keep_tex: true
   bookdown::html_document2:
@@
     toc_float: yes
+  bookdown::pdf_document2:
+    citation_package: natbib
+    keep_tex: yes
+    number_sections: yes
+    toc: no
+    toc_depth: 2
   bookdown::word_document2:
@@
     toc_depth: 2
-titlepage: yes
-bibliography: "mybib"
+citation_package: natbib
+classoption: titlepage
+biblio-style: apalike
 documentclass: article
-classoption: titlepage
 fontsize: 12pt
-citation_package: natbib
-biblio-style: "apalike"
 header-includes:
@@
 - \usepackage{endfloat}
-abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common notions of the assumed advantage in survival of belonging to a high social class.
-The issue is analysed according to gender and age group (adults and elderly) and in relation to the development of economic inequality.
-The results show that high social class is not always favourable for survival.
-Men in the elite category, particularly in working age, had higher mortality compared to others during a large part of the studied period; a male mortality class reversal appears at a surprisingly late date, while the social gradient among women conforms to the expected pattern.
-We suggest that health-related behaviour is decisive not only in later but earlier phases of the mortality transition as well.
-The results implicate that the association between social class and health is more complex than is assumed in many of the dominant theories in demography and epidemiology.\\par
-
-\\textbf{Keywords:} Class reversal in mortality; Cox regression; Gender differences; Inverse probability weighting; Mortality transition; Northern Sweden; Social inequality; 19th to 21 century
-"
+bibliography: mybib
+abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
+  notions of the assumed advantage in survival of belonging to a high social class.
+  The issue is analysed according to gender and age group (adults and elderly) and
+  in relation to the development of economic inequality. The results show that high
+  social class is not always favourable for survival. Men in the elite category, particularly
+  in working age, had higher mortality compared to others during a large part of the
+  studied period; a male mortality class reversal appears at a surprisingly late date,
+  while the social gradient among women conforms to the expected pattern. We suggest
+  that health-related behaviour is decisive not only in later but earlier phases of
+  the mortality transition as well. The results implicate that the association between
+  social class and health is more complex than is assumed in many of the dominant
+  in mortality; Cox regression; Gender differences; Inverse probability weighting;
+  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century "
+  theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
+titlepage: yes
 ---
@@
 The results are discussed in relation to the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
+The study is unique in its long time-perspective and its utilization of historical micro-data of a population of sufficient size for the necessary analyses.
 
@@
 Individuals are anonymized and as the two databases are not
-linked they are treated as separate units.
+linked, they are treated as separate units.
 This prevents us from following individuals between the two databases throughout their lives.
@@
 Female labour is much better covered in the Linnaeus database, and it is difficult to define households in the same way as in Poplink. Extramarital cohabitation became common and female
-labour force participation developed as the norm, in Sweden during this
+labour force participation developed as the norm in Sweden during this
 period. The husband's occupation as signifier of social class became,
@@
 \item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
-\item {\tt middle}, Hisclass 3, 4, 5, and 6. Lower managers,
-farm hands and lower professionals, clerical and sales personell.
+\item {\tt middle}, Hisclass 3, 4, 5, and 6. Lower managers, lower professionals, clerical and sales personell.
 \item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
@@
 century that a distinct social gradient appears, characterised by better survival for higher social classes.
-The difference between the worker and the farmer group, that towards the end of the 20th century has become marginal in numbers has been small in most periods.
+The difference between the worker and the farmer group, that towards the end of the 20th century has become marginal in numbers, has been small in most periods.
 
@@
 A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country.
-These findings are contrary to what is usually assumed and what other studies have shown.
+These findings are contrary to what is usually assumed.
 The studies usually referred to when discussing the historically large social inequalities in health are mainly based on data from England and Wales [@ant67; @wowi95; @pamuk85].
@@
 
-The surprising results make it reasonable to consider the validity and possible weaknesses in the study.
+The surprising results in the present study make it reasonable to consider the validity and possible weaknesses in the study.
 Some of the limitations have been addressed in the presentation of the variables: the fact that social class of women was defined according to household until 1950 but was defined individually from 1976, different definitions of presence during the same two periods and some differences in parishes included at different periods.
@@
 The period thereafter is characterized by increasing economic differences, coinciding with increasing health differences [@rowa08; @rowa09].
-We do not have access to the exact development of income inequality in our regions, but the available information based on the 1930 census and the data from 1970 onwards indicate that inequality has been lower than in most other parts of Sweden but that it has followed the general development of first diminishing and later increasing inequality.
+We do not have access to the exact development of income inequality in our regions, but there is no reason to believe that it has not followed the general development of first diminishing and later increasing inequality, but maybe at somewhat lower levels of inequality.
 The long-term relationship between economic inequality and health is still a puzzle requiring further studies.
 
+The WHO Commission on Social Determinants of Health framework provides a holistic model on how socio-economic position influences health.
+Social class works through intermediate determinants (material circumstances, behaviours, biological factors and psycho-social factors) related to exposure and vulnerability [@who10].
+People are situated within socio-economic and political contexts, as well as prevailing societal value systems.
 In an overview of explanations of the persistence of health inequalities (primarily focusing on high-income countries with extensive welfare arrangements), @mack12 identifies a couple of theories, emphasizing different mechanisms and circumstances such as psycho-social pathways, material conditions, cultural capital, etc.
@@
 The results presented here prove that the most wealthy and privileged groups were not able to avoid the most common health risks in historical societies.
+
 While the McKeown thesis has been refuted by several scholars [@szreter88, @colgrove02], it has played and is still playing a central role within epidemiology and social medicine.
-
-One of these leading theories in social determinants of health is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of different disadvantages connected to social position.
+One of these leading theories is the hypothesis by @liph95 stating that social conditions are fundamental causes of disease and not only expressions of specific disadvantages connected to social position.
 The criticism of the arguments by McKeown forced them to address following question:  "If Colgrove is right about the McKeown thesis, social epidemiology is left with a gaping hole in its explanatory repertoire and a challenge to a cherished principle about the importance of social factors of health" [@liph02].
-This led Link and Phelan to incorporate alternative ways of how social conditions determine health, but still with the understanding that access to more resources guarantees better survival chances.
 
@@
 The historical perspective shows the complexity of how health and mortality are socially constructed, and illustrates the need to contextualize our analyses.
-The dominant theories are not necessarily wrong but may need to be further developed and refined.
-The WHO Commission on Social Determinants of Health framework provides a holistic model on how socio-economic position influences health.
-Social class works through intermediate determinants (material circumstances, behaviours, biological factors and psycho-social factors) related to exposure and vulnerability [@who10].
-People are situated within socio-economic and political contexts, as well as prevailing societal value systems.
-
 We must thus consider other and potentially more important explanations than the pure advantage of being in a high social position, reflecting that the embodiment can be materialized in different ways.

### 2018-09-18 article3.Rmd
A change in the title

@@
 ---
-title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
+title: "Is high social class always beneficial for survival? A study of northern Sweden 1801--2013"
 author: Sören Edvinsson\footnote{Corresponding author; Phone number +46907866040, email soren.edvinsson\@umu.se} and Göran Broström

### 2018-09-18 article3.Rmd
Revising acknowledgement

@@
 Living Conditions* at the Centre for Demographic and Ageing Research
-(CEDAR) at Umeå University, financed by the Swedish Research Council (grant
-No 349-2008-6592) and the project *Old age, health and inequality:
-Mortality from a life course perspective in Sweden 1986--2009*, financed by
-the Swedish Research Council (grant No 421-2012-892). We thank the
+(CEDAR) at Umeå University, financed by the Swedish Research Council under grant
+No 349-2008-6592 and grant No 421-2012-892. We thank the
 Demographic Data Base, Umeå University, for giving us access to data and

### 2018-09-18 marginal.Rmd
Changing reference

@@
 ---
-title: "Is high social class always beneficial for survival? Northern Sweden 1801--2013"
+title: "Is high social class always beneficial for survival? A study of northern Sweden 1801--2013"
 author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com}
@@
 A study of mortality and social class in Sweden during the period 1960--1968 found that self-employed men and non-manual male workers in fact had higher mortality than manual workers in total as well as in cardiovascular mortality [@vano89].
-A recent study by @erdr17 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country.
+A recent study by @drer18 on the social pattern of mortality, based on 1880--1910 census data linked to death registers, shows that the unexpected pattern found in the present study is not unique, but rather seems to have characterised all parts of the country.
 These findings are contrary to what is usually assumed.

### 2018-09-18 marginal.Rmd
fix yaml error

@@
     toc_depth: 2
+  bookdown::html_document2:
+    number_sections: yes
+    toc: yes
+    toc_depth: 2
+    toc_float: yes
   bookdown::word_document2:
@@
 - \usepackage{endfloat}
+- \usepackage{setspace}
 bibliography: mybib
@@
   social class and health is more complex than is assumed in many of the dominant
+    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
   in mortality; Cox regression; Gender differences; Inverse probability weighting;
-  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century "
-  theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
+  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."
+
 titlepage: yes
 ---
+
+\doublespacing
 

### 2018-09-19 marginal.Rmd
add doublespacing to setspace

@@
 - \usepackage{endfloat}
-- \usepackage{setspace}
+- \usepackage[doublespacing]{setspace}
 bibliography: mybib

### 2018-09-19 popstud_anon.Rmd
first commit for Pop. Stud.

--- 2018-09-18 marginal.Rmd
@@
 - \usepackage{endfloat}
-- \usepackage{setspace}
+- \usepackage[doublespacing]{setspace}
+- \usepackage{sectsty}
+- \subsectionfont{\normalfont\itshape}
 bibliography: mybib
@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs/art3d-',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs_popstud/',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
 library(knitr)
@@
 # ##dr
-knitr::kable(dr, caption = "Single term deletions, ages 30-64.")
+knitr::kable(dr, caption = "Single term deletions, ages 30-64.", digits = c(0, 1, 2, 4))
 # ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."),
@@
 drold <- drop1(fittold, test = "Chisq")
-knitr::kable(drold, caption = "Single term deletions, ages 65-99.")
+knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
 ##drold

### 2018-09-19 popstud_open.Rmd
first commit for Pop. Stud.

--- 2018-09-19 popstud_anon.Rmd
@@
 ---
-title: "Is high social class always beneficial for survival? A study of northern Sweden 1801--2013"
-author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com}
-  and Y
-date: '`r Sys.Date()`'
+title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"}
+author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se (corresponding author)}
+  and Göran Broström \footnote{email goran.brostrom@umu.se}
+date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
 output:

### 2018-09-20 popstud_open.Rmd
introduce aa in town names
cleaning, new colors, captions, ...

@@
 
-(ref:cities) The Skellefteå and Umeå regions in Sweden.
+(ref:cities) The Skellefte\aa\ and Ume\aa\ regions in Sweden.
 
@@
 
-```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4}
+<!--
+```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4,cache=FALSE}
+cls <- rainbow(5, s = 1.0)
+cls <- cls[c(2, 3, 1, 5, 4)]
+cls[1] <- "black"
+oldcols <- c("red", "darkgreen", "blue", "green", "black")
 xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
-```
-
-```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
-xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+```
+
+```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4,cache=FALSE}
+xx2 <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
 ##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 ##xxx <- rbind(xx, xy, xz)
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
+yy2 <- round(prop.table(xx2, 1) * 100, 1)
+barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+```
+-->
+
+```{r plottaalla,fig.height = 8, out.height = "18cm", fig.cap="Exposure by social class (from the top: elite, middle, workers, farmers, none), time period, and sex.", fig.scap="Exposure by social class, time period, and sex"}
+cls <- rainbow(5, s = 1.0)
+cls <- cls[c(2, 3, 1, 5, 4)]
+cls[1] <- "black"
+
+oldpar <- par(mfrow = c(2, 1))
+barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Men")
+barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Women")
+par(oldpar)
 ```
@@
 
-```{r colorc}
+```{r colorc,cache=FALSE}
 ## Colors for plotting strata:
-scol <- c("black", "green", "blue", "darkgreen", "red")
+##scol <- c("black", "green", "blue", "darkgreen", "red")
+scol <- cls[5:1]
 ## Common for all groups:
@@
 
-```{r getske3064, eval = FALSE}
+```{r getske3064, eval = FALSE, cache = FALSE}
 cap1w <-  "Cumulative hazards for social class by time period, ages 30--64, women."
@@
 
-```{r probswabs, fig.cap = "Absolute risk of dying before the age of 65, given survival to age 40, by period and social class.", fig.scap = "Absolute risk of dying before 65.", fig.height = 8, out.height = "14cm"}
+```{r probswabs, fig.cap = "Probability of dying before the age of 65, given survival to age 40, by period and social class.", fig.scap = "Probability of dying before 65.", fig.height = 10, out.height = "18cm", cache = FALSE}
 source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
+oldpar <- par(mfrow = c(2, 1), lwd = 1.5)
 midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
@@
 plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(0.0, 0.55),
-     xlab = "Period", xlim = c(1800, 2015), ##log = "y",
+     xlab = "", xlim = c(1800, 2015), ##log = "y",
      ylab = "P(die before 65 | alive at 40", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
 
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 }
-legend("bottomleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1, bty = "n")
 refclm <- 1 ##pdm[3, ]
-plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(0.0, 0.55), xlab = "Period",
+plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(0.0, 0.55), xlab = "",
      ylab = "P(die before 65 | alive at 40)", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
      xlim = c(1800, 2015))#, log = "y")
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 }
-legend("bottomleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1, bty = "n")
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
@@
 ```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and social class.
-Reference is worker.", fig.scap = "Relative risk of dying before 65.", fig.height = 8, out.height = "14cm"}
+Reference is worker.", fig.scap = "Relative risk of dying before 65.", fig.height = 10, out.height = "18cm", cache = FALSE}
 source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
+oldpar <- par(mfrow = c(2, 1), lwd = 1.5)
 midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
 refclw <- pdw[3, ]
-plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(1/3, 3),
-     xlab = "Period", xlim = c(1800, 2015), log = "y",
+plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(1/10, 5),
+     xlab = "", xlim = c(1800, 2015), log = "y",
      ylab = "Relative risk", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
 
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 }
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("bottom", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
 refclm <- pdm[3, ]
-plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(1/3, 3), xlab = "Period",
+plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(1/10, 5), xlab = "",
      ylab = "Relative risk", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
      xlim = c(1800, 2015), log = "y")
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 }
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("bottom", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
@@
 
-```{r cumswtab, message=FALSE, cache = TRUE, eval = TRUE}
+```{r cumswtab, message=FALSE, eval = TRUE}
 fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
@@
 # ##dr
-knitr::kable(dr, caption = "Single term deletions, ages 30-64.", digits = c(0, 1, 2, 4))
-# ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."),
-  ##    type = "latex", comment = FALSE)
+```
+
+```{r cumswtabout, message = FALSE, cache = FALSE, results='asis'}
+##knitr::kable(dr, caption = "Significance tests, ages 40-64.", digits = c(0, 1, 2, 4))
+print(xtable(dr,label = "tab:cumswtab", caption = "ANOVA, ages 40--64."),
+      type = "latex", comment = FALSE)
 ```
@@
 We use the *expected remaining years of life at 65 up to age 100* (reported as mean age at death) as the summary measure, since the relative risk is a not so good measure when risks are high.
-First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is strong.
-
-```{r cumswtabold, cap="Table", cache = TRUE, eval = TRUE}
+First, looking at Table \@ref(tab:cumswtaboldout), it is obvious that the statistical significance regarding differences between groups over time is strong.
+
+```{r cumswtabold}
 # vb6599$marital_status <- vb6599$civst
@@
 drold <- drop1(fittold, test = "Chisq")
-knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
+```
+
+```{r cumswtaboldout, cap = "Table", cache = FALSE, results='asis'}
+##knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
 ##drold
-##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--99."),
-  ##    type = "latex", comment = FALSE)
+print(xtable(drold,label = "tab:cumswtabold", caption = "ANOVA, ages 65--99."),
+      type = "latex", comment = FALSE)
 ```
@@
 
-```{r cumsw2abs, fig.cap = "Mean age at death for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2abs, fig.cap = "Mean age at death for a 65-year-old, by social class and period.", fig.scap = "Remaining life at 65", fig.height = 10, out.height = "18cm",cache=FALSE}
 ##source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
+oldpar <- par(mfrow = c(2, 1), lwd = 1.5)
 midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
 mm <- 0 ##resw[3, ]
-plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(75, 88), xlab = "Period",
+plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(73, 88), xlab = "",
      ylab = "Average death age", axes = FALSE, main = "Women, 65-99")
 
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
 mm <- 0 ##resm[3, ]
-plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(73, 85), xlab = "Period",
+plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(73, 88), xlab = "",
      ylab = "Average death age", axes = FALSE, main = "Men, 65-99")
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 }
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
 ##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
@@
 
-```{r cumsw2, fig.cap = "Excess years of life (compared to worker) before 100 for a 65-year-old, by social class and period.", fig.scap = "Differences in expected years of life at 65", fig.height = 8, out.height = "14cm"}
+```{r cumsw2, fig.cap = "Excess years of life (compared to worker) before 100 for a 65-year-old, by social class and period.", fig.scap = "Excess years of life at 65 relative to worker", fig.height = 10, out.height = "18cm",cache=FALSE}
 ##source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
+oldpar <- par(mfrow = c(2, 1), lwd = 1.5)
 midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
 mm <- resw[3, ]
-plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "Period",
+plot(midper, resw[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "",
      ylab = "Excess years", axes = FALSE, main = "Women, 65-99")
 
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)
@@
 
-legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
+legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
 mm <- resm[3, ]
-plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "Period",
+plot(midper, resm[1, ] - mm, col = scol[1], type = "b", ylim = c(-3, 5), xlab = "",
      ylab = "Excess years", axes = FALSE, main = "Men, 65-99")
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
+axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
 axis(2, las = 1)

### 2018-09-26 popstud_open.Rmd
add orcid

@@
 ---
-title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"}
-author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se (corresponding author)}
-  and Göran Broström \footnote{email goran.brostrom@umu.se}
+title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:}
+author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
+  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894}
 date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden

### 2018-09-26 popstud_open.Rmd
Revision of introduction

@@
 The results are discussed in relation to the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.
-The study is unique in its long time-perspective and its utilization of historical micro-data of a population of sufficient size for the necessary analyses.
+The study is unique in its long time-perspective and its utilization of historical micro-data of a sufficiently large and socially diverse population for analyses of a central issue in health research.
 
@@
 The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
-Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
 The period 1801--1950 is covered by the database Poplink [@wee16].

### 2018-09-26 popstud_open.Rmd
fix Appendix and Figure 3

@@
 
-```{r plottaalla,fig.height = 8, out.height = "18cm", fig.cap="Exposure by social class (from the top: elite, middle, workers, farmers, none), time period, and sex.", fig.scap="Exposure by social class, time period, and sex"}
+```{r plottaalla,fig.height = 8, out.height = "18cm", fig.cap="Relative exposure by social class (from the top: elite, middle, workers, farmers, none), time period, and sex.", fig.scap="Exposure by social class, time period, and sex"}
 cls <- rainbow(5, s = 1.0)
@@
 oldpar <- par(mfrow = c(2, 1))
+barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Women")
 barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Men")
-barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Women")
 par(oldpar)
@@
 
+# Appendix {-}
+
+Figure 3 is shown as absolute numbers, exposures for women in Table \@ref(tab:womenexp), and exposures for men in Table \@ref(tab:menexp).
+
+```{r exposuretab, message = FALSE, cache = FALSE, results = 'asis'}
+##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
+ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
+print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure in years by hisclass and period, women.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
+```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
+print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure in years by hisclass and period, men.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
+Then number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).
+
+```{r deathtab, message = FALSE, cache = FALSE, results = 'asis'}
+##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
+deaths <- with(vb, tapply(event, list(period, hisclass, sex), sum))
+print(xtable(deaths[,,2], label = "tab:womendeaths", caption = "Number of deaths by hisclass and period, women.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
+```{r deathtab2, message = FALSE, cache = FALSE, results = 'asis'}
+print(xtable(deaths[,,1], label = "tab:mendeaths", caption = "Number of deaths by hisclass and period, men.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
 # References

### 2018-09-26 future.Rmd
first commit

--- 2018-09-26 popstud_open.Rmd
@@
 ---
-title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:}
-author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
-  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894}
+title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"}
+author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se}
+  and Göran Broström \footnote{email goran.brostrom@umu.se (corresponding author)}
 date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
@@
 bibliography: mybib
-abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
-  notions of the assumed advantage in survival of belonging to a high social class.
-  The issue is analysed according to gender and age group (adults and elderly) and
-  in relation to the development of economic inequality. The results show that high
-  social class is not always favourable for survival. Men in the elite category, particularly
-  in working age, had higher mortality compared to others during a large part of the
-  studied period; a male mortality class reversal appears at a surprisingly late date,
-  while the social gradient among women conforms to the expected pattern. We suggest
-  that health-related behaviour is decisive not only in later but earlier phases of
-  the mortality transition as well. The results implicate that the association between
-  social class and health is more complex than is assumed in many of the dominant
-    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
+abstract: "This is a test version with analyses using tabular data and pch models!! Focusing on two regions in northern Sweden 1801--2013, we challenge common
+  notions of the assumed advantage in survival of belonging to a high social class. \\par\n\\textbf{Keywords:} Class reversal
   in mortality; Cox regression; Gender differences; Inverse probability weighting;
@@
 vb1$period <- factor(vb1$period)
+vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = seq(40, 65, by = 1))
 ##
@@
 vb2$period <- factor(vb2$period)
+vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = seq(65, 100, by = 1))
 vb <- rbind(vb1, vb2)
 vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
+vbTab <- rbind(vb1Tab, vb2Tab) # NEW
 ```
@@
 for (i in 1:nper){
-    dat <- vb1[vb1$sex == "male" & vb1$period == labb[i], ]
-    dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    x <- survfit(fit)
-    resm[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
-    pdm[, i] <- 1 - x$surv[cumsum(x$strata)]
+    dat <- vb1Tab[vb1Tab$sex == "male" & vb1Tab$period == labb[i], ]
+    ##dat <- vb1[vb1$sex == "male" & vb1$period == labb[i], ]
+    ##dat$wght <- getWeights(dat)
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##x <- survfit(fit)
+    for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 1))
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        pdm[j, i] <- ppch(65, cuts = fit$cuts, levels = haz)
+    }
+    ##resm[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
+    ##pdm[, i] <- 1 - x$surv[cumsum(x$strata)]
 }
@@
 for (i in 1:nper){
-    dat <- vb1[vb1$sex == "female" & vb1$period == labb[i], ]
-    dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    x <- survfit(fit)
-    resw[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
-    pdw[, i] <- 1 - x$surv[cumsum(x$strata)]
+    dat <- vb1Tab[vb1Tab$sex == "female" & vb1Tab$period == labb[i], ]
+    ##dat <- vb1[vb1$sex == "female" & vb1$period == labb[i], ]
+    ##dat$wght <- getWeights(dat)
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##x <- survfit(fit)
+    for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 1))
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        pdw[j, i] <- ppch(65, cuts = fit$cuts, levels = haz)
+    }
+    ##resw[, i] <- summary(x, rmean = 65)$table[, "*rmean"]
+    ##pdw[, i] <- 1 - x$surv[cumsum(x$strata)]
 }

### 2018-09-27 popstud_open.Rmd
fix figure numbers on p. 8

@@
 
-Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
+Figure \@ref(fig:plottaalla) shows the distribution
 (per cent) of exposure time according to class for men and women,
@@
 The proportion with missing values among women is a larger problem, however;
-see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
+see Figure \@ref(fig:plottaalla). Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th

### 2018-09-27 marginal.Rmd
what?

@@
 ```{r cumswtab, message=FALSE, cache = TRUE, eval = TRUE}
-fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
-
+##fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
+source("R/makeTab.R")
+vb1Tab <- makeTab(vb1)
+source("~/R/Testing/pch.R")
+Tfit <- pch(event ~ offset(log(exposure)) + age + hisclass * period * sex + civst + urban, seq(40, 65, by = 5), vb1Tab)
+Tdr <- drop1(Tfit, test = "Chisq")
+knitr::kable(Tdr, caption = "ANOVA, ages 40-65.")
 # vb3064$social_class <- vb3064$hisclass
@@
 #             data = vb3064, family = poisson)
-dr <- drop1(fitt, test = "Chisq")
+##dr <- drop1(fitt, test = "Chisq")
 # ##dr
-knitr::kable(dr, caption = "Single term deletions, ages 30-64.")
+##knitr::kable(dr, caption = "Single term deletions, ages 30-64.")
 # ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."),

### 2018-09-29 future.Rmd
scaling down data

@@
 library(knitr)
-opts_chunk$set(engine='R',fig.path='figs_popstud/',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
+opts_chunk$set(engine='R',fig.path='figs_popstud/',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=FALSE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
 library(knitr)
@@
 
-```{r readalldata}
+```{r readalldata, eval = FALSE}
 source("R/get.file.R")
-skume <- get.file("skume")
-##load("data/skume.rda")
+##skume <- get.file("skume")
+load("data/skume.rda")
 skume <- skume[order(skume$id, skume$enter), ]
@@
 ##################################################
-foli <- get.file("foli")
-##load("data/foli.rda")
+##foli <- get.file("foli")
+load("data/foli.rda")
 foli$civst[is.na(foli$civst)] <- "unmarried"
@@
 vbTab <- rbind(vb1Tab, vb2Tab) # NEW
+```
+
+```{r nydata}
+vbTab <- readRDS("data/vbTab.rds")
+vb1Tab <- readRDS("data/vb1Tab.rds")
+vb2Tab <- readRDS("data/vb2Tab.rds")
 ```
@@
 #
-pn.exp <- round(sum(skume1$exit - skume1$enter) + sum(skume2$exit - skume2$enter))
-pn.d <-  as.integer(sum(skume1$event) + sum(skume2$event))
+popl <- vbTab[vbTab$period %in% levels(vbTab$period)[1:6],]
+pn.exp <- sum(popl$exposure)
+##round(sum(skume1$exit - skume1$enter) + sum(skume2$exit - skume2$enter))
+pn.d <-  as.integer(sum(popl$event)) ##as.integer(sum(skume1$event) + sum(skume2$event))
 #
@@
 #
-n.exp <- round(sum(foli1$exit - foli1$enter) + sum(foli2$exit -foli2$enter))
-n.d <-  as.integer(sum(foli1$event) + sum(foli2$event))
+foli <- vbTab[vbTab$period %in% levels(vbTab$period)[7:10], ]
+##n.exp <- round(sum(foli1$exit - foli1$enter) + sum(foli2$exit ##-foli2$enter))
+n.exp <- sum(foli$exposure)
+n.d <-  as.integer(sum(foli$event))
 
@@
 
-<!--
 ```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4,cache=FALSE}
@@
 oldcols <- c("red", "darkgreen", "blue", "green", "black")
-xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
-##xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xx <- with(vbTab[vbTab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##xy <- with(skume[skume$sex == "male", ], tapply(exposure, ##list(period = period, hisclass = hisclass), sum))
 ##xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+##barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
 ```
@@
 ```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4,cache=FALSE}
-xx2 <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+xx2 <- with(vbTab[vbTab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
 ##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 yy2 <- round(prop.table(xx2, 1) * 100, 1)
-barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
-```
--->
+##barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+```
 
@@
 
-<!--
-
 ```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4}
-xx <- with(vb, tapply(exposure, list(period = period, civst = civst), sum))
+xx <- with(vbTab, tapply(exposure, list(period = period, civst = civst), sum))
 ##xy <- with(skume, tapply(exposure, list(period = period, civst = civst), sum))
@@
 ```
--->
 
@@
 
-labb <- levels(vb$period)
+labb <- levels(vbTab$period)
 nper <- length(labb)
@@
 ```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4}
-orr <- aggregate(x = vb$exposure, by = vb[, c("period", "civst", "hisclass")], FUN = sum)
+orr <- aggregate(x = vbTab$exposure, by = vbTab[, c("period", "civst", "hisclass")], FUN = sum)
 
@@
 ```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4}
-orr <- aggregate(x = vb$exposure, by = vb[, c("period", "urban", "hisclass")], FUN = sum)
+orr <- aggregate(x = vbTab$exposure, by = vbTab[, c("period", "urban", "hisclass")], FUN = sum)
 
@@
 ```{r exweight,echo=TRUE,echo = FALSE}
-vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
+vb2_2005_men <- vb2Tab[vb2Tab$sex == "male" & vb2Tab$period == "2005-2013", ]
 fit1 <- glm(I(hisclass == "elite") ~ urban * civst, data = vb2_2005_men, family = binomial)
@@
 
-```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4}
+```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4,eval = FALSE}
 par(cex.axis = 0.8)
@@
 
-```{r estmed}
+```{r estmed, eval = FALSE}
 knitr::kable(round(medi, 1), caption = "Estimated medians, men 65-99, 2005-2013.")
@@
 
-```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4}
+```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4, eval = FALSE}
 source("R/getWeights.R")
@@
 
-```{r explife40}
+```{r explife40,eval=FALSE}
 avew <- round(summary(survfit(fitw), rmean = 100)$table[, "*rmean"], 2)
@@
 source("R/getWeights.R")
-labb <- levels(vb$period)
+labb <- levels(vbTab$period)
 nper <- length(labb)
-his <- levels(vb1$hisclass)
+his <- levels(vb1Tab$hisclass)
 nhis <- length(his)
@@
 ```{r cumswtab, message=FALSE, eval = TRUE}
-fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
+##fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)
 
@@
 # vb3064$age_group <- vb3064$age
-# fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
-#             data = vb3064, family = poisson)
+fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = 40:65,
+             data = vb1Tab)
 dr <- drop1(fitt, test = "Chisq")

### 2018-09-29 future.Rmd
Tab for old

@@
 
-fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
+##fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex ##+ civst + urban, data = vb2)
+fittold <- tpchreg(event ~ offset(log(exposure)) + age + hisclass * period * sex + civst + urban, cuts = 65:100, data = vb2Tab)
 drold <- drop1(fittold, test = "Chisq")
@@
 nper <- length(labb)
-his <- levels(vb1$hisclass)
+his <- levels(vb2Tab$hisclass)
 nhis <- length(his)
@@
 for (i in 1:nper){
-    dat <- vb2[vb2$sex == "male" & vb2$period == labb[i], ]
+    dat <- vb2Tab[vb2Tab$sex == "male" & vb2Tab$period == labb[i], ]
     dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+    for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
+                       cuts = 65:100, 
+                       data = dat[dat$hisclass == his[i], ])
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        resm[j, i] <- mpch(levels = haz, cuts = 65:100)
+    }
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
 }
@@
 for (i in 1:nper){
-    dat <- vb2[vb2$sex == "female" & vb2$period == labb[i], ]
+    dat <- vb2Tab[vb2Tab$sex == "female" & vb2Tab$period == labb[i], ]
     dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+        for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
+                       cuts = 65:100, 
+                       data = dat[dat$hisclass == his[i], ])
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        resw[j, i] <- mpch(levels = haz, cuts = 65:100)
+    }
+
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
 }

### 2018-10-01 popstud_open.Rmd
Some final corrections

@@
 ---
-title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:}
+title: Is high social class always beneficial for survival?
+A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:This work was founded by the Swedish Research Council under grants No 349-2008-6592 No 421-2012-892}
 author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
@@
 (per cent) of exposure time according to class for men and women,
-respectively.
+respectively (see also Appendix).
 The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women.
@@
 
-The proportion with missing values among women is a larger problem, however;
-see Figure \@ref(fig:plottaalla). Corresponding to what we found for men, the
+The proportion with missing values among women is a larger problem, however. Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th
@@
 
+# References
+
 # Appendix {-}
 
-Figure 3 is shown as absolute numbers, exposures for women in Table \@ref(tab:womenexp), and exposures for men in Table \@ref(tab:menexp).
+The exposure in years for men and women respectively are shown in Table \@ref(tab:womenexp), and Table \@ref(tab:menexp).
 
@@
 
-Then number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).
+The number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).
 
@@
 ```
-
-# References

### 2018-10-01 popstud_open.Rmd
Corrections in appendix

@@
 ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
-print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure in years by hisclass and period, women.", digits = 0), 
+print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure in years by social class and period, women.", digits = 0), 
       type = "latex", comment = FALSE)
@@
 ```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
-print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure in years by hisclass and period, men.", digits = 0), 
+print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure in years by social class and period, men.", digits = 0), 
       type = "latex", comment = FALSE)
@@
 deaths <- with(vb, tapply(event, list(period, hisclass, sex), sum))
-print(xtable(deaths[,,2], label = "tab:womendeaths", caption = "Number of deaths by hisclass and period, women.", digits = 0), 
+print(xtable(deaths[,,2], label = "tab:womendeaths", caption = "Number of deaths by social class and period, women.", digits = 0), 
       type = "latex", comment = FALSE)
@@
 ```{r deathtab2, message = FALSE, cache = FALSE, results = 'asis'}
-print(xtable(deaths[,,1], label = "tab:mendeaths", caption = "Number of deaths by hisclass and period, men.", digits = 0), 
+print(xtable(deaths[,,1], label = "tab:mendeaths", caption = "Number of deaths by social class and period, men.", digits = 0), 
       type = "latex", comment = FALSE)

### 2018-10-02 popstud_open.Rmd
cosmetics

@@
 title: Is high social class always beneficial for survival?
-A study of northern Sweden 1801--2013 \footnote{Short; "Is high social class beneficial for survival?"} \footnote{Funding:This work was founded by the Swedish Research Council under grants No 349-2008-6592 No 421-2012-892}
-author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author)}
-  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894}
+A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"} \footnote{Funded by the Swedish Research Council under grants 349-2008-6592 and 421-2012-892.}
+author: Sören Edvinsson \footnote{email soren.edvinsson@umu.se, orcid 0000-0001-7439-002X (corresponding author).}
+  and Göran Broström \footnote{email goran.brostrom@umu.se, orcid 0000-0002-2865-7894.}
 date: Centre for Demographic and Ageing Research, Umeå University, Umeå, Sweden
@@
 Figure \@ref(fig:plottaalla) shows the distribution
-(per cent) of exposure time according to class for men and women,
-respectively (see also Appendix).
+(per cent) of exposure times according to class for women and men (see also Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively, in the Appendix).
 The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women.
@@
 
-The proportion with missing values among women is a larger problem, however. Corresponding to what we found for men, the
+The proportion of missing values among women is a larger problem, however. Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th
@@
 
-The social classes are compared by the *expected years of life
-before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40, respectively.
+The social classes are compared by the *expected years of life before 100* for a person aged 65, and the *relative risk of dying before age 65* for a person aged 40.
 The reason for using different comparative measures for the two age groups is the fact that mortality is very low in the younger age group compared to the older, at least during recent periods.
@@
 ##knitr::kable(dr, caption = "Significance tests, ages 40-64.", digits = c(0, 1, 2, 4))
-print(xtable(dr,label = "tab:cumswtab", caption = "ANOVA, ages 40--64."),
+print(xtable(dr,label = "tab:cumswtabout", caption = "ANOVA, ages 40--64."),
       type = "latex", comment = FALSE)
@@
 ##drold
-print(xtable(drold,label = "tab:cumswtabold", caption = "ANOVA, ages 65--99."),
+print(xtable(drold,label = "tab:cumswtaboldout", caption = "ANOVA, ages 65--99."),
       type = "latex", comment = FALSE)
@@
 
-# References
-
 # Appendix {-}
 
-The exposure in years for men and women respectively are shown in Table \@ref(tab:womenexp), and Table \@ref(tab:menexp).
+The exposure times in years for women and men are shown in Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively.
 
@@
 ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
-print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure in years by social class and period, women.", digits = 0), 
+print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure time in years by social class and period, women.", digits = 0), 
       type = "latex", comment = FALSE)
@@
 ```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
-print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure in years by social class and period, men.", digits = 0), 
+print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure time in years by social class and period, men.", digits = 0), 
       type = "latex", comment = FALSE)

### 2018-10-08 popstud_open.Rmd
Small change

@@
 
-The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
+The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden (Demografiska databasen, CEDAR, Umeå universitet (2015).
+U15014.
+http://dx.doi.org/10.17197/U15014).
 The period 1801--1950 is covered by the database Poplink [@wee16].
@@
 
-This work was undertaken within the research programme *Ageing and
-Living Conditions* at the Centre for Demographic and Ageing Research
-(CEDAR) at Umeå University, financed by the Swedish Research Council (grant
-No 349-2008-6592) and Edvinsson was economically supported by the Swedish Research Council via the project *Old age, health and inequality:
-Mortality from a life course perspective in Sweden 1986--2009* (grant No 421-2012-892). We thank the
-Demographic Data Base, Umeå University, for giving us access to data and
+We thank the Demographic Data Base, Umeå University, for giving us access to data and
 Jonas Helgertz at the Department of Economic History, Lund University, for

### 2018-10-08 popstud_open.Rmd
abstract and acknowledgement

@@
 bibliography: mybib
-abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
+titlepage: yes
+---
+
+\doublespacing
+
+```{r time1, echo = FALSE}
+time1 <- Sys.time() # Time the document rendering
+```
+
+```{r setup, include=FALSE}
+library(knitr)
+opts_chunk$set(engine='R',fig.path='figs_popstud/',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
+library(knitr)
+library(eha)
+library(xtable)
+library(skum)
+source("R/oneSpell.R")
+options(show.signif.stars = FALSE)
+```
+
+```{r readalldata}
+source("R/get.file.R")
+skume <- get.file("skume")
+##load("data/skume.rda")
+skume <- skume[order(skume$id, skume$enter), ]
+skume$id <- paste("pop", skume$id, sep = "_")
+##
+skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
+skume1 <- oneSpell(skume1)
+##
+skume2 <- age.window(skume, c(65, 100))
+skume2 <- oneSpell(skume2)
+##################################################
+foli <- get.file("foli")
+##load("data/foli.rda")
+foli$civst[is.na(foli$civst)] <- "unmarried"
+foli <- foli[order(foli$id, foli$enter), ]
+foli$AR <- foli$ULORSAK <- NULL
+##
+foli1 <- age.window(foli, c(40, 65))
+foli1 <- oneSpell(foli1)
+##
+foli2 <- age.window(foli, c(65, 100))
+foli2 <- oneSpell(foli2)
+##################################################
+source("CreData/R/addPer.R")
+skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
+skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
+##
+foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
+foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
+##################################################
+vb1 <- rbind(skume1, foli1)
+vb1$period <- factor(vb1$period)
+##
+vb2 <- rbind(skume2, foli2)
+vb2$period <- factor(vb2$period)
+vb <- rbind(vb1, vb2)
+vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
+```
+
+\section*{Acknowledgement}
+
+We thank the Demographic Data Base, Umeå University, for giving us access to data and
+Jonas Helgertz at the Department of Economic History, Lund University, for
+providing us with scripts for creating Hisclass from Swedish census
+data.
+
+\begin{abstract} 
+Focusing on two regions in northern Sweden 1801--2013, we challenge common
   notions of the assumed advantage in survival of belonging to a high social class.
@@
   social class and health is more complex than is assumed in many of the dominant
-    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
+    theories in demography and epidemiology.
+
+    \textbf{Keywords:} Class reversal
   in mortality; Cox regression; Gender differences; Inverse probability weighting;
-  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."
-
-titlepage: yes
----
-
-\doublespacing
-
-```{r time1, echo = FALSE}
-time1 <- Sys.time() # Time the document rendering
-```
-
-```{r setup, include=FALSE}
-library(knitr)
-opts_chunk$set(engine='R',fig.path='figs_popstud/',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
-library(knitr)
-library(eha)
-library(xtable)
-library(skum)
-source("R/oneSpell.R")
-options(show.signif.stars = FALSE)
-```
-
-```{r readalldata}
-source("R/get.file.R")
-skume <- get.file("skume")
-##load("data/skume.rda")
-skume <- skume[order(skume$id, skume$enter), ]
-skume$id <- paste("pop", skume$id, sep = "_")
-##
-skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
-skume1 <- oneSpell(skume1)
-##
-skume2 <- age.window(skume, c(65, 100))
-skume2 <- oneSpell(skume2)
-##################################################
-foli <- get.file("foli")
-##load("data/foli.rda")
-foli$civst[is.na(foli$civst)] <- "unmarried"
-foli <- foli[order(foli$id, foli$enter), ]
-foli$AR <- foli$ULORSAK <- NULL
-##
-foli1 <- age.window(foli, c(40, 65))
-foli1 <- oneSpell(foli1)
-##
-foli2 <- age.window(foli, c(65, 100))
-foli2 <- oneSpell(foli2)
-##################################################
-source("CreData/R/addPer.R")
-skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
-skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
-##
-foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
-foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
-##################################################
-vb1 <- rbind(skume1, foli1)
-vb1$period <- factor(vb1$period)
-##
-vb2 <- rbind(skume2, foli2)
-vb2$period <- factor(vb2$period)
-vb <- rbind(vb1, vb2)
-vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
-```
+  Mortality transition; Northern Sweden; Social inequality; 19th to 21th century.
+
+\end{abstract}
 
@@
 
-From Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong in our statistical model of the probability of dying.
+From Table \@ref(tab:cumswtabout) it is obvious that the *statistical significance* is strong in our statistical model of the probability of dying.
 The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**.

### 2018-10-08 popstud_open.Rmd
fix abstract and acknowledgement

@@
 \section*{Acknowledgement}
-
+\thispagestyle{empty}
 We thank the Demographic Data Base, Umeå University, for giving us access to data and
@@
 yy2 <- round(prop.table(xx2, 1) * 100, 1)
-barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+##barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
 ```
@@
 
-```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4}
+```{r civstskum, fig.cap = "Exposure by marital status (from the top: never married, married and previously married) and time period", fig.scap='Exposure by marital status and time period.',out.height = "14cm",fig.height=4,eval = FALSE}
 xx <- with(vb, tapply(exposure, list(period = period, civst = civst), sum))
@@
 
-```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4}
+```{r assesciv, fig.cap = "Fraction ever married by hisclass and period.", fig.height = 4,eval=FALSE}
 orr <- aggregate(x = vb$exposure, by = vb[, c("period", "civst", "hisclass")], FUN = sum)
@@
 
-```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4}
+```{r assesurb, fig.cap = "Fraction urban by hisclass and period.", fig.height = 4,eval=FALSE}
 orr <- aggregate(x = vb$exposure, by = vb[, c("period", "urban", "hisclass")], FUN = sum)
@@
 
-```{r exweight,echo=TRUE,echo = FALSE}
+```{r exweight,echo=TRUE,echo = FALSE,eval=FALSE}
 vb2_2005_men <- vb2[vb2$sex == "male" & vb2$period == "2005-2013", ]
@@
 
-```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4}
+```{r testfit, fig.cap = "Survival by hisclass, men 2005-2013, unweighted and weighted analysis.", fig.height = 4,eval=FALSE}
 par(cex.axis = 0.8)
@@
 
-```{r estmed}
+```{r estmed,eval=FALSE}
 knitr::kable(round(medi, 1), caption = "Estimated medians, men 65-99, 2005-2013.")
@@
 
-```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4}
+```{r forty64, fig.cap = "Adjusted survival curves, men in ages 40-65 and 65-100, years 2005-2013.", fig.height = 4,eval=FALSE}
 source("R/getWeights.R")
@@
 ```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass.
-Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm"}
+Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm",eval=FALSE}
 source("R/plotED.glm.R")

### 2018-10-09 popstud_open.Rmd
Small changes

@@
 
-The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden (Demografiska databasen, CEDAR, Umeå universitet (2015).
-U15014.
-http://dx.doi.org/10.17197/U15014).
+The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden [@ddb15].
 The period 1801--1950 is covered by the database Poplink [@wee16].
@@
 Figure \@ref(fig:plottaalla) shows the distribution
-(per cent) of exposure times according to class for women and men (see also Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively, in the Appendix).
+(per cent) of exposure times according to class for women and men (see also Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively, as well as Table \@ref(tab:womendeaths), and \@ref(tab:mendeaths) for number of deaths during the different periods in the Appendix).
 The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women.

### 2018-10-09 popstud_anon.Rmd
copied from 'open' and modified

@@
 ---
-title: "Is high social class always beneficial for survival? A study of northern Sweden 1801--2013"
-author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com}
-  and Y
-date: '`r Sys.Date()`'
+title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"} 
+author: X (corresponding author) and Y
 output:
@@
 bibliography: mybib
-abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
+titlepage: yes
+---
+
+\doublespacing
+
+```{r time1, echo = FALSE}
+time1 <- Sys.time() # Time the document rendering
+```
+
+```{r setup, include=FALSE}
+library(knitr)
+opts_chunk$set(engine='R',fig.path='figs_popstud/',tidy=FALSE,fig.height=8.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,warning = FALSE)##$
+library(knitr)
+library(eha)
+library(xtable)
+library(skum)
+source("R/oneSpell.R")
+options(show.signif.stars = FALSE)
+```
+
+```{r readalldata}
+source("R/get.file.R")
+skume <- get.file("skume")
+##load("data/skume.rda")
+skume <- skume[order(skume$id, skume$enter), ]
+skume$id <- paste("pop", skume$id, sep = "_")
+##
+skume1 <- age.window(skume, c(40, 65))  ###### OJOJOJ!!!!!!!!!!!!!!!!!!!!!!!!!!! (40, 65)
+skume1 <- oneSpell(skume1)
+##
+skume2 <- age.window(skume, c(65, 100))
+skume2 <- oneSpell(skume2)
+##################################################
+foli <- get.file("foli")
+##load("data/foli.rda")
+foli$civst[is.na(foli$civst)] <- "unmarried"
+foli <- foli[order(foli$id, foli$enter), ]
+foli$AR <- foli$ULORSAK <- NULL
+##
+foli1 <- age.window(foli, c(40, 65))
+foli1 <- oneSpell(foli1)
+##
+foli2 <- age.window(foli, c(65, 100))
+foli2 <- oneSpell(foli2)
+##################################################
+source("CreData/R/addPer.R")
+skume1 <- addPer(skume1, cuts = seq(1801, 1951, by = 25))
+skume2 <- addPer(skume2, cuts = seq(1801, 1951, by = 25))
+##
+foli1 <- addPer(foli1, cuts = c(1976, 1986, 1996, 2005, 2014))
+foli2 <- addPer(foli2, cuts = c(1976, 1986, 1996, 2005, 2014))
+##################################################
+vb1 <- rbind(skume1, foli1)
+vb1$period <- factor(vb1$period)
+##
+vb2 <- rbind(skume2, foli2)
+vb2$period <- factor(vb2$period)
+vb <- rbind(vb1, vb2)
+vb$exposure <- vb$exit - vb$enter # For statistics and figures purposes
+```
+
+\begin{abstract} 
+Focusing on two regions in northern Sweden 1801--2013, we challenge common
   notions of the assumed advantage in survival of belonging to a high social class.
@@
 
-The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden.
-Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time.
+The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden [@ddb15].
 The period 1801--1950 is covered by the database Poplink [@wee16].
@@
 
-Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
-(per cent) of exposure time according to class for men and women,
-respectively.
+Figure \@ref(fig:plottaalla) shows the distribution
+(per cent) of exposure times according to class for women and men (see also Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively, as well as Table \@ref(tab:womendeaths), and \@ref(tab:mendeaths) for number of deaths during the different periods in the Appendix).
 The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women.
@@
 yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
-```
-
-```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
-xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
+##barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+```
+
+```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4,cache=FALSE}
+xx2 <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
 ##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
@@
 ##xxx <- rbind(xx, xy, xz)
-yy <- round(prop.table(xx, 1) * 100, 1)
-barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
-```
-
-The proportion with missing values among women is a larger problem, however;
-see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
+yy2 <- round(prop.table(xx2, 1) * 100, 1)
+##barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
+```
+-->
+
+```{r plottaalla,fig.height = 8, out.height = "18cm", fig.cap="Relative exposure by social class (from the top: elite, middle, workers, farmers, none), time period, and sex.", fig.scap="Exposure by social class, time period, and sex"}
+cls <- rainbow(5, s = 1.0)
+cls <- cls[c(2, 3, 1, 5, 4)]
+cls[1] <- "black"
+
+oldpar <- par(mfrow = c(2, 1))
+barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Women")
+barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Men")
+par(oldpar)
+```
+
+The proportion of missing values among women is a larger problem, however. Corresponding to what we found for men, the
 proportion without social position is very low until the middle of the 20th
@@
 # ##dr
-knitr::kable(dr, caption = "Single term deletions, ages 30-64.", digits = c(0, 1, 2, 4))
-# ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."),
-  ##    type = "latex", comment = FALSE)
+```
+
+```{r cumswtabout, message = FALSE, cache = FALSE, results='asis'}
+##knitr::kable(dr, caption = "Significance tests, ages 40-64.", digits = c(0, 1, 2, 4))
+print(xtable(dr,label = "tab:cumswtabout", caption = "ANOVA, ages 40--64."),
+      type = "latex", comment = FALSE)
 ```
@@
 drold <- drop1(fittold, test = "Chisq")
-knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
+```
+
+```{r cumswtaboldout, cap = "Table", cache = FALSE, results='asis'}
+##knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
 ##drold
-##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--99."),
-  ##    type = "latex", comment = FALSE)
+print(xtable(drold,label = "tab:cumswtaboldout", caption = "ANOVA, ages 65--99."),
+      type = "latex", comment = FALSE)
 ```
@@
 
-# References
+# Appendix {-}
+
+The exposure times in years for women and men are shown in Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively.
+
+```{r exposuretab, message = FALSE, cache = FALSE, results = 'asis'}
+##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
+ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
+print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure time in years by social class and period, women.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
+```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
+print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure time in years by social class and period, men.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
+The number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).
+
+```{r deathtab, message = FALSE, cache = FALSE, results = 'asis'}
+##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
+deaths <- with(vb, tapply(event, list(period, hisclass, sex), sum))
+print(xtable(deaths[,,2], label = "tab:womendeaths", caption = "Number of deaths by social class and period, women.", digits = 0), 
+      type = "latex", comment = FALSE)
+```
+
+```{r deathtab2, message = FALSE, cache = FALSE, results = 'asis'}
+print(xtable(deaths[,,1], label = "tab:mendeaths", caption = "Number of deaths by social class and period, men.", digits = 0), 
+      type = "latex", comment = FALSE)
+```

### 2018-10-10 future.Rmd
remove some errors

--- 2018-09-29 future.Rmd
@@
 vb1$period <- factor(vb1$period)
-vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = seq(40, 65, by = 1))
+vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = seq(40, 65, by = 5))
 ##
@@
 vb2$period <- factor(vb2$period)
-vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = seq(65, 100, by = 1))
+vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = seq(65, 100, by = 5))
 vb <- rbind(vb1, vb2)
@@
     for (j in seq_along(his)){
-        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 1))
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 5))
         haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
@@
 # vb3064$age_group <- vb3064$age
-fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = 40:65,
+fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = seq(40, 65, by = 5),
              data = vb1Tab)
@@
 
-fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
+##fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex ##+ civst + urban, data = vb2)
+fittold <- tpchreg(event ~ offset(log(exposure)) + age + hisclass * period * sex + civst + urban, cuts = seq(65, 100, by = 5), data = vb2Tab)
 drold <- drop1(fittold, test = "Chisq")
@@
 for (i in 1:nper){
-    dat <- vb2[vb2$sex == "male" & vb2$period == labb[i], ]
+    dat <- vb2Tab[vb2Tab$sex == "male" & vb2Tab$period == labb[i], ]
     dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+    for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
+                       cuts = seq(65, 100, by = 5), 
+                       data = dat[dat$hisclass == his[i], ])
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        resm[j, i] <- mpch(levels = haz, cuts = 65:100)
+    }
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
 }
@@
 for (i in 1:nper){
-    dat <- vb2[vb2$sex == "female" & vb2$period == labb[i], ]
+    dat <- vb2Tab[vb2Tab$sex == "female" & vb2Tab$period == labb[i], ]
     dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+        for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
+                       cuts = seq(65, 100, by = 5), 
+                       data = dat[dat$hisclass == his[i], ])
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        resw[j, i] <- mpch(levels = haz, cuts = 65:100)
+    }
+
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
 }

### 2018-10-11 future.Rmd
error tracking

--- 2018-09-29 future.Rmd
@@
 vb1$period <- factor(vb1$period)
-vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = seq(40, 65, by = 1))
+cuts1 <- seq(40, 65, by = 5) ############################## NOTE!!
+vb1Tab <- toTpch(vb1[, !(colnames(vb1) %in% c("id", "birthdate"))], cuts = cuts1)
 ##
@@
 vb2$period <- factor(vb2$period)
-vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = seq(65, 100, by = 1))
+cuts2 <- c(seq(65, 90, by = 5), 100) ###################### NOTE!!
+vb2Tab <- toTpch(vb2[, !(colnames(vb2) %in% c("id", "birthdate"))], cuts = cuts2)
 vb <- rbind(vb1, vb2)
@@
 vbTab <- rbind(vb1Tab, vb2Tab) # NEW
+saveRDS(vbTab, file = "data/vbTab.rds")
+saveRDS(vb1Tab, file = "data/vb1Tab.rds")
+saveRDS(vb2Tab, file = "data/vb2Tab.rds")
 ```
@@
 vb2Tab <- readRDS("data/vb2Tab.rds")
+cuts1 <- seq(40, 65, by = 5) ############################## NOTE: fragile!!
+cuts2 <- c(seq(65, 90, by = 5), 100) ###################### NOTE: fragile!!
 ```
@@
     for (j in seq_along(his)){
-        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = seq(40, 65, by = 1))
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, data = dat[dat$hisclass == his[j], ], cuts = cuts1)
         haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
@@
 
-<!--
-```{r cumsw, fig.cap = "Differences in expected years of survival until the age of 65, given survival to age 40, by period and hisclass.
-Reference is the working class.", fig.scap = "Differences in expected years of life at age 40", fig.height = 8, out.height = "14cm"}
-source("R/plotED.glm.R")
-oldpar <- par(mfrow = c(2, 1))
-midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
-refclw <- resw[3, ]
-plot(midper, resw[1, ] - refclw, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period", xlim = c(1810, 2015),
-     ylab = "Years", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)
-
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
-axis(2, las = 1)
-axis(4, las = 1)
-abline(h = 65)
-box()
-for (i in 2:5){
-    lines(midper, resw[i, ] - refclw, col = scol[i], type = "b", pch = i, lty = i)
-}
-legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
-refclm <- resm[3, ]
-plot(midper, resm[1, ] - refclm, col = scol[1], type = "b", ylim = c(-3, 3), xlab = "Period",
-     ylab = "Years", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1, xlim = c(1810, 2015))
-axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
-axis(2, las = 1)
-axis(4, las = 1)
-abline(h = 65)
-box()
-for (i in 2:5){
-    lines(midper, resm[i, ] - refclm, col = scol[i], type = "b", lty = i, pch = i)
-}
-legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
-##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
-  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
-##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
-  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
-par(oldpar)
-```
-
--->
 Differences in mortality in the age group 40--64 are getting very small as we move forward in time: Almost all forty-year-olds today survive beyond 65, regardless of social class.
@@
 # vb3064$age_group <- vb3064$age
-fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = 40:65,
+fitt <- tpchreg(event ~ offset(log(exposure)) + age + urban + civst + sex * hisclass * period, cuts = cuts1,
              data = vb1Tab)
@@
 
-fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
+##fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex ##+ civst + urban, data = vb2)
+fittold <- tpchreg(event ~ offset(log(exposure)) + age + hisclass * period * sex + civst + urban, cuts = cuts2, data = vb2Tab)
 drold <- drop1(fittold, test = "Chisq")
@@
 for (i in 1:nper){
-    dat <- vb2[vb2$sex == "male" & vb2$period == labb[i], ]
+    dat <- vb2Tab[vb2Tab$sex == "male" & vb2Tab$period == labb[i], ]
     dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+    for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
+                       cuts = cuts2, 
+                       data = dat[dat$hisclass == his[i], ])
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        cat("lenght(haz) = ", length(haz), "\n")
+        resm[j, i] <- mpch(levels = haz, cuts = fit$cuts)
+    }
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##resm[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
 }
@@
 for (i in 1:nper){
-    dat <- vb2[vb2$sex == "female" & vb2$period == labb[i], ]
+    dat <- vb2Tab[vb2Tab$sex == "female" & vb2Tab$period == labb[i], ]
     dat$wght <- getWeights(dat)
-    fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
-    resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
+        for (j in seq_along(his)){
+        fit <- tpchreg(event ~ offset(log(exposure)) + age, 
+                       cuts = cuts2, 
+                       data = dat[dat$hisclass == his[i], ])
+        haz <- c(0, fit$hazards, fit$hazards[length(fit$hazards)])
+        resw[j, i] <- mpch(levels = haz, cuts = fit$cuts)
+    }
+
+    ##fit <- coxph(Surv(enter, exit, event) ~ strata(hisclass), weights = wght, data = dat)
+    ##resw[, i] <- summary(survfit(fit), rmean = 100)$table[, "*rmean"]
 }
